### 2014-03-03 notation.Rmd
Initial commit.

# Mathematical Notation

## $Z_{(h, i, j)}$ 
$Z_{(h, i, j)}$ is the methylation state in chromosome$_h$ at C$_i$ of sample$_j$.
$Z_{(h, i, j)} = 1$ if methylated.
$Z_{(h, i, j)} = 0$ if unmethylated.
The $Z_{(h, i, j)}$ are hidden states.

## $M_{(i, j)}$ and $U_{(i, j)}$
$M_{(i, j)} = \sum_{h = 1}^{h = H} Z_{(h, i, j)}$ is the number of chromosomes that are methylated at C$_i$ in sample$_j$.
$U_{(i, j)}=  \sum_{h = 1}^{h = H} Z_{(h, i, j)}$ = number of chromosomes that are unmethylated at C$_i$ in sample$_j$.

## $B_{(i, j)}$
$B_{(i, j)}= \frac{M_{(i, j)}}{M_{(i, j)} + U_{(i, j)}}$.
There are also M-values, defined as $M_{(i, j)} = logit(B_{(i, j)})$

## $h$
$h$ is typically unknown. At best we know that $\widehat{Z_{(h, i, j)}}$ have the same $h$, i.e. they were estimated from the same read.

## $i$
$\mathcal{I}_j$, the set of all $i$ in sample $j$, vary over $h$ and $j$
The $\{i\}$ are not equally spaced in the genome. 
The concept of IPD

## $j$
$|J| = |\{j\}|$ is typically small, e.g.  $|J| <= 3$ for methylC-seq data. However, for microarray data might have $|J| \approx 100-1000$

## m-tuples

## NIC

## Estimates of $M, U, Z, B$
__Question__: Use lower cases letters, i.e. $m, u, z, \beta$, or _hatted_ versions, i.e. $\hat{M}, \hat{U}, \hat{Z}, \hat{\beta}$?

## Summarisations
Some studies summarise methylation at the promoter- or gene-level. For example, they might summarise methylation at feature $f$ by the average $= \beta_{j}^{(f)} = \frac{1}{|f|}\sum_{i \in f} \beta_{(i, j)}$, where $|f|$ is the number of methylation measurements in feature $f$.

## One-group tests of differential methylation
The idea is we have one group of methylomes that we expect to be similar at the vast majority of positions. We are interested in identifying those positions where they differ, _epi-alleles_, in Emma Whitelaw's terminology.

Consider the $i^{th}$ CpG. We want to test $H_{0}: B_{(i, 1)} = \ldots = B_{(i, n)} \equiv B_{(i, 0)}$ vs. $H_{A}:$ not all the $B_{(i, j)}$ are equal.

The model for a single CpG is very simple:
$$
M_{(i, j)} | (B_{(i, j)}, N_{(i, j)}) = Binomial(N_{(i, j)}, B_{(i, j)})
$$

Then, the test is simply a chi-square test that the proprotions are equal, i.e. `prop.test(m, n)`.

One problem is that we will be doing millions of such tests (one for each CpG). It will be therefore worthwhile using some shrinkage of improve power. E.g. might impose a prior on $B_{(i, j)}$:
$$
B_{(i, j)} = Beta(\mu_{i}, \theta_{i})
$$

The hyperparameters $\mu$ and $\theta$ are estimated from the data _ala_ empirical Bayes.

### Questions
1. Should there prior on $B_{(i, j)}$ be over $j$, i.e.:
$$
B_{(i, j)} = Beta(\mu_{(i, j)}, \theta_{(i, j)})
$$

## Other people's notation
Feng _et al._ use subscript $j$ to denote groups and subscript $k$ to denote replicates within groups (Feng, H., Conneely, K. N., & Wu, H. (2014). A Bayesian hierarchical model to detect differentially methylated loci from single nucleotide resolution sequencing data. Nucleic Acids Research, gku154. doi:10.1093/nar/gku154). They also have a nice hierarchical model for estimating the level of DNA methylation at a given CpG from bisulfite-sequencing data.

### 2014-03-22 Rmarkdown/dinucleotide_distribution.Rmd
Started section on assays for studying DNA methylation. Started writing R code for counting CpGs in a genome

Distribution of dinucleotides in the human and mouse reference genomes
========================================================

CpGs are underrepresented in mammalian genomes. This appendix demonstrates this result using the human reference genome (build hg19) and mouse reference genome (build mm10). All results are computed at the chromosome-level as well as being summarised genome-wide.

Firstly, the frequency of each nucleotide is computed. Secondly, the expected frequency of each dinucleotide is computed under the _independence model_. Under the independence model, the frequency of each dinucleotide is simply the product of the respective observed nucleotide frequencies. Finally, these expected frequencies are compared to the observed values from the respective reference genomes.

I only use the autosomes, sex chromosomes and mitochondrial chromosome in all analyses, that is, I do not consider the "random", "unplaced" and "alternate haplotype" contigs in the reference genomes.

Human genome 
========================================================

Step 1: Compute the nucleotide frequencies
--------------------------------------------------------
```{r}
#### Load necessary packages ####
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg19)

#### A function to compute the GC-% of a BSgenome object ####
gcPercentage <- function(seqnames, genome){
  ## data.frame to store results
  gc <- data.frame(seqnames = seqnames, gc_count = NA_real_, seqlength = NA_real_, gc_perc = NA_real_)

  for (i in seq_along(seqnames)){
    subject <- genome[[seqnames[i]]]
    cat(">>> Finding all hits in chromosome", seqnames[i], "...\n")
    gc[i, 'gc_count'] <- letterFrequency(x = subject, letters = 'CG')
    gc[i, 'seqlength'] <- length(subject)
    gc[i, 'gc_perc'] <- gc[i, 'gc_count'] / gc[i, 'seqlength']
  }

  ## Compute genome-wide summary
  gc_gw <- data.frame(seqnames = 'genome', gc_count = sum(gc[, 'gc_count']), seqlength = sum(gc[, 'seqlength']), gc_perc = sum(gc[, 'gc_count']) / sum(gc[, 'seqlength']))

  ## Return chromosome-level and genome-wide results in a single data.frame
  res <- rbind(gc, gc_gw)
  return(res)
}

#### Compute GC-% ####
seqnames <- paste0('chr', c(1:22, 'X', 'Y', 'M'))
human_gc_perc <- gcPercentage(seqnames, Hsapiens)
human_gc_perc

nucleotide_frequencies <- do.call('rbind', lapply(seqnames, function(seqname, genome){alphabetFrequency(genome[[seqname]])}, genome = Hsapiens)) # TODO: Check (and remove?) mask
```

Step 2: Compute expected number of CpGs under an independence model
--------------------------------------------------------

Software details
========================================================

### 2014-04-23 latex/phd_thesis.tex
Re-jigged section on BSmooth Wrote section on BiSeq Added master .tex, .bib and .bst files.

\documentclass[11pt,a4paper,oneside]{book}	%fleqn,

\usepackage{longtable,rotating,amsmath,epic, graphics}

\usepackage{multirow, setspace, fullpage, graphicx, amssymb}

\usepackage{epsfig, moreverb}

\usepackage{natbib, color}

\usepackage{tocbibind}

\usepackage{easymat}

\usepackage[pdftex,
            pdfauthor={Your Name},
            pdftitle={Thesis},
            pdfsubject={Dissertation},
            pdfkeywords={microarray gene expression statistical models regression},
            pdfproducer={Latex with hyperref, or other system},
            pdfcreator={pdflatex, or other tool}]{hyperref}

\settocname{Table of Contents}

\bibliographystyle{natbib} % natbib

\bibpunct{[}{]},a{},

\renewcommand\baselinestretch{1.5} %line spacing is 1.5

% convenient macros for lists

\newcommand\BI{\begin{itemize}}

\newcommand\EI{\end{itemize}}

\newcommand\BE{\begin{enumerate}}

\newcommand\EE{\end{enumerate}}

\newcommand\BD{\begin{description}}

\newcommand\ED{\end{description}}

\newcommand\BF{\begin{figure}}

\newcommand\EF{\end{figure}}

\newcommand\BT{\begin{table}}

\newcommand\ET{\end{table}}

\newcommand\I\item

\newcommand\h[1]{\Large\textbf{#1}\vspace{12pt}}

\newcommand\N\newpage

\newcommand\BC{\begin{quotation}\small\noindent}

\newcommand\EC{\end{quotation}}

\newcommand\beq{\begin{equation}}

\newcommand\eeq{\end{equation}}

\newcommand{\ba}{{\bf a}}

\newcommand{\bb}{{\bf b}}

\newcommand{\bc}{{\bf c}}

\newcommand{\bd}{{\bf d}}

\newcommand{\be}{{\bf e}}

\newcommand{\bg}{{\bf g}}

\newcommand{\bh}{{\bf h}}

\newcommand{\bm}{{\bf m}}

\newcommand{\bn}{{\bf n}}

\newcommand{\bp}{{\bf p}}

\newcommand{\bq}{{\bf q}}

\newcommand{\br}{{\bf r}}

\newcommand{\bs}{{\bf s}}

\newcommand{\bt}{{\bf t}}

\newcommand{\bu}{{\bf u}}

\newcommand{\bv}{{\bf v}}

\newcommand{\bw}{{\bf w}}

\newcommand{\bx}{{\bf x}}

\newcommand{\by}{{\bf y}}

\newcommand{\bz}{{\bf z}}

\newcommand{\bC}{{\bf C}}

\newcommand{\bD}{{\bf D}}

\newcommand{\bH}{{\bf H}}

\newcommand{\bI}{{\bf I}}

\newcommand{\bM}{{\bf M}}

\newcommand{\bP}{{\bf P}}

\newcommand{\bT}{{\bf T}}

\newcommand{\bU}{{\bf U}}

\newcommand{\bV}{{\bf V}}

\newcommand{\bW}{{\bf W}}

\newcommand{\bX}{{\bf X}}

\newcommand{\bY}{{\bf Y}}

\newcommand{\bZ}{{\bf Z}}

\newcommand{\bone}{\mbox{\boldmath $1$}}

\newcommand{\balpha}{\mbox{\boldmath $\alpha$}}

\newcommand{\bbeta}{\mbox{\boldmath $\beta$}}

\newcommand{\bdelta}{\mbox{\boldmath $\delta$}}

\newcommand{\bepsilon}{\mbox{\boldmath $\epsilon$}}

\newcommand{\bgamma}{\mbox{\boldmath $\gamma$}}

\newcommand{\blambda}{\mbox{\boldmath $\lambda$}}

\newcommand{\bmu}{\mbox{\boldmath $\mu$}}

\newcommand{\bomea}{\mbox{\boldmath $\omega$}}

\newcommand{\bpi}{\mbox{\boldmath $\pi$}}

\newcommand{\bphi}{\mbox{\boldmath $\phi$}}

\newcommand{\brho}{\mbox{\boldmath $\rho$}}

\newcommand{\bsigma}{\mbox{\boldmath $\sigma$}}

\newcommand{\bSigma}{\mbox{\boldmath $\Sigma$}}

\newcommand{\btau}{\mbox{\boldmath $\tau$}}

\newcommand{\btheta}{\mbox{\boldmath $\theta$}}

\newcommand{\bpsi}{\mbox{\boldmath $\psi$}}

\newcommand{\bxi}{\mbox{\boldmath $\xi$}}

\newcommand{\bzeta}{\mbox{\boldmath $\zeta$}}

\newcommand{\bfeta}{\mbox{\boldmath $\eta$}}

\newcommand{\n}{\frac{1}{n}}

\newcommand{\E}{{\rm I\!E}}

\newcommand{\D}{{\rm I\!D}}

\newcommand{\R}{{\rm I\!R}}

\newcommand{\ls}{{\cal R}}

\newcommand{\var}{{\rm var}}

\newcommand{\cov}{{\rm cov}}

\setlength{\oddsidemargin}{5mm}

\setlength{\evensidemargin}{5mm}

\setlength{\textwidth}{150mm}

\setlength{\marginparsep}{0pt}

\setlength{\marginparwidth}{0pt}

\setlength{\topmargin}{0pt}

\setlength{\headsep}{0pt}

\setlength{\textheight}{225mm}

\setlength{\footskip}{1.5cm}

\setlength{\parskip}{4pt}

%\definecolor{sapgreen}{rgb}{.1,.4,.1}

\definecolor{brick}{rgb}{.5,.0,.0}

\definecolor{darkblue}{rgb}{.0,.0,.5}

%\definecolor{black}{rgb}{0,1,0}

\definecolor{gray}{gray}{.5}

\pagestyle{plain}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Starting

\title{\huge \bf Thesis title\\}

\author{{\bf \Large Peter Francis Hickey}\\\\
Submitted in total fulfilment of the requirements\\
of the degree of Doctor of Philosophy\\\\
November 2014\\\\
Department of Mathematics and Statistics\\
The University of Melbourne\\\\}

\date{The Walter and Eliza Hall Institute of Medical Research}

\maketitle

\frontmatter

\chapter{Abstract}

% should be between 300 - 500 words

Microarrays are a popular tool for measuring relative gene expression between RNA populations for thousands of genes simultaneously...

\chapter{Declaration}

{\bf This is to certify that}

\BI

\I[(i)] the thesis comprises only my original work towards

the PhD except where indicated in the Preface,

\I[(ii)] due acknowledgement has been made in

the text to all other material used,

\I[(iii)] the thesis is less than 100,000 words in length,

exclusive of tables, maps, bibliographies and appendices.

\EI

\vspace{2cm}

{\large\bf Signed,\\\\\\\\}

\chapter{Preface}

* Some data in my thesis are mapped to hg18 and some to hg19.
* Include summary of notation?

The data sets used in this thesis were provided by the investigators listed below.\\\\

\chapter{Acknowledgments}

I am grateful to the following people for their support during my PhD...

\tableofcontents

\listoffigures

\listoftables

\chapter{List of Abbreviations}

\begin{tabular}{ll}

bp & base pair \\
kb & kilobase = $1000$ bp \\
Mb & megabase = $1,000,000$ bp \\
Gb & gigabase = $1,000,000,000$ bp \\
WGBS & whole genome bisulfite sequencing \\
RRBS & reduced representation bisulfite sequencing \\
FDR & false discovery rate \\
DMR & differentially methylated region \\
DMC & differentially methylated cytosine \\

\end{tabular}

\chapter{List of Symbols}

\begin{tabular}{ll}

i.e. & that is\\

\end{tabular}

\mainmatter

% each chapter is in a separate .tex file

\include{introduction} % introduction.tex

\include{a_statistical_model_of_methlyC-seq} % a_statistical_model_of_methlyC-seq.tex

%\include{backgroundcorrect} % backgroundcorrect.tex

% etc.

\backmatter

\chapter{Concluding Comments}

\label{chapter:Conclusions}

% 2 - 4 pages

In this thesis, several fundamental low-level analysis issues for microarrays have been considered.

\bibliography{phd_thesis}

\chapter{Appendix A}

\label{chapter:AppendixCode}

Insert supplementary figures/R code here.

\end{document}

### 2014-04-28 latex/phd_thesis.tex
Finished description of DSS. Described MOABS and DMAP. Describe inference framework and simulation model in Lacey et al. Need to fix - LaTeX doesn't compile

@@
 \bibpunct{[}{]},a{},
+
+% TODO: Remove all unused macros
 
@@
 DMC & differentially methylated cytosine \\
+HMM & hidden Markov model \\
+CNV & copy number variant \\
 

### 2014-05-15 latex/phd_thesis.tex
Began section on parameter estimation.

--- 2014-04-23 latex/phd_thesis.tex
@@
 DMC & differentially methylated cytosine \\
+HMM & hidden Markov model \\
+CNV & copy number variant \\
+SE & single-end sequencing \\
+PE & paired-end sequencing \\
 

### 2014-08-12 latex/phd_thesis.tex
Began section on typical analysis of (whole genome) bisulfite sequencing by moving some stuff from introduction. Then wrote about challenges of PCR duplicate detection.
Added filtering of wgbs data. Closes #7
Experimental design for identifying differential methylation. One-group DMC testing.
Began section on calling DMCs in two-group experiments

--- 2014-04-23 latex/phd_thesis.tex
@@
 
+\newcommand{\R}{{\textbf{\textsf{R}}}}}
+
 \setlength{\oddsidemargin}{5mm}
@@
 DMC & differentially methylated cytosine \\
+HMM & hidden Markov model \\
+CNV & copy number variant \\
+SE & single-end sequencing \\
+PE & paired-end sequencing \\
+indel & insertion or deletion \\
 
@@
 
+\include{wgbs_analysis} % wgbs_analysis.tex
+
 \include{a_statistical_model_of_methlyC-seq} % a_statistical_model_of_methlyC-seq.tex
-
-%\include{backgroundcorrect} % backgroundcorrect.tex
-
-% etc.
 

### 2014-09-18 Rmarkdown/comethylation.Rmd
Started on comethylation chapter

---
title: "Co-methylation"
author: "Peter Hickey"
date: "18 September 2014"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
---

## Chapter overview

Co-methylation is the dependence of DNA methylation at different methylation loci. As such, it is a kind of spatial dependence where the "space" is the genome. In this chapter I describe two measures of co-methylation, namely, _within-fragment co-methylation_ and _correlation of $\beta$-values_. Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci.

Throughout this chapter I focus on CpGs although the methods described are also applicable to non-CpG methylation loci. The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples I am studying.

## Within-fragment co-methylation

Within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$, here assuming a single DNA fragment for each chromosome in the sample (i.e. no fragmentation of the DNA). I would like to learn how, for example, $Z_{h, i}$ affects $Z_{h, i'}$ and develop a simple, tractable model of this dependence structure. In particular, I am interested in how this dependence varies as a function of the distance between methylation loci.

As every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, my studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).

In studying within-fragment co-methylation, I assume that all copies of each chromosome have the same dependence structure (__TODO: Although not the same mean structure? Make this mathematically precise__). The data are sampled from a population containing multiple copies of each chromosome and if each copy has a different dependence structure then we cannot hope to say anything about these structures since these copies are not identifiable.

If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci. However, current technologies for bisulfite-sequencing mean that we are limited to studying reads that, at most, span about $200-250$ bp and so are limited to studying co-methylation of methylation loci that are very close to one another.

### Using pairs of methylation loci

The simplest summary of within-fragment co-methylation uses pairs (2-tuples) of methylation loci. Consider a single pair of CpGs in the reference genome and all reads that contain both CpGs of the pair (see Figure \ref{reads_containing_pair_of_CpGs}).

\begin{figure}[h!]
  \centering
    \includegraphics[width=0.5\textwidth]{../figures/reads_containing_pair_of_CpGs.pdf}
      \caption{All reads containing a pair of CpGs. The CpGs are highlighted by the short blue bars along the bottom of the figure. Each long grey (resp. green) bar is a read mapped to the forward (resp. reverse) strand. Those reads highlighted in yellow contain both CpGs of the pair. Data are from the IMR90-iPSC sample.}
      \label{reads_containing_pair_of_CpGs}
\end{figure}

We tabulate the methylation states of each of these reads as shown in Table \ref{reads_containing_pair_of_CpGs_counts}. The `methtuple` software can perform this tabulation for all pairs of methylation loci in a sample.

The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the log-odds ratio of this table, $LOR_{(i, i'), j} = log_2 \Big (\frac{MM_{i, i'} \times UU_{i, i'}}{MU_{i, i'} \times UM_{i, i'}} \Big)$. For the data shown in Figure \ref{reads_containing_pair_of_CpGs} the $LOR =$ `r round(log2((1.5 * 4.5) / (0.5 * 2.5)), 1)`.

\begin{table}[h]
  \centering

  \begin{tabular}{lcccl}
                           & \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
                           &                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5} 
\multirow{2}{*}{First CpG} & \textbf{M}           & 1              & 2             & 3              \\
                           & \textbf{U}           & 0              & 4             & 4              \\ \cline{2-5} 
                           & \textbf{Total}       & 1              & 6             & 7             
\end{tabular}
\caption{Counts of each methylation pattern for reads shown in Figure \ref{reads_containing_pair_of_CpGs}}
\label{reads_containing_pair_of_CpGs_counts}
\end{table}

A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs on the same DNA fragment (_co-methylation_), a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG on the same DNA fragment and a value close to zero is evidence that the methylation state at each CpG on the same DNA fragment are independent of one another. More precisely, we may wish to compute a confidence interval for $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.

In theory we can compute $LOR$ for all pairs of methylation loci. In practice, using current technology, we can only compute it for pairs of methylation loci that are at most separated by 200-250 bp.

#### Constructing pairs of methylation loci

In addition to the constraints imposed by the sequencing technology on the construction of pairs of methylation loci, there are user-specified constraints.

For a sequence containing $k$ methlyation loci there are $\binom{k}{2}$ pairs, which is $\mathcal{O}(k^{2})$ as $\lim_{k \to \infty}$. In contrast, there $k - 1$ pairs of _neighbouring_ methylation loci (i.e., pairs where the _number of intervening methylation loci_ equal to zero, $NIML = 0$), which is $\mathcal{O}(k)$ as $\lim_{k \to \infty}$.

Apart from this difference in the number of pairs, the interpretation of co-methylation at neighbouring loci is far simpler than for non-neighbouring loci and so it may be preferable to focus on neighbouring loci. For example, if we observe $LOR >> 0$ for a pair of CpGs, $(CpG_{1}, CpG_{5})$, with $NIML > 0$ then we need to take care to ensure that this is due to a genuine dependence between these two CpGs above and beyond the pairwise dependencies $(CpG_{1}, CpG_{2}), \ldots, (CpG_{4}, CpG_{5})$.

The `methtuple` software by default tries to only construct pairs of neighbouring methylation loci[^all-combinations]. However, because `methtuple` processes each read independently, filters methylation loci on user-specified criteria and is unaware of the reference genome it will occasionally create pairs that are not truly neighbouring. Therefore, if strictly neighbouring pairs are required, non-neighbouring pairs should be post-hoc filtered (__TODO: e.g. using the `whatFunction` in `myRMethTuplePackage`__).

[^all-combinations]: `methtuple` also provides the option `--all-combinations` to specifically create all $\binom{k}{m}$ $m$-tuples. However the computational requirements are generally prohibitive for $m > 2$.

A further consideration is whether to collapse CpGs on opposite strands into the same methylation loci. By default, `methtuple` does not collapse CpGs across strands but this option is available via the `--strand-collapse` option or as a post-hoc step in __TODO: e.g. using the `whatFunction` in `myRMethTuplePackage`__. While CpG methylation is typically strand-symmetric it may be better to not collapse across strands until there is evidence that this will not be harmful, e.g., data from the strand-specific CpGs look similar.

### Using higher-order $m$-tuples of methylation loci

__TODO__

Complexities:
* $2^{m - 1}$ $LOR$s
* Incorporating distance

### Results of co-methylation analyses

### Markov models of within-fragment co-methylation

The

A $k$-order Markov model of within-fragment co-methylation assumes that the dependence of $Z_{h, i}$ on the "history" of the process, $(Z_{h, 1}, \ldots, Z_{h, i - 1})$ is through the $k$ most recent observations, $Z_{h, i - k}, \ldots, Z_{h, i - 1}$. This could be generalised via a Markov random field model of within-fragment co-methylation to assume that the dependence of $Z_{h, i}$ on the "rest" of the process $Z_{h, 1}, \ldots, Z_{h, i - 1}, Z_{h, i + 1}, Z_{N_{loci}}$ is through the k-nearest observations, $Z_{h, i - k_{1}}, \ldots, Z_{h, i - 1}, Z_{h, i + 1}, \ldots, Z_{h, i + k_{2}}$, where $k_{2} - k_{1} = k$.

Regardless of whether we use a Markov chain or a Markov random field, a complication is that $Z_{h}$ is unevenly spaced along the genome. I use the intra-pair distance ($IPD$) to measure the distance between observations along $Z_{h}$. This distance measure introduces additional dimension(s) to our model.

### 2014-09-29 Rmarkdown/comethylation.Rmd
Visualising co-methylation. Using higher-order m-tuples to explore co-methylation.

@@
 author: "Peter Hickey"
-date: "18 September 2014"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
       in_header: ../latex/header.tex
+  html_document:
+    keep_md: true
 ---
@@
 
-Co-methylation is the dependence of DNA methylation at different methylation loci.
-As such, it is a kind of spatial dependence where the "space" is the genome.
+Co-methylation is the dependence of DNA methylation at different methylation loci within a sample.
+As such, it is a kind of spatial dependence where the "space" is the genome of the sample.
 In this chapter I describe two measures of co-methylation, namely, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
@@
 
-Throughout this chapter I focus on CpGs although the methods described are also applicable to non-CpG methylation loci.
+I begin by explaining various methods for estimating co-methylation and their associated challenges.
+I then use these methods to analyse more than fifty whole-genome bisulfite-sequencing libraries to study the variation of co-methylation within and between different cell types.
+
+Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
 The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples I am studying.
@@
 Within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
-It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$, here assuming a single DNA fragment for each chromosome in the sample (i.e.
-no fragmentation of the DNA).
-I would like to learn how, for example, $Z_{h, i}$ affects $Z_{h, i'}$ and develop a simple, tractable model of this dependence structure.
-In particular, I am interested in how this dependence varies as a function of the distance between methylation loci.
+It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
+For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure.
+We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
 
-As every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, my studying the methylation patterns at m-tuples.
+As every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, by studying the methylation patterns at m-tuples.
 Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.
@@
 If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci.
-However, current technologies for bisulfite-sequencing mean that we are limited to studying reads that, at most, span about $200-250$ bp and so are limited to studying co-methylation of methylation loci that are very close to one another.
+However, current technologies for bisulfite-sequencing mean that we are limited to studying reads that, at most, span about $200-250$ bp and so are limited to studying co-methylation of methylation loci that are very close to one another along the chromosome.
 
@@
 
-A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs on the same DNA fragment (_co-methylation_), a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG on the same DNA fragment and a value close to zero is evidence that the methylation state at each CpG on the same DNA fragment are independent of one another.
+A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs in the pair on the same DNA fragment (_co-methylation_), a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG in the pair on the same DNA fragment and a value close to zero is evidence that the methylation state at each CpG in the pair on the same DNA fragment are independent of one another.
 More precisely, we may wish to compute a confidence interval for $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.
@@
 
-In addition to the constraints imposed by the sequencing technology on the construction of pairs of methylation loci, there are user-specified constraints.
+There are additional considerations in constructing pairs of methylation loci on top of those constraints imposed by the sequencing technology.
 
-For a sequence containing $k$ methlyation loci there are $\binom{k}{2}$ pairs, which is $\mathcal{O}(k^{2})$ as $\lim_{k \to \infty}$.
-In contrast, there $k - 1$ pairs of _neighbouring_ methylation loci (i.e., pairs where the _number of intervening methylation loci_ equal to zero, $NIML = 0$), which is $\mathcal{O}(k)$ as $\lim_{k \to \infty}$.
+For a sequence containing $k$ methylation loci there are $\binom{k}{2}$ pairs, which is $\mathcal{O}(k^{2})$ as $\lim_{k \to \infty}$.
+In contrast, there are $k - 1$ pairs of _neighbouring_ methylation loci (i.e., pairs where the _number of intervening methylation loci_ equal to zero, $NIML = 0$), which is $\mathcal{O}(k)$ as $\lim_{k \to \infty}$.
 
 Apart from this difference in the number of pairs, the interpretation of co-methylation at neighbouring loci is far simpler than for non-neighbouring loci and so it may be preferable to focus on neighbouring loci.
-For example, if we observe $LOR >> 0$ for a pair of CpGs, $(CpG_{1}, CpG_{5})$, with $NIML > 0$ then we need to take care to ensure that this is due to a genuine dependence between these two CpGs above and beyond the pairwise dependencies $(CpG_{1}, CpG_{2}), \ldots, (CpG_{4}, CpG_{5})$.
+For example, if we observe $LOR >> 0$ for a pair of CpGs, $(CpG_{1}, CpG_{5})$, with $NIML = 3 > 0$ then we need to take care to ensure that this is due to a genuine dependence between these two CpGs above and beyond the pairwise dependencies $(CpG_{1}, CpG_{2}), \ldots, (CpG_{4}, CpG_{5})$.
 
 The `methtuple` software by default tries to only construct pairs of neighbouring methylation loci[^all-combinations].
-However, because `methtuple` processes each read independently, filters methylation loci on user-specified criteria and is unaware of the reference genome it will occasionally create pairs that are not truly neighbouring.
+However, because `methtuple` processes each read independently, filters methylation loci on user-specified criteria and is unaware of the reference genome, it will occasionally create pairs that are not truly neighbouring.
 Therefore, if strictly neighbouring pairs are required, non-neighbouring pairs should be post-hoc filtered (__TODO: e.g.
@@
 
-A further consideration is whether to collapse CpGs on opposite strands into the same methylation loci.
-By default, `methtuple` does not collapse CpGs across strands but this option is available via the `--strand-collapse` option or as a post-hoc step in __TODO: e.g.
+Furthermore, while CpG methylation is typically strand-symmetric, it may be preferable to compute co-methylation separately for each strand.
+This will double the number of CpG 2-tuples.
+By default, `methtuple` does not collapse CpGs across strands, but this option is available via the `--strand-collapse` option or as a post-hoc step in __TODO: e.g.
 using the `whatFunction` in `myRMethTuplePackage`__.
-While CpG methylation is typically strand-symmetric it may be better to not collapse across strands until there is evidence that this will not be harmful, e.g., data from the strand-specific CpGs look similar.
+
+#### Visualisations
+
+For a typical whole-genome bisulfite-sequencing experiment of human DNA, tens of millions of neighbouring CpG pairs are sequenced (and hundreds of millions of non-neighbouring CpG pairs).
+For every pair of CpGs, $(i, i')$, in the $j^{th}$ sample we know the $IPD$ and have computed a log-odds ratio, $LOR$_{(i, i'), j}$.
+Each such pair may have further annotations, such as the genomic context of the pair (e.g.
+within a CpG island, within a gene body, within a repetitive region, etc.).
+We need a simple way to summarise these data so that we might explore the effect of each of these variables on within-fragment co-methylation.
+
+For any given $IPD$ there are hundreds of thousands of CpG pairs and so there is a _distribution_ of co-methylation for any given $IPD$.
+We can visualise the distribution for a given $IPD$ using classical visualisation techniques such as histograms, kernel density estimates and boxplots (__TODO: Example figures of each of these__).
+These summaries may be stratified by one or more annotations (__TODO: Example figures with stratification__).
+
+What these plots don't give the reader is a sense of how $IPD$ affects within-fragment co-methylation.
+__FIGURE__ and __FIGURE__ (__TODO:__ two plots of co-methylation as boxplots, one with outliers plotted and one without) show initial attempts to create a visualisation that shows the relationship between $LOR$ and $IPD$.
+Unfortunately, these plots are rather "busy".
+
+Instead, I plot selected quantiles of each $IPD$-specific $LOR$ distribution as a function of $IPD$ (__FIGURE__).
+These convey the same information as the series of boxplots in a simpler and cleaner fashion.
+I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$.
+These quantile plots may also be stratified by additional annotations (__FIGURE__).
+
+The simplicity and interpretability of the log-odds ratio make the above described method an attractive one for studying within-fragment co-methylation.
+However, by using 2-tuples we are limited to exploring "one-step" co-methylation, i.e., one-step dependencies of the stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
+In the next section I describe the difficulties of using higher order $m-tuples$ ($m > 2$) to explore within-fragment co-methylation.
 
@@
 
+In general, to explore $m$-step dependencies we need to use $(m + 1)$-tuples.
+We can generalise the use of a two-way contingency table for studying co-methylation of a 2-tuple to the use of an $m$-way contingency table for co-methylation of an m-tuple (see Table ~\ref{reads_containing_three_CpGs_counts}).
+
+__TODO: Add real numbers to table__
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}cccc@{}}
+\toprule
+           &            & \multicolumn{2}{c}{Third CpG} \\ \cmidrule(l){3-4}
+First CpG  & Second CpG & \textbf{M}    & \textbf{U}    \\ \midrule
+\textbf{M} & \textbf{M} &      4        &      0        \\
+\textbf{}  & \textbf{U} &      2        &      0        \\
+\textbf{U} & \textbf{M} &      0        &      1        \\
+\textbf{}  & \textbf{U} &      1        &      1        \\ \bottomrule
+\end{tabular}
+\caption{Example three-way table summarising read counts at a 3-tuple.}
+\label{reads_containing_three_CpGs_counts}
+\end{table}
+
+     chr strand pos1 pos2 pos3 MMM MMU MUM MUU UMM UMU UUM UUU cov
+ 4: chr1      +  489  493  497   4   0   2   0   0   1   1   0   8
+
+However, as we increase $m$ we have more parameters to estimate - an $m$-way table has $\binom{m}{2}$ two-way dependencies, $\binom{k}{3}$ three-way dependencies, $\ldots$, and $\binom{m}{m} = 1$ $m$-way dependency.
+
+Furthermore, as we increase $m$ we have fewer data points with which to estimate these parameters, since we only use reads that contain all $m$ methylation loci in the $m$-tuple (__TODO: Bar chart of how many CpGs per read__).
+This makes the estimates quite noisy unless some form of regularisation is applied, such as a parametric model of the dependencies or a non-parametric, data-driven smoothing of the estimates.
+I explore the use of the Expectation-Maximisation algorithm in __SECTION__ to incorporate all reads that overlap at least one methylation loci in the $m$-tuple.
+
+#### An EM algorithm to estimate co-methylation
+
+__TODO:__ Discuss with Terry the idea of fitting a log-linear model to the counts of methylation states using all reads that overlap at least one methylation loci in the tuple.
+Effectively, this is a log-linear model with missing data.
+
+#### Visualisations
+
+Aside from estimating this increased number of parameters, we must also interpret these with respect to $IPD$ and other annotations.
+The $IPD$ vector for $m$-tuples has $(m - 1)$ elements meaning we now have to explore the effect of $IPD$ on $LOR$ over an $(m - 1)$-dimensional _grid_ rather than along a 1-dimensional line.
+Again, the sparsity of the data make this difficult - approximately $20,000 - 30,000$ unique $IPD$s are observed for 3-tuples - but even more difficult is creating a clear and coherent visualisation of this data.
+
+As is clear from __FIGURE__, these is very difficult for 3-tuples, let alone for higher-order $m$-tuples.
+We have an $(m - 1)$ grid of $IPD$s and a _distribution_ of $LOR$ for each $IPD$.
+In the case of 2-tuples I used lines to summarise the quantiles of the $LOR$ distributions, but this is not effective when $m > 2$.
+Instead, we must create separate figures for each quantile, which makes it difficult to interpret the effects of $IPD$ on $LOR$.
+
+## Results of co-methylation analyses
+
 __TODO__
 
-Complexities:
-* $2^{m - 1}$ $LOR$s
-* Incorporating distance
+## A Markov model of within-fragment co-methylation
 
-### Results of co-methylation analyses
-
-### Markov models of within-fragment co-methylation
-
-The
+__Possibly better off in the chapter on simulation__
 

### 2014-09-29 R/EM_co-methylation.R
Preliminary code to explore EM algorithm for estimating co-methylation

@@ -1,0 +1,16 @@
+#### Exploring EM to estimate within-fragment co-methylation ####
+# Peter Hickey
+# 29/09/2014
+
+#### Some example data ####
+n <- 1000
+x <- data.frame(l1 = sample(c(0L, 1L, NA_integer_), size = n,
+                            prob = c(0.65, 0.3, 0.1), replace = TRUE),
+                l2 = sample(c(0L, 1L, NA_integer_), size = n,
+                            prob = c(0.65, 0.3, 0.1), replace = TRUE),
+                l3 = sample(c(0L, 1L, NA_integer_), size = n,
+                            prob = c(0.65, 0.3, 0.1), replace = TRUE)
+                )
+xx <- xtabs(~ ., data = x, exclude = NULL, na.action = na.pass)
+
+ll <- loglm(~ l1 + l2 + l3, xx)

### 2014-09-29 Rmarkdown/comethylation.Rmd
Started introduction and background to co-methylation.

@@
 
-## Within-fragment co-methylation
+\subsection{Previous work}\label{subsec:previous work}
+
+I define co-methylation is the dependence of DNA methylation at different methylation loci within a sample.
+In other words, it is the dependence or correlation structure of DNA methylation data within a sample.
+This is slightly different to \citeauthor{Schatz:2004cha} who defined it as, "the presence of methylation over a stretch of neighboring CpG positions" and more in line with \citeauthor{Eckhardt:2006gh} who defined it as, "the relationship between the degree of methylation over distance".
+
+There have been several attempts to explore the relationship between methylation events at different loci within an individual.
+However, frequently the exact question being asked and the methodology have not been well-defined.
+Moreover, to the best of my knowledge there is no software available that implements any of the proposed methods.
+
+From a biological perspective, by studying co-methylation we hope to better understand the regulation of DNA methylation.
+From a statistical perspective, more powerful and efficient analysis methods can be developed if this correlation structure is better understood, accounted for and exploited.
+
+An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation.
+For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci.
+This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays (__CITE__).
+
+### Correlations of aggregate measurements of methylation
+
+Most analyses of co-methylation have been based on aggregate measurements of methylation, such as $\beta$-values \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Hebestreit:2013ko, Lacey:2013iy, Liu:dy}.
+This is true of studies using both microarray data, where only these type of measurements are available, and sequencing data, where base-resolution measurements are also available.
+Very often the exact analysis is not well-described but it can generally be thought of as studying the autocorrelation of the $\beta$-values along the genome.
+
+__UP TO HERE__
+
+__TODO__
+
+* Describe some analyses using autocorrelation of beta-values.
+* More recently, within-fragment analyses. Describe concept (move from next section to this section).
+* Other definitions/methods
+  * Lister's mCG correlations
+  * co-methylation across samples
+  * "vertical" correlations of \citet{Capra:2014uh}
+
+The method I developed, which is described in Section \ref{subsec:wf_cometh}, is superior because of __TODO__.
+
+\subsection{Within-fragment co-methylation}\label{subsec:wf_cometh}
 
@@
 However, current technologies for bisulfite-sequencing mean that we are limited to studying reads that, at most, span about $200-250$ bp and so are limited to studying co-methylation of methylation loci that are very close to one another along the chromosome.
+Obviously, we cannot study within-fragment co-methylation from microarray data since these produce measurements of methylation that are aggregated across many hundreds or thousands of cells.
 
-### Using pairs of methylation loci
+\subsubsection{Using pairs of methylation loci}\label{subsec:pairs}
 
@@
 In theory we can compute $LOR$ for all pairs of methylation loci. In practice, using current technology, we can only compute it for pairs of methylation loci that are at most separated by 200-250 bp.
+
+#### Comparison of LOR to correlations
+
+__TODO: Compare use of log-odds ratio to use of correlation coefficient and justify using LOR__
 
@@
 
+## Correlations of $\beta$-values
+
 ## Results of co-methylation analyses

### 2014-09-30 Rmarkdown/comethylation.Rmd
Discussion of Eckhardt et al.

@@
     keep_md: true
+bibliography: ../latex/phd_thesis.bib
 ---
@@
 Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
-The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples I am studying.
+The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples that I am studying.
 
-\subsection{Previous work}\label{subsec:previous work}
+\section{Previous work}\label{subsec:previous work}
 
-I define co-methylation is the dependence of DNA methylation at different methylation loci within a sample.
+I define co-methylation as the dependence of DNA methylation at different methylation loci within a sample.
 In other words, it is the dependence or correlation structure of DNA methylation data within a sample.
-This is slightly different to \citeauthor{Schatz:2004cha} who defined it as, "the presence of methylation over a stretch of neighboring CpG positions" and more in line with \citeauthor{Eckhardt:2006gh} who defined it as, "the relationship between the degree of methylation over distance".
+This is slightly different to \cite{Schatz:2004cha} who defined it as, "the presence of methylation over a stretch of neighbouring CpG positions"; it is more in line with \cite{Eckhardt:2006gh} who defined it as, "the relationship between the degree of methylation over distance".
 
@@
 However, frequently the exact question being asked and the methodology have not been well-defined.
-Moreover, to the best of my knowledge there is no software available that implements any of the proposed methods.
+Moreover, to the best of my knowledge, there is no software available that implements any of the proposed methods.
 
@@
 
-Most analyses of co-methylation have been based on aggregate measurements of methylation, such as $\beta$-values \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Hebestreit:2013ko, Lacey:2013iy, Liu:dy}.
+To date, most analyses of co-methylation have been based on aggregate measurements of methylation, such as $\beta$-values \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Hebestreit:2013ko, Lacey:2013iy, Liu:dy}.
 This is true of studies using both microarray data, where only these type of measurements are available, and sequencing data, where base-resolution measurements are also available.
@@
 
+#### \cite{Eckhardt:2006gh}
+
+\cite{Eckhardt:2006gh} is an important paper in the field and has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}.
+\cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.
+PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data.
+Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays.
+Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.
+
+As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance".
+They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $ >2,000$ bp" and reference Figure 3a in support of this claim.
+However, what is plotted in Figure 3a is not "correlation" in the usual sense of the word.
+
+In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-priting this figure__).
+The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp.
+Therefore, it is not possible that the "percentage identical methylation" was obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.
+Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed.
+An email discussion with the senior author, Stephan Beck, was unable to clarify the definition.
+
+My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
+I will refer to our best guess of "percentage identical methylation" as  $PIM^{*}$ and discuss some of its properties in the next section.
+
+##### $PIM^{*}$
+
+Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $z_{i, h}$ is sampled, that is, $Pr(z_{h, i} = 1) = p_{h}$.
+For two distinct CpGs sequenced on two (potentially) distinct amplicons, $z_{h, i}$ and  $z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
+Not only may the two CpGs be from distinct amplicons, but the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels.
+This most general setting is described by $z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(z_{h, i}, z_{h', i'}) = c$.
+In this setting we can compute the theoretical $PIM^{*}$ as follows:
+
+\begin{align*}
+PIM^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
+        &= Pr(z_{h, i} = 0, z_{h', i'} = 0) + Pr(z_{h, i} = 1, z_{h', i'} = 1) \\
+        &= Pr(z_{h', i'} = 0 | z_{h, i} = 0) Pr(z_{h, i} = 0) +
+           Pr(z_{h', i'} = 1 | z_{h, i} = 1) Pr(z_{h, i} = 1) \\
+        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
+           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
+        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
+\end{align*}
+
 __UP TO HERE__
+
+__TODO: Write up proof in appendix and link to it__
+
+__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__
+
+In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases:
+
+__TODO: Derive results for special cases and illustrate them with a plot__
+
+1. $z_{h, i}$ Bernoulli($p_{h}$) and $z_{h, i'}$ Bernoulli($p_{h'}$) and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
+2. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'}$, and $cor(z_{h, i}, z_{h', i'}) \neq c$.
+3. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'} > 0.5$, and $cor(z_{h, i}, z_{h', i'}) \neq c$.
+4. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'}$, and $cor(z_{h, i}, z_{h', i'}) = c$.
+
+__TODO START: Re-write and simplify these special case results__
+For case 4 we have:
+
+\begin{align}
+\label{eq:pim_iid}
+PIM_{1}^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
+          &= Pr(Pr(z_{h, i} = 0, z_{h', i'} = 0)) + Pr(Pr(z_{h, i} = 1, z_{h', i'} = 1)) \\
+          &= (1 - p)^2 + p^2 \\
+\end{align}
+
+By differentiating $PIM_{ind}^{*}$ with respect to $p$ we see that $PIM_{iid} \geq 0.5$.
+
+For case 2 we have:
+
+\begin{align}
+\label{eq:pim_inid}
+PIM_{2}^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
+          &= Pr(Pr(z_{h, i} = 0, z_{h', i'} = 0)) + Pr(Pr(z_{h, i} = 1, z_{h', i'} = 1)) \\
+          &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \\
+\end{align}
+
+where $p_{h}$ (resp.
+$p_{h'}$) is the genome-wide average CpG methylation level for the sample from which the $h^{th}$ (resp.
+$h'^{th}) arose.
+Unlike $PIM_{iid}$, $PIM_{inid}$ may be less than 0.5.
+In fact, $0 \leq PIM_{inid} \leq 1$ for $0 \leq p_{h}, p_{h'} \leq 1$.
+__TODO END: Re-write and simplify these special case results__
+
+The plot of "percentage identical methylation" in Figure 3a of \cite{Eckhardt:2006gh} asymptotes at 0.4, which is less than the theoretical lowerbound under the assumption of indepdence \eqref{eq:pim_ind}.
+This is likely due to the fact that the amplicons, and thus the $z_{h, i}$, come from a variety of samples and genomic contexts with different average CpG methylation levels, $p$.
+Alternatively, it may be that it is not $PIM^{*}$ that is plotted in Figure 3a or that the methylation states of two CpGs separated by a large distance are anti-correlated, resulting in $PIM^{*} < 0.5$ (__TODO: Ref equation__).
 
@@
   * co-methylation across samples
-  * "vertical" correlations of \citet{Capra:2014uh}
+  * "vertical" correlations of \cite{Capra:2014uh}
 

### 2014-09-30 R/Eckhardt.R

@@ -1,0 +1,66 @@
+#### Trying to understand Eckhart et al. (2006) ####
+# Peter Hickey
+# 30/09/2014
+# There are attempts to understand and derive the "percentage identical
+# methylation" measurement that is shown in Figure 3a as evidence of
+# co-methylation.
+
+#### TODOs ####
+# Tidy up into a .Rmd file
+
+n1 <- 100
+n2 <- 100
+prob <- c(0.5, 0.5)
+x <- sample(c(0, 1), size = n1, replace = TRUE, prob = prob)
+y <- sample(c(0, 1), size = n2, replace = TRUE, prob = prob)
+# x <- c(rep(0, n1), rep(1, n2))
+# y <- c(rep(1, n1), rep(0, n2))
+
+z <- matrix(rep(NA, n1 * n2), ncol = n2)
+
+for (i in seq_len(n1)) {
+  for (j in seq_len(n2)) {
+    z[i, j]  <- x[i] == y[j]
+  }
+}
+
+sum(z) / length(z)
+
+p <- seq(0, 1, by = 0.01)
+pp <- expand.grid(p, p)
+ppp <- (1 - pp$Var1) * (1 - pp$Var2) + pp$Var1 * pp$Var2
+
+
+f <- function(p, c) {
+  ifelse(pmax(1 - 1/ p, 1 - 1 / (1 - p)) <= c,
+         2 * p * (1 - c) * (p - 1) + 1,
+         NA)
+}
+
+f(seq(0, 1, 0.01), 0)
+f(seq(0, 1, 0.01), 0.5)
+f(seq(0, 1, 0.01), 1)
+f(seq(0, 1, 0.01), -0.5)
+f(seq(0, 1, 0.01), -1)
+
+g <- function(p1, p2, c) {
+  # TODO: What is the sufficient condition on the value of c?
+  2 * p1 * p2 + 2 * c * p1 * (1 - p1) - (p1 + p2) + 1
+}
+
+g(pp$Var1, pp$Var2, 0)
+g(p, p, 0)
+stopifnot(all.equal(f(p, 0), g(p, p, 0)))
+g(pp$Var1, pp$Var2, 0.5)
+g(p, p, 0.5)
+stopifnot(all.equal(f(p, 0.5), g(p, p, 0.5)))
+g(pp$Var1, pp$Var2, 1)
+stopifnot(all.equal(f(p, 1), g(p, p, 1)))
[…]

### 2014-09-30 Rmarkdown/appendix.Rmd
Proof of the probability that two dependent Bernoulli random variables are identical

---
title: "Appendix"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

## The probability that two dependent Bernoulli random variables are identical

\cite{Lindqvist:1978fm} wrote a brief note on Bernoulli trials with dependence. Building on earlier work by \cite{Klotz:1973if}, Lindqvist parameterises the Bernoulli process $X_{1}, X_{2}, \ldots$ on $\{0, 1\}$ by the parameters $p = Pr(X_{i} = 1)$ and $c = cor(X_{i - 1}, X_{i})$ and shows that the transition matrix is given by

\begin{equation*}
\label{eq:transition_matrix_id}
\Pi =
 \begin{pmatrix}
  (1 - p) + cp & p (1 - c) \\
  (1 - p) (1 - c) & p + c (1 - p)
 \end{pmatrix}
\end{equation*}

From this we can compute the joint distribution,

\begin{equation*}
Pr(X_{1} = x_{1}, \ldots, X_{n} = x_{n}) = Pr(X_{1} = x_{1}) Pr(X_{2} = x_{2} | X_{2} = x_{2}) \cdots Pr(X_{n} = x_{n} | X_{n - 1} = x_{n - 1})
\end{equation*}

In particular, in the case $n = 2$ we can compute the probability that two dependent but identically distributed Bernoulli random variables are equal, $Pr(X_{1} = X_{2}) = Pr(X_{1} = 0, X_{2} = 0) + Pr(X_{1} = 1, X_{2} = 1)$.

To extend the above result to the probability that two dependent but __non-identically__ distributed Bernoulli random variables are equal simply requires that we derive the appropriate transition matrix.

Let $X_{i} \eqd Bernoulli(p_{i})$ for $i = 1, 2$. The transition matrix, $\Pi = \bigl(\begin{smallmatrix}
Pr(X_{2} = x_{2} | X_{1} = x_{1}) \end{smallmatrix} \bigr)$, is given by

\begin{equation*}
\label{eq:transition_matrix_nid}
\Pi =
 \begin{pmatrix}
  (1 - p_{2}) + c p_{1} & p{2} - c p_{1} \\
  \big [ (1 - p_{2}) + c p_{1} \big ] (1 - p_{1}) & \big [ p_{2} + c (1 - p_{1}) \big ] p_{1}
 \end{pmatrix}
\end{equation*}

We can then compute the desired probability

\begin{align*}
Pr(X_{1} = X_{2}) &= Pr(X_{1} = 0, X_{2} = 0) + Pr(X_{1} = 1, X_{2} = 1) \\
                  &= Pr(X_{2} = 0 | X_{1} = 0) Pr(X_{0} = 0) +
                     Pr(X_{2} = 1 | X_{1} = 1) Pr(X_{0} = 1) \\
                  &= \big [ (1 - p_{2}) + c p_{1} \big ] (1 - p_{1}) + 
                     \big [ p_{2} + c ( 1 - p_{1}) \big ] p_{1} \\
                  &= (1 - p_{1}) (1 - p_{2}) + c p_{1} \big [1 - p_{1}] +
                     p_{1} p_{2} + c p_{1} (1 - p_{1}) \\
                  &= (1 - p_{1}) (1 - p_{2}) + p_{1} p_{2} + 2 c p_{1} (1 - p_{1})
\end{align*}

### 2014-09-30 Rmarkdown/comethylation.Rmd
Further work on Eckhardt et al.

@@
 My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
-I will refer to our best guess of "percentage identical methylation" as  $PIM^{*}$ and discuss some of its properties in the next section.
-
-##### $PIM^{*}$
+I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.
 
@@
 
-__UP TO HERE__
-
-__TODO: Write up proof in appendix and link to it__
+This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp.
+$z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp.
+$X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.
 
@@
 
-In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases:
+In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:
 
-__TODO: Derive results for special cases and illustrate them with a plot__
+1.
+$PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
+2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
+3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.
 
-1. $z_{h, i}$ Bernoulli($p_{h}$) and $z_{h, i'}$ Bernoulli($p_{h'}$) and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
-2. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'}$, and $cor(z_{h, i}, z_{h', i'}) \neq c$.
-3. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'} > 0.5$, and $cor(z_{h, i}, z_{h', i'}) \neq c$.
-4. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'}$, and $cor(z_{h, i}, z_{h', i'}) = c$.
+As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$.
+In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp.
+$< 0.5$) for $c > 0$ (resp.
+$c < 0$).
+__TODO: Illustrate these special cases with a plot.__
 
-__TODO START: Re-write and simplify these special case results__
-For case 4 we have:
+To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$.
+If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp.
+However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.
 
-\begin{align}
-\label{eq:pim_iid}
-PIM_{1}^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
-          &= Pr(Pr(z_{h, i} = 0, z_{h', i'} = 0)) + Pr(Pr(z_{h, i} = 1, z_{h', i'} = 1)) \\
-          &= (1 - p)^2 + p^2 \\
-\end{align}
+Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$.
+I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.
 
-By differentiating $PIM_{ind}^{*}$ with respect to $p$ we see that $PIM_{iid} \geq 0.5$.
+Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis.
+Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.
 
-For case 2 we have:
+#### Other co-methylation analyses using aggregate measurements
 
-\begin{align}
-\label{eq:pim_inid}
-PIM_{2}^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
-          &= Pr(Pr(z_{h, i} = 0, z_{h', i'} = 0)) + Pr(Pr(z_{h, i} = 1, z_{h', i'} = 1)) \\
-          &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \\
-\end{align}
+#### Previous within-fragment co-methylation analyses
 
-where $p_{h}$ (resp.
-$p_{h'}$) is the genome-wide average CpG methylation level for the sample from which the $h^{th}$ (resp.
-$h'^{th}) arose.
-Unlike $PIM_{iid}$, $PIM_{inid}$ may be less than 0.5.
-In fact, $0 \leq PIM_{inid} \leq 1$ for $0 \leq p_{h}, p_{h'} \leq 1$.
-__TODO END: Re-write and simplify these special case results__
+* More recently, within-fragment analyses. Describe concept (move from next section to this section).
 
-The plot of "percentage identical methylation" in Figure 3a of \cite{Eckhardt:2006gh} asymptotes at 0.4, which is less than the theoretical lowerbound under the assumption of indepdence \eqref{eq:pim_ind}.
-This is likely due to the fact that the amplicons, and thus the $z_{h, i}$, come from a variety of samples and genomic contexts with different average CpG methylation levels, $p$.
-Alternatively, it may be that it is not $PIM^{*}$ that is plotted in Figure 3a or that the methylation states of two CpGs separated by a large distance are anti-correlated, resulting in $PIM^{*} < 0.5$ (__TODO: Ref equation__).
+#### Other definitions
 
-__TODO__
-
-* Describe some analyses using autocorrelation of beta-values.
-* More recently, within-fragment analyses. Describe concept (move from next section to this section).
-* Other definitions/methods
 * Lister's mCG correlations

### 2014-09-30 R/Eckhardt.R

@@ -26,7 +26,7 @@ for (i in seq_len(n1)) {
 
 sum(z) / length(z)
 
-p <- seq(0, 1, by = 0.01)
+p <- seq(0.01, 0.99, by = 0.01)
 pp <- expand.grid(p, p)
 ppp <- (1 - pp$Var1) * (1 - pp$Var2) + pp$Var1 * pp$Var2
 
@@ -37,11 +37,11 @@ f <- function(p, c) {
          NA)
 }
 
-f(seq(0, 1, 0.01), 0)
-f(seq(0, 1, 0.01), 0.5)
-f(seq(0, 1, 0.01), 1)
-f(seq(0, 1, 0.01), -0.5)
-f(seq(0, 1, 0.01), -1)
+f(p, 0)
+f(p, 0.5)
+f(p, 1)
+f(p, -0.5)
+f(seq(p, -1)
 
 g <- function(p1, p2, c) {
   # TODO: What is the sufficient condition on the value of c?

### 2014-10-01 Rmarkdown/comethylation.Rmd
Reviewed co-methylation work in Cokus (2008), Chodavarapu (2010), Li (2010) and Lyko (2010).

@@
 
+#### \cite{Cokus:2008fc}
+
+\cite{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing.
+They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_.
+They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis].
+The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance.
+To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.
+
+[^arabidopsis]: Many plant species, such as _A. thaliana_, have much higher levels of non-CG methylation than do mammals.
+
+\cite{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci.
+From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin".
+They also identified correlations between different methylation contexts, e.g.
+CG vs.
+CHG, which "suggest[s] complex interactions between the different types of methylation".
+
+The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only).
+Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier transform analysis of the autocorrelation.
+
+[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}
+
+The $167$ nucleotide periodicity is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al.
+hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.
+
+The $10$ nucleotide periodicity is equal to the length of one helical DNA turn.
+From this result Cokus et al.
+hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites 8-10 apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_) (__TODO: If I want a picture explaining this then Figure 3b from \cite{Jurkowska:2011dt} might form the basis for my own illustration).
+However, in a later paper by the same group, this $10$ nucleotide periodicity was also observed for CG and CHG methylation, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}.
+
+[^comethylation_def2]: This is referred to as "co-methylation", here meaning the methylation of multiple loci in a single event, by \cite{Jurkowska:2011dt}.
+This article reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.
+
+The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})
+
+#### \cite{Chodavarapu:2010iqa}
+
+\cite{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (HSF1).
+The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq.
+The nucleosome start sites anchor the methylation observations to a common grid and is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes.
+A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
+Again, the observed periodicity is confirmed by analysing a Fourier transform of these plots.
+
+In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the HSF1 sample, which confirmed the $10$ nucleotide periodicity.
+This led \cite{Chodavarapu:2010iqa} to propose an extended version of their earlier hypothesis \cite{Cokus:2008fc}: namely, "that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.
+
+#### \cite{Li:2010fb}
+
+\cite{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
+The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $> 1,000$ bp", citing their results as being consistent with those in \cite{Eckhardt:2006gh}.
+In addition to this genome-wide analysis, \cite{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements.
+Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.
+
+Unfortunately, just as in \cite{Eckhardt:2006gh}, \cite{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available.
+All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions.
+So, once again, I have been forced to make my best guess as to what these figures show and how it was computed.
+
+Supplementary Figure 6a plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp.
+Based on the "$n = 18,936,995$" quoted in the header of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "istance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs.
+If all pairs were used this $n$ would be much higher since there are more than twenty million CpGs on one strand of the haploid human genome.
+To emphasise, the correlation shown by \cite{Li:2010fb} is not the same as the autocorrelation computed in \cite{Cokus:2008fc} since the autocorrelation uses all pairs of CpGs, regardless of the number of intervening CpGs.
+
+The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \cite{Cokus:2008fc} (highlighted in Supplementary Figure 7).
+This is further evidence that nucleosome spacing effects the strength of co-methylation.
+However, \cite{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \cite{Cokus:2008fc} and \cite{Chodavarapu:2010iqa}.
+
+#### \cite{Lyko:2010dr}
+
+In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation (Supplementary Figure 7).
+They report that the "correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
+This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report?
+Check my data any such increases__).
+__TODO: Is this the fastest decay of correlation in any species?__ However, I am concerned that the maximum autocorrelation is less $0.015$, which doesn't seem to make sense and is very close to zero.
+This may simply be due to a mislabelled or "normalised" value being plotted as the "correlation".
+(__TODO: I have emailed the senior author, Ryszard Maleszka, to try to resolve this concern__).
+
+Additionally, \cite{Lyko:2010dr} identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__).
+They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.
+
+Since the authors specifically describe their analysis as an "autocorrelation", I believe it uses all pairs of CpGs regardless of the number of intervening CpGs.
+
 #### Other co-methylation analyses using aggregate measurements
 
-#### Previous within-fragment co-methylation analyses
+* \cite{Hebestreit:2013ko, Lacey:2013iy, Liu:dy}
+
+\paragraph{Previous within-fragment co-methylation analyses}
+\label{sec:previous_within-fragment_analyses}
 
 * More recently, within-fragment analyses. Describe concept (move from next section to this section).
+* Cokus
 

### 2014-10-02 Rmarkdown/comethylation.Rmd
Reviewed Lacey (2013), Hebestreit (2013) and Liu (2014).

@@
 
-### Correlations of aggregate measurements of methylation
+\subsection{Correlations of aggregate measurements of methylation}
+\label{sec:previous_cor_of_agg_methylation}
 
-To date, most analyses of co-methylation have been based on aggregate measurements of methylation, such as $\beta$-values \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Hebestreit:2013ko, Lacey:2013iy, Liu:dy}.
-This is true of studies using both microarray data, where only these type of measurements are available, and sequencing data, where base-resolution measurements are also available.
-Very often the exact analysis is not well-described but it can generally be thought of as studying the autocorrelation of the $\beta$-values along the genome.
+To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
+These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).
+
+Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.
 
@@
 
-\cite{Eckhardt:2006gh} is an important paper in the field and has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}.
+\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples.
+Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}.
 \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.
@@
 
-#### Other co-methylation analyses using aggregate measurements
+#### \cite{Lacey:2013iy}
 
-* \cite{Hebestreit:2013ko, Lacey:2013iy, Liu:dy}
+\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2).
+The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
 
-\paragraph{Previous within-fragment co-methylation analyses}
+Caution is warranted in extrapolating these results to statements about genome-wide co-methylation.
+Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs (__TODO: Check number__) in the haploid human genome.
+Secondly, these data are from a single sample from a single tissue, which cannot capture the sample-to-sample nor tissue-to-tissue variability in co-methylation.
+Finally, RRBS targets CpG dense regions of the genome with low levels of methylation and so these data do not capture co-methylation outside of these regions, where the majority of the CpGs are found and which are highly methylated.
+
+[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (__TODO: Cite and link to appendix__).
+
+\subsection{Within-fragment co-methylation analyses}
 \label{sec:previous_within-fragment_analyses}
@@
 * Cokus
+* \cite{Lacey:2009by}
+* \cite{Paliwal:2013fk}
 
-#### Other definitions
+\subsection{Other related work}
+\label{sec:previous_other_related_work}
+
+Other definitions of co-methylation and related work:
 
@@
 
+#### \cite{Hebestreit:2013ko}
+
+The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}.
+This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.
+
+#### \cite{Liu:dy}
+
+The focus of this study on identify clusters of correlated CpGs and relating methylation changes at these clusters to SNP genotypes.
+The authors compute the correlation of pairs of $\beta$-values across samples, which is analogous to estimating the linkage disequilibrium between a pair of single nucleotide polymorphisms (SNPs) across a set of samples.
+This is different to computing autocorrelations of $\beta$-values, which are computed within, rather than between, samples.
+An autocorrelation analysis implicitly assumes stationarity along the genome whereas this analysis assumes stationarity across samples (__TODO: Discuss with Terry__).
+
+The discovery cohort contains 247 human samples typed on the Illumina HumanMethylation450 array.
+After normalisation and other pre-processing steps, they identify the top-$25\%$ most variably methylated CpG sites (vCpGs) in the 247 samples.
+They then compute Pearson correlations of methylation levels for all pairs of vCpGs across the samples.
+
+From this they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set".
+However, they also note that "methylation codependency (co-methylation) itself varies across the genome", as is to be expected.
+
 The method I developed, which is described in Section \ref{subsec:wf_cometh}, is superior because of __TODO__.
 
-\subsection{Within-fragment co-methylation}\label{subsec:wf_cometh}
+\section{Correlations of $\beta$-values}\label{subsec:beta_cor}
+
+__TODO__
+
+\section{Within-fragment co-methylation}\label{subsec:wf_cometh}
 
@@
 
-\subsubsection{Using pairs of methylation loci}\label{subsec:pairs}
+\subsection{Using pairs of methylation loci}\label{subsec:pairs}
 
@@
 
-#### Comparison of LOR to correlations
+### Comparison of LOR to correlations
 
@@
 
-#### Constructing pairs of methylation loci
+### Constructing pairs of methylation loci
 
@@
 
-#### Visualisations
+### Visualisations
 
@@
 
-### Using higher-order $m$-tuples of methylation loci
+## Using higher-order $m$-tuples of methylation loci
 
@@
 
-#### An EM algorithm to estimate co-methylation
+### An EM algorithm to estimate co-methylation
 
@@
 Instead, we must create separate figures for each quantile, which makes it difficult to interpret the effects of $IPD$ on $LOR$.
-
-## Correlations of $\beta$-values
 

### 2014-10-02 Rmarkdown/appendix.Rmd
Further work on Eckhardt et al.
Some analysis of Lacey (2013).

@@
 \end{align*}
+
+__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__
+
+## Analysis of \cite{Lacey:2013iy}
+
+I want to estimate the number of CpGs used by \cite{Lacey:2013iy} to estimate the correlation of methylation levels.
+They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, MTCTL2.
+Unfortunately, the data are not publicly available.
+Instead, I use data from ENCODE of a skeletal muscle myotubes cell line that had RRBS performed in triplicate (HSMMtube).
+
+Firstly, I download the data and unzip it:
+```{r, cache = TRUE, message = FALSE}
+library(R.utils)
+library(data.table)
+library(dplyr)
+
+tmp <- tempdir()
+url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
+files <- paste0(url, 'wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep', 
+                     1:3, '.bed.gz')
+dests <- paste0(tmp,'/rep', 1:3, '.bed.gz')
+x <- mapply(function(f, d) {
+  download.file(f, d)
+  gunzip(filename = d, overwrite = TRUE)
+  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t')
+  }, files, dests, SIMPLIFY = FALSE)
+```
+
+Then I count the number of CpGs on chromosome 11 in each sample:
+```{r}
+y <- lapply(x, function(xx) {
+  xx <- tbl_dt(xx)
+  xx %>% 
+    group_by(V1) %>% 
+    summarise(n())
+})
+
+rbind_all(y) %>% filter(V1 == 'chr11')
+```
+
+### TODOs
+
+* Cite data source and publication

### 2014-10-03 Rmarkdown/comethylation.Rmd
Reviewed within-fragment co-methylation of Cokus (2008).

@@
 
-#### \cite{Cokus:2008fc}
+#### \cite{Cokus:2008fc}\label{sec:Cokus}
 
@@
 
-* More recently, within-fragment analyses. Describe concept (move from next section to this section).
-* Cokus
+Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules.
+I call this within-fragment co-methylation, where fragment refers to a fragment of DNA.
+More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
+It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
+For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure.
+We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
+
+Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads.
+That is, by studying the methylation patterns at m-tuples.
+Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.
+
+[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).
+
+If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci.
+However, the most commonly used technology for high-throughput bisulfite-sequencing, which uses Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum.
+Assembling these reads into longer contigs will not solve this problem since this cannot guarantee that the assembled reads actually originated from the same DNA fragment.
+
+Long-read sequencing technologies, such as Pacific Biosciences (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__).
+An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).
+
+[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons.
+These include the fragility of large DNA fragments treated with sodium bisulfite (__TODO: Discuss with someone from AGRF__)
+
+To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides.
+There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.
+Here I review this previously published work.
+
+#### \cite{Cokus:2008fc}
+
+As already discussed in Section \ref{sec:Cokus}, \cite{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_.
+They find this same periodicity "when individual reads [were] examined directly".
+
+The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence.
+For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci.
+Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated.
+This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification].
+They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes.
+\cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
+This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages ().
+There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
+
+[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs.
+The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
+Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.
+
+I consider these results to use a restricted definition of within-fragment "co-methylation".
+These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment.
+Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$.
+From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$.
+It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
+While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to  estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
+
+Furthermore, by aggregating all pairs of CHH loci separated by a common distance and then estimating these probabilities, there is an implicit assumption that all such pairs on the same chromosome have the same conditional probability, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$.
+There is good reason to suspect that this is not true and that there exists significant pair-to-pair variability.
+For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for "co-methylation" of these same loci.
+
+__PREVIOUS WORK__
+
 * \cite{Lacey:2009by}
@@
 
-__TODO__
+__TODO: A spiel on how my method is better than existing ones.__
 
@@
 
-Within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
-It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
-For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure.
-We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
-
-As every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, by studying the methylation patterns at m-tuples.
-Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.
-
-[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).
+__TODO: A spiel on how my method is better than existing ones.__
 
@@
 The data are sampled from a population containing multiple copies of each chromosome and if each copy has a different dependence structure then we cannot hope to say anything about these structures since these copies are not identifiable.
-
-If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci.
-However, current technologies for bisulfite-sequencing mean that we are limited to studying reads that, at most, span about $200-250$ bp and so are limited to studying co-methylation of methylation loci that are very close to one another along the chromosome.
-Obviously, we cannot study within-fragment co-methylation from microarray data since these produce measurements of methylation that are aggregated across many hundreds or thousands of cells.
 

### 2014-10-03 Rmarkdown/comethylation.Rmd
Reviewed within-fragment co-methylation from Lacey (2009)

@@
 
+#### \cite{Lacey:2009by}
+
+\cite{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans.
+The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field where the methylation state of one loci, $Z_{h, i}$, depends on that of its neighbours either side, $Z_{h, i - 1}$ and $Z_{h, i + 1}$.
+The physical distances between these neighbours, the $IPD$s, are not considered.
+
+The neighbouring sites model requires an analysis of within-fragment co-methylation in order to estimate the transition probabilities.
+They estimate these parameters from two small regions, each approximately $200$ nucleotides in length and containing 13 and 14 CpGs, from 10 human ovarian carcinoma samples and a pool of somatic controls.
+These regions were sequenced using hairpin-bisulfite PCR, an assay that measures DNA methylation on both strands of a double-stranded DNA molecule[^hairpin_bs].
+For each region, $12-16$ clones were sequenced per sample[^clones]
+
+[^hairpin_bs]: This is unlike traditional bisulfite-sequencing, which only measures methylation along a single DNA strand.
+
+[^clones]: Essentially, $12-16$ reads were generated per sample.
+
+For each of these regions, \cite{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours.
+In other words, they found significant within-fragment co-methylation.
+They also found that the overall level of methylation of the CpGs in a region influenced the strength of the within-fragment co-methylation.
+
+In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a miniscule fraction of the human genome.
+While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.
+
 __PREVIOUS WORK__
 
-* \cite{Lacey:2009by}
 * \cite{Paliwal:2013fk}

### 2014-10-03 Rmarkdown/comethylation.Rmd
Tidying.

@@
 
-Long-read sequencing technologies, such as Pacific Biosciences (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__).
+Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__).
 An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).
@@
 While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.
-
-__PREVIOUS WORK__
-
-* \cite{Paliwal:2013fk}
 

### 2014-10-07 Rmarkdown/comethylation.Rmd
Review co-methylation of Lister (2009).
TODOs on re-structuring of chapter.

@@
 
+__TODO: Break out review of co-methylation into its own chapter.__
+
 Co-methylation is the dependence of DNA methylation at different methylation loci within a sample.
@@
 \section{Previous work}\label{subsec:previous work}
+
+__TODO: Incorporate material in "Correlations" section of `a_statistical_model_of_methylC-seq.md`.__
 
@@
 These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).
+
+The analysis in \cite{Lister:2009hy} takes a different approach.
+It too uses within-sample aggregate methylation levels, but rather than estimating correlations of the methylation levels, instead looks for patterns in the spacing of "methylcytosines" along the genome.
 
@@
 
+#### \cite{Lister:2009hy}
+
+\cite{Lister:2009hy} report an $8-10$ nucleotide periodicity in the occurence "methylcytosines".
+Recall that \cite{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a Binomial test (see Section \ref{sec:criticism_of_methylcytosines} for details and a criticism of this concept).
+Hence, while this may appear to be a within-fragment analysis of co-methylation, it in fact uses aggregate methylation levels.
+However, unlike the methods described in Section \ref{sec:previous_cor_of_agg_methylation}, the analysis of these aggregate methylation levels is not based on a correlation of these levels, but rather on unnormalised counts, which leads to problems.
+
+\cite{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$.
+These counts are plotted as a bar chart, with $IPD$ along the x-axis and count on the y-axis, along with a cubic spline fit to the counts.
+This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (Supplementary Figure 9 \cite{Lister:2009hy}).
+
+[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Supplementary Figure 9, but I am not certain.
+
+An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots.
+This periodicity is very weak several contexts and not at all present for exonic CpGs.
+The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts since there are very few "methylcytosines" for these contexts in these human samples.
+Furthermore, in all nine contexts the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.
+
+Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites".
+However, I suspect that the results in these nine plots are driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines.
+That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis.
+To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$.
+If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s.
+Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.
+
+In fact, there are well known periodicities in the distribution of nucleotides in genomes.
+For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \cite{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \cite{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.
+
+#### \cite{Landan:2012kp}
+
+__TODO: Review co-methylation work in this paper__
+
 \subsection{Within-fragment co-methylation analyses}
@@
 Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules.
-I call this within-fragment co-methylation, where fragment refers to a fragment of DNA.
+I call this within-fragment co-methylation, where a "fragment" refers to a fragment of DNA.
 More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
@@
 
-\subsection{Other related work}
-\label{sec:previous_other_related_work}
+\subsection{Between-sample co-methylation}
+\label{sec:between_sample_cometh}
 
-Other definitions of co-methylation and related work:
+As is clear from the above, there are multiple definitions of within-sample co-methylation.
+An fundamentally different concept to within-sample co-methylation is between-sample co-methylation, which is the correlation of methylation levels between a single pair of loci across a set of samples rather than the correlation of pairs of loci within a single sample.
+(__CITE__) estimate various forms of between-sample co-methylation and are reviewed below.
 
-* Lister's mCG correlations
-* co-methylation across samples
 * "vertical" correlations of \cite{Capra:2014uh}
+* MethLD of \cite{Liu:dy}
 
-#### \cite{Hebestreit:2013ko}
-
-The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}.
-This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.
+__TODO: Comment on utility of within-sample vs. between-sample co-methylation__
 
@@
 
-The method I developed, which is described in Section \ref{subsec:wf_cometh}, is superior because of __TODO__.
+\subsection{Exploiting co-methylation in analyses of differential methylation}
+\label{sec:exploiting_cometh}
+
+\cite{Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Sofer:2013bk} have each sought to exploit co-methylation, both the within-sample and between-sample forms, in developing more powerful and efficient ways to identify differential methylation.
+
+#### \cite{Hebestreit:2013ko}
+
+The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}.
+This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.
 

### 2014-10-08 Rmarkdown/comethylation.Rmd
Split co-methylation chapter.
Now split into “old stuff” (review of previous work) and “new stuff” (my work)

--- 2014-10-03 Rmarkdown/comethylation.Rmd
@@
 
-Co-methylation is the dependence of DNA methylation at different methylation loci within a sample.
-As such, it is a kind of spatial dependence where the "space" is the genome of the sample.
-In this chapter I describe two measures of co-methylation, namely, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
+In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
 Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci.
-
-I begin by explaining various methods for estimating co-methylation and their associated challenges.
-I then use these methods to analyse more than fifty whole-genome bisulfite-sequencing libraries to study the variation of co-methylation within and between different cell types.
+I apply these methods to more than fifty whole-genome bisulfite-sequencing libraries to study the between-sample variability of co-methylation.
 
@@
 The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples that I am studying.
-
-\section{Previous work}\label{subsec:previous work}
-
-I define co-methylation as the dependence of DNA methylation at different methylation loci within a sample.
-In other words, it is the dependence or correlation structure of DNA methylation data within a sample.
-This is slightly different to \cite{Schatz:2004cha} who defined it as, "the presence of methylation over a stretch of neighbouring CpG positions"; it is more in line with \cite{Eckhardt:2006gh} who defined it as, "the relationship between the degree of methylation over distance".
-
-There have been several attempts to explore the relationship between methylation events at different loci within an individual.
-However, frequently the exact question being asked and the methodology have not been well-defined.
-Moreover, to the best of my knowledge, there is no software available that implements any of the proposed methods.
-
-From a biological perspective, by studying co-methylation we hope to better understand the regulation of DNA methylation.
-From a statistical perspective, more powerful and efficient analysis methods can be developed if this correlation structure is better understood, accounted for and exploited.
-
-An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation.
-For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci.
-This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays (__CITE__).
-
-\subsection{Correlations of aggregate measurements of methylation}
-\label{sec:previous_cor_of_agg_methylation}
-
-To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
-These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).
-
-Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.
-
-#### \cite{Eckhardt:2006gh}
-
-\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples.
-Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}.
-\cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.
-PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data.
-Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays.
-Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.
-
-As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance".
-They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $ >2,000$ bp" and reference Figure 3a in support of this claim.
-However, what is plotted in Figure 3a is not "correlation" in the usual sense of the word.
-
-In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-priting this figure__).
-The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp.
-Therefore, it is not possible that the "percentage identical methylation" was obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.
-Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed.
-An email discussion with the senior author, Stephan Beck, was unable to clarify the definition.
-
-My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
-I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.
-
-Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $z_{i, h}$ is sampled, that is, $Pr(z_{h, i} = 1) = p_{h}$.
-For two distinct CpGs sequenced on two (potentially) distinct amplicons, $z_{h, i}$ and  $z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
-Not only may the two CpGs be from distinct amplicons, but the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels.
-This most general setting is described by $z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(z_{h, i}, z_{h', i'}) = c$.
-In this setting we can compute the theoretical $PIM^{*}$ as follows:
-
-\begin{align*}
-PIM^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
-        &= Pr(z_{h, i} = 0, z_{h', i'} = 0) + Pr(z_{h, i} = 1, z_{h', i'} = 1) \\
-        &= Pr(z_{h', i'} = 0 | z_{h, i} = 0) Pr(z_{h, i} = 0) +
-           Pr(z_{h', i'} = 1 | z_{h, i} = 1) Pr(z_{h, i} = 1) \\
-        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
-           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
-        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
-\end{align*}
-
-This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp.
-$z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp.
-$X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.
-
-__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__
-
-In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:
-
-1.
-$PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
-2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
-3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.
-
-As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$.
-In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp.
-$< 0.5$) for $c > 0$ (resp.
-$c < 0$).
-__TODO: Illustrate these special cases with a plot.__
-
-To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$.
-If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp.
-However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.
-
-Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$.
-I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.
-
-Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis.
-Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.
-
-#### \cite{Cokus:2008fc}\label{sec:Cokus}
-
-\cite{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing.
-They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_.
-They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis].
-The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance.
-To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.
-
-[^arabidopsis]: Many plant species, such as _A. thaliana_, have much higher levels of non-CG methylation than do mammals.
-
-\cite{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci.
-From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin".
-They also identified correlations between different methylation contexts, e.g.
-CG vs.
-CHG, which "suggest[s] complex interactions between the different types of methylation".
-
-The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only).
-Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier transform analysis of the autocorrelation.
-
-[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}
-
-The $167$ nucleotide periodicity is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al.
-hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.
-
-The $10$ nucleotide periodicity is equal to the length of one helical DNA turn.
-From this result Cokus et al.
-hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites 8-10 apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_) (__TODO: If I want a picture explaining this then Figure 3b from \cite{Jurkowska:2011dt} might form the basis for my own illustration).
-However, in a later paper by the same group, this $10$ nucleotide periodicity was also observed for CG and CHG methylation, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}.
-
-[^comethylation_def2]: This is referred to as "co-methylation", here meaning the methylation of multiple loci in a single event, by \cite{Jurkowska:2011dt}.
-This article reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.
-
-The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})
-
-#### \cite{Chodavarapu:2010iqa}
-
-\cite{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (HSF1).
-The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq.
-The nucleosome start sites anchor the methylation observations to a common grid and is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes.
-A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
-Again, the observed periodicity is confirmed by analysing a Fourier transform of these plots.
-
-In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the HSF1 sample, which confirmed the $10$ nucleotide periodicity.
-This led \cite{Chodavarapu:2010iqa} to propose an extended version of their earlier hypothesis \cite{Cokus:2008fc}: namely, "that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.
-
-#### \cite{Li:2010fb}
-
-\cite{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
-The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $> 1,000$ bp", citing their results as being consistent with those in \cite{Eckhardt:2006gh}.
-In addition to this genome-wide analysis, \cite{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements.
-Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.
-
-Unfortunately, just as in \cite{Eckhardt:2006gh}, \cite{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available.
-All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions.
-So, once again, I have been forced to make my best guess as to what these figures show and how it was computed.
-
-Supplementary Figure 6a plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp.
-Based on the "$n = 18,936,995$" quoted in the header of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "istance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs.
-If all pairs were used this $n$ would be much higher since there are more than twenty million CpGs on one strand of the haploid human genome.
-To emphasise, the correlation shown by \cite{Li:2010fb} is not the same as the autocorrelation computed in \cite{Cokus:2008fc} since the autocorrelation uses all pairs of CpGs, regardless of the number of intervening CpGs.
-
-The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \cite{Cokus:2008fc} (highlighted in Supplementary Figure 7).
-This is further evidence that nucleosome spacing effects the strength of co-methylation.
-However, \cite{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \cite{Cokus:2008fc} and \cite{Chodavarapu:2010iqa}.
-
-#### \cite{Lyko:2010dr}
-
-In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation (Supplementary Figure 7).
-They report that the "correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
-This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report?
-Check my data any such increases__).
-__TODO: Is this the fastest decay of correlation in any species?__ However, I am concerned that the maximum autocorrelation is less $0.015$, which doesn't seem to make sense and is very close to zero.
-This may simply be due to a mislabelled or "normalised" value being plotted as the "correlation".
-(__TODO: I have emailed the senior author, Ryszard Maleszka, to try to resolve this concern__).
-
-Additionally, \cite{Lyko:2010dr} identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__).
-They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.
-
-Since the authors specifically describe their analysis as an "autocorrelation", I believe it uses all pairs of CpGs regardless of the number of intervening CpGs.
-
-#### \cite{Lacey:2013iy}
-
-\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2).
-The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
-
-Caution is warranted in extrapolating these results to statements about genome-wide co-methylation.
-Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs (__TODO: Check number__) in the haploid human genome.
-Secondly, these data are from a single sample from a single tissue, which cannot capture the sample-to-sample nor tissue-to-tissue variability in co-methylation.
-Finally, RRBS targets CpG dense regions of the genome with low levels of methylation and so these data do not capture co-methylation outside of these regions, where the majority of the CpGs are found and which are highly methylated.
-
-[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (__TODO: Cite and link to appendix__).
-
-\subsection{Within-fragment co-methylation analyses}
-\label{sec:previous_within-fragment_analyses}
-
-Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules.
-I call this within-fragment co-methylation, where fragment refers to a fragment of DNA.
-More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
-It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
-For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure.
-We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
-
-Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads.
-That is, by studying the methylation patterns at m-tuples.
-Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.
-
-[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).
-
-If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci.
-However, the most commonly used technology for high-throughput bisulfite-sequencing, which uses Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum.
-Assembling these reads into longer contigs will not solve this problem since this cannot guarantee that the assembled reads actually originated from the same DNA fragment.
-
-Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__).
-An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).
-
-[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons.
-These include the fragility of large DNA fragments treated with sodium bisulfite (__TODO: Discuss with someone from AGRF__)
-
-To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides.
-There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.
-Here I review this previously published work.
-
-#### \cite{Cokus:2008fc}
-
-As already discussed in Section \ref{sec:Cokus}, \cite{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_.
-They find this same periodicity "when individual reads [were] examined directly".
-
-The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence.
-For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci.
-Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated.
-This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification].
-They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes.
-\cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
-This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages ().
-There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
-
-[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs.
-The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
-Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.
-
-I consider these results to use a restricted definition of within-fragment "co-methylation".
-These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment.
-Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$.
-From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$.
-It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
-While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to  estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
-
-Furthermore, by aggregating all pairs of CHH loci separated by a common distance and then estimating these probabilities, there is an implicit assumption that all such pairs on the same chromosome have the same conditional probability, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$.
-There is good reason to suspect that this is not true and that there exists significant pair-to-pair variability.
-For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for "co-methylation" of these same loci.
-
-#### \cite{Lacey:2009by}
-
-\cite{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans.
-The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field where the methylation state of one loci, $Z_{h, i}$, depends on that of its neighbours either side, $Z_{h, i - 1}$ and $Z_{h, i + 1}$.
-The physical distances between these neighbours, the $IPD$s, are not considered.
-
-The neighbouring sites model requires an analysis of within-fragment co-methylation in order to estimate the transition probabilities.
-They estimate these parameters from two small regions, each approximately $200$ nucleotides in length and containing 13 and 14 CpGs, from 10 human ovarian carcinoma samples and a pool of somatic controls.
-These regions were sequenced using hairpin-bisulfite PCR, an assay that measures DNA methylation on both strands of a double-stranded DNA molecule[^hairpin_bs].
-For each region, $12-16$ clones were sequenced per sample[^clones]
-
-[^hairpin_bs]: This is unlike traditional bisulfite-sequencing, which only measures methylation along a single DNA strand.
-
-[^clones]: Essentially, $12-16$ reads were generated per sample.
-
-For each of these regions, \cite{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours.
-In other words, they found significant within-fragment co-methylation.
-They also found that the overall level of methylation of the CpGs in a region influenced the strength of the within-fragment co-methylation.
-
-In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a miniscule fraction of the human genome.
-While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.
-
-\subsection{Other related work}
-\label{sec:previous_other_related_work}
-
-Other definitions of co-methylation and related work:
-
-* Lister's mCG correlations
-* co-methylation across samples
-* "vertical" correlations of \cite{Capra:2014uh}
-
-#### \cite{Hebestreit:2013ko}
-
-The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}.
-This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.
-
-#### \cite{Liu:dy}
-
-The focus of this study on identify clusters of correlated CpGs and relating methylation changes at these clusters to SNP genotypes.
-The authors compute the correlation of pairs of $\beta$-values across samples, which is analogous to estimating the linkage disequilibrium between a pair of single nucleotide polymorphisms (SNPs) across a set of samples.
-This is different to computing autocorrelations of $\beta$-values, which are computed within, rather than between, samples.
-An autocorrelation analysis implicitly assumes stationarity along the genome whereas this analysis assumes stationarity across samples (__TODO: Discuss with Terry__).
-
-The discovery cohort contains 247 human samples typed on the Illumina HumanMethylation450 array.
-After normalisation and other pre-processing steps, they identify the top-$25\%$ most variably methylated CpG sites (vCpGs) in the 247 samples.
-They then compute Pearson correlations of methylation levels for all pairs of vCpGs across the samples.
-
-From this they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set".
-However, they also note that "methylation codependency (co-methylation) itself varies across the genome", as is to be expected.
-
-The method I developed, which is described in Section \ref{subsec:wf_cometh}, is superior because of __TODO__.
 

### 2014-10-08 Rmarkdown/comethylation_review.Rmd
Split co-methylation chapter.
Now split into “old stuff” (review of previous work) and “new stuff” (my work)

---
title: "Co-methylation"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

## Chapter overview

Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.

\section{What is co-methylation and why study it?}\label{subsec:co-methylation_definition}

I define co-methylation as the dependence structure of DNA methylation data. This broad definition encompasses several concepts that have variously been described as "co-methylation", or similar, some of which are closely related and others that are disparate. My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" (\cite{Schatz:2004cha}) or "the relationship between the degree of methylation over distance" (\cite{Eckhardt:2006gh}); it also includes the "correlation of two [loci] across many sample" (\cite{Akulenko:2013fl}) and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighboring cell types" (\cite{Capra:2014uh}). A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated (e.g. \cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010drm, Landan:2012kp}). Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context. The lowest level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__. At a higher level is the _correlation of aggregate methylation levels_, such as the autocorrelation of $\beta$-values within a sample.

Between-sample co-methylation is the relationship between methylation levels at two or more loci across a set of samples[^concordance]. For example, \cite{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. Furthermore, measure of within-sample co-methylation can themselves be compared between samples, as is discussed in Chapter \ref{chap:co-methylation}.

[^concordance]: One "between-sample correlation of methylation levels" measure that does not fall under my definition of co-methylation is the following. Consider a pair of samples, $j$ and $j'$, and compute $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$. This measure often reported as evidence for the "concordance" or "replicability" of methylation levels between replicates (e.g., when comparing the genome-wide methylation levels of a cell line passaged 4 versus 5 times \cite{Lister:2009hy} or when comparing different technologies and assays \cite{Zhang:2013uu, Stevens:2013hv, Hansen:2011gu}) but is not the same as co-methylation.

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays (__CITE__).

There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data. For example, if co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis tools.

A common difficulty in translating biological observations into mathematical or statistical language are vague, ambiguous or otherwise insufficiently detailed descriptions of the analyses performed. This is frequently coupled with a lack of software to implement the proposed method. These deficiencies can bring into question the results and certainly hinder the development of new analysis methods and software. Many of the papers reviewed in this chapter suffer from these deficiencies and in these cases I have made my best attempt at deciphering the methods, often via correspondence with the authors.

\section{Correlations of aggregate measurements of methylation}
\label{sec:previous_cor_of_agg_methylation}

To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).

The analysis in \cite{Lister:2009hy} takes a different approach. It too uses within-sample aggregate methylation levels, but rather than estimating correlations of the methylation levels, instead looks for patterns in the spacing of "methylcytosines" along the genome.

Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.

### \cite{Eckhardt:2006gh}

\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance". They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $ >2,000$ bp" and reference Figure 3a in support of this claim. However, what is plotted in Figure 3a is not "correlation" in the usual sense of the word.

In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-priting this figure__). The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp. Therefore, it is not possible that the "percentage identical methylation" was obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp. Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed. An email discussion with the senior author, Stephan Beck, was unable to clarify the definition.

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $z_{i, h}$ is sampled, that is, $Pr(z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $z_{h, i}$ and  $z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels. This most general setting is described by $z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(z_{h, i}, z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{*}$ as follows:

\begin{align*}
PIM^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
        &= Pr(z_{h, i} = 0, z_{h', i'} = 0) + Pr(z_{h, i} = 1, z_{h', i'} = 1) \\
        &= Pr(z_{h', i'} = 0 | z_{h, i} = 0) Pr(z_{h, i} = 0) +
           Pr(z_{h', i'} = 1 | z_{h, i} = 1) Pr(z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp. $z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp. $X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.

__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__

In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:

1. $PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.

As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$. In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp. $< 0.5$) for $c > 0$ (resp. $c < 0$). __TODO: Illustrate these special cases with a plot.__

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.

Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$. I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.

Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis. Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.

### \cite{Cokus:2008fc}\label{sec:Cokus}

\cite{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing. They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_. They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis]. The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance. To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.

[^arabidopsis]: Many plant species, such as _A. thaliana_, have much higher levels of non-CG methylation than do mammals.

\cite{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci. From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin". They also identified correlations between different methylation contexts, e.g. CG vs. CHG, which "suggest[s] complex interactions between the different types of methylation".

The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only). Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier transform analysis of the autocorrelation.

[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}

The $167$ nucleotide periodicity is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al. hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.

The $10$ nucleotide periodicity is equal to the length of one helical DNA turn. From this result Cokus et al. hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites 8-10 apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_) (__TODO: If I want a picture explaining this then Figure 3b from \cite{Jurkowska:2011dt} might form the basis for my own illustration). However, in a later paper by the same group, this $10$ nucleotide periodicity was also observed for CG and CHG methylation, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}.

[^comethylation_def2]: This is referred to as "co-methylation", here meaning the methylation of multiple loci in a single event, by \cite{Jurkowska:2011dt}. This article reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.

The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})

### \cite{Chodavarapu:2010iqa}

\cite{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (HSF1). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of these plots.

In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the HSF1 sample, which confirmed the $10$ nucleotide periodicity. This led \cite{Chodavarapu:2010iqa} to propose an extended version of their earlier hypothesis \cite{Cokus:2008fc}: namely, "that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.

### \cite{Li:2010fb}

\cite{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation". The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $> 1,000$ bp", citing their results as being consistent with those in \cite{Eckhardt:2006gh}. In addition to this genome-wide analysis, \cite{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements. Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.

Unfortunately, just as in \cite{Eckhardt:2006gh}, \cite{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available. All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions. So, once again, I have been forced to make my best guess as to what these figures show and how it was computed.

Supplementary Figure 6a plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp. Based on the "$n = 18,936,995$" quoted in the header of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "istance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs. If all pairs were used this $n$ would be much higher since there are more than twenty million CpGs on one strand of the haploid human genome. To emphasise, the correlation shown by \cite{Li:2010fb} is not the same as the autocorrelation computed in \cite{Cokus:2008fc} since the autocorrelation uses all pairs of CpGs, regardless of the number of intervening CpGs.

The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \cite{Cokus:2008fc} (highlighted in Supplementary Figure 7). This is further evidence that nucleosome spacing effects the strength of co-methylation. However, \cite{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \cite{Cokus:2008fc} and \cite{Chodavarapu:2010iqa}.

### \cite{Lyko:2010dr}

In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation (Supplementary Figure 7). They report that the "correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away". This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report? Check my data any such increases__). __TODO: Is this the fastest decay of correlation in any species?__ However, I am concerned that the maximum autocorrelation is less $0.015$, which doesn't seem to make sense and is very close to zero. This may simply be due to a mislabelled or "normalised" value being plotted as the "correlation". (__TODO: I have emailed the senior author, Ryszard Maleszka, to try to resolve this concern__).

Additionally, \cite{Lyko:2010dr} identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__). They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.

Since the authors specifically describe their analysis as an "autocorrelation", I believe it uses all pairs of CpGs regardless of the number of intervening CpGs.

### \cite{Lacey:2013iy}

\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2). The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".

Caution is warranted in extrapolating these results to statements about genome-wide co-methylation. Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs (__TODO: Check number__) in the haploid human genome. Secondly, these data are from a single sample from a single tissue, which cannot capture the sample-to-sample nor tissue-to-tissue variability in co-methylation. Finally, RRBS targets CpG dense regions of the genome with low levels of methylation and so these data do not capture co-methylation outside of these regions, where the majority of the CpGs are found and which are highly methylated.

[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (__TODO: Cite and link to appendix__).

### \cite{Lister:2009hy}

\cite{Lister:2009hy} report an $8-10$ nucleotide periodicity in the occurence "methylcytosines". Recall that \cite{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a Binomial test (see Section \ref{sec:criticism_of_methylcytosines} for details and a criticism of this concept). Hence, while this may appear to be a within-fragment analysis of co-methylation, it in fact uses aggregate methylation levels. However, unlike the methods described in Section \ref{sec:previous_cor_of_agg_methylation}, the analysis of these aggregate methylation levels is not based on a correlation of these levels, but rather on unnormalised counts, which leads to problems.

\cite{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$. These counts are plotted as a bar chart, with $IPD$ along the x-axis and count on the y-axis, along with a cubic spline fit to the counts. This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (Supplementary Figure 9 \cite{Lister:2009hy}).

[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Supplementary Figure 9, but I am not certain.

An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak several contexts and not at all present for exonic CpGs. The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts since there are very few "methylcytosines" for these contexts in these human samples. Furthermore, in all nine contexts the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.

Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites". However, I suspect that the results in these nine plots are driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines. That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis. To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$. If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s. Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.

In fact, there are well known periodicities in the distribution of nucleotides in genomes. For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \cite{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \cite{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.

### \cite{Landan:2012kp}

__TODO: Review co-methylation work in this paper__

\section{Within-fragment co-methylation}
\label{sec:previous_within-fragment_analyses}

Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules. I call this within-fragment co-methylation, where a "fragment" refers to a fragment of DNA. More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$. For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure. We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.

Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads. That is, by studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).

If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci. However, the most commonly used technology for high-throughput bisulfite-sequencing, which uses Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum. Assembling these reads into longer contigs will not solve this problem since this cannot guarantee that the assembled reads actually originated from the same DNA fragment.

Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__). An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).

[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons. These include the fragility of large DNA fragments treated with sodium bisulfite (__TODO: Discuss with someone from AGRF__)

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides. There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels. Here I review this previously published work.

### \cite{Cokus:2008fc}

As already discussed in Section \ref{sec:Cokus}, \cite{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_. They find this same periodicity "when individual reads [were] examined directly".

The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence. For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci. This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages (). There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.

I consider these results to use a restricted definition of within-fragment "co-methylation". These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment. Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$. From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to  estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

Furthermore, by aggregating all pairs of CHH loci separated by a common distance and then estimating these probabilities, there is an implicit assumption that all such pairs on the same chromosome have the same conditional probability, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$. There is good reason to suspect that this is not true and that there exists significant pair-to-pair variability. For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for "co-methylation" of these same loci.

### \cite{Lacey:2009by}

\cite{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans. The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field where the methylation state of one loci, $Z_{h, i}$, depends on that of its neighbours either side, $Z_{h, i - 1}$ and $Z_{h, i + 1}$. The physical distances between these neighbours, the $IPD$s, are not considered.

The neighbouring sites model requires an analysis of within-fragment co-methylation in order to estimate the transition probabilities. They estimate these parameters from two small regions, each approximately $200$ nucleotides in length and containing 13 and 14 CpGs, from 10 human ovarian carcinoma samples and a pool of somatic controls. These regions were sequenced using hairpin-bisulfite PCR, an assay that measures DNA methylation on both strands of a double-stranded DNA molecule[^hairpin_bs]. For each region, $12-16$ clones were sequenced per sample[^clones]

[^hairpin_bs]: This is unlike traditional bisulfite-sequencing, which only measures methylation along a single DNA strand.

[^clones]: Essentially, $12-16$ reads were generated per sample.

For each of these regions, \cite{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours. In other words, they found significant within-fragment co-methylation. They also found that the overall level of methylation of the CpGs in a region influenced the strength of the within-fragment co-methylation.

In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a miniscule fraction of the human genome. While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.

\section{Between-sample co-methylation}
\label{sec:between_sample_cometh}

As is clear from the above, there are multiple definitions of within-sample co-methylation. An fundamentally different concept to within-sample co-methylation is between-sample co-methylation, which is the correlation of methylation levels between a single pair of loci across a set of samples rather than the correlation of pairs of loci within a single sample. (__CITE__) estimate various forms of between-sample co-methylation and are reviewed below.

* "vertical" correlations of \cite{Capra:2014uh}
* MethLD of \cite{Liu:dy}

__TODO: Comment on utility of within-sample vs. between-sample co-methylation__

### \cite{Liu:dy}

The focus of this study on identify clusters of correlated CpGs and relating methylation changes at these clusters to SNP genotypes. The authors compute the correlation of pairs of $\beta$-values across samples, which is analogous to estimating the linkage disequilibrium between a pair of single nucleotide polymorphisms (SNPs) across a set of samples. This is different to computing autocorrelations of $\beta$-values, which are computed within, rather than between, samples. An autocorrelation analysis implicitly assumes stationarity along the genome whereas this analysis assumes stationarity across samples (__TODO: Discuss with Terry__).

The discovery cohort contains 247 human samples typed on the Illumina HumanMethylation450 array. After normalisation and other pre-processing steps, they identify the top-$25\%$ most variably methylated CpG sites (vCpGs) in the 247 samples. They then compute Pearson correlations of methylation levels for all pairs of vCpGs across the samples.

From this they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set". However, they also note that "methylation codependency (co-methylation) itself varies across the genome", as is to be expected.

\section{Exploiting co-methylation in analyses of differential methylation}
\label{sec:exploiting_cometh}

\cite{Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Sofer:2013bk} have each sought to exploit co-methylation, both the within-sample and between-sample forms, in developing more powerful and efficient ways to identify differential methylation.

### \cite{Hebestreit:2013ko}

The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}. This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.

### 2014-10-13 Rmarkdown/comethylation_review.Rmd
Drafting thesis bootcamp plan

@@
 ---
-title: "Co-methylation"
+title: "Co-methylation review"
 author: "Peter Hickey"

### 2014-10-17 Rmarkdown/methsim.Rmd
Drafted introduction of methsim chapter

---
title: "A simulation model of DNA methylation data"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

# Chapter overview

# Introduction

Simulation studies play a vital role in the development, validation and benchmarking of methods in applied statistics. Simulation is necessary when there is a lack of real datasets with well-known outcomes that we would otherwise use to assess a method's performance. While simulated data can never fully capture the richness of "real" data, we can learn a lot about a method by studying how it performs when applied to simulated data.

While we may sometimes be able to design an experiment with known outcomes via the use of controls, both positive and negative, this is not always possible in genomics. One reason for this is the cost of performing a well-designed control experiment. Why spend your time and money on an experiment where you "know the answer" when you could be spending that same time and money on investigating something new? Another reason is that in genomics we are often working with cutting-edge technology that is so new that we lack the technological and biological knowledge to even design such a control experiment.

The key advantage of using simulated data over real data is that we know the truth _a priori_. Moreover, we can manipulate the truth via parameters in the simulation model and can examine how a method performs under a variety of scenarios. And this manipulation is cheap, a mere matter of changing parameters and re-running a piece of software, which means that we can investigate a broad range of plausible scenarios.

Simulation studies can provide many insights into the performance of a method. At the most basic level, if a method fails to work or performs poorly when applied to simulated data, then it is very unlikely to work well when applied to real data. We can also identify which scenarios are "easy", ones where most methods are able to identify the truth, and scenarios under which certain methods perform better than others.

Simulation studies can also tell us about the plausibility of hypothesised stochastic or mechanistic models of a phenomenon. By simulating data from the proposed model and comparing it to the real data we can identify shortcomings in the model that we can later refine.

In the field of genomics, almost all papers proposing a new method will include a simulation study and perhaps a comparison to existing methods, if these exist. There are a few key criteria when designing a simulation method:

1. Realism: The simulated data must be "similar" to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes "similar enough".
2. Cost: It should be fast and cheap to simulate data. The most common use of simulation models in applied statistics is to repeatedly generate datasets under a range of parameter settings. This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to analyse the full space of scenarios. An exception to this rule may occur when the simulation model is used to test a proposed mechanistic or stochastic model of a phenomenon, such as in studies of molecular dynamics. Even then, however, the cost of a simulation should be less than the cost of performing the equivalent experiments, otherwise the simulation is generally not worth the effort.
3. Usability: There __must__ be a software implementation. Simulation models exist to be simulated from; a simulation model without an implementation is next to useless. The implementation should give the user easy access to the key  parameters and have sensible default settings. The output of the software should be in a standard format or readily convertible to a simple, manipulable format.

The above discussion refers to the general use of simulation models, particularly in applied statistics. I now turn my attention to the need for simulation models in the study of DNA methylation. Whole-genome bisulfite-sequencing remains an expensive assay, costing a few thousand dollars per sample (human) at the time of writing. Understandably, studies using WGBS have been from well-funded research groups focusing on "sexy" biology rather than on efforts to generate extensively-validated control datasets.

The statistical methodology to analyse such experiments lags somewhat behind the ability to generate the data, as is common in genomics. Nonetheless, more statistical methods are being published seeking to address common research questions such as identifying differential methylation. However, because of the lack of "gold-standard" datasets, it is not clear which of these perform well for common experimental designs and research questions. Many of these published methods feature a simulation study in the accompanying paper, but, as we will shortly see, these simulations are woefully inadequate, unrealistic and vary between papers. In order to perform a proper benchmarking of these methods we require a realistic simulation model, something that I have attempted to address with my software, `methsim`, which I describe in this chapter.

# Background + literature review

# Methodology

# Implementation

# Case study

# Discussion of findings

# Conclusions

### 2014-10-17 Rmarkdown/methsim.Rmd
Reviewed methods for simulating BS-seq reads

@@
 
+Simulated bisulfite-sequencing data may be used in the comparison of strategies for the alignment and processing of bisulfite-converted reads, or in the comparison of downstream analysis methods.
+Simulation methods for bisulfite-sequencing data may choose to simulate the individual sequencing reads, unaligned or aligned, or some summary measure of methylation, such as $\beta$-values.
+The former is required for the comparison of different alignment strategies whereas the latter is typically used by developers of downstream analysis methods, such as differential methylation callers, since this is often all that is required to run these methods.
+Moreover, simulated reads must be aligned and processed before they can be used in comparisons of downstream analysis methods, which is often not of direct relevance in comparisons of these tools.
+
+## Methods for simulating bisulfite-sequencing reads
+
+All of the available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies. These are not suitable for comparisons of downstream analysis methods.
+
+`Sherman` ([http://www.bioinformatics.babraham.ac.uk/projects/sherman/](http://www.bioinformatics.babraham.ac.uk/projects/sherman/)) is software to simulate bisulfite-sequencing reads, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts as `FASTQ` files.
+The simulated reads are designed for comparing the performance of different alignment strategies.
+It has many parameters, of which the ones relevant to our discussion of simulating realistic DNA methylation data are `-CG` and `-CH`, the bisulfite conversion rates for CG and CH methylation loci, respectively.
+These are set by the user with values between $0$ and $100$.
+Reads are simulated by sampling from the user-specified reference genome.
+When a read contains a CG (resp.
+CH) locus, it is randomly assigned as being converted (unmethylated) with probability `-CG` / 100 (resp.
+`-CH` / 100).
+
+While suitable for comparing alignment strategies, `Sherman` produces data that is not suitable for use in comparing downstream analysis methods.
+All CG (resp.
+CH) loci have an average $\beta$-value of `-CG` (resp.
+`-CG`) regardless of the genomic context, which we know to be inconsistent with real data.
+Furthermore, the methylation state of each methylation locus is independent, which is clearly inconsistent with the strong co-methylation observed in real data.
+
+`FastqToBS` ([http://users.dimi.uniud.it/~nicola.prezza/projects.html](http://users.dimi.uniud.it/~nicola.prezza/projects.html)) uses the same strategy as `Sherman`.
+
+`DNemulator` ([http://www.cbrc.jp/dnemulator/README.html](http://www.cbrc.jp/dnemulator/README.html)) uses a slightly more sophisticated simulation strategy to simulate `FASTQ` files for use in comparing alignment strategies for bisulfite-sequencing data.
+`DNemulator` does this with three separate routines, `fasta-methly-sim`, `fasta-polymorph` and `fasta-bisulf-sim`:
+
+1.
+`fasta-methyl-sim` converts cytosines in the reference genome (`FASTA` file) to a character indicating the methylation level of that locus: `C` represents $0\%$ methylated, `c` represents $10\%$ methylated, `d` represents $20\%$ methylated, `v` represents $50\%$ methylated and `t` represents $100\%$ methylated.
+Each of these conversions has a different probability in the CG and CH contexts.
+2. `fasta-polymorph` simulates a polymorphic, diploid genome based on the modified reference sequence created by `fasta-methyl-sim`.
+3. `fasta-bisulf-sim` simulates reads by sampling from the simulated genome created by `fasta-polymorph`. Read are simulated with bisulfite-conversion error and sequencing error.
+
+The reads simulated by `DNemulator` will result in $\beta$-values that have more context-dependence than those resulting from reads generated by `Sherman`, `FastqToBS` or `BSsim`.
+However, methylation events are still generated independently of one another, which means there is no co-methylation in the simulated data.
+Therefore, reads simulated by `DNemulator`, while suitable for comparing alignment strategies, are not suitable for comparing downstream analysis methods.
+
+`BSsim` ([http://122.228.158.106/BSSim/](http://122.228.158.106/BSSim/) and used in \cite{Xie:2014ez}), uses a similar strategy to `DNemulator`.
+
+## Methods for $\beta$-values
+
+Methods to describe:
+
+- [ ] \cite{Feng:2014iq}
+- [ ] \cite{Lacey:2013iy}
+- [ ] \cite{Sofer:2013bk}
+- [ ] \cite{Xu:2013eg}
+- [ ] \cite{Chen:2014jb}
+- [ ] \cite{Park:2014ho}
+- [ ] \cite{Dolzhenko:2014bo}
+- [ ] \cite{Xie:2014ez}
+- [ ] Others?
+
 # Methodology
+
+`methsim` is designed for simulating data for downstream analyses, and uses a hybrid strategy that is capable of creating raw, unaligned reads or summary measures, such as $\beta$-values, from the same simulation run.
 

### 2014-10-18 Rmarkdown/methsim.Rmd
Reviewed methods for simulating aggregate methylation levels

@@
 
+A simulation may be model-based or based on sampling of real data, roughly corresponding to parametric and non-parametric simulation models, respectively.
+It is often simpler to simulate from a model-based simulation, particularly if the model is a well-studied parametric distribution.
+However, this simplicity often comes at the cost of increased assumptions, whose validity may be questionable.
+Of course, the parameters in any model-based simulation should be estimated from real data, although this may not use formal estimation procedures such as maximum likelihood.
+
+In contrast, simulations based on sampling of real data may reduce the number of assumptions required.
+However, care must be taken in selecting the units to be sampled so that the sampling process is efficient and so that the sampled data don't grossly distort within- and between-sample dependencies.
+
 ## Methods for simulating bisulfite-sequencing reads
 
-All of the available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies. These are not suitable for comparisons of downstream analysis methods.
+All of the available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies and are model-based.
+These are not suitable for comparisons of downstream analysis methods.
 
@@
 
-## Methods for $\beta$-values
+## Methods for simulating aggregate methylation levels
 
-Methods to describe:
+Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section.
+Generally, the simulation method is not a major feature of the paper but is there to support claims about the performance of the proposed method.
+Consequently, the simulation model is often only briefly described and a software implementation is rarely made available.
 
-- [ ] \cite{Feng:2014iq}
-- [ ] \cite{Lacey:2013iy}
-- [ ] \cite{Sofer:2013bk}
-- [ ] \cite{Xu:2013eg}
-- [ ] \cite{Chen:2014jb}
-- [ ] \cite{Park:2014ho}
-- [ ] \cite{Dolzhenko:2014bo}
-- [ ] \cite{Xie:2014ez}
-- [ ] Others?
+The most widely studied downstream analysis problem is that of identifying differential methylation.
+Differential methylation is identified by comparing summary measures of methylation, such as $\beta$-values, between two or more groups.
+It is therefore generally not necessary to simulate individual reads but is sufficient to simulate these summary measures directly.
+As the `methsim` software is focused on simulating data for comparing methods for calling differential methylation, I review methods designed for a similar purpose.
+
+A common framework for simulating data for a $k$-group experiment is the following hierarchical model:
+
+1. Simulate the unobserved group-specific true methylation levels, $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$.
+  - For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}}$.
+  - Differentially methylated loci are simulated by setting loci with $B_{i, j_{k}} \neq B_{i, j_{k'}}$
+2. Simulate the observed sample-specific sequencing depths, $u_{i, j} + m_{i, j}$.
+3. Simulate the observed sample-specific methylation levels, $\beta_{i, j_{k}}$ from some model incorporating both $B_{i, j_{k}}$ and $u_{i, j} + m_{i, j}$.
+
+A popular choice of parametric model is the Beta-Binomial distribution \cite[e.g.,]{Feng:2014iq, Lacey:2013iy, Xu:2013eg, Chen:2014jb, Dolzhenko:2014bo}.
+Under this model the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j}, \phi_{i, j})$ distribution, and the $\beta$-values are simulated from a Binomial distribution, $\beta_{i, j} = Binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
+
+### Simulating sequence depth
+
+The sequencing depths, $m_{i, j} + u_{i, j}$, are typically sampled from real data \cite[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, although a rounded Normal distribution \cite{Xu:2013eg} and a rounded mixture of Gamma distributions have also been used \cite{Lacey:2013iy}.
+
+The most sophisticated approach to simulation of sequencing depth in bisulfite-sequencing experiments is the RRBS simulation of \cite{Lacey:2013iy}.
+In addition to using a mixture of distributions to capture both the low-coverage and high-coverage modes observed in RRBS sequencing coverage, the authors model the correlation of sequencing depth across samples for a given region.
+They do this by using a Normal copula to make the set of sequencing depths a jointly dependent set of random variables.
+While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the benefits are unclear.
+Moreover, it is simpler to include such correlations by a sensible sampling of real data.
+Specifically, this can be achieved by sampling sequencing depth across samples for a given region rather than sampling across the full set of sequencing depths.
+
+### Simulating $B_{i, j}$
+
+Under the Beta-Binomial model, $\mu_{i, j}$ is the mean and $\phi_{i, j}$ the dispersion of the Beta distribution[^beta_parameterisation].
+The means, $\mu_{i, j}$, are group-specific while the dispersions, $\phi_{i, j}$, are sample specific.
+If $\mu_{i, j}$ are allowed to be sample-specific then there would be no group-structure in the subsequent $\beta$-values.
+Conversely, if the dispersions are group-specific then there would be little within-group variability in the subsequent $\beta$-values.
+
+[^beta_parameterisation]: Note that this is different to the standard parameterisation of the Beta distribution, which is described by two shape parameters, $\alpha$ and $\beta$.
+The relationship between the two parameterisations is $\mu = \frac{\alpha}{\alpha + \beta}$ and $\phi = \frac{1}{\alpha + \beta + 1}$ \cite{Feng:2014iq}.
+
+Other distributions may be used instead of the Beta distribution.
+For example, \cite{Xie:2014ez} also use a Normal distribution and a mixture of Normal distributions and \cite{Xu:2013eg} also use the truncated Normal and a mixture of truncated Normal distributions.
+An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, i.e., sample the observed $\beta_{i, j}$ from a real experiment.
+
+While the simulation of differentially methylated loci is straightforward - simply vary $B_{i, j_{i}}$ across the $k$ groups - the simulation of differentially methylated regions is more difficult.
+In principal we would simply simulate runs of differentially methylated loci, however, it must be noted that this requires careful choice of parameters.
+Such parameters include the length of DMRs, the minimal number of loci they must contain, the maximal intra-pair distances of loci within a DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should __FIGURE__ be considered one DMR or two DMRs?
+(__TODO: Figure with 4 DMCs, then 1 non-DMC and then 5 DMCs).
+These decisions ultimately have to be made by the user of the simulation software based on the types of events they are interested in.
+For example, DMRs in cancer can be very long whereas others may be interested in identifying very short DMRs overlapping transcription factor binding sites.
+
+__TODO: Comment on independence of $B_{i, j}$ in almost all methods.__
+
+### Simulating positions of methylation loci
+
+All of the described methods, except \cite{Lacey:2013iy}, simulate methylation events based on a reference genome.
+\cite{Lacey:2013iy} propose a hidden Markov model for the intra-pair distances of CpGs in human RRBS data and simulate methylation loci based on this model.
+I do not think a model-based strategy is useful here, in fact I think it is counterproductive, and I prefer a sampling-based strategy, i.e., the use of methylation loci from a reference genome.
+While the set of methylation loci do vary across samples within a species due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed, particularly when the main aim of the simulation model is to realistically simulate methylation levels and not the location of the methylation loci.
+Furthermore, if sequence variation is required then this can be achieved by sampling from this set of loci or by sampling from the set of methylation loci from a group of samples.
+
+### Other model-based simulations
+
+A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \cite[e.g.,]{Chen:2014jb, Chen:2013eh}.
+These models are designed for simulating microarray data and not sequencing data, since they do not include the variability caused by variation in sequencing depth.
+
+## Methods based entirely on sampling real data
+
+* Xie (2013)
 

### 2014-10-18 Rmarkdown/methsim.Rmd
Reviewed use of co-methylation in parametric simulation model

@@
 Consequently, the simulation model is often only briefly described and a software implementation is rarely made available.
+In fact, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} has a software implementation available.
 
@@
 Under the Beta-Binomial model, $\mu_{i, j}$ is the mean and $\phi_{i, j}$ the dispersion of the Beta distribution[^beta_parameterisation].
-The means, $\mu_{i, j}$, are group-specific while the dispersions, $\phi_{i, j}$, are sample specific.
-If $\mu_{i, j}$ are allowed to be sample-specific then there would be no group-structure in the subsequent $\beta$-values.
-Conversely, if the dispersions are group-specific then there would be little within-group variability in the subsequent $\beta$-values.
+Both the means, $\mu_{i, j}$, and the dispersions, $\phi_{i, j}$, are group-specific and are allowed to vary across methylation loci ($i$).
+The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e., the _biological variability_ of DNA methylation.
 
@@
 
+As noted by \cite{Feng:2014iq}, the Beta distribution is a very flexible distribution with support on $[0, 1]$ and has "long been a natural choice to model binomial proportions", particularly as a conjugate prior, as it is used in the empirical Bayes model of \cite{Feng:2014iq}.
+The Beta-Binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or overdispersed Binomail distributions
+
 Other distributions may be used instead of the Beta distribution.
-For example, \cite{Xie:2014ez} also use a Normal distribution and a mixture of Normal distributions and \cite{Xu:2013eg} also use the truncated Normal and a mixture of truncated Normal distributions.
-An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, i.e., sample the observed $\beta_{i, j}$ from a real experiment.
+For example, \cite{Xie:2014ez} consider a Normal distribution and a mixture of Normal distributions and \cite{Xu:2013eg} consider the truncated Normal and a mixture of truncated Normal distributions.
 
-While the simulation of differentially methylated loci is straightforward - simply vary $B_{i, j_{i}}$ across the $k$ groups - the simulation of differentially methylated regions is more difficult.
-In principal we would simply simulate runs of differentially methylated loci, however, it must be noted that this requires careful choice of parameters.
+An alternative to specifying a parametric distribution for the $B_{i, j}$ would be to sample these from real data, i.e., sample the observed $\beta_{i, j}$.
+However, without further perturbation, such an approach would have insufficient within-group variability since all samples in that group have the same "true" methylation level.
+
+While the simulation of differentially methylated loci is straightforward - simply vary the mean of the distribution of $B_{i, j_{i}}$ across the $k$ groups - the simulation of differentially methylated regions is more complex.
+In principal we would simply simulate a differentially methylated region by simulating runs of differentially methylated loci.
+However, it must be noted that this requires careful choice of parameters.
 Such parameters include the length of DMRs, the minimal number of loci they must contain, the maximal intra-pair distances of loci within a DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should __FIGURE__ be considered one DMR or two DMRs?
@@
 
-__TODO: Comment on independence of $B_{i, j}$ in almost all methods.__
+### Simulating co-methylation
+
+A key feature ignored by all the above methods, with the honourable exception of \cite{Lacey:2013iy}, is co-methylation.
+The methylation states of neighbouring loci are highly dependent and, consequently, the $B_{i, j}$ of nearby loci are highly dependent.
+To be clear, almost all of the published simulation studies of bisulfite-sequencing data have failed to model a very important feature of DNA methylation data, co-methylation.
+
+As demonstrated in Chapter \ref{chap:co-methylation}, the physical process of co-methylation is at the level of individual DNA fragments, not at the level of aggregate methylation levels.
+The observed correlations of $\beta$-values are driven by this within-fragment co-methylation.
+However, a simulation model of DNA methylation data need not induce correlations at the level of individual DNA fragments.
+Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation but it may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
+This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependence.
+This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.
+
+\cite{Lacey:2013iy} take a different approach, but one that still results in correlated $B_{i, j}$.
+To begin, they compute $\beta$-values from chromosome 11 for a single normal myotube cell line that was sequenced with RRBS.
+They then fit a Gaussian variogram to these $\beta$-values, which shows "a strong correlation for sites in close proximity, decaying to near independence at distances beyong 3000 bp".
+To simulate spatially correlated $B_{i, j}$ they use an iterative process:
+
+1. Simulate $B_{i, j}$ from a Beta distribution with parameters estimated from the chromosome 11 MTCTL2 data. These are estimated under an assumption of independence.
+2. Induce correlation amongst the $B_{i, j}$ (across $i$) by a transformation of the $B_{i, j}$.
+
+The second step uses a method published by \cite{Zaykin:2002ko}.
+The transformed values, $B^{*}$, are created by the transformation $B^{*}_{i, j} = 1 - \Phi{C \Phi - 1(1 - B)}$, where $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, which is estimated from the fitted Gaussian variogram, and where $\Phi(\dot)$ denotes the standard normal distribution function.
+
+\cite{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).
+__TODO: Try these scripts, see how long they take__.
+The running times reported in the orignal publication are shown in __TABLE__.
+__TODO: To simulate a single RRBS dataset would take HOW LONG__.
+The simulation scripts could be made more user-friendly by creating an R package, rather than supplying the software as a collection of scripts, and would benefit from some documentation.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+Number of samples & Number of sites & Running time (seconds) \\ \midrule
+8                 & 1000            & \textless 1            \\
+8                 & 5000            & 13                     \\
+8                 & 10000           &                        \\ \bottomrule
+\end{tabular}
+\caption{Reported running times for simulation methodology of \cite{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
+\label{tab:Lacey_runtime}
+\end{table}
 
@@
 
+A simulation can be based entirely on sampling real data[^synthetic].
+This type of simulated data is attractive because through careful sampling it can capture heterogeneity and correlations that may be too difficult to parameterise in a model.
+At the same time, if the sampling units are poorly chosen or the sampling strategy is incorrect then it may ignore these same features, or worse, introduce artefacts into the simulated data.
+
+[^synthetic]: This may also be referred to as creating a synthetic dataset.
+
+__UP TO HERE__
+
 * Xie (2013)

### 2014-10-18 Rmarkdown/methsim.Rmd
Reviewed sampling-based simulation strategies.

@@
 __TODO: Try these scripts, see how long they take__.
-The running times reported in the orignal publication are shown in __TABLE__.
+The running times reported in the orignal publication are shown in __TABLE__ for a single simulation.
 __TODO: To simulate a single RRBS dataset would take HOW LONG__.
@@
 These models are designed for simulating microarray data and not sequencing data, since they do not include the variability caused by variation in sequencing depth.
+Neither of these include co-methylation in their simulation model.
 
@@
 This type of simulated data is attractive because through careful sampling it can capture heterogeneity and correlations that may be too difficult to parameterise in a model.
-At the same time, if the sampling units are poorly chosen or the sampling strategy is incorrect then it may ignore these same features, or worse, introduce artefacts into the simulated data.
+At the same time, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features, or worse, introduce artefacts into the simulated data.
 
@@
 
-__UP TO HERE__
+Sampling methods are most easily implemented at the level of $\beta$-values.
+Sampling reads is reads is far more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis.
+Down-sampling is only of interest to examine the effects of sequencing coverage on downstream analyses.
 
-* Xie (2013)
+\cite{Sofer:2013bk} use a sampling-based simulation method in the development of their `Aclust` software for identifying differential methylation from microarray data.
+The idea is adapted from a simulation method for array comparative genomic hybridization experiments \cite{Gaile:2007bq}.
+\cite{Sofer:2013bk} sample "blocks" of CpGs, a region of the genome where all CpGs are within 10 kb of the next, to "generate (spatial) correlation-preserved methylation data".
+By sampling blocks rather than individual CpGs, this sampling scheme preserves the correlation structure between CpGs occuring in the same block.
+Two blocks are unlikely to be correlated since there is little evidence that CpGs separated by more than 10 kb have spatially correlated methylation levels.
+
+\cite{Sofer:2013bk} sample from a dataset of 539 Illumina 450k microarrays and select a small number of "target" CpGs, which are CpGs whose methylation level is highly variable across the 539 samples.
+If a block does not contain a "target" then it is sampled uniformly at random from the 539 samples.
+However, if a block does contain a "target" then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the "target" CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the "target" CpG[^cases_and_controls].
+This is essentially weighted sampling of the real data to induce differential methylation.
+
+[^cases_and_controls] The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methlyated for "controls" at "target" CpGs.
+
+Due to the correlation structure of the $\beta$-values, the co-methylation, it is likely, although not guaranteed, that other CpGs in the blocks containing targets also display differential methylation.
 

### 2014-10-18 Rmarkdown/methsim.Rmd
Started explanation of methsim.

@@
 
-[^cases_and_controls] The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methlyated for "controls" at "target" CpGs.
+[^cases_and_controls] The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at "target" CpGs.
 
@@
 
-`methsim` is designed for simulating data for downstream analyses, and uses a hybrid strategy that is capable of creating raw, unaligned reads or summary measures, such as $\beta$-values, from the same simulation run.
+`methsim` is is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements.
+My intention in developing `methsim` was to design simulation software that could generate realistic data for use in developing and comparing differential methylation callers.
+In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments.
+An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or sequencing reads.
+
+Simulating data with `methsim` involves 3 steps:
+
+1. Simulate the true methylome of each sample.
+2. Sample reads, including sequencing error and bisulfite-conversion error.
+3. Construct the output, be it reads, methylation patterns at m-tuples or $\beta$-values.
+
+`methsim` requires an input dataset from which to estimate key parameters.
+For the input dataset, `methsim` requires the methylation patterns at various sized m-tuples, which can be produced by the `methtuple` software.
+
+## Simulating the truth
+
+As we have seen, DNA methylation is highly heterogeneous along the genome.
+Nonetheless, there are clear regions of "similarity", regions where we might hope to model by a simple parametric model.
+`methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together.
+The idea of segmenting a highly heterogenous stochastic process into a series of locally homogeneous prcoesses is not new.
+A hidden Markov model is an example of such a process.
+While the entire process may highly heterogeneous, conditional on the hidden states the process is homogeneous.
+Hidden Markov models, and other models assuming local similarity in spite of global heterogeneity, have been used with great success in bioinformatics (__CITE__).
+
+`methsim` takes the following approach to simulating the true underlying methylome for each sample:
+
+1. Segment the methylome into regions of similarity.
+2. Within each region simulate the methylation states from a simple parametric model.
+3. "Stitch" the resulting regions together.
+
+Actually, the stitching together of segments is not strictly a separate step and can be incorporated into step 2.
+
+### Segmenting the methylome
+
+\cite{Burger:2013kq} developed the R/Bioconductor package, `MethylSeekR`, to discover regulatory motifs from bisulfite-sequencing data.
+`MethylSeekR` segments the genome into unmethylated regions (_UMR_s), lowly-methylated regions (_LMR_s) and partially methylated regions[^pmds] (_PMR_s).
+It does this using a two-stage algorithm applied to the $\beta$-values from a sample:
+
+[^pmds]: Partially methylated regions are more commonly known as partially methylated domains, but I refer to them as regions for consistency with UMRs and LMRs.
+
+1.
+Identify partially methylated regions.
+A summary statistic, $\alpha$, which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs.
+Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify _PMR_s and non-_PMR_s
+2.
+Identify UMRs and LMRs.
+The PMRs are masked from the genome and simple heuristics are used to identify UMRs and LMRs based on the average $\beta$-values in a window and the number of CpGs in the window.
+
+`methsim` uses the output of `MethylSeekR` to define "regions of similarity", namely UMRs, LMRs, PMRs and the 'rest of the genome'.
+The 'rest of the genome' roughly corresponds to 'mostly methylated regions' (_MMR_s), although \cite{Burger:2013kq} do not describe it as such.
+(__TODO: Figure showing 'rest of genome' is MMRs__)
+
+While a tailored algorithm may improve the segmentation process, the segmentation produced by `MethylSeekR` is a reasonable approximation to finding "regions of similarity".
+This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)).
+It should be noted that `methsim` does perform some post-processing of the `MethylSeekR` output so that the segmentation is disjoint and exhaustive.
+
+The segmentation is a property of the sample, not of any particular haploid methylome in the sample.
+
+### Simulating a single haploid methylome
+
+As discussed in Chapter \ref{chap:statmodel}, whole-genome bisulfite sequencing  measures DNA methylation from individual DNA fragments from a pool of 100s or 1000s of cells.
+The methylation states of each copy of the haploid genome can be represented by a pair of vectors, $Z_{h, j}^{+} = (Z_{h, 1, j}^{+}, \ldots, Z_{h, n_{loci}^{+}, j})$, for the positive strand, and $Z_{h, j}^{-} = (Z_{h, 1, j}^{-}, \ldots, Z_{h, n_{loci}^{-}, j})$, for the negative strand.
+For simplicity, I will focus on CpG methylation and assume that the methylation state of all CpGs are strand-symmetric so that each haploid genome can be represented by a single vector, $Z_{h, j} = (Z_{h, 1, j}, \ldots, Z_{h, n_{loci}, j})$.
+
+I propose to model $Z_{h, j}$ as a two-state, one-step Markov chain.
+This allows me to incorporate a model of within-fragment co-methylation into my simulation.
+The choice of a one-step chain is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.
+
+In Chapter \ref{chap:co-methylation} I showed that the strength of within-fragment co-methylation varies as a function of the intra-pair distance and by the genomic context.
+I also showed that for a pair of CpGs separated by given IPD, and in the same genomic context, that there is a large amount of variability in the strength of within-fragment co-methylation.
+
+For these reasons I allow the transition probabilities to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous.
+In particular, I allow the transition probabilities, $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} = b | Z_{h, i, j} = a)$, to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region for this pair of loci[^context], $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} | Z_{h, i, j}, r_{i}, IPD(i, i + 1))$, where $r_{i}$ is the region type (UMR, LMR, PMR or MMR) of the $i^{th}$ methylation locus.
+
+[^context]: Actually, it only depends on the context of the $i^{th}$ locus.
+Most pairs of loci, $(i, i + 1)$ will lie in the same genomic feature.
+For pairs that span the boundary of two different features I have arbitrarily chosen to use the context of the $i^{th}$ locus.
+
+The above-described model, being spatially inhomogeneous, is not particularly amenable to analytical calculations.
+However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome.
+For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h, j)}$, to estimate or otherwise assign.
+
+`methsim` assigns these transition probabilities based on empirical the distributions of  $\beta_{i, j}$ and $LOR_{IPD(i, i + 1), j}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$.
+Call these empirical distributions $\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, respectively.
+These empirical distributions are shown for a single sample in __FIGURE__.
+(__TODO: Do these plots for a single sample__).
+
+__TODO: Describe how a transition probability can be computed from a marginal probability and a log odds ratio__
+
+To simulate $Z_{h, j}$ the algorithm proceeds as follows (in `R` pseudocode):
+
+```{r, echo = TRUE, eval = FALSE}
+n <- length(z)
+u <- runif(n, 0, 1)
+# Sample from region-specific beta-value distributions
+b <- sample(BETA, n)
+# Sample from region-specific LOR distributions
+l <- sample(LOR, n - 1)
+
+# Sample the first observation from the "stationary/marginal" distribution
+if (u[1] < b[1]) {
+  z[1] <- 0
+} else {
+  z[1] <- 1
+}
+
+# Sample the rest of the z
+for (i in 2:n) {
+  if (z[i] == 0) {
+    # Compute transition probability
+    # Compute z[i]
+
+  } else {
+    # Compute transition probability
+    # Compute z[i]
+  }
+}
+```
+
+### Simulating a single sample
+
+Given a set of transition matrices, $\Big { P_{i^{(h, j)} \Big }$, we can readily simulate a realisation of the methylation states along a haploid genome, $Z_{h, j}$.
+Naively, we might simulate a single sample by simply generating a realisation for each $h$ of $Z_{h, j}$.
+However, although there is cell-to-cell heterogeneity in the methylation patterns, there is still a degree of dependence for the $i^{th}$ methylation state across the $h$ haploid genomes in the $j^{th}$ sample.
+After all, the cells are from the same individual and are broadly of the same "type".
+
+What this means is that we want some degree of dependence between the set of transition matrices for haploid methylome $h$, $\Big { P_{i^{(h, j)} \Big }$, and for those for haploid methylome $h'$, $\Big { P_{i^{(h', j)} \Big }$.
+This is asking quite a lot of our simple simulation model.
+Instead, `methsim` uses a heuristic procedure to achieve a similar result.
+
+To begin, we assume that all $H$ haploid genomes in the sample have the same co-methylation structure.
+That is equivalent to defining $\Big { P_{i^{(h, j)} \Big } := $\Big { P_{i^{(j)} \Big }$ for all $h$.
+__TODO: Comment on this assumption__
+
+Naively, we might simulate a sample by simulating $H$ realisations of the Markov chain with transition matrices $\Big { P_{i^{(j)} \Big }$.
+However, this gives no control over the dependence between the methylation states at the $i^{th}$ locus over the $H$ haploid genomes in the $j^{th}$ sample.
+In fact, such a model will have the "correct" co-methylation structure but "incorrect" $\beta$-values.
+__TODO: Check this and demonstrate using `methsim`__
+
+Instead, `methsim` simulates $R << H$ realisations of the Markov chain with transition probabilities given by $\Big { P_{i^{(j)} \Big }$.
+It then weights these realisations according to the observed heterogeneity for each region.
+__TODO: Describe the inuition of this weighting scheme as being that $w_{r}$ of the cells have identical methylation states for that region.
+
+### Simulating multiple samples from the same group
+
+### Simulating differential methylation
+
+## Parameter estimation
 

### 2014-10-19 Rmarkdown/avy.Rmd
Background to Avy

---
title: "Epialleles in Agouti viable yellow mice"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

# Chapter overview

The agouti viable yellow ($A^{vy}$) mouse model is perhaps the archetype of _heritable_ epigenetic modifications resulting in phenotypic variation. The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins this model, recently performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ locus, with the aim of identifying epialleles. An epiallele is a region of the genome where samples are genetically identical yet epigenetically variable.

Due to a somewhat unusual experimental design, Professor Whitelaw requested the assistance of Professor Terry Speed and I in analysing the data. As with many experiments, there was pressure to quickly develop methods and implement these as software to analyse the data. This chapter describes the experiment and how we addressed these challenges.

# Introduction + literature review

"The Agouti gene is responsible for determining whether a mammal's coat is banded (_agouti_) or of a solid color (_non-agouti_)" [TODO: re-word. Wikipedia!]. In mice, there are several well-studied mutants where differences in the expression of the agouti protein influences multiple, seemingly unrelated phenotypic traits (_pleiotropy_). Specifically, yellow mice, those with high expression of the agouti protein, are obese, diabetic and have an increased susceptibility to tumours.

One of these agouti-associated obesity mutations is the agouti viable yellow allele ($A^{vy}$), discovered by \cite{Duhl:1994bx}. The cause of the $A^{vy}$ mutation in the agouti protein was identified by a team led by Professor Emma Whitelaw \cite{Morgan:1999ds} as being due to methylation differences upstream of the agouti gene. Moreover, these epigenetic-driven phenotypic differences display variable expressivity (differences between individuals), variegation (mosaicism between cells of the same individual), are mitically heritable and display parent-of-origin effects (distinct from classic parental imprinting).

For the $A^{vy}$ allele, there is an intracisternal A particle (IAP) insertion  approximately 100 kb upstream of the agouti gene. The IAP contains a cryptic promoter that can induce ectopic (unusual) expression of the agouti gene \cite{Rakyan:2002wz}. Methylation of this IAP prevents activation of the cryptic promoter and so the agouti gene is inactive, ultimately resulting in a mouse with a _pseudoagouti[^pseudoagouti_vs_agouti]_ coat. Conversely, when the IAP is unmethylated the mouse has a yellow coat since the cryptic promoter is activated and the agouti protein is expressed. Mice with intermediate levels methylation[^intermediate_methylation] at the IAP have a mottled coat due to mosaicism between cells.

[^pseudoagouti_vs_agouti]: These mice are properly described as pseudoagouti since they are heterozygous for the wildtype agouti allele ($A^{vy}/a$) but are phenotypically indistinguishable from true agouti mice, which are homozygous for the wildtype gene ($a/a$).

[^intermediate_methylation]: Intermediate methylation referring to the average methylation of the IAP across a pool of cells from the individual mouse.

The $A^{vy}$ allele is homozygous lethal, only heterozygous mice ($A^{vy}/a$) and homozygous wildtype mice ($a/a$) are viable, and so the mouse model is properly referred to as the agouti viable yellow mouse (__TODO: Raykan (2012) refer to homozygous $A^{vy}/A^{vy}$ mice, but I thought this was embryonically lethal???__).

An epiallele is a region of the genome that is genetically identical yet epigenetically variable. Epialleles are of most interest when they also confer phenotypic variabilitiy. The $A^{vy}$ allele is the archetype of a _metastable_ epiallele, which \cite{Rakyan:2002wz} define as an allele "at which the epigenetic state can switch and establishment is a probabilistic event". Importantly, such metastable epialleles are mitotically heritable (__TODO: Are there non-metastable epialleles?__).

The $A^{vy}$ epiallele was discoverable because much was known about its probable location and it could be verified through functional assays applied to the offspring of extensive mouse breedings (__TODO: Clarify__). The agouti gene was known to cause differences in coat colour and so it was the obvious candidate. Differences in expression of this gene were not due to genetic variability since the mice were isogenic, and environmental differences were minimal since all mice were raised in a similar laboratory environment. This pointed to the possibility that an epiallele was responsible for the phenotypic variability. I know describe a general strategy for identifying epialleles when there is a strong candidate gene. This is not exactly the strategy employed by \cite{Morgan:1999ds} but is a

## A general strategy for identifying epialleles of known genes

Suppose we have a gene whose expression is known to cause the phenotypic variation we are interested in. Furthermore, suppose that we suspect this difference in gene expression is due to epigenetic variation, such as when the variation in gene expression is observed in genetically identical animals raised in similar environments. These days we could perform bisulfite-sequencing of several samples with different values of the phenotype and look for methylation differences that associate with the phenotype.

The obvious first place to look is at the promoter of the gene, to look for the classical methylation/expression relationship described in Chapter \ref{chap:intro}. Things are rarely so simple and we must often widen our search. Cryptic promoters, such as that in the IAP insertion of the $A^{vy}$ locus, are good candidates (__TODO: How easy is it to identify cryptic promoters__). Enhancers, which can be many kilobases upstream or downstream of a gene, are more challenging to identify since the search must be further widened.

While challenging, this is the "easiest" experimental design from which to identify epialleles since we have a known phenotype and a candidate region(s) in which to search. But what if we want to perform a screen for _putative_ epialleles? A putative epiallele is any region of the genome where the samples display differential methylation. It is identified based solely on differential methylation and not on any association with phenotypic differences. Rather, these putative epialleles may be studied in greater detail in larger cohorts to see whether they associate with an as yet undetermined phenotype. This is the type of experiment that Emma Whitelaw came to Terry and I with to ask for assistance in analysing.

# Description of Emma's experiment

Extra evidence that a putative epiallele might be real:

* It's a retrotransposon. IAP is a retrotransposon and Morgan (1999) state "Because retrotransposons are abundant in mammalian genomes, this type of (epigenetic) inheritance may be common."

# Methodology

# Discussion of findings

# Conclusion

### 2014-10-19 Rmarkdown/avy.Rmd
Pre-break commit

@@
 ---
-title: "Epialleles in Agouti viable yellow mice"
+title: "Identifying epialleles in isogenic mice"
 author: "Peter Hickey"
@@
 
-The agouti viable yellow ($A^{vy}$) mouse model is perhaps the archetype of _heritable_ epigenetic modifications resulting in phenotypic variation.
-The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins this model, recently performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ locus, with the aim of identifying epialleles.
-An epiallele is a region of the genome where samples are genetically identical yet epigenetically variable.
+The agouti viable yellow ($A^{vy}$) mouse model is the archetypical _epiallele_, a genetically identical region where epigenetic variability result in phenotypic variation.
+The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins the $A^{vy}$ model, recently performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ locus, with the aim of identifying additional epialleles.
 
@@
 This chapter describes the experiment and how we addressed these challenges.
+It also discusses general considerations for identifying epialleles.
 
@@
 
-One of these agouti-associated obesity mutations is the agouti viable yellow allele ($A^{vy}$), discovered by \cite{Duhl:1994bx}.
+One of these agouti-associated obesity mutations is the agouti viable yellow allele ($A^{vy}$), discovered by (__TODO: Either 1960s [according to http://books.google.com.au/books?id=V6wNB2cGxMwC&lpg=PT250&dq=IAP%20near%20avy&pg=PT250#v=onepage&q=IAP%20near%20avy&f=false] or \cite{Duhl:1994bx}).
 The cause of the $A^{vy}$ mutation in the agouti protein was identified by a team led by Professor Emma Whitelaw \cite{Morgan:1999ds} as being due to methylation differences upstream of the agouti gene.
@@
 It is identified based solely on differential methylation and not on any association with phenotypic differences.
-Rather, these putative epialleles may be studied in greater detail in larger cohorts to see whether they associate with an as yet undetermined phenotype.
+Rather, these putative epialleles may be followed-up in larger cohorts to see whether they associate with a yet-to-be-determined phenotype.
 This is the type of experiment that Emma Whitelaw came to Terry and I with to ask for assistance in analysing.
 
-# Description of Emma's experiment
+# Description of experiment
 
-Extra evidence that a putative epiallele might be real:
+Five mice on a C57Bl/6 background underwent whole-genome bisulfite-sequencing.
+Two of the five mice are heterozygous for the $A^{vy}$ locus, one yellow and one pseudoagouti, while the remaining mice are homozygous wildtype agouti (Table \ref{tab:whitelaw_data}).
 
-* It's a retrotransposon. IAP is a retrotransposon and Morgan (1999) state "Because retrotransposons are abundant in mammalian genomes, this type of (epigenetic) inheritance may be common."
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+Sample code & Mouse number & Agouti genotype           \\ \midrule
+sample1     & mouse \#56   & $A^{vy}/a$ (yellow)       \\
+sample2     & mouse \#57   & $a/a$                     \\
+sample3     & mouse \#58   & $a/a$                     \\
+sample4     & mouse \#65   & $A^{vy}/a$ (pseudoagouti) \\
+sample5     & mouse \#66   &                           \\ \bottomrule
+\end{tabular}
+\caption{Description of mice in Whitelaw experiment.}
+\label{tab:whitelaw_data}
+\end{table}
+
+The aim of the study is identify putative epialleles, regions that differ in methylation between individuals but without any genetic cause (the mice are inbred and isogenic[^isogenic], so we assume methylation is not affected by genetic variants).
+
+[^isogenic]: Technically, the mice are isogenic except for a 3 Mb region surrounding the $A^{vy}$ allele.
+The $A^{vy}$ allele is inherited from a distant ancestor who was the offspring of mating between a C57Bl/6 mouse ($a$) with a C3H mouse ($A^{vy}$).
+It is inherited as part of the 3 Mb chunk of DNA from the C3H genome.
+
+While any differential methylation identified in this analysis is by definition a candidate epiallele, certain features are make stronger candidates.
+The Whitelaw are particularly interested in differential methylation of retrotransposons.
+Two of the best examples of metastable epialleles, $A^{vy}$ and $Axin^{Fu}$, are due to associations with retroviral elements \cite{Rakyan:2002wz}.
+To be of interest, the epiallele should also have a functional consequence, but this cannot be determined from this experiment alone.
+Rather, this will be determined by downstream analyses incorporating additional data and experiments of the putative epialleles.
+
+My involvement in the project was at a relatively late stage and focused on developing a method to analyse these data.
+I was to write a software implemention of this method but the final analysis would be performed by Harald Oey, a bioinformatician in the Whitelaw group.
+Consequently, this chapter focuses on the statistical details of the analysis rather than the biological results and their interpretation.
+
+## Data
+
+Harald aligned the reads to the mouse reference genome (mm9) using Bismark \cite{Krueger:2011eb}.
+__TODO: Check pre- and post-processing of reads.__ I received from Harald five tables, one for each sample, containing the number of methylated and unmethylated reads at each CpG.
+An example is shown below:
+
+```
+chr     start   end     beta    M       U
+chr2    3000163 3000164 0.7     7       3
+chr2    3000214 3000215 0.916666666666667       11      1
+chr2    3000221 3000222 1       12      0
+```
+
+I only received the data for chromosome 2, since these were sufficient for me to develop the methodology and its implementation.
+
+The mouse reference genome is based on the C57Bl/6 strain and does not contain the $A^{vy}$ allele.
+Therefore we cannot measure methylation at the IAP insertion in the $A^{vy}$ allele from the sequencing data.
+However, because the $A^{vy}$ allele is inserted in chromosome 2, there may be some "bleedover" into the region surrounding the insertion (chr2:154,760,839-154,760,969) (__TODO: This is far shorter than 3 Mb, why?__)
+
+__TODO: Comment on sequencing coverage__
 
@@
 
-# Discussion of findings
+The basic question is to identify loci where not all of the five mice have the same methylation level.
+There is no grouping structure of the samples.
+While we could conceivably group the mice by coat colour (yellow vs.
+agouti and pseudoagouti) or by genotype at the agouti locus ($a/a$ vs.
+$A^{vy} vs.
+a$) or by _epigenotype_ at the agouti locus ("yellow" vs.
+pseudoagouti vs.
+agouti), the Whitelaw group were interested in _any_ methylation differences between _any_ subset of the five mice[^n=1].
+This effectively makes this a one group experiment.
+
+[^n=1]: Both the "coat colour" and "agouti epigenotype" contrasts result in groups with $n = 1$, meaning we would be unable to estimate their corresponding within-group variances.
+
+This is the first example I had seen of a bisulfite-sequencing experiment with one-group.
+Surprisingly, this rules out all(?) published methods for identifying differential methylation, since these are all designed for identifying differential methylation in experiments with two or more groups.
+
+There is also the question of what sized "features" we are interested in testing for differential methylation.
+Harald and Emma told me that they were interested in both differentially methylated CpGs and differentially methylated regions, where the regions were not defined _a priori_.
+I took the commonly used approach of testing individual CpGs for differential methylation and then building runs of DMCs to define differentially methylated regions.
+
+My algorithm for calling differentially methylated regions is adapted from that of \cite{Hansen:2011gu}.
+I call my implementation `runDMC` (runs of differentially methylated cytosines) and it proceeds as follows:
+
+1. Identify candidate differentially methylated cytosines (_cDMCs_).
+2. Declare contiguous runs of _cDMCs_ where the $IPD$ between _cDMCs_ is less than some threshold to be candidate differentially methylated regions (_cDMRs_).
+3. Summarise _cDMRs_ and apply filters.
+
+## Step 1: Identifying candidate differentially methylated cytosines
+
+At each CpG, $i$, we are testing the following hypothesis:
+
+\begin{equation*}
+
+H_{0}:& \beta_{i, j} = \beta_{i, 0}, j = 1, \ldots, 5 \\
+  vs.& \\
+H_{1}:& \exists j = 1, \ldots, 5 s.t. \beta_{i, j} \neq \beta_{i, 0}
+
+\end{equation*}
+
+where $\beta_{i, 0}$ is estimated from the data. For each CpG we can summarise the data by a $5 \times 2$ contingency table:
+
+\begin{table}[h]
+\centering
+\begin{tabular}{lll}
+\hline
+Sample code & Methylated reads & Unmethylated reads \\ \hline
+sample1     & $m_{i, 1}$       & $u_{i, 1}$         \\
+sample2     & $m_{i, 2}$       & $u_{i, 2}$         \\
+sample3     & $m_{i, 3}$       & $u_{i, 3}$         \\
+sample4     & $m_{i, 4}$       & $u_{i, 4}$         \\
+sample5     & $m_{i, 5}$       & $u_{i, 5}$         \\ \hline
+\end{tabular}
+\caption{A $5 \times 2$ contingency table summarising the number of methylated and unmethylated reads for each sample, $j$, at the $i^{th}$ CpG.}
+\label{tab:5x2}
+\end{table}
+
+This amounts to the classical test of differences in proportions for a group of samples.
+I used the simple $\chi^{2}$ test, as implemented by the `prop.test` function in `R` (__CITE__).
+This compares the observed and expected counts of methylated and unmethylated reads under the assumption that all 5 samples have equal methylation levels at this CpG.
+In fact, I wrote a simple wrapper around `prop.test`, called `propTest` (__TODO: Include code in appendix__), so that only the the value of the $\chi^{2}$ test statistic, the degrees of freedom and the resulting P-value were retained, since the rest of the object returned by `prop.test` was not required.
+`propTest` can also parallelise the computation by chromosome for a faster runtime.
+
+__TODO: Plot and comment on the distribution of P-values__
+
+A CpG is declared a _cDMC_ if the P-value is in the lower $5\%$ of all P-values across the genome.
+A stricter criteria is required if the aim of the analysis is to identify differentially methylated cytosines.
+However, this first step is rather designed to identify sites for the `runDMC` algorithm (described below).
+
+The question of
+
+## Identifying differentially methylated regions
+
+The `runDMC` algorithm is similar to one proposed in \cite{Hansen:2011gu}, which is implemented in the `bsseq` R/Bioconductor package (__CITE__).
+`bsseq` defines differentially methylated region as runs of CpGs with a t-statistic greater than an empirically chosen threshold.
+Because `bsseq` is designed for two-group experiment, it also requires that the t-tatistic has the same sign for all CpGs in the region, indicating that the effect is in the same direction for all CpGs in the DMR.
+
+Rather than using the $\chi^{2}$ test-statistics, `runDMC` uses the corresponding P-values.
+The reason is that for some CpGs only a subset of the samples have sequencing data, hence the test statistic has fewer degrees of freedom.
+By using the P-value rather than the raw $\chi^{2}$ test statistic, we effectively normalise for the degrees of freedom.
+
+`runDMC` is written in `C++` and interfaces with `R` via the `Rcpp` package (__CITE__).
+The `C++` version is some $25$ times faster than the equivalent code written in `R` (__TODO: Include benchmark data__).
+The user passes it an object containing the set of CpGs, each annotated by whether it is a _cDMC_, and specifies __UP TO HERE__
+
+# Results
+
+# Discussion
 

### 2014-10-19 Rmarkdown/avy.Rmd
Section-level tidying

@@
 
-# Chapter overview
+## Chapter overview
 
-The agouti viable yellow ($A^{vy}$) mouse model is the archetypical _epiallele_, a genetically identical region where epigenetic variability result in phenotypic variation.
-The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins the $A^{vy}$ model, recently performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ locus, with the aim of identifying additional epialleles.
+The agouti viable yellow ($A^{vy}$) allele is the archetypical _epiallele_, a genetically identical region where epigenetic variability result in phenotypic variation.
+The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins the $A^{vy}$ mouse model,  performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ allele, with the aim of identifying additional epialleles.
 
@@
 
-# Introduction + literature review
+## Introduction + literature review
 
@@
 
-## A general strategy for identifying epialleles of known genes
+### A general strategy for identifying epialleles of known genes
 
@@
 
-# Description of experiment
+## Description of experiment
 
@@
 
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
 \begin{table}[h]
@@
 
-## Data
+### Data
 
@@
 
-# Methodology
+## Methodology
 
@@
 This is the first example I had seen of a bisulfite-sequencing experiment with one-group.
-Surprisingly, this rules out all(?) published methods for identifying differential methylation, since these are all designed for identifying differential methylation in experiments with two or more groups.
+Surprisingly, this rules out all(?) published methods for identifying differential methylation, since these are all designed for identifying differential methylation in experiments between two or more groups.
 
@@
 Harald and Emma told me that they were interested in both differentially methylated CpGs and differentially methylated regions, where the regions were not defined _a priori_.
-I took the commonly used approach of testing individual CpGs for differential methylation and then building runs of DMCs to define differentially methylated regions.
+For defining DMRs I took the commonly used approach of testing individual CpGs for differential methylation and then defining DMRs as runs of DMCs.
 
-My algorithm for calling differentially methylated regions is adapted from that of \cite{Hansen:2011gu}.
+My algorithm for defining differentially methylated regions is adapted from that of \cite{Hansen:2011gu}.
 I call my implementation `runDMC` (runs of differentially methylated cytosines) and it proceeds as follows:
@@
 
-## Step 1: Identifying candidate differentially methylated cytosines
+### Step 1: Identifying candidate differentially methylated cytosines
 
@@
 
-\begin{equation*}
-
-H_{0}:& \beta_{i, j} = \beta_{i, 0}, j = 1, \ldots, 5 \\
-  vs.& \\
-H_{1}:& \exists j = 1, \ldots, 5 s.t. \beta_{i, j} \neq \beta_{i, 0}
-
-\end{equation*}
+$$
+\begin{aligned}
+H_{0, i}: & \forall j \; B_{i, j} = B_{i, 0} \\
+& vs.  \\
+H_{1, i}: & \exists j: \; B{i, j} \neq B_{i, 0}
+\end{aligned}
+$$
 
@@
 
-The question of
+__TODO: Some summaries of the cDMCs, e.g., top-ranked CpGs__
 
-## Identifying differentially methylated regions
+__TODO: A discussion of FDR correction for single-loci tests__
+
+### Identifying differentially methylated regions
 
@@
 The `C++` version is some $25$ times faster than the equivalent code written in `R` (__TODO: Include benchmark data__).
-The user passes it an object containing the set of CpGs, each annotated by whether it is a _cDMC_, and specifies __UP TO HERE__
+The user passes `runDMC` an object containing the set of CpGs, each annotated by whether it is a _cDMC_, and specifies the maximal $IPD$ of CpGs and number of _non-cDMCs_ before a run is "broken".
+The output is an object with the start and end co-ordinates of each run, as well as some basic summary statistics such as the number of _cDMCs_ and _non-cDMCs_.
+These cDMRs are then annotated and filtered in the final stage.
 
-# Results
+__TODO: Some summaries of the `runDMC` output, e.g., number of runs, lengths,
 
-# Discussion
+### Summarising and filtering candidate differentially methylated regions
 
-# Conclusion
+The vast majority of "runs" returned by `runDMC` are a single _cDMC_.
+Obviously we want to filter these out as they are single loci not runs.
+For example, we may require a run to contain at least 3 _cDMCs_.
+
+We may want to summarise each _cDMR_ by a statistic that measures how different the entire region is between the sample, not just the individual CpGs.
+However, care should be taken in the interpretation of such a statistic.
+We know _a priori_ that the CpGs in an cDMR are 'significantly' differentially methylated and so we expect the region to also be 'significantly' differentially methylated.
+Nevertheless, a summary statistic may be useful in ranking candidate DMRs.
+
+One potential summary statistic is the Mantel-Haenszel (_MH_) statistic of the CpGs in the _cDMR_.
+The Mantel-Haenszel statistic can be used to test that two nominal variables are conditional independent in each stratum.
+In other words,
+
+> The Mantel-Haenszel test tests the null hypothesis that two nominal variables [M and U] are conditionally independent in each stratum [at each CpG], assuming that there is no three-way interaction [assuming that the effect is consistent across all CpGs in the region] (From the help page for the `R` implementation of the Mantel-Haenszel test (see ``?mantelhaen.test`) [with my comments in square brackets])
+
+Testing the conditional independence of $M$ and $U$ _at a single CpG_ is equivalent to testing that $\beta_{j} = \beta_{0}$ for all $j = 1, \ldots, 5$ mice at that CpG.
+Therefore, the Mantel-Haenszel test is equivalent to testing  the null hypothesis that all mice have the same methylation level at each CpG ($\beta_{0, i}$) in the region, although this may vary from CpG to CpG, i.e.
+$\beta_{0, i}$ need not equal $\beta_{0, i + 1}$).
+
+In our case, with 2 categories ($M$ and $U$) and 5 mice, the Mantel-Haenszel statistic has an approximate $\chi^{2}((5 - 1) \times (2 - 1) \times(2 - 1)) = \chi^{2}(4)$ distribution under the null hypothesis of no conditional independence of $M$ and $U$ across the sites[^df].
+Furthermore, the Mantel-Haenszel statistic is the optimal statistic for the alternative hypothesis stating a consistent (same direction) difference across all CpGs in the region \cite{Somes:1986uk}.
+We may with to rank regions based on the value of the Mantel-Haenszel statistic.
+
+[^df]: If only a $J < 5$ of the mice have non-zero coverage across all CpGs in the region, then the degrees of freedom are $J - 1$.
+
+There are some important caveats mentioned above that warrant further discussion.
+One of these, the assumption of no three-way interaction, which is required for the validity of the null distribution of the Mantel-Haenzel statistic, is related to the idea of "consistency" for a DMR.
+
+By "consistency" I mean that the difference in methylation across the region occurs within the same (subset) of mice.
+This is best illustrated by describing an "inconsistent" DMR.
+Suppose we have a region containing five CpGs, each of which is differentially methylated.
+However, suppose that at the first CpG it is `sample1` that is "different", at the second CpG it is `sample2` that is "different", etc.
+Then this region is not differentially methylated in any single subset of mice, which is the standard interpretation of a DMR.
+It must be emphasised that the `runDMC` algorithm provides no guarantees that a _cDMR_ is consistent since it simply builds runs of _cDMCs_.
+
+Returning to the Mantel-Haenszel statistic, the assumption of no three-way interaction means no interaction between mouse, methylation level and CpG across the region.
+Another way to think of this is that no three-way interaction means that there is no CpG-to-CpG heterogeneity across the region in the interaction between mouse and methylation level, i.e., that any difference in methylation between the mice is _consistent_ across the CpGs.
+__TODO: Discuss interpretation with Terry__.
+
+We could of course test whether there is evidence of a three-way interaction.
+Specifically, we could fit a log-linear model to the tables in each region and compare the "full" model (mouse $\times$ CpG $\times$ methylation level) against the reduced "all main effects and two-way interactions" model (mouse $\times$ CpG + mouse $\times$ methylation level + CpG $\times$ methylation level) via a likelihood ratio test.
+Again, we do not use the P-values to strictly rule regions as "significant" or "non-significant", but rather to highlight regions  that may be inconsistent.
+
+__TODO: Is this what we want?__
+
+```{r, eval = FALSE, echo = FALSE}
+# Testing whether no 3-way interaction is the same thing as consistency
+
+# Some data where each CpG is sig but region is not consistent
+x <- matrix(c(3, 40, 30, 40, 30, 40, 30, 40, 30, 40,
+              30, 40, 3, 40, 30, 40, 30, 40, 30, 40,
+              30, 40, 30, 40, 3, 40, 30, 40, 30, 40,
+              30, 40, 30, 40, 30, 40, 3, 40, 30, 40,
+              30, 40, 30, 40, 30, 40, 30, 40, 3, 40),
+              ncol = 10)
+
+tx <- array(x, dim = c(2, 5, 5))
+
+# Every table is significant
+for (i in 1:5) {
+  print(chisq.test(tx[, , i]))
+}
+
+# MH test is non-significant
+mantelhaen.test(tx)
+
+# But that's because no three-way interaction assumption is violated?
+llx <- loglin(table = tx, margin = list(c(1, 2), c(1, 3), c(2, 3)))
+1 - pchisq(llx$lrt, llx$df)
+
+# Try with some permuted data: Every CpG is DMC and region is consistent
+y <- rbind(rep(c(3, 40), 5), matrix(rep(c(30, 40), 20), ncol = 10))
+
+ty <- array(y, dim = c(2, 5, 5))
+
+# Every table is significant
+for (i in 1:5) {
+  print(chisq.test(ty[, , i]))
+}
+
+# MH test is highly significant
+mantelhaen.test(ty)
+
+# And there's no 3-way interaction
+lly <- loglin(table = ty, margin = list(c(1, 2), c(1, 3), c(2, 3)))
+1 - pchisq(lly$lrt, lly$df)
+```
+
+## Results
+
+## Discussion
+
+* Analysis doesn't incorporate co-methylation
+* Consistency
+* Experiment is a bit of a fishing expedition?
+
+## Conclusions

### 2014-10-19 Rmarkdown/methsim.Rmd
Section-level tidying

@@
 
-# Chapter overview
+## Chapter overview
 
-# Introduction
+## Introduction
 
@@
 
-# Background + literature review
+## Background + literature review
 
@@
 
-## Methods for simulating bisulfite-sequencing reads
+### Methods for simulating bisulfite-sequencing reads
 
@@
 
-## Methods for simulating aggregate methylation levels
+### Methods for simulating aggregate methylation levels
 
@@
 Consequently, the simulation model is often only briefly described and a software implementation is rarely made available.
-In fact, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} has a software implementation available.
+In fact, of the methods reviewed in this section, only that of \cite{Lacey:2013iy} has a software implementation available.
 
@@
 
-### Simulating sequence depth
+#### Simulating sequence depth
 
@@
 
-### Simulating $B_{i, j}$
+#### Simulating $B_{i, j}$
 
@@
 
-### Simulating co-methylation
+#### Simulating co-methylation
 
@@
 This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependence.
-This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.
+This idea takes its inspiration from \cite{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.
 
@@
 
-### Simulating positions of methylation loci
+#### Simulating positions of methylation loci
 
@@
 
-### Other model-based simulations
+#### Other model-based simulations
 
@@
 
-## Methods based entirely on sampling real data
+### Methods based entirely on sampling real data
 
@@
 
-## Simulating the truth
+### Simulating the truth
 
@@
 
-### Segmenting the methylome
+#### Segmenting the methylome
 
@@
 
-### Simulating a single haploid methylome
+#### Simulating a single haploid methylome
 
@@
 However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome.
-For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h, j)}$, to estimate or otherwise assign.
+For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h, j)}}$, to estimate or otherwise assign.
 
@@
 
-### Simulating a single sample
+#### Simulating a single sample
 
-Given a set of transition matrices, $\Big { P_{i^{(h, j)} \Big }$, we can readily simulate a realisation of the methylation states along a haploid genome, $Z_{h, j}$.
+Given a set of transition matrices, $\Big \{ P_{i^{(h, j)}} \Big \}$, we can readily simulate a realisation of the methylation states along a haploid genome, $Z_{h, j}$.
 Naively, we might simulate a single sample by simply generating a realisation for each $h$ of $Z_{h, j}$.
@@
 
-What this means is that we want some degree of dependence between the set of transition matrices for haploid methylome $h$, $\Big { P_{i^{(h, j)} \Big }$, and for those for haploid methylome $h'$, $\Big { P_{i^{(h', j)} \Big }$.
+What this means is that we want some degree of dependence between the set of transition matrices for haploid methylome $h$, $\Big \{ P_{i^{(h, j)}} \Big \}$, and for those for haploid methylome $h'$, $\Big \{ P_{i^{(h', j)}} \Big \}$.
 This is asking quite a lot of our simple simulation model.
@@
 To begin, we assume that all $H$ haploid genomes in the sample have the same co-methylation structure.
-That is equivalent to defining $\Big { P_{i^{(h, j)} \Big } := $\Big { P_{i^{(j)} \Big }$ for all $h$.
+That is equivalent to defining $\Big \{ P_{i^{(h, j)}} \Big \} := \Big \{ P_{i^{(j)}} \Big \}$ for all $h$.
 __TODO: Comment on this assumption__
 
-Naively, we might simulate a sample by simulating $H$ realisations of the Markov chain with transition matrices $\Big { P_{i^{(j)} \Big }$.
+Naively, we might simulate a sample by simulating $H$ realisations of the Markov chain with transition matrices $\Big \{ P_{i^{(j)}} \Big \}$.
 However, this gives no control over the dependence between the methylation states at the $i^{th}$ locus over the $H$ haploid genomes in the $j^{th}$ sample.
@@
 
-Instead, `methsim` simulates $R << H$ realisations of the Markov chain with transition probabilities given by $\Big { P_{i^{(j)} \Big }$.
+Instead, `methsim` simulates $R << H$ realisations of the Markov chain with transition probabilities given by $\Big \{ P_{i^{(j)}} \Big \}$.
 It then weights these realisations according to the observed heterogeneity for each region.
@@
 
-### Simulating multiple samples from the same group
+#### Simulating multiple samples from the same group
 
-### Simulating differential methylation
+#### Simulating differential methylation
 
-## Parameter estimation
+### Parameter estimation
 
-# Implementation
+## Implementation
 
-# Case study
+## Case study
 
-# Discussion of findings
+## Discussion of findings
 
-# Conclusions
+## Conclusions

### 2014-10-19 Rmarkdown/datasets.Rmd
Began draft of description of datasets used in these

---
title: "Datasets used in thesis"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

## Chapter overview

I have analysed more than 50 whole-genome bisulfite-sequencing datasets during the course of my PhD studies. This brief chapter documents those that I use in my thesis.

## Lister data

The _Lister data_ refer to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}. The Lister data were the largest publicly available human WGBS datasets until quite recently.

The WGBS libraries from the \cite{Lister:2009hy} paper were the first published WGBS libraries from mammalian DNA. A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal fibroblasts (_FF_), with those from a pluripotent tissue, embryonic stem cells (_H1_). Each tissue was run in duplicate. While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing. In any case, the published analyses pool these technical duplicates, which ignores all between-replicate variability.

The WGBS libraries from the \cite{Lister:2011kg} include some created and published by other groups (see __TABLE__). These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells. Unfortunately, there are no replicates for any of these samples.

__TODO: Table of datasets__

There are a four "mini datasets" within the Lister data that I make some use of in my thesis. These are described below and illustrated in __FIGURE__.

The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper. The _ADS_ sample, a human adipose tissue cell line, is the "founder" of this mini dataset. The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line. Specifically, the _ADS-adipose_ sample is purified adipocytes from _ADS_ while the _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.

The second mini dataset I refer to as the _Lister-FF_ data and includes samples _FF_, _FF-iPSC\_6.9_, _FF-iPSC\_19.7_, _FF-iPSC\_19.11_ and _FF-iPSC\_11.11+BMP4_, all from the 2011 paper. The _FF_ sample, a foreskin fibroblasts cell line, is the "founder" of this this mini dataset. The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from _FF_. In fact, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are subclones derived from _FF-iPSC\_19_, whose methylome was not sequenced. I believe _FF-iPSC\_6.9_ is an independently derived iPSC cell line from _FF_, although this isn't made clear in the original publication. The _FF-iPSC\_19.11+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _FF-iPSC\_19.11_ by growing a clone of it in bone morphogenic protein 4 (BMP4).

The third mini dataset I refer to as the _Lister-IMR90_ data and includes samples _IMR90\_r1_ (2009), _IMR90\_r2_ (2009) and _IMR90-iPSC_ (2011). The _IMR90-iPSC_ sample is an induced pluripotent stem cell line derived from the _IMR90_ cell line.

The final mini dataset I refer to as the _Lister-H1_ data and includes samples _H1\_r1_ (2009), _H1\_r2_ (2009) and _H1+BMP4_ (2011). The _H1+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _H1_ by growing a clone of it in bone morphogenic protein 4 (BMP4).

The aligned reads for the \cite{Lister:2009hy} data were downloaded from [http://neomorph.salk.edu/human_methylome/](http://neomorph.salk.edu/human_methylome/). The aligned reads for the \cite{Lister:2011kg} data were downloaded from [http://neomorph.salk.edu/ips_methylomes/](http://neomorph.salk.edu/ips_methylomes/). I wrote Python scripts to convert these files to `SAM` files, these scripts are available from [https://github.com/PeteHaitch/Lister2BAM](https://github.com/PeteHaitch/Lister2BAM). These `SAM` files were then converted to `BAM` files and duplicate reads were marked using Picard's `MarkDuplictes` routine [http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/).

## EPISCOPE data

__TODO: Check permission to use data in thesis__

## Whitelaw data

I describe this dataset in detail in Chapter \ref{chap:avy}.

__TODO: Check permission to use data in thesis__

## Ziller data

## Seisenberger data

### 2014-10-28 Rmarkdown/datasets.Rmd
Described EPISCOPE, Ziller and Seisenberger data

@@
 
-I have analysed more than 50 whole-genome bisulfite-sequencing datasets during the course of my PhD studies. This brief chapter documents those that I use in my thesis.
+I have analysed more than 50 whole-genome bisulfite-sequencing datasets during the course of my PhD studies. This short chapter documents those that I use in my thesis.
 
@@
 The WGBS libraries from the \cite{Lister:2009hy} paper were the first published WGBS libraries from mammalian DNA.
-A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal fibroblasts (_FF_), with those from a pluripotent tissue, embryonic stem cells (_H1_).
+A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (`IMR90`), with those from a pluripotent tissue, embryonic stem cells (`H1`).
 Each tissue was run in duplicate.
 While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing.
-In any case, the published analyses pool these technical duplicates, which ignores all between-replicate variability.
+In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.
 
@@
 These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells.
-Unfortunately, there are no replicates for any of these samples.
+There are no replicates for any of these samples.
 
@@
 
-The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper.
-The _ADS_ sample, a human adipose tissue cell line, is the "founder" of this mini dataset.
-The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line.
-Specifically, the _ADS-adipose_ sample is purified adipocytes from _ADS_ while the _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.
+The first I refer to as the _Lister-ADS_ data and includes samples `ADS`, `ADS-adipose` and `ADS-iPSC`, all from the 2011 paper.
+The `ADS` sample, a human adipose tissue cell line, is the "founder" of this mini dataset.
+The `ADS-adipose` and `ADS-iPSC` are both derived from the `ADS` cell line.
+The `ADS-adipose` sample are "adipocytes derived from the `ADS` cells through adipogenic differentiation conditions".
+The `ADS-iPSC` cell line is an induced pluripotent stem cell line derived from `ADS`.
 
-The second mini dataset I refer to as the _Lister-FF_ data and includes samples _FF_, _FF-iPSC\_6.9_, _FF-iPSC\_19.7_, _FF-iPSC\_19.11_ and _FF-iPSC\_11.11+BMP4_, all from the 2011 paper.
-The _FF_ sample, a foreskin fibroblasts cell line, is the "founder" of this this mini dataset.
-The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from _FF_.
-In fact, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are subclones derived from _FF-iPSC\_19_, whose methylome was not sequenced.
-I believe _FF-iPSC\_6.9_ is an independently derived iPSC cell line from _FF_, although this isn't made clear in the original publication.
-The _FF-iPSC\_19.11+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _FF-iPSC\_19.11_ by growing a clone of it in bone morphogenic protein 4 (BMP4).
+The second mini dataset I refer to as the _Lister-FF_ data and includes samples `FF`, `FF-iPSC\_6.9`, `FF-iPSC\_19.7`, `FF-iPSC\_19.11` and `FF-iPSC\_11.11+BMP4`, all from the 2011 paper.
+The `FF` sample, a foreskin fibroblast cell line, is the "founder" of this this mini dataset.
+The `FF-iPSC\_6.9`, `FF-iPSC\_19.7` and `FF-iPSC\_19.11` are all induced pluripotent stem cell lines derived from `FF`.
+In fact, `FF-iPSC\_19.7` and `FF-iPSC\_19.11` are subclones derived from `FF-iPSC\_19`, whose methylome was not sequenced.
+I believe `FF-iPSC\_6.9` is an independently derived iPSC cell line from `FF`, although this isn't made clear in the original publication.
+The `FF-iPSC\_19.11+BMP4` sample is a trophoblast cell line derived by _in vitro_ differentiating the `FF-iPSC\_19.11` by growing a clone of it in bone morphogenic protein 4 (_BMP4_).
 
-The third mini dataset I refer to as the _Lister-IMR90_ data and includes samples _IMR90\_r1_ (2009), _IMR90\_r2_ (2009) and _IMR90-iPSC_ (2011).
-The _IMR90-iPSC_ sample is an induced pluripotent stem cell line derived from the _IMR90_ cell line.
+The third mini dataset I refer to as the _Lister-IMR90_ data and includes samples `IMR90\_r1` (2009), `IMR90\_r2` (2009) and `IMR90-iPSC` (2011).
+The `IMR90-iPSC` sample is an induced pluripotent stem cell line derived from the `IMR90` cell line.
 
-The final mini dataset I refer to as the _Lister-H1_ data and includes samples _H1\_r1_ (2009), _H1\_r2_ (2009) and _H1+BMP4_ (2011).
-The _H1+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _H1_ by growing a clone of it in bone morphogenic protein 4 (BMP4).
+The final mini dataset I refer to as the _Lister-H1_ data and includes samples `H1\_r1` (2009), `H1\_r2` (2009) and `H1+BMP4` (2011).
+The `H1+BMP4` sample is a trophoblast cell line derived by _in vitro_ differentiating the `H1` by growing a clone of it in bone morphogenic protein 4 (BMP4).
 
@@
 
-__TODO: Check permission to use data in thesis__
+The EPISCOPE data were kindly provided to me by Professor Susan Clark (Garvan Institute of Medical Research, Sydney) and Dr Peter Molloy (CSIRO Animal, Food and Health Sciences).
+The data are from three human donors across four different tissues, for a total of 12 whole-genome bisulfite-sequencing libraries.
+
+The four tissues are:
+
+* BUF: Buffy coat layer, which are leukocytes and platelets derived by centrifugation of a whole blood sample.
+* SA: Subcutaneous adidose tissue, which is fat found just below the skin.
+Unlike visceral fat, "subcutaneous fat is not related to many of the classic obesity-related pathologies, such as heart disease, cancer, and stroke" (__TODO: Get better reference than wikipedia__).
+* VA: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs (stomach, liver, intestines, kidneys, etc.) (__TODO: Get better reference than wikipedia__).
+* VAT: Visceral adipocytes, which are derived from the VA tissue.
+
+The data are summarised in __TABLE__
+
+__TODO: Table of datasets__
+__TODO: Get age of donors__
+
+The sequencing data for these 12 samples were processed and aligned by Aaron Statham (Garvan Institute of Medical Research).
+Each sample was aligned to the human reference genome (hg19) using Bismark (v0.8.3) with the Bowtie2 backend.
+The default alignment options were used, except that the maximum insert size for valid paired-end alignments was set to 1000 instead of 500 (`-X 1000`).
+An example command is shown below:
+
+```
+bismark -p 4 --bowtie2 -X 1000 --unmapped --ambiguous --gzip --bam -o SAMPLE_NAME  REF_GENOME -1 READ_1.fq -2 READ_2.fq
+```
+
+Duplicate reads had been removed from the `BAM` files that I received.
 
@@
 
-__TODO: Check permission to use data in thesis__
-
 ## Ziller data
 
+The Ziller are a subset of the data used in \cite{Ziller:2013fs}.
+Specifically, I use a convenience sample of 21 (__TODO: Check final numbers__) WGBS libraries.
+These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research), who already had aligned and processed BAM files for this subset of samples.
+
+There are samples from the frontal cortex from two "normal" women (`Frontal_cortex_normal_1` and `Frontal_cortex_normal_2`) and from two women who had Alzheimer's disease (`Frontal_cortex_AD_1` and `Frontal_cortex_AD_2`); a sample from a human liver carcinoma cell line (`HepG2_cell_line`); a new sample from the IMR90 lung fibroblast cell line (`IMR90_cell_line`); and samples from a matched tumour-normal pair for colon cancer (`Colon_Tumor_Primary` and `Colon_Tumor_Primary`).
+Full details of the Ziller data used in my thesis are described in __TABLE__.
+Note that I consider the samples described as `replicates` to be technical replicates, since each replicate comes from the same biological sample.
+
+__TODO: Table of datasets. Only include those samples with properly processed files (i.e., exclude truncated files)__
+
 ## Seisenberger data
+
+The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \cite{Seisenberger:2012ko}.
+Like the Ziller data, these were a convenience sample provided to me by a colleague.
+I thank Felix Krueger (Babraham Institute) who processed and aligned the reads, and sent me the `BAM` files.
+
+I have the data for three samples, detailed in __TABLE__.
+Developmentally, the samples are ordered `J1_1`, `E6.5_epiblast_1` and `E16.5_male_1`.
+Both the `E6.5_epiblast_1` and `E16.5_male_1` samples are from a pool of $10-30$ embryos.
+
+__TODO: Table of datasets__

### 2014-10-28 Rmarkdown/biology_background.Rmd
Split intro and biology background. Completed draft of first half of biology background chapter.

---
title: "Biology background"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\section{Chapter Overview}\label{sec:biology_background_chapter_overview}

- [x] DONE

In this chapter I summarise the basic biology necessary for
understanding my thesis. I then introduce DNA methylation, describe some of its biological roles and discuss the types of questions asked in experiments studying DNA methylation. Finally, I describe assays for studying DNA methylation, with a particular focus on the whole-genome bisulfite-sequencing assay known as _methylC-seq_.

\section{From DNA to genomes}\label{ref:from_dna_to_genomes}

- [x] DONE

The genome is the genetic material of an organism. In most organisms, the genome is encoded by deoxyribonucleic acid, _DNA_. __FIGURE__ beautifully illustrates the hierarchy that takes us from DNA through to the genome.

In eukaryotes[^eukaryotes], which includes plants, fungi and animals, the DNA is wound around repeating units of eight _histone_ protein cores to form _nucleosomes_, which are the fundamental unit of _eukaryotic chromatin_. Chromatin compactly packages the DNA into _chromosomes_, so that the organism's complete nuclear DNA, its nuclear genome, might fit into the nucleus of its cells[^extra_nuclear_genomes].

[^eukaryotes]: Eukaryotes are organisms composed of one or more cells with a distinct nucleus and cytoplasm. Most of a eukaryotic cell's DNA is contained in the nucleus \cite{Alberts:2007vj}.

[^extra_nuclear_genomes]: Not all of an organism's genome is present in the nucleus of a cell. Important exceptions are mitochondrial DNA (mtDNA) found in animals, plants and fungi and chloroplast DNA (ctDNA) found in plants.

\subsection{DNA}\label{sec:DNA}

The double helix is the most common, and most famous, structure of DNA (__REFERENCE__). In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_. One strand is called the 5' strand (pronounced "5 prime strand") - also known as the sense strand, Crick strand or top strand - and the other strand called the 3' strand (pronounced "3 prime strand") - also known as the antisense strand, Watson strand or bottom strand. Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_). These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix.

[__FIGURE OF DNA BASE  PAIRING__](http://upload.wikimedia.org/wikipedia/commons/4/4c/DNA_Structure%2BKey%2BLabelled.pn_NoBB.png).

A gene is a sequence of DNA that is _transcribed_ to produce a functional product called ribonucleic acid or _RNA_ (Source: [http://sandwalk.blogspot.com.au/2007/01/what-is-gene.html](http://sandwalk.blogspot.com.au/2007/01/what-is-gene.html)). RNA may in turn be translated into a protein sequence or perform other roles in the regulation of gene expression (__CITE__). It is important to note here than not all DNA is transcribed into RNA. That is not to say that untranscribed DNA is unimportant. For instance, there are untranscribed _regulatory sequences_ of DNA that determine whether a nearby gene is transcribed. There is also _junk DNA_, _garbage DNA_, __etc.__ (__SOURCE__). Conversely, not all transcribed DNA is a gene. DNA transcription is "permissive" and there are many DNA sequences that are transcribed by accident or in error.

DNA is able to self-replicate so that, for example, eukaryotic cells created during _mitosis_ contain the same DNA as the "parent" cell. During DNA replication, the two strands are separated and each strand's _complementary_ DNA sequence is copied by an enzyme called _DNA polymerase_. It is because the two strands of DNA are complementary that ensures the daughter cell contains the same DNA sequence as the parent cell[^dna_replication].

[^dna_replication]: This of course ignores errors in the replication process. Such errors are very rare events but because DNA replication happens __HOW FREQUENTLY__ these events do occur. There are error-correcting processes that reduce the chance that such an error is retained in the daughter sequence, however, these are not perfect. Hence errors in DNA replication are one source of what are known as _mutations_ in the DNA.

\subsection{Nucleosomes and chromatin}\label{sec:nucleosome_and_chromatin}

Basic descriptions of nucleosomes and chromatin are ripe with analogies. In one such popular analogy, the histones are the 'spool' around which the DNA 'thread' is wrapped to form nucleosomes (__CITE__). These nucleosomes themselves have the appearance of "beads on a string" when viewed under an electron microscope (__CITE__).

The core of a nucleosome consists of four pairs of histones, H2A, H2B, H3 and H4, which are consequently known as the _core histones_. A fifth histone, H1/H5, is known as the _linker histone_. Each of these histones has a 'tail' consisting of a string of amino acids. These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins (__CITE__) (see below Section \ref{epigenetics} on epigenetics).

The nucleosomes are interconnected by _linker DNA_ to form the macromolecule  called chromatin. The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone (__CITE: http://www.nature.com/scitable/topicpage/dna-packaging-nucleosomes-and-chromatin-310#__).

Chromatin is frequently described as either _heterochromatin_ or _euchromatin_. Heterochromatin, or closed chromatin, is more tightly packed than euchromatin, or open chromatin. Heterochromatin is associated with transcriptionally repressed regions of the genome because the machinery required to translate DNA to RNA is unable to physically access the DNA. In contrast, euchromatin is associated with transcriptionally active regions of the genome since its more open nature allows easier access for the translational machinery.

The same region of an individual's genome may vary between heterochromatin and euchromatin states at different stages of the organism's life and in different cells of the organism. This is one mechanism which by gene expression is regulated.

Chromatin is further coiled up into various literally-named structures, such as the 30-nanometre and 250-nanometre fibres, and, ultimately, packaged into _chromosomes_. The set of chromosomes makes up an individual's _genome_.

\subsection{Genomes}\label{sec:genomes}

As is clear from the above description, the genome is a complex three-dimensional structure. Nonetheless, in bioinformatics and computational biology, the genome is mostly considered as a single-stranded, one-dimensional string of the nucleobases _A_, _C_, _G_ and _T_.

While eukaryotic genomes share the above-described features, and indeed share many regions of common DNA sequence, eukaryotic genomes come in many shapes and sizes. I only discuss the genomes of three organisms relevant to my thesis: Human (_Homo sapiens_), Mouse (_Mus musculus_) and Arabidopsis (_Arabidopsis thaliana_).

\subsubsection{Human genome}\label{sec:human_genome}

Humans are _diploid_ organisms, meaning that we have two copies of each chromosome in a typical cell[^somatic_cell]. We inherit one chromosome of each pair from our mother and one from our father. A typical human cell has 23 pairs of nuclear chromosomes, 22 _autosomes_ and 1 pair of _sex chromosomes_, as well as hundreds or thousands of copy of the small mitochondrial chromosome[^mtDNA].

[^somatic_cell]: A sperm or egg cell is haploid and has a single (recombined) copy of each chromosome.

[^mtDNA]: The mitochondrial DNA is maternally inherited.

The length of a chromosome is typically reported as the number of DNA base pairs (bp) in a single copy of that chromosome and the _haploid_ length of a genome is the sum of these chromosome lengths. The haploid human genome is approximately 3.3 billion base pairs long.

A _reference genome_ was completed by the Human Genome Project in 2003 (__CITE__). This reference genome does not represent the genome of any one  human since it uses DNA donated by several different people. Rather, a reference genome is a kind of map or scaffold that can be used to identify similarities and differences between individual genomes.

The human genome has obvious uses in medical research and biotechnology, but is also used to learn about evolution and human history, such as migration. Every person, except genetically identical twins, has their own unique genome. However, the genomes of any two randomly selected people are identical at __WHAT PERCENTAGE__ of sites. The vast majority, __WHAT PERCENTAGE__, of the human genome is made up of non-coding DNA. This is similar to other organisms with similarly large genomes (__CITE__). Much of the non-coding DNA is junk DNA. Indeed, over __WHAT PERCENTAGE__ of the human genome is comprised of repetitive DNA.

\subsubsection{Mouse genome}\label{sec:mouse_genome}

Mice are also diploid organisms, but have 19 pairs of autosomes, one pair of sex chromosomes and a mitochondrial chromosome. The mouse genome is slightly smaller than the human genome, at 2.8 billion base pairs long. Like the human genome, there is a mouse reference genome (__CITE__). It is based on several female mice from the _C57BL/6J_ strain, an important strain of mouse widely used in medical research.

Mouse strains used in medical research are highly inbred due to years of concerted mating programs. This reduced genetic variability, and the control that researchers have over it, make these mice a very powerful tool in identifying the biological cause of a diverse range of phenotypes.

\subsubsection{Arabidopsis Thaliana genome}\label{sec:human_genome}

Arabidopsis is a small flowering plant. The most widely studied species is _A. thaliana_, which is diploid and has one of the smallest genomes known in the plant kingdom (__CITE__). The genome of _A. thaliana_ is 135 million base pairs long and is packaged as five pairs of chromosomes (__CITE__). The small size of its genome, coupled with its small physical size, rapid development and its tolerance to both cross-pollination and self-pollination[^cross_and_self_pollination], make _A. thaliana_ an ideal organism for studying plant development.

[^cross_and_self_pollination]: The advantage in genetic studies of using a plant that is tolerant of self-pollination is that recessive mutations can be quickly made homozygous and thus expressed. Cross-pollination allows the creation of new strains with multiple mutations.

\subsection{Genetic variation}\label{sec:genetic_variation}

A key question in biology, perhaps _the_ key question, is what determines an organism's _phenotype_. An individual's phenotype is "the set of observable characteristics of an individual resulting from the interaction of its genotype with the environment" (__CITE: currently using google's definition__). Two simple examples of phenotypes are height and weight. Both have a genetic component, e.g., the offspring of tall parents are on average taller than those offspring of short parents, but there are also environmental components, such as the contribution of diet to weight. In medical research, a person's phenotype might be whether she is affected by a particular disease. It might also be some proxy, such as her blood pressure or the expression levels of particular genes.

Variation between individual's genomes, be it at single nucleotides[^snvs_snps] or across larger regions[^indels], is one important source of phenotypic variation. Importantly, genetically-driven phenotypic variation is frequently _heritable_, meaning that phenotypes can be passed on from one generation to the generation via the genome.

[^snvs_snps]: An individual nucleotide that is variable in the population is called a single nucletoide variant (_SNV_). A SNV that is frequently variable in the population, say at least $1\%$ of the population are variable at that site, is called a single nucleotide polymorphism (SNP). All SNPs are SNVs but the converse is not true.

[^indels]: __TODO: Define__

Environmental variation has clear influence on certain phenotypes. However, it often can be difficult to determine whether phenotypic variation is driven by genetic varation or environment variation, particularly in humans where genetically similar individuals typically grow up in similar environments.

The above discussion has been about phenotypic variation in a population. But there is also phenotypic variation within the individual. If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell) and a cone cell (a photoreceptor in the retina) all have identical genomes. In fact, all cells in an organism, excluding the gametes, have an identical genome yet play very different roles. This is due to different genes being active in different cells.

_Epigenetics_, which I describe in the next section, plays a role in determining both the between-individual phenotypic variation as well as the within-individual phenotypic variation.

\section{Epigenetics}\label{sec:epigenetics}

- [x] DONE

"Epigenetics" is a real Humpty Dumpty phrase: each author seems to believe that, "When I use a word, ... it means just what I choose it to mean - neither more nor less" [@carroll1897through]. I am sure to continue in this grand tradition.

Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" (Source: [http://en.wikipedia.org/wiki/Epigenetics](http://en.wikipedia.org/wiki/Epigenetics)). He meant epigenetics as the study of how _genotypes_ give rise to phenotypes during development [@waddington1957strategy].

A popular contemporary definition of epigenetics is attributed to the epigeneticist (__CHECK SOURCE AND QUOTE__) Art Riggs: "[epigenetics is] the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" (p1 @russo1996epigenetic). This is in line with the _epi_ prefix being derived from the Greek word for "upon, near to or in addition" (__INSERT OXFORD DICTIONARY REFERENCE__), hence the idea that epigenetics encodes information "on top of" the DNA sequence. However, it is quite different to Waddington's original definition.

[^mitotic_meiotic]: Mitotically heritable means heritable during cell division and meitotically heritable means heritable during sexual reproduction.

More recently, the definition of epigenetics has taken on a "more biochemical flavour" ([@Daxinger:2010bu]). It is used to desribe marks whose heritably is yet to be proven, such as histone modifications (__SOURCE__). Sir Adrian Bird, an esteemed British geneticist, attempts to unite these definitions in the following ([@Bird:2007em]):

> The following could be a unifying definition of epigenetic events: the structural adaptation of chromosomal regions so as to register, signal or perpetuate altered activity states. This definition is inclusive of chromosomal marks [e.g. histone marks], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify.

Regardless of which definition you subscribe to, an _epigenetic mark_ is the modification that causes this "epigenetic change". Actually, "causes" may be too strong a phrase as much of current epigenetics research is in identifying associations rather than causations and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.

The _epigenome_ of a cell is the set of epigenetics marks present on the cell's genome. In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual. Indeed, we can identify variation in a single epigenetic mark within cells of the same cell type from the same individual (__EXAMPLE__).

Methylation, acetylation, ubiquination and phosphorylation of histones, which alter the chromatin structure, are examples of the "more biochemical flavour" of recent epigenetics. But the classic epigenetic mark is DNA methylation, which I describe in the next section and whose variability I have studied in my thesis.

- Are histone modifications epigenetic? (http://www.nature.com/news/2008/081022/full/4551023a.html)

## DNA methylation

- [ ] DONE

DNA methylation is a chemical modification of DNA that can impart information on top of the DNA sequence. It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying]. It therefore fits into Art Riggs' definition of an epigenetic modification [@russo1996epigenetic]. So pervasive is DNA methylation that [@Lister:2009boa] dubbed it the "fifth base" of the DNA code.

[^dnam_copying]: In practice this copying is not as faithful as, say, the copying of DNA from the parent to the daughter strand. Furthermore, the faithfulness of this copying will be different in different conditions, such as in a healthy, well-differentiated liver cell compared to a cancerous liver cell. Nevertheless, the copying of DNA methylation is faithful enough for most biologists to consider it as a mitotically heritable mark, most of the time (__TODO__: Is this just true of 5mC in mammals or more generally?)

What is referred to as DNA methylation is typically methylation of the cytosine nucleotide. Cytosine methylation is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm], however, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably as is standard in the literature.

[^other_DNAm]: Two additional forms of DNA methylation, N6-Methyladenine (m6A) and N4-methylcytosine (m4C), are primarily found in prokaryotes (__SOURCE: http://en.wikipedia.org/wiki/DNA_methyltransferase#cite_note-21__) (or is it bacterial (__SOURCE: http://www.ncbi.nlm.nih.gov/pubmed/16479578__))

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__SOURCE__). In 1925, Johnson and Coghill isolated this methylated nucleotide as a product of hydrolysis of tuberculinic acid [@Johnson:1925js]. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings [@VISCHER:1949ty].

In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated 5-methylcytosine from nucleic acid prepared from cow thymus [@HOTCHKISS:1948va]. Using paper chromatography, Hodgkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.

[^tuberculosis]: _Mycobacterium tuberculosis_ was then known as _Tubercle bacillus_.

In the next section I focus on DNA methylation in mammals, in particular humans, but include brief descriptions of DNA methylation in other organisms.

### DNA methylation in mammals
DNA methylation is essential for normal development in mammals, such as human and mouse (__REFERENCE__). My thesis focuses on DNA methylation in humans (_Homo sapiens_) but much of the material in this section is true for other mammals.

The typical site of DNA methylation is at the C5 carbon position of a cytosine nucleotide. This modified cytosine is called _5-Methylcytosine_ or _5mC_. [__FIGURE__](http://en.wikipedia.org/wiki/File:5-Methylcytosine.svg) shows the structure of 5mC.

At the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated. For double-stranded DNA, (__TODO: Clarify; only CG methylation can be strand symmetric. CHG and CHH are not strand symmetric motifs__) methylation is usually symmetric across the two strands (_strand symmetric_) although _hemimethylation_ does occur. Within a diploid cell a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other. While the former that is more common, examples of the latter case, such as _allele-specific methylation_ and _genomic imprinting_, are of considerable biological interest (__SOURCE/EXAMPLES__).

In mammalian genomes, most cytosines are unmethylated except for those at the _CpG dinucleotide_, which is usually strand-symmetric. I explain the importance of CpG dinucleotides in the next section.

### CpG dinucleotides

A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__). The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-Methylcytosine (__REFERENCE__). I will use the abbreviation _mCpG_ to refer to a methylated CpG, _uCpG_ to refer to an unmethylated CpG and simply _CpG_ when the methylation state is irrelevant.

CpGs are underrepresented in the human genome. The GC-content of the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is approximately __WHAT %__. Under a random sampling model of nucleotides we would expect approximately __WHAT %__ of the human genome to be comprised of CpGs. In fact, only __WHAT %__ of the dinucleotides are CpGs. __FIGURE__ shows the frequency of each dinucleotide in the human reference genome (hg19)[^cpg_frequency].

[^cpg_frequency]: See __APPENDIX__ for the `R` code used to compute these values for the human reference genome (build hg19).

One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines ([@Scarano:1967ue]). Thus, over time, many mCpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).

There are many other evolutionary pressures on the distribution of nucleotides in a genome. One effect of this is that the distribution of CpGs is far from uniform. I discuss this distribution in the next section.

### CpG islands and other sandy metaphors

One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next. __FIGURE__ is a kernel density plot of the distances between neighbouring CpGs  and __FIGURE__ is the empirical cumulative distribution function of the distances between neighbouring CpGs for the human reference genome (hg19)[^cpg_distance].

[^cpg_distance]: See __APPENDIX__ for the `R` code used to generate these figures.

As can be seen from __FIGURE__, there are a subset of CpGs that are within __HOW MANY__ bases of the next CpG.

The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_). Again, there are evolutionary arguments for why these CGIs form (__REFERENCE__).

CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__REFERENCE__). [@GardinerGarden:1987we] gave the classical definition of a CGI:

> For the purpose of this survey, regions of DNA with a moving average of %G+C over 50 and Obs/Exp CpG over 0.6 have been classed as CpG-rich regions. CpG-rich regions over 200 bp in length are unlikely to have occurred by chance alone, so, as a working definition, have been labelled as CpG islands.

Other authors have since provided alternative definitions (__REFERENCES__ Adrian Bird, Rob Klose, Hao Wu, etc.)

[@Wu:2010do] developed a Hidden Markov Model to define CGIs. In my thesis I have used the predicted CGIs from this model. These predicted CGIs are available from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/).

The "sandy/beachy" metaphors have been continued (i.e. stretched to breaking point) with various authors defining CGI shores, CGI shelves, CpG canyons, CpG deserts and CpGs in the "open sea". CGI shores [@Irizarry:2008hg] are defined as regions within 2kb of CGIs. They regions have been demonstrated to have an increased variability of CpG methylation [@Irizarry:2008hg]. CGI shelves (__WHO__) are defined as __WHAT__ and __WHY ARE THEY USEFUL__. CpG deserts are __WHAT__, CpG canyons are __WHAT__ and the "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.

### Non-CpG methylation

In humans, cytosine methylation in most cell types is found almost exclusively at CpGs. There are, however, certain cell types with widespread non-CpG methylation. Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H stands for any nucleotide except G (__SOURCE IUPAC__). The rule-of-thumb is that non-CpG methylation is rare in somatic cells but common in pluripotent cells. Of course, there are exceptions to every rule, especially in biology.

To give a few examples, [@Lister:2009hy] found that in a somatic cell line (_fibroblasts_ or skin cells) $99.98\%$ of methylcytosines occured at CpGs whereas as in an embryonic stem cell line $24.5\%$ of methylcytosines occured in a non-CpG context. A subsequent paper from the same group ([@Lister:2011kg]) extended this result when they reported that, more generally, non-CpG methylation accounts for $20-30\%$ of methylcytosines in _pluripotent_ cell lines (which includes embryonic stem cell lines along with induced pluripotent stem cell lines, see __SECTION___). An exception to the rule is provided by [@Lister:2013et], where they found that neurons, a somatic and not a pluripotent type of cell, also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).

Non-CpG methylation is less well studied than CpG methylation. This is partly due to the design of popular assays for studying cytosine methylation; for example the popular Illumina 27k and 450k beadchips (see __SECTION__) only measure cytosine methylation at CpGs __CHECK__. However, recent technological advances (see __SECTION__) mean that cytosine methylation can be routinely assayed regardless of the sequence context. The biological role of non-CpG methylation is less well understood than that of CpG methylation (__SOURCE__).

### Modifications of a modification

Methylation is not the only chemical modification of cytosine nucleobase, although it is by far the most common. Further modifications, listed approximately from most frequent to least frequent, are 5-hydroxymethylcytosine (5hmC), 5-formylcytosine (5fC) and 5-carboxylcytosine (5caC) (__TODO:__ cite 10.1038/nrg3772).

The biological significance of these marks is still being determined, in part because the assays for studying these are still in development (__SOURCE__) and because their relatively scarcity means that experiments to detect these modifications are more difficult and expensive.

One genome-wide study of 5hmC found that less than $ 1\%$ of all assayed cytosines in mouse fetal cortex and adult cortex cells were hydroxymethylated [@Lister:2013et]. Most of the 5hmC was detected in the CpG context and, although the genome-wide level of 5hmC was low, the authors reported significant levels of 5hmC at particular cytosines in the genome.

[@Kriaucionis:2009bm] and [@Tahiliani:2009kl] discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC. This suggested a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.

### Writers, readers and erasers

A common analogy used in describing DNA methylation is that of "writers", "readers" and "erasers" [@Moore:2013in]. Writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes. These DNMTs are commonly split into two groups, namely the maintainence methyltransferase (thought to be DNMT1) and the _de novo_ methyltransferases (thought to be DNMT3a and DNMT3b)[^DNMT]. DNA methylation is not preserved by the DNA replication machinery and so it is the role of DNMT1 to restore the methylation pattern on the daughter strand of DNA following DNA replication. In contrast, DNMT3a and DNMT3b lay down new methylation marks and are particularly active during development when there are widespread changes in DNA methylation (__SOURCE__). DNMT1 and DNMT3b appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal (__SOURCE: http://labs.genetics.ucla.edu/fan/papers/npp2012112a.pdf__) whereas mouse knockouts for DNMT3a are runted but survive for ~4 weeks after birth (__SOURCE: http://labs.genetics.ucla.edu/fan/papers/npp2012112a.pdf__).

[^DNMT]: DNMT2, now known as TRDMT1, was once thought to be a DNA methyltransferase, however, __http://www.sciencemag.org/content/311/5759/395__ showed that it in fact methylates a small RNA and not DNA. Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity. Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__SOURCE__).

[^KO]: A knockout mouse for gene $X$ is a mouse that has been genetically engineered to remove or otherwise inactivate gene $X$. Mouse knockouts can be either heterozygous knockouts (one copy still of the gene is still present/active) or homozygous knockouts (both copies of gene absent/inactive).

The readers of DNA methylation recognise methylated DNA. These readers can recruit additional proteins to the site of the methylated cytosine to perform a variety of functions related to gene expression. For example, the _MBD_ (methyl-CpG-binding domain) group of proteins bind to DNA containing a methylated CpG which then suppress gene expression by preventing transcription factor binding at that site (__SOURCE__). Another group, the _UHRF_ (ubiquitin-like, containing PHD and RING finger domain) proteins, help DNMT1 methylate _hemimethylated_ DNA, which is DNA where one strand is methylated and the other is not (e.g. the parent strand compared to the daughter strand following DNA replication).

The removal or erasure of DNA methylation, _demethylation_, is characterised as _passive_ loss or _active_ removal. Passive loss occurs when the maintainence methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division. This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions. This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of precise targeted demethylation (__EXAMPLE__) nor the two stages of rapid global demethylation that occur during development [@Wu:2014gw].

Active demethylation is currently a hot topic in epigenetics research. Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation. These mechanisms were recently reviewed by [@Wu:2014gw], which I now briefly summarise:

1. The direct removal of the methyl group from 5mC is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a 5mC -> 5hmC -> 5fC -> 5caC reaction. The removal of 5hmC, 5fC or 5caC is biochemically "easier" than the removal of 5mC and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery (__FIGURE OF TET-CYCLE__).

One question raised by point (3) is whether 5hmC, 5fC and 5caC are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.

### Function of DNA methylation: "normal", cancer and development

### DNA methylation in other organisms and kingdoms

#### Insects {-}

* Honeybees
* Fruit fly? DNA methylation's existance in other organisms, such as the common fruit fly (_Drosophila Melanogaster_), is an area of contention although recent studies suggest it is present albeit at very small quantities and only in very limited contexts [@Takayama:2014jp].
* General

#### Plants {-}

* Arabadopsis
* Non-CG methylation

#### Fungi {-}

#### Bacteria {-}

#### Archaea {-}

## Assays for studying DNA methylation

- [ ] DONE

A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays [@Laird:2010iv]. Therefore, almost all assays of DNA methylation require one of the following _pre-treatments_ of the DNA:

1. _Enzyme digestion_
2. _Affinity enrichment_
3. _Sodium bisulfite conversion_

Each of these pre-treatments can then be combined with a second step of amplification or hybridisation. For example,

1. Gel-based analysis
2. Sanger sequencing
3. Microarray hybridisation
4. Massively parallel sequencing

An exception to this classification scheme are a new class of assay that seek to directly "read" whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA. For example, both [@Laszlo:2013kf, Schreiber:2013in] measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine was methylated while __PACBIO AND ONT DO WHAT?__.

Almost all assays of DNA methylation measure a "population average" from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay (see __FIGURE__). Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist but are still in development and not in widespread use (__SOURCE/EXAMPLES__).

The _resolution_ and _throughput_ are two key variables when choosing which assay to use for an experiment. The resolution of an assay is the scale on which DNA methylation can be measured[^resolution]. For example, a high resolution assay allows a researcher to quantify the level of DNA methylation at a single base whereas a low resolution assay might only allow for qualitative assessment (i.e. presence or absence) of DNA methylation at larger regions, such as CpG islands.

[^resolution]: Depending on the experiment and its aims, the resolution of an assay might instead be defined as the scale on which DNA methylation can be quantified or that which allows inference to address a specific hypothesis.

The throughput of an assay can be quantified in two ways. The first is per-sample throughput, which is how many measurements of DNA methylation are made per-sample[^throughput1]. This is typically what people mean when they describe an assay as being "high-throughput" or "low-throughput" and is the definition I use in the title of my thesis. Depending on your definition of "high", a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample. The second definition of throughput is per-dollar, that is, "how many samples can I process within my budget?". To avoid confusion, __I should refer to the latter definition as something else__.

[^throughput1]: This might reasonably be argued as being a definition of resolution.

The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-dollar throughput. Experiments that use an assay with high-resolution and high per-sample throughput generally have fewer samples than experiments using a lower resolution assay or an assay with lower per-sample throughput.

In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays. In particular, I will describe in detail the "gold standard" assay of DNA methylation, _whole-genome bisulfite-sequencing_ (WGBS), that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.

### Enzyme digestion assays
_Restriction endonucleases_ are an important technique in molecular biology. These enzymes can preferrentially "cut/cleave/digest" DNA at particular sequence motifs. The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_. The methylation of a position in the recognication motif can inhibit a restriction enzyme from cleaving the DNA. This can be used to design an assay to infer the methylation state of a DNA fragment.

For example, the recognition site of the restriction enzyme _HpaII_ is `CCGC`. However, _HpaII_ will only digest DNA when the second cytosine in the motif is unmethylated. The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome (__SOURCE__).

Another popular restriction enzyme for studying DNA methylation is _McrBC_, "an enzyme with the unusual and desirable property of cutting methylated DNA promiscuously (recognition sequence $R^{m} C(N)_{55– 103} R^{m} C$)" [@Irizarry:2008hg].

Assays based on restriction enzyme were some of the first developed for studying DNA methylation. They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches [@Laird:2010iv]. Restriction enzyme assays have a relatively low resolution (__REASONS/SOURCE__) but may also be used as a first-pass to enrich a sample for methylated or unmethylated regions which are then assayed using a higher resolution technique.

### Affinity enrichment assays
Affinity enrichment assays compare measurements between an "enriched" version and an "input" (control) version of the same sample to infer the presence or absence of DNA methylation. This is very similar to the idea behind chromatin immunoprecipitation followed by microarray hybridisation (ChIP-chip) or sequencing (ChIP-seq). Enrichment may be done using a methylation-specific antibody, such as __EXAMPLE__, or a __unmethylation-specific(?)__ antibody, such as __EXAMPLE__.

Some examples of affinity enrichment assays for DNA methylation are the microarray-based MeDIP, mDIP and mCIP (all based on methylated DNA immunoprecipitation) and their sequencing-based relatives, MeDIP-seq and mDIP-seq. These are all low resolution assays since they are based on the enrichment of regional differences between the enriched and input samples. Furthermore, the bioinformatic analysis of data from these assays is complicated by the varying CpG density along the genome, which leads to different enrichment affinities for different regions of the genome. However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation [@Laird:2010iv}].

### Sodium bisulfite conversion assays
In the 1980s, __Hayatsu OR Wang et al.__ discovered that when _denatured_ (__DEFINE?__) DNA is treated with sodium bisulfite (__CHEMICAL SYMBOL__), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines (__SOURCE__). The methylated cytosines are said to be "protected" from conversion to uracils. When the bisulfite treated is amplified by PCR the uracils are subsequently converted to thymines.

This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite, which are sometimes referred to as bisulfite-conversion assays (__FROMMER; CLARK__). These assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil (thymine) following the bisulfite treatment (__see FIGURE__).

#### Analysis platforms combined with bisulfite conversion {-}

Initial experiments based on the sodium bisulfite pre-treatment of DNA used Sanger sequencing of cloned PCR products, a very laborious task, which restricted experiments to studying a limited number of short segments of DNA (__HOW MANY & LONG?__). Although subsequent enhancements in the automation of Sanger sequencing improved the throughput of these assays, it was never going to be able to deliver a cost-effective, genome-scale assay of DNA methylation. The development of hybridisation microarrays provided an alternative way of analysing bisulfite-treated DNA that could provide cheap, genome-wide measurements of DNA methylation(__SOURCE__).

Microarrays contain thousands, even millions, of short _oligonucleotide_ probes. Each probe is is designed to hybridise to a particular DNA sequence and emits a signal, such as flurousence, that can be measured to infer the strength of the hybridisation. Therefore, an (idealised) way to analyse DNA methylation with a microarray is to hybridise bisulfite-converted DNA to a microarray that contains probes for both the methylated and unmethylated versions of all sequences of interest. The relative methylation of each sequence can be inferred from the relative intensities of the "methylated probe" to the "unmethylated probe". Such an idealised experiment brushes over many complications including [@Laird:2010iv]:

* The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a largely 3-base alphabet) leads to decreased hybridisation specificity
* Sequences containing multiple cytosines require multiple probe versions in order to assay all possible methylation patterns
* This approach requires the design of organism-specific microarrays

The Illumina 450k (__INSERT FULL NAME__) provides a modern implementation of the idea of bisulfite pre-treatment of DNA followed by microarray hybridisation for studying human samples. This array assays __HOW MANY__ CpGs (__IN WHAT TYPE OF REGIONS__) and uses two different probe designs (__WHY?__). There are still substantial bioinformatics challenges in analysing data from these assays (__CITE: SNPs in probes, normalisation, etc.__).

As mentioned in __SECTION__, genomics research was revolutionised by the development of cheap high-throughput sequencing technology and the study of DNA methylation was no exception. In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology [@Cokus:2008fc, Lister:2008bh]. [@Cokus:2008fc] termed their approach _BS-seq_ while [@Lister:2008bh] called theirs _methylC-seq_. These techniques, along with methods that these with targeting of particular regions of the genome, are described further in __SECTION__.

One final approach to mention is the use of mass spectrometry to analyse bisulfite-treated DNA, as implemented in Sequenom's EpiTYPER (__SOURCE__). This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and is frequently used to validate findings discovered using other platforms (__SOURCE__).

### DNA kinetics assays
__TODO__

### Pros and cons of bisulfite-conversion assays
Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation. Currently, multiple cells are required as input although the minimum number required is being lowered and efforts are well produce single-cell, whole-genome DNA methylation maps using bisulfite-treated DNA (__SOURCE__).

A key advantage of bisulfite-conversion assays over enzyme digest and affinity enrichment assays is that it provide base pair resolution of DNA methylation. In fact, single-molecule, base pair resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].

[^microarrays]: Microarray hybridisation can provide base pair resolution but not single molecule resolution. The signal from a microarray-based experiment a sample-wide average since, for a given locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.

A recently discovered disadvantage is that bisulfite-conversion assays are unable to distinguish between 5mC and 5hmC because both modifications similarly protect cytosine from deamination to uracil (__SEE [@Yu:2012ej] for references__). In effect, the detection of 5mC, and all subsequent inference, is confounded with that of 5hmC. For most experiments this isn't much of a problem - most cells have very low levels of 5hmC and so there is little confounding - however, in certain experiments this needs a more careful approach. To address this issue, [@Booth:2012fh] developed _oxidative bisulfite-sequencing_ (oxBS-seq) and [@Yu:2012ej] developed _Tet-assisted bisulfite sequencing_ (TAB-seq) to separate 5mC and 5hmC detection.

oxBS-seq specifically measures 5mC. The input DNA is oxidated by potassium perruthenate (KRuO$_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment. Only 5mC is protected from conversion during the bisulfite-treatment, which effectively means that only 5mC remains to be detected at the sequencing stage. The level of 5hmC can be estimated by performing traditional bisulfite-sequencing and then "subtracting" the oxBS-seq signal (5mC) from the bisulfite-sequencing signal (5mC + 5hmC).

TAB-seq takes the opposite approach to oxBS-seq by specifically measuring 5hmC. The input DNA is treated with a $\beta$-glucosyltransferase, which converts 5hmC to $\beta$-glucosyl-5-hydroxymethylcytosine (5gmC), followed by TET oxidation. Only 5gmC is protected from TET oxidation, which effectively means that only 5hmC remains to be detected at the sequencing stage. The level of 5mC can be estimated by performing traditional bisulfite-sequencing and then "subtracting" the TAB-seq signal (5hmC) from the bisulfite-sequencing signal (5mC + 5hmC).

A disadvantage of bisulfite-conversion assays is that they require knowledge of the underyling DNA sequence in order to infer the methylation states of cytosines. This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome. When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample (__SOURCE__). __FIGURE__ describes such an example.

The bisulfite-treatment of DNA can introduce biases and other problems. Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_. _PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence [@Warnecke:1997eh]. _Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC (__SOURCE__). Finally, bisulfite-treatment can lead to _DNA degredation_, which, as the name suggets, leads to a loss in DNA quality or quantity and can make it particularly hard to sequence long DNA molecules (__SOURCE: Anecdotal__).

## Whole-genome bisulfite-sequencing

- [ ] DONE

### Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq

* "RRBS’s ability to interrogate a locus is dependent on its MspI-cut-site (CCGG) density and consequently measures 10%–15% of the CpGs in the human genome (Bock et al. 2010; Harris et al. 2010)" \cite{Stevens:2013hv}
* See \cite{Stevens:2013hv} for MRE-seq details and a good summary of DNA methylation assays

## General TODOs
* Nucleotide, residue, (nucleo)base; what's the difference and which term should I use?
	* A nucleotide is composed of a nucleobase (also termed a nitrogenous base), a five-carbon sugar (either ribose or 2-deoxyribose), and one or more phosphate groups \[Coghill, Anne M.; Garson, Lorrin R., ed. (2006). The ACS style guide: effective communication of scientific information (3rd ed.). Washington, D.C.: American Chemical Society. p. 244. ISBN 978-0-8412-3999-9.)\]
* Should I give more detail on Rafa's CGI HMM?
* Define eukaryotes, prokaryotes and viruses
* Hyphenation or not of "bisulfite-sequencing"?
  - Yes, hyphenate.
* General term for methylC-seq, BS-seq, etc.
  - "whole-genome bisulfite-sequencing (WGBS)"
* Is it most accurate to say that restriction enzymes "cleave", "cut" or "digest" DNA?
* Define/describe DNA sequencing and the genomics revolution in the section describing DNA (or reference genome section?)
	* Define "library"
* Fix up reference formating
* Note that when I refer to a read I mean either a single-end read or a paired-end read. I will use to read_1 and read_2 if I want to refer a particular read that makes up a paired-end read.
* \cite{Zhang:2013uu} credit Laurent et al. (2010) with BS-seq and Cokus with WGBS.

### 2014-10-29 Rmarkdown/biology_background.Rmd
Update epigenetics section.

@@
 
-"Epigenetics" is a real Humpty Dumpty phrase: each author seems to believe that, "When I use a word, ...
-it means just what I choose it to mean - neither more nor less" [@carroll1897through].
+"Epigenetics" is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ...
+it means just what I choose it to mean - neither more nor less" \cite{carroll1897through}.
 I am sure to continue in this grand tradition.
 
-Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" (Source: [http://en.wikipedia.org/wiki/Epigenetics](http://en.wikipedia.org/wiki/Epigenetics)).
-He meant epigenetics as the study of how _genotypes_ give rise to phenotypes during development [@waddington1957strategy].
+Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" \cite{Waddington:2012bu}.
+Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \cite{Waddington:2012bu}.
 
-A popular contemporary definition of epigenetics is attributed to the epigeneticist (__CHECK SOURCE AND QUOTE__) Art Riggs: "[epigenetics is] the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" (p1 @russo1996epigenetic).
-This is in line with the _epi_ prefix being derived from the Greek word for "upon, near to or in addition" (__INSERT OXFORD DICTIONARY REFERENCE__), hence the idea that epigenetics encodes information "on top of" the DNA sequence.
+A popular contemporary definition of epigenetics is attributed to the epigeneticist Art Riggs that epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \cite[p1]{russo1996epigenetic}.
+The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence.
 However, it is quite different to Waddington's original definition.
@@
 
-More recently, the definition of epigenetics has taken on a "more biochemical flavour" ([@Daxinger:2010bu]).
-It is used to desribe marks whose heritably is yet to be proven, such as histone modifications (__SOURCE__).
-Sir Adrian Bird, an esteemed British geneticist, attempts to unite these definitions in the following ([@Bird:2007em]):
+More recently, the definition of epigenetics has taken on a "more biochemical flavour" \cite{Daxinger:2010bu} to include marks whose heritability is yet to be established.
+The heritability, or lack thereof, of histone modifications means that many epigeneticists do not consider these to be truly epigenetic \cite{Berger:2009ex}, and describing them as such is a sure-fire way to annoy a good percentage of your audience \cite{Ledford:2008ew}.
+
+Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following ([@Bird:2007em]):
 
@@
 
-The _epigenome_ of a cell is the set of epigenetics marks present on the cell's genome.
+The _epigenome_ of a cell is the set of epigenetic marks present on the cell's genome.
 In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual.
-Indeed, we can identify variation in a single epigenetic mark within cells of the same cell type from the same individual (__EXAMPLE__).
+Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual (__TODO: cite examples or upcoming chapters__).
+Epigenomics is the study of the epigenome, analogous to genomics being the study of the genome.
+However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is much evidence that genetic variation can drive epigenetic variation (__CITE__).
 
-Methylation, acetylation, ubiquination and phosphorylation of histones, which alter the chromatin structure, are examples of the "more biochemical flavour" of recent epigenetics.
-But the classic epigenetic mark is DNA methylation, which I describe in the next section and whose variability I have studied in my thesis.
-
-- Are histone modifications epigenetic? (http://www.nature.com/news/2008/081022/full/4551023a.html)
+One mark that most authors agree is an epigenetic mark is DNA methylation, which I describe in the next section and whose study is the focus of my thesis.
 

### 2014-10-29 Rmarkdown/biology_background.Rmd
End of day commit

@@
 
-Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following ([@Bird:2007em]):
+Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following \cite{Bird:2007em}:
 
@@
 
-## DNA methylation
+\section{DNA methylation}\label{sec:dna_methylation}
 
-- [ ] DONE
+- [x] DONE
 
@@
 It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying].
-It therefore fits into Art Riggs' definition of an epigenetic modification [@russo1996epigenetic].
-So pervasive is DNA methylation that [@Lister:2009boa] dubbed it the "fifth base" of the DNA code.
+It therefore fits into Art Riggs' afore-mentioned definition of an epigenetic modification \cite[p1]{russo1996epigenetic}.
+DNA methylation is found in bacteria, fungi, plants and animals.
 
@@
 
-What is referred to as DNA methylation is typically methylation of the cytosine nucleotide.
-Cytosine methylation is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm], however, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably as is standard in the literature.
+A major reason that DNA methylation is studied is that it is essential for normal development in mammals \cite{Li:1992bp}.
+It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression (__CITE__), silencing of transposable elements (__CITE__), X chromosome inactivation (__CITE__) and tumorigenesis (__CITE__).
+However, DNA methylation is not found in all organisms, for example, it not detectable in yeast, and so is not essential for life.
+Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation (__TODO: examples and citations__).
 
-[^other_DNAm]: Two additional forms of DNA methylation, N6-Methyladenine (m6A) and N4-methylcytosine (m4C), are primarily found in prokaryotes (__SOURCE: http://en.wikipedia.org/wiki/DNA_methyltransferase#cite_note-21__) (or is it bacterial (__SOURCE: http://www.ncbi.nlm.nih.gov/pubmed/16479578__))
+When people speak of DNA methylation they are generally referring to methylation of the cytosine nucleobase.
+Even more specifically, they are referring to the molecule _5-methylcytosine_.
+5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm].
+However, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably, as is standard in the literature.
+The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{@Lister:2009boa}.
+
+[^other_DNAm]: Two additional forms of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C).
+N6-methyladenine is a methylated form of adenine, which is found in mRNA \cite[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \cite{Ratel:2006dv}.
+N4-methylcytosine has also been detected in bacterial DNA \cite[e.g.,]{Ehrlich:1985bv, Ratel:2006dv}.
 
@@
 Ruppel, first identified a methylated nucleic acid in 1898.
-Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__SOURCE__).
-In 1925, Johnson and Coghill isolated this methylated nucleotide as a product of hydrolysis of tuberculinic acid [@Johnson:1925js].
-However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings [@VISCHER:1949ty].
-
-In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated 5-methylcytosine from nucleic acid prepared from cow thymus [@HOTCHKISS:1948va].
-Using paper chromatography, Hodgkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.
+Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__TODO: Can only find first page of this paper and it's in German.
+Need to verify__).
+In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}.
+However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{@VISCHER:1949ty}.
 
@@
 
-In the next section I focus on DNA methylation in mammals, in particular humans, but include brief descriptions of DNA methylation in other organisms.
+In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated _5-methylcytosine_ from nucleic acid prepared from cow thymus \cite{HOTCHKISS:1948va}.
+Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.
 
-### DNA methylation in mammals
+\subsection{DNA methylation in mammals}\label{sec:dna_methylation_in_mammals}
+
+- [x] Done
+
 DNA methylation is essential for normal development in mammals, such as human and mouse (__REFERENCE__).
-My thesis focuses on DNA methylation in humans (_Homo sapiens_) but much of the material in this section is true for other mammals.
+My thesis focuses on DNA methylation in humans but much of the material in this section is true for other mammals.
 
@@
 [__FIGURE__](http://en.wikipedia.org/wiki/File:5-Methylcytosine.svg) shows the structure of 5mC.
+It is common to describe a cytosine as being "methylated" or "unmethylated", however, care must be taken in defining these terms.
 
-At the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
-For double-stranded DNA, (__TODO: Clarify; only CG methylation can be strand symmetric.
-CHG and CHH are not strand symmetric motifs__) methylation is usually symmetric across the two strands (_strand symmetric_) although _hemimethylation_ does occur.
-Within a diploid cell a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other.
-While the former that is more common, examples of the latter case, such as _allele-specific methylation_ and _genomic imprinting_, are of considerable biological interest (__SOURCE/EXAMPLES__).
+At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
+Double-stranded DNA, at least at _palindromic_ methylation loci[^palindromic_sequence], is generally symmetrically methylated, i.e., the loci on each strand are both methylated or both unmethylated.
+However, _hemimethylation_, where the methylation loci on one strand is methylated and its partner on the opposite strand is unmethylated, can and does occur.
 
-In mammalian genomes, most cytosines are unmethylated except for those at the _CpG dinucleotide_, which is usually strand-symmetric. I explain the importance of CpG dinucleotides in the next section.
+[^palindromic_sequence]: "A palindromic sequence is a nucleic acid sequence (DNA or RNA) that is the same whether read 5' (five-prime) to 3' (three prime) on one strand or 5' to 3' on the complementary strand with which it forms a double helix" (__TODO: current source is wikipedia, find and properly cite__).
 
-### CpG dinucleotides
+Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other.
+While the former that is more common, examples of the latter case, such as _allele-specific methylation_ and _genomic imprinting_, are of considerable biological interest (__TODO: sources and examples__).
 
+In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_.
 A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__).
 The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation.
-Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-Methylcytosine (__REFERENCE__).
-I will use the abbreviation _mCpG_ to refer to a methylated CpG, _uCpG_ to refer to an unmethylated CpG and simply _CpG_ when the methylation state is irrelevant.
+A CpG is a palindromic sequence and is generally symmetrically methylated.
+Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-methylcytosine (__REFERENCE__).
+
+\subsection{CpG dinucleotides}\label{cpg_dinucleotides}
+
+- [x] Done
 
@@
 
-One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines ([@Scarano:1967ue]).
-Thus, over time, many mCpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).
+One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{@Scarano:1967ue}.
+Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).
 
@@
 One effect of this is that the distribution of CpGs is far from uniform.
-I discuss this distribution in the next section.
+In fact, CpGs tend to form clusters, which are termed CpG islands.
 
-### CpG islands and other sandy metaphors
+\subsection{CpG islands and other sandy metaphors}\label{sec:cpg_islands}
+
+- [ ] Done
 
@@
 
-The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_). Again, there are evolutionary arguments for why these CGIs form (__REFERENCE__).
+The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_).
+Again, there are evolutionary arguments for why these CGIs form (__TODO: sources__).
 
-CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__REFERENCE__).
-[@GardinerGarden:1987we] gave the classical definition of a CGI:
+CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__).
+The classical definition of a CpG island, given by \cite{GardinerGarden:1987we} and used by, for example, the UCSC genome browser, is a region of the genome where:
 
-> For the purpose of this survey, regions of DNA with a moving average of %G+C over 50 and Obs/Exp CpG over 0.6 have been classed as CpG-rich regions.
-CpG-rich regions over 200 bp in length are unlikely to have occurred by chance alone, so, as a working definition, have been labelled as CpG islands.
+1. The moving average of GC-content is greater than $50\%$, and
+2. The observed-to-expected ratio of CpGs is greater than $0.6$, and
+3. The region is longer than 200 bp.
 
-Other authors have since provided alternative definitions (__REFERENCES__ Adrian Bird, Rob Klose, Hao Wu, etc.)
+This definition was refined by \cite{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands.
+More recently, \cite{@Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
 
-[@Wu:2010do] developed a Hidden Markov Model to define CGIs.
-In my thesis I have used the predicted CGIs from this model.
-These predicted CGIs are available from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/).
+An alternative definition, and one that pre-dates the definition of \cite{GardinerGarden:1987we}, is based on the identification of unmethylated regions of the genome, which are typically CpG-dense.
+Such regions were previously called HpaII tiny fragment islands, or _HTF islands_, and named after the restriction enzyme used to identify them \cite{Cooper:1983ev, Bird:1985gj}.
 
-The "sandy/beachy" metaphors have been continued (i.e.
-stretched to breaking point) with various authors defining CGI shores, CGI shelves, CpG canyons, CpG deserts and CpGs in the "open sea".
-CGI shores [@Irizarry:2008hg] are defined as regions within 2kb of CGIs.
-They regions have been demonstrated to have an increased variability of CpG methylation [@Irizarry:2008hg].
-CGI shelves (__WHO__) are defined as __WHAT__ and __WHY ARE THEY USEFUL__.
-CpG deserts are __WHAT__, CpG canyons are __WHAT__ and the "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.
+The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpG canyons and CpGs in the "open sea":
 
-### Non-CpG methylation
+- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
+- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
+
+__UP TO HERE__
+
+- CpG canyons are __WHAT__ and the
+- "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.
+- methylation deserts: 100 Kbp windows with the lowest 1% methylation level in the sample of interest
+
+Another metaphor, based on methylation levels rather than CpG density, are methylation deserts.
+These regions are sample-dependent and were defined by \cite{Li:2012gg} as 100 Kbp windows with the lowest 1% methylation level in the given sample.
+
+\subsection{Non-CpG methylation}\label{sec:non_cpg_methylation}
+
+- [ ] Done
 
@@
 
-### Modifications of a modification
+\subsection{Modifications of a modification}\label{sec:modifications_of_a_modification}
+
+- [ ] Done
 
@@
 
-### Writers, readers and erasers
+\subsection{Writers, readers and erasers}\label{sec:writers_reads_and_erasers}
 
@@
 
-### Function of DNA methylation: "normal", cancer and development
+\subsection{Function of DNA methylation: "normal", cancer and development}\label{sec:function_of_dna_methylation}
 
-### DNA methylation in other organisms and kingdoms
+- [ ] Done
 
-#### Insects {-}
+\subsection{DNA methylation in other organisms}\label{sec:dna_methylation_in_other_organisms}
+
+- [ ] Done
+
+__TODO: Don't want too much detail here__
+
+\subsubsection{Insect}
+
+- [ ] Done
 
@@
 
-#### Plants {-}
+\subsubsection{Plants}
+
+- [ ] Done
 
@@
 
-#### Fungi {-}
+\subsubsection{Fungi}
 
-#### Bacteria {-}
+- [ ] Done
 
-#### Archaea {-}
+\subsubsection{Bacteria}
 
-## Assays for studying DNA methylation
+- [ ] Done
+
+\subsubsection{Archaea}
+
+- [ ] Done
+
+\section{Assays for studying DNA methylation}\label{sec:dna_methylation_assays}
 
@@
 
-### Enzyme digestion assays
+\subsection{Enzyme digestion assays}\label{sec:enzyme_digestion_assays}
+
+- [ ] Done
+
 _Restriction endonucleases_ are an important technique in molecular biology.
@@
 
-### Affinity enrichment assays
+\subsection{Affinity enrichment assays}\label{sec:affinity_enrichment_assays}
+
+- [ ] Done
+
 Affinity enrichment assays compare measurements between an "enriched" version and an "input" (control) version of the same sample to infer the presence or absence of DNA methylation.
@@
 
-### Sodium bisulfite conversion assays
+\subsection{Sodium bisulfite conversion assays}\label{sec:bs_assays}
+
+- [ ] Done
+
 In the 1980s, __Hayatsu OR Wang et al.__ discovered that when _denatured_ (__DEFINE?__) DNA is treated with sodium bisulfite (__CHEMICAL SYMBOL__), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines (__SOURCE__).
@@
 
-#### Analysis platforms combined with bisulfite conversion {-}
+\subsubsection{Analysis platforms combined with bisulfite conversion}\label{sec:bs_platforms}
+
+- [ ] Done
 
@@
 
-### DNA kinetics assays
-__TODO__
+\subsection{DNA kinetics assays}\label{sec:dna_kinetics_assays}
 
-### Pros and cons of bisulfite-conversion assays
+- [ ] Done
+
+\subsection{Pros and cons of bisulfite-conversion assays}\label{sec:pros_and_cons_of_bs_assays
+}
+
+- [ ] Done
+
 Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation.
@@
 
-## Whole-genome bisulfite-sequencing
+\section{Whole-genome bisulfite-sequencing}\label{sec:wgbs}
 
@@
 
-### Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq
+\subsection{Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq}\label{sec:targeted_bsseq}
+
+- [ ] Done
 

### 2014-10-29 Rmarkdown/biology_background.Rmd
CpG islands and other sandy metaphors

@@
 
-- [ ] Done
+- [x] Done
 
@@
 The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_).
-Again, there are evolutionary arguments for why these CGIs form (__TODO: sources__).
-
-CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__).
+CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__).
 The classical definition of a CpG island, given by \cite{GardinerGarden:1987we} and used by, for example, the UCSC genome browser, is a region of the genome where:
 
-1. The moving average of GC-content is greater than $50\%$, and
+1. The (moving) average of GC-content is greater than $50\%$, and
 2. The observed-to-expected ratio of CpGs is greater than $0.6$, and
@@
 
-The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpG canyons and CpGs in the "open sea":
+The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
+
+CpG island shores, shelves and the open sea are all defined with respect to CpG islands. Specifically,
 
@@
 - CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
+- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \cite{Sandoval:2011dj}.
 
-__UP TO HERE__
-
-- CpG canyons are __WHAT__ and the
-- "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.
-- methylation deserts: 100 Kbp windows with the lowest 1% methylation level in the sample of interest
-
-Another metaphor, based on methylation levels rather than CpG density, are methylation deserts.
-These regions are sample-dependent and were defined by \cite{Li:2012gg} as 100 Kbp windows with the lowest 1% methylation level in the given sample.
+Other metaphors, these based on methylation levels rather than CpG density, are methylation deserts \cite{Li:2012gg} and CpG canyons \cite{Jeong:2014bla}.
+Because these regions are defined with respect to methylation levels rather than DNA sequence, these are generally identified in a tissue-specific manner.
 

### 2014-10-29 Rmarkdown/biology_background.Rmd
Section on non-CpG methylation

@@
 There are, however, certain cell types with widespread non-CpG methylation.
-Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H stands for any nucleotide except G (__SOURCE IUPAC__).
+Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H stands for any nucleotide except G (__TODO: cite__).
 The rule-of-thumb is that non-CpG methylation is rare in somatic cells but common in pluripotent cells.
@@
 
-To give a few examples, [@Lister:2009hy] found that in a somatic cell line (_fibroblasts_ or skin cells) $99.98\%$ of methylcytosines occured at CpGs whereas as in an embryonic stem cell line $24.5\%$ of methylcytosines occured in a non-CpG context.
-A subsequent paper from the same group ([@Lister:2011kg]) extended this result when they reported that, more generally, non-CpG methylation accounts for $20-30\%$ of methylcytosines in _pluripotent_ cell lines (which includes embryonic stem cell lines along with induced pluripotent stem cell lines, see __SECTION___).
-An exception to the rule is provided by [@Lister:2013et], where they found that neurons, a somatic and not a pluripotent type of cell, also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
+To give a few examples, \cite{Lister:2009hy} found that in a _fibroblast_ cell line that $99.98\%$ of cytosines that displayed statistically significant evidence of methylation occured at CpGs.
+In contrast, in an embryonic stem cell line, they found that $24.5\%$ of cytosines that displayed statistically significant evidence of methylation occured in a non-CpG context.
+However, it should also be noted that these non-CpG loci that were methylated had on average a much lower level of methylation than their CpG counterparts.
 
-Non-CpG methylation is less well studied than CpG methylation.
-This is partly due to the design of popular assays for studying cytosine methylation; for example the popular Illumina 27k and 450k beadchips (see __SECTION__) only measure cytosine methylation at CpGs __CHECK__.
-However, recent technological advances (see __SECTION__) mean that cytosine methylation can be routinely assayed regardless of the sequence context.
-The biological role of non-CpG methylation is less well understood than that of CpG methylation (__SOURCE__).
+A subsequent paper from the same group extended this result when they reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines \cite{Lister:2011kg}.
+Pluripotent cells includes embryonic stem cells along with induced pluripotent stem cell lines.
+
+An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \cite{Lister:2013et}, where they found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
+
+Overall, non-CpG methylation is less well studied and less well understood than CpG methylation.
+This is partly due to sampling bias - commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips (__TODO: ref section on technology__), only measure CpG methylation.
+However, recent technological advances (__TODO: ref section on technology__) mean that cytosine methylation can be routinely assayed regardless of the sequence context.
 

### 2014-10-30 Rmarkdown/biology_background.Rmd
Completed modifications of a modification

@@
 
-- [ ] Done
+- [x] Done
 
-Methylation is not the only chemical modification of cytosine nucleobase, although it is by far the most common.
-Further modifications, listed approximately from most frequent to least frequent, are 5-hydroxymethylcytosine (5hmC), 5-formylcytosine (5fC) and 5-carboxylcytosine (5caC) (__TODO:__ cite 10.1038/nrg3772).
+Methylation is not the only chemical modification of cytosines, although it is by far the most common.
+These additional modifications may be thought of as 'modifications of a modification' since they extend 5-methylcytosine.
+Listed approximately from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \cite{Plongthongkum:2014jg}.
+The biological significance of these marks is still being determined, in part because the assays for studying these are still in development and because their relatively scarcity in the genome means that experiments to detect these modifications are more difficult and expensive.
 
-The biological significance of these marks is still being determined, in part because the assays for studying these are still in development (__SOURCE__) and because their relatively scarcity means that experiments to detect these modifications are more difficult and expensive.
+One genome-wide study of 5hmC found that less than $1\%$ of all assayed cytosines in mouse fetal cortex and adult cortex cells displayed any statistically significant evidence of hydroxymethylation \cite{Lister:2013et}.
+These cytosines appeared to be restricted to the CpG context and had only very low levels of 5hmC.
 
-One genome-wide study of 5hmC found that less than $ 1\%$ of all assayed cytosines in mouse fetal cortex and adult cortex cells were hydroxymethylated [@Lister:2013et].
-Most of the 5hmC was detected in the CpG context and, although the genome-wide level of 5hmC was low, the authors reported significant levels of 5hmC at particular cytosines in the genome.
-
-[@Kriaucionis:2009bm] and [@Tahiliani:2009kl] discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC.
+\cite{Kriaucionis:2009bm} and \cite{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC.
 This suggested a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.

### 2014-10-30 Rmarkdown/biology_background.Rmd
Writer, reads and erasers of DNAm. Bisulfite-conversion

@@
 
-- [x] Done
+- [x] DONE
 
@@
 
-A common analogy used in describing DNA methylation is that of "writers", "readers" and "erasers" [@Moore:2013in].
+- [x] DONE
+
+A common analogy used in describing DNA methylation uses 'writers', 'readers' and 'erasers' \cite{Moore:2013in}.
 Writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
@@
 DNA methylation is not preserved by the DNA replication machinery and so it is the role of DNMT1 to restore the methylation pattern on the daughter strand of DNA following DNA replication.
-In contrast, DNMT3a and DNMT3b lay down new methylation marks and are particularly active during development when there are widespread changes in DNA methylation (__SOURCE__).
-DNMT1 and DNMT3b appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal (__SOURCE: http://labs.genetics.ucla.edu/fan/papers/npp2012112a.pdf__) whereas mouse knockouts for DNMT3a are runted but survive for ~4 weeks after birth (__SOURCE: http://labs.genetics.ucla.edu/fan/papers/npp2012112a.pdf__).
+In contrast, DNMT3a and DNMT3b lay down new methylation marks and are particularly active during development when there are widespread changes in DNA methylation (__TODO: cite__).
+DNMT1 and DNMT3b appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \cite{Li:1992bp} whereas mouse knockouts for DNMT3a are runted but survive for ~4 weeks after birth \cite{Li:1992bp}.
 
-[^DNMT]: DNMT2, now known as TRDMT1, was once thought to be a DNA methyltransferase, however, __http://www.sciencemag.org/content/311/5759/395__ showed that it in fact methylates a small RNA and not DNA.
+[^DNMT]: DNMT2, now known as TRDMT1, was once thought to be a DNA methyltransferase, however, it was shown to in fact methylate a small RNA and not DNA \cite{Goll:2006kb}.
 Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity.
-Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__SOURCE__).
+Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__TODO: cite__).
 
@@
 These readers can recruit additional proteins to the site of the methylated cytosine to perform a variety of functions related to gene expression.
-For example, the _MBD_ (methyl-CpG-binding domain) group of proteins bind to DNA containing a methylated CpG which then suppress gene expression by preventing transcription factor binding at that site (__SOURCE__).
-Another group, the _UHRF_ (ubiquitin-like, containing PHD and RING finger domain) proteins, help DNMT1 methylate _hemimethylated_ DNA, which is DNA where one strand is methylated and the other is not (e.g.
-the parent strand compared to the daughter strand following DNA replication).
+For example, the methyl-CpG-binding domain (_MBD_) group of proteins bind to DNA containing a methylated CpG, which can then suppress gene expression by preventing transcription factor binding at that site \cite{Nan:1993gm}.
+Another group, the ubiquitin-like, containing PHD and RING finger domain (_UHRF_) proteins, help _DNMT1_ methylate hemimethylated DNA, such as the daughter strand created during DNA replication \cite{Sharif:2007ft, Bostick:2007jh}.
 
 The removal or erasure of DNA methylation, _demethylation_, is characterised as _passive_ loss or _active_ removal.
-Passive loss occurs when the maintainence methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division.
+Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division.
 This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions.
-This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of precise targeted demethylation (__EXAMPLE__) nor the two stages of rapid global demethylation that occur during development [@Wu:2014gw].
+This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of precise targeted demethylation (__TODO: example__) nor the two stages of rapid global demethylation that occur during development \cite{Wu:2014gw}.
 
@@
 Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation.
-These mechanisms were recently reviewed by [@Wu:2014gw], which I now briefly summarise:
+These mechanisms were recently reviewed by \cite{Wu:2014gw}, which I briefly summarise:
 
-1. The direct removal of the methyl group from 5mC is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
+1. The direct removal of the methyl group from _5mC_ is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
 2.
@@
 5mC oxidation-dependent active DNA demethylation.
-This follows from the observation that the TET enzymes can iteratively oxidate a 5mC -> 5hmC -> 5fC -> 5caC reaction.
-The removal of 5hmC, 5fC or 5caC is biochemically "easier" than the removal of 5mC and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery (__FIGURE OF TET-CYCLE__).
+This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ -> _5hmC_ -> _5fC_ -> _5caC_ reaction.
+The removal of _5hmC_, _5fC_ or _5caC_ is biochemically "easier" than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery (__TODO: Figure of TET cycle__).
 
-One question raised by point (3) is whether 5hmC, 5fC and 5caC are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
+One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
 
@@
 
-- [ ] Done
+- [ ] DONE
 
@@
 
-- [ ] DONE
+- [x] DONE
 
-A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays [@Laird:2010iv].
+A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as the _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays \cite{Laird:2010iv}.
 Therefore, almost all assays of DNA methylation require one of the following _pre-treatments_ of the DNA:
@@
 An exception to this classification scheme are a new class of assay that seek to directly "read" whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA.
-For example, both [@Laszlo:2013kf, Schreiber:2013in] measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine was methylated while __PACBIO AND ONT DO WHAT?__.
+For example, both \cite{Laszlo:2013kf} and \cite{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine was methylated.
+Similarly, __TODO: what does Pacbio and Oxford Nanopore do?__
 
@@
 Hundreds or thousands of cells are required in order to have sufficient material as input for the assay.
-Assays that require only a single cell as input do exist but are still in development and not in widespread use (__SOURCE/EXAMPLES__).
+Assays that require only a single cell as input do exist but are still in development and not in widespread use (__TODO: sources and examles__).
 
@@
 
-[^resolution]: Depending on the experiment and its aims, the resolution of an assay might instead be defined as the scale on which DNA methylation can be quantified or that which allows inference to address a specific hypothesis.
+[^resolution]: Depending on the experiment and its aims, the resolution of an assay might instead be defined as the scale on which DNA methylation can be quantified or the scale at which allows inference to address a specific hypothesis.
 
@@
 Depending on your definition of "high", a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample.
-The second definition of throughput is per-dollar, that is, "how many samples can I process within my budget?".
-To avoid confusion, __I should refer to the latter definition as something else__.
+The second definition of throughput is  related to cost, be it money or time, i.e., "how many samples can I process?".
 
@@
 
-The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-dollar throughput.
+The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-cost throughput.
 Experiments that use an assay with high-resolution and high per-sample throughput generally have fewer samples than experiments using a lower resolution assay or an assay with lower per-sample throughput.
@@
 
-- [ ] Done
+- [x] DONE
 
 _Restriction endonucleases_ are an important technique in molecular biology.
-These enzymes can preferrentially "cut/cleave/digest" DNA at particular sequence motifs.
+These enzymes can preferentially "cut/cleave/digest" DNA at particular sequence motifs.
 The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_.
-The methylation of a position in the recognication motif can inhibit a restriction enzyme from cleaving the DNA.
+The methylation of a position in the recognition motif can inhibit a restriction enzyme from cleaving the DNA.
 This can be used to design an assay to infer the methylation state of a DNA fragment.
@@
 However, _HpaII_ will only digest DNA when the second cytosine in the motif is unmethylated.
-The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome (__SOURCE__).
-
-Another popular restriction enzyme for studying DNA methylation is _McrBC_, "an enzyme with the unusual and desirable property of cutting methylated DNA promiscuously (recognition sequence $R^{m} C(N)_{55– 103} R^{m} C$)" [@Irizarry:2008hg].
+The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome \cite{Khulan:2006gx}.
 
 Assays based on restriction enzyme were some of the first developed for studying DNA methylation.
-They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches [@Laird:2010iv].
-Restriction enzyme assays have a relatively low resolution (__REASONS/SOURCE__) but may also be used as a first-pass to enrich a sample for methylated or unmethylated regions which are then assayed using a higher resolution technique.
+They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \cite{Laird:2010iv}.
+Restriction enzyme assays have a relatively low resolution (__TODO: reasons and source__) but may also be used as a first-pass to enrich a sample for methylated or unmethylated regions which are then assayed using a higher resolution technique.
 
@@
 
-- [ ] Done
+- [x] DONE
 
-Affinity enrichment assays compare measurements between an "enriched" version and an "input" (control) version of the same sample to infer the presence or absence of DNA methylation.
-This is very similar to the idea behind chromatin immunoprecipitation followed by microarray hybridisation (ChIP-chip) or sequencing (ChIP-seq).
-Enrichment may be done using a methylation-specific antibody, such as __EXAMPLE__, or a __unmethylation-specific(?)__ antibody, such as __EXAMPLE__.
+Affinity enrichment assays compare measurements between an 'enriched' version and an 'input' (control) version of the same sample to infer the presence or absence of DNA methylation.
+This may use antibody immunoprecipitation or methyl-binding proteins.
+Some examples of affinity enrichment assays for DNA methylation are the microarray-based MeDIP, mDIP and mCIP and their sequencing-based relatives, MeDIP-seq and mDIP-seq.
 
-Some examples of affinity enrichment assays for DNA methylation are the microarray-based MeDIP, mDIP and mCIP (all based on methylated DNA immunoprecipitation) and their sequencing-based relatives, MeDIP-seq and mDIP-seq.
 These are all low resolution assays since they are based on the enrichment of regional differences between the enriched and input samples.
 Furthermore, the bioinformatic analysis of data from these assays is complicated by the varying CpG density along the genome, which leads to different enrichment affinities for different regions of the genome.
-However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation [@Laird:2010iv}].
+However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation \cite{Laird:2010iv}.
 
@@
 
-- [ ] Done
+- [x] DONE
 
-In the 1980s, __Hayatsu OR Wang et al.__ discovered that when _denatured_ (__DEFINE?__) DNA is treated with sodium bisulfite (__CHEMICAL SYMBOL__), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines (__SOURCE__).
+In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \cite{Shapiro:1970im, Hayatsu:1970jv}.
 The methylated cytosines are said to be "protected" from conversion to uracils.
-When the bisulfite treated is amplified by PCR the uracils are subsequently converted to thymines.
+This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq}, which are referred to as bisulfite-conversion assays.
 
-This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite, which are sometimes referred to as bisulfite-conversion assays (__FROMMER; CLARK__).
-These assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil (thymine) following the bisulfite treatment (__see FIGURE__).
+When bisulfite-treated is amplified by PCR, the uracils are converted to thymines.
+Thus, these assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (__FIGURE__).
 
@@
 
-- [ ] Done
+__UP TO HERE__
+
+- [ ] DONE
 

### 2014-11-06 Rmarkdown/comethylation_review.Rmd
Completed modifications of a modification
Updated co-methylation review

@@
 
-## Chapter overview
+\section{Introduction}
 
@@
 This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation.
-In this chapter I review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}.
+In this chapter I first review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}.
 I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
@@
 Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.
-The lowest level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__.
+The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__.
 At a higher level is the _correlation of aggregate methylation levels_, such as the autocorrelation of $\beta$-values within a sample.
 
-Between-sample co-methylation is the relationship between methylation levels at two or more loci across a set of samples[^concordance].
+Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance].
 For example, \cite{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples.
-Furthermore, measure of within-sample co-methylation can themselves be compared between samples, as is discussed in Chapter \ref{chap:co-methylation}.
+Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.
 
@@
 For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci.
-This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays (__CITE__).
+This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.
 
@@
 The analysis in \cite{Lister:2009hy} takes a different approach.
-It too uses within-sample aggregate methylation levels, but rather than estimating correlations of the methylation levels, instead looks for patterns in the spacing of "methylcytosines" along the genome.
+While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the autnors instead looks for patterns in the spacing of "methylcytosines" along the genome.
 
-Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.
+In reviewing this previous work, I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.
 
-### \cite{Eckhardt:2006gh}
+\subsection{\cite{Eckhardt:2006gh}}\label{sec:eckhardt}
 
-\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples.
+\cite{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples.
 Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}.
@@
 
-In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-priting this figure__).
+In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-printing this figure__).
+Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].
 The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp.
-Therefore, it is not possible that the "percentage identical methylation" was obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.
-Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed.
-An email discussion with the senior author, Stephan Beck, was unable to clarify the definition.
+Therefore, it is impossible for the "percentage identical methylation" to have been obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.
 
-My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
+[^beck]: An email discussion with the senior author, Stephan Beck, was unable to clarify the definition or method.
+
+My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of identical methylation calls at different CpGs from (generally) different amplicons, that is, $Pr(Z_{h, i} = Z_{h', i'})$.
 I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.
 
-Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $z_{i, h}$ is sampled, that is, $Pr(z_{h, i} = 1) = p_{h}$.
-For two distinct CpGs sequenced on two (potentially) distinct amplicons, $z_{h, i}$ and  $z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$.
-Not only may the two CpGs be from distinct amplicons, but the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels.
-This most general setting is described by $z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(z_{h, i}, z_{h', i'}) = c$.
+Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$.
+For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$.
+Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels.
+This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$.
 In this setting we can compute the theoretical $PIM^{*}$ as follows:
@@
 \begin{align*}
-PIM^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
-        &= Pr(z_{h, i} = 0, z_{h', i'} = 0) + Pr(z_{h, i} = 1, z_{h', i'} = 1) \\
-        &= Pr(z_{h', i'} = 0 | z_{h, i} = 0) Pr(z_{h, i} = 0) +
-           Pr(z_{h', i'} = 1 | z_{h, i} = 1) Pr(z_{h, i} = 1) \\
+PIM^{*} &= Pr(Z_{h, i} = Z_{h', i'}) \\
+        &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
+        &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) +
+           Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
         &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
@@
 
-This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp.
-$z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp.
-$X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.
+This result is proved in __TODO: Cite and link to appendix__.
 
@@
 1.
-$PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
-2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
-3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.
+$PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
+3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
 
@@
 If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp.
-However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.
+However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible, even supposing $c > 0$.
 
@@
 
-### \cite{Cokus:2008fc}\label{sec:Cokus}
+\subsection{\cite{Cokus:2008fc}}\label{sec:Cokus}
 
@@
 
-### \cite{Chodavarapu:2010iqa}
+\subsection{\cite{Chodavarapu:2010iqa}}{sec:chodavarapu}
 
@@
 
-### \cite{Li:2010fb}
+\subsection{\cite{Li:2010fb}}\label{sec:li}
 
@@
 
-### \cite{Lyko:2010dr}
+\subsection{\cite{Lyko:2010dr}}\label{sec:lyko}
 
-In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation (Supplementary Figure 7).
-They report that the "correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
+In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation.
+They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
 This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report?
 Check my data any such increases__).
-__TODO: Is this the fastest decay of correlation in any species?__ However, I am concerned that the maximum autocorrelation is less $0.015$, which doesn't seem to make sense and is very close to zero.
-This may simply be due to a mislabelled or "normalised" value being plotted as the "correlation".
-(__TODO: I have emailed the senior author, Ryszard Maleszka, to try to resolve this concern__).
+__TODO: Is this the fastest decay of correlation in any species?__
 
-Additionally, \cite{Lyko:2010dr} identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__).
+\cite{Lyko:2010dr} also identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__).
 They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.
 
-Since the authors specifically describe their analysis as an "autocorrelation", I believe it uses all pairs of CpGs regardless of the number of intervening CpGs.
+The data supporting these results, Supplementary Figure 7 of the original publication, appears weak at first glance since the maximum value of the autocorrelation is less than $0.015$.
+I believe that this is due an error in the computation of the autocorrelation.
+Specifically, what I suspect has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been assigned a $\beta$-value of zero.
 
-### \cite{Lacey:2013iy}
+To demonstrate this, let $\beta^{*} = (\beta^{*}_{i'}, \ldots, \beta^{*}_{n'})$ be this "augmented" vector of $\beta$-values:
+
+$$
+\beta^{*}_{i'} = \left\{
+  \begin{array}{ll}
+    0 & \text{if } i' \text{ is not a CpG}\\
+    \beta_{i} & \text{if } i' \text{ is a CpG}
+  \end{array}
+\right.
+$$
+
+Using the `FF` chromosome 22 data, I computed and plotted (__FIGURE__) the correlations of all pairs of CpG $\beta$-values as well as the autocorrelation of the augmented vector $\beta^{*}$.
+From these plots we see that the autocorrelation of $\beta^{*}$ is a scaled version of the correlations of all pairs of CpG $\beta$-values.
+Therefore, while the autocorrelations are incorrectly computed, conclusions based on the relative height of peaks should be approximately correct.
+
+```{r}
+# TODO: The above described analysis and plots
+
+```
+
+\subsection{\cite{Lacey:2013iy}}\label{sec:lacey_2013}
 
 \cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2).
-The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
+The empirical and fitted variograms asymptote for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
 
@@
 
-### \cite{Lister:2009hy}
+\subsection{\cite{Lister:2009hy}}\label{sec:lister_2009}
 
@@
 An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots.
-This periodicity is very weak several contexts and not at all present for exonic CpGs.
-The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts since there are very few "methylcytosines" for these contexts in these human samples.
-Furthermore, in all nine contexts the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.
+This periodicity is very weak in several contexts and is absent for exonic CpGs.
+The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts, since there are very few "methylcytosines" for these contexts in these human samples.
+Furthermore, in all nine contexts, the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.
 
 Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites".
-However, I suspect that the results in these nine plots are driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines.
+However, I suspect that the results in these nine plots are in fact driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines.
 That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis.
@@
 For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \cite{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \cite{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.
-
-### \cite{Landan:2012kp}
-
-__TODO: Review co-methylation work in this paper__
 
@@
 
-Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads.
-That is, by studying the methylation patterns at m-tuples.
+Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, by studying the methylation patterns at m-tuples.
 Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.
@@
 
-Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__).
+Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lower throughput of these technologies, and the reliability with which they can identify methylcytosines, create new complications (__CITE__).
 An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).
@@
 There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.
-Here I review this previously published work.
 
-### \cite{Cokus:2008fc}
+\subsection{\cite{Cokus:2008fc}}\label{sec:cokus}
 
@@
 The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence.
-For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci.
+For the within-fragment co-methylation results (Supplementary Figure 14 of \cite{Cokus:2008fc}) they first identify all reads containing multiple CHH loci.
 Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated.
@@
 \cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
-This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages ().
+This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages.
 There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
 
-[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs.
+[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read, since a read may contain multiple pairs.
 The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
@@
 I consider these results to use a restricted definition of within-fragment "co-methylation".
-These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment.
+To explain, these results tell us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment, but only about the probability that both CHH loci are methylated.
 Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$.
+
 From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$.
@@
 
-### \cite{Lacey:2009by}
+\subsection{\cite{Lacey:2009by}}\label{sec:lacey_2009}
 
@@
 
-In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a miniscule fraction of the human genome.
-While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.
+In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a minuscule fraction of the human genome.
+While the methods and models are quite interesting, the data are so limited that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.
+
+\subsection{\cite{Landan:2012kp}}\label{sec:landan}
+
+\cite{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.
+The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs.
+They use ultradeep (median coverage of each CpG in the target $> 10,000 \times$) bisulfite-sequencing of a 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
+They not only use this information to estimate co-methylation but also to estimate the rate of _epipolymorphisms_ and _epialleles_.
+
+As part of their analysis they compute "methylation linkage diagrams".
+For each pair of CpGs in their targeted regions, they create a $2 \times 2$ contingency table.
+The general form of the table is shown below in Table \ref{tab:landan}.
+They then compute the Pearson correlation of each table[^cor_equivalence], which estimates the correlation of methylation events at that pair of CpGs.
+Importantly, these are within-read measurements from a single sample and so these are estimates of within-read, within-sample co-methylation.
+
+[^cor_equivalence]: Since this is a correlation of two binary variables, it is equivalent to computing the _phi_ coefficient of each table (__TODO: Find a better source than Wikipedia and my own numerical calculations.
+The algebraic proof isn't hard, just tedious.__).
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}ccccc@{}}
+\cmidrule(l){3-5}
+ &  & \multicolumn{3}{c}{Second CpG} \\ \cmidrule(l){3-5}
+ &  & m & u & Total \\ \cmidrule(l){2-5}
+\multirow{3}{*}{First CpG} & m & $n_{m, m}$ & $n_{m, u}$ &  \\
+ & u & $n_{u, m}$ & $n_{u, u}$ &  \\
+ & Total &  &  & n \\ \cmidrule(l){2-5}
+\end{tabular}
+\caption{Summary of read counts at a CpG 2-tuple.}
+\label{tab:landan}
+\end{table}
+
+```{r}
+# Equivalence of Pearson correlations and Phi coefficient for binary table.
+set.seed(333)
+z <- array(rpois(n = 4, lambda = 20), dim = c(2, 2),
+           dimnames = list(c('M', 'U'), c('M', 'U')))
+n11 <- z[1, 1]
+n10 <- z[1, 2]
+n01 <- z[2, 1]
+n00 <- z[2, 2]
+n <- sum(z)
+n1dot <- n11 + n10
+ndot1 <- n11 + n01
+x <- c(rep(1, n11), rep(1, n10), rep(0, n01), rep(0, n00))
+y <- c(rep(1, n11), rep(0, n10), rep(1, n01), rep(0, n00))
+plot(x, y)
+cor(x, y)
+m1 <- (n11 + n10) / n
+m2 <- (n11 + n01) / n
+v1 <- m1 * (1 - m1)
+v2 <- m2 * (1 - m2)
+(n11 / n - m1 * m2) / sqrt(v1 * v2)
+all.equal(cor(x, y), (n11 / n - m1 * m2) / sqrt(v1 * v2))
+all.equal(cor(x, y),
+          (n11 * n00 - n10 * n01) /
+            sqrt(((n1dot * (n - n1dot) * ndot1 * (n - ndot1)))))
+all.equal(GenomicRanges::phicoef(z), cor(x, y))
+```
+
+These correlations are then plotted like a linkage disequilibrium heatmap to identify blocks of CpGs with significant co-methylation. From the analysis of this small set of regions they conclude that:
+
+> "[the] correlation between the methylation states of pairs of CpGs was generally very low.
+This lack of correlation suggests that methylation dynamics are typically independent for different CpGs, making the methylation state of one CpG (whether high or low) uninformative on the methylation state of nearby CpGs."
+
+As we will see in \ref{chap:co-methylation}, this is contradicted by my analyses of co-methylation.
 
@@
 
-As is clear from the above, there are multiple definitions of within-sample co-methylation.
-An fundamentally different concept to within-sample co-methylation is between-sample co-methylation, which is the correlation of methylation levels between a single pair of loci across a set of samples rather than the correlation of pairs of loci within a single sample.
-(__CITE__) estimate various forms of between-sample co-methylation and are reviewed below.
+As is clear from the above, there are multiple definitions of within-sample co-methylation, each capturing variations on the idea of the correlation of methylation levels at pairs of loci within a single sample.
+An almost orthogonal concept is between-sample co-methylation, which is the correlation of methylation levels of a single pair of loci between a set of samples.
 
-* "vertical" correlations of \cite{Capra:2014uh}
-* MethLD of \cite{Liu:dy}
+Since it is a loci-specific, population-level correlation, between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci.
+Briefly, for two loci, $(i, i')$, the correlation of aggregate methylation levels is computed across samples using, for example, the Pearson correlation coefficient, $r = cor(\bm{\beta_{i}}, \bm{\beta_{i'}})$, where $\bm{\beta_{i}} = (\beta_{i, 1}, \ldots, \beta_{i, j}, \ldots, \beta_{i, n})$.
 
-__TODO: Comment on utility of within-sample vs. between-sample co-methylation__
+The concept of between-sample co-methylation leads to the attractive idea of identifying "tag-loci", whose methylation level is a good proxy for the methylation level across a larger region.
+These tag-loci could act like tag-SNPs used in genome-wide association studies, which would allow researchers to assay a small subset of loci in the methylome and capture much of the methylation variability.
+Unfortunately, unlike linkage disequilibrium, we cannot assume that between-sample co-methylation is identical across different tissues from the same individuals since DNA methylation, unlike DNA sequence, varies across tissues.
+Therefore, at best, we might hope to generate tissue-specific maps of between-sample co-methylation.
 
-### \cite{Liu:dy}
+In order to exploit between-sample co-methylation in this way, we also need to know the size of these "co-methylation blocks", that is, the length of the regions over which between-sample co-methylation is sufficiently strong.
+The longer these blocks are, the more sparsely we can afford to sample individual methylomes.
 
-The focus of this study on identify clusters of correlated CpGs and relating methylation changes at these clusters to SNP genotypes.
-The authors compute the correlation of pairs of $\beta$-values across samples, which is analogous to estimating the linkage disequilibrium between a pair of single nucleotide polymorphisms (SNPs) across a set of samples.
-This is different to computing autocorrelations of $\beta$-values, which are computed within, rather than between, samples.
-An autocorrelation analysis implicitly assumes stationarity along the genome whereas this analysis assumes stationarity across samples (__TODO: Discuss with Terry__).
+An example of between-sample co-methylation is given by \cite{Liu:dy}.
+The authors compute the correlation of pairs of $\beta$-values across samples to identify clusters of correlated CpGs and relate methylation changes at these clusters to nearby SNP genotypes.
+The resolution of their data is limited since it uses Illumina HumanMethylation450 array, but they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set".
+They also note that, "methylation codependency [co-methylation] itself varies across the genome", as is to be expected.
+No software is provided to implement the method.
 
-The discovery cohort contains 247 human samples typed on the Illumina HumanMethylation450 array.
-After normalisation and other pre-processing steps, they identify the top-$25\%$ most variably methylated CpG sites (vCpGs) in the 247 samples.
-They then compute Pearson correlations of methylation levels for all pairs of vCpGs across the samples.
+The `coMET` `R` package ([http://epigen.kcl.ac.uk/comet/](http://epigen.kcl.ac.uk/comet/)) can be used to estimate and plot between-sample co-methylation heatmaps from Illumina HumanMethylation450 array data.
+However, it is as yet unpublished and is still  in development at the time of writing.
 
-From this they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set".
-However, they also note that "methylation codependency (co-methylation) itself varies across the genome", as is to be expected.
-
-\section{Exploiting co-methylation in analyses of differential methylation}
+\section{Exploiting co-methylation in downstream analyses}
 \label{sec:exploiting_cometh}
 
-\cite{Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Sofer:2013bk} have each sought to exploit co-methylation, both the within-sample and between-sample forms, in developing more powerful and efficient ways to identify differential methylation.
+It has been difficult to properly account for co-methylation in downstream analyses due to the variety of definitions, and the varying rigour and reproducibility of the published analyses.
+Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data.
+Some authors have already sought to do this.
 
-### \cite{Hebestreit:2013ko}
+For example, the `bsseq` R/Bioconductor package for identifying differentially methylated regions from WGBS data \cite{Hansen:2011gu, Hansen:2012gr} cite \cite{Eckhardt:2006gh} as evidence that methylation levels at nearby CpGs are correlated.
+`bsseq` uses local binomial likelihood smoothing to exploit the spatial correlation of $\beta$-values.
+Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed estimates of the $\beta$-values are, on average, more accurate than the raw $\beta$-values.
+This is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error.
 
-The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}.
-This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.
+The R/Bioconductor package `BiSeq` \cite{Hebestreit:2013ko} also smooths the $\beta$-values, albeit slightly differently to `bsseq` since it is designed for identifying differentially methylated regions from RRBS data.
+The authors also cite \cite{Eckhardt:2006gh} as justification for smoothing the $\beta$-values.
+
+To summarise, both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis and are motivated by an analysis of within-sample co-methylation.
+Whenever using statistical smoothing, an important parameter is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother.
+The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide.
+`BiSeq` uses a much smaller default window size of 80 bp, a decision that is again driven by its focus on RRBS data which has a higher CpG density than WGBS data.
+Ideally, a data-driven window size would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice.
+A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs.
+the rest of the genome (see Chapter \ref{chap:co-methyaltion}).
+
+`BiSeq` also estimates and tries to account for the spatial correlation between test statistics of differential methylation at neighbouring loci.
+This is not the same as looking for correlations between the $\beta$-values, but is rather based on looking for correlations amongst the _differences_ in $\beta$-values, since these differences are what the test statistics of differential methylation are based upon.
+Specifically, `BiSeq` fits a semivariogram to the test statistics of differentially methylated CpGs.
+This correlations estimated using the semivariogram are in turn used in estimates of the standard deviations when identifying differentially methylated regions.
+
+More generally, methods such as `SLIM` \cite{Wang:2011cw} and `comb-p` \cite{Pedersen:2012vl} attempt to account for the correlation amongst test statistics in genomics data.
+They do this by analysing the distribution of P-values rather than the distribution of the test statistics themselves or the underlying data.
+This makes these methods quite general but more application-specific methods may be more powerful.
+`comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \cite{Pedersen:2012vl}.
+
+`Aclust` is notable for its use of between-sample co-methylation in identifying differentially methylated regions.
+Unlike other methods for detecting DMRs, which test for DMCs and then cluster these DMCs into DMRs, `Aclust` clusters loci prior to testing for differential methylation.
+That is, it only tests for DMRs and not DMCs.
+This has the advantage of reducing the dimensionality of the data.
+Clusters are formed using a modified form of agglomerative nested clustering, where the modification restricts clusters to comprise neighbouring loci.
+The distance metric used in the clustering is based on an estimate of the between-sample co-methylation for each pair of adjacent loci.
+
+\cite{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it.
+They study the dynamics of DNA methylation in differentiating cell lineages.
+For each locus there is a correlation between its methylation state in the precursor cell and its dependent cell types, which they call "vertical" correlations in contrast to the "horizontal" correlations of within-sample co-methylation.
+In essence, \cite{Capra:2014hx} are bringing well-established methods for the study of DNA sequence evolution to the study of DNA methylation dynamics.
+In this specialised setting, where "between-samples" means "between samples in a lineage", they show that a method based on "vertical" correlations is better at imputing the methylation level of CpGs than a method based on "horizontal" within-sample co-methylation.
+
+__TODO: Are methods using correlations of P-values or test-statistics based on within- or between-sample co-methylation?__
+
+## TODOs
+
+- An autocorrelation analysis implicitly assumes stationarity along the genome whereas between-sample co-methylation assumes stationarity across samples (__TODO: Discuss with Terry__).

### 2014-11-06 latex/phd_thesis.tex
Add comethylation chapter to thesis.
Discussion of Eckhardt et al.
Fixing up latex errors
Split co-methylation chapter.
Now split into “old stuff” (review of previous work) and “new stuff” (my work)
Updated Makefile to add new chapters
Updated Makefile
Removed unused macros

@@
 \documentclass[11pt,a4paper,oneside]{book}	%fleqn,
+\usepackage{etex}
 
-\usepackage{longtable,rotating,amsmath,epic, graphics}
+\usepackage{longtable,rotating, amsmath, epic, graphics, bm}
 
@@
 \usepackage{easymat}
+
+\usepackage{booktabs}
 
@@
 
-% convenient macros for lists
-
-\newcommand\BI{\begin{itemize}}
-
-\newcommand\EI{\end{itemize}}
-
-\newcommand\BE{\begin{enumerate}}
-
-\newcommand\EE{\end{enumerate}}
-
-\newcommand\BD{\begin{description}}
-
-\newcommand\ED{\end{description}}
-
-\newcommand\BF{\begin{figure}}
-
-\newcommand\EF{\end{figure}}
-
-\newcommand\BT{\begin{table}}
-
-\newcommand\ET{\end{table}}
-
-\newcommand\I\item
-
-\newcommand\h[1]{\Large\textbf{#1}\vspace{12pt}}
-
-\newcommand\N\newpage
-
-\newcommand\BC{\begin{quotation}\small\noindent}
-
-\newcommand\EC{\end{quotation}}
-
-\newcommand\beq{\begin{equation}}
-
-\newcommand\eeq{\end{equation}}
-
-\newcommand{\ba}{{\bf a}}
-
-\newcommand{\bb}{{\bf b}}
-
-\newcommand{\bc}{{\bf c}}
-
-\newcommand{\bd}{{\bf d}}
-
-\newcommand{\be}{{\bf e}}
-
-\newcommand{\bg}{{\bf g}}
-
-\newcommand{\bh}{{\bf h}}
-
-\newcommand{\bm}{{\bf m}}
-
-\newcommand{\bn}{{\bf n}}
-
-\newcommand{\bp}{{\bf p}}
-
-\newcommand{\bq}{{\bf q}}
-
-\newcommand{\br}{{\bf r}}
-
-\newcommand{\bs}{{\bf s}}
-
-\newcommand{\bt}{{\bf t}}
-
-\newcommand{\bu}{{\bf u}}
-
-\newcommand{\bv}{{\bf v}}
-
-\newcommand{\bw}{{\bf w}}
-
-\newcommand{\bx}{{\bf x}}
-
-\newcommand{\by}{{\bf y}}
-
-\newcommand{\bz}{{\bf z}}
-
-\newcommand{\bC}{{\bf C}}
-
-\newcommand{\bD}{{\bf D}}
-
-\newcommand{\bH}{{\bf H}}
-
-\newcommand{\bI}{{\bf I}}
-
-\newcommand{\bM}{{\bf M}}
-
-\newcommand{\bP}{{\bf P}}
-
-\newcommand{\bT}{{\bf T}}
-
-\newcommand{\bU}{{\bf U}}
-
-\newcommand{\bV}{{\bf V}}
-
-\newcommand{\bW}{{\bf W}}
-
-\newcommand{\bX}{{\bf X}}
-
-\newcommand{\bY}{{\bf Y}}
-
-\newcommand{\bZ}{{\bf Z}}
-
-\newcommand{\bone}{\mbox{\boldmath $1$}}
-
-\newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
-
-\newcommand{\bbeta}{\mbox{\boldmath $\beta$}}
-
-\newcommand{\bdelta}{\mbox{\boldmath $\delta$}}
-
-\newcommand{\bepsilon}{\mbox{\boldmath $\epsilon$}}
-
-\newcommand{\bgamma}{\mbox{\boldmath $\gamma$}}
-
-\newcommand{\blambda}{\mbox{\boldmath $\lambda$}}
-
-\newcommand{\bmu}{\mbox{\boldmath $\mu$}}
-
-\newcommand{\bomea}{\mbox{\boldmath $\omega$}}
-
-\newcommand{\bpi}{\mbox{\boldmath $\pi$}}
-
-\newcommand{\bphi}{\mbox{\boldmath $\phi$}}
-
-\newcommand{\brho}{\mbox{\boldmath $\rho$}}
-
-\newcommand{\bsigma}{\mbox{\boldmath $\sigma$}}
-
-\newcommand{\bSigma}{\mbox{\boldmath $\Sigma$}}
-
-\newcommand{\btau}{\mbox{\boldmath $\tau$}}
-
-\newcommand{\btheta}{\mbox{\boldmath $\theta$}}
-
-\newcommand{\bpsi}{\mbox{\boldmath $\psi$}}
-
-\newcommand{\bxi}{\mbox{\boldmath $\xi$}}
-
-\newcommand{\bzeta}{\mbox{\boldmath $\zeta$}}
-
-\newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
-
-\newcommand{\n}{\frac{1}{n}}
-
-\newcommand{\E}{{\rm I\!E}}
-
-\newcommand{\D}{{\rm I\!D}}
-
-\newcommand{\R}{{\rm I\!R}}
-
-\newcommand{\ls}{{\cal R}}
-
 \newcommand{\var}{{\rm var}}
@@
 
-\newcommand{\R}{{\textbf{\textsf{R}}}}}
+\def\eqd{\,{\buildrel d \over =}\,}
+
+%\newcommand{\R}{{\textbf{\textsf{R}}}}}
 
@@
 
-\BI
+\begin{itemize}
 
-\I[(i)] the thesis comprises only my original work towards
+\item [(i)] the thesis comprises only my original work towards
 
@@
 
-\I[(ii)] due acknowledgement has been made in
+\item [(ii)] due acknowledgement has been made in
 
@@
 
-\I[(iii)] the thesis is less than 100,000 words in length,
+\item [(iii)] the thesis is less than 100,000 words in length,
 
@@
 
-\EI
+\end{itemize}
 
@@
 
+\include{biology_background} % biology_background.tex
+
 \include{wgbs_analysis} % wgbs_analysis.tex
@@
 \include{a_statistical_model_of_methlyC-seq} % a_statistical_model_of_methlyC-seq.tex
+
+\include{avy} % avy.tex
+
+\include{comethylation_review} % comethylation_review.tex
+
+\include{comethylation} % comethylation.tex
+
+\include{methsim} % methim.tex
 

### 2014-11-06 Rmarkdown/appendix.Rmd
Updated appendix

@@
 \cite{Lindqvist:1978fm} wrote a brief note on Bernoulli trials with dependence. 
-Building on earlier work by \cite{Klotz:1973if}, Lindqvist parameterises the Bernoulli process $X_{1}, X_{2}, \ldots$ on $\{0, 1\}$ by the parameters $p = Pr(X_{i} = 1)$ and $c = cor(X_{i - 1}, X_{i})$ and shows that the transition matrix is given by
+Building on earlier work by \cite{Klotz:1973if}, Lindqvist parameterises the 
+Bernoulli process $X_{1}, X_{2}, \ldots$ on $\{0, 1\}$ by the parameters 
+$p = Pr(X_{i} = 1)$ and $c = cor(X_{i - 1}, X_{i})$ and shows that the 
+transition matrix is given by
 
@@
 
+provided that $max(1 - \frac{1}{p}, 1 - \frac{1}{1 - p}) \leq c \leq 1$.
+
 From this we can compute the joint distribution,
@@
 
-In particular, in the case $n = 2$ we can compute the probability that two dependent but identically distributed Bernoulli random variables are equal, $Pr(X_{1} = X_{2}) = Pr(X_{1} = 0, X_{2} = 0) + Pr(X_{1} = 1, X_{2} = 1)$.
+In particular, in the case $n = 2$ we can compute the probability that two 
+dependent but identically distributed Bernoulli random variables are equal.
 
-To extend the above result to the probability that two dependent but __non-identically__ distributed Bernoulli random variables are equal simply requires that we derive the appropriate transition matrix.
-
-Let $X_{i} \eqd Bernoulli(p_{i})$ for $i = 1, 2$. The transition matrix, $\Pi = \bigl(\begin{smallmatrix}
-Pr(X_{2} = x_{2} | X_{1} = x_{1}) \end{smallmatrix} \bigr)$, is given by
+To extend the above result to the probability that two dependent and 
+__non-identically__ distributed Bernoulli random variables are equal, simply 
+requires that we derive the appropriate transition matrix. Switching notation 
+to that used in Chapter \ref{chap:co-methylation_review}, let 
+$Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h', i'} \eqd Bernoulli(p_{h'})$. The 
+transition matrix, 
+$\Pi = \bigl(\begin{smallmatrix} 
+Pr(Z_{h', i'} = z_{h', i'} | Z_{h, i} = z_{h, i}) 
+\end{smallmatrix} \bigr)$, is given by
 
@@
  \begin{pmatrix}
-  (1 - p_{2}) + c p_{1} & p{2} - c p_{1} \\
-  \big [ (1 - p_{2}) + c p_{1} \big ] (1 - p_{1}) & \big [ p_{2} + c (1 - p_{1}) \big ] p_{1}
+  (1 - p_{h'}) + c p_{h} & p_{h'} - c p_{h} \\
+  \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) & \big [ p_{h'} + c (1 - p_{h}) \big ] p_{h}
  \end{pmatrix}
@@
 \begin{align*}
-Pr(X_{1} = X_{2}) &= Pr(X_{1} = 0, X_{2} = 0) + Pr(X_{1} = 1, X_{2} = 1) \\
-                  &= Pr(X_{2} = 0 | X_{1} = 0) Pr(X_{0} = 0) +
-                     Pr(X_{2} = 1 | X_{1} = 1) Pr(X_{0} = 1) \\
-                  &= \big [ (1 - p_{2}) + c p_{1} \big ] (1 - p_{1}) + 
-                     \big [ p_{2} + c ( 1 - p_{1}) \big ] p_{1} \\
-                  &= (1 - p_{1}) (1 - p_{2}) + c p_{1} \big [1 - p_{1}] +
-                     p_{1} p_{2} + c p_{1} (1 - p_{1}) \\
-                  &= (1 - p_{1}) (1 - p_{2}) + p_{1} p_{2} + 2 c p_{1} (1 - p_{1})
+Pr(Z_{h, i} = Z_{h', i'}) &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
+                  &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(X_{0} = 0) +
+                     Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(X_{0} = 1) \\
+                  &= \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) +
+                     \big [ p_{h'} + c ( 1 - p_{h}) \big ] p_{h} \\
+                  &= (1 - p_{h}) (1 - p_{h'}) + c p_{h} \big [1 - p_{h}] +
+                     p_{h} p_{h'} + c p_{h} (1 - p_{h}) \\
+                  &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
 \end{align*}
@@
 Firstly, I download the data and unzip it:
-```{r, cache = TRUE, message = FALSE}
+```{r, cache = FALSE, message = FALSE}
 library(R.utils)
@@
 library(dplyr)
+print(getwd())
+```
 
-tmp <- tempdir()
+```{r, cache = FALSE, message = FALSE}
+dl_dir <- 'downloads/'
 url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
-files <- paste0(url, 'wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep', 
+files <- paste0('wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep',
                 1:3, '.bed.gz')
-dests <- paste0(tmp,'/rep', 1:3, '.bed.gz')
+dests <- paste0(dl_dir, files)
 x <- mapply(function(f, d) {
-  download.file(f, d)
+  if (!file.exists(gsub(".gz", "", paste0(dl_dir, f)))) {
+    download.file(url = paste0(url, f), d)
     gunzip(filename = d, overwrite = TRUE)
-  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t')
+    }
+  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t', 
+        verbose = FALSE)
   }, files, dests, SIMPLIFY = FALSE)

### 2014-11-06 Rmarkdown/wgbs_analysis.Rmd
Converted webs analysis chapter from md to Rmd

---
title: "Analysis of bisulfite-sequencing data"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\section{Chapter overview}\label{sec:wgbs_analysis_chapter_overview}

This chapter outlines the steps in a bioinformatics analysis of a whole-genome bisulfite-sequencing dataset, with a focus on experiments studying _differential methylation_. Throughout this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol (__CHECK__). Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some additional tweaks.

There are four fundamental steps in the bioinformatics analysis:

1. Data quality control checks
2. Read mapping and post-processing of mapped reads
3. Methylation calling
4. Identifying differential methylation

__FIGURE__ provides a more detailed view of these four steps (__Figure should be a flow chart with the various sub-steps included, e.g. Inference = DMC, DMR, ASM, meth-SNP detection, etc.__)

Steps 1, 2 and 4 will be familiar to anyone who analyses high-thoughput sequencing data, but each requires a "twist" to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation but there are similarities to variant calling from DNA-seq.

In this chapter I describe steps 1-2 in broad terms. An excellent review of these fundamental steps, including comparisons of different software, is given by \citet{Krueger:2012ks}. My thesis has focused on steps 3 and 4 of this process. In this chapter I review previous work on steps 3-4 and defer my contributions on these topics to subsequent chapters. While this chapter focuses on analysing bisulfite-sequencing data, in particular whole-genome bisulfite-sequencing data, some parts __SECTION Cellular heterogeneity, Inference, OTHERS???__ will also be applicable to methylation microarrays (__CHECK AFTER I'VE WRITTEN IT__).

\section{Data quality control checks}\label{sec:data_qc}

The first step in any analysis of high-throughput sequencing data is to perform a quality control check of the data. Much of this is done visually by comparing summary graphs of the current sample(s) to previous 'good' samples. As such, much of data quality control checking relies on the judgement of the analyst.

The `FastQC` software (__CITE__) is a very useful tool for performing this first step. It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output is not so useful, or at least must be cautiously interpreted, for bisulfite-sequencing data. For example, `FastQC` will report a __warning/error__ if the GC-percentage, the percentage of read bases that are guanine or cytosine, is __less than what value__. Due to the bisulfite-treatment, bisulfite-sequencing data will naturally have a very low percentage of cytosine bases (__around what%__) and therefore a low GC-percentage. Therefore, a low GC-percentage is no cause for concern.

More important is the identification and removal of contaminating sequences. FastQC will screen a subset of the reads against a list of known, common contaminants. Examples of such contaminants include the sequencing adapters, PCR primers and __another example__.

Illumina adapter sequences are ligated to each DNA molecule in the library in order to perform Illumina sequencing. The sequencer can "read into" the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those created in WGBS libraries. This means that some reads are a chimera containing sequence of interest (from the sample) and junk (from the adapters). This junk needs to be removed for two reasons:

1. Reads containing adapter contamination will generally not map to the reference genome, meaning these reads go to waste.
2. If they do map, then this will result incorrect inferences - the garbage in, garbage out maxim.

Using a tool such as `Trim Galore!` (__CITE__) or `trimmomatic` (__CITE__), the reads can be _trimmed_ to remove these contaminants. Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.

\section{Read mapping and post-processing of mapped reads}\label{sec:mapping_and_post-processing}

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are now mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.

These bisulfite-sequencing read mappers take one of two approaches:

1. "Methylation-aware" mismatch penalties.
2. _In silico_ bisulfite-conversion of reads and reference genomes.

While "methylation-aware" mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads (__CITE FELIX__). This biases downstream inference and means that these mappers are generally less popular. I will instead focus on the _in silico_ bisulfite-conversion mappers, called as such because of a key step taken by these in the mapping process.

_In silico_ bisulfite-conversion mappers convert all cytosines to thymines (resp. guanines to adenines) of the forward (resp. reverse) strand from the reference genome. They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines. The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as Bowtie2 (__CITE__) or BWA (__CITE__). This process is illustrated in __FIGURE__.

[^2_reads]: Two versions are made because we don't know _a priori_ from which of the two strands the read originated.

Depending on the exact settings used, the mapper reports the "best" location of each read with respect to the two reference genomes. It reports the original sequence of the read in the output file so that the methylation status of each position can be inferred by comparing it to the corresponding reference sequence.

_In silico_ bisulfite-conversion mappers avoid the bias inherent in the "methylation-aware" mappers because all reads, regardless of methylation status, "look the same" to the mapper. However, they do suffer from a slight loss in mapping efficiency (__CITE FELIX__).

__TABLE__ lists some popular bisulfite-sequencing read mappers with the underlying mapping software listed in brackets:

* Bismark (Bowtie1 or Bowtie2, __CITE__)
* BWA-meth (BWA, __CITE__)
* BSMAP (SOAP, __CITE__)
* __OTHERS__

Each of these aligners can report the output in the standard `SAM` format (__CITE__), although each mapper does so in a slightly different way, which it difficult to develop downstream analysis tools.

\subsection{PCR duplicates}\label{sec:pcr_duplicates}

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that result in some molecules being over- or under-represented compared to their "true" frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiply-sequenced fragments are called _PCR duplicates_.

In bisulfite-sequencing data, PCR duplicates containing methylation locus $i$ result in $\beta_{i}$ being a biased estimate of $B_{i}$. This is because the sequenced reads do not accurately represent the true methylation levels of the sample.

Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify __possible__ PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]). Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` library (__CITE__), the `rmdup` function that is a part of the `SAMtools` library (__CITE__) and `SAMBLASTER` \citep{Faust:2014hf}. These software use mapped reads from a `SAM/BAM` file as input. There are subtle differences in how each method calls PCR duplicates. As an example, this is how `Picard`'s `MarkDuplicates` function works:

[^pcr_dup]: The distinction between possible PCR duplicates and true PCR duplicates is rarely made, probably because the phrase is so clunky. Possible PCR duplicates are almost always referred to as PCR duplicates with the implicit assumption that the reader is aware that these very likely include falst positive calls. Consistent with the literature, I will use the term PCR duplicates when I refer to reads identified as PCR duplicates by some software. I will use _true_ PCR duplicates when I need to distinguish the two concepts.

> Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair. When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment. You can thus think of it as determining "if all the bases from the read were aligned, where would the 5' most base have been aligned". It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair. "Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15 (Source: [http://sourceforge.net/apps/mediawiki/picard/index.php?title=Main_Page#Q:_How_does_MarkDuplicates_work.3F](http://sourceforge.net/apps/mediawiki/picard/index.php?title=Main_Page#Q:_How_does_MarkDuplicates_work.3F)).

Using these software will result in false positive calls because reads may satisfy this (or similar) criteria yet come from separate DNA fragments in the original sample. The false positive rate is a particular problem when a small "genome", which might be a targetted region of a genome such as exons or CGIs, is sequenced at high coverage. This can be thought of as an example of the pigeonhole princple, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).

We could make this mathematically more precise, but it doesn't give us a simple answer to the question, "should we remove possible PCR duplicates from bisulfite-sequencing data?". The unsatisfactory answer is, "it depends". A rough guide is that provided the average/median sequencing coverage of the "genome" is less than the fragment length[^pcr_fragment_length] then we expect few false positive calls.

[^pcr_fragment_length]: For single-end data, the "fragment length" in these calculations is the read length.

In practice, this means that for whole-genome sequencing data we can be fairly confident that possible PCR duplicates are _true_ PCR duplicates. However, for targetted sequencing, such as RRBS or amplicon sequencing, we are much less confident and may remove some of our signal if we remove possible PCR duplicates. Instead, for RRBS we might exclude regions with an "abnormally" high sequencing coverage \citep{Krueger:2012ks}. For amplicon sequencing we often can't afford to exclude possible PCR duplicates if, for example, the aim is to identify rare epialleles by very deep sequencing of a small region.

\subsection{M-bias}\label{sec:m-bias}

Ideally, the probability that a base is called methylated should be independent of the sequencing cycle. \citet{Hansen:2011gu} found that this is not the case and that in fact there is considerable bias towards the start (5') and end (3') of reads. They called this bias the _M-bias_.

M-bias can be identified by plotting the read-position methylation level ($rpml$), which is the proportion of reads that are methylated at each read-position, as a function of read-position (see __FIGURE__). These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for `read_1` and `read_2`. If there is no M-bias then this plot should be a horizontal line. A "bend" or "spike" in this plot is evidence of M-bias. Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same height for `read_1` and `read_2`, although we see this is often not quite the case (see __FIGURE__).

If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is "batch aware". For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with.

The strongest source of M-bias in Illumina WGBS data is at the 5' end of `read_2`, which sequences the 3' end of the DNA fragment. Because the DNA fragment is often shorter than the sum of the read lengths, the 3' end of the fragment often contains adapter sequence and other "junk" sequence. The adapter sequence may contain cytosine bases, which will be misinterpreted as evidence of methylation \citep{Krueger:2012ks}. Similarly, "fill-in cytosines" are used in the construction of RRBS libraries to repair the ends of DNA fragments after cleavage by MspI; these would also be misinterpreted as evidence of methylation \citep{Krueger:2012ks}. Another source of M-bias is incomplete or uneven bisulfite-conversion (__CITE__).

\subsubsection{Estimating M-bias}\label{sec:estimating_m-bias}

Estimating the M-bias and incorporating the results into the methylation calling can be done using two different strategies:

1. Compute the M-bias from the filtered data, then call methylation events incorporating both the filters and the M-bias. This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor`, `comethylation` and `Bis-SNP`.
2. Call methylation events from the filtered data but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls. This is the approach taken by `BSmooth`.

I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `comethylation`. In theory, the first pass over the `SAM/BAM` could be "skipped" by computing the M-bias simultaneously with sorting the `SAM/BAM` and marking PCR duplicates. For example, using (idealised) Unix pipes:

```
# 'align' is read mapping software
# 'mark_duplicates' is PCR duplicate marking software. It creates 'out.bam'.
# 'mbias' computes mbias. It produces 'mbias.out' and (passes through) 'out.bam'.
align ref.idx r1.fq r2.fq | mark_duplicates | mbias --filters
# call_meth is methylation calling software. It produces 'call_meth.out' based on 'mbias.out' and 'out.bam'
call_meth --filters mbias.out out.bam
```

A somewhat subtle point is that M-bias should only be estimated from reads that are actually going to be used for methylation calling. Suppose M-bias is highly correlated with some other quality metric, such as base quality, so that positions with M-bias also have low base quality[^bq_mbias]. If you already intend to ignore all positions with a base quality less than 20 in your methylation calling then it makes sense to also ignore these positions when computing M-bias; otherwise you will overestimate the effect of M-bias and unnecessarily exclude read positions in your methylation calling. Unfortunately, the software I use for estimating M-bias, `bismark_methylation_extractor`, does not allow the user to exclude certain reads or read-positions from its calculations.

[^bq_mbias]: This is very often the case since the 3' end of reads are typically of lower quality and also frequently suffer from M-bias.

```{r}
# TODO: Figure of M-bias computed from raw vs. filtered data
```

```{r}
# TODO: Figure showing several M-bias curves with different patterns, e.g. 5' and 3' junk, within-read spikes/troughs, etc.
```

\subsection{Pre-trimming reads destroys the one-to-one relationship between read-position and sequencing cycle}\label{sec:pre-trimming_reads}

Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read position. This causes a minor problem when computing M-bias because we no longer know whether the read position is identical to the sequencing cycle. Soft- or hard-clipping reads of their adapter sequence __during__ the alignment avoids this issue, because the clipping information (should be) preserved in the `CIGAR` string[^cigar].

[^cigar]: Many downstream tools, including the current version of `comethylation`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads. The current version of `comethylation` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip). __FIX THIS__. This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice. __TODO: Update what changes are made to `comethylation`. The `bwa-meth` and `LAST` alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark`, cannot soft-clip reads and so require that reads are pre-trimmed (__CITE `bwa-meth` pre-print or paper__).__

The M-bias plot is based on the read-position from the aligned data and not the sequencing cycle (which isn't directly available in the `SAM/BAM` file). If the reads have been pre-trimmed, then each read-position in the M-bias plot will therefore contain data from multiple sequencing cycles, which can amplify or mask M-bias signals.

For example, suppose we performed $100$ bp single-end sequencing and pre-trimmed the first 20 bp of $90\%$ of the reads. Then, read-position $80$ will comprised of $10\%$ sequencing cycle $80$ and $90\%$ sequencing cycle $100$. It is very likely the sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$. When using `bismark_methylation_extractor` this may necessitate that read-positions $80-100$ are ignored in downstream analyses, which would unnecessarily throw away $20\%$ of the data!

This problem cannot be avoided if reads are pre-trimmed because the trimming information is not preserved. \cite{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming]. However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be "sequencing cycle" rather than "read-position".

[^read_trimming]: Generally, this would also require that methylation calling is performed separately on each read length because most methylation callers are unable to deal with different M-bias profiles for different read lengths.

\subsection{Identifying M-bias}\label{sec:identifying_m-bias}

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e. read-positions that are "too far away" from the vast majority of read-positions. A minor problem with this approach is that the user may (unconsciously) use a different cutoff for each sample. For example, it is more difficult to determine an appropriate cutoff when there is a gradual slope in the M-bias plot than when there is a clear jump (__TODO: An example where there last few read-positions are more than `cutoff` from the median but due to a gradual slope vs. an example where a single position is more than `cutoff` from the median due to a spike__).

This motivated me to include a function, `trim.mbias` (__TODO: Check names used in final version. In particular, `trim.mbias` needs to be re-named to reflect that it identifies, rather than trims, M-bias.__), to identify M-bias in a more systematic way. `trim.mbias`, along with the related functions `read.mbias`[^read.mbias] and `plot.mbias`[^plot.mbias], are part of my `cometh` package.

[^read.mbias]: `read.mbias` reads in the `M-bias` files produced by `Bismark` and returns these as a `data.frame`.
[^plot.mbias]: `plot.mbias` creates a plot of M-bias stratified by the methylation type and, if the data are paired-end, whether the read is `read_1` or `read_2`.

`trim.mbias` identifies read-positions that display M-bias. This is done separately for each methylation type because the level of methylation varies widely between CG and non-CG methylation types. For paired-end sequencing experiments, it is also done separately for each of `read_1` and `read_2` because M-bias is very different for these two mates (see __FIGURE__) and also because `read_2` often has a slightly lower average level of methylation than does `read_1` (see __FIGURE and provide explanation__).

To do this, `trim.mbias` uses a simple normalisation of the read-position methylation levels ($rpml$). Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels (for example, $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in `read_1`).

`trim.mbias` identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$. While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types. I tend to use a value of `cutoff = 0.02` (__TODO: confirm__), meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $nrpml_{CpG} < 0.73$ or $nrpml_{CpG} > 0.77$ is flagged as showing evidence of M-bias. It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types. I tend to exclude read-positions that display evidence of M-bias in the methylation type I am working with, which is typically CpG methylation.

The results of `trim.bias` are returned as a `data.frame`:

```{r}
# TODO: Insert code snippet showing a call to `trim.mbias`
```

The results can also be visualised by specifying the `plot = raw` or `plot = normalised` arguments. `plot = raw` gives the same M-bias plot as `plot.mbias` except that outliers are now flagged. `plot = normalised` plots the $nrpml$ at each read position along with the a horizontal line indicating the chosen `cutoff`.

```{r}
# TODO: Insert code snippet showing plots returned by `trim.bias` with both `plot = raw` and `plot = normalised`
```

\subsubsection{What to do about M-bias?}\label{sec:what_to_do_about_m-bias}

Now that we've found read-positions with M-bias, what can we do about it? Typically, read-positions showing evidence of M-bias are excluded when calling methylation events. In fact, a slightly cruder procedure is often used whereby all read-positions preceding or following those with M-bias are excluded. Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method. Generally, this is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads.

However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. __FIGURE__ shows an example of such a spike that occurred at __WHICH__ read-position, likely due to a boost in the laser for sequencing cycles $101-150$ (__CHECK: If can't find a source then rephrase__). Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `comethylation` software is more flexible by allowing the user to exclude individual read-positions or ranges of read-positions via the `--ignore-read1-positions` and `--ignore-read2-positions` flags.

```{r, echo = FALSE}
# TODO: M-bias plot with an internal spike
```

Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read positions entirely. In several datasets I have analysed, more than __HOW MUCH__ of the entire sequencing data is lost to M-bias.

One idea is to weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute $\beta$-values, and use these in downstream inferences, without much loss in interpretation. An additional downside to this approach is an increased computational complexity and cost, and this is why I have not further pursued this idea.

\subsection{Other biases}\label{sec:other_biases}

There are several other sources of bias when making methylation calls. Some are what I call technical biases - those driven by the sequencing protocol and processing of the data - while others I call biological biases - those driven by measuring something other than what you thought you were measuring.

Apart from M-bias, technical biases include sequencing and alignment errors, which can both give rise to spurious or otherwise incorrect methylation calls in obvious ways.

Biological biases include cellular heterogeneity \citep{Jaffe:2014gy} and sequence variation at or nearby methylation loci, such as CpGs that are also SNPs or with SNPs nearby. I discuss the latter in the section __METHYLATION CALLING__.

\subsubsection{Cellular heterogeneity}\label{sec:cellular_heterogeneity}

Strictly speaking, cellular heterogeneity exists once you study DNA from more than a single cell, as shown in a recent papers using single-cell bisulfite-sequencing \citep{Guo:2013iha,Smallwood:2014kn}, however in this context it means when you study DNA from a sample that is made up of different types of cells.

For example, many studies of DNA methylation use whole blood as the  sample tissue due to the ease with which it can be obtained. However, whole blood contains a mixture of cell types, each of which has a distinct methylation profile. \citet{Houseman:2012km} exploited this fact to estimate the cell composition of whole blood samples, in particular the differences in white blood cell (leukocyte) proportions between different sub-populations (e.g. cases and controls).

In contrast to \citet{Houseman:2012km} where this cellular heterogeneity was useful, in studies exploring the relationship between DNA methylation changes and phenotypes, this cellular heterogeneity can seriously bias the results and must be properly accounted for in any analysis \citep{Jaffe:2014gy,Houseman:2014ds}. \citet{Jaffe:2014gy} provide evidence that several reported relationships between age and DNA methylation are likely due to changes in the cell composition of whole blood with age and not due to DNA methylation changes _per se_.

Methods to estimate the cellular heterogeneity bias and adjust for it are available (e.g. \citet{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}), although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data. This is not to say that these problems don't exist for sequencing data, merely that these have not been as well-explored.

\section{Methylation calling}\label{sec:methylation_calling}

Methylation calling is the process of calling each sequenced methylation locus as being either methylated or unmethylated[^no_call], as well as determining the _context_ or _type_ of each methylation event (i.e., CpG, CHG or CHH) based on the sequencing data and a reference DNA sequence. In principle, this is a simple process, however, this belies some complications, which I discuss in this section.

[^no_call]: A third possibility is making the call that the "methylation locus" is not in fact a methylation locus. For example, if the sequenced base at a cytosine in the reference sequence is an _A_ or a _G_ then this may be evidence that the position is not in fact a methylation locus.

Most bisulfite-sequencing alignment software either performs methylation calling during the alignment process, as done by `Bismark`[^bismark_methylation_extractor], or as a separate step after the alignment and post-processing of the `SAM/BAM` file. An example of the latter is `Bis-SNP` \citep{Liu:2012ge}, which performs methylation calling, and also variant genotyping, from bisulfite-sequencing data aligned with the user's choice of alignment software.

[^bismark_methylation_extractor]: `Bismark` also includes a program called `bismark_methylation_extractor`, which, as the name suggests, extracts the methylation calls from the `SAM/BAM` file. So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` or `comethylation` is required to make use of the methylation calls.

Both `Bismark` and `Bis-SNP` perform methylation calling of 1-tuples. In contrast, `comethylation` can perform methylation calling at m-tuples. `MethPat` (__CITE__) is also capable of performing methylation calling at m-tuples.

\subsubsection{A criticism of "identifying methylcytosines"}{sec:criticism_of_methylcytosines}

__TODO: Need a better title for this section__

\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced an idea "to identify the presence of a methylated cytosine" [supplementary material]. This idea was also used in subsequent, and similarly high-profile, publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013]. I believe this idea is an unrequired source of confusion, as I will explain.

[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.

It is necessary to use scare quotes to distinguish the term "methylcytosine", as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from methylcytosine, the standard definition meaning a 5-methylcytosine. I will first describe the idea and how it differs from methylation calling. Then I will describe how it is used to (1) estimate the number of methylcytosines in the sample, and (2) as a kind of screening procedure that determines whether each cytosines will be used in downstream analyses. Finally, I will explain why (1) is a bad estimator and why I think (2) is unnecessary or misguided.

A "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}. Thus, the idea of identifying "methylcytosines" is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample. To do so, Lister et al. compare each $\beta$-value to a (coverage-specific and sample-specific) cutoff that determines whether the cytosine is classified as a "methylcytosine" or not.

The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}. The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy} and \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}. Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done (__TODO: Discuss with Terry__). What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.

The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to "error", where "error" is a combination of the estimated sequencing error and the estimated bisulfite-converstion error. Mathematically, for each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated "error". Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".

[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.

I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure (__CITE__). This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A. thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).

The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}. It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.

Now to why I don't think this is a very useful procedure. Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample. These are two different things. If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$. A simple example makes this clear.

Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy. Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy. We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy. From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".

Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values. For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates. A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".

Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}. While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation. For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated. If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.

\subsection{Methylated or unmethylated?}\label{sec:methylated_or_unmethylated}

All bisulfite-sequencing assays use _reference-based_ methylation calling. That is, they require the specification of a DNA reference sequence that the aligned bisulfite-sequencing data are compared against to infer the methylation state of each sequenced locus. Care must be taken to correctly handle the orientation and strand of the alignment. __FIGURE__ gives several examples (__TODO: Figure explaining reference-based methylation calling for reads aligned to forward and reverse strands__).

When using reference-based methylation calling, the position of the methylation locus is with respect to the reference genome, since then all samples will use a common set of co-ordinates. Furthermore, unless the sample of the genome is fully known, methylation loci in insertions cannot be distinguished from genetic variation since there is no reference sequence to compare them against.

Each aligned read may be used for methylation calling, although each read should be subjected to some filtering to remove low-quality reads and low-quality bases. Standard filters include removing reads marked as possible PCR duplicates, reads with a mapping quality (`mapQ`) below a threshold and bases with a base quality below a threshold, and ignoring read-positions where there is evidence of M-bias.

The reference sequence is typically the reference genome used in the alignment step, which implicitly defines $\mathcal{I}_{j} :=\mathcal{I}_{rg}, j = 1, \ldots, N$, where $\mathcal{I}_{rg}$ is the set of methylation loci in the reference genome. Due to DNA variation between the sample and the reference genome, this assumption is obviously not true. A more refined approach incorporates genetic differences between the sample and the reference genome. This can be done in several ways:

* Whole-genome sequencing or genotyping of the sample
* Calling genetic variations directly from the bisulfite-sequencing data
* Excluding sites of known genetic variation

\subsubsection{Whole-genome sequencing or genotyping of the sample}\label{sec:wgs_or_genotyping}

The gold-standard is to perform whole-genome DNA sequencing of each sample. This data is then used to form a set of sample-specific methylation loci, $\mathcal{I}_{j}$. However, this approach is also very expensive due to the extra sequencing requirements. A cheaper alternative is to genotype the sample on a genome-wide SNP microarray. This will give very accurate, very cheap genotypes at a large number of loci ($500,000 - $5,000,000$), however, by definition it cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.

\subsubsection{Calling genetic variations directly from the bisulfite-sequencing data}\label{sec:wgbs_variant_calling}

The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci, $\mathcal{I}_{j} := \mathcal{I}_{j}^{Bis-SNP}$, at which to call methylation events. It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina (_Lister-style_) whole-genome bisulfite-sequencing protocol.

Certain variants, in particular (heterozygous) $C>T$ SNPs, are more difficult to accurately genotype than others. Unfortunately, $C>T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants. It is possible, although difficult, to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand - if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see __FIGURE__). Other substitutions are more readily called, and insertion and deletion events (InDels) may also be called. In case it isn't clear, `Bis-SNP` actually provides three important pieces of information that make it almost as good as having whole genome DNA sequencing data on the same sample:

1. Reference-specific methylation loci, i.e., cytosines in the reference genome that are mutated to non-cytosine nucleotides in the sample's genome.
2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine nucleotides in the reference genome.
3. Other genetic variants that may be used in additional analyses, such as identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).

Genotype calls made using `Bis-SNP` are less accurate than those from equivalent-depth whole-genome DNA sequencing because of the reduced complexity of bisulfite-converted DNA. However, we essentially get DNA variation "for free" by using `Bis-SNP`, which makes it my preferred approach for methylation calling at 1-tuples.

\subsubsection{Excluding sites of known genetic variation}\label{sec:known_genetic_variation}

The third approach is to call methylation loci using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population. For example, exclude all cytosines in the reference genome that are also SNPs in dbSNP (__CITE__). This approach can only exclude sites from $\mathcal{I}_{j}$, it cannot augment $\mathcal{I}_{j}$ by adding sample-specific methylation loci.

This is a conservative approach, as it will remove loci regardless of whether the sample has a genetic variant at that position or not, but it may be good enough. It also obviously requires a database of known variation for the organism being studied, which is the case for commonly studied organisms such as humans and mice. Finally, post-hoc filtering of known genetic variants is arguably a bare minimum requirement for methylation calling in order to prevent spurious methylation calls contaminating the dataset.

To remove those reference-specific methylation loci that are __not__ found in databases of known genetic variation, we might identify loci in the sample that display a large number of non-C/T bases (resp. non-G/A bases) at a C (resp. G) on the forward (resp. reverse) strand of the reference genome.

\subsection{Determining the context or methylation type}\label{sec:methylation_context}

In addition to determining whether a cytosine is methylated or unmethylated, we also want to determine the _context_ of the cytosine, also known as the _methylation type_. That is, we want to determine whether the cytosine is a CG, CHG or a CHH.

This is done by examining the two nucleotides upstream of the cytosine. It can be done based on the reference sequence, as is done in `Bismark` and `comethylation`, or from the reads themselves. The obvious difficulty with using the reads themselves is if the cytosine occurs at the last or second last position of the read, in which case the context may not be unambiguously determined from the the read alone. Instead, the context may be refined by initially using the reference genome context and then correcting for any sample-specific genetic variants in the two downstream bases.

A further complication occurs when there is a genetic variant in the two nucleotides upstream of the methylation locus. We would like to use the two upstream nucleotides from the sample to infer the sample-specific methylation context, either inferred from each read separately or from a variant calling procedure such as `Bis-SNP`. However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome.

\section{Identifying differential methylation}\label{sec:identifying_differential_methylation}

Identifying differential methylation means to identify sites (differentially methylated cytosines or _DMC_s) or regions (differentially methylated regions or _DMR_s) in the genome that have different levels of methylation between two or more conditions. Much of the methodology for identifying differential methylation has focused on two-group experiments and a restricted definition: differential methylation is the difference in __average__ methylation between __two__ groups. This is analogous to identifying differential gene expression, which focuses on identifying genes that have different __average__ expression levels between two groups.

While bisulfite-sequencing experiments with multiple groups have been performed and analysed (e.g. \citet{Lister:2009hy, Hansen:2011gu, Hansen:2013eo}), these have typically been done using a series of pair-wise comparisons or by comparing each sample in turn against some 'baseline' sample. This is because the analysis of a pair-wise comparison is relatively simple and because it is also generally easier to interpret a two-group comparison than a multi-group comparison. However, as the number of groups increases it becomes more difficult to interpret multiple pair-wise comparisons.

Like any experimental questions, good experimental design is essential in order to reliably infer differential methylation. Identifying DMCs is relatively straightforward statistical problem since it amounts to the well-worn problem of testing the difference of means. However, identifying DMRs, which may or may not build upon an initial scan for DMCs, is a far more challenging statistical problem owing the high spatial dependence of methylation at neighbouring cytosines and the uneven spacing of cytosines throughout the genome.

\subsection{Experimental design}\label{sec:experimental_design}

In any experiment of differential methylation we want the DMCs and DMRs to be _biological_ and _replicable_; it's no good if all the differences are simply due to technical artefacts or random fluctuations. Key to ensuring biological relevance good experimental design, such as the use of _replicates_ in each experimental group. A distinction is often made between _technical replicates_ and _biological replicates_. Roughly, biological replicates are experimental units that all underwent the same treatment and are used to estimate the within-group variability of the treatment. Technical replicates are repeated measurements of the same experimental unit, perhaps with slight variations in the sample preparation, and are used to estimate the variability of the sample preparation and measurement process.

As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one. For example, \cite{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine). I contend that these are more like technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].

[^cell_passaging]: In the case of IMR90 cell line, the first replicate, IMR90_r1, underwent 4 cell passages and the second replicate, IMR90_r2, underwent 5 cell passages. In the caase of the H1 cell line, the first replicate, H1_r1, underwent 25 passages in the first media and 9 passages in the second media, and the second replicate, H1_r2, underwent 27 passages in the first media and 5 passages in the second media.

Moreover, early experiments with whole-genome bisulfite-sequencing frequently had no replicates of any kind, as is the case of the H9 and IMR90-iPSC cell lines in \cite{Lister:2009hy}, and once these replicates were judged to be broadly consistent these data were pooled prior to analysis, thus ignoring all within-group variability.

Ideally, variability between technical replicates should be orthogonal to the biological question, but this rarely occurs. Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.

\subsection{DMCs}\label{sec:dmcs}

There are reports of differential methylation at individual CpGs resulting in a phenotypic difference (__CITE EXAMPLES; Emma Whitelaw has some I think__). Nonetheless, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from WGBS experiments with sample sizes you can count on one or two hands and average sequencing depth of $10-30\times$.

Identifying DMCs boils down to identifying differences in means. As such, it can be readily framed as a "stand-alone" test or cast into a regression framework. There is an enormous body of statistical literature on testing for differences in means (__CITE examples__). Most methods to detect DMCs from WGBS data are parametric and frequently use the binomial model described in __SECTION__.

Regardless of the test used, all analyses must pay a large multiple-hypothesis testing penalty. Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data. but the number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments (e.g. approximately 1 million tests in genome-wide association studies and 20,000 tests in studies of differential gene expression).

However, when identifying DMCs, the _effective_ number of tests is fewer, perhaps very much so, due to the high degree of dependence of methylation at neighbouring methylation loci, which in turn ensures a high degree of dependence amongst the tests.

__TODO: What techniques can be used to do FDR, multiple hypothesis testing? Are standard tools applicable in such high-dependence settings?__

\subsubsection{One-group tests}\label{sec:one_group_dmcs}

The one-group WGBS experiment is uncommon and the $A^{vy}$ experiment (__SECTION__) is the only example that I know of. In a one-group experiment a DMC is one where at least one of the samples has a different methylation level to the group average. Mathematically, we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$, where $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$). This can be tested in a number of ways as I will demonstrate.

The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown below (__TODO: Prettify table with `booktabs`__)

\begin{table}[h]
\centering
\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
\label{my-label}
\begin{tabular}{c c c}
\hline
             & m          & u          \\ \hline
$Sample_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
$\vdots$     & $\vdots$   & $\vdots$   \\
$Sample_{n}$ & $m_{i, n}$ & $u_{i, n}$ \\
\end{tabular}
\end{table}

Such contingency tables can be analysed in many different ways. Here I apply a few methods to highlight their similarities and differences. Firstly, here are some example data for a single CpG:

```{r}
x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
names(dimnames(x)) <- c("sample", "meth_state")
x
```

We see that while `sample_1, ..., sample_4` are nearly completely methylated at this CpG, `sample_5` has approximately $50\%$ methylation, which may be indicative of allele-specific methylation.

In my analysis of the $A^{vy}$ experiment I used the `prop.test` function in R. As documented in the help page for `prop.test`:

> `prop.test` can be used for testing the null that the proportions (probabilities of success) in several groups are the same

in R to test the equality of the proportions ($\beta$-values) across the five mice.

```{r}
prop.test(x)
```

As expected, this test returns a very small P-value, `r print(format.pval(prop.test(x)$p.value), quote = FALSE)`.

What might not be clear from its name is that `prop.test(x)` is just performing Pearson's $\chi^{2}$-test of the null hypothesis that the joint distribution of the cell counts in the $2 \times n$ contingency table is the product of the row and column marginals. This can be seen by examining the source code or by comparing the output to `chisq.test(x)`: (__TODO: Check by hand__)

```{r}
chisq.test(x)
```

Instead of using the "stand-alone" tests, `prop.test` and `chisq.test`, we could explicitly frame this as a log-linear model and test the goodness-of-fit of the "no interaction" model using either the likelihood ratio statistic (`lrt`) or the Pearson's $\chi^{2}$ test statistic (`pearson`). Formally, the regression model is (__TODO: Write down the regression model__):

This model can be fit using the `loglin` function in R or using the arguably more user-friendly `loglm` (from the `MASS` package) or `glm` functions:

```{r}
# Using loglin
loglin(x, margin = list(1, 2))
# Using loglm
library(MASS)
loglm(formula =  ~ sample + meth_state, data = x)
# Using glm (1)
glm(x ~ 1, family = binomial)
# Using glm (2)
glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2])
# TODO: Write sentence summarising output.
```

A key advantage of using regression is the option to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects. __TODO: How to do this using `loglin`?__

More complex regression analyses are possible, such as empirical Bayes or fully Bayesian models. I describe these in more detail in discussing two-group and multi-group experiments, but they may also be applied to one-group experiments.

A proper software implementation needs to take care of issues such as incomplete tables and non-convergence of the iterative proportional scaling algorithm used by `loglin`. Speed is also important when performing this many tests and the regression based approaches are much slower, at least when naively applied, due to additional checks they make of the data:

```{r}
# Basic benchmarking
library(microbenchmark)
print(microbenchmark(prop.test(x), chisq.test(x), loglin(x, margin = list("sample", "meth_state")), loglm(formula = ~ sample + meth_state, data = x), glm(x ~ 1, family = binomial), glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2]), glm.fit(x = matrix(, nrow(x), 0L), y = x, family = binomial()), times = 100), order = "median")
```

\subsubsection{Two-group experiments}\label{sec:two_group_dmcs}

In a two group experiment, for each methylation locus we are testing the hypothesis that the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, is different to that in the second group, $\bar{\beta_{i}^{(2)}}$.

If there are no replicates in either group[^2x2] then we have a $2 \times 2$ contingency table:

[^2x2]: I'd argue that this is a two-sample experiment rather than a two-group experiment since each group has $n = 1$.

(__TODO: Prettify table with `booktabs`__)

\begin{table}[h]
\centering
\caption{$2 \times 2$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a two-group experiment with no replicates in either group.}
\label{my-label}
\begin{tabular}{c c c}
\hline
             & m          & u          \\ \hline
$Group_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
$Group_{2}$ & $m_{i, 2}$ & $u_{i, 2}$ \\
\end{tabular}
\end{table}

This $2 \times 2$ table is then tested for independence of the rows and columns. As the sample size for each table is often small, Fisher's exact test is generally preferable since, unlike Pearson's $\chi^{2}$ test, the P-values are _exact_, meaning they do not rely on large-sample approximations (__CITE: Agresti__). Fisher's exact test is a _conditional_ test because the null distribution (a hypergeometric distribution) is derived by conditioning on both the marginal totals. __CITE: Agresti__ discusses the arguments for and against conditional and unconditional tests of independence in the $2 \times 2$ table.

The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation. Unfortunately, some authors have used Fisher's exact test to analyse experiments including replicates (__TODO: Highlight papers that have inappropriately used Fisher's exact test; does Lister 2013 inappropriately pool then test?). This is incorrect because it effectively aggregates all the counts within each group and so ignores all within-group biological variability \citep{Hansen:2012gr}.

Instead, a test that incorporates the within-group variability should be used.  For example, we could use a t-test to compare the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$. Several methods have been proposed for testing for DMCs. These methods differ considerably in how they model and transform the data as well as what test statistic is used to identify DMCs.

\paragraph{Transforming}\label{sec:transforming}

Smoothing of the $\beta$-values has been used advocated by several authors to remove variation due to low sequencing coverage and to exploit the spatial correlation of DNA methylation \citep{Hansen:2011gu, Hebestreit:2013ko, Gokhman:2014c}. Both `BSmooth` \citep{Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} smooth the $\beta$-values using a binomial local likelihood smoother. \citet{Lister:2011kg, Gokhman:2014cp} used a simpler moving average in windows along the genome.

\paragraph{Modeling}\label{sec:modeling}

`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages. Broadly, under the Beta-Binomial model the "true" methylation level at the $i^{th}$ methylation locus in the $k^{th}$ group is modeled as a Beta random variable, $B_{i, k} = Beta(\mu_{i, k}, \theta_{i, k})$. The observed number of methylated reads at the $i^{th}$ methylation locus, in the $j^{th}$ sample is modeled as a Binomial random variable, $M_{i, j, k} | B_{i, k}, n_{i, j, k} = Binomial(n_{i, j, k}, B_{i, k}), which is conditional on the unobserved $B_{i, k}$ and the observed sequencing coverage, $n_{i, j, k}$.

\paragraph{Estimating and testing}\label{sec:estimating_and_testing}

`DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation. The test used to identify DMCs is variously a Wald test (`BiSeq`, `DSS`), likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).

\subsubsection{Multi-group experiments}\label{sec:multigroup_dmcs}

\subsection{DMRs}

There are two very different strategies for identifying differentially methylated regions:

1. Use regions that are defined _a priori_, which are then tested for differential methylation. Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g. 3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
2. Use regions that are defined based on the data, which are then tested for differential methylation. Valid statistical inference of such regions is __very challenging__.

\subsubsection{Using _a priori_ regions}\label{sec:a_prior_regions}

The methylation level is summarised at the region level and then compared across groups. For example, the average of the $\beta$-values across the region may be compared between groups using a t-test.

__TODO: Describe the above mathematically__

This is qualitatively similar to identifying differentially methylated cytosines.

__PROS__

* Simple to implement
* Valid statistical properties

__CONS__

* Doesn't account for differential CpG density across regions
* What is the right "unit" - the CGI, the promoter, a tile???
* Doesn't account for spatial correlation of methylation

\subsubsection{Using data-driven regions}\label{sec:data_driven_regions}

__PROS__

* Can discover the right "unit".
* Might account for spatial correlation of methylation

__CONS__

* Loss of statistical properties in testing, e.g. FDR control
* Bloody hard

\section{Other downstream analyses}\label{sec:other_downstream_analyses}

Bisulfite-sequencing data are used to address a variety of questions apart from identifying differential methylation. Not all these analyses require statistical inference. For example, if we are interested in knowing whether the promoter of a particular gene is methylated in our sample then it may be sufficient to simply plot the level of methylation at each CpG in the region.

However, once we move to experiments with multiple samples, multiple regions and more complex questions then statistical methods are generally required to distinguish the signal from the noise. In this section I briefly describe other

__TODO: Any other important downstream analyses worth mentioning?__

\subsection{Variably methylated regions and differential variability}\label{sec:vmrs}

\subsection{Allele-specific methylation}\label{sec:asm}

\subsection{Methylation entropy}\label{sec:methylation_entropy}

\subsection{Epipolymorphism}\label{sec:epipolymorphism}

\subsection{Epiallele detection}\label{sec:epialleles}

\subsection{Integrating DNA methylation data with other genomes data}\label{sec:data_integration}

## TODOs
* See https://github.com/brentp/clustermodel/README.md for a nice, brief summary of the main DMR calling approaches.
* Replace "R" with "\\R"

### 2014-11-08 Rmarkdown/biology_background.Rmd
Finished first draft of biology background chapter.

@@
 In this chapter I summarise the basic biology necessary for
-understanding my thesis.
-I then introduce DNA methylation, describe some of its biological roles and discuss the types of questions asked in experiments studying DNA methylation.
-Finally, I describe assays for studying DNA methylation, with a particular focus on the whole-genome bisulfite-sequencing assay known as _methylC-seq_.
+understanding my thesis. I introduce DNA methylation, describe some of its biological roles and assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
 
@@
 It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression (__CITE__), silencing of transposable elements (__CITE__), X chromosome inactivation (__CITE__) and tumorigenesis (__CITE__).
+
 However, DNA methylation is not found in all organisms, for example, it not detectable in yeast, and so is not essential for life.
@@
 5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm].
-However, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably, as is standard in the literature.
-The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{@Lister:2009boa}.
+However, I will continue to use the term _DNA methylation_ to describe the more specific 5mC, as is standard in the literature.
 
@@
 In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}.
-However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{@VISCHER:1949ty}.
+However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{VISCHER:1949ty}.
 
@@
 Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.
-
-\subsection{DNA methylation in mammals}\label{sec:dna_methylation_in_mammals}
-
-- [x] DONE
-
-DNA methylation is essential for normal development in mammals, such as human and mouse (__REFERENCE__).
-My thesis focuses on DNA methylation in humans but much of the material in this section is true for other mammals.
 
@@
 
+\subsection{DNA methylation in mammals}\label{sec:dna_methylation_in_mammals}
+
+- [x] DONE
+The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{Lister:2009boa}.
 In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_.
@@
 A CpG is a palindromic sequence and is generally symmetrically methylated.
-Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-methylcytosine (__REFERENCE__).
+Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine (__REFERENCE__).
 
@@
 
-One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{@Scarano:1967ue}.
+One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{Scarano:1967ue}.
 Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).
@@
 This definition was refined by \cite{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands.
-More recently, \cite{@Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
+More recently, \cite{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
 
@@
 
-- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
+- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}.
+CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
+Shelves have a comparatively low density of CpGs.
 - CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
@@
 
-Overall, non-CpG methylation is less well studied and less well understood than CpG methylation.
+Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation.
 This is partly due to sampling bias - commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips (__TODO: ref section on technology__), only measure CpG methylation.
 However, recent technological advances (__TODO: ref section on technology__) mean that cytosine methylation can be routinely assayed regardless of the sequence context.
+
+Non-CpG methylation is very common in other organisms, such as plants.
+For example, \cite{Lister:2008bh} found that $45\%$ of "methylcytosines" (see __SECTION__ for the definition and a criticism) in the widely-studied _Arabidopsis Thaliana_ are at CHG or CHH loci.
+However, they also found that the level of methylation at non-CpG loci is typically lower than that observed at CpG dinucleotides.
 
@@
 One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
-
-\subsection{Function of DNA methylation: "normal", cancer and development}\label{sec:function_of_dna_methylation}
-
-- [ ] DONE
-
-\subsection{DNA methylation in other organisms}\label{sec:dna_methylation_in_other_organisms}
-
-- [ ] DONE
-
-__TODO: Don't want too much detail here__
-
-\subsubsection{Insect}
-
-- [ ] DONE
-
-* Honeybees
-* Fruit fly?
-DNA methylation's existance in other organisms, such as the common fruit fly (_Drosophila Melanogaster_), is an area of contention although recent studies suggest it is present albeit at very small quantities and only in very limited contexts [@Takayama:2014jp].
-* General
-
-\subsubsection{Plants}
-
-- [ ] DONE
-
-* Arabadopsis
-* Non-CG methylation
-
-\subsubsection{Fungi}
-
-- [ ] DONE
-
-\subsubsection{Bacteria}
-
-- [ ] DONE
-
-\subsubsection{Archaea}
-
-- [ ] DONE
 
@@
 The methylated cytosines are said to be "protected" from conversion to uracils.
-This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq}, which are referred to as bisulfite-conversion assays.
+This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq, Clark:1994wc}, which are referred to as bisulfite-conversion assays.
 
@@
 Thus, these assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (__FIGURE__).
-
-\subsubsection{Analysis platforms combined with bisulfite conversion}\label{sec:bs_platforms}
-
-__UP TO HERE__
-
-- [ ] DONE
 
@@
 Microarrays contain thousands, even millions, of short _oligonucleotide_ probes.
-Each probe is is designed to hybridise to a particular DNA sequence and emits a signal, such as flurousence, that can be measured to infer the strength of the hybridisation.
+Each probe is is designed to hybridise to a particular DNA sequence and emits a fluorescent signal that can be measured to infer the strength of the hybridisation.
 Therefore, an (idealised) way to analyse DNA methylation with a microarray is to hybridise bisulfite-converted DNA to a microarray that contains probes for both the methylated and unmethylated versions of all sequences of interest.
 The relative methylation of each sequence can be inferred from the relative intensities of the "methylated probe" to the "unmethylated probe".
-Such an idealised experiment brushes over many complications including [@Laird:2010iv]:
+Such an idealised experiment brushes over many complications including \cite{Laird:2010iv}:
 
-* The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a largely 3-base alphabet) leads to decreased hybridisation specificity
-* Sequences containing multiple cytosines require multiple probe versions in order to assay all possible methylation patterns
-* This approach requires the design of organism-specific microarrays
+- The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a largely 3-base alphabet) leads to decreased hybridisation specificity
+- Sequences containing multiple cytosines require multiple probe versions in order to assay all possible methylation patterns
+- This approach requires the design of organism-specific microarrays
 
-The Illumina 450k (__INSERT FULL NAME__) provides a modern implementation of the idea of bisulfite pre-treatment of DNA followed by microarray hybridisation for studying human samples.
-This array assays __HOW MANY__ CpGs (__IN WHAT TYPE OF REGIONS__) and uses two different probe designs (__WHY?__).
-There are still substantial bioinformatics challenges in analysing data from these assays (__CITE: SNPs in probes, normalisation, etc.__).
+The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k_) provides a modern implementation of this type of assay for studying human DNA.
+This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \cite{Stirzaker:2014cu}.
+There are well-established methods and software for analysing Illumina 450k data such as `minfi` \cite{Aryee:2014by}
 
-As mentioned in __SECTION__, genomics research was revolutionised by the development of cheap high-throughput sequencing technology and the study of DNA methylation was no exception.
-In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology [@Cokus:2008fc, Lister:2008bh].
-[@Cokus:2008fc] termed their approach _BS-seq_ while [@Lister:2008bh] called theirs _methylC-seq_.
-These techniques, along with methods that these with targeting of particular regions of the genome, are described further in __SECTION__.
+Genomics research was revolutionised by the development of cheap high-throughput sequencing technology, and the study of DNA methylation was no exception.
+In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology \cite{Cokus:2008fc, Lister:2008bh}.
+Whole-genome bisulfite-sequencing is the gold-standard for analysing genome-wide DNA methylation data.
 
-One final approach to mention is the use of mass spectrometry to analyse bisulfite-treated DNA, as implemented in Sequenom's EpiTYPER (__SOURCE__).
-This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and is frequently used to validate findings discovered using other platforms (__SOURCE__).
+\cite{Cokus:2008fc} termed their approach _BS-seq_ while \cite{Lister:2008bh} called theirs _methylC-seq_.
+From a bioinformatics perspective, the main difference is that the _BS-seq_ protocol produces sequencing reads from four bisulfite-converted DNA strands - the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) - which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results.
+In contrast, the _methylC-seq_ protocol only produces sequencing reads from two bisulfite-converted strands - _OT_ and _OB_ - and so only requires mapping to two different _in silico_ converted reference genomes, followed by a merge.
+The methylC-seq protocol is now the standard whole-genome bisulfite-sequencing protocol, due in part to the simpler bioinformatics analysis.
 
-\subsection{DNA kinetics assays}\label{sec:dna_kinetics_assays}
+Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples.
+Furthermore, approximately $40\%$ of reads will not contain any CpGs, which, depending on the experiment, might be considered a gross waste of money.
+Several assays have been developed to perform targeted high-throughput bisulfite-sequencing.
+The targeted nature of these assays are their obvious advantage and disadvantage - only a subset of the genome needs be sequenced but you only obtain information about methylation in those regions.
+Depending on the experiment this tradeoff may be worthwhile, and these assays have been successfully used in a number of studies (__CITE__).
 
-- [ ] DONE
+Some examples of these targeted bisulfite-sequencing are:
+- Reduced representation bisulfite-sequencing (_RRBS_, \cite{Meissner:2005eb}): RRBS uses restriction enzymes to first select regions of the genome with a high CpG density (based on the Msp-I cleavage motif), which are subsequently treated with sodium-bisulfite and sequenced.
+- NimbleGen's SeqCap Epi Enrichment System and Agilent's SureSelectxt Human Methyl-Seq: Both of these commercial products use DNA hybridisation to enrich the sequencing library for pre-defined regions of interest.
+This enriched library is then bisulfite-converted and sequenced.
 
-\subsection{Pros and cons of bisulfite-conversion assays}\label{sec:pros_and_cons_of_bs_assays
+One final class of bisulfite-conversion assays does not use microarrays or high-throughput sequencing.
+Sequenon's EpiTYPER \cite{Ehrich:2006tt} uses mass spectrometry to analyse DNA methylation from bisulfite-converted DNA.
+This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and may be used to validate findings discovered using other platforms \cite{Laird:2010iv}.
+
+\subsubsection{Pros and cons of bisulfite-conversion assays}\label{sec:pros_and_cons_of_bs_assays
 }
 
-- [ ] DONE
+- [x] DONE
 
-Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation.
-Currently, multiple cells are required as input although the minimum number required is being lowered and efforts are well produce single-cell, whole-genome DNA methylation maps using bisulfite-treated DNA (__SOURCE__).
+Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation since cytosine methylation can be detected at single-base resolution \cite{Stirzaker:2014cu}.
+However, almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1-5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
+The minimal amount of DNA is being reduced with each technological advance.
+Furthermore, single-cell bisulfite-sequencing assays have recently been reported \cite{Guo:2013ih, Smallwood:2014kn}[^not_genome_wide]
+
+[^not_genome_wide]: These are not yet proper genome-wide assays.
+The technique of \cite{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
+And the technique of \cite{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \cite{Smallwood:2014kn}.
 
@@
 
-A recently discovered disadvantage is that bisulfite-conversion assays are unable to distinguish between 5mC and 5hmC because both modifications similarly protect cytosine from deamination to uracil (__SEE [@Yu:2012ej] for references__).
+A recently discovered disadvantage is that bisulfite-conversion assays are unable to distinguish 5hmC from 5mC because hydroxymethyl similarly protects a cytosine from deamination to uracil.
 In effect, the detection of 5mC, and all subsequent inference, is confounded with that of 5hmC.
 For most experiments this isn't much of a problem - most cells have very low levels of 5hmC and so there is little confounding - however, in certain experiments this needs a more careful approach.
-To address this issue, [@Booth:2012fh] developed _oxidative bisulfite-sequencing_ (oxBS-seq) and [@Yu:2012ej] developed _Tet-assisted bisulfite sequencing_ (TAB-seq) to separate 5mC and 5hmC detection.
+To address this issue, \cite{Booth:2012fh} developed _oxidative bisulfite-sequencing_ (oxBS-seq) and \cite{Yu:2012ej} developed _Tet-assisted bisulfite sequencing_ (TAB-seq) for separate 5mC and 5hmC detection.
 
 oxBS-seq specifically measures 5mC.
-The input DNA is oxidated by potassium perruthenate (KRuO$_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment.
+The input DNA is oxidated by potassium perruthenate (RKRuO_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment.
 Only 5mC is protected from conversion during the bisulfite-treatment, which effectively means that only 5mC remains to be detected at the sequencing stage.
@@
 
-A disadvantage of bisulfite-conversion assays is that they require knowledge of the underyling DNA sequence in order to infer the methylation states of cytosines.
+There is also potential confounding of 5C, 5mC and 5hmC levels with 5fC and 5caC, although this has received less attention since 5fC and 5caC are believed to exist at far lower quantities than 5mC and 5hmC.
+Nonetheless, sequencing assays of 5fC and 5caC exist based on the idea of treating the DNA with a chemical and performing 'signal subtraction' with traditional bisulfite-sequencing or another assay \cite{Wu:2014jj}.
+
+Another disadvantage of bisulfite-conversion assays is that they require knowledge of the underlying DNA sequence in order to infer the methylation states of cytosines.
 This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome.
-When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample (__SOURCE__).
-__FIGURE__ describes such an example.
+When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \cite{Liu:2012ge}.
+__FIGURE__ illustrates such an example.
 
@@
 Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_.
-_PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence [@Warnecke:1997eh].
-_Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC (__SOURCE__).
-Finally, bisulfite-treatment can lead to _DNA degredation_, which, as the name suggets, leads to a loss in DNA quality or quantity and can make it particularly hard to sequence long DNA molecules (__SOURCE: Anecdotal__).
+_PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \cite{Warnecke:1997eh}.
+_Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC.
+Finally, treating DNA with sodium bisulfite damages the DNA, resulting in the fragmentation of long molecules.
+This limits the size of the fragments that can be studied using bisulfite-conversion assays to $\approx 500$ bp \cite{Ecker:2010et}
 
-\section{Whole-genome bisulfite-sequencing}\label{sec:wgbs}
+\subsection{DNA kinetics assays}\label{sec:dna_kinetics_assays}
 
-- [ ] DONE
+- [x] DONE
 
-\subsection{Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq}\label{sec:targeted_bsseq}
+Ideally, an assay of DNA methylation could naturally 'read' methylated cytosines as a distinct signal from unmethylated cytosines.
+This idea forms the basis of assays using the kinetics of DNA to infer the presence of DNA methylation and other DNA modifications.
+While very exciting, these assays do not yet scale to studying genome-wide DNA methylation levels in mammalian-sized genomes.
 
-- [ ] DONE
-
-* "RRBS’s ability to interrogate a locus is dependent on its MspI-cut-site (CCGG) density and consequently measures 10%–15% of the CpGs in the human genome (Bock et al.
-2010; Harris et al.
-2010)" \cite{Stevens:2013hv}
-* See \cite{Stevens:2013hv} for MRE-seq details and a good summary of DNA methylation assays
+The only commercially available assay in this class is the Pacific Biosciences SMRT technology \cite{Flusberg:2010hra}.
+It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.
 

### 2014-11-08 Rmarkdown/biology_background.Rmd
Fixed latex errors and missing references.

@@
 These readers can recruit additional proteins to the site of the methylated cytosine to perform a variety of functions related to gene expression.
-For example, the methyl-CpG-binding domain (_MBD_) group of proteins bind to DNA containing a methylated CpG, which can then suppress gene expression by preventing transcription factor binding at that site \cite{Nan:1993gm}.
+For example, the methyl-CpG-binding domain (_MBD_) group of proteins bind to DNA containing a methylated CpG, which can then suppress gene expression by preventing transcription factor binding at that site \cite{Nan:1993gma}.
 Another group, the ubiquitin-like, containing PHD and RING finger domain (_UHRF_) proteins, help _DNMT1_ methylate hemimethylated DNA, such as the daughter strand created during DNA replication \cite{Sharif:2007ft, Bostick:2007jh}.
@@
 This can be used to design an assay to infer the methylation state of a DNA fragment.
+
+__TODO: Is it most accurate to say that restriction enzymes "cleave", "cut" or "digest" DNA?__
 
@@
 oxBS-seq specifically measures 5mC.
-The input DNA is oxidated by potassium perruthenate (RKRuO_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment.
+The input DNA is oxidated by potassium perruthenate ($RKRuO_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment.
 Only 5mC is protected from conversion during the bisulfite-treatment, which effectively means that only 5mC remains to be detected at the sequencing stage.
@@
 It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.
-
-## General TODOs
-* Nucleotide, residue, (nucleo)base; what's the difference and which term should I use?
-	* A nucleotide is composed of a nucleobase (also termed a nitrogenous base), a five-carbon sugar (either ribose or 2-deoxyribose), and one or more phosphate groups \[Coghill, Anne M.; Garson, Lorrin R., ed.
-(2006).
-The ACS style guide: effective communication of scientific information (3rd ed.).
-Washington, D.C.: American Chemical Society.
-p.
-244.
-ISBN 978-0-8412-3999-9.)\]
-* Should I give more detail on Rafa's CGI HMM?
-* Define eukaryotes, prokaryotes and viruses
-* Hyphenation or not of "bisulfite-sequencing"?
-  - Yes, hyphenate.
-* General term for methylC-seq, BS-seq, etc.
-  - "whole-genome bisulfite-sequencing (WGBS)"
-* Is it most accurate to say that restriction enzymes "cleave", "cut" or "digest" DNA?
-* Define/describe DNA sequencing and the genomics revolution in the section describing DNA (or reference genome section?)
-	* Define "library"
-* Fix up reference formating
-* Note that when I refer to a read I mean either a single-end read or a paired-end read. I will use to read_1 and read_2 if I want to refer a particular read that makes up a paired-end read.
-* \cite{Zhang:2013uu} credit Laurent et al. (2010) with BS-seq and Cokus with WGBS.

### 2014-11-08 Rmarkdown/wgbs_analysis.Rmd
Fixed latex errors and missing references.

@@
 ---
-title: "Analysis of bisulfite-sequencing data"
+title: "Bioinformatics analysis of bisulfite-sequencing data"
 author: "Peter Hickey"

### 2014-11-15 Rmarkdown/wgbs_analysis.Rmd
Moved stuff from stat model to webs analysis chapter

@@
 
-While "methylation-aware" mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads (__CITE FELIX__).
+While _methylation-aware_ mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads (__CITE FELIX__).
 This biases downstream inference and means that these mappers are generally less popular.
@@
 Each of these aligners can report the output in the standard `SAM` format (__CITE__), although each mapper does so in a slightly different way, which it difficult to develop downstream analysis tools.
+
+Read mapping is not perfect and produces both false positive and false negative results.
+False positives are due to reads mapped to the wrong location and reads mapped to multiple locations with equal mapping scores.
+False negatives are reads that are not mapped to any location; these reads are effectively lost from any downstream analysis.
+The parameter settings used by the mapping software determine the false positive and false negative rates.
+
+There are biological and technical reasons why mapping against a reference genome can produce these errors.
+Biologically, if the sample contains sequences that are too genetically divergent from the reference genome then these sequences will be difficult, even impossible, to map.
+A particularly problematic class of sequences are those from repetitive regions of the genome.
+These repetitive sequences will map to multiple locations in the reference genome equally well.
+Furthermore, the number of times these repetitive sequences occur differs between the reference genome and sample's genome.
+
+Technically, reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences.
+Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \cite{Flusberg:2010hra}, which produces longer reads.
+
+Another source of technical error in read mapping is really due to sequencing error.
+A sequencing error can transform a uniquely mapping read to one that maps equally well to multiple locations or, worse still, a read that maps uniquely to a single, but incorrect, location.
+Sequencing errors can also corrupt a read so badly that it no longer can be mapped, which leads to that read being a false negative.
+In practice, most people try to mitigate these problems through their choice of parameters used by the read mapping software.
+
+Ideally, mapping software assigns the degree of confidence it has that the read is correctly mapped via a mapping quality score (`mapQ`).
+In theory, reads might be down-weighted in downstream analyses based on the mapping quality score.
+However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful.
+`Bismark` \cite{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).
+
+The above problems are general challenges of read mapping and are not specific to bisulfite-sequencing data, although the reduced complexity of bisulfite-sequencing reads exacerbates these issues.
+The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data.
+Repetitive sequences, such as LINEs and SINEs, are frequently methylated and so are of interest to researchers studying DNA methylation.
+The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__CHECK IN MY DATA AND REPORT COVERAGE__).
 
@@
 Each aligned read may be used for methylation calling, although each read should be subjected to some filtering to remove low-quality reads and low-quality bases.
-Standard filters include removing reads marked as possible PCR duplicates, reads with a mapping quality (`mapQ`) below a threshold and bases with a base quality below a threshold, and ignoring read-positions where there is evidence of M-bias.
+When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from the estimation of $M$ and $U$[^read_filters].
+Strictly speaking each sequenced nucleotide is assigned a weight, however, in practice, filter are normally first applied to reads and then to all nucleotides within 'surviving' reads.
+In my own `methtuple` software I routinely use the following filters.
+
+[^read_filters]: A read that is not used to estimate $M$ and $U$ may still be used in other analyses, such as estimating copy number variation, and vice-versa.
+
+A read survives if:
+
+1. Read is mapped (SE, PE) and mapped in the expected orientation (PE only).
+2. Read is not marked as a PCR duplicate.
+3. Read has a mapping quality score greater than some threshold.
+
+A sequenced base survives if:
+
+1. Read-position of base means that it is unlikely to be affected by M-bias.
+2. The base quality score greater than some threshold.
+3.
+The base is a 'methylation-induced mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a "non-methylation-induced mismatch" (e.g., the sequenced base is an A or a G at a C in the reference sequence).
 

### 2014-11-15 Rmarkdown/a_statistical_model_of_methlyC-seq.Rmd
Moved stat model chapter to Rmarkdown.

---
title: "A statistical framework for analysing bisulfite-sequencing data"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\section{Notes from outline.md (plus other notes added along the way)}

* Chapter Overview
* Notation
  * See `notation.Rmd`
* Estimating $Z$, $M$, $U$ and $\beta$
  * See `notation.Rmd`
  * Current approaches
* Statistical properties of $z$, $m$, $u$ and $\beta$
  * See `notation.Rmd`
  * Marginal distributions
  * Dependence/correlation over $i$ (see also __Chapter X__)
    * Li _et al._ compute $\beta$ values and compute correlations within a sample as a function of genomic distance.
    * Arabidopsis compute $\beta$ values and compute correlations within a sample as a function of genomic distance.
    * Lister (2009) Fig. 3 and Sup. Fig. 9
    * Lister _et al._ compute both within-read and across-read correlations or dependencies of DNA methylation.
    * Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.
  * Dependence/correlation over $j$ (?)
    * Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.

\section{Chapter overview}\label{sec:statmodel_chapter_overview}

- [x] DONE

In this chapter I set out a statistical framework for analysing bisulfite sequencing data. I begin by explaining the various levels of stochasticity in a methylC-seq experiment. Following this, I define the mathematical notation that I use throughout my thesis. I describe key variables, common estimators of these and their statistical properties, and statistical models used in downstream analyses.

To begin, I consider the simple experiment of performing methylC-seq on a single sample. Once I have described the framework for a single sample, I extend it to multiple samples. The ideas here are simple, although when extended to their full generality the notation becomes messy.

\section{One sample}\label{sec:one_sample}

- [x] DONE

Even with only a single sample, there are several levels of variation to consider, which have a hierarchical structure. I separate this structure into two stages, as illustrated in __CARTOON FIGURE OF METHYLC-SEQ EXPERIMENT__:

1. Pre-sequencing
2. Post-sequencing

\subsection{Single locus analysis}\label{sec:one_sample_single_locus}

- [x] DONE

As described in Chapter \ref{chap:wgbs_analysis}, most analyses to date of WGBS data have focused on the methylation level at individual loci averaged across all cells in the sample. For this reason, I initially describe the framework in terms of single loci. However, bisulfite sequencing data contain much more information than provided by these univariate, marginal summaries of the data. In Section \ref{sec:one_sample_mtuples} I introduce _m-tuples_ that can summarise some of this extra information and which I use extensively in Chapter \ref{chap:comethylation}.

\subsubsection{Pre-sequencing}\label{sec:one_sample_pre_sequencing}

- [x] DONE

A methylation locus is a single cytosine, that is, a CpG, CHG or CHH. The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus with respect to the forward strand, e.g. chr1:723,461. I frequently refer to loci by the subscript $i$ rather than by $pos_{i}$. This means that the distance between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome and, for a small number of instances, that the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes. Generally, $N_{loci}$ is not known exactly, although estimates may be made based from the sample or a reference genome, but this is no great concern.

The methylation state of a locus can vary within a sample due to the fact that DNA for each sample is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile. Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states. Therefore, it is also necessary to consider the next level down in the hierarchy; the DNA fragments within the sample.

Suppose that in the pool of DNA fragments for the sample that there are $H_{i}$ fragments containing the $i^{th}$ methylation locus. In general, $H_{i}$ is unknown and will vary from locus to locus within a sample[^H_i]. Note that the value of $H_{i}$ is determined following the library preparation, including PCR amplificatin of the DNA; therefore, it can give a grossly distorted picture of the true representation of the cells. I denote by $\mathcal{H}_{i}$ the set of all fragments containing the $i^{th}$ locus.

[^H_i]: Knowing $H_{i}$ would require knowing: (1) the number of cells used as input (which might only be known to within an order of magnitude), (2) the ploidy of each cell (generally known) and (3) the number of PCR cycles (generally known). But the real problem is that none of the steps in creating the pool of DNA fragments is perfect, in particular, PCR introduces biases -- some molecules are preferrentially amplified while others "drop out". So even if we knew (1), (2) and (3) we cannot simply multiply these together to compute $H_{i}$, although this might at least give us a rough estimate.

Although we do not know the number of fragments in the pool, we can define (and measure) the methylation state of a locus on a single DNA fragment. I denote by the indicator random variable, $Z_{h, i}$, the methylation state of $i^{th}$ methylation locus on the $h^{th}$ DNA fragment containing the $i^{th}$ methylation locus:

\begin{equation*}
Z_{h, i} = \left\{
  \begin{array}{l l}
    1 & \quad \text{if methylated on the } h^{th} \text{ fragment}\\
    0 & \quad \text{if unmethylated on the } h^{th} \text{ fragment}
  \end{array} \right.
\end{equation*}

By summing over the number of fragments containing the $i^{th}$ locus, we obtain the number of fragments that are methylated at the $i^{th}$ locus ($M_{i}$) and unmethylated at the $i^{th}$ locus ($U_{i}$):

\begin{align*}
  M_{i} &= \sum_{h = 1}^{H = H_{i}} Z_{h, i} = |\{Z: Z \in \mathcal{H}_{i}, Z = 1 \}| \\
  U_{i} &= \sum_{h = 1}^{H = H_{i}} (1 - Z_{h, i}) = |\{Z: Z \in \mathcal{H}_{i}, Z = 0 \}|
\end{align*}

Related to these is the proportion of fragments that methylated at the $i^{th}$ locus:
\begin{equation*}
  B_{i} = \frac{M_{i}}{M_{i} + U_{i}}
\end{equation*}

Again, I emphasise that $H_{i, j}, Z_{h, i, j}, M_{i, j}, U_{i, j} \text{ and } B_{i, j}$ are unobservable. However, by sequencing the pools of DNA fragments we aim to estimate these variables.

\subsubsection{Post-sequencing}\label{sec:one_sample_post_sequencing}

- [x] DONE

We do not sequence every fragment in the pool. Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments. We have a large number(~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments. Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to $H_{i}$.

At this point I ignore reads containing no methylation loci as these are not relevant to this discussion and I make three simplifying assumptions:

1. We perform single-end sequencing.
2. Sequencing is performed without error.
3. Read mapping is perfect.

The effect of the first assumption is minor. When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without "gaps", i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$. However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur (__FIGURE__):

1. There may be gaps due to the insert size being longer than the sum of the read lengths, e.g. $(i, i + 1, i + 3, i + 4))$. In effect, we have missing data for any intervening methylation loci; the $(i + 2)^{th}$ loci in this example.
2. Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g. read_1 gives us $(i, i + 1)$ and read_2 gives us $(i + 1, i + 2, i + 3)$. In this example we must use only one of read_1 or read_2 as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.

I discuss the implications of the latter two assumptions in Section \ref{sec:sequencing_and_read_mapping_errors}.

Each read measures the methylation state of one or more loci from a single DNA fragment. I denote by $\mathcal{R}_i$ the set of all reads containing the $i^{th}$ locus. Therefore, the number of reads containing the $i^{th}$ locus is $|\mathcal{R}_{i}|$, where $|\mathcal{R}_{i}| \leq H_{i}$ with strict inequality for almost all $i$.

A single read containing the $i^{th}$ locus is denoted by $z: z \in \mathcal{R}_{i}$; the observed methylation state is given by:

\begin{equation*}
z: z \in \mathcal{R}_{i} = \left\{
  \begin{array}{l l}
    1 & \quad \text{if methylated at the } i^{th} \text{ locus}\\
    0 & \quad \text{if unmethylated at the } i^{th} \text{ locus}
  \end{array} \right.
\end{equation*}

By summing over the number of reads containing the $i^{th}$ locus we obtain the number of reads that are methylated at the $i^{th}$ locus ($m_{i}$) and unmethylated at the $i^{th}$ locus ($u_{i}$):

\begin{align*}
  m_{i} &= \sum_{z: z \in \mathcal{R}_{i}} z \\
        &= |\{z: z \in \mathcal{R}_{i}, z = 1 \}| \\
  u_{i} &= \sum_{z: z \in \mathcal{R}_{i}} (1 - z) \\
      &= |\{z: z \in \mathcal{R}_{i}, z = 0 \}
\end{align*}

Similarly, we obtain the proportion of reads that are methylated at the $i^{th}$ locus as:
\begin{equation*}
  \beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}
\end{equation*}

These are the so-called $\beta$-values, which are commonly interpreted as an estimate of the proportion of cells in the sample that are methylated at the $i^{th}$ locus. I discuss this interpretation, and other estimators of the  methylation level at a locus, in __SECTION__. I now move from single locus summaries to multi-loci summaries using m-tuples.

\subsection{m-tuples: Multiple loci analysis}\label{sec:one_sample_mtuples}

I define an m-tuple to be a tuple of methylation loci, where m = $1, 2, \ldots$ is the size of the tuple. For example, a CpG 3-tuple is 3 CpGs. A CHH 1-tuple is a single CHH; 1-tuples are the basis of single locus analyses of bisulfite-sequencing data. We can learn more from bisulfite-sequencing data by using m-tuples with m $> 1$.

An observation on an m-tuple is the methylation pattern from a single read that overlaps the m-tuple. I require that the observation is from a single read as this ensures that each observation ultimately comes from a single cell[^chimeric_reads]. For example, suppose we have a read containing 3 CpGs -- the first two CpGs are methylated while the last one is unmethylated. From this read we can observe 3 $\times$ CpG 1-tuples, 2 $\times$ CpG 2-tuples and 1 $\times$ CpG 3-tuple. We can have multiple observations on an m-tuple by observing multiple reads, each containing the m-tuple. __FIGURE__ illustrates this example.

[^chimeric_reads]: This ignores chimeric reads, which depending on the sequencing protocol may be a problem.

Note that in __FIGURE__ I have not constructed a 2-tuple using the first CpG and the last CpG. In general, I focus on m-tuples where the m methylation loci are adjacent. That is, I focus on m-tuples where the _number of intervening loci_ is zero ($NIL = 0$). There are 3 reasons for this:

1. Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ m-tuples, provided that we restrict ourselves to those m-tuples with $NIL = 0$. In contrast, if we allow $NIL \geq 0$ then there are $\binom{l}{m} \sim$ m-tuples.
2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}
3. Measurability: We cannot observe m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology. This is true even when $NIL = 0$ but is especially the case if we allow $NIL \geq 0$.

Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.

For each m-tuple, I define the _intra-pair distance_ (IPD) as vector containing the $m - 1$ pair-wise distances (measured in bp) between adjacent methylation loci in the m-tuple. For example, the 2-tuple (chr7:145, 163) has $IPD = (163 - 145) = (18)$. The 5-tuple (chr2:560, 570, 572, 588, 612) has $IPD = (570 - 560, 572 - 570, 588 - 572, 612 - 588) = (10, 2, 16, 24)$. The IPD vector of a 1-tuple is not defined. In Chapter \ref{chap:co-methylation} I use the IPDs, along with other features such as the genomic context, to define 'similar' m-tuples.

\subsubsection{Pre-sequencing}

- [x] DONE

Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$. The remaining definitions are analogous to those for single methylation loci, that is, when m $= 1$.

I denote by the vector of indicator random variable, $Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)}$, the methylation pattern of the m-tuple $(i, i + 1, \ldots, i + m - 1)$ on the $h^{th}$ DNA fragment containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$:

\begin{equation*}
Z_{h, i} = \left\{
  \begin{array}{l l}
    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus on the } h^{th} \text{ fragment}\\
    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th}) \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus on the } h^{th} \text{ fragment} \\
    \vdots \\
    (1, 1, \ldots, 1) & \quad \text{if methylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus on the } h^{th} \text{ fragment}
   \end{array} \right.
\end{equation*}

$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$ and $\mathcal{R}_{(i, i + 1, i + m - 1)}$ denotes the set of all reads containing the m-tuple $(i, i + 1, i + m - 1)$.

There are $2^{\text{m}}$ possible methylation patterns at an m-tuple. Rather than denote a methylation pattern by a m-vector of zeros and ones, I also write these using $U$ and $M$; for example, the possible methylation patterns at a 2-tuple are $MM, MU, UM$ and $UU$.

Analogously to the definition of $M_{i}$ and $U_{i}$ and $m_{i}$ and $u_{i}$ for the case where $\text{m} = 1$, we have when $\text{m} = 2$:

\begin{align*}
  MM_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (1, 1)\}| \\
  MU_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (1, 0)\}| \\
  UM_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (0, 1)\}| \\
  UU_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (0, 0)\}| \\
\end{align*}

We could extend the $B_{i}$ values to m-tuples, although the intuitive interpretation of these as the average methylation level is lost. Instead, it reflects the relative frequencies of each methylation pattern. Here are the definitions for $\text{m} = 2$:

\begin{align*}
  B_{(i, i + 1)}^{MM} &= \frac{MM_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}} \\
  B_{(i, i + 1)}^{MU} &= \frac{MU_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}}  \\
  B_{(i, i + 1)}^{UM} &= \frac{UM_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}}  \\
  B_{(i, i + 1)}^{UU} &= \frac{UU_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}}
\end{align*}

The definitions for $\text{m} > 3$ follow in the obvious manner.

Again, I emphasise that $H_{(i, i + 1, i + \text{m} - 1), j}, Z_{h, (i, i + 1, i + \text{m} - 1), j}, B_{(i, i + 1, i + \text{m} - 1), j}$ and the set of methylation patterns are unobservable. However, by sequencing the pools of DNA fragments we aim to estimate these variables.

\subsubsection{Post-sequencing}

- [x] DONE

The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_i$. A single read containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, and the observed methylation state is given by:

\begin{equation*}
z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)} = \left\{
  \begin{array}{l l}
    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus}\\
    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th}) \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
    \vdots \\
    (1, 1, \ldots, 1) & \quad \text{if methylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus}
      \end{array} \right.
\end{equation*}

Note that we do not know from which haplotype ($h$) each read came from, only that all methylation loci in the read came from the same DNA fragment.

By counting over the number of reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ we obtain the number of reads contain each methylation pattern at that m-tuple. Here are the definitions for $\text{m} = 2$:

\begin{align*}
  mm_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (1, 1)\}| \\
  mu_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (1, 0)\}| \\
  um_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (0, 1)\}| \\
  uu_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (0, 0)\}|
\end{align*}

The definitions for $\text{m} > 2$ follow in the obvious manner.

We can extend the $\beta_{i}$ values to m-tuples. Here are the definitions for $\text{m} = 2$:

\begin{align*}
  \beta_{(i, i + 1)}^{mm} &= \frac{mm_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
  \beta_{(i, i + 1)}^{mu} &= \frac{MU_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
  \beta_{(i, i + 1)}^{um} &= \frac{UM_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
  \beta_{(i, i + 1)}^{uu} &= \frac{UU_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}}
\end{align*}

The definitions for $\text{m} > 3$ follow in the obvious manner.

Finally, \cite{Landan:2012kp} compute the average methylation of a read containing the m-tuple $(i, \ldots, i + m - 1)$. For each read, $z \in \mathcal{R}_{(i, i + 1, i + 2, i + 3)}$, the average methylation of the read, $\zeta_{z}$, is defined as the proportion of methylation loci in the read that are methylated. Thus, $\zeta_{z} = 0, \frac{1}{m}, \frac{2}{m}, \ldots, 1$.

\subsection{Some complications for a single sample}\label{sec:one_sample_complications}

- [x] DONE

I now discuss some complications and how this framework might accommodate these issues in practice.

\subsubsection{What is $\mathcal{I}$?}\label{sec:what_is_I}

- [x] DONE

As mentioned in Chapter \ref{chap:wgbs_analysis}, studies using bisulfite-conversion assays rely on either a reference genome or, less commonly, separate DNA sequencing of the sample that is assayed. Different analysis strategies lead to different definitions of $\mathcal{I}$, which are approximations to the 'true' set of methylation loci in the sample, $\mathcal{I}^{truth}$. Listed here are definitions of $\mathcal{I}$ from least closely matching to most closely matching $\mathcal{I}^{truth}$:

1. $\mathcal{I}^{ref}$: Defined by the set of methylation loci in the reference genome. This ignore all genetic variation between the sample and the reference.
2. $\mathcal{I}^{refFilter}$: Defined by filtering out problematic loci from $\mathcal{I}^{ref}$. A conservative approach that removes many sites of genetic variation between the sample and the reference as well as sites that do not display genetic variation between the sample and the reference. Misses sample-specific methylation loci.
3. $\mathcal{I}^{BisSNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}. Identifies sample-specific methylation loci and removes reference-specific methylation loci. The best that can be done if only bisulfite-sequencing data is available. Genetic variation detection is not as good as that from DNA-seq data. This is the approach I favour.
4. $\mathcal{I}^{WGS}$: Defined by identifying all methylation loci from _whole-genome sequencing_ (WGS) of the sample's genome. The gold standard. All methylation loci are defined with respect to the sample's genome. The only differences between $\mathcal{I}^{WGS}$ and the "truth" are due to sequencing errors and variant calling errors of the WGS data.

\subsubsection{Genetic heterozygosity at a methylation locus}\label{sec:heterozygosity}

- [x] DONE

In a diploid cell there are are sites where, for example, the maternal chromosome is a CpG whereas the paternal chromosome is an ApG. In effect, the maternal and paternal chromosomes within the sample have different $\mathcal{I}^{truth}$. The number of these heterozygous methylation loci is often small enough not to worry about. However, in some studies, such as those of allele-specific methylation, these loci can be very important.

Parent-specific methylation loci can be identified by calling heterozygous genetic variants using `Bis-SNP` or from WGS of the sample. In practice, the existance of parent-specific methylation loci is often ignored.

\subsubsection{Sequencing and read mapping errors}\label{sec:sequencing_and_read_mapping_errors}

- [x] DONE

Sequencing errors from the Illumina technology are typically base substitutions, for example, a _G_ being read as a _T_. These errors occur more frequently at the 3' end of the reads. Other sequencing technologies have different error profiles. In DNA sequencing, the base quality scores are used to filter out or down-weight likely sequencing errors. However, because the base quality scoring algorithm is tuned for DNA-seq where all 4 nucleotides appear frequently, the base quality scores of methylC-seq data may not be as reliable since the relative frequencies of the 4 nucleotides is quite different.

There are two negative effects of sequencing errors. Firstly, if a methylated cytosine on the forward strand, which should be a C in the read, is erroneously read as another base -- A, G or T -- then the methylation signal is lost. If the erroneous call is an A or a G then at least we might be able to infer that a sequencing error occured. However, if the erroneous basecall is a T then we might instead mistakenly conclude that the cytosine was unmethylated.

Secondly, sequencing errors make read mapping more difficult. These issues are discussed in Chapter \ref{chap:wgbs_analysis}. In terms of the statistical framework described in this chapter, the effect of both sequencing errors and read mapping errors are to introduce errors into $m$ and $u$ and all resulting statistics.

\section{Multiple samples}\label{sec:multiple_samples}

- [x] DONE

To move from a single sample to multiple samples simply requires an additional subscript, $j = 1, \ldots, n$, where $n$ is the number of samples. For example, $\mathcal{I}_{j}$ is the set of methylation loci in the $j^{th}$ sample and $\beta_{i, j}$ is the beta value for the $i^{th}$ locus in the $j^{th}$ sample.

This defines the three levels in the hierarchy of a typical experiment -- individual molecules ($h$), individual methylation loci ($i$) and individual samples ($j$). A fourth level is how the samples relate in terms of an outcome of interest, such as phenotype. This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between pre-defined groups of samples, or the aim of the experiment might be to discover this level, such as in a clustering analysis.

A common experiment of the first type is the two-group design in which $n_{1}$ samples are from group $1$ and $n_{2}$ samples are from group $2$ ($n_{1} + n_{2} = n$). This can be represented by a design matrix $X = [X_{j}]$, where $X_{j} = 1$ if the sample is from group $1$ and $X_{j} = 0$ if the sample is from group $2$.

More generally, we can study multi-group experiments via a suitable definition of the design matrix $X$. We can also include covariates in the standard way by allowing $X_{j}$ to be a row vector, $X_{j} = (x_{1, j}, \ldots, x_{P, j})$, where $x_{p, j}$ encodes the information on the $p^{th}$ covariate for the $j^{th}$ sample.

The second type of experiment is one that aims to cluster individuals based on methylation patterns. This can be thought of as an experiment where the outcome of interest is unknown and we wish to discover it by clustering together samples with 'similar' methylation patterns. We might also wish to then correlate these clusters with some external information, such as disease-status.

\subsection{Some complications with multiple samples}

- [x] DONE

In addition to the complications of the previous section, we now have sample-to-sample variation that must be addressed by this framework.

\subsubsection{}$\mathcal{I}$}

- [x] DONE

Each individual has their own set of methylation loci, that is, $\mathcal{I}_{j}$ differs for all $j$. Furthermore, sequencing coverage varies from sample-to-sample. This means that even if the samples have exactly the same $\mathcal{I}_{j}$, i.e., the samples are genetically identical, each sample will have a different set of loci with 'sufficient' sequencing coverage. Loci without sufficient sequencing coverage are effectively missing data.

In practice, we might choose to study $\mathcal{I}^{common} = \cap_{j} \mathcal{I}_{j}$ or some other intersection of the $\mathcal{I}_{j}$, such as all methylation loci present in at least some fraction of the $n$ samples.

A conservative analysis might only analyse those loci where at least some fraction of the $n$ samples have sufficient sequencing coverage. A less conservative analysis might try to impute the missing values based on methylation levels at neighbouring loci.

\section{Parameter estimation}\label{sec:parameter_estimation}

- [x] DONE

In this section I summarise current techniques for parameter estimation from WGBS data. I do not describe the processing of the raw data. When necessary, I have 'translated' the original work into my notation to make these methods more readily comparable. All methods estimate parameters from files containing the aligned reads.

\subsection{Estimating $M$, $U$}\label{sec:estimating_m_and_u}

- [x] DONE

The simplest and most commonly used estimators of $M_{i}$ and $U_{i}$ are $m_{i}$ and $u_{i}$, that is, the number of reads that are methylated and unmethylated at the $i^{th}$ locus (e.g., \cite{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg, Hansen:2011gu}). $M_{i}$ and $U_{i}$ may be estimated per-strand or aggregated across strands in the case of palindromic methylation loci. CpG methylation is typically aggregated across strands.

Not all reads, nor positions within reads, are equally treated in this counting process. Essentially, each sequenced nucleotide is assigned a weight. By far the most used weighting scheme is a set of filters that assigns each sequenced nucleotide a weight of 0 (observation excluded) or 1 (observation included). More complicated weighting schemes might include weighting by mapping quality and by base quality, although this would likely lead to fractional estimates of $M$ and $U$, which may not be desirable.

`methtuple` uses the same 'filter and count' method to estimate $MM_{(i, i + 1)}$, $MU_{(i, i + 1)}$, etc., that is, the frequency of methylation patterns at m-tuples.

\subsection{Estimating $B$}

- [ ] DONE

Most analyses of bisulfite-sequencing data have focused on estimating the average methylation level at individual methylation loci, $B_{i, j}$. The simplest estimator is $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$, which has been widely used \citet[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}). Under a Binomial model for the number of methylated reads at a locus, $M_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, where $n_{i, j}$ is the sequencing coverage of the $i^{th}$ locus in the $j^{th}$ sample, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.

Recently, more sophisticated methods have been proposed to model the average methylation level. These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and the smoothing-methods proposed by \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}. The aim of these methods isn't necessarily to estimate $B_{i, j}$ - it is typically to identify differential methylation (discussed below) - but estimates of $B_{i, j}$ can be obtained if so desired.

\subsubsection{Beta-binomial models}

- [x] DONE

Several papers have proposed the beta-binomial distribution as a natural model "for describing methylation levels of an individual site across replicates" \cite{Dolzhenko:2014bo}. The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing. Another way to think of this is that the Beta component models the biological variability of the data, while the Binomial component models the sampling variability. This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data. For example, the `edgeR` software \cite{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.

An attractive feature of the beta-binomial distribution is that it can be motivated by, and analysed, using Bayesian statistics, including empirical Bayes methods, or frequentist techniques such as maximum likelihood. For example, the software `DSS` \cite{Feng:2014iq} and `MOABS` \citet{Sun:2014fk} both use the beta-binomial distribution in an empirical Bayes analysis of differential methylation from bisulfite-sequencing data. In contrast, `RADmeth` \cite{Dolzhenko:2014bo} uses the beta-binomial model in a maximum likelihood framework to address the same problem.

\subsubsection{Smoothing $\beta$-values}

- [x] DONE

BSmooth, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$. Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$. Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation, which is discussed in Chapters \ref{chap:co-methylation}.

Smoothing is particularly powerful for loci with low sequencing coverage, where the denominator $m_{i, j} + u_{i, j}$ is small and the corresponding standard error of $\beta_{i, j}$ is large. The smoothed $\beta$-values, and not the raw $\beta$-values, are then generally used in all downstream analyses.

__TODO: Discuss with Terry - is smoothing of the raw $\beta$-values still useful when you have high-coverage sequencing data?__

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$. The smoothing is local to exploit co-methylation, which is a local phenomenon.

In both `bsseq` and `BiSeq` the raw $\beta$-values are weighted according to the binomial likelihood and the kernel function. The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the kernel gives greater weight to those $\beta_{i, j}$ near the centre of the window. \citet{Lacey:2013iy} note that loci with very high sequencing coverage will strongly influence the smoother, potentially biasing estimates at neighbouring loci with low coverage.

`bsseq` assumes that for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$. In contrast, `BiSeq` first creates clusters of CpGs and only assumes that the underlying methylation level is smooth at positions within each cluster.

Whenever smoothing is used, a key parameter is the bandwidth, which is the size of the window in which observations are included at each iteration of the smoother. `bsseq` uses a much larger window size than `BiSeq`; the default window size in `bsseq` is one that contains at least 70 CpGs and is at least 2000kb wide, whereas the default window size in `BiSeq` is 80bp, regardless of CpG-density. This is due to `BiSeq` being developed for RRBS data, which has a high CpG-density per window, whereas `bsseq` was developed for WGBS data, which has a lower and more variable CpG density per window.

Another 'parameter' choice when smoothing is the choice of kernel, although this is generally less important than the choice of bandwidth. `bsseq` uses a tricube kernel and `BiSeq` uses a triangular kernel.

\citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} compare the smoothing results of `BiSeq` to `bsseq`. Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`, but the comparison is selective and limited to a few regions. Furthermore, the comparisons are made using RRBS data, both real and simulated, and, `bsseq` is designed from WGBS[^rrbs_sim].

[^rrbs_sim]: Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} altered the default `bsseq` parameters to try to make them comparable to `BiSeq`. \citet{Hebestreit:2013ko} changed the default minimum window size to 80bp but still required at least 20 CpGs per window. \citet{Lacey:2013iy} kept the default minimum window size of 2000 bp but reduced the minimum number of CpGs per window to 50 from the default of 70.

\section{Statistical properties of $\beta$-values}\label{sec:beta}

- [x] DONE

Under the binomial model, $M_{i, j} \eqd Bin(m_{i, j} + u_{i, j}, B_{i, j})$ , $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt(\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}})$ \citep{Hansen:2012gr}. The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample. In this section I discuss this interpretation and statistical properties of this estimator.

\subsection{Distributions}

- [x] DONE

In a study involving multiple samples, the set of $\beta$-values can be summarised as a matrix where each row is a locus and each column is a sample. This matrix might be visualised to learn about the distribution of methylation levels, either row-wise, to learn about the variability across samples, or column-wise, to learn about the variability within samples.

Restricting our attention to CpGs, __FIGURE__ shows the column-wise summaries of the EPISCOPE data, the genome-wide distributions of $\beta$-values within each sample. What is immediately clear is that these distributions are bimodal. This bimodality is driven by the fact that CpGs in CpG islands are generally unmethylated whereas those outside of CpG islands are generally methylated, which we can see when we stratify CpGs by their CpG island status.

```{r}
# TODO: Show a bunch of plots of beta-values for different samples
```

```{r}
# TODO: Show a bunch of plots of beta-values stratified by CGI-status for different samples
```

To do the same plots for the row-wise summaries would be rather tedious since there are many, many more rows (loci) than columns (samples) in the matrix of $\beta$-values. However, a few examples are shown in __FIGURE__. These row-wise summaries are often modelled by the beta distribution \cite[e.g.,]{Hebestreit:2013ko, Lacey:2013iy}. The Beta distribution is a flexible 2-parameter distribution on $[0, 1]$. It can be unimodal, "U"-shaped or "J"-shaped, depending on the choice of parameters. The Beta distribution also includes the Uniform and arcsine distributions as special cases (__TODO: better source than wikipedia__).

The analysis of the row-wise distributions leads to the analysis of differential methylation and differential variability, which are discussed in __SECTION__.

\subsection{Correlations}

- [x] DONE

Many researchers have observed that DNA methylation is spatially correlated along the genome, \cite[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}). I call this correlation of methylation levels _co-methylation_.

Just as we can explore the distribution of DNA methylation levels within a sample or between samples, so too can we explore co-methylation within a sample or between samples. I examine this in detail in Chapter \ref{chap:co-methylation}

\subsection{Biases}

- [x] DONE

The natural interpretation of $\beta_{i, j}$ as the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample will be biased if the probability of sequencing a fragment with a methylated site is different from the probability of sequencing a fragment with an unmethylated site. In fact, it has been shown that methylated DNA is overrepresented in bisulfite-sequencing data due, with the problem exacerbated by higher rounds of PCR amplification and dependent on the bisulfite-conversion protocol \cite{Ji:2014fq}. PCR amplification can result in overreprestation of one of the DNA strands in bisulfite-sequencing data \cite{Warnecke:1997eh}.

Lab-based solutions to overcome these biases exist for targeted bisulfite-sequencing, but are technically difficult and expensive to extend to whole-genome studies. \cite{Ji:2014fq} propose potential computational corrections for these biases but as yet these have not been implemented.

\subsection{Transformations}

__UP TO HERE__

- [ ] DONE

$\beta$-values are the _de facto_ standard unit for reporting methylation levels due to their natural interpretation as the average level of methylation at the locus. However, they are not necessarily the best unit for statistical inference. This is because a $\beta$-value is an estimate of a proportion and there are a well-known challenges when working with proportion data, such as:

1. Proportions are bound between 0 and 1, inclusive.
2. The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$. Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
3. We need to more than just the $\beta$-value to have a sense of how precise an estimate it is. Essentially, we need to also know the sequencing coverage of the methylation loci. Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100$ $u = 300$. Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision. Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.

To address (2), proportion data are often transformed via a variance stabilisation transformation. The aim is to make the variance (approximately) independent of the mean. Popular variance stabilisation transformations include:

- The arcsine transformation, $\arcsin{\sqrt{\beta}}$ \citep{ANSCOMBE:1948bw}. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
- The "averaged arcsine" transformation, $\arcsin{\sqrt{\frac{m}{m + u + 1}}} + \arcsin{\sqrt{\frac{m + 1}{m + u + 1}}}$ \citep{Freeman:1950bh}. This transformation does not have a unique inverse \citep{Nunes:2009vj}.

The use of variance stabilising transformations for proportion data seem to have fallen out of favour (e.g. [http://www.esajournals.org/doi/full/10.1890/10-0340.1](http://www.esajournals.org/doi/full/10.1890/10-0340.1)). The now favoured approach is generalised linear models (__CITE__), in particular the logistic regression model (__CITE__?).

__TODO: Cite beta-values vs. M-value paper and discuss__

\section{Downstream analyses}\label{sec:downstream}

- [ ] DONE

\subsection{Regression models}\label{sec:regression_models}

- [ ] DONE

* Logistic regression
  * M-values
  * The logit transformation, $\text{logit} (\frac{\beta}{1 - \beta})$. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
  * The probit transformation, $\Phi^{-1}(\beta)$, where $\Phi$ is the standard Normal cumulative distribution function. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
* Beta regression

\cite{Du:2010dc} show improved inferences when using Logit transformed $\beta$-values, which they call M-values. This transformation reduces the heteroscedasticity of $\beta$-values near 0 and 1.

\section{General TODOs}

* Autocorrelation or correlation?
* Discuss distribution of coverage? Perhaps in context of sequencing bias?

### 2014-11-22 Rmarkdown/biology_background.Rmd
Corrected citations
Begin polishing biology background

@@
 
-- [x] DONE
-
 In this chapter I summarise the basic biology necessary for
@@
 
-- [x] DONE
-
 The genome is the genetic material of an organism.
 In most organisms, the genome is encoded by deoxyribonucleic acid, _DNA_.
-__FIGURE__ beautifully illustrates the hierarchy that takes us from DNA through to the genome.
-
 In eukaryotes[^eukaryotes], which includes plants, fungi and animals, the DNA is wound around repeating units of eight _histone_ protein cores to form _nucleosomes_, which are the fundamental unit of _eukaryotic chromatin_.
 Chromatin compactly packages the DNA into _chromosomes_, so that the organism's complete nuclear DNA, its nuclear genome, might fit into the nucleus of its cells[^extra_nuclear_genomes].
+
+The material in this chapter can be found in any good introductory molecular biology text, such as \cite{Alberts:2007vj}.
 
@@
 
-The double helix is the most common, and most famous, structure of DNA (__REFERENCE__).
+The double helix is the most common, and most famous, structure of DNA.
 In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_.
@@
 These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix.
+This is illustrated in Figure \ref{fig:double_helix}.
 
-[__FIGURE OF DNA BASE  PAIRING__](http://upload.wikimedia.org/wikipedia/commons/4/4c/DNA_Structure%2BKey%2BLabelled.pn_NoBB.png).
+![Simple diagram of double-stranded DNA showing complementary base pairings.
+By Forluvoft (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File%3ADNA_simple2.svg](http://commons.wikimedia.org/wiki/File%3ADNA_simple2.svg)\label{fig:double_helix}](../figures/DNA_simple2.pdf)
 
-A gene is a sequence of DNA that is _transcribed_ to produce a functional product called ribonucleic acid or _RNA_ (Source: [http://sandwalk.blogspot.com.au/2007/01/what-is-gene.html](http://sandwalk.blogspot.com.au/2007/01/what-is-gene.html)).
-RNA may in turn be translated into a protein sequence or perform other roles in the regulation of gene expression (__CITE__).
-It is important to note here than not all DNA is transcribed into RNA.
-That is not to say that untranscribed DNA is unimportant.
+A gene is a sequence of DNA that is _transcribed_ to produce a functional product in the form of ribonucleic acid, _RNA_.
+RNA may in turn be translated into a protein sequence or perform other roles in the regulation of gene expression.
+It is important to note here than not all DNA is transcribed into RNA, which is not to say that untranscribed DNA is unimportant.
 For instance, there are untranscribed _regulatory sequences_ of DNA that determine whether a nearby gene is transcribed.
-There is also _junk DNA_, _garbage DNA_, __etc.__ (__SOURCE__).
+There is also _junk DNA_ that is of little consequence to the organism \cite{Alberts:2007vj}.
 Conversely, not all transcribed DNA is a gene.
-DNA transcription is "permissive" and there are many DNA sequences that are transcribed by accident or in error.
+DNA transcription is _permissive_ and there are many DNA sequences that are transcribed by accident or in error.
 
-DNA is able to self-replicate so that, for example, eukaryotic cells created during _mitosis_ contain the same DNA as the "parent" cell.
+DNA is able to self-replicate.
+This means that, for example, eukaryotic cells created during _mitosis_ contain the same DNA as the 'parent' cell.
 During DNA replication, the two strands are separated and each strand's _complementary_ DNA sequence is copied by an enzyme called _DNA polymerase_.
@@
 [^dna_replication]: This of course ignores errors in the replication process.
-Such errors are very rare events but because DNA replication happens __HOW FREQUENTLY__ these events do occur.
+Such errors are very rare events but because DNA replication happens so frequently these events do occur.
 There are error-correcting processes that reduce the chance that such an error is retained in the daughter sequence, however, these are not perfect.
@@
 
-Basic descriptions of nucleosomes and chromatin are ripe with analogies.
-In one such popular analogy, the histones are the 'spool' around which the DNA 'thread' is wrapped to form nucleosomes (__CITE__).
-These nucleosomes themselves have the appearance of "beads on a string" when viewed under an electron microscope (__CITE__).
-
 The core of a nucleosome consists of four pairs of histones, H2A, H2B, H3 and H4, which are consequently known as the _core histones_.
@@
 Each of these histones has a 'tail' consisting of a string of amino acids.
-These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins (__CITE__) (see below Section \ref{epigenetics} on epigenetics).
+These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \cite{Alberts:2007vj} (discussed in Section \ref{sec:epigenetics} on epigenetics).
 
 The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin.
-The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone (__CITE: http://www.nature.com/scitable/topicpage/dna-packaging-nucleosomes-and-chromatin-310#__).
+The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \cite{Annunziato:2008ue}.
 
@@
 
+Basic descriptions of nucleosomes and chromatin are ripe with analogies.
+In one such popular analogy, the histones are the 'spool' around which the DNA 'thread' is wrapped to form nucleosomes.
+The chromatin then has the appearance of "beads on a string" when viewed under an electron microscope \cite{Alberts:2007vj}.
 Chromatin is further coiled up into various literally-named structures, such as the 30-nanometre and 250-nanometre fibres, and, ultimately, packaged into _chromosomes_.
@@
 The length of a chromosome is typically reported as the number of DNA base pairs (bp) in a single copy of that chromosome and the _haploid_ length of a genome is the sum of these chromosome lengths.
-The haploid human genome is approximately 3.3 billion base pairs long.
+The haploid human genome is approximately 3 billion base pairs long (Golden Path Length [http://asia.ensembl.org/Homo_sapiens/Location/Genome?r=1](http://asia.ensembl.org/Homo_sapiens/Location/Genome?r=1)).
 
-A _reference genome_ was completed by the Human Genome Project in 2003 (__CITE__).
-This reference genome does not represent the genome of any one  human since it uses DNA donated by several different people.
+A human _reference genome_ was jointly completed by the International Human Genome Consortium and Celera Genomics in 2003 \cite{Venter:2001fq,Lander:2001hk}.
+This reference genome does not represent the genome of any one human since it uses DNA donated by several different people \cite{Venter:2001fq,Lander:2001hk}.
 Rather, a reference genome is a kind of map or scaffold that can be used to identify similarities and differences between individual genomes.
 
-The human genome has obvious uses in medical research and biotechnology, but is also used to learn about evolution and human history, such as migration.
-Every person, except genetically identical twins, has their own unique genome.
-However, the genomes of any two randomly selected people are identical at __WHAT PERCENTAGE__ of sites.
-The vast majority, __WHAT PERCENTAGE__, of the human genome is made up of non-coding DNA.
-This is similar to other organisms with similarly large genomes (__CITE__).
-Much of the non-coding DNA is junk DNA.
-Indeed, over __WHAT PERCENTAGE__ of the human genome is comprised of repetitive DNA.
+The human genome has obvious uses in medical research and biotechnology, but is also used to learn about evolution and human history, such as human migration patterns \cite{Hellenthal:2014ky}.
+Every person, even a _monozygotic_ ('identical') twin \cite{Hellenthal:2014ky}, has their own unique genome.
+However, genomes of any two randomly selected people are identical at approximately $99.9\%$ of sites.
+Furthermore, the vast majority, $98\%$ by some estimates \cite{Elgar:2008dm}, of the human genome is made up of non-coding DNA and upwards of $50\%$ is repetitive sequence \cite{Treangen:2012dn}.
 
@@
 Mice are also diploid organisms, but have 19 pairs of autosomes, one pair of sex chromosomes and a mitochondrial chromosome.
-The mouse genome is slightly smaller than the human genome, at 2.8 billion base pairs long.
-Like the human genome, there is a mouse reference genome (__CITE__).
+The mouse genome is slightly smaller than the human genome, at 2.7 billion base pairs long (Golden Path Length [http://asia.ensembl.org/Mus_musculus/Location/Genome?r=1](http://asia.ensembl.org/Mus_musculus/Location/Genome?r=1)).
+Like the human genome, there is a mouse reference genome \cite{MouseGenomeSequencingConsortium:2002fn}.
 It is based on several female mice from the _C57BL/6J_ strain, an important strain of mouse widely used in medical research.
@@
 
-\subsubsection{Arabidopsis Thaliana genome}\label{sec:human_genome}
+\subsubsection{Arabidopsis thaliana genome}\label{sec:arabidopsis_genome}
 
@@
 The most widely studied species is _A.
-thaliana_, which is diploid and has one of the smallest genomes known in the plant kingdom (__CITE__).
-The genome of _A.
-thaliana_ is 135 million base pairs long and is packaged as five pairs of chromosomes (__CITE__).
+thaliana_, which is diploid and has one of the smallest genomes known in the plant kingdom at 135 million base pairs long, which is packaged as five pairs of chromosomes ([http://plants.ensembl.org/Arabidopsis_thaliana/Info/Index](http://plants.ensembl.org/Arabidopsis_thaliana/Info/Index)).
 The small size of its genome, coupled with its small physical size, rapid development and its tolerance to both cross-pollination and self-pollination[^cross_and_self_pollination], make _A.
@@
 
-A key question in biology, perhaps _the_ key question, is what determines an organism's _phenotype_.
-An individual's phenotype is "the set of observable characteristics of an individual resulting from the interaction of its genotype with the environment" (__CITE: currently using google's definition__).
+A key question in biology, perhaps _the_ key question, is what determines an individual's _phenotype_.
+An individual's phenotype is the set of its observable characteristics resulting from the interaction of its genotype with the environment.
 Two simple examples of phenotypes are height and weight.
-Both have a genetic component, e.g., the offspring of tall parents are on average taller than those offspring of short parents, but there are also environmental components, such as the contribution of diet to weight.
+Both have a genetic component, e.g., the offspring of tall parents are on average taller than the offspring of short parents, but also have environmental components, such as the contribution of diet to weight.
 In medical research, a person's phenotype might be whether she is affected by a particular disease.
@@
 
-[^indels]: __TODO: Define__
+[^indels]: An indel is a short insertion or deletion of DNA sequence in an organism, usually relative to a reference genome.
 
 Environmental variation has clear influence on certain phenotypes.
-However, it often can be difficult to determine whether phenotypic variation is driven by genetic varation or environment variation, particularly in humans where genetically similar individuals typically grow up in similar environments.
+However, it often can be difficult to determine whether phenotypic variation is driven by genetic varation or environment variation, particularly in humans where genetically similar individuals typically also grow up in similar environments.
 
@@
 If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell) and a cone cell (a photoreceptor in the retina) all have identical genomes.
-In fact, all cells in an organism, excluding the gametes, have an identical genome yet play very different roles.
+In fact, all cells in an organism, excluding the gametes, have an identical genome yet play very different biological roles.
 This is due to different genes being active in different cells.
@@
 
-- [x] DONE
-
-"Epigenetics" is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ...
+_Epigenetics_ is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ...
 it means just what I choose it to mean - neither more nor less" \cite{carroll1897through}.
@@
 
-A popular contemporary definition of epigenetics is attributed to the epigeneticist Art Riggs that epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \cite[p1]{russo1996epigenetic}.
+A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs - epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \cite[p1]{russo1996epigenetic}.
 The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence.
@@
 
-> The following could be a unifying definition of epigenetic events: the structural adaptation of chromosomal regions so as to register, signal or perpetuate altered activity states.
+> "The following could be a unifying definition of epigenetic events: the structural adaptation of chromosomal regions so as to register, signal or perpetuate altered activity states.
 This definition is inclusive of chromosomal marks [e.g.
-histone marks], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify.
+histone marks], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify."
 
 Regardless of which definition you subscribe to, an _epigenetic mark_ is the modification that causes this "epigenetic change".
-Actually, "causes" may be too strong a phrase as much of current epigenetics research is in identifying associations rather than causations and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.
+Actually, "causes" may be too strong a phrase, as much of current epigenetics research is in identifying associations rather than causations, and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.
 
@@
 In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual.
-Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual (__TODO: cite examples or upcoming chapters__).
-Epigenomics is the study of the epigenome, analogous to genomics being the study of the genome.
-However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is much evidence that genetic variation can drive epigenetic variation (__CITE__).
+Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual as well will see in later chapters.
+_Epigenomics_ is the study of the epigenome, analogous to genomics being the study of the genome.
+However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is evidence that the epigenetic variation is associated with genetic variation \cite{McVicker:2013ih,Bell:2011fm,Zhang:2010bg,vanEijk:2012hc}
 
@@
 
-- [x] DONE
-
 DNA methylation is a chemical modification of DNA that can impart information on top of the DNA sequence.
 It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying].
-It therefore fits into Art Riggs' afore-mentioned definition of an epigenetic modification \cite[p1]{russo1996epigenetic}.
+It therefore fits into Arthur Riggs' afore-mentioned definition of an epigenetic modification \cite[p1]{russo1996epigenetic}.
 DNA methylation is found in bacteria, fungi, plants and animals.
@@
 [^dnam_copying]: In practice this copying is not as faithful as, say, the copying of DNA from the parent to the daughter strand.
-Furthermore, the faithfulness of this copying will be different in different conditions, such as in a healthy, well-differentiated liver cell compared to a cancerous liver cell.
-Nevertheless, the copying of DNA methylation is faithful enough for most biologists to consider it as a mitotically heritable mark, most of the time (__TODO__: Is this just true of 5mC in mammals or more generally?)
+Furthermore, the faithfulness of this copying will be different in different conditions, such as in a healthy liver cell compared to a cancerous liver cell.
+Nevertheless, the copying of DNA methylation is faithful enough for most biologists to consider it as a mitotically heritable mark, most of the time.
 
 A major reason that DNA methylation is studied is that it is essential for normal development in mammals \cite{Li:1992bp}.
-It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression (__CITE__), silencing of transposable elements (__CITE__), X chromosome inactivation (__CITE__) and tumorigenesis (__CITE__).
+It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \cite{Razin:1991vj}, silencing of transposable elements\cite{Jones:2001eg}, X chromosome inactivation \cite{Mohandas:1981uf} and tumorigenesis \cite{Ehrlich:2002kz}.
 
-However, DNA methylation is not found in all organisms, for example, it not detectable in yeast, and so is not essential for life.
-Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation (__TODO: examples and citations__).
+However, DNA methylation is not found in all organisms, for example, it not detectable in yeast \cite{Capuano:2014cu}, and so is not essential for life.
+Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation\cite{Capuano:2014cu}.
 
@@
 Ruppel, first identified a methylated nucleic acid in 1898.
-Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__TODO: Can only find first page of this paper and it's in German.
-Need to verify__).
+Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide \cite{Ruppel:1899hc}.
 In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}.
@@
 This modified cytosine is called _5-Methylcytosine_ or _5mC_.
-[__FIGURE__](http://en.wikipedia.org/wiki/File:5-Methylcytosine.svg) shows the structure of 5mC.
-It is common to describe a cytosine as being "methylated" or "unmethylated", however, care must be taken in defining these terms.
+Figure \ref{fig:5mC} shows the structure of 5mC.
 
+![Chemical structure of 5-methylcytosine.
+"5-methylcytosine".
+By Yikrazuul (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg](http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg)\label{fig:5mC}](../figures/5-methylcytosine.png)
+
+It is common to describe a cytosine as being 'methylated' or 'unmethylated', however, care must be taken in defining these terms.
 At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
@@
 
-[^palindromic_sequence]: "A palindromic sequence is a nucleic acid sequence (DNA or RNA) that is the same whether read 5' (five-prime) to 3' (three prime) on one strand or 5' to 3' on the complementary strand with which it forms a double helix" (__TODO: current source is wikipedia, find and properly cite__).
+[^palindromic_sequence]: A palindromic DNA sequence is one that is identical on when read in the 5' to 3' direction on both the original strand and the complementary strand of the double helix.
+For example, _CG_ is a palindromic sequence.
 
 Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other.
-While the former that is more common, examples of the latter case, such as _allele-specific methylation_ and _genomic imprinting_, are of considerable biological interest (__TODO: sources and examples__).
+While the former is more common, examples of the latter case, such as _allele-specific methylation_ \cite{Shoemaker:2010jz} and _genomic imprinting_ \cite{Li:1993ea}, are of considerable biological interest.
 
@@
 
-- [x] DONE
 The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{Lister:2009boa}.
 In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_.
-A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__).
-The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation.
+A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence.
+The 'p' stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation.
 A CpG is a palindromic sequence and is generally symmetrically methylated.
-Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine (__REFERENCE__).
+Approximately 70% of CpGs are methylated in mammals \cite{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.
 
@@
 
-- [x] DONE
-
 CpGs are underrepresented in the human genome.
-The GC-content of the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is approximately __WHAT %__.
-Under a random sampling model of nucleotides we would expect approximately __WHAT %__ of the human genome to be comprised of CpGs.
-In fact, only __WHAT %__ of the dinucleotides are CpGs.
-__FIGURE__ shows the frequency of each dinucleotide in the human reference genome (hg19)[^cpg_frequency].
-
-[^cpg_frequency]: See __APPENDIX__ for the `R` code used to compute these values for the human reference genome (build hg19).
+The GC-content of each the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is between $41\%$.
+If the nucleotides were uniformly distributed across the human genome we would expect about $4.1\%$ of dinucleotides to be Cpgs.
+Instead, only $1\%$ of dinucleotides are CpGs.
+See the Appendix for details of these calculations and a figure comparing the observed:expected ratio of each dinucleotide in the human reference genome (hg19).
 
 One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{Scarano:1967ue}.
-Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).
-
+Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinuclotide_distribution}).
 There are many other evolutionary pressures on the distribution of nucleotides in a genome.
@@
 \subsection{CpG islands and other sandy metaphors}\label{sec:cpg_islands}
+
+__UP TO HERE__
 
@@
 
-The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_).
+The 20-40% of CpGs that are unmethylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_).
 CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__).
@@
 The minimal amount of DNA is being reduced with each technological advance.
-Furthermore, single-cell bisulfite-sequencing assays have recently been reported \cite{Guo:2013ih, Smallwood:2014kn}[^not_genome_wide]
+Furthermore, single-cell bisulfite-sequencing assays have recently been reported \cite{Guo:2013ih, Smallwood:2014kn}[^not_genome_wide].
 

### 2014-11-22 Rmarkdown/appendix.Rmd
Updated appendex

@@
       in_header: ../latex/header.tex
+    keep_tex: no
   html_document:
-    keep_md: true
+    keep_md: yes
 bibliography: ../latex/phd_thesis.bib
 ---
+
+## CpGs are underrepresented in the human genome
+
+This short R script demonstrates that CpGs are underrepresented in the human reference genome (build hg19).
+
+Firstly, the frequency of each nucleotide is computed.
+Secondly, the expected frequency of each dinucleotide is computed under the _independence model_.
+Under the independence model, the frequency of each dinucleotide is simply the product of the respective observed nucleotide frequencies.
+Finally, these expected frequencies are compared to the observed values from the respective reference genomes.
+
+I only use the autosomes, sex chromosomes and mitochondrial chromosome in all analyses, that is, I do not consider the "random", "unplaced" and "alternate haplotype" contigs in the reference genomes.
+
+All results are computed at the chromosome-level as well as being summarised genome-wide.
+
+### Compute the nucleotide and dinucleotide frequencies
+
+```{r, cache = TRUE}
+library(BSgenome.Hsapiens.UCSC.hg19)
+library(knitr)
+# Nucleotide frequencies
+nf <- bsapply(new("BSParams", X = Hsapiens, FUN = letterFrequency, 
+            exclude = c("rand", "Un", "hap"), simplify = TRUE), 
+            letters = c("A", "C", "G", "T"))
+# Pretty print the results
+kable(data.frame(percentage = round(100 * rowSums(nf) / 
+                                      sum(as.numeric(nf)), 1)))
+
+# GC-percentage
+round(100 * sum(nf[c("C", "G"), ]) / sum(as.numeric(nf)), 1)
+
+# Dinucleotide frequencies
+df <- bsapply(new("BSParams", X = Hsapiens, FUN = dinucleotideFrequency, 
+            exclude = c("rand", "Un", "hap"), simplify = TRUE))
+# Pretty print the results
+kable(data.frame(percentage = round(100 * rowSums(df) / 
+                                      sum(as.numeric(df)), 1)))
+
+# CpG percentage
+round(100 * rowSums(df) / sum(as.numeric(df)), 1)["CG"]
+```
+
+### Plot the observed:expected ratio for each dinucleotide
+
+```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution"}
+library(ggplot2)
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
+edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
+x <- data.frame(Dinucleotide = names(rowSums(df)),
+                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))), 
+                Expected = as.vector(edf))
+ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) + 
+  geom_point(size = 5) + ylim(c(0, 1.5)) + 
+  ggtitle("CpGs are underrepresented in human reference genome (hg19)") + 
+  thesis_theme +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  geom_hline(yintercept = 1, linetype = 2)
+```
 
@@
 
+## Software details
+
+```{r}
+devtools::session_info()
+```
+
 ### TODOs

### 2014-11-22 Rmarkdown/dinucleotide_distribution/dinucleotide_distribution.Rmd
Described methylKit and (most of) DSS
Updated appendex

--- 2014-03-22 Rmarkdown/dinucleotide_distribution.Rmd
@@
-Distribution of dinucleotides in the human and mouse reference genomes
-========================================================
+---
+title: "Distribution of dinucleotides in the human and mouse reference genomes"
+author: Peter Hickey
+date: 22 April 2014
+output: pdf_document
+---
+
+## Distribution of dinucleotides in the human and mouse reference genomes
 
@@
 
-Human genome 
-========================================================
+### Human genome
 
-Step 1: Compute the nucleotide frequencies
---------------------------------------------------------
+#### Step 1: Compute the nucleotide frequencies
 ```{r}
 #### Load necessary packages ####
-library(BSgenome)
 library(BSgenome.Hsapiens.UCSC.hg19)
@@
 
-Step 2: Compute expected number of CpGs under an independence model
---------------------------------------------------------
+#### Step 2: Compute expected number of CpGs under an independence model
 
-Software details
-========================================================
+### Software details
+
+```{r}
+devtools::session_info()
+```

### 2014-11-22 Rmarkdown/a_statistical_model_of_methlyC-seq.Rmd
Refactor webs analysis and stat model chapters

@@
 
-\section{Notes from outline.md (plus other notes added along the way)}
-
-* Chapter Overview
-* Notation
-  * See `notation.Rmd`
-* Estimating $Z$, $M$, $U$ and $\beta$
-  * See `notation.Rmd`
-  * Current approaches
-* Statistical properties of $z$, $m$, $u$ and $\beta$
-  * See `notation.Rmd`
-  * Marginal distributions
-  * Dependence/correlation over $i$ (see also __Chapter X__)
-    * Li _et al._ compute $\beta$ values and compute correlations within a sample as a function of genomic distance.
-    * Arabidopsis compute $\beta$ values and compute correlations within a sample as a function of genomic distance.
-    * Lister (2009) Fig. 3 and Sup. Fig. 9
-    * Lister _et al._ compute both within-read and across-read correlations or dependencies of DNA methylation.
-    * Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.
-  * Dependence/correlation over $j$ (?)
-    * Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.
-
 \section{Chapter overview}\label{sec:statmodel_chapter_overview}
@@
 Once I have described the framework for a single sample, I extend it to multiple samples.
-The ideas here are simple, although when extended to their full generality the notation becomes messy.
+The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.
 
@@
 
-Even with only a single sample, there are several levels of variation to consider, which have a hierarchical structure.
-I separate this structure into two stages, as illustrated in __CARTOON FIGURE OF METHYLC-SEQ EXPERIMENT__:
+Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment.
+This variation has a hierarchical structure, which I first separate into two levels, as illustrated in __CARTOON FIGURE OF METHYLC-SEQ EXPERIMENT__:
 
@@
 
-__UP TO HERE__
+- [x] DONE
 
-- [ ] DONE
-
-$\beta$-values are the _de facto_ standard unit for reporting methylation levels due to their natural interpretation as the average level of methylation at the locus.
+$\beta$-values are the _de facto_ standard unit for reporting methylation levels due to their natural interpretation as an estimate of the average level of methylation at the locus.
 However, they are not necessarily the best unit for statistical inference.
@@
 
-1. Proportions are bound between 0 and 1, inclusive.
-2.
+1.
 The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$.
 Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
-3.
+2.
 We need to more than just the $\beta$-value to have a sense of how precise an estimate it is.
@@
 Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.
+3. Proportions are bound between 0 and 1, inclusive.
 
-To address (2), proportion data are often transformed via a variance stabilisation transformation.
-The aim is to make the variance (approximately) independent of the mean.
+To address (1), proportion data are often transformed via a variance stabilisation transformation.
+The aim is to make the variance independent of the mean, at least approximately.
 Popular variance stabilisation transformations include:
 
-- The arcsine transformation, $\arcsin{\sqrt{\beta}}$ \citep{ANSCOMBE:1948bw}. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
+- The arcsine transformation, $\arcsin({\sqrt{\frac{m + 1}{m + u + 1}}})$ \citep{ANSCOMBE:1948bw}. A small value, in this case 1, is added to both $m$ and $u$ to avoid $\beta = 0, 1$.
 - The "averaged arcsine" transformation, $\arcsin{\sqrt{\frac{m}{m + u + 1}}} + \arcsin{\sqrt{\frac{m + 1}{m + u + 1}}}$ \citep{Freeman:1950bh}.
-This transformation does not have a unique inverse \citep{Nunes:2009vj}.
+One problem with this transformation is that it does not have a unique inverse \citep{Nunes:2009vj}.
 
-The use of variance stabilising transformations for proportion data seem to have fallen out of favour (e.g.
-[http://www.esajournals.org/doi/full/10.1890/10-0340.1](http://www.esajournals.org/doi/full/10.1890/10-0340.1)).
-The now favoured approach is generalised linear models (__CITE__), in particular the logistic regression model (__CITE__?).
+However, the general use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \cite[e.g.,]{Warton:2011gm}.
 
-__TODO: Cite beta-values vs. M-value paper and discuss__
+One transformation that remains popular, at least in the analysis of DNA methylation microarray data, is the logit-transformation, also known as $\mathcal{M}$-values.
+An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2}(\frac{\beta}{1 - \beta} = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero.
+Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \cite[e.g.]{Smyth:2005ta}.
+
+\cite{Du:2010dc} advocate for the use of $\mathcal{M}$-values in statistical analyses of methylation levels from microarray data.
+The main reason for this is that $\mathcal{M}$-values are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values.
+As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformatics and genomics researchers makes it a favourable choice.
+Nonetheless, \cite{Du:2010dc} also advocate that the results of analyses should be reported as $\beta$-values, owing to their "more intuitive biological interpretation".
+
+As with $\beta$-values, $\mathcal{M}$-values derived from bisulfite-sequencing data generally cannot be directly analysed due to the variable sequencing coverage across loci.
+However, the variable sequencing coverage can be accounted for using the afore-mentioned beta-binomial models and related regression models, which are discussed in the next section.
 
@@
 
-\subsection{Regression models}\label{sec:regression_models}
-
-- [ ] DONE
-
-* Logistic regression
-  * M-values
-  * The logit transformation, $\text{logit} (\frac{\beta}{1 - \beta})$. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
-  * The probit transformation, $\Phi^{-1}(\beta)$, where $\Phi$ is the standard Normal cumulative distribution function. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
-* Beta regression
-
-\cite{Du:2010dc} show improved inferences when using Logit transformed $\beta$-values, which they call M-values. This transformation reduces the heteroscedasticity of $\beta$-values near 0 and 1.
+- Don't rehash the details of Chapter WGBS analysis. Rather summarise these questions as statistical problems in the framework I outline in this chapter.
+- DMC: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
+- DMR: Identify runs of DMCs
+- DVC: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
+- DVR: Identify runs of DVCs
 

### 2014-11-22 Rmarkdown/wgbs_analysis.Rmd
Refactor webs analysis and stat model chapters

@@
 This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling.
-This is the approach used by `bismark_methylation_extractor`, `comethylation` and `Bis-SNP`.
+This is the approach used by `bismark_methylation_extractor`, `methtuple` and `Bis-SNP`.
 2.
@@
 
-I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `comethylation`.
+I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple`.
 In theory, the first pass over the `SAM/BAM` could be "skipped" by computing the M-bias simultaneously with sorting the `SAM/BAM` and marking PCR duplicates.
@@
 
-[^cigar]: Many downstream tools, including the current version of `comethylation`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads.
-The current version of `comethylation` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip).
+[^cigar]: Many downstream tools, including the current version of `methtuple`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads.
+The current version of `methtuple` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip).
 __FIX THIS__.
 This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice.
-__TODO: Update what changes are made to `comethylation`.
+__TODO: Update what changes are made to `methtuple`.
 The `bwa-meth` and `LAST` alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark`, cannot soft-clip reads and so require that reads are pre-trimmed (__CITE `bwa-meth` pre-print or paper__).__
@@
 Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
-My `comethylation` software is more flexible by allowing the user to exclude individual read-positions or ranges of read-positions via the `--ignore-read1-positions` and `--ignore-read2-positions` flags.
+My `methtuple` software is more flexible by allowing the user to exclude individual read-positions or ranges of read-positions via the `--ignore-read1-positions` and `--ignore-read2-positions` flags.
 
@@
 [^bismark_methylation_extractor]: `Bismark` also includes a program called `bismark_methylation_extractor`, which, as the name suggests, extracts the methylation calls from the `SAM/BAM` file.
-So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` or `comethylation` is required to make use of the methylation calls.
+So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` or `methtuple` is required to make use of the methylation calls.
 
 Both `Bismark` and `Bis-SNP` perform methylation calling of 1-tuples.
-In contrast, `comethylation` can perform methylation calling at m-tuples.
-`MethPat` (__CITE__) is also capable of performing methylation calling at m-tuples.
-
-\subsubsection{A criticism of "identifying methylcytosines"}{sec:criticism_of_methylcytosines}
-
-__TODO: Need a better title for this section__
-
-\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced an idea "to identify the presence of a methylated cytosine" [supplementary material].
-This idea was also used in subsequent, and similarly high-profile, publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013].
-I believe this idea is an unrequired source of confusion, as I will explain.
-
-[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
-
-It is necessary to use scare quotes to distinguish the term "methylcytosine", as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from methylcytosine, the standard definition meaning a 5-methylcytosine.
-I will first describe the idea and how it differs from methylation calling.
-Then I will describe how it is used to (1) estimate the number of methylcytosines in the sample, and (2) as a kind of screening procedure that determines whether each cytosines will be used in downstream analyses.
-Finally, I will explain why (1) is a bad estimator and why I think (2) is unnecessary or misguided.
-
-A "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
-Thus, the idea of identifying "methylcytosines" is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
-To do so, Lister et al.
-compare each $\beta$-value to a (coverage-specific and sample-specific) cutoff that determines whether the cytosine is classified as a "methylcytosine" or not.
-
-The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy} and \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
-Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done (__TODO: Discuss with Terry__).
-What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-
-The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to "error", where "error" is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
-Mathematically, for each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated "error".
-Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
-
-[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
-
-I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure (__CITE__).
-This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
-thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
-
-The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
-It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
-
-Now to why I don't think this is a very useful procedure.
-Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
-These are two different things.
-If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
-A simple example makes this clear.
-
-Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
-Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
-We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
-From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
-
-Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
-For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
-A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
-
-Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
-While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
-For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
-If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
+In contrast, `methtuple` can perform methylation calling at m-tuples.
+`methpat` (__CITE__) is also capable of performing methylation calling at m-tuples.
 
@@
 This is done by examining the two nucleotides upstream of the cytosine.
-It can be done based on the reference sequence, as is done in `Bismark` and `comethylation`, or from the reads themselves.
+It can be done based on the reference sequence, as is done in `Bismark` and `methtuple`, or from the reads themselves.
 The obvious difficulty with using the reads themselves is if the cytosine occurs at the last or second last position of the read, in which case the context may not be unambiguously determined from the the read alone.
@@
 However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome.
+
+\subsection{A criticism of "identifying methylcytosines"}{sec:criticism_of_methylcytosines}
+
+__TODO: Need a better title for this section__
+
+\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" [supplementary material].
+This idea was also used in subsequent, and similarly high-profile, publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013].
+I believe this idea is an unrequired source of confusion, as I will explain.
+
+[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
+
+It is necessary to use scare quotes to distinguish the term "methylcytosine", as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from methylcytosine, the standard definition meaning a 5-methylcytosine.
+I will first describe the idea and how it differs from methylation calling.
+Then I will describe how it is used to (1) estimate the number of methylcytosines in the sample, and (2) as a kind of screening procedure that determines whether each cytosines will be used in downstream analyses.
+Finally, I will explain why (1) is a bad estimator and why I think (2) is unnecessary or misguided.
+
+A "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
+Thus, the idea of identifying "methylcytosines" is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
+To do so, Lister et al.
+compare each $\beta$-value to a (coverage-specific and sample-specific) cutoff that determines whether the cytosine is classified as a "methylcytosine" or not.
+
+The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
+The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy} and \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
+Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done (__TODO: Discuss with Terry__).
+What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
+
+The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to "error", where "error" is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
+Mathematically, for each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated "error".
+Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
+
+[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
+
+I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure (__CITE__).
+This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
+thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
+
+The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
+It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
+
+Now to why I don't think this is a very useful procedure.
+Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
+These are two different things.
+If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
+A simple example makes this clear.
+
+Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
+Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
+We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
+From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
+
+Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
+For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
+A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
+
+Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
+While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
+For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
+If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
 

### 2014-11-25 Rmarkdown/biology_background.Rmd
Polishing biology background

@@
 title: "Biology background"
-author: "Peter Hickey"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
 
-![Chemical structure of 5-methylcytosine.
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.3\textwidth]{../figures/5-methylcytosine.png}
+\caption{Chemical structure of 5-methylcytosine.
 "5-methylcytosine".
-By Yikrazuul (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg](http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg)\label{fig:5mC}](../figures/5-methylcytosine.png)
+By Yikrazuul (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg](http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg)}
+\label{fig:5mC}
+\end{figure}
 
@@
 
-__UP TO HERE__
+One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next, the _intra-pair distances_ (IPDs).
+__FIGURE__ is the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19).
+We see that approximately $70\%$ of CpGs are within 100 bp of the next CpG.
+__FIGURE__ also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome.
+By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more "close" pairs and more "distant" pairs than we would expect by chance.
 
-- [x] DONE
+__FIGURE__ plots a frequency polygon of the observed IPDs. From this we can see there is a cluster of CpGs with $IPD < 10$, which largely correspond to what are called _CpG islands_ (CGIs).
 
-One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next.
-__FIGURE__ is a kernel density plot of the distances between neighbouring CpGs  and __FIGURE__ is the empirical cumulative distribution function of the distances between neighbouring CpGs for the human reference genome (hg19)[^cpg_distance].
-
-[^cpg_distance]: See __APPENDIX__ for the `R` code used to generate these figures.
-
-As can be seen from __FIGURE__, there are a subset of CpGs that are within __HOW MANY__ bases of the next CpG.
-
-The 20-40% of CpGs that are unmethylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_).
-CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__).
-The classical definition of a CpG island, given by \cite{GardinerGarden:1987we} and used by, for example, the UCSC genome browser, is a region of the genome where:
+Within these CpG islands are the 20-40% of CpGs that are frequently unmethylated in mammalian genomes.
+CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \cite{Wu:2010do}.
+The classical definition of a CpG island, given by \cite{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:
 
@@
 
-The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
+These "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
+CpG island shores, shelves and the open sea are all defined with respect to CpG islands.
+Specifically,
 
-CpG island shores, shelves and the open sea are all defined with respect to CpG islands. Specifically,
-
-- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}.
+- CpG island shores are regions within 2 kb of CpG islands \cite{Irizarry:2008hg}.
 CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
 Shelves have a comparatively low density of CpGs.
-- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
+- CpG island shelves are defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
 - The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \cite{Sandoval:2011dj}.
 
-Other metaphors, these based on methylation levels rather than CpG density, are methylation deserts \cite{Li:2012gg} and CpG canyons \cite{Jeong:2014bla}.
+Other metaphors, these based on methylation levels rather than CpG density, include methylation deserts \cite{Li:2012gg} and CpG canyons \cite{Jeong:2014bla}.
 Because these regions are defined with respect to methylation levels rather than DNA sequence, these are generally identified in a tissue-specific manner.
@@
 
-- [x] DONE
-
-In humans, cytosine methylation in most cell types is found almost exclusively at CpGs.
+In humans, cytosine methylation in most cell types is found almost exclusively at CpGs \cite{Jones:2012jr}.
 There are, however, certain cell types with widespread non-CpG methylation.
-Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H stands for any nucleotide except G (__TODO: cite__).
-The rule-of-thumb is that non-CpG methylation is rare in somatic cells but common in pluripotent cells.
+Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H is the IUPAC code for any nucleotide except G ([http://www.bioinformatics.org/sms2/iupac.html](http://www.bioinformatics.org/sms2/iupac.html)).
+The rule-of-thumb for mammalian genomes is that non-CpG methylation is rare in somatic cells but common in pluripotent cells.
 Of course, there are exceptions to every rule, especially in biology.
@@
 
-A subsequent paper from the same group extended this result when they reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines \cite{Lister:2011kg}.
+A subsequent paper from the same group extended this result.
+\cite{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines .
 Pluripotent cells includes embryonic stem cells along with induced pluripotent stem cell lines.
@@
 Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation.
-This is partly due to sampling bias - commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips (__TODO: ref section on technology__), only measure CpG methylation.
-However, recent technological advances (__TODO: ref section on technology__) mean that cytosine methylation can be routinely assayed regardless of the sequence context.
+This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips, only measure CpG methylation.
+However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context.
+This discussed further in Section sec:dna_methylation_assays.
 
 Non-CpG methylation is very common in other organisms, such as plants.
-For example, \cite{Lister:2008bh} found that $45\%$ of "methylcytosines" (see __SECTION__ for the definition and a criticism) in the widely-studied _Arabidopsis Thaliana_ are at CHG or CHH loci.
+For example, \cite{Lister:2008bh} found that in the widely-studied _Arabidopsis Thaliana_ that $45\%$ of cytosines that displayed statistically significant evidence of methylation are at CHG or CHH loci.
 However, they also found that the level of methylation at non-CpG loci is typically lower than that observed at CpG dinucleotides.
@@
 
-- [x] DONE
-
 Methylation is not the only chemical modification of cytosines, although it is by far the most common.
 These additional modifications may be thought of as 'modifications of a modification' since they extend 5-methylcytosine.
-Listed approximately from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \cite{Plongthongkum:2014jg}.
+Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \cite{Plongthongkum:2014jg}.
 The biological significance of these marks is still being determined, in part because the assays for studying these are still in development and because their relatively scarcity in the genome means that experiments to detect these modifications are more difficult and expensive.
@@
 \cite{Kriaucionis:2009bm} and \cite{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC.
-This suggested a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.
+This suggests a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.
 
@@
 
-- [x] DONE
-
-A common analogy used in describing DNA methylation uses 'writers', 'readers' and 'erasers' \cite{Moore:2013in}.
-Writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
+A common analogy used in describing epigentic marks uses 'writers', 'readers' and 'erasers' \cite{Moore:2013in}.
+In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
 
 In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes.
-These DNMTs are commonly split into two groups, namely the maintainence methyltransferase (thought to be DNMT1) and the _de novo_ methyltransferases (thought to be DNMT3a and DNMT3b)[^DNMT].
-DNA methylation is not preserved by the DNA replication machinery and so it is the role of DNMT1 to restore the methylation pattern on the daughter strand of DNA following DNA replication.
-In contrast, DNMT3a and DNMT3b lay down new methylation marks and are particularly active during development when there are widespread changes in DNA methylation (__TODO: cite__).
-DNMT1 and DNMT3b appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \cite{Li:1992bp} whereas mouse knockouts for DNMT3a are runted but survive for ~4 weeks after birth \cite{Li:1992bp}.
-
-[^DNMT]: DNMT2, now known as TRDMT1, was once thought to be a DNA methyltransferase, however, it was shown to in fact methylate a small RNA and not DNA \cite{Goll:2006kb}.
-Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity.
-Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__TODO: cite__).
+The DNMTs are commonly split into two groups, namely the maintenance methyltransferase and the _de novo_ methyltransferases.
+DNA methylation is not preserved by the DNA replication machinery and so it is the role of the maintainence methyltransferases to restore the methylation pattern on the daughter strand of DNA following DNA replication.
+In mammals, _DNMT1_ is known as the maintenance DNA methyltransferase.
+_DNMT3a_ and _DNMT3b_ are known as the _de nove_ methyltransferases, although these are also required for the maintenance of DNA methylation \cite{Jones:2009ji}.
+Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \cite{Li:1992bp}.
+In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \cite{Li:1992bp}.
 
@@
 Mouse knockouts can be either heterozygous knockouts (one copy still of the gene is still present/active) or homozygous knockouts (both copies of gene absent/inactive).
+
+_DNMT2_, now known as _TRDMT1_, was once thought to be a DNA methyltransferase but was shown to in fact methylate a small RNA and not DNA \cite{Goll:2006kb}.
+Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity.
+Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__TODO: cite__).
 
@@
 
-The removal or erasure of DNA methylation, _demethylation_, is characterised as _passive_ loss or _active_ removal.
+The removal or erasure of DNA methylation, called _demethylation_, is characterised as _passive_ loss or _active_ removal.
 Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division.
 This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions.
-This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of precise targeted demethylation (__TODO: example__) nor the two stages of rapid global demethylation that occur during development \cite{Wu:2014gw}.
+This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of local tissue-specific differences in DNA methylation \cite{Irizarry:2009ky} nor the two stages of rapid global demethylation that occur during development \cite{Wu:2014gw}.
 
@@
 This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ -> _5hmC_ -> _5fC_ -> _5caC_ reaction.
-The removal of _5hmC_, _5fC_ or _5caC_ is biochemically "easier" than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery (__TODO: Figure of TET cycle__).
+The removal of _5hmC_, _5fC_ or _5caC_ is biochemically "easier" than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.
 
 One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
+This is an active area of investigation.
 
 \section{Assays for studying DNA methylation}\label{sec:dna_methylation_assays}
-
-- [x] DONE
 
@@
 
-An exception to this classification scheme are a new class of assay that seek to directly "read" whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA.
-For example, both \cite{Laszlo:2013kf} and \cite{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine was methylated.
-Similarly, __TODO: what does Pacbio and Oxford Nanopore do?__
+An exception to this classification scheme are a new class of assay that seek to directly 'read' whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA.
+For example, both \cite{Laszlo:2013kf} and \cite{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine is methylated.
 
 Almost all assays of DNA methylation measure a "population average" from a pool of hundreds or thousands of cells.
-For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay (see __FIGURE__).
+For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay.
 Hundreds or thousands of cells are required in order to have sufficient material as input for the assay.
-Assays that require only a single cell as input do exist but are still in development and not in widespread use (__TODO: sources and examles__).
+Assays that require only a single cell as input do exist \cite[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.
 
-The _resolution_ and _throughput_ are two key variables when choosing which assay to use for an experiment.
+The _resolution_ and _throughput_ of an assay are two key variables when choosing which to use for an experiment.
 The resolution of an assay is the scale on which DNA methylation can be measured[^resolution].
@@
 The first is per-sample throughput, which is how many measurements of DNA methylation are made per-sample[^throughput1].
-This is typically what people mean when they describe an assay as being "high-throughput" or "low-throughput" and is the definition I use in the title of my thesis.
-Depending on your definition of "high", a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample.
-The second definition of throughput is  related to cost, be it money or time, i.e., "how many samples can I process?".
+This is typically what people mean when they describe an assay as being 'high-throughput' or 'low-throughput' and is the definition I use in the title of my thesis.
+Depending on your definition of 'high', a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample.
+The second definition of throughput is  related to cost, be it money or time, i.e., 'how many samples can I afford to analyse?''.
 
@@
 In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays.
-In particular, I will describe in detail the "gold standard" assay of DNA methylation, _whole-genome bisulfite-sequencing_ (WGBS), that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.
+In particular, I will describe in detail the 'gold standard; assay of DNA methylation, _whole-genome bisulfite-sequencing_ (WGBS), that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.
 
@@
 
-- [x] DONE
-
 _Restriction endonucleases_ are an important technique in molecular biology.
-These enzymes can preferentially "cut/cleave/digest" DNA at particular sequence motifs.
+These enzymes can preferentially 'cut', 'cleave' or 'digest' DNA at or near to particular sequence motifs.
 The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_.
@@
 This can be used to design an assay to infer the methylation state of a DNA fragment.
-
-__TODO: Is it most accurate to say that restriction enzymes "cleave", "cut" or "digest" DNA?__
 
@@
 They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \cite{Laird:2010iv}.
-Restriction enzyme assays have a relatively low resolution (__TODO: reasons and source__) but may also be used as a first-pass to enrich a sample for methylated or unmethylated regions which are then assayed using a higher resolution technique.
+Restriction enzyme assays have a relatively low resolution.
 
 \subsection{Affinity enrichment assays}\label{sec:affinity_enrichment_assays}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
 In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \cite{Shapiro:1970im, Hayatsu:1970jv}.
-The methylated cytosines are said to be "protected" from conversion to uracils.
+The methylated cytosines are said to be 'protected' from conversion to uracils.
 This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq, Clark:1994wc}, which are referred to as bisulfite-conversion assays.
@@
 When bisulfite-treated is amplified by PCR, the uracils are converted to thymines.
-Thus, these assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (__FIGURE__).
+These bisulfite-conversion assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (Figure \ref{fig:bisulfite-conversion}).
 
-Initial experiments based on the sodium bisulfite pre-treatment of DNA used Sanger sequencing of cloned PCR products, a very laborious task, which restricted experiments to studying a limited number of short segments of DNA (__HOW MANY & LONG?__).
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/bisulfite-conversion.pdf}
+\caption{The effect of bisulfite-treatment of DNA.
+The double-stranded DNA is denatured and each strand undergoes bisulfite-treatment.
+Methylated cytosines remain as cytosines while unmethylated cytosines become uraciles.
+These bisulfite-converted DNA strands then undergo PCR amplification which converts the uracils to thymines.
+Note that while there are four possible PCR products, some bisulfite-sequencing protocols do not sample PCR products from the \textit{CTOT} or \textit{CTOB} strands.
+\textit{OT} = original top strand; \textit{OB} = original bottom strand; \textit{CTOT} = complementary to the original top strand; \textit{CTOB} = complementary to the original bottom strand.
+This figure is adapted from \citet{Krueger:2012ks}.}
+\label{fig:bisulfite-conversion}
+\end{figure}
+
+Initial experiments based on the sodium bisulfite pre-treatment of DNA used Sanger sequencing of cloned PCR products, a very laborious task, which restricted experiments to studying a limited number of short segments of DNA.
 Although subsequent enhancements in the automation of Sanger sequencing improved the throughput of these assays, it was never going to be able to deliver a cost-effective, genome-scale assay of DNA methylation.
-The development of hybridisation microarrays provided an alternative way of analysing bisulfite-treated DNA that could provide cheap, genome-wide measurements of DNA methylation(__SOURCE__).
+The development of hybridisation microarrays provided an alternative way of analysing bisulfite-treated DNA that could provide cheap, genome-wide measurements of DNA methylation.
 
@@
 Therefore, an (idealised) way to analyse DNA methylation with a microarray is to hybridise bisulfite-converted DNA to a microarray that contains probes for both the methylated and unmethylated versions of all sequences of interest.
-The relative methylation of each sequence can be inferred from the relative intensities of the "methylated probe" to the "unmethylated probe".
+The relative methylation of each sequence can be inferred from the relative intensities of the 'methylated probe' to the 'unmethylated probe'.
 Such an idealised experiment brushes over many complications including \cite{Laird:2010iv}:
@@
 This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \cite{Stirzaker:2014cu}.
-There are well-established methods and software for analysing Illumina 450k data such as `minfi` \cite{Aryee:2014by}
+There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed in \cite{Dedeurwaerder:2014bt}.
 
@@
 The targeted nature of these assays are their obvious advantage and disadvantage - only a subset of the genome needs be sequenced but you only obtain information about methylation in those regions.
-Depending on the experiment this tradeoff may be worthwhile, and these assays have been successfully used in a number of studies (__CITE__).
+Depending on the experiment, this tradeoff may be worthwhile, and these assays have been successfully used in a number of studies such as the BLUEPRINT Epigenome project ([http://www.blueprint-epigenome.eu/](http://www.blueprint-epigenome.eu/)) and the NIH Roadmap Epigenomics Mapping Consortium ([http://www.roadmapepigenomics.org/](http://www.roadmapepigenomics.org))
 
@@
 One final class of bisulfite-conversion assays does not use microarrays or high-throughput sequencing.
-Sequenon's EpiTYPER \cite{Ehrich:2006tt} uses mass spectrometry to analyse DNA methylation from bisulfite-converted DNA.
+Sequenom's EpiTYPER \cite{Ehrich:2006tt} uses mass spectrometry to analyse DNA methylation from bisulfite-converted DNA.
 This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and may be used to validate findings discovered using other platforms \cite{Laird:2010iv}.
@@
 
-- [x] DONE
-
-Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation since cytosine methylation can be detected at single-base resolution \cite{Stirzaker:2014cu}.
-However, almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1-5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
-The minimal amount of DNA is being reduced with each technological advance.
-Furthermore, single-cell bisulfite-sequencing assays have recently been reported \cite{Guo:2013ih, Smallwood:2014kn}[^not_genome_wide].
-
-[^not_genome_wide]: These are not yet proper genome-wide assays.
-The technique of \cite{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
-And the technique of \cite{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \cite{Smallwood:2014kn}.
-
-A key advantage of bisulfite-conversion assays over enzyme digest and affinity enrichment assays is that it provide base pair resolution of DNA methylation.
+Bisulfite-conversion assays are considered the 'gold standard' for studying DNA methylation since cytosine methylation can be detected at single-base resolution \cite{Stirzaker:2014cu}.
 In fact, single-molecule, base pair resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].
@@
 The signal from a microarray-based experiment a sample-wide average since, for a given locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.
+
+Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1-5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
+However, the minimal amount of DNA is being reduced with each technological advance.
+Recently, single-cell bisulfite-sequencing assays have been reported \cite{Guo:2013ih, Smallwood:2014kn}, although they are not yet a 'commercial' sequencing assay.
+Furthermore, these are not yet proper genome-wide assays.
+The technique of \cite{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
+And the technique of \cite{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \cite{Smallwood:2014kn}.
 
@@
 Only 5mC is protected from conversion during the bisulfite-treatment, which effectively means that only 5mC remains to be detected at the sequencing stage.
-The level of 5hmC can be estimated by performing traditional bisulfite-sequencing and then "subtracting" the oxBS-seq signal (5mC) from the bisulfite-sequencing signal (5mC + 5hmC).
+The level of 5hmC can be estimated by performing traditional bisulfite-sequencing and then 'subtracting' the oxBS-seq signal (5mC) from the bisulfite-sequencing signal (5mC + 5hmC).
 
@@
 Only 5gmC is protected from TET oxidation, which effectively means that only 5hmC remains to be detected at the sequencing stage.
-The level of 5mC can be estimated by performing traditional bisulfite-sequencing and then "subtracting" the TAB-seq signal (5hmC) from the bisulfite-sequencing signal (5mC + 5hmC).
+The level of 5mC can be estimated by performing traditional bisulfite-sequencing and then 'subtracting' the TAB-seq signal (5hmC) from the bisulfite-sequencing signal (5mC + 5hmC).
 
@@
 When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \cite{Liu:2012ge}.
-__FIGURE__ illustrates such an example.
+Figure \ref{fig:bisulfite_CT_SNP.pdf} illustrates such an example.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
+\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inferrence about the methylation state of the sample's genome.
+The locus in orange is a cytosine in the refererence genome but a thymine in the sample's genome.
+Because the read is compared against the reference genome it is incorrectly inferred to be an unmethylated cytosine.
+The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome.
+Because the read is compared against the reference genome it is incorrectly inferred to be a thymine.}
+\label{fig:bisulfite_CT_SNP.pdf}
+\end{figure}
 
@@
 _PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \cite{Warnecke:1997eh}.
-_Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC.
+_Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment.
 Finally, treating DNA with sodium bisulfite damages the DNA, resulting in the fragmentation of long molecules.
@@
 \subsection{DNA kinetics assays}\label{sec:dna_kinetics_assays}
-
-- [x] DONE
 
@@
 It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.
+
+\section{Summary}
+
+DNA methylation is an epigenetic modification with important roles in many biological systems.
+In mammals, CpGs are frequently methylated while non-CpG sites are less frequently methylated.
+Most unmethylated CpGs are found in CpG islands, which are important regulatory elements in the genome.
+
+Whole-genome bisulfite-sequencing provides single-nucleotide resolution data and is the gold-standard assay for studying DNA methylation.
+Methods for analysing whole-genome bisulfite-sequencing data are the focus of my thesis.
+The following chapter describes the bioinformatics analysis of a bisulfite-sequencing experiment, including its many statistical and computational challenges.

### 2014-11-25 Rmarkdown/appendix.Rmd
Updated appendix

@@
 
-## CpGs are underrepresented in the human genome
-
-This short R script demonstrates that CpGs are underrepresented in the human reference genome (build hg19).
+## CpGs are underrepresented in the human genome (hg19)
 
@@
 ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) + 
-  geom_point(size = 5) + ylim(c(0, 1.5)) + 
+  geom_point(size = 5) + coord_cartestian(ylim = c(0, 1.5)) + 
   ggtitle("CpGs are underrepresented in human reference genome (hg19)") + 
@@
   geom_hline(yintercept = 1, linetype = 2)
+```
+
+## Distances between CpGs in the human reference genome (hg19)
+
+I extract the co-ordinates of all CpGs in the human reference genome (hg19) and compute the intra-pair distances (IPD) between adjacent CpGs.
+I then plot the empirical cumulative distribution function of IPD.
+These are compared to the distances we would expect if CpGs were uniformly distributed across the genome.
+
+I only use the autosomes, sex chromosomes and mitochondrial chromosome in all analyses, that is, I do not consider the "random", "unplaced" and "alternate haplotype" contigs in the reference genomes.
+
+```{r cpg_ipds, cache = TRUE}
+library(BSgenome.Hsapiens.UCSC.hg19)
+set.seed(666)
+
+# Observed IPDs
+cpgs <- bsapply(new("BSParams", X = Hsapiens, FUN = matchPattern, 
+                    exclude = c("rand", "Un", "hap"), simplify = TRUE),
+                pattern = "CG")
+d <- unlist(lapply(cpgs, start), use.names = FALSE)
+ipd <- diff(d)
+# Drop negative distances, which correspond to CpGs on different chromosomes
+ipd <- ipd[ipd > 0]
+
+# Expected IPDs if CpGs were uniformly distributed on genome
+d_sim <- bsapply(new("BSParams", X = Hsapiens, FUN = function(x) {
+  cpg_freq <- countPattern("CG", x)
+  sort(sample(x = length(x), size = cpg_freq, replace = FALSE))
+}, exclude = c("rand", "Un", "hap"), simplify = TRUE))
+ipd_sim <- diff(unlist(d_sim, use.names = FALSE))
+# Drop negative distances, which correspond to CpGs on different chromosomes
+ipd_sim <- ipd_sim[ipd_sim > 0]
+
+# Drop negative values, which are due to CpGs being on different chromosomes
+ipd_df <- data.frame(IPD = c(ipd, ipd_sim), 
+                  Source = c(rep("Observed", length(ipd)), 
+                             rep("Expected", length(ipd_sim))))
+```
+
+### Empirical cumulative distribution plot
+
+```{r cpg_ipds_cdf}
+library(ggplot2)
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
+
+ggplot(aes(x = IPD, colour = Source), data = ipd_df) + stat_ecdf() + thesis_theme + 
+  coord_cartesian(xlim = c(0, 1000)) + 
+  xlab("Distance between CpGs (bp)\n(Axis truncated at 1000)") + 
+  ggtitle("Human reference genome (hg19)")
+```
+
+### Histogram
+
+```{r cpg_ipds_hist}
+library(ggplot2)
+library(scales)
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
+
+ipd_table <- table(ipd)
+ipd_sim_table <- table(ipd_sim)
+ipd_table_df <- data.frame(IPD = c(as.numeric(names(ipd_table)), 
+                                   as.numeric(names(ipd_sim_table))), 
+                           Count = c(ipd_table, ipd_sim_table), 
+                           Source = c(rep("Observed", length(ipd_table)), 
+                                      rep("Expected", length(ipd_sim_table))))
+
+ggplot(aes(x = IPD, y = Count), 
+       data = subset(ipd_table_df, Source == "Observed")) + 
+  geom_line() + coord_cartesian(xlim = c(0, 300)) + 
+  xlab("Distance between CpGs (bp)\n (Axis truncated at 300)") + 
+  ggtitle("Human reference genome (hg19)") + 
+  scale_y_continuous(labels = comma)
 ```

### 2014-11-26 Rmarkdown/a_statistical_model_of_methlyC-seq.Rmd
Split WGBS analysis into “bioinformatics” and “statistics”

@@
 ---
-title: "A statistical framework for analysing bisulfite-sequencing data"
+title: "Statistical analysis of whole-genome bisulfite-sequencing data"
 author: "Peter Hickey"
@@
 
+\subsection{A criticism of ``identifying methylcytosines''}{sec:criticism_of_methylcytosines}
+
+\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
+This idea was also used in subsequent publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013] with similarly high profiles.
+I believe that this definition of a "methylcytosine" is an unnecessary source of confusion, as I will explain.
+
+[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
+
+It is necessary to use scare quotes to distinguish the term "methylcytosine", as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from methylcytosine, the standard definition meaning a 5-methylcytosine.
+I will first describe the idea and how it differs from methylation calling.
+Then I will describe how it is used to (1) estimate the number of methylcytosines in the sample, and (2) as a kind of screening procedure that determines whether each cytosines will be used in downstream analyses.
+Finally, I will explain why (1) is a bad estimator and why I think (2) is unnecessary or misguided.
+
+A "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
+Thus, the idea of identifying "methylcytosines" is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
+To do so, \citeauthor{Lister:2008bh} compare each the proportion of methylated reads, the $\beta$-value, to a coverage-specific (and sample-specific) cutoff that determines whether the cytosine is classified as a "methylcytosine" or not.
+
+The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
+The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy} and \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
+Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done.
+What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
+
+The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
+Mathematically, for each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated "error".
+Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
+
+[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
+
+I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure (__CITE__).
+This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
+thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
+
+The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
+It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
+
+Now to why I don't think this is a very useful procedure.
+Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
+These are two different things.
+If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
+A simple example makes this clear.
+
+Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
+Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
+We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
+From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
+
+Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
+For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
+A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
+
+Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
+While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
+For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
+If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
+
 \section{Statistical properties of $\beta$-values}\label{sec:beta}
@@
 
+\section{Dump from previous chapter}
+
+\section{Identifying differential methylation}\label{sec:identifying_differential_methylation}
+
+Identifying differential methylation means to discover sites (differentially methylated cytosines or _DMC_s) or regions (differentially methylated regions or _DMR_s) in the genome that have different levels of methylation between two or more conditions.
+Much of the methodology for identifying differential methylation has focused on two-group experiments and a somewhat restricted definition: differential methylation is the difference in __average__ methylation between __two__ groups.
+This is analogous to identifying differential gene expression, which focuses on identifying genes that have different __average__ expression levels between two groups.
+
+While bisulfite-sequencing experiments with multiple groups have been performed and analysed \citep[e.g.,]{Lister:2009hy, Hansen:2011gu, Hansen:2013eo}), these have typically been done using a series of pair-wise comparisons or by comparing each sample in turn against some 'baseline' sample.
+This is because the analysis of a pair-wise comparison is relatively simple and because it is also generally easier to interpret a two-group comparison than a multi-group comparison.
+However, as the number of groups increases it becomes more difficult to interpret multiple pair-wise comparisons.
+
+Like when trying to answer any scientific question, the experimental design is very important in order to reliably infer differential methylation.
+Identifying DMCs is relatively straightforward statistical problem, since it amounts to the well-studied problem of testing the difference of means.
+However, identifying DMRs, which may or may not build upon an initial scan for DMCs, is a far more challenging statistical problem owing to the high spatial dependence of methylation at neighbouring cytosines (discussed in Chapter \ref{chap:co-methylation}) and the uneven spacing of cytosines throughout the genome.
+
+\subsection{Experimental design}\label{sec:experimental_design}
+
+In any experiment of differential methylation we want the DMCs and DMRs to be both _biologically_ and _statistically_ significant; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
+Key to ensuring biological relevance is good experimental design, such as the use of _replicates_ in each experimental group.
+A distinction is often made between _technical replicates_ and _biological replicates_.
+Briefly, biological replicates are experimental units that all undergo the same treatment and are used to estimate the within-group variability of the treatment.
+Technical replicates are repeated measurements of the same experimental unit, perhaps with slight variations in the sample preparation, and are used to estimate the variability of the sample preparation and measurement process.
+
+As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one.
+For example, \cite{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine).
+I contend that these are better defined as technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].
+
+[^cell_passaging]: In the case of IMR90 cell line, the first replicate, IMR90_r1, underwent 4 cell passages and the second replicate, IMR90_r2, underwent 5 cell passages.
+In the case of the H1 cell line, the first replicate, H1_r1, underwent 25 passages in the first media and 9 passages in the second media, and the second replicate, H1_r2, underwent 27 passages in the first media and 5 passages in the second media.
+
+Unfortunately, early experiments with whole-genome bisulfite-sequencing frequently had no replicates of any kind, as is the case of the H9 and IMR90-iPSC cell lines in \cite{Lister:2009hy}.
+Moreover, even when replicates were performed they were frequently pooled pooled prior to analysis, thus ignoring all within-group variability.
+
+Ideally, variability between technical replicates should be orthogonal to the biological question, but this rarely occurs.
+Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.
+Again, this emphasises the importance of good experimental design.
+
+\subsection{DMCs}\label{sec:dmcs}
+
+There have been reports of differential methylation at individual CpGs resulting in a phenotypic difference \cite{Furst:2012gh}.
+However, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from WGBS experiments with sample sizes you can count on one or two hands and average sequencing depth of $10-30\times$.
+
+Identifying DMCs boils down to identifying differences in means.
+As such, it can be readily framed as a 'stand-alone' test, such as a t-test, or cast into a regression framework.
+There is an enormous body of statistical literature on testing for differences in means.
+
+Regardless of the test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty.
+Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data.
+but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments (e.g., approximately 1 million tests in genome-wide association studies and 20,000 tests in studies of differential gene expression).
+
+However, when identifying DMCs, the _effective_ number of tests is generally fewer due to the high degree of dependence of methylation at neighbouring methylation loci, which ensures a high degree of dependence amongst the tests.
+
+__TODO: What techniques can be used to do FDR, multiple hypothesis testing? Are standard tools applicable in such high-dependence settings?__
+
+\subsubsection{One-group tests}\label{sec:one_group_dmcs}
+
+The one-group WGBS experiment is uncommon and the $A^{vy}$ experiment described in Chapter \ref{chap:Avy} is the only example that I know of.
+In a one-group experiment a DMC is one where at least one of the samples has a different methylation level to the group average.
+Mathematically, we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$, where $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
+This can be tested in a number of ways as I will demonstrate.
+
+The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown below (__TODO: Prettify table with `booktabs`__)
+
+\begin{table}[h]
+\centering
+\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
+\label{my-label}
+\begin{tabular}{c c c}
+\hline
+& m          & u          \\ \hline
+$Sample_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
+$\vdots$     & $\vdots$   & $\vdots$   \\
+$Sample_{n}$ & $m_{i, n}$ & $u_{i, n}$ \\
+\end{tabular}
+\end{table}
+
+Such contingency tables can be analysed in many different ways. Here I apply a few methods to highlight their similarities and differences. Firstly, here are some example data for a single CpG:
+
+```{r}
+x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
+names(dimnames(x)) <- c("sample", "meth_state")
+x
+```
+
+We see that while `sample_1, ..., sample_4` are nearly completely methylated at this CpG, `sample_5` has approximately $50\%$ methylation, which may be indicative of allele-specific methylation.
+
+In my analysis of the $A^{vy}$ experiment I used the `prop.test` function in R. As documented in the help page for `prop.test`:
+
+> `prop.test` can be used for testing the null that the proportions (probabilities of success) in several groups are the same
+
+in R to test the equality of the proportions ($\beta$-values) across the five mice.
+
+```{r}
+prop.test(x)
+```
+
+As expected, this test returns a very small P-value, `r print(format.pval(prop.test(x)$p.value), quote = FALSE)`.
+
+What might not be clear from its name is that `prop.test(x)` is just performing Pearson's $\chi^{2}$-test of the null hypothesis that the joint distribution of the cell counts in the $2 \times n$ contingency table is the product of the row and column marginals.
+This can be seen by examining the source code or by comparing the output to `chisq.test(x)`: (__TODO: Check by hand__)
+
+```{r}
+chisq.test(x)
+```
+
+Instead of using the "stand-alone" tests, `prop.test` and `chisq.test`, we could explicitly frame this as a log-linear model and test the goodness-of-fit of the "no interaction" model using either the likelihood ratio statistic (`lrt`) or the Pearson's $\chi^{2}$ test statistic (`pearson`).
+Formally, the regression model is (__TODO: Write down the regression model__):
+
+This model can be fit using the `loglin` function in R or using the arguably more user-friendly `loglm` (from the `MASS` package) or `glm` functions:
+
+```{r}
+# Using loglin
+loglin(x, margin = list(1, 2))
+# Using loglm
+library(MASS)
+loglm(formula =  ~ sample + meth_state, data = x)
+# Using glm (1)
+glm(x ~ 1, family = binomial)
+# Using glm (2)
+glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2])
+# TODO: Write sentence summarising output.
+```
+
+A key advantage of using regression is the option to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
+__TODO: How to do this using `loglin`?__
+
+More complex regression analyses are possible, such as empirical Bayes or fully Bayesian models.
+I describe these in more detail in discussing two-group and multi-group experiments, but they may also be applied to one-group experiments.
+
+A proper software implementation needs to take care of issues such as incomplete tables and non-convergence of the iterative proportional scaling algorithm used by `loglin`.
+Speed is also important when performing this many tests and the regression based approaches are much slower, at least when naively applied, due to additional checks they make of the data:
+
+```{r}
+# Basic benchmarking
+library(microbenchmark)
+print(microbenchmark(prop.test(x), chisq.test(x), loglin(x, margin = list("sample", "meth_state")), loglm(formula = ~ sample + meth_state, data = x), glm(x ~ 1, family = binomial), glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2]), glm.fit(x = matrix(, nrow(x), 0L), y = x, family = binomial()), times = 100), order = "median")
+```
+
+\subsubsection{Two-group experiments}\label{sec:two_group_dmcs}
+
+In a two group experiment, for each methylation locus we are testing the hypothesis that the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, is different to that in the second group, $\bar{\beta_{i}^{(2)}}$.
+
+If there are no replicates in either group[^2x2] then we have a $2 \times 2$ contingency table:
+
+[^2x2]: I'd argue that this is a two-sample experiment rather than a two-group experiment since each group has $n = 1$.
+
+(__TODO: Prettify table with `booktabs`__)
+
+\begin{table}[h]
+\centering
+\caption{$2 \times 2$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a two-group experiment with no replicates in either group.}
+\label{my-label}
+\begin{tabular}{c c c}
+\hline
+& m          & u          \\ \hline
+$Group_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
+$Group_{2}$ & $m_{i, 2}$ & $u_{i, 2}$ \\
+\end{tabular}
+\end{table}
+
+This $2 \times 2$ table is then tested for independence of the rows and columns.
+As the sample size for each table is often small, Fisher's exact test is generally preferable since, unlike Pearson's $\chi^{2}$ test, the P-values are _exact_, meaning they do not rely on large-sample approximations (__CITE: Agresti__).
+Fisher's exact test is a _conditional_ test because the null distribution (a hypergeometric distribution) is derived by conditioning on both the marginal totals.
+__CITE: Agresti__ discusses the arguments for and against conditional and unconditional tests of independence in the $2 \times 2$ table.
+
+The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
+Unfortunately, some authors have used Fisher's exact test to analyse experiments including replicates (__TODO: Highlight papers that have inappropriately used Fisher's exact test; does Lister 2013 inappropriately pool then test?).
+This is incorrect because it effectively aggregates all the counts within each group and so ignores all within-group biological variability \citep{Hansen:2012gr}.
+
+Instead, a test that incorporates the within-group variability should be used.
+For example, we could use a t-test to compare the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$.
+Several methods have been proposed for testing for DMCs.
+These methods differ considerably in how they model and transform the data as well as what test statistic is used to identify DMCs.
+
+\paragraph{Transforming}\label{sec:transforming}
+
+Smoothing of the $\beta$-values has been used advocated by several authors to remove variation due to low sequencing coverage and to exploit the spatial correlation of DNA methylation \citep{Hansen:2011gu, Hebestreit:2013ko, Gokhman:2014c}.
+Both `BSmooth` \citep{Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} smooth the $\beta$-values using a binomial local likelihood smoother.
+\citet{Lister:2011kg, Gokhman:2014cp} used a simpler moving average in windows along the genome.
+
+\paragraph{Modeling}\label{sec:modeling}
+
+`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
+Broadly, under the Beta-Binomial model the "true" methylation level at the $i^{th}$ methylation locus in the $k^{th}$ group is modeled as a Beta random variable, $B_{i, k} = Beta(\mu_{i, k}, \theta_{i, k})$.
+The observed number of methylated reads at the $i^{th}$ methylation locus, in the $j^{th}$ sample is modeled as a Binomial random variable, $M_{i, j, k} | B_{i, k}, n_{i, j, k} = Binomial(n_{i, j, k}, B_{i, k}), which is conditional on the unobserved $B_{i, k}$ and the observed sequencing coverage, $n_{i, j, k}$.
+
+\paragraph{Estimating and testing}\label{sec:estimating_and_testing}
+
+`DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
+The test used to identify DMCs is variously a Wald test (`BiSeq`, `DSS`), likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).
+
+\subsubsection{Multi-group experiments}\label{sec:multigroup_dmcs}
+
+\subsection{DMRs}
+
+There are two very different strategies for identifying differentially methylated regions:
+
+1.
+Use regions that are defined _a priori_, which are then tested for differential methylation.
+Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g.
+3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
+2. Use regions that are defined based on the data, which are then tested for differential methylation. Valid statistical inference of such regions is __very challenging__.
+
+\subsubsection{Using _a priori_ regions}\label{sec:a_prior_regions}
+
+The methylation level is summarised at the region level and then compared across groups. For example, the average of the $\beta$-values across the region may be compared between groups using a t-test.
+
+__TODO: Describe the above mathematically__
+
+This is qualitatively similar to identifying differentially methylated cytosines.
+
+__PROS__
+
+* Simple to implement
+* Valid statistical properties
+
+__CONS__
+
+* Doesn't account for differential CpG density across regions
+* What is the right "unit" - the CGI, the promoter, a tile???
+* Doesn't account for spatial correlation of methylation
+
+\subsubsection{Using data-driven regions}\label{sec:data_driven_regions}
+
+__PROS__
+
+* Can discover the right "unit".
+* Might account for spatial correlation of methylation
+
+__CONS__
+
+* Loss of statistical properties in testing, e.g. FDR control
+* Bloody hard
+
+\section{Other downstream analyses}\label{sec:other_downstream_analyses}
+
+Bisulfite-sequencing data are used to address a variety of questions apart from identifying differential methylation.
+Not all these analyses require statistical inference.
+For example, if we are interested in knowing whether the promoter of a particular gene is methylated in our sample then it may be sufficient to simply plot the level of methylation at each CpG in the region.
+
+However, once we move to experiments with multiple samples, multiple regions and more complex questions then statistical methods are generally required to distinguish the signal from the noise.
+In this section I briefly describe other
+
+__TODO: Any other important downstream analyses worth mentioning?__
+
+\subsection{Variably methylated regions and differential variability}\label{sec:vmrs}
+
+\subsection{Allele-specific methylation}\label{sec:asm}
+
+\subsection{Methylation entropy}\label{sec:methylation_entropy}
+
+\subsection{Epipolymorphism}\label{sec:epipolymorphism}
+
+\subsection{Epiallele detection}\label{sec:epialleles}
+
+\subsection{Integrating DNA methylation data with other genomes data}\label{sec:data_integration}
+
+\subsection{General computational considerations}
+
+- Using integers instead of floats/doubles
+
 \section{General TODOs}
@@
 * Discuss distribution of coverage? Perhaps in context of sequencing bias?
+* See https://github.com/brentp/clustermodel/README.md for a nice, brief summary of the main DMR calling approaches.

### 2014-11-26 Rmarkdown/wgbs_analysis.Rmd
Split WGBS analysis into “bioinformatics” and “statistics”

@@
 ---
-title: "Bioinformatics analysis of bisulfite-sequencing data"
-author: "Peter Hickey"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+title: "Bioinformatics analysis of whole-genome bisulfite-sequencing data"
 output:
@@
 
-This chapter outlines the steps in a bioinformatics analysis of a whole-genome bisulfite-sequencing dataset, with a focus on experiments studying _differential methylation_.
-Throughout this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay.
-All data used in my thesis were generated using this protocol (__CHECK__).
-Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some additional tweaks.
-
-There are four fundamental steps in the bioinformatics analysis:
+This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges.
+There are four fundamental steps in the analysis of whole-genome bisulfite-sequencing data:
 
@@
 3. Methylation calling
-4. Identifying differential methylation
+4. Downstream analyses
 
-__FIGURE__ provides a more detailed view of these four steps (__Figure should be a flow chart with the various sub-steps included, e.g. Inference = DMC, DMR, ASM, meth-SNP detection, etc.__)
+Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data.
+Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq.
+There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis.
 
-Steps 1, 2 and 4 will be familiar to anyone who analyses high-thoughput sequencing data, but each requires a "twist" to work with bisulfite-sequencing data.
-Step 3 is obviously unique to assays of DNA methylation but there are similarities to variant calling from DNA-seq.
+This chapter focuses on steps 1-3 and concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data.
+The following chapter addresses step 4.
 
-In this chapter I describe steps 1-2 in broad terms.
-An excellent review of these fundamental steps, including comparisons of different software, is given by \citet{Krueger:2012ks}.
-My thesis has focused on steps 3 and 4 of this process.
-In this chapter I review previous work on steps 3-4 and defer my contributions on these topics to subsequent chapters.
-While this chapter focuses on analysing bisulfite-sequencing data, in particular whole-genome bisulfite-sequencing data, some parts __SECTION Cellular heterogeneity, Inference, OTHERS???__ will also be applicable to methylation microarrays (__CHECK AFTER I'VE WRITTEN IT__).
+In this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay.
+All data used in my thesis were generated using this protocol.
+Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.
 
@@
 
-The `FastQC` software (__CITE__) is a very useful tool for performing this first step.
+The `FastQC` software is a very useful tool for performing this first step ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)).
 It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination.
 `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data.
-This means that some of its output is not so useful, or at least must be cautiously interpreted, for bisulfite-sequencing data.
-For example, `FastQC` will report a __warning/error__ if the GC-percentage, the percentage of read bases that are guanine or cytosine, is __less than what value__.
-Due to the bisulfite-treatment, bisulfite-sequencing data will naturally have a very low percentage of cytosine bases (__around what%__) and therefore a low GC-percentage.
-Therefore, a low GC-percentage is no cause for concern.
+This means that some of its output must be interpreted with caution for bisulfite-sequencing data.
+For example, `FastQC` will report a warning (resp.
+error) if the relative frequency of the four nucleotides differ by more than $10\%$ (resp.
+$20\%$).
+As noted in the `FastQC` documentation, some a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.
 
 More important is the identification and removal of contaminating sequences.
-FastQC will screen a subset of the reads against a list of known, common contaminants.
-Examples of such contaminants include the sequencing adapters, PCR primers and __another example__.
-
+`FastQC` will screen a subset of the reads against a list of known, common contaminants such as adapter sequences.
 Illumina adapter sequences are ligated to each DNA molecule in the library in order to perform Illumina sequencing.
-The sequencer can "read into" the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those created in WGBS libraries.
-This means that some reads are a chimera containing sequence of interest (from the sample) and junk (from the adapters).
-This junk needs to be removed for two reasons:
+The sequencer can 'read into' the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those created in bisulfite-sequencing libraries.
+This means that some reads are a chimera that contain the sequence of interest (from the sample) and junk (from the adapters).
+This contamination needs to be removed for two reasons:
 
 1. Reads containing adapter contamination will generally not map to the reference genome, meaning these reads go to waste.
-2. If they do map, then this will result incorrect inferences - the garbage in, garbage out maxim.
+2. If they do map, then this will result incorrect inferences - the 'garbage in, garbage out' maxim.
 
-Using a tool such as `Trim Galore!` (__CITE__) or `trimmomatic` (__CITE__), the reads can be _trimmed_ to remove these contaminants.
+Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \cite{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants.
 Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.
@@
 Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.
+Several review articles have tried to summarise and compare the various approaches \cite{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
 
@@
 
-1. "Methylation-aware" mismatch penalties.
+1. Methylation-aware mismatch penalties.
 2. _In silico_ bisulfite-conversion of reads and reference genomes.
 
-While _methylation-aware_ mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads (__CITE FELIX__).
+While methylation-aware mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads \cite{Krueger:2012ks}.
 This biases downstream inference and means that these mappers are generally less popular.
@@
 They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines.
-The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as Bowtie2 (__CITE__) or BWA (__CITE__).
-This process is illustrated in __FIGURE__.
+The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as `Bowtie1` \cite{Langmead:2009fv}, `Bowtie2` \cite{Langmead:2012jh} or `bwa`  \cite{Li:2009fi, Li:2010bl}[^non_directional].
 
@@
 
-Depending on the exact settings used, the mapper reports the "best" location of each read with respect to the two reference genomes.
+[^non_directional]: If the data were generated using a non-directional protocol, then each read of the CT-read and the GA-read are mapped to both of the CT-genome and GA-genome, resulting in four mapping steps per read.
+
+Depending on the exact settings used, the mapper reports the 'best' location of each read with respect to the two reference genomes.
 It reports the original sequence of the read in the output file so that the methylation status of each position can be inferred by comparing it to the corresponding reference sequence.
 
-_In silico_ bisulfite-conversion mappers avoid the bias inherent in the "methylation-aware" mappers because all reads, regardless of methylation status, "look the same" to the mapper.
-However, they do suffer from a slight loss in mapping efficiency (__CITE FELIX__).
+_In silico_ bisulfite-conversion mappers avoid the bias inherent in the methylation-aware mappers because all reads, regardless of methylation status, 'look the same' to the mapper.
+However, they do suffer from a slight loss in mapping efficiency \cite{Krueger:2012ks}.
 
-__TABLE__ lists some popular bisulfite-sequencing read mappers with the underlying mapping software listed in brackets:
+Table \ref{tab:bs-seq_mappers} list some popular bisulfite-sequencing read mappers, which have been selected to highlight the variety of back-ends used by these software.
 
-* Bismark (Bowtie1 or Bowtie2, __CITE__)
-* BWA-meth (BWA, __CITE__)
-* BSMAP (SOAP, __CITE__)
-* __OTHERS__
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+Name      & Reference                                            & Underlying mapping software \\ \midrule
+Bismark   & \cite{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
+bwa-meth  & \cite{Pedersen:2014wt}                             & bwa-mem                     \\
+BSMAP     & \cite{Xi:2009bd}                                   & SOAP                        \\
+Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
+\end{tabular}
+\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of back-ends used by these software.}
+\label{tab:bs-seq_mappers}
+\end{table}
 
-Each of these aligners can report the output in the standard `SAM` format (__CITE__), although each mapper does so in a slightly different way, which it difficult to develop downstream analysis tools.
+Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \cite{Li:2009ka}.
+However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way.
+This makes it difficult to develop downstream analysis tools.
 
@@
 The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data.
-Repetitive sequences, such as LINEs and SINEs, are frequently methylated and so are of interest to researchers studying DNA methylation.
-The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__CHECK IN MY DATA AND REPORT COVERAGE__).
+Repetitive sequences, such as LINEs and SINEs, are frequently methylated to prevent transcription and so are of interest to researchers studying DNA methylation.
+The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__TODO: CHECK IN MY DATA AND REPORT COVERAGE__).
 
@@
 PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work.
-Unfortunately, it can introduce biases into the library that result in some molecules being over- or under-represented compared to their "true" frequency.
+Unfortunately, it can introduce biases into the library that result in some molecules being over-represented or under-represented compared to their true frequency.
 This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome.
-These multiply-sequenced fragments are called _PCR duplicates_.
+These multiple-sequenced fragments are called _PCR duplicates_.
 
-In bisulfite-sequencing data, PCR duplicates containing methylation locus $i$ result in $\beta_{i}$ being a biased estimate of $B_{i}$.
+In bisulfite-sequencing data, PCR duplicates containing the a methylation locus can result in a biased estimate of the methylation level at that locus.
 This is because the sequenced reads do not accurately represent the true methylation levels of the sample.
@@
 Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate.
-However, it is relatively easy to identify __possible__ PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]).
-Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` library (__CITE__), the `rmdup` function that is a part of the `SAMtools` library (__CITE__) and `SAMBLASTER` \citep{Faust:2014hf}.
+However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]).
+Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \cite{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
 These software use mapped reads from a `SAM/BAM` file as input.
@@
 
-[^pcr_dup]: The distinction between possible PCR duplicates and true PCR duplicates is rarely made, probably because the phrase is so clunky.
-Possible PCR duplicates are almost always referred to as PCR duplicates with the implicit assumption that the reader is aware that these very likely include falst positive calls.
+[^pcr_dup]: The distinction between suspected PCR duplicates and true PCR duplicates is rarely made, probably because the phrase is so clunky.
+Suspected PCR duplicates are almost always referred to as PCR duplicates with the implicit assumption that the reader is aware that these very likely include false positive calls.
 Consistent with the literature, I will use the term PCR duplicates when I refer to reads identified as PCR duplicates by some software.
@@
 It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair.
-"Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15 (Source: [http://sourceforge.net/apps/mediawiki/picard/index.php?title=Main_Page#Q:_How_does_MarkDuplicates_work.3F](http://sourceforge.net/apps/mediawiki/picard/index.php?title=Main_Page#Q:_How_does_MarkDuplicates_work.3F)).
+"Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15 ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).
 
 Using these software will result in false positive calls because reads may satisfy this (or similar) criteria yet come from separate DNA fragments in the original sample.
-The false positive rate is a particular problem when a small "genome", which might be a targetted region of a genome such as exons or CGIs, is sequenced at high coverage.
-This can be thought of as an example of the pigeonhole princple, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).
+The false positive rate is a particular problem when a small 'genome' is sequenced at high coverage, where the 'genome' might be a targetted region of a genome such as exons or CpG islands.
+This can be thought of as an example of the pigeonhole principle, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).
 
-We could make this mathematically more precise, but it doesn't give us a simple answer to the question, "should we remove possible PCR duplicates from bisulfite-sequencing data?".
-The unsatisfactory answer is, "it depends".
-A rough guide is that provided the average/median sequencing coverage of the "genome" is less than the fragment length[^pcr_fragment_length] then we expect few false positive calls.
+We could make this mathematically more precise, but it doesn't give us a simple answer to the question, 'should we remove possible PCR duplicates from bisulfite-sequencing data?'.
+The unsatisfactory answer is, 'it depends'.
+A rule of thumb is that provided the average or median sequencing coverage of the 'genome' is less than the fragment length[^pcr_fragment_length] then we expect few false positive calls.
 
-[^pcr_fragment_length]: For single-end data, the "fragment length" in these calculations is the read length.
+[^pcr_fragment_length]: For single-end data, the 'fragment length' in these calculations is the read length.
 
@@
 \citet{Hansen:2011gu} found that this is not the case and that in fact there is considerable bias towards the start (5') and end (3') of reads.
-They called this bias the _M-bias_.
+They called this _M-bias_.
 
-M-bias can be identified by plotting the read-position methylation level ($rpml$), which is the proportion of reads that are methylated at each read-position, as a function of read-position (see __FIGURE__).
-These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for `read_1` and `read_2`.
+M-bias can be identified by plotting the read-position methylation level ($rpml$), which is the proportion of reads that are methylated at each read-position, as a function of read-position.
+These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for _read 1_ and _read 2_.
 If there is no M-bias then this plot should be a horizontal line.
-A "bend" or "spike" in this plot is evidence of M-bias.
-Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same height for `read_1` and `read_2`, although we see this is often not quite the case (see __FIGURE__).
+A 'bend' or 'spike' in this plot is evidence of M-bias.
+Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same height for `read_1` and `read_2`, although we see this is often not quite the case.
+Figure \ref{fig:ADS_M-bias} shows the M-bias plot for the ADS sample from the Lister (2011) dataset.
 
-If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is "batch aware".
+```{r m-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:ADS_M-bias", fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset.
+Each of read 1 (R1) and read 2 (R2) are plotted separately.
+We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", fig.width = 8}
+library(dplyr)
+library(ggplot2)
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
+read.mbias <- function(file) {
+
+  # Echo some information
+  message(paste0("Reading ", file, " ..."))
+
+  # Read the file.
+  x <- read.table(file, sep = '\n', as.is = TRUE)
+
+  # Find 'header' lines, which are dispersed throughout the file.
+  header_lines <- sapply(X = x, function(xx) {
+    grepl(pattern = 'context', x = xx)
+    })
+    y <- vector(mode = "list", length = sum(header_lines))
+    names(y) <- x[header_lines]
+
+    # z indexes the lines containing M-bias values.
+    z <- sapply(X = x, function(xx) {
+      grep(pattern = "context|^=|position", x= xx, invert = TRUE)
+      })
+      dz <- diff(z)
+      next_context_idx <- c(which(dz > 1), nrow(z))
+      colnames_idx <- sapply(X = x, function(xx) {
+        grep(pattern = "position", x = xx)
+        })[1]
+        colnames <- strsplit(x = x[colnames_idx, ], split = "\t")[[1]]
+
+        # Loop over the lines in the file.
+        for (i in seq_along(next_context_idx)) {
+          # Note the "+4" to skip the intermediate 'header' rows
+          start_idx <- ifelse(i == 1, 4, z[next_context_idx[i - 1]] + 4)
+          end_idx <- z[next_context_idx[i]]  
+          yy <- x[seq(from = start_idx, to = end_idx, by = 1), ]
+          y[[i]] <- do.call("rbind", lapply(yy, function(yyy, colnames) {
+            tmp <- strsplit(yyy, split = "\t")
+            val <- data.frame(as.integer(tmp[[1]][1]),
+            as.integer(tmp[[1]][2]),
+            as.integer(tmp[[1]][3]),
+            as.numeric(tmp[[1]][4]),
+            as.integer(tmp[[1]][5]))
+            colnames(val) <- colnames
+            return(val)
+            }, colnames = colnames))
+          }
+          y <- do.call("rbind", y)
+          y$Context <- strtrim(rownames(y), width = 3)
+
+          # Add R1/R2 annotation if paired-end data
+          if (grepl(pattern = "R1", x = rownames(y)[1])){
+            y$Read <- ifelse(grepl(pattern = "R1", x = rownames(y)), "R1", "R2")
+            } else{
+              y$Read <- "SE"
+            }
+            rownames(y) <- NULL
+
+            # Simplify column names
+            colnames(y) <- c('read_pos', 'M', 'U', 'perc_M', 'total',
+            'methylation_type', 'read')
+
+            # Re-order columns
+            y <- y[, c('read', 'read_pos', 'methylation_type', 'M', 'U', 'total',
+            'perc_M')]
+
+            # Replace CpG with CG
+            y$methylation_type <- ifelse(y$methylation_type == 'CpG', 'CG',
+            y$methylation_type)
+
+            # Convert to an MBias object.
+            return(y)
+          }
+#files <- paste0("../M-bias/", list.files("../M-bias/"))
+# ADS
+files <- "../M-bias/ADS.M-bias.txt"
+sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
+mbias <- lapply(files, read.mbias)
+mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
+                      do.call("rbind", mbias)))
+mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
+mbias %>%
+  filter(sample == "ADS") %>%
+  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
+  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: ADS") +
+  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
+  thesis_theme +
+  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
+
+```
+
+If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
 For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs.
@@
 
-The strongest source of M-bias in Illumina WGBS data is at the 5' end of `read_2`, which sequences the 3' end of the DNA fragment.
-Because the DNA fragment is often shorter than the sum of the read lengths, the 3' end of the fragment often contains adapter sequence and other "junk" sequence.
+The strongest source of M-bias in Illumina WGBS data is at the 5' end of _read 2_, which sequences the 3' end of the DNA fragment.
+Because the DNA fragment is often shorter than the sum of the read lengths, the 3' end of the fragment often contains adapter sequence and other 'junk' sequence.
 The adapter sequence may contain cytosine bases, which will be misinterpreted as evidence of methylation \citep{Krueger:2012ks}.
 Similarly, "fill-in cytosines" are used in the construction of RRBS libraries to repair the ends of DNA fragments after cleavage by MspI; these would also be misinterpreted as evidence of methylation \citep{Krueger:2012ks}.
-Another source of M-bias is incomplete or uneven bisulfite-conversion (__CITE__).
+Another source of M-bias is incomplete or uneven bisulfite-conversion.
 
-\subsubsection{Estimating M-bias}\label{sec:estimating_m-bias}
+\subsubsection{Computing M-bias}\label{sec:estimating_m-bias}
 
-Estimating the M-bias and incorporating the results into the methylation calling can be done using two different strategies:
+Computing the M-bias and incorporating the results into the methylation calling can be done using two different strategies:
 
 1.
-Compute the M-bias from the filtered data, then call methylation events incorporating both the filters and the M-bias.
+Compute the M-bias from the aligned reads, then call methylation events.
+The methylation calling should include filters to remove the detected M-bias (along with any other additional filters).
 This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling.
-This is the approach used by `bismark_methylation_extractor`, `methtuple` and `Bis-SNP`.
+This is the approach used by `bismark_methylation_extractor` \citep{Krueger:2011eb} and `Bis-SNP` \citep{Liu:2012ge}.
+([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
 2.
-Call methylation events from the filtered data but retain the read-position of each methylation event.
+Call methylation events but retain the read-position of each methylation event.
 Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias.
 This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls.
-This is the approach taken by `BSmooth`.
+This is the approach taken by `BSmooth` \citep{Hansen:2012gr}.
 
-I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple`.
-In theory, the first pass over the `SAM/BAM` could be "skipped" by computing the M-bias simultaneously with sorting the `SAM/BAM` and marking PCR duplicates.
-For example, using (idealised) Unix pipes:
-
-```
-# 'align' is read mapping software
-# 'mark_duplicates' is PCR duplicate marking software. It creates 'out.bam'.
-# 'mbias' computes mbias. It produces 'mbias.out' and (passes through) 'out.bam'.
-align ref.idx r1.fq r2.fq | mark_duplicates | mbias --filters
-# call_meth is methylation calling software. It produces 'call_meth.out' based on 'mbias.out' and 'out.bam'
-call_meth --filters mbias.out out.bam
-```
+I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
+In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.
 
@@
 
-```{r}
-# TODO: Figure of M-bias computed from raw vs. filtered data
-```
+\subsubsection{Pre-trimming reads destroys the one-to-one relationship between read-position and sequencing cycle}\label{sec:pre-trimming_reads}
 
-```{r}
-# TODO: Figure showing several M-bias curves with different patterns, e.g. 5' and 3' junk, within-read spikes/troughs, etc.
-```
+Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read position.
+This causes a minor problem when computing M-bias because we no longer know whether the read position is identical to the sequencing cycle.
+Soft-clipping or hard-clipping reads of their adapter sequence _during_ the alignment avoids this issue, because the clipping information (should be) preserved in the `CIGAR` string[^cigar].
 
-\subsection{Pre-trimming reads destroys the one-to-one relationship between read-position and sequencing cycle}\label{sec:pre-trimming_reads}
-
-Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read position.
-This causes a minor problem when computing M-bias because we no longer know whether the read position is identical to the sequencing cycle.
-Soft- or hard-clipping reads of their adapter sequence __during__ the alignment avoids this issue, because the clipping information (should be) preserved in the `CIGAR` string[^cigar].
-
-[^cigar]: Many downstream tools, including the current version of `methtuple`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads.
-The current version of `methtuple` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip).
-__FIX THIS__.
+[^cigar]: Not all software properly handles the information in the `CIGAR` string, particularly for soft-clipped reads.
 This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice.
-__TODO: Update what changes are made to `methtuple`.
-The `bwa-meth` and `LAST` alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark`, cannot soft-clip reads and so require that reads are pre-trimmed (__CITE `bwa-meth` pre-print or paper__).__
+The `bwa-meth` \citep{Pedersen:2014wt} and `LAST` \citep{Kieibasa:2011do} alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark` \citep{Krueger:2011eb}, cannot soft-clip reads and so require that reads are pre-trimmed.
 
@@
 For example, suppose we performed $100$ bp single-end sequencing and pre-trimmed the first 20 bp of $90\%$ of the reads.
-Then, read-position $80$ will comprised of $10\%$ sequencing cycle $80$ and $90\%$ sequencing cycle $100$.
-It is very likely the sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$.
-When using `bismark_methylation_extractor` this may necessitate that read-positions $80-100$ are ignored in downstream analyses, which would unnecessarily throw away $20\%$ of the data!
+Then, read-position $80$ will comprise $10\%$ sequencing cycle $80$ and $90\%$ sequencing cycle $100$.
+It is very likely that sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$.
+As most methylation callers exclude all read positions beyond a certain point, e.g., beyond read position 80, the would unnecessarily throw away $20\%$ of the data.
+`methtuple` is able to exclude specific read-positions, which allows it to deal with 'spikes' in the M-bias plot, like that shown in Figure \ref{fig:E18VA_M-bias}.
 
-This problem cannot be avoided if reads are pre-trimmed because the trimming information is not preserved.
+```{r M-bias_spike, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:E18VA_M-bias", fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset.
+Each of read 1 (R1) and read 2 (R2) are plotted separately.
+For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike.
+For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types.
+The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1.
+This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8}
+#files <- paste0("../M-bias/", list.files("../M-bias/"))
+# E18VA
+files <- "../M-bias/E18VA.M-bias.txt"
+sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
+mbias <- lapply(files, read.mbias)
+mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
+                      do.call("rbind", mbias)))
+mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
+mbias %>%
+  filter(sample == "E18VA") %>%
+  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
+  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
+  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
+  thesis_theme +
+  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
+```
+
+The loss of the one-to-one relationship between sequencing cycle and read-position cannot be avoided if reads are pre-trimmed because the trimming information is not preserved.
 \cite{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming].
-However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be "sequencing cycle" rather than "read-position".
+However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be 'sequencing cycle' rather than 'read-position', thus avoiding the issue entirely.
 
@@
 
-\subsection{Identifying M-bias}\label{sec:identifying_m-bias}
+\subsubsection{Identifying M-bias}\label{sec:identifying_m-bias}
 
-In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e.
-read-positions that are "too far away" from the vast majority of read-positions.
+In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions.
 A minor problem with this approach is that the user may (unconsciously) use a different cutoff for each sample.
@@
 an example where a single position is more than `cutoff` from the median due to a spike__).
+This motivated me to include functionality in my `MethylationTuples` R package ([https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples))  to perform more systematic processing of M-bias results.
+These are implemented in the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.
 
-This motivated me to include a function, `trim.mbias` (__TODO: Check names used in final version.
-In particular, `trim.mbias` needs to be re-named to reflect that it identifies, rather than trims, M-bias.__), to identify M-bias in a more systematic way.
-`trim.mbias`, along with the related functions `read.mbias`[^read.mbias] and `plot.mbias`[^plot.mbias], are part of my `cometh` package.
+For each sample, M-bias is computed separately for each methylation type because the level of methylation varies widely between CG and non-CG methylation types.
+For paired-end sequencing experiments, it is also done separately for each of _read 1_ and _read 2_ because M-bias is very different for these two mates and also because _read 2_ often has a slightly lower average level of methylation than does _read 1_ (e.g., see Figure \ref{fig:ADS_M-bias}).
 
-[^read.mbias]: `read.mbias` reads in the `M-bias` files produced by `Bismark` and returns these as a `data.frame`.
-[^plot.mbias]: `plot.mbias` creates a plot of M-bias stratified by the methylation type and, if the data are paired-end, whether the read is `read_1` or `read_2`.
+To do this, I use a simple normalisation of the read-position methylation levels ($rpml$).
+Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels (for example, $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read 1_).
 
-`trim.mbias` identifies read-positions that display M-bias.
-This is done separately for each methylation type because the level of methylation varies widely between CG and non-CG methylation types.
-For paired-end sequencing experiments, it is also done separately for each of `read_1` and `read_2` because M-bias is very different for these two mates (see __FIGURE__) and also because `read_2` often has a slightly lower average level of methylation than does `read_1` (see __FIGURE and provide explanation__).
-
-To do this, `trim.mbias` uses a simple normalisation of the read-position methylation levels ($rpml$).
-Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels (for example, $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in `read_1`).
-
-`trim.mbias` identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$.
+The `trim` method identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$.
 While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types.
-I tend to use a value of `cutoff = 0.02` (__TODO: confirm__), meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $nrpml_{CpG} < 0.73$ or $nrpml_{CpG} > 0.77$ is flagged as showing evidence of M-bias.
+I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $nrpml_{CpG} < 0.72$ or $nrpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias.
 It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types.
@@
 
-The results of `trim.bias` are returned as a `data.frame`:
+The `plot` method has a `type` argument, so that the 'raw' M-bias or the 'normalised' M-bias can be visualised.
 
 ```{r}
-# TODO: Insert code snippet showing a call to `trim.mbias`
-```
-
-The results can also be visualised by specifying the `plot = raw` or `plot = normalised` arguments.
-`plot = raw` gives the same M-bias plot as `plot.mbias` except that outliers are now flagged.
-`plot = normalised` plots the $nrpml$ at each read position along with the a horizontal line indicating the chosen `cutoff`.
-
-```{r}
-# TODO: Insert code snippet showing plots returned by `trim.bias` with both `plot = raw` and `plot = normalised`
+# TODO: Insert code snippet showing plots returned by `plot(MBias)` with both `type = raw` and `type = normalised`
 ```
@@
 However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude.
-__FIGURE__ shows an example of such a spike that occurred at __WHICH__ read-position, likely due to a boost in the laser for sequencing cycles $101-150$ (__CHECK: If can't find a source then rephrase__).
+Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101.
 Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
@@
 
-```{r, echo = FALSE}
-# TODO: M-bias plot with an internal spike
-```
-
 Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read positions entirely.
-In several datasets I have analysed, more than __HOW MUCH__ of the entire sequencing data is lost to M-bias.
+In the datasets I have analysed, I've had to ignore up to 30 read-positions per read due to M-bias.
 
@@
 A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability.
-However, we could still compute $\beta$-values, and use these in downstream inferences, without much loss in interpretation.
-An additional downside to this approach is an increased computational complexity and cost, and this is why I have not further pursued this idea.
+However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation.
+The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.
 
@@
 Apart from M-bias, technical biases include sequencing and alignment errors, which can both give rise to spurious or otherwise incorrect methylation calls in obvious ways.
-
-Biological biases include cellular heterogeneity \citep{Jaffe:2014gy} and sequence variation at or nearby methylation loci, such as CpGs that are also SNPs or with SNPs nearby.
-I discuss the latter in the section __METHYLATION CALLING__.
+Biological biases include cellular heterogeneity and sequence variation at or nearby to methylation loci.
+I discuss the latter in __TODO: WHERE???__
 
@@
 
-Strictly speaking, cellular heterogeneity exists once you study DNA from more than a single cell, as shown in a recent papers using single-cell bisulfite-sequencing \citep{Guo:2013iha,Smallwood:2014kn}, however in this context it means when you study DNA from a sample that is made up of different types of cells.
+Strictly speaking, cellular heterogeneity exists once you study DNA from more than a single cell, as shown in a recent papers using single-cell bisulfite-sequencing \citep{Guo:2013iha,Smallwood:2014kn}.
+However, in this context it means when you study DNA from a sample that is made up of different _types_ of cells.
 
@@
 
-In contrast to \citet{Houseman:2012km} where this cellular heterogeneity was useful, in studies exploring the relationship between DNA methylation changes and phenotypes, this cellular heterogeneity can seriously bias the results and must be properly accounted for in any analysis \citep{Jaffe:2014gy,Houseman:2014ds}.
-\citet{Jaffe:2014gy} provide evidence that several reported relationships between age and DNA methylation are likely due to changes in the cell composition of whole blood with age and not due to DNA methylation changes _per se_.
+In contrast to \citet{Houseman:2012km}, where this cellular heterogeneity was useful, this cellular heterogeneity can seriously bias the results and must be properly accounted for in any study exploring the relationship between differences in DNA methylation and a phenotype \citep{Jaffe:2014gy,Houseman:2014ds}.
+For example, \citet{Jaffe:2014gy} provide evidence that several reported relationships between age and DNA methylation are likely due to changes in the cell composition of whole blood with age and not due to DNA methylation changes _per se_.
 
-Methods to estimate the cellular heterogeneity bias and adjust for it are available (e.g.
-\citet{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}), although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data.
+Methods to estimate the cellular heterogeneity bias and adjust for it are available \citep[e.g.,]{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}, although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data.
 This is not to say that these problems don't exist for sequencing data, merely that these have not been as well-explored.
@@
 
-[^no_call]: A third possibility is making the call that the "methylation locus" is not in fact a methylation locus.
-For example, if the sequenced base at a cytosine in the reference sequence is an _A_ or a _G_ then this may be evidence that the position is not in fact a methylation locus.
+[^no_call]: A third possibility is making the call that the 'methylation locus' is not in fact a methylation locus.
+For example, if the sequenced base at a cytosine in the reference sequence is an adenine or a guanine then this may be evidence that the position is not in fact a methylation locus.
 
@@
 [^bismark_methylation_extractor]: `Bismark` also includes a program called `bismark_methylation_extractor`, which, as the name suggests, extracts the methylation calls from the `SAM/BAM` file.
-So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` or `methtuple` is required to make use of the methylation calls.
+So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` is required to make use of the methylation calls.
+`methtuple` may be used as a drop-in replacement for `bismark_methylation_extractor` and provides several extra features, discussed later in this section.
 
-Both `Bismark` and `Bis-SNP` perform methylation calling of 1-tuples.
-In contrast, `methtuple` can perform methylation calling at m-tuples.
-`methpat` (__CITE__) is also capable of performing methylation calling at m-tuples.
+Most methylation callers, such as `Bismark` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
+In contrast, `methtuple` can perform methylation calling at m-tuples; an m-tuple is $m$ adjacent methylation loci, where $m = 1, 2, \ldots$.
+m-tuples are discussed in greater detail in Chapter \ref{chap:a_statistical_model_of_wgbs} and are used extensively in Chapter \ref{chap:comethylation}.
+One of the few other methylation callers that can perform calling at m-tuples is `methpat` ([https://github.com/bjpop/methpat](https://github.com/bjpop/methpat)).
 
@@
 Care must be taken to correctly handle the orientation and strand of the alignment.
-__FIGURE__ gives several examples (__TODO: Figure explaining reference-based methylation calling for reads aligned to forward and reverse strands__).
 
 When using reference-based methylation calling, the position of the methylation locus is with respect to the reference genome, since then all samples will use a common set of co-ordinates.
-Furthermore, unless the sample of the genome is fully known, methylation loci in insertions cannot be distinguished from genetic variation since there is no reference sequence to compare them against.
+Some loci cannot be typed using a reference-based approach.
+For example, unless the genome of the sample is fully known, methylation loci in insertions cannot be distinguished from genetic variation since there is no reference sequence to compare them against.
 
@@
 When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from the estimation of $M$ and $U$[^read_filters].
-Strictly speaking each sequenced nucleotide is assigned a weight, however, in practice, filter are normally first applied to reads and then to all nucleotides within 'surviving' reads.
+Strictly speaking each sequenced nucleotide is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all nucleotides within 'surviving' reads.
 In my own `methtuple` software I routinely use the following filters.
-
-[^read_filters]: A read that is not used to estimate $M$ and $U$ may still be used in other analyses, such as estimating copy number variation, and vice-versa.
 
@@
 
-1. Read is mapped (SE, PE) and mapped in the expected orientation (PE only).
+1. Read is mapped (single-end or paire-end) and mapped in the expected orientation (paired-end only).
 2. Read is not marked as a PCR duplicate.
@@
 3.
-The base is a 'methylation-induced mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a "non-methylation-induced mismatch" (e.g., the sequenced base is an A or a G at a C in the reference sequence).
+The base is a 'bisulfite mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a 'non-bisulfite mismatch; (e.g., the sequenced base is an A or a G at a C in the reference sequence).
 
-The reference sequence is typically the reference genome used in the alignment step, which implicitly defines $\mathcal{I}_{j} :=\mathcal{I}_{rg}, j = 1, \ldots, N$, where $\mathcal{I}_{rg}$ is the set of methylation loci in the reference genome.
-Due to DNA variation between the sample and the reference genome, this assumption is obviously not true.
-A more refined approach incorporates genetic differences between the sample and the reference genome.
+[^read_filters]: A read that is not used to estimate $M$ and $U$ may still be used in other analyses, such as estimating copy number variation.
+
+The reference sequence is typically the reference genome used in the alignment step, although there are obviously differences between the sample's genome and the reference genome.
+This reference-based approach can be refined to incorporate genetic differences between the sample and the reference genome.
 This can be done in several ways:
 
-* Whole-genome sequencing or genotyping of the sample
-* Calling genetic variations directly from the bisulfite-sequencing data
-* Excluding sites of known genetic variation
+- Whole-genome sequencing or genotyping of the sample
+- Calling genetic variations directly from the bisulfite-sequencing data
+- Excluding sites of known genetic variation
 
@@
 The gold-standard is to perform whole-genome DNA sequencing of each sample.
-This data is then used to form a set of sample-specific methylation loci, $\mathcal{I}_{j}$.
+This data is then used to form a set of sample-specific methylation loci.
 However, this approach is also very expensive due to the extra sequencing requirements.
 A cheaper alternative is to genotype the sample on a genome-wide SNP microarray.
-This will give very accurate, very cheap genotypes at a large number of loci ($500,000 - $5,000,000$), however, by definition it cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.
+This will give very accurate, very cheap genotypes at a large number of loci ($500,000 - $5,000,000$).
+However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.
 
@@
 
-The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci, $\mathcal{I}_{j} := \mathcal{I}_{j}^{Bis-SNP}$, at which to call methylation events.
-It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina (_Lister-style_) whole-genome bisulfite-sequencing protocol.
+The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci at which to call methylation events.
+It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina ('_Lister-style_') whole-genome bisulfite-sequencing protocol.
 
@@
 Unfortunately, $C>T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants.
-It is possible, although difficult, to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand - if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see __FIGURE__).
-Other substitutions are more readily called, and insertion and deletion events (InDels) may also be called.
+Fortunately, it is possible to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand - if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
+Other base substitutions are more readily detected, and insertion and deletion events (indels) may also be called.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
+\caption{`Bis-SNP` is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
+For all three loci, the reads mapped to the forward strand contain a thymine.
+However, it is the base on the reverse strand that reveal the true genotype.
+When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location.
+This is only possible with bisulfite-data generated using the directional protocol.
+This figure is adapted from \cite{Liu:2012ge}}
+\label{fig:Bis-SNP.pdf}
+\end{figure}
+
 In case it isn't clear, `Bis-SNP` actually provides three important pieces of information that make it almost as good as having whole genome DNA sequencing data on the same sample:
@@
 
-Genotype calls made using `Bis-SNP` are less accurate than those from equivalent-depth whole-genome DNA sequencing because of the reduced complexity of bisulfite-converted DNA.
-However, we essentially get DNA variation "for free" by using `Bis-SNP`, which makes it my preferred approach for methylation calling at 1-tuples.
+Genotype calls made using `Bis-SNP` are less accurate than those from equal-depth whole-genome DNA sequencing because of the reduced complexity of bisulfite-converted DNA.
+However, we essentially get to measure DNA variation 'for free' by using `Bis-SNP`, which makes it my preferred approach for incoporating genetic variation into methylation calling.
+The genetic variant calls made by `Bis-SNP` can also be used to _post-hoc_ filter methylation calls made by other software.
+I use this approach to filter methylation calls at m-tuples made with `methtuple`.
 
@@
 The third approach is to call methylation loci using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population.
-For example, exclude all cytosines in the reference genome that are also SNPs in dbSNP (__CITE__).
-This approach can only exclude sites from $\mathcal{I}_{j}$, it cannot augment $\mathcal{I}_{j}$ by adding sample-specific methylation loci.
+For example, exclude all cytosines in the reference genome that are also SNPs in dbSNP \citep{Sherry:2001gh}.
+This approach can only exclude sites from consideration, it cannot add sample-specific methylation loci.
 
-This is a conservative approach, as it will remove loci regardless of whether the sample has a genetic variant at that position or not, but it may be good enough.
+This is a conservative approach, as it will remove loci regardless of whether the sample has a genetic variant at that position or not, but it may be a good enough method in some cases.
 It also obviously requires a database of known variation for the organism being studied, which is the case for commonly studied organisms such as humans and mice.
@@
 
-\subsection{A criticism of "identifying methylcytosines"}{sec:criticism_of_methylcytosines}
+\section{\texttt{methtuple}}
 
-__TODO: Need a better title for this section__
-
-\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" [supplementary material].
-This idea was also used in subsequent, and similarly high-profile, publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013].
-I believe this idea is an unrequired source of confusion, as I will explain.
-
-[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
-
-It is necessary to use scare quotes to distinguish the term "methylcytosine", as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from methylcytosine, the standard definition meaning a 5-methylcytosine.
-I will first describe the idea and how it differs from methylation calling.
-Then I will describe how it is used to (1) estimate the number of methylcytosines in the sample, and (2) as a kind of screening procedure that determines whether each cytosines will be used in downstream analyses.
-Finally, I will explain why (1) is a bad estimator and why I think (2) is unnecessary or misguided.
-
-A "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
-Thus, the idea of identifying "methylcytosines" is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
-To do so, Lister et al.
-compare each $\beta$-value to a (coverage-specific and sample-specific) cutoff that determines whether the cytosine is classified as a "methylcytosine" or not.
-
-The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy} and \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
-Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done (__TODO: Discuss with Terry__).
-What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-
-The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to "error", where "error" is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
-Mathematically, for each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated "error".
-Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
-
-[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
-
-I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure (__CITE__).
-This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
-thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
-
-The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
-It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
-
-Now to why I don't think this is a very useful procedure.
-Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
-These are two different things.
-If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
-A simple example makes this clear.
-
-Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
-Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
-We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
-From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
-
-Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
-For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
-A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
-
-Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
-While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
-For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
-If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
-
-\section{Identifying differential methylation}\label{sec:identifying_differential_methylation}
-
-Identifying differential methylation means to identify sites (differentially methylated cytosines or _DMC_s) or regions (differentially methylated regions or _DMR_s) in the genome that have different levels of methylation between two or more conditions.
-Much of the methodology for identifying differential methylation has focused on two-group experiments and a restricted definition: differential methylation is the difference in __average__ methylation between __two__ groups.
-This is analogous to identifying differential gene expression, which focuses on identifying genes that have different __average__ expression levels between two groups.
-
-While bisulfite-sequencing experiments with multiple groups have been performed and analysed (e.g.
-\citet{Lister:2009hy, Hansen:2011gu, Hansen:2013eo}), these have typically been done using a series of pair-wise comparisons or by comparing each sample in turn against some 'baseline' sample.
-This is because the analysis of a pair-wise comparison is relatively simple and because it is also generally easier to interpret a two-group comparison than a multi-group comparison.
-However, as the number of groups increases it becomes more difficult to interpret multiple pair-wise comparisons.
-
-Like any experimental questions, good experimental design is essential in order to reliably infer differential methylation.
-Identifying DMCs is relatively straightforward statistical problem since it amounts to the well-worn problem of testing the difference of means.
-However, identifying DMRs, which may or may not build upon an initial scan for DMCs, is a far more challenging statistical problem owing the high spatial dependence of methylation at neighbouring cytosines and the uneven spacing of cytosines throughout the genome.
-
-\subsection{Experimental design}\label{sec:experimental_design}
-
-In any experiment of differential methylation we want the DMCs and DMRs to be _biological_ and _replicable_; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
-Key to ensuring biological relevance good experimental design, such as the use of _replicates_ in each experimental group.
-A distinction is often made between _technical replicates_ and _biological replicates_.
-Roughly, biological replicates are experimental units that all underwent the same treatment and are used to estimate the within-group variability of the treatment.
-Technical replicates are repeated measurements of the same experimental unit, perhaps with slight variations in the sample preparation, and are used to estimate the variability of the sample preparation and measurement process.
-
-As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one.
-For example, \cite{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine).
-I contend that these are more like technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].
-
-[^cell_passaging]: In the case of IMR90 cell line, the first replicate, IMR90_r1, underwent 4 cell passages and the second replicate, IMR90_r2, underwent 5 cell passages.
-In the caase of the H1 cell line, the first replicate, H1_r1, underwent 25 passages in the first media and 9 passages in the second media, and the second replicate, H1_r2, underwent 27 passages in the first media and 5 passages in the second media.
-
-Moreover, early experiments with whole-genome bisulfite-sequencing frequently had no replicates of any kind, as is the case of the H9 and IMR90-iPSC cell lines in \cite{Lister:2009hy}, and once these replicates were judged to be broadly consistent these data were pooled prior to analysis, thus ignoring all within-group variability.
-
-Ideally, variability between technical replicates should be orthogonal to the biological question, but this rarely occurs.
-Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.
-
-\subsection{DMCs}\label{sec:dmcs}
-
-There are reports of differential methylation at individual CpGs resulting in a phenotypic difference (__CITE EXAMPLES; Emma Whitelaw has some I think__).
-Nonetheless, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from WGBS experiments with sample sizes you can count on one or two hands and average sequencing depth of $10-30\times$.
-
-Identifying DMCs boils down to identifying differences in means.
-As such, it can be readily framed as a "stand-alone" test or cast into a regression framework.
-There is an enormous body of statistical literature on testing for differences in means (__CITE examples__).
-Most methods to detect DMCs from WGBS data are parametric and frequently use the binomial model described in __SECTION__.
-
-Regardless of the test used, all analyses must pay a large multiple-hypothesis testing penalty.
-Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data.
-but the number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments (e.g.
-approximately 1 million tests in genome-wide association studies and 20,000 tests in studies of differential gene expression).
-
-However, when identifying DMCs, the _effective_ number of tests is fewer, perhaps very much so, due to the high degree of dependence of methylation at neighbouring methylation loci, which in turn ensures a high degree of dependence amongst the tests.
-
-__TODO: What techniques can be used to do FDR, multiple hypothesis testing? Are standard tools applicable in such high-dependence settings?__
-
-\subsubsection{One-group tests}\label{sec:one_group_dmcs}
-
-The one-group WGBS experiment is uncommon and the $A^{vy}$ experiment (__SECTION__) is the only example that I know of.
-In a one-group experiment a DMC is one where at least one of the samples has a different methylation level to the group average.
-Mathematically, we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$, where $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
-This can be tested in a number of ways as I will demonstrate.
-
-The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown below (__TODO: Prettify table with `booktabs`__)
-
-\begin{table}[h]
-\centering
-\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
-\label{my-label}
-\begin{tabular}{c c c}
-\hline
-             & m          & u          \\ \hline
-$Sample_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
-$\vdots$     & $\vdots$   & $\vdots$   \\
-$Sample_{n}$ & $m_{i, n}$ & $u_{i, n}$ \\
-\end{tabular}
-\end{table}
-
-Such contingency tables can be analysed in many different ways. Here I apply a few methods to highlight their similarities and differences. Firstly, here are some example data for a single CpG:
-
-```{r}
-x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
-names(dimnames(x)) <- c("sample", "meth_state")
-x
-```
-
-We see that while `sample_1, ..., sample_4` are nearly completely methylated at this CpG, `sample_5` has approximately $50\%$ methylation, which may be indicative of allele-specific methylation.
-
-In my analysis of the $A^{vy}$ experiment I used the `prop.test` function in R. As documented in the help page for `prop.test`:
-
-> `prop.test` can be used for testing the null that the proportions (probabilities of success) in several groups are the same
-
-in R to test the equality of the proportions ($\beta$-values) across the five mice.
-
-```{r}
-prop.test(x)
-```
-
-As expected, this test returns a very small P-value, `r print(format.pval(prop.test(x)$p.value), quote = FALSE)`.
-
-What might not be clear from its name is that `prop.test(x)` is just performing Pearson's $\chi^{2}$-test of the null hypothesis that the joint distribution of the cell counts in the $2 \times n$ contingency table is the product of the row and column marginals.
-This can be seen by examining the source code or by comparing the output to `chisq.test(x)`: (__TODO: Check by hand__)
-
-```{r}
-chisq.test(x)
-```
-
-Instead of using the "stand-alone" tests, `prop.test` and `chisq.test`, we could explicitly frame this as a log-linear model and test the goodness-of-fit of the "no interaction" model using either the likelihood ratio statistic (`lrt`) or the Pearson's $\chi^{2}$ test statistic (`pearson`).
-Formally, the regression model is (__TODO: Write down the regression model__):
-
-This model can be fit using the `loglin` function in R or using the arguably more user-friendly `loglm` (from the `MASS` package) or `glm` functions:
-
-```{r}
-# Using loglin
-loglin(x, margin = list(1, 2))
-# Using loglm
-library(MASS)
-loglm(formula =  ~ sample + meth_state, data = x)
-# Using glm (1)
-glm(x ~ 1, family = binomial)
-# Using glm (2)
-glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2])
-# TODO: Write sentence summarising output.
-```
-
-A key advantage of using regression is the option to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
-__TODO: How to do this using `loglin`?__
-
-More complex regression analyses are possible, such as empirical Bayes or fully Bayesian models.
-I describe these in more detail in discussing two-group and multi-group experiments, but they may also be applied to one-group experiments.
-
-A proper software implementation needs to take care of issues such as incomplete tables and non-convergence of the iterative proportional scaling algorithm used by `loglin`.
-Speed is also important when performing this many tests and the regression based approaches are much slower, at least when naively applied, due to additional checks they make of the data:
-
-```{r}
-# Basic benchmarking
-library(microbenchmark)
-print(microbenchmark(prop.test(x), chisq.test(x), loglin(x, margin = list("sample", "meth_state")), loglm(formula = ~ sample + meth_state, data = x), glm(x ~ 1, family = binomial), glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2]), glm.fit(x = matrix(, nrow(x), 0L), y = x, family = binomial()), times = 100), order = "median")
-```
-
-\subsubsection{Two-group experiments}\label{sec:two_group_dmcs}
-
-In a two group experiment, for each methylation locus we are testing the hypothesis that the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, is different to that in the second group, $\bar{\beta_{i}^{(2)}}$.
-
-If there are no replicates in either group[^2x2] then we have a $2 \times 2$ contingency table:
-
-[^2x2]: I'd argue that this is a two-sample experiment rather than a two-group experiment since each group has $n = 1$.
-
-(__TODO: Prettify table with `booktabs`__)
-
-\begin{table}[h]
-\centering
-\caption{$2 \times 2$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a two-group experiment with no replicates in either group.}
-\label{my-label}
-\begin{tabular}{c c c}
-\hline
-             & m          & u          \\ \hline
-$Group_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
-$Group_{2}$ & $m_{i, 2}$ & $u_{i, 2}$ \\
-\end{tabular}
-\end{table}
-
-This $2 \times 2$ table is then tested for independence of the rows and columns.
-As the sample size for each table is often small, Fisher's exact test is generally preferable since, unlike Pearson's $\chi^{2}$ test, the P-values are _exact_, meaning they do not rely on large-sample approximations (__CITE: Agresti__).
-Fisher's exact test is a _conditional_ test because the null distribution (a hypergeometric distribution) is derived by conditioning on both the marginal totals.
-__CITE: Agresti__ discusses the arguments for and against conditional and unconditional tests of independence in the $2 \times 2$ table.
-
-The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
-Unfortunately, some authors have used Fisher's exact test to analyse experiments including replicates (__TODO: Highlight papers that have inappropriately used Fisher's exact test; does Lister 2013 inappropriately pool then test?).
-This is incorrect because it effectively aggregates all the counts within each group and so ignores all within-group biological variability \citep{Hansen:2012gr}.
-
-Instead, a test that incorporates the within-group variability should be used.
-For example, we could use a t-test to compare the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$.
-Several methods have been proposed for testing for DMCs.
-These methods differ considerably in how they model and transform the data as well as what test statistic is used to identify DMCs.
-
-\paragraph{Transforming}\label{sec:transforming}
-
-Smoothing of the $\beta$-values has been used advocated by several authors to remove variation due to low sequencing coverage and to exploit the spatial correlation of DNA methylation \citep{Hansen:2011gu, Hebestreit:2013ko, Gokhman:2014c}.
-Both `BSmooth` \citep{Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} smooth the $\beta$-values using a binomial local likelihood smoother.
-\citet{Lister:2011kg, Gokhman:2014cp} used a simpler moving average in windows along the genome.
-
-\paragraph{Modeling}\label{sec:modeling}
-
-`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
-Broadly, under the Beta-Binomial model the "true" methylation level at the $i^{th}$ methylation locus in the $k^{th}$ group is modeled as a Beta random variable, $B_{i, k} = Beta(\mu_{i, k}, \theta_{i, k})$.
-The observed number of methylated reads at the $i^{th}$ methylation locus, in the $j^{th}$ sample is modeled as a Binomial random variable, $M_{i, j, k} | B_{i, k}, n_{i, j, k} = Binomial(n_{i, j, k}, B_{i, k}), which is conditional on the unobserved $B_{i, k}$ and the observed sequencing coverage, $n_{i, j, k}$.
-
-\paragraph{Estimating and testing}\label{sec:estimating_and_testing}
-
-`DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
-The test used to identify DMCs is variously a Wald test (`BiSeq`, `DSS`), likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).
-
-\subsubsection{Multi-group experiments}\label{sec:multigroup_dmcs}
-
-\subsection{DMRs}
-
-There are two very different strategies for identifying differentially methylated regions:
-
-1.
-Use regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g.
-3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
-2. Use regions that are defined based on the data, which are then tested for differential methylation. Valid statistical inference of such regions is __very challenging__.
-
-\subsubsection{Using _a priori_ regions}\label{sec:a_prior_regions}
-
-The methylation level is summarised at the region level and then compared across groups. For example, the average of the $\beta$-values across the region may be compared between groups using a t-test.
-
-__TODO: Describe the above mathematically__
-
-This is qualitatively similar to identifying differentially methylated cytosines.
-
-__PROS__
-
-* Simple to implement
-* Valid statistical properties
-
-__CONS__
-
-* Doesn't account for differential CpG density across regions
-* What is the right "unit" - the CGI, the promoter, a tile???
-* Doesn't account for spatial correlation of methylation
-
-\subsubsection{Using data-driven regions}\label{sec:data_driven_regions}
-
-__PROS__
-
-* Can discover the right "unit".
-* Might account for spatial correlation of methylation
-
-__CONS__
-
-* Loss of statistical properties in testing, e.g. FDR control
-* Bloody hard
-
-\section{Other downstream analyses}\label{sec:other_downstream_analyses}
-
-Bisulfite-sequencing data are used to address a variety of questions apart from identifying differential methylation.
-Not all these analyses require statistical inference.
-For example, if we are interested in knowing whether the promoter of a particular gene is methylated in our sample then it may be sufficient to simply plot the level of methylation at each CpG in the region.
-
-However, once we move to experiments with multiple samples, multiple regions and more complex questions then statistical methods are generally required to distinguish the signal from the noise.
-In this section I briefly describe other
-
-__TODO: Any other important downstream analyses worth mentioning?__
-
-\subsection{Variably methylated regions and differential variability}\label{sec:vmrs}
-
-\subsection{Allele-specific methylation}\label{sec:asm}
-
-\subsection{Methylation entropy}\label{sec:methylation_entropy}
-
-\subsection{Epipolymorphism}\label{sec:epipolymorphism}
-
-\subsection{Epiallele detection}\label{sec:epialleles}
-
-\subsection{Integrating DNA methylation data with other genomes data}\label{sec:data_integration}
-
-## TODOs
-* See https://github.com/brentp/clustermodel/README.md for a nice, brief summary of the main DMR calling approaches.
-* Replace "R" with "\\R"
+__TODO__

### 2014-11-26 Rmarkdown/appendix.Rmd
Typo in R code

@@
 ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) + 
-  geom_point(size = 5) + coord_cartestian(ylim = c(0, 1.5)) + 
+  geom_point(size = 5) + coord_cartesian(ylim = c(0, 1.5)) + 
   ggtitle("CpGs are underrepresented in human reference genome (hg19)") + 

### 2014-11-26 Rmarkdown/biology_background.Rmd
Minor wording change.
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.

@@
 
-\section{Chapter Overview}\label{sec:biology_background_chapter_overview}
+\chapter{Biology background}\label{chap:biology_background}
+
+\section{Chapter overview}\label{sec:biology_background_chapter_overview}
 
@@
 
-[^other_DNAm]: Two additional forms of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C).
+[^other_DNAm]: Two additional examples of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C).
 N6-methyladenine is a methylated form of adenine, which is found in mRNA \cite[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \cite{Ratel:2006dv}.

### 2014-11-26 Rmarkdown/comethylation_review.Rmd
Corrected citations
Added TODO
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.

@@
 ---
+
+\chapter{Co-methylation review}\label{chap:comethylation_review}
 
@@
 Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated (e.g.
-\cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010drm, Landan:2012kp}).
+\cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}).
 Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.
@@
 - An autocorrelation analysis implicitly assumes stationarity along the genome whereas between-sample co-methylation assumes stationarity across samples (__TODO: Discuss with Terry__).
+- Comment on http://nar.oxfordjournals.org/content/40/22/11490.long

### 2014-11-26 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.

--- 2014-11-26 Rmarkdown/wgbs_analysis.Rmd
@@
 ---
+
+\chapter{Bioinformatics analysis of whole-genome bisulfite-sequencing data}\label{chap:wgbs_bioinformatics_analysis}
 
@@
 Each of read 1 (R1) and read 2 (R2) are plotted separately.
-We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", fig.width = 8}
+We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", fig.width = 8, dev = "pdf"}
 library(dplyr)
@@
 The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1.
-This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8}
+This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8, dev = "pdf"}
 #files <- paste0("../M-bias/", list.files("../M-bias/"))

### 2014-11-26 latex/phd_thesis.tex
Fixed latex errors and missing references.
Updated main files.
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.

@@
             pdfcreator={pdflatex, or other tool}]{hyperref}
+
+% pandoc
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{booktabs}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\newenvironment{Shaded}{}{}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
+\newcommand{\RegionMarkerTok}[1]{{#1}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
+\newcommand{\NormalTok}[1]{{#1}}
 
@@
 
+\setcounter{tocdepth}{1}
+
 \begin{document}
@@
 
-\title{\huge \bf Thesis title\\}
+\title{\huge \bf The statistical analysis of high-throughput assays
+for studying DNA methylation.\\}
 
@@
 
-Microarrays are a popular tool for measuring relative gene expression between RNA populations for thousands of genes simultaneously...
+\textbf{TODO}
 
@@
 
-\include{wgbs_analysis} % wgbs_analysis.tex
+\include{wgbs_bioinformatics_analysis} % wgbs_bioinformatics_analysis.tex
 
-\include{a_statistical_model_of_methlyC-seq} % a_statistical_model_of_methlyC-seq.tex
+\include{wgbs_statistical_analysis} % wgbs_statistical_analysis.tex
 
-\include{avy} % avy.tex
+\include{datasets} % datasets.tex
+
+\include{Avy_epialleles} % Avy_epialleles.tex
 
@@
 
-\chapter{Appendix A}
+\include{appendix} % appendix.tex
 

### 2014-11-27 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Section on methtuple

@@
           }
+
 #files <- paste0("../M-bias/", list.files("../M-bias/"))
@@
   scale_color_discrete(guide = guide_legend(title = "Methylation type"))
-
 ```
@@
 
-Computing the M-bias and incorporating the results into the methylation calling can be done using two different strategies:
+Estimating the M-bias and incorporating the results into the methylation calling can be done using two different strategies:
 
@@
 This is the approach used by `bismark_methylation_extractor` \citep{Krueger:2011eb} and `Bis-SNP` \citep{Liu:2012ge}.
-([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
 2.
@@
 
-I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
+I find the first strategy conceptually simpler, and easier to program, particularly when I start dealing with m-tuples, and so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
 In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.
@@
 It is very likely that sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$.
-As most methylation callers exclude all read positions beyond a certain point, e.g., beyond read position 80, the would unnecessarily throw away $20\%$ of the data.
-`methtuple` is able to exclude specific read-positions, which allows it to deal with 'spikes' in the M-bias plot, like that shown in Figure \ref{fig:E18VA_M-bias}.
 
@@
 Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
-My `methtuple` software is more flexible by allowing the user to exclude individual read-positions or ranges of read-positions via the `--ignore-read1-positions` and `--ignore-read2-positions` flags.
 
@@
 [^bismark_methylation_extractor]: `Bismark` also includes a program called `bismark_methylation_extractor`, which, as the name suggests, extracts the methylation calls from the `SAM/BAM` file.
-So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` is required to make use of the methylation calls.
-`methtuple` may be used as a drop-in replacement for `bismark_methylation_extractor` and provides several extra features, discussed later in this section.
-
-Most methylation callers, such as `Bismark` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
-In contrast, `methtuple` can perform methylation calling at m-tuples; an m-tuple is $m$ adjacent methylation loci, where $m = 1, 2, \ldots$.
-m-tuples are discussed in greater detail in Chapter \ref{chap:a_statistical_model_of_wgbs} and are used extensively in Chapter \ref{chap:comethylation}.
-One of the few other methylation callers that can perform calling at m-tuples is `methpat` ([https://github.com/bjpop/methpat](https://github.com/bjpop/methpat)).
+So while `Bismark` annotates each base as methylated or unmethylated during the alignment, a secondary step using `bismark_methylation_extractor` is required to make the methylation calls.
 
@@
 Strictly speaking each sequenced nucleotide is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all nucleotides within 'surviving' reads.
-In my own `methtuple` software I routinely use the following filters.
+In my own work, I routinely use the following filters.
 
@@
 
-1. Read is mapped (single-end or paire-end) and mapped in the expected orientation (paired-end only).
-2. Read is not marked as a PCR duplicate.
-3. Read has a mapping quality score greater than some threshold.
+1. The read is mapped (single-end or paire-end) and mapped in the expected orientation (paired-end only).
+2. The read is not marked as a PCR duplicate.
+3. The read has a mapping quality score greater than some threshold.
 
@@
 
-1. Read-position of base means that it is unlikely to be affected by M-bias.
+1. The read-position of the base means that it is unlikely to be affected by M-bias.
 2. The base quality score greater than some threshold.
@@
 
-\section{\texttt{methtuple}}
+\section{\texttt{methtuple}}\label{sec:methtuple}
 
-__TODO__
+Most methylation callers, such as `Bismark` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
+The output file is a table, where each row records the co-ordinates of a cytosine and the number of methylated and unmethylated reads at that position.
+Table \ref{tab:1-tuples} is representative of the type of data returned by these programs.
+The file format is generally tab-delimited plain text, the Browser Extensible Data (BED) format or the Variant Call Format (VCF).
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lllll@{}}
+\toprule
+Chromosome & Strand & Position & M & U \\ \midrule
+chr1       & +      & 100      & 7 & 1 \\
+chr1       & -      & 101      & 5 & 2 \\
+chr2       & +      & 400      & 0 & 3 \\
+chr2       & +      & 450      & 1 &   \\ \bottomrule
+\end{tabular}
+\caption{My caption}
+\label{tab:1-tuples}
+\end{table}
+
+While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data.
+Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
+Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_analysis}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
+In order to study these concepts, we need software that can perform methylation calling at m-tuples, which was not available when I began my PhD, and so I wrote `methtuple`.
+
+[^chimeric_reads]: This ignores chimeric reads, which are created when two DNA fragments ligate to one another during the library preparation.
+Certain bisulfite-sequencing protocols frequently produce chimeric reads.
+For example, using the post-bisulfite adapter tagging (PBAT) protocol \citep{Miura:2012fr} with a low input amount of DNA results in a huge number of chimeric reads (personal communication from Felix Krueger).
+The standard whole-genome bisulfite-sequencing protocol is not known to suffer from this issue.
+
+\subsection{m-tuples}\label{sec:m-tuples}
+
+I define an m-tuple to be a tuple of $m = 1, 2, \ldots$ methylation loci.
+I refer to $m$ as the _size_ of the tuple.
+In principle, the $m$ loci that make up an m-tuple could come from anywhere in the genome, but it makes more sense to require that the $m$ loci be close to one another.
+In fact, I generally require that an m-tuple consists of $m$ _adjacent_ methylation loci[^adjacent].
+An equivalent way of describing an m-tuple as comprising adjacent methylation loci is that the number of intervening loci is zero ($NIL = 0$).
+There are three reasons that I focus on m-tuples with $NIL = 0$:
+
+1.
+Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ m-tuples, provided that we restrict ourselves to those m-tuples with $NIL = 0$.
+In contrast, if we allow $NIL \geq 0$ then there are $\binom{l}{\text{m}} \sim$ m-tuples.
+Obviously, $\binom{l}{\text{m}} \geq l - \text{m} + 1$, with strict inequality if $m \neq 1, l$.
+2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}
+3.
+Measurability: We cannot observe m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology.
+This is true even when $NIL = 0$ but is more of an issue if we allow $NIL \geq 0$.
+
+[^adjacent]: Two methylation loci are adjacent if there is no methylation loci in between the pair.
+For example, `CGCG` and `CGTTACG` both contain two adjacent CpGs (the intervening `TTA` in the second sequence does not include a CpG).
+In contrast, the first and last CpG in the sequence `CGTCGTCG` are __not__ adjacent, since the intervening sequencing, `TCGT` include a CpG.
+Note that in situations where we are only interested in studying CpGs, we define 'methylation loci' to mean 'CpGs'.
+Therefore the sequence `CGTCTTCG` contains two adjacent methylation loci; while there is a `C` in the intervening sequence, `TCTT`, it is a CHH not a CpG.
+
+Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
+The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combination` flag is set[^nil0].
+
+[^nil0]: Actually, while `methtuple` tries to produce m-tuples with $NIL = 0$ it can't guarantee this because it would require parsing the reference genome sequence.
+This is only really an issue with paired-end sequencing and is made clear in the example.
+Some _post-hoc_ filtering of the m-tuples will generally be required in order to remove those m-tuples with $NIL > 0$.
+
+Just as each methylation call at a 1-tuple comes from a single read, each methylation call at an m-tuple comes from a single read.
+For an m-tuple there are $2^{m}$ possible methylations calls.
+For example, at a 1-tuple there are $2^{1}$ possible methylation calls -- $M$ or $U$; at a 3-tuple there are $2^{3} = 8$ possible methylation calls -- $MMM, MMU, MUM, MUU, UMM, UMU, UUM$ or $UUU$.
+
+For each m-tuple, I also define the intra-pair distance (_IPD_) as the vector containing the $m - 1$ pair-wise distances (measured in bp) between methylation loci in the m-tuple.
+For example, the 2-tuple (chr7:+:145, chr7:+:163) has $IPD = (163 - 145) = (18)$.
+The 5-tuple (chr2:-:560, chr2:-:570, chr2:-:572, chr2:-:588, chr2:-:612) has $IPD = (570 - 560, 572 - 570, 588 - 572, 612 - 588) = (10, 2, 16, 24)$.
+The IPD vector of a 1-tuple is undefined.
+In Chapter \ref{chap:co-methylation} I use the IPDs, along with other features such as the genomic context, to define 'similar' m-tuples.
+
+To illustrate several of the above-mentioned concepts, suppose we sequence a region of the genome containing five methylation loci with three paired-end reads (`A`, `B` and `C`):
+
+```
+ref: 1    2   3 4 5
+A_1: |----->
+A_2:         <------|
+B_1:
+B_2: |----->   <----|
+C_1:    |----->
+C_2:      <------|
+```
+
+If we are interested in 1-tuples, then we would obtain the following from each read by running `methtuple`:
+
+```
+A: {1}, {2}, {3}, {4}, {5}
+B: {1}, {2}, {4}, {5}
+C: {2}, {3}, {4}
+```
+
+The result is identical regardless of whether the `--all-combinations` flag is set.
+
+If we are interested in 3-tuples, then we would obtain the following from each read by running `methtuple` in its default mode:
+
+```
+A: {1, 2, 3}, {2, 3, 4}, {3, 4, 5}
+B: {1, 2, 4}, {2, 4, 5}
+C: {2, 3, 4}
+```
+
+Things to note:
+
+- Read-pair `A` sequences all three (= 5 - 3 + 1) adjacent 3-tuples
+- Read-pair `B` sequences none of the adjacent 3-tuples but does 'erroneously' construct two 3-tuples with $NIL > 0$.
+This happens because m-tuples are created independently from each read-pair; effectively, read-pair `B` is unaware of methylation locus `3`.
+Depending on the downstream analysis, you may want to _post-hoc_ filter out these m-tuples with $NIL > 0$.
+- The twice-sequenced methylation loci, `2` and `3`, in read-pair `C` are not double counted.
+
+However, if we were to run `methtuple` with `--all-combinations` then we would obtain:
+
+```
+A: {1, 2, 3}, {2, 3, 4}, {3, 4, 5}, {1, 2, 4}, {1, 2, 5}, {1, 3, 4}, {1, 3, 5}, {1, 4, 5}, {2, 3, 5}, {2, 4, 5}
+B: {1, 2, 4}, {2, 4, 5}, {1, 2, 5}, {1, 4, 5}
+C: {2, 3, 4}
+```
+
+\subsection{Motivation}
+
+`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software to perform methylation calling at m-tuples from bisulfite-sequencing data.
+Table \ref{tab:other_m-tuple_methylation_callers} lists other software that can perform methylation calling at m-tuples.
+However, none of these existed when I started writing `methtuple` nor do any of them do exactly what I require.
+To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome data.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lllll@{}}
+\toprule
+Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
+\code\{methclone\} & \cite\{Li:2014ei\}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf\{TODO: Try installing.
+Suspect you can't get actual methylation calls.\} \\
+\code\{methpat\}   & \url\{https://github.com/bjpop/methpat\} & Output of \code\{bismark\_methylation\_extractor\} & Yes (amplicons)                  &                                                                                  \\
+\code\{DMEAS\}     & \cite\{He:2013cj\}                       & Output of \code\{bismark\_methylation\_extractor\} &                                  & Windows operating system only.
+Perl code only available as PDF file.
+\\ \bottomrule
+\end{tabular}
+\caption{My caption}
+\label{tab:other_m-tuple_methylation_callers}
+\end{table}
+
+\subsection{Methods}
+
+`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`.
+The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program.
+There are many useful options to filter reads and read-positions.
+Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting these overlapping bases.
+`methtuple` also allows the user to filter out specific read positions rather than wholesale filtering of the ends of reads.
+This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA}.
+Such a 'spike' can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.
+
+`methtuple` is written in Python and uses the `pysam` ([https://github.com/pysam-developers/pysam/](https://github.com/pysam-developers/pysam/)) module to parse the `BAM` file.
+It is compatible with both Python2 and Python3.
+
+`methtuple` is currently limited to processing files produced by `Bismark` due to its reliance on the `Bismark`-specific tags `XM`, the "methylation call string",  `XR`, the "read conversion state for the alignment", and `XG`, the "genome conversion state for the alignment" ([http://www.bioinformatics.bbsrc.ac.uk/projects/bismark/Bismark_User_Guide_v0.13.0.pdf](http://www.bioinformatics.bbsrc.ac.uk/projects/bismark/Bismark_User_Guide_v0.13.0.pdf)).
+It could be extended to work with other bisulfite-sequencing aligners.
+However, due to the eccentricities of each aligner, such an extension would have to be aligner-specific and therefore a considerable undertaking.
+Each extension would require that tags analogous to the `XR`, `XG` and `XM` tags can be generated from the given `BAM` file.
+In the case of the `XM` tag, this would likely require that the reference genome is parsed in parallel with the `BAM` file, adding considerable computational overhead.
+Perhaps the easiest option would be to add a script that 'Bismark-ifies' the original `BAM` file.
+Since all my data are aligned with Bismark, or were converted to Bismark's `BAM` format, I have not yet had need to pursue this line of work.
+
+\subsection{Performance}
+
+I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples from the EPISCOPE, Lister, Seisenberger and Ziller datasets.
+To improve the performance I process split each sample by chromosome and process each chromosome in parallel.
+This is implemented in a helper script ([https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh](https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh)) that makes extensive use of `GNU parallel` \citep{Tange:2011ty}.
+
+Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets.
+For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see __APPENDIX__ for details of these machines).
+
+The running time of \texttt{methtuple} is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its ploidy.
+The running time is largely independent of the tuple size (\texttt{-m}).
+The variation in running times within a chromosome is due to the number of reads generated per sample and the length of the reads, where the length of a paired-end read is defined as the sum of the mates' lengths.
+Samples with more reads take longer to process and samples sequenced with longer reads take longer because these contain more m-tuples.
+
+The maximum memory usage is not strictly proportional to chromosome length.
+It is instead driven by the number and density of CpGs on the chromosome.
+For example, \texttt{chr19}, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than \texttt{chr18}, which has less than half the CpG density of \texttt{chr19} (see __APPENDIX__).
+The relationship between the maximum memory usage and the tuple size (\texttt{-m}) is complex; more data have to be retained as \texttt{-m} increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
+Therefore, memory usage is relatively constant across values of \texttt{-m} for a given chromosome.
+The obvious exception is for tuple size \texttt{2ac}, which used the \texttt{--all-combinations} flag.
+This means that all 2-tuples with $NIL \geq 0$ were computed and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the enormous increase in memory usage.
+
+```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \code{GNU time} converted from seconds to minutes.
+The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
+library(dplyr)
+library(ggplot2)
+library(stringr)
+library(gridExtra)
+
+# facet_wrap_labeller() from http://stackoverflow.com/a/19298442
+facet_wrap_labeller <- function(gg.plot, labels = NULL) {
+  #works with R 3.0.1 and ggplot2 0.9.3.1
+
+  g <- ggplotGrob(gg.plot)
+  gg <- g$grobs      
+  strips <- grep("strip_t", names(gg))
+
+  for(ii in seq_along(labels))  {
+    modgrob <- getGrob(gg[[strips[ii]]], "strip.text", 
+                       grep=TRUE, global=TRUE)
+    gg[[strips[ii]]]$children[[modgrob$name]] <- editGrob(modgrob, 
+                                                          label = labels[ii])
+  }
+
+  g$grobs <- gg
+  class(g) = c("arrange", "ggplot",class(g)) 
+  g
+}
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
+# Time
+time <- lapply(paste0("../data/methtuple_performance/time/", 
+                      list.files("../data/methtuple_performance/time/")), 
+               read.table)
+names(time) <- gsub(".time", "", 
+                    list.files("../data/methtuple_performance/time/"))
+n <- str_split(names(time), "\\.")
+time <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1), 
+                                             times = sapply(time, nrow)),
+                                         levels = 
+                                           paste0("chr", 
+                                                  c(1:22, "X", "Y", "M", "L"))),
+                   "Tuple" = rep(sapply(n, "[", 2),
+                                      times = sapply(time, nrow)),
+                   "Time" = unlist(time))
+
+gg_time <- time %>%
+  ggplot(aes(x = Chromosome, y = Time / 60)) + geom_boxplot() +
+  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Running time (minutes)") + 
+  ggtitle("Running time of methtuple across 49 samples")
+facet_wrap_labeller(gg_time, paste0("Tuple size: ", unique(time$Tuple)))
+```
+
+```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_memory", fig.cap = "; the suffix 'ac' means that the option \texttt{--all-combinations} was set.
+This option creates all tuples with $NIL \geq 0$ which is far greater than the equivalent tuples with $NIL = 0$.
+The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
+These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865})", fig.width = 8, dev = "pdf"}
+# Memory
+memory <- lapply(paste0("../data/methtuple_performance/memory/", 
+                      list.files("../data/methtuple_performance/memory/")), 
+               read.table)
+names(memory) <- gsub(".memory", "", 
+                    list.files("../data/methtuple_performance/memory/"))
+n <- str_split(names(memory), "\\.")
+# NB: Need to divide reported memory usage by 4 due to bug in 'time'
+# (https://bugzilla.redhat.com/show_bug.cgi?id=703865)
+memory <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1), 
+                                             times = sapply(memory, nrow)),
+                                         levels = 
+                                           paste0("chr", 
+                                                  c(1:22, "X", "Y", "M", "L"))),
+                   "Tuple" = rep(sapply(n, "[", 2),
+                                      times = sapply(memory, nrow)),
+                   "Memory" = unlist(memory) / 4)
+gg_memory <- memory %>%
+  ggplot(aes(x = Chromosome, y = Memory / 10^6)) + geom_boxplot() +
+  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Memory usage (GB)") + 
+  ggtitle("Maximum memory usage of methtuple across 49 samples")
+facet_wrap_labeller(gg_memory, paste0("Tuple size: ", unique(memory$Tuple)))
+```
+
+\subsection{Availability}
+
+`methtuple` is open-source software release under the MIT licence and available from [https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple).

### 2014-11-27 Rmarkdown/appendix.Rmd
Updated appendix

@@
   geom_hline(yintercept = 1, linetype = 2)
+```
+
+### Frequency of CpGs on each chromosome in the human reference genome (hg19)
+
+Chromosome 19 has the highest CpG density of the autosomes in the human reference genome (hg19):
+```{r cpg_frequency}
+sort(df['CG', ] / colSums(df) * 100, decreasing = TRUE)
 ```
@@
 
+## Computing details
+
+All computational work was performed on one of the Bioinformatics Division's HP Blade servers.
+These are shared-use, shared-memory machines.
+The basic specifications are shown in Table \ref{tab:unices}.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lllll@{}}
+\toprule
+Machine name & Processors                           & Cores per processor & Total number of cores & RAM    \\ \midrule
+unix88       & $4\times$ Intel Xeon X7350 @ 2.93GHz & 4                   & 16                    & 128 GB \\
+unix301      & $4\times$ AMD Opteron 8435 @ 2.6GHz  & 6                   & 24                    & 256 GB \\
+unix302      & $4\times$ AMD Opteron 6174 @ 2.2GHz  & 12                  & 48                    & 512 GB \\
+unix303      & $4\times$ AMD Opteron 6176 @ 2.3GHz  & 12                  & 48                    & 512 GB \\
+unix305      & $4\times$ AMD Opteron 6276 @ 2.3GHz  & 16                  & 64                    &        \\ \bottomrule
+\end{tabular}
+\caption{Bioinformatics Division server specifications.}
+\label{tab:unices}
+\end{table}
+
 ## Software details

### 2014-11-28 Rmarkdown/wgbs_statistical_analysis.Rmd
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.
Keep on truckin

--- 2014-11-26 Rmarkdown/a_statistical_model_of_methlyC-seq.Rmd
@@
 title: "Statistical analysis of whole-genome bisulfite-sequencing data"
-author: "Peter Hickey"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
 
+\chapter{Statistical analysis of whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_analysis}
+
 \section{Chapter overview}\label{sec:statmodel_chapter_overview}
@@
 
-In this chapter I set out a statistical framework for analysing bisulfite sequencing data.
+I set out a statistical framework for analysing bisulfite-sequencing data.
 I begin by explaining the various levels of stochasticity in a methylC-seq experiment.
 Following this, I define the mathematical notation that I use throughout my thesis.
-I describe key variables, common estimators of these and their statistical properties, and statistical models used in downstream analyses.
+I describe key variables, common estimators of these and their statistical properties.
+The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.
 
-To begin, I consider the simple experiment of performing methylC-seq on a single sample.
-Once I have described the framework for a single sample, I extend it to multiple samples.
-The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.
+In the second half of the chapter I use this framework to discuss downstream analyses of whole-genome bisulfite-sequencing data.
 
@@
 
-- [x] DONE
+Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment.
+This variation has a hierarchical structure, which is illustrated in __FIGURE__ and can be separated into pre-sequencing and post-sequencing sources of variation.
 
-Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment.
-This variation has a hierarchical structure, which I first separate into two levels, as illustrated in __CARTOON FIGURE OF METHYLC-SEQ EXPERIMENT__:
-
-1. Pre-sequencing
-2. Post-sequencing
-
-\subsection{Single locus analysis}\label{sec:one_sample_single_locus}
-
-- [x] DONE
-
-As described in Chapter \ref{chap:wgbs_analysis}, most analyses to date of WGBS data have focused on the methylation level at individual loci averaged across all cells in the sample.
-For this reason, I initially describe the framework in terms of single loci.
-However, bisulfite sequencing data contain much more information than provided by these univariate, marginal summaries of the data.
-In Section \ref{sec:one_sample_mtuples} I introduce _m-tuples_ that can summarise some of this extra information and which I use extensively in Chapter \ref{chap:comethylation}.
-
-\subsubsection{Pre-sequencing}\label{sec:one_sample_pre_sequencing}
-
-- [x] DONE
+\subsection{Pre-sequencing}\label{sec:one_sample_pre_sequencing}
 
@@
 The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus with respect to the forward strand, e.g.
-chr1:723,461.
+chr1:+:666.
 I frequently refer to loci by the subscript $i$ rather than by $pos_{i}$.
 This means that the distance between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome and, for a small number of instances, that the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes.
-Generally, $N_{loci}$ is not known exactly, although estimates may be made based from the sample or a reference genome, but this is no great concern.
+Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome.
+Not knowing the exact value of $N_{loci}$ is no great concern.
 
-The methylation state of a locus can vary within a sample due to the fact that DNA for each sample is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile.
+The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state.
+As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile.
 Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states.
@@
 In general, $H_{i}$ is unknown and will vary from locus to locus within a sample[^H_i].
-Note that the value of $H_{i}$ is determined following the library preparation, including PCR amplificatin of the DNA; therefore, it can give a grossly distorted picture of the true representation of the cells.
+Note that the value of $H_{i}$ is determined following the library preparation, including PCR amplification of the DNA; therefore, it can give a grossly distorted picture of the true representation of the cells.
 I denote by $\mathcal{H}_{i}$ the set of all fragments containing the $i^{th}$ locus.
@@
 [^H_i]: Knowing $H_{i}$ would require knowing: (1) the number of cells used as input (which might only be known to within an order of magnitude), (2) the ploidy of each cell (generally known) and (3) the number of PCR cycles (generally known).
-But the real problem is that none of the steps in creating the pool of DNA fragments is perfect, in particular, PCR introduces biases -- some molecules are preferrentially amplified while others "drop out".
+But the real problem is that none of the steps in creating the pool of DNA fragments is perfect.
+In particular, PCR introduces biases -- some molecules are preferentially amplified while others 'drop out'.
 So even if we knew (1), (2) and (3) we cannot simply multiply these together to compute $H_{i}$, although this might at least give us a rough estimate.
@@
 Although we do not know the number of fragments in the pool, we can define (and measure) the methylation state of a locus on a single DNA fragment.
-I denote by the indicator random variable, $Z_{h, i}$, the methylation state of $i^{th}$ methylation locus on the $h^{th}$ DNA fragment containing the $i^{th}$ methylation locus:
+I denote by the indicator random variable, $Z_{h, i}$, the methylation state of $i^{th}$ methylation locus on the $h^{th}$ DNA fragment:
 
@@
 
-Related to these is the proportion of fragments that methylated at the $i^{th}$ locus:
+From these we can compute the proportion of fragments that methylated at the $i^{th}$ locus:
+
 \begin{equation*}
@@
 
-Again, I emphasise that $H_{i, j}, Z_{h, i, j}, M_{i, j}, U_{i, j} \text{ and } B_{i, j}$ are unobservable. However, by sequencing the pools of DNA fragments we aim to estimate these variables.
+The above definitions can be extended to individual methylation loci, 1-tuples, to m-tuples. Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.
 
-\subsubsection{Post-sequencing}\label{sec:one_sample_post_sequencing}
-
-- [x] DONE
-
-We do not sequence every fragment in the pool.
-Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments.
-We have a large number(~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
-Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to $H_{i}$.
-
-At this point I ignore reads containing no methylation loci as these are not relevant to this discussion and I make three simplifying assumptions:
-
-1. We perform single-end sequencing.
-2. Sequencing is performed without error.
-3. Read mapping is perfect.
-
-The effect of the first assumption is minor.
-When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without "gaps", i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
-However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur (__FIGURE__):
-
-1.
-There may be gaps due to the insert size being longer than the sum of the read lengths, e.g.
-$(i, i + 1, i + 3, i + 4))$.
-In effect, we have missing data for any intervening methylation loci; the $(i + 2)^{th}$ loci in this example.
-2.
-Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g.
-read_1 gives us $(i, i + 1)$ and read_2 gives us $(i + 1, i + 2, i + 3)$.
-In this example we must use only one of read_1 or read_2 as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
-
-I discuss the implications of the latter two assumptions in Section \ref{sec:sequencing_and_read_mapping_errors}.
-
-Each read measures the methylation state of one or more loci from a single DNA fragment.
-I denote by $\mathcal{R}_i$ the set of all reads containing the $i^{th}$ locus.
-Therefore, the number of reads containing the $i^{th}$ locus is $|\mathcal{R}_{i}|$, where $|\mathcal{R}_{i}| \leq H_{i}$ with strict inequality for almost all $i$.
-
-A single read containing the $i^{th}$ locus is denoted by $z: z \in \mathcal{R}_{i}$; the observed methylation state is given by:
-
-\begin{equation*}
-z: z \in \mathcal{R}_{i} = \left\{
-  \begin{array}{l l}
-    1 & \quad \text{if methylated at the } i^{th} \text{ locus}\\
-    0 & \quad \text{if unmethylated at the } i^{th} \text{ locus}
-  \end{array} \right.
-\end{equation*}
-
-By summing over the number of reads containing the $i^{th}$ locus we obtain the number of reads that are methylated at the $i^{th}$ locus ($m_{i}$) and unmethylated at the $i^{th}$ locus ($u_{i}$):
-
-\begin{align*}
-  m_{i} &= \sum_{z: z \in \mathcal{R}_{i}} z \\
-        &= |\{z: z \in \mathcal{R}_{i}, z = 1 \}| \\
-  u_{i} &= \sum_{z: z \in \mathcal{R}_{i}} (1 - z) \\
-      &= |\{z: z \in \mathcal{R}_{i}, z = 0 \}
-\end{align*}
-
-Similarly, we obtain the proportion of reads that are methylated at the $i^{th}$ locus as:
-\begin{equation*}
-  \beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}
-\end{equation*}
-
-These are the so-called $\beta$-values, which are commonly interpreted as an estimate of the proportion of cells in the sample that are methylated at the $i^{th}$ locus.
-I discuss this interpretation, and other estimators of the  methylation level at a locus, in __SECTION__.
-I now move from single locus summaries to multi-loci summaries using m-tuples.
-
-\subsection{m-tuples: Multiple loci analysis}\label{sec:one_sample_mtuples}
-
-I define an m-tuple to be a tuple of methylation loci, where m = $1, 2, \ldots$ is the size of the tuple.
-For example, a CpG 3-tuple is 3 CpGs.
-A CHH 1-tuple is a single CHH; 1-tuples are the basis of single locus analyses of bisulfite-sequencing data.
-We can learn more from bisulfite-sequencing data by using m-tuples with m $> 1$.
-
-An observation on an m-tuple is the methylation pattern from a single read that overlaps the m-tuple.
-I require that the observation is from a single read as this ensures that each observation ultimately comes from a single cell[^chimeric_reads].
-For example, suppose we have a read containing 3 CpGs -- the first two CpGs are methylated while the last one is unmethylated.
-From this read we can observe 3 $\times$ CpG 1-tuples, 2 $\times$ CpG 2-tuples and 1 $\times$ CpG 3-tuple.
-We can have multiple observations on an m-tuple by observing multiple reads, each containing the m-tuple.
-__FIGURE__ illustrates this example.
-
-[^chimeric_reads]: This ignores chimeric reads, which depending on the sequencing protocol may be a problem.
-
-Note that in __FIGURE__ I have not constructed a 2-tuple using the first CpG and the last CpG.
-In general, I focus on m-tuples where the m methylation loci are adjacent.
-That is, I focus on m-tuples where the _number of intervening loci_ is zero ($NIL = 0$).
-There are 3 reasons for this:
-
-1.
-Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ m-tuples, provided that we restrict ourselves to those m-tuples with $NIL = 0$.
-In contrast, if we allow $NIL \geq 0$ then there are $\binom{l}{m} \sim$ m-tuples.
-2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}
-3.
-Measurability: We cannot observe m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology.
-This is true even when $NIL = 0$ but is especially the case if we allow $NIL \geq 0$.
-
-Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
-
-For each m-tuple, I define the _intra-pair distance_ (IPD) as vector containing the $m - 1$ pair-wise distances (measured in bp) between adjacent methylation loci in the m-tuple.
-For example, the 2-tuple (chr7:145, 163) has $IPD = (163 - 145) = (18)$.
-The 5-tuple (chr2:560, 570, 572, 588, 612) has $IPD = (570 - 560, 572 - 570, 588 - 572, 612 - 588) = (10, 2, 16, 24)$.
-The IPD vector of a 1-tuple is not defined.
-In Chapter \ref{chap:co-methylation} I use the IPDs, along with other features such as the genomic context, to define 'similar' m-tuples.
-
-\subsubsection{Pre-sequencing}
-
-- [x] DONE
-
-Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.
-The remaining definitions are analogous to those for single methylation loci, that is, when m $= 1$.
-
-I denote by the vector of indicator random variable, $Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)}$, the methylation pattern of the m-tuple $(i, i + 1, \ldots, i + m - 1)$ on the $h^{th}$ DNA fragment containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$:
+I denote by the vector of indicator random variables, $Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)}$, the methylation pattern of the m-tuple $(i, i + 1, \ldots, i + m - 1)$ on the $h^{th}$ DNA fragment containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$:
 
@@
 There are $2^{\text{m}}$ possible methylation patterns at an m-tuple.
-Rather than denote a methylation pattern by a m-vector of zeros and ones, I also write these using $U$ and $M$; for example, the possible methylation patterns at a 2-tuple are $MM, MU, UM$ and $UU$.
+Rather than describe a methylation pattern by a m-vector of zeros and ones, I also write these using $U$ and $M$; for example, the possible methylation patterns at a 2-tuple are $MM, MU, UM$ and $UU$.
 
-Analogously to the definition of $M_{i}$ and $U_{i}$ and $m_{i}$ and $u_{i}$ for the case where $\text{m} = 1$, we have when $\text{m} = 2$:
+Analogous to the definition of $M_{i}$ and $U_{i}$ for 1-tuples ($\text{m} = 1$), we have when $\text{m} = 2$:
 
@@
 
-The definitions for $\text{m} > 3$ follow in the obvious manner.
+The definitions for $\text{m} > 2$ follow in the obvious manner.
 
@@
 
-\subsubsection{Post-sequencing}
+\subsection{Post-sequencing}\label{sec:one_sample_post_sequencing}
 
-- [x] DONE
+In a whole-genome bisulfite-sequencing experiment we do not sequence every DNA fragment in the pool.
+Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments.
+We have a large number(~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
+Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to $H_{i}$.
 
-The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_i$.
+At this point I ignore reads that do not contain any methylation loci as these are not relevant to this discussion and I make three simplifying assumptions:
+
+1. Sequencing is performed without error.
+2. Read mapping is perfect.
+3. We perform single-end sequencing.
+
+The effect of the first two assumptions are discussed in Section \ref{sec:mapping_and_post-processing}.
+The effect of the third assumption is minor.
+When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
+However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur (__FIGURE__):
+
+1.
+There may be gaps due to the insert size being longer than the sum of the read lengths, e.g.
+$(i, i + 1, i + 3, i + 4))$.
+In effect, we have missing data for any intervening methylation loci; the $(i + 2)^{th}$ loci in this example.
+2.
+Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g.
+read_1 gives us $(i, i + 1)$ and read_2 gives us $(i + 1, i + 2, i + 3)$.
+In this example we must use only one of read_1 or read_2 as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
+
+Each read measures the methylation state of one or more loci from a single DNA fragment.
+I denote by $\mathcal{R}_i$ the set of all reads containing the $i^{th}$ locus.
+Therefore, the number of reads containing the $i^{th}$ locus is $|\mathcal{R}_{i}|$, where $|\mathcal{R}_{i}| \leq H_{i}$ with strict inequality for almost all $i$.
+
+A single read containing the $i^{th}$ locus is denoted $z: z \in \mathcal{R}_{i}$ and the observed methylation state is indicated by:
+
+\begin{equation*}
+z: z \in \mathcal{R}_{i} = \left\{
+  \begin{array}{l l}
+    1 & \quad \text{if methylated at the } i^{th} \text{ locus}\\
+    0 & \quad \text{if unmethylated at the } i^{th} \text{ locus}
+  \end{array} \right.
+\end{equation*}
+
+By summing over the number of reads containing the $i^{th}$ locus we obtain the number of reads that are methylated at the $i^{th}$ locus ($m_{i}$) and unmethylated at the $i^{th}$ locus ($u_{i}$):
+
+\begin{align*}
+  m_{i} &= \sum_{z: z \in \mathcal{R}_{i}} z \\
+        &= |\{z: z \in \mathcal{R}_{i}, z = 1 \}| \\
+  u_{i} &= \sum_{z: z \in \mathcal{R}_{i}} (1 - z) \\
+      &= |\{z: z \in \mathcal{R}_{i}, z = 0 \}
+\end{align*}
+
+From these we can compute the proportion of reads that are methylated at the $i^{th}$ locus as:
+
+\begin{equation*}
+  \beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}
+\end{equation*}
+
+These are the so-called $\beta$-values, which are commonly interpreted as an estimate of the proportion of cells in the sample that are methylated at the $i^{th}$ locus.
+I discuss this interpretation, and other estimators of the  methylation level at a locus, in Section \ref{sec:estimating_B}.
+
+Again, these definitions can be extended from 1-tuples to m-tuples.
+The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$.
 A single read containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, and the observed methylation state is given by:
@@
 
-The definitions for $\text{m} > 2$ follow in the obvious manner.
-
 We can extend the $\beta_{i}$ values to m-tuples. Here are the definitions for $\text{m} = 2$:
@@
 
-The definitions for $\text{m} > 3$ follow in the obvious manner.
+The definitions for $\text{m} > 2$ follow in the obvious manner.
 
-Finally, \cite{Landan:2012kp} compute the average methylation of a read containing the m-tuple $(i, \ldots, i + m - 1)$.
-For each read, $z \in \mathcal{R}_{(i, i + 1, i + 2, i + 3)}$, the average methylation of the read, $\zeta_{z}$, is defined as the proportion of methylation loci in the read that are methylated.
+One final definition is the average methylation of each read, which is used by some authors, including \cite{Landan:2012kp}.
+For each read, $z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, the average methylation of the read, $\zeta_{z}$, is defined as the proportion of methylation loci in the read that are methylated.
 Thus, $\zeta_{z} = 0, \frac{1}{m}, \frac{2}{m}, \ldots, 1$.
@@
 \subsection{Some complications for a single sample}\label{sec:one_sample_complications}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
-As mentioned in Chapter \ref{chap:wgbs_analysis}, studies using bisulfite-conversion assays rely on either a reference genome or, less commonly, separate DNA sequencing of the sample that is assayed.
+As mentioned in Chapter \ref{chap:wgbs_bioinformatics_analysis}, studies using bisulfite-conversion assays rely on either a reference genome or, less commonly, separate DNA sequencing of the sample that is assayed.
 Different analysis strategies lead to different definitions of $\mathcal{I}$, which are approximations to the 'true' set of methylation loci in the sample, $\mathcal{I}^{truth}$.
@@
 A conservative approach that removes many sites of genetic variation between the sample and the reference as well as sites that do not display genetic variation between the sample and the reference.
-Misses sample-specific methylation loci.
+This approach cannot identify sample-specific methylation loci.
 3.
-$\mathcal{I}^{BisSNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}.
+$\mathcal{I}^{Bis-SNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}.
 Identifies sample-specific methylation loci and removes reference-specific methylation loci.
-The best that can be done if only bisulfite-sequencing data is available.
-Genetic variation detection is not as good as that from DNA-seq data.
-This is the approach I favour.
+This is the best approach if only bisulfite-sequencing data are available.
 4.
@@
 All methylation loci are defined with respect to the sample's genome.
-The only differences between $\mathcal{I}^{WGS}$ and the "truth" are due to sequencing errors and variant calling errors of the WGS data.
+The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth} are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.
 
@@
 
-- [x] DONE
+The genome of a diploid organism has sites that are heterozygous due to differences between the maternal and paternal chromosomes.
+Such heterozygous loci are sometimes also methylation loci.
+For example, a locus where the maternal chromosome is a CpG and the paternal chromosome is an ApG.
+In effect, the maternal and paternal chromosomes within the sample have different $\mathcal{I}^{truth}$.
 
-In a diploid cell there are are sites where, for example, the maternal chromosome is a CpG whereas the paternal chromosome is an ApG.
-In effect, the maternal and paternal chromosomes within the sample have different $\mathcal{I}^{truth}$.
-The number of these heterozygous methylation loci is often small enough not to worry about.
-However, in some studies, such as those of allele-specific methylation, these loci can be very important.
-
-Parent-specific methylation loci can be identified by calling heterozygous genetic variants using `Bis-SNP` or from WGS of the sample.
-In practice, the existance of parent-specific methylation loci is often ignored.
-
-\subsubsection{Sequencing and read mapping errors}\label{sec:sequencing_and_read_mapping_errors}
-
-- [x] DONE
-
-Sequencing errors from the Illumina technology are typically base substitutions, for example, a _G_ being read as a _T_.
-These errors occur more frequently at the 3' end of the reads.
-Other sequencing technologies have different error profiles.
-In DNA sequencing, the base quality scores are used to filter out or down-weight likely sequencing errors.
-However, because the base quality scoring algorithm is tuned for DNA-seq where all 4 nucleotides appear frequently, the base quality scores of methylC-seq data may not be as reliable since the relative frequencies of the 4 nucleotides is quite different.
-
-There are two negative effects of sequencing errors.
-Firstly, if a methylated cytosine on the forward strand, which should be a C in the read, is erroneously read as another base -- A, G or T -- then the methylation signal is lost.
-If the erroneous call is an A or a G then at least we might be able to infer that a sequencing error occured.
-However, if the erroneous basecall is a T then we might instead mistakenly conclude that the cytosine was unmethylated.
-
-Secondly, sequencing errors make read mapping more difficult.
-These issues are discussed in Chapter \ref{chap:wgbs_analysis}.
-In terms of the statistical framework described in this chapter, the effect of both sequencing errors and read mapping errors are to introduce errors into $m$ and $u$ and all resulting statistics.
+The number of these genetically heterozygous methylation loci is often small enough not to worry about.
+However, in some studies, such as those of allele-specific methylation, these loci can be very important and might be identified by calling heterozygous genetic variants using `Bis-SNP` or from WGS of the sample.
+In practice, the existence of such loci is often ignored.
 
@@
 
-- [x] DONE
-
-To move from a single sample to multiple samples simply requires an additional subscript, $j = 1, \ldots, n$, where $n$ is the number of samples.
+From a purely notational perspective, the move from a single sample to multiple samples simply requires an additional subscript, $j = 1, \ldots, n$, where $n$ is the number of samples.
+This defines the three levels in the hierarchy of a typical experiment -- DNA fragments ($h$), methylation loci ($i$) and samples ($j$).
 For example, $\mathcal{I}_{j}$ is the set of methylation loci in the $j^{th}$ sample and $\beta_{i, j}$ is the beta value for the $i^{th}$ locus in the $j^{th}$ sample.
 
-This defines the three levels in the hierarchy of a typical experiment -- individual molecules ($h$), individual methylation loci ($i$) and individual samples ($j$).
-A fourth level is how the samples relate in terms of an outcome of interest, such as phenotype.
-This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between pre-defined groups of samples, or the aim of the experiment might be to discover this level, such as in a clustering analysis.
+A fourth level is how the sample's relate to one another, such as the treatment group of each sample.
+This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between tumour and normal tissue.
+Conversely, the aim of the experiment might be to _discover_ this grouping, such as in a clustering analysis.
 
-A common experiment of the first type is the two-group design in which $n_{1}$ samples are from group $1$ and $n_{2}$ samples are from group $2$ ($n_{1} + n_{2} = n$).
-This can be represented by a design matrix $X = [X_{j}]$, where $X_{j} = 1$ if the sample is from group $1$ and $X_{j} = 0$ if the sample is from group $2$.
-
-More generally, we can study multi-group experiments via a suitable definition of the design matrix $X$.
+We can define this fourth level using a design matrix $X = [X_{j}]$.
+For example, in a two-group experiment $X_{j} = 1$ if the sample is from group $1$ and $X_{j} = 0$ if the sample is from group $2$.
 We can also include covariates in the standard way by allowing $X_{j}$ to be a row vector, $X_{j} = (x_{1, j}, \ldots, x_{P, j})$, where $x_{p, j}$ encodes the information on the $p^{th}$ covariate for the $j^{th}$ sample.
 
-The second type of experiment is one that aims to cluster individuals based on methylation patterns.
-This can be thought of as an experiment where the outcome of interest is unknown and we wish to discover it by clustering together samples with 'similar' methylation patterns.
-We might also wish to then correlate these clusters with some external information, such as disease-status.
+\subsection{Some complications with multiple samples}\label{sec{multi_sample_complications}}
 
-\subsection{Some complications with multiple samples}
+In addition to the complications of the Section \ref{sec:one_sample_complications}, we now have sample-to-sample variation that must be addressed by this framework.
 
-- [x] DONE
+\subsubsection{$\mathcal{I}$}
 
-In addition to the complications of the previous section, we now have sample-to-sample variation that must be addressed by this framework.
-
-\subsubsection{}$\mathcal{I}$}
-
-- [x] DONE
-
-Each individual has their own set of methylation loci, that is, $\mathcal{I}_{j}$ differs for all $j$.
+Each sample has their own set of methylation loci, that is, $\mathcal{I}_{j}$ differs between $j$.
 Furthermore, sequencing coverage varies from sample-to-sample.
@@
 
-- [x] DONE
-
-In this section I summarise current techniques for parameter estimation from WGBS data.
-I do not describe the processing of the raw data.
+The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}}$.
+To do so it is necessary to estimate $M$ and $U$.
+In this section I review various methods for estimating these key parameters.
 When necessary, I have 'translated' the original work into my notation to make these methods more readily comparable.
-All methods estimate parameters from files containing the aligned reads.
+I have suppressed $j$ subscript when referring to a single sample.
 
@@
 
-- [x] DONE
-
-The simplest and most commonly used estimators of $M_{i}$ and $U_{i}$ are $m_{i}$ and $u_{i}$, that is, the number of reads that are methylated and unmethylated at the $i^{th}$ locus (e.g., \cite{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg, Hansen:2011gu}).
+The simplest and most commonly used estimators of $M_{i}$ and $U_{i}$ are $m_{i}$ and $u_{i}$, that is, the number of reads that are methylated and unmethylated at the $i^{th}$ locus, subject to some filtering of the reads and read-positions \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg, Hansen:2011gu}).
 $M_{i}$ and $U_{i}$ may be estimated per-strand or aggregated across strands in the case of palindromic methylation loci.
@@
 
-Not all reads, nor positions within reads, are equally treated in this counting process.
-Essentially, each sequenced nucleotide is assigned a weight.
-By far the most used weighting scheme is a set of filters that assigns each sequenced nucleotide a weight of 0 (observation excluded) or 1 (observation included).
-More complicated weighting schemes might include weighting by mapping quality and by base quality, although this would likely lead to fractional estimates of $M$ and $U$, which may not be desirable.
+`methtuple` uses this simple method to estimate $MM_{(i, i + 1)}$, $MU_{(i, i + 1)}$, etc., that is, the frequency of methylation patterns at m-tuples.
 
-`methtuple` uses the same 'filter and count' method to estimate $MM_{(i, i + 1)}$, $MU_{(i, i + 1)}$, etc., that is, the frequency of methylation patterns at m-tuples.
+\subsection{Estimating $B$}\label{sec:estimating_B}
 
-\subsection{Estimating $B$}
+Most analyses of bisulfite-sequencing data have focused on estimating the average methylation level at individual methylation loci, $B_{i}$.
+The simplest estimator is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
 
-- [ ] DONE
+Under a Binomial model for the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample, $M_{i, j} | n_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.
+It should be noted that this is conditional on the observed sequencing coverage, $n_{i, j}$.
 
-Most analyses of bisulfite-sequencing data have focused on estimating the average methylation level at individual methylation loci, $B_{i, j}$.
-The simplest estimator is $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$, which has been widely used \citet[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
-Under a Binomial model for the number of methylated reads at a locus, $M_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, where $n_{i, j}$ is the sequencing coverage of the $i^{th}$ locus in the $j^{th}$ sample, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.
-
-Recently, more sophisticated methods have been proposed to model the average methylation level.
+Recently, more sophisticated methods have been proposed to estimate or model the average methylation level.
 These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and the smoothing-methods proposed by \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
-The aim of these methods isn't necessarily to estimate $B_{i, j}$ - it is typically to identify differential methylation (discussed below) - but estimates of $B_{i, j}$ can be obtained if so desired.
+The aim of these methods isn't necessarily to estimate $B_{i, j}$ - rather, it is typically a step along the way to identify differential methylation - but estimates of $B_{i, j}$ can be obtained if so desired.
 
@@
 
-- [x] DONE
-
 Several papers have proposed the beta-binomial distribution as a natural model "for describing methylation levels of an individual site across replicates" \cite{Dolzhenko:2014bo}.
 The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing.
-Another way to think of this is that the Beta component models the biological variability of the data, while the Binomial component models the sampling variability.
+Another way to think of this is that the 'beta' component models the biological variability of the data, while the 'binomial' component models the sampling variability of sequencing.
 This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data.
@@
 
-An attractive feature of the beta-binomial distribution is that it can be motivated by, and analysed, using Bayesian statistics, including empirical Bayes methods, or frequentist techniques such as maximum likelihood.
+An attractive feature of the beta-binomial distribution is that it can be motivated by, and analysed with, Bayesian methods, including empirical Bayes methods, or frequentist techniques such as maximum likelihood.
 For example, the software `DSS` \cite{Feng:2014iq} and `MOABS` \citet{Sun:2014fk} both use the beta-binomial distribution in an empirical Bayes analysis of differential methylation from bisulfite-sequencing data.
@@
 
-\subsubsection{Smoothing $\beta$-values}
+\subsubsection{Smoothing $\beta$-values}\label{sec:smoothing_beta-values}
 
-- [x] DONE
-
-BSmooth, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$.
+`BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$.
 Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$.
-Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation, which is discussed in Chapters \ref{chap:co-methylation}.
+Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation} and \ref{chap:co-methylation_review}).
 
 Smoothing is particularly powerful for loci with low sequencing coverage, where the denominator $m_{i, j} + u_{i, j}$ is small and the corresponding standard error of $\beta_{i, j}$ is large.
-The smoothed $\beta$-values, and not the raw $\beta$-values, are then generally used in all downstream analyses.
-
-__TODO: Discuss with Terry - is smoothing of the raw $\beta$-values still useful when you have high-coverage sequencing data?__
+The smoothed $\beta$-values, rather than the raw $\beta$-values, are then generally used in all downstream analyses.
 
 Both `bsseq` and `BiSeq` use a binomial local likelihood smoother.
-In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$.
-The smoothing is local to exploit co-methylation, which is a local phenomenon.
+In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j}$ \eqd Binom(n_{i, j}, B_{i, j})$.
+The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.
 
-In both `bsseq` and `BiSeq` the raw $\beta$-values are weighted according to the binomial likelihood and the kernel function.
+In both `bsseq` and `BiSeq` the raw $\beta$-values are weighted according to the binomial likelihood and a kernel function.
 The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the kernel gives greater weight to those $\beta_{i, j}$ near the centre of the window.
-\citet{Lacey:2013iy} note that loci with very high sequencing coverage will strongly influence the smoother, potentially biasing estimates at neighbouring loci with low coverage.
+\citet{Lacey:2013iy} note that loci with very high sequencing coverage will strongly influence the smoother, potentially biasing estimates at neighbouring loci with lower coverage.
 
@@
 `bsseq` uses a much larger window size than `BiSeq`; the default window size in `bsseq` is one that contains at least 70 CpGs and is at least 2000kb wide, whereas the default window size in `BiSeq` is 80bp, regardless of CpG-density.
-This is due to `BiSeq` being developed for RRBS data, which has a high CpG-density per window, whereas `bsseq` was developed for WGBS data, which has a lower and more variable CpG density per window.
+This is due to `BiSeq` being developed for RRBS data, which has a high CpG-density per window, whereas `bsseq` was developed for whole-genome data, which has a more variable, and lower on average, CpG density per window.
 
@@
 \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} compare the smoothing results of `BiSeq` to `bsseq`.
-Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`, but the comparison is selective and limited to a few regions.
-Furthermore, the comparisons are made using RRBS data, both real and simulated, and, `bsseq` is designed from WGBS[^rrbs_sim].
+Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`, but the comparison is a little unfair since both use RRBS data, for which `bsseq` is not designed[^rrbs_sim].
 
@@
 
-\subsection{A criticism of ``identifying methylcytosines''}{sec:criticism_of_methylcytosines}
-
-\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
-This idea was also used in subsequent publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013] with similarly high profiles.
-I believe that this definition of a "methylcytosine" is an unnecessary source of confusion, as I will explain.
-
-[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
-
-It is necessary to use scare quotes to distinguish the term "methylcytosine", as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from methylcytosine, the standard definition meaning a 5-methylcytosine.
-I will first describe the idea and how it differs from methylation calling.
-Then I will describe how it is used to (1) estimate the number of methylcytosines in the sample, and (2) as a kind of screening procedure that determines whether each cytosines will be used in downstream analyses.
-Finally, I will explain why (1) is a bad estimator and why I think (2) is unnecessary or misguided.
-
-A "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
-Thus, the idea of identifying "methylcytosines" is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
-To do so, \citeauthor{Lister:2008bh} compare each the proportion of methylated reads, the $\beta$-value, to a coverage-specific (and sample-specific) cutoff that determines whether the cytosine is classified as a "methylcytosine" or not.
-
-The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy} and \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
-Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done.
-What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-
-The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
-Mathematically, for each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated "error".
-Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
-
-[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
-
-I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure (__CITE__).
-This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
-thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
-
-The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
-It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
-
-Now to why I don't think this is a very useful procedure.
-Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
-These are two different things.
-If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
-A simple example makes this clear.
-
-Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
-Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
-We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
-From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
-
-Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
-For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
-A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
-
-Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
-While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
-For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
-If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
-
 \section{Statistical properties of $\beta$-values}\label{sec:beta}
 
-- [x] DONE
-
-Under the binomial model, $M_{i, j} \eqd Bin(m_{i, j} + u_{i, j}, B_{i, j})$ , $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt(\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}})$ \citep{Hansen:2012gr}.
+Under the binomial model, ($M_{i, j} | n_{i, j}) \eqd Bin(n_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt(\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}})$ \citep{Hansen:2012gr}.
 The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample.
@@
 
-- [x] DONE
-
 In a study involving multiple samples, the set of $\beta$-values can be summarised as a matrix where each row is a locus and each column is a sample.
+Some values will be missing, either because there was insufficient sequencing coverage to estimate a $\beta$-value or because that locus is not a cytosine for that sample.
 This matrix might be visualised to learn about the distribution of methylation levels, either row-wise, to learn about the variability across samples, or column-wise, to learn about the variability within samples.
 
-Restricting our attention to CpGs, __FIGURE__ shows the column-wise summaries of the EPISCOPE data, the genome-wide distributions of $\beta$-values within each sample.
+Restricting our attention to CpGs, __FIGURE__ shows the column-wise summaries of the EPISCOPE data, that is, the genome-wide distributions of $\beta$-values for each sample.
 What is immediately clear is that these distributions are bimodal.
@@
 
-```{r}
-# TODO: Show a bunch of plots of beta-values for different samples
-```
+__UP TO HERE__
 
-```{r}
-# TODO: Show a bunch of plots of beta-values stratified by CGI-status for different samples
+__TODO: Using the EPISCOPE data create the following:__
+
+- [ ] Genome-wide density
+  - [ ] Stratified by CGI
+- [ ] Differences (or correlations) of $\beta$-values between strands
+
+```{r, eval = FALSE, echo = FALSE}
+# Run on one of the unices
+setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
+library(MethylationTuples)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
+
+x <- readRDS('EPISCOPE_1_tuples_strand_collapsed.rds')
+
+cgi <- read.table('~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt', header = TRUE, stringsAsFactors = FALSE)
+cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
+cgi_status <- overlapsAny(x, cgi)
+
+beta <- methLevel(x, min_cov = 10L)
+
+cn <- colnames(beta)
+beta <- data_frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
+           Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
+           beta = as.vector(beta),
+           CGI = rep(cgi_status, ncol(beta)))
+
+a <- beta[sample(nrow(beta), 100000), ]
+
+ggplot(aes(x = beta), data = a) + 
+  geom_density(aes(linetype = Tissue)) + 
+  facet_grid(. ~ Individual) +
+  thesis_theme
+
+ggplot(aes(x = beta), data = a) + 
+  geom_density(aes(linetype = Tissue)) + 
+  facet_grid(CGI ~ Individual) +
+  thesis_theme
+
 ```
@@
 - DVR: Identify runs of DVCs
+
+\subsection{What is the genome-wide methylation level?}
+
+This is a simple question with a somewhat complex answer.
+
+- Average (weighted) $\beta$-values.
+- Lister's method
+
+\subsubsection{A criticism of ``identifying methylcytosines''}{sec:criticism_of_methylcytosines}
+
+__TODO: Simplify and shorten__
+
+\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
+This idea was also used in subsequent publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013] with similarly high profiles.
+I believe that this definition of a "methylcytosine" is an unnecessary source of confusion, as I will explain.
+
+[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
+
+It is necessary to use scare quotes to distinguish the term 'methylcytosine', as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from the standard definition meaning a 5-methylcytosine.
+A 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
+Thus, the idea of identifying 'methylcytosines' is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
+
+The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
+The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy}.
+\citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
+Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done.
+What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
+
+The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
+Effectively, \citeauthor{Lister:2008bh} are testing the null hypothesis that $\beta_{i, j} = 0$ against the one-side alternative, $\beta_{i, j} > 0$.
+Any site where the null hypothesis is rejected is declared a 'methylcytosine'.
+For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
+Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
+
+[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
+I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated.
+
+The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
+It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
+
+This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
+thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
+This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.
+
+Now to why I don't think this is a very useful procedure.
+Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
+These are two different things.
+If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
+A simple example makes this clear.
+
+Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
+Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
+We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
+From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
+
+Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
+For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
+A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
+
+Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
+While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
+For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
+If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
 

### 2014-12-01 Rmarkdown/wgbs_statistical_analysis.Rmd
Re-factor wgbs_statistical_analysis chapter.

@@
 
-- [x] DONE
-
 I set out a statistical framework for analysing bisulfite-sequencing data.
-I begin by explaining the various levels of stochasticity in a methylC-seq experiment.
-Following this, I define the mathematical notation that I use throughout my thesis.
+I begin by explaining the various levels of stochasticity and then define the mathematical notation that I use throughout my thesis.
 I describe key variables, common estimators of these and their statistical properties.
@@
 
+- Chapter overview
+  - One sample
+    - Pre-sequencing
+    - Post-sequencing
+    - Some complications for a single sample
+      - What is $\mathcal{I}$?
+      - Genetic heterozygosity at a methylation locus
+  - Multiple samples
+    - Some complications with multiple samples
+      - What is $\mathcal{I}$?
+  - Parameter estimation
+    - Estimating $M$, $U$
+    - Estimating $B$
+      - Beta-binomial models
+      - Smoothing
+    - Statistical properties of $\beta$-values
+      - Empirical distributions
+      * Write a function(s) to do the following and apply it to the EPISCOPE, Lister, Seisenberger and Ziller datasets:
+        - Genome-wide distribution of CpG $\beta$-values
+        * Kernel density estimates
+        * Stratify by CGI status, genic, intragenic, repeats
+        * Correlation across strands
+        * Effect of sequencing coverage threshold (`min_cov`)
+      - What is the genome-wide methylation level?
+        * Various figures get bandied about; what's a good summary?
+        - A criticism of ``identifying methylcytosines''
+      - Correlations
+      - Biases
+      - Transformations
+  - Downstream analyses
+    - Differential methylation
+      - Experimental design considerations
+      - DMCs: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
+        - One-group experiments
+        - Two-group experiments
+        - Multi-group experiments
+      - DMRs: Identify runs of DMCs
+        - A priori vs. data-driven regions
+        - One-group experiments
+        - Two-group experiments
+        - Multi-group experiments
+    - Differential variability
+      - DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
+      - DVRs: Identify runs of DVCs
+    - Allele-specific methylation
+    - Methylation entropy
+    - Epipolymorphism
+    - Epiallele detection
+    - Integrating DNA methylation data with other genomes data
+
 \section{One sample}\label{sec:one_sample}
 
-Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment.
-This variation has a hierarchical structure, which is illustrated in __FIGURE__ and can be separated into pre-sequencing and post-sequencing sources of variation.
+Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment. I find it convenient to separate these into:
+
+1. Pre-sequencing sources of variation
+2. Sequencing and post-sequencing sources of variation.
+
+I describe these sources of variation, which are illustrated in Figure \ref{fig:bs-seq_protocol}.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
+\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment.
+Each sample starts as a population of cells, with potentially different methylation patterns.
+Each DNA fragment is coloured by its originating cell, although this is unknown in practice.
+Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
+\label{fig:bs-seq_protocol}
+\end{figure}
 
@@
 A methylation locus is a single cytosine, that is, a CpG, CHG or CHH.
-The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus with respect to the forward strand, e.g.
-chr1:+:666.
-I frequently refer to loci by the subscript $i$ rather than by $pos_{i}$.
-This means that the distance between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome and, for a small number of instances, that the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes.
+The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., chr1:+:666.
 Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome.
 Not knowing the exact value of $N_{loci}$ is no great concern.
+
+It is sometimes convenient to refer to loci by the subscript $i$ rather than by $pos_{i}$.
+It is important to remember that the number of base pairs between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome.
+Also, depending on whether the data are stranded, a pair of loci may be on opposite strands.
+In a small number of instances the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes, e.g., the last CpG on chromosome 1 and the first CpG on chromosome 2.
 
@@
 Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states.
-Therefore, it is also necessary to consider the next level down in the hierarchy; the DNA fragments within the sample.
+Therefore, it is also necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.
 
@@
 
-The above definitions can be extended to individual methylation loci, 1-tuples, to m-tuples. Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.
+The above definitions can be extended from individual methylation loci, 1-tuples, to m-tuples. Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.
 
@@
 
-\subsubsection{$\mathcal{I}$}
+\subsubsection{What is $\mathcal{I}$?}
 
@@
 
-The simplest and most commonly used estimators of $M_{i}$ and $U_{i}$ are $m_{i}$ and $u_{i}$, that is, the number of reads that are methylated and unmethylated at the $i^{th}$ locus, subject to some filtering of the reads and read-positions \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg, Hansen:2011gu}).
-$M_{i}$ and $U_{i}$ may be estimated per-strand or aggregated across strands in the case of palindromic methylation loci.
-CpG methylation is typically aggregated across strands.
-
-`methtuple` uses this simple method to estimate $MM_{(i, i + 1)}$, $MU_{(i, i + 1)}$, etc., that is, the frequency of methylation patterns at m-tuples.
+It is rare to need direct estimates of $M_{i}$ or $U_{i}$.
+In order to estimate $M_{i}$ and $U_{i}$, the absolute number of methylated and unmethylated DNA fragments at the $i^{th}$ locus, we would also require an estimate of $|\mathcal{H}_{i}|$, the number of DNA fragments containing the $i^{th}$ locus.
+Rather, we are generally interested in estimating the proportion of reads that are methylated, $B_{i} = \frac{M_{i}}{M_{i} + U_{i}}$.
 
@@
 
-Most analyses of bisulfite-sequencing data have focused on estimating the average methylation level at individual methylation loci, $B_{i}$.
-The simplest estimator is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
+The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
+The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).
 
 Under a Binomial model for the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample, $M_{i, j} | n_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.
-It should be noted that this is conditional on the observed sequencing coverage, $n_{i, j}$.
 
 Recently, more sophisticated methods have been proposed to estimate or model the average methylation level.
-These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and the smoothing-methods proposed by \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
-The aim of these methods isn't necessarily to estimate $B_{i, j}$ - rather, it is typically a step along the way to identify differential methylation - but estimates of $B_{i, j}$ can be obtained if so desired.
+These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
 
@@
 
-Several papers have proposed the beta-binomial distribution as a natural model "for describing methylation levels of an individual site across replicates" \cite{Dolzhenko:2014bo}.
+Several papers have proposed the beta-binomial distribution since, as noted by \cite{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" .
 The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing.
@@
 
-\subsection{Distributions}
+\subsection{Empirical distributions}
+
+- TODO: Lots of figures and discussion. See chapter plan.
 
@@
 
-Restricting our attention to CpGs, __FIGURE__ shows the column-wise summaries of the EPISCOPE data, that is, the genome-wide distributions of $\beta$-values for each sample.
+\subsubsection{Genome-wide distribution of $\beta$-values}
+
+Restricting our attention to CpGs, Figure \ref{fig:EPISCOPE_beta_distribution} shows the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each EPISCOPE sample.
 What is immediately clear is that these distributions are bimodal.
-This bimodality is driven by the fact that CpGs in CpG islands are generally unmethylated whereas those outside of CpG islands are generally methylated, which we can see when we stratify CpGs by their CpG island status.
+Figure \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} shows that this bimodality is largely driven by the fact that CpGs in CpG islands are mostly unmethylated whereas those outside of CpG islands are generally methylated.
+Intermediate $\beta$-values, those between $0.2-0.8$, are evidence of heterogeneity.
 
-__UP TO HERE__
+Both Figures \ref{fig:EPISCOPE_beta_distribution} and  \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} also reveal that the genome-wide distributions of $\beta$-values for a given tissue are very similar between the three individuals.
+The three adipose-derived tissues, _SA_, _VA_ and _VAT_, have similar genome-wide profiles while the blood-derived tissue, _BUF_, has more intermediate $\beta$-values, potentially indicating more cellular heterogeneity.
 
-__TODO: Using the EPISCOPE data create the following:__
-
-- [ ] Genome-wide density
-  - [ ] Stratified by CGI
-- [ ] Differences (or correlations) of $\beta$-values between strands
-
-```{r, eval = FALSE, echo = FALSE}
+```{r EPISCOPE_genome_wide_beta-value_distributions, eval = FALSE, echo = FALSE}
 # Run on one of the unices
@@
 library(tidyr)
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
+thesis_theme <- theme_classic(base_size = 20)
+thesis_theme <- theme_bw(base_size = 20)
 
-x <- readRDS('EPISCOPE_1_tuples_strand_collapsed.rds')
+x <- readRDS("EPISCOPE_1_tuples_strand_collapsed.rds")
 
-cgi <- read.table('~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt', header = TRUE, stringsAsFactors = FALSE)
+cgi <- read.table("~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt",
+                  header = TRUE, stringsAsFactors = FALSE)
 cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
@@
 cn <- colnames(beta)
-beta <- data_frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
+beta <- data.frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
                    Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
                    beta = as.vector(beta),
-           CGI = rep(cgi_status, ncol(beta)))
+                   CGI = rep(cgi_status, times = ncol(beta)),
+                   stringsAsFactors = FALSE)
+beta <- tbl_df(beta)
 
-a <- beta[sample(nrow(beta), 100000), ]
+#a <- beta[sample(nrow(beta), 100000), ]
 
-ggplot(aes(x = beta), data = a) + 
-  geom_density(aes(linetype = Tissue)) + 
-  facet_grid(. ~ Individual) +
-  thesis_theme
+g <- ggplot(aes(x = beta), data = beta) +
+  geom_density(size = 1.5) +
+  facet_grid(Tissue ~ Individual, labeller = label_both) +
+  thesis_theme + scale_colour_brewer(palette = "Dark2") +
+  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
+  xlab(expression(paste(beta, "-values", sep = "")))
+ggsave("EPISCOPE_beta_distribution.pdf", g, height = 9, width = 16)
 
-ggplot(aes(x = beta), data = a) + 
-  geom_density(aes(linetype = Tissue)) + 
-  facet_grid(CGI ~ Individual) +
-  thesis_theme
-
+g <- ggplot(aes(x = beta), data = beta) +
+  geom_density(aes(colour = CGI), size = 2) +
+  facet_grid(Tissue ~ Individual, labeller = label_both) +
+  thesis_theme + scale_colour_brewer(palette = "Dark2") +
+  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
+  xlab(expression(paste(beta, "-values", sep = "")))
+ggsave("EPISCOPE_beta_distribution_CGI_stratified.pdf", g, height = 9,
+       width = 16)
 ```
 
-To do the same plots for the row-wise summaries would be rather tedious since there are many, many more rows (loci) than columns (samples) in the matrix of $\beta$-values.
-However, a few examples are shown in __FIGURE__.
-These row-wise summaries are often modelled by the beta distribution \cite[e.g.,]{Hebestreit:2013ko, Lacey:2013iy}.
-The Beta distribution is a flexible 2-parameter distribution on $[0, 1]$.
-It can be unimodal, "U"-shaped or "J"-shaped, depending on the choice of parameters.
-The Beta distribution also includes the Uniform and arcsine distributions as special cases (__TODO: better source than wikipedia__).
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
+Most CpGs are highly unmethylated or highly methylated, with the buffy coat samples being the most extreme example of this bimodality.
+'Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+Abbreviations: _BUF_ = buffy coat; _SA_ = subcutaneous adipose; _VA_ = visceral adipose; _VAT_ = adipocytes derived from visceral adipose.}
+\label{fig:EPISCOPE_beta_distribution}
+\end{figure}
 
-The analysis of the row-wise distributions leads to the analysis of differential methylation and differential variability, which are discussed in __SECTION__.
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution_CGI_stratified.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Most CpGs outside of CpG islands are methylated whereas most CpGs within CpG islands are unmethylated.
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+Abbreviations: _BUF_ = buffy coat; _SA_ = subcutaneous adipose; _VA_ = visceral adipose; _VAT_ = adipocytes derived from visceral adipose; _CGI_ = CpG island.}
+\label{fig:EPISCOPE_beta_distribution_CGI_stratified}
+\end{figure}
+
+\subsubsection{Strand-specific summaries}
+
+CpG $\beta$-values are often computed by aggregating the $m$ and $u$ counts across the forward and reverse strands.
+This assumes that the two strands are indeed symmetrically methylated.
+To investigate this, for each CpG we can compute the $\beta$-values using only the forward-strand reads, $\beta^{+}$, and only the reverse-strand reads, $\beta^{-}$.
+
+```{r EPISCOPE_strand_specific_beta-value_distributions, eval = FALSE, echo = FALSE}
+# Run on one of the unices
+setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
+library(MethylationTuples)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+library(RColorBrewer)
+
+thesis_theme <- theme_classic(base_size = 20)
+thesis_theme <- theme_bw(base_size = 20)
+
+y <- readRDS("EPISCOPE_1_tuples.rds")
+y_fwd <- y[strand(y) == "+"]
+y_rev <- y[strand(y) == "-"]
+
+ol <- findOverlaps(unstrand(y_fwd), shift(unstrand(y_rev), -1L), type = "equal")
+beta_fwd <- methLevel(y_fwd, min_cov = 5L)[queryHits(ol), ]
+beta_rev <- methLevel(y_rev, min_cov = 5L)[subjectHits(ol), ]
+
+cors <- matrix(unlist(mclapply(1:10, function(min_cov, y_fwd, y_rev, ol) {
+  beta_fwd <- methLevel(y_fwd, min_cov = min_cov)[queryHits(ol), ]
+  beta_rev <- methLevel(y_rev, min_cov = min_cov)[subjectHits(ol), ]
+  sapply(seq_len(ncol(y)), function(i, x, y) {
+    cor(x[, i], y[, i], use = "complete.obs")
+    }, x = beta_fwd, y = beta_rev)
+  }, y_fwd = y_fwd, y_rev = y_rev, ol = ol, mc.cores = 10)), ncol = 10)
+
+beta_pairs <- data.frame(Individual = rep(substr(cn, 1, 3),
+                                          each = nrow(beta_fwd)),
+                         Tissue = rep(substr(cn, 4, 6), each = nrow(beta_fwd)),
+                         beta_fwd = as.vector(beta_fwd),
+                         beta_rev = as.vector(beta_rev),
+                         stringsAsFactors = FALSE)
+beta_pairs <- tbl_df(beta_pairs)
+
+beta_pairs %>% group_by(Individual, Tissue) %>%
+  summarise(cor(beta_fwd, beta_rev, use = "complete.obs"))
+
+mypar <- function(a = 1, b = 1, brewer.n = 8, brewer.name = "Dark2",
+                  cex.lab = 2, cex.main = 2, cex.axis = 1.5,
+                  mar = c(5.1, 5.1, 3.5, 2.1), mgp = c(3, 1, 0), ...) {
+    par(mar = mar, mgp = mgp, cex.lab = cex.lab, cex.main = cex.main,
+        cex.axis = cex.axis)
+    par(mfrow = c(a, b), ...)
+    palette(brewer.pal(brewer.n, brewer.name))
+}
+pdf("EPISCOPE_beta_correlations_across_strand.pdf", width = 16, height = 9)
+mypar(4, 3)
+sapply(seq_len(ncol(beta_fwd)), function(i, x, y, ns) {
+  smoothScatter(x[, i], y[, i],
+                xlab = expression(beta["+"]),
+                ylab = expression(beta["-"]),
+                main = paste0(ns[i], ": cor = ", round(cor(x[, i], y[, i],
+                                                        use = "complete.obs"),
+                                                    2)))
+  }, x = beta_fwd, y = beta_rev, ns = colnames(beta_fwd))
+dev.off()
+```
+
+\subsection{Interpretation}
+
+In papers about DNA methylation there is often a sentence like the following (given by \cite{Laird:2003kw}): "About 70% of the CpG dinucleotides in the mammalian genome are methylated".
+In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
+However, as can be seen from __FIGURE__, the extreme bimodality of the $\beta$-value distribution means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
+To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
+
+The "$70\%$" statement can also be interpreted as an estimate of the probability that a cytosine randomly sampled from a haploid copy of a mammalian genome is a methylcytosine.
+Because in a whole-genome bisulfite-sequencing experiment we measure methylation from a pool of cells, not a single haploid copy of the genome, I do not think it makes sense to describe a particular cytosine in the genome as being a 'methylcytosine'.
+That is, it only makes sense to describe an observation of a cytosine, $z_{i}$, as being methylated or unmethylated and not to describe a cytosine, $i$, as being a methylated or unmethylated.
+However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have sought to use the latter definition, which I believe to be an unnecessary source of confusion.
+
+[^Lister_2013]: It is worth noting that this concept was not used in more recent paper from the same group \citep{Lister:2013et}.
+
+\citet{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
+In the language of these papers, a 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
+This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$.
+This can be thought of as testing the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
+
+Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows.
+For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
+Any site with an FDR-adjusted P-value below a threshold was declared a 'methylcytosine'[^fdr]
+
+[^mathematical_description]: The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy}.
+\citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}
+
+[^fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
+I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated.
+This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
+thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
+This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.
+
+The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error rates estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
+It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
+
+The proportion of cytosines that are identified by this procedure as 'methylcytosines' is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous.
+For example, suppose we had a sample where the true methylation level of every CpG, $B_{i}$, was $0.2$.
+Given sufficient sequencing coverage, every CpG in the genome would be declared a 'methylcytosine' when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.
+
+Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values.
+For example, in \citet[supplementary figure 2a]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates.
+A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.
+
+In summary, for bisulfite-sequencing experiments where the DNA for each sample comes from multiple cells I do not think it makes sense, nor is it useful, to refer to individual cytosines, $i$, as being methylated or unmethylated.
+Instead, it is better to summarise the methylation level at a CpG by a $\beta$-value since this has a natural interpretation as the estimated proportion of haploid genomes in the sample that are methylated at that CpG.
+Unfortunately, $\beta$-values are not without their own issues, as discussed below.
 
 \subsection{Correlations}
-
-- [x] DONE
 
@@
 Just as we can explore the distribution of DNA methylation levels within a sample or between samples, so too can we explore co-methylation within a sample or between samples.
-I examine this in detail in Chapter \ref{chap:co-methylation}
+I examine this in detail in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
 
 \subsection{Biases}
-
-- [x] DONE
 
@@
 \subsection{Transformations}
-
-- [x] DONE
 
@@
 
-\section{Downstream analyses}\label{sec:downstream}
+\section{Downstream analyses}\label{sec:downstream_analyses}
 
-- [ ] DONE
+As is clear from the length of Chapter \ref{chap:wgbs_bioinformatics_analysis} alone, there is a considerable amount of bioinformatics work before we even consider analysing the data to address the biological question of interest.
+I call these the 'downstream' analyses, and there are many that might be considered.
+Using the above statistical framework, I explain the most commonly asked questions.
 
-- Don't rehash the details of Chapter WGBS analysis. Rather summarise these questions as statistical problems in the framework I outline in this chapter.
-- DMC: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
-- DMR: Identify runs of DMCs
-- DVC: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
-- DVR: Identify runs of DVCs
+\subsection{Differential methylation}
 
-\subsection{What is the genome-wide methylation level?}
-
-This is a simple question with a somewhat complex answer.
-
-- Average (weighted) $\beta$-values.
-- Lister's method
-
-\subsubsection{A criticism of ``identifying methylcytosines''}{sec:criticism_of_methylcytosines}
-
-__TODO: Simplify and shorten__
-
-\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
-This idea was also used in subsequent publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013] with similarly high profiles.
-I believe that this definition of a "methylcytosine" is an unnecessary source of confusion, as I will explain.
-
-[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.
-
-It is necessary to use scare quotes to distinguish the term 'methylcytosine', as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from the standard definition meaning a 5-methylcytosine.
-A 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
-Thus, the idea of identifying 'methylcytosines' is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.
-
-The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy}.
-\citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}.
-Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done.
-What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.
-
-The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
-Effectively, \citeauthor{Lister:2008bh} are testing the null hypothesis that $\beta_{i, j} = 0$ against the one-side alternative, $\beta_{i, j} > 0$.
-Any site where the null hypothesis is rejected is declared a 'methylcytosine'.
-For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
-Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".
-
-[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
-I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated.
-
-The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
-It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
-
-This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
-thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
-This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.
-
-Now to why I don't think this is a very useful procedure.
-Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample.
-These are two different things.
-If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$.
-A simple example makes this clear.
-
-Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy.
-Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy.
-We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy.
-From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".
-
-Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values.
-For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
-A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".
-
-Another example is that only cytosines identified as "methylcytosines" in at least one sample were used in differential methylation analyses \citep{Lister:2009hy, Lister:2011kg}.
-While there are often good reasons for screening loci prior to differential testing \citep[e.g.]{Bourgon:2010cr}, this could potentially bias results when looking for regions of differential methylation.
-For example, suppose there are ten cytosines in a row where the first three and last three are differentially methylated.
-If we ignore the four intervening cytosines, which are not differentially methylated, then we falsely conclude that the entire region is differentially methylated rather than it being two smaller DMRs.
-
-\section{Dump from previous chapter}
-
-\section{Identifying differential methylation}\label{sec:identifying_differential_methylation}
-
-Identifying differential methylation means to discover sites (differentially methylated cytosines or _DMC_s) or regions (differentially methylated regions or _DMR_s) in the genome that have different levels of methylation between two or more conditions.
-Much of the methodology for identifying differential methylation has focused on two-group experiments and a somewhat restricted definition: differential methylation is the difference in __average__ methylation between __two__ groups.
-This is analogous to identifying differential gene expression, which focuses on identifying genes that have different __average__ expression levels between two groups.
-
-While bisulfite-sequencing experiments with multiple groups have been performed and analysed \citep[e.g.,]{Lister:2009hy, Hansen:2011gu, Hansen:2013eo}), these have typically been done using a series of pair-wise comparisons or by comparing each sample in turn against some 'baseline' sample.
-This is because the analysis of a pair-wise comparison is relatively simple and because it is also generally easier to interpret a two-group comparison than a multi-group comparison.
-However, as the number of groups increases it becomes more difficult to interpret multiple pair-wise comparisons.
-
-Like when trying to answer any scientific question, the experimental design is very important in order to reliably infer differential methylation.
-Identifying DMCs is relatively straightforward statistical problem, since it amounts to the well-studied problem of testing the difference of means.
-However, identifying DMRs, which may or may not build upon an initial scan for DMCs, is a far more challenging statistical problem owing to the high spatial dependence of methylation at neighbouring cytosines (discussed in Chapter \ref{chap:co-methylation}) and the uneven spacing of cytosines throughout the genome.
+A common downstream analysis is the identification of differential methylation, for which many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
+This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
 
@@
 
-In any experiment of differential methylation we want the DMCs and DMRs to be both _biologically_ and _statistically_ significant; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
+In any analysis of differential methylation we want the DMCs and DMRs to be both _biologically_ and _statistically_ significant; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
 Key to ensuring biological relevance is good experimental design, such as the use of _replicates_ in each experimental group.
@@
 
-Unfortunately, early experiments with whole-genome bisulfite-sequencing frequently had no replicates of any kind, as is the case of the H9 and IMR90-iPSC cell lines in \cite{Lister:2009hy}.
-Moreover, even when replicates were performed they were frequently pooled pooled prior to analysis, thus ignoring all within-group variability.
+Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind, or pooled the replicates prior to analysis, thus ignoring all within-group variability, as is the case of the H1 and IMR90 cell lines in \cite{Lister:2009hy}.
 
-Ideally, variability between technical replicates should be orthogonal to the biological question, but this rarely occurs.
+Ideally, technical variability is orthogonal to the biological variability, but this rarely occurs.
 Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.
-Again, this emphasises the importance of good experimental design.
+Again, this emphasises the importance of good experimental design such as randomisation, replication and the use of control, both positive and negative.
 
@@
 
-There have been reports of differential methylation at individual CpGs resulting in a phenotypic difference \cite{Furst:2012gh}.
-However, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from WGBS experiments with sample sizes you can count on one or two hands and average sequencing depth of $10-30\times$.
+There have been reports of differential methylation at individual CpGs resulting in a phenotypic difference \citep[e.g.,]{Furst:2012gh}.
+However, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome experiments with sample sizes you can count on one or two hands and an average sequencing depth of $10-30\times$.
+Nonetheless, identifying DMCs, even if not an end in itself, is an important step in identifying DMRs.
 
-Identifying DMCs boils down to identifying differences in means.
-As such, it can be readily framed as a 'stand-alone' test, such as a t-test, or cast into a regression framework.
-There is an enormous body of statistical literature on testing for differences in means.
+Identifying DMCs boils down to identifying differences in the mean level of methylation, for which there is an enormous body of statistical literature.
+The problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.
 
@@
 but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments (e.g., approximately 1 million tests in genome-wide association studies and 20,000 tests in studies of differential gene expression).
-
-However, when identifying DMCs, the _effective_ number of tests is generally fewer due to the high degree of dependence of methylation at neighbouring methylation loci, which ensures a high degree of dependence amongst the tests.
-
-__TODO: What techniques can be used to do FDR, multiple hypothesis testing? Are standard tools applicable in such high-dependence settings?__
-
-\subsubsection{One-group tests}\label{sec:one_group_dmcs}
-
-The one-group WGBS experiment is uncommon and the $A^{vy}$ experiment described in Chapter \ref{chap:Avy} is the only example that I know of.
-In a one-group experiment a DMC is one where at least one of the samples has a different methylation level to the group average.
-Mathematically, we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$, where $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
-This can be tested in a number of ways as I will demonstrate.
-
-The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown below (__TODO: Prettify table with `booktabs`__)
+Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.
 
@@
 \centering
-\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
-\label{my-label}
-\begin{tabular}{c c c}
-\hline
-& m          & u          \\ \hline
+\begin{tabular}{@{}ll@{}}
+\toprule
+Method                                       & Used by                                                                            \\ \midrule
+\cite{Benjamini:1995ws}                      & \cite{Akalin:2012cm, Jaffe:2012gx, Lacey:2013iy, Rijlaarsdam:2014hp, Sofer:2013bk} \\
+\cite{Wang:2011cw}                           & \cite{Akalin:2012cm}                                                               \\
+"False discovery rate correction/adjustment" & \cite{Dolzhenko:2014bo, Gokhman:2014cp, Lister:2009hy, Xie:2014ez}                 \\
+\cite{Storey:2007cp}                         & \cite{Jaffe:2012gx}                                                                \\
+\cite{Benjamini:2001ho}                      & \cite{Sofer:2013bk, Hebestreit:2013ko}                                             \\
+"Bonferonni adjustment"                      & \cite{Feng:2014iq}                                                                 \\ \bottomrule
+\end{tabular}
+\caption{Methods proposed for adjusting for the multiple hypothesis testing performed in an analysis of DMCs.
+Several papers use describe their analysing as performing a "false discovery rate adjustment" or "false discovery rate correction" without explicitly stating what they are doing or citing a reference.
+One paper uses an ultra-conservative Bonferonni correction.}
+\label{tab:multiple_testing}
+\end{table}
+
+One thing to note, however, is that the _effective_ number of tests is fewer than the actual number of tests.
+This is because the methylation levels at neighbouring loci are correlated (see Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}), which means that tests of differential methylation are generally positively correlated, thus reducing the effective number of independent tests.
+The classical Benjamini-Hochberg procedure also controls the false discovery rate under certain forms of positive dependence \citep{Benjamini:2001ho}.
+(__TODO: Do these techniques work in the high-dependence setting of DMC testing?__)
+
+\subsubsection{One-group experiments}
+
+While an uncommon design, I did work on a one-group whole-genome bisulfite-sequencing experiment during my PhD.
+The data, courtesy of Harry Oey and Emma Whitelaw (La Trobe University), came from whole-genome bisulfite-sequencing of five _isogenic_ mice, meaning that these mice were genetically identical (or close to it).
+
+In a one-group experiment a DMC is a cytosine where at least one of the samples has a different methylation level to the group average.
+The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown in Table \ref{tab:one-group_dmc_testing}.
+
+\begin{table}[h]
+\centering
+\resizebox{\textwidth}{!}{%
+  \begin{tabular}{@{}lll@{}}
+  \toprule
+  & m          & u          \\ \midrule
   $Sample_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
   $\vdots$     & $\vdots$   & $\vdots$   \\
-$Sample_{n}$ & $m_{i, n}$ & $u_{i, n}$ \\
+  $Sample_{n}$ & $m_{i, n}$ &            \\ \bottomrule
   \end{tabular}
+}
+\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
+\label{tab:one-group_dmc_testing}
 \end{table}
 
-Such contingency tables can be analysed in many different ways. Here I apply a few methods to highlight their similarities and differences. Firstly, here are some example data for a single CpG:
+Such a contingency table can be analysed in many different ways.
+Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j.
+$\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
+
+To the best of my knowledge, none of the software specifically written for analysing differential methylation analyses is able to test this hypothesis.
+This is because these software are almost all designed for two-group experiments.
+However, this hypothesis is readily testing using routine statistical methods.
+As an example, consider the following data:
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+& m    & u    \\ \midrule
+$Sample_{1}$ & $38$ & $1$  \\
+$Sample_{2}$ & $79$ & $2$  \\
+$Sample_{3}$ & $59$ & $1$  \\
+$Sample_{4}$ & $69$ & $2$  \\
+$Sample_{5}$ & $44$ & $46$ \\ \bottomrule
+\end{tabular}
+\caption{An example of a differentially methylated cytosine in a one-group experiment.}
+\label{tab:one-group_example}
+\end{table}
+
+We could analyse this table using a simple stand-alone test, such as Pearson's $\chi^{2}$ test.
+Alternatively, we could frame this as a generalised linear model, either a log-linear model or a logistic regression.
+The hypothesis is then tested by computing the deviance of the model[^deviance]
+
+[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.
+
+assessing the goodness-of-fit of the 'no interaction' model (equivalently, comparing the saturated model to .
+The following R code shows the equivalence of these approaches for a CpG where the fifth sample is differentially methylated.
 
 ```{r}
+# The data
 x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
@@
 x
+
+# Using Pearson's chi-square test
+chisq.test(x)
+
+# Fitting a log-linear model to perform iterative proportional fitting to the
+# contingency table.
+loglin(x, margin = list("sample", "meth_state"))
+# An alternative formulation using loglm(), a convenient wrapper of loglin(),
+# from the MASS package.
+library(MASS)
+loglm(formula =  ~ sample + meth_state, data = x)
+
+# Fitting a generalised linear model with a binomial link using two different
+# formulations
+# (1) Using the array of frequencies
+glm(x ~ 1, family = binomial)
+# (2) Using relative frequencies (i.e., the beta-values) and weights (i.e.,
+# the sequencing coverage).
+glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[, 1] + x[, 2])
 ```
 
-We see that while `sample_1, ..., sample_4` are nearly completely methylated at this CpG, `sample_5` has approximately $50\%$ methylation, which may be indicative of allele-specific methylation.
+A key advantage of using either of the generalised linear models is the option to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
+More complex regression analyses are also possible, such as empirical Bayes or fully Bayesian models.
 
-In my analysis of the $A^{vy}$ experiment I used the `prop.test` function in R. As documented in the help page for `prop.test`:
+\subsubsection{Two-group experiments}\label{sec:two_group_dmcs}
 
-> `prop.test` can be used for testing the null that the proportions (probabilities of success) in several groups are the same
+__UP TO HERE__
 
-in R to test the equality of the proportions ($\beta$-values) across the five mice.
-
-```{r}
-prop.test(x)
-```
-
-As expected, this test returns a very small P-value, `r print(format.pval(prop.test(x)$p.value), quote = FALSE)`.
-
-What might not be clear from its name is that `prop.test(x)` is just performing Pearson's $\chi^{2}$-test of the null hypothesis that the joint distribution of the cell counts in the $2 \times n$ contingency table is the product of the row and column marginals.
-This can be seen by examining the source code or by comparing the output to `chisq.test(x)`: (__TODO: Check by hand__)
-
-```{r}
-chisq.test(x)
-```
-
-Instead of using the "stand-alone" tests, `prop.test` and `chisq.test`, we could explicitly frame this as a log-linear model and test the goodness-of-fit of the "no interaction" model using either the likelihood ratio statistic (`lrt`) or the Pearson's $\chi^{2}$ test statistic (`pearson`).
-Formally, the regression model is (__TODO: Write down the regression model__):
-
-This model can be fit using the `loglin` function in R or using the arguably more user-friendly `loglm` (from the `MASS` package) or `glm` functions:
-
-```{r}
-# Using loglin
-loglin(x, margin = list(1, 2))
-# Using loglm
-library(MASS)
-loglm(formula =  ~ sample + meth_state, data = x)
-# Using glm (1)
-glm(x ~ 1, family = binomial)
-# Using glm (2)
-glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2])
-# TODO: Write sentence summarising output.
-```
-
-A key advantage of using regression is the option to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
-__TODO: How to do this using `loglin`?__
-
-More complex regression analyses are possible, such as empirical Bayes or fully Bayesian models.
-I describe these in more detail in discussing two-group and multi-group experiments, but they may also be applied to one-group experiments.
-
-A proper software implementation needs to take care of issues such as incomplete tables and non-convergence of the iterative proportional scaling algorithm used by `loglin`.
-Speed is also important when performing this many tests and the regression based approaches are much slower, at least when naively applied, due to additional checks they make of the data:
-
-```{r}
-# Basic benchmarking
-library(microbenchmark)
-print(microbenchmark(prop.test(x), chisq.test(x), loglin(x, margin = list("sample", "meth_state")), loglm(formula = ~ sample + meth_state, data = x), glm(x ~ 1, family = binomial), glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[,1] + x[, 2]), glm.fit(x = matrix(, nrow(x), 0L), y = x, family = binomial()), times = 100), order = "median")
-```
+\section{Dump from previous chapter}
 
@@
 
+- [x] DELETE ME. This is covered elsewhere. Make sure all points in the below are addressed in the above.
+
 Smoothing of the $\beta$-values has been used advocated by several authors to remove variation due to low sequencing coverage and to exploit the spatial correlation of DNA methylation \citep{Hansen:2011gu, Hebestreit:2013ko, Gokhman:2014c}.
@@
 \subsection{DMRs}
+
+__TODO: FDRs of region__
 

### 2014-12-01 Rmarkdown/datasets.Rmd
End of day commit
Added TODO
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.
Misc. notes.

@@
 
-## Chapter overview
+\chapter{Datasets used in thesis}\label{chap:datasets}
+
+\section{Chapter overview}
 
 I have analysed more than 50 whole-genome bisulfite-sequencing datasets during the course of my PhD studies. This short chapter documents those that I use in my thesis.
+
+__TODO: Describe processing of data__
 
@@
 __TODO: Table of datasets__
+
+## CpG islands
+
+I have used CpG island definitions from the hidden Markov model proposed by \cite{@Wu:2010do}.
+These predicted CpG islands were downloaded from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/) on 29/10/2014.
+
+## Misc. notes
+
+- `Bis-SNP` run without `--trim5` and `--trim5` flags.
+This is a bit unfortunate but it was because the `bissnp_easy_usage.pl` script, which I used to run `Bis-SNP`, didn't support these.
+To re-genotype all samples would have taken too long and it shouldn't greatly affect my analyses.
+- Percentage of CpGs that are in Rafa's CGIs

### 2014-12-01 Rmarkdown/appendix.Rmd
Moved mathematical argument to appendix.

@@
 
+## Equivalence of tests to identify DMCs in a one-group experiment
+
+The computational equivalence of these approaches are shown in the following R code:
+
+```{r}
+# The data
+x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
+names(dimnames(x)) <- c("sample", "meth_state")
+x
+
+# Using Pearson's chi-square test
+chisq.test(x)
+
+# Fitting a log-linear model to perform iterative proportional fitting to the
+# contingency table.
+loglin(x, margin = list("sample", "meth_state"))
+# An alternative formulation using loglm(), a convenient wrapper of loglin(),
+# from the MASS package.
+library(MASS)
+loglm(formula =  ~ sample + meth_state, data = x)
+
+# Fitting a generalised linear model with a binomial link using two different
+# formulations
+# (1) Using the array of frequencies
+glm(x ~ 1, family = binomial)
+# (2) Using relative frequencies (i.e., the beta-values) and weights (i.e.,
+  # the sequencing coverage).
+  glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[, 1] + x[, 2])
+```
+
+The saturated log-linear model can be written as $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{a, b}^{meth_state,sample}$, where $F_{a, b}$ is the value of the $(a, b)$ element in the $n \times 2$ contingency table.
+The null hypothesis that the cytosine is not a DMC is then tested by testing for an association between `sample` and `meth_state`, i.e., whether the parameter $\lambda_{a, b}^{meth_state, sample}$ is significantly different from zero.
+
+This log-linear model can be reformulated as a logistic regression by noting that the log-odds (logits) for each row are $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \log(F_{a, 1}) - \log(F_{a, 2})$. Therefore,
+
+\begin{align*}
+\log(\frac{F_{a, 1}}{F_{a, 2}}) &= \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{1}^{sample} + \lambda_{a, 1} \\
+&- (\lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{2}^{sample} + \lambda_{a, 2}) \\
+&= \underbrace{(\lambda_{1}^{sample} - \lambda_{2}^{sample})}_{\alpha} + \underbrace{(\lambda_{a, 1} - \lambda_{a, 2})}_{\gamma_{i}}
+\end{align*}
+
+The expression $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha + \gamma_{i}$ is a logistic regression model.
+We can test the null hypothesis that the cytosine is not a DMC by testing for a `sample`-effect on the log-odds, i.e., whether the parameter $\gamma_{i}$ is significantly different from zero.
+
+Regardless of whether we use the log-linear model or the logistic model, to test the null hypothesis that the cytosine is not a DMC we can compute the deviance[^deviance] of the reduced model and compare it to a $\chi^{2}$ random variable with $(2 - 1) \times (n - 1)$ degrees of freedom.
+For the log-linear regression the reduced model is $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample}$ and for the logistric regression the reduced model is $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha$.
+
+[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.
+
 ## The probability that two dependent Bernoulli random variables are identical

### 2014-12-01 Rmarkdown/wgbs_statistical_analysis.Rmd
Finished DMC section and started DMR section

@@
 
-\subsubsection{Beta-binomial models}
+\subsubsection{Beta-binomial models}\label{sec:beta-binomial_models}
 
@@
 
-\subsection{Transformations}
+\subsection{Transformations}\label{sec:transformations}
 
@@
 Alternatively, we could frame this as a generalised linear model, either a log-linear model or a logistic regression.
-The hypothesis is then tested by computing the deviance of the model[^deviance]
+The equivalence of these methods is shown in the Appendix.
 
-[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.
-
-assessing the goodness-of-fit of the 'no interaction' model (equivalently, comparing the saturated model to .
-The following R code shows the equivalence of these approaches for a CpG where the fifth sample is differentially methylated.
-
-```{r}
-# The data
-x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
-names(dimnames(x)) <- c("sample", "meth_state")
-x
-
-# Using Pearson's chi-square test
-chisq.test(x)
-
-# Fitting a log-linear model to perform iterative proportional fitting to the
-# contingency table.
-loglin(x, margin = list("sample", "meth_state"))
-# An alternative formulation using loglm(), a convenient wrapper of loglin(),
-# from the MASS package.
-library(MASS)
-loglm(formula =  ~ sample + meth_state, data = x)
-
-# Fitting a generalised linear model with a binomial link using two different
-# formulations
-# (1) Using the array of frequencies
-glm(x ~ 1, family = binomial)
-# (2) Using relative frequencies (i.e., the beta-values) and weights (i.e.,
-# the sequencing coverage).
-glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[, 1] + x[, 2])
-```
-
-A key advantage of using either of the generalised linear models is the option to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
+A key advantage of using either of the generalised linear models is the capability to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
 More complex regression analyses are also possible, such as empirical Bayes or fully Bayesian models.
 
-\subsubsection{Two-group experiments}\label{sec:two_group_dmcs}
+\subsubsection{Multi-group experiments}\label{sec:two_group_dmcs}
+
+Due to the cost of whole-genome bisulfite-sequencing, early experiments using this technology to study differential methylation compared one sample against another \citep[e.g.,]{Lister:2009hy, Berman:2012ga}.
+While this might be considered a two-group experiment, one without replicates, I believe this is better thought of as a one-group experiment with two samples.
+Indeed, the analysis of such an experiment uses techniques from Section \label{sec:one_group_dmcs}, e.g., Pearson's $\chi^{2}$ test or Fisher's exact test, which is better suited to the small sample sizes routinely seen in each $2 \times 2$ table.
+
+The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
+Unfortunately, some authors have still used Fisher's exact test to analyse such experiments.
+For example, \cite{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data the two replicates they have for each sample.
+This is incorrect, because such an analysis effectievly ignores all within-group variability \citep{Hansen:2012gr}.
+Instead, a test that incorporates the within-group variability should be used.
+
+In a two-group experiment this might be a t-test comparing the the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$.
+More generally, we can use a regression model to allow for multiple groups (i.e., ANOVA) and to incorporate additional covariates.
+
+Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment.
+Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_model}), and/or perform some transformation, such as smoothing of the $\beta$-values (Section \ref{transformation}).
+
+`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
+`DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
+`BiSeq` and `methlySig` also perform spatial smoothing of the data.
+
+The test used to identify DMCs in these regression models is variously a Wald test (`BiSeq`, `DSS`), a likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).
+
+Not all methods for identifying differential methylation are designed for identifying DMCs.
+Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs and, in the case of `bsseq`.
+
+\subsection{DMRs}
+
+There are two very different strategies for identifying differentially methylated regions:
+
+1.
+Usinb regions that are defined _a priori_, which are then tested for differential methylation.
+Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g.
+3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
+2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
+
+The former is simple to analyse but is limited in its ability to discover new differentially methylated regions.
+The latter offers the opportunity to identify previously unknown regions but valid statistical inference of these regions is __very challenging__.
 
@@
 \section{Dump from previous chapter}
-
-\subsubsection{Two-group experiments}\label{sec:two_group_dmcs}
-
-In a two group experiment, for each methylation locus we are testing the hypothesis that the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, is different to that in the second group, $\bar{\beta_{i}^{(2)}}$.
-
-If there are no replicates in either group[^2x2] then we have a $2 \times 2$ contingency table:
-
-[^2x2]: I'd argue that this is a two-sample experiment rather than a two-group experiment since each group has $n = 1$.
-
-(__TODO: Prettify table with `booktabs`__)
-
-\begin{table}[h]
-\centering
-\caption{$2 \times 2$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a two-group experiment with no replicates in either group.}
-\label{my-label}
-\begin{tabular}{c c c}
-\hline
-& m          & u          \\ \hline
-$Group_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
-$Group_{2}$ & $m_{i, 2}$ & $u_{i, 2}$ \\
-\end{tabular}
-\end{table}
-
-This $2 \times 2$ table is then tested for independence of the rows and columns.
-As the sample size for each table is often small, Fisher's exact test is generally preferable since, unlike Pearson's $\chi^{2}$ test, the P-values are _exact_, meaning they do not rely on large-sample approximations (__CITE: Agresti__).
-Fisher's exact test is a _conditional_ test because the null distribution (a hypergeometric distribution) is derived by conditioning on both the marginal totals.
-__CITE: Agresti__ discusses the arguments for and against conditional and unconditional tests of independence in the $2 \times 2$ table.
-
-The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
-Unfortunately, some authors have used Fisher's exact test to analyse experiments including replicates (__TODO: Highlight papers that have inappropriately used Fisher's exact test; does Lister 2013 inappropriately pool then test?).
-This is incorrect because it effectively aggregates all the counts within each group and so ignores all within-group biological variability \citep{Hansen:2012gr}.
-
-Instead, a test that incorporates the within-group variability should be used.
-For example, we could use a t-test to compare the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$.
-Several methods have been proposed for testing for DMCs.
-These methods differ considerably in how they model and transform the data as well as what test statistic is used to identify DMCs.
-
-\paragraph{Transforming}\label{sec:transforming}
-
-- [x] DELETE ME. This is covered elsewhere. Make sure all points in the below are addressed in the above.
-
-Smoothing of the $\beta$-values has been used advocated by several authors to remove variation due to low sequencing coverage and to exploit the spatial correlation of DNA methylation \citep{Hansen:2011gu, Hebestreit:2013ko, Gokhman:2014c}.
-Both `BSmooth` \citep{Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} smooth the $\beta$-values using a binomial local likelihood smoother.
-\citet{Lister:2011kg, Gokhman:2014cp} used a simpler moving average in windows along the genome.
-
-\paragraph{Modeling}\label{sec:modeling}
-
-`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
-Broadly, under the Beta-Binomial model the "true" methylation level at the $i^{th}$ methylation locus in the $k^{th}$ group is modeled as a Beta random variable, $B_{i, k} = Beta(\mu_{i, k}, \theta_{i, k})$.
-The observed number of methylated reads at the $i^{th}$ methylation locus, in the $j^{th}$ sample is modeled as a Binomial random variable, $M_{i, j, k} | B_{i, k}, n_{i, j, k} = Binomial(n_{i, j, k}, B_{i, k}), which is conditional on the unobserved $B_{i, k}$ and the observed sequencing coverage, $n_{i, j, k}$.
-
-\paragraph{Estimating and testing}\label{sec:estimating_and_testing}
-
-`DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
-The test used to identify DMCs is variously a Wald test (`BiSeq`, `DSS`), likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).
-
-\subsubsection{Multi-group experiments}\label{sec:multigroup_dmcs}
 

### 2014-12-02 Rmarkdown/wgbs_statistical_analysis.Rmd
Re-jigging of WGBS statistical analysis chapter

@@
 All methylation loci are defined with respect to the sample's genome.
-The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth} are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.
+The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth}$ are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.
 
@@
 Both `bsseq` and `BiSeq` use a binomial local likelihood smoother.
-In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j}$ \eqd Binom(n_{i, j}, B_{i, j})$.
+In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$.
 The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.
@@
 
-\subsection{Experimental design}\label{sec:experimental_design}
+\subsubsection{Experimental design}\label{sec:experimental_design}
 
@@
 
-\subsection{DMCs}\label{sec:dmcs}
+\subsubsection{DMCs}\label{sec:dmcs}
 
@@
 
-\subsubsection{One-group experiments}
+\paragraph{One-group experiments}\label{sec:one_group_dmcs}
 
@@
 
-\subsubsection{Multi-group experiments}\label{sec:two_group_dmcs}
+\paragraph{Multi-group experiments}\label{sec:two_group_dmcs}
 
@@
 
-\subsection{DMRs}
+\subsubsection{DMRs}
+
+A differentially methylated region is a region of the genome where there are multiple cytosines with evidence of differential methylation.
+Importantly, not all cytosines in the region must be genome-wide significant DMCs.
+Rather, the idea is that a DMR might capture a weaker but broader difference in methylation.
+Such a signal may well have a clearer biological interpretation and, indeed, might be expected such as in differences in methylation at a gene promoter.
+
+__UP TO HERE: Take care of TODOs and then move onto other downstream analyses.__
 
@@
 1.
-Usinb regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g.
-3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
+Using regions that are defined _a priori_, which are then tested for differential methylation.
+Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__), gene promoters (__CITE__) or predefined bins \citep[e.g., 100 bp bins]{Park:2014ho}.
 2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
 
-The former is simple to analyse but is limited in its ability to discover new differentially methylated regions.
-The latter offers the opportunity to identify previously unknown regions but valid statistical inference of these regions is __very challenging__.
+The former is much simpler to analyse but is limited in its ability to discover new differentially methylated regions.
+The latter offers the opportunity to identify previously unknown regions but valid statistical inference of these regions is very challenging.
 
-__UP TO HERE__
-
-\section{Dump from previous chapter}
-
-\subsection{DMRs}
-
-__TODO: FDRs of region__
-
-There are two very different strategies for identifying differentially methylated regions:
-
-1.
-Use regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g.
-3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
-2. Use regions that are defined based on the data, which are then tested for differential methylation. Valid statistical inference of such regions is __very challenging__.
-
-\subsubsection{Using _a priori_ regions}\label{sec:a_prior_regions}
-
-The methylation level is summarised at the region level and then compared across groups. For example, the average of the $\beta$-values across the region may be compared between groups using a t-test.
-
-__TODO: Describe the above mathematically__
-
-This is qualitatively similar to identifying differentially methylated cytosines.
-
-__PROS__
-
-* Simple to implement
-* Valid statistical properties
-
-__CONS__
+__TODO: Further cons of a priori regions__
 
@@
 
-\subsubsection{Using data-driven regions}\label{sec:data_driven_regions}
-
-__PROS__
+__TODO: Further pros of data-driven regions__
 
@@
 
-__CONS__
+The most common type of algorithm for identifying data-driven DMRs is based on detecting clusters of DMCs (__CITE: e.g., Lister, Hansen, DSS?, Berman?, etc.).
+This initial identification of DMCs will typically use a relaxed statistical significance threshold.
+For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
+This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
 
-* Loss of statistical properties in testing, e.g. FDR control
-* Bloody hard
+[^a_prior]: In contrast, DMR analyses using _a priori_ regions do not need to first test for DMCs, although this single-locus analysis may still be performed.
 
-\section{Other downstream analyses}\label{sec:other_downstream_analyses}
+__TODO: Other data-driven DMR detection strategies__
+- bump-hunting
+- Mixed effect model
+- GEE, i.e., `Aclust` distinguishes itself by using data-driven regions to identify DMRs but without first scanning the genome for DMCs.
+- probe aggregation, e.g., `comb-P`
 
-Bisulfite-sequencing data are used to address a variety of questions apart from identifying differential methylation.
-Not all these analyses require statistical inference.
-For example, if we are interested in knowing whether the promoter of a particular gene is methylated in our sample then it may be sufficient to simply plot the level of methylation at each CpG in the region.
+Finally, __TODO: Do these methods actually test for differential methylation at these regions?
+If they do, then a test of a DMR requires a statistic that summarises the evidence that the region is differentially methylated.
+What about FDRs?__
 
-However, once we move to experiments with multiple samples, multiple regions and more complex questions then statistical methods are generally required to distinguish the signal from the noise.
-In this section I briefly describe other
+For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region.
+For example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage.
+These averages are then compared between samples.
 
-__TODO: Any other important downstream analyses worth mentioning?__
+\paragraph{One-group experiments}\label{sec:one_group_dmrs}
+
+For a region containing $c$ methylation loci, construct the $n \times 2 \times c$ contingency table, i.e., each 'slice' of the table is a $n \times 2$ table containing the data for a single locus.
+Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions (__CITE: http://www.biostathandbook.com/cmh.html__).
+Instead, we could fit the log-linear model with `sample`, `meth_type` and `locus` as the set of factors.
+The saturated model is
+
+\begin{equation*}
+\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus} + \lambda_{a, b, c}^{meth_state, sample, locus}
+\end{equation*}
+
+We would first need to test whether there is a significant three-way interaction between `meth_type`, `sample` and `locus` by computing the deviance of the model without the three-way interaction
+
+\begin{equation*}
+\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
+\end{equation*}
+
+On the one hand, if there is no three-way interaction, then we can test for an interaction between `meth_type` and `sample`, i.e., test for a difference in methylation level across the loci, by comparing the model
+
+\begin{equation*}
+\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
+\end{equation*}
+
+to the model
+
+\begin{equation*}
+\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
+\end{equation*}
+
+This test is analogous to the generalised Cochran-Mantel-Haenszel test \citep{Landis:1978vk,MANTEL:1978iv,Birch:1965uj} of the $n \times 2 \times c$ table \citep[p231]{agresti1990categorical}.
+The advantage of the Cochran-Mantel-Haenszel is its applicability for sparse datasets for which the asymptotic theory does not hold for the test based on the log-linear regression  \citep[p232]{agresti1990categorical}.
+
+On the other hand, if there is a significant three-way interaction between `meth_type`, `sample` and `locus` then this requires more careful modelling to determine the nature of this interaction and whether the region is a consistent DMR.
+
+\paragraph{Multi-group experiments}\label{sec:multi_group_dmrs}
+
+__TODO__
+
+\paragraph{Consistency of DMRs}\label{sec:consistency_of_dmrs}
+
+While not often explicitly stated, it is desirable that a DMR is _consistent_.
+The definition of consistency is best illustrated by a pathological example describing a region that is inconsistent.
+Suppose we have a one-group experiment with five samples[^one_by_five] and are studying a region with five CpGs, all of which are DMCs.
+However, the first CpG is differentially methylated in the first sample, the second CpG is differentially methylated in the second sample, etc.
+While the entire region is 'differentially methylated', the difference is not observed consistently in any one sample.
+A consistent region is therefore one where the differences are 'in the same direction'.
+
+[^one_by_five]: Equivalently, a five group experiment without replicates.
+
+Consistent DMRs are most easily identified in two-group experiments, where we can use the sign of the test statistic to determine the 'direction' of the signal.
+This is exactly what is done by, for example, \cite{Hansen:2011gu}.
+
+In one-group experiments, the idea of consistency is similar to testing the homogeneity of odds ratios for all $n \times 2$ contigency tables in the region.
+However, __TODO.
+What about a region that is consistent but the odds ratios increases across the region__
+
+\section{Dump from previous chapter}
 
@@
 
-\subsection{General computational considerations}
-
-- Using integers instead of floats/doubles
-
 \section{General TODOs}
 
-* Autocorrelation or correlation?
 * Discuss distribution of coverage? Perhaps in context of sequencing bias?
-* See https://github.com/brentp/clustermodel/README.md for a nice, brief summary of the main DMR calling approaches.

### 2014-12-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Keep on truckin
Misc. notes.
Fixed some syntax errors

@@
 This is only possible with bisulfite-data generated using the directional protocol.
-This figure is adapted from \cite{Liu:2012ge}}
-\label{fig:Bis-SNP.pdf}
+This figure is adapted from \cite{Liu:2012ge}.}
+\label{fig:Bis-SNP}
 \end{figure}
@@
 
+__TODO: Size of m-tuples histogram__
+
 \subsection{Motivation}
@@
 Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
-\code\{methclone\} & \cite\{Li:2014ei\}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf\{TODO: Try installing.
+\texttt\{methclone\} & \cite\{Li:2014ei\}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf\{TODO: Try installing.
 Suspect you can't get actual methylation calls.\} \\
-\code\{methpat\}   & \url\{https://github.com/bjpop/methpat\} & Output of \code\{bismark\_methylation\_extractor\} & Yes (amplicons)                  &                                                                                  \\
-\code\{DMEAS\}     & \cite\{He:2013cj\}                       & Output of \code\{bismark\_methylation\_extractor\} &                                  & Windows operating system only.
+\texttt\{methpat\}   & \url\{https://github.com/bjpop/methpat\} & Output of \texttt\{bismark\_methylation\_extractor\} & Yes (amplicons)                  &                                                                                  \\
+\texttt\{DMEAS\}     & \cite\{He:2013cj\}                       & Output of \texttt\{bismark\_methylation\_extractor\} &                                  & Windows operating system only.
 Perl code only available as PDF file.
@@
 
-```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \code{GNU time} converted from seconds to minutes.
+```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes.
 The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
@@
 
-```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_memory", fig.cap = "; the suffix 'ac' means that the option \texttt{--all-combinations} was set.
-This option creates all tuples with $NIL \geq 0$ which is far greater than the equivalent tuples with $NIL = 0$.
-The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
-These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865})", fig.width = 8, dev = "pdf"}
+```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_memory", fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
+These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
+The suffix 'ac' means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
 # Memory

### 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
Described BiSeq and data-driven DMRs

@@
 
-In the second half of the chapter I use this framework to discuss downstream analyses of whole-genome bisulfite-sequencing data.
+As is clear from the length of Chapter \ref{chap:wgbs_bioinformatics_analysis} alone, there is a considerable amount of bioinformatics work before we even consider analysing the data to address the biological question of interest.
+I call these the 'downstream' analyses, and there are many that might be considered.
+Using the above statistical framework, I explain the statistical question of identifying differential methylation.
+I also discuss other downstream analyses.
 
@@
 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: _BUF_ = buffy coat; _SA_ = subcutaneous adipose; _VA_ = visceral adipose; _VAT_ = adipocytes derived from visceral adipose.}
+Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose.}
 \label{fig:EPISCOPE_beta_distribution}
@@
 Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: _BUF_ = buffy coat; _SA_ = subcutaneous adipose; _VA_ = visceral adipose; _VAT_ = adipocytes derived from visceral adipose; _CGI_ = CpG island.}
+Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose; \em{CGI} = CpG island.}
 \label{fig:EPISCOPE_beta_distribution_CGI_stratified}
@@
 
-\section{Downstream analyses}\label{sec:downstream_analyses}
-
-As is clear from the length of Chapter \ref{chap:wgbs_bioinformatics_analysis} alone, there is a considerable amount of bioinformatics work before we even consider analysing the data to address the biological question of interest.
-I call these the 'downstream' analyses, and there are many that might be considered.
-Using the above statistical framework, I explain the most commonly asked questions.
-
-\subsection{Differential methylation}
+\section{Differential methylation}
 
@@
 
-\subsubsection{Experimental design}\label{sec:experimental_design}
+\subsection{Experimental design}\label{sec:experimental_design}
 
@@
 
-\subsubsection{DMCs}\label{sec:dmcs}
+\subsection{DMCs}\label{sec:dmcs}
 
@@
 
-\subsubsection{DMRs}
+\subsection{DMRs}
 
@@
 
-__UP TO HERE: Take care of TODOs and then move onto other downstream analyses.__
-
 There are two very different strategies for identifying differentially methylated regions:
@@
 Using regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__), gene promoters (__CITE__) or predefined bins \citep[e.g., 100 bp bins]{Park:2014ho}.
+Such regions might be CpG islands \citep[e.g.,]{Huang:1999fs,Doi:2009ky}[^array], MspI restriction fragments \citep[e.g.,]{Stockwell:2014fq}, any other predefined genomic feature, such as a gene promoters or transcription factor binding sites, or general predefined bins \citep[e.g., 100 bp bins]{Park:2014ho}.
 2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
 
+[^array]: Both these examples are from microarray studies, but the same idea can be applied to sequencing studies.
+
 The former is much simpler to analyse but is limited in its ability to discover new differentially methylated regions.
-The latter offers the opportunity to identify previously unknown regions but valid statistical inference of these regions is very challenging.
+It is also hampered because the correct unit or scale for differential methylation may not be known for the experiment.
 
-__TODO: Further cons of a priori regions__
+The latter offers the opportunity to identify previously unknown regions that are subject to differential methylation.
+Included in this is the opportunity to discover the scale over which differential methylation acts.
+However, valid statistical inference of these regions is far more challenging.
 
-* Doesn't account for differential CpG density across regions
-* What is the right "unit" - the CGI, the promoter, a tile???
-* Doesn't account for spatial correlation of methylation
+\subsection{Using _a priori_ regions}
 
-__TODO: Further pros of data-driven regions__
+The idea of testing for differential methylation is relatively straightforward.
+For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region, for example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage.
+These averages are then compared between samples.
+I describe basic implementations of this for both one-group and multi-group experiments, and then detail a more sophisticated analysis implemented in `BiSeq` \citep{Hebestreit:2013ko}.
 
-* Can discover the right "unit".
-* Might account for spatial correlation of methylation
-
-The most common type of algorithm for identifying data-driven DMRs is based on detecting clusters of DMCs (__CITE: e.g., Lister, Hansen, DSS?, Berman?, etc.).
-This initial identification of DMCs will typically use a relaxed statistical significance threshold.
-For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
-This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
-
-[^a_prior]: In contrast, DMR analyses using _a priori_ regions do not need to first test for DMCs, although this single-locus analysis may still be performed.
-
-__TODO: Other data-driven DMR detection strategies__
-- bump-hunting
-- Mixed effect model
-- GEE, i.e., `Aclust` distinguishes itself by using data-driven regions to identify DMRs but without first scanning the genome for DMCs.
-- probe aggregation, e.g., `comb-P`
-
-Finally, __TODO: Do these methods actually test for differential methylation at these regions?
-If they do, then a test of a DMR requires a statistic that summarises the evidence that the region is differentially methylated.
-What about FDRs?__
-
-For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region.
-For example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage.
-These averages are then compared between samples.
-
-\paragraph{One-group experiments}\label{sec:one_group_dmrs}
+\subsubsection{One-group experiments}\label{sec:one_group_dmrs}
 
@@
 
-\paragraph{Multi-group experiments}\label{sec:multi_group_dmrs}
+\subsubsection{Multi-group experiments}\label{sec:multi_group_dmrs}
 
@@
 
-\paragraph{Consistency of DMRs}\label{sec:consistency_of_dmrs}
+\subsubsection{A more complex example}\label{sec:a_more_complex_example}
+
+`BiSeq` is designed for RRBS data and tests for differential methylation at "CpG clusters".
+These CpG clusters are based on an initical scan of the genome looking for CpGs that are within a given distance of one another and that have sufficient sequencing coverage across the set of samples.
+While there is clearly a data-dependent component to these cluster definitions, I reserve the use of 'data-driven regions' for those which are based on the methylation levels of loci rather than their genomic co-ordinates.
+\citep{Hebestreit:2013ko} alternatively recommend using the target regions of the assay, such as MspI fragments in the case of RRBS, instead of creating these CpG clusters.
+
+One the clusters are defined, the $\beta$-values in each cluster are smoothed using a local binomial likelihood smoother.
+For each CpG `BiSeq` fits a beta regression model[^not_beta-binomial], which is tested for evidence of differential methylation at that cytosine.
+Based on these P-values, `BiSeq` then use a hierarchical testing procedure to control the false discovery rate at both the cluster-level and locus-level.
+This method is based on several papers by Yoav Benjamini and colleagues \citet{Benjamini:1997fh, Benjamini:2001ho, Benjamini:2006gd, Benjamini:2007cm}.
+It aims to first control the false discovery rate at the cluster-level and then refine the signal by trimming non-DMCs from those clusters that have been declared as differentially methylated.
+Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_ (see Section \ref{sec:consistency_of_dmrs}).
+
+[^not_beta-binomial]: This is different to the beta-binomial regression framework described in Section \ref{sec:beta-binomial_models}.
+
+Based on my understanding of this procedure, only cytosines within differentially methylated clusters can be declared DMCs since it is only those cytosines that are tested in the second-step of the hierarchical testing algorithm.
+Therefore, isolated DMCs will not be detected by this procedure.
+
+\subsection{Using data-driven regions}
+
+The most common type of algorithm for identifying data-driven DMRs, and arguably also the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
+The initial identification of DMCs will typically use a relaxed statistical significance threshold.
+For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
+This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
+
+These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
+Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
+
+The challenges of valid statistical inference at such data-driven regions are not unique to the problem of testing for differentially methylated regions.
+Similar problems arise in the analysis of chromatin immunoprecipitation sequencing (ChIP-seq) experiments \citep{Schwartzman:2011to, Lun:2014gn} and in the field of signal processing \citep{Schwartzman:2011vg}.
+It boils down to the problem of how to assign statistical significance to regions or peaks using the same data that were used to define these regions or peaks.
+These regions are already enriched for differences, which biases the interpretation the results of tests for differences at these regions.
+One way, and I would argue the best way, to avoid this issue is to test these regions using a separate dataset, which completely avoids the issue of statistical 'double-dipping'.
+If the sample size is large enough, then permutation testing may also be appropriate.
+For example, \cite{Hansen:2013eo} permute the group labels of their samples and re-ran their analysis to determine for each of the observed regions "how often we see another block of similar length and effect size anywhere in the genome and in any of the permutations".
+The chief limitation of the permutation strategy is the restricted number of permutations that are possible from small sample sizes and the time and computational resources it takes to perform analyse the data for a single permutation, which are often substantial.
+For example, \cite{Hansen:2013eo} only performed nine permutations.
+
+A somewhat less _ad hoc_ strategy for identify differentially methylated regions is called "bump-hunting" \citep{Jaffe:2012gx}.
+Initially developed for methylation microarrays, and now available for broader use in the R/Bioconductor package `bumphunter` ([http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html](http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html)), bump-hunting is quite a general statistical problem.
+For the case of identifying DMRs, the idea is to plot the value of a statistic used to test for DMCs as a function of the position in the genome[^smoothing] and then to identify bumps in this plot.
+Bumps are identified as contiguous regions of the genome where the signal is above some threshold and may include an error term to account for the spatial correlation of the signal.
+The significance of these peaks may be assessed with a permutation strategy.
+
+Another alternative for combining individual loci into data-driven DMRs is based on the loci-specific P-values.
+These methods can be thought of extensions to Fisher's method for combining P-values \citep{Fisher:1936vv}, by attempting to account for the correlation of tests at nearby methylation loci.
+`comb-p` \citep{Pedersen:2012vl} uses the Stouffer-Liptak-Kechris \citep{Stouffer:1949ua, Kechris:2010fw, Zaykin:2011jq,} correction for spatially correlated P-values.
+`SLIM` \citep{Wang:2011cw} is used by `methylKit` \citep{Akalin:2012cm} to do a similar correction.
+
+Finally, there are a class of methods that turn the problem of identifying data-driven DMRs on its head by constructing the regions without first testing the individual loci for differential methylation.
+Then, only once these regions are defined, these methods test for differential methylation.
+This is different to using _a priori_ regions, since the regions are still data-defined, but not with respect to differential methylation.
+The only example I know of such a method is `Aclust` \citep{Sofer:2013bk}.
+`Aclust` first clusters CpGs into candidate regions by performing agglomerative nested clustering of the between-sample co-methylation.
+Briefly, this is the correlation of methylation levels at two loci across the samples[^between_sample_co-methylation].
+These clusters are then tested for differential methylation using generalised estimating equations.
+
+[^between_sample_co-methylation]: See Chapter \ref{co-methylation_review} for further details of between-sample co-methylation.
+
+\subsubsection{Consistency of DMRs}\label{sec:consistency_of_dmrs}
 
@@
 
-\section{Dump from previous chapter}
+\section{Other downstream analyses}
+
+\subsection{PMDs}
+
+- Lister
+- Berman
 
 \subsection{Variably methylated regions and differential variability}\label{sec:vmrs}
+
+- Belinda
+- Jaffe
 

### 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
Re-jigged DMRs and added PMD section.

@@
 I also discuss other downstream analyses.
-
-- Chapter overview
-  - One sample
-    - Pre-sequencing
-    - Post-sequencing
-    - Some complications for a single sample
-      - What is $\mathcal{I}$?
-      - Genetic heterozygosity at a methylation locus
-  - Multiple samples
-    - Some complications with multiple samples
-      - What is $\mathcal{I}$?
-  - Parameter estimation
-    - Estimating $M$, $U$
-    - Estimating $B$
-      - Beta-binomial models
-      - Smoothing
-    - Statistical properties of $\beta$-values
-      - Empirical distributions
-      * Write a function(s) to do the following and apply it to the EPISCOPE, Lister, Seisenberger and Ziller datasets:
-        - Genome-wide distribution of CpG $\beta$-values
-        * Kernel density estimates
-        * Stratify by CGI status, genic, intragenic, repeats
-        * Correlation across strands
-        * Effect of sequencing coverage threshold (`min_cov`)
-      - What is the genome-wide methylation level?
-        * Various figures get bandied about; what's a good summary?
-        - A criticism of ``identifying methylcytosines''
-      - Correlations
-      - Biases
-      - Transformations
-  - Downstream analyses
-    - Differential methylation
-      - Experimental design considerations
-      - DMCs: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
-        - One-group experiments
-        - Two-group experiments
-        - Multi-group experiments
-      - DMRs: Identify runs of DMCs
-        - A priori vs. data-driven regions
-        - One-group experiments
-        - Two-group experiments
-        - Multi-group experiments
-    - Differential variability
-      - DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
-      - DVRs: Identify runs of DVCs
-    - Allele-specific methylation
-    - Methylation entropy
-    - Epipolymorphism
-    - Epiallele detection
-    - Integrating DNA methylation data with other genomes data
 
@@
 
-A common downstream analysis is the identification of differential methylation, for which many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
+A common downstream analysis is the identification of differential methylation.
 This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
+Many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
+All of these methods are designed for two-group experiments, with some being general enough to extend to multiple-group experiments[^multi-group_extension].
+I describe in greater detail methods for identifying differential methylation in a one-group experiment, admitedly an unusual design but one that I faced during my PhD.
+
+[^multi-group_extension]: The extension to multiple-group experiments isn't always implemented in the software, even if discussed in the paper.
 
@@
 Regardless of the test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty.
-Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data.
-but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments (e.g., approximately 1 million tests in genome-wide association studies and 20,000 tests in studies of differential gene expression).
+Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments[^multiple_testing].
 Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.
+
+[^multiple_testing]; For example, there are approximately 20,000 tests in studies of differential gene expression and one million tests in genome-wide association studies.
 
@@
 
-\paragraph{One-group experiments}\label{sec:one_group_dmcs}
+\subsubsection{One-group experiments}\label{sec:one_group_dmcs}
 
@@
 
-\paragraph{Multi-group experiments}\label{sec:two_group_dmcs}
+\subsubsection{Multi-group experiments}\label{sec:multi_group_dmcs}
 
@@
 Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment.
-Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_model}), and/or perform some transformation, such as smoothing of the $\beta$-values (Section \ref{transformation}).
+Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (see Section \ref{sec:beta-binomial_model}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (see Section \ref{transformation}).
 
@@
 Not all methods for identifying differential methylation are designed for identifying DMCs.
-Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs and, in the case of `bsseq`.
+Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs.
 
@@
 These averages are then compared between samples.
-I describe basic implementations of this for both one-group and multi-group experiments, and then detail a more sophisticated analysis implemented in `BiSeq` \citep{Hebestreit:2013ko}.
+I describe a basic implementation of this for a one-group experiment and then detail a more sophisticated method that is implemented in `BiSeq` \citep{Hebestreit:2013ko}.
 
@@
 
-\subsubsection{Multi-group experiments}\label{sec:multi_group_dmrs}
-
-__TODO__
-
-\subsubsection{A more complex example}\label{sec:a_more_complex_example}
+\subsubsection{A more sophisticated multi-group analysis}\label{sec:a_more_complex_example}
 
@@
 
-The most common type of algorithm for identifying data-driven DMRs, and arguably also the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
+The most common type of algorithm for identifying data-driven DMRs, also arguably the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
 The initial identification of DMCs will typically use a relaxed statistical significance threshold.
@@
 
-\subsection{PMDs}
+There are many other questions that can be addressed using bisulfite-sequencing data.
+Like differential methylation, some of these might be also be tackled with other methylation assays, such as microarrays.
+However, as we will see, some can only be answered by assays that have single nucleotide resolution, which is where bisulfite-sequencing can prove particularly powerful.
 
-- Lister
-- Berman
+\subsection{Partially methylated domains}
+
+Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$.
+An example of such a region is shown in __FIGURE__ (__TODO: Find a PMD and plot it__).
+
+PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}.
+Using a simple sliding window algorithm, \citeauthors{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb.
+They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample.
+This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.
+
+A subsequent study found that PMDs are a common feature of somatic cell lines and that they comprise $> 30\%$ of the genome \citep{Lister:2011kg}.
+Perhaps even more intriguingly, across four somatic cell lines profiled with whole-genome bisulfite-sequencing, \citeauthor{Lister:2011kg} found a large amount of these genomes ($664$ Mb) comprised shared PMDs.
+
+PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_).
+However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}.
+Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg}.
+
+More sophisticated methods have also been proposed to identify these PMDs.
+`methylSeekR` \citep{Burger:2013kq} is one such algorithm.
+It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions.
+While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that methylation microarrays cannot provide.
 
@@
 - Jaffe
+- DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
+- DVRs: Identify runs of DVCs
 

### 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
Added section on differential variability

@@
 
-- Belinda
-- Jaffe
-- DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
-- DVRs: Identify runs of DVCs
+Differential methylation is an analysis of the difference in means and seeks to identify differentially methylated regions (DMRs).
+Analogously, we can analyse differences in the variability of methylation profiles, which is known as differential variability, and those regions known as variably methylated regions (VMRs).
+It has been hypothesised that increased variability in DNA methylation indicates an epigentic _plasticity_ that may be highly relevant to common diseases, in particular, cancer \citep{Feinberg:2010dy}.
+
+In a one-group experiment, a variably methylated region is one that has increased variability compared to 'similar' regions elsewhere in the genome.
+In a two-group experiment, a differentially variable locus is one where the variation in one group is significantly larger than that in the other group.
+\citet{Jaffe:2011hy} first developed formal statistical tests for both the one-group and two-group experiments.
+These methods were developed for us with data from the CHARM array (see __SECTION__).
+
+Statistical tests of variations are known to be more difficult than tests of means and require larger sample sizes.
+A more subtle difficulty is in dealing with outliers.
+An outlier in one group will greatly increase the variation in that same group, but this does not necessarily mean that the locus is differentially methylated.
+It is not hard to envisage an example where the two groups in fact have very similar variability once the outlier is excluded.
+Tests of differential variability that are based on the F-test \citep[e.g.,]{Hansen:2011gu} or Bartlett's test \citep[e.g.,]{Teschendorff:2012hea} will be susceptible to calling such loci as differentially variable.
+By contrast, \citet{Phipson:2014ji} designed `DiffVar` to test for differential variability that is deliberately robust to such outliers by using Levene's test \citep{Olkin:1960uu}.
+Which approach is preferable will depend on the user's perspective as to whether such loci with outliers should be called as variably methylated.
 

### 2014-12-03 Rmarkdown/biology_background.Rmd
Minor fixes
Added TODO re CHARM

@@
 
+__TODO: Describe CHARM \citep{Irizarry:2008hg}, although not yet sure which sub-section it should appear in__.
+
 \subsection{Sodium bisulfite conversion assays}\label{sec:bs_assays}
@@
 - Reduced representation bisulfite-sequencing (_RRBS_, \cite{Meissner:2005eb}): RRBS uses restriction enzymes to first select regions of the genome with a high CpG density (based on the Msp-I cleavage motif), which are subsequently treated with sodium-bisulfite and sequenced.
+- Extended reduced representation bisulfite-sequencing, also known as enhanced reduced representation bisulfite-sequencing (eRRBS, \cite{Akalin:2012he}). A modified version of the RRBS protocol.
 - NimbleGen's SeqCap Epi Enrichment System and Agilent's SureSelectxt Human Methyl-Seq: Both of these commercial products use DNA hybridisation to enrich the sequencing library for pre-defined regions of interest.

### 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
Section on allele-specific methylation.

@@
 The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
-Unfortunately, some authors have still used Fisher's exact test to analyse such experiments.
+Unfortunately, some authors have still used Fisher's exact test \citep{Fisher:1922bx} to analyse such experiments.
 For example, \cite{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data the two replicates they have for each sample.
@@
 
+In a diploid cell, allele-specific methylation occurs when only one of the parental chromosomes is methylated at a particular locus, where the locus may be an individual cytosine or a broader region.
+One particular form of allele-specific methylation occurs at imprinted genes where the silenced copy of the gene is heavily methylated and the active copy is lightly methylated \citep{Shoemaker:2010jz}.
+However, it is now apparent that allele-specific methylation is far more prevalent than at just these imprinted regions \citep{Tycko:2010hc, Shoemaker:2010jz}.
+
+The obvious method to detect allele-specific methylation from bisulfite-sequencing relies on the read containing a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus.
+The heterozygous variant allows the reads to be separated by the observed allele[^phase], which can then be used to test for allele-specific methylation.
+For example, \citet{Shoemaker:2010jz} construct a $2 \times 2$ contingency table, like that shown in Table \ref{tab:asm}, and test for an association between the allele and the methylation state using Fisher's exact test \citep{Fisher:1922bx}.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+& Allele 1 & Allele 2 \\ \midrule
+m &          &          \\
+u &          &          \\ \bottomrule
+\end{tabular}
+\caption{$2 \times 2$ table used to test for allele-specific methylation.}
+\label{tab:asm}
+\end{table}
+
+Such an approach, while straightforward, is also obviously limited to a subset of the available methylation loci.
+Both \citet{Fang:2012ci} and \cite{Peng:2012dh} published methods to detect allele-specfic methylation that do not require nearby heterozygous genetic variants nearby to the methylation locus of interest.
+These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data.
+Unfortunately, there is no publicly available software implementing the method proposed by \cite{Peng:2012dh} and so I do not discuss it further.
+
+\citet{Fang:2012ci} uses reads containing multiple methylation loci and looks for regions of the genome where there are two distinct methylation patterns at the read-level that occur at roughly equal proportions, indicating one pattern comes from one allele and the other pattern from the other allele.
+The likelihood of allele-specific methylation is computed using an expectation-maximisation algorithm, which assigns reads to one of the two possible alleles.
+Neighbouring regions displaying allele-specific methylation are then joined together.
+While not mentioned in the paper, the proposed method is now available in the `MethPipe` software ([http://smithlabresearch.org/software/methpipe/](http://smithlabresearch.org/software/methpipe/)).
+
+[^phase]: This does not give parent-specificity unless the phase of the genotype is also known.
+
 \subsection{Methylation entropy}\label{sec:methylation_entropy}

### 2014-12-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Split wgbs_statistical_analysis into wgbs_statistical_framework and wgbs_downstream_analyses

--- 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
@@
 ---
-title: "Statistical analysis of whole-genome bisulfite-sequencing data"
+title: "Downstream analyses of whole-genome bisulfite-sequencing data"
 output:
@@
     fig_caption: yes
+    keep_tex: yes
     includes:
@@
 
-\chapter{Statistical analysis of whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_analysis}
+\chapter{Downstream analyses of whole-genome bisulfite-sequencing data}
 
-\section{Chapter overview}\label{sec:statmodel_chapter_overview}
+This chapter discusses various statistical analyses of bisulfite-sequencing data, which I call _downstream analyses_.
 
-I set out a statistical framework for analysing bisulfite-sequencing data.
-I begin by explaining the various levels of stochasticity and then define the mathematical notation that I use throughout my thesis.
-I describe key variables, common estimators of these and their statistical properties.
-The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.
+The first half of the chapter deals with analyses that are based on aggregate methylation levels, such as $\beta$-values.
+I explain the statistical questions behind the common downstream analyses of identifying differential methylation, variable methylation and partially methylated domains.
 
-As is clear from the length of Chapter \ref{chap:wgbs_bioinformatics_analysis} alone, there is a considerable amount of bioinformatics work before we even consider analysing the data to address the biological question of interest.
-I call these the 'downstream' analyses, and there are many that might be considered.
-Using the above statistical framework, I explain the statistical question of identifying differential methylation.
-I also discuss other downstream analyses.
+The second half of the chapter reviews downstream analyses that are based on methylation patterns at m-tuples.
+I introduce `MethylationTuples`, an R package I am developing for managing, analysing and visualising methylation patterns at m-tuples, which is complementary to `methtuples` --- `methtuples` extracts the methylation patterns at m-tuples and `MethylationTuples` provides a framework for analysing these data, which is currently lacking for this type of data.
 
-\section{One sample}\label{sec:one_sample}
-
-Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment. I find it convenient to separate these into:
-
-1. Pre-sequencing sources of variation
-2. Sequencing and post-sequencing sources of variation.
-
-I describe these sources of variation, which are illustrated in Figure \ref{fig:bs-seq_protocol}.
-
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
-\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment.
-Each sample starts as a population of cells, with potentially different methylation patterns.
-Each DNA fragment is coloured by its originating cell, although this is unknown in practice.
-Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
-\label{fig:bs-seq_protocol}
-\end{figure}
-
-\subsection{Pre-sequencing}\label{sec:one_sample_pre_sequencing}
-
-A methylation locus is a single cytosine, that is, a CpG, CHG or CHH.
-The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., chr1:+:666.
-Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome.
-Not knowing the exact value of $N_{loci}$ is no great concern.
-
-It is sometimes convenient to refer to loci by the subscript $i$ rather than by $pos_{i}$.
-It is important to remember that the number of base pairs between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome.
-Also, depending on whether the data are stranded, a pair of loci may be on opposite strands.
-In a small number of instances the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes, e.g., the last CpG on chromosome 1 and the first CpG on chromosome 2.
-
-The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state.
-As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile.
-Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states.
-Therefore, it is also necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.
-
-Suppose that in the pool of DNA fragments for the sample that there are $H_{i}$ fragments containing the $i^{th}$ methylation locus.
-In general, $H_{i}$ is unknown and will vary from locus to locus within a sample[^H_i].
-Note that the value of $H_{i}$ is determined following the library preparation, including PCR amplification of the DNA; therefore, it can give a grossly distorted picture of the true representation of the cells.
-I denote by $\mathcal{H}_{i}$ the set of all fragments containing the $i^{th}$ locus.
-
-[^H_i]: Knowing $H_{i}$ would require knowing: (1) the number of cells used as input (which might only be known to within an order of magnitude), (2) the ploidy of each cell (generally known) and (3) the number of PCR cycles (generally known).
-But the real problem is that none of the steps in creating the pool of DNA fragments is perfect.
-In particular, PCR introduces biases -- some molecules are preferentially amplified while others 'drop out'.
-So even if we knew (1), (2) and (3) we cannot simply multiply these together to compute $H_{i}$, although this might at least give us a rough estimate.
-
-Although we do not know the number of fragments in the pool, we can define (and measure) the methylation state of a locus on a single DNA fragment.
-I denote by the indicator random variable, $Z_{h, i}$, the methylation state of $i^{th}$ methylation locus on the $h^{th}$ DNA fragment:
-
-\begin{equation*}
-Z_{h, i} = \left\{
-  \begin{array}{l l}
-    1 & \quad \text{if methylated on the } h^{th} \text{ fragment}\\
-    0 & \quad \text{if unmethylated on the } h^{th} \text{ fragment}
-  \end{array} \right.
-\end{equation*}
-
-By summing over the number of fragments containing the $i^{th}$ locus, we obtain the number of fragments that are methylated at the $i^{th}$ locus ($M_{i}$) and unmethylated at the $i^{th}$ locus ($U_{i}$):
-
-\begin{align*}
-  M_{i} &= \sum_{h = 1}^{H = H_{i}} Z_{h, i} = |\{Z: Z \in \mathcal{H}_{i}, Z = 1 \}| \\
-  U_{i} &= \sum_{h = 1}^{H = H_{i}} (1 - Z_{h, i}) = |\{Z: Z \in \mathcal{H}_{i}, Z = 0 \}|
-\end{align*}
-
-From these we can compute the proportion of fragments that methylated at the $i^{th}$ locus:
-
-\begin{equation*}
-  B_{i} = \frac{M_{i}}{M_{i} + U_{i}}
-\end{equation*}
-
-The above definitions can be extended from individual methylation loci, 1-tuples, to m-tuples. Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.
-
-I denote by the vector of indicator random variables, $Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)}$, the methylation pattern of the m-tuple $(i, i + 1, \ldots, i + m - 1)$ on the $h^{th}$ DNA fragment containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$:
-
-\begin{equation*}
-Z_{h, i} = \left\{
-  \begin{array}{l l}
-  (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus on the } h^{th} \text{ fragment}\\
-  (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th}) \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus on the } h^{th} \text{ fragment} \\
-  \vdots \\
-  (1, 1, \ldots, 1) & \quad \text{if methylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus on the } h^{th} \text{ fragment}
-  \end{array} \right.
-  \end{equation*}
-
-$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$ and $\mathcal{R}_{(i, i + 1, i + m - 1)}$ denotes the set of all reads containing the m-tuple $(i, i + 1, i + m - 1)$.
-
-There are $2^{\text{m}}$ possible methylation patterns at an m-tuple.
-Rather than describe a methylation pattern by a m-vector of zeros and ones, I also write these using $U$ and $M$; for example, the possible methylation patterns at a 2-tuple are $MM, MU, UM$ and $UU$.
-
-Analogous to the definition of $M_{i}$ and $U_{i}$ for 1-tuples ($\text{m} = 1$), we have when $\text{m} = 2$:
-
-\begin{align*}
-  MM_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (1, 1)\}| \\
-  MU_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (1, 0)\}| \\
-  UM_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (0, 1)\}| \\
-  UU_{(i, i + 1)} &= |\{Z: Z \in \mathcal{H}_{(i, i + 1)}, Z = (0, 0)\}| \\
-  \end{align*}
-
-We could extend the $B_{i}$ values to m-tuples, although the intuitive interpretation of these as the average methylation level is lost.
-Instead, it reflects the relative frequencies of each methylation pattern.
-Here are the definitions for $\text{m} = 2$:
-
-  \begin{align*}
-  B_{(i, i + 1)}^{MM} &= \frac{MM_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}} \\
-  B_{(i, i + 1)}^{MU} &= \frac{MU_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}}  \\
-  B_{(i, i + 1)}^{UM} &= \frac{UM_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}}  \\
-  B_{(i, i + 1)}^{UU} &= \frac{UU_{(i, i + 1)}}{MM_{(i, i + 1)} + MU_{(i, i + 1)} + UM_{(i, i + 1)} + UU_{(i, i + 1)}}
-  \end{align*}
-
-The definitions for $\text{m} > 2$ follow in the obvious manner.
-
-Again, I emphasise that $H_{(i, i + 1, i + \text{m} - 1), j}, Z_{h, (i, i + 1, i + \text{m} - 1), j}, B_{(i, i + 1, i + \text{m} - 1), j}$ and the set of methylation patterns are unobservable.
-However, by sequencing the pools of DNA fragments we aim to estimate these variables.
-
-\subsection{Post-sequencing}\label{sec:one_sample_post_sequencing}
-
-In a whole-genome bisulfite-sequencing experiment we do not sequence every DNA fragment in the pool.
-Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments.
-We have a large number(~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
-Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to $H_{i}$.
-
-At this point I ignore reads that do not contain any methylation loci as these are not relevant to this discussion and I make three simplifying assumptions:
-
-1. Sequencing is performed without error.
-2. Read mapping is perfect.
-3. We perform single-end sequencing.
-
-The effect of the first two assumptions are discussed in Section \ref{sec:mapping_and_post-processing}.
-The effect of the third assumption is minor.
-When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
-However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur (__FIGURE__):
-
-1.
-There may be gaps due to the insert size being longer than the sum of the read lengths, e.g.
-$(i, i + 1, i + 3, i + 4))$.
-In effect, we have missing data for any intervening methylation loci; the $(i + 2)^{th}$ loci in this example.
-2.
-Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g.
-read_1 gives us $(i, i + 1)$ and read_2 gives us $(i + 1, i + 2, i + 3)$.
-In this example we must use only one of read_1 or read_2 as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
-
-Each read measures the methylation state of one or more loci from a single DNA fragment.
-I denote by $\mathcal{R}_i$ the set of all reads containing the $i^{th}$ locus.
-Therefore, the number of reads containing the $i^{th}$ locus is $|\mathcal{R}_{i}|$, where $|\mathcal{R}_{i}| \leq H_{i}$ with strict inequality for almost all $i$.
-
-A single read containing the $i^{th}$ locus is denoted $z: z \in \mathcal{R}_{i}$ and the observed methylation state is indicated by:
-
-\begin{equation*}
-z: z \in \mathcal{R}_{i} = \left\{
-  \begin{array}{l l}
-    1 & \quad \text{if methylated at the } i^{th} \text{ locus}\\
-    0 & \quad \text{if unmethylated at the } i^{th} \text{ locus}
-  \end{array} \right.
-\end{equation*}
-
-By summing over the number of reads containing the $i^{th}$ locus we obtain the number of reads that are methylated at the $i^{th}$ locus ($m_{i}$) and unmethylated at the $i^{th}$ locus ($u_{i}$):
-
-\begin{align*}
-  m_{i} &= \sum_{z: z \in \mathcal{R}_{i}} z \\
-        &= |\{z: z \in \mathcal{R}_{i}, z = 1 \}| \\
-  u_{i} &= \sum_{z: z \in \mathcal{R}_{i}} (1 - z) \\
-      &= |\{z: z \in \mathcal{R}_{i}, z = 0 \}
-\end{align*}
-
-From these we can compute the proportion of reads that are methylated at the $i^{th}$ locus as:
-
-\begin{equation*}
-  \beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}
-\end{equation*}
-
-These are the so-called $\beta$-values, which are commonly interpreted as an estimate of the proportion of cells in the sample that are methylated at the $i^{th}$ locus.
-I discuss this interpretation, and other estimators of the  methylation level at a locus, in Section \ref{sec:estimating_B}.
-
-Again, these definitions can be extended from 1-tuples to m-tuples.
-The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$.
-A single read containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, and the observed methylation state is given by:
-
-\begin{equation*}
-z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)} = \left\{
-  \begin{array}{l l}
-    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus}\\
-    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th}) \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
-    \vdots \\
-    (1, 1, \ldots, 1) & \quad \text{if methylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus}
-      \end{array} \right.
-\end{equation*}
-
-Note that we do not know from which haplotype ($h$) each read came from, only that all methylation loci in the read came from the same DNA fragment.
-
-By counting over the number of reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ we obtain the number of reads contain each methylation pattern at that m-tuple.
-Here are the definitions for $\text{m} = 2$:
-
-\begin{align*}
-  mm_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (1, 1)\}| \\
-  mu_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (1, 0)\}| \\
-  um_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (0, 1)\}| \\
-  uu_{(i, i + 1)} &= |\{z: z \in \mathcal{R}_{(i, i + 1)}, z = (0, 0)\}|
-\end{align*}
-
-We can extend the $\beta_{i}$ values to m-tuples. Here are the definitions for $\text{m} = 2$:
-
-\begin{align*}
-  \beta_{(i, i + 1)}^{mm} &= \frac{mm_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{mu} &= \frac{MU_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{um} &= \frac{UM_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{uu} &= \frac{UU_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}}
-\end{align*}
-
-The definitions for $\text{m} > 2$ follow in the obvious manner.
-
-One final definition is the average methylation of each read, which is used by some authors, including \cite{Landan:2012kp}.
-For each read, $z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, the average methylation of the read, $\zeta_{z}$, is defined as the proportion of methylation loci in the read that are methylated.
-Thus, $\zeta_{z} = 0, \frac{1}{m}, \frac{2}{m}, \ldots, 1$.
-
-\subsection{Some complications for a single sample}\label{sec:one_sample_complications}
-
-I now discuss some complications and how this framework might accommodate these issues in practice.
-
-\subsubsection{What is $\mathcal{I}$?}\label{sec:what_is_I}
-
-As mentioned in Chapter \ref{chap:wgbs_bioinformatics_analysis}, studies using bisulfite-conversion assays rely on either a reference genome or, less commonly, separate DNA sequencing of the sample that is assayed.
-Different analysis strategies lead to different definitions of $\mathcal{I}$, which are approximations to the 'true' set of methylation loci in the sample, $\mathcal{I}^{truth}$.
-Listed here are definitions of $\mathcal{I}$ from least closely matching to most closely matching $\mathcal{I}^{truth}$:
-
-1. $\mathcal{I}^{ref}$: Defined by the set of methylation loci in the reference genome. This ignore all genetic variation between the sample and the reference.
-2.
-$\mathcal{I}^{refFilter}$: Defined by filtering out problematic loci from $\mathcal{I}^{ref}$.
-A conservative approach that removes many sites of genetic variation between the sample and the reference as well as sites that do not display genetic variation between the sample and the reference.
-This approach cannot identify sample-specific methylation loci.
-3.
-$\mathcal{I}^{Bis-SNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}.
-Identifies sample-specific methylation loci and removes reference-specific methylation loci.
-This is the best approach if only bisulfite-sequencing data are available.
-4.
-$\mathcal{I}^{WGS}$: Defined by identifying all methylation loci from _whole-genome sequencing_ (WGS) of the sample's genome.
-The gold standard.
-All methylation loci are defined with respect to the sample's genome.
-The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth}$ are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.
-
-\subsubsection{Genetic heterozygosity at a methylation locus}\label{sec:heterozygosity}
-
-The genome of a diploid organism has sites that are heterozygous due to differences between the maternal and paternal chromosomes.
-Such heterozygous loci are sometimes also methylation loci.
-For example, a locus where the maternal chromosome is a CpG and the paternal chromosome is an ApG.
-In effect, the maternal and paternal chromosomes within the sample have different $\mathcal{I}^{truth}$.
-
-The number of these genetically heterozygous methylation loci is often small enough not to worry about.
-However, in some studies, such as those of allele-specific methylation, these loci can be very important and might be identified by calling heterozygous genetic variants using `Bis-SNP` or from WGS of the sample.
-In practice, the existence of such loci is often ignored.
-
-\section{Multiple samples}\label{sec:multiple_samples}
-
-From a purely notational perspective, the move from a single sample to multiple samples simply requires an additional subscript, $j = 1, \ldots, n$, where $n$ is the number of samples.
-This defines the three levels in the hierarchy of a typical experiment -- DNA fragments ($h$), methylation loci ($i$) and samples ($j$).
-For example, $\mathcal{I}_{j}$ is the set of methylation loci in the $j^{th}$ sample and $\beta_{i, j}$ is the beta value for the $i^{th}$ locus in the $j^{th}$ sample.
-
-A fourth level is how the sample's relate to one another, such as the treatment group of each sample.
-This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between tumour and normal tissue.
-Conversely, the aim of the experiment might be to _discover_ this grouping, such as in a clustering analysis.
-
-We can define this fourth level using a design matrix $X = [X_{j}]$.
-For example, in a two-group experiment $X_{j} = 1$ if the sample is from group $1$ and $X_{j} = 0$ if the sample is from group $2$.
-We can also include covariates in the standard way by allowing $X_{j}$ to be a row vector, $X_{j} = (x_{1, j}, \ldots, x_{P, j})$, where $x_{p, j}$ encodes the information on the $p^{th}$ covariate for the $j^{th}$ sample.
-
-\subsection{Some complications with multiple samples}\label{sec{multi_sample_complications}}
-
-In addition to the complications of the Section \ref{sec:one_sample_complications}, we now have sample-to-sample variation that must be addressed by this framework.
-
-\subsubsection{What is $\mathcal{I}$?}
-
-Each sample has their own set of methylation loci, that is, $\mathcal{I}_{j}$ differs between $j$.
-Furthermore, sequencing coverage varies from sample-to-sample.
-This means that even if the samples have exactly the same $\mathcal{I}_{j}$, i.e., the samples are genetically identical, each sample will have a different set of loci with 'sufficient' sequencing coverage.
-Loci without sufficient sequencing coverage are effectively missing data.
-
-In practice, we might choose to study $\mathcal{I}^{common} = \cap_{j} \mathcal{I}_{j}$ or some other intersection of the $\mathcal{I}_{j}$, such as all methylation loci present in at least some fraction of the $n$ samples.
-
-A conservative analysis might only analyse those loci where at least some fraction of the $n$ samples have sufficient sequencing coverage.
-A less conservative analysis might try to impute the missing values based on methylation levels at neighbouring loci.
-
-\section{Parameter estimation}\label{sec:parameter_estimation}
-
-The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}}$.
-To do so it is necessary to estimate $M$ and $U$.
-In this section I review various methods for estimating these key parameters.
-When necessary, I have 'translated' the original work into my notation to make these methods more readily comparable.
-I have suppressed $j$ subscript when referring to a single sample.
-
-\subsection{Estimating $M$, $U$}\label{sec:estimating_m_and_u}
-
-It is rare to need direct estimates of $M_{i}$ or $U_{i}$.
-In order to estimate $M_{i}$ and $U_{i}$, the absolute number of methylated and unmethylated DNA fragments at the $i^{th}$ locus, we would also require an estimate of $|\mathcal{H}_{i}|$, the number of DNA fragments containing the $i^{th}$ locus.
-Rather, we are generally interested in estimating the proportion of reads that are methylated, $B_{i} = \frac{M_{i}}{M_{i} + U_{i}}$.
-
-\subsection{Estimating $B$}\label{sec:estimating_B}
-
-The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
-The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).
-
-Under a Binomial model for the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample, $M_{i, j} | n_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.
-
-Recently, more sophisticated methods have been proposed to estimate or model the average methylation level.
-These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
-
-\subsubsection{Beta-binomial models}\label{sec:beta-binomial_models}
-
-Several papers have proposed the beta-binomial distribution since, as noted by \cite{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" .
-The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing.
-Another way to think of this is that the 'beta' component models the biological variability of the data, while the 'binomial' component models the sampling variability of sequencing.
-This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data.
-For example, the `edgeR` software \cite{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.
-
-An attractive feature of the beta-binomial distribution is that it can be motivated by, and analysed with, Bayesian methods, including empirical Bayes methods, or frequentist techniques such as maximum likelihood.
-For example, the software `DSS` \cite{Feng:2014iq} and `MOABS` \citet{Sun:2014fk} both use the beta-binomial distribution in an empirical Bayes analysis of differential methylation from bisulfite-sequencing data.
-In contrast, `RADmeth` \cite{Dolzhenko:2014bo} uses the beta-binomial model in a maximum likelihood framework to address the same problem.
-
-\subsubsection{Smoothing $\beta$-values}\label{sec:smoothing_beta-values}
-
-`BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$.
-Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$.
-Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation} and \ref{chap:co-methylation_review}).
-
-Smoothing is particularly powerful for loci with low sequencing coverage, where the denominator $m_{i, j} + u_{i, j}$ is small and the corresponding standard error of $\beta_{i, j}$ is large.
-The smoothed $\beta$-values, rather than the raw $\beta$-values, are then generally used in all downstream analyses.
-
-Both `bsseq` and `BiSeq` use a binomial local likelihood smoother.
-In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$.
-The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.
-
-In both `bsseq` and `BiSeq` the raw $\beta$-values are weighted according to the binomial likelihood and a kernel function.
-The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the kernel gives greater weight to those $\beta_{i, j}$ near the centre of the window.
-\citet{Lacey:2013iy} note that loci with very high sequencing coverage will strongly influence the smoother, potentially biasing estimates at neighbouring loci with lower coverage.
-
-`bsseq` assumes that for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$.
-In contrast, `BiSeq` first creates clusters of CpGs and only assumes that the underlying methylation level is smooth at positions within each cluster.
-
-Whenever smoothing is used, a key parameter is the bandwidth, which is the size of the window in which observations are included at each iteration of the smoother.
-`bsseq` uses a much larger window size than `BiSeq`; the default window size in `bsseq` is one that contains at least 70 CpGs and is at least 2000kb wide, whereas the default window size in `BiSeq` is 80bp, regardless of CpG-density.
-This is due to `BiSeq` being developed for RRBS data, which has a high CpG-density per window, whereas `bsseq` was developed for whole-genome data, which has a more variable, and lower on average, CpG density per window.
-
-Another 'parameter' choice when smoothing is the choice of kernel, although this is generally less important than the choice of bandwidth.
-`bsseq` uses a tricube kernel and `BiSeq` uses a triangular kernel.
-
-\citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} compare the smoothing results of `BiSeq` to `bsseq`.
-Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`, but the comparison is a little unfair since both use RRBS data, for which `bsseq` is not designed[^rrbs_sim].
-
-[^rrbs_sim]: Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} altered the default `bsseq` parameters to try to make them comparable to `BiSeq`.
-\citet{Hebestreit:2013ko} changed the default minimum window size to 80bp but still required at least 20 CpGs per window.
-\citet{Lacey:2013iy} kept the default minimum window size of 2000 bp but reduced the minimum number of CpGs per window to 50 from the default of 70.
-
-\section{Statistical properties of $\beta$-values}\label{sec:beta}
-
-Under the binomial model, ($M_{i, j} | n_{i, j}) \eqd Bin(n_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt(\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}})$ \citep{Hansen:2012gr}.
-The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample.
-In this section I discuss this interpretation and statistical properties of this estimator.
-
-\subsection{Empirical distributions}
-
-- TODO: Lots of figures and discussion. See chapter plan.
-
-In a study involving multiple samples, the set of $\beta$-values can be summarised as a matrix where each row is a locus and each column is a sample.
-Some values will be missing, either because there was insufficient sequencing coverage to estimate a $\beta$-value or because that locus is not a cytosine for that sample.
-This matrix might be visualised to learn about the distribution of methylation levels, either row-wise, to learn about the variability across samples, or column-wise, to learn about the variability within samples.
-
-\subsubsection{Genome-wide distribution of $\beta$-values}
-
-Restricting our attention to CpGs, Figure \ref{fig:EPISCOPE_beta_distribution} shows the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each EPISCOPE sample.
-What is immediately clear is that these distributions are bimodal.
-Figure \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} shows that this bimodality is largely driven by the fact that CpGs in CpG islands are mostly unmethylated whereas those outside of CpG islands are generally methylated.
-Intermediate $\beta$-values, those between $0.2-0.8$, are evidence of heterogeneity.
-
-Both Figures \ref{fig:EPISCOPE_beta_distribution} and  \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} also reveal that the genome-wide distributions of $\beta$-values for a given tissue are very similar between the three individuals.
-The three adipose-derived tissues, _SA_, _VA_ and _VAT_, have similar genome-wide profiles while the blood-derived tissue, _BUF_, has more intermediate $\beta$-values, potentially indicating more cellular heterogeneity.
-
-```{r EPISCOPE_genome_wide_beta-value_distributions, eval = FALSE, echo = FALSE}
-# Run on one of the unices
-setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
-library(MethylationTuples)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-thesis_theme <- theme_classic(base_size = 20)
-thesis_theme <- theme_bw(base_size = 20)
-
-x <- readRDS("EPISCOPE_1_tuples_strand_collapsed.rds")
-
-cgi <- read.table("~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt",
-                  header = TRUE, stringsAsFactors = FALSE)
-cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
-cgi_status <- overlapsAny(x, cgi)
-
-beta <- methLevel(x, min_cov = 10L)
-
-cn <- colnames(beta)
-beta <- data.frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
-                   Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
-                   beta = as.vector(beta),
-                   CGI = rep(cgi_status, times = ncol(beta)),
-                   stringsAsFactors = FALSE)
-beta <- tbl_df(beta)
-
-#a <- beta[sample(nrow(beta), 100000), ]
-
-g <- ggplot(aes(x = beta), data = beta) +
-  geom_density(size = 1.5) +
-  facet_grid(Tissue ~ Individual, labeller = label_both) +
-  thesis_theme + scale_colour_brewer(palette = "Dark2") +
-  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
-  xlab(expression(paste(beta, "-values", sep = "")))
-ggsave("EPISCOPE_beta_distribution.pdf", g, height = 9, width = 16)
-
-g <- ggplot(aes(x = beta), data = beta) +
-  geom_density(aes(colour = CGI), size = 2) +
-  facet_grid(Tissue ~ Individual, labeller = label_both) +
-  thesis_theme + scale_colour_brewer(palette = "Dark2") +
-  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
-  xlab(expression(paste(beta, "-values", sep = "")))
-ggsave("EPISCOPE_beta_distribution_CGI_stratified.pdf", g, height = 9,
-       width = 16)
-```
-
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
-Most CpGs are highly unmethylated or highly methylated, with the buffy coat samples being the most extreme example of this bimodality.
-'Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose.}
-\label{fig:EPISCOPE_beta_distribution}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution_CGI_stratified.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
-Only CpGs with at least $10\times$ sequencing coverage are included.
-Most CpGs outside of CpG islands are methylated whereas most CpGs within CpG islands are unmethylated.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose; \em{CGI} = CpG island.}
-\label{fig:EPISCOPE_beta_distribution_CGI_stratified}
-\end{figure}
-
-\subsubsection{Strand-specific summaries}
-
-CpG $\beta$-values are often computed by aggregating the $m$ and $u$ counts across the forward and reverse strands.
-This assumes that the two strands are indeed symmetrically methylated.
-To investigate this, for each CpG we can compute the $\beta$-values using only the forward-strand reads, $\beta^{+}$, and only the reverse-strand reads, $\beta^{-}$.
-
-```{r EPISCOPE_strand_specific_beta-value_distributions, eval = FALSE, echo = FALSE}
-# Run on one of the unices
-setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
-library(MethylationTuples)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-library(RColorBrewer)
-
-thesis_theme <- theme_classic(base_size = 20)
-thesis_theme <- theme_bw(base_size = 20)
-
-y <- readRDS("EPISCOPE_1_tuples.rds")
-y_fwd <- y[strand(y) == "+"]
-y_rev <- y[strand(y) == "-"]
-
-ol <- findOverlaps(unstrand(y_fwd), shift(unstrand(y_rev), -1L), type = "equal")
-beta_fwd <- methLevel(y_fwd, min_cov = 5L)[queryHits(ol), ]
-beta_rev <- methLevel(y_rev, min_cov = 5L)[subjectHits(ol), ]
-
-cors <- matrix(unlist(mclapply(1:10, function(min_cov, y_fwd, y_rev, ol) {
-  beta_fwd <- methLevel(y_fwd, min_cov = min_cov)[queryHits(ol), ]
-  beta_rev <- methLevel(y_rev, min_cov = min_cov)[subjectHits(ol), ]
-  sapply(seq_len(ncol(y)), function(i, x, y) {
-    cor(x[, i], y[, i], use = "complete.obs")
-    }, x = beta_fwd, y = beta_rev)
-  }, y_fwd = y_fwd, y_rev = y_rev, ol = ol, mc.cores = 10)), ncol = 10)
-
-beta_pairs <- data.frame(Individual = rep(substr(cn, 1, 3),
-                                          each = nrow(beta_fwd)),
-                         Tissue = rep(substr(cn, 4, 6), each = nrow(beta_fwd)),
-                         beta_fwd = as.vector(beta_fwd),
-                         beta_rev = as.vector(beta_rev),
-                         stringsAsFactors = FALSE)
-beta_pairs <- tbl_df(beta_pairs)
-
-beta_pairs %>% group_by(Individual, Tissue) %>%
-  summarise(cor(beta_fwd, beta_rev, use = "complete.obs"))
-
-mypar <- function(a = 1, b = 1, brewer.n = 8, brewer.name = "Dark2",
-                  cex.lab = 2, cex.main = 2, cex.axis = 1.5,
-                  mar = c(5.1, 5.1, 3.5, 2.1), mgp = c(3, 1, 0), ...) {
-    par(mar = mar, mgp = mgp, cex.lab = cex.lab, cex.main = cex.main,
-        cex.axis = cex.axis)
-    par(mfrow = c(a, b), ...)
-    palette(brewer.pal(brewer.n, brewer.name))
-}
-pdf("EPISCOPE_beta_correlations_across_strand.pdf", width = 16, height = 9)
-mypar(4, 3)
-sapply(seq_len(ncol(beta_fwd)), function(i, x, y, ns) {
-  smoothScatter(x[, i], y[, i],
-                xlab = expression(beta["+"]),
-                ylab = expression(beta["-"]),
-                main = paste0(ns[i], ": cor = ", round(cor(x[, i], y[, i],
-                                                        use = "complete.obs"),
-                                                    2)))
-  }, x = beta_fwd, y = beta_rev, ns = colnames(beta_fwd))
-dev.off()
-```
-
-\subsection{Interpretation}
-
-In papers about DNA methylation there is often a sentence like the following (given by \cite{Laird:2003kw}): "About 70% of the CpG dinucleotides in the mammalian genome are methylated".
-In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
-However, as can be seen from __FIGURE__, the extreme bimodality of the $\beta$-value distribution means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
-To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
-
-The "$70\%$" statement can also be interpreted as an estimate of the probability that a cytosine randomly sampled from a haploid copy of a mammalian genome is a methylcytosine.
-Because in a whole-genome bisulfite-sequencing experiment we measure methylation from a pool of cells, not a single haploid copy of the genome, I do not think it makes sense to describe a particular cytosine in the genome as being a 'methylcytosine'.
-That is, it only makes sense to describe an observation of a cytosine, $z_{i}$, as being methylated or unmethylated and not to describe a cytosine, $i$, as being a methylated or unmethylated.
-However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have sought to use the latter definition, which I believe to be an unnecessary source of confusion.
-
-[^Lister_2013]: It is worth noting that this concept was not used in more recent paper from the same group \citep{Lister:2013et}.
-
-\citet{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
-In the language of these papers, a 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
-This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$.
-This can be thought of as testing the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.
-
-Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows.
-For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
-Any site with an FDR-adjusted P-value below a threshold was declared a 'methylcytosine'[^fdr]
-
-[^mathematical_description]: The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy}.
-\citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}
-
-[^fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$.
-I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated.
-This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
-thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
-This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.
-
-The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error rates estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
-It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
-
-The proportion of cytosines that are identified by this procedure as 'methylcytosines' is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous.
-For example, suppose we had a sample where the true methylation level of every CpG, $B_{i}$, was $0.2$.
-Given sufficient sequencing coverage, every CpG in the genome would be declared a 'methylcytosine' when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.
-
-Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values.
-For example, in \citet[supplementary figure 2a]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates.
-A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.
-
-In summary, for bisulfite-sequencing experiments where the DNA for each sample comes from multiple cells I do not think it makes sense, nor is it useful, to refer to individual cytosines, $i$, as being methylated or unmethylated.
-Instead, it is better to summarise the methylation level at a CpG by a $\beta$-value since this has a natural interpretation as the estimated proportion of haploid genomes in the sample that are methylated at that CpG.
-Unfortunately, $\beta$-values are not without their own issues, as discussed below.
-
-\subsection{Correlations}
-
-Many researchers have observed that DNA methylation is spatially correlated along the genome, \cite[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}).
-I call this correlation of methylation levels _co-methylation_.
-
-Just as we can explore the distribution of DNA methylation levels within a sample or between samples, so too can we explore co-methylation within a sample or between samples.
-I examine this in detail in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
-
-\subsection{Biases}
-
-The natural interpretation of $\beta_{i, j}$ as the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample will be biased if the probability of sequencing a fragment with a methylated site is different from the probability of sequencing a fragment with an unmethylated site.
-In fact, it has been shown that methylated DNA is overrepresented in bisulfite-sequencing data due, with the problem exacerbated by higher rounds of PCR amplification and dependent on the bisulfite-conversion protocol \cite{Ji:2014fq}.
-PCR amplification can result in overreprestation of one of the DNA strands in bisulfite-sequencing data \cite{Warnecke:1997eh}.
-
-Lab-based solutions to overcome these biases exist for targeted bisulfite-sequencing, but are technically difficult and expensive to extend to whole-genome studies.
-\cite{Ji:2014fq} propose potential computational corrections for these biases but as yet these have not been implemented.
-
-\subsection{Transformations}\label{sec:transformations}
-
-$\beta$-values are the _de facto_ standard unit for reporting methylation levels due to their natural interpretation as an estimate of the average level of methylation at the locus.
-However, they are not necessarily the best unit for statistical inference.
-This is because a $\beta$-value is an estimate of a proportion and there are a well-known challenges when working with proportion data, such as:
-
-1.
-The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$.
-Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
-2.
-We need to more than just the $\beta$-value to have a sense of how precise an estimate it is.
-Essentially, we need to also know the sequencing coverage of the methylation loci.
-Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100$ $u = 300$.
-Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision.
-Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.
-3. Proportions are bound between 0 and 1, inclusive.
-
-To address (1), proportion data are often transformed via a variance stabilisation transformation.
-The aim is to make the variance independent of the mean, at least approximately.
-Popular variance stabilisation transformations include:
-
-- The arcsine transformation, $\arcsin({\sqrt{\frac{m + 1}{m + u + 1}}})$ \citep{ANSCOMBE:1948bw}. A small value, in this case 1, is added to both $m$ and $u$ to avoid $\beta = 0, 1$.
-- The "averaged arcsine" transformation, $\arcsin{\sqrt{\frac{m}{m + u + 1}}} + \arcsin{\sqrt{\frac{m + 1}{m + u + 1}}}$ \citep{Freeman:1950bh}.
-One problem with this transformation is that it does not have a unique inverse \citep{Nunes:2009vj}.
-
-However, the general use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \cite[e.g.,]{Warton:2011gm}.
-
-One transformation that remains popular, at least in the analysis of DNA methylation microarray data, is the logit-transformation, also known as $\mathcal{M}$-values.
-An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2}(\frac{\beta}{1 - \beta} = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero.
-Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \cite[e.g.]{Smyth:2005ta}.
-
-\cite{Du:2010dc} advocate for the use of $\mathcal{M}$-values in statistical analyses of methylation levels from microarray data.
-The main reason for this is that $\mathcal{M}$-values are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values.
-As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformatics and genomics researchers makes it a favourable choice.
-Nonetheless, \cite{Du:2010dc} also advocate that the results of analyses should be reported as $\beta$-values, owing to their "more intuitive biological interpretation".
-
-As with $\beta$-values, $\mathcal{M}$-values derived from bisulfite-sequencing data generally cannot be directly analysed due to the variable sequencing coverage across loci.
-However, the variable sequencing coverage can be accounted for using the afore-mentioned beta-binomial models and related regression models, which are discussed in the next section.
-
-\section{Differential methylation}
+\section{Differential methylation}\label{sec:differential_methylation}
 
@@
 
-\section{Other downstream analyses}
+\section{Other downstream analyses using $\beta$-values}
 
-There are many other questions that can be addressed using bisulfite-sequencing data.
-Like differential methylation, some of these might be also be tackled with other methylation assays, such as microarrays.
-However, as we will see, some can only be answered by assays that have single nucleotide resolution, which is where bisulfite-sequencing can prove particularly powerful.
+Two other common downstream analyses based on $\beta$-values are the identification of variable methylation and partially methylated domains.
 
-\subsection{Partially methylated domains}
-
-Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$.
-An example of such a region is shown in __FIGURE__ (__TODO: Find a PMD and plot it__).
-
-PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}.
-Using a simple sliding window algorithm, \citeauthors{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb.
-They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample.
-This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.
-
-A subsequent study found that PMDs are a common feature of somatic cell lines and that they comprise $> 30\%$ of the genome \citep{Lister:2011kg}.
-Perhaps even more intriguingly, across four somatic cell lines profiled with whole-genome bisulfite-sequencing, \citeauthor{Lister:2011kg} found a large amount of these genomes ($664$ Mb) comprised shared PMDs.
-
-PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_).
-However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}.
-Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg}.
-
-More sophisticated methods have also been proposed to identify these PMDs.
-`methylSeekR` \citep{Burger:2013kq} is one such algorithm.
-It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions.
-While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that methylation microarrays cannot provide.
-
-\subsection{Variably methylated regions and differential variability}\label{sec:vmrs}
+\subsection{Variable methyation}\label{sec:variable_methylation}
 
@@
 Which approach is preferable will depend on the user's perspective as to whether such loci with outliers should be called as variably methylated.
+
+\subsection{Partially methylated domains}
+
+Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$.
+An example of such a region is shown in __FIGURE__ (__TODO: Find a PMD and plot it__).
+
+PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}.
+Using a simple sliding window algorithm, \citeauthor{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb.
+They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample.
+This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.
+
+A subsequent study found that PMDs are a common feature of somatic cell lines and that they comprise $> 30\%$ of the genome \citep{Lister:2011kg}.
+Perhaps even more intriguingly, across four somatic cell lines profiled with whole-genome bisulfite-sequencing, \citeauthor{Lister:2011kg} found a large amount of these genomes ($664$ Mb) comprised shared PMDs.
+
+PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_).
+However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}.
+Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg}.
+
+More sophisticated methods have also been proposed to identify these PMDs.
+`methylSeekR` \citep{Burger:2013kq} is one such algorithm.
+It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions.
+While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that methylation microarrays cannot provide.
+
+\section{Downstream analyses of methylation patterns at m-tuples}\label{sec:downstream_analyses_of_m-tuples}
+
+The above-described analyses do not specifically exploit the fact that bisulfite-sequencing reads can measure methylation patterns at m-tuples.
+Yet much can be learnt my studying exactly that.
+To begin, I review the different types of questions that can be addressed when we make use of this extra information that is only available from bisulfite-sequencing data.
+
+\subsection{Methylation entropy}\label{sec:methylation_entropy}
+
+__UP TO HERE__
+
+- `QDMR` \citep{Zhang:2011dp} and `CpG_MPs` \citep{Su:2012hl}, developed by the same research group, are methods for identifying DMRs using a form of methylation entropy from MeDIP-chip and BS-seq, respectively.
+- \citet{Xie:2011cy} introduced the concept of methylation entropy for BS-seq.
+
+While mathematically based on the idea of Shannon entropy \citep{Shannon:IH1_srIe}, the interpretation of methylation entropy has much in common with the physical concept of "disordered/heterogeneity"
 
@@
 
-\subsection{Methylation entropy}\label{sec:methylation_entropy}
-
 \subsection{Epipolymorphism}\label{sec:epipolymorphism}
@@
 
-\subsection{Integrating DNA methylation data with other genomes data}\label{sec:data_integration}
+\section{`MethylationTuples`}\label{sec:MethylationTuples}
 
-\section{General TODOs}
+A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are somewhat lacking.
+This in part motivated me to begin developing `MethylationTuples`, an R package for managing, analysing and visualising methylation patterns at m-tuples.
 
-* Discuss distribution of coverage? Perhaps in context of sequencing bias?
+- I'm using to address co-methylation
+- Could be used for entropy, epipolymorphism and epialleles.
+- Could implement Fang or Peng's method for ASM

### 2014-12-04 Rmarkdown/wgbs_statistical_framework.Rmd
Split wgbs_statistical_analysis into wgbs_statistical_framework and wgbs_downstream_analyses

--- 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
@@
 ---
-title: "Statistical analysis of whole-genome bisulfite-sequencing data"
+title: "Statistical framework for whole-genome bisulfite-sequencing data"
 output:
@@
 
-\chapter{Statistical analysis of whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_analysis}
+\chapter{Statistical framework for whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}
 
-\section{Chapter overview}\label{sec:statmodel_chapter_overview}
+\section{Chapter overview}\label{sec:wgbs_statistical_framework_overview}
 
-I set out a statistical framework for analysing bisulfite-sequencing data.
-I begin by explaining the various levels of stochasticity and then define the mathematical notation that I use throughout my thesis.
-I describe key variables, common estimators of these and their statistical properties.
-The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.
+In this chapter I set out a statistical framework for analysing bisulfite-sequencing data.
+The ideas here are simple, however, they have not yet been put into a unified mathematical framework.
+My intention in doing so is to clarify several subtleties that, in my experience, are potential sources of confusion.
 
-As is clear from the length of Chapter \ref{chap:wgbs_bioinformatics_analysis} alone, there is a considerable amount of bioinformatics work before we even consider analysing the data to address the biological question of interest.
-I call these the 'downstream' analyses, and there are many that might be considered.
-Using the above statistical framework, I explain the statistical question of identifying differential methylation.
-I also discuss other downstream analyses.
+Beginning with a single sample, I explain the various sources of variation in this statistical model and introduce the mathematical notation that I use throughout my thesis.
+I then extend this framework to multiple samples.
+Finally, I describe key variables, common estimators of these and their statistical properties by analysing __HOW MANY__ whole-genome bisulfite-sequencing samples.
 
@@
 
-\section{Differential methylation}
+\section{Summary}
 
-A common downstream analysis is the identification of differential methylation.
-This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
-Many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
-All of these methods are designed for two-group experiments, with some being general enough to extend to multiple-group experiments[^multi-group_extension].
-I describe in greater detail methods for identifying differential methylation in a one-group experiment, admitedly an unusual design but one that I faced during my PhD.
-
-[^multi-group_extension]: The extension to multiple-group experiments isn't always implemented in the software, even if discussed in the paper.
-
-\subsection{Experimental design}\label{sec:experimental_design}
-
-In any analysis of differential methylation we want the DMCs and DMRs to be both _biologically_ and _statistically_ significant; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
-Key to ensuring biological relevance is good experimental design, such as the use of _replicates_ in each experimental group.
-A distinction is often made between _technical replicates_ and _biological replicates_.
-Briefly, biological replicates are experimental units that all undergo the same treatment and are used to estimate the within-group variability of the treatment.
-Technical replicates are repeated measurements of the same experimental unit, perhaps with slight variations in the sample preparation, and are used to estimate the variability of the sample preparation and measurement process.
-
-As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one.
-For example, \cite{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine).
-I contend that these are better defined as technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].
-
-[^cell_passaging]: In the case of IMR90 cell line, the first replicate, IMR90_r1, underwent 4 cell passages and the second replicate, IMR90_r2, underwent 5 cell passages.
-In the case of the H1 cell line, the first replicate, H1_r1, underwent 25 passages in the first media and 9 passages in the second media, and the second replicate, H1_r2, underwent 27 passages in the first media and 5 passages in the second media.
-
-Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind, or pooled the replicates prior to analysis, thus ignoring all within-group variability, as is the case of the H1 and IMR90 cell lines in \cite{Lister:2009hy}.
-
-Ideally, technical variability is orthogonal to the biological variability, but this rarely occurs.
-Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.
-Again, this emphasises the importance of good experimental design such as randomisation, replication and the use of control, both positive and negative.
-
-\subsection{DMCs}\label{sec:dmcs}
-
-There have been reports of differential methylation at individual CpGs resulting in a phenotypic difference \citep[e.g.,]{Furst:2012gh}.
-However, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome experiments with sample sizes you can count on one or two hands and an average sequencing depth of $10-30\times$.
-Nonetheless, identifying DMCs, even if not an end in itself, is an important step in identifying DMRs.
-
-Identifying DMCs boils down to identifying differences in the mean level of methylation, for which there is an enormous body of statistical literature.
-The problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.
-
-Regardless of the test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty.
-Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments[^multiple_testing].
-Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.
-
-[^multiple_testing]; For example, there are approximately 20,000 tests in studies of differential gene expression and one million tests in genome-wide association studies.
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}ll@{}}
-\toprule
-Method                                       & Used by                                                                            \\ \midrule
-\cite{Benjamini:1995ws}                      & \cite{Akalin:2012cm, Jaffe:2012gx, Lacey:2013iy, Rijlaarsdam:2014hp, Sofer:2013bk} \\
-\cite{Wang:2011cw}                           & \cite{Akalin:2012cm}                                                               \\
-"False discovery rate correction/adjustment" & \cite{Dolzhenko:2014bo, Gokhman:2014cp, Lister:2009hy, Xie:2014ez}                 \\
-\cite{Storey:2007cp}                         & \cite{Jaffe:2012gx}                                                                \\
-\cite{Benjamini:2001ho}                      & \cite{Sofer:2013bk, Hebestreit:2013ko}                                             \\
-"Bonferonni adjustment"                      & \cite{Feng:2014iq}                                                                 \\ \bottomrule
-\end{tabular}
-\caption{Methods proposed for adjusting for the multiple hypothesis testing performed in an analysis of DMCs.
-Several papers use describe their analysing as performing a "false discovery rate adjustment" or "false discovery rate correction" without explicitly stating what they are doing or citing a reference.
-One paper uses an ultra-conservative Bonferonni correction.}
-\label{tab:multiple_testing}
-\end{table}
-
-One thing to note, however, is that the _effective_ number of tests is fewer than the actual number of tests.
-This is because the methylation levels at neighbouring loci are correlated (see Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}), which means that tests of differential methylation are generally positively correlated, thus reducing the effective number of independent tests.
-The classical Benjamini-Hochberg procedure also controls the false discovery rate under certain forms of positive dependence \citep{Benjamini:2001ho}.
-(__TODO: Do these techniques work in the high-dependence setting of DMC testing?__)
-
-\subsubsection{One-group experiments}\label{sec:one_group_dmcs}
-
-While an uncommon design, I did work on a one-group whole-genome bisulfite-sequencing experiment during my PhD.
-The data, courtesy of Harry Oey and Emma Whitelaw (La Trobe University), came from whole-genome bisulfite-sequencing of five _isogenic_ mice, meaning that these mice were genetically identical (or close to it).
-
-In a one-group experiment a DMC is a cytosine where at least one of the samples has a different methylation level to the group average.
-The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown in Table \ref{tab:one-group_dmc_testing}.
-
-\begin{table}[h]
-\centering
-\resizebox{\textwidth}{!}{%
-  \begin{tabular}{@{}lll@{}}
-  \toprule
-  & m          & u          \\ \midrule
-  $Sample_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
-  $\vdots$     & $\vdots$   & $\vdots$   \\
-  $Sample_{n}$ & $m_{i, n}$ &            \\ \bottomrule
-  \end{tabular}
-}
-\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
-\label{tab:one-group_dmc_testing}
-\end{table}
-
-Such a contingency table can be analysed in many different ways.
-Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j.
-$\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
-
-To the best of my knowledge, none of the software specifically written for analysing differential methylation analyses is able to test this hypothesis.
-This is because these software are almost all designed for two-group experiments.
-However, this hypothesis is readily testing using routine statistical methods.
-As an example, consider the following data:
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}lll@{}}
-\toprule
-& m    & u    \\ \midrule
-$Sample_{1}$ & $38$ & $1$  \\
-$Sample_{2}$ & $79$ & $2$  \\
-$Sample_{3}$ & $59$ & $1$  \\
-$Sample_{4}$ & $69$ & $2$  \\
-$Sample_{5}$ & $44$ & $46$ \\ \bottomrule
-\end{tabular}
-\caption{An example of a differentially methylated cytosine in a one-group experiment.}
-\label{tab:one-group_example}
-\end{table}
-
-We could analyse this table using a simple stand-alone test, such as Pearson's $\chi^{2}$ test.
-Alternatively, we could frame this as a generalised linear model, either a log-linear model or a logistic regression.
-The equivalence of these methods is shown in the Appendix.
-
-A key advantage of using either of the generalised linear models is the capability to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
-More complex regression analyses are also possible, such as empirical Bayes or fully Bayesian models.
-
-\subsubsection{Multi-group experiments}\label{sec:multi_group_dmcs}
-
-Due to the cost of whole-genome bisulfite-sequencing, early experiments using this technology to study differential methylation compared one sample against another \citep[e.g.,]{Lister:2009hy, Berman:2012ga}.
-While this might be considered a two-group experiment, one without replicates, I believe this is better thought of as a one-group experiment with two samples.
-Indeed, the analysis of such an experiment uses techniques from Section \label{sec:one_group_dmcs}, e.g., Pearson's $\chi^{2}$ test or Fisher's exact test, which is better suited to the small sample sizes routinely seen in each $2 \times 2$ table.
-
-The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
-Unfortunately, some authors have still used Fisher's exact test to analyse such experiments.
-For example, \cite{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data the two replicates they have for each sample.
-This is incorrect, because such an analysis effectievly ignores all within-group variability \citep{Hansen:2012gr}.
-Instead, a test that incorporates the within-group variability should be used.
-
-In a two-group experiment this might be a t-test comparing the the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$.
-More generally, we can use a regression model to allow for multiple groups (i.e., ANOVA) and to incorporate additional covariates.
-
-Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment.
-Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (see Section \ref{sec:beta-binomial_model}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (see Section \ref{transformation}).
-
-`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
-`DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
-`BiSeq` and `methlySig` also perform spatial smoothing of the data.
-
-The test used to identify DMCs in these regression models is variously a Wald test (`BiSeq`, `DSS`), a likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).
-
-Not all methods for identifying differential methylation are designed for identifying DMCs.
-Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs.
-
-\subsection{DMRs}
-
-A differentially methylated region is a region of the genome where there are multiple cytosines with evidence of differential methylation.
-Importantly, not all cytosines in the region must be genome-wide significant DMCs.
-Rather, the idea is that a DMR might capture a weaker but broader difference in methylation.
-Such a signal may well have a clearer biological interpretation and, indeed, might be expected such as in differences in methylation at a gene promoter.
-
-There are two very different strategies for identifying differentially methylated regions:
-
-1.
-Using regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands \citep[e.g.,]{Huang:1999fs,Doi:2009ky}[^array], MspI restriction fragments \citep[e.g.,]{Stockwell:2014fq}, any other predefined genomic feature, such as a gene promoters or transcription factor binding sites, or general predefined bins \citep[e.g., 100 bp bins]{Park:2014ho}.
-2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
-
-[^array]: Both these examples are from microarray studies, but the same idea can be applied to sequencing studies.
-
-The former is much simpler to analyse but is limited in its ability to discover new differentially methylated regions.
-It is also hampered because the correct unit or scale for differential methylation may not be known for the experiment.
-
-The latter offers the opportunity to identify previously unknown regions that are subject to differential methylation.
-Included in this is the opportunity to discover the scale over which differential methylation acts.
-However, valid statistical inference of these regions is far more challenging.
-
-\subsection{Using _a priori_ regions}
-
-The idea of testing for differential methylation is relatively straightforward.
-For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region, for example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage.
-These averages are then compared between samples.
-I describe a basic implementation of this for a one-group experiment and then detail a more sophisticated method that is implemented in `BiSeq` \citep{Hebestreit:2013ko}.
-
-\subsubsection{One-group experiments}\label{sec:one_group_dmrs}
-
-For a region containing $c$ methylation loci, construct the $n \times 2 \times c$ contingency table, i.e., each 'slice' of the table is a $n \times 2$ table containing the data for a single locus.
-Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions (__CITE: http://www.biostathandbook.com/cmh.html__).
-Instead, we could fit the log-linear model with `sample`, `meth_type` and `locus` as the set of factors.
-The saturated model is
-
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus} + \lambda_{a, b, c}^{meth_state, sample, locus}
-\end{equation*}
-
-We would first need to test whether there is a significant three-way interaction between `meth_type`, `sample` and `locus` by computing the deviance of the model without the three-way interaction
-
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{equation*}
-
-On the one hand, if there is no three-way interaction, then we can test for an interaction between `meth_type` and `sample`, i.e., test for a difference in methylation level across the loci, by comparing the model
-
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{equation*}
-
-to the model
-
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{equation*}
-
-This test is analogous to the generalised Cochran-Mantel-Haenszel test \citep{Landis:1978vk,MANTEL:1978iv,Birch:1965uj} of the $n \times 2 \times c$ table \citep[p231]{agresti1990categorical}.
-The advantage of the Cochran-Mantel-Haenszel is its applicability for sparse datasets for which the asymptotic theory does not hold for the test based on the log-linear regression  \citep[p232]{agresti1990categorical}.
-
-On the other hand, if there is a significant three-way interaction between `meth_type`, `sample` and `locus` then this requires more careful modelling to determine the nature of this interaction and whether the region is a consistent DMR.
-
-\subsubsection{A more sophisticated multi-group analysis}\label{sec:a_more_complex_example}
-
-`BiSeq` is designed for RRBS data and tests for differential methylation at "CpG clusters".
-These CpG clusters are based on an initical scan of the genome looking for CpGs that are within a given distance of one another and that have sufficient sequencing coverage across the set of samples.
-While there is clearly a data-dependent component to these cluster definitions, I reserve the use of 'data-driven regions' for those which are based on the methylation levels of loci rather than their genomic co-ordinates.
-\citep{Hebestreit:2013ko} alternatively recommend using the target regions of the assay, such as MspI fragments in the case of RRBS, instead of creating these CpG clusters.
-
-One the clusters are defined, the $\beta$-values in each cluster are smoothed using a local binomial likelihood smoother.
-For each CpG `BiSeq` fits a beta regression model[^not_beta-binomial], which is tested for evidence of differential methylation at that cytosine.
-Based on these P-values, `BiSeq` then use a hierarchical testing procedure to control the false discovery rate at both the cluster-level and locus-level.
-This method is based on several papers by Yoav Benjamini and colleagues \citet{Benjamini:1997fh, Benjamini:2001ho, Benjamini:2006gd, Benjamini:2007cm}.
-It aims to first control the false discovery rate at the cluster-level and then refine the signal by trimming non-DMCs from those clusters that have been declared as differentially methylated.
-Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_ (see Section \ref{sec:consistency_of_dmrs}).
-
-[^not_beta-binomial]: This is different to the beta-binomial regression framework described in Section \ref{sec:beta-binomial_models}.
-
-Based on my understanding of this procedure, only cytosines within differentially methylated clusters can be declared DMCs since it is only those cytosines that are tested in the second-step of the hierarchical testing algorithm.
-Therefore, isolated DMCs will not be detected by this procedure.
-
-\subsection{Using data-driven regions}
-
-The most common type of algorithm for identifying data-driven DMRs, also arguably the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
-The initial identification of DMCs will typically use a relaxed statistical significance threshold.
-For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
-This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
-
-These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
-Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
-
-The challenges of valid statistical inference at such data-driven regions are not unique to the problem of testing for differentially methylated regions.
-Similar problems arise in the analysis of chromatin immunoprecipitation sequencing (ChIP-seq) experiments \citep{Schwartzman:2011to, Lun:2014gn} and in the field of signal processing \citep{Schwartzman:2011vg}.
-It boils down to the problem of how to assign statistical significance to regions or peaks using the same data that were used to define these regions or peaks.
-These regions are already enriched for differences, which biases the interpretation the results of tests for differences at these regions.
-One way, and I would argue the best way, to avoid this issue is to test these regions using a separate dataset, which completely avoids the issue of statistical 'double-dipping'.
-If the sample size is large enough, then permutation testing may also be appropriate.
-For example, \cite{Hansen:2013eo} permute the group labels of their samples and re-ran their analysis to determine for each of the observed regions "how often we see another block of similar length and effect size anywhere in the genome and in any of the permutations".
-The chief limitation of the permutation strategy is the restricted number of permutations that are possible from small sample sizes and the time and computational resources it takes to perform analyse the data for a single permutation, which are often substantial.
-For example, \cite{Hansen:2013eo} only performed nine permutations.
-
-A somewhat less _ad hoc_ strategy for identify differentially methylated regions is called "bump-hunting" \citep{Jaffe:2012gx}.
-Initially developed for methylation microarrays, and now available for broader use in the R/Bioconductor package `bumphunter` ([http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html](http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html)), bump-hunting is quite a general statistical problem.
-For the case of identifying DMRs, the idea is to plot the value of a statistic used to test for DMCs as a function of the position in the genome[^smoothing] and then to identify bumps in this plot.
-Bumps are identified as contiguous regions of the genome where the signal is above some threshold and may include an error term to account for the spatial correlation of the signal.
-The significance of these peaks may be assessed with a permutation strategy.
-
-Another alternative for combining individual loci into data-driven DMRs is based on the loci-specific P-values.
-These methods can be thought of extensions to Fisher's method for combining P-values \citep{Fisher:1936vv}, by attempting to account for the correlation of tests at nearby methylation loci.
-`comb-p` \citep{Pedersen:2012vl} uses the Stouffer-Liptak-Kechris \citep{Stouffer:1949ua, Kechris:2010fw, Zaykin:2011jq,} correction for spatially correlated P-values.
-`SLIM` \citep{Wang:2011cw} is used by `methylKit` \citep{Akalin:2012cm} to do a similar correction.
-
-Finally, there are a class of methods that turn the problem of identifying data-driven DMRs on its head by constructing the regions without first testing the individual loci for differential methylation.
-Then, only once these regions are defined, these methods test for differential methylation.
-This is different to using _a priori_ regions, since the regions are still data-defined, but not with respect to differential methylation.
-The only example I know of such a method is `Aclust` \citep{Sofer:2013bk}.
-`Aclust` first clusters CpGs into candidate regions by performing agglomerative nested clustering of the between-sample co-methylation.
-Briefly, this is the correlation of methylation levels at two loci across the samples[^between_sample_co-methylation].
-These clusters are then tested for differential methylation using generalised estimating equations.
-
-[^between_sample_co-methylation]: See Chapter \ref{co-methylation_review} for further details of between-sample co-methylation.
-
-\subsubsection{Consistency of DMRs}\label{sec:consistency_of_dmrs}
-
-While not often explicitly stated, it is desirable that a DMR is _consistent_.
-The definition of consistency is best illustrated by a pathological example describing a region that is inconsistent.
-Suppose we have a one-group experiment with five samples[^one_by_five] and are studying a region with five CpGs, all of which are DMCs.
-However, the first CpG is differentially methylated in the first sample, the second CpG is differentially methylated in the second sample, etc.
-While the entire region is 'differentially methylated', the difference is not observed consistently in any one sample.
-A consistent region is therefore one where the differences are 'in the same direction'.
-
-[^one_by_five]: Equivalently, a five group experiment without replicates.
-
-Consistent DMRs are most easily identified in two-group experiments, where we can use the sign of the test statistic to determine the 'direction' of the signal.
-This is exactly what is done by, for example, \cite{Hansen:2011gu}.
-
-In one-group experiments, the idea of consistency is similar to testing the homogeneity of odds ratios for all $n \times 2$ contigency tables in the region.
-However, __TODO.
-What about a region that is consistent but the odds ratios increases across the region__
-
-\section{Other downstream analyses}
-
-There are many other questions that can be addressed using bisulfite-sequencing data.
-Like differential methylation, some of these might be also be tackled with other methylation assays, such as microarrays.
-However, as we will see, some can only be answered by assays that have single nucleotide resolution, which is where bisulfite-sequencing can prove particularly powerful.
-
-\subsection{Partially methylated domains}
-
-Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$.
-An example of such a region is shown in __FIGURE__ (__TODO: Find a PMD and plot it__).
-
-PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}.
-Using a simple sliding window algorithm, \citeauthors{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb.
-They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample.
-This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.
-
-A subsequent study found that PMDs are a common feature of somatic cell lines and that they comprise $> 30\%$ of the genome \citep{Lister:2011kg}.
-Perhaps even more intriguingly, across four somatic cell lines profiled with whole-genome bisulfite-sequencing, \citeauthor{Lister:2011kg} found a large amount of these genomes ($664$ Mb) comprised shared PMDs.
-
-PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_).
-However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}.
-Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg}.
-
-More sophisticated methods have also been proposed to identify these PMDs.
-`methylSeekR` \citep{Burger:2013kq} is one such algorithm.
-It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions.
-While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that methylation microarrays cannot provide.
-
-\subsection{Variably methylated regions and differential variability}\label{sec:vmrs}
-
-- Belinda
-- Jaffe
-- DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
-- DVRs: Identify runs of DVCs
-
-\subsection{Allele-specific methylation}\label{sec:asm}
-
-\subsection{Methylation entropy}\label{sec:methylation_entropy}
-
-\subsection{Epipolymorphism}\label{sec:epipolymorphism}
-
-\subsection{Epiallele detection}\label{sec:epialleles}
-
-\subsection{Integrating DNA methylation data with other genomes data}\label{sec:data_integration}
+__TODO__
 

### 2014-12-04 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Fixed syntax errors

@@
 This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges.
+I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay.
+All data used in my thesis were generated using this protocol.
+Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.
+
 There are four fundamental steps in the analysis of whole-genome bisulfite-sequencing data:
@@
 There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis.
+This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:downstream_analyses}.
 
-This chapter focuses on steps 1-3 and concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data.
-The following chapter addresses step 4.
-
-In this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay.
-All data used in my thesis were generated using this protocol.
-Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.
+The chapter concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data.
 
@@
 # ADS
-files <- "../M-bias/ADS.M-bias.txt"
-sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
+files <- "../data/M-bias/ADS.M-bias.txt"
+sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
 mbias <- lapply(files, read.mbias)
@@
 # E18VA
-files <- "../M-bias/E18VA.M-bias.txt"
-sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
+files <- "../data/M-bias/E18VA.M-bias.txt"
+sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
 mbias <- lapply(files, read.mbias)
@@
 Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
-\texttt\{methclone\} & \cite\{Li:2014ei\}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf\{TODO: Try installing.
-Suspect you can't get actual methylation calls.\} \\
-\texttt\{methpat\}   & \url\{https://github.com/bjpop/methpat\} & Output of \texttt\{bismark\_methylation\_extractor\} & Yes (amplicons)                  &                                                                                  \\
-\texttt\{DMEAS\}     & \cite\{He:2013cj\}                       & Output of \texttt\{bismark\_methylation\_extractor\} &                                  & Windows operating system only.
+\texttt{methclone} & \cite{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing.
+Suspect you can't get actual methylation calls.} \\
+\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
+\texttt{DMEAS}     & \cite{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only.
 Perl code only available as PDF file.

### 2014-12-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Added methylation entropy

@@
 
-__UP TO HERE__
+There are several definitions and uses of _methylation entropy_ in the literature.
+\citet{Xie:2011cy} introduced the concept for bisulfite-sequencing data as a way of quantifying the variability of methylation patterns at an m-tuple.
+For a given 4-tuple, $(i, \ldots, i + 3)$, \citet{Xie:2011cy} analyse the set of reads aligned to it, $\{z: z \in \mathcal{R}_{(i, \ldots, i + 3} \}$, to compute the methylatione entropy.
+On the one hand, if we only observe a single unique methylation pattern then the m-tuple has the minimum methylation entropy of zero.
+On the other hand, if we observe all possible $2^{4} = 16$ methylation patterns then the m-tuple has the maximum methylation entropy of 1.
+Depending on the frequency of the patterns, we obtain intermediate methylation values ($0 < \text{methylation entropy} < 1)$.
+Of course, methylation entropy need not be limited to 4-tuples.
+The choice of the size of the tuples depends on the regions of interest, the read lengths and the sequencing coverage.
 
-- `QDMR` \citep{Zhang:2011dp} and `CpG_MPs` \citep{Su:2012hl}, developed by the same research group, are methods for identifying DMRs using a form of methylation entropy from MeDIP-chip and BS-seq, respectively.
-- \citet{Xie:2011cy} introduced the concept of methylation entropy for BS-seq.
+The natural interpretation of methylation entropy as a measure of 'disorder' allows us to quantify how heterogeneous DNA methylation is at a locus \citep{Xie:2011cy, He:2013cj}[^other_definitions].
+These methylation entropies can be analysed to identify heterogeneous regions of the genome or perhaps tested for an association with a phenotype.
 
-While mathematically based on the idea of Shannon entropy \citep{Shannon:IH1_srIe}, the interpretation of methylation entropy has much in common with the physical concept of "disordered/heterogeneity"
+[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles}.
+Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since this uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.
+
+While \citet{Xie:2011cy} do not provide software implementing the method, a subsequent paper from some of the same researchers introduces the `DMEAS` software to do so \cite{He:2013cj} .
+Unfortunately, `DMEAS` is only available as a Windows binary or as a Perl script that itself is only available as PDF file, greatly limiting its usefulness.
+
+While I have not yet done so, it is not difficult to add the calculation of methylation entropies to the `MethylationTuples` software.
+The input for such an analysis of methylation entropy is a list of m-tuples and the associated frequencies of each methylation pattern, exactly what `methtuple` produces.
+The formula for computing methylation entropies is a simple sum of ratios and log-ratios, easily implemented in R, which `MethylationTuples` is written in.
 
@@
 
+__UP TO HERE__
+
 \subsection{Epiallele detection}\label{sec:epialleles}
+
+- methclone
+- Conrad (MPFE)
+- How it differs to Emma Whitelaw's definition
 
@@
 
-A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are somewhat lacking.
-This in part motivated me to begin developing `MethylationTuples`, an R package for managing, analysing and visualising methylation patterns at m-tuples.
+A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are rather lacking.
+`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples.
+While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), `MethylationTuples`, in conjunction with `methtuple`, can also be used to do many of the above analyses.
 
-- I'm using to address co-methylation
+- I'm using it to address co-methylation
 - Could be used for entropy, epipolymorphism and epialleles.
-- Could implement Fang or Peng's method for ASM
+- Could implement Fang or Peng's method for ASM, but less readily the "heterozygous SNP" method

### 2014-12-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Much of the work on epialleles and epipolymorphism

@@
 Many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
-All of these methods are designed for two-group experiments, with some being general enough to extend to multiple-group experiments[^multi-group_extension].
-I describe in greater detail methods for identifying differential methylation in a one-group experiment, admitedly an unusual design but one that I faced during my PhD.
+
+During my PhD I needed to analyse data from one-group experiment for differential methylation, admittedly an unusual design.
+Almost all these existing methods are designed for two-group experiments, with some being general enough to extend to multiple-group experiments[^multi-group_extension].
+Since I spent time working on the one-group cases, and also because a review article for the two-group case was recently published \citep{Robinson:2014jz}, I pay somewhat disproportianate attention to the one-group scenario in what follows.
 
@@
 
-\subsection{Epipolymorphism}\label{sec:epipolymorphism}
+\subsection{Epialleles and epipolymorphism}\label{sec:epialleles_and_epipolymorphism}
 
-__UP TO HERE__
+A DNA sequence may have multiple epigenetic states.
+For example, the cytosine in the sequence `TCGA` may be methylated or unmethylated; each of the methylated and unmethylated versions of that sequence is an _epiallele_.
+Restricting our attention to CpG methylation, a sequence with $m$ CpGs has $2^{m}$ potential epialleles.
+Therefore, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical[^epiallele_definition].
+An epiallele may also be described as an _epimutation_, if it sufficiently rare.
 
-\subsection{Epiallele detection}\label{sec:epialleles}
+[^epiallele_definition]\cite{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
+While most would agree that is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive.
+After all, we refer to alternate forms of a genetic sequence as alleles regardless of whether we know of a phenotypic consequence of the variant.
 
-- methclone
-- Conrad (MPFE)
-- How it differs to Emma Whitelaw's definition
+Epipolymorphism, analogous to genetic polymorphism, is the presence of multiple epialleles in a population.
+However, unlike genetic polymorphism, where the population is typically a set of chromosomes from multiple individuals, the population here is often within an individual, even within a tissue within an individual.
+The 'epipolymorphism' of a locus has also been defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}.
+
+\subsubsection{Examples}
+
+Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom.
+Even then, the number of such epialleles is small --- a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.
+
+In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \cite{Rakyan:2002wz}.
+The poster child for the potential important of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}.
+Mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic and have an increased suscpetibility to tumours compared to those mice with a methylated version of the allele, which have a _pseudoagouti[^pseudoagouti_vs_agouti]_ (brown) coat and none of the associated health defects.
+
+[^pseudoagouti_vs_agouti]: These mice are properly described as _pseudoagouti_ rather than agouti.
+They are heterozygous for the wildtype agouti allele ($A^{vy}/a$) but are phenotypically indistinguishable from true agouti mice, which are homozygous for the wildtype gene ($a/a$).
+
+In humans, there have been several interesting studies using epialleles to infer the clonality and evolution of cancer \citep{Siegmund:2009kk, Li:2014ei}, as well as the evolution of methylation dynamics and the epipolymorphism of various loci in an immortalised cell line \citep{Landan:2012kp}.
+
+Finally, although there is much interest, if not outright hype, surrounding transgenerational epialleles, these are difficult to prove outside of very controlled experimental conditions \cite{Daxinger:2012kl}.
+
+Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing.
+For example, __FIGURE__ shows the number of methylation patterns observed at 4-tuples with a frequency greater than __WHAT__ from the __WHICH__ sample.
+There has been some progress in developing software to extract and analyse these epialleles, which I now describe.
+
+\subsubsection{Methods}
+
+The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error.
+Another difficulty, perhaps unavoidable with the present technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele.
+
+__UP TO HERE: Read the `methclone` paper and say somethign intelligent about it. Then, finish the Methods section and move onto the `MethylationTuples` section.__
+
+`methclone` \citep{Li:2014ei}
+
+- `methclone`
+- `MPFE`
+- `methpat`
+- ASM-methods are arguably epiallele detection methods
+- Methods based solely on methylation patterns don't check that the sequence is identical.
 
@@
 
-A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are rather lacking.
+A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are rather lacking or, at the least, very much tailored toward a particular analysis.
+What has been lacking is a general method for extracting methylation patterns at m-tuples and software for manipulating, analysing and visualising the results.
+I have made significant progress towards the first goal with `methtuple` (see Section \ref{sec:methtuple}) and now introduce `MethylationTuples` ([https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples)) to address the second missing link.
+
 `MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples.
-While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), `MethylationTuples`, in conjunction with `methtuple`, can also be used to do many of the above analyses.
+While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are also well suited to analysing methylation entropy, allele-specific methylation, epialleles and epipolymorphism
 
-- I'm using it to address co-methylation
-- Could be used for entropy, epipolymorphism and epialleles.
-- Could implement Fang or Peng's method for ASM, but less readily the "heterozygous SNP" method
+- Could implement Fang or Peng's method for ASM, but less readily the "heterozygous SNP" method (could have a different `sample` for each parental chromosome)
+
+__TODO: Things to mention__
+- `GenomicTuples`
+  - Package build on existing class-leading infrastructure
+- Describe utility functions, e.g., `collapseStrand`
+- Analytical methods currently implemented in the package
+- Compatability with other packages, e.g., `bsseq`.
+- Challenges
+- Future directions

### 2014-12-05 Rmarkdown/wgbs_downstream_analyses.Rmd
Described methods for epialleles. Started section on MethylationTuples

@@
 The 'epipolymorphism' of a locus has also been defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}.
+\citeauthor{Landan:2012kp} the epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} (f_{p}^{2})$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
+
+[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism under a model that assumes sampling with replacement or, equivalently, an infinite population size.
+While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.
 
@@
 
+`methclone` \citep{Li:2014ei} is software to estimate the frequency of epialleles at m-tuples and to identify "shifts" in these distributions between two samples.
+`methclone` is limited to pairwise comparisons.
+The input is a pair of `BAM` files created with the `Bismark` aligner, one for each sample.
+__TODO: Can you get output files with the frequency of each methylation pattern at each m-tuple?__ For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($\60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples.
+The "foreground" combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples while the "background" combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two stages [samples]".
+The difference in these combinatorial entropies, $\delta S = S - \tilde{S}$, a kind of observed-to-expected log-ratio, is used to identify shifts in the epiallele distribution between a pair of samples.
+A $\delta S = 0$ corresponds to no change and a $\delta S = -144$ corresponds to maximal difference in entropy.
+It isn't clear whether the range of $\delta S$ depends on the size of the m-tuples nor is it clear how to choose the threshold at which to declare a significant shift in the distribution of epialleles.
+The example in the paper shows results for a threshold of $-60, -70$ and $-80$, but these choices appear somewhat arbitrary.
+
 The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error.
 Another difficulty, perhaps unavoidable with the present technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele.
+`methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies, which does not attempt to account for these potential sources of bias.
+In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.
 
-__UP TO HERE: Read the `methclone` paper and say somethign intelligent about it. Then, finish the Methods section and move onto the `MethylationTuples` section.__
+`MPFE` is designed to estimate the frequency of epialleles by maximising a multinomial likelihood that includes error terms for both incomplete bisulfite-conversion and sequencing error.
+The maximisation of this likelihood is computatationally demanding, as evidenced by the need for a "fast" algorithm that approximates the likelihood.
+`MPFE` is designed for amplicon bisulfite-sequencing and may not scale to whole-genome data.
+The input is a file containing the number of times each methylation pattern was observed at that m-tuple.
+Unfortunately, `MPFE` does not provide a way to create this table.
 
-`methclone` \citep{Li:2014ei}
-
-- `methclone`
-- `MPFE`
-- `methpat`
-- ASM-methods are arguably epiallele detection methods
-- Methods based solely on methylation patterns don't check that the sequence is identical.
+Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,]{Fang:2012ciPeng:2012dh}, might also adapted to identify epialleles and their associated frequencies.
+It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which strictly speaking is a requirement for the m-tuple to be an epipolymorphic locus.
 
@@
 
-A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are rather lacking or, at the least, very much tailored toward a particular analysis.
-What has been lacking is a general method for extracting methylation patterns at m-tuples and software for manipulating, analysing and visualising the results.
+A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are often lacking or very much tailored toward a particular analysis.
+What has been lacking is:
+
+1. A general method for extracting methylation patterns at m-tuples,
+2. Software for manipulating, analysing and visualising these methylatoin patterns.
+
 I have made significant progress towards the first goal with `methtuple` (see Section \ref{sec:methtuple}) and now introduce `MethylationTuples` ([https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples)) to address the second missing link.
 
-`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples.
-While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are also well suited to analysing methylation entropy, allele-specific methylation, epialleles and epipolymorphism
+\subsection{Design}
 
+`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples released under a Artistic-2.0 licence.
+While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are also well-suited to analysing methylation entropy, allele-specific methylation, epialleles and epipolymorphism.
+
+`MethylationTuples` is written to work with the Bioconductor project \citep{Gentleman:2004ju}[^bioc].
+Bioconductor makes extensive use of R's S4 object system and encourages developers to re-use existing Bioconductor infrastructure.
+From a developer's perspective, this avoids the need to re-invent the wheel.
+And from the user's perspective, it helps avoid multiple versions of the 'wheel', each that acts slightly differently and that may not be as well-tested.
+
+[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to [https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples).
+
+Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \cite{Lawrence:2013hi, Lawrence:2014gy}.
+However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges].
+Therefore, I wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)).
+In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations.
+For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., `type = 'equal'`).
+Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.
+
+[^tuples_vs_ranges]: The difference between a genomic range and a genomic tuple  can be thought of as the difference between an interval and a set.
+Namely, an interval includes the co-ordinates in between the start and end whereas a set only includes those co-ordinates listed in the set.
+For example, the genomic interval `chr3:+:[10, 12]` includes the co-ordinates `chr3:10`, `chr3:11` and `chr3:12` on the forward strand, whereas the genomic 2-tuple `chr3:+:{10, 12}` only includes the co-ordinates `chr3:10` and `chr3:12` on the forward strand.
+
+In the `MethylationTuples` package, I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated frequencies of the methylation patterns.
+The `MethPat` class extends the `SummarizedExperiment` class, also defined in the `GenomicRanges` package, but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples.
+A `MethPat` object is as a matrix-like object, where rows represent m-tuples of interest and columns represent samples.
+Data from multiple samples can be stored in a single `MethPat` object.
+Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size ($\text{m}$).
+__FIGURE__ is an schematic of the `MethPat` class.
+
+\subsection{Available methods}
+
+- Describe utility functions, e.g., `collapseStrand`
+- Analytical methods currently implemented in the package
 - Could implement Fang or Peng's method for ASM, but less readily the "heterozygous SNP" method (could have a different `sample` for each parental chromosome)
 
-__TODO: Things to mention__
-- `GenomicTuples`
-  - Package build on existing class-leading infrastructure
-- Describe utility functions, e.g., `collapseStrand`
-- Analytical methods currently implemented in the package
-- Compatability with other packages, e.g., `bsseq`.
-- Challenges
-- Future directions
+\subsection{Compatability with other Bioconductor packages}
+
+- e.g., `bsseq`, `GenomicRanges`, `BiSeq`, `methyVisual`
+
+\subsection{Computational challenges}
+
+\subsection{Future directions}

### 2014-12-05 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Added example of methtuple output

@@
 
-Most methylation callers, such as `Bismark` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
+Most methylation callers, such as `bismark_methylation_extractor` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
 The output file is a table, where each row records the co-ordinates of a cytosine and the number of methylated and unmethylated reads at that position.
@@
 
-\subsection{Methods}
+\subsection{Methods}\label{sec:methtuple_methods}
 
@@
 Since all my data are aligned with Bismark, or were converted to Bismark's `BAM` format, I have not yet had need to pursue this line of work.
+
+The output format of `methtuple` is tab-delimited plain text, optionally compressed with `gzip` or `bzip2`. An example of the output is shown below for 3-tuples:
+
+```
+chr     strand  pos1    pos2    pos3    MMM     MMU     MUM     MUU     UMM     UMU     UUM     UUU
+chr1    +       469     471     484     0       0       4       1       1       0       0       0
+chr1    +       471     484     489     1       0       0       0       2       2       1       0
+
+```
+
+__TODO: Show example of `methtuple` output__
 

### 2014-12-05 Rmarkdown/wgbs_downstream_analyses.Rmd
Methods and challenges of MethylationTuples

@@
     fig_caption: yes
-    keep_tex: yes
     includes:
@@
 The input is a pair of `BAM` files created with the `Bismark` aligner, one for each sample.
-__TODO: Can you get output files with the frequency of each methylation pattern at each m-tuple?__ For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($\60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples.
+__TODO: Can you get output files with the frequency of each methylation pattern at each m-tuple?__ For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples.
 The "foreground" combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples while the "background" combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two stages [samples]".
@@
 `MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples released under a Artistic-2.0 licence.
+R is a very popular language for analysing data, particularly in bioinformatics, and is my computational mother tongue, hence my choice of R as the language.
+To improve the performance of the package are written in `C++`, making use of the wonderful `Rcpp` package \cite{Eddelbuettel:2011to,Eddelbuettel:2013if},
+
 While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are also well-suited to analysing methylation entropy, allele-specific methylation, epialleles and epipolymorphism.
@@
 A `MethPat` object is as a matrix-like object, where rows represent m-tuples of interest and columns represent samples.
-Data from multiple samples can be stored in a single `MethPat` object.
 Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size ($\text{m}$).
-__FIGURE__ is an schematic of the `MethPat` class.
+Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing data on $n$ samples for 3-tuples.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/MethPat.pdf}
+\caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples.
+Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the frequencies of the methylation patterns (grey box) are aligned.
+The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, $S_{n}$.
+Some sample may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
+\label{fig:MethPat}
+\end{figure}
 
@@
 
-- Describe utility functions, e.g., `collapseStrand`
-- Analytical methods currently implemented in the package
-- Could implement Fang or Peng's method for ASM, but less readily the "heterozygous SNP" method (could have a different `sample` for each parental chromosome)
+The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples.
+A `MethPat` object can be constructed using the `MethPat` constructor function or from the output files of `methtuple` via the `readMethtuple` function.
+
+[^methtuple_output]: See Section \ref{sec:methtuple_methods} for example of the `methtuple` output format.
+
+The `MethPat` object provides fast subsetting by rows (m-tuples), including subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps` methods.
+Several other useful utility functions are (__TODO: Update list of implemented methods__):
+
+- `collapseStrand`: Collapse strand-specific data by aggregating the counts. Only applicable to CpG methylation loci.
+- `combine`: Combine multiple `MethPat` objects into one.
+- `filterOutVariants`: Remove m-tuples that contain a known variant. Variants must be provided as a VCF.
+- `getCoverage`: Compute the sequencing coverage of the m-tuples in each sample.
+- `IPD`: Compute the IPD vector for each m-tuple.
+- `methLevel`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
+- `plotBetaDensity`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
+- `tuples`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.
+
+These are in addition to the many useful utility functions inherited from the `SummarizedExperiment` class.
+
+In the `MethPat` class, its associated methods and other utility functions, the `MethylationTuples` package provides a toolbox for manipulating methylation patterns at m-tuples.
+Aside from providing the necessary infrastructure to analyse methylation patterns at m-tuples, `MethylationTuples` currently includes specific methods to analyse and visualise co-methylation (Chapter \ref{chap:co-methylation}) with the `cometh` and `methLevelCor` functions.
+I also plan to add methods for estimating epialleles and epipolymorphism.
+And it is my hope that `MethylationTuples` will provide a useful foundation on which others can implement their own methods for analysing methylation patterns at m-tuples.
 
@@
 
-- e.g., `bsseq`, `GenomicRanges`, `BiSeq`, `methyVisual`
+Being based on core Bioconductor functionality, `MethylationTuples` is highly compatible with existing packages.
+In particular, `MethPat` objects containing 1-tuples are readily coerced for use in differential methylation calling packages, e.g., `bsseq` and `BiSeq`, or to identify partially methylated domains with `MethylSeekR`.
+I also make extensive use of `MethylationTuples` in my `methsim` software (Chapter \ref{chap:methsim})
 
-\subsection{Computational challenges}
+\subsection{Computational challenges and future directions}
 
-\subsection{Future directions}
+The challenges of working with large datasets in R are well-known (__CITE__).
+These are in large part due to R being designed as an 'in-memory' application (__CITE__).
+More generally with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity - a more efficient way of storing the data may be less convenient to work with and _vice versa_.
+
+I have used the `MethylationTuples` package to analyses datasets containing up to 19 whole-genome bisulfite-sequencing samples (the Ziller data), albeit using a server with $512$ Gb of (shared) memory.
+I have found the `MethPat` class to be a very convenient representation of the data, but one that does not coincide with the most efficient storage.
+This is particularly true as the size of the m-tuples increases and the data become sparser.
+However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.
+Shown in __TABLE__ is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
+
+The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases.
+It may be possible to improve this by using a different approach to the internal storage of the data.
+For example, it may be possible to create an index so that the matrices storing the frequencies of methylation patterns can be re-ordered and stored as run-length encodings, a very space-efficient storage scheme.
+
+In the short term, my aim for `MethylationTuples` is to release a version to Bioconductor with the core infrastructure for manipulating data of methylation patterns at m-tuples.
+In the longer term, I would like to extend the set of downstream analyses of m-tuples that are available in the package and for other users to do the same with their own methods.

### 2014-12-06 Rmarkdown/appendix.Rmd
Minor changes to appendix

@@
 
-```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution"}
+```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution", dev = "pdf"}
 library(ggplot2)
@@
 
-```{r cpg_ipds_cdf}
+```{r cpg_ipds_cdf, dev = "pdf"}
 library(ggplot2)
@@
 
-### Histogram
+### Frequency polygon
 
-```{r cpg_ipds_hist}
+```{r cpg_ipds_hist, dev = "pdf"}
 library(ggplot2)
@@
 
-The computational equivalence of these approaches are shown in the following R code:
+The computational equivalence of using Pearson's $\chi^{2}$ test, the deviance of a log-linear model and the deviance of a logistic regression are shown in the following R code:
 

### 2014-12-06 Rmarkdown/biology_background.Rmd
whitespace

--- 2014-11-26 Rmarkdown/biology_background.Rmd
@@
 
-The material in this chapter can be found in any good introductory molecular biology text, such as \cite{Alberts:2007vj}.
+The material in this chapter can be found in any good introductory molecular biology text, such as \citet{Alberts:2007vj}.
 
-[^eukaryotes]: Eukaryotes are organisms composed of one or more cells with a distinct nucleus and cytoplasm. Most of a eukaryotic cell's DNA is contained in the nucleus \cite{Alberts:2007vj}.
+[^eukaryotes]: Eukaryotes are organisms composed of one or more cells with a distinct nucleus and cytoplasm. Most of a eukaryotic cell's DNA is contained in the nucleus \citep{Alberts:2007vj}.
 
@@
 For instance, there are untranscribed _regulatory sequences_ of DNA that determine whether a nearby gene is transcribed.
-There is also _junk DNA_ that is of little consequence to the organism \cite{Alberts:2007vj}.
+There is also _junk DNA_ that is of little consequence to the organism \citep{Alberts:2007vj}.
 Conversely, not all transcribed DNA is a gene.
@@
 Each of these histones has a 'tail' consisting of a string of amino acids.
-These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \cite{Alberts:2007vj} (discussed in Section \ref{sec:epigenetics} on epigenetics).
+These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \citep{Alberts:2007vj} and are discussed in Section \ref{sec:epigenetics} on epigenetics.
 
 The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin.
-The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \cite{Annunziato:2008ue}.
+The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
 
@@
 In one such popular analogy, the histones are the 'spool' around which the DNA 'thread' is wrapped to form nucleosomes.
-The chromatin then has the appearance of "beads on a string" when viewed under an electron microscope \cite{Alberts:2007vj}.
+The chromatin then has the appearance of "beads on a string" when viewed under an electron microscope \citep{Alberts:2007vj}.
 Chromatin is further coiled up into various literally-named structures, such as the 30-nanometre and 250-nanometre fibres, and, ultimately, packaged into _chromosomes_.
@@
 
-A human _reference genome_ was jointly completed by the International Human Genome Consortium and Celera Genomics in 2003 \cite{Venter:2001fq,Lander:2001hk}.
-This reference genome does not represent the genome of any one human since it uses DNA donated by several different people \cite{Venter:2001fq,Lander:2001hk}.
+A human _reference genome_ was jointly completed by the International Human Genome Consortium and Celera Genomics in 2003 \citep{Venter:2001fq,Lander:2001hk}.
+This reference genome does not represent the genome of any one human since it uses DNA donated by several different people \citep{Venter:2001fq,Lander:2001hk}.
 Rather, a reference genome is a kind of map or scaffold that can be used to identify similarities and differences between individual genomes.
 
-The human genome has obvious uses in medical research and biotechnology, but is also used to learn about evolution and human history, such as human migration patterns \cite{Hellenthal:2014ky}.
-Every person, even a _monozygotic_ ('identical') twin \cite{Hellenthal:2014ky}, has their own unique genome.
+The human genome has obvious uses in medical research and biotechnology, but is also used to learn about evolution and human history, such as human migration patterns \citep{Hellenthal:2014ky}.
+Every person, even a _monozygotic_ ('identical') twin, has their own unique genome \citep{Bruder:2008kd}.
 However, genomes of any two randomly selected people are identical at approximately $99.9\%$ of sites.
-Furthermore, the vast majority, $98\%$ by some estimates \cite{Elgar:2008dm}, of the human genome is made up of non-coding DNA and upwards of $50\%$ is repetitive sequence \cite{Treangen:2012dn}.
+Furthermore, the vast majority, $98\%$ by some estimates \citep{Elgar:2008dm}, of the human genome is made up of non-coding DNA and upwards of $50\%$ is repetitive sequence \citep{Treangen:2012dn}.
 
@@
 The mouse genome is slightly smaller than the human genome, at 2.7 billion base pairs long (Golden Path Length [http://asia.ensembl.org/Mus_musculus/Location/Genome?r=1](http://asia.ensembl.org/Mus_musculus/Location/Genome?r=1)).
-Like the human genome, there is a mouse reference genome \cite{MouseGenomeSequencingConsortium:2002fn}.
+Like the human genome, there is a mouse reference genome \citep{MouseGenomeSequencingConsortium:2002fn}.
 It is based on several female mice from the _C57BL/6J_ strain, an important strain of mouse widely used in medical research.
@@
 _Epigenetics_ is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ...
-it means just what I choose it to mean - neither more nor less" \cite{carroll1897through}.
+it means just what I choose it to mean - neither more nor less" \citep{carroll1897through}.
 I am sure to continue in this grand tradition.
 
-Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" \cite{Waddington:2012bu}.
-Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \cite{Waddington:2012bu}.
+Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" \citep{Waddington:2012bu}.
+Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \citep{Waddington:2012bu}.
 
-A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs - epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \cite[p1]{russo1996epigenetic}.
+A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs - epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \citep[p1]{russo1996epigenetic}.
 The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence.
@@
 
-More recently, the definition of epigenetics has taken on a "more biochemical flavour" \cite{Daxinger:2010bu} to include marks whose heritability is yet to be established.
-The heritability, or lack thereof, of histone modifications means that many epigeneticists do not consider these to be truly epigenetic \cite{Berger:2009ex}, and describing them as such is a sure-fire way to annoy a good percentage of your audience \cite{Ledford:2008ew}.
+More recently, the definition of epigenetics has taken on a "more biochemical flavour" \citep{Daxinger:2010bu} to include marks whose heritability is yet to be established.
+The heritability, or lack thereof, of histone modifications means that many epigeneticists do not consider these to be truly epigenetic \citep{Berger:2009ex}, and describing them as such is a sure-fire way to annoy a good percentage of your audience \citep{Ledford:2008ew}.
 
-Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following \cite{Bird:2007em}:
+Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following \citep{Bird:2007em}:
 
@@
 _Epigenomics_ is the study of the epigenome, analogous to genomics being the study of the genome.
-However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is evidence that the epigenetic variation is associated with genetic variation \cite{McVicker:2013ih,Bell:2011fm,Zhang:2010bg,vanEijk:2012hc}
+However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is evidence that the epigenetic variation is associated with genetic variation \citep{Zhang:2010bg, Bell:2011fm, vanEijk:2012hc, McVicker:2013ih}
 
@@
 It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying].
-It therefore fits into Arthur Riggs' afore-mentioned definition of an epigenetic modification \cite[p1]{russo1996epigenetic}.
+It therefore fits into Arthur Riggs' afore-mentioned definition of an epigenetic modification \citep[p1]{russo1996epigenetic}.
 DNA methylation is found in bacteria, fungi, plants and animals.
@@
 
-A major reason that DNA methylation is studied is that it is essential for normal development in mammals \cite{Li:1992bp}.
-It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \cite{Razin:1991vj}, silencing of transposable elements\cite{Jones:2001eg}, X chromosome inactivation \cite{Mohandas:1981uf} and tumorigenesis \cite{Ehrlich:2002kz}.
+A major reason that DNA methylation is studied is that it is essential for normal development in mammals \citep{Li:1992bp}.
+It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \citep{Razin:1991vj}, silencing of transposable elements\citep{Jones:2001eg}, X chromosome inactivation \citep{Mohandas:1981uf} and tumorigenesis \citep{Ehrlich:2002kz}.
 
-However, DNA methylation is not found in all organisms, for example, it not detectable in yeast \cite{Capuano:2014cu}, and so is not essential for life.
-Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation\cite{Capuano:2014cu}.
+However, DNA methylation is not found in all organisms, for example, it not detectable in yeast \citep{Capuano:2014cu}, and so is not essential for life.
+Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation \citep{Capuano:2014cu}.
 
@@
 [^other_DNAm]: Two additional examples of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C).
-N6-methyladenine is a methylated form of adenine, which is found in mRNA \cite[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \cite{Ratel:2006dv}.
-N4-methylcytosine has also been detected in bacterial DNA \cite[e.g.,]{Ehrlich:1985bv, Ratel:2006dv}.
+N6-methyladenine is a methylated form of adenine, which is found in mRNA \citep[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \citep{Ratel:2006dv}.
+N4-methylcytosine has also been detected in bacterial DNA \citep[e.g.,]{Ehrlich:1985bv, Ratel:2006dv}.
 
@@
 Ruppel, first identified a methylated nucleic acid in 1898.
-Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide \cite{Ruppel:1899hc}.
-In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}.
-However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{VISCHER:1949ty}.
+Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide \citep{Ruppel:1899hc}.
+In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \citep{Johnson:1925js}.
+However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \citep{VISCHER:1949ty}.
 
@@
 
-In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated _5-methylcytosine_ from nucleic acid prepared from cow thymus \cite{HOTCHKISS:1948va}.
+In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated _5-methylcytosine_ from nucleic acid prepared from cow thymus \citep{HOTCHKISS:1948va}.
 Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.
@@
 Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other.
-While the former is more common, examples of the latter case, such as _allele-specific methylation_ \cite{Shoemaker:2010jz} and _genomic imprinting_ \cite{Li:1993ea}, are of considerable biological interest.
+While the former is more common, examples of the latter case, such as _allele-specific methylation_ \citep{Shoemaker:2010jz} and _genomic imprinting_ \citep{Li:1993ea}, are of considerable biological interest.
 
@@
 
-The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{Lister:2009boa}.
+The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \citep{Lister:2009boa}.
 In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_.
@@
 A CpG is a palindromic sequence and is generally symmetrically methylated.
-Approximately 70% of CpGs are methylated in mammals \cite{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.
+Approximately 70% of CpGs are methylated in mammals \citep{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.
 
@@
 
-One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{Scarano:1967ue}.
+One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}.
 Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinuclotide_distribution}).
@@
 
-__FIGURE__ plots a frequency polygon of the observed IPDs. From this we can see there is a cluster of CpGs with $IPD < 10$, which largely correspond to what are called _CpG islands_ (CGIs).
-
+__FIGURE OBSERVED VS EXPECTED__ also shows that there is a cluster of CpGs with $IPD < 10$.
+These largely correspond to CpGs that lie within what are called _CpG islands_ (CGIs).
 Within these CpG islands are the 20-40% of CpGs that are frequently unmethylated in mammalian genomes.
-CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \cite{Wu:2010do}.
-The classical definition of a CpG island, given by \cite{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:
+CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}.
+The classical definition of a CpG island, given by \citep{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:
 
@@
 
-This definition was refined by \cite{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands.
-More recently, \cite{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
+This definition was refined by \citet{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands.
+More recently, \citet{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
+I use this definition in my thesis.
 
-An alternative definition, and one that pre-dates the definition of \cite{GardinerGarden:1987we}, is based on the identification of unmethylated regions of the genome, which are typically CpG-dense.
-Such regions were previously called HpaII tiny fragment islands, or _HTF islands_, and named after the restriction enzyme used to identify them \cite{Cooper:1983ev, Bird:1985gj}.
+An alternative definition, and one that pre-dates the definition of \citet{GardinerGarden:1987we}, is based on the identification of unmethylated regions of the genome, which are typically CpG-dense.
+Such regions were previously called HpaII tiny fragment islands, or _HTF islands_, and named after the restriction enzyme used to identify them \citep{Cooper:1983ev, Bird:1985gj}.
 
@@
 
-- CpG island shores are regions within 2 kb of CpG islands \cite{Irizarry:2008hg}.
-CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
+- CpG island shores are regions within 2 kb of CpG islands and have been demonstrated to have an increased variability of CpG methylation \citep{Irizarry:2008hg}.
 Shelves have a comparatively low density of CpGs.
-- CpG island shelves are defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
-- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \cite{Sandoval:2011dj}.
+- CpG island shelves are defined as regions within 2 kb of a CpG island shore \citep{Bibikova:2011fw}.
+- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \citep{Sandoval:2011dj}.
 
-Other metaphors, these based on methylation levels rather than CpG density, include methylation deserts \cite{Li:2012gg} and CpG canyons \cite{Jeong:2014bla}.
+Other metaphors, these based on methylation levels rather than CpG density, include methylation deserts \citep{Li:2012gg} and CpG canyons \citep{Jeong:2014bla}.
 Because these regions are defined with respect to methylation levels rather than DNA sequence, these are generally identified in a tissue-specific manner.
@@
 
-In humans, cytosine methylation in most cell types is found almost exclusively at CpGs \cite{Jones:2012jr}.
+In humans, cytosine methylation in most cell types is found almost exclusively at CpGs \citep{Jones:2012jr}.
 There are, however, certain cell types with widespread non-CpG methylation.
@@
 
-To give a few examples, \cite{Lister:2009hy} found that in a _fibroblast_ cell line that $99.98\%$ of cytosines that displayed statistically significant evidence of methylation occured at CpGs.
+To give a few examples, \citet{Lister:2009hy} found that in a _fibroblast_ cell line that $99.98\%$ of cytosines that displayed statistically significant evidence of methylation occured at CpGs.
 In contrast, in an embryonic stem cell line, they found that $24.5\%$ of cytosines that displayed statistically significant evidence of methylation occured in a non-CpG context.
@@
 A subsequent paper from the same group extended this result.
-\cite{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines .
+\citet{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines .
 Pluripotent cells includes embryonic stem cells along with induced pluripotent stem cell lines.
 
-An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \cite{Lister:2013et}, where they found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
+An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, where they found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
 
@@
 Non-CpG methylation is very common in other organisms, such as plants.
-For example, \cite{Lister:2008bh} found that in the widely-studied _Arabidopsis Thaliana_ that $45\%$ of cytosines that displayed statistically significant evidence of methylation are at CHG or CHH loci.
+For example, \citet{Lister:2008bh} found that in the widely-studied _Arabidopsis Thaliana_ that $45\%$ of cytosines that displayed statistically significant evidence of methylation are at CHG or CHH loci.
 However, they also found that the level of methylation at non-CpG loci is typically lower than that observed at CpG dinucleotides.
@@
 These additional modifications may be thought of as 'modifications of a modification' since they extend 5-methylcytosine.
-Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \cite{Plongthongkum:2014jg}.
+Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \citep{Plongthongkum:2014jg}.
 The biological significance of these marks is still being determined, in part because the assays for studying these are still in development and because their relatively scarcity in the genome means that experiments to detect these modifications are more difficult and expensive.
 
-One genome-wide study of 5hmC found that less than $1\%$ of all assayed cytosines in mouse fetal cortex and adult cortex cells displayed any statistically significant evidence of hydroxymethylation \cite{Lister:2013et}.
+One genome-wide study of 5hmC found that less than $1\%$ of all assayed cytosines in mouse fetal cortex and adult cortex cells displayed any statistically significant evidence of hydroxymethylation \citep{Lister:2013et}.
 These cytosines appeared to be restricted to the CpG context and had only very low levels of 5hmC.
 
-\cite{Kriaucionis:2009bm} and \cite{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC.
+\citet{Kriaucionis:2009bm} and \citet{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC.
 This suggests a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.
@@
 
-A common analogy used in describing epigentic marks uses 'writers', 'readers' and 'erasers' \cite{Moore:2013in}.
+A common analogy used in describing epigentic marks uses 'writers', 'readers' and 'erasers' \citep{Moore:2013in}.
 In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
@@
 In mammals, _DNMT1_ is known as the maintenance DNA methyltransferase.
-_DNMT3a_ and _DNMT3b_ are known as the _de nove_ methyltransferases, although these are also required for the maintenance of DNA methylation \cite{Jones:2009ji}.
-Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \cite{Li:1992bp}.
-In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \cite{Li:1992bp}.
+_DNMT3a_ and _DNMT3b_ are known as the _de nove_ methyltransferases, although these are also required for the maintenance of DNA methylation \citep{Jones:2009ji}.
+Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \citep{Li:1992bp}.
+In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \citep{Li:1992bp}.
 
@@
 
-_DNMT2_, now known as _TRDMT1_, was once thought to be a DNA methyltransferase but was shown to in fact methylate a small RNA and not DNA \cite{Goll:2006kb}.
+_DNMT2_, now known as _TRDMT1_, was once thought to be a DNA methyltransferase but was shown to in fact methylate a small RNA and not DNA \citep{Goll:2006kb}.
 Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity.
-Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__TODO: cite__).
+Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b \citep{Jurkowska:2011dt}.
 
@@
 These readers can recruit additional proteins to the site of the methylated cytosine to perform a variety of functions related to gene expression.
-For example, the methyl-CpG-binding domain (_MBD_) group of proteins bind to DNA containing a methylated CpG, which can then suppress gene expression by preventing transcription factor binding at that site \cite{Nan:1993gma}.
-Another group, the ubiquitin-like, containing PHD and RING finger domain (_UHRF_) proteins, help _DNMT1_ methylate hemimethylated DNA, such as the daughter strand created during DNA replication \cite{Sharif:2007ft, Bostick:2007jh}.
+For example, the methyl-CpG-binding domain (_MBD_) group of proteins bind to DNA containing a methylated CpG, which can then suppress gene expression by preventing transcription factor binding at that site \citep{Nan:1993gma}.
+Another group, the ubiquitin-like, containing PHD and RING finger domain (_UHRF_) proteins, help _DNMT1_ methylate hemimethylated DNA, such as the daughter strand created during DNA replication \citep{Sharif:2007ft, Bostick:2007jh}.
 
@@
 This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions.
-This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of local tissue-specific differences in DNA methylation \cite{Irizarry:2009ky} nor the two stages of rapid global demethylation that occur during development \cite{Wu:2014gw}.
+This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of local tissue-specific differences in DNA methylation \citep{Irizarry:2009ky} nor the two stages of rapid global demethylation that occur during development \citep{Wu:2014gw}.
 
@@
 Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation.
-These mechanisms were recently reviewed by \cite{Wu:2014gw}, which I briefly summarise:
+These mechanisms were recently reviewed by \citet{Wu:2014gw}, which I briefly summarise:
 
@@
 
-A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as the _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays \cite{Laird:2010iv}.
+A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as the _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays \citep{Laird:2010iv}.
 Therefore, almost all assays of DNA methylation require one of the following _pre-treatments_ of the DNA:
@@
 An exception to this classification scheme are a new class of assay that seek to directly 'read' whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA.
-For example, both \cite{Laszlo:2013kf} and \cite{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine is methylated.
+For example, both \citet{Laszlo:2013kf} and \citet{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine is methylated.
 
@@
 Hundreds or thousands of cells are required in order to have sufficient material as input for the assay.
-Assays that require only a single cell as input do exist \cite[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.
+Assays that require only a single cell as input do exist \citep[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.
 
@@
 However, _HpaII_ will only digest DNA when the second cytosine in the motif is unmethylated.
-The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome \cite{Khulan:2006gx}.
+The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome \citep{Khulan:2006gx}.
 
 Assays based on restriction enzyme were some of the first developed for studying DNA methylation.
-They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \cite{Laird:2010iv}.
+They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \citep{Laird:2010iv}.
 Restriction enzyme assays have a relatively low resolution.
@@
 Furthermore, the bioinformatic analysis of data from these assays is complicated by the varying CpG density along the genome, which leads to different enrichment affinities for different regions of the genome.
-However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation \cite{Laird:2010iv}.
+However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation \citep{Laird:2010iv}.
+
+__TODO: Describe CHARM \citep{Irizarry:2008hg}, although not yet sure which sub-section it should appear in__.
 
@@
 
-In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \cite{Shapiro:1970im, Hayatsu:1970jv}.
+In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \citep{Shapiro:1970im, Hayatsu:1970jv}.
 The methylated cytosines are said to be 'protected' from conversion to uracils.
-This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq, Clark:1994wc}, which are referred to as bisulfite-conversion assays.
+This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \citep{Frommer:1992wq, Clark:1994wc}, which are referred to as bisulfite-conversion assays.
 
@@
 The relative methylation of each sequence can be inferred from the relative intensities of the 'methylated probe' to the 'unmethylated probe'.
-Such an idealised experiment brushes over many complications including \cite{Laird:2010iv}:
+Such an idealised experiment brushes over many complications including \citep{Laird:2010iv}:
 
@@
 The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k_) provides a modern implementation of this type of assay for studying human DNA.
-This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \cite{Stirzaker:2014cu}.
-There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed in \cite{Dedeurwaerder:2014bt}.
+This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \citep{Stirzaker:2014cu}.
+There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed by \citey{Dedeurwaerder:2014bt}.
 
 Genomics research was revolutionised by the development of cheap high-throughput sequencing technology, and the study of DNA methylation was no exception.
-In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology \cite{Cokus:2008fc, Lister:2008bh}.
+In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology \citep{Cokus:2008fc, Lister:2008bh}.
 Whole-genome bisulfite-sequencing is the gold-standard for analysing genome-wide DNA methylation data.
 
-\cite{Cokus:2008fc} termed their approach _BS-seq_ while \cite{Lister:2008bh} called theirs _methylC-seq_.
+\citet{Cokus:2008fc} termed their approach _BS-seq_ while \citet{Lister:2008bh} called theirs _methylC-seq_.
 From a bioinformatics perspective, the main difference is that the _BS-seq_ protocol produces sequencing reads from four bisulfite-converted DNA strands - the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) - which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results.
@@
 Some examples of these targeted bisulfite-sequencing are:
-- Reduced representation bisulfite-sequencing (_RRBS_, \cite{Meissner:2005eb}): RRBS uses restriction enzymes to first select regions of the genome with a high CpG density (based on the Msp-I cleavage motif), which are subsequently treated with sodium-bisulfite and sequenced.
+- Reduced representation bisulfite-sequencing (_RRBS_, \citet{Meissner:2005eb}): RRBS uses restriction enzymes to first select regions of the genome with a high CpG density (based on the Msp-I cleavage motif), which are subsequently treated with sodium-bisulfite and sequenced.
+- Extended reduced representation bisulfite-sequencing, also known as enhanced reduced representation bisulfite-sequencing (eRRBS, \citet{Akalin:2012he}). A modified version of the RRBS protocol.
 - NimbleGen's SeqCap Epi Enrichment System and Agilent's SureSelectxt Human Methyl-Seq: Both of these commercial products use DNA hybridisation to enrich the sequencing library for pre-defined regions of interest.
@@
 One final class of bisulfite-conversion assays does not use microarrays or high-throughput sequencing.
-Sequenom's EpiTYPER \cite{Ehrich:2006tt} uses mass spectrometry to analyse DNA methylation from bisulfite-converted DNA.
-This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and may be used to validate findings discovered using other platforms \cite{Laird:2010iv}.
+Sequenom's EpiTYPER uses mass spectrometry to analyse DNA methylation from bisulfite-converted DNA \citep{Ehrich:2006tt}.
+This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and may be used to validate findings discovered using other platforms \citep{Laird:2010iv}.
 
@@
 
-Bisulfite-conversion assays are considered the 'gold standard' for studying DNA methylation since cytosine methylation can be detected at single-base resolution \cite{Stirzaker:2014cu}.
+Bisulfite-conversion assays are considered the 'gold standard' for studying DNA methylation since cytosine methylation can be detected at single-base resolution \citep{Stirzaker:2014cu}.
 In fact, single-molecule, base pair resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].
@@
 However, the minimal amount of DNA is being reduced with each technological advance.
-Recently, single-cell bisulfite-sequencing assays have been reported \cite{Guo:2013ih, Smallwood:2014kn}, although they are not yet a 'commercial' sequencing assay.
+Recently, both \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although they are not yet a 'commercial' sequencing assay.
 Furthermore, these are not yet proper genome-wide assays.
-The technique of \cite{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
-And the technique of \cite{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \cite{Smallwood:2014kn}.
+The technique of \citet{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
+And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \citep{Smallwood:2014kn}.
 
@@
 For most experiments this isn't much of a problem - most cells have very low levels of 5hmC and so there is little confounding - however, in certain experiments this needs a more careful approach.
-To address this issue, \cite{Booth:2012fh} developed _oxidative bisulfite-sequencing_ (oxBS-seq) and \cite{Yu:2012ej} developed _Tet-assisted bisulfite sequencing_ (TAB-seq) for separate 5mC and 5hmC detection.
+To address this issue, \citet{Booth:2012fh} developed _oxidative bisulfite-sequencing_ (oxBS-seq) and \citet{Yu:2012ej} developed _Tet-assisted bisulfite sequencing_ (TAB-seq) for separate 5mC and 5hmC detection.
 
@@
 There is also potential confounding of 5C, 5mC and 5hmC levels with 5fC and 5caC, although this has received less attention since 5fC and 5caC are believed to exist at far lower quantities than 5mC and 5hmC.
-Nonetheless, sequencing assays of 5fC and 5caC exist based on the idea of treating the DNA with a chemical and performing 'signal subtraction' with traditional bisulfite-sequencing or another assay \cite{Wu:2014jj}.
+Nonetheless, sequencing assays of 5fC and 5caC exist based on the idea of treating the DNA with a chemical and performing 'signal subtraction' with traditional bisulfite-sequencing or another assay \citep{Wu:2014jj}.
 
@@
 This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome.
-When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \cite{Liu:2012ge}.
+When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \citep{Liu:2012ge}.
 Figure \ref{fig:bisulfite_CT_SNP.pdf} illustrates such an example.
@@
 Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_.
-_PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \cite{Warnecke:1997eh}.
+_PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \citep{Warnecke:1997eh}.
 _Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment.
 Finally, treating DNA with sodium bisulfite damages the DNA, resulting in the fragmentation of long molecules.
-This limits the size of the fragments that can be studied using bisulfite-conversion assays to $\approx 500$ bp \cite{Ecker:2010et}
+This limits the size of the fragments that can be studied using bisulfite-conversion assays to $\approx 500$ bp \citep{Ecker:2010et}
 
@@
 
-The only commercially available assay in this class is the Pacific Biosciences SMRT technology \cite{Flusberg:2010hra}.
+The only commercially available assay in this class is the Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}.
 It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.

### 2014-12-06 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Number of CpGs per read

@@
 
-Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \cite{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants.
+Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \citep{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants.
 Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.
@@
 Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.
-Several review articles have tried to summarise and compare the various approaches \cite{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
+Several review articles have tried to summarise and compare the various approaches \citep{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
 
@@
 
-While methylation-aware mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads \cite{Krueger:2012ks}.
+While methylation-aware mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads \citep{Krueger:2012ks}.
 This biases downstream inference and means that these mappers are generally less popular.
@@
 They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines.
-The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as `Bowtie1` \cite{Langmead:2009fv}, `Bowtie2` \cite{Langmead:2012jh} or `bwa`  \cite{Li:2009fi, Li:2010bl}[^non_directional].
+The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as `Bowtie1` \citep{Langmead:2009fv}, `Bowtie2` \citep{Langmead:2012jh} or `bwa`  \citep{Li:2009fi, Li:2010bl}[^non_directional].
 
@@
 _In silico_ bisulfite-conversion mappers avoid the bias inherent in the methylation-aware mappers because all reads, regardless of methylation status, 'look the same' to the mapper.
-However, they do suffer from a slight loss in mapping efficiency \cite{Krueger:2012ks}.
+However, they do suffer from a slight loss in mapping efficiency \citep{Krueger:2012ks}.
 
@@
 Name      & Reference                                            & Underlying mapping software \\ \midrule
-Bismark   & \cite{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
-bwa-meth  & \cite{Pedersen:2014wt}                             & bwa-mem                     \\
-BSMAP     & \cite{Xi:2009bd}                                   & SOAP                        \\
+Bismark   & \citet{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
+bwa-meth  & \citet{Pedersen:2014wt}                             & bwa-mem                     \\
+BSMAP     & \citet{Xi:2009bd}                                   & SOAP                        \\
 Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
@@
 
-Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \cite{Li:2009ka}.
+Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \citep{Li:2009ka}.
 However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way.
@@
 Technically, reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences.
-Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \cite{Flusberg:2010hra}, which produces longer reads.
+Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}, which produces longer reads.
 
@@
 However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful.
-`Bismark` \cite{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).
+`Bismark` \citep{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).
 
@@
 The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data.
-Repetitive sequences, such as LINEs and SINEs, are frequently methylated to prevent transcription and so are of interest to researchers studying DNA methylation.
-The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__TODO: CHECK IN MY DATA AND REPORT COVERAGE__).
+Repetitive sequences, such as LINEs and SINEs, are typically methylated in order to prevent transcription and are often of interest to researchers studying DNA methylation.
+The low mapping efficiencies of these regions means that there is often limited or less reliable data for these elements from bisulfite-sequencing data.
 
@@
 However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]).
-Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \cite{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
+Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
 These software use mapped reads from a `SAM/BAM` file as input.
@@
 The loss of the one-to-one relationship between sequencing cycle and read-position cannot be avoided if reads are pre-trimmed because the trimming information is not preserved.
-\cite{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming].
+\citet{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming].
 However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be 'sequencing cycle' rather than 'read-position', thus avoiding the issue entirely.
@@
 In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions.
-A minor problem with this approach is that the user may (unconsciously) use a different cutoff for each sample.
-For example, it is more difficult to determine an appropriate cutoff when there is a gradual slope in the M-bias plot than when there is a clear jump (__TODO: An example where there last few read-positions are more than `cutoff` from the median but due to a gradual slope vs.
-an example where a single position is more than `cutoff` from the median due to a spike__).
-This motivated me to include functionality in my `MethylationTuples` R package ([https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples))  to perform more systematic processing of M-bias results.
-These are implemented in the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.
+It is rather a subjective decision to make.
+Two problems that I found were in maintaining consistency across samples and in determining where to draw the line when there is no dramatic 'spike' in the M-bias.
+For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both read 1 and read 2, as well as a big problem at read-position 101 in read 2.
+What is less clear is where to draw the line on the gradual decay downwards toward the end of read 1 and towards the start of read 2.
+This motivated me to write a few simple functions to do perform more systematic processing of M-bias results.
+These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.
 
@@
 Biological biases include cellular heterogeneity and sequence variation at or nearby to methylation loci.
-I discuss the latter in __TODO: WHERE???__
 
@@
 This is only possible with bisulfite-data generated using the directional protocol.
-This figure is adapted from \cite{Liu:2012ge}.}
+This figure is adapted from \citet{Liu:2012ge}.}
 \label{fig:Bis-SNP}
@@
 
-__TODO: Size of m-tuples histogram__
+The number of CpGs per read for the Lister data is shown in Figure \ref{fig:Lister_cpgs_per_read}.
+Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples.
+This can be seen by comparing, for example, the _ADS_ and _HSF1_ samples.
+Samples sequenced more deeply will have more reads per m-tuple, although this can't be seen in these plots since they are normalised by sequencing depth.
+
+```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data."}
+library(dplyr)
+library(ggplot2)
+library(scales)
+thesis_theme <- theme_classic(base_size = 20)
+thesis_theme <- theme_bw(base_size = 20)
+dataset <- "Lister"
+hist <- lapply(paste0("../data/observable_m-tuples/", dataset, "/",
+                      list.files(paste0("../data/observable_m-tuples/",
+                                        dataset), "\\.CG\\_per\\_read\\.hist")),
+               read.table, header = TRUE)
+sn <- gsub("\\.CG\\_per\\_read\\.hist", "",
+           list.files(paste0("../data/observable_m-tuples/",
+                             dataset), "\\.CG\\_per\\_read\\.hist"))
+sn <- gsub("DM\\_", "", sn)
+hist_df <- bind_rows(hist) %>%
+  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
+  group_by(Sample) %>%
+  mutate(Proportion = count / sum(as.numeric(count)))
+hist_df %>%
+  ggplot(aes(x = n, y = Proportion * 100)) +
+  geom_bar(stat = "identity") +
+  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
+  coord_cartesian(xlim = c(-1, 8.5)) +
+  scale_y_continuous(label = comma) +
+  xlab("Number of CpGs per read\n(Axis truncated at 8)") +
+  ylab("Percentage of total reads") +
+  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
+  thesis_theme
+```
 
@@
 Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
-\texttt{methclone} & \cite{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing.
+\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing.
 Suspect you can't get actual methylation calls.} \\
 \texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
-\texttt{DMEAS}     & \cite{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only.
+\texttt{DMEAS}     & \citet{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only.
 Perl code only available as PDF file.
@@
 
-__TODO: Show example of `methtuple` output__
-
 \subsection{Performance}
@@
 Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets.
-For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see __APPENDIX__ for details of these machines).
+For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).
 

### 2014-12-06 Rmarkdown/wgbs_downstream_analyses.Rmd
MethPat memory usage

@@
 As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one.
-For example, \cite{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine).
+For example, \citet{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine).
 I contend that these are better defined as technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].
@@
 
-Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind, or pooled the replicates prior to analysis, thus ignoring all within-group variability, as is the case of the H1 and IMR90 cell lines in \cite{Lister:2009hy}.
+Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind, or pooled the replicates prior to analysis, thus ignoring all within-group variability, as is the case of the H1 and IMR90 cell lines in \citet{Lister:2009hy}.
 
@@
 Method                                       & Used by                                                                            \\ \midrule
-\cite{Benjamini:1995ws}                      & \cite{Akalin:2012cm, Jaffe:2012gx, Lacey:2013iy, Rijlaarsdam:2014hp, Sofer:2013bk} \\
-\cite{Wang:2011cw}                           & \cite{Akalin:2012cm}                                                               \\
-"False discovery rate correction/adjustment" & \cite{Dolzhenko:2014bo, Gokhman:2014cp, Lister:2009hy, Xie:2014ez}                 \\
-\cite{Storey:2007cp}                         & \cite{Jaffe:2012gx}                                                                \\
-\cite{Benjamini:2001ho}                      & \cite{Sofer:2013bk, Hebestreit:2013ko}                                             \\
-"Bonferonni adjustment"                      & \cite{Feng:2014iq}                                                                 \\ \bottomrule
+\citet{Benjamini:1995ws}                      & \citet{Akalin:2012cm, Jaffe:2012gx, Lacey:2013iy, Rijlaarsdam:2014hp, Sofer:2013bk} \\
+\citet{Wang:2011cw}                           & \citet{Akalin:2012cm}                                                               \\
+"False discovery rate correction/adjustment" & \citet{Dolzhenko:2014bo, Gokhman:2014cp, Lister:2009hy, Xie:2014ez}                 \\
+\citet{Storey:2007cp}                         & \citet{Jaffe:2012gx}                                                                \\
+\citet{Benjamini:2001ho}                      & \citet{Sofer:2013bk, Hebestreit:2013ko}                                             \\
+"Bonferonni adjustment"                      & \citet{Feng:2014iq}                                                                 \\ \bottomrule
 \end{tabular}
@@
 Unfortunately, some authors have still used Fisher's exact test \citep{Fisher:1922bx} to analyse such experiments.
-For example, \cite{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data the two replicates they have for each sample.
+For example, \citet{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data the two replicates they have for each sample.
 This is incorrect, because such an analysis effectievly ignores all within-group variability \citep{Hansen:2012gr}.
@@
 For a region containing $c$ methylation loci, construct the $n \times 2 \times c$ contingency table, i.e., each 'slice' of the table is a $n \times 2$ table containing the data for a single locus.
-Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions (__CITE: http://www.biostathandbook.com/cmh.html__).
+Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions \citep{Good:1987ce}.
 Instead, we could fit the log-linear model with `sample`, `meth_type` and `locus` as the set of factors.
@@
 The initial identification of DMCs will typically use a relaxed statistical significance threshold.
-For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
+For example, \citet{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
 This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
@@
 If the sample size is large enough, then permutation testing may also be appropriate.
-For example, \cite{Hansen:2013eo} permute the group labels of their samples and re-ran their analysis to determine for each of the observed regions "how often we see another block of similar length and effect size anywhere in the genome and in any of the permutations".
+For example, \citet{Hansen:2013eo} permute the group labels of their samples and re-ran their analysis to determine for each of the observed regions "how often we see another block of similar length and effect size anywhere in the genome and in any of the permutations".
 The chief limitation of the permutation strategy is the restricted number of permutations that are possible from small sample sizes and the time and computational resources it takes to perform analyse the data for a single permutation, which are often substantial.
-For example, \cite{Hansen:2013eo} only performed nine permutations.
+For example, \citet{Hansen:2013eo} only performed nine permutations.
 
@@
 Consistent DMRs are most easily identified in two-group experiments, where we can use the sign of the test statistic to determine the 'direction' of the signal.
-This is exactly what is done by, for example, \cite{Hansen:2011gu}.
+This is exactly what is done by, for example, \citet{Hansen:2011gu}.
 
-In one-group experiments, the idea of consistency is similar to testing the homogeneity of odds ratios for all $n \times 2$ contigency tables in the region.
-However, __TODO.
-What about a region that is consistent but the odds ratios increases across the region__
+In one-group experiments, the idea of consistency is similar, but not quite identical, to testing the homogeneity of odds ratios for all $n \times 2$ contingency tables in the region.
+For example, a region may be consistent but the strength of the effect may vary across the region, thus the odds ratios may vary.
 
@@
 
-While \citet{Xie:2011cy} do not provide software implementing the method, a subsequent paper from some of the same researchers introduces the `DMEAS` software to do so \cite{He:2013cj} .
+While \citet{Xie:2011cy} do not provide software implementing the method, a subsequent paper from some of the same researchers introduces the `DMEAS` software to do so \citep{He:2013cj} .
 Unfortunately, `DMEAS` is only available as a Windows binary or as a Perl script that itself is only available as PDF file, greatly limiting its usefulness.
@@
 Such an approach, while straightforward, is also obviously limited to a subset of the available methylation loci.
-Both \citet{Fang:2012ci} and \cite{Peng:2012dh} published methods to detect allele-specfic methylation that do not require nearby heterozygous genetic variants nearby to the methylation locus of interest.
+Both \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require nearby heterozygous genetic variants nearby to the methylation locus of interest.
 These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data.
-Unfortunately, there is no publicly available software implementing the method proposed by \cite{Peng:2012dh} and so I do not discuss it further.
+Unfortunately, there is no publicly available software implementing the method proposed by \citet{Peng:2012dh} and so I do not discuss it further.
 
@@
 
-[^epiallele_definition]\cite{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
+[^epiallele_definition]\citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
 While most would agree that is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive.
@@
 
-In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \cite{Rakyan:2002wz}.
+In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}.
 The poster child for the potential important of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}.
@@
 
-Finally, although there is much interest, if not outright hype, surrounding transgenerational epialleles, these are difficult to prove outside of very controlled experimental conditions \cite{Daxinger:2012kl}.
+Finally, although there is much interest, if not outright hype, surrounding transgenerational epialleles, these are difficult to prove outside of very controlled experimental conditions \citep{Daxinger:2012kl}.
 
@@
 R is a very popular language for analysing data, particularly in bioinformatics, and is my computational mother tongue, hence my choice of R as the language.
-To improve the performance of the package are written in `C++`, making use of the wonderful `Rcpp` package \cite{Eddelbuettel:2011to,Eddelbuettel:2013if},
+To improve the performance of the package, peaks are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
 
@@
 
-Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \cite{Lawrence:2013hi, Lawrence:2014gy}.
+Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}.
 However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges].
@@
 The `MethPat` object provides fast subsetting by rows (m-tuples), including subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps` methods.
-Several other useful utility functions are (__TODO: Update list of implemented methods__):
+Several other useful utility functions are:
 
@@
 - `methLevel`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
+- `MBias` and associated methods and functions: Helper functions to perform a more systematic analysis of M-bias (see Section \ref{sec:identifying_m-bias}).
 - `plotBetaDensity`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
@@
 
-The challenges of working with large datasets in R are well-known (__CITE__).
-These are in large part due to R being designed as an 'in-memory' application (__CITE__).
-More generally with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity - a more efficient way of storing the data may be less convenient to work with and _vice versa_.
+The challenges of working with large datasets in R are well-known.
+These are in large part due to R being designed as an 'in-memory' application and it's implementation of 'copy-on-modify' semantics \citep{wickham2014advanced}.
+More generally with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity --- a more efficient way of storing the data may be less convenient to work with and _vice versa_.
 
@@
 I have found the `MethPat` class to be a very convenient representation of the data, but one that does not coincide with the most efficient storage.
-This is particularly true as the size of the m-tuples increases and the data become sparser.
+This is more of a problem as the size of the m-tuples increases and the data become sparser.
 However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.
-Shown in __TABLE__ is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
+Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
+It also highlights the sparsity of the data as the size of the tuples grows --- most entries are $0$, meaning that particular methylation pattern was not observed in that particular sample, or `NA`, meaning that that m-tuple was not observed in that particular sample.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lllll@{}}
+\toprule
+& \texttt\{pryr::object\_size(x)\} & \texttt\{nrow\} & Number of assays & Percentage of \texttt\{NA\} and $0$ values \\ \midrule
+1-tuples                                 & $5.9$ GB                        & $56,348,522$    & $2$              & $28\%$                                     \\
+2-tuples                                 & $20.1$ GB                        & $100,586,237$   & $4$              & $80\%$                                     \\
+2-tuples (\texttt\{--all-combinations\}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
+3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
+4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             &                                            \\ \bottomrule
+\end{tabular}
+\caption{Size of \texttt{MethPat} objects for the \em{EPISCOPE} data.
+All m-tuples are stranded.
+The size of the \code{MethPat} object is computed using the \texttt{object_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
+The number of rows (\texttt{nrow}) corresponds to the number of m-tuples in the object.
+The number of assays is $2^{\text{m}}$, where $m$ is the size of the m-tuples.
+The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an `NA` value means that that m-tuple was not observed in that particular sample.}
+\label{tab:methpat_sparsity}
+\end{table}
 

### 2014-12-06 Rmarkdown/wgbs_statistical_framework.Rmd
replace \cite with \citet or \citep

@@
 I then extend this framework to multiple samples.
-Finally, I describe key variables, common estimators of these and their statistical properties by analysing __HOW MANY__ whole-genome bisulfite-sequencing samples.
+Finally, I describe key variables, common estimators of these and their statistical properties by analysing 51 whole-genome bisulfite-sequencing samples.
 
@@
 When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
-However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur (__FIGURE__):
+However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur:
 
@@
 
-One final definition is the average methylation of each read, which is used by some authors, including \cite{Landan:2012kp}.
+One final definition is the average methylation of each read, which is used by \citet{Landan:2012kp}.
 For each read, $z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, the average methylation of the read, $\zeta_{z}$, is defined as the proportion of methylation loci in the read that are methylated.
@@
 3.
-$\mathcal{I}^{Bis-SNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}.
+$\mathcal{I}^{Bis-SNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \citep{Liu:2012ge}.
 Identifies sample-specific methylation loci and removes reference-specific methylation loci.
@@
 Recently, more sophisticated methods have been proposed to estimate or model the average methylation level.
-These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
+These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
 
@@
 
-Several papers have proposed the beta-binomial distribution since, as noted by \cite{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" .
+Several papers have proposed the beta-binomial distribution since, as noted by \citet{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" .
 The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing.
@@
 This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data.
-For example, the `edgeR` software \cite{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.
+For example, the `edgeR` software \citep{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.
 
 An attractive feature of the beta-binomial distribution is that it can be motivated by, and analysed with, Bayesian methods, including empirical Bayes methods, or frequentist techniques such as maximum likelihood.
-For example, the software `DSS` \cite{Feng:2014iq} and `MOABS` \citet{Sun:2014fk} both use the beta-binomial distribution in an empirical Bayes analysis of differential methylation from bisulfite-sequencing data.
-In contrast, `RADmeth` \cite{Dolzhenko:2014bo} uses the beta-binomial model in a maximum likelihood framework to address the same problem.
+For example, the software `DSS` \citep{Feng:2014iq} and `MOABS` \citep{Sun:2014fk} both use the beta-binomial distribution in an empirical Bayes analysis of differential methylation from bisulfite-sequencing data.
+In contrast, `RADmeth` \citep{Dolzhenko:2014bo} uses the beta-binomial model in a maximum likelihood framework to address the same problem.
 
@@
 
-In papers about DNA methylation there is often a sentence like the following (given by \cite{Laird:2003kw}): "About 70% of the CpG dinucleotides in the mammalian genome are methylated".
+In papers about DNA methylation there is often a sentence like this quote from \citet{Laird:2003kw}: "About 70% of the CpG dinucleotides in the mammalian genome are methylated".
 In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
@@
 \citet{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
-In the language of these papers, a 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
+In the language of these papers, a "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
 This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$.
@@
 Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values.
-For example, in \citet[supplementary figure 2a]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates.
+For example, in \citet[supplementary figure 2a of]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates.
 A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.
@@
 
-Many researchers have observed that DNA methylation is spatially correlated along the genome, \cite[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}).
+Many researchers have observed that DNA methylation is spatially correlated along the genome, \citep[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}.
 I call this correlation of methylation levels _co-methylation_.
@@
 The natural interpretation of $\beta_{i, j}$ as the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample will be biased if the probability of sequencing a fragment with a methylated site is different from the probability of sequencing a fragment with an unmethylated site.
-In fact, it has been shown that methylated DNA is overrepresented in bisulfite-sequencing data due, with the problem exacerbated by higher rounds of PCR amplification and dependent on the bisulfite-conversion protocol \cite{Ji:2014fq}.
-PCR amplification can result in overreprestation of one of the DNA strands in bisulfite-sequencing data \cite{Warnecke:1997eh}.
+In fact, it has been shown that methylated DNA is overrepresented in bisulfite-sequencing data due, with the problem exacerbated by higher rounds of PCR amplification and dependent on the bisulfite-conversion protocol \citep{Ji:2014fq}.
+PCR amplification can result in overreprestation of one of the DNA strands in bisulfite-sequencing data \citep{Warnecke:1997eh}.
 
 Lab-based solutions to overcome these biases exist for targeted bisulfite-sequencing, but are technically difficult and expensive to extend to whole-genome studies.
-\cite{Ji:2014fq} propose potential computational corrections for these biases but as yet these have not been implemented.
+\citet{Ji:2014fq} propose potential computational corrections for these biases but as yet these have not been implemented.
 
@@
 
-However, the general use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \cite[e.g.,]{Warton:2011gm}.
+However, the general use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \citep{Warton:2011gm}.
 
@@
 An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2}(\frac{\beta}{1 - \beta} = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero.
-Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \cite[e.g.]{Smyth:2005ta}.
+Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \citep[e.g., ]{Smyth:2005ta}.
 
-\cite{Du:2010dc} advocate for the use of $\mathcal{M}$-values in statistical analyses of methylation levels from microarray data.
+\citet{Du:2010dc} advocate for the use of $\mathcal{M}$-values in statistical analyses of methylation levels from microarray data.
 The main reason for this is that $\mathcal{M}$-values are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values.
 As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformatics and genomics researchers makes it a favourable choice.
-Nonetheless, \cite{Du:2010dc} also advocate that the results of analyses should be reported as $\beta$-values, owing to their "more intuitive biological interpretation".
+Nonetheless, \citet{Du:2010dc} also advocate that the results of analyses should be reported as $\beta$-values, owing to their "more intuitive biological interpretation".
 

### 2014-12-06 data/observable_m-tuples/histogram_of_observable_m-tuples.Rmd
CpGs per read data

---
title: "Observable m-tuples"
author: "Peter Hickey"
date: "12/6/2014"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(scales)
thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)
```

# EPISCOPE

```{r}
dataset <- "EPISCOPE"
hist <- lapply(paste0(dataset, "/",
                      list.files(dataset, "\\.CG\\_per\\_read\\.hist")), 
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "", 
           list.files(dataset, "\\.CG\\_per\\_read\\.hist"))
hist_df <- bind_rows(hist) %>% 
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
g <- hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ Sample, ncol = 4, nrow = 3) +
  coord_cartesian(xlim = c(-1, 8.5)) + 
  scale_y_continuous(label = comma) + 
  xlab("Number of CpGs per read\n(Axis truncated at 8)") + 
  ylab("Percentage of total reads") + 
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
ggsave(paste0(dataset, "_CpGs_per_read.pdf"), g, width = 16, height = 9)
```

# Lister

```{r}
dataset <- "Lister"
hist <- lapply(paste0(dataset, "/",
                      list.files(dataset, "\\.CG\\_per\\_read\\.hist")), 
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "", 
           list.files(dataset, "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("DM\\_", "", sn)
hist_df <- bind_rows(hist) %>% 
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
g <- hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
  coord_cartesian(xlim = c(-1, 8.5)) + 
  scale_y_continuous(label = comma) + 
  xlab("Number of CpGs per read\n(Axis truncated at 8)") + 
  ylab("Percentage of total reads") + 
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
ggsave(paste0(dataset, "_CpGs_per_read.pdf"), g, width = 16, height = 9)
```

# Seisenberger

```{r}
dataset <- "Seisenberger"
hist <- lapply(paste0(dataset, "/",
                      list.files(dataset, "\\.CG\\_per\\_read\\.hist")), 
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "", 
           list.files(dataset, "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("DM\\_", "", sn)
hist_df <- bind_rows(hist) %>% 
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
g <- hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ Sample, ncol = 3, nrow = 1) +
  coord_cartesian(xlim = c(-1, 8.5)) + 
  scale_y_continuous(label = comma) + 
  xlab("Number of CpGs per read\n(Axis truncated at 8)") + 
  ylab("Percentage of total reads") + 
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
ggsave(paste0(dataset, "_CpGs_per_read.pdf"), g, width = 16, height = 9)
```

# Ziller

```{r}
dataset <- "Ziller"
hist <- lapply(paste0(dataset, "/",
                      list.files(dataset, "\\.CG\\_per\\_read\\.hist")), 
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "", 
           list.files(dataset, "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("\\.rmdup", "", sn)
hist_df <- bind_rows(hist) %>% 
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
g <- hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
  coord_cartesian(xlim = c(-1, 8.5)) + 
  scale_y_continuous(label = comma) + 
  xlab("Number of CpGs per read\n(Axis truncated at 8)") + 
  ylab("Percentage of total reads") + 
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
ggsave(paste0(dataset, "_CpGs_per_read.pdf"), g, width = 16, height = 9)
```

### 2014-12-06 Rmarkdown/appendix.Rmd
fixed some formatting and Makefile issues

@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, message = FALSE}
 library(BSgenome.Hsapiens.UCSC.hg19)
@@
 
-```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution", dev = "pdf"}
+```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution", dev = "pdf", message = FALSE}
 library(ggplot2)
@@
 
-```{r cpg_ipds, cache = TRUE}
+```{r cpg_ipds, cache = FALSE, message = FALSE}
 library(BSgenome.Hsapiens.UCSC.hg19)
@@
 
-```{r cpg_ipds_cdf, dev = "pdf"}
+```{r cpg_ipds_cdf, dev = "pdf", message = FALSE}
 library(ggplot2)

### 2014-12-06 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
fixed some formatting and Makefile issues

@@
 Each of read 1 (R1) and read 2 (R2) are plotted separately.
-We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", fig.width = 8, dev = "pdf"}
+We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", dev = "pdf", outwidth = "0.8\\linewidth"}
 library(dplyr)
@@
 The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1.
-This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8, dev = "pdf"}
+This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", dev = "pdf", outwidth = "0.8\\linewidth"}
 #files <- paste0("../M-bias/", list.files("../M-bias/"))
@@
 
-```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data."}
+```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data.", outwidth = "0.8\\linewidth"}
 library(dplyr)
@@
 library(scales)
-thesis_theme <- theme_classic(base_size = 20)
-thesis_theme <- theme_bw(base_size = 20)
+thesis_theme <- theme_classic(base_size = 12)
+thesis_theme <- theme_bw(base_size = 12)
 dataset <- "Lister"
@@
 ```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes.
-The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
+The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", dev = "pdf", outwidth = "0.8\\linewidth"}
 library(dplyr)
@@
 These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
-The suffix 'ac' means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
+The suffix 'ac' means that the option \texttt{--all-combinations} was set.", dev = "pdf", outwidth = "0.8\\linewidth"}
 # Memory

### 2014-12-06 Rmarkdown/wgbs_downstream_analyses.Rmd
fixed some formatting and Makefile issues

@@
 Such a contingency table can be analysed in many different ways.
-Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j.
+Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$.
 $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
@@
 Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the frequencies of the methylation patterns (grey box) are aligned.
-The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, $S_{n}$.
+The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, S_{n}$.
 Some sample may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
@@
 \toprule
-& \texttt\{pryr::object\_size(x)\} & \texttt\{nrow\} & Number of assays & Percentage of \texttt\{NA\} and $0$ values \\ \midrule
+& \texttt{pryr::object\_size(x)} & \texttt{nrow} & Number of assays & Percentage of \texttt{NA} and $0$ values \\ \midrule
 1-tuples                                 & $5.9$ GB                        & $56,348,522$    & $2$              & $28\%$                                     \\
 2-tuples                                 & $20.1$ GB                        & $100,586,237$   & $4$              & $80\%$                                     \\
-2-tuples (\texttt\{--all-combinations\}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
+2-tuples (\texttt{--all-combinations}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
 3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
@@
 All m-tuples are stranded.
-The size of the \code{MethPat} object is computed using the \texttt{object_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
+The size of the \texttt{MethPat} object is computed using the \texttt{object\_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
 The number of rows (\texttt{nrow}) corresponds to the number of m-tuples in the object.
 The number of assays is $2^{\text{m}}$, where $m$ is the size of the m-tuples.
-The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an `NA` value means that that m-tuple was not observed in that particular sample.}
+The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
 \label{tab:methpat_sparsity}

### 2014-12-07 Rmarkdown/wgbs_statistical_framework.Rmd
TODOs for wgbs_statistical_framework

@@
 
-Restricting our attention to CpGs, Figure \ref{fig:EPISCOPE_beta_distribution} shows the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each EPISCOPE sample.
+Restricting our attention to CpGs, __FIGURES__ show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively.
+__FIGURES__ show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
+
 What is immediately clear is that these distributions are bimodal.
-Figure \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} shows that this bimodality is largely driven by the fact that CpGs in CpG islands are mostly unmethylated whereas those outside of CpG islands are generally methylated.
-Intermediate $\beta$-values, those between $0.2-0.8$, are evidence of heterogeneity.
+The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels.
+The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice.
+Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}.
 
-Both Figures \ref{fig:EPISCOPE_beta_distribution} and  \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} also reveal that the genome-wide distributions of $\beta$-values for a given tissue are very similar between the three individuals.
-The three adipose-derived tissues, _SA_, _VA_ and _VAT_, have similar genome-wide profiles while the blood-derived tissue, _BUF_, has more intermediate $\beta$-values, potentially indicating more cellular heterogeneity.
+The only samples with significant intermediate methylation are either somatic cell lines --- __LIST SAMPLES__ --- or cancer cells lines or tissue --- __LIST SAMPLES__.
+__EXPLANATION FOR CANCER__.
+__EPISCOPE DATA SHOWS ITS NOT SOMATIC CELLS BUT CELL LINES THAT GET INTERMEDIATE METHYLATION__.
+__HYPOTHESIS FOR SOMATIC CELL LINES__
 
-```{r EPISCOPE_genome_wide_beta-value_distributions, eval = FALSE, echo = FALSE}
-# Run on one of the unices
-setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
-library(MethylationTuples)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-thesis_theme <- theme_classic(base_size = 20)
-thesis_theme <- theme_bw(base_size = 20)
+- Most samples are hypermethylated
 
-x <- readRDS("EPISCOPE_1_tuples_strand_collapsed.rds")
+__ALL THE GENOME-WIDE FIGURES__
 
-cgi <- read.table("~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt",
-                  header = TRUE, stringsAsFactors = FALSE)
-cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
-cgi_status <- overlapsAny(x, cgi)
-
-beta <- methLevel(x, min_cov = 10L)
-
-cn <- colnames(beta)
-beta <- data.frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
-                   Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
-                   beta = as.vector(beta),
-                   CGI = rep(cgi_status, times = ncol(beta)),
-                   stringsAsFactors = FALSE)
-beta <- tbl_df(beta)
-
-#a <- beta[sample(nrow(beta), 100000), ]
-
-g <- ggplot(aes(x = beta), data = beta) +
-  geom_density(size = 1.5) +
-  facet_grid(Tissue ~ Individual, labeller = label_both) +
-  thesis_theme + scale_colour_brewer(palette = "Dark2") +
-  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
-  xlab(expression(paste(beta, "-values", sep = "")))
-ggsave("EPISCOPE_beta_distribution.pdf", g, height = 9, width = 16)
-
-g <- ggplot(aes(x = beta), data = beta) +
-  geom_density(aes(colour = CGI), size = 2) +
-  facet_grid(Tissue ~ Individual, labeller = label_both) +
-  thesis_theme + scale_colour_brewer(palette = "Dark2") +
-  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
-  xlab(expression(paste(beta, "-values", sep = "")))
-ggsave("EPISCOPE_beta_distribution_CGI_stratified.pdf", g, height = 9,
-       width = 16)
-```
+__CGI FIGURES__
 
@@
 
-```{r EPISCOPE_strand_specific_beta-value_distributions, eval = FALSE, echo = FALSE}
-# Run on one of the unices
-setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
-library(MethylationTuples)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-library(RColorBrewer)
-
-thesis_theme <- theme_classic(base_size = 20)
-thesis_theme <- theme_bw(base_size = 20)
-
-y <- readRDS("EPISCOPE_1_tuples.rds")
-y_fwd <- y[strand(y) == "+"]
-y_rev <- y[strand(y) == "-"]
-
-ol <- findOverlaps(unstrand(y_fwd), shift(unstrand(y_rev), -1L), type = "equal")
-beta_fwd <- methLevel(y_fwd, min_cov = 5L)[queryHits(ol), ]
-beta_rev <- methLevel(y_rev, min_cov = 5L)[subjectHits(ol), ]
-
-cors <- matrix(unlist(mclapply(1:10, function(min_cov, y_fwd, y_rev, ol) {
-  beta_fwd <- methLevel(y_fwd, min_cov = min_cov)[queryHits(ol), ]
-  beta_rev <- methLevel(y_rev, min_cov = min_cov)[subjectHits(ol), ]
-  sapply(seq_len(ncol(y)), function(i, x, y) {
-    cor(x[, i], y[, i], use = "complete.obs")
-    }, x = beta_fwd, y = beta_rev)
-  }, y_fwd = y_fwd, y_rev = y_rev, ol = ol, mc.cores = 10)), ncol = 10)
-
-beta_pairs <- data.frame(Individual = rep(substr(cn, 1, 3),
-                                          each = nrow(beta_fwd)),
-                         Tissue = rep(substr(cn, 4, 6), each = nrow(beta_fwd)),
-                         beta_fwd = as.vector(beta_fwd),
-                         beta_rev = as.vector(beta_rev),
-                         stringsAsFactors = FALSE)
-beta_pairs <- tbl_df(beta_pairs)
-
-beta_pairs %>% group_by(Individual, Tissue) %>%
-  summarise(cor(beta_fwd, beta_rev, use = "complete.obs"))
-
-mypar <- function(a = 1, b = 1, brewer.n = 8, brewer.name = "Dark2",
-                  cex.lab = 2, cex.main = 2, cex.axis = 1.5,
-                  mar = c(5.1, 5.1, 3.5, 2.1), mgp = c(3, 1, 0), ...) {
-    par(mar = mar, mgp = mgp, cex.lab = cex.lab, cex.main = cex.main,
-        cex.axis = cex.axis)
-    par(mfrow = c(a, b), ...)
-    palette(brewer.pal(brewer.n, brewer.name))
-}
-pdf("EPISCOPE_beta_correlations_across_strand.pdf", width = 16, height = 9)
-mypar(4, 3)
-sapply(seq_len(ncol(beta_fwd)), function(i, x, y, ns) {
-  smoothScatter(x[, i], y[, i],
-                xlab = expression(beta["+"]),
-                ylab = expression(beta["-"]),
-                main = paste0(ns[i], ": cor = ", round(cor(x[, i], y[, i],
-                                                        use = "complete.obs"),
-                                                    2)))
-  }, x = beta_fwd, y = beta_rev, ns = colnames(beta_fwd))
-dev.off()
-```
+__TODO: Report tables of strand-correlations with different coverages__
 

### 2014-12-07 Rmarkdown/wgbs_downstream_analyses.Rmd
Typos and fixes

@@
 The input is a pair of `BAM` files created with the `Bismark` aligner, one for each sample.
-__TODO: Can you get output files with the frequency of each methylation pattern at each m-tuple?__ For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples.
+For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples[^methclone_output].
 The "foreground" combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples while the "background" combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two stages [samples]".
@@
 The example in the paper shows results for a threshold of $-60, -70$ and $-80$, but these choices appear somewhat arbitrary.
+
+[^methclone_output]: It is not clear from the paper or documentaiton whether `methclone` returns a file containing the number of methylation patterns at each m-tuple, _a ala_ `methtuple`.
+Unfortunately, I haven't been able to install `methclone`, so have not been able to test this out.
 

### 2014-12-07 Rmarkdown/wgbs_statistical_framework.Rmd
Genome-wide beta distributions

@@
 
-Restricting our attention to CpGs, __FIGURES__ show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively.
-__FIGURES__ show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
+Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity, fig:Lister_plotBetaDensity, fig:Seisenberger_plotBetaDensity, fig:Zilelr_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively.
+Figures \ref{fig:EPISCOPE_plotBetaFreqPoly, fig:Lister_plotBetaFreqPoly, fig:Seisenberger_plotBetaFreqPoly, fig:Zilelr_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
 
@@
 The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice.
-Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}.
+Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.
 
-The only samples with significant intermediate methylation are either somatic cell lines --- __LIST SAMPLES__ --- or cancer cells lines or tissue --- __LIST SAMPLES__.
-__EXPLANATION FOR CANCER__.
-__EPISCOPE DATA SHOWS ITS NOT SOMATIC CELLS BUT CELL LINES THAT GET INTERMEDIATE METHYLATION__.
-__HYPOTHESIS FOR SOMATIC CELL LINES__
+Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _SRR949208_ and _SRR949209_ from the Ziller dataset --- or cancer cells lines and tissue --- _SRR949206_, _SRR949207_, _SRR949210_, _SRR949211_, _SRR949212_, _SRR949213_ and _SRR949215_ from the Ziller dataset.
+Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the Seisenberger dataset also displays greater levels of intermediate methylation.
+This sample was also created by pooling DNA from multiple mice, which may explain the extra variability in the $\beta$-value distribution.
 
-- Most samples are hypermethylated
+It has been previously observed that cancer samples have highly variable DNA methylation \citep{Hansen:2011gu}, which, combined with the possibility of multiple sub-clones, explains these intermediate $\beta$-values in the cancer samples.
+The explanation for the somatic cell lines is less clear.
+Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation.
+This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_.
+Naively, a cell line is a 'pure' cell population, however, the DNA methylation data clearly reveal widespread differences between cells in these populations.
 
-__ALL THE GENOME-WIDE FIGURES__
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
+Densities were normalised so that the maximum value for each sample is 1.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
+'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+\label{fig:EPISCOPE_plotBetaDensity}
+\end{figure}
 
-__CGI FIGURES__
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data.
+Densities were normalised so that the maximum value for each sample is 1.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
+'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+\label{fig:Lister_plotBetaDensity}
+\end{figure}
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data.
+Densities were normalised so that the maximum value for each sample is 1.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
+'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+\label{fig:Seisenberger_plotBetaDensity}
+\end{figure}
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaDensity.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
+Densities were normalised so that the maximum value for each sample is 1.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
+'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+\label{fig:Ziller_plotBetaDensity}
+\end{figure}
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+\label{fig:EPISCOPE_plotBetaFreqPoly}
+\end{figure}
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+\label{fig:Lister_plotBetaFreqPoly}
+\end{figure}
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+\label{fig:Seisenberger_plotBetaFreqPoly}
+\end{figure}
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaFreqPoly.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+\label{fig:Ziller_plotBetaFreqPoly}
+\end{figure}
+
+__TODO: DESCRIBE CGI FIGURES__
 
 \begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
-Most CpGs are highly unmethylated or highly methylated, with the buffy coat samples being the most extreme example of this bimodality.
-'Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose.}
-\label{fig:EPISCOPE_beta_distribution}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:EPISCOPE_plotBetaDensity_CGI}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution_CGI_stratified.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
+\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Most CpGs outside of CpG islands are methylated whereas most CpGs within CpG islands are unmethylated.
 Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose; \em{CGI} = CpG island.}
-\label{fig:EPISCOPE_beta_distribution_CGI_stratified}
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:Lister_plotBetaDensity_CGI}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:Seisenberger_plotBetaDensity_CGI}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaDensity_CGI.pdf}
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:Ziller_plotBetaDensity_CGI}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:EPISCOPE_plotBetaFreqPoly_CGI}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:Lister_plotBetaFreqPoly_CGI}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:Seisenberger_plotBetaFreqPoly_CGI}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaFreqPoly_CGI.pdf}
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Zilelr data, stratified by whether the CpG is in a CpG island.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Only CpGs with at least $10\times$ sequencing coverage are included.
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
+Abbreviations: \em{CGI} = CpG island.}
+\label{fig:Ziller_plotBetaFreqPoly_CGI}
 \end{figure}

### 2014-12-07 Rmarkdown/wgbs_statistical_framework.Rmd
CGI beta distributions

@@
 It has been previously observed that cancer samples have highly variable DNA methylation \citep{Hansen:2011gu}, which, combined with the possibility of multiple sub-clones, explains these intermediate $\beta$-values in the cancer samples.
+
 The explanation for the somatic cell lines is less clear.
 Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation.
+Likewise, the various frontal cortex samples in the Ziller dataset --- _SRR949193_, _SRR949194_, _SRR949195_, _SRR949196_, _SRR949197_, _SRR949198_, _SRR949199_, _SRR949201_, _SRR949202_, _SRR949203_ and _SRR949205_ --- which includes both 'normal' and Alzheimer's samples, have very low levels of intermediate methylation.
 This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_.
@@
 
-__TODO: DESCRIBE CGI FIGURES__
+The bimodality of the genome-wide $\beta$-value distributions is driven by the fact that most CpGs is CpG islands are unmethylated whereas those outside of the islands are mostly methylated.
+Figures \ref{fig:EPISCOPE_plotBetaDensity_CGI, fig:Lister_plotBetaDensity_CGI, fig:Seisenberger_plotBetaDensity_CGI, fig:Ziller_plotBetaDensity_CGI} show the kernel density plots of the $\beta$-value distributions stratified by CpG island status for the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
+These distributions are normalised so that each density has a maximum value of 1.
+
+These plots show that CpG islands have a more strictly bimodal distribution than do the non-islands.
+While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample, except for the _E16.5\_male\_1_ sample.
+The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated.
+Theseplots also show that most of the intermediate methylation occurs outside of the CpG islands.
+
+Because these densities are normalised, these plots don't show the proportion of CpGs in CpG islands.
+Therefore, Figures \ref{fig:EPISCOPE_plotBetaFreqPoly_CGI, fig:Lister_plotBetaFreqPoly_CGI, fig:Seisenberger_plotBetaFreqPoly_CGI, fig:Ziller_plotBetaFreqPoly_CGI} show the same data but with $\beta$-values grouped into $0.01$-width bins and plotted against the percentage of total CpGs in each bin.
+These plots highlight that the majority of unmethylated CpGs occur in CpG islands and that most of the intermediate methylation occurs outside of CpG islands, owing to most CpGs being outside of an island.
 
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{CGI} = CpG island.}
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:EPISCOPE_plotBetaDensity_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{CGI} = CpG island.}
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Lister_plotBetaDensity_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{CGI} = CpG island.}
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Seisenberger_plotBetaDensity_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.
-Abbreviations: \em{CGI} = CpG island.}
+Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Ziller_plotBetaDensity_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
-Abbreviations: \em{CGI} = CpG island.}
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
 \label{fig:EPISCOPE_plotBetaFreqPoly_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
-Abbreviations: \em{CGI} = CpG island.}
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
 \label{fig:Lister_plotBetaFreqPoly_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
-Abbreviations: \em{CGI} = CpG island.}
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
 \label{fig:Seisenberger_plotBetaFreqPoly_CGI}
@@
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.
-Abbreviations: \em{CGI} = CpG island.}
+Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
 \label{fig:Ziller_plotBetaFreqPoly_CGI}

### 2014-12-07 Rmarkdown/wgbs_statistical_framework.Rmd
Strand correlations, summary

@@
 CpG $\beta$-values are often computed by aggregating the $m$ and $u$ counts across the forward and reverse strands.
-This assumes that the two strands are indeed symmetrically methylated.
-To investigate this, for each CpG we can compute the $\beta$-values using only the forward-strand reads, $\beta^{+}$, and only the reverse-strand reads, $\beta^{-}$.
+On average, this doubles the sequencing coverage of each CpG but presupposes that the two strands are indeed symmetrically methylated.
+To investigate the validity of this assumption we can compute the correlation of strand-specific $\beta$-values.
+Tables \ref{tab:EPISCOPE_strand_correlations, tab:Lister_strand_correlations, tab:Seisenberger_strand_correlations, tab:Ziller_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
 
-__TODO: Report tables of strand-correlations with different coverages__
+There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the low strand-correlations at these cutoffs.
+Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 - 0.9$.
+However, there are some notable exceptions.
+The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset; _J1\_1_ from the Seisenberger dataset_ --- all have far lower strand correlations, $r = 0.5 - 0.7$.
+This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.
+The other Seisenberger samples --- _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ --- also have lower strand correlations, but these may be due to the fact that these data are from pooled samples.
+
+Overall, it seems that non-embryonic stem cell samples have very high correlations of CpG $\beta$-values across strands, which means that these data can safely be combined across strands.
+However, it is a good idea to first check this assumption prior to combining data across strands.
+
+```{r EPISCOPE_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+library(xtable)
+x <- read.table("../data/strand_correlation/EPISCOPE_strand_correlations.txt",
+stringsAsFactors = FALSE)
+colnames(x) <- 1:10
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the EPISCOPE dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:EPISCOPE_strand_correlations")
+align(xt) <- rep("c", ncol(x) + 1)
+digits(xt) <- 2
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+```
+
+```{r Lister_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+x <- read.table("../data/strand_correlation/Lister_strand_correlations.txt",
+stringsAsFactors = FALSE)
+colnames(x) <- 1:10
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Lister dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Lister_strand_correlations")
+align(xt) <- rep("c", ncol(x) + 1)
+digits(xt) <- 2
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+```
+
+```{r Seisenberger_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+x <- read.table("../data/strand_correlation/Seisenberger_strand_correlations.txt",
+stringsAsFactors = FALSE)
+colnames(x) <- 1:10
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Seisenberger dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Seisenberger_strand_correlations")
+align(xt) <- rep("c", ncol(x) + 1)
+digits(xt) <- 2
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+```
+
+```{r Ziller_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
+stringsAsFactors = FALSE)
+colnames(x) <- 1:10
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Ziller dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_strand_correlations")
+align(xt) <- rep("c", ncol(x) + 1)
+digits(xt) <- 2
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+```
 
@@
 In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
-However, as can be seen from __FIGURE__, the extreme bimodality of the $\beta$-value distribution means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
+However, as can be seen from Figures \ref{EPISCOPE_plotBetaFreqPoly, Lister_plotBetaFreqPoly, Seisenberger_plotBetaFreqPoly, Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
 To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
@@
 Because in a whole-genome bisulfite-sequencing experiment we measure methylation from a pool of cells, not a single haploid copy of the genome, I do not think it makes sense to describe a particular cytosine in the genome as being a 'methylcytosine'.
-That is, it only makes sense to describe an observation of a cytosine, $z_{i}$, as being methylated or unmethylated and not to describe a cytosine, $i$, as being a methylated or unmethylated.
+That is, it only makes sense to describe an observation of a cytosine, $z_{i}$, as being methylated or unmethylated and not to describe a cytosine, really a position in the genome, $i$, as being a methylated or unmethylated.
 However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have sought to use the latter definition, which I believe to be an unnecessary source of confusion.
@@
 
-__TODO__
+In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data.
+I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured.
+Using this framework, I have described common estimators of DNA methylation and some of their statistical properties.
+
+I have highlighted the genome-wide similarities and differences between a diverse set of whole-genome bisulfite-sequencing samples.
+I have shown that the CpG islands drive a strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
+
+Somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples.
+The may reflect that the DNA methylome is well-established in these samples and relatively static.
+The increased level of partial methylation in somatic cell lines compared to somatic tissue samples may be attributable to the development of sub-clones during the culturing of the sample.
+This is consistent with the high epipolymorphism observed in a study that tracked the dynamics of DNA methylation in an _in vitro_ evolutionary cell culture system \citep{Landan:2012kp}.
+Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strand to increase the sequencing coverage of each CpG.
+
+Induced pluripotent stem cell lines also appear to have a quite strictly regulated methylome, with little intermediate methylation.
+This is likely a consequence of the fact that during the induction of pluripotency, the methylome of the sample is 'reset' \citep{Stricker:2013kl} thereby resulting in a homogeneous population of cells.
+
+The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_), the hypervariable progenitor germ cells (_E16.5\_male\_1_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_).
+__TODO: SOmethign about Seisenberger being difficult to interpret__
 

### 2014-12-08 Rmarkdown/biology_background.Rmd
Typos and fixes
Included more figures in biology_background

@@
 One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next, the _intra-pair distances_ (IPDs).
-__FIGURE__ is the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19).
+Figure \ref{fig:IPD_ECDF_hg19} is a plot of the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19).
 We see that approximately $70\%$ of CpGs are within 100 bp of the next CpG.
-__FIGURE__ also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome.
-By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more "close" pairs and more "distant" pairs than we would expect by chance.
+Figure \ref{fig:IPD_ECDF_hg19} also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome.
+By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more 'close' pairs and more 'distant' pairs than we would expect by chance.
 
-__FIGURE OBSERVED VS EXPECTED__ also shows that there is a cluster of CpGs with $IPD < 10$.
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/IPD_ECDF_hg19.pdf}
+\caption{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19).
+The observed distances are contrasts with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
+\label{fig:IPD_ECDF_hg19}
+\end{figure}
+
+Figure \ref{fig:IPD_frequency_polygon_hg19} is an alternative way to visualise these data, which plots the percentage of pairs of CpGs with a given IPD.
+Figure \ref{fig:IPD_frequency_polygon_hg19} shows that there is a cluster of CpGs with $IPD < 10$.
 These largely correspond to CpGs that lie within what are called _CpG islands_ (CGIs).
-Within these CpG islands are the 20-40% of CpGs that are frequently unmethylated in mammalian genomes.
+
+\begin{figure}[htbp]
+\includegraphics[width=\textwidth]{../figures/IPD_frequency_polygon_hg19.pdf}
+\caption{The frequency of the observed distances between adjacent CpGs in the human reference genome (hg19).}
+\label{fig:IPD_frequency_polygon_hg19}
+\end{figure}
+
+CpG islands contain the 20-40% of CpGs that are frequently unmethylated in mammalian genomes.
 CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}.
@@
 
-These "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
+These 'sandy/beachy' metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
 CpG island shores, shelves and the open sea are all defined with respect to CpG islands.
@@
 
-__TODO: Describe CHARM \citep{Irizarry:2008hg}, although not yet sure which sub-section it should appear in__.
-
 \subsection{Sodium bisulfite conversion assays}\label{sec:bs_assays}
@@
 This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \citep{Stirzaker:2014cu}.
-There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed by \citey{Dedeurwaerder:2014bt}.
+There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed by \citet{Dedeurwaerder:2014bt}.
 

### 2014-12-08 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
density of cogs

@@
 It is instead driven by the number and density of CpGs on the chromosome.
-For example, \texttt{chr19}, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than \texttt{chr18}, which has less than half the CpG density of \texttt{chr19} (see __APPENDIX__).
+For example, \texttt{chr19}, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than \texttt{chr18}, which has less than half the CpG density of \texttt{chr19} (see Table \ref{tab:cpg_density_hg19}).
 The relationship between the maximum memory usage and the tuple size (\texttt{-m}) is complex; more data have to be retained as \texttt{-m} increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
@@
 
+```{r autosomal_cpg_density_hg19, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+library(xtable)
+x <- read.table("../data/wgbs_bioinformatics_analysis/CpG_density_hg19.txt",
+                header = TRUE, stringsAsFactors = FALSE)
+xt <- xtable(x, caption = "Percentage of CpGs per chromosome in the human reference genome (hg19).", label = "tab:cpg_density_hg19")
+align(xt) <- rep("c", ncol(x) + 1)
+digits(xt) <- 2
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, include.rownames = FALSE)
+```
+
 \subsection{Availability}

### 2014-12-08 Rmarkdown/wgbs_statistical_framework.Rmd
summary of wgbs_statistical_framework

@@
 
-There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the low strand-correlations at these cutoffs.
-Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 - 0.9$.
-However, there are some notable exceptions.
-The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset; _J1\_1_ from the Seisenberger dataset_ --- all have far lower strand correlations, $r = 0.5 - 0.7$.
+There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the low strand-correlations at these lower coverage cutoffs.
+Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 - 0.9$, however, there are some notable exceptions.
+
+The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$.
+The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$).
 This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.
-The other Seisenberger samples --- _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ --- also have lower strand correlations, but these may be due to the fact that these data are from pooled samples.
 
-Overall, it seems that non-embryonic stem cell samples have very high correlations of CpG $\beta$-values across strands, which means that these data can safely be combined across strands.
-However, it is a good idea to first check this assumption prior to combining data across strands.
+All three Seisenberger samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell _J1\_1_.
+However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.
+
+Overall, it seems that with the exception of embryonic stem cells, most samples have very highly correlated strand-specific CpG $\beta$-values, which means that these data can safely be combined across strands.
+However, it remains a good idea to first check this assumption prior to combining data across strands.
 
@@
 ```
-
 \subsection{Interpretation}
@@
 I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured.
-Using this framework, I have described common estimators of DNA methylation and some of their statistical properties.
+By using a common statistical framework we can better understand how different statistical methods relate to one another.
+Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties.
+I then examined the empirical distributions of these variables across a diverse set of 51 whole-genome bisulfite-sequencing samples.
 
-I have highlighted the genome-wide similarities and differences between a diverse set of whole-genome bisulfite-sequencing samples.
-I have shown that the CpG islands drive a strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
+Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
+The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_).
+The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA.
+Nonetheless, some of the increased variability in the Seisenberger data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.
 
-Somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples.
+In contrast, the somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples.
 The may reflect that the DNA methylome is well-established in these samples and relatively static.
@@
 This is consistent with the high epipolymorphism observed in a study that tracked the dynamics of DNA methylation in an _in vitro_ evolutionary cell culture system \citep{Landan:2012kp}.
-Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strand to increase the sequencing coverage of each CpG.
+Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.
 
@@
 This is likely a consequence of the fact that during the induction of pluripotency, the methylome of the sample is 'reset' \citep{Stricker:2013kl} thereby resulting in a homogeneous population of cells.
+These samples also have highly correlated strand-specific $\beta$-values, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.
 
-The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_), the hypervariable progenitor germ cells (_E16.5\_male\_1_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_).
-__TODO: SOmethign about Seisenberger being difficult to interpret__
-
-\section{General TODOs}
-
-* Discuss distribution of coverage? Perhaps in context of sequencing bias?
+All of the above highlights the considerable variability of DNA methylation data, both between samples and within a sample, and the care with which $\beta$-values must be interpreted.
+While an attractively simple measure, analyses based on $\beta$-values are based on the 'average' behaviour, where the averaging is over many sources of variation.
+Analyses based on $\beta$-values also do not make full use of the information available in bisulfite-sequencing data, as we shall see in Chapter \ref{chap:wgbs_downstream_analyses}.

### 2014-12-08 Rmarkdown/appendix.Rmd
Removed code from appendix related to biology_background

--- 2014-12-06 Rmarkdown/appendix.Rmd
@@
 ---
-
-## CpGs are underrepresented in the human genome (hg19)
-
-Firstly, the frequency of each nucleotide is computed.
-Secondly, the expected frequency of each dinucleotide is computed under the _independence model_.
-Under the independence model, the frequency of each dinucleotide is simply the product of the respective observed nucleotide frequencies.
-Finally, these expected frequencies are compared to the observed values from the respective reference genomes.
-
-I only use the autosomes, sex chromosomes and mitochondrial chromosome in all analyses, that is, I do not consider the "random", "unplaced" and "alternate haplotype" contigs in the reference genomes.
-
-All results are computed at the chromosome-level as well as being summarised genome-wide.
-
-### Compute the nucleotide and dinucleotide frequencies
-
-```{r, cache = TRUE}
-library(BSgenome.Hsapiens.UCSC.hg19)
-library(knitr)
-# Nucleotide frequencies
-nf <- bsapply(new("BSParams", X = Hsapiens, FUN = letterFrequency,
-            exclude = c("rand", "Un", "hap"), simplify = TRUE),
-            letters = c("A", "C", "G", "T"))
-# Pretty print the results
-kable(data.frame(percentage = round(100 * rowSums(nf) /
-                                      sum(as.numeric(nf)), 1)))
-
-# GC-percentage
-round(100 * sum(nf[c("C", "G"), ]) / sum(as.numeric(nf)), 1)
-
-# Dinucleotide frequencies
-df <- bsapply(new("BSParams", X = Hsapiens, FUN = dinucleotideFrequency,
-            exclude = c("rand", "Un", "hap"), simplify = TRUE))
-# Pretty print the results
-kable(data.frame(percentage = round(100 * rowSums(df) /
-                                      sum(as.numeric(df)), 1)))
-
-# CpG percentage
-round(100 * rowSums(df) / sum(as.numeric(df)), 1)["CG"]
-```
-
-### Plot the observed:expected ratio for each dinucleotide
-
-```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution", dev = "pdf"}
-library(ggplot2)
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
-edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
-x <- data.frame(Dinucleotide = names(rowSums(df)),
-                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))),
-                Expected = as.vector(edf))
-ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) +
-  geom_point(size = 5) + coord_cartesian(ylim = c(0, 1.5)) +
-  ggtitle("CpGs are underrepresented in human reference genome (hg19)") +
-  thesis_theme +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  geom_hline(yintercept = 1, linetype = 2)
-```
-
-### Frequency of CpGs on each chromosome in the human reference genome (hg19)
-
-Chromosome 19 has the highest CpG density of the autosomes in the human reference genome (hg19):
-```{r cpg_frequency}
-sort(df['CG', ] / colSums(df) * 100, decreasing = TRUE)
-```
-
-## Distances between CpGs in the human reference genome (hg19)
-
-I extract the co-ordinates of all CpGs in the human reference genome (hg19) and compute the intra-pair distances (IPD) between adjacent CpGs.
-I then plot the empirical cumulative distribution function of IPD.
-These are compared to the distances we would expect if CpGs were uniformly distributed across the genome.
-
-I only use the autosomes, sex chromosomes and mitochondrial chromosome in all analyses, that is, I do not consider the "random", "unplaced" and "alternate haplotype" contigs in the reference genomes.
-
-```{r cpg_ipds, cache = TRUE}
-library(BSgenome.Hsapiens.UCSC.hg19)
-set.seed(666)
-
-# Observed IPDs
-cpgs <- bsapply(new("BSParams", X = Hsapiens, FUN = matchPattern,
-                    exclude = c("rand", "Un", "hap"), simplify = TRUE),
-                pattern = "CG")
-d <- unlist(lapply(cpgs, start), use.names = FALSE)
-ipd <- diff(d)
-# Drop negative distances, which correspond to CpGs on different chromosomes
-ipd <- ipd[ipd > 0]
-
-# Expected IPDs if CpGs were uniformly distributed on genome
-d_sim <- bsapply(new("BSParams", X = Hsapiens, FUN = function(x) {
-  cpg_freq <- countPattern("CG", x)
-  sort(sample(x = length(x), size = cpg_freq, replace = FALSE))
-}, exclude = c("rand", "Un", "hap"), simplify = TRUE))
-ipd_sim <- diff(unlist(d_sim, use.names = FALSE))
-# Drop negative distances, which correspond to CpGs on different chromosomes
-ipd_sim <- ipd_sim[ipd_sim > 0]
-
-# Drop negative values, which are due to CpGs being on different chromosomes
-ipd_df <- data.frame(IPD = c(ipd, ipd_sim),
-                  Source = c(rep("Observed", length(ipd)),
-                             rep("Expected", length(ipd_sim))))
-```
-
-### Empirical cumulative distribution plot
-
-```{r cpg_ipds_cdf, dev = "pdf"}
-library(ggplot2)
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
-
-ggplot(aes(x = IPD, colour = Source), data = ipd_df) + stat_ecdf() + thesis_theme +
-  coord_cartesian(xlim = c(0, 1000)) +
-  xlab("Distance between CpGs (bp)\n(Axis truncated at 1000)") +
-  ggtitle("Human reference genome (hg19)")
-```
-
-### Frequency polygon
-
-```{r cpg_ipds_hist, dev = "pdf"}
-library(ggplot2)
-library(scales)
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
-
-ipd_table <- table(ipd)
-ipd_sim_table <- table(ipd_sim)
-ipd_table_df <- data.frame(IPD = c(as.numeric(names(ipd_table)),
-                                   as.numeric(names(ipd_sim_table))),
-                           Count = c(ipd_table, ipd_sim_table),
-                           Source = c(rep("Observed", length(ipd_table)),
-                                      rep("Expected", length(ipd_sim_table))))
-
-ggplot(aes(x = IPD, y = Count),
-       data = subset(ipd_table_df, Source == "Observed")) +
-  geom_line() + coord_cartesian(xlim = c(0, 300)) +
-  xlab("Distance between CpGs (bp)\n (Axis truncated at 300)") +
-  ggtitle("Human reference genome (hg19)") +
-  scale_y_continuous(labels = comma)
-```
 

### 2014-12-08 Rmarkdown/datasets.Rmd
Almost finished datasets

--- 2014-10-19 Rmarkdown/datasets.Rmd
@@
 
-## Chapter overview
+\chapter{Datasets used in thesis}\label{chap:datasets}
 
-I have analysed more than 50 whole-genome bisulfite-sequencing datasets during the course of my PhD studies. This brief chapter documents those that I use in my thesis.
+\section{Chapter overview}
 
-## Lister data
+This chapter briefly documents the 51 whole-genome bisulfite-sequencing samples that I use in my thesis. The `BAM` files for each sample underwent the same basic processing:
 
-The _Lister data_ refer to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
+1. Genetic variants were called using `Bis-SNP (v0.82.2)` via the `bissnp_easy_usage.pl` script.
+2.
+M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set.
+These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an average CpG $\beta$-value more than $0.03$ away from the median were excluded from methylation calling (read 1 and read 2 analysed separately).
+3.
+CpG methylation calling was performed using `methtuple (v1.4.0)` for m-tuples $\text{m} = 1, \ldots, 8$.
+CpG 2-tuples were called both with and without the `--all-combinations` flag; all other tuple sizes were called without the `--all-combinations` flag.
+4.
+A `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples, 3-tuples and 4-tuples.
+I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this large sizes to be generally useful.
+
+__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e., going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $275$ GB alone.
+
+\section{Lister dataset}\label{sec:Lister_dataset}
+
+The _Lister data_ refers to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
 The Lister data were the largest publicly available human WGBS datasets until quite recently.
+All samples are of human DNA.
+
+\subsection{Sample descriptions}
 
 The WGBS libraries from the \cite{Lister:2009hy} paper were the first published WGBS libraries from mammalian DNA.
-A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal fibroblasts (_FF_), with those from a pluripotent tissue, embryonic stem cells (_H1_).
+A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (`IMR90`), with those from a pluripotent tissue, embryonic stem cells (`H1`).
 Each tissue was run in duplicate.
 While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing.
-In any case, the published analyses pool these technical duplicates, which ignores all between-replicate variability.
+In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.
 
-The WGBS libraries from the \cite{Lister:2011kg} include some created and published by other groups (see __TABLE__).
+The WGBS libraries from the \cite{Lister:2011kg} include some created and published by other groups (see Table \ref{tab:Lister_dataset}).
 These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells.
-Unfortunately, there are no replicates for any of these samples.
+There are no replicates for any of the 2011 samples.
 
-__TODO: Table of datasets__
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}cccccc@{}}
+\toprule
+Sample              & Tissue type             & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
+ADS                 & Adipose                 & PE         & $75$ bp     & $23\times$    & \citet{Lister:2011kg}       \\
+ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp     & $24\times$    & \citet{Lister:2011kg}       \\
+ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp     & $26\times$    & \citet{Lister:2011kg}       \\
+FF                  & Foreskin fibroblasts    & SE         & $85$ bp     & $16\times$    & \citet{Lister:2011kg}       \\
+FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp     & $10\times$    & \citet{Lister:2011kg}       \\
+FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
+FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp     & $8\times$     & \citet{Lister:2011kg}       \\
+FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp     & $17\times$    & \citet{Lister:2011kg}       \\
+IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
+IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
+IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
+H1\_r1              & ESC                     & SE         & $85$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
+H1\_r2              & ESC                     & SE         & $85$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
+H1+BMP4             & IVD from H1             & SE         & $85$ bp     & $33\times$    & \citet{Lister:2011kg}       \\
+H9                  & ESC                     & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
+H9\_Laurent         & ESC                     & PE         & $75$ bp     & $8\times$     & \citet{Laurent:2010iw}      \\
+HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
+\end{tabular}
+\caption{Samples in Lister dataset.
+The reported read lengths are prior to any trimming of the reads.
+Abbreviations: \em{iPSC} = induced pluripotent stem cell; \em{ESC} = embryonic stem cell; \em{IVD} = \em{in vitro} differentiated from pluripotent cell line \em{PE} = single-end sequencing; \em{PE} = paired-end sequencing.}
+\label{tab:Lister_dataset}
+\end{table}
 
-There are a four "mini datasets" within the Lister data that I make some use of in my thesis. These are described below and illustrated in __FIGURE__.
-
+There are a four 'mini datasets' within the Lister data that I make some use of in my thesis.
 The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper.
-The _ADS_ sample, a human adipose tissue cell line, is the "founder" of this mini dataset.
+The _ADS_ sample, a human adipose tissue cell line, is the 'founder' of this mini dataset.
 The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line.
-Specifically, the _ADS-adipose_ sample is purified adipocytes from _ADS_ while the _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.
+The _ADS-adipose_ sample are "adipocytes derived from the `ADS` cells through adipogenic differentiation conditions".
+The _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.
 
 The second mini dataset I refer to as the _Lister-FF_ data and includes samples _FF_, _FF-iPSC\_6.9_, _FF-iPSC\_19.7_, _FF-iPSC\_19.11_ and _FF-iPSC\_11.11+BMP4_, all from the 2011 paper.
-The _FF_ sample, a foreskin fibroblasts cell line, is the "founder" of this this mini dataset.
-The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from _FF_.
+The _FF_ sample, a foreskin fibroblast cell line, is the 'founder' of this this mini dataset.
+The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from `FF`.
 In fact, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are subclones derived from _FF-iPSC\_19_, whose methylome was not sequenced.
 I believe _FF-iPSC\_6.9_ is an independently derived iPSC cell line from _FF_, although this isn't made clear in the original publication.
-The _FF-iPSC\_19.11+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _FF-iPSC\_19.11_ by growing a clone of it in bone morphogenic protein 4 (BMP4).
+The _FF-iPSC\_19.11+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _FF-iPSC\_19.11_ by growing a clone of it in bone morphogenic protein 4 (_BMP4_).
 
 The third mini dataset I refer to as the _Lister-IMR90_ data and includes samples _IMR90\_r1_ (2009), _IMR90\_r2_ (2009) and _IMR90-iPSC_ (2011).
-The _IMR90-iPSC_ sample is an induced pluripotent stem cell line derived from the _IMR90_ cell line.
+The _IMR90-iPSC_ sample is an induced pluripotent stem cell line derived from the _IMR90_cell line.
 
@@
 
+\subsection{Creation of \texttt{BAM} files}
+
 The aligned reads for the \cite{Lister:2009hy} data were downloaded from [http://neomorph.salk.edu/human_methylome/](http://neomorph.salk.edu/human_methylome/).
 The aligned reads for the \cite{Lister:2011kg} data were downloaded from [http://neomorph.salk.edu/ips_methylomes/](http://neomorph.salk.edu/ips_methylomes/).
-I wrote Python scripts to convert these files to `SAM` files, these scripts are available from [https://github.com/PeteHaitch/Lister2BAM](https://github.com/PeteHaitch/Lister2BAM).
-These `SAM` files were then converted to `BAM` files and duplicate reads were marked using Picard's `MarkDuplictes` routine [http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/).
+These samples were aligned against the hg18 build of the human reference genome.
 
-## EPISCOPE data
+I wrote Python scripts to convert these files to `SAM` files.
+These scripts are available from [https://github.com/PeteHaitch/Lister2BAM](https://github.com/PeteHaitch/Lister2BAM).
+These `SAM` files were then converted to `BAM` files with `SAMtools` \citep{Li:2009ka} and duplicate reads were marked using Picard's `MarkDuplicates` routine [http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/).
 
-__TODO: Check permission to use data in thesis__
+\section{EPISCOPE dataset}\label{sec:EPISCOPE_dataset}
 
-## Whitelaw data
+The EPISCOPE data were kindly provided to me by Professor Susan Clark (Garvan Institute of Medical Research, Sydney) and Dr Peter Molloy (CSIRO Animal, Food and Health Sciences).
+This dataset is not yet published.
 
-I describe this dataset in detail in Chapter \ref{chap:avy}.
+\subsection{Sample descriptions}
 
-__TODO: Check permission to use data in thesis__
+The data are from three human donors across four different tissues, for a total of 12 whole-genome bisulfite-sequencing libraries. The four tissues are:
 
-## Ziller data
+- BUF: Buffy coat layer, which are leukocytes and platelets derived by centrifugation of a whole blood sample.
+- SA: Subcutaneous adidose tissue, which is fat found just below the skin.
+Unlike visceral fat, "subcutaneous fat is not related to many of the classic obesity-related pathologies, such as heart disease, cancer, and stroke" (__TODO: Get better reference than wikipedia__).
+- VA: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs (stomach, liver, intestines, kidneys, etc.) (__TODO: Get better reference than wikipedia__).
+- VAT: Visceral adipocytes, which are derived from the VA tissue.
 
-## Seisenberger data
+The data are summarised in Table \ref{tab:EPISCOPE_dataset}.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}cccc@{}}
+\toprule
+Sample & Sequencing & Read length & Ave. coverage \\ \midrule
+E13BUF & PE         & $101$ bp    & $8\times$     \\
+E13SA  & PE         & $101$ bp    & $28\times$    \\
+E13VA  & PE         & $150$ bp    & $27\times$    \\
+E13VAT & PE         & $101$ bp    & $25\times$    \\
+E18BUF & PE         & $101$ bp    & $21\times$    \\
+E18SA  & PE         & $101$ bp    & $25\times$    \\
+E18VA  & PE         & $150$ bp    & $36\times$    \\
+E18VAT & PE         & $101$ bp    & $26\times$    \\
+E23BUF & PE         & $101$ bp    & $12\times$    \\
+E23SA  & PE         & $101$ bp    & $29\times$    \\
+E23VA  & PE         & $101$ bp    & $32\times$    \\
+E23VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
+\end{tabular}
+\caption{Samples in EPISCOPE dataset. The reported read lengths are prior to any trimming of the reads.}
+\label{tab:EPISCOPE_dataset}
+\end{table}
+
+\subsection{Creation of \texttt{BAM} files}
+
+The sequencing data for these 12 samples were processed and aligned by Aaron Statham (Garvan Institute of Medical Research).
+Each sample was aligned to the human reference genome (hg19) using Bismark (v0.8.3) with the Bowtie2 backend.
+The default alignment options were used, except that the maximum insert size for valid paired-end alignments was set to 1000 instead of 500 (`-X 1000`).
+An example command is shown below:
+
+```
+bismark -p 4 --bowtie2 -X 1000 --unmapped --ambiguous --gzip --bam -o SAMPLE_NAME REF_GENOME -1 READ_1.fq -2 READ_2.fq
+```
+
+Duplicate reads had already been removed from the `BAM` files that I received.
+
+\section{Seisenberger dataset}\label{sec:Seisenberger_dataset}
+
+The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \cite{Seisenberger:2012ko}.
+These were a convenience sample provided to me by a colleague.
+I thank Felix Krueger (Babraham Institute) who processed and aligned the reads, and sent me the `BAM` files.
+
+\subsection{Sample descriptions}
+
+I have the data for only three samples from the original publication, detailed in Table \ref{tab:Seisenberger_dataset}.
+The _J1\_1_ sample is from an embryonic stem cell line while both the _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ samples are from a pool of $10-30$ embryos.
+Developmentally, the samples are ordered _J1\_1_ (embryonic stem cell), _E6.5\_epiblast\_1_ (embryonic day 6.5 epiblast) and _E16.5\_male\_1_ (embryonic day 16.5 male progenitor germ cells).
+
+I believe that the samples I received labelled _J1\_1_ and _E16.5\_male\_1_  in fact correspond to _J1\_2_ and _E16.5\_male\_2_, respectively, i.e., the second  replicate rather than the first.
+The data I received are all $100$ bp paired-end sequences, which matches replicate 2 rather than replicate 1 for both of these samples \citet[Supplementary Table 1 of]{Seisenberger:2012ko}.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}clcccl@{}}
+\toprule
+Sample            & Tissue                     & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
+J1\_1             & ESC                        & PE         & $100$ bp    & $12\times$    & \citet{Seisenberger:2012ko} \\
+E6.5\_epiblast\_1 & Epiblast                   & PE         & $101$ bp    & $13\times$    & \citet{Seisenberger:2012ko} \\
+E16.5\_male\_1    & Male progenitor germ cells & PE         & $150$ bp    & $12\times$    & \citet{Seisenberger:2012ko} \\ \bottomrule
+\end{tabular}
+\caption{Samples in Seisenberger dataset. The reported read lengths are prior to any trimming of the reads.}
+\label{tab:Seisenberger_dataset}
+\end{table}
+
+\subsection{Creation of \texttt{BAM} files}
+
+The sequencing data for these 3 samples were processed and aligned by Felix Krueger (Babraham Institute).
+Each sample was aligned to the mouse reference genome (GRCm38/mm10) using Bismark (v0.7.12) with the Bowtie1 backend.
+The default alignment options were used.
+
+\section{Ziller dataset}\label{sec:Ziller_dataset}
+
+The Ziller are a subset of the data used in \cite{Ziller:2013fs}.
+Specifically, I use a convenience sample of 19 (__TODO: Check final numbers__) WGBS libraries.
+These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research), who already had aligned and processed BAM files for these 19 libraries.
+These libraries in fact represent 19 sequencing runs of eight biological samples, which I was not aware of until the data had already been processed.
+Therefore, I have _post hoc_ merged these sequencing runs.
+This unfortunately reduces the power to detect genetic variants since the coverage of individual sequencing runs is lower than merged data, but does not adversely affect methylation calling since the number of methylated and unmethylated reads can be summed across sequencing runs.
+
+The eight biological samples are as follows: frontal cortex from two 'normal' women donors (_Frontal\_cortex\_normal\_1_ and Frontal\_cortex\_normal\_2_) and from two women who had Alzheimer's disease (_Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_); a sample from a human liver carcinoma cell line (\HepG2\_cell\_line_); a new sample from the _IMR90_ lung fibroblast cell line (_IMR90\_cell\_line_); and samples from a colon cancer matched tumour-normal pair (_Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_).
+
+Table \ref{tab:Ziller_dataset_sequencing_runs} summarises the data for the 19 sequencing runs and __TABLE__ summarises the data for the _post hoc_ merged data.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}cccccc@{}}
+\toprule
+Sample    & Tissue                     & Sequencing & Read length & Ave. coverage & First published       \\ \midrule
+SRR949193 & Frontal\_cortex\_normal\_1 & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
+SRR949194 & Frontal\_cortex\_normal\_1 & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
+SRR949195 & Frontal\_cortex\_normal\_1 & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
+SRR949196 & Frontal\_cortex\_normal\_2 & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
+SRR949197 & Frontal\_cortex\_normal\_2 & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
+SRR949198 & Frontal\_cortex\_normal\_2 & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
+SRR949199 & Frontal\_cortex\_AD\_1     & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
+SRR949201 & Frontal\_cortex\_AD\_1     & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
+SRR949202 & Frontal\_cortex\_AD\_2     & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
+SRR949203 & Frontal\_cortex\_AD\_2     & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
+SRR949206 & HepG2\_cell\_line          & PE         & 101         & $2\times$     & \citet{Ziller:2013fs} \\
+SRR949207 & HepG2\_cell\_line          & PE         & 101         & $1\times$     & \citet{Ziller:2013fs} \\
+SRR949208 & IMR90\_cell\_line          & PE         & 101         & $1\times$     & \citet{Ziller:2013fs} \\
+SRR949209 & IMR90\_cell\_line          & PE         & 101         & $3\times$     & \citet{Ziller:2013fs} \\
+SRR949210 & Colon\_Tumor\_Primary      & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\
+SRR949211 & Colon\_Tumor\_Primary      & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\
+SRR949212 & Colon\_Tumor\_Primary      & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
+SRR949213 & Colon\_Tumor\_Primary      & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\
+SRR949215 & Colon\_Primary\_Normal     & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\ \bottomrule
+\end{tabular}
+\caption{Sequencing runs in Ziller dataset. The reported read lengths are prior to any trimming of the reads.}
+\label{tab:Ziller_dataset_sequencing_runs}
+\end{table}
+
+__TODO: Table after post-hoc merge__
+
+\section{CpG islands}
+
+I have used the CpG island definition from the hidden Markov model proposed by \citet{Wu:2010do}.
+The predicted CpG islands for the human reference genome (both hg18 and hg19) were downloaded from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/) on 29/10/2014.
+I used the supplied software to create the predicted CpG islands for the mouse reference genome (mm10).

### 2014-12-08 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Switched to tabular. Proof reading.

@@
 \centering
-\begin{tabular}{@{}lll@{}}
+\begin{tabulary}{\textwidth}{LLL}
 \toprule
@@
 Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
-\end{tabular}
+\end{tabulary}
 \caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of back-ends used by these software.}
@@
 Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate.
-However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]).
+However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as 'PCR duplicates'[^pcr_dup]).
 Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
@@
 
-> Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair.
+> "Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair.
 When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment.
@@
 It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair.
-"Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15 ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).
+"Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15" ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).
 
@@
 
-```{r m-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:ADS_M-bias", fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset.
+```{r ADS_M-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset.
 Each of read 1 (R1) and read 2 (R2) are plotted separately.
@@
 
-```{r M-bias_spike, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:E18VA_M-bias", fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset.
+```{r E18VA_M-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset.
 Each of read 1 (R1) and read 2 (R2) are plotted separately.
@@
 
-Strictly speaking, cellular heterogeneity exists once you study DNA from more than a single cell, as shown in a recent papers using single-cell bisulfite-sequencing \citep{Guo:2013iha,Smallwood:2014kn}.
+Strictly speaking, cellular heterogeneity exists once you study DNA from more than a single cell, as shown in a recent papers using single-cell bisulfite-sequencing \citep{Guo:2013ih, Smallwood:2014kn}.
 However, in this context it means when you study DNA from a sample that is made up of different _types_ of cells.
@@
 A cheaper alternative is to genotype the sample on a genome-wide SNP microarray.
-This will give very accurate, very cheap genotypes at a large number of loci ($500,000 - $5,000,000$).
+This will give very accurate, very cheap genotypes at a large number of loci ($500,000 - 5,000,000$).
 However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.
@@
 The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci at which to call methylation events.
-It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina ('_Lister-style_') whole-genome bisulfite-sequencing protocol.
+It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina ('Lister-style') whole-genome bisulfite-sequencing protocol.
 
@@
 \includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
-\caption{`Bis-SNP` is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
+\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
 For all three loci, the reads mapped to the forward strand contain a thymine.
@@
 Chromosome & Strand & Position & M & U \\ \midrule
-chr1       & +      & 100      & 7 & 1 \\
-chr1       & -      & 101      & 5 & 2 \\
-chr2       & +      & 400      & 0 & 3 \\
-chr2       & +      & 450      & 1 &   \\ \bottomrule
+chr1       & $+$      & 100      & 7 & 1 \\
+chr1       & $-$      & 101      & 5 & 2 \\
+chr2       & $+$      & 400      & 0 & 3 \\
+chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
 \end{tabular}
-\caption{My caption}
+\caption{Example of output for methylation calling at 1-tuples.}
 \label{tab:1-tuples}
@@
 Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
-Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_analysis}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
+Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
 In order to study these concepts, we need software that can perform methylation calling at m-tuples, which was not available when I began my PhD, and so I wrote `methtuple`.
@@
 For each m-tuple, I also define the intra-pair distance (_IPD_) as the vector containing the $m - 1$ pair-wise distances (measured in bp) between methylation loci in the m-tuple.
-For example, the 2-tuple (chr7:+:145, chr7:+:163) has $IPD = (163 - 145) = (18)$.
-The 5-tuple (chr2:-:560, chr2:-:570, chr2:-:572, chr2:-:588, chr2:-:612) has $IPD = (570 - 560, 572 - 570, 588 - 572, 612 - 588) = (10, 2, 16, 24)$.
+For example, the 2-tuple (`chr7:+:145`, `chr7:+:163`) has $IPD = (163 - 145) = (18)$.
+The 5-tuple (`chr2:-:560`, `chr2:-:570`, `chr2:-:572`, `chr2:-:588`, `chr2:-:612`) has $IPD = (570 - 560, 572 - 570, 588 - 572, 612 - 588) = (10, 2, 16, 24)$.
 The IPD vector of a 1-tuple is undefined.
@@
 To illustrate several of the above-mentioned concepts, suppose we sequence a region of the genome containing five methylation loci with three paired-end reads (`A`, `B` and `C`):
+
+__TODO: Convert the following code chunks to figures in Illustrator__
 
@@
 
-```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data.", outwidth = "0.8\\linewidth"}
+```{r Lister_cpgs_per_read, dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data.", out.width = "0.8\\linewidth"}
 library(dplyr)
@@
 \centering
-\begin{tabular}{@{}lllll@{}}
+\begin{tabulary}{\textwidth}{LLLLL}
 \toprule
@@
 \\ \bottomrule
-\end{tabular}
-\caption{My caption}
+\end{tabulary}
+\caption{Other software for methylation calling at m-tuples.}
 \label{tab:other_m-tuple_methylation_callers}
@@
 
-```
+```{sh}
 chr     strand  pos1    pos2    pos3    MMM     MMU     MUM     MUU     UMM     UMU     UUM     UUU
@@
 
-```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes.
-The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", dev = "pdf", outwidth = "0.8\\linewidth"}
+```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes.
+The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", dev = "pdf", out.width = "0.8\\linewidth"}
 library(dplyr)
@@
 
-```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_memory", fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
+```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
 These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
-The suffix 'ac' means that the option \texttt{--all-combinations} was set.", dev = "pdf", outwidth = "0.8\\linewidth"}
+The suffix 'ac' means that the option \texttt{--all-combinations} was set.", dev = "pdf", out.width = "0.8\\linewidth"}
 # Memory

### 2014-12-08 Rmarkdown/datasets.Rmd
Switched to longtable

@@
 
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}cccccc@{}}
+\begin{longtable}{@{}cccccc@{}}
 \toprule
@@
 HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
-\end{tabular}
 \caption{Samples in Lister dataset.
@@
 \label{tab:Lister_dataset}
-\end{table}
+\end{longtable}
 
@@
 
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}cccc@{}}
+\begin{longtable}{@{}cccc@{}}
 \toprule
@@
 E23VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
-\end{tabular}
 \caption{Samples in EPISCOPE dataset. The reported read lengths are prior to any trimming of the reads.}
 \label{tab:EPISCOPE_dataset}
-\end{table}
+\end{longtable}
 
@@
 
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}clcccl@{}}
+\begin{longtable}{@{}clcccl@{}}
 \toprule
@@
 E16.5\_male\_1    & Male progenitor germ cells & PE         & $150$ bp    & $12\times$    & \citet{Seisenberger:2012ko} \\ \bottomrule
-\end{tabular}
 \caption{Samples in Seisenberger dataset. The reported read lengths are prior to any trimming of the reads.}
 \label{tab:Seisenberger_dataset}
-\end{table}
+\end{longtable}
 
@@
 
-The eight biological samples are as follows: frontal cortex from two 'normal' women donors (_Frontal\_cortex\_normal\_1_ and Frontal\_cortex\_normal\_2_) and from two women who had Alzheimer's disease (_Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_); a sample from a human liver carcinoma cell line (\HepG2\_cell\_line_); a new sample from the _IMR90_ lung fibroblast cell line (_IMR90\_cell\_line_); and samples from a colon cancer matched tumour-normal pair (_Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_).
+The eight biological samples are as follows: frontal cortex from two 'normal' women donors (_Frontal\_cortex\_normal\_1_ and Frontal\_cortex\_normal\_2_) and from two women who had Alzheimer's disease (_Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_); a sample from a human liver carcinoma cell line (HepG2\_cell\_line_); a new sample from the _IMR90_ lung fibroblast cell line (_IMR90\_cell\_line_); and samples from a colon cancer matched tumour-normal pair (_Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_).
 
@@
 SRR949215 & Colon\_Primary\_Normal     & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\ \bottomrule
-\end{tabular}
 \caption{Sequencing runs in Ziller dataset. The reported read lengths are prior to any trimming of the reads.}
 \label{tab:Ziller_dataset_sequencing_runs}
-\end{table}
+\end{longtable}
 

### 2014-12-09 Rmarkdown/appendix.Rmd
Re-jigged appendix

@@
 
-## Equivalence of tests to identify DMCs in a one-group experiment
+\section{Equivalence of tests to identify DMCs in a one-group experiment}\label{app:one_group_equivalence}
+
+\subsection{Mathematical equivalence}
+
+The saturated log-linear model can be written as $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{a, b}^{meth_state,sample}$, where $F_{a, b}$ is the value of the $(a, b)$ element in the $n \times 2$ contingency table.
+The null hypothesis that the cytosine is not a DMC is then tested by testing for an association between `sample` and `meth_state`, i.e., whether the parameter $\lambda_{a, b}^{meth_state, sample}$ is significantly different from zero.
+
+This log-linear model can be reformulated as a logistic regression by noting that the log-odds (logits) for each row are $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \log(F_{a, 1}) - \log(F_{a, 2})$. Therefore,
+
+\begin{align*}
+\log(\frac{F_{a, 1}}{F_{a, 2}}) &= \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{1}^{sample} + \lambda_{a, 1} \\
+&- (\lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{2}^{sample} + \lambda_{a, 2}) \\
+&= \underbrace{(\lambda_{1}^{sample} - \lambda_{2}^{sample})}_{\alpha} + \underbrace{(\lambda_{a, 1} - \lambda_{a, 2})}_{\gamma_{i}}
+\end{align*}
+
+The expression $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha + \gamma_{i}$ is a logistic regression model.
+We can test the null hypothesis that the cytosine is not a DMC by testing for a `sample`-effect on the log-odds, i.e., whether the parameter $\gamma_{i}$ is significantly different from zero.
+
+Regardless of whether we use the log-linear model or the logistic model, to test the null hypothesis that the cytosine is not a DMC we can compute the deviance[^deviance] of the reduced model and compare it to a $\chi^{2}$ random variable with $(2 - 1) \times (n - 1)$ degrees of freedom.
+For the log-linear regression the reduced model is $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample}$ and for the logistric regression the reduced model is $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha$.
+
+[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.
+
+\subsection{Computational equiavelence}
 
@@
 
-The saturated log-linear model can be written as $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{a, b}^{meth_state,sample}$, where $F_{a, b}$ is the value of the $(a, b)$ element in the $n \times 2$ contingency table.
-The null hypothesis that the cytosine is not a DMC is then tested by testing for an association between `sample` and `meth_state`, i.e., whether the parameter $\lambda_{a, b}^{meth_state, sample}$ is significantly different from zero.
-
-This log-linear model can be reformulated as a logistic regression by noting that the log-odds (logits) for each row are $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \log(F_{a, 1}) - \log(F_{a, 2})$. Therefore,
-
-\begin{align*}
-\log(\frac{F_{a, 1}}{F_{a, 2}}) &= \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{1}^{sample} + \lambda_{a, 1} \\
-&- (\lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{2}^{sample} + \lambda_{a, 2}) \\
-&= \underbrace{(\lambda_{1}^{sample} - \lambda_{2}^{sample})}_{\alpha} + \underbrace{(\lambda_{a, 1} - \lambda_{a, 2})}_{\gamma_{i}}
-\end{align*}
-
-The expression $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha + \gamma_{i}$ is a logistic regression model.
-We can test the null hypothesis that the cytosine is not a DMC by testing for a `sample`-effect on the log-odds, i.e., whether the parameter $\gamma_{i}$ is significantly different from zero.
-
-Regardless of whether we use the log-linear model or the logistic model, to test the null hypothesis that the cytosine is not a DMC we can compute the deviance[^deviance] of the reduced model and compare it to a $\chi^{2}$ random variable with $(2 - 1) \times (n - 1)$ degrees of freedom.
-For the log-linear regression the reduced model is $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample}$ and for the logistric regression the reduced model is $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha$.
-
-[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.
-
-## The probability that two dependent Bernoulli random variables are identical
+\section{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}
 
@@
 
-## Analysis of \cite{Lacey:2013iy}
+\section{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
 
-I want to estimate the number of CpGs used by \cite{Lacey:2013iy} to estimate the correlation of methylation levels.
-They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, MTCTL2.
-Unfortunately, the data are not publicly available.
-Instead, I use data from ENCODE of a skeletal muscle myotubes cell line that had RRBS performed in triplicate (HSMMtube).
+I want to estimate the number of CpGs used by \citet{Lacey:2013iy} to estimate the correlation of methylation levels.
+They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, _MTCTL2_.
+Unfortunately, there is no URL provided to these data.
+\citet{Lacey:2013iy} cite their data as having come from \citet{Tsumagari:2013gn}, which includes a sample "Normal Human Skeletal Muscle Myoblasts", _HSMM_ ([http://genome-mirror.duhs.duke.edu/ENCODE/cellTypes.html](http://genome-mirror.duhs.duke.edu/ENCODE/cellTypes.html), [http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/](http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/)).
+I believe this to be the same cell line.
+In any case, for my purposes I simply needed some comparable RRBS data, not necessarily the exact same sample as  _MTCTL2_.
+
+There are approximately $60,000$ CpGs on chromosome 11 in these comparable RRBS data. The following R code shows how I computed this using three RRBS replicates of the _HSMM_ sample.
 
@@
 library(data.table)
-library(dplyr)
-print(getwd())
-```
-
-```{r, cache = FALSE, message = FALSE}
-dl_dir <- 'downloads/'
+dl_dir <- "../data/appendix/"
 url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
@@
 
-Then I count the number of CpGs on chromosome 11 in each sample:
+Then I count the number of CpGs on chromosome 11 in each replicate:
 ```{r}
-y <- lapply(x, function(xx) {
-  xx <- tbl_dt(xx)
-  xx %>%
-    group_by(V1) %>%
-    summarise(n())
-})
-
-rbind_all(y) %>% filter(V1 == 'chr11')
+y <- rbindlist(x)
+y[V1 == "chr11", .N, by = list(V4)]
 ```
 
-## Computing details
+\section{Computing details}\label{app:computing_details}
 

### 2014-12-09 Rmarkdown/comethylation_review.Rmd
Minor fixes
Began revising co-methylation review

--- 2014-11-06 Rmarkdown/comethylation_review.Rmd
@@
 ---
-title: "Co-methylation review"
 author: "Peter Hickey"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
 
+\chapter{A critical review of co-methylation}\label{chap:co-methylation_review}
+
 \section{Introduction}
@@
 This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation.
-In this chapter I first review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}.
+In this chapter I critically review these definitions and put them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}.
 I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
@@
 I define co-methylation as the dependence structure of DNA methylation data.
-This broad definition encompasses several concepts that have variously been described as "co-methylation", or similar, some of which are closely related and others that are disparate.
-My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" (\cite{Schatz:2004cha}) or "the relationship between the degree of methylation over distance" (\cite{Eckhardt:2006gh}); it also includes the "correlation of two [loci] across many sample" (\cite{Akulenko:2013fl}) and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighboring cell types" (\cite{Capra:2014uh}).
+This broad definition encompasses several concepts that have variously been described as 'co-methylation', or similar.
+My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}) or "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; it also includes the "correlation of two [loci] across many sample" \citep{Akulenko:2013fl} and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
 A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.
@@
 Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample.
-Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated (e.g.
-\cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010drm, Landan:2012kp}).
+Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}.
 Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.
+
 The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__.
-At a higher level is the _correlation of aggregate methylation levels_, such as the autocorrelation of $\beta$-values within a sample.
+At a higher level is the _correlation of aggregate methylation levels_, such as the spatial correlation of $\beta$-values within a sample.
 
 Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance].
-For example, \cite{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples.
+For example, \citet{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples.
 Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.
 
-[^concordance]: One "between-sample correlation of methylation levels" measure that does not fall under my definition of co-methylation is the following.
-Consider a pair of samples, $j$ and $j'$, and compute $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$.
-This measure often reported as evidence for the "concordance" or "replicability" of methylation levels between replicates (e.g., when comparing the genome-wide methylation levels of a cell line passaged 4 versus 5 times \cite{Lister:2009hy} or when comparing different technologies and assays \cite{Zhang:2013uu, Stevens:2013hv, Hansen:2011gu}) but is not the same as co-methylation.
+[^concordance]: One measure of between-sample correlation of methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$.
+This measure is often reported as evidence for the "concordance" of methylation levels between replicates, but is not what I consider a form of co-methylation.
 
@@
 An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation.
-For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci.
+For example, we might identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci.
 This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.
@@
 There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data.
-For example, if co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis tools.
+If co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis methods.
 
@@
 
-To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
-These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).
+To date, most analyses of co-methylation have been based on analysing correlations of aggregate levels of methylation, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
+An exception is the analysis in \citet{Lister:2009hy}.
+While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the authors instead look for patterns in the spacing of 'methylcytosines' along the genome[^Lister_methylcytosines].
 
-The analysis in \cite{Lister:2009hy} takes a different approach.
-While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the autnors instead looks for patterns in the spacing of "methylcytosines" along the genome.
+[^Lister_methylcytosines]: See Section \ref{sec:beta-value_interpretation} for a critique of this definition of 'methylcytosine'.
 
-In reviewing this previous work, I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.
+In reviewing this previous work, I have been forced to make my best guess about some of the methods since they are often not sufficiently described in the original publication.
 
-\subsection{\cite{Eckhardt:2006gh}}\label{sec:eckhardt}
+\subsection{\citet{Eckhardt:2006gh}}\label{sec:eckhardt}
 
-\cite{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples.
-Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}.
-\cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.
+\citet{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples.
+Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \citep{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \citep{Hansen:2011gu}.
+\citet{Eckhardt:2006gh} performed PCR-amplicon bisulfite-sequencing of $2,524$ amplicons on human chromosomes $6$, $20$ and $22$.
 PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data.
@@
 As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance".
-They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $ >2,000$ bp" and reference Figure 3a in support of this claim.
-However, what is plotted in Figure 3a is not "correlation" in the usual sense of the word.
+They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $>2,000$ bp" and reference the plot reprinted in Figure \ref{fig:Eckhardt} in support of this claim.
+In Figure \ref{fig:Eckhardt}, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp, which is not a correlation in the usual sense of the word.
+Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].
 
-In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-printing this figure__).
-Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/ng1909-F3a.jpg}
+\caption{Correlation of DNA methylation with spatial distance.
+(a) Correlation between comethylation and spatial distance.
+Orange dots represent CpG methylation values aggregated and averaged over $25,000$ individual measurements.
+Gray dots represent CpG methylation values based on resampling of random CpG positions.
+Blue dots indicate CpG methylation values based on resampling of amplicon positions.
+At distances $>1,000$ bp, we did not detect any correlation between CpG methylation and spatial distance.
+Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Eckhardt:2006gh}), copyright (2006)}
+\label{fig:Eckhardt}
+\end{figure}
+
+[^beck]: An email discussion with the senior author, Stephan Beck, was unable to clarify the definition or method.
+
 The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp.
@@
 
-[^beck]: An email discussion with the senior author, Stephan Beck, was unable to clarify the definition or method.
-
 My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of identical methylation calls at different CpGs from (generally) different amplicons, that is, $Pr(Z_{h, i} = Z_{h', i'})$.
-I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.
+I will refer to my best guess of "percentage identical methylation" as $PIM^{\ast}$, and now turn to some of its properties.
 
@@
 For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$.
-Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels.
+Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels.
 This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$.
-In this setting we can compute the theoretical $PIM^{*}$ as follows:
+In this setting we can compute the theoretical $PIM^{\ast}$ as follows:
 
 \begin{align*}
-PIM^{*} &= Pr(Z_{h, i} = Z_{h', i'}) \\
+PIM^{\ast} &= Pr(Z_{h, i} = Z_{h', i'}) \\
         &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
@@
 
-This result is proved in __TODO: Cite and link to appendix__.
+This result is proven in Appendix \ref{app:bernoulli_equal}.
 
@@
 
-In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:
+In general, $PIM^{\ast} \in [0, 1]$ since it is a probability, which is already unlike a correlation coefficient that takes values in $\[-1, 1\]$.
+There are a few important special cases that are worth discussing:
 
 1.
-$PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
-2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
-3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+$PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
+3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
 
-As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$.
-In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp.
+As for $PIM^{\ast}$, $PIM^{\ast}_{inid} \in [0, 1]$.
+In contrast, it is easy to show by differentiation that $PIM^{\ast}_{iid} \geq 0.5$ and that $PIM^{\ast}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp.
 $< 0.5$) for $c > 0$ (resp.
 $c < 0$).
-__TODO: Illustrate these special cases with a plot.__
 
-To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$.
+To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{\ast}_{iid}$.
 If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp.
-However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible, even supposing $c > 0$.
+However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{\ast} < 0.5$ are possible, even supposing $c > 0$.
 
-Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$.
-I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.
+Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure \ref{fig:Eckhardt} is mathematically equivalent to $PIM^{\ast}$.
+I believe this to be true but due to the aforementioned inadequacies of the method description it is not possible to verify this derivation.
 
-Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis.
-Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.
+Quite apart from what exactly is plotted in Figure \ref{fig:Eckhardt}, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis.
+Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure \ref{fig:Eckhardt} does not represent any one sample but a pool of samples.
 
-\subsection{\cite{Cokus:2008fc}}\label{sec:Cokus}
+\subsection{\citet{Cokus:2008fc}}\label{sec:Cokus}
 
-\cite{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing.
+\citet{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing.
 They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_.
@@
 
-\cite{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci.
+\citet{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci.
 From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin".
-They also identified correlations between different methylation contexts, e.g.
-CG vs.
+They also identified correlations between different methylation contexts, e.g., CG vs.
 CHG, which "suggest[s] complex interactions between the different types of methylation".
@@
 The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only).
-Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier transform analysis of the autocorrelation.
+Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.
 
-[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}
+[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}
 
@@
 From this result Cokus et al.
-hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites 8-10 apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_) (__TODO: If I want a picture explaining this then Figure 3b from \cite{Jurkowska:2011dt} might form the basis for my own illustration).
-However, in a later paper by the same group, this $10$ nucleotide periodicity was also observed for CG and CHG methylation, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}.
-
-[^comethylation_def2]: This is referred to as "co-methylation", here meaning the methylation of multiple loci in a single event, by \cite{Jurkowska:2011dt}.
-This article reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.
-
+hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8-10$ nucleotides apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_).
 The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})
 
-\subsection{\cite{Chodavarapu:2010iqa}}{sec:chodavarapu}
+[^comethylation_def2]: This is also referred to as "co-methylation" by \citet{Jurkowska:2011dt}, here meaning the methylation of multiple loci in a single event.
+\citet{Jurkowska:2011dt} reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.
 
-\cite{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (HSF1).
+\subsection{\citet{Chodavarapu:2010iqa}}{sec:chodavarapu}
+
+\citet{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (_HSF1_).
 The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq.
-The nucleosome start sites anchor the methylation observations to a common grid and is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes.
+The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes.
 A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
-Again, the observed periodicity is confirmed by analysing a Fourier transform of these plots.
+Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.
 
-In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the HSF1 sample, which confirmed the $10$ nucleotide periodicity.
-This led \cite{Chodavarapu:2010iqa} to propose an extended version of their earlier hypothesis \cite{Cokus:2008fc}: namely, "that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.
+In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ nucleotide periodicity.
+In fact, these results revealed that the $10$ nucleotide is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA.
+This led \citet{Chodavarapu:2010iqa} to discard their earlier hypothesis as to the source of the $10$ nucleotide periodicity \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.
 
-\subsection{\cite{Li:2010fb}}\label{sec:li}
+\subsection{\citet{Li:2010fb}}\label{sec:li}
 
-\cite{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
-The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $> 1,000$ bp", citing their results as being consistent with those in \cite{Eckhardt:2006gh}.
-In addition to this genome-wide analysis, \cite{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements.
+\citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
+The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", citing their results as being consistent with those in \citet{Eckhardt:2006gh}.
+In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements.
 Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.
 
-Unfortunately, just as in \cite{Eckhardt:2006gh}, \cite{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available.
+Unfortunately, just as in \citet{Eckhardt:2006gh}, \citet{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available.
 All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions.
@@
 
-Supplementary Figure 6a plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp.
-Based on the "$n = 18,936,995$" quoted in the header of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "istance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs.
-If all pairs were used this $n$ would be much higher since there are more than twenty million CpGs on one strand of the haploid human genome.
-To emphasise, the correlation shown by \cite{Li:2010fb} is not the same as the autocorrelation computed in \cite{Cokus:2008fc} since the autocorrelation uses all pairs of CpGs, regardless of the number of intervening CpGs.
+Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp.
+Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs.
+If all pairs were used this $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome.
+To emphasise, the correlation shown by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.
 
-The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \cite{Cokus:2008fc} (highlighted in Supplementary Figure 7).
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/journal.pbio.1000533.s006.eps}
+\caption{Co-methylation patterns for different genomic features. A reproduction of Supplementary Figure 6 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
+\label{fig:Li6}
+\end{figure}
+
+The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \citet{Cokus:2008fc} (reproduced in Figure \ref{fig:Li7}).
 This is further evidence that nucleosome spacing effects the strength of co-methylation.
-However, \cite{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \cite{Cokus:2008fc} and \cite{Chodavarapu:2010iqa}.
+However, \citet{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/Figure_S7.eps}
+\caption{Transformation of methylation correlation at nearby CpG cytosines. A reproduction of Supplementary Figure 7 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
+\label{fig:Li7}
+\end{figure}
 
@@
 
-In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation.
+In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \citet{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation.
 They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
-This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report?
-Check my data any such increases__).
-__TODO: Is this the fastest decay of correlation in any species?__
+This is the first, and, as far as I am aware, only report of an organism in which the correlation of methylation levels first increases before decreasing.
 
-\cite{Lyko:2010dr} also identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__).
-They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.
-
-The data supporting these results, Supplementary Figure 7 of the original publication, appears weak at first glance since the maximum value of the autocorrelation is less than $0.015$.
-I believe that this is due an error in the computation of the autocorrelation.
+However, there is a something strange about this result, shown in Figure \ref{fig:Lyko}, namely, the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}).
+I strongly suspect that the autocorrelation has been incorrectly computed.
 Specifically, what I suspect has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been assigned a $\beta$-value of zero.
 
-To demonstrate this, let $\beta^{*} = (\beta^{*}_{i'}, \ldots, \beta^{*}_{n'})$ be this "augmented" vector of $\beta$-values:
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/Lyko_Figure_S7.pdf}
+\caption{Periodicity of methylation patterns.
+(A) Autocorrelation of CpG methylation status over 1 kb.
+(B) Autocorrelation over 100 bp.
+Figures A and B show that the correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away.
+CpGs within a neighborhood of 2 bp to 100 bp are thus more likely to share the same methylation status than more distant CpGs.
+(C) Fourrier transform of autocorrelation showing a clear periodicity peak at 33 cycles per 100 bp (every 3 bp).
+(D) Distribution of codon position of mCs, and distribution of methylation level depending on the position.
+These two panels indicate that the distance between methylated CpGs is often a multiple of three and that the methylated cytosine corresponds most frequently to the first nucleotide of an arginine codon.
+A reproduction of Supplementary Figure 7 from \citet{Lyko:2010dr} under the Creative Common Attribution (CCBY) licence.}
+\label{fig:Lyko}
+\end{figure}
 
-$$
-\beta^{*}_{i'} = \left\{
+While I don't have the data used in \citet{Lyko:2010dr}, I believe an example using the chromosome 22 data from the _FF_ sample demostrates the point.
+Let $\mathbf{\beta^{\ast}}$ be the augmented vector of $\beta$-values, where:
+
+\begin{equation*}
+\beta^{\ast}_{i'} = \left\{
   \begin{array}{ll}
@@
   \right.
-$$
+\end{equation*}
 
-Using the `FF` chromosome 22 data, I computed and plotted (__FIGURE__) the correlations of all pairs of CpG $\beta$-values as well as the autocorrelation of the augmented vector $\beta^{*}$.
-From these plots we see that the autocorrelation of $\beta^{*}$ is a scaled version of the correlations of all pairs of CpG $\beta$-values.
-Therefore, while the autocorrelations are incorrectly computed, conclusions based on the relative height of peaks should be approximately correct.
+__FIGURE__ is a plot of the autocorrelation of the augmented vector, $\mathbf{\beta^{\ast}}$.
 
-```{r}
-# TODO: The above described analysis and plots
+__TODO: Include plot of this for FF data on chr22.
+This brings into question the other results based on this plot, e.g., the 3bp periodicity, which may reflect nucleotide periodicities rather than methylation periodicities.__
+__TODO: What does this plot look like__
+__TODO: What are the implications for the results in Lyko__
 
-```
+\subsection{\citet{Lacey:2013iy}}\label{sec:lacey_2013}
 
-\subsection{\cite{Lacey:2013iy}}\label{sec:lacey_2013}
-
-\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2).
-The empirical and fitted variograms asymptote for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
+\citet{Lacey:2013iy} explore co-methylation in their paper on modelling and analysing reduced representation bisulfite-sequencing data (RRBS).
+To do so, \citeauthors{Lacey:2013iy} fit a Gaussian variogram model to the CpG $\beta$-values from chromosome 11 of a normal myotube cell line (_MTCTL2_).
+The empirical and fitted variograms asymptote for CpGs separated by around $3000$ bp (Figure 1d of \citet{Lacey:2013iy}, unable to be reproduced due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
 
 Caution is warranted in extrapolating these results to statements about genome-wide co-methylation.
-Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs (__TODO: Check number__) in the haploid human genome.
+Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs in the haploid human genome.
 Secondly, these data are from a single sample from a single tissue, which cannot capture the sample-to-sample nor tissue-to-tissue variability in co-methylation.
@@
 
-[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (__TODO: Cite and link to appendix__).
+[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (see Appendix \ref{app:lacey}).
 
-\subsection{\cite{Lister:2009hy}}\label{sec:lister_2009}
+\subsection{\citet{Lister:2009hy}}\label{sec:lister_2009}
 
-\cite{Lister:2009hy} report an $8-10$ nucleotide periodicity in the occurence "methylcytosines".
-Recall that \cite{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a Binomial test (see Section \ref{sec:criticism_of_methylcytosines} for details and a criticism of this concept).
-Hence, while this may appear to be a within-fragment analysis of co-methylation, it in fact uses aggregate methylation levels.
-However, unlike the methods described in Section \ref{sec:previous_cor_of_agg_methylation}, the analysis of these aggregate methylation levels is not based on a correlation of these levels, but rather on unnormalised counts, which leads to problems.
+\citet{Lister:2009hy} report an $8-10$ nucleotide periodicity in the occurrence "methylcytosines".
+Recall that \citet{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a Binomial test (see Section \ref{sec:beta-value_interpretation} for details and a criticism of this concept).
+Hence, while this may appear to be a within-fragment analysis of co-methylation with it's reference to "methylcytosines", it in fact  based on $\beta$-values.
+Furthermore, unlike the methods described in Section \ref{sec:previous_cor_of_agg_methylation}, the analysis is not based on a correlation of these $\beta$-values, but rather on unnormalised counts of a binary classification of $\beta$-values, which leads to problems.
 
-\cite{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$.
+\citet{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$.
 These counts are plotted as a bar chart, with $IPD$ along the x-axis and count on the y-axis, along with a cubic spline fit to the counts.
-This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (Supplementary Figure 9 \cite{Lister:2009hy}).
+This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (reproduced in Figure \ref{fig:Lister_2009_Sup_Fig_9}).
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.pdf}
+\caption{Spacing of Adjacent Methylcytosines in Different Contexts.
+Prevalence of mCHG/mCHH sites (y-axis) as a function of the number of bases between adjacent mCHG/mCHH sites (x-axis) based on all non redundant pair-wise distances up to $50$ nt in exons, introns and random sequences.
+The blue line represents smoothing by cubic splines.
+Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Lister:2009hy}), copyright (2009)}
+\label{fig:Lister_2009_Sup_Fig_9}
+\end{figure}
 
@@
 The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts, since there are very few "methylcytosines" for these contexts in these human samples.
-Furthermore, in all nine contexts, the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.
+Furthermore, the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.
 
-Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites".
+Based on these nine plots, \citet{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \citet{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites".
 However, I suspect that the results in these nine plots are in fact driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines.
@@
 In fact, there are well known periodicities in the distribution of nucleotides in genomes.
-For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \cite{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \cite{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.
+For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.
 
@@
 
-Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules.
-I call this within-fragment co-methylation, where a "fragment" refers to a fragment of DNA.
+As discussed in Section \ref{sec:m-tuples}, bisulfite-sequencing allows for the study of methylation patterns at m-tuples from individual DNA fragments.
+Since each read comes from a single haplotype from a single cell[^chimera], these data can be used to study co-methylation at the level of individual DNA molecules.
+I argue that this is the more biologically meaningful level at which to study co-methylation, since this is the same scale as the physical process that lays down and maintains DNA methylation.
+
+[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).
+
+I call this _within-fragment co-methylation_, where 'fragment' refers to a fragment of DNA.
 More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
@@
 We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
-
-Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, by studying the methylation patterns at m-tuples.
-Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.
-
-[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).
+There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.
 
 If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci.
-However, the most commonly used technology for high-throughput bisulfite-sequencing, which uses Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum.
-Assembling these reads into longer contigs will not solve this problem since this cannot guarantee that the assembled reads actually originated from the same DNA fragment.
+However, the most commonly used technology for high-throughput bisulfite-sequencing, using Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum.
+We are therefore limited to studying co-methylation between methylation loci separated by at most $200-250$ bp.
+Note that assembling these reads into longer contigs will not solve this problem since the assembly cannot guarantee that the assembled reads actually originated from the same cell or haplotype.
 
-Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lower throughput of these technologies, and the reliability with which they can identify methylcytosines, create new complications (__CITE__).
-An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).
+Long-read sequencing technologies, such as Pacific Biosciences, Roche 454 and Oxford Nanopore, would overcome this limitation but the lower throughput of these technologies, and the reliability with which they can identify methylcytosines, create new complications.
+An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo].
 
-[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons.
-These include the fragility of large DNA fragments treated with sodium bisulfite (__TODO: Discuss with someone from AGRF__)
+[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons, including the fragility of large DNA fragments treated with sodium bisulfite.
 
-To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides.
-There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.
+To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200-250$ nucleotides.
 
-\subsection{\cite{Cokus:2008fc}}\label{sec:cokus}
+\subsection{\citet{Cokus:2008fc}}\label{sec:cokus_within_fragment}
 
-As already discussed in Section \ref{sec:Cokus}, \cite{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_.
-They find this same periodicity "when individual reads [were] examined directly".
+As already discussed in Section \ref{sec:Cokus}, \citet{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_.
+They also reported they they find this same periodicity "when individual reads [were] examined directly".
 
 The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence.
-For the within-fragment co-methylation results (Supplementary Figure 14 of \cite{Cokus:2008fc}) they first identify all reads containing multiple CHH loci.
+For the within-fragment co-methylation results (reproduced in Figure \ref{fig:Cokus_2008_Sup_Fig_14}) \citeauthor{Cokus:2008fc} first identify all reads containing multiple CHH loci.
 Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated.
@@
 They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes.
-\cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
-This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages.
+\citeauthor{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
+This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of normality, and a running average of these averages.
 There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
 
-[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read, since a read may contain multiple pairs.
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/Cokus_2008_Sup_Fig_14.pdf}
+\caption{Within-read probability of additional methylation of CHH sites within a given distance from a methylated CHH site.
+Data were derived from individual BS-Seq reads.
+The x-axis indicates the distance between the two cytosines.
+The y-axis indicates the probability of methylation of CHH sites within the given distance from another methylated CHH site.
+Each point is the mean value from averaging the probability from each of the five Arabidopsis chromosomes, and the blue line is a running average of these mean values.
+Error bars represent 95% confidence intervals via critical values of Student $t$ distributions.
+Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Cokus:2008fc}), copyright (2008)}
+\label{fig:Cokus_2008_Sup_Fig_14}
+\end{figure}
+
+[^cokus_clarification]: A read may contain multiple pairs of CHH loci and it is not clear whether \citeauthor{Cokus:2008fc} consider all pairs of CHH loci or only the first pair of CHH loci in a read.
 The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
@@
 
-I consider these results to use a restricted definition of within-fragment "co-methylation".
+I consider these results to be based on a restricted definition of within-fragment co-methylation.
 To explain, these results tell us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment, but only about the probability that both CHH loci are methylated.
-Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$.
 
+Specifically, \citet{Cokus:2008fc} only estimate the marginal probability that a CHH is methylated, $Pr(Z_{h, i} = 1)$, and the probability that two CHH loci on the same fragment are both methylated, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $i < i'$.
 From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$.
@@
 There is good reason to suspect that this is not true and that there exists significant pair-to-pair variability.
-For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for "co-methylation" of these same loci.
+For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for the co-methylation of these same loci.
+This aggregation may also introduction artificial associations into the $2 \times 2$ contingency table from which these probabilities are estimated \citep{Good:1987ce}.
+
+__UP TO HERE__
 
@@
 - An autocorrelation analysis implicitly assumes stationarity along the genome whereas between-sample co-methylation assumes stationarity across samples (__TODO: Discuss with Terry__).
+- Re: Autocorrelation analyses of the $\mathbf{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$.
+The interpretation of these analyses is complicated by the non-stationarity of $\mathbf{\beta}$, meaning that the strength of the dependence is affected not only by the distance between loci but also by their positions in the genome.
+- Comment on http://nar.oxfordjournals.org/content/40/22/11490.long
+- Not accounting for autocorrelation of errors, which manifest as autocorrelation of the residuals: "While it does not bias the OLS coefficient estimates, the standard errors tend to be underestimated (and the t-scores overestimated) when the autocorrelations of the errors at low lags are positive." [http://en.wikipedia.org/wiki/Autocorrelation](http://en.wikipedia.org/wiki/Autocorrelation)

### 2014-12-10 Rmarkdown/comethylation_review.Rmd
Revised co-methylation review.

@@
 To date, most analyses of co-methylation have been based on analysing correlations of aggregate levels of methylation, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
-An exception is the analysis in \citet{Lister:2009hy}.
-While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the authors instead look for patterns in the spacing of 'methylcytosines' along the genome[^Lister_methylcytosines].
-
-[^Lister_methylcytosines]: See Section \ref{sec:beta-value_interpretation} for a critique of this definition of 'methylcytosine'.
-
-In reviewing this previous work, I have been forced to make my best guess about some of the methods since they are often not sufficiently described in the original publication.
+These analyses have an implicit assumption of stationarity, that is, that the correlation between two loci depends only on the distance between them and not the actual position of the loci in the genome.
 
@@
 They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
-This is the first, and, as far as I am aware, only report of an organism in which the correlation of methylation levels first increases before decreasing.
+However, this analysis is fundamentally flawed.
 
-However, there is a something strange about this result, shown in Figure \ref{fig:Lyko}, namely, the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}).
+Figure \ref{fig:Lyko} reproduces the results for the co-methylation analysis.
+Note that the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}).
 I strongly suspect that the autocorrelation has been incorrectly computed.
-Specifically, what I suspect has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been assigned a $\beta$-value of zero.
+Specifically, what I believe has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been artificially assigned a $\beta$-value of zero.
+It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation.
 
@@
 
-While I don't have the data used in \citet{Lyko:2010dr}, I believe an example using the chromosome 22 data from the _FF_ sample demostrates the point.
-Let $\mathbf{\beta^{\ast}}$ be the augmented vector of $\beta$-values, where:
+While I don't have the data used in \citet{Lyko:2010dr}, I believe an example using the chromosome 22 data from the _FF_ sample demonstrates the point.
+Let $\bm{\beta^{\ast}}$ be the augmented vector of $\beta$-values, where:
 
@@
 
-__FIGURE__ is a plot of the autocorrelation of the augmented vector, $\mathbf{\beta^{\ast}}$.
+Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$.
+We see that it is remarkably similar to that panel (a) of Figure \ref{fig:Lyko} --- the correlation first increases then decreases and the maximum correlation is very small, close to zero.
 
-__TODO: Include plot of this for FF data on chr22.
-This brings into question the other results based on this plot, e.g., the 3bp periodicity, which may reflect nucleotide periodicities rather than methylation periodicities.__
-__TODO: What does this plot look like__
-__TODO: What are the implications for the results in Lyko__
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star.pdf}
+\caption{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \em{FF} sample.}
+\label{fig:FF_chr21_acf_beta_star}
+\end{figure}
+
+Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}).
+Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp periodicity is more likely due to the spacing of nucleotides in the genome than anything to do with their methylation state.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
+\caption{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \em{FF} sample.
+Fourier transformation computed and plotted using the \texttt{spectrum} function in R.
+The period is defined as 1 / frequency.
+The maximum frequency is approximately $0.33$, which corresponds to a period of approximately $3$ bp.}
+\label{fig:FF_chr21_acf_beta_star_fourier}
+\end{figure}
+
+All of this casts serious doubt on the validity and interpretation of the co-methylation results in \citet{Lyko:2010dr}.
 
@@
 Hence, while this may appear to be a within-fragment analysis of co-methylation with it's reference to "methylcytosines", it in fact  based on $\beta$-values.
-Furthermore, unlike the methods described in Section \ref{sec:previous_cor_of_agg_methylation}, the analysis is not based on a correlation of these $\beta$-values, but rather on unnormalised counts of a binary classification of $\beta$-values, which leads to problems.
+However, unlike the other methods described in this section, the analysis is not based on a correlation of these $\beta$-values, but rather on patterns in the spacing of 'methylcytosines' along the genome, which leads to problems.
 
@@
 
-[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).
+[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process.
 
@@
 
-__UP TO HERE__
+\subsection{\citet{Lacey:2009by}}\label{sec:lacey_2009}
 
-\subsection{\cite{Lacey:2009by}}\label{sec:lacey_2009}
-
-\cite{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans.
+\citet{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans.
 The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field where the methylation state of one loci, $Z_{h, i}$, depends on that of its neighbours either side, $Z_{h, i - 1}$ and $Z_{h, i + 1}$.
@@
 
-For each of these regions, \cite{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours.
+For each of these regions, \citet{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours.
 In other words, they found significant within-fragment co-methylation.
@@
 
-\subsection{\cite{Landan:2012kp}}\label{sec:landan}
+\subsection{\citet{Landan:2012kp}}\label{sec:landan}
 
-\cite{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.
+\citet{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.
 The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs.
 They use ultradeep (median coverage of each CpG in the target $> 10,000 \times$) bisulfite-sequencing of a 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
-They not only use this information to estimate co-methylation but also to estimate the rate of _epipolymorphisms_ and _epialleles_.
+They also use these data to estimate co-methylation.
 
@@
 
-[^cor_equivalence]: Since this is a correlation of two binary variables, it is equivalent to computing the _phi_ coefficient of each table (__TODO: Find a better source than Wikipedia and my own numerical calculations.
-The algebraic proof isn't hard, just tedious.__).
+[^cor_equivalence]: Since this is a correlation of two binary variables, it is equivalent to computing the _phi_ coefficient of each $2 \times 2$ table.
 
@@
 
-```{r}
-# Equivalence of Pearson correlations and Phi coefficient for binary table.
-set.seed(333)
-z <- array(rpois(n = 4, lambda = 20), dim = c(2, 2),
-           dimnames = list(c('M', 'U'), c('M', 'U')))
-n11 <- z[1, 1]
-n10 <- z[1, 2]
-n01 <- z[2, 1]
-n00 <- z[2, 2]
-n <- sum(z)
-n1dot <- n11 + n10
-ndot1 <- n11 + n01
-x <- c(rep(1, n11), rep(1, n10), rep(0, n01), rep(0, n00))
-y <- c(rep(1, n11), rep(0, n10), rep(1, n01), rep(0, n00))
-plot(x, y)
-cor(x, y)
-m1 <- (n11 + n10) / n
-m2 <- (n11 + n01) / n
-v1 <- m1 * (1 - m1)
-v2 <- m2 * (1 - m2)
-(n11 / n - m1 * m2) / sqrt(v1 * v2)
-all.equal(cor(x, y), (n11 / n - m1 * m2) / sqrt(v1 * v2))
-all.equal(cor(x, y),
-          (n11 * n00 - n10 * n01) /
-            sqrt(((n1dot * (n - n1dot) * ndot1 * (n - ndot1)))))
-all.equal(GenomicRanges::phicoef(z), cor(x, y))
-```
-
 These correlations are then plotted like a linkage disequilibrium heatmap to identify blocks of CpGs with significant co-methylation. From the analysis of this small set of regions they conclude that:
@@
 As is clear from the above, there are multiple definitions of within-sample co-methylation, each capturing variations on the idea of the correlation of methylation levels at pairs of loci within a single sample.
-An almost orthogonal concept is between-sample co-methylation, which is the correlation of methylation levels of a single pair of loci between a set of samples.
+A different concept is between-sample co-methylation, which is the correlation of methylation levels of a single pair of loci between a set of samples.
 
 Since it is a loci-specific, population-level correlation, between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci.
-Briefly, for two loci, $(i, i')$, the correlation of aggregate methylation levels is computed across samples using, for example, the Pearson correlation coefficient, $r = cor(\bm{\beta_{i}}, \bm{\beta_{i'}})$, where $\bm{\beta_{i}} = (\beta_{i, 1}, \ldots, \beta_{i, j}, \ldots, \beta_{i, n})$.
+Briefly, for two loci, $(i, i')$, the correlation of aggregate methylation levels is computed across samples using, for example, the Pearson correlation coefficient, $r = cor(\bm{\beta_{i}}, \bm{\beta_{i'}})$, where $\bm{\beta_{i}} = (\beta_{i, 1}, \ldots, \beta_{i, N_{loci}})$.
 
@@
 These tag-loci could act like tag-SNPs used in genome-wide association studies, which would allow researchers to assay a small subset of loci in the methylome and capture much of the methylation variability.
+This idea has been explored by \citet{Barrera:2012cq} with respect to the use of _HpaII_ sites as proxies for the methylation level of CpG islands.
+
 Unfortunately, unlike linkage disequilibrium, we cannot assume that between-sample co-methylation is identical across different tissues from the same individuals since DNA methylation, unlike DNA sequence, varies across tissues.
@@
 
-In order to exploit between-sample co-methylation in this way, we also need to know the size of these "co-methylation blocks", that is, the length of the regions over which between-sample co-methylation is sufficiently strong.
+In order to exploit between-sample co-methylation in this way, we also need to know the size of these co-methylation blocks, that is, the length of the regions over which between-sample co-methylation is sufficiently strong.
 The longer these blocks are, the more sparsely we can afford to sample individual methylomes.
 
-An example of between-sample co-methylation is given by \cite{Liu:dy}.
+An example of between-sample co-methylation is given by \citet{Liu:dy}.
 The authors compute the correlation of pairs of $\beta$-values across samples to identify clusters of correlated CpGs and relate methylation changes at these clusters to nearby SNP genotypes.
@@
 
-It has been difficult to properly account for co-methylation in downstream analyses due to the variety of definitions, and the varying rigour and reproducibility of the published analyses.
+It has been difficult to properly account for co-methylation in downstream analyses due to the variety of definitions, the varying rigour and reproducibility of the published analyses and the lack of software to estimate co-methylation..
 Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data.
@@
 
-For example, the `bsseq` R/Bioconductor package for identifying differentially methylated regions from WGBS data \cite{Hansen:2011gu, Hansen:2012gr} cite \cite{Eckhardt:2006gh} as evidence that methylation levels at nearby CpGs are correlated.
+For example, the `bsseq` R/Bioconductor package for identifying differentially methylated regions from WGBS data \citep{Hansen:2011gu, Hansen:2012gr} cite \citet{Eckhardt:2006gh} as evidence that methylation levels at nearby CpGs are correlated.
 `bsseq` uses local binomial likelihood smoothing to exploit the spatial correlation of $\beta$-values.
@@
 
-The R/Bioconductor package `BiSeq` \cite{Hebestreit:2013ko} also smooths the $\beta$-values, albeit slightly differently to `bsseq` since it is designed for identifying differentially methylated regions from RRBS data.
-The authors also cite \cite{Eckhardt:2006gh} as justification for smoothing the $\beta$-values.
+The R/Bioconductor package `BiSeq` \citep{Hebestreit:2013ko} also smooths the $\beta$-values, albeit slightly differently to `bsseq` since it is designed for identifying differentially methylated regions from RRBS data.
+The authors also cite \citet{Eckhardt:2006gh} as justification for smoothing the $\beta$-values.
 
@@
 Specifically, `BiSeq` fits a semivariogram to the test statistics of differentially methylated CpGs.
-This correlations estimated using the semivariogram are in turn used in estimates of the standard deviations when identifying differentially methylated regions.
+The correlations estimated using the semivariogram are in turn used in estimates of the standard deviations when identifying differentially methylated regions.
 
-More generally, methods such as `SLIM` \cite{Wang:2011cw} and `comb-p` \cite{Pedersen:2012vl} attempt to account for the correlation amongst test statistics in genomics data.
+More generally, methods such as `SLIM` \citep{Wang:2011cw} and `comb-p` \citep{Pedersen:2012vl} attempt to account for the correlation amongst test statistics in genomics data.
 They do this by analysing the distribution of P-values rather than the distribution of the test statistics themselves or the underlying data.
 This makes these methods quite general but more application-specific methods may be more powerful.
-`comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \cite{Pedersen:2012vl}.
+`comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \citep{Pedersen:2012vl}.
 
-`Aclust` is notable for its use of between-sample co-methylation in identifying differentially methylated regions.
+These methods exploiting correlations amongst the test statistics or their P-values are not directly based on co-methylation estimates.
+However, what drives the correlations amongst the test statistics is a complex mixture of within-sample and between-sample co-methylation.
+
+`Aclust` \citep{Sofer:2013bk} is notable for its use of between-sample co-methylation to identify differentially methylated regions.
 Unlike other methods for detecting DMRs, which test for DMCs and then cluster these DMCs into DMRs, `Aclust` clusters loci prior to testing for differential methylation.
@@
 
-\cite{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it.
+\citet{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it.
 They study the dynamics of DNA methylation in differentiating cell lineages.
 For each locus there is a correlation between its methylation state in the precursor cell and its dependent cell types, which they call "vertical" correlations in contrast to the "horizontal" correlations of within-sample co-methylation.
-In essence, \cite{Capra:2014hx} are bringing well-established methods for the study of DNA sequence evolution to the study of DNA methylation dynamics.
+\citet{Capra:2014hx} are bringing well-established methods for the study of DNA sequence evolution to the study of DNA methylation dynamics.
 In this specialised setting, where "between-samples" means "between samples in a lineage", they show that a method based on "vertical" correlations is better at imputing the methylation level of CpGs than a method based on "horizontal" within-sample co-methylation.
 
-__TODO: Are methods using correlations of P-values or test-statistics based on within- or between-sample co-methylation?__
+\section{Summary}
 
-## TODOs
+While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult.
+As I have shown, there are problems with most, if not all, the previous analyses, ranging from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation.
+Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.
 
-- An autocorrelation analysis implicitly assumes stationarity along the genome whereas between-sample co-methylation assumes stationarity across samples (__TODO: Discuss with Terry__).
-- Re: Autocorrelation analyses of the $\mathbf{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$.
-The interpretation of these analyses is complicated by the non-stationarity of $\mathbf{\beta}$, meaning that the strength of the dependence is affected not only by the distance between loci but also by their positions in the genome.
-- Comment on http://nar.oxfordjournals.org/content/40/22/11490.long
-- Not accounting for autocorrelation of errors, which manifest as autocorrelation of the residuals: "While it does not bias the OLS coefficient estimates, the standard errors tend to be underestimated (and the t-scores overestimated) when the autocorrelations of the errors at low lags are positive." [http://en.wikipedia.org/wiki/Autocorrelation](http://en.wikipedia.org/wiki/Autocorrelation)
+It is clearly inappropriate to treat methylation at nearby loci as independent and methods have been proposed to account for or exploit this in various ways.
+These methods would be better informed by a more rigorous and reproducible analysis of co-methylation, which I give in Chapter \ref{chap:co-methylation}.

### 2014-12-10 Rmarkdown/appendix.Rmd
Updated appendix

@@
 
-\section{Equivalence of tests to identify DMCs in a one-group experiment}\label{app:one_group_equivalence}
+\section{Chapter \ref{chap:wgbs_downstream_analyses}}
 
-\subsection{Mathematical equivalence}
+\subsection{Equivalence of tests to identify DMCs in a one-group experiment}\label{app:one_group_equivalence}
+
+\subsubsection{Mathematical equivalence}
 
@@
 
-\subsection{Computational equiavelence}
+\subsubsection{Computational equiavelence}
 
@@
 ```
+\section{Chapter \ref{chap:co-methylation_review}}\label{app:co-methylation_review}
 
-\section{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}
+\subsection{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}
 
@@
 
-\section{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
+\subection{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
 
@@
 
-```{r}
-devtools::session_info()
-```
+__TODO__
 
-### TODOs
-
-* Cite data source and publication
+- R version
+- `MethylationTuples`
+- `methsim`
+- `methtuples`
+- others

### 2014-12-10 Rmarkdown/wgbs_downstream_analyses.Rmd
Typos and latex errors
Updated appendix

@@
 
-\chapter{Downstream analyses of whole-genome bisulfite-sequencing data}
+\chapter{Downstream analyses of whole-genome bisulfite-sequencing data}\label{chap:wgbs_downstream_analyses}
 
@@
 Alternatively, we could frame this as a generalised linear model, either a log-linear model or a logistic regression.
-The equivalence of these methods is shown in the Appendix.
+The equivalence of these methods is shown in Appendex \ref{app:one_group_equivalence}.
 
@@
 There has been some progress in developing software to extract and analyse these epialleles, which I now describe.
+(Alternatively, cite Landan et al.)
 

### 2014-12-10 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Updated figures in wgbs_bioinformatics_analysis

@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
@@
 Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
-The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combination` flag is set[^nil0].
+The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combinations` flag is set[^nil0].
 
@@
 
-To illustrate several of the above-mentioned concepts, suppose we sequence a region of the genome containing five methylation loci with three paired-end reads (`A`, `B` and `C`):
+To illustrate several of the above-mentioned concepts, suppose we sequence a region of the genome containing five methylation loci with three paired-end reads (`A`, `B` and `C`), shown in Figure \ref{fig:methtuple_example1}.
 
-__TODO: Convert the following code chunks to figures in Illustrator__
+\begin{figure}[h!]
+\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
+\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
+\label{fig:methtuple_example1}
+\end{figure}
 
-```
-ref: 1    2   3 4 5
-A_1: |----->
-A_2:         <------|
-B_1:
-B_2: |----->   <----|
-C_1:    |----->
-C_2:      <------|
-```
-
-If we are interested in 1-tuples, then we would obtain the following from each read by running `methtuple`:
-
-```
-A: {1}, {2}, {3}, {4}, {5}
-B: {1}, {2}, {4}, {5}
-C: {2}, {3}, {4}
-```
-
+If we are interested in 1-tuples, Figure \ref{fig:methtuple_example2} shows what we would obtain from each read by running `methtuple`.
 The result is identical regardless of whether the `--all-combinations` flag is set.
 
-If we are interested in 3-tuples, then we would obtain the following from each read by running `methtuple` in its default mode:
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
+\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
+\label{fig:methtuple_example2}
+\end{figure}
 
-```
-A: {1, 2, 3}, {2, 3, 4}, {3, 4, 5}
-B: {1, 2, 4}, {2, 4, 5}
-C: {2, 3, 4}
-```
-
-Things to note:
+If we are interested in 3-tuples, Figure \ref{fig:methtuple_example3} shows what we would obtain from each read by running `methtuple` in its default mode. A few things to note:
 
 - Read-pair `A` sequences all three (= 5 - 3 + 1) adjacent 3-tuples
-- Read-pair `B` sequences none of the adjacent 3-tuples but does 'erroneously' construct two 3-tuples with $NIL > 0$.
+- Read-pair `B` sequences none of the adjacent 3-tuples but does 'erroneously' construct two 3-tuples from pairs of non-adjacent loci.
 This happens because m-tuples are created independently from each read-pair; effectively, read-pair `B` is unaware of methylation locus `3`.
-Depending on the downstream analysis, you may want to _post-hoc_ filter out these m-tuples with $NIL > 0$.
-- The twice-sequenced methylation loci, `2` and `3`, in read-pair `C` are not double counted.
+Depending on the downstream analysis, you may want to _post-hoc_ filter out these m-tuples with non-adjacent loci.
+- The twice-sequenced methylation loci in read-pair `C`, `2` and `3`, are not double counted.
 
-However, if we were to run `methtuple` with `--all-combinations` then we would obtain:
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
+\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
+\label{fig:methtuple_example3}
+\end{figure}
 
-```
-A: {1, 2, 3}, {2, 3, 4}, {3, 4, 5}, {1, 2, 4}, {1, 2, 5}, {1, 3, 4}, {1, 3, 5}, {1, 4, 5}, {2, 3, 5}, {2, 4, 5}
-B: {1, 2, 4}, {2, 4, 5}, {1, 2, 5}, {1, 4, 5}
-C: {2, 3, 4}
-```
+Finally, Figure \ref{fig:methtuple_example4} shows the output if we were to analyse 3-tuples but with the `--all-combinations` flag set.
 
-The number of CpGs per read for the Lister data is shown in Figure \ref{fig:Lister_cpgs_per_read}.
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
+\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
+\label{fig:methtuple_example4}
+\end{figure}
+
+With current sequencing technology we are limited to extracting m-tuples that span no more than $200-250$ bp.
+This obviously affects the size of m-tuples that we can study.
+Figure \ref{fig:Lister_CpGs_per_read} shows the number of CpGs per read for the Lister dataset (see Chapter \ref{chap:dataset} for a description of the Lister dataset).
 Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples.
@@
 
-```{r Lister_cpgs_per_read, dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data.", out.width = "0.8\\linewidth"}
-library(dplyr)
-library(ggplot2)
-library(scales)
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
-dataset <- "Lister"
-hist <- lapply(paste0("../data/observable_m-tuples/", dataset, "/",
-                      list.files(paste0("../data/observable_m-tuples/",
-                                        dataset), "\\.CG\\_per\\_read\\.hist")),
-               read.table, header = TRUE)
-sn <- gsub("\\.CG\\_per\\_read\\.hist", "",
-           list.files(paste0("../data/observable_m-tuples/",
-                             dataset), "\\.CG\\_per\\_read\\.hist"))
-sn <- gsub("DM\\_", "", sn)
-hist_df <- bind_rows(hist) %>%
-  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
-  group_by(Sample) %>%
-  mutate(Proportion = count / sum(as.numeric(count)))
-hist_df %>%
-  ggplot(aes(x = n, y = Proportion * 100)) +
-  geom_bar(stat = "identity") +
-  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
-  coord_cartesian(xlim = c(-1, 8.5)) +
-  scale_y_continuous(label = comma) +
-  xlab("Number of CpGs per read\n(Axis truncated at 8)") +
-  ylab("Percentage of total reads") +
-  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
-  thesis_theme
-```
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
+\caption{Number of CpGs per read for the Lister dataset.}
+\label{fig:Lister_CpGs_per_read}
+\end{figure}
 
@@
 Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
-\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing.
-Suspect you can't get actual methylation calls.} \\
+\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Suspect you can't get actual methylation calls.} \\
 \texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
@@
 
-The output format of `methtuple` is tab-delimited plain text, optionally compressed with `gzip` or `bzip2`. An example of the output is shown below for 3-tuples:
+The output format of `methtuple` is tab-delimited plain text, optionally compressed with `gzip` or `bzip2`. Figure \ref{fig:3-tuples} shows an example of the output for 3-tuples.
 
-```{sh}
-chr     strand  pos1    pos2    pos3    MMM     MMU     MUM     MUU     UMM     UMU     UUM     UUU
-chr1    +       469     471     484     0       0       4       1       1       0       0       0
-chr1    +       471     484     489     1       0       0       0       2       2       1       0
-
-```
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
+\caption{Example output of \texttt{methtuple} for 3-tuples.}
+\label{fig:3-tuples}
+\end{figure}
 

### 2014-12-10 Rmarkdown/comethylation_review.Rmd
Updated appendix
Latex fixes

@@
 
-__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__
-
-In general, $PIM^{\ast} \in [0, 1]$ since it is a probability, which is already unlike a correlation coefficient that takes values in $\[-1, 1\]$.
+In general, $PIM^{\ast} \in {[0, 1]}$ since it is a probability, which is already unlike a correlation coefficient that takes values in ${[{-1}, 1]}$.
 There are a few important special cases that are worth discussing:
@@
 \centering
-\includegraphics[width=0.8\textwidth]{../figures/journal.pbio.1000533.s006.eps}
+\includegraphics[width=0.8\textwidth]{../figures/{journal.pbio.1000533.s006}.pdf}
 \caption{Co-methylation patterns for different genomic features. A reproduction of Supplementary Figure 6 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{../figures/Figure_S7.eps}
+\includegraphics[width=0.8\textwidth]{../figures/Figure_S7.pdf}
 \caption{Transformation of methylation correlation at nearby CpG cytosines. A reproduction of Supplementary Figure 7 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
@@
 \citet{Lacey:2013iy} explore co-methylation in their paper on modelling and analysing reduced representation bisulfite-sequencing data (RRBS).
-To do so, \citeauthors{Lacey:2013iy} fit a Gaussian variogram model to the CpG $\beta$-values from chromosome 11 of a normal myotube cell line (_MTCTL2_).
+To do so, \citeauthor{Lacey:2013iy} fit a Gaussian variogram model to the CpG $\beta$-values from chromosome 11 of a normal myotube cell line (_MTCTL2_).
 The empirical and fitted variograms asymptote for CpGs separated by around $3000$ bp (Figure 1d of \citet{Lacey:2013iy}, unable to be reproduced due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
@@
 Each point is the mean value from averaging the probability from each of the five Arabidopsis chromosomes, and the blue line is a running average of these mean values.
-Error bars represent 95% confidence intervals via critical values of Student $t$ distributions.
+Error bars represent $95\%$ confidence intervals via critical values of Student $t$ distributions.
 Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Cokus:2008fc}), copyright (2008)}

### 2014-12-10 Rmarkdown/datasets.Rmd
Latex fixes

@@
 
-\begin{longtable}{@{}cccccc@{}}
+\begin{table}[h!]
+\centering
+\begin{tabulary}{\textwidth}{LLLLLL}
 \toprule
@@
 HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
+\end{tabulary}
 \caption{Samples in Lister dataset.
@@
 \label{tab:Lister_dataset}
-\end{longtable}
+\end{table}
 
@@
 
-\begin{longtable}{@{}cccc@{}}
+\begin{table}[h!]
+\centering
+\begin{tabulary}{\textwidth}{LLLL}
 \toprule
@@
 E23VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
+\end{tabulary}
 \caption{Samples in EPISCOPE dataset. The reported read lengths are prior to any trimming of the reads.}
 \label{tab:EPISCOPE_dataset}
-\end{longtable}
+\end{table}
 
@@
 The default alignment options were used, except that the maximum insert size for valid paired-end alignments was set to 1000 instead of 500 (`-X 1000`).
-An example command is shown below:
-
-```
-bismark -p 4 --bowtie2 -X 1000 --unmapped --ambiguous --gzip --bam -o SAMPLE_NAME REF_GENOME -1 READ_1.fq -2 READ_2.fq
-```
-
 Duplicate reads had already been removed from the `BAM` files that I received.
@@
 
-\begin{longtable}{@{}clcccl@{}}
+\begin{table}[h!]
+\centering
+\begin{tabulary}{\textwidth}{LLLLL}
 \toprule
-Sample            & Tissue                     & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
-J1\_1             & ESC                        & PE         & $100$ bp    & $12\times$    & \citet{Seisenberger:2012ko} \\
-E6.5\_epiblast\_1 & Epiblast                   & PE         & $101$ bp    & $13\times$    & \citet{Seisenberger:2012ko} \\
-E16.5\_male\_1    & Male progenitor germ cells & PE         & $150$ bp    & $12\times$    & \citet{Seisenberger:2012ko} \\ \bottomrule
-\caption{Samples in Seisenberger dataset. The reported read lengths are prior to any trimming of the reads.}
+Sample            & Tissue                     & Sequencing & Read length & Ave. coverage             \\ \midrule
+J1\_1             & ESC                        & PE         & $100$ bp    & $12\times$     \\
+E6.5\_epiblast\_1 & Epiblast                   & PE         & $101$ bp    & $13\times$     \\
+E16.5\_male\_1    & Male progenitor germ cells & PE         & $150$ bp    & $12\times$     \\ \bottomrule
+\end{tabulary}
+\caption{Samples in Seisenberger dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Seisenberger:2012ko}.}
 \label{tab:Seisenberger_dataset}
-\end{longtable}
+\end{table}
 
@@
 
-Table \ref{tab:Ziller_dataset_sequencing_runs} summarises the data for the 19 sequencing runs and __TABLE__ summarises the data for the _post hoc_ merged data.
+Table \ref{tab:Ziller_dataset_sequencing_runs} summarises the data for the 19 individual sequencing runs[^Ziller_merged_ave_cov].
 
-\begin{longtable}{@{}cccccc@{}}
+[^Ziller_merged_ave_cov]: The average sequencing coverage of the _post-hoc_ merged samples are approximately the sums of the average sequencing coverage for the corresponding individual sequencing runs.
+
+\begin{table}[h!]
+\centering
+\begin{tabulary}{\textwidth}{LLLLL}
 \toprule
-Sample    & Tissue                     & Sequencing & Read length & Ave. coverage & First published       \\ \midrule
-SRR949193 & Frontal\_cortex\_normal\_1 & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
-SRR949194 & Frontal\_cortex\_normal\_1 & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
-SRR949195 & Frontal\_cortex\_normal\_1 & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
-SRR949196 & Frontal\_cortex\_normal\_2 & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
-SRR949197 & Frontal\_cortex\_normal\_2 & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
-SRR949198 & Frontal\_cortex\_normal\_2 & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
-SRR949199 & Frontal\_cortex\_AD\_1     & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
-SRR949201 & Frontal\_cortex\_AD\_1     & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
-SRR949202 & Frontal\_cortex\_AD\_2     & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
-SRR949203 & Frontal\_cortex\_AD\_2     & PE         & 101         & $10\times$    & \citet{Ziller:2013fs} \\
-SRR949206 & HepG2\_cell\_line          & PE         & 101         & $2\times$     & \citet{Ziller:2013fs} \\
-SRR949207 & HepG2\_cell\_line          & PE         & 101         & $1\times$     & \citet{Ziller:2013fs} \\
-SRR949208 & IMR90\_cell\_line          & PE         & 101         & $1\times$     & \citet{Ziller:2013fs} \\
-SRR949209 & IMR90\_cell\_line          & PE         & 101         & $3\times$     & \citet{Ziller:2013fs} \\
-SRR949210 & Colon\_Tumor\_Primary      & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\
-SRR949211 & Colon\_Tumor\_Primary      & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\
-SRR949212 & Colon\_Tumor\_Primary      & PE         & 101         & $9\times$     & \citet{Ziller:2013fs} \\
-SRR949213 & Colon\_Tumor\_Primary      & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\
-SRR949215 & Colon\_Primary\_Normal     & PE         & 101         & $8\times$     & \citet{Ziller:2013fs} \\ \bottomrule
-\caption{Sequencing runs in Ziller dataset. The reported read lengths are prior to any trimming of the reads.}
+Sample    & Tissue                     & Sequencing & Read length & Ave. coverage        \\ \midrule
+SRR949193 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$ \\
+SRR949194 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$    \\
+SRR949195 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$    \\
+SRR949196 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
+SRR949197 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
+SRR949198 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
+SRR949199 & Frontal\_cortex\_AD\_1     & PE         & 101 bp         & $9\times$     \\
+SRR949201 & Frontal\_cortex\_AD\_1     & PE         & 101 bp         & $9\times$     \\
+SRR949202 & Frontal\_cortex\_AD\_2     & PE         & 101 bp         & $10\times$    \\
+SRR949203 & Frontal\_cortex\_AD\_2     & PE         & 101 bp         & $10\times$    \\
+SRR949206 & HepG2\_cell\_line          & PE         & 101 bp         & $2\times$     \\
+SRR949207 & HepG2\_cell\_line          & PE         & 101 bp         & $1\times$     \\
+SRR949208 & IMR90\_cell\_line          & PE         & 101 bp         & $1\times$     \\
+SRR949209 & IMR90\_cell\_line          & PE         & 101 bp         & $3\times$     \\
+SRR949210 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $8\times$     \\
+SRR949211 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $8\times$     \\
+SRR949212 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $9\times$     \\
+SRR949213 & Colon\_Tumor\_Primary      & PE         & 101 bp        & $8\times$     \\
+SRR949215 & Colon\_Primary\_Normal     & PE         & 101 bp         & $8\times$     \\ \bottomrule
+\end{tabulary}
+\caption{Sequencing runs in Ziller dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Ziller:2013fs}.}
 \label{tab:Ziller_dataset_sequencing_runs}
-\end{longtable}
-
-__TODO: Table after post-hoc merge__
+\end{table}
 

### 2014-12-10 Rmarkdown/wgbs_statistical_framework.Rmd
Typos and latex errors
Updated appendix
Latex fixes

@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
@@
   \begin{array}{l l}
-  (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus on the } h^{th} \text{ fragment}\\
-  (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th}) \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus on the } h^{th} \text{ fragment} \\
+  (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
+  & \quad \text{ locus on the } h^{th} \text{ fragment}\\
+  (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big ) \\
+  & \quad \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus on the } h^{th} \text{ fragment} \\
   \vdots \\
-  (1, 1, \ldots, 1) & \quad \text{if methylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus on the } h^{th} \text{ fragment}
+  (1, 1, \ldots, 1) & \quad \text{if methylated at the } \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
+  & \quad \text{ locus on the } h^{th} \text{ fragment}
   \end{array} \right.
@@
   \begin{array}{l l}
-    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus}\\
-    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th}) \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
+    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \\
+    & \quad \Big ( i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big) \text{ locus}\\
+    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \\
+    \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big) \text{ locus} \\
+      & \quad \text{and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
     \vdots \\
-    (1, 1, \ldots, 1) & \quad \text{if methylated at the } i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th}) \text{ locus}
+    (1, 1, \ldots, 1) & \quad \text{if methylated at the } \\
+    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \text{ locus}
       \end{array} \right.
@@
 
-- TODO: Lots of figures and discussion. See chapter plan.
+__TODO: Update discussion of results in light of Ziller merged data__
 
@@
 
-Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity, fig:Lister_plotBetaDensity, fig:Seisenberger_plotBetaDensity, fig:Zilelr_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively.
-Figures \ref{fig:EPISCOPE_plotBetaFreqPoly, fig:Lister_plotBetaFreqPoly, fig:Seisenberger_plotBetaFreqPoly, fig:Zilelr_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
+Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively.
+Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_merged_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
@@
 
-\begin{figure}[htbp]
-\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaDensity.pdf}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
 \caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
@@
 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
-\label{fig:Ziller_plotBetaDensity}
+\label{fig:Ziller_merged_plotBetaDensity}
 \end{figure}
 
-\begin{figure}[htbp]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[htbp]
-\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaFreqPoly.pdf}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
 \caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
@@
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
-\label{fig:Ziller_plotBetaFreqPoly}
+\label{fig:Ziller_merged_plotBetaFreqPoly}
 \end{figure}
@@
 The bimodality of the genome-wide $\beta$-value distributions is driven by the fact that most CpGs is CpG islands are unmethylated whereas those outside of the islands are mostly methylated.
-Figures \ref{fig:EPISCOPE_plotBetaDensity_CGI, fig:Lister_plotBetaDensity_CGI, fig:Seisenberger_plotBetaDensity_CGI, fig:Ziller_plotBetaDensity_CGI} show the kernel density plots of the $\beta$-value distributions stratified by CpG island status for the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
+Figures \ref{fig:EPISCOPE_plotBetaDensity_CGI}, \ref{fig:Lister_plotBetaDensity_CGI}, \ref{fig:Seisenberger_plotBetaDensity_CGI} and \ref{fig:Ziller_merged_plotBetaDensity_CGI} show the kernel density plots of the $\beta$-value distributions stratified by CpG island status for the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
 These distributions are normalised so that each density has a maximum value of 1.
@@
 Because these densities are normalised, these plots don't show the proportion of CpGs in CpG islands.
-Therefore, Figures \ref{fig:EPISCOPE_plotBetaFreqPoly_CGI, fig:Lister_plotBetaFreqPoly_CGI, fig:Seisenberger_plotBetaFreqPoly_CGI, fig:Ziller_plotBetaFreqPoly_CGI} show the same data but with $\beta$-values grouped into $0.01$-width bins and plotted against the percentage of total CpGs in each bin.
+Therefore, Figures \ref{fig:EPISCOPE_plotBetaFreqPoly_CGI}, \ref{fig:Lister_plotBetaFreqPoly_CGI}, \ref{fig:Seisenberger_plotBetaFreqPoly_CGI} and \ref{fig:Ziller_merged_plotBetaFreqPoly_CGI} show the same data but with $\beta$-values grouped into $0.01$-width bins and plotted against the percentage of total CpGs in each bin.
 These plots highlight that the majority of unmethylated CpGs occur in CpG islands and that most of the intermediate methylation occurs outside of CpG islands, owing to most CpGs being outside of an island.
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaDensity_CGI.pdf}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
 \caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data, stratified by whether the CpG is in a CpG island.
@@
 Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
-\label{fig:Ziller_plotBetaDensity_CGI}
+\label{fig:Ziller_merged_plotBetaDensity_CGI}
 \end{figure}
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaFreqPoly_CGI.pdf}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
 \caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Zilelr data, stratified by whether the CpG is in a CpG island.
@@
 Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
-\label{fig:Ziller_plotBetaFreqPoly_CGI}
+\label{fig:Ziller_merged_plotBetaFreqPoly_CGI}
 \end{figure}
@@
 To investigate the validity of this assumption we can compute the correlation of strand-specific $\beta$-values.
-Tables \ref{tab:EPISCOPE_strand_correlations, tab:Lister_strand_correlations, tab:Seisenberger_strand_correlations, tab:Ziller_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
+Tables \ref{tab:EPISCOPE_strand_correlations}, \ref{tab:Lister_strand_correlations}, \ref{tab:Seisenberger_strand_correlations}, \ref{tab:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
 
@@
 
-```{r Ziller_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
+```{r Ziller_merged_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+x <- read.table("../data/strand_correlation/Ziller_merged_strand_correlations.txt",
 stringsAsFactors = FALSE)
@@
 xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Ziller dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_strand_correlations")
+Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_merged_strand_correlations")
 align(xt) <- rep("c", ncol(x) + 1)
@@
 ```
-\subsection{Interpretation}
+
+\subsection{Interpretation}\label{sec:beta-value_interpretation}
 

### 2014-12-10 latex/phd_thesis.tex
Minor fixes
Removed Avy chapter from thesis.
Split wgbs_statistical_analysis into wgbs_statistical_framework and wgbs_downstream_analyses
Added tabulary support
Latex fixes

@@
 
-\usepackage{booktabs}
+\usepackage{tabulary}
 
@@
 
-\include{wgbs_statistical_analysis} % wgbs_statistical_analysis.tex
-
 \include{datasets} % datasets.tex
 
-\include{Avy_epialleles} % Avy_epialleles.tex
+\include{wgbs_statistical_framework} % wgbs_statistical_framework.tex
+
+\include{wgbs_downstream_analyses} % wgbs_downstream_analyses.tex
 
@@
 
-\chapter{Concluding Comments}
-
-\label{chapter:Conclusions}
+\chapter{Concluding Comments}\label{chapter:Conclusions}
 
 % 2 - 4 pages
+\textbf{TODO}
 
-In this thesis, several fundamental low-level analysis issues for microarrays have been considered.
+\include{appendix} % appendix.tex
 
@@
 
-\include{appendix} % appendix.tex
-
-\label{chapter:AppendixCode}
-
-Insert supplementary figures/R code here.
-
 \end{document}

### 2014-12-10 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits to wgbs_bioinformatics_analysis

@@
 
-This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges.
+This chapter and Chapter \ref{chap:wgbs_downstream_analyses} describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges.
 I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay.
@@
 There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis.
-This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:downstream_analyses}.
+This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:wgbs_downstream_analyses}.
 
@@
 
-Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \citep{Li:2009ka}.
+Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or its binary equivalent, `BAM` \citep{Li:2009ka}.
 However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way.
@@
 However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful.
-`Bismark` \citep{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).
+`Bismark` \citep{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1`, released in April 2014).
 
@@
 It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair.
-"Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15" ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).
+"Best" is defined as the read pair having the highest sum of base qualities as bases with $Q >= 15$" ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).
 
@@
 
-If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
+If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should probably be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
 For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs.
@@
 
-I find the first strategy conceptually simpler, and easier to program, particularly when I start dealing with m-tuples, and so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
+I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
 In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.
@@
 Suppose M-bias is highly correlated with some other quality metric, such as base quality, so that positions with M-bias also have low base quality[^bq_mbias].
-If you already intend to ignore all positions with a base quality less than 20 in your methylation calling then it makes sense to also ignore these positions when computing M-bias; otherwise you will overestimate the effect of M-bias and unnecessarily exclude read positions in your methylation calling.
-Unfortunately, the software I use for estimating M-bias, `bismark_methylation_extractor`, does not allow the user to exclude certain reads or read-positions from its calculations.
+If you already intend to ignore all positions with a base quality less than 20 in your methylation calling then it makes sense to also ignore these positions when computing M-bias, otherwise you will overestimate the effect of M-bias and unnecessarily exclude read positions in your methylation calling.
+Unfortunately, the software I use for estimating M-bias, `bismark_methylation_extractor` \cite{Krueger:2011eb}, does not allow the user to exclude certain reads or read-positions from its calculations.
 
@@
 It is rather a subjective decision to make.
-Two problems that I found were in maintaining consistency across samples and in determining where to draw the line when there is no dramatic 'spike' in the M-bias.
+Two problems that I found were:
+
+1. Maintaining consistency across samples
+2. Determining where to draw the line when there is no dramatic 'spike' in the M-bias.
+
 For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both read 1 and read 2, as well as a big problem at read-position 101 in read 2.
@@
 To do this, I use a simple normalisation of the read-position methylation levels ($rpml$).
-Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels (for example, $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read 1_).
+Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels, i.e., $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read 1_.
 
@@
 While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types.
-I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $nrpml_{CpG} < 0.72$ or $nrpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias.
+I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $rpml_{CpG} < 0.72$ or $rpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias.
 It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types.
@@
 Typically, read-positions showing evidence of M-bias are excluded when calling methylation events.
-In fact, a slightly cruder procedure is often used whereby all read-positions preceding or following those with M-bias are excluded.
+In fact, a slightly cruder procedure is often used whereby the entire ends of reads are removed.
+For example, for a $100$ bp read with M-bias observed at positions $1-4, 9, 94, 96-100$, then positions $1-9$ and $94-100$ would be ignored when methylation calling.
 Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method.
-Generally, this is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads.
 
+Generally, this strategy is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads.
 However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude.
 Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101.
-Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
+Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
+My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded.
 
+In the datasets I have analysed I've had to ignore up to 30 read-positions per read due to M-bias.
 Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read positions entirely.
-In the datasets I have analysed, I've had to ignore up to 30 read-positions per read due to M-bias.
-
-One idea is to weight methylation calls by the level of M-bias.
+One idea is to inversely weight methylation calls by the level of M-bias.
 A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability.
@@
 There are several other sources of bias when making methylation calls.
-Some are what I call technical biases - those driven by the sequencing protocol and processing of the data - while others I call biological biases - those driven by measuring something other than what you thought you were measuring.
+Some are what I call technical biases - those driven by the sequencing protocol and processing of the data --- while others I call biological biases - those driven by measuring something other than what you thought you were measuring.
 
@@
 However, whole blood contains a mixture of cell types, each of which has a distinct methylation profile.
-\citet{Houseman:2012km} exploited this fact to estimate the cell composition of whole blood samples, in particular the differences in white blood cell (leukocyte) proportions between different sub-populations (e.g.
-cases and controls).
+\citet{Houseman:2012km} exploited this fact to estimate the cell composition of whole blood samples, in particular the differences in white blood cell (leukocyte) proportions between different groups, e.g., cases and controls.
 
@@
 The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci at which to call methylation events.
-It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina ('Lister-style') whole-genome bisulfite-sequencing protocol.
+It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina whole-genome bisulfite-sequencing protocol.
 
-Certain variants, in particular (heterozygous) $C>T$ SNPs, are more difficult to accurately genotype than others.
+Certain variants, in particular heterozygous $C>T$ SNPs, are more difficult to accurately genotype than others.
 Unfortunately, $C>T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants.
-Fortunately, it is possible to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand - if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
+Fortunately, it is possible to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand --- if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
 Other base substitutions are more readily detected, and insertion and deletion events (indels) may also be called.
@@
 \begin{figure}[h!]
+\centering
+\centering
 \includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
@@
 \end{tabular}
-\caption{Example of output for methylation calling at 1-tuples.}
+\caption{Example of output for methylation calling at 1-tuples. Most CpGs will have a measurement from both the positive and negative strands, but not always.}
 \label{tab:1-tuples}
@@
 \begin{figure}[h!]
+\centering
+\centering
 \includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
@@
 \begin{figure}[h!]
+\centering
 \includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
@@
 \begin{figure}[h!]
+\centering
 \includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
@@
 \begin{figure}[h!]
+\centering
 \includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
@@
 \begin{figure}[h!]
+\centering
 \includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
@@
 \begin{figure}[h!]
+\centering
 \includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
@@
 I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples from the EPISCOPE, Lister, Seisenberger and Ziller datasets.
-To improve the performance I process split each sample by chromosome and process each chromosome in parallel.
+To improve the performance I  split each sample by chromosome and process each chromosome in parallel.
 This is implemented in a helper script ([https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh](https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh)) that makes extensive use of `GNU parallel` \citep{Tange:2011ty}.

### 2014-12-10 Rmarkdown/biology_background.Rmd
Edits to biology_background

@@
 
-Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following \citep{Bird:2007em}:
+Sir Adrian Bird, an esteemed British geneticist, and someone who is in favour of the broader definition, attempts to unite these definitions in the following \citep{Bird:2007em}:
 
@@
 It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying].
-It therefore fits into Arthur Riggs' afore-mentioned definition of an epigenetic modification \citep[p1]{russo1996epigenetic}.
+It therefore fits into Arthur Riggs' aforementioned definition of an epigenetic modification \citep[p1]{russo1996epigenetic}.
 DNA methylation is found in bacteria, fungi, plants and animals.
@@
 
-1. _Enzyme digestion_
-2. _Affinity enrichment_
-3. _Sodium bisulfite conversion_
+1. Enzyme digestion
+2. Affinity enrichment
+3. Sodium bisulfite conversion
 
@@
 
-Almost all assays of DNA methylation measure a "population average" from a pool of hundreds or thousands of cells.
+Almost all assays of DNA methylation measure a population average from a pool of hundreds or thousands of cells.
 For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay.
@@
 Some examples of these targeted bisulfite-sequencing are:
+
 - Reduced representation bisulfite-sequencing (_RRBS_, \citet{Meissner:2005eb}): RRBS uses restriction enzymes to first select regions of the genome with a high CpG density (based on the Msp-I cleavage motif), which are subsequently treated with sodium-bisulfite and sequenced.

### 2014-12-10 Rmarkdown/datasets.Rmd
Editing

@@
 M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set.
-These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an average CpG $\beta$-value more than $0.03$ away from the median were excluded from methylation calling (read 1 and read 2 analysed separately).
+These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level more than $0.03$ from the median, i.e., $|$nrplml_{CpG}$ - median(rplm_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
 3.
@@
 4.
-A `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples, 3-tuples and 4-tuples.
-I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this large sizes to be generally useful.
+A `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples (2-tuples with `--all-combinations`), 3-tuples and 4-tuples.
+I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.
 
@@
 
-The _Lister data_ refers to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
+The _Lister dataset_ refers to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
 The Lister data were the largest publicly available human WGBS datasets until quite recently.
@@
 
-- BUF: Buffy coat layer, which are leukocytes and platelets derived by centrifugation of a whole blood sample.
-- SA: Subcutaneous adidose tissue, which is fat found just below the skin.
+- _BUF_: Buffy coat layer, which are leukocytes and platelets derived by centrifugation of a whole blood sample.
+- _SA_: Subcutaneous adidose tissue, which is fat found just below the skin.
 Unlike visceral fat, "subcutaneous fat is not related to many of the classic obesity-related pathologies, such as heart disease, cancer, and stroke" (__TODO: Get better reference than wikipedia__).
-- VA: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs (stomach, liver, intestines, kidneys, etc.) (__TODO: Get better reference than wikipedia__).
-- VAT: Visceral adipocytes, which are derived from the VA tissue.
+- _VA_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs (stomach, liver, intestines, kidneys, etc.) (__TODO: Get better reference than wikipedia__).
+- _VAT_: Visceral adipocytes, which are derived from the VA tissue.
 
@@
 The sequencing data for these 12 samples were processed and aligned by Aaron Statham (Garvan Institute of Medical Research).
-Each sample was aligned to the human reference genome (hg19) using Bismark (v0.8.3) with the Bowtie2 backend.
+Each sample was aligned to the human reference genome (hg19) using `Bismark` (`v0.8.3`) with the `Bowtie2` backend.
 The default alignment options were used, except that the maximum insert size for valid paired-end alignments was set to 1000 instead of 500 (`-X 1000`).
@@
 The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \cite{Seisenberger:2012ko}.
-These were a convenience sample provided to me by a colleague.
-I thank Felix Krueger (Babraham Institute) who processed and aligned the reads, and sent me the `BAM` files.
+These were a convenience sample provided to me by a colleague, Felix Krueger (Babraham Institude).
+I thank Felix who send my the `BAM` files containing processed and aligned reads.
 
@@
 I have the data for only three samples from the original publication, detailed in Table \ref{tab:Seisenberger_dataset}.
-The _J1\_1_ sample is from an embryonic stem cell line while both the _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ samples are from a pool of $10-30$ embryos.
+The _J1\_1_ sample is from an embryonic stem cell line while both the _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ samples are derived from pools of $10-30$ embryos.
 Developmentally, the samples are ordered _J1\_1_ (embryonic stem cell), _E6.5\_epiblast\_1_ (embryonic day 6.5 epiblast) and _E16.5\_male\_1_ (embryonic day 16.5 male progenitor germ cells).
@@
 I believe that the samples I received labelled _J1\_1_ and _E16.5\_male\_1_  in fact correspond to _J1\_2_ and _E16.5\_male\_2_, respectively, i.e., the second  replicate rather than the first.
-The data I received are all $100$ bp paired-end sequences, which matches replicate 2 rather than replicate 1 for both of these samples \citet[Supplementary Table 1 of]{Seisenberger:2012ko}.
+The data I received are all $100$ bp paired-end sequences, which matches replicate 2 rather than replicate 1 for both of these samples \citet[Supplementary Table 1]{Seisenberger:2012ko}.
 
@@
 The sequencing data for these 3 samples were processed and aligned by Felix Krueger (Babraham Institute).
-Each sample was aligned to the mouse reference genome (GRCm38/mm10) using Bismark (v0.7.12) with the Bowtie1 backend.
+Each sample was aligned to the mouse reference genome (GRCm38/mm10) using `Bismark` (`v0.7.12`) with the `Bowtie1` backend.
 The default alignment options were used.
@@
 These libraries in fact represent 19 sequencing runs of eight biological samples, which I was not aware of until the data had already been processed.
-Therefore, I have _post hoc_ merged these sequencing runs.
+Therefore, I have _post hoc_ merged the processed data from sequencing runs.
 This unfortunately reduces the power to detect genetic variants since the coverage of individual sequencing runs is lower than merged data, but does not adversely affect methylation calling since the number of methylated and unmethylated reads can be summed across sequencing runs.
 
-The eight biological samples are as follows: frontal cortex from two 'normal' women donors (_Frontal\_cortex\_normal\_1_ and Frontal\_cortex\_normal\_2_) and from two women who had Alzheimer's disease (_Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_); a sample from a human liver carcinoma cell line (HepG2\_cell\_line_); a new sample from the _IMR90_ lung fibroblast cell line (_IMR90\_cell\_line_); and samples from a colon cancer matched tumour-normal pair (_Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_).
+The eight biological samples are as follows: frontal cortex from two 'normal' women donors (_Frontal\_cortex\_normal\_1_ and _Frontal\_cortex\_normal\_2_) and from two women who had Alzheimer's disease (_Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_); a sample from a human liver carcinoma cell line (_HepG2\_cell\_line_); a new sample from the _IMR90_ lung fibroblast cell line (_IMR90\_cell\_line_); and samples from a colon cancer matched tumour-normal pair (_Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_).
 

### 2014-12-10 Rmarkdown/wgbs_statistical_framework.Rmd
Editing

@@
 I then extend this framework to multiple samples.
-Finally, I describe key variables, common estimators of these and their statistical properties by analysing 51 whole-genome bisulfite-sequencing samples.
+
+Finally, I describe key variables, common estimators of these and their statistical properties by analysing __HOW MANY__ whole-genome bisulfite-sequencing samples.
 
@@
 \begin{figure}[h!]
+\centering
 \includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
@@
 A methylation locus is a single cytosine, that is, a CpG, CHG or CHH.
-The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., chr1:+:666.
+The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., `chr1:+:666`.
 Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome.
@@
 
-$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$ and $\mathcal{R}_{(i, i + 1, i + m - 1)}$ denotes the set of all reads containing the m-tuple $(i, i + 1, i + m - 1)$.
+$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$.
 
@@
 1.
-There may be gaps due to the insert size being longer than the sum of the read lengths, e.g.
-$(i, i + 1, i + 3, i + 4))$.
-In effect, we have missing data for any intervening methylation loci; the $(i + 2)^{th}$ loci in this example.
+There may be gaps due to the insert size being longer than the sum of the read lengths, e.g., $(i, i + 1, i + 3, i + 4))$.
+In effect, we have missing data for any intervening methylation loci, the $(i + 2)^{th}$ loci in this example.
 2.
 Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g.
-read_1 gives us $(i, i + 1)$ and read_2 gives us $(i + 1, i + 2, i + 3)$.
-In this example we must use only one of read_1 or read_2 as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
+_read 1_ gives us $(i, i + 1)$ and _read 2_ gives us $(i + 1, i + 2, i + 3)$.
+In this example we must use only one of _read 1_ or _read 2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
 
@@
     (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \\
-    & \quad \Big ( i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big) \text{ locus}\\
+    & \quad \Big ( i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big) \\
+    & \quad \text{ locus}\\
     (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \\
-    \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big) \text{ locus} \\
+    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big) \\
+    & \quad \text{ locus} \\
       & \quad \text{and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
@@
     (1, 1, \ldots, 1) & \quad \text{if methylated at the } \\
-    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \text{ locus}
+    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
+    & \quad \text{ locus}
       \end{array} \right.
@@
 
-Note that we do not know from which haplotype ($h$) each read came from, only that all methylation loci in the read came from the same DNA fragment.
+Note that we do not know from which haplotype, $h$, each read came from, only that all methylation loci in the read came from the same DNA fragment.
 
@@
 This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between tumour and normal tissue.
-Conversely, the aim of the experiment might be to _discover_ this grouping, such as in a clustering analysis.
+Alternatively, the aim of the experiment might be to _discover_ this grouping, such as in a clustering analysis.
 
@@
 
-In addition to the complications of the Section \ref{sec:one_sample_complications}, we now have sample-to-sample variation that must be addressed by this framework.
+In addition to the complications of the Section \ref{sec:one_sample_complications}, we now have sample-to-sample variation that must be addressed within this framework.
 
@@
 
-The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
+The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,][]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
 The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).
@@
 Recently, more sophisticated methods have been proposed to estimate or model the average methylation level.
-These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
+These methods, which are still based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
 
@@
 Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$.
-Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation} and \ref{chap:co-methylation_review}).
+Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
 
@@
 Both `bsseq` and `BiSeq` use a binomial local likelihood smoother.
-In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$.
+In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$.
 The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.
@@
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
-'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:EPISCOPE_plotBetaDensity}
@@
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
-'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Lister_plotBetaDensity}
@@
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
-'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Seisenberger_plotBetaDensity}
@@
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
-'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Ziller_merged_plotBetaDensity}
@@
 The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated.
-Theseplots also show that most of the intermediate methylation occurs outside of the CpG islands.
+These plots also show that most of the intermediate methylation occurs outside of the CpG islands.
 
@@
 To investigate the validity of this assumption we can compute the correlation of strand-specific $\beta$-values.
-Tables \ref{tab:EPISCOPE_strand_correlations}, \ref{tab:Lister_strand_correlations}, \ref{tab:Seisenberger_strand_correlations}, \ref{tab:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
+Tables \ref{tab:EPISCOPE_strand_correlations}, \ref{tab:Lister_strand_correlations}, \ref{tab:Seisenberger_strand_correlations} and \ref{tab:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
 
@@
 
-The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$.
+Some of the embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$.
 The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$).
@@
 In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
-However, as can be seen from Figures \ref{EPISCOPE_plotBetaFreqPoly, Lister_plotBetaFreqPoly, Seisenberger_plotBetaFreqPoly, Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
+However, as can be seen from Figures \ref{EPISCOPE_plotBetaFreqPoly}, \ref{Lister_plotBetaFreqPoly}, \ref{Seisenberger_plotBetaFreqPoly} and \ref{Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
 To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
@@
 
-\citet{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material).
-In the language of these papers, a "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}.
+\citeauthor{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" \citep[Suppmenetary Material]{Lister:2008bh}.
+In the language of these papers, a "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[Supplementary Material]{Lister:2009hy}.
 This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$.
@@
 Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows.
-For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
-Any site with an FDR-adjusted P-value below a threshold was declared a 'methylcytosine'[^fdr]
+For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Pr(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
+Any site with an FDR-adjusted P-value below a threshold was declared a "methylcytosine"[^fdr]
 
@@
 
-The proportion of cytosines that are identified by this procedure as 'methylcytosines' is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous.
+The proportion of cytosines that are identified by this procedure as "methylcytosines" is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous.
 For example, suppose we had a sample where the true methylation level of every CpG, $B_{i}$, was $0.2$.
-Given sufficient sequencing coverage, every CpG in the genome would be declared a 'methylcytosine' when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.
+Given sufficient sequencing coverage, every CpG in the genome would be declared a "methylcytosine" when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.
 
 Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values.
-For example, in \citet[supplementary figure 2a of]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates.
+For example, in \citet[Supplementary Figure 2a of][]{Lister:2009hy} the authors use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
 A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.
@@
 The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$.
-Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
+Taking the derivative of this, we see that the maximum standard error, $\sqrt{\frac{0.25}{m + u}}$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
 2.
@@
 Essentially, we need to also know the sequencing coverage of the methylation loci.
-Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100$ $u = 300$.
+Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100, u = 300$.
 Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision.
@@
 One transformation that remains popular, at least in the analysis of DNA methylation microarray data, is the logit-transformation, also known as $\mathcal{M}$-values.
-An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2}(\frac{\beta}{1 - \beta} = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero.
+An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2} \Big ( \frac{\beta}{1 - \beta} \Big ) = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero.
 Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \citep[e.g., ]{Smyth:2005ta}.

### 2014-12-11 Rmarkdown/biology_background.Rmd
CHARM

@@
 They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \citep{Laird:2010iv}.
-Restriction enzyme assays have a relatively low resolution.
+One such genome-wide assays is CHARM (_comprehensive high-throughput arrays for relative methylation_, \citet{Irizarry:2008hg}).
+CHARM combines a methylation fractionation step --- be it _MeDIP_, _HpaII_, or as in the original publication, _McrBC_ --- with a tiling array and analysis techniques that leverage regional DNA methylation levels.
+
+While restriction enzyme assays do not typically provide single nucleotide resolution data, `methylCRF` \citep{Stevens:2013hv} is able to infer single CpG methylation levels by combining data from a restriction enzyme based assay (MRE-seq) with one based on affinity enrichment (MeDIP-seq) in a sophisticated statistical analysis.
 

### 2014-12-11 Rmarkdown/comethylation_review.Rmd
Editing

@@
 This broad definition encompasses several concepts that have variously been described as 'co-methylation', or similar.
-My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}) or "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; it also includes the "correlation of two [loci] across many sample" \citep{Akulenko:2013fl} and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
+My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha} or "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; it also includes the "correlation of two [loci] across many sample" \citep{Akulenko:2013fl} and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
 A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.
@@
 Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample.
-Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}.
+Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}.
 Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.
 
-The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__.
+The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation_.
 At a higher level is the _correlation of aggregate methylation levels_, such as the spatial correlation of $\beta$-values within a sample.
@@
 
-\subsection{\citet{Chodavarapu:2010iqa}}{sec:chodavarapu}
+\subsection{\citet{Chodavarapu:2010iqa}\label{sec:chodavarapu}
 
@@
 \includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
-\caption{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \em{FF} sample.
+\caption{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample.
 Fourier transformation computed and plotted using the \texttt{spectrum} function in R.
@@
  &  & m & u & Total \\ \cmidrule(l){2-5}
-\multirow{3}{*}{First CpG} & m & $n_{m, m}$ & $n_{m, u}$ &  \\
- & u & $n_{u, m}$ & $n_{u, u}$ &  \\
- & Total &  &  & n \\ \cmidrule(l){2-5}
+\multirow{3}{*}{First CpG} & m & $n_{m, m}$ & $n_{m, u}$ & $n_{m, .} \\
+ & u & $n_{u, m}$ & $n_{u, u}$ & $n_{u, .}$ \\
+ & Total & $n_{., m}$  & $n_{, u}$  & n \\ \cmidrule(l){2-5}
 \end{tabular}
@@
 
-As we will see in \ref{chap:co-methylation}, this is contradicted by my analyses of co-methylation.
+As we will see in Chapter \ref{chap:co-methylation}, this is contradicted by my analyses of co-methylation.
 
@@
 A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs.
-the rest of the genome (see Chapter \ref{chap:co-methyaltion}).
+the rest of the genome (see Chapter \ref{chap:co-methylation}).
 

### 2014-12-11 Rmarkdown/wgbs_downstream_analyses.Rmd
Editing

@@
 This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
-Many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
+Many methods have been proposed \citep[e.g.,]{Akalin:2012cm, Chen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
 
@@
 As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one.
-For example, \citet{Lister:2009hy} state that, "_For each cell type, two __biological__ replicates were performed with cells of different passage number_" (emphasis mine).
+For example, \citet{Lister:2009hy} state that, "For each cell type, two __biological__ replicates were performed with cells of different passage number" (emphasis mine).
 I contend that these are better defined as technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].
@@
 
-[^multiple_testing]; For example, there are approximately 20,000 tests in studies of differential gene expression and one million tests in genome-wide association studies.
+[^multiple_testing]: For example, there are approximately 20,000 tests in studies of differential gene expression and one million tests in genome-wide association studies.
 
@@
 \centering
-\begin{tabular}{@{}ll@{}}
+\begin{tabulary}{\textwidth}{LL}
 \toprule
@@
 \citet{Wang:2011cw}                           & \citet{Akalin:2012cm}                                                               \\
-"False discovery rate correction/adjustment" & \citet{Dolzhenko:2014bo, Gokhman:2014cp, Lister:2009hy, Xie:2014ez}                 \\
+``False discovery rate correction/adjustment'' & \citet{Dolzhenko:2014bo, Gokhman:2014cp, Lister:2009hy, Xie:2014ez}                 \\
 \citet{Storey:2007cp}                         & \citet{Jaffe:2012gx}                                                                \\
@@
 "Bonferonni adjustment"                      & \citet{Feng:2014iq}                                                                 \\ \bottomrule
-\end{tabular}
+\end{tabulary}
 \caption{Methods proposed for adjusting for the multiple hypothesis testing performed in an analysis of DMCs.
@@
 \centering
-\resizebox{\textwidth}{!}{%
   \begin{tabular}{@{}lll@{}}
@@
   \end{tabular}
-}
 \caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
@@
 Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$.
-$\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).
+$\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$.
 
@@
 While this might be considered a two-group experiment, one without replicates, I believe this is better thought of as a one-group experiment with two samples.
-Indeed, the analysis of such an experiment uses techniques from Section \label{sec:one_group_dmcs}, e.g., Pearson's $\chi^{2}$ test or Fisher's exact test, which is better suited to the small sample sizes routinely seen in each $2 \times 2$ table.
+Indeed, the analysis of such an experiment uses techniques from Section \ref{sec:one_group_dmcs}, e.g., Pearson's $\chi^{2}$ test or Fisher's exact test \citep{Fisher:1922bx}, which is better suited to the small sample sizes routinely seen in each $2 \times 2$ table.
 
 The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
-Unfortunately, some authors have still used Fisher's exact test \citep{Fisher:1922bx} to analyse such experiments.
-For example, \citet{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data the two replicates they have for each sample.
+Unfortunately, some authors have still used Fisher's exact test to analyse such experiments.
+For example, \citet{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data from the two replicates they have for each sample.
 This is incorrect, because such an analysis effectievly ignores all within-group variability \citep{Hansen:2012gr}.
@@
 
-In a two-group experiment this might be a t-test comparing the the average methylation level in the first group, $\bar{\beta_{i}^{(1)}}$, to that in the second group, $\bar{\beta_{i}^{(2)}}$.
+In a two-group experiment this might be a t-test comparing the the average methylation level in the first group, $\widebar{\beta_{i}^{(1)}}$, to that in the second group, $\widebar{\beta_{i}^{(2)}}$.
 More generally, we can use a regression model to allow for multiple groups (i.e., ANOVA) and to incorporate additional covariates.
@@
 Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment.
-Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (see Section \ref{sec:beta-binomial_model}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (see Section \ref{transformation}).
+Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (see Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (see Section \ref{sec:transformations}).
 
@@
 Rather, the idea is that a DMR might capture a weaker but broader difference in methylation.
-Such a signal may well have a clearer biological interpretation and, indeed, might be expected such as in differences in methylation at a gene promoter.
+Such a signal may well have a clearer biological interpretation, such as in differences in methylation at a gene promoter.
 
@@
 Using regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands \citep[e.g.,]{Huang:1999fs,Doi:2009ky}[^array], MspI restriction fragments \citep[e.g.,]{Stockwell:2014fq}, any other predefined genomic feature, such as a gene promoters or transcription factor binding sites, or general predefined bins \citep[e.g., 100 bp bins]{Park:2014ho}.
+Such regions might be CpG islands \citep[e.g.,][]{Huang:1999fs,Doi:2009ky}[^array], MspI restriction fragments \citep[e.g.,][]{Stockwell:2014fq}, any other predefined genomic feature, such as a gene promoters or transcription factor binding sites, or general predefined bins \citep[e.g., 100 bp bins used by][]{Park:2014ho}.
 2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
@@
 
-\subsection{Using _a priori_ regions}
+\subsection{Using \em{a priori} regions}
 
@@
 Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions \citep{Good:1987ce}.
-Instead, we could fit the log-linear model with `sample`, `meth_type` and `locus` as the set of factors.
+Instead, we could fit the log-linear model with `sample`, `meth_state` and `locus` as the set of factors.
 The saturated model is
 
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus} + \lambda_{a, b, c}^{meth_state, sample, locus}
-\end{equation*}
+\begin{align*}
+\log(F_{a, b, c}) &= \lambda_{0} \\
+  &+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
+  &+ \lambda_{a, b}^{meth\_state, sample} + \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus} \\
+  &+ \lambda_{a, b, c}^{meth\_state, sample, locus}
+\end{align*}
 
-We would first need to test whether there is a significant three-way interaction between `meth_type`, `sample` and `locus` by computing the deviance of the model without the three-way interaction
+We would first need to test whether there is a significant three-way interaction between `meth_state`, `sample` and `locus` by computing the deviance of the model without the three-way interaction
 
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{equation*}
+\begin{align*}
+\log(F_{a, b, c}) &= \lambda_{0} \\
+&+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
+&+ \lambda_{a, b}^{meth\_state, sample} + \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus}
+\end{align*}
 
-On the one hand, if there is no three-way interaction, then we can test for an interaction between `meth_type` and `sample`, i.e., test for a difference in methylation level across the loci, by comparing the model
+On the one hand, if there is no three-way interaction, then we can test for an interaction between `meth_state` and `sample`, i.e., test for a difference in methylation level across the loci, by comparing the model
 
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{equation*}
+\begin{align*}
+\log(F_{a, b, c}) &= \lambda_{0} \\
+&+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
+&+ \lambda_{a, b}^{meth\_state, sample} + \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus}
+\end{align*}
 
@@
 
-\begin{equation*}
-\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{equation*}
+\begin{align*}
+\log(F_{a, b, c}) &= \lambda_{0} \\
+&+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
+&+ \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus}
+\end{align*}
 
@@
 
-On the other hand, if there is a significant three-way interaction between `meth_type`, `sample` and `locus` then this requires more careful modelling to determine the nature of this interaction and whether the region is a consistent DMR.
+On the other hand, if there is a significant three-way interaction between `meth_state`, `sample` and `locus` then this requires more careful modelling to determine the nature of this interaction and whether the region is a consistent DMR.
 
@@
 
-The most common type of algorithm for identifying data-driven DMRs, also arguably the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
+The most common type of algorithm for identifying data-driven DMRs, also arguably the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
 The initial identification of DMCs will typically use a relaxed statistical significance threshold.
@@
 These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
-Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
+Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
 
@@
 These methods can be thought of extensions to Fisher's method for combining P-values \citep{Fisher:1936vv}, by attempting to account for the correlation of tests at nearby methylation loci.
-`comb-p` \citep{Pedersen:2012vl} uses the Stouffer-Liptak-Kechris \citep{Stouffer:1949ua, Kechris:2010fw, Zaykin:2011jq,} correction for spatially correlated P-values.
+`comb-p` \citep{Pedersen:2012vl} uses the Stouffer-Liptak-Kechris \citep{Stouffer:1949ua, Kechris:2010fw, Zaykin:2011jq} correction for spatially correlated P-values.
 `SLIM` \citep{Wang:2011cw} is used by `methylKit` \citep{Akalin:2012cm} to do a similar correction.
@@
 
-[^between_sample_co-methylation]: See Chapter \ref{co-methylation_review} for further details of between-sample co-methylation.
+[^between_sample_co-methylation]: See Chapter \ref{chap:co-methylation_review} for further details of between-sample co-methylation.
 
@@
 \citet{Jaffe:2011hy} first developed formal statistical tests for both the one-group and two-group experiments.
-These methods were developed for us with data from the CHARM array (see __SECTION__).
+These methods were developed for us with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}).
 
-Statistical tests of variations are known to be more difficult than tests of means and require larger sample sizes.
+Statistical tests of variations are well known to be more difficult than tests of means and require larger sample sizes.
 A more subtle difficulty is in dealing with outliers.
@@
 It is not hard to envisage an example where the two groups in fact have very similar variability once the outlier is excluded.
-Tests of differential variability that are based on the F-test \citep[e.g.,]{Hansen:2011gu} or Bartlett's test \citep[e.g.,]{Teschendorff:2012hea} will be susceptible to calling such loci as differentially variable.
+Tests of differential variability that are based on the F-test \citep[e.g.,][]{Hansen:2011gu} or Bartlett's test \citep[e.g.,][]{Teschendorff:2012hea} will be susceptible to calling such loci as differentially variable.
 By contrast, \citet{Phipson:2014ji} designed `DiffVar` to test for differential variability that is deliberately robust to such outliers by using Levene's test \citep{Olkin:1960uu}.
@@
 Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$.
-An example of such a region is shown in __FIGURE__ (__TODO: Find a PMD and plot it__).
-
 PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}.
@@
 It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions.
-While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that methylation microarrays cannot provide.
+While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that whole-genome bisulfite-sequencing provides
 
@@
 
-The above-described analyses do not specifically exploit the fact that bisulfite-sequencing reads can measure methylation patterns at m-tuples.
+None of the aforementioned analyses specifically exploit the fact that bisulfite-sequencing reads measure methylation patterns at multiple loci, i.e., at m-tuples.
 Yet much can be learnt my studying exactly that.
-To begin, I review the different types of questions that can be addressed when we make use of this extra information that is only available from bisulfite-sequencing data.
+
+In this section I review the different types of questions that can be addressed when we make use of this extra information that is only available from bisulfite-sequencing data.
 
@@
 
-[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles}.
+[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles_and_epipolymorphism}.
 Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since this uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.
@@
 Unfortunately, `DMEAS` is only available as a Windows binary or as a Perl script that itself is only available as PDF file, greatly limiting its usefulness.
-
-While I have not yet done so, it is not difficult to add the calculation of methylation entropies to the `MethylationTuples` software.
-The input for such an analysis of methylation entropy is a list of m-tuples and the associated frequencies of each methylation pattern, exactly what `methtuple` produces.
-The formula for computing methylation entropies is a simple sum of ratios and log-ratios, easily implemented in R, which `MethylationTuples` is written in.
 
@@
 & Allele 1 & Allele 2 \\ \midrule
-m &          &          \\
-u &          &          \\ \bottomrule
+m &  m_{1}        &  m_{2}       \\
+u &  u_{1}        &  u_{2}       \\ \bottomrule
 \end{tabular}
@@
 
-[^epiallele_definition]\citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
+[^epiallele_definition]: \citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
 While most would agree that is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive.
@@
 The 'epipolymorphism' of a locus has also been defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}.
-\citeauthor{Landan:2012kp} the epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} (f_{p}^{2})$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
+\citeauthor{Landan:2012kp} define the epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
 
@@
 The "foreground" combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples while the "background" combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two stages [samples]".
-The difference in these combinatorial entropies, $\delta S = S - \tilde{S}$, a kind of observed-to-expected log-ratio, is used to identify shifts in the epiallele distribution between a pair of samples.
-A $\delta S = 0$ corresponds to no change and a $\delta S = -144$ corresponds to maximal difference in entropy.
-It isn't clear whether the range of $\delta S$ depends on the size of the m-tuples nor is it clear how to choose the threshold at which to declare a significant shift in the distribution of epialleles.
+The difference in these combinatorial entropies, $\Delta S = S - \tilde{S}$, a kind of observed-to-expected log-ratio, is used to identify shifts in the epiallele distribution between a pair of samples.
+A $\Delta S = 0$ corresponds to no change and a $\Delta S = -144$ corresponds to maximal difference in entropy.
+It isn't clear whether the range of $\Delta S$ depends on the size of the m-tuples nor is it clear how to choose the threshold at which to declare a significant shift in the distribution of epialleles.
 The example in the paper shows results for a threshold of $-60, -70$ and $-80$, but these choices appear somewhat arbitrary.
@@
 
-Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,]{Fang:2012ciPeng:2012dh}, might also adapted to identify epialleles and their associated frequencies.
+Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,][]{Fang:2012ci, Peng:2012dh}, might also adapted to identify epialleles and their associated frequencies.
 It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which strictly speaking is a requirement for the m-tuple to be an epipolymorphic locus.
 
-\section{`MethylationTuples`}\label{sec:MethylationTuples}
+\section{\texttt{MethylationTuples}}\label{sec:MethylationTuples}
 
-A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are often lacking or very much tailored toward a particular analysis.
+A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are often lacking or, at best, very much tailored toward a particular analysis.
 What has been lacking is:
@@
 R is a very popular language for analysing data, particularly in bioinformatics, and is my computational mother tongue, hence my choice of R as the language.
-To improve the performance of the package, peaks are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
+To improve the performance of the package, parts are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
 
@@
 In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations.
-For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., `type = 'equal'`).
+For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., `type = \`equal'`).
 Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.
@@
 The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, S_{n}$.
-Some sample may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
+Some samples may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
 \label{fig:MethPat}
@@
 
-I have used the `MethylationTuples` package to analyses datasets containing up to 19 whole-genome bisulfite-sequencing samples (the Ziller data), albeit using a server with $512$ Gb of (shared) memory.
+I have used the `MethylationTuples` package to analyses datasets containing up to 17 whole-genome bisulfite-sequencing samples (the Lister data), albeit using a server with $512$ Gb of (shared) memory.
 I have found the `MethPat` class to be a very convenient representation of the data, but one that does not coincide with the most efficient storage.
@@
 Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
-It also highlights the sparsity of the data as the size of the tuples grows --- most entries are $0$, meaning that particular methylation pattern was not observed in that particular sample, or `NA`, meaning that that m-tuple was not observed in that particular sample.
+It also highlights the sparsity of the data as the size of the tuples grows --- most entries are $0$ (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).
 
@@
 3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
-4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             &                                            \\ \bottomrule
-\end{tabular}
+4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
+\end{tabulary}
 \caption{Size of \texttt{MethPat} objects for the \em{EPISCOPE} data.

### 2014-12-11 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Removed R chunks

@@
 
-```{r ADS_M-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset.
+\begin{figure}[h!]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
+\caption{M-bias plot for the ADS sample from the Lister (2011) dataset.
 Each of read 1 (R1) and read 2 (R2) are plotted separately.
-We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", dev = "pdf", outwidth = "0.8\\linewidth"}
-library(dplyr)
-library(ggplot2)
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
-read.mbias <- function(file) {
-
-  # Echo some information
-  message(paste0("Reading ", file, " ..."))
-
-  # Read the file.
-  x <- read.table(file, sep = '\n', as.is = TRUE)
-
-  # Find 'header' lines, which are dispersed throughout the file.
-  header_lines <- sapply(X = x, function(xx) {
-    grepl(pattern = 'context', x = xx)
-    })
-  y <- vector(mode = "list", length = sum(header_lines))
-  names(y) <- x[header_lines]
-
-  # z indexes the lines containing M-bias values.
-  z <- sapply(X = x, function(xx) {
-    grep(pattern = "context|^=|position", x= xx, invert = TRUE)
-    })
-  dz <- diff(z)
-  next_context_idx <- c(which(dz > 1), nrow(z))
-  colnames_idx <- sapply(X = x, function(xx) {
-    grep(pattern = "position", x = xx)
-    })[1]
-  colnames <- strsplit(x = x[colnames_idx, ], split = "\t")[[1]]
-
-  # Loop over the lines in the file.
-  for (i in seq_along(next_context_idx)) {
-    # Note the "+4" to skip the intermediate 'header' rows
-    start_idx <- ifelse(i == 1, 4, z[next_context_idx[i - 1]] + 4)
-    end_idx <- z[next_context_idx[i]]  
-    yy <- x[seq(from = start_idx, to = end_idx, by = 1), ]
-    y[[i]] <- do.call("rbind", lapply(yy, function(yyy, colnames) {
-      tmp <- strsplit(yyy, split = "\t")
-      val <- data.frame(as.integer(tmp[[1]][1]),
-                        as.integer(tmp[[1]][2]),
-                        as.integer(tmp[[1]][3]),
-                        as.numeric(tmp[[1]][4]),
-                        as.integer(tmp[[1]][5]))
-      colnames(val) <- colnames
-      return(val)
-      }, colnames = colnames))
-    }
-  y <- do.call("rbind", y)
-  y$Context <- strtrim(rownames(y), width = 3)
-
-  # Add R1/R2 annotation if paired-end data
-  if (grepl(pattern = "R1", x = rownames(y)[1])){
-    y$Read <- ifelse(grepl(pattern = "R1", x = rownames(y)), "R1", "R2")
-    } else{
-      y$Read <- "SE"
-      }
-  rownames(y) <- NULL
-
-  # Simplify column names
-  colnames(y) <- c('read_pos', 'M', 'U', 'perc_M', 'total',
-                   'methylation_type', 'read')
-
-  # Re-order columns
-  y <- y[, c('read', 'read_pos', 'methylation_type', 'M', 'U', 'total',
-             'perc_M')]
-
-  # Replace CpG with CG
-  y$methylation_type <- ifelse(y$methylation_type == 'CpG', 'CG',
-                               y$methylation_type)
-
-  # Convert to an MBias object.
-  return(y)
-  }
-
-#files <- paste0("../M-bias/", list.files("../M-bias/"))
-# ADS
-files <- "../data/M-bias/ADS.M-bias.txt"
-sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
-mbias <- lapply(files, read.mbias)
-mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
-                      do.call("rbind", mbias)))
-mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
-mbias %>%
-  filter(sample == "ADS") %>%
-  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
-  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: ADS") +
-  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
-  thesis_theme +
-  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
-```
+We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.}
+\label{fig:ADS_M-bias}
+\end{figure}
 
@@
 
-```{r E18VA_M-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset.
+\begin{figure}[h!]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
+\caption{M-bias plot for the E18VA sample from the EPISCOPE dataset.
 Each of read 1 (R1) and read 2 (R2) are plotted separately.
@@
 The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1.
-This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", dev = "pdf", outwidth = "0.8\\linewidth"}
-#files <- paste0("../M-bias/", list.files("../M-bias/"))
-# E18VA
-files <- "../data/M-bias/E18VA.M-bias.txt"
-sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
-mbias <- lapply(files, read.mbias)
-mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
-                      do.call("rbind", mbias)))
-mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
-mbias %>%
-  filter(sample == "E18VA") %>%
-  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
-  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
-  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
-  thesis_theme +
-  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
-```
+This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
+\end{figure}
 
@@
 
-Most methylation callers, such as `bismark_methylation_extractor` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
-The output file is a table, where each row records the co-ordinates of a cytosine and the number of methylated and unmethylated reads at that position.
-Table \ref{tab:1-tuples} is representative of the type of data returned by these programs.
-The file format is generally tab-delimited plain text, the Browser Extensible Data (BED) format or the Variant Call Format (VCF).
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}lllll@{}}
-\toprule
-Chromosome & Strand & Position & M & U \\ \midrule
-chr1       & $+$      & 100      & 7 & 1 \\
-chr1       & $-$      & 101      & 5 & 2 \\
-chr2       & $+$      & 400      & 0 & 3 \\
-chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
-\end{tabular}
-\caption{Example of output for methylation calling at 1-tuples. Most CpGs will have a measurement from both the positive and negative strands, but not always.}
-\label{tab:1-tuples}
-\end{table}
-
-While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data.
-Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
-Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
-In order to study these concepts, we need software that can perform methylation calling at m-tuples, which was not available when I began my PhD, and so I wrote `methtuple`.
-
-[^chimeric_reads]: This ignores chimeric reads, which are created when two DNA fragments ligate to one another during the library preparation.
-Certain bisulfite-sequencing protocols frequently produce chimeric reads.
-For example, using the post-bisulfite adapter tagging (PBAT) protocol \citep{Miura:2012fr} with a low input amount of DNA results in a huge number of chimeric reads (personal communication from Felix Krueger).
-The standard whole-genome bisulfite-sequencing protocol is not known to suffer from this issue.
+`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software I wrote to perform methylation calling at m-tuples.
+In order to explain how it works and my motivation for writing it, I first need to explain what an m-tuple is.
 
@@
 
-`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software to perform methylation calling at m-tuples from bisulfite-sequencing data.
+Most methylation callers, such as `bismark_methylation_extractor` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
+The output file is a table, where each row records the co-ordinates of a cytosine and the number of methylated and unmethylated reads at that position.
+Table \ref{tab:1-tuples} is representative of the type of data returned by these programs.
+The file format is generally tab-delimited plain text, the Browser Extensible Data (BED) format or the Variant Call Format (VCF).
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lllll@{}}
+\toprule
+Chromosome & Strand & Position & M & U \\ \midrule
+chr1       & $+$      & 100      & 7 & 1 \\
+chr1       & $-$      & 101      & 5 & 2 \\
+chr2       & $+$      & 400      & 0 & 3 \\
+chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
+\end{tabular}
+\caption{Example of output for methylation calling at 1-tuples. Most CpGs will have a measurement from both the positive and negative strands, but not always.}
+\label{tab:1-tuples}
+\end{table}
+
+While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data.
+Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
+Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
+
+[^chimeric_reads]: This ignores chimeric reads, which are created when two DNA fragments ligate to one another during the library preparation.
+Certain bisulfite-sequencing protocols frequently produce chimeric reads.
+For example, using the post-bisulfite adapter tagging (PBAT) protocol \citep{Miura:2012fr} with a low input amount of DNA results in a huge number of chimeric reads (personal communication from Felix Krueger).
+The standard whole-genome bisulfite-sequencing protocol is not known to suffer from this issue.
+
+In order to study these concepts, we need software that can perform methylation calling at m-tuples.
 Table \ref{tab:other_m-tuple_methylation_callers} lists other software that can perform methylation calling at m-tuples.
-However, none of these existed when I started writing `methtuple` nor do any of them do exactly what I require.
+However, none of these existed when I started writing `methtuple`.
+Even now, none of these do exactly what I require and have more general limitations.
 To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome data.
@@
 \centering
-\begin{tabulary}{\textwidth}{LLLLL}
+\begin{tabulary}{\textwidth}{LLLL}
 \toprule
-Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
-\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Suspect you can't get actual methylation calls.} \\
-\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
-\texttt{DMEAS}     & \citet{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only.
-Perl code only available as PDF file.
-\\ \bottomrule
+Software           & Reference                                & Input                                              & Limitations                                                                \\ \midrule
+methclone & \citet{Li:2014ei}                       & \texttt{Bismark} \texttt{BAM}                                  & Unable to install \\
+methpat  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bsm} & Designed for amplicons not whole-genome data.
+\\
+DMEAS    & \citet{He:2013cj}                       & Output of \texttt{bsm} & Windows operating system only. Perl code only available as PDF file.             \\
+\bottomrule
 \end{tabulary}
-\caption{Other software for methylation calling at m-tuples.}
+\caption{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bsm} =  \texttt{bismark\_methylation\_extractor}.}
 \label{tab:other_m-tuple_methylation_callers}
@@
 
-```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes.
-The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", dev = "pdf", out.width = "0.8\\linewidth"}
-library(dplyr)
-library(ggplot2)
-library(stringr)
-library(gridExtra)
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
+\caption{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes.
+The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.
+Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
+\label{fig:methtuple_runtime}
+\end{figure}
 
-# facet_wrap_labeller() from http://stackoverflow.com/a/19298442
-facet_wrap_labeller <- function(gg.plot, labels = NULL) {
-  #works with R 3.0.1 and ggplot2 0.9.3.1
-
-  g <- ggplotGrob(gg.plot)
-  gg <- g$grobs
-  strips <- grep("strip_t", names(gg))
-
-  for(ii in seq_along(labels))  {
-    modgrob <- getGrob(gg[[strips[ii]]], "strip.text",
-                       grep=TRUE, global=TRUE)
-    gg[[strips[ii]]]$children[[modgrob$name]] <- editGrob(modgrob,
-                                                          label = labels[ii])
-    }
-
-  g$grobs <- gg
-  class(g) = c("arrange", "ggplot",class(g))
-  g
-  }
-thesis_theme <- theme_classic(base_size = 12)
-thesis_theme <- theme_bw(base_size = 12)
-# Time
-time <- lapply(paste0("../data/methtuple_performance/time/",
-                      list.files("../data/methtuple_performance/time/")),
-               read.table)
-names(time) <- gsub(".time", "",
-                    list.files("../data/methtuple_performance/time/"))
-n <- str_split(names(time), "\\.")
-time <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1),
-                                             times = sapply(time, nrow)),
-                                         levels =
-                                           paste0("chr",
-                                                  c(1:22, "X", "Y", "M", "L"))),
-                   "Tuple" = rep(sapply(n, "[", 2),
-                                 times = sapply(time, nrow)),
-                   "Time" = unlist(time))
-
-gg_time <- time %>%
-  ggplot(aes(x = Chromosome, y = Time / 60)) + geom_boxplot() +
-  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Running time (minutes)") +
-  ggtitle("Running time of methtuple across 49 samples")
-facet_wrap_labeller(gg_time, paste0("Tuple size: ", unique(time$Tuple)))
-```
-
-```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
-These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
-The suffix 'ac' means that the option \texttt{--all-combinations} was set.", dev = "pdf", out.width = "0.8\\linewidth"}
-# Memory
-memory <- lapply(paste0("../data/methtuple_performance/memory/",
-                        list.files("../data/methtuple_performance/memory/")),
-                 read.table)
-names(memory) <- gsub(".memory", "",
-                      list.files("../data/methtuple_performance/memory/"))
-n <- str_split(names(memory), "\\.")
-# NB: Need to divide reported memory usage by 4 due to bug in 'time'
-# (https://bugzilla.redhat.com/show_bug.cgi?id=703865)
-memory <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1),
-                                               times = sapply(memory, nrow)),
-                                           levels =
-                                             paste0("chr",
-                                                    c(1:22, "X", "Y", "M", "L"))),
-                     "Tuple" = rep(sapply(n, "[", 2),
-                                   times = sapply(memory, nrow)),
-                     "Memory" = unlist(memory) / 4)
-gg_memory <- memory %>%
-  ggplot(aes(x = Chromosome, y = Memory / 10^6)) + geom_boxplot() +
-  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Memory usage (GB)") +
-  ggtitle("Maximum memory usage of methtuple across 49 samples")
-facet_wrap_labeller(gg_memory, paste0("Tuple size: ", unique(memory$Tuple)))
-```
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
+\caption{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
+These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
+The suffix `ac' means that the option \texttt{--all-combinations} was set.
+Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
+\label{fig:methtuple_runtime}
+\end{figure}
 

### 2014-12-15 Rmarkdown/datasets.Rmd
Misc. editing

@@
 
-This chapter briefly documents the 51 whole-genome bisulfite-sequencing samples that I use in my thesis. The `BAM` files for each sample underwent the same basic processing:
+This chapter briefly documents the 40 whole-genome bisulfite-sequencing samples that I use in my thesis. The `BAM` files for each sample underwent the same basic processing:
 
@@
 
-__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e., going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $275$ GB alone.
+__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e., going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $> 275$ GB alone.
 
@@
 - _SA_: Subcutaneous adidose tissue, which is fat found just below the skin.
-Unlike visceral fat, "subcutaneous fat is not related to many of the classic obesity-related pathologies, such as heart disease, cancer, and stroke" (__TODO: Get better reference than wikipedia__).
-- _VA_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs (stomach, liver, intestines, kidneys, etc.) (__TODO: Get better reference than wikipedia__).
+Unlike visceral adipose tissue, subcutaneous adipose tissue is thought to be protective against obesity-related metabolic dysfunction \citep{Chau:2014do}.
+- _VA_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs and is associated with metabolic dysfunction \citep{Chau:2014do}.
 - _VAT_: Visceral adipocytes, which are derived from the VA tissue.
@@
 The Ziller are a subset of the data used in \cite{Ziller:2013fs}.
-Specifically, I use a convenience sample of 19 (__TODO: Check final numbers__) WGBS libraries.
+Specifically, I use a convenience sample of 8 WGBS libraries.
 These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research), who already had aligned and processed BAM files for these 19 libraries.

### 2014-12-15 Rmarkdown/wgbs_statistical_framework.Rmd
Editing
Misc. editing

@@
 
-\subsubsection{Genetic heterozygosity at a methylation locus}\label{sec:heterozygosity}
+\subsubsection{Genetic heterozygosity at a methylation locus} \label{sec:heterozygosity}
 
@@
 
-__TODO: Update discussion of results in light of Ziller merged data__
-
 In a study involving multiple samples, the set of $\beta$-values can be summarised as a matrix where each row is a locus and each column is a sample.
@@
 
-Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively.
+Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
 Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_merged_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
 
-What is immediately clear is that these distributions are bimodal.
+What is immediately clear is that these distributions are bimodal --- most CpGs are highly methylated or lowly methylated.
 The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels.
@@
 
-Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _SRR949208_ and _SRR949209_ from the Ziller dataset --- or cancer cells lines and tissue --- _SRR949206_, _SRR949207_, _SRR949210_, _SRR949211_, _SRR949212_, _SRR949213_ and _SRR949215_ from the Ziller dataset.
+Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _IMR90\_cell\_line_ from the Ziller dataset --- or cancer cells lines and tissue --- _HepG2\_cell\_line_, _Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_ from the Ziller dataset.
 Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the Seisenberger dataset also displays greater levels of intermediate methylation.
@@
 Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation.
-Likewise, the various frontal cortex samples in the Ziller dataset --- _SRR949193_, _SRR949194_, _SRR949195_, _SRR949196_, _SRR949197_, _SRR949198_, _SRR949199_, _SRR949201_, _SRR949202_, _SRR949203_ and _SRR949205_ --- which includes both 'normal' and Alzheimer's samples, have very low levels of intermediate methylation.
+Likewise, the various frontal cortex samples in the Ziller dataset --- _Frontal\_cortex\_normal\_1_, _Frontal\_cortex\_normal\_2_, _Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_ --- which includes both 'normal' and Alzheimer's samples, have very low levels of intermediate methylation.
 This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_.
@@
 Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties.
-I then examined the empirical distributions of these variables across a diverse set of 51 whole-genome bisulfite-sequencing samples.
+I then examined the empirical distributions of these variables across a diverse set of 40 whole-genome bisulfite-sequencing samples.
 
 Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
-The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_).
+The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_).
 The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA.

### 2014-12-15 Rmarkdown/comethylation.Rmd
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.
Minor fixes
Spiel on correlations of beta-values

@@
 ---
-title: "Co-methylation"
 author: "Peter Hickey"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
 
+\chapter{Co-methylation}\label{chap:co-methylation}
+
 ## Chapter overview
@@
 Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci.
-I apply these methods to more than fifty whole-genome bisulfite-sequencing libraries to study the between-sample variability of co-methylation.
+I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.
 
@@
 
-__TODO: A spiel on how my method is better than existing ones.__
+Chapter \ref{chap:co-methylation_review} documented several previous analyses based on 'correlations' of aggregate methylation levels, such as $\beta$-values.
+What was lacking from these was a clear description and implementation of the proposed method.
+Here I describe two simple methods based on correlations of $\beta$-values.
+Both of these are implemented in the `methLevelCor` function that is part of the `MethylationTuples` software.
+
+The aim of this analysis is to address the question of 'how are aggregate methylation levels correlated as a function of the distance between loci?'. There are two steps in this analysis:
+
+1. Define the pairs of methylation loci.
+2. For each $IPD$, compute the correlation of $\beta$-values for all pairs of methylation loci separated by that $IPD$.
+
+Step (2) may be further stratified by the genomic context of the pair of CpGs, such as whether they are in a CpG island, or other variables of interest.
+
+The same analysis could be performed using $\mathcal{M}$-values rather than $\beta$-values, and this is available as an option in the `methLevelCor` function.
+
+\subsection{Constructing the pairs}
+
+I consider two different strategies for creating pairs of CpGs.
+Both strategies are based on CpGs in the reference genome after having filtered out those reference-specific CpGs.
+That is, I begin with $\mathcal{I}^{ref}$ and, based on the `Bis-SNP` output, I filter out loci that are not CpGs in the sample's genome.
+Rather than actually remove these sites, I simply set the corresponding $\beta$-value to `NA`.
+I also set the $\beta$-values of 'missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.
+
+The first strategy creates all pairs of adjacent CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context.
+For a chromosome with $N$ CpGs there are $(N - 1)$ pairs.
+I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have zero intervening loci^[nil0].
+
+[^nil]: Recall that this is with respect to the reference genome and not the sample genome.
+
+The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context.
+In practice, the second strategy only uses pairs for a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy.
+This is more like a traditional autocorrelation analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$, albeit of a vector of unequally spaced observations and with care taken to only use pairs on the same chromosome and strand.
+I refer to this as the $NIL \geq 0$ strategy.
+Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.
+
+The interpretation of correlations computed using the $NIL \geq 0$ is complicated because the set of pairs are more heterogeneous than those under the $NIL = 0$ strategy.
+For this reason I prefer the $NIL = 0$ strategy, however, it may be useful to consider and contrast the two strategies, which is what I have done in my thesis.
+
+\subsection{Stratifying by a genomic feature}
+
+Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two distinct regions of the genome that have very different methylation dynamics.
+In other words, the vector of $\beta$-values for a given sample is highly non-stationary --- the covariance and correlation between two loci depends on more than just the $IPD$.
+An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values.
+To do this, we can stratify our analysis by a genomic feature.
+While this does not eliminate this non-stationarity, it may help identify factors that drive some of the heterogeneity.
+
+For any genomic feature, a pair of loci may be inside, outside or spanning the boundary of the feature.
+__FIGURE__ illustrates the 'feature status' of some pairs of loci.
+Note that a pair where each loci is in a different CpG island is declared to be inside the feature, i.e., elements need not be in the same feature but rather in the same type of feature.
+The genomic feature should partition the genome so that each loci is either inside or outside of the feature.
+Since the number of 'spanning pairs' is substantially fewer than those inside or outside of the feature, and the boundaries of the feature are oftentimes fuzzy, I have excluded these 'spanning pairs' in the results shown below.
+
+\subsection{Choice of correlation coefficient}
+
+The three commonly used correlation coefficients are Pearson's $r$ (__CITE__), Spearman's $\rho$ (__CITE__) and Kendall's $\tau$ (__CITE__).
+Briefly, Pearson's product-moment correlation is designed to detect linear relationships between two variables.
+However, it is not robust to outliers, which motivates the use of Spearman's and Kendall's rank-based correlation coefficients.
+Spearman's $\rho$ is simply the Pearson product-moment correlation of the ranks of the data while Kendall's $\tau$ is based on the relative frequency of concordantly ranked pairs.
+Owing to the size of whole-genome bisulfite-sequencing data, I have chosen to only compute Pearson's $r$ and Spearman's $\rho$, which are much faster to compute than Kendall's $\tau$.
+
+\subsection{Results}
+
+For all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets, I compute the Pearson and Spearman correlations using both the $NIL = 0$ and $NIL \geq 0$ strategies.
+I compute these genome-wide and after stratifying CpG pairs by whether they are in a CpG island.
+
+__TODO: Results__
 

### 2014-12-16 Rmarkdown/comethylation.Rmd
Minor edits to co-methylation

@@
 Spearman's $\rho$ is simply the Pearson product-moment correlation of the ranks of the data while Kendall's $\tau$ is based on the relative frequency of concordantly ranked pairs.
+
 Owing to the size of whole-genome bisulfite-sequencing data, I have chosen to only compute Pearson's $r$ and Spearman's $\rho$, which are much faster to compute than Kendall's $\tau$.
+The `methLevelCor` function in the `MethylationTuples` will compute a confidence interval, $95\%$ by default, for the Pearson correlation.
+This feature is not yet available when using Spearman's or Kendall's correlation coefficient.
 
@@
 I compute these genome-wide and after stratifying CpG pairs by whether they are in a CpG island.
+These analyses are performed separately for each strand and also for strand-collapsed $\beta$-values.
 

### 2014-12-19 Rmarkdown/comethylation.Rmd
Updates on co-methylation and co-methylation review

@@
 Chapter \ref{chap:co-methylation_review} documented several previous analyses based on 'correlations' of aggregate methylation levels, such as $\beta$-values.
-What was lacking from these was a clear description and implementation of the proposed method.
-Here I describe two simple methods based on correlations of $\beta$-values.
-Both of these are implemented in the `methLevelCor` function that is part of the `MethylationTuples` software.
+What was lacking from these were clear descriptions of the proposed methods and software implementations.
+
+\subsection{Methods}
+
+Here I describe two simple methods based on correlations of $\beta$-values. Both of these are implemented in the `methLevelCor` function that is part of the `MethylationTuples` software.
 
@@
 
-\subsection{Constructing the pairs}
+\subsubsection{Constructing the pairs}\label{sec:beta_pair_construction}
 
@@
 
-\subsection{Stratifying by a genomic feature}
+\subsubsection{Stratifying by a genomic feature}\label{sec:pair_stratification}
 
@@
 
-\subsection{Choice of correlation coefficient}
+\subsubsection{Choice of correlation coefficient}
 
@@
 
-__TODO: A spiel on how my method is better than existing ones.__
+The key advantage of studying within-fragment co-methylation is that measures the dependence of DNA methylation at the scale on which the biological process acts.
+I have developed a method based on odds-ratios to quantify the strength of co-methylation for a pair of methylation loci in a single sample.
 
-In studying within-fragment co-methylation, I assume that all copies of each chromosome have the same dependence structure (__TODO: Although not the same mean structure?
-Make this mathematically precise__).
-The data are sampled from a population containing multiple copies of each chromosome and if each copy has a different dependence structure then we cannot hope to say anything about these structures since these copies are not identifiable.
+\subsection{Methods}
 
-\subsection{Using pairs of methylation loci}\label{subsec:pairs}
+Here I describe a simple method based on analysing methylation patterns at 2-tuples[^higher_order_m-tuples].
+For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type.
+This method is implemented in the `cometh` function that is part of the `MethylationTuples` software.
 
-The simplest summary of within-fragment co-methylation uses pairs (2-tuples) of methylation loci.
-Consider a single pair of CpGs in the reference genome and all reads that contain both CpGs of the pair (see Figure \ref{reads_containing_pair_of_CpGs}).
+[^higher_order_m-tuples]: Extensions to m-tuples with $\text{m} > 2$ are discussed in __SECTION__.
 
-\begin{figure}[h!]
-  \centering
-    \includegraphics[width=0.5\textwidth]{../figures/reads_containing_pair_of_CpGs.pdf}
-      \caption{All reads containing a pair of CpGs.
-The CpGs are highlighted by the short blue bars along the bottom of the figure.
-Each long grey (resp.
-green) bar is a read mapped to the forward (resp.
-reverse) strand.
-Those reads highlighted in yellow contain both CpGs of the pair.
-Data are from the IMR90-iPSC sample.}
-      \label{reads_containing_pair_of_CpGs}
-\end{figure}
+The aim of this analysis is to address the question of 'how are methylation states on individual DNA fragments correlated as a function of the distance between loci?'.
+There are two steps in this analysis:
 
-We tabulate the methylation states of each of these reads as shown in Table \ref{reads_containing_pair_of_CpGs_counts}.
-The `methtuple` software can perform this tabulation for all pairs of methylation loci in a sample.
+1. Define the 2-tuples.
+2. For each 2-tuple, estimates the dependence between the methylation state at the first locus and the second locus. For this I use the logarithm of the odds ratio ($LOR$).
 
-The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the log-odds ratio of this table, $LOR_{(i, i'), j} = log_2 \Big (\frac{MM_{i, i'} \times UU_{i, i'}}{MU_{i, i'} \times UM_{i, i'}} \Big)$.
-For the data shown in Figure \ref{reads_containing_pair_of_CpGs} the $LOR =$ `r round(log2((1.5 * 4.5) / (0.5 * 2.5)), 1)`.
+Step (2) may be further stratified by the genomic context of the pair of CpGs, such as whether they are in a CpG island, or other variables of interest.
+This is done in the same manner as for pairs of $\beta$-values (Section \ref{sec:pair_stratification}).
 
-\begin{table}[h]
-  \centering
+Finally, a similar analysis could be performed using a different measure of dependence, which I discuss in Section \ref{sec:choice_of_dependence_measure}.
 
-  \begin{tabular}{lcccl}
-                           & \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
-                           &                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
-\multirow{2}{*}{First CpG} & \textbf{M}           & 1              & 2             & 3              \\
-                           & \textbf{U}           & 0              & 4             & 4              \\ \cline{2-5}
-                           & \textbf{Total}       & 1              & 6             & 7
-\end{tabular}
-\caption{Counts of each methylation pattern for reads shown in Figure \ref{reads_containing_pair_of_CpGs}}
-\label{reads_containing_pair_of_CpGs_counts}
-\end{table}
+\subsubsection{Constructing the 2-tuples}
 
-A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs in the pair on the same DNA fragment (_co-methylation_), a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG in the pair on the same DNA fragment and a value close to zero is evidence that the methylation state at each CpG in the pair on the same DNA fragment are independent of one another.
+The construction of 2-tuples is very similar to constructing the pairs of methylation levels, described in Section \ref{sec:beta_pair_construction}.
+However, we are now limited to analysing co-methylation of loci that can be captured within a single read.
+In practice, this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD \lessapprox 200$.
+
+Again, we can employ either the $NIL = 0$ or the $NIL \geq 0$ strategy.
+Both strategies start with CpGs in the reference genome that, for each sample, are then filtered to remove reference-specific loci.
+Rather than actually remove these loci, I simply set the corresponding counts of the methylation patterns to `NA`.
+
+As in Section \ref{subsec:beta_cor} I consider the $NIL = 0$ and $NIL \geq 0$ strategies for constructing pairs.
+The $NIL \geq 0$ strategy corresponds to using the `--all-combination` flag in `methtuple`.
+Again, the $NIL = 0$ strategy is simpler, both conceptually and computationally, and so I prefer it.
+For example, if we have a pair with $NIL = 3$ then we need to take care to ensure that this $LOR$ is due to a genuine dependence between these two CpGs above and beyond the pairwise dependencies $(CpG_{1}, CpG_{2}), \ldots, (CpG_{4}, CpG_{5})$.
+
+\subsubsection{Choice of dependence measure}\label{sec:choice_of_dependence_measure}
+
+Recall that for each 2-tuple and sample we can summarise the methylation patterns by a $2 \times 2$ contingency table (__FIGURE__).
+The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the logarithm of the odds ratio of this table, $LOR_{(i, i'), j} = log_2 \Big (\frac{MM_{i, i'} \times UU_{i, i'}}{MU_{i, i'} \times UM_{i, i'}} \Big)$.
+I use base-2 logarithms due to their widespread use and familiarity within biology.
+In the `cometh` function I also add a small offset (default `offset = 1`) to avoid problems arising from zero counts, which frequently occur.
+
+A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs in the pair on the same DNA fragment, a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG in the pair on the same DNA fragment, and a value close to zero is evidence that the methylation state at each CpG in the pair on the same DNA fragment are independent of one another.
 More precisely, we may wish to compute a confidence interval for $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.
 
-In theory we can compute $LOR$ for all pairs of methylation loci. In practice, using current technology, we can only compute it for pairs of methylation loci that are at most separated by 200-250 bp.
+__TODO: Comparison of LORs and Pearson correlations of Landan__
 
-### Comparison of LOR to correlations
+__TODO: Discussion of Kraemer's work, in particular, the fact that non-zero LOR are not informative for the strength of the correlation__
 
-__TODO: Compare use of log-odds ratio to use of correlation coefficient and justify using LOR__
+__TODO: \cite{Bonett:2007bd} is a reference that Pearson correlation of 2x2 is phi coefficient__
 
-### Constructing pairs of methylation loci
+\subsection{Results}
 
-There are additional considerations in constructing pairs of methylation loci on top of those constraints imposed by the sequencing technology.
+$LOR$s are only computed for 2-tuples with sufficient sequencing coverage (here, $5 \times$ for stranded 2-tuples and $10 \times$ for unstranded 2-tuples).
+For any given $IPD$ there are hundreds of thousands of CpG pairs, and so there is a distribution of co-methylation for any given $IPD$.
+We could visualise the distribution for a given $IPD$ using classical visualisation techniques such as histograms, kernel density estimates and boxplots.
+These summaries may be stratified by one or more annotations.
+However, what these $IPD$-specific plots don't give the reader is a sense of how $IPD$ affects within-fragment co-methylation.
 
-For a sequence containing $k$ methylation loci there are $\binom{k}{2}$ pairs, which is $\mathcal{O}(k^{2})$ as $\lim_{k \to \infty}$.
-In contrast, there are $k - 1$ pairs of _neighbouring_ methylation loci (i.e., pairs where the _number of intervening methylation loci_ equal to zero, $NIML = 0$), which is $\mathcal{O}(k)$ as $\lim_{k \to \infty}$.
+Instead, for each sample I plot selected quantiles --- $10\%, 25\%, 50\%, 75\% and 90\%$ --- of each $IPD$-specific $LOR$ distribution as a function of $IPD$.
+I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$.
+These quantile plots may also be stratified by additional annotations.
 
-Apart from this difference in the number of pairs, the interpretation of co-methylation at neighbouring loci is far simpler than for non-neighbouring loci and so it may be preferable to focus on neighbouring loci.
-For example, if we observe $LOR >> 0$ for a pair of CpGs, $(CpG_{1}, CpG_{5})$, with $NIML = 3 > 0$ then we need to take care to ensure that this is due to a genuine dependence between these two CpGs above and beyond the pairwise dependencies $(CpG_{1}, CpG_{2}), \ldots, (CpG_{4}, CpG_{5})$.
+Key points:
 
-The `methtuple` software by default tries to only construct pairs of neighbouring methylation loci[^all-combinations].
-However, because `methtuple` processes each read independently, filters methylation loci on user-specified criteria and is unaware of the reference genome, it will occasionally create pairs that are not truly neighbouring.
-Therefore, if strictly neighbouring pairs are required, non-neighbouring pairs should be post-hoc filtered (__TODO: e.g.
-using the `whatFunction` in `myRMethTuplePackage`__).
+- LOR decreases with IPD
+- For a given IPD, a wide variation in LOR
+- LOR does not go to zero over observable distance
+- Variation is fairly constant across IPD
+- Noise at the end is due to limited data
+- CGIs are more co-methylated than non-CGI
+- SE data can't cut the mustard
+- Intriguing mode/plateua around 150-200 bp (NIL >= 0)
+  - SE data suggests some is technical, e.g., incomplete removal of M-bias
+  - Also, IMR90 PE data doesn't have it
+- LOR vs. Pearson
+- NIL = 0 vs. NIL >= 0
+- iPSC vs. somatic
+- Plot histogram of LOR and histogram of r
+- Look for tables with LOR > 0 and r = 0
 
-[^all-combinations]: `methtuple` also provides the option `--all-combinations` to specifically create all $\binom{k}{m}$ $m$-tuples.
-However the computational requirements are generally prohibitive for $m > 2$.
+An example quantile plot for the __WHICH SAMPLE__ is shown in __FIGURE__.
+The vast majority of $LOR$s are positive, meaning that most 2-tuples are co-methylated.
+The strength of co-methylation decays roughly as $IPD$ increases, which can be seen by the downward shift in the quantiles as a function of $IPD$.
+I
 
-Furthermore, while CpG methylation is typically strand-symmetric, it may be preferable to compute co-methylation separately for each strand.
-This will double the number of CpG 2-tuples.
-By default, `methtuple` does not collapse CpGs across strands, but this option is available via the `--strand-collapse` option or as a post-hoc step in __TODO: e.g.
-using the `whatFunction` in `myRMethTuplePackage`__.
+__FIGURES__ show that the two DNA strands have near identical $LOR$ distributions.
+This is not to discount the possibility of 2-tuples with strand-specific differences in co-methylation, however, it does mean that it is generally safe to aggregate counts across strands.
+
+__FIGURES__
+
+__TODO__
+
+__OLD WRITING BELOW THIS LINE__
 
@@
 
-__TODO__
-
 ## A Markov model of within-fragment co-methylation

### 2014-12-19 Rmarkdown/comethylation_review.Rmd
Latex fixes
Updates on co-methylation and co-methylation review

@@
 The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs.
-They use ultradeep (median coverage of each CpG in the target $> 10,000 \times$) bisulfite-sequencing of a 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
+They use ultra-deep (median coverage of each CpG in the target $> 10,000 \times$) bisulfite-sequencing of a 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
 They also use these data to estimate co-methylation.
@@
  &  & \multicolumn{3}{c}{Second CpG} \\ \cmidrule(l){3-5}
- &  & m & u & Total \\ \cmidrule(l){2-5}
-\multirow{3}{*}{First CpG} & m & $n_{m, m}$ & $n_{m, u}$ & $n_{m, .} \\
- & u & $n_{u, m}$ & $n_{u, u}$ & $n_{u, .}$ \\
- & Total & $n_{., m}$  & $n_{, u}$  & n \\ \cmidrule(l){2-5}
+ &  & $m$ & $u$ & Total \\ \cmidrule(l){2-5}
+\multirow{3}{*}{First CpG} & $m$ & $n_{m, m}$ & $n_{m, u}$ & $n_{m, .}$ \\
+ & $u$ & $n_{u, m}$ & $n_{u, u}$ & $n_{u, .}$ \\
+ & Total & $n_{., m}$  & $n_{, u}$  & $n$ \\ \cmidrule(l){2-5}
 \end{tabular}

### 2014-12-21 Rmarkdown/comethylation.Rmd
Discussing correlations of beta-values

@@
 For any genomic feature, a pair of loci may be inside, outside or spanning the boundary of the feature.
-__FIGURE__ illustrates the 'feature status' of some pairs of loci.
+Figure \ref{fig:feature_status} illustrates the 'feature status' of some pairs of loci.
 Note that a pair where each loci is in a different CpG island is declared to be inside the feature, i.e., elements need not be in the same feature but rather in the same type of feature.
@@
 Since the number of 'spanning pairs' is substantially fewer than those inside or outside of the feature, and the boundaries of the feature are oftentimes fuzzy, I have excluded these 'spanning pairs' in the results shown below.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/feature_status.pdf}
+\caption{Schematic illustrating the 'feature status' of four pairs of methylation loci.
+The circles represent loci and the grey bars a genomic feature, such as CpG islands.
+Each pair's 'feature status' is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside).
+Note that the rightmost pair is 'inside' even though the two loci that comprise the pair are in different elements of the genomic feature.}
+\label{fig:feature_status}
+\end{figure}
 
@@
 
+\subsubsection{Interpretation of correlation}\label{sec:interpretation_of_correlation}
+
+Before computing correlations and trying to interpret these, it is a good idea to look at the data from which these correlations are computed.
+Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200, 2000$ with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample.
+It highlights two major problems.
+
+Firstly, the bimodality of the distribution of $\beta$-values means that the points concentrate in the corners of the plot --- $(0, 0), (0, 1), (1, 0)$ and $(1, 1)$.
+A correlation coefficient is not well suited to summarising these distributions of points.
+Secondly, when using the $NIL = 0$ strategy, as the $IPD$ increases, the number of points in each scatterplot decreases and so the correlations are very unstable.
+
+In summary, the correlations of pairs of $\beta$-values should be cautiously interpreted, bearing in mind that these are rather crude summaries of the underyling scatterplots that hide many details.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/ADS_beta_scatterplots.pdf}
+\caption{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the \em{ADS} sample.
+$\beta$-values are computed using strand-collapsed counts.
+$n$ is the number of pairs in each scatterplot.
+The Spearman correlation of each scatterplot is also reported.
+Individual plots created using the \texttt{smoothScatter} function in R.}
+\label{fig:ADS_beta_scatterplots}
+\end{figure}
+
 \subsection{Results}
@@
 These analyses are performed separately for each strand and also for strand-collapsed $\beta$-values.
+
+\subsubsection{Can we collapse by strand?}
+
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_merged_spearman_beta_cors_min_cov_5} show that the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
+Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
+\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\label{fig:EPISCOPE_pearson_beta_cors_min_cov_5}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
+\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\label{fig:Lister_pearson_beta_cors_min_cov_5}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
+\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\label{fig:Seisenberger_pearson_beta_cors_min_cov_5}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
+\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\label{fig:Ziller_merged_pearson_beta_cors_min_cov_5}
+\end{figure}
+
+\subsubsection{Choice of correlation coefficient}
+
+Figures \ref{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} compares the Pearson correlation coefficient to the Spearman correlation coefficient.
+The two correlation coefficient give qualitatively similar results, however, the Spearman correlation is consistently smaller in magnitude than the Pearson correlation.
+
+I have elected to use Spearman's correlation coefficient in what follows because it is more robust to outliers than Pearson's correlation \citep{Kraemer:2006te}.
+However, this does not remove the general limitations of using correlation coefficients (discussed in Section \ref{sec:interpretation_of_correlation}).
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\label{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\label{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\label{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\label{fig:Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\subsubsection{$NIL = 0$ versus $NIL \geq 0$}
+
+__UP TO HERE__
+
+__Key points__
+
+- NIL = 0 vs. NIL >= 0
+- CGI vs. non-CGI
+- periodicities
 

### 2014-12-21 Rmarkdown/comethylation_review.Rmd
Updated coMET reference

@@
 
-The `coMET` `R` package ([http://epigen.kcl.ac.uk/comet/](http://epigen.kcl.ac.uk/comet/)) can be used to estimate and plot between-sample co-methylation heatmaps from Illumina HumanMethylation450 array data.
-However, it is as yet unpublished and is still  in development at the time of writing.
+A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$.
+Then, compute $R = cor(X^{T})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci.
+This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data.
 

### 2014-12-21 Rmarkdown/comethylation.Rmd
NIL = 0 beta correlations

@@
 
-\subsubsection{$NIL = 0$ versus $NIL \geq 0$}
+\subsubsection{$NIL = 0$}
+
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} show the Spearman correlations of strand-collapsed $\beta$-values for $NIL = 0$ pairs of CpGs.
+Two things immediately stand out.
+Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$ --- pairs of CpGs separated larger distances are on average less correlated than pairs of CpGs separated by shorter distances.
+Secondly, the estimated correlations violently jump around for larger $IPD$s.
+The latter is simply due to the aforementioned lack of $NIL = 0$ pairs at larger $IPD$s with which to estimate these correlations.
+Therefore, it is the trend, rather than the individual values, that should be analysed in these plots.
+
+The are some consistent patterns in comparing these trends across samples.
+The correlations very quickly drop well below $0.5$, within approximately $50$ bp for all the EPISCOPE and Seisenberger samples and most of the Lister and Ziller samples.
+Samples that are exceptions to this trend --- _ADS_, _ADS-adipose_, _ADS-iPSC_, _FF_, _IMR90\_r1_, _IMR90\_r2_, _IMR90\_cell\_line_ and _HepG2\_cell\_line_ --- have a slower decay in these correlations and remain more highly correlated over $IPD = 2, \ldots, 1500$.
+These same samples are also notable because they have significant amounts of intermediate methylation (Section \ref{sec:genome-wide_distribution_of_beta_values}).
+I believe this is what drives the stronger and slowly decaying correlations.
+The scatterplots of $\beta$-values for these samples will have more $\beta$-values in the middle of the scatterplot, and fewer in the corners, hence the stronger correlations.
+
+Most samples have a correlation close to $1$ for neighbouring CpGs ($IPD = 2$).
+The notable exceptions are the embryonic stem cell replicates, _H1\_r1_ and _H1\_r2_ (Lister dataset), which begin with a maximum correlation of only around $0.5$.
+This is not observed in the other embryonic stem cells _H9_, _H9\_Laurent_, _HSF1_ and _J1\_1_), which makes it difficult to know whether $\beta$-values of embryonic stem cells are truly less correlated than other cell types.
+
+The trend of the correlations for all these samples plateaus out close to zero, slightly higher for those samples with significant intermediate methylation.
+However, it should not be concluded from this that the methylation of CpGs separated by more than a few hundred base pairs are uncorrelated.
+Recall that these results are only for $NIL = 0$ pairs of CpGs, which are increasingly rare for larger $IPD$s.
+These results tell us about the 'first-order' correlations of $\beta$-values, which decay very quickly but whose importance also decreases quite quickly owing to the vast majority of CpGs being within $100$ bp of another CpG (e.g., Figure \ref{fig:IPD_ECDF_hg19} for humans).
+
+One final aspect of these genome-wide plots bears mentioning.
+There are hints of an approximately $150-200$ bp periodicity in these plots.
+This is a similar scale to the previously reported periodicities in correlations of $\beta$-values (see Chapter \ref{chap:co-methylation_review}).
+While this periodicity can readily be detected by eye, it is somewhat more difficult to verify using traditional methods based on Fourier analysis of the signal.
+The chief reason is that the data are very noisy beyond the second or third putative cycle due to the limited data at these larger $IPD$s.
+
+To address this problem, we might try smoothing the raw correlations, for example, by using LOESS (__CITE__), and then looking for periodicities in the smoothed values.
+However, this is only helpful when the data are corrupted by a small amount of noise, certainly not the extreme noise we see for $IPD > 1000$ in most samples.
+Furthermore, smoothing introduces problems of its own, such as choosing the smoothing bandwidth in such a way as to not introduce or destroy real periodicities.
+Even if we can automatically choose these smoothing parameters in an unbiased manner, we are still stuck with only observing a handful of periods of this size from $NIL = 0$ pairs.
+So while these periodicities are 'obvious' in the plots, they are rather more difficult to detect without visualising the raw data.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\label{fig:Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+The above are genome-wide plots, all pairs of CpGs with a given $IPD$ are treated identically.
+However, there are good reasons to expect that the correlations of $\beta$-values, like the $\beta$-values themselves, might be influenced by their genomic context.
+To investigate this, we can stratify pairs of CpGs pairs by a genomic feature and repeat the analysis.
+The obvious feature to stratify on is CpG islands.
+
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning].
+Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.
+
+The message from these plots is clear, $\beta$-values of $NIL = 0$ pairs within CpG islands are much more correlated than those pairs outside of islands.
+Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside islands is less influenced by $IPD$.
+In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD.
+From these plots we can now see that the apparent sharp decline in correlations over the first $50$ bp from the genome-wide data is really driven by a shift from pairs that are frequently inside CpG islands to pairs that are mostly outside of islands.
+
+[^spanning]: Pairs spanning the boundary are excluded. There are few of these pairs.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\label{fig:Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\subsubsection{$NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}
+
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} are the plots of CpG $\beta$-value correlations against $IPD$ for $NIL \geq 0$ pairs.
+In contrast to the $NIL = 0$ results, the correlations here are relatively stable, at least for $IPD = 2, \ldots, 2000$[^ipd_choice].
+This is because there are many more $NIL \geq 0$ pairs than there are $NIL = 0$ pairs.
+
+[^ipd_choice]: These correlations could be computed for $IPD > 2000$.
+However, constructing all $NIL \geq 0$ pairs requires much more time and memory than constructing all $NIL = 0$ pairs because there are many more such pairs.
+For this reason, and for compatability with the $NIL = 0$ results, I have focused on $IPD = 2, \ldots, 2000$ for the $NIL \geq 0$ pairs.
 
@@
 
-__Key points__
+- Complications of interpreting NIL >= 0 plots
+- Periodicities or lack thereof
+- CGI vs. non-CGI
 
-- NIL = 0 vs. NIL >= 0
-- CGI vs. non-CGI
-- periodicities
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
+\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
+\end{figure}
 
-__TODO: Results__
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
+\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
+\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
+\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
+\label{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\subsection{Summary}
+
+The $NIL = 0$ results measure the 'first-order' or 'one-step' correlations of $\beta$-values whereas the $NIL \geq 0$ measure a mixture of different orders of correlations of $\beta$-values.
+For this reason I find the $NIL = 0$ plots are easier to interpret, however, which is more relevant will depend on the application.
 

### 2014-12-21 Rmarkdown/comethylation.Rmd
Correlations of NIL >= 0 beta-values

@@
 \includegraphics[width=\textwidth]{../figures/ADS_beta_scatterplots.pdf}
-\caption{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the \em{ADS} sample.
+\caption{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the ADS sample.
 $\beta$-values are computed using strand-collapsed counts.
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_merged_spearman_beta_cors_min_cov_5} show that the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show that the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
 Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.
@@
 \caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
-\label{fig:EPISCOPE_pearson_beta_cors_min_cov_5}
+\label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
 \end{figure}
@@
 \caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
-\label{fig:Lister_pearson_beta_cors_min_cov_5}
+\label{fig:Lister_spearman_beta_cors_min_cov_5}
 \end{figure}
@@
 \caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
-\label{fig:Seisenberger_pearson_beta_cors_min_cov_5}
+\label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
 \end{figure}
@@
 \caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
-\label{fig:Ziller_merged_pearson_beta_cors_min_cov_5}
+\label{fig:Ziller_spearman_beta_cors_min_cov_5}
 \end{figure}
@@
 
-Figures \ref{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} compares the Pearson correlation coefficient to the Spearman correlation coefficient.
+Figures \ref{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} compares the Pearson correlation coefficient to the Spearman correlation coefficient.
 The two correlation coefficient give qualitatively similar results, however, the Spearman correlation is consistently smaller in magnitude than the Pearson correlation.
@@
 \caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
-\label{fig:Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
+\label{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
 \end{figure}
@@
 \caption{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
-\label{fig:Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
 \end{figure}
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning].
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning].
 Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.
@@
 Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside islands is less influenced by $IPD$.
-In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD.
+In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD$.
 From these plots we can now see that the apparent sharp decline in correlations over the first $50$ bp from the genome-wide data is really driven by a shift from pairs that are frequently inside CpG islands to pairs that are mostly outside of islands.
@@
 \caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
-\label{fig:Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
+\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
 \end{figure}
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} are the plots of CpG $\beta$-value correlations against $IPD$ for $NIL \geq 0$ pairs.
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} are the plots of CpG $\beta$-value correlations against $IPD$ for $NIL \geq 0$ pairs.
 In contrast to the $NIL = 0$ results, the correlations here are relatively stable, at least for $IPD = 2, \ldots, 2000$[^ipd_choice].
@@
 
-__UP TO HERE__
-
-- Complications of interpreting NIL >= 0 plots
-- Periodicities or lack thereof
-- CGI vs. non-CGI
+When we aggregate over as many factors as we are in these genome-wide $NIL \geq 0$ plots, we find that there is little difference in the correlation curves between samples.
+All samples have a very high correlation, close to $1$, for small $IPD$s that steadily decays to a lower correlation of $0.2-0.4$ by $IPD = 2000$.
 
@@
 \caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
-\label{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
+\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
 \end{figure}
+
+Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].
+
+[^hepg2]: The minor exception is the _HepG2\_cell\_line_ where the CGI and non-CGI lines intersect by $IPD = 1500$, but we have already seen that this sample has an unusual methylome compared to the other samples.
+
+__TODO: How often are points in these plots from different CGIs? Compute distribution of CGI widths and gaps between them__
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
+\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
+\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
+\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
+\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+__TODO: Periodicities__
+- The periodicities (10? and 167ish) are more noticeable in the non-island pairs than the pairs in CpG islands. Possible reasons?
+- A shorter periodicity over the first 100 bp?
 
@@
 The $NIL = 0$ results measure the 'first-order' or 'one-step' correlations of $\beta$-values whereas the $NIL \geq 0$ measure a mixture of different orders of correlations of $\beta$-values.
-For this reason I find the $NIL = 0$ plots are easier to interpret, however, which is more relevant will depend on the application.
+For this reason I find the $NIL = 0$ plots are easier to interpret, however, they do not tell the whole story.
+
+By comparing the $NIL = 0$ and $NIL \geq 0$ results we see that the strength of the correlation depends on not just the distance between the two CpGs but the number of intervening CpGs.
+This indicates that these correlations are more than 'first-order'.
+Moreover, the differences in correlations between CpG islands and non-islands indicates that the 'order' of these correlations itself varies across the genome.
+All this means that the vector of $\beta$-values for a single sample forms a complicated and highly non-stationary process with variable-order correlations.
+
+This is valuable information in trying to account for correlated $\beta$-values in downstream analyses.
+However, it is based on multiple levels of aggregation, such as aggregation of cells to form $\beta$-values and such aggregations may hide the source of these correlated $\beta$-values.
+To better understand what leads to correlated $\beta$-values, I investigated within-fragment co-methylation.
 
@@
 
-The key advantage of studying within-fragment co-methylation is that measures the dependence of DNA methylation at the scale on which the biological process acts.
+The key advantage of studying within-fragment co-methylation is that it measures the dependence of DNA methylation at the scale on which the biological process acts.
 I have developed a method based on odds-ratios to quantify the strength of co-methylation for a pair of methylation loci in a single sample.
@@
 
-Finally, a similar analysis could be performed using a different measure of dependence, which I discuss in Section \ref{sec:choice_of_dependence_measure}.
+A similar analysis could be performed using a different measure of dependence, which I discuss in Section \ref{sec:choice_of_dependence_measure}.
 
@@
 However, we are now limited to analysing co-methylation of loci that can be captured within a single read.
-In practice, this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD \lessapprox 200$.
+With current technology this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD \lessapprox 250$.
 

### 2014-12-22 Rmarkdown/comethylation.Rmd
Discussion of LORs

@@
 
-__TODO: How often are points in these plots from different CGIs? Compute distribution of CGI widths and gaps between them__
-
 \begin{figure}[h!]
@@
 
-__TODO: Periodicities__
-- The periodicities (10? and 167ish) are more noticeable in the non-island pairs than the pairs in CpG islands. Possible reasons?
-- A shorter periodicity over the first 100 bp?
+The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands. __TODO: Discuss more__
 
@@
 
-Recall that for each 2-tuple and sample we can summarise the methylation patterns by a $2 \times 2$ contingency table (__FIGURE__).
-The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the logarithm of the odds ratio of this table, $LOR_{(i, i'), j} = log_2 \Big (\frac{MM_{i, i'} \times UU_{i, i'}}{MU_{i, i'} \times UM_{i, i'}} \Big)$.
+Recall that for each 2-tuple and sample we can summarise the methylation patterns by a $2 \times 2$ contingency table, like that shown in Table \ref{tab:2x2}.
+The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the logarithm of the odds ratio of this table, $LOR_{(i, i'), j} = \log_{2} \Big (\frac{mm_{i, i'} \times uu_{i, i'}}{mu_{i, i'} \times um_{i, i'}} \Big)$.
 I use base-2 logarithms due to their widespread use and familiarity within biology.
 In the `cometh` function I also add a small offset (default `offset = 1`) to avoid problems arising from zero counts, which frequently occur.
+For the example in Table \ref{tab:2x2} with `offset = 1` we have $LOR = \log_{2} \Big (\frac{(1 + 1) \times (4 + 1)}{(0 + 1) \times (4 + 1)} \Big) = 1.74$.
 
-A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs in the pair on the same DNA fragment, a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG in the pair on the same DNA fragment, and a value close to zero is evidence that the methylation state at each CpG in the pair on the same DNA fragment are independent of one another.
-More precisely, we may wish to compute a confidence interval for $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.
+From a Bayesian perspective, the offset may be interpreted as a prior count from a table with no association, i.e., a table with $LOR = 0$.
+The offset then has the effect of shrinking the raw estimate of the $LOR$, with more shrinkage applied for larger values of `offset`.
+Analyses using different offsets should be compared with caution.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{lcccl}
+                           & \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
+                           &                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
+\multirow{2}{*}{First CpG} & \textbf{M}           & 1              & 2             & 3              \\
+                           & \textbf{U}           & 0              & 4             & 4              \\ \cline{2-5}
+                           & \textbf{Total}       & 1              & 6             & 7
+\end{tabular}
+\caption{Example $2 \times 2$ contingency table summarising methylation patterns at a 2-tuple.}
+\label{tab:2x2}
+\end{table}
+
+A positive $LOR$ is evidence that the methylation states of two CpGs on the same DNA fragment are likely to be the identical, a negative $LOR$ is evidence that the methylation states of two CpGs on the same DNA fragment are likely to be the different, and a value close to zero is evidence that the methylation states of two CpGs on the same DNA fragment are independent of one another.
+More precisely, we may wish to compute a confidence interval for the $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.
+
+There are several alternative measures of association for a $2 \times 2$ table of binary variables such as __EXAMPLES__
 
 __TODO: Comparison of LORs and Pearson correlations of Landan__
+
+- Plot histogram of LOR and histogram of r
+
+- Look for tables with LOR > 0 and r = 0 (and vice versa?)
 
@@
 - iPSC vs. somatic
-- Plot histogram of LOR and histogram of r
-- Look for tables with LOR > 0 and r = 0
 
@@
 __TODO__
+
+\section{Summary}
+
+- Discuss 1D nature of $IPD$s vs. $4D$ structure of genome (3D + time)
 

### 2014-12-23 Rmarkdown/comethylation.Rmd
Rough draft of co-methylation chapter complete-ish.

@@
 
+\FloatBarrier
 \section{Correlations of $\beta$-values}\label{subsec:beta_cor}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 
+\FloatBarrier
 \subsection{Results}
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show that the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
 Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 
+\FloatBarrier
 \subsubsection{Choice of correlation coefficient}
@@
 
+\FloatBarrier
 \subsubsection{$NIL = 0$}
@@
 
+\FloatBarrier
 \subsubsection{$NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}
@@
 
-Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].
+Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].
 
@@
 
-The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands. __TODO: Discuss more__
+The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands.
+This may be due to  different regulation of DNA methylation outside of CpG islands or simply because CpG islands have such consistently high correlations that these periodicities are not evident.
 
-\subsection{Summary}
-
-The $NIL = 0$ results measure the 'first-order' or 'one-step' correlations of $\beta$-values whereas the $NIL \geq 0$ measure a mixture of different orders of correlations of $\beta$-values.
-For this reason I find the $NIL = 0$ plots are easier to interpret, however, they do not tell the whole story.
-
-By comparing the $NIL = 0$ and $NIL \geq 0$ results we see that the strength of the correlation depends on not just the distance between the two CpGs but the number of intervening CpGs.
-This indicates that these correlations are more than 'first-order'.
-Moreover, the differences in correlations between CpG islands and non-islands indicates that the 'order' of these correlations itself varies across the genome.
-All this means that the vector of $\beta$-values for a single sample forms a complicated and highly non-stationary process with variable-order correlations.
-
-This is valuable information in trying to account for correlated $\beta$-values in downstream analyses.
-However, it is based on multiple levels of aggregation, such as aggregation of cells to form $\beta$-values and such aggregations may hide the source of these correlated $\beta$-values.
-To better understand what leads to correlated $\beta$-values, I investigated within-fragment co-methylation.
-
+\FloatBarrier
 \section{Within-fragment co-methylation}\label{subsec:wf_cometh}
@@
 
-Here I describe a simple method based on analysing methylation patterns at 2-tuples[^higher_order_m-tuples].
+Here I describe a simple method based on analysing methylation patterns at 2-tuples.
 For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type.
 This method is implemented in the `cometh` function that is part of the `MethylationTuples` software.
-
-[^higher_order_m-tuples]: Extensions to m-tuples with $\text{m} > 2$ are discussed in __SECTION__.
 
@@
 
+\FloatBarrier
 \subsubsection{Choice of dependence measure}\label{sec:choice_of_dependence_measure}
+
+As discussed, I use the logarithm of the odds ratio to summarise within-fragment co-methylation. Here I explain my reasons for choosing this measure of association.
 
@@
 
-There are several alternative measures of association for a $2 \times 2$ table of binary variables such as __EXAMPLES__
+There are several alternative measures of association for a $2 \times 2$ table of binary variables.
+An interesting review of these is given by \citet{Kraemer:2006te}.
+The Pearson correlation coefficient, also known as the phi coefficient when applied to binary $2 \times 2$ tables, was used by \citet{Landan:2012kp} to analyse within-fragment co-methylation.
+There are a few reasons that I prefer $LOR$ to Pearson's correlation.
 
-__TODO: Comparison of LORs and Pearson correlations of Landan__
+Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' (__CITE: quote from \cite{Kraemer:2004el} but references two other papers.
+Find original quote__).
+Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints (__CITE: Plackett?__).
+In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities (__CITE: Kraemer HC.
+A measure of 2 × 2 association with stable variance and approximately normal small sample distribution: planning cost-effective studies.__), which I find undesirable.
 
-- Plot histogram of LOR and histogram of r
+Secondly, the empirical distribution of $LOR$ for all 2-tuples in a sample  (Figure \ref{fig:EPISCOPE_lor_histogram}) is much nicer than that for Pearson's correlation coefficient (Figure \ref{fig:EPISCOPE_pearson_histogram}).
+Specifically, the distributions of $LOR$ for each sample are approximately unimodal and symmetric whereas the distributions of Pearson's correlation are multimodal, asymmetric, and clearly constrained by the underlying marginal probabilities.
+It is far easier to work with symmetric, unimodal distributions, such as when wanting to summarise these by measures of central tendency and spread, which is important in what follows.
 
-- Look for tables with LOR > 0 and r = 0 (and vice versa?)
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_histogram.pdf}
+\caption{Histograms of $LOR$ for 2-tuples from the EPISCOPE data (binwidth $= 0.5$).}
+\label{fig:EPISCOPE_lor_histogram}
+\end{figure}
 
-__TODO: Discussion of Kraemer's work, in particular, the fact that non-zero LOR are not informative for the strength of the correlation__
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_histogram.pdf}
+\caption{Histograms of Pearson correlation coefficient for 2-tuples from the EPISCOPE data (binwidth $= 0.05$).}
+\label{fig:EPISCOPE_lor_histogram}
+\end{figure}
 
-__TODO: \cite{Bonett:2007bd} is a reference that Pearson correlation of 2x2 is phi coefficient__
-
-\subsection{Results}
-
-$LOR$s are only computed for 2-tuples with sufficient sequencing coverage (here, $5 \times$ for stranded 2-tuples and $10 \times$ for unstranded 2-tuples).
-For any given $IPD$ there are hundreds of thousands of CpG pairs, and so there is a distribution of co-methylation for any given $IPD$.
-We could visualise the distribution for a given $IPD$ using classical visualisation techniques such as histograms, kernel density estimates and boxplots.
-These summaries may be stratified by one or more annotations.
-However, what these $IPD$-specific plots don't give the reader is a sense of how $IPD$ affects within-fragment co-methylation.
-
-Instead, for each sample I plot selected quantiles --- $10\%, 25\%, 50\%, 75\% and 90\%$ --- of each $IPD$-specific $LOR$ distribution as a function of $IPD$.
-I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$.
-These quantile plots may also be stratified by additional annotations.
-
-Key points:
-
-- LOR decreases with IPD
-- For a given IPD, a wide variation in LOR
-- LOR does not go to zero over observable distance
-- Variation is fairly constant across IPD
-- Noise at the end is due to limited data
-- CGIs are more co-methylated than non-CGI
-- SE data can't cut the mustard
-- Intriguing mode/plateua around 150-200 bp (NIL >= 0)
-  - SE data suggests some is technical, e.g., incomplete removal of M-bias
-  - Also, IMR90 PE data doesn't have it
-- LOR vs. Pearson
-- NIL = 0 vs. NIL >= 0
-- iPSC vs. somatic
-
-An example quantile plot for the __WHICH SAMPLE__ is shown in __FIGURE__.
-The vast majority of $LOR$s are positive, meaning that most 2-tuples are co-methylated.
-The strength of co-methylation decays roughly as $IPD$ increases, which can be seen by the downward shift in the quantiles as a function of $IPD$.
-I
-
-__FIGURES__ show that the two DNA strands have near identical $LOR$ distributions.
-This is not to discount the possibility of 2-tuples with strand-specific differences in co-methylation, however, it does mean that it is generally safe to aggregate counts across strands.
-
-__FIGURES__
-
-__TODO__
-
-\section{Summary}
-
-- Discuss 1D nature of $IPD$s vs. $4D$ structure of genome (3D + time)
-
-__OLD WRITING BELOW THIS LINE__
-
-### Visualisations
-
-For a typical whole-genome bisulfite-sequencing experiment of human DNA, tens of millions of neighbouring CpG pairs are sequenced (and hundreds of millions of non-neighbouring CpG pairs).
-For every pair of CpGs, $(i, i')$, in the $j^{th}$ sample we know the $IPD$ and have computed a log-odds ratio, $LOR$_{(i, i'), j}$.
-Each such pair may have further annotations, such as the genomic context of the pair (e.g.
-within a CpG island, within a gene body, within a repetitive region, etc.).
-We need a simple way to summarise these data so that we might explore the effect of each of these variables on within-fragment co-methylation.
-
-For any given $IPD$ there are hundreds of thousands of CpG pairs and so there is a _distribution_ of co-methylation for any given $IPD$.
-We can visualise the distribution for a given $IPD$ using classical visualisation techniques such as histograms, kernel density estimates and boxplots (__TODO: Example figures of each of these__).
-These summaries may be stratified by one or more annotations (__TODO: Example figures with stratification__).
-
-What these plots don't give the reader is a sense of how $IPD$ affects within-fragment co-methylation.
-__FIGURE__ and __FIGURE__ (__TODO:__ two plots of co-methylation as boxplots, one with outliers plotted and one without) show initial attempts to create a visualisation that shows the relationship between $LOR$ and $IPD$.
-Unfortunately, these plots are rather "busy".
-
-Instead, I plot selected quantiles of each $IPD$-specific $LOR$ distribution as a function of $IPD$ (__FIGURE__).
-These convey the same information as the series of boxplots in a simpler and cleaner fashion.
-I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$.
-These quantile plots may also be stratified by additional annotations (__FIGURE__).
-
-The simplicity and interpretability of the log-odds ratio make the above described method an attractive one for studying within-fragment co-methylation.
-However, by using 2-tuples we are limited to exploring "one-step" co-methylation, i.e., one-step dependencies of the stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
-In the next section I describe the difficulties of using higher order $m-tuples$ ($m > 2$) to explore within-fragment co-methylation.
-
-## Using higher-order $m$-tuples of methylation loci
-
-In general, to explore $m$-step dependencies we need to use $(m + 1)$-tuples.
-We can generalise the use of a two-way contingency table for studying co-methylation of a 2-tuple to the use of an $m$-way contingency table for co-methylation of an m-tuple (see Table ~\ref{reads_containing_three_CpGs_counts}).
-
-__TODO: Add real numbers to table__
+Finally, and related to the second point, there are $2 \times 2$ tables that are clear evidence of co-methylation that have a Pearson correlation close to zero.
+An example is shown in Table \ref{tab:2x2_big_lor_small_cor}.
+This table has a $LOR = 3.3$ and Pearson's correlation of $-0.025$, which is due to the table being dominated by one cell's count, $uu = 39$.
+This occurs for all tables where one cell dominates.
+The problem is only partially fixed by adding an offset.
+For example, adding $1$ to all cells gives a Pearson's correlation of $0.29$, which is still a low correlation.
 
@@
 \centering
-\begin{tabular}{@{}cccc@{}}
-\toprule
-           &            & \multicolumn{2}{c}{Third CpG} \\ \cmidrule(l){3-4}
-First CpG  & Second CpG & \textbf{M}    & \textbf{U}    \\ \midrule
-\textbf{M} & \textbf{M} &      4        &      0        \\
-\textbf{}  & \textbf{U} &      2        &      0        \\
-\textbf{U} & \textbf{M} &      0        &      1        \\
-\textbf{}  & \textbf{U} &      1        &      1        \\ \bottomrule
+\begin{tabular}{lcccl}
+& \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
+&                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
+\multirow{2}{*}{First CpG} & \textbf{M}           & 0              & 1             & 1            \\
+& \textbf{U}           & 1              & 39             & 40             \\ \cline{2-5}
+& \textbf{Total}       & 1              & 40             & 41
 \end{tabular}
-\caption{Example three-way table summarising read counts at a 3-tuple.}
-\label{reads_containing_three_CpGs_counts}
+\caption{A $2 \times 2$ contingency table with a large $LOR$ but small Pearson's correlation.}
+\label{tab:2x2_big_lor_small_cor}
 \end{table}
 
-     chr strand pos1 pos2 pos3 MMM MMU MUM MUU UMM UMU UUM UUU cov
- 4: chr1      +  489  493  497   4   0   2   0   0   1   1   0   8
+\FloatBarrier
+\subsection{Results}
 
-However, as we increase $m$ we have more parameters to estimate - an $m$-way table has $\binom{m}{2}$ two-way dependencies, $\binom{k}{3}$ three-way dependencies, $\ldots$, and $\binom{m}{m} = 1$ $m$-way dependency.
+For all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets I compute the $LOR$ of all 2-tuples using both the $NIL = 0$ and $NIL \geq 0$ strategies.
+I compute these genome-wide and after stratifying CpG 2-tuples by whether they are in a CpG island.
+These analyses are performed separately for each strand and also for strand-collapsed counts of methylation patterns.
 
-Furthermore, as we increase $m$ we have fewer data points with which to estimate these parameters, since we only use reads that contain all $m$ methylation loci in the $m$-tuple (__TODO: Bar chart of how many CpGs per read__).
-This makes the estimates quite noisy unless some form of regularisation is applied, such as a parametric model of the dependencies or a non-parametric, data-driven smoothing of the estimates.
-I explore the use of the Expectation-Maximisation algorithm in __SECTION__ to incorporate all reads that overlap at least one methylation loci in the $m$-tuple.
+For any given $IPD$ there are hundreds of thousands of CpG pairs, and so there is a _distribution_ of co-methylation for any given $IPD$.
+For a given $IPD$ we can visualise the distribution of $LOR$ using a standard plot such as a histogram, kernel density estimate or boxplot.
+However, these $IPD$-specific plots don't give a sense of how $IPD$ affects within-fragment co-methylation.
+For this reason I instead plot selected quantiles --- $10\%, 25\%, 50\%, 75\% and 90\%$ --- of each $IPD$-specific $LOR$ distribution as a function of $IPD$.
+I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$[^quantile].
+These quantile plots may also be stratified by additional annotations such as genomic features.
 
-### An EM algorithm to estimate co-methylation
+[^quantile]: Because the distributions of $LOR$ are approximately symmetric and unimodal, these quantiles give a fair summary of each of these distributions.
+The same would not be true of the highly asymmetric and multimodal distributions of Pearson's correlations, hence one reason for my preference of $LOR$ over Pearson's correlation.
 
-__TODO:__ Discuss with Terry the idea of fitting a log-linear model to the counts of methylation states using all reads that overlap at least one methylation loci in the tuple.
-Effectively, this is a log-linear model with missing data.
+\FloatBarrier
+\subsubsection{Can we collapse by strand?}
 
-### Visualisations
+Figures \ref{fig:EPISCOPE_lor_min_cov_5}, \ref{fig:Lister_lor_min_cov_5}, \ref{fig:Seisenberger_lor_min_cov_5} and \ref{fig:Ziller_lor_min_cov_5} show the $LOR$ quantile plots for $NIL = 0$ CpGs 2-tuples stratified by strand.
+Since the distributions of $LOR$ are very similar for the two strands, I decided to use the strand-collapsed counts of methylation patterns in what follows.
+This is not to discount the possibility of 2-tuples with strand-specific differences in co-methylation, however, it does mean that it is generally safe to aggregate counts across strands.
 
-Aside from estimating this increased number of parameters, we must also interpret these with respect to $IPD$ and other annotations.
-The $IPD$ vector for $m$-tuples has $(m - 1)$ elements meaning we now have to explore the effect of $IPD$ on $LOR$ over an $(m - 1)$-dimensional _grid_ rather than along a 1-dimensional line.
-Again, the sparsity of the data make this difficult - approximately $20,000 - 30,000$ unique $IPD$s are observed for 3-tuples - but even more difficult is creating a clear and coherent visualisation of this data.
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_5.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the EPISCOPE dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:EPISCOPE_lor_min_cov_5}
+\end{figure}
 
-As is clear from __FIGURE__, these is very difficult for 3-tuples, let alone for higher-order $m$-tuples.
-We have an $(m - 1)$ grid of $IPD$s and a _distribution_ of $LOR$ for each $IPD$.
-In the case of 2-tuples I used lines to summarise the quantiles of the $LOR$ distributions, but this is not effective when $m > 2$.
-Instead, we must create separate figures for each quantile, which makes it difficult to interpret the effects of $IPD$ on $LOR$.
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_5.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the Lister dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Lister_lor_min_cov_5}
+\end{figure}
 
-## Results of co-methylation analyses
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_5.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the Seisenberger dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Seisenberger_lor_min_cov_5}
+\end{figure}
 
-## A Markov model of within-fragment co-methylation
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_5.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the Ziller dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Ziller_lor_min_cov_5}
+\end{figure}
 
-__Possibly better off in the chapter on simulation__
+\FloatBarrier
+\subsubsection{$NIL = 0$}
 
-A $k$-order Markov model of within-fragment co-methylation assumes that the dependence of $Z_{h, i}$ on the "history" of the process, $(Z_{h, 1}, \ldots, Z_{h, i - 1})$ is through the $k$ most recent observations, $Z_{h, i - k}, \ldots, Z_{h, i - 1}$.
-This could be generalised via a Markov random field model of within-fragment co-methylation to assume that the dependence of $Z_{h, i}$ on the "rest" of the process $Z_{h, 1}, \ldots, Z_{h, i - 1}, Z_{h, i + 1}, Z_{N_{loci}}$ is through the k-nearest observations, $Z_{h, i - k_{1}}, \ldots, Z_{h, i - 1}, Z_{h, i + 1}, \ldots, Z_{h, i + k_{2}}$, where $k_{2} - k_{1} = k$.
+Figures \ref{fig:fig:EPISCOPE_lor_min_cov_10_strand_collapsed}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed} show the $LOR$ quantile plots for strand-collapsed $NIL = 0$ CpGs 2-tuples.
+Like for the correlations of $\beta$-values, two things immediately stand out.
+Firstly, for all samples the trend is that the strength of the within-fragment co-methylation decays as a function of $IPD$ --- 2-tuples with a larger $IPD$ are on average less co-methylated than2-tuples with a smaller $IPD$.
+Secondly, the estimated $LOR$s violently jump around for larger $IPD$s.
+The latter is simply due to the aforementioned lack of $NIL = 0$ pairs at larger $IPD$s with which to estimate these $LOR$s.
+Therefore, it is the trend in these plots that should be analysed.
 
-Regardless of whether we use a Markov chain or a Markov random field, a complication is that $Z_{h}$ is unevenly spaced along the genome.
-I use the intra-pair distance ($IPD$) to measure the distance between observations along $Z_{h}$.
-This distance measure introduces additional dimension(s) to our model.
+It is also obvious that longer paired-end reads are far more preferable to shorter single-end reads when studying within-fragment co-methylation.
+Even when using paired-end reads, getting to $IPD = 150$ with $NIL = 0$ pairs is challenging.
+This is unfortunate because it doesn't allow us to investigate the potential nucleosome-related $\approx 150-170$ bp periodicity of methylation levels using $NIL = 0$ pairs.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the EPISCOPE dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the Lister dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Lister_lor_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the Seisenberger dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the Ziller dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Ziller_lor_min_cov_10_strand_collapsed}
+\end{figure}
+
+The are some consistent patterns in comparing these quantile plots across samples.
+Those 2-tuples with a small $IPD$ have a median $LOR = 3-4$ meaning that the loci in these 2-tuples are $8-16\times$ more likely to have the same rather than opposite methylation states.
+However, we also see than for any given $IPD$ that there is considerable variation in $LOR$.
+For example, for all samples and $IPD$s, the $10\%$ and $90\%$ quantiles span $3-5$ (base 2) orders of magnitude.
+
+For each sample the $LOR$ distributions are consistently centred away from zero, meaning that over the observable $IPD$s the vast majority of 2-tuples are co-methylated.
+As $IPD$ increases, the CpG $NIL = 0$ 2-tuples become less co-methylated on average.
+However, the width of each sample's $LOR$ distributions remains relatively constant, so the variation in $LOR$ doesn't greatly depend on $IPD$.
+
+While the similarities between samples are obvious, the differences are somewhat harder to detect from these genome-wide summaries.
+The ADS and FF mini datasets provide us with an opportunity to discuss some potential biological differences in within-fragment co-methylation.
+In Figure \ref{fig:Lister_lor_min_cov_10_strand_collapsed} we see that the _ADS-iPSC_ sample has a consistently higher level of co-methylation than the _ADS_ and _ADS-adipose_ samples.
+Likewise, the _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ have consistently higher levels of co-methylation than the _FF_  sample, although not the _FF-iPSC\_19.11+BMP4_ sample.
+
+I can think of two possible interpretations of this result.
+Firstly, iPSCs may have a more regulated methylome than do somatic cells.
+The fact that these same iPSCs have very little intermediate methylation could be seen as support of this interpretation.
+However, this same lack of intermediate methylation may simply reflect that the iPSC samples are a more homogeneous population of cells.
+It would be possible to have a sample with no within-fragment co-methylation but where all $\beta$-values are 0 or 1, e.g., a sample where all cells are perfect clones of a cell whose CpG methylation states are independent of one another.
+
+A second interpretation of this result is that it is due to 'reseting' of the methylome upon induction of pluripotency \citep{Lister:2011kg}.
+As a population of cells develops from a single cell, there are errors in copying the DNA methylation patterns from mother to daughter cell.
+These errors disrupt the co-methylation of 2-tuples.
+If the iPSC and iPSC-derived cell lines have had less (developmental) time between the laying down of the original DNA methylation pattern and the point at which they were sequenced, then there is a lower probability that the co-methylation has been disrupted by replication errors.
+However, high levels of within-fragment co-methylation does not on its own imply that the sample is homogeneous.
+It would be possible to have a sample with high levels of within-fragment co-methylation that also has high levels of intermediate methylation, e.g., a sample with multiple distinct clonal subpopulations where each subpopulation has high levels of co-methylation but  also has a distinct methylome.
+
+Taking the co-methylation results and the distribution of $\beta$-values together, I conclude that for the ADS and FF mini datasets the iPSC and iPSC-derived samples truly have higher levels of co-methylation and are also more homogeneous than their precursor cells.
+
+The above analyses are all based on the genome-wide distributions of within-fragment co-methylation.
+Again, we can stratify these by whether the CpG 2-tuples are inside or outside of a CpG island.
+Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed_CGI} are plots of $LOR$ distributions for $NIL = 0$ CpG 2-tuples stratified by whether the pair is inside or outside of a CpG island[^spanning].
+Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.
+
+Again, the message from these plots is clear, $NIL = 0$ CpG 2-tuples inside CpG islands have higher levels of within-fragment co-methylation on average than those 2-tuples outside of islands[^hepg2_2].
+For each sample, the difference between the CpG island and non-island distributions is a vertical translation.
+For many samples this vertical shift means that the median $LOR$ for islands is approximately equal to the $75\%$ percentile for non-islands.
+The width of these distributions is not greatly affected by the CpG island status.
+
+[^hepg2_2]: Again, the minor exception is the _HepG2\_cell\_line_ where the CGI and non-CGI distributions approximately overlap by $IPD = 50$, but we have already seen that this sample has an unusual methylome compared to the other samples.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the EPISCOPE dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the Lister dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Lister_lor_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the Seisenberger dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the Ziller dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Ziller_lor_min_cov_10_strand_collapsed_CGI}
+\end{figure}
+
+\FloatBarrier
+\subsubsection{$NIL \geq 0$}\label{sec:cometh_nil_geq0}
+
+Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac} and \ref{fig:Ziller_merged_lor_min_cov_10_strand_collapsed_ac} show the $LOR$ quantile plots for strand-collapsed $NIL \geq 0$ CpGs 2-tuples.
+By using $NIL \geq 0$ 2-tuples we can explore the effect of within-fragment co-methylation at greater $IPD$s, potentially beyond the crucial $IPD = \approx 150-170$ bp distance.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed_ac.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the EPISCOPE dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed_ac.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the Lister dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Lister_lor_min_cov_10_strand_collapsed_ac}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed_ac.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the Seisenberger dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed_ac.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the Ziller dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Ziller_lor_min_cov_10_strand_collapsed_ac}
+\end{figure}
+
+These plots appear to show that there is a dramatic increase in within-fragment co-methylation at precisely these $IPD$s for paired-end samples from the Lister dataset.
+However, I am very skeptical as to the biological relevance of these results.
+
+There are three reasons for my skepticism.
+Firstly, similar results are not observed for any of the EPISCOPE samples nor any of the Ziller samples[^mouse_vs_man].
+Secondly, the Lister samples with single-end sequencing have a similar spike in the estimated within-fragment co-methylation at the edge of the observable $IPD$s.
+However, for single-end data this corresponds to an $IPD \approx 60-80$ bp and we see no evidence of an increase in within-fragment co-methylation at these $IPD$s in any of the paired-end samples.
+This leads to my final reason, and what I think is the source of this artefact.
+Namely, estimates of within-fragment co-methylation are highly susceptible bias due to the incomplete removal of M-bias.
+This is particularly true for larger $IPD$s where the loci in the 2-tuples come from the ends of the reads, which are where M-bias is most prevalent.
+
+[^mouse_vs_man]: I exclude the Seisenberger data simply because these are of mice rather than humans.
+The Seisenberger data also display a dramatic increase in within-fragment co-methylattion but since this increase occurs at the edge of the observable $IPD$s, I am concerned it suffers from the same bias as the Lister data.
+
+As discussed in Section \ref{sec:m-bias}, the incomplete removal of M-bias is a real problem for reads that have undergone pre-trimming because this destroys the relationship between read-position and sequencing cycle.
+__FIGURE__ shows that there is considerable variation in the read lengths of the _ADS_ sample, indicative of aggressive pre-trimming of these reads (__TODO: Histogram of read lengths__).
+
+To investigate my suspicion that these results are due to the incomplete removal of M-bias, I used an aggressive filtering strategy that attempts to only use read-positions that are unlikely to suffer from M-bias.
+I applied this to the _ADS_ sample.
+In this analysis I ignored read-positions $1-20$ and $55-80$ of both _read 1_ and _read 2_.
+Unfortunately, this is a situation where the M-bias filtering strategy of `methtuple` is a disadvantage.
+`methtuple` is designed to filter out specific read-positions and cannot 'exclude the first/last $n$ bases'.
+For example, using this aggressive strategy will have a detrimental effect for a read that has been pre-trimmed to $30$ bp, of which the last $10$ bp are affected by M-bias, because this filtering strategy will __retain__ read-positions $21-30$, which are exactly those affected by M-bias.
+Nevertheless, it is the best that can be done with the current version of the software[^methtuple_refactoring] and with such aggressively pre-trimmed data.
+
+[methtuple_refactoring]: We have now seen that both the 'filter specific read-positions' and 'filter the first/last $n$ read-positions' can lead to undesirable behaviour and so switching strategies is no saviour.
+The real solution is not to pre-trim reads and to instead use an aligner such as `bwa-meth` that can softclip these positions.
+Unfortunately, `bwa-meth` was only available after my data were already aligned and after `methtuple` was written.
+To use `methtuple` with `bwa-meth` output requires significant changes to `methtuple` because the `bwa-meth` output does not include anything analogous to the XM-tag of `Bismark`, which `methtuple` uses to do methylation calling.
+
+The results for the aggressively trimmed _ADS_ sample are shown in __FIGURE__.
+
+__TODO: Comment on these results and how we should interpret the Lister co-methylation results (particularly NIL >= 0)__
+
+__TODO: Summarise the M-bias issue__
+
+Comparing CpG islands to non-islands we again, Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac_CGI}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_ac_CGI}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac_CGI} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed_ac_CGI}, see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the EPISCOPE dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the Lister dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Lister_lor_min_cov_10_strand_collapsed_ac_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the Seisenberger dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
+\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the Ziller dataset.
+The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
+Both the x-axis and y-axis have been truncated for clarity.}
+\label{fig:Ziller_lor_min_cov_10_strand_collapsed_ac_CGI}
+\end{figure}
+
+\FloatBarrier
+\section{Summary}
+
+Correlations of $\beta$-values and estimates of within-fragment co-methylation are complementary methods to better understand the complex dependence structure of bisulfite-sequencing data.
+Which is more relevant depends on the question at hand.
+Within-fragment co-methylation seems closer to the biology because it measures the dependence of DNA methylation on the scale that the DNA methyltransferases act.
+However, the most common downstream analyses are based on $\beta$-values and so these correlations may be more relevant to analyses of DNA methylation data.
+Ultimately, the two measures are intertwined since the correlations of $\beta$-values are driven by co-methylation on individual DNA fragments.
+This idea is explored in greater detail in Chapter \ref{chap:methsim}.
+
+A major challenge in interpreting within-fragment co-methylation is its susceptibility to M-bias or, rather, the difficulty of accounting for M-bias in pre-trimmed data.
+Since we can do nothing about data that has already been pre-trimmed, we are forced to be cautious in our interpretation of within-fragment co-methylation for 2-tuples with larger $IPD$s.
+This is particularly relevant to the increase in within-fragment co-methylation for $NIL \geq 0$ CpG 2-tuples with $IPD \approx 150-170$.
+This result is tantalisingly similar to the periodicity we see in the correlations of $\beta$-values and the approximate spacing of nucleosomes in human DNA (see Section \ref{sec:nucleosome_and_chromatin}).
+However, I do not feel comfortable concluding that this is truly what we are observing because of the potential for these systematic biases to create an artificial signal.
+Furthermore, we do not detect this signal any of the EPISCOPE samples, which are some of the best quality data I have and, being samples from human tissues rather than cell lines, are arguably more biologically relevant.
+
+The presented analyses only measure pairwise dependencies.
+More specifically, the $NIL = 0$ results measure the 'first-order' or 'one-step' correlations of $\beta$-values whereas the $NIL \geq 0$ measure a mixture of different orders of dependences but all are done in a pairwise manner.
+I find the $NIL = 0$ results easier interpret but they clearly do not tell the whole story.
+
+By comparing the $NIL = 0$ and $NIL \geq 0$ results we see that the strength of the dependency itself depends on not just the distance between the two CpGs but the number of intervening CpGs.
+This indicates that these dependencies are more than 'first-order'.
+Moreover, the differences in dependencies between CpG islands and non-islands indicates that the 'order' of these dependencies itself varies across the genome.
+All this means that the DNA methylation measurements for a single sample form a complicated and highly non-stationary process with variable order dependencies.
+
+In theory, we could explore these higher-order dependencies of within-fragment co-methylation by analysing m-tuples with $\text{m} > 2$.
+However, the complexity of the challenge increases exponentially with m.
+For example, even when analysing 3-tuples there are four $LOR$s for each $2 \times 2 \times 2$ contingency table and the $IPD$ vector now has two dimensions, meaning that the results have to be visualised over a surface.
+
+Another complexity that these analyses haved brushed over is the fact that the genome is not a 1D line but is rather a complex 3D structure whose shape also varies in time.
+Therefore, the relevant distance between two loci is not necessarily the number of nucleotides between them but may instead be the Euclidean distance in 3D space.
+Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the 3D structure of the genome, such as those provided by chromatin conformation capture technologies \citep{Dekker:2013hi}.
+
+The results of this chapter clearly demonstrate that DNA methylation has a highly complex dependence structure.
+These dependencies exist on individual DNA fragments and also manifest as correlations of aggregate measures of methylation, such as $\beta$-values.
+Bisulfite-sequencing give the highest resolution picture of this dependence structure.
+Nonetheless, this is still a rather limited picture, particularly for within-fragment co-methylation.
+Data from longer reads will be immensely useful in developing a better model of within-fragment co-methylation.
+In particular, reads that can span multiple nucleosomes will allow for co-methylation to be better tied in with and expand upon existing results on the relationship between nucleosomes and DNA methylation.
+Until such data are available, we can try to learn more about plausible models of DNA co-methylation through computational methods such as the simulation method described in the next chapter.

### 2014-12-23 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Figure fixing.

@@
 Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
-My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded.
+My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded, however, this is not without its own problems, as we shall see in Section \ref{sec:cometh_nil_geq0}.
 

### 2014-12-23 Rmarkdown/comethylation.Rmd
Minor fixes

@@
 In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
-Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci.
+Using both these measures I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci.
 I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.
@@
 Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
-The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples that I am studying.
+I do not explore co-methylation of non-CpG methylation loci because these loci are rarely methylated in the samples that I am studying.
 
@@
 What was lacking from these were clear descriptions of the proposed methods and software implementations.
+In this chapter I try to rectify the former and with the `MethylationTuples` I make available a software implementation of these methods.
 
@@
 For a chromosome with $N$ CpGs there are $(N - 1)$ pairs.
-I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have zero intervening loci^[nil0].
+I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have no intervening loci ($NIL$ = number of intervening loci)^[nil0].
 
-[^nil]: Recall that this is with respect to the reference genome and not the sample genome.
+[^nil0]: Recall that this is with respect to the reference genome and not the sample genome.
 
 The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context.
-In practice, the second strategy only uses pairs for a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy.
+In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy.
 This is more like a traditional autocorrelation analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$, albeit of a vector of unequally spaced observations and with care taken to only use pairs on the same chromosome and strand.
@@
 The interpretation of correlations computed using the $NIL \geq 0$ is complicated because the set of pairs are more heterogeneous than those under the $NIL = 0$ strategy.
-For this reason I prefer the $NIL = 0$ strategy, however, it may be useful to consider and contrast the two strategies, which is what I have done in my thesis.
+For this reason I prefer the $NIL = 0$ strategy, however, it is useful to consider and contrast the two strategies, which is what I have done in my thesis.
 
@@
 
-Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two distinct regions of the genome that have very different methylation dynamics.
-In other words, the vector of $\beta$-values for a given sample is highly non-stationary --- the covariance and correlation between two loci depends on more than just the $IPD$.
+Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two regions of the genome that have very different methylation dynamics.
+In other words, the vector of $\beta$-values for a given sample is highly non-stationary --- the correlation between two loci depends on more than just the $IPD$.
 An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values.
@@
 
-The three commonly used correlation coefficients are Pearson's $r$ (__CITE__), Spearman's $\rho$ (__CITE__) and Kendall's $\tau$ (__CITE__).
+Three commonly used correlation coefficients are Pearson's $r$ \citep{Pearson:1895vt}, Spearman's $\rho$ \citep{Spearman:1904eg} and Kendall's $\tau$ \citep{KENDALL:1938bv}.
 Briefly, Pearson's product-moment correlation is designed to detect linear relationships between two variables.
@@
 
-To address this problem, we might try smoothing the raw correlations, for example, by using LOESS (__CITE__), and then looking for periodicities in the smoothed values.
+To address this problem, we might try smoothing the raw correlations, for example, by using LOESS \citep{Cleveland:1979fu}, and then looking for periodicities in the smoothed values.
 However, this is only helpful when the data are corrupted by a small amount of noise, certainly not the extreme noise we see for $IPD > 1000$ in most samples.
@@
 
-Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' (__CITE: quote from \cite{Kraemer:2004el} but references two other papers.
-Find original quote__).
-Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints (__CITE: Plackett?__).
-In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities (__CITE: Kraemer HC.
-A measure of 2 × 2 association with stable variance and approximately normal small sample distribution: planning cost-effective studies.__), which I find undesirable.
+Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' \citep{Kraemer:2004el}.
+Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints \citep{plackett1981analysis}.
+In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities \citep{kraemer1986measure}, which I find undesirable.
 

### 2014-12-23 Rmarkdown/wgbs_downstream_analyses.Rmd
Latex fixes
Minor fixes

@@
 & Allele 1 & Allele 2 \\ \midrule
-m &  m_{1}        &  m_{2}       \\
-u &  u_{1}        &  u_{2}       \\ \bottomrule
+$m$ &  $m_{1}$        &  $m_{2}$       \\
+$u$ &  $u_{1}$        &  $u_{2}$       \\ \bottomrule
 \end{tabular}
@@
 
-Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing.
-For example, __FIGURE__ shows the number of methylation patterns observed at 4-tuples with a frequency greater than __WHAT__ from the __WHICH__ sample.
+Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing \citep{Landan:2012kp}.
 There has been some progress in developing software to extract and analyse these epialleles, which I now describe.
-(Alternatively, cite Landan et al.)
 

### 2014-12-23 Rmarkdown/wgbs_statistical_framework.Rmd
Added section label
Figure fixing.
Minor fixes

@@
 
-Finally, I describe key variables, common estimators of these and their statistical properties by analysing __HOW MANY__ whole-genome bisulfite-sequencing samples.
+Finally, I describe key variables, common estimators of these and their statistical properties by analysing 40 whole-genome bisulfite-sequencing samples.
 
@@
 
-\subsubsection{Genome-wide distribution of $\beta$-values}
+\subsubsection{Genome-wide distribution of $\beta$-values}\label{sec:genome-wide_distribution_of_beta_values}
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}

### 2014-12-23 Rmarkdown/comethylation.Rmd
Still need to think about NIL >= 0 co-methylation results

@@
 
+__TODO: Re-write everything between @@@@@.
+I'm really not yet sure if this result is biology or an artefact.
+My instinct says artefact but I haven't yet been able to prove this either way to my satisfaction__
+@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
 These plots appear to show that there is a dramatic increase in within-fragment co-methylation at precisely these $IPD$s for paired-end samples from the Lister dataset.
@@
 As discussed in Section \ref{sec:m-bias}, the incomplete removal of M-bias is a real problem for reads that have undergone pre-trimming because this destroys the relationship between read-position and sequencing cycle.
-__FIGURE__ shows that there is considerable variation in the read lengths of the _ADS_ sample, indicative of aggressive pre-trimming of these reads (__TODO: Histogram of read lengths__).
+__FIGURE__ shows that while there is considerable variation in the read lengths of the _ADS_ sample, which is indicative of aggressive pre-trimming of these reads, most reads are full length or close to it (__TODO: Histogram of read lengths__).
 
@@
 __TODO: Summarise the M-bias issue__
+@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
 

### 2014-12-23 Rmarkdown/datasets.Rmd
Figure fixing.
Typo

@@
 M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set.
-These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level more than $0.03$ from the median, i.e., $|$nrplml_{CpG}$ - median(rplm_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
+These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level more than $0.03$ from the median, i.e., $|nrplml_{CpG} - median(rplm_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
 3.
@@
 
-\begin{table}[h!]
+\begin{table}[!htb]
 \centering

### 2015-02-18 Rmarkdown/biology_background.Rmd
Figure fixing.
Added figure of dinucleotide frequencies in hg19

@@
 The GC-content of each the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is between $41\%$.
-If the nucleotides were uniformly distributed across the human genome we would expect about $4.1\%$ of dinucleotides to be Cpgs.
+If the nucleotides were uniformly distributed across the human genome we would expect about $4.1\%$ of dinucleotides to be CpGs.
 Instead, only $1\%$ of dinucleotides are CpGs.
-See the Appendix for details of these calculations and a figure comparing the observed:expected ratio of each dinucleotide in the human reference genome (hg19).
 
 One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}.
-Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinuclotide_distribution}).
+Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}).
 There are many other evolutionary pressures on the distribution of nucleotides in a genome.
@@
 In fact, CpGs tend to form clusters, which are termed CpG islands.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=0.3\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
+\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19).
+The expected frequency is computed under an 'independence' model, based on the observed frequencies of each nucleotide.}
+\label{fig:dinucleotide_frequency_hg19}
+\end{figure}
 
@@
 \caption{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19).
-The observed distances are contrasts with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
+The observed distances are contrasted with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
 \label{fig:IPD_ECDF_hg19}

### 2015-02-18 Rmarkdown/comethylation.Rmd
A few TODOs

--- 2014-12-23 Rmarkdown/comethylation.Rmd
@@
 \chapter{Co-methylation}\label{chap:co-methylation}
+
+# TODOs
+
+- Try out Matthew Stephen's adaptive hierarchical shrinkage method on the estimated log odds ratios (see [https://github.com/stephens999/ash](https://github.com/stephens999/ash)).
 
@@
 
+__TODO: I need to re-write this section (bold).
+I'm really not yet sure if this result is biology or an artefact.
+My instinct says artefact but I haven't yet been able to prove this either way to my satisfaction
 These plots appear to show that there is a dramatic increase in within-fragment co-methylation at precisely these $IPD$s for paired-end samples from the Lister dataset.
@@
 As discussed in Section \ref{sec:m-bias}, the incomplete removal of M-bias is a real problem for reads that have undergone pre-trimming because this destroys the relationship between read-position and sequencing cycle.
-__FIGURE__ shows that there is considerable variation in the read lengths of the _ADS_ sample, indicative of aggressive pre-trimming of these reads (__TODO: Histogram of read lengths__).
+FIGURE shows that while there is considerable variation in the read lengths of the _ADS_ sample, which is indicative of aggressive pre-trimming of these reads, most reads are full length or close to it (TODO: Histogram of read lengths).
 
@@
 `methtuple` is designed to filter out specific read-positions and cannot 'exclude the first/last $n$ bases'.
-For example, using this aggressive strategy will have a detrimental effect for a read that has been pre-trimmed to $30$ bp, of which the last $10$ bp are affected by M-bias, because this filtering strategy will __retain__ read-positions $21-30$, which are exactly those affected by M-bias.
+For example, using this aggressive strategy will have a detrimental effect for a read that has been pre-trimmed to $30$ bp, of which the last $10$ bp are affected by M-bias, because this filtering strategy will _retain_ read-positions $21-30$, which are exactly those affected by M-bias.
 Nevertheless, it is the best that can be done with the current version of the software[^methtuple_refactoring] and with such aggressively pre-trimmed data.
@@
 
-The results for the aggressively trimmed _ADS_ sample are shown in __FIGURE__.
+The results for the aggressively trimmed _ADS_ sample are shown in FIGURE.
 
-__TODO: Comment on these results and how we should interpret the Lister co-methylation results (particularly NIL >= 0)__
-
-__TODO: Summarise the M-bias issue__
+- TODO: Comment on these results and how we should interpret the Lister co-methylation results (particularly NIL >= 0)
+- TODO: Summarise the M-bias issue
+__
 

### 2015-02-18 Rmarkdown/comethylation.Rmd
A few minor typos

@@
 
-Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].
+Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].
 
@@
 
-Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' \citep{Kraemer:2004el}.
+Firstly, the ``possible range of values of the [log] OR is not influenced by the marginal distributions of the variables involved'' \citep{Kraemer:2004el}.
 Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints \citep{plackett1981analysis}.
@@
 \caption{Histograms of Pearson correlation coefficient for 2-tuples from the EPISCOPE data (binwidth $= 0.05$).}
-\label{fig:EPISCOPE_lor_histogram}
+\label{fig:EPISCOPE_pearson_histogram}
 \end{figure}
@@
 
-Figures \ref{fig:fig:EPISCOPE_lor_min_cov_10_strand_collapsed}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed} show the $LOR$ quantile plots for strand-collapsed $NIL = 0$ CpGs 2-tuples.
+Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed} show the $LOR$ quantile plots for strand-collapsed $NIL = 0$ CpGs 2-tuples.
 Like for the correlations of $\beta$-values, two things immediately stand out.
@@
 
-A second interpretation of this result is that it is due to 'reseting' of the methylome upon induction of pluripotency \citep{Lister:2011kg}.
+A second interpretation of this result is that it is due to 'reseting' of the methylome upon induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.
 As a population of cells develops from a single cell, there are errors in copying the DNA methylation patterns from mother to daughter cell.
@@
 
-Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac} and \ref{fig:Ziller_merged_lor_min_cov_10_strand_collapsed_ac} show the $LOR$ quantile plots for strand-collapsed $NIL \geq 0$ CpGs 2-tuples.
+Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed_ac} show the $LOR$ quantile plots for strand-collapsed $NIL \geq 0$ CpGs 2-tuples.
 By using $NIL \geq 0$ 2-tuples we can explore the effect of within-fragment co-methylation at greater $IPD$s, potentially beyond the crucial $IPD = \approx 150-170$ bp distance.

### 2015-02-18 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
A few minor typos

@@
 This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
+\label{fig:E18VA_M-bias}
 \end{figure}
@@
 This obviously affects the size of m-tuples that we can study.
-Figure \ref{fig:Lister_CpGs_per_read} shows the number of CpGs per read for the Lister dataset (see Chapter \ref{chap:dataset} for a description of the Lister dataset).
+Figure \ref{fig:Lister_CpGs_per_read} shows the number of CpGs per read for the Lister dataset (see Chapter \ref{chap:datasets} for a description of the Lister dataset).
 Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples.
@@
 `methtuple` also allows the user to filter out specific read positions rather than wholesale filtering of the ends of reads.
-This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA}.
+This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}.
 Such a 'spike' can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.
@@
 Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
-\label{fig:methtuple_runtime}
+\label{fig:methtuple_memory}
 \end{figure}

### 2015-02-18 Rmarkdown/wgbs_downstream_analyses.Rmd
A few minor typos

@@
 \citet{Benjamini:2001ho}                      & \citet{Sofer:2013bk, Hebestreit:2013ko}                                             \\
-"Bonferonni adjustment"                      & \citet{Feng:2014iq}                                                                 \\ \bottomrule
+``Bonferonni adjustment''                      & \citet{Feng:2014iq}                                                                 \\ \bottomrule
 \end{tabulary}
 \caption{Methods proposed for adjusting for the multiple hypothesis testing performed in an analysis of DMCs.
-Several papers use describe their analysing as performing a "false discovery rate adjustment" or "false discovery rate correction" without explicitly stating what they are doing or citing a reference.
+Several papers use describe their analysing as performing a ``false discovery rate adjustment'' or ``false discovery rate correction'' without explicitly stating what they are doing or citing a reference.
 One paper uses an ultra-conservative Bonferonni correction.}
@@
 However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}.
-Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg}.
+Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.
 

### 2015-02-18 Rmarkdown/wgbs_statistical_framework.Rmd
A few minor typos

@@
 Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
-Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_merged_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
+Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
 
@@
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
-\label{fig:Ziller_merged_plotBetaFreqPoly}
+\label{fig:Ziller_plotBetaFreqPoly}
 \end{figure}
@@
 In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
-However, as can be seen from Figures \ref{EPISCOPE_plotBetaFreqPoly}, \ref{Lister_plotBetaFreqPoly}, \ref{Seisenberger_plotBetaFreqPoly} and \ref{Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
+However, as can be seen from Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
 To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
@@
 Induced pluripotent stem cell lines also appear to have a quite strictly regulated methylome, with little intermediate methylation.
-This is likely a consequence of the fact that during the induction of pluripotency, the methylome of the sample is 'reset' \citep{Stricker:2013kl} thereby resulting in a homogeneous population of cells.
+This is likely a consequence of the fact that during the induction of pluripotency, the methylome of the sample is 'reset' \citep{Lister:2011kg, Stricker:2013kl} thereby resulting in a homogeneous population of cells.
 These samples also have highly correlated strand-specific $\beta$-values, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.

### 2015-02-18 Rmarkdown/80-word_summary.Rmd
Removed markdown files or converted them to Markdown

---
title: "80-word summary"
output:
  pdf_document:
    fig_caption:
    - yes
    includes:
      in_header:
      - ../latex/header.tex
  html_document:
    keep_md:
    - yes
bibliography: ../latex/phd_thesis.bib
---

# The statistical analysis of data from high-throughput assays for studying DNA methylation

I develop a statistical framework for modelling and analysing DNA methylation data. In particular, I focus on the complex dependence structure of these data. I describe how this can be incorporated into statistical analyses of bisulfite-sequencing (BS-seq) experiments and develop software to simulate realistic BS-seq data by incorporating this dependence.

I also report results from my analysis of more than fifty BS-seq samples, both publicly available and those collected by collaborators, which have informed the methods developed in my thesis.

### 2015-02-18 Rmarkdown/TODO.Rmd
Removed markdown files or converted them to Markdown

---
title: "TODOs"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

Those of highest priority are tagged, __URGENT__.

## PhD handbook

TODOs according to the [Unimelb PhD handbook](http://gradresearch.unimelb.edu.au/handbooks/phd/) (as of 07/10/2014).

- [ ] Read [Preparation of Graduate Research Theses Procedure (MPF1263)](https://policy.unimelb.edu.au/MPF1263).
- [ ] Read [Examination of Graduate Research Students Procedure (MPF1262)](https://policy.unimelb.edu.au/MPF1262).
- [ ] Read [Examination of Graduate Research Students Policy (MPF1207)](https://policy.unimelb.edu.au/MPF1207).

### [Copyright](http://gradresearch.unimelb.edu.au/handbooks/phd/research.html#copyright)

> Before you contemplate publication of your thesis either online or in print, it is important that you understand both your rights and obligations under copyright. Some publishers require that you sign an author or publishing agreement. Before signing any agreement, you should make sure that you fully understand the terms and your rights under the agreement. __If your thesis includes copyright material created by other people, you must ensure that you have the right to publish this material. While the fair dealing provisions of the Copyright Act allow you to reproduce a limited amount of third party copyright works in your thesis for the purpose of study and research, you cannot rely upon these provisions when you publish the work to the world. If you intend to publish your work, you will generally need to seek permission from the copyright owner as there are only limited provisions in the Copyright Act that allow material to be published without permission.__ Visit the University's Copyright website for more information.

- [ ] What does this mean for the use of figures published in articles, textbooks or on websites?
  - [ ] Further details on copyright [http://unimelb.edu.au/copyright/research.html](http://unimelb.edu.au/copyright/research.html)

### [Ethics approval](http://gradresearch.unimelb.edu.au/handbooks/phd/research.html#ethics)

> __All projects requiring the use of human participants, animals, genetically modified organisms or hazardous biological agents are subject to review and approval by the appropriate University committee prior to commencement of the work. Approval will not be granted retrospectively.__ Further information about research ethics and the requirements of the Human Research Ethics Committee, Animal Welfare Committee or Gene Technology and Biosafety Committee is available in departments or from the Office of Research Ethics and Integrity.

- [ ] Clarify what, if any, ethics statement I need in my thesis.
  - I am not directly performing any experiments as part of my PhD. However, I am using data from experiments that would require such ethics approval. In particular, the collaborations with Sue Clark (human samples) and Emma Whitelaw (mouse samples).

### [Editing of theses](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#editing)

> The use of third party editorial assistance, either paid or voluntary, must be acknowledged in the Preface and is limited to the guidelines adopted by the University.

- [ ] Acknowledge all proofreaders in Preface.

### [Completion seminar](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#completion)

> Completion of the seminar requirement will be signed off by the supervisors and head of department on the Completion Seminar section of the [Submission of Thesis: Statement by Candidate, Supervisor and Chairperson of Examiners](http://gradresearch.unimelb.edu.au/docs/exams/GS6B_Submission_Form.pdf) form and will include the date and location of the completion seminar.

- [ ] Complete [Submission of Thesis: Statement by Candidate, Supervisor and Chairperson of Examiners](http://gradresearch.unimelb.edu.au/docs/exams/GS6B_Submission_Form.pdf)

### [Word limit](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#wordlimit)

> Candidates should aim to write a thesis of 80,000 words.

> The word limit is exclusive of words in tables, maps, bibliographies and appendices. Footnotes are included as part of the word limit. Appendices must be limited to supporting material genuinely subsidiary to the main argument of the thesis.

> Candidates may write up to 100,000 words without seeking special permission from the RHD Committee.

- [ ] Find a `wordcount` function for LaTeX or markdown files.

### [Preparation of a thesis that contains published work](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#published)

> Thesis chapters may be comprised of work that is written specifically for the thesis, reprints or other reproductions of published works or an adaptation of published work..

- I don't expect this be an issue for me as I am not re-using any published work.

> Publications, such as reprints of journal articles, published creative writing and catalogues and documentation of public performances or exhibited work, may also be included in an appropriate form in the appendix, __including for example as DVD, CD or URLs, if not included as a chapter__.
- [ ] Include all URLs of software in appendix.

### [Format of the thesis](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#format)

> A thesis should include the following formatting:
* International Standard Paper Size A4 (297 x 210mm)
* 1.5 spacing and presented in a clear and legible font and would normally be expected to be double-sided
* Left and right margins of no less than 30mm and page numbers that appear inside the margins
* Pages that are numbered consecutively and clearly
* Folding diagrams or charts arranged so as to open to the top and right.

- [ ] Check the spacing and margins of my LaTeX pdf.

#### Order of contents

> A thesis follows the following order:
* Title page
* Abstract
* Declaration
* Preface (if applicable)
* Acknowledgments
* Table of contents
* List of tables, figures and illustrations (if list items are fewer than 10 in number, this is not necessary)
* Main text
* Bibliography or List of References
* Appendices.

- [ ] Check order of my thesis.

#### Title page

> A thesis must be preceded by a title page. The title page of the thesis should show the:
* Title of the thesis
* Full name of the author (as it appears in the Student Portal)
* Degree for which submitted (see below)
* Month and year
* Name of the department or faculty in which the research was carried out.

> Candidates who have pursued a course of study by research alone, including combined Masters/PhD degrees, shall state on the title page: "Submitted in total fulfilment of the requirements of the degree of Doctor of Philosophy".

- [ ] Check title page fits criteria.

#### Abstract

> The title page must be followed by an abstract of 300–500 words in English (in the case of creative arts the Abstract must include a description of the form and presentation of the creative work).

#### Declaration

> The following declaration page, signed by the candidate:

> This is to certify that:
* the thesis comprises only my original work towards the PhD except where indicated in the Preface,
* due acknowledgement has been made in the text to all other material used,
* the thesis is fewer than 100 000 words in length, exclusive of tables, maps, bibliographies and appendices OR the thesis is [number of words] as approved by the Research Higher Degrees Committee.

#### Preface

> If applicable, a Preface page includes a statement of:
* Work carried out in collaboration indicating the nature and proportion of the contribution of others and in general terms the portions of the work which the candidate claims as original
* Work submitted for other qualifications
* Work carried out prior to PhD candidature enrolment
* any third party editorial assistance, either paid or voluntary (as limited to the Editing of Research Theses by Professional Editors guidelines) and/or
* Where a substantially unchanged multi-author paper is included in the thesis a statement prepared by the candidate explaining the contributions of all involved. A signed copy by all authors must be included with the submission form.

#### Thesis preparation and binding

> Theses should normally be bound using thermal binding.

> Thermally bound theses must be used with covers strong enough to resist damage by bending or knocking. Twin-ring, spring-back and spiral binders are not acceptable, as theses bound in these ways frequently do not survive travel through the post. Thermal binding should not exceed 300 pages (absolute maximum thickness of 35mm per copy).

> If there are photographs or charts which need to be included in the thesis, facilities are available for colour laser printing and photocopying in the Graduate Student Association (GSA)and a scanner is available in the Graduate Presentation and Publishing Centre. The GSA also provides a thermal binding and photocopying service (refer to [The Graduate Centre](http://gradresearch.unimelb.edu.au/handbooks/phd/intro.html#centre)).

### [Notice of intention to submit thesis](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#intention)

> Three months prior to the submission date of the thesis, the candidate is required to submit an 80-word summary of the thesis argument.

> The 80-word summary should be submitted via the on-line [form](http://gradresearch.unimelb.edu.au/exams/candidates.html#summary).

> In addition to the 80-word summary the following information is required:
* Candidate's name
* Student number
* Department
* Supervisors' names (if required an Application for Change of Supervisor(s) form is available from my.unimelb)
* Thesis title
* Current mailing address
* Estimated thesis submission date (this is the date that you intend to submit your thesis for examination and it may not necessarily be the Expected Thesis Submission Date that appears in my.unimelb).

> Note: If more than three months lapses between submission of the 80-word summary and the thesis for examination, it is recommended that the candidate notifies the Examinations Office two weeks before the thesis is to be submitted to avoid delays with the examination. The Examinations Office can be notified by email (msgr-exams@unimelb.edu.au).

- [x] Submit intention to submit form.
- [x] Find out the latest date a thesis can be submitted in 2014.

MGSR replied:

> We will be accepting submissions up until around 11am on 24 December 2014, however, we strongly suggest submitting at least a week prior so that your thesis can be sent to examiners, if they have been nominated, prior to the Christmas break.

So, I have nominated 17/12/2014 as the date that I will submit my thesis.

### [Thesis submission](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#submission)

> Three copies of the thesis should be submitted for examination with the candidate's name clearly marked on the spine, with the family name in capitals.

> The form also includes a section where the candidate (in consultation with the supervisor) may request that the examiners maintain the confidentiality of the thesis contents. Such a request may be required where there are research contracts with outside organisations or other issues related to intellectual property.

- [ ] Check with Sue Clark and Emma Whitelaw whether sections of my thesis related to their work need to remain confidential.

### [Final form of the thesis](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#finalform)

> On completion of the examination two copies of the thesis in permanent hard cover binding, incorporating any necessary amendments or revisions, must be submitted to the Chair of Examiners to be approved. Candidates who commenced from January 2007 are also required to submit an electronic copy of the thesis to the University of Melbourne instutional repository (refer to Making your thesis available on the University of Melbourne institutional repository p. 34). Addenda of any length are not accepted. One copy must be printed on archival quality paper (paper should be approximately 80–100gsm) available from the Co-op Bookshop (in the Baillieu Library Building) or stationery suppliers. The words "Produced on archival quality paper" should be printed on the title page of this copy.

> The University logo is not permitted to be used in the thesis.

> The following must be printed on the spine of the thesis:
* Full name of the author
* Title of the thesis (abbreviated if necessary)
* Degree (ie PhD)
* Year of initial submission, or in the case of rewrites the year of resubmission.

> There are no colour restrictions for the thesis cover. The archival quality copy is for the Baillieu Library and the second copy for the Department Library.

### [Making your thesis available on the University of Melbourne institutional repository](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#eprints)

> A digital copy of a completed higher degree thesis must be submitted to the repository (via the Library).

> Candidates can elect to have the entirety of their thesis on open access, or only the citation and metadata. Candidates are required to discuss with their supervisor any Intellectual Property implications prior to lodging their thesis online.

> If the candidate agrees to allow the full text of their thesis to be made available, the necessary permissions for third party copyright material must be dealt with (refer to Copyright section of this handbook and the Copyright Office webpage).

- [ ] Select open access option (provided Sue Clark and Emma Whitelaw don't require confidentiality).
- [ ] Check rules on self-publishing.

### [Bibliography format](http://gradresearch.unimelb.edu.au/handbooks/phd/thesis.html#resources)

> [See the Citing Styles tab](http://www.unimelb.libguides.com/content.php?pid=87165&sid=3005272) for style guides and manuals, and for general referencing information.

- [ ] What style should I use for my bibliography?

### [Amendment to thesis](http://gradresearch.unimelb.edu.au/handbooks/phd/examination.html#amendment)

> Candidates are not permitted to make any changes to their thesis not specified by the examiners (or implied in their reports). Candidates are required to consult with their supervisor and the Chair of Examiners in this matter.

## 3.5 year candidature review

TODOs for 3.5 year candidature review, which is due by November 15.

- [x] Complete progress review
- [x] Organise progress review meeting __URGENT__
  - [x] Emailed Terry
  - [x] Emailed Peter
  - [x] Emailed Gordon

## Fellowship applications

TODOs for fellowship applications for 2015.

- [ ] Check out the [Joint Academic Scholarships Online Network](http://jason.edu.au/)
- [x] __URGENT:__ Consult notes from meeting with Lynne Hartley

### 2015-02-18 Rmarkdown/bootcamp_plan.Rmd
Removed markdown files or converted them to Markdown

---
title: "Bootcamp plan"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

# Thesis outline

## Thesis title

Statistical analysis of high-throughput assays for studying DNA methylation

## Research questions

* What DNA methylation is
* What bisulfite-sequencing is and how people use it
* Why statistics is needed
* What questions I have addressed
* How my work is useful

# Chapter outlines

For each chapter I outline my plans and add checkboxes for what is and is not drafted.

## Preamble

Do last.

## Introduction

- [x] Chapter overview
- [ ] DNA
  - [ ] basic structure, mitotic replication, DNA -> nucleosomes -> chromatin -> chromosomes -> genome, DNA -> RNA -> protein.
- [ ] Epigenetics
- [x] DNA methylation
  - Mostly done.
- [x] Assays for studying DNA methylation
  - Mostly done.
- [ ] WGBS
- [ ] Outline of thesis

## Bioinformatics analysis of bisulfite-sequencing data

This chapter is mostly complete. Need to re-work the intro a little to be more general than differential methylation experiments.

Put this in context with Krueger et al.'s Nature Methods paper and any other papers that explain the "pipeline" for a bioinformatics analysis of BS-seq data.

- [x] Chapter overview
- [x] Data quality control checks
- [x] Read mapping and post-processing of mapped reads
- [x] Methylation calling
- [ ] Downstream analyses

## A statistical framework for analysing bisulfite-sequencing data

I should emphasise that this chapter is also useful because this framework is lacking, which makes it difficult to compare methods.

This chapter should focus on the statistical framework and not the bioinformatics problems (these should go in the previous chapter) and so this chapter requires some re-working.

- [x] Chapter overview
- [ ] One sample
- [ ] _n_ samples
- [ ] Parameter estimation
- [ ] Statistical properties of $\beta$
- [ ] Downstream analyses

## Datasets used in thesis

This chapter is a good candidate to write during the bootcamp.

## Differential methylation in Agouti viable yellow mice

This chapter is a good candidate to write during the bootcamp.

- [ ] Chapter overview
- [ ] Introduction
- [ ] Background + literature review
- [ ] Methodology
- [ ] Discussion of findings
- [ ] Conclusion

## Co-methylation

I want to combine the "Co-methylation review" and current "Co-methylation" chapter into a single chapter.

- [ ] Chapter overview
- [ ] Introduction
- [ ] Background
- [ ] Literature review
  - [ ] Correlations of aggregate methylation
  - [ ] Within-fragment co-methylation
  - [ ] Between sample-comethylation
- [ ] Methodology
  - [ ] Correlations of $\beta$-values
  - [ ] Within-fragme co-methylation
- [ ] Implementation
- [ ] Case studies
  - [ ] Lister data
  - [ ] EPISCOPE
  - [ ] Ziller
- [ ] Discussion of findings
  - [ ] Biology
  - [ ] Limitations of using 2-tuples
  - [ ] Accounting for and exploiting co-methylation in analyses of differential methylation
- [ ] Conclusions

## Co-methylation review

This chapter is mostly complete. Just need to review some old papers and decide whether this should really be a stand-alone chapter or a part of the next chapter.

- [x] Chapter overview
- [x] What is co-methylation and why study it
- [ ] Correlations of aggregate methylation
    - Mostly done, still need to review Landan et al.
- [x] Within-fragment co-methylation
- [ ] Between sample co-methylation
  - Still to review Capra and Kotska (2014) and Liu et al. (2014)
- [ ] Exploiting co-methylation in analyses of differential methylation

## Co-methylation

- [x] Chapter overview
- [ ] Correlations of beta-values
- [ ] Within-fragment co-methylation
  - Partially written
- [ ] Using higher-order m-tuples
- [ ] Results of co-methylation analyses

## A simulation model of DNA methylation data

This chapter is a good candidate to write during the bootcamp.

- [ ] Chapter overview
- [ ] Introduction
- [ ] Background + literature review
- [ ] Methodology
- [ ] Implementation
- [ ] Case study
- [ ] Discussion of findings
- [ ] Conclusion

## Concluding remarks

## Bibliography

## Appendices

### 2015-02-18 Rmarkdown/outline.Rmd
Removed markdown files or converted them to Markdown

---
title: "Thesis outline"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

# Thesis outline
1. Preamble
2. Introduction
3. Statistical model of methylC-seq data
4. Standard analysis of methylC-seq data
5. m-tuples
6. Correlation of $\beta$-values
7. Simulation model: Part I
8. Simulation model: Part II
9. Data analysis - EPISCOPE
10. Concluding Remarks
11. Bibliography
12. Appendices

## Chapter outlines
### Preamble
* Abstract
* Preface
	* Data sets: Who made them and which chapters are they used in
	* Analyses: Who performed them
	* Code: Who wrote it
	* Publications: Any arising from chapters of the thesis
* Acknowledgements
	* Supervisors
	* Specific
		* Collaborators
		* WEHI Bioinformatics, especially Melanie Bahlo
		* Felix Krueger, Ryan Lister, Kasper Hansen, Sue Clarke, Rafael Irizarry, Mark Robinson
	* IT, especially Keith and Jakub
	* Other WEHI: Communications, etc.
	* General
		* Family
		* Friends
	* Financial
		* APA, VLSCI, EMBL-PhD, Melanie Bahlo, conference subsidies
* Table of Contents
* List of Figures
* List of Tables
* List of Abbreviations

### Introduction
* Chapter Overview
* DNA and DNA methylation
	* DNA
	* DNA methlyation and its function
	* Types of DNA methylation
* Assays for studying DNA methylation
	* Bisulfite-sequencing
	* Overview of assays (chronological?)
		* Sanger bisulfite-sequencing
			* Can't distinguish mC and hmC
		* Microarrays
		* Enrichment assays
			* methylC-seq, BS-seq, RRBS, PBAT
	* Why use methylC-seq
* Description of methylC-seq
	* History
	* Protocol
	* Data
	* Design of methylC-seq experiments
* Standard analyses
	* Brief, non-mathematical (see also Chapter 4)
* Other biology
	* Primary tisue
	* Cell lines
		* IPS cells
		* ES cells  
* Thesis outline

### Statistical model of methylC-seq data
* Chapter Overview
* Notation
	* See `notation.Rmd`
* Estimating $Z$, $M$, $U$ and $\beta$
	* See `notation.Rmd`
	* Current approaches
* Statistical properties of $z$, $m$, $u$ and $\beta$
	* See `notation.Rmd`
	* Marginal distributions
	* Dependence/correlation over $i$ (see also __Chapter X__)
		* Li _et al._ compute $\beta$ values and compute correlations within a sample as a funciton of genomic distance.
		* Arabidopsis compute $\beta$ values and compute correlations within a sample as a funciton of genomic distance.
		* Lister _et al._ compute both within-read and across-read correlations or dependencies of DNA methylation.
		* Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.
	* Dependence/correlation over $j$ (?)
		* Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.

### Standard analysis of methylC-seq data
* Chapter Overview
* QC, mapping, post-processing
* Methylation calling
	* Reference-based calling
* Biases
	* M-bias
	* Cell-composition, age, global hypo- or hyper-methylation, SNPs, alignment artefacts
* Univariate tests of $\beta_i$

### m-tuples
* Chapter Overview
* Previous work
	* Amos Tanay
	* Lister _et al._
	* _Arabidopsis_
	* Others
* Within-fragment co-methylation
	* Concept
	* Pros
	* Cons
	* An early reference to comethylation (http://nar.oxfordjournals.org/content/32/21/e167.full)
* Methods
	* `comethylation.py`
	* `cometh`
* Results
	* Lister data
	* Seisenberg data
	* Other?
* Discussion

### Correlation of $\beta$-values
* Chapter Overview
* Previous work
	* _Arabidopsis_
	* _PMBC_
	* Other
	* What are they actually computing?
* Methods
	* `comethylation.py`
	* `cometh`
	* $NIC = 0$ vs. $NIC >= 0$
* Results
	* Lister data
	* Seisenberg data
	* Other
* Discussion

### Simulation model: Part I
* Chapter Overview
* Previous work
	* Simulators for comparing aligners
		* SHERMAN, LAST, etc.
	* Simulators for comparing DMR finders
		* Michelle Lacey, `bsseq`, `BiSeq`, etc.
* Limitations of previous simulation models
* First simple model

### Simulation model: Part II
* Chapter Overview
* A more complex model
* Methods
	* `methsim`
* Results
	* Comparison to real data
	* Limitations
* Discussion

### Data analysis - EPISCOPE
* Chapter overview
* Introduction
* Methods
	* Pre-processing
		* `Bismark`
		* `MarkDuplicates`
	* `comethylation.py`
	* `cometh`
* Results
* Discussion

### Concluding Remarks
### Bibliography
### Appendices

### Others(?)
#### Differential methylation calling
#### Simulation-based inference
* See Brian Ripley's lecture notes (email from Terry)
	* "_The basic idea is quite simple – simulate data from one or more plausible models (or for a parametric model, at a range of plausible parameter values), apply the same (or similar) procedure to the simulated datasets as was applied to the original data, and then analyse the results._"
#### Nick Wong's data

## Proposed chapters in 2-year PhD review
1. Introduction
2. Public datasets used in thesis
3. Pre-processing methylC-seq data and technical biases
4. Statistical model of methylC-seq data
5. Within haplotype dependence structure
6. Across haplotype dependence structure
7. Simulation of methlC-seq data
8. Data analysis
9. Conclusions

## Questions
* Formal literature review chapter or just intersperse lit review at beginning of relevant chapter(s)?
* Terminology (see Comethylation GitHub issue #48)

### 2015-02-18 Rmarkdown/paper_review.Rmd
Removed markdown files or converted them to Markdown

---
title: "Paper reviews"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

These are my reviews of important papers in the field. I do not want to reproduce these verbatim in my thesis but rather summarise these down to key points of similarity and difference between the methods.

## \cite{Eckhardt:2006gh}

\cite{Eckhardt:2006gh} studied DNA methylation patterns on chromosome 6, 20 and 22 in 43 human samples from 12 different tissues. Each of the 43 samples was made up from a pool of up to 3 sex- and age-matched samples.

This study was very labour-intensive work and low-throughput by contemporary standards because it predated the era of cheap, "high-throughput" sequencing (e.g. Illumina sequencing). Instead, the authors used Sanger sequencing of bisulfite-converted PCR amplicons. Briefly, they designed primers for bisulfite-treated DNA from chromosomes 6, 20 and 22. Some of these PCR amplicons were then subcloned into a vector. The non-cloned PCR amplicons and the cloned PCR-amplicons (up to 20 clones per amplicon) were Sanger sequenced using ABI 3730 capillary sequencers. All this results in a fairly sparse sample by modern standards from each chromosome and sample.

Nonetheless, they were perhaps the first to study co-methylation, which they defined as "the relationship between the degree of methylation over distance". The method by which they did this isn't clearly stated in the paper and what follows is my interpretation of what is reported. For a variety of distances between 0 and 20 kb they sampled 25,000 pairs of sequenced fragments (e.g. $z_{h, i}$ and $z_{h', i'}$) where two CpGs were separated by the given IPD (i.e. $pos_{i'} - pos_{i}$ = IPD). For each pair at that distance they recorded whether the methylation measurements were identical, that is, both methylated or both unmethylated (i.e. $\mathbf{1}_{z_{i, h} = z_{i', h'}}$). Then, for each distance, they computed the percentage of pairs that were in agreement, i.e. $\frac{1}{25000}\sum_{25,000 pairs}\mathbf{1}_{z_{i, h} = z_{i', h'}}$.

__TODO: Discuss interpetation with Terry__
__TODO: What are the grey and blue dots in Fig. 3c?__

Based on this analysis they concluded that there was "a significant correlation for co-methylation over short distances ($\leq$ 1,000 bp), [but] it deteriorated rapidly for distances $>$ 2,000 bp".

__TODO: Fix indicator function__

A useful quote regarding mosaicism and epipolymorphism:

> we conclude that the majority (490%) of the observed heterogeneous methylation is caused by mosaicism, although we cannot exclude the additional possibility of heteroge- neous tissue sampling.

## \cite{Cokus:2008fc}
\cite{Cokus:2008fc} published the BS-seq protocol for performing WGBS. Most of the results in this paper used BS-seq of wild-type _Arabidopsis Thaliana_ and DNMT mutants. They also report limited results from low-coverage BS-seq of mouse germ cells but these are not relevant to my discussion.

The authors used the simple $m$ and $u$ read-counting estimators of $M$ and $U$, subject to some filtering and other post-processing of the reads. Most analyses were restricted to loci with at least 5x sequencing coverage.

They visualised the levels of DNA methylation across a variety of genomic elements such as protein coding genes, pseudogenes and repeats. They did not perform an analysis of differential methylation. The authors were also interested in the dependence structure of DNA methylation. For the most part they did this by studying the $\beta$-values but they also performed a "within-read" analysis.

To look at the correlation of "average" methylation, they performed an autocorrelation analysis of the $\beta$-values. It is not clear from the paper whether they restricted these analyses to pairs of methylation loci with $NIL = 0$ or whether they allowed $NIL \geq 0$. They did separate autocorrelation analyses for CpG, CHG and CHH methylation.

To look at the dependence of methylation events from the same DNA fragment, they estimated the "within-read probability of additional methylation of CHH sites within a given distance from a methylated CHH site". They only report the "within-read" results for CHH methylation. This amounts to estimating the conditional probability $Pr(Z_{h, i + s} | Z_{h, i})$ and plotting it as a function of genomic distance between the two loci, $pos_{i + 1} - pos{i}$. It is not clear from the methods whether $s = 1$, i.e. the conditional probability that the __next__ CHH is methylated (which amounts to using pairs of CHH loci with $NIL = 0$), or whether $s \geq 1$, i.e. the conditional probability that __any__ upstream CHH is methylated (which amounts to using pairs of CHH loci with $NIL \geq 0$). Furthermore, these results are limited by the short read lengths (31nt) available at the time of the study.

The authors reported a 10 bp periodicity of CHH methylation, which could be detected in both the $\beta$-value and "within-read" analyses. They noted that this periodicity is consistent with one helical DNA turn and that the mammalian Dnmt3a, which is the homologue to the main enzyme controlling asymmetric (e.g. CHH) methylation in _Arabidopsis_, _DRM2_, had recently been shown to methylate CpG cites $~8-10$ nucleotides apart.

Autocorrelation analyses of the $\beta$-values also identified a periodicity of $~167$nt for CpG, CHG and CHH methylation, although it was strongest for CHG methylation. The authors noted that this "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin". From this they postulated that nucleosomes or histone modifications might dictate access to the DNA by DNMTs and that methylated DNA might be more "compact" than unmethylated DNA.

Both the $~10$nt and $~167$nt periodicities can be seen by eye and were confirmed by Fourier analysis of the $\beta$-autocorrelation function.

## \cite{Lister:2008bh}

\cite{Lister:2008bh} published the methylC-seq protocol for performing WGBS. Like \cite{Cokus:2008fc}, \cite{Lister:2008bh} studied wild-type and DNMT-defective mutant _Arabidopsis Thaliana_ strains. Also like \cite{Cokus:2008fc}, much of the analysis is by visual comparison of summarised data.

The authors used the simple $m$ and $u$ read-counting estimators of $M$ and $U$, subject to some filtering and other post-processing of the reads. To identify methylcytosines in each sample, the authors used a simple Binomial model, $Bin(m_{i, j} + u_{i, j}, B_{i, j})$ . They defined $\epsilon_{j}$ as the "error rate" for the $j^{th}$ sample, which is the sequencing error rate + the bisulfite non-conversion rate and is assumed constant across all loci. They estimated $\epsilon_{j}$ from reads aligned to the unmethylated chloroplast genome of each sample and obtained values between $1.3-3.2\%$ per sample.

For each locus and sample the authors tested the hypothesis $H_{0}: B_{i, j} = \epsilon{j}$ vs. $H_{1}: B_{i, j} > \epsilon{j}$. This can be thought of as testing the hypothesis "what is the minimum $m_{i, j}$ I would have to observe to conclude that these weren't all due to "errors"" (__CHECK WITH TERRY, in particular is the alternative one- or two-sided?__). For each sample, all loci were ranked by the false discovery rate-adjusted (FDR-adjusted) P-value from this test (__Not clear from text which FDR procedure was used__). All loci with an FDR < 0.05 were called as methylcytosines.

The authors did not perform any statistical tests of differential methylation nor did they report any results on the dependence structure of DNA methylation.

## \citet{Lister:2009hy}

\cite{Lister:2009hy} was a landmark paper in the study of DNA methylation in humans from WGBS data. The authors studied 4 samples: 2 cell types (IMR90 and H1) and 2 "biological" replicates per cell type. However, most of the results reported in \cite{Lister:2009hy} were from analyses of data pooled across "biological" replicates. This completely ignores all biological variability and in general isn't a good idea. In the following description of their statistical analyses, all references to "samples" means "pooled samples".

The authors used the simple $m$ and $u$ read-counting estimators of $M$ and $U$, subject to some filtering of the reads. The authors used the Binomial model from \cite{Lister:2008bh} to identify methylcytosines.  The bisulfite non-conversion rate was estimated from a spike-in control that was included with each sample for sequencing[^spike_in]. It is not clear how they estimated the sequencing error rate. They reported that $\epsilon_{\text{IMR90}} = 0.005$ and $\epsilon_{\text{IMR90}} = 0.0024$. All loci with an FDR < 0.01 were called as methylcytosines.

[^spike_in]: The spike-in control was unmethylated cl857 Sam7 Lambda DNA. This "lambda phage" is a common choice of spike-in control in bisulfite-conversion assays.

With this set of methylcytosines they sought to identify differential methylation and partially methylated domains (PMDs). They used a two-stage analysis to identify differential methylation:

1. Identify differentially methylated cytosines (DMCs)
2. Identify differentially methylated regions (DMRs)

The first step used a Binomial model for each sample, namely $Bin(m_{i, \text{IMR90}} + u_{i, \text{IMR90}}, B_{i, \text{IMR90}})$ and $Bin(m_{i, \text{H1}} + u_{i, \text{H1}}, B_{i, \text{H1}})$. The authors then used a two-tailed Fisher's exact test of the hypothesis $B_{i, \text{IMR90}} = B_{i, \text{H1}}$ vs. $B_{i, \text{IMR90}} \neq B_{i, \text{H1}}$. They performed this test at all loci that were called as methylcytosines in at least one of the cell types and that had at least 3 reads in at least one of the samples. Cytosines were called as differentially methylated if the FDR-adjusted P-value was less than 0.05.

The second step was to identify DMRs, that is, regions containing multiple cytosines that display differential methylation between the IMR90 and H1 samples. In fact, \cite{Lister:2009hy} only sought to "find regions of the genome enriched for sites of higher levels of DNA methylation in IMR90 relative to H1, as identified by Fisher's Exact Test" \cite[Supplementary Material, pp. 26]{Lister:2009hy}. This was only performed for CpG methlylation loci.

The authors used a heuristic approach based on a 1kb sliding window approach and 100bp step size. If the window contained at least 4 differentially methylated cytosines then it was extended in 1kb increments until a 1kb increment was reached that contained less than 4 differentially methylated cytosines. Once the extension had terminated, a region was declared to be a DMR if it contained at least 10 differentially methylated cytosines and was at least 2kb in length. The authors did not report a sensitivity analysis of any of these parameters.

A sliding window approach was also used to identify partially methylated domains (PMDs). This was only performed for CpG methlylation loci. A larger window size of 10kb and step size of 10kb were used. If the window contained 10 mCpGs, each covered by at least 5 reads, and the __average__ $\beta$-value in the region was less than 0.7 then the region was incremented by 10kb. The extension was terminated once the next increment had an average $\beta$-value greater than 0.7 or less than 10 mCpGs and the region was called a PMD.

\cite{Lister:2009hy} also investigated what at first appears to be a "within-read" measure of co-methylation (Fig. 3g, 3h; Supp. Fig. 9). However, on a closer reading I do not think it is a truly "within-read" measure. Specifically, they tabulated the number of mCG (resp. mCHG or mCHH) sites 1-50n bp downstream of a mCG (resp. mCHG or mCHH). I believe that they define methylcytosines based on their binomial test, which means these tabulations are "across-reads" rather than "within-reads". To truly do this as a "within-read" analysis you would tabulate $z_{i, j} = (1, 1)$ for a variety of IPD. The fact that they used 50 bp reads and measured over a 1-50 bp window makes it easy to misinterpret their results as being "within-read".

With that in mind, these results suggest a 8-10 bp periodicity in the co-occurence of methylcytosines. This co-occurence is clearest for CHG and CHH loci in intronic sequences. Other loci and contexts do not show this behaviour or not to the same extend. For example, the graph of mCG co-occurence in exonic sequences is dominated by a 3nt cycle, presumably due to codon structure and selective pressures on coding sequences (Sup. Fig. 9).

To summarise, I take the "co-methylation" results of \cite{Lister:2009hy} with a grain of salt for several reasons:

1. The details of the method are not very clear from the paper. Aside from the "within-read" vs. "across-read" issue, it is not clear whether they use pairs of cytosines with $NIL = 0$ or with $NIL \geq 0$ in their analysis.
2. Estimates of periodicity are based on visual inference from a cubic spline smoothing of the co-occurence patterns. They did not perform the more meaningful Fourier analysis.
3. They only look at the co-occurence of methylcytosines and not the co-occurence of unmethylated cytosines.
4. The number of observations per distance is very small in some contexts. For example, less than 100 observations per IPD are used in the graph of mCHH co-occurence in the "random" context (Sup. Fig. 9).

## \citet{Lister:2011kg}

\cite{Lister:2011kg} is an extension of \cite{Lister:2009hy}. Here the authors studied 15 methylC-seq datas, 4 of which were from previous publications. These samples came a variety of tissues but can be classified as being either from a cell line that is embryonic stem cell (ESC), induced pluripotent stem cell (iPSC), differentiated cell or _in vitro_ differentiated from pluripotent cell (IVD).

\cite{Lister:2011kg} used similar analysis methods as they did in \cite{Lister:2009hy}. Namely, methylcytosines were identified using the Binomial test and DMRs were identified using a sliding window approach. The details of the DMR finder are more complicated due to the larger sample size and multiple comparisons made between the 4 classes of cell type. For two-group comparisons, the average (smoothed) $\beta$-values in each window were tested for a mean difference using a Wilcoxon test. For multi-group comparisons, the Wilcoxon test was replaced by a Kruskall-Wallis one-way analysis of variance. The authors corrected the resulting P-values using the Benjamini-Hochberg method (__CITE__). Putative DMRs were those with an adjusted P-value < 0.01 and were also required to have a mean-difference greater than some threshold.

A sliding window approach was also used to identify partially methylated domains. The authors did not investigate the dependence structure of DNA methylation.

[Mark van de Wiel](http://www.few.vu.nl/~MA.van.de.Wiel/) raised some concerns about the statistical analysis of these experiments [on Nature's website](http://www.nature.com/nature/journal/v471/n7336/full/nature09798.html#comments).

## \citet{Lister:2013et}
\cite{Lister:2013et} used methylC-seq to study 5mc and TAB-seq to study 5hmC in neurons and glial cells from the frontal cortex of human and mouse samples. The authors reported that non-CpG methylation was the dominant form of 5mC in neurons but not in glial cells.

The analysis of 5mC used a different strategy to that of \cite{Lister:2009hy, Lister:2011kg}. Firstly, they did not perform an initial screen for "methylcytosines" using the Binomial test. Secondly, differentially methylated cytosines were identified using a test of the $J \times 2$ contingency table.  Specifically, for the $i^{th}$ CpG a $J \times 2$ table was constructed where row $j$ was the counts of methylated and unmethylated reads for the $j^{th}$ sample, $(m_{i, j}, u_{i, j})$. This table was tested for goodness-of-fit using a root-mean test (http://tygert.com/chi.pdf). DMRs were again constructed using a sliding window approach and were subjected to _post-hoc_ filters.

## \cite{Li:2010fb}
\cite{Li:2010fb} report the methylome of a single sample. They used WGBS to study 5mC from peripheral blood mononuclear cells (PBMCs) from an Asian man whose genome had also been used to create the Han Chinese reference genome.

The authors used the simple $m$ and $u$ read-counting estimators of $M$ and $U$, subject to some filtering of the reads. The authors used the Binomial model from \cite{Lister:2008bh} to identify methylcytosines. As they only had the one sample, many of the analyses were descriptive. For example, they looked at the distribution of $\beta$-values 20 different genomic elements such as CGIs, UTRs and repetitive sequences.

The authors looked at the autocorrelation of $\beta$-values across a range of genomic elements. They reported a $~170$nt periodicity in the autocorrelation plot of CpG methylation, similar to that found by \cite{Cokus:2008fc} in _Arabidopsis_. They did not find evidence of any smaller periodicities such as the 8-10 bp periodicity reported by \cite{Cokus:2008fc} and \cite{Lister:2009hy}.

The $~170$nt periodicity can be seen in the autocorrelation plots and was confirmed by Fourier analysis. They found that the autocorrelation of $\beta$-values is stronger when considering pairs of $\beta$-values from the same DNA strand than those from opposite strands. Furthermore, they found this autocorrelation was stronger in some genomic elements than others.

The authors compared the PBMC methylome to the IMR90 methylome from \cite{Lister:2009hy} to identify tissue-specific DMRs (tDMRs). DMR testing was done by forming regions containing 5 CpGs and comparing methylation levels between PBMC and IMR90 using Fisher's exact test, presumably by aggregating read counts across all CpGs in the window. Regions with a P-value $< 10^{-20}$, at least a two-fold difference in methylation between PBMC and IMR90 were declared  tDMRs. Neighbouring tDMRs were joined if they were 'consistent'.

Because the genome of the individual had already been sequenced, this allowed them to study allele-specific methylation (ASM). They extracted all reads overlapping a heterozygous SNP and then looked at methylation levels from reads containing each allele.

## \cite{Hansen:2011gu} and \cite{Hansen:2012gr}

 \cite{Hansen:2012gr} describes a method for identifying DMRs between two groups of samples. This method was first used in \cite{Hansen:2011gu} to identify DMRs in colon cancer tumours. As part of their study, \cite{Hansen:2011gu} performed WGBS on 3 colon cancer samples and their matched normal mucosa, as well as of 2 colon adenomas[^adenoma]. The statistical method they developed, which they call `BSmooth`, are implemented in the R/Bioconductor package `bsseq`. I will describe `BSmooth` and then explain how it was used, in a modified form, in \cite{Hansen:2011gu}.

[^adenoma]: An adenoma is a benign tumour formed from glandular structures in epithelial tissue (__SOURCE: google.com__)

`BSmooth` proposes the following general framework for performing a two-group differential methylation analysis:

1. Smooth raw $\beta$-values.
2. Compute at each methylation locus a test statistic that quantifies the difference between the two groups (in the case of `Bsmooth`, the mean difference in methylation level between the two groups).
3. Find contiguous runs of extreme test statistics; call these putative DMRs.
4. Filter the list of putative DMRs to produce the final list of candidate DMRs. These filters might be biologically-motivated or be designed to remove known false-positive calls from the DMR-calling algorithm.

 \cite{Hansen:2012gr} show that because `BSmooth` explicitly accounts for biological variability it outperforms an analysis using Fisher's exact test of data that has been pooled by group. `Bsmooth` can also make use of the paired design of the experiment described in \cite{Hansen:2011gu}.

### Step 1 {-}

The analysis begins with a table for each sample of the simple $m$ and $u$ read-counting estimators of $M$ and $U$, subject to some filtering of the reads, for each CpG. For each sample, the resulting "raw" $\beta$-values are smoothed in order to remove noise due to low sequencing coverage.

* Is smoothing of the raw $\beta$-values still useful when you have high-coverage sequencing data (__DISCUSS WITH TERRY__)?

A binomial local likelihood smoother was chosen because they model $M_{i, j}$ as $Binom(M_{i, j} + U_{i, j}, B_{i})$ and it is "local" because "methylation levels are strongly correlated across the genome" \citep{Hansen:2012gr}. The authors cite \citet{Eckhardt:2006gh} as evidence that DNA methylation levels are similar at proximal CpGs).

Under the binomial model for $M_{i, j}$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estiamtor of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt(\frac{\beta_{i, j}(1 - \beta_{i, j})}{M_{i, j} + U_{i, j}})$ (__CHECK WITH TERRY: the standard error should be defined in terms of estimates not parameters, i.e. $\beta$ instead of $B$, correct?__). `BSmooth` assumes that for each sample that the underlying methylation level, $f_{j}(i)$, is a smoothly varying function of the position in the genome, $i$.

A window size for smoothing, $w_{i}$, is defined for each $i$. The default window size is defined as one that contains at least 70 CpGs and is at least 2 kb wide, that is, $w_{i} = (pos_{i}, pos_{i'})$, where:
\begin{equation*}
pos_{i'} = \left\{
  \begin{array}{l l}
    pos_{i} + 2000 & \quad \text{if } NIL(pos_{i}, \ldots, pos_{i} + 2000) \geq 70 \\
    pos_{i} + min(\omega: NIL(pos_{i}, \ldots, pos_{i} + \omega) = 70 - 1 = 69) & \quad \text{otherwise}
  \end{array} \right.
\end{equation*}
Note that the end of the window may not be a CpG if $w_{i} = 2000$ but will be if $w_{i} > 2000$.

The raw $\beta$-values are weighted according to the binomial likelihood and a tricube kernel. The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the tricube kernal gives greater weight to those $\beta_{i, j}$ near the centre of the window.

### Step 2 {-}

\citet{Hansen:2012gr} describe `BSmooth` in the context of a simple two-group linear model. The model could be extended to deal with multiple groups but this has not been done.

Let $X_{j} = 1$ if the $j^{th}$ sample is a case and $X_{j} = 0$ if a control. There are $n_{0}$ controls and $n_{1}$ cases ($N = n_{0} + n_{1}$). `BSmooth` assumes that the samples within each group are biological replicates. Within each window, $\log(\frac{f_{j}(i)}{1 - f_{j}(i)})$ is approximated by a second degree polynomial. `BSmooth` transforms the resulting model parameters(__?__) to fit the model $f_{j}(i) = \alpha_{i} + \beta_{i} X_{i} + \epsilon_{i, j}$. The model allows for locus-dependent variation, $\sigma_{i}^{2}$, that, like $f_{j}(i)$, is assumed to be a a smoothly varying function of the position in the genome, $i$.

__DISCUSS WITH TERRY: Are they fitting the model to $\log(\frac{f_{j}(i)}{1 - f_{j}(i)})$, i.e. a logistic model, and then transforming or fitting directly to $f_{j}(i)$, i.e. a linear model?__

In this model, $\alpha_{i}$ represents the base level of methylation at the $i^{th}$ locus and $\beta_{i}$ is the true difference between the two groups. The $\epsilon_{i, j}$ represent biological variability within the locus-dependent variance, $\sigma_{j}^{2}$.

`BSmooth` fits this model to the __smoothed $\beta$-values__. The parameters are estimated as empirical averages:
\begin{align*}
	\hat{\alpha}_{i} &= \frac{1}{N} \sum_{j = 1}^{N} f_{j}(i) \\
	\hat{\beta}_{i}  &= \frac{1}{n_{1}} \sum_{j: X_j = 1} f_{j}(i) - \frac{1}{n_{0}} \sum_{j: X_j = 0} f_{j}(i).
\end{align*}

To estimate $\sigma_{i}$, firstly the within-group standard deviations are computed. Secondly, in an attempt to "improve precision", these standard deviations are truncated at the $75^{th}$ percentile and then smoothed using a running mean (window size of 101 observations). I denote the resulting estimate of $\sigma_{i}$ by $\hat{\sigma}_{i}$.

__TODO: it's not clear from the text whether smoothing+flooring are done on the within-group sds or the grand sd__.

Finally, a t-statistic is computed at each locus as $t_{i} = \frac{\beta_{i}}{\hat{\sigma}_{i} \sqrt(1 / n_{0} + 1 / n_{1})}$. This tests the hypothesis that the average methylation level at this locus is the equal in both groups, or, formally, $H_{0}: \beta_{i}^{0} = \beta_{i}^{0}$ vs. $H_{1}: \beta_{i}^{0} \neq \beta_{i}^{0}$, where $\beta_{i}^{k}$ is the average methylation level in group $k = 0, 1$. __TODO: Don't use $\beta$ in model definition as this gets confused with $\beta$-values__

The authors recommend that t-statistics are only computed at loci with "some coverage in most or all samples".

### Step 3 {-}

Regions with $|t_{i}| > q_{t}^{0.95}$, where $q_{t}^{0.95}$ is the 95th percentile of the empirical distribution of $\{t_i\}_{i = 1}^{M}$, and where all differences are in the same direction, are called putative DMRs.

### Step 4 {-}

 \citet{Hansen:2012gr} recommend _post-hoc_ filtering of these putative DMRs. These filters include an algorithm to filter out very small DRMs and those that did not display a large mean-difference ($> 0.1$) across the DMR; an algorithm to merge neighbouring DMRs if they were within a pre-defined distance of another DMR and displayed a consistent methylation-difference pattern; and an algorithm to clasify the final list of DMRs according to biological significance.

### `BSmooth` use in \citet{Hansen:2011gu} {-}

All WGBS samples were sequenced at very low coverage ($~5 \times$), which initially motivated the smoothing-approach taken by `BSmooth`. The authors only report looking for DMRs in the tumour-normal comparison and not the normal-adenoma or adenoma-tumour comparisons. However, they compared the adenoma methylation profiles to those of the tumour and normals at several DMRs.

Unlike in 'normal' samples, DNA methylation is highly variable in tumour samples. This means that in a two-group experiment, the within-group variation in the tumour samples will often dwarf the between-group variation. To avoid this, \citet{Hansen:2011gu} use only the normal samples to estimate $\sigma_{i}$. They argue that estimating the standard deviations from the normal samples is equivalent to assuming that the tumour samples are not biological replicates (__DISCUSS ARGUMENT WITH TERRY. Also, do they not do a Welch t-test because this does not account for pairing? Is there a "paired Welch t-test"?__).

 \citet{Hansen:2011gu} used two different window sizes for two different analyses -- a large window size of to detect low frequency changes and a small window size of to detect high frequency changes. The large window size required 500 CpGs per window and a minimum window of 2 kb and the small window size required 70 CpGs per window and a minimum window size of 1 kb[^bsmooth_typo1]. They also used a lower t-statistic cutoff for the low-frequency analysis.

[^bsmooth_typo1]: The supplementary material of \cite{Hansen:2011gu} says that the large window size was 2 kb but I think this is a typo and they meant 20 kb. They add to the confusion in \citet{Hansen:2012gr} where they suggest that the small window size is $\geq2$ kb with at $\geq70$ CpGs and that the large window is $\geq40$ kb and $\geq500$ CpGs.

The analysis also includes a correction factor to allow for high-frequency DMRs within low-frequency DMRs, such as a small hypermethylated region within a larger hypomethylated block. To do this, they modified the linear model to be $f_{j}(i) = \alpha_{i} + \beta_{i}^{s} X_{i} + \beta_{i}^{l} X_{i} + \epsilon_{i, j}$, where $\beta_{i}^{s}$ represents small DMRs and $\beta_{i}^{l}$ represents large DMRs. $\beta_{i}^{l}$ is much more slowly varying than $\beta_{i}^{s}$.

Similarly, the t-statistics are separated into $t_{i} = t_{i}^{s} + t_{i}^{l}$. Any site with differential methylation should have a large value of $|t_{i}|$. However, in order to identify small DMRs there needs to be a correction for $t_{i}^{l}$. \citet{Hansen:2011gu} estimate $t_{i}^{l}$ by linearly interpolating $t_{i}$ across all genomic positions, not just methylation loci. This function is evaluation on a 2kb grid and smoothed using a Huber likelihood model in 25 kb windows[^bsmooth_typo2]. The smoothed function is then evaluated at each methylation loci and these values are taken as estimates of $t_{i}^{l}$. They then use the 'corrected' values of $t_{i}^{s} = t_{i} - t_{i}^{l}$ to identify small DMRs, using the method described in steps 2-4.

[bsmooth_typo2]: \citet{Hansen:2012gr} reports that this is a 50 kb window

### `BSmooth` use in \citet{Hansen:2013eo}
__TODO__

## \citet{Hebestreit:2013ko}

`BiSeq`, published in \citet{Hebestreit:2013ko}, is another R/Bioconductor package that performs DMR detection and uses smoothing of the raw $\beta$-values. It It is designed for use with CpG methylation from RRBS data and takes a 5-step approach to DMR detection:

1. Define CpG clusters
2. Smooth methylation data within CpG clusters
3. Model and test group effect within CpG clusters
4. Apply hierarchical testing procedure:
	* Test CpG clusters for differential methylation and control weighted FDR on clusters
	* Trim rejected CpG clusters and control FDR on single CpG sites
5. Define DMR boundaries

### Step 1 {-}

A CpG cluster is defined as a group of CpGs with the following properies (default parameter values in brackets):

1. All CpGs have 'sufficient' sequencing coverage in 'most' samples (75% of samples have 'sufficient' coverage)
2. All CpGs are within a maximum $IPD$ of the next CpG ($100$ bp)
3. There are at least a minimum number of CpGs in the cluster ($20$ CpGs)

These properties are used to define the cluster boundaries but all $\beta$-values within the cluster boundaries are used in subsequent analyses.

### Step 2 {-}

Within each cluster, raw $\beta$-values are smoothed using a binomial local likelihood with a triangular kernel. This is in contrast to `BSmooth` where smoothing is done across the entire genome and not separately within clusters. Also in constrast to `BSmooth`, `BiSeq` uses a much smaller smoothing window of 80 bp, which is fixed across the genome.

### Step 3 {-}

\citet{Hebestreit:2013ko} use the beta distribution to model the smoothed $\beta$-values, that is, smoothed $\beta_{i} = Beta(\mu_{i}, \phi_{i})$. Under this model,
\begin{align*}
E(\beta_{i}) &= \mu_{i} \\
Var(\beta_{i}) &= \frac{\mu_{i} (1 - \mu_{i})}{1 + \phi_{i}}.
\end{align*}

They claim this is the appropriate distribution since $\beta \in [0, 1]$ for both the smoothed and raw $\beta$-values[^biseq_typo1]. This assumption is debatable, as discussed in __SECTION__.

[^biseq_typo1]: Not $\beta \in (0, 1)$ as claimed in the paper.

\citet{Hebestreit:2013ko} develop a regression model of the smoothed $\beta$-values. `BiSeq` regresses the smoothed $\beta$-values against $P$ covariates via a probit link function, $g(\mu) = \Phi^{-1}(\mu)$, where $\Phi^{-1}(.)$ is the standard Normal cumulative distribution function. Thus, $g(\mu_{i}) = X_{i} \beta_{i}$, where $X_{i,j} = (X_{i, 1}, \ldots, X_{x, P})^{T}$ is a $P$-vector of covariates and $\beta_{i} = (\beta_{i, 1}, \ldots, \beta_{i, P})^{T}$ is a $P-vector$ of regression parameters.

__TODO: Don't use $\beta$ in model definition as this gets confused with $\beta$-values__

Maximum likelihood estimators of $\beta$ and $\phi$ are obtained via numerical methods since no closed form solutions are available[^biseq_typo2]. While $\mu_{i}$ is presumably allowed to vary for each $i$, it is not clear whether $\phi_{i}$ is allowed to vary across loci. However, \citet{Hebestreit:2013ko} do suggest that $\phi_{i}$ might be estimated separately within each group when doing a tumour-normal comparison to allow for the increased within-group methylation variability of tumour samples \citep{Hansen:2011gu}.

[^biseq_typo2]: __CHECK WITH TERRY: is $\beta$ or $\mu$ really the regression parameter? I suppose it doesn't really matter since one is a function of the other.

`BiSeq` performs a Wald test (__CITE__) at each CpG as a preliminary test of differential methylation.

### Step 4 {-}

`BiSeq` uses a hierarchical FDR procedure developed by __CITE BENJAMINI and HELLER__. This procedure first tests for significance at the cluster-level and then at CpGs within significant clusters. This reduces the number of hypotheses since only loci within significant clusters are tested, therefore increasing the statistical power.

As\citeauthor{Hebestreit:2013ko} say, the first step is to "$\ldots$ detect CpG clusters containing at least one differentially methylated location and to control a size-weighted FDR on clusters". For each CpG the P-value from the Wald test is transformed to a z-score, $z_{i}$. Each cluster is then assigned an average z-score, $\bar{Z}_{c}$, where $c = 1, \ldots, C$ is the cluster index. The standard deviation of $\bar{Z}_{c}$, $\hat{sigma}_{\bar{Z}_{c}}$, is estimated using an approach that accounts for the correlation of P-values between two loci $i$ and $i'$. This correlation is estimated from a semivariogram of the $z_{i}$. It is recommended that the variogram is estimated from re-sampled data to ensure that the null hypothesis holds and thus that variance of the z-scores follow the standard Normal distribution.

For each cluster, a P-value is obtained as $\Phi^{-1}(\frac{\bar{Z}_{c}}{\hat{sigma}_{\bar{Z}_{c}}})$. Those clusters with a P-value satisfying the Benjamini-Heller FDR criteria are retained for further investigation.

The second step is to trim the clusters of CpGs that are not themselves differentially methylated. A P-value is computed for each CpG within the significant clusters; this P-value is conditional on the cluster being rejected. These conditional P-values are then subjected to a secondary FDR criteria, proposed by Benjamini and Heller, and those sites with P-values greater than the cutoff are trimmed from the clusters.

__DISCUSS WITH TERRY: I don't fully understand how this FDR procedure works. E.g. it would surely miss any significant CpG within an insignificant cluster, rigth?__

### Step 5 {-}

The result of steps 1-4 are significantly differentially methylated CpGs. While `BiSeq` uses clusters to find these DMCs, these clusters are not themselves the DMRs. Rather, DMRs are defined as adjacent DMCs that occur within the same cluster and that display consistent methylation differences, that is, the DMCs are not switching from positive to negative differences.

__NOTE TO SELF: I like the FDR procedure but I'm not convince beta regression is necessary. Perhaps we can get away with the cheaper linear/logistic regression but make use of the cluster + FDR approach?__

## \citet{Akalin:2012cm}
\citet{Akalin:2012cm} developed `methylKit`, an `R` package for processing bisulfite-sequencing data (__LINK TO SOFTWARE__). The methods described in the paper are applicable to WGBS, although all examples in the paper use RRBS data and the accompanying website for the software suggests that WGBS data might be too large for the software to handle [`v0.9.2` https://code.google.com/p/methylkit/](https://code.google.com/p/methylkit/). `methylKit` contains many useful utility functions for processing and visualising bisulfite-sequencing data.

\citet{Akalin:2012cm} does not describe a method for _identifying_ DMRs, rather it describes two methods for _testing_ pre-defined regions

\citet{Akalin:2012cm} describe two methods implemented in `methylKit` for testing for differential methylation at either individual cytosines or at __pre-defined__ regions (using some aggregated/average measure of methylation in the region). It does not describe a method for _identifying_ DMRs, such as post-hoc combining DMCs into regions.

The tests of differential methylation are logistic regression of the raw $\beta$-values or Fisher's exact test of a table of $m$ and $u$ counts. The default in `methylKit` is to use logistic regression if there are multiple samples per group and Fisher's exact test otherwise. \citet{Akalin:2012cm}

Once DMC/DMR testing has been performed, P-values are corrected for multiple hypothesis testing. `methylKit` implements both the sliding linear model (SLIM), which adjusts for correlations amongst nearby P-values \citep{Wang:2011cw}, and the Benjamini-Hochberg (__CITE__) method for P-value correction.

## \citet{Feng:2014iq}
\citet{Feng:2014iq} describe an empirical Bayes hierarchical model to identify DMCs from bisulfite sequencing data. The method is implemented in the R/Bioconductor package, `DSS` (__CITE__).

Let $k = 1, \ldots, K$ denote which group the $j^{th}$ sample belongs to. \citep{Feng:2014iq} describe a case-control experiment (i.e. $K = 2$), and only consider CpGs. `DSS` uses the following beta-binomial model[^dss_notation]:

[^dss_notation]: Note that in order to make the notation in this section consistent with the general framework described in this chapter, there are subtle differences to the index parameters used in \citet{Feng:2014iq}.

\begin{align*}
M_{i, j, k} | B_{i, k}, M_{i, j, k} + U_{i, j, k} &= Binomial(M_{i, j, k} + U_{i, j, k}, B_{i, k}) \\
B_{i, k} &= Beta(\mu_{i, k}, \theta_{i, k}) \\
\theta_{i, k} &= log-Normal(m_{0, k}, r_{0, k}^{2})
\end{align*}

In words, for each CpG the true level of methylation, the prior parameters and the hyperparameters are assumed to be identical within each group.

Parameter estimates are obtained by the following algorithm:

1. For each CpG, estimate the dispersion parameter, $\theta_{i, k}$, using a method  of moments estimator.
2. Estimate the hyperparameters, $m_{0, k}$ and $r_{0, k}^{2}$, as the mean and variance, respectively, of the empirical distribution of the logarithm of the $\theta_{i, k}$.
3. Estimate the group-wise mean level of methylation, $\mu_{i, k}$, by $\hat{\mu}_{i, k} = \sum_{j: j \in group_{k}} \frac{m_{i, j, k}}{m_{i, j, k} + u_{i, j, k}}$.
4. For each CpG, compute relevant quantities, such as the mean and standard deviation, of the conditional posterior distribution of the dispersion parameters, $Pr(\theta_{i, k} | m_{i, j, k}, m_{i, j, k} + u_{i, j, k}, \mu_{i, k})$. In practice, these quantities are computed using the Newton-Raphson method after plugging in estimates of $m_{0, k}, r_{0, k}^{2}$ and $\mu_{i, k}$.

The effect of the this procedure is to shrink the CpG-, group-wise dispersion estimates, $\theta_{i, k}$, towards the group-wise prior mean, $m_{0, k}$. In this form, each CpG site is allowed to have its own group-wise dispersion although `DSS` also allows the user to select a common dispersion across groups. \citet{Feng:2014iq} do not describe any further moderating of the $\theta_{i, k}$, such as the abundance-dependent trended dispersion estimates available in the differential gene expression analysis software, `edgeR` (__CITE__).

Now that all parameters have been estimated, each CpG is tested for differential methylation. Namely, for a two-group experiment, at each CpG `DSS` uses a Wald test of the hypothesis $H_{0}: \mu_{i, 1} = \mu_{i, 2}$ vs. $H_{0}: \mu_{i, 1} \neq \mu_{i, 2}$. Due to the hierarchical structure of the model, it is not straightforward to derive the null distribution of the resulting test statistics. Instead, \citet{Feng:2014iq} use simulation studies to argue that the test statistics can be safely approximated by the standard Normal distribution.

While \citet{Feng:2014iq} focuses an empirical Bayes model for detecting differential methylation at individual CpGs, the authors propose a simple thresholding algorithm to identify DMRs. This algorithm calls DMRs as regions greater than a specified minimum size (100bp), containing a specified minimal number of CpGs (3) and containing multiple DMCs that satisfy a P-value threshold (at least 80% of DMCs with P-values < a user-specific minimal P-value). As the authors note, there is no accounting for the spatial correlation amongst P-values and say that improved DMR calling is left for future work.

## \cite{Sun:2014fk}

The `MOABS` software, published in \citet{Sun:2014fk}, uses a Beta-Binomial empirical Bayes hierarchical model of methylation at individual CpGs. The model is therefore similar in spirit to that proposed in \citet{Feng:2014iq}, although the model is not as well described[^moabs_errors1]. For example, it is not made explicit how the hyperparameters of the Beta distribution are estimated.

[^moabs_errors1]: The EB description is poorly written and I think is wrong in several places, which makes it confusing. (__DISCUSS WITH TERRY__)

For a two-group experiment, `MOABS` does not perform hypothesis testing to identify DMCs. Rather, for each methylation loci, `MOABS` estimates a $95\%$ posterior probability interval, $PI(a, b)$ (__TODO__ Is this a credible interval in Bayesian parlance?) of the difference in mean methylation levels between group 1 and 2[^moabs_errors2].

[^moabs_errors2]: The paper calls this a confidence interval rather than a posterior probability interval. This seems to mix up the hypothesis-testing and Bayesian frameworks.

They then define a _credible methylation difference_ ($CDIF$) as:

\begin{equation*}
PI(a, b) = \left\{
  \begin{array}{l l}
    a & \quad \text{if } a \geq 0 \\
    0 & \quad \text{if } a < 0 < b \\
    b & \quad \text{if } b \leq 0 \\
  \end{array} \right.
\end{equation*}

The CDIF effectively converts all posterior intervals to a single number. \cite{Sun:2014fk} write, "In practice, $CDIF$ represents the conservative estimation of the true methylation difference, i.e. for $97.5\%$ of chance the absolute value of true methylation difference is greater than or equal to that of $CDIF$." They interpret $CDIF = 0$ to mean that there is no significant difference in methylation betweent the two groups[^moabs_errors3].

[^moabs_errors3]: Again, this seems to mix up the hypothesis-testing and Bayesian frameworks.

DMCs are those cytosines with a $CDIF > \text{cutoff}$, where $\text{cutoff}$ is chosen so as to control the FDR. The FDR-based $\text{cutoff}$ is estimated by permutating sample labels and re-running the analysis (__CHECK WITH TERRY: Necessary to do this permutation multiple times?__).

The procedure could also be used to identify methylation changes at pre-defined regions, rather than individual methylation loci, by aggregating/averaging methylation levels across each region.

For a two-group experiment, `MOABS` uses a hidden Markov model (HMM) to group cytosines into DMRs. Note that all cytosines, not just DMCs, are used by the DMR-detection HMM. To do this, `MOABS` uses a first-order HMM where the observed values are the group-wise difference in mean methylation at each locus, $B_{i}^{2} - B_{i}^{1}$. There are 3 hidden states -- hypomethylation ($B_{i}^{2} - B_{i}^{1} < -\text{cutoff}$), no difference (|$B_{i}^{2} - B_{i}^{1}| < \text{cutoff}$) and hypermethylation ($B_{i}^{2} - B_{i}^{1} > \text{cutoff}$). The $\text{cutoff}$ is the same at that used for identifying DMCs (__DISCUSS WITH TERRY: I think this is true. Is this a good idea?__).

`MOABS` is written in `C++` (__TODO: Is it a command line program or is there an R interface?__) and also provides several other useful functions including:

* A first-order, 2-state HMM to detect hypomethylated and hypermethylated regions from a single sample
* A script to downsample BAM files in regions containing predicted copy number variants (CNVs). \cite{Sun:2014fk} claim that "this will result in CNV bias free `BAM` files for downstream analysis".

## \citet{Stockwell:2014fq}

`DMAP`, published in \citet{Stockwell:2014fq}, does not allow detection of DMCs, rather it focuses solely on identifying DMRs. `DMAP` provides a function to test for differential methylation in sliding windows, _a la_ \cite{Lister:2009hy}, which the authors recommend for analysing WGBS data. For analysing RRBS data, `DMAP` proposes the use of the MspI fragments as the natural unit for calling DMRs. MspI fragments are 40-220 bp long and are identified by scanning the reference genome for the recognition motif, _C'CGG_. `DMAP` includes an algorithm for resolving methylation calls from reads that overlap adjacent MspI fragments.

Within each window, whether it be a fixed-width sliding window or based on MspI fragments, a Fisher's exact test is performed between two samples[^dmap]. If there are more than two samples then the minimum P-value from all pair-wise comparisons is reported as evidence of differential methylation. `DMAP` also provides a $\chi^2$-test for inter-individual variability of DNA methylation in multiple samples.

[^dmap]: It is not clear whether `DMAP` only uses the Fisher's exact test for two __samples__ or also for two __groups__, e.g., by first aggregating all read counts within each group and then doing a Fisher's exact test.

`DMAP` can also perform an F-test to compare DNA methylation between two groups, for example, cases and controls. This test is implemented outside of a regression framework and so does not allow for additional covariates to be included in the model.

Regardless of which test is used, DMRs are called as those regions with a P-values below a given threshold and satisfying other criteria such as a minimum number of CpGs in the window.

## \citet{Lacey:2013iy}

\cite{Lacey:2013iy} details a modeling and inference framework for RRBS data, as well as a method for simulating RRBS data. Here I only describe the modelin framework and reserve the description and discussion of the simulation framework for __CHAPTER__.

The modeling framework is based on a two-group experiment. At each methylation loci the authors fit a logistic regression model of $\beta_i$ against a indicator variable denoting from which group the sample comes. The raw P-values from the $\chi^2$ test of the residuals of this model are then transformed by a procedure dubbed `QUASASS` (quantile adjustment for score and sample size). The authors claim that `QUASASS` adjusts for the differences in read coverage and sample size that would otherwise reduce the sensitivity in low-coverage regions and reduce the specificity in high-coverage regions. __DISCUSS WITH TERRY: What is `QUASASS` actually doing?__. The authors recommend that `QUASASS`-adjusted P-values, rather than the raw P-values, are used in all downstream analyses.

To identify DMRs, \citeauthor{Lacey:2013iy} seek to model the distribution of P-values at neighbouring methylation loci. They use a Uniform Product distribution to model the P-values under the null hypothesis of no differential methylation at each loci. This model assumes that P-values at neighbouring loci are independent, which seems unlikely given the spatial correlation of $\beta$ values.

DMRs are regions where the first methylation loci has a P-value $\leq 0.05$ and are then built up using an iterative procedure. This iterative algorithm continues to add neighbouring methylation loci (up to a user specific threshold with default value of $50$) and computes the P-value of the region under the Uniform Product model. The maximum number of additional loci that retains a region-wise P-value $\leq 0.05$ is defined as the end of the region provided the following conditions are also met:

* That the end site by itself has a P-value $< 0.05$
* All differences in methylation are 'in the same direction'
* All methylation loci in the region are in "sufficiently close proximity"

The final list of putative DMRs is then pruned to remove regions containing only a single loci and overlapping DMRs are merged.

\citet{Lacey:2013iy} state that an advantage of this iterative procedure is that, "DMRs can include internal sites that are not significantly differentially methylated" as well as sites that are missing data.

An R package implementing the algorithm is available from [http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/).

### Simulation model {-}

__This section to move to chapter on simulating WGBS data.__

It is important to note that \citet{Lacey:2013iy} seek to simulate RRBS data and not WGBS data. This has several implications. Firstly, the RRBS assay enriches for small, CpG-dense regions of the genome such as CGIs. As such the distribution of IPDs is skewed towards zero and is less bimodal than that from WGBS data (__CHECK__), although it still has a long right-tail. Secondly, these CpG-dense regions are enriched for methylation loci with methylation levels near zero. Finally, the sequencing coverage of loci is right-skewed, even bimodal, in RRBS data.

All parameters used in the simulation model were estimated from a single normal myotube cell line (MTCTL2; __CITE Tsumagari et al., 2013__). \citeauthor{Lacey:2013iy} state that this sample is representative of the methylation profiles that they are typically interested in as part of their research emphasis. It is therefore unclear how generalisable these parameters estimates are.

Rather than simulate data at methylation loci from a reference genome, the authors model the distribution of IPDs for CpGs. Specifically, they model IPDs in human RRBS data. For this they use a two-component Normal mixture distribution, where one component models CpG-dense regions of the genome and the second component models CpG-sparse regions of the genome. They further impose a hidden Markov model on the distribution of IPDs so that a small IPD is more likely to be followed by another small IPD, in an attempt to model the clusters of CpG sites observed in RRBS data. It is unclear to me what advantage, if any, modelling the distribution of IPDs has over simply taking the empirical distribution of IPDs from a reference genome.

Unlike `methsim`, \citeauthor{Lacey:2013iy} simulate $\beta$-values and sequencing coverage rather than simulating the reads themselves. They use a mixture of two gamma distributions to model the distribution of log-transformed sequencing coverage. This model is applied independently to bins of methylation loci (bin width = $50$ bp) and thus all loci within each bin have the same sequencing coverage for a given sample.  As this model is independently applied to each bin, it does not quite account for the fact that sequencing coverage is highly correlated at loci separated by short distances ($< 50-200$ bp) due to each read potentially containing multiple methylation loci/covering multiple bins (__TODO: show correlation of sequencing coverage elsewhere in thesis and reference here__). They did, however, model the correlation of sequencing depth at each site across samples (median correlation $= 0.8$) via a Normal copula. The simulated values are then exponentiated (as the model is of logarithmic coverage) and rounded to the nearest integer (as the model produces non-integer variables).

The simulation procedure aims to capture the spatial correlation of $\beta$-values. This correlation was estimated by fitting a Guassian model to the empirical variogram for sites randomly selected from chromosome 11 of the MTCTL2 data. \citet{Lacey:2013iy} report that the variogram shows "a strong correlation for sites in close proximity, decaying to near independence at distances beyong 3000 bp"[^lacey]. The spatially correlated $\beta$-values were then simulated via a two-step process. Firstly, a sample of independent $\beta$-values were sampled from a Beta distribution, with parameters estimated from chr11 of the MTCTL2 data assuming independence of loci. Secondly, correlations were induced via the model proposed by __CITE Zaykin et al. and describe the model__, which uses the correlations estimated from the fitted variogram to induce correlations in the Beta random variates. The parameters of the underlying Beta distribution are then re-estimated to better approximate the distribution of $\beta$-values from the MTCTL2 chromosome 11 data. This algorithm was run several times to arrive at the final parameter values.

[^lacey]: My results show a non-zero correlation beyond 3 kb, which is hardly independence.

They also argue, however, that the correlations of differences in $\beta$-values between two groups will be less than correlations of the $\beta$-values themselves, and provide some evidence for this. To do this they fit exponential variogram models to the raw and `QUASASS`-adjusted P-values from non-differentially methylated loci and report $r \leq 0.4$ for sites with $IPD > 50$ bp and $r < 0.2$ for sites with $IPD > 1000$ bp. It is not clear whether only pairs of sites with $NIL = 0$ where used in this analysis.

\citet{Lacey:2013iy} also propose an algorithm for inducing DMRs in a two-group experiment, where the first $\frac{n}{2}$ samples are cases and the second $\frac{n}{2}$ samples are controls. The DMR-construction algorithm is as follows:

1. Specify the proportion of the methylation loci that are in DMRs, the expected difference in methylation levels for loci in these DMRs ($\delta_{B}$) and the length of these DMRs.
2. Regions satisfying the minimum length requirement are identified from the data and a subset are sampled so that the constraint on the proportion of methylation loci in DMRs is satisfied.
3. Compute the median $B_{i, j}$ for control samples in each region.
	* If the median $B_{i, j}$ in the controls is less than $\delta_{B}$ then the region is assigned as hypermethylated relative to the controls.
	* If the median $B_{i, j}$ in the controls is greater than $1 - \delta_{B}$ then the region is assigned as hypomethylated relative to the controls.
	* Otherwise the region is randomly assigned as hypomethylated or hypermethylated in controls.
4. If the region is assigned as hypomethylated, then the methylation levels of the cases in that region are replaced with $min(B_{i, j} - \delta_{B}, 0.01)$. If the region is hypermethylated, then the methylation levels of the cases in that region are replaced with $max(B_{i, j} + \delta_{B}, 0.99)$.

Under this model all DMRs are of the same length, although they may contain a different number of methylation loci, and all DMRs have the same expected difference in methylation levels between cases and controls.

To summarise, \citeauthor{Lacey:2013iy} propose the following algorithm to simulate RRBS data, including DMRs, for $n_{loci}$ loci and $n$ samples:

1. Simulate $n_{loci} - 1$ IPDs. All samples are assumed to have the same set of resulting methylation loci, i.e. $\mathcal{I}_{1} = \ldots = \mathcal{I}_{n}$.
2. Bin the $n_loci$ loci in 50 bp bins. For each bin, simulate sequencing coverage across the $n$ samples using the Normal copula model of sequencing coverage. Let $n_{i, j} = m_{i, j} + u_{i, j}$ by the simulated sequencing coverage at the $i^{th}$ locus for the $j^{th}$ sample.
3. Simulate spatially correlated methylation levels, $B_{i}$. __The same $B$-values are used for all samples, i.e. $B_{i, j} = B_{i}$.__
4. Construct DMRs by modifying the methylation levels $B_{i, j}$ of cases within these regions.
5. Generate observed $\beta$-values, $\beta_{i, j} = \frac{m_{i, j}}{n_{i, j}}$ using the binomial model, $m_{i, j} = Bin(n_{i, j}, B_{i, j})$, where outside of DMRs $B_{i, j} = B_{i}$ and inside DMRs $B_{i} = B_{i, case}$ or $B_{i} = B_{i, control}$, as appropriate.

## \citet{Akulenko:2013fl}

\citet{Akulenko:2013fl} investigated the co-methylation of genes between samples in a study of $344$ samples from The Cancer Genome Atlas (TCGA). Most of these samples ($317/344$) were breast cancer samples while the remainder ($27/344$) were from matched normal tissue. This study used data from the Illumina Infinium HumanMethylation27k BeadChip (__CITE__), which contains approximately $27,578$ probes that measure single-CpG DNA methylation at CpGs in $14,475$ (__CITE: Original paper not Akulenko__). Most (__HOW MANY__) of these probes are in gene promoters.

After pre-processing of the raw data, each gene was assigned the average $\beta$-value of all $\beta$-values within that gene for each sample. This reduced the number of $\beta$-values from 25,578 to 13,313. After some further filtering to remove probes that were known to be prone to technical artefacts, they computed the Pearson correlation ($r$) of all pairs of gene-level $\beta$-values ($88,611,328$ pairs) using all 344 samples.

This analysis identified $13,643$ pairs of $\beta$-values with $|r| > 0.9$ and $377,547$ pairs of $\beta$-values with $|r| > 0.75$. The overwhelming majority ($377360/377547, 99.95\%$) of pairs with $|r| > 0.75$ were driven by outlier observations, whereby a small number of samples had $\beta$-values greatly different to the rest of the samples.

\citeauthor{Akulenko:2013fl} also investigated how these pair-wise correlations vary as a function of the genomic distance between the two probes in each pair. They reported that "co-methylation level only weakly anti-correlated with genomic distance ($r = -0.29$)", however, this analysis only used a tiny subset of the dataset. Namely, they only performed this analysis for the 74 of the 187 "significant" pairs of $\beta$-values where both genes in the pair were on the same chromosome.

There are several limitations to this study:

1. The low resolution of the 27k array, which leads to summarising a gene's methylation level from measurements of one or two CpGs.
2. They only used a tiny subset of the available data to explore the relationship between co-methylation and genomic distance.
3. Once the "outliers" were removed, almost all the 'co-methylation' disappears. I think these outlier samples need closer investigation rather than just removal.

## \citet{Sofer:2013bk}

The [`Aclust`](http://www.hsph.harvard.edu/tamar-sofer/packages/) software, published in \citet{Sofer:2013bk}, uses generalised estimating equations to detect differential methylation. `Aclust` first clusters sites and then tests for exposure effects on clusters (e.g. case/control status). This in contrast to most other DMR methods, which test for exposure effects at individual methylation loci and then cluster loci. While `Aclust` was designed for analysing methylation array data, in principle it should also work for bisulfite-sequencing data.

`Aclust` uses the following model, where $E_{j}$ denotes the exposure of the $j^{th}$ sample and $X_{j}$, a $p \times n$ matrix, denotes $p$ additional covariates for each of the $n$ samples, such as the batch or age of the sample[^aclust_notation]:

[^aclust_notation]: Note that this differs from the original notation, particularly in the use of index variables.

\begin{equation*}
\beta_{i, j} = \alpha_{i} + E_{j} \alpha_{E_{i}} + X_{j}^{T} \alpha_{X_{i, j}} + \epsilon_{i, j}.
\end{equation*}

This model allows each of the $i = 1, \ldots, n_{loci}$ methylation loci to have a unique baseline methylation value ($\alpha_{i}$), an exposure effect ($\alpha_{E_{i}}$) and covariate effects ($\alpha_{X_{i}}$). Note that the covariates themselves can vary across loci. The errors, $\epsilon_{i, j}$, are assumed to follow a zero-mean distribution. The test of interest is generally of $H_{0}: \alpha_{E_{i}} = 0$ vs. $H_{1}: \alpha_{E_{i}} \neq 0$.

Rather than test every loci, however, `Aclust` first clusters the loci and instead fits a related model to the clusters themselves. Suppose that loci $i = 1, 2, 3$ are determined to be in a cluster, $c$. Then, the model for cluster $c$ is:

\begin{equation*}
\beta_{i, j} = \alpha_{i} + E_{j} \alpha_{E_{c}} + X_{j}^{T} \alpha_{X_{i, j}} + \epsilon_{i, j}.
\end{equation*}

This model allows for each of the $i = 1, 2, 3$ loci in the $c^{th}$ cluster to have a unique baseline methylation value ($\alpha_{i}$) and covariate effects ($\alpha_{X_{i}}$). However, the exposure effect, $E_{c}$, is assumed to be constant for all loci in the cluster. The errors, $\epsilon_{i, j}$, are assumed to follow a zero-mean distribution with a covariance matrix[^aclust_typo]. The covariance matrix could include covariances between errors $\epsilon_{i, j}$ and $\epsilon_{i, j'}$. The test of interest is generally of $H_{0}: \alpha_{E_{c}} = 0$ vs. $H_{1}: \alpha_{E_{c}} \neq 0$.

[^aclust_type]: The $\epsilon$ are defined in two different ways in the paper (p2).

`Aclust` performs a form of agglomorative nested clustering (__CITE Izenman, 2008__) of adjacent methylation loci. Initially, each cluster, $c_{l}$, is comprised of a single methylation loci. Then, clusters are iteratively merged if the _distance metric_ of the two clusters is less than a minimum value, $\bar{\mathcal{D}}$, (default 0.25) and the $IPD$ of the last site of the first cluster and the first site of the second cluster is less than a minimum distance (default 1 kb). The algorithm terminates when no more clusters can be merged.

The distance metric for two methylation loci, $(i, i')$, is $dist(i, i') = 1 - cor(\{(\beta_{i, j}, \beta_{i', j})\}_{j = 1}^{n})$, with the Spearman correlation as the default. The distance metric between two clsuters may use _single_, _average_ or _complete_ distance of the all $dist(i, i')$, where $i$ is from the first cluster and $i'$ is from the second cluster. The _single_ distance requires that only at least one pair of methylation loci have $dist(i, i') < \bar{\mathcal{D}}$, the _average_ distance requires that the mean distance between all pairs of loci is $< \bar{\mathcal{D}}$, and the _complete_ distance requires that the distance between all pairs of loci is $< \bar{\mathcal{D}}$. For a fixed $\bar{\mathcal{D}}$, the _complete_ distance produces smaller clusters than the _average_ distance, than in turn produces smaller clusters than the _single_ distance. The authors recommend _average_ or _complete_ linkage for identifying DMRs, based on their simulation study.

The authors also recommend an initial merging step prior to clustering, which merges all loci within a given $IPD$ (default 99 bp). This step is recommended in order to avoid `Aclust` breaking larger clusters of generally highly correlated loci into multiple smaller clusters due to an intervening methylation loci that is not as correlated with the rest of the loci in the larger cluster. Clusters may be filtered out if they do not contain a minimum number of methylation loci.

Once the final set of clusters is formed, the effect of the exposure variable on each cluster is tested within the GEE framework. `Aclust` uses a robust sandwich variance estimator and requires the user to supply a (possibly mis-specified) working covariance matrix. A Wald test is used to test for the effect of the exposure. Finally, the resulting P-values are corrected for multiple testing via the Benjamini-Hochberg procedure (__CITE__).

### Simulation methodology {-}

\citeauthor{Sofer:2013bk} adapt the methodology of __CITE: Gaile et al.__ to simulate correlated $\beta$-values in a two-group experiment. They simulate Illumina 450k beadchip data with $n_{1} = 40$ cases and $n_{2} = 40$ controls. The simulated data are based on a block re-sampling of data from 539 breast invasive adenocarcinoma samples assayed on the Illumina 450k beandchip obtained from The Cancer Genome Atlas (TCGA). By sampling blocks of CpGs from the same sample rather than sampling individual CpGs, the correlations of the $\beta$-values are preserved.

Firstly, they segment the genome into blocks ($n_{blocks} = 2861$) based on the distance between assayed CpGs, where a new block is formed if the next CpG is more than 10 kb from the previous CpG. Singleton CpGs were added to the nearest block. This segmentation means that methylation levels at CpGs in the same block are correlated, although this is imposed somewhat indirectly by segmenting on $IPD$ rather than the correlations themselves.

Based on the 539 TCGA samples, they then selected a small number of CpGs (10), called targets, that displayed substantial variability across the 539 samples, had "substantial correlation with neighbouring sites" and were not in the same block. They then simulated sample-specific blocks of $\beta$-values by sampling these from the 539 TCGA samples. If a block contained one of the targets, then sampling was done in such a way that one group had that block preferentially sampled from those samples' blocks with a high $\beta$-value at that target and the other group had that block preferentially sampled from those samples' blocks with a low $\beta$-value at that target. This induces differential methylation between the two-groups at that target and, more generally, across several CpGs in the block due to the correlation amongst $\beta$-values. Blocks did not contain targets were sampled from the 539 TCGA samples without any sampling weights.

Due to the correlation structure of the $\beta$-values, it is likely, though not guaranteed, that other CpGs in the blocks containing targets also display differential methylation.

## \citet{Su:2012hl}

\citet{Su:2012hl} describes the `CpG_MPs` software (http://bioinfo.hrbmu.edu.cn/CpG\_MPs) for identifying differential methylation patterns from bisulfite sequencing data. `CpG_MPs` scans the genome for contiguous runs of $\beta$-values $\leq 0.3$ ($\geq 0.7$) and calls these as unmethylated (methylated) "hotspots" (regions)). It then extends these regions by adding neighbouring CpGs, provided that there is no more than 1 Cpg with a $\beta$-values $> 0.3$ ($< 0.7$) in the extension. Finally, `CpG_MPs` computes summary statistics, such as the mean and standard deviation, of the $\beta$-values in these hotspots (__I don't know when these values are even used by `CpG_Mps`__).

`CpG_MPs` also assigns methylated hotspots a value of $1$ and unmethylated hotspots a value of $-1$. Hotspots that overlap across samples are called an overlapping region (OR) and are assigned a score, $-1 \leq u \leq 1$, based on an aggregation of the hotspot scores. Overlapping regions are labelled as "conservatively unmethylated regions" if $u = -1$, DMRs if $ -1 < u < 1$ and "conservatively methylated regions" if $u = 1$. __Then there's a whole bunch of other weird arbitrary cutoffs, definitions and use of Shannon Entropy to further refine these regions__.

## \citet{Liu:dy}

\citet{Liu:dy} identified clusters of CpGs that had correlated $\beta$-values and where the level of methylation in the cluster was influenced by genetic variation (SNPs), which they called these _GeMes_. They used from three studies ($n = 247, n = 91$ and $n= 305$, respectively), where DNA methylation was measured using the Illumina 450k beadchip and each sample was also genotyped using a SNP array. I only describe what they discovered about $\beta$-value correlations and do not discuss the algorithm for finding GeMes or the GeMes themselves.

Spatial correlations of $\beta$-values along the genome across multiple samples are similar to linkage disequilibrium (LD) between SNPs. \citeauthor{Liu:dy} explicitly compared the strength of these correlations in the same individuals. They reported that the $\beta$-value correlations, which started at around $0.5$, were reduced to less than $0.25$ when $IPD > 500$ bp and were more-or-less zero when $IPD > 2000$ bp. In contrast, SNP LD correlations in the same individuals, which also started at around $0.5$, were reduced by half when $IPD > 3000$ bp. Spatial correlations of $\beta$-values used only the top $25\%$ most variably methylated CpGs and were smoothed using a cubic spline. They also drew heat maps, like LD block maps, of $\beta$-value correlations to display their results.

## \citet{Xie:2011cy} and \citet{He:2013cj}

\citet{Xie:2011cy} define _methylation entropy_ (ME) to quantify the heterogeneity of methylation patterns at m-tuples within a single sample. `DMEAS`, published in \citet{He:2013cj} and available from [http://sourceforge.net/projects/dmeas/files/](http://sourceforge.net/projects/dmeas/files/), is software that implements an algorithm to compute ME.

m-tuples can have identical $ME$ but very different average levels of methylation ($\beta$-values) at each of the $m$ loci. For example, an m-tuple that has only fully methylated reads and an m-tuple that has only fully unmethylated reads both have an $ME = 0$ because there is no heterogeneity at either m-tuple. Similarly, m-tuples can have different $\beta$-values but identical $ME$.

ME is a modified version of Shannon entropy and is defined as:

\begin{equation*}
ME = \frac{e}{m} \sum_{p = 1}^{2^{m}} -\frac{n_{p}}{N} \log_{10}(frac{n_{p}}{N}),
\end{equation*}

where $m$ is the size of the m-tuple, $N$ is the number of reads covering the m-tuple and $n_{p}$ is the number of reads with the $p^{th}$ methylation pattern[^dmeas_problem]. $e$ is never properly defined in the paper but is described as the "entropy for code bit". In a supplementary file available on the `DMEAS` website, it is stated that $e = \frac{\ln(10)}{\ln(2)}$ (__CITE http://ufpr.dl.sourceforge.net/project/dmeas/DMEAS%20user%20guide.pdf__).

[^dmeas_problem]: The description of the `DMEAS` software in \citet{He:2013cj} allows for methylation loci with an unknown methylation state. Under this model there are now $3^{m}$ possible methylation patterns per m-tuple and the summatation index should be updated accordingly.

I think that this means it is the entropy value of the $p^{th}$ methylation pattern, where an unmethylated CpG is 0 and methylated CpG is $1$[^dmeas_problem2]. This amounts to $e$ being equal to the number of methylated CpGs in the m-tuple. For example, $e = 4$ if the pattern contains 4 methylated CpGs and $e = 2$ if the pattern contains 2 methylated CpGs. This is then normalised to account for the size of the m-tuple, $0 \leq \frac{e}{m} \leq 1$.

[^dmeas_problem2]: The `DMEAS` software further defines a CpG of unknown methylation state as 2. This seems to me like it would screw things up because $ME > 1$ is possible if an m-tuple contains a CpG with unknown methylation state.

ME is minimal ($ME = 0$) when all reads mapping to an m-tuple have the same methylation pattern and is maximal ($ME = 1$) when all $2^{m}$ methylation patterns are observed at equal frequency.

The statistical significance of an observed $ME$ is assessed using a permutation test. The null hypothesis is that the observed methylation patterns are purely stochastic. Specifically, for each m-tuple they simulate $10,000$ datasets with the same $N$ and average $\beta$-value as the observed data and compute the $ME$. The observed $ME$ is compared to this permutation distribution to compute a P-value. They also devise a test of "allele-specific methylation", although strictly speaking this is a test of whether there are two or more methylation subpopulations in the sample because it does not make use of genetic data.

As described  in \citet{He:2013cj}, `DMEAS` has very limited functionality because it can only examine 4-tuples of CpGs with $\geq 16$x sequencing coverage. Furthermore, `DMEAS` is only available as compiled software for Windows or as a Perl script that is provided in a `PDF` document.

## \citet{Li:2013gn}

[`eDMR`](https://github.com/ShengLi/edmr), published in \citet{Li:2013gn}, is software to identify DMRs from two-group experiments using bisulfite sequencing data. The publication focuses on CpG methylation from enrichment assays, such as RRBS, and it is unclear whether the software scales to WGBS data. The example in the dataset is an RRBS experiment of ten accute myeloid leukemia (AML) samples and five normal bone marrow (NBM) samples.

`DMR` is provided as an `R` package and is intended to be compatible with the `methylKit` R package. Indeed, it first uses `methylKit` to identify DMCs using Fisher's exact test at CpGs with at least $10\times$ coverage in at least 3 NBM samples. By using Fisher's exact test they are ignoring all within-group variability of methylation levels.

`eDMR` "aims to optimize the threshold for determining DNA methylation regions and to perform statistical significance testing". The first step is to define clusters of CpGs. They do this using this somewhat convoluted algorithm:

1. Compute the empirical distribution of $IPD$s for all CpGs with $\geq 10\times$ coverage[^edmr_problem1]
2. Fit a mixture of Normal distributions to this empirical distribution.
3. Determine the best separation point between the two components of the mixture distribution. This step incorporates a cost function, the role of which I don't really understand. This separation point, $D$, is used to define the boundaries of regions. $D = 183.5$ in the AML vs. NBM example.
4. Scan the genome for CpGs with $IPD > D$[^edmr_problem2]. The first CpGs of such a pair is the end of the previous region and the second CpG of such a pair is the start of the next region. The average level of methylation within each region in each group is the average of the $\beta$-values within that region in each group.
5. eDMRs are then defined as those regions with satisfying the following default criteria: contains at least 1 DMC, contains at least 3 CpGs, and the absolute mean methylation difference is greater than $20\%$. The exact cut-offs can be specified by the user.

[^edmr_problem1]: It is not clear for how many samples these sites must have $\geq 10\times$ coverage.
[^edmr_problem2]: It's not clear if all CpGs are used or just those with sequencing coverage $\geq 10 \times$.

Each eDMR is then assigned a P-value by combining the P-values from the Fisher's test used to identify DMCs in that eDMR. P-values are combined using the Stouffer-Liptak method (__CITE__). These final P-values are then adjusted using, I believe, the Benjamini-Hochberg FDR procedure.

## \citet{Lyko:2010dr}

__This should go in the co-methylation section.__

\citet{Lyko:2010dr} performed WGBS of honey bee (_Apis mellifera_) brains. It was known that young bees fed large amounts of royal jelly developed into queens (fertile females) whereas those fed smaller amounts develop into drones (males) or workers (infertile females). Furthermore, \citet{Kucharski:2008gu} had shown that a similar effect could be achieved by silencing expression of the DNA methyltransferase, DNMT3. Therefore,\citet{Lyko:2010dr} investigated the hypothesis that honey bees fed large amounts of royal jelly (queens) have different brain methylomes to those bees fed smaller amounts of royal jelly (workers).

As a part of their study, \citeauthor{Lyko:2010dr} investigated autocorrelation of $\beta$-values at CpGs in honey bees. It is not clear whether they only considered pairs with $NIL=0$ or whether they used all pairs ($NIL \geq 0$).

They found that

> As in the human and Arabidopsis genomes, methylation in Apis shows evidence of periodicity, although due to a much lower density of modified CpGs in this species the periodicity of 10 nucleotides (one helical DNA turn) is not obvious. However, a 3-base periodic pattern is clearly detectable, reflecting a preferential methylation of CpGs occupying the first and second position of the arginine codons.

However, the absolute value of these correlations are __tiny__, with the maximum being less than $0.015$. I suspect there is an error in these plots (Supplementary Figure 7, __DISCUSS WITH TERRY__).

They did not stratify their analysis by genomic elements, such as CGIs.

## \citet{Peng:2012dh}

\citet{Peng:2012dh} descibe an algorithm for identifying allele specific methylation (ASM). Their method assigns each read to one of $N_{h}$ "epigenomes" based on the pattern of methylation along the read. Each "epigenome" corresponds to what I call a haplotype, and the model assumes that the number of haplotypes in the sample is equal to the ploidy of the sample, e.g., $N_{h} = 2$ for a human sample.

Basically, \citeauthor{Peng:2012dh} assume a $H$-component mixture model and assign reads to each component, i.e. haplotype, using an EM algorithm. The output is a $N_{loci} \times N_{H}$ matrix, $A$, where $A[i, h]$ is the EM-estimate of methylation at the $i^{th}$ loci in the $h^{th}$ haplotype.

Unlike other methods for detecting ASM, this algorithm does not rely on methylation loci being close to heterozygous SNPs and therefore, at least, theoretically, can detect ASM in any region of the genome.

In practice, this method will only work well in regions with a high density of methylation loci and where the $N_{h}$ epigenomes are quite distinct from one another. The example described in the paper focuses on PMDs, regions where we know there are multiple methylation patterns, in _Arabidopsis_, an organism with a "reasonably high" density of methylation loci.
Furthermore, the example dataset is rather artificial because it is a synthetic dataset made by combining reads from two _Arabidopsis_ methylomes.

__There is no software available that implements the proposed algorithm.__

## \citet{Qu:2013ji}

`MLML`, published in \citet{Qu:2013ji}, is software to jointly estimate 5mC and 5hmC levels from samples sequenced with any two of BS-seq, oxBS-seq and TAB-seq. Depending on the exact combination of technologies, the estimation of either 5mC or 5hmC levels is based on a "subtraction" of read counts between the two experiments. A naive "subtraction" can result in estimates that are $< 0$ or $> 1$.

`MLML` implements an EM algorithm to estimate both 5mC and 5hmC levels under a Binomial-mixture likelihood. The resulting estimates are guaranteed to be "non-negative, and never sum over one". __It's not immediately clear why `MLML` does not guarantee that the estimates sum to exactly one.__

## \citet{TricheJr:2013tj}

\citet{TricheJr:2013tj} investigated different regression models for identifying DMCs and DMRs from Illumina methylation beadchip data. They report that a Beta-regression model[^beta_regression] "yielded more validated hits" than models that used Wilcoxon-Mann-Whitney tests, Student's t-tests or Welch's t-tests.

[^beta_regression]: A "Beta regression model" means explicitly modelling the $\beta$-values as Beta random variables. This is distinct from a "regression of the $\beta$-values, where other distributional assumptions or transformations of the $\beta$-values might be used in a standard linear model framework.

Beta regression (__CITE Ferrari and Cribari-Neto__) does not fit into the generalise linear model framework (__CHECK WITH TERRY__) because the mean and variance terms are coupled to both depend on covariates.  A software implementation of the proposed Beta regression model is __AVAILABLE WHERE__.

\citeauthor{TricheJr:2013tj} caution that when the number of samples per group is small ($< 25$) or when the design is unbalanced that control of the Type-I error rate may be lost.

## \citet{Xu:2013eg}

\citet{Xu:2013eg} propose a test to identify DMCs from bisulfite-sequencing data in two-group experiments. Their model is based on a method for the analysis of clustered binary data, published in __CITE Rao and Scott, 1992__.

For the $i^{th}$ individual, the methylation level at the $j^{th}$ locus is modelled as a binomial random variable, $M_{i, j} = Bin(M_{i, j} + M_{i, j}, B_{i, j}$. Within each group, the observed $\beta$-values are adjusted by a group-specific "design effect". The design effect for the $i^{th}$ locus in the $l^{th}$ group, $d_{i, l}$, is the ratio of the estimated $\beta$-value variability while accounting for within-group variability to without accounting for within-group variability (__I THINK__). The design effect is used to correct both the average withing-group count of methylated reads and the average within-group sequencing coverage.

The test statistic for the $i^{th}$ locus is then a $\chi^2$ statistic on 1 degree of freedom comparing the each group's design-adjusted average $\beta$-value to the design-adjusted grand average $\beta$-value. __This model seems convoluted and I wonder whether it fits into a simpler GLM analaysis of deviance framework__.

__There is no software available that implements the proposed algorithm.__

### Simulation study {-}

\citet{Xu:2013eg} includes a simple simulation study. The "true" group-wise methylation levels at each methylation loci, $B_{i, l}$, are simulated from a beta, a (truncated) Normal or a (truncated) Normal-mixture distribution. Thus all samples within each group have the same "true" methylation level and the methylation level at each locus is independent. Sequencing coverage for each individuals was sampled from a (rounded) Normal distribution.

## \citet{Zhang:2012id}

\citet{Zhang:2012id} propose a Dirichlet process beta mixture model (DPBMM) to cluster methylation profiles. The Dirichlet process effectively means there is an 'infinite' number of components to the mixture distribution. The resulting model is analytically intractable and so they propose a Gibbs sampler to compute the relevant posterior distributions. DPBMM does not require that the number of clusters be pre-specified as this is estimated as part of the Gibbs sampler.

MATLAB code implementing the algorithm is available from [https://sites.google.com/ site/bdpmmmethy/home](https://sites.google.com/ site/bdpmmmethy/home).

## \citet{Zhang:2011dp}

`QDMR`, published in \citet{Zhang:2011dp}, is another method for identifying DMRs based on a modified form of Shannon Entropy. `QDMR` uses $\beta$-values and is so applicable to both microarray- and sequencing-based bisulfite-treatment assays. The publication uses MeDIP-chip and RRBS data.

The methylation entropy of a locus is computed using all samples, without regard to an 'group' variable. The methylation entropy of a region, rather than a single methylation locus, can be computed by using the sample-wise 'averages' of $\beta$-value across the region.

The `QDMR` software is available from [http://bioinfo.hrbmu.edu.cn/qdmr](http://bioinfo.hrbmu.edu.cn/qdmr).

## \citet{Landan:2012kp}

\citet{Landan:2012kp} investigated the stochastic variability of methylation patterns at CpG 4-tuples. Borrowing concepts from population genetics, they call a group of adjacent methylation loci an _allele_ and each possible methylation patterns at the allele an _epiallele_. In my framework, an allele is an m-tuple and an epiallele is a (section) of a haplotype.

\citeauthor{Landan:2012kp} defined the _epipolymorphism_ of an allele as "the probability that two epialleles randomly sampled from the [allele] differ from each other". Mathematically, the epipolymorphism of a 4-tuple is defined as $p_{poly} = 1 - \sum_{l = 1}^{l = 2^{4}} p_{l}$, where $p_{l}$ is the relative frequency of the $l^{th}$ epiallele in the sample. They also defined the average methylation level of a 4-tuple, $\zeta_{(i, i + 1, i + 2, i + 3)}$, as follows:

1. Compute the methylation level of each read overlapping that 4-tuple = $0, 0.25, 0.5, 0.75, 1$.
2. Compute the average methylation content of all reads overlapping that 4-tuple, $\zeta_{(i, i + 1, i + 2, i + 3)}$. $\zeta_{(i, i + 1, i + 2, i + 3)}$ is equivalent to the average $\beta$-value in that m-tuple, $\frac{1}{4} \sum_{i' = i}^{i' = i + 3} \beta_{i'}$, __provided that all reads overlapping any of the methylation loci in that m-tuple overlap all methylation loci in that m-tuple.__

They then plotted the level of epipolymorphism against the average level of methylation for all 4-tuples (with "suffient" coverage) and compared the resulting curves to two theoretical model of methylation variability. These two models were _bimodal epipolymorphism_ and _maximum epipolymorphism_. The bimodal epipolymorphism model assumes that each read, $z$, has $\zeta_{z} = 0$ or $\zeta{z} = 1$, that is, each read is either fully unmethylated or fully methylated. Mathematically, 'bimodal epipolymorphism' can be written as $2m (1 - m)$, where $m$ is the average methylation level of the 4-tuple. The maximum epipolymorphism model was modelled as $1 - (1 - 2(m(1 - m)))^4$ (__TODO: Figure out what the maximum model corresponds to__).

In the ENCODE data they found that differentiated, somatic tissues had a higher frequency of epipolymorphism than did embryonic cells. From this they concluded that germline and pluropotent cells are able to establish or maintain a more coherent methylation state and that somatic cells accumulate substantial stochastic variation during somatic development.

\citeauthor{Landan:2012kp} also found that DMRs identified between the H1 embryonic stem cell line and somatic cell lines had a higher degree of epipolymorphism than non-DMRs[^landan_dmrs]. They found a similar pattern in the bisulfite tumour vs. normal padlock probe bisulfite sequencing data from \cite{Hansen:2011gu}.

[^landan_dmrs]: Sample-specific DMRs were simply defined as regions with $\geq 20\%$ difference in average methylation between the (duplicate-averaged) H1 methylome and the sample of interest. No details are provided on how large the regions had to be to qualify as DMRs.

\citeauthor{Landan:2012kp} then studied the evolution of DNA methylation patterns in an _in vitro_ model system tracking immortalised fibroblasts over 300 generation Two lines, A and B, were tracked and sampled at several time points for profiling with MeDIP-seq. Unfortunately, at least from my perspective, MeDIP-seq does not provide single-base-resolution of DNA methylation and so I do not discuss these result any further.

They did, however, also profile 45 cancer-related CGIs in the the model system at multiple time points with "ultra-deep" ($> 10,000 \times$ coverage) bisulfite-sequencing. By sampling the over time they were able to investigate the evolution of stochastic epipolymorphism. Using this data they also analysed the correlation of DNA methylation at pairs of CpGs in these 45 regions. They visualised these using plots analogous to linkdage-disequilibrium plots. They did not look at how these correlations vary as a function of distance.

From the analysis of this small set of regions they concluded that:

> [the] correlation between the methylation states of pairs of CpGs was generally very low. This lack of correlation suggests that methylation dynamics are typically independent for different CpGs, making the methylation state of one CpG (whether high or low) uninformative on the methylation state of nearby CpGs.

__My co-methylation results and extensive $\beta$-value correlations suggest otherwise.__

The correlation of methylation at a pair of CpGs was computed from the $2 \times 2$ contigency table, shown __BELOW__.

__TODO: Represent as contingency table.__ Let $n_{11}$ be the number of reads that are methylated at both CpGs, $n_{00}$ be the number of reads that are unmethylated at both CpGs and $n_{01}$ and $n_{10}$ be the number of reads that have different methylation states at each CpG ($N = n_{00} + n{01} + n{10} + n_{11}$). The average methylation level of each CpG is $m_{1} = \frac{n_{11} + n_{10}}{N}$ and $m_{2} = \frac{n_{11} + n_{01}}{N}$, respectively. The variance of methylation at each CpG is $v_{1} = m_{1} \times (1 - m_{1})$ and $v_{2} = m_{2} \times (1 - m_{2})$, respectively. The correlation of methylation at the pair of CpGs is then $\frac{n_{11} / N - m_{1} \times m_{2}}{\sqrt(v_{1} + v_{2})}$. __TODO: Is this the same as correlation of $beta$-values using only those reads that overlap both CpGs?__

## \citet{Zhang:2013uu} and \citet{Stevens:2013hv}

`M&M`, published in \citet{Zhang:2013uu}, is a statistical framework for jointly analysing MeDIP-seq and MRE-seq to detect DMRs in a two-group experiment. MeDIP-seq is an enrichment-assay and MRE-seq is a restriction enzyme assay. `M&M` is a window-based algorithm (default = 500 bp non-overlapping windows). `M&M` was compared to `MEDIP`, which only uses MeDIP-seq data, and WGBS of the same samples. Each method was run on biological duplicates to estimate the corresponding false positive rate.

`methylCRF`, published in \citet{Stevens:2013hv}, is a software implementation that extends the `M&M` framework. Specifically, it uses a conditional random field model of counts-per-window whereas `M&M` assumes independence across windows. It gives single-base estimates for DNA methylation, despite the input being data from enrichment-assays. \citeauthor{Stevens:2013hv} provide evidence that these predicts single-base estimates are very close to observed values from WGBS of the same sample.

\citet{Stevens:2013hv} plot $\beta_{i}$ vs. $\beta_{i+1}$ for $IPD \leq 750$ bp to argue that the distribution of methylation is highly non-random (Fig 2b).

## \citet{Hon:2013jra}

### Summary of method for identifying tsDMRs

The authors aim to identify tissue-specific differentially methylated regions (tsDMRs). They study 17 different tissues from mouse, although they exclude the placental methylome from most analyses, which reduces the total number of tissues to 16.

To identify tsDMRs they use a Hidden Markov Model (HMM) on the $\chi^2$ test statistics of different methylation levels across the 16 tissues. Specifically, the $\chi^2$ statistic was calculated as follows:

> For a given genomic interval, let $m_{t}$ be the number of methylated cytosines sequenced and $d_{t}$ be the depth of sequencing for tissue $t$. To capture the deviation of the methylomes from that expected if all tissues were uniformly methylated, we used a $\chi^2$ test statistic. We denoted the uniform methylation level in this interval as $f = \sum m_{t} / \sum d_{t}$. Then, the expected number of methylated cytosines sequenced for each tissue was $e_{t} = f d_{t}$. Thus, the $\chi^2$ test statistic was calculated by $\chi^2 = \sum(m_{t} - e_{t})^{2} / e_{t}$, with the degrees of freedom equal to one less than the total number of tissues.

It is not clear that they correctly adjusted the degrees of freedom when the number of available tissues (i.e. those with sufficient sequencing coverage) was less than 16 for a given region.

> To identify tsDMRs, we calculated $\chi^2$ values for each CpG unit, __defined as three consecutive CpGs__

The genomic intervals used in their analyses are groups of 3 CpGs, which they at times refer to as _CpG units_. I assume that these CpG units do not overlap but this isn't made clear. This $\chi^2$ test statistic differs from that used in my analysis of the _Avy_ methylome data because I test at every CpG whereas they test at every group of 3 CpGs.

Then, a Hidden Markov Model (HMM) was used to identify putative tsDMRs:

> ..., [we] employed an HMM as implemented by [`pmtk3`](https://github.com/probml/pmtk3) to segment the genome, using methods as previously described[^58] with several alterations. Briefly, we trained a four-state HMM, with each state consisting of a mixture of two Gaussians, with the Baun-Welch [sic] algorithm. States were estimated using the forward-backward algorithm, with the highest-valued $\chi^2$ state denoting a prefiltered set of tsDMRs[^wot].

[^58]: Dixon, J.R. et al. Topological domains in mammalian genomes identified by analysis of chromatin interactions. Nature 485, 376–380 (2012).

[^wot]: "with the highest-valued $\chi^2$ state denoting a prefiltered set of tsDMRs"; what does this even mean?

__There is not enough detail to reproduce the HMM used in their analysis__. They don't describe what the four states in the HMM actually are. Nor do they describe the "several alterations" made to the method in Dixon et al. Moreover, based on a quick scan, Dixon et al. only contains a generic description of a 3-state HMM, which isn't very helpful in trying to understand the methods in the present paper. In summary, the HMM isn't described in sufficient detail to reproduce it.

The putative tsDMRs were then filtered according to variation across tissues:

> To select for tsDMRs with a large magnitude of variance in tissue methylation, we determined the intersection point of tissue-specific and non–tissue specific standard deviation distributions (Fig. 2e, right) and removed sites with smaller standard deviation than this intersection point (here $11.88\%$ mCG). The final set of tsDMRs was represented by three genomic loci: the left and right CpG boundaries were called by the HMM, along with the central CpG site having the highest $\chi^2$ value, which represented an estimate of the methylated base showing the greatest tissue specificity.

No particular comments on this step.

They report an FDR for each region:

> To estimate the false discovery rate, we repeated this analysis on ten random permutations of the data set, each of which consisted of random assignment of methylated base calls within each tissue while maintaining baselevel sequencing depth.

They claim that the intial HMM segmentation identifies DMRs at the unbelievable (in the literal sense of the word) FDR value of $3.96 \times 10^{-5}$. My instinct is that the number of permutations is insufficient to accurately estimate the FDR and I take the reported FDRs with a large handful of salt.

Given a putative tsDMR, the tissues driving that variability eredentified as those with the highest specificity relative to the Shannon entropy of the region. This basically amounts to checking which of the tissues had the "most different" methylation levels within putative tsDMRs.

### Other questions

#### Are these tissues from the same mouse? Do they have multiple mice per tissue? How much of the variation is mouse-specific (sequence-driven) rather than tissue-driven? {-}

The Methods says that, "Mouse tissues were isolated from a female C57Bl/6 mouse at least 4 weeks old (Charles River) at 14.5 d of pregnancy". I _think_ this means a single mouse, although the wording is somewhat ambiguous. If multiple mice were used it does not say how many mice were used nor which tissues were collected from which mouse. Later on they say that, "Previously published mouse methylomes (ES and NPC) were downloaded and remapped". So, at least two of the tissues were from different mice.

While they did use methylomes from Cast/129 and 129/Cast hybrids to replicate some of the putative "cortex vestigal enhancers" identified in the C57Bl/6 methylomes, if the majority of the tissues are $n=1$ from a single mouse then there is no sense of the variability of these tsDMRs across mice.

### How could this be adapted to the _Avy_ methylome data?

The basic idea would be to run a HMM on the $\chi^2$ statistics. There are two key things to consider:

1. What goes into the HMM, i.e. what statistic?
2. What sort of HMM?

#### What goes into the HMM? {-}

You could do this on CpG-specific $\chi^2$ statistics (what we have now) or on $\chi^2$ statistics computed from groups of CpGs (e.g. groups of 3 as done in the present paper). There's no reason to believe that groups of 3 is the "correct" number to use if grouping, and I think the amount of grouping would depend on the region's characteristics (e.g. CpG density, spatial correlation of methylation of the region).

They suggest grouping because they have low coverage (~$8\times$) sequencing data, whereas the _Avy_ methylome data has a higher sequencing coverage, so in that sense grouping may be unnecessary for the _Avy_ methylome data.

#### What sort of HMM? {-}

As discussed above, the HMM used in this paper is not described in sufficient detail to replicate it. In particular, I don't know what the four states are. But if I was to try to use a HMM for this problem in the _Avy_ methylome data I would start with two states: "no difference in methylation across the 5 samples" vs. "difference in methylation across the 5 samples".

There still remain several other important choices to parameterise the HMM such as the transition probabilities and emission distributions, which again aren't described in sufficient detail in the present paper.

In summary, implementing a HMM would require a fair bit of exploratory work to identify a good parameterisation.

## \citet{Knijnenburg:2014ke}

\citeauthor{Knijnenburg:2014ke} propose a segmentation algorithm, `MSR` (multiscale signal representation), to represent a vast array of genomic signals that are measured on different scales. Such signals include DNA methylation.

An example of segmentation DNA methylation data, comparing a colon tumor to its matched normal mucosa, uses the log-ratio of tumor to normal DNA methylation levels. While this is a "differential methylation" signal, as the authors' note, `MSR` does not explicitly test for differential methylation as its purpose is genome segmentation:

> The `MSR` method is complementary to other segmentation techniques, such as `Segway` and `ChromHMM`: whereas the `MSR` segments one genomic signal at multiple scales, these HMM-based models use multiple genomic signals to provide one segmentation, thereby dividing the genome into a number of functional states called chromatin domains.

This example, like all others in the paper, is effectively an $n=1$ analysis. __It is not clear how this method would be generalised to include biological variation.__

While the authors propose a method to test for significant signal, it is not clear whether or how this method accounts for the fact that the same dataset is used to discover and test the features. This score, called _SFC_ (significant fold change), seems similar in spirit to the CDIF proposed in \cite{Sun:2014fk}.

The code is written in Matlab and available from [https://github.com/tknijnen/msr/](https://github.com/tknijnen/msr/). It includes a runtime environment that allows users who don't have Matlab installed to run the `MSR` software.

## \citet{Hovestadt:2014fm}

\citeauthor{Hovestadt:2014fm} performed WGBS on 34 human and five murine medulloblastoma tumors, and eight human and three murine normals. They are interested in integrating DNA methylation with other measures, such as RNA expression and histone modifications. Much of the analysis of the WGBS data is "novel", and is based on binning and thresholding. To identify sub-group-specific differential methylation they do the following:

> For all patterns described earlier, the average methylation level per gene and sample was calculated for the entire WGBS cohort. Analysis of variance was used to determine genes differentially methylated between all four medulloblastoma subgroups and combined controls (adjusted P value , 0.001). The Benjamini–Hochberg procedure was used to adjust for multiple testing. Subgroup-specific differentially methylated genes were determined by applying a post-hoc test on genes previously determined as being differentially methylated. (R package: `multcomp`, functions: `mcp` and `glht`). Individual medulloblastoma subgroups were required to be significant against all other subgroups combined, and against control samples separately (P value , 0.001; in essence eight comparisons were made). This testing procedure was deemed appropriate because of large sample sizes and continuous methylation levels resulting of averaging of multiple CpGs.

## \citet{Ushijima:2003gqa}

This paper uses a clever strategy of starting with a single cell, and then growing clonal populations to track the fidelity of DNA methylation. As it is an older paper, they used Sanger bisulfite-sequencing, which means they only studied a small number of regions (albeit at base pair resolution).

> We determined the methylation status of each CpG site on each DNA molecule obtained from clonal populations of normal human mammary epithelial cells. Methylation pattern error rates (MPERs) were calculated based upon the deviation from the methylation patterns that should be obtained if the cells had 100% fidelity in replicating the methylation pattern. Unmethylated CGIs in the promoter regions of five genes showed MPERs of 0.018–0.032 errors/site/21.6 generations, and the fidelity of methylation pattern was calculated as 99.85%–99.92%/site/generation. In contrast, unmethylated CGIs outside the promoter regions showed MPERs more than twice as high (P < 0.01). Methylated regions, including a CGI in the MAGE-A3 promoter and DMR of the H19 gene, showed much lower MPERs than unmethylated CGIs. These showed that errors in methylation pattern were mainly due to de novo methylations in unmethylated regions.

For some of the regions, they identified two major clonal sub-populations, which they interpreted as being the initial methylation pattern of each allele in the original single cell.

They also show evidence that in their experiment the the fidelity errors occur at least an order of magnitude more frequently than failures of the bisulfite conversion process:

> These [bisulfite conversion efficiency] values showed that the MPERs in CGIs in the promoter regions are 10-fold more than the unconversion rates.

In theory, could estimate epiallele frequencies from Figure 3. The analysis methods are very simple proportions of 'CpGs that changed methylation state'.

## \citet{Capra:2014uh}

This is the first paper I've seen that uses phylogenetic approaches to studying DNA methylation dynamics. Specifically, they seek to infer DNA methylation changes from RRBS data of the haematopoetic lineage. The main result of interest is that they try to incorporate the "vertical" correlation of DNA methylation dynamics between precursor and dependent instead of the "horizontal" correlation of neighbouring CpGs within the same sample. The vertical approach beats the (simple) horizontal approach in inferring missing data.

This approach discretisation of $\beta$-values into lowly-, intermediate- and highly-methylated categories.

Putting DNA methylation into a phylogenetic framework gives (in principle) access to a whole raft of existing analysis methodologies.

## \citet{Lee:2014cv}

Figure 3 describes several scenarios that would lead to DNA methylation heterogeneity.

## \citet{Gokhman:2014cp}

This paper compares DNA methylomes from a Denisovan sample and a Neandertal to methylomes from present day humans. They exploit the natural deamination of cytosines (mC -> thymine and C -> uracil) to infer the methylation state of cytosines from the C->T conversion ratios. __IMPORTANT:__ These C->T ratios are __not__ the same as $\beta$-values, but they can be scaled by a number, $\frac{1}{p_{m}}$, so that they have a similar interpretation.

They smooth these C->T ratios using a sliding window, where the window size is chosen as that which maximises the correlation with methylomes from present day humans (43-89 CpGs, depending on Denisovan or Neandartal and autosome or X-chromosome).

The authors advocate smoothing of DNA methylation:

> Such noise reduction is advisable in methylation samples, as smoothing over neighboring CpGs increases accuracy for low-coverage positions and complies with the high correlation in methylation between adjacent positions and the mostly regional function of DNA methylation

and cite the `BSmooth` paper in support of this position.

They use a rather weird way of removing potential PCR duplicates:

> To remove PCR duplicates that are characterized by abnormally high coverage, we fitted the (strand-specific) coverage distribution to a mixture of (two) Gaussians. Positions with a coverage that was higher than three standard deviations from the mean of the main distribution were removed from the analyses. In the Neandertal, the threshold was set to 46 reads and in the Denisovan the threshold was set to 30 reads.

I don't really understand why they used this and not the standard Picard `MarkDuplicates`-type approach; perhaps it's because the sequencing protocol removes uracils, which would lead to variation in read lengths?

### Identification of DMRs

There's a convoluted procedure to estimate what amounts to the average $\beta$-value of each window, and its associated standard deviation, which are used to form a z-stat. P-values are FDR-adjusted and only windows with FDR $< 0.05$ are retained. Furthermore, only windows with a greater than $60\%$ change in methylation are retained for further analysis.

## \citet{Rijlaarsdam:2014hp}

`DMRforPairs` tries to find DMRs in 1 vs. 1 comparisons - an ambitious aim. I understand the desire for such comparisons but I wouldn't believe the results of most $n = 1$ studies.

## \citep{Chen:2014jb}

This paper proposes a non-parametric test (_Cuzik_ test) of differences in DNA methylation __between multiple groups__, that also accounts for the well-known association of age and DNA methylation. The model requires that the age variables is discretised into bins. There is no discussion of how to choose bins (they use 5-year bins in the real data analysis). Furthermore, my intuition is that leaving age as a continuous variable is preferable.

The Cuzik test assumes there is a trend of differential methylation with age, either increasing ($W_{1}$) or decreasing ($W_{2}$). As the trend is unknown, the authors propose using $W = max(W_{1}, W_{2})$ and then make use of an asymptotic bound to obtain approximate P-values. The description of the Cuzik test sounds rather different [on Wikipedia](http://en.wikipedia.org/wiki/Cuzick%E2%80%93Edwards_test) - I'm not sure what to make of that.

The authors make the reasonable observation that if there is a trend (even if it's not truly linear) then there test, which is designed to detect such a trend, has more power than do tests that aren't designed to detect a trend.

### Simulation study

$\beta$-values are simulated from marginal distributions (Uniform, truncated Normal or Beta), which do not include the spatial correlation of $\beta$-values. They specifically simulate scenarios where there is a trend-with-age of the $\beta$-values, thus ensuring there test looks good, and do not consider a setting where there is no trend.

## \citep{Park:2014ho}

This paper describes the `methylSig` software. `methylSig` uses a Beta-Binomial model to test for DMCs across an arbitrary number of groups (although the proposed test is a likelihood ratio test of group $k$ vs. group $k'$). The dispersion parameter in the Beta-Binomial model, $\theta_{i}$ (assumed constant across groups), is treated as a nuisance parameter that is first estimated and then plugged into the likelihoods to estimate (using maximum likelihood) the parameters of interest, $\mu_{i, k}$ and $\mu_{i, k'}$. For small sample sizes, they use the $t$-distribution, rather than the 'natural' $\chi^{2}_{1}$-distribution of a likelihood ratio statistic. The dispersion parameters are not shrunk, e.g. via Empirical Bayes.

They also propose incorporating local information, via kernel smoothing, to estimate the parameters $\mu_{i}$ and $\theta_{i}$.

The examples all use ERRBS (enhanced RRBS) data and so it isn't clear whether `methylSig` scales to WGBS data.

To increase power, the authors suggest "tiling" observations. Tiling amounts to pooling data, i.e. summing all measurements within a tile (25 bp for ERRBS data, larger for WGBS)) and then using the tiles as the observations rather than the CpGs.

The plots of the results are very busy and I don't like them much.

### Identifying differentially methylated transcription factors

Rather than looking for general DMRs, the authors focus on identifying differentially methylated transcription factors. There are two approaches:

1. Test which TFs contain a significant number of DMCs.
2. Test which TFs are hypo- or hyper-methylated (combining all CpGs in the TF).

### Simulation study

The simulation study is based on ERRBS data from a study of acute myeloid leukemia. It uses the coverage distribution, the locations of CpGs and the estimated dispersions from the real data. However, the $\beta$-values are simulated from Beta-Binomial distributions with group-specific means to induce DMCs. This model does not incorporate correlation amongst the $\beta$-values, although the simulation models allows for DMCs to cluster into DMRs, which is similar to inducing correlations amongst the $\beta$-values, albeit restricted to CpGs in DMRs.

The authors advocate for using `methylSig` with dispersions estimated locally via kernel smoothing. They note that `BSmooth` has low power to detect "independent DMCs", but this is not surprising because `BSmooth` is designed to detect DMRs.

## \citet{Smallwood:2014kn}

This is the first paper that I know of with single-cell bisulfite sequencing data (there is also a scRRBS protocol). The technique, _scBS-seq_, is a variation on PBAT-seq. The single-cell DNA underwent five cycles of random priming, extension and purifying, which means there were up to five 'copies' of each allele.

It is __not a genome-wide technique__, since it only gives "accurate" results for $48.4\%$ of CpGs.

They profile ovulated metaphase II oocytes (MII) because:

> MIIs are an excellent model for technical assessment as they: (i) can be individually handpicked to ensure that only one cell is processed; (ii) represent a highly homogeneous population, which allows discrimination between technical and biological variability; and (iii) present a distinct DNA methylome comprising large-scale hypermethylated and hypomethylated domains.

They contrast the MIIs with embryonic stem cells (ESCs):

> ESCs grown in serum conditions exist in a state of dynamic equilibrium characterized by transcriptional heterogeneity and stochastic switching of transcriptional states9–12, and emerging evidence from immunofluorescence and locus-specific studies suggests that 5mC heterogeneity exists in ESCs13.

A fascinating fact from the supplementary material, which I don't understand:

> The overall higher mapping efficiency of oocytes versus ESCs can be explained by the amount of DNA in each cells (4n for MII oocytes and 2n for ESCs), resulting in a relatively lower contribution of spurious sequences in MIIs.

They sequenced multiple single cells (51 across from four different 'populations') at very shallow depth ($\approx 20 \times 10^{6}$ 100bp PE reads/sample). Furthermore, mapping efficiences were very low ($\approx 3.9 \times 10^{6}$ reads/sample or $20\%$), which is largely due to low-complexity sequences (specifically, poly-Ts that are inherant to the protocol). They also sequenced "bull cell mass" samples, i.e. pools of the same cells. They could largely recapitulate the $\beta$-values of the pools using 12 single-cells.

## \citet{Robinson:2014tx}

Mark has written a review of methods to detect DMRs. It is fairly high-level (as it should be) and could basically act as my description of calling DMCs and DMRs! For example:

> It is therefore no surprise that BB assumptions are made in several recently proposed packages, such as BiSeq [30], MOABS [29], DSS [28] and RADMeth [33]. Similarly, empirical Bayes (EB) methods fit naturally for modeling and inference across many types of genomic data and DNA methylation assays are no different. MOABS and DSS both implement hierarchical models and use the whole dataset to estimate the hyperparameters of the beta distribution; RADMeth and BiSeq use standard maximum likelihood without any moderation. While BiSeq and RADMeth do not moderate parameter estimates, they provide facilities for complex designs through design matrices, which MOABS and DSS do not currently offer.

Mark notes that neither Fisher's exact test, nor logistic regression, control for within-group variability:

> While this strategy (Fisher's exact test) may be sufficient in comparing cell lines, we stress that the use of FET should be avoided in the general case; most systems have inherent biological variation and FET does not account for it.

> Likewise, using the binomial distribution, such as a logistic regression framework (e.g., methylKit; 32), also does not facilitate estimation of biological variability, unless an overdispersion term is used.

This quote fits with my intuition:

> Ultimately, our intuition suggests that moderated t/F-statistics on the normalized log-ratios of intensities seems most rigorous. -- Highlighted 25/07/2014

Mark emphasies the fact that there is a __big__ difference between methods that work on pre-defined regions and methods that seek to identify such regions:

> ... one must distinguish between methods that operate on predefined regions, with those that define regions of DM. The latter is considerably more difficult because ensuring control of the false discovery rate (FDR) at the region-level is non-trivial; in particular, controlling false discoveries at the site-level does not give a direct way to controlling false discoveries at the region-level when the region itself is also to be defined.

## \citet{Liu:2012ge}

I believe `Bis-SNP` is the best methylation calling software due to it's comprehensiveness. Unfortunately, it can only call methylation at 1-tuples and I need to be able to call methylation at m-tuples. `Bis-SNP` is built on `GATK (v1)` and so includes base-quality recalibration and indel realignment. Importantly, `Bis-SNP` works with any `BAM` file of bisulfite-sequencing data, which means that the user can choose his/her favourite alignment software.

As the authors state:

> Bis-SNP is a practical tool that can both (1) improve DNA methylation calling accuracy by detecting SNPs at cytosines and adjacent positions, and (2) identify heterozygous SNPs that can be used to investigate mono-allelic DNA methylation and polymorphisms in cis-regulatory sequences.

Figure 1 explains how it performs SNP calling from bisulfite-sequencing data.

Calling "C-strand" SNPs depends on the underlying methylation state ($\beta$, which defaults to the standard $\beta$-values) and bisulfite-conversion errors (underconversion = $\alpha$, which defaults to $0.25$) and overconversion = $\gamma$, which defaults to $0$).

Rather than filter out positions from an M-bias plot, `Bis-SNP` "walks" from the 5' end of each read and excludes all positions prior to the first $T$ that is mapped to a reference $C$. This is to remove the so-called "5' bias" of the Illumina bisulfite-sequencing but will do nothing for 3' bias.

## \citet{Dolzhenko:2014bo}

> Existing methods based on Fisher’s Exact Test and HMMs are appropriate for comparing a pair of samples at a time (coming either directly from the experiment or obtained by pooling other samples); however, they lack the ability to account for variability of methylation levels between replicates.

> Unlike BSmooth, [BiSeq] can be used for experiments that go beyond comparing two groups of samples, but it requires a set of candidate regions that may exhibit differential methylation.

> The beta- binomial, which has first been used for modeling WGBS proportions by Molaro and others [17], is a natural choice for describing methylation levels of an individual site across replicates as it can account for both sampling and epigenetic variability.

> The beta-binomial regression is fit separately for each target site...[parameters] are estimated using the method of maximum likelihood

> [To find DMRs] the p-values are transformed using weighted Z test (also known as Stouffer-Liptak test), employing an approach proposed by Kechris and others [23].

Correlations of z-scores are computed using the method implemented in `comb-P`.

> We recommend calculating correlation and subsequently combining the pvalues of sites located within 200 bp. In our experience correlation typically becomes much weaker beyond this point, so the users do not generally need to alter this parameter. However, it may be appropriate to increase the value of this parameter when analyzing very noisy data.

Sounds fair.

> As explained by Rice [33], Fisher’s method is best suited for testing the existence of at least one significant test among the ones being combined, while the Z test is more appropriate in situations requiring the consensus among all of the combined tests, suggesting that the Z test is more appropriate for our purposes.

### Some discussion of Rice

> In a wide variety of biological applications there is a need to combine the results from independent tests for which the raw data cannot be pooled.

True, but the tests in BS-seq are __not independent__. So, what are the consequences of this?

> Fisher's testing procedure represents a test against broad alternatives. It specifically tests whether at least one component test is significant, and can yield a significant combined test statistic when the component tests, on balance, strongly support $H_0$. This is an undesirable characteristic when asking whether a group of tests collectively supports the same $H_0$.

### Simulation

6 vs. 6 two-group experiment.

1. Sample coverage from real data
2. Sample number of methylated reads from a binomial distribution
3. Simulate methylation levels from distributions in [Epigenome-wide association studies for common human diseases](http://www.nature.com/nrg/journal/v12/n8/full/nrg3000.html). These distributions are hypothetical.

__No correlations of methylation levels in simulation.__

## \citet{Xie:2014ez}

A Bayesian test to replace the "Lister method" of identifying methylcytosines. I still don't think identifying methylcytosines is particularly useful.

## \cite{Zheng:2014ii}

`MethylPurify`

> It relies on the following characteristics of the DNA methylome data: (1) all CpG cytosines within a short genomic interval (approximately 300 bp) in a pure cell population share similar methylation levels which are either mostly methylated or mostly unmethylated [36]; (2) the number of bisulfite reads mapped to each genomic interval to tumor and normal cells are in accordance with their relative compositions in the mixture, subject to standard sampling noise.

Then uses a mixture model to estimate tumour and normal components.

### Simulation

They use the [`BSsim`](http://122.228.158.106/BSSim) software for simulation.

## \citet{Huh:2014ki}

`Bis-class` is a Bayesian method for clasifying sites as methylated or unmethylated. The authors claim it is most useful for lowly methylated genomes with low coverage, which is where simpler methods struggle.

The classifier incorporates information about the global and local methylation levels in the sample via a prior distribution. The "localness" is incorporated via a kernel with the window size set to be the distance where the spatial correlation between methylation levels drops below 0.2. __TODO: Do they describe how they compute these correlations__

## \citet{Chodavarapu:2010iqa}

> We identified 10-base periodicities in the DNA methylation status of nucleosome-bound DNA (Arabidopsis) and found that nucleosomal DNA was more highly methylated than flanking DNA. These results indicate that nucleosome positioning influences DNA methylation patterning throughout the genome and that DNA methyltransferases preferentially target nucleosome-bound DNA. We also observed similar trends in human nucleosomal DNA, indicating that the relationships between nucleosomes and DNA methyltransferases are conserved.

> Similar to findings in animal and fungal high-throughput nucleosome sequencing studies5–8, we found 10-base periodicities in WW (W 5 A or T) dinucleotides, and SS (S 5 G or C) dinucleotides that were 5 bases out of phase with the WW dinucleotides (Fig. 1d and Supplementary Figs 3–5).

> Notably, all three types of methylation showed a 10-base periodicity on nucleosomal DNA, which was in phase with the WW dinucleotides and out of phase with the SS dinucleotides (Figs 1d, 2a–c and Supplementary Fig. 6). These methylation preferences were not correlated with preferences for CG, CHG or CHH sequences at these locations (Supplementary Fig. 7). Because DNA methyltransferases access the major groove, this methylation would be on DNA that is on the outside of the nucleosome (minor groove facing the histones) and thus more accessible to the DNA methyltransferases. This in turn indicates that DNA might be in part methylated as the DNA is still bound to nucleosomes, leading to the observed 10-base pair periodicity.

> We previously reported a 10-nucleotide periodicity in CHH methylation data when performing autocorrelation analysis on the methylation pattern of the whole genome3. Our previous interpretation was that the structure of the DRM2 enzyme might be responsible for this pattern, because the orthologous Dnmt3 enzymes in mammals are known to act as heteromeric complexes in which two methyltransferase active sites have a spacing equivalent to roughly 10 nucleotides of DNA13. However, the current data in which we see 10-base pair periodicities for all types of methylation indicate a more general explanation: that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation for all DNA methyltransferases. Nucleosomal preferences could also partially explain the sequence preferences that we observed previously for CHG and CHH methylation3. Highly methylated cytosines tended to be followed immediately by A/T but not C, consistent with our finding that DNA methylation is out of phase with CC dinucleotides.

> This suggests the possibility that DNA methylation, which frequently exists in the transcribed regions of active genes26,28, could have a conserved role in exon definition or splicing regulation.

## \cite{Law:2010jr}

> the symmetric CG and CHG contexts (in which H = A, T or C) and the asymmetric CHH context

__This is not correct, CHG is not symmetric__. For example, a CCG on the + strand is a CGG on the - strand, i.e. a CG motif not a CHG motif.

```sh
+   CAG		CCG		CTG
-   GCT		GGC		GAC
```

> a model in which the two DNMT3A active sites are separated by approximately one helical DNA turn, which suggests that each tetrameric complex could simultaneously methylate two cytosine residues at a defined spacing of 8–10 bp

> In vivo, the spacing of CG dinucleotides at many DMRs is also consistent with an ~8–10 bp periodicity51, as is the finding that CG dinucleotides at an 8 bp spacing are overrepresented across the human genome68,69 and, to a lesser extent, across the mouse genome68. As DNMT3A seems to be a non-processive DNA methyltransferase70, the formation of an oligomer could help to explain the observed periodic pattern of DNA methylation.

## \cite{Jurkowska:2011dt}

> Active genes usually show hypomethylation around the transcriptional start site (TSS) and high levels of methylation in the gene body, which increase with gene expression.[30, 31, 46] The methylation of the gene body is believed to block aberrant transcription initiation inside the gene and therefore to help in avoiding the production of truncated mRNAs and proteins.

> Dnmt3L hence has a dual, gender-specific role: it is essentialfor the establishment of genomic imprints in oocytes and forthe silencing of the dispersed repeated sequences in malegerm cells.

> In the crystal structure, the active sites of the two central Dnmt3a subunits are separated by one DNA helical turn, which corresponds to around 10 bps, suggesting that two CG sites on __opposite strands__ separated by 10 bps could be methylated by Dnmt3a in one binding event. Additionally, Dnmt3a polymerises on DNA in vitro, forming Dnmt3a-DNA filaments, which positions the adjacent methyltransferase molecules that interact with the same DNA strand at a distance of 8–10 bps,[169] suggesting that these sites could become comethylated by Dnmt3a (Figure 3 B). Indeed, in vitro methylation experiments demonstrated that there is a correlation of methylation between two sites localised ~ 10 bps apart both in the same strands and in opposite strands.

> Additionally, the presence of 5–10 % of hemimethylated sites in the genome[210, 211] cannot be explained by the simple copy model alone, in which every hemimethylated CG site is faithfully and quickly remethylated by Dnmt1 after each round of replication. These considerations led to the development of stochastic models for DNA methylation,[212,213] based on differential equations taking into account the probabilities for methylation of a hemimethylated site (fhm) and methylation of an unmethylated site (fum).

> In summary, it appears that the “maintenance DNA methylation” refers to the preservation of average levels of DNA methylation at certain regions, but not to an accurate copying of site-specific DNA methylation patterns.

## \cite{Lacey:2009by}

> Excessively long context ranges might render the model ineffective, while choosing multiple breakpoints along a sequence would assume a piecewise model that is not likely to be biologically realistic.

I should address the "piecewise model ... unrealistic" issue in my discussion of `methsim`.

## \cite{Morgan:1999ds}

People already knew that differences in agouti expression resulted in phenotypic differences. What they didn't know was what causes the differences in expression. This is what Emma identified:

> In viable yellow (Avy/a) mice, transcription originating in an intra-cisternal A particle (IAP) retrotransposon inserted upstream of the agouti gene (A) causes ectopic expression of agouti protein, resulting in yellow fur, obesity, diabetes and increased susceptibility to tumours (Neomorphic agouti mutations in obese yellow mice, 1994).

## `methclone` \citep{Li:2014ei}

- Landan = changes at a global scale, methclone = changes between or within samples (at a locus?)
- Re Landan's epipolymorphism: "the change in epipolymorphism of a sample or locus does not reveal the underlying clonality."
- "Methclone calculates the combinatorial entropy (ΔS) change of epialleles at one locus and out- puts the loci with a ranked list of epiallele changes de- fined by the entropy change (Figure 1), from no change (0) to maximum difference in entropy (-144)."
- Can `methclone` output the counts of methylation at epialleles?
- __TODO: Try installing on my own machine__
- Used 4-tuples in the paper and applied to eRRBS data. Required > 60 reads per-locus, per-sample.
- Don't really like the colour scheme. There's no obvious relationship between colour and methylation pattern.
- The problem of looking for epiallele shifts is that of looking for chances in the probability vector of a multinomial distribution.
- In the definition of $\delta S$, is there any sense of a $0000$ pattern being more divergent than a $1110$ pattern from a $1111$ pattern?
- The $\delta S$ cutoffs seem arbitrarily chosen?
- "Interestingly, this indi- cates that the number of different epialleles (eloci) between unrelated individuals are rather stable, whereas the number of eloci are much more dynamic within the same person during a tumor’s treatment and progression."
- Investigated various sized m-tuples, $m = 1, \ldots, 10$. 4-tuples were the best in the sense that there was sufficient data to estimate these.
- Uses "combinatorial entropy" (defined in cited papers).
- Can be used to compare the shift at two time points. Does not allow for replicates.
- The number of epialleles is normalised by the library size, i.e., not the sequencing coverage at that loci, and further normalised by a constant (200) to "control the entropy dynamic range".
- Entropy is minimal if there is only one pattern
- Entropy is maximal if there are all $2^{m}$ possible patterns and these all occur with equal frequency.

# Review of simulation methods

### \cite{Feng:2014iq}

1. Simulate group-specific $B_{i, j}$ from a Beta distribution with hyperparameters given by:
  - Group-specific means sampled from real data.
  - Group-specific dispersion parameters sampled from real data.
2. Sample sample-specific $u_{i, j} + m_{i, j}$ from same RRBS experiment.
3. Simulate sample-specfic $\beta_{i, j}$ from Beta-Binomial distribution that incorporates (1) and (2).

### \cite{Lacey:2013iy}

Simulate RRBS and not WGBS. All parameters estimated from a single sample.

0. Simulate CpGs from a point-process (HMM).
1. Simulate group-specific $B_{i, j}$.
  - (a) Simulate independent Beta-distributed RVs
  - (b) Induce correlations via variogram and re-estimate Beta-parameters.
  - Iterate (a) and (b) until stable.
2. Simulate sequencing-coverage as Gamma-mixture across bins. Includes across sample correlations of sequencing depth.
3. Simulate sample-specfic $\beta_{i, j}$ from Beta-Binomial distribution that incorporates (1) and (2).

### \cite{Sofer:2013bk}

Simulate microarray data by sampling from 539 TCGA 450k arrays.

A. Sample blocks of probes to preserve spatial correlation.
B. Induce differential methylation by weighted sampling of blocks containing "targets" and hopefully across other CpGs in the block since they are generally correlated.

### \cite{Xu:2013eg}

1. Simulate group-specific $B_{i, j_{k}}$
  - Beta
  - Truncated Normal
  - Truncated Normal-mixture
2. Simulate sample-specific $u_{i, j} + m_{i, j}$ from a round Normal distribution.
3. Simulate sample-specific $\beta_{i, j}$ from Binomial distribution that incorporates (1) and (2).

### \cite{Chen:2014jb}

No simulation of true values or sequencing coverage, but directly simulate $\beta$-values.

1. Simulate group-specific $B_{i, j_{k}}$
  - Uniform
  - Truncated Normal
  - Beta

Differential methylation is from a trend-with-category model.

### \cite{Chen:2014jb}

Simulate ERRBS and not WGBS.

1. Simulate group-specific $B_{i, j_{k}}$ from Beta distribution
2. Sample sample-specific $u_{i, j} + m_{i, j}$ from ERRBS experiment.
3. Simulate sample-specific $\beta_{i, j}$ from Binomial distribution that incorporates (1) and (2).

### \cite{Dolzhenko:2014bo}

1. Simulate group-specific $B_{i, j_{k}}$ from Beta distribution
2. Sample sample-specific $u_{i, j} + m_{i, j}$ from real data.
3. Simulate sample-specific $\beta_{i, j}$ from Binomial distribution that incorporates (1) and (2).

### \cite{Chen:2013eh}

No simulation of true values or sequencing coverage, but directly simulate $\beta$-values.

1. Simulate group-specific $B_{i, j_{k}}$
  - Beta
  - Normal distribution
  - Normal-mixture

### 2015-03-18 Rmarkdown/paper_review.Rmd
Reviewed WGBSsuite

@@
 
+## \cite{Rackham:2015bv}
+
+This was the "oh, crap" paper, published when I was writing up the last chapter of my thesis (which is on simulating WGBS data and my `methsim` software).
+
+`WGBSsuite`:
+
+- Simulates single-base resolution WGBS data
+- "Several" parameters may be derived from real data
+- Simulation uses two __dependent__ HMMs (dependent only through the methylation state, i.e., 'methylated', 'de-methylated' and 'transition' states).
+
+The authors highlight that previous simulation methods have been too simple or designed from RRBS data \citep{Lacey:2013iy}.
+
+An important difference between `WGBSsuite` and `methsim` is that `methsim` simulates individual reads whereas `WGBSsuite` does not and can not.
+Instead, `WGBSsuite` simulates the average methylation level at each CpG, which should be faster and simpler to do.
+
+Finally, for better or worse, I'd argue `methsim` is more data-driven since almost all parameters are sampled from real data.
+
+### Algorithm
+
+1.
+Simulate CpG positions using a homogeneous first-order HMM with three states (CGI, shores and deserts).
+According to the transition matrix, it is impossible to go from a CGI to a desert without going via shore (probably reasonable).
+2.
+Simulate the methylation state of each CpG using an inhomogeneous first-order HMM with four states ('methylated', 'transition1', 'transition2' and 'de-methylated').
+The inhomogeneity is to induce some spatial dependence amongst nearby CpGs such that the transition probabilities are modulated by the distance between CpGs.
+According to the transition matrix, it is impossible to go from methylated to de-methylated without going via transition2.
+Similarly, it is impossible to go from de-methylated to methylated without going via transition1.
+It is also impossible to go between the two transition states without going via one of methylated or de-methylated.
+The modulation of transition probabilities as a function of distance is via an exponential decay.
+The rate of decay depends only on the distance between CpGs and not their context (e.g., identical between CGI and desert for a given distance).
+3. Simulate read depth from a Poisson distribution.
+4.
+Simulate read counts from a binomial or (truncated) negative binomial distribution.
+That is, the binomial (or truncated negative binomial) distribution is the emission distribution of the HMM in (2).
+The probability of success is identical within all 'methylated' states, 'de-methylated' states, and within 'transition1' and 'transition2' states, except for some random Gaussian noise (i.i.d.
+for each CpG).
+The average methylation level in 'de-methylated' regions is defined as $p_{de-methylated} = median(\{ \beta_{i}: 0 \leq \beta_{i} < 0.5 \})$; for 'methylated' regions it is defined as $p_{methylated} = median( \{ \beta_{i} : 0.5 \leq \beta_{i} \geq \leq 1 \}); and for 'transition1' and 'transition2' regions it is defined as $p_{transition} = \frac{p_{methylated} + p_{de-methyalted}}{2}.
+$\beta_{i}$ are computed from real data.
+5. DMRs are induced by defining the expected difference in average methylation between cases and controls for randomly sampled regions.
+
+- I don't really understand the need or desire for (1); just use the locations of CpGs (and consequently CGIs, shores, etc.) from the genome you are interested in.
+- There's no discussion of running time or computational requirements in the main paper.
+- All HMMs are first-order.
+
+### Miscellaneous comments
+
+> DNA sequencing following bisulphite treatment results in sets of reads that either map to a methylated or un-methylated versions of the genome.
+
+This isn't how bisulfite-aware mappers typically work. If they did, it would only be possible to map fully methylated or fully unmethylated reads.
+
+> We acknowledge that this simulator does not deal with sample mixtures, an experimental setup that is common when methylation profiles are generated from complex heterogenous tissues (i.e., where the underlying signal is likely to originate from multiple cell-types).
+
+This is hard. I think I get some of the way there in `methsim` by having multiple haplotypes of varying frequencies and having more than three states in the underlying average methylation level.
+
+> the distance between $CpG_{t}$ and $CpG_{t+1}$ is drawn from an exponential emission distribution with a fixed rate $r$ (rate = 1/mean) different for each state ($r_{island}$, $r_{desert}$ and $r_{shore}$).
+In the CpG island state this rate favours short distances (default = $0.025$), the desert state long distances (default = $10^{−4}$) and the CpG shore state medium length distances (default = $0.05$).
+
+It would be nice to have some justification for these defaults, e.g., a comparison with distances in real data.
+For example, this means that the average distance between CpGs in a CGI is 40 bp, which on first impression seems high to me but may be entirely appropriate.
+
+> where $\pi_{s}(t)$ = as $exp{−b \log(min(1, l_{t} − l_{t−1})}, s = 1,...,4$.
+$l_{t}$ is the location of the $t^{th}$ CpG site, $a_{s} \in [0,1]$ is the baseline probability of transition whereas the coefficient $b$ (default $= 1.895e−02$) controls how much $\pi_{s}(t)$ depends on the distance between two consecutive CpGs.
+
+Again, it would be nice to see some justification for this default value.
+
+> [Stuff on "phase difference"]
+
+On first reading I don't really understand this. I think it's to induce between-group variability, i.e., DMRs.
+
 # Review of simulation methods

### 2015-03-19 Rmarkdown/TODO.Rmd
Updated TODOs

@@
 
-- [ ] Find a `wordcount` function for LaTeX or markdown files.
+- [x] Find a `wordcount` function for LaTeX or markdown files.
+  - Use `texcount` ([http://stackoverflow.com/questions/2974954/correct-word-count-of-a-latex-document](http://stackoverflow.com/questions/2974954/correct-word-count-of-a-latex-document)).
 

### 2015-03-19 Rmarkdown/methsim.Rmd
Complete re-write of the Makefile, which required modifications of all .Rmd files and the main .tex file.
Editing and new stuff in methsim chapter.

@@
 
-## Chapter overview
+\chapter{A simulation model of DNA methylation data}\label{chap:methsim}
 
-## Introduction
+\section{Chapter overview}\label{sec:methsim_overview}
 
-Simulation studies play a vital role in the development, validation and benchmarking of methods in applied statistics.
-Simulation is necessary when there is a lack of real datasets with well-known outcomes that we would otherwise use to assess a method's performance.
-While simulated data can never fully capture the richness of "real" data, we can learn a lot about a method by studying how it performs when applied to simulated data.
+In this chapter I argue for the utility, even necessity, of methodology and software for simulating DNA methylation data.
+I present the `methsim` software to simulate DNA methylation data.
+`methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software.
+I describe the statistical model behind `methsim` and its software implementation.
+Finally, I provide examples of how `methsim` can be used to learn about DNA methylation and how it might be used to develop simulation studies to compare competing analysis methods.
 
-While we may sometimes be able to design an experiment with known outcomes via the use of controls, both positive and negative, this is not always possible in genomics.
-One reason for this is the cost of performing a well-designed control experiment.
-Why spend your time and money on an experiment where you "know the answer" when you could be spending that same time and money on investigating something new?
-Another reason is that in genomics we are often working with cutting-edge technology that is so new that we lack the technological and biological knowledge to even design such a control experiment.
+1. Introduction
+  - Why simulation?
+2. Literature review
+  - Simulating reads vs. $\beta$-values
+  -
+3. Methods
+  - Implementation
+4. Case study
+5. Conclusions
 
-The key advantage of using simulated data over real data is that we know the truth _a priori_.
+\section{Introduction}\label{sec:methsim_introduction}
+
+- [x] DONE
+
+It can be difficult to design an experiment that can be used for validating or benchmarking different analysis strategies.
+In fact, the huge variety of experimental factors, and their possible values, can make such a task infeasible.
+But perhaps the bigger hurdle is that there are not attractive experiments to perform; why spend your time and money on an experiment where you "know the answer" when you could be spending that same time and money on investigating something new?
+In these scenarios and more, simulation studies play a vital role in applied statistics, where they can be used in the development, validation and benchmarking of different analysis methodologies.
+
+The key advantage of simulated data is that we know the truth _a priori_.
 Moreover, we can manipulate the truth via parameters in the simulation model and can examine how a method performs under a variety of scenarios.
@@
 We can also identify which scenarios are "easy", ones where most methods are able to identify the truth, and scenarios under which certain methods perform better than others.
+While simulated data can never fully capture the richness of "real" data, we can learn a lot about a method by studying how it performs when applied to simulated data.
 
-Simulation studies can also tell us about the plausibility of hypothesised stochastic or mechanistic models of a phenomenon.
-By simulating data from the proposed model and comparing it to the real data we can identify shortcomings in the model that we can later refine.
+Simulation methods may also be used to learn about the plausibility of hypothesised models of a phenomenon, be they mechanistic or stochastic.
+By simulating data from the proposed model, and comparing it to the real data, we can identify hypotheses that are incompatible with reality.
+This can also help identify shortcomings in the model so that it may be refined in an iterative manner.
 
-In the field of genomics, almost all papers proposing a new method will include a simulation study and perhaps a comparison to existing methods, if these exist.
 There are a few key criteria when designing a simulation method:
@@
 
-The above discussion refers to the general use of simulation models, particularly in applied statistics.
-I now turn my attention to the need for simulation models in the study of DNA methylation.
-Whole-genome bisulfite-sequencing remains an expensive assay, costing a few thousand dollars per sample (human) at the time of writing.
-Understandably, studies using WGBS have been from well-funded research groups focusing on "sexy" biology rather than on efforts to generate extensively-validated control datasets.
+There is a lack of gold-standard datasets in the field of DNA methylation for the benchmarking and validation of analysis methods.
+Therefore, the development of realistic, cheap and user-friendly software to simulate DNA methylation data will be of benefit.
 
-The statistical methodology to analyse such experiments lags somewhat behind the ability to generate the data, as is common in genomics.
-Nonetheless, more statistical methods are being published seeking to address common research questions such as identifying differential methylation.
-However, because of the lack of "gold-standard" datasets, it is not clear which of these perform well for common experimental designs and research questions.
-Many of these published methods feature a simulation study in the accompanying paper, but, as we will shortly see, these simulations are woefully inadequate, unrealistic and vary between papers.
-In order to perform a proper benchmarking of these methods we require a realistic simulation model, something that I have attempted to address with my software, `methsim`, which I describe in this chapter.
+\section{Literature review}\label{sec:methsim_lit_review}
 
-## Background + literature review
+- [x] DONE
 
-Simulated bisulfite-sequencing data may be used in the comparison of strategies for the alignment and processing of bisulfite-converted reads, or in the comparison of downstream analysis methods.
-Simulation methods for bisulfite-sequencing data may choose to simulate the individual sequencing reads, unaligned or aligned, or some summary measure of methylation, such as $\beta$-values.
-The former is required for the comparison of different alignment strategies whereas the latter is typically used by developers of downstream analysis methods, such as differential methylation callers, since this is often all that is required to run these methods.
-Moreover, simulated reads must be aligned and processed before they can be used in comparisons of downstream analysis methods, which is often not of direct relevance in comparisons of these tools.
-
-A simulation may be model-based or based on sampling of real data, roughly corresponding to parametric and non-parametric simulation models, respectively.
+The simulation methods described here can be thought to lie on a spectrum with model-based methods at one end and re-sampling based methods at the other end[^model-based].
 It is often simpler to simulate from a model-based simulation, particularly if the model is a well-studied parametric distribution.
 However, this simplicity often comes at the cost of increased assumptions, whose validity may be questionable.
-Of course, the parameters in any model-based simulation should be estimated from real data, although this may not use formal estimation procedures such as maximum likelihood.
-
-In contrast, simulations based on sampling of real data may reduce the number of assumptions required.
+Simulations based on sampling of real data may reduce the number of assumptions required.
 However, care must be taken in selecting the units to be sampled so that the sampling process is efficient and so that the sampled data don't grossly distort within- and between-sample dependencies.
 
-### Methods for simulating bisulfite-sequencing reads
+Any procedure for simulating DNA methylation data should obviously be tailored to its purpose.
+For example, if the study is comparing alignment software for bisulfite-sequencing data then the simulation software should produce `FASTQ` files with realistic base quality score profiles and sequencing errors.
+On the other hand, it may be sufficient to simulate some aggregate data, such as $\beta$-values, when the simulated data are to be used for comparing methods to identify differential methylation.
 
-All of the available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies and are model-based.
+[^model-based]: Of course, the parameters in any model-based simulation should be based on real data, although this may not use a formal estimation procedure such as maximum likelihood.
+
+\subsection{Methods for simulating bisulfite-sequencing reads}
+
+- [x] DONE
+
+All of the currently available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies and are model-based.
 These are not suitable for comparisons of downstream analysis methods.
@@
 The simulated reads are designed for comparing the performance of different alignment strategies.
-It has many parameters, of which the ones relevant to our discussion of simulating realistic DNA methylation data are `-CG` and `-CH`, the bisulfite conversion rates for CG and CH methylation loci, respectively.
-These are set by the user with values between $0$ and $100$.
+`Sherman` has many parameters, of which the ones relevant to our discussion of simulating realistic DNA methylation data are `-CG` and `-CH`, the bisulfite conversion rates for CG and CH methylation loci, respectively.
+These are set by the user with values between $0$ and $100$ ($\%$).
 Reads are simulated by sampling from the user-specified reference genome.
@@
 CH) loci have an average $\beta$-value of `-CG` (resp.
-`-CG`) regardless of the genomic context, which we know to be inconsistent with real data.
+`-CG`) regardless of the genomic context, which we know to be incompatible with real data.
 Furthermore, the methylation state of each methylation locus is independent, which is clearly inconsistent with the strong co-methylation observed in real data.
-
-`FastqToBS` ([http://users.dimi.uniud.it/~nicola.prezza/projects.html](http://users.dimi.uniud.it/~nicola.prezza/projects.html)) uses the same strategy as `Sherman`.
 
@@
 
-The reads simulated by `DNemulator` will result in $\beta$-values that have more context-dependence than those resulting from reads generated by `Sherman`, `FastqToBS` or `BSsim`.
+The reads simulated by `DNemulator` will result in $\beta$-values that have more context-dependence than those resulting from reads generated by `Sherman`.
 However, methylation events are still generated independently of one another, which means there is no co-methylation in the simulated data.
@@
 
-`BSsim` ([http://122.228.158.106/BSSim/](http://122.228.158.106/BSSim/) and used in \cite{Xie:2014ez}), uses a similar strategy to `DNemulator`.
+Other software for simulating individual bisulfite-sequencing reads are `FastqToBS` ([http://users.dimi.uniud.it/~nicola.prezza/projects.html](http://users.dimi.uniud.it/~nicola.prezza/projects.html)), which uses a similar strategy as `Sherman`, and `BSsim` ([http://122.228.158.106/BSSim/](http://122.228.158.106/BSSim/), and used in \citet{Xie:2014ez}), which has a similar strategy to `DNemulator`.
 
-### Methods for simulating aggregate methylation levels
+\subsection{Methods for simulating aggregate methylation levels}
+
+- [x] DONE
+
+The most widely studied downstream analysis problem is that of identifying differential methylation, which is done by comparing summary measures of methylation, such as $\beta$-values, between two or more groups.
+It is therefore generally not necessary to simulate individual reads, but is sufficient to directly simulate these aggregated measures, when simulating data for these studies.
 
 Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section.
-Generally, the simulation method is not a major feature of the paper but is there to support claims about the performance of the proposed method.
+Generally, such a simulation method exists to support claims about the performance of the proposed method and is not a major feature of the paper.
 Consequently, the simulation model is often only briefly described and a software implementation is rarely made available.
-In fact, of the methods reviewed in this section, only that of \cite{Lacey:2013iy} has a software implementation available.
+In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available[^WGBSsuite].
 
-The most widely studied downstream analysis problem is that of identifying differential methylation.
-Differential methylation is identified by comparing summary measures of methylation, such as $\beta$-values, between two or more groups.
-It is therefore generally not necessary to simulate individual reads but is sufficient to simulate these summary measures directly.
-As the `methsim` software is focused on simulating data for comparing methods for calling differential methylation, I review methods designed for a similar purpose.
+[^WGBSsuite]: As I was writing this chapter, the `WGBSsuite` software was published \citep{Rackham:2015bv}.
+`WGBSsuite` is the only published software dedicated to simulating whole-genome bisulfite sequencing data for the purpose of comparing methods for identifying differential methylation.
 
-A common framework for simulating data for a $k$-group experiment is the following hierarchical model:
+Fortunately, these simulation methods follow a common framework, even if the details differ:
 
-1. Simulate the unobserved group-specific true methylation levels, $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$.
-  - For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}}$.
-  - Differentially methylated loci are simulated by setting loci with $B_{i, j_{k}} \neq B_{i, j_{k'}}$
-2. Simulate the observed sample-specific sequencing depths, $u_{i, j} + m_{i, j}$.
-3. Simulate the observed sample-specific methylation levels, $\beta_{i, j_{k}}$ from some model incorporating both $B_{i, j_{k}}$ and $u_{i, j} + m_{i, j}$.
+0. [Optional] Simulate the locations of methylation loci.
+1. Simulate the unobserved group-specific true methylation levels.
+2. Simulate the observed sample-specific sequencing depths.
+3. Simulate the observed sample-specific methylation levels, e.g., the $\beta$-values.
 
-A popular choice of parametric model is the Beta-Binomial distribution \cite[e.g.,]{Feng:2014iq, Lacey:2013iy, Xu:2013eg, Chen:2014jb, Dolzhenko:2014bo}.
-Under this model the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j}, \phi_{i, j})$ distribution, and the $\beta$-values are simulated from a Binomial distribution, $\beta_{i, j} = Binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
+A popular choice of parametric model in this framework is the beta-binomial model \citep[e.g.,]{Feng:2014iq, Lacey:2013iy, Xu:2013eg, Chen:2014jb, Dolzhenko:2014bo}.
+This model, and others, are now discussed.
 
-#### Simulating sequence depth
+\subsubsection{Simulating methylation loci}
 
-The sequencing depths, $m_{i, j} + u_{i, j}$, are typically sampled from real data \cite[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, although a rounded Normal distribution \cite{Xu:2013eg} and a rounded mixture of Gamma distributions have also been used \cite{Lacey:2013iy}.
+- [x] DONE
 
-The most sophisticated approach to simulation of sequencing depth in bisulfite-sequencing experiments is the RRBS simulation of \cite{Lacey:2013iy}.
-In addition to using a mixture of distributions to capture both the low-coverage and high-coverage modes observed in RRBS sequencing coverage, the authors model the correlation of sequencing depth across samples for a given region.
+\citet{Lacey:2013iy} and \citet{Rackham:2015bv} are notable in that they choose to simulate the locations of CpGs rather than simply using their locations in a reference genome.
+Both papers use hidden Markov models to simulate genomes with regions of high and low CpG density.
+
+I do not think this a useful or necessary step.
+In fact, I think it is counterproductive.
+While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed.
+If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome.
+And it is counterproductive because the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci.
+Furthermore, almost all downstream analyses are _reference-based_ (__TODO: Cite relevant section__), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.
+
+\subsubsection{Simulating $B_{i, j}$}
+
+- [x] DONE
+
+Recall that $B_{i, j}$ is the underlying _true_ methylation level at locus $i$ in sample $j$.
+In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific.
+That is, we want to specify $B_{i, j^{k}}$, where $j^{k}$ indicates that sample $j$ comes from group $k$.
+For non-differentially methylated loci all $k$ groups have identical $B_{i, j^{k}} = B_{i, j^{0}}$; differentially methylated loci are simulated by setting $B_{i, j^{k}} \neq B_{i, j^{k'}}$ for some $k \neq k'$.
+
+Under the beta-binomial model, the true methylation level, $B^{i, j}$, is assumed to follow a $Beta(\mu_{i, j^{k}}, \phi_{i, j^{k}})$ distribution, where $\mu_{i, j^{k}}$ is the mean and $\phi_{i, j^{k}}$ is the dispersion of the beta distribution[^beta_parameterisation].
+Both the means, $\mu_{i, j^{k}}$, and the dispersions, $\phi_{i, j^{k}}$, are group-specific and are allowed to vary across methylation loci ($i$).
+The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e., the within-group _biological variability_ of DNA methylation.
+
+[^beta_parameterisation]: Note that this is different to the standard parameterisation of the Beta distribution, which is described by two shape parameters, $\alpha$ and $\beta$.
+The relationship between the two parameterisations is $\mu = \frac{\alpha}{\alpha + \beta}$ and $\phi = \frac{1}{\alpha + \beta + 1}$ \citep{Feng:2014iq}.
+
+As noted by \citet{Feng:2014iq}, the beta distribution is a very flexible distribution with support on $[0, 1]$ and has "long been a natural choice to model binomial proportions", particularly as a conjugate prior, as it is used in the empirical Bayes model of \citet{Feng:2014iq}.
+The beta-binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or overdispersed binomial distributions
+
+Other distributions may be used instead of the beta distribution for modelling $B_{i, j}$.
+For example, \citet{Xie:2014ez} consider both a single Gaussian distribution and a mixture of Gaussian distributions, while \citet{Xu:2013eg} consider both a truncated Gaussian and a mixture of truncated Gaussian distributions.
+
+An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$.
+`WGBSsuite`\citep{Rackham:2015bv} uses a modified form of this approach.
+In `WGBSsuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated").
+Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset.
+For example, the average methylation level for all "de-methylated" regions is defined as $B_{'de-methylated'} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$.
+Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e., $B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$.
+Care needs to be taken that $0 \leq B_{i, j} \leq 1$.
+
+\subsubsection{Simulating sequence depth}
+
+- [x] DONE
+
+The sequencing depth at each methylation loci, $m_{i, j} + u_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.
+
+The most sophisticated approach to simulation of sequencing depth in bisulfite-sequencing experiments is given by \citet{Lacey:2013iy}.
+In addition to using a mixture of distributions to capture both the low-coverage and high-coverage modes observed in RRBS sequencing coverage, \citet{Lacey:2013iy} model the correlation of sequencing depth across samples for a given region.
 They do this by using a Normal copula to make the set of sequencing depths a jointly dependent set of random variables.
 While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the benefits are unclear.
-Moreover, it is simpler to include such correlations by a sensible sampling of real data.
-Specifically, this can be achieved by sampling sequencing depth across samples for a given region rather than sampling across the full set of sequencing depths.
+Moreover, it is simpler, and likely more computationally efficient, to include such correlations by a sensible sampling of sequencing depths from real data.
 
-#### Simulating $B_{i, j}$
+\subsubsection{Simulating the observed methylation levels}
 
-Under the Beta-Binomial model, $\mu_{i, j}$ is the mean and $\phi_{i, j}$ the dispersion of the Beta distribution[^beta_parameterisation].
-Both the means, $\mu_{i, j}$, and the dispersions, $\phi_{i, j}$, are group-specific and are allowed to vary across methylation loci ($i$).
-The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e., the _biological variability_ of DNA methylation.
+- [x] DONE
 
-[^beta_parameterisation]: Note that this is different to the standard parameterisation of the Beta distribution, which is described by two shape parameters, $\alpha$ and $\beta$.
-The relationship between the two parameterisations is $\mu = \frac{\alpha}{\alpha + \beta}$ and $\phi = \frac{1}{\alpha + \beta + 1}$ \cite{Feng:2014iq}.
+The final step is to simulate the read counts, $M_{i, j}$ and $U_{i, j}$.
+These are based on the true underlying methylation level, $B_{i, j}$, and the sequencing depth, $m_{i, j} + u_{i, j}$.
+In the beta-binomial model this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
+In addition to binomial read sampling, the `WGBSsuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.
 
-As noted by \cite{Feng:2014iq}, the Beta distribution is a very flexible distribution with support on $[0, 1]$ and has "long been a natural choice to model binomial proportions", particularly as a conjugate prior, as it is used in the empirical Bayes model of \cite{Feng:2014iq}.
-The Beta-Binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or overdispersed Binomail distributions
+\subsubsection{Simulating differentially methylated regions}
 
-Other distributions may be used instead of the Beta distribution.
-For example, \cite{Xie:2014ez} consider a Normal distribution and a mixture of Normal distributions and \cite{Xu:2013eg} consider the truncated Normal and a mixture of truncated Normal distributions.
+- [ ] DONE
 
-An alternative to specifying a parametric distribution for the $B_{i, j}$ would be to sample these from real data, i.e., sample the observed $\beta_{i, j}$.
-However, without further perturbation, such an approach would have insufficient within-group variability since all samples in that group have the same "true" methylation level.
+As we have seen, the simulation of a differentially methylated locus is straightforward; for a given locus ($i$) simply vary $B_{i, j_{i}}$ across the $k$ groups.
+The simulation of a differentially methylated region is more complex.
 
-While the simulation of differentially methylated loci is straightforward - simply vary the mean of the distribution of $B_{i, j_{i}}$ across the $k$ groups - the simulation of differentially methylated regions is more complex.
-In principal we would simply simulate a differentially methylated region by simulating runs of differentially methylated loci.
+In principal we can simulate a differentially methylated region by simply simulating runs of differentially methylated loci.
 However, it must be noted that this requires careful choice of parameters.
-Such parameters include the length of DMRs, the minimal number of loci they must contain, the maximal intra-pair distances of loci within a DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should __FIGURE__ be considered one DMR or two DMRs?
+Such parameters include the length of the DMR, the minimal number of loci it must contain, the maximal intra-pair distances of loci within the DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should __FIGURE__ be considered one DMR or two DMRs?
 (__TODO: Figure with 4 DMCs, then 1 non-DMC and then 5 DMCs).
-These decisions ultimately have to be made by the user of the simulation software based on the types of events they are interested in.
-For example, DMRs in cancer can be very long whereas others may be interested in identifying very short DMRs overlapping transcription factor binding sites.
+Such decisions ultimately have to be made by the user of the simulation software based on the types of events they are interested in.
 
-#### Simulating co-methylation
+__TODO: Comment on how the various methods simulate DMRs, not just DMCs.
 
-A key feature ignored by all the above methods, with the honourable exception of \cite{Lacey:2013iy}, is co-methylation.
+\subsection{Simulating co-methylation}
+
+- [ ] DONE
+
+A key feature ignored by all the above methods, with the honourable exceptions of \citet{Lacey:2013iy} and \citet{Rackham:2015bv}, is co-methylation.
 The methylation states of neighbouring loci are highly dependent and, consequently, the $B_{i, j}$ of nearby loci are highly dependent.
-To be clear, almost all of the published simulation studies of bisulfite-sequencing data have failed to model a very important feature of DNA methylation data, co-methylation.
+To be clear, the majority of studies using simulated DNA methylation data have failed to model a very important feature of DNA methylation data, co-methylation.
 
@@
 However, a simulation model of DNA methylation data need not induce correlations at the level of individual DNA fragments.
-Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation but it may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
-This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependence.
-This idea takes its inspiration from \cite{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.
+Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation, but it still may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
+This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependent.
+This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.
 
-\cite{Lacey:2013iy} take a different approach, but one that still results in correlated $B_{i, j}$.
+\citet{Lacey:2013iy} take a different approach, but one that still results in correlated $B_{i, j}$.
 To begin, they compute $\beta$-values from chromosome 11 for a single normal myotube cell line that was sequenced with RRBS.
@@
 
-The second step uses a method published by \cite{Zaykin:2002ko}.
+The second step uses a method published by \citet{Zaykin:2002ko}.
 The transformed values, $B^{*}$, are created by the transformation $B^{*}_{i, j} = 1 - \Phi{C \Phi - 1(1 - B)}$, where $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, which is estimated from the fitted Gaussian variogram, and where $\Phi(\dot)$ denotes the standard normal distribution function.
 
-\cite{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).
+__TODO: Review how `WGBSsuite` induce co-methylation__
+
+__TODO: Move this running time stuff elsewhere__
+\citet{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).
 __TODO: Try these scripts, see how long they take__.
-The running times reported in the orignal publication are shown in __TABLE__ for a single simulation.
+The running times reported in the original publication are shown in __TABLE__ for a single simulation.
 __TODO: To simulate a single RRBS dataset would take HOW LONG__.
@@
 \end{tabular}
-\caption{Reported running times for simulation methodology of \cite{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
+\caption{Reported running times for simulation methodology of \citet{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
 \label{tab:Lacey_runtime}
@@
 
-#### Simulating positions of methylation loci
+\subsubsection{Other model-based simulations}
 
-All of the described methods, except \cite{Lacey:2013iy}, simulate methylation events based on a reference genome.
-\cite{Lacey:2013iy} propose a hidden Markov model for the intra-pair distances of CpGs in human RRBS data and simulate methylation loci based on this model.
-I do not think a model-based strategy is useful here, in fact I think it is counterproductive, and I prefer a sampling-based strategy, i.e., the use of methylation loci from a reference genome.
-While the set of methylation loci do vary across samples within a species due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed, particularly when the main aim of the simulation model is to realistically simulate methylation levels and not the location of the methylation loci.
-Furthermore, if sequence variation is required then this can be achieved by sampling from this set of loci or by sampling from the set of methylation loci from a group of samples.
-
-#### Other model-based simulations
+- [ ] DONE
 
@@
 
-### Methods based entirely on sampling real data
+__TODO: Also cite Jaffe (2012).__
+
+\subsection{Methods based on re-sampling real data}
+
+- [ ] DONE
 
@@
 
-\cite{Sofer:2013bk} use a sampling-based simulation method in the development of their `Aclust` software for identifying differential methylation from microarray data.
-The idea is adapted from a simulation method for array comparative genomic hybridization experiments \cite{Gaile:2007bq}.
-\cite{Sofer:2013bk} sample "blocks" of CpGs, a region of the genome where all CpGs are within 10 kb of the next, to "generate (spatial) correlation-preserved methylation data".
+\citet{Sofer:2013bk} use a sampling-based simulation method in the development of their `Aclust` software for identifying differential methylation from microarray data.
+The idea is adapted from a simulation method for array comparative genomic hybridization experiments \citet{Gaile:2007bq}.
+\citet{Sofer:2013bk} sample "blocks" of CpGs, a region of the genome where all CpGs are within 10 kb of the next, to "generate (spatial) correlation-preserved methylation data".
 By sampling blocks rather than individual CpGs, this sampling scheme preserves the correlation structure between CpGs occuring in the same block.
@@
 
-\cite{Sofer:2013bk} sample from a dataset of 539 Illumina 450k microarrays and select a small number of "target" CpGs, which are CpGs whose methylation level is highly variable across the 539 samples.
+\citet{Sofer:2013bk} sample from a dataset of 539 Illumina 450k microarrays and select a small number of "target" CpGs, which are CpGs whose methylation level is highly variable across the 539 samples.
 If a block does not contain a "target" then it is sampled uniformly at random from the 539 samples.
@@
 
-## Methodology
+\subsection{Summary}
 
-`methsim` is is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements.
+- [ ] DONE
+
+- Summarise existing methods and software.
+
+\section{Methods}
+
+- [ ] DONE
+
+`methsim` is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements.
 My intention in developing `methsim` was to design simulation software that could generate realistic data for use in developing and comparing differential methylation callers.
 In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments.
-An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or sequencing reads.
+An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or entire sequencing reads.
 
@@
 
-### Simulating the truth
+\subsection{Simulating the truth}
+
+- [ ] DONE
 
@@
 
-#### Segmenting the methylome
+\subsubsection{Segmenting the methylome}
 
-\cite{Burger:2013kq} developed the R/Bioconductor package, `MethylSeekR`, to discover regulatory motifs from bisulfite-sequencing data.
+- [ ] DONE
+
+\citet{Burger:2013kq} developed the R/Bioconductor package, `MethylSeekR`, to discover regulatory motifs from bisulfite-sequencing data.
 `MethylSeekR` segments the genome into unmethylated regions (_UMR_s), lowly-methylated regions (_LMR_s) and partially methylated regions[^pmds] (_PMR_s).
@@
 `methsim` uses the output of `MethylSeekR` to define "regions of similarity", namely UMRs, LMRs, PMRs and the 'rest of the genome'.
-The 'rest of the genome' roughly corresponds to 'mostly methylated regions' (_MMR_s), although \cite{Burger:2013kq} do not describe it as such.
+The 'rest of the genome' roughly corresponds to 'mostly methylated regions' (_MMR_s), although \citet{Burger:2013kq} do not describe it as such.
 (__TODO: Figure showing 'rest of genome' is MMRs__)
@@
 
-#### Simulating a single haploid methylome
+\subsubsection{Simulating a single haploid methylome}
+
+- [ ] DONE
 
@@
 
-#### Simulating a single sample
+\subsubsection{Simulating a single sample}
+
+- [ ] DONE
 
@@
 
-#### Simulating multiple samples from the same group
+\subsubsection{Simulating multiple samples from the same group}
 
-#### Simulating differential methylation
+- [ ] DONE
 
-### Parameter estimation
+\subsubsection{Simulating differential methylation}
 
-## Implementation
+- [ ] DONE
 
-## Case study
+\subsection{Parameter estimation}
 
-## Discussion of findings
+- [ ] DONE
 
-## Conclusions
+\subsection{Implementation}
+
+- [ ] DONE
+
+\section{Case study}
+
+- [ ] DONE
+
+\section{Conclusions}
+
+- [ ] DONE
+
+# TODO:
+
+- [ ] Decide on a variable for sequencing depth; $m_{i, j} + u_{i, j}$ is kinda confusing. Perhaps $d_{i, j} / D_{i, j}$?

### 2015-03-19 latex/phd_thesis.tex
Missing includes
Temporarily remove chapter from draft
Figure fixing.
Minor fixes
Added methsim chapter to main .tex file.

@@
 \usepackage{tabulary}
+
+\usepackage{mathabx}
+
+\usepackage{placeins}
 
@@
 of the degree of Doctor of Philosophy\\\\
-November 2014\\\\
+March 2015\\\\
 Department of Mathematics and Statistics\\
@@
 
-* Some data in my thesis are mapped to hg18 and some to hg19.
-* Include summary of notation?
+\begin{itemize}
+  \item Some data in my thesis are mapped to hg18 and some to hg19.
+  \item Include summary of notation?
+\end{itemize}
 

### 2015-03-23 Rmarkdown/methsim.Rmd
Added figure of “one DMR or two DMRs” Started Methods.

@@
 Consequently, the simulation model is often only briefly described and a software implementation is rarely made available.
-In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available[^WGBSsuite].
+In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available[^WGBSSuite].
+As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}.
+`WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data when comparing methods for identifying differential methylation.
+`WGBSSuite` is available for download as a collection of R scripts[^WGBSSuite].
 
-[^WGBSsuite]: As I was writing this chapter, the `WGBSsuite` software was published \citep{Rackham:2015bv}.
-`WGBSsuite` is the only published software dedicated to simulating whole-genome bisulfite sequencing data for the purpose of comparing methods for identifying differential methylation.
+[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software.
+Furthermore, no licence file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code.
+(__TODO: Email the developer, Owen Rackham, asking him to add a licence.
+He is considering my request.__)
 
@@
 An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$.
-`WGBSsuite`\citep{Rackham:2015bv} uses a modified form of this approach.
-In `WGBSsuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated").
+`WGBSSuite`\citep{Rackham:2015bv} uses a modified form of this approach.
+In `WGBSSuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated").
 Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset.
@@
 In the beta-binomial model this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
-In addition to binomial read sampling, the `WGBSsuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.
+In addition to binomial read sampling, the `WGBSSuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.
 
@@
 
-- [ ] DONE
+- [x] DONE
 
-As we have seen, the simulation of a differentially methylated locus is straightforward; for a given locus ($i$) simply vary $B_{i, j_{i}}$ across the $k$ groups.
+As we have seen, the simulation of a differentially methylated locus is straightforward; for a given locus ($i$) simply vary $B_{i, j_{k}}$ across the $k$ groups.
 The simulation of a differentially methylated region is more complex.
@@
 However, it must be noted that this requires careful choice of parameters.
-Such parameters include the length of the DMR, the minimal number of loci it must contain, the maximal intra-pair distances of loci within the DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should __FIGURE__ be considered one DMR or two DMRs?
-(__TODO: Figure with 4 DMCs, then 1 non-DMC and then 5 DMCs).
+Such parameters include the length of the DMR, the minimal number of loci it must contain, the maximal intra-pair distances of loci within the DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should Figure \ref{fig:dmr_break} be considered one DMR or two DMRs?
 Such decisions ultimately have to be made by the user of the simulation software based on the types of events they are interested in.
 
-__TODO: Comment on how the various methods simulate DMRs, not just DMCs.
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/dmr_break.pdf}
+\caption{A hypothetical region in a study of differential methylation in a two-group experiment.
+Plotted is the difference in $\beta$-value between the two groups ($\delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}).
+The first three methylation loci are DMLs, as are the last five methylation loci.
+However, the fourth locus is not a DML.
+Should this region be considered as two distinct DMRs or as a single DMR?}
+\label{fig:dmr_break}
+\end{figure}
 
@@
 
-- [ ] DONE
+- [x] DONE
 
-A key feature ignored by all the above methods, with the honourable exceptions of \citet{Lacey:2013iy} and \citet{Rackham:2015bv}, is co-methylation.
+A key feature ignored by the majority of simulation methods, with the notable exceptions of \citet{Lacey:2013iy} and \citet{Rackham:2015bv}, is co-methylation.
 The methylation states of neighbouring loci are highly dependent and, consequently, the $B_{i, j}$ of nearby loci are highly dependent.
@@
 Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation, but it still may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
-This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependent.
-This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.
+For example, under the Beta-Binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent.
+This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure via a lag-1 autoregressive model of the simulated $\beta$-values.
 
-\citet{Lacey:2013iy} take a different approach, but one that still results in correlated $B_{i, j}$.
+\citet{Lacey:2013iy} take a different approach, but one that still results in correlated $B_{i, j_{k}}$ across $i$, within group $k$.
 To begin, they compute $\beta$-values from chromosome 11 for a single normal myotube cell line that was sequenced with RRBS.
@@
 
-1. Simulate $B_{i, j}$ from a Beta distribution with parameters estimated from the chromosome 11 MTCTL2 data. These are estimated under an assumption of independence.
-2. Induce correlation amongst the $B_{i, j}$ (across $i$) by a transformation of the $B_{i, j}$.
+1. Simulate $B_{i, j_{k}}$ from a Beta distribution with parameters estimated from the chromosome 11 MTCTL2 data. These are estimated under an assumption of independence.
+2. Induce correlation amongst the $B_{i, j_{k}}$ (across $i$) by a transformation of the $B_{i, j_{k}}$.
 
 The second step uses a method published by \citet{Zaykin:2002ko}.
-The transformed values, $B^{*}$, are created by the transformation $B^{*}_{i, j} = 1 - \Phi{C \Phi - 1(1 - B)}$, where $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, which is estimated from the fitted Gaussian variogram, and where $\Phi(\dot)$ denotes the standard normal distribution function.
+The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi{C \Phi - 1(1 - B_{j_{k}})}$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\dot)$ denotes the standard normal distribution function.
 
-__TODO: Review how `WGBSsuite` induce co-methylation__
-
-__TODO: Move this running time stuff elsewhere__
-\citet{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).
-__TODO: Try these scripts, see how long they take__.
-The running times reported in the original publication are shown in __TABLE__ for a single simulation.
-__TODO: To simulate a single RRBS dataset would take HOW LONG__.
-The simulation scripts could be made more user-friendly by creating an R package, rather than supplying the software as a collection of scripts, and would benefit from some documentation.
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}lll@{}}
-\toprule
-Number of samples & Number of sites & Running time (seconds) \\ \midrule
-8                 & 1000            & \textless 1            \\
-8                 & 5000            & 13                     \\
-8                 & 10000           &                        \\ \bottomrule
-\end{tabular}
-\caption{Reported running times for simulation methodology of \citet{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
-\label{tab:Lacey_runtime}
-\end{table}
+`WGBSSuite` induces spatial correlation amongst the $B_{i, j_{k}}$ in a less direct manner.
+Recall that each methylation locus is assigned one of the four underlying states ("de-methylated", "1st transition", "2nd transition", "methylated") via a hidden Markov model.
+The transition matrix of this hidden Markov model ensures that neighbouring loci, $i, (i+1)$, are more likely to be assigned the same state.
+Furthermore, since all loci within each of the four states are assigned the same (perturbed) underlying methylation level, $B_{i, j_{k}} = B_{region-type} + \epsilon_{i, j_{k}}$, neighbouring loci have similar methylation levels.
+Note that the $IPD$ only plays a direct role in the initial segmentation of the genome, not in the assigning of the $B_{i, j_{k}}$.
 
@@
 
-- [ ] DONE
+- [x] DONE
 
-A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \cite[e.g.,]{Chen:2014jb, Chen:2013eh}.
-These models are designed for simulating microarray data and not sequencing data, since they do not include the variability caused by variation in sequencing depth.
-Neither of these include co-methylation in their simulation model.
-
-__TODO: Also cite Jaffe (2012).__
+A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \citet[e.g.,]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
+These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth.
+Of these methods, only \citet{Jaffe:2012gx} simulate correlation amongst the $B_{i, j_{k}}$.
 
@@
 
-- [ ] DONE
+- [x] DONE
 
-A simulation can be based entirely on sampling real data[^synthetic].
-This type of simulated data is attractive because through careful sampling it can capture heterogeneity and correlations that may be too difficult to parameterise in a model.
-At the same time, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features, or worse, introduce artefacts into the simulated data.
+A simulation may also be based entirely on re-sampling of real data[^synthetic].
+This type of simulated data is attractive because through careful sampling it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model.
+At the same time, however, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features or, worse still, introduce artefacts into the simulated data.
 
@@
 
-Sampling methods are most easily implemented at the level of $\beta$-values.
-Sampling reads is reads is far more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis.
-Down-sampling is only of interest to examine the effects of sequencing coverage on downstream analyses.
+Re-sampling methods are most easily implemented at the level of $\beta$-values.
+Re-sampling reads is reads is more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis.
+Down-sampling is only really of interest to examine the effects of sequencing coverage on downstream analyses.
 
-\citet{Sofer:2013bk} use a sampling-based simulation method in the development of their `Aclust` software for identifying differential methylation from microarray data.
-The idea is adapted from a simulation method for array comparative genomic hybridization experiments \citet{Gaile:2007bq}.
+\citet{Sofer:2013bk} use a re-sampling based simulation method in the development of their `Aclust` software for identifying differential methylation from microarray data.
+The idea is adapted from \citet{Gaile:2007bq}, which is a simulation method for array comparative genomic hybridization experiments.
 \citet{Sofer:2013bk} sample "blocks" of CpGs, a region of the genome where all CpGs are within 10 kb of the next, to "generate (spatial) correlation-preserved methylation data".
-By sampling blocks rather than individual CpGs, this sampling scheme preserves the correlation structure between CpGs occuring in the same block.
+By sampling blocks rather than individual CpGs, this sampling scheme preserves the correlation structure between CpGs occurring in the same block.
 Two blocks are unlikely to be correlated since there is little evidence that CpGs separated by more than 10 kb have spatially correlated methylation levels.
@@
 
-- [ ] DONE
+- [x] DONE
 
-- Summarise existing methods and software.
+My initial interest in simulation methods was to explore models of co-methylation, particularly at the level of individual DNA fragments. This requires two key capabilities:
+
+1. Simulation of individual reads.
+2. Simulation of co-methylation.
+
+None of the published simulation methods satisfied both these requirements. This motivated the development of `methsim`.
+
+`methsim` is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements.
+In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments.
+An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or entire sequencing reads.
+While my initial motivation in developing `methsim` was to explore models of co-methylation, I realised that it could also be used in the development and comparison of methods for the downstream analysis of DNA methylation data.
 
@@
 
-- [ ] DONE
-
-`methsim` is specifically designed to model the co-methylation structure of bisulfite-sequencing data by simulating individual DNA fragments rather than directly simulating summary methylation measurements.
-My intention in developing `methsim` was to design simulation software that could generate realistic data for use in developing and comparing differential methylation callers.
-In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments.
-An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or entire sequencing reads.
+- [x] DONE
 
@@
 1. Simulate the true methylome of each sample.
-2. Sample reads, including sequencing error and bisulfite-conversion error.
+2. Simulate reads, including sequencing error and bisulfite-conversion error, by sampling from the true methylome.
 3. Construct the output, be it reads, methylation patterns at m-tuples or $\beta$-values.
@@
 
-- [ ] DONE
+- [x] DONE
 
@@
 `methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together.
-The idea of segmenting a highly heterogenous stochastic process into a series of locally homogeneous prcoesses is not new.
+The idea of segmenting a globally heterogenous stochastic process into a series of locally homogeneous processes is not new.
 A hidden Markov model is an example of such a process.
 While the entire process may highly heterogeneous, conditional on the hidden states the process is homogeneous.
-Hidden Markov models, and other models assuming local similarity in spite of global heterogeneity, have been used with great success in bioinformatics (__CITE__).
+Hidden Markov models, and other models assuming local similarity in spite of global heterogeneity, have been used with great success in bioinformatics.
 
@@
 
-Actually, the stitching together of segments is not strictly a separate step and can be incorporated into step 2.
+In practice, the stitching together of segments is not strictly a separate step and can be incorporated into step 2.
 
@@
 
-- [ ] DONE
+- [x] DONE
 
-\citet{Burger:2013kq} developed the R/Bioconductor package, `MethylSeekR`, to discover regulatory motifs from bisulfite-sequencing data.
-`MethylSeekR` segments the genome into unmethylated regions (_UMR_s), lowly-methylated regions (_LMR_s) and partially methylated regions[^pmds] (_PMR_s).
+`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the methylome into these regions of similarity.
+Initially published to discover regulatory motifs from bisulfite-sequencing data, `MethylSeekR` segments the genome into unmethylated regions (_UMR_s), lowly-methylated regions (_LMR_s) and partially methylated regions[^pmds] (_PMR_s).
 It does this using a two-stage algorithm applied to the $\beta$-values from a sample:
@@
 Identify partially methylated regions.
-A summary statistic, $\alpha$, which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs.
+A summary statistic, $\alpha$, (__TODO: What is $\alpha$__) which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs.
 Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify _PMR_s and non-_PMR_s
@@
 
-`methsim` uses the output of `MethylSeekR` to define "regions of similarity", namely UMRs, LMRs, PMRs and the 'rest of the genome'.
-The 'rest of the genome' roughly corresponds to 'mostly methylated regions' (_MMR_s), although \citet{Burger:2013kq} do not describe it as such.
-(__TODO: Figure showing 'rest of genome' is MMRs__)
+`methsim` uses the functionality of `MethylSeekR` to partition the genome into "regions of similarity", namely `UMR`, `LMR`, `PMR` or `other`[^other].
+While a tailored algorithm may improve the segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting into "regions of similarity".
+This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)).
+It should be noted that `methsim` also performs some post-processing of the `MethylSeekR` output so that the resulting segmentation is disjoint and exhaustive.
 
-While a tailored algorithm may improve the segmentation process, the segmentation produced by `MethylSeekR` is a reasonable approximation to finding "regions of similarity".
-This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)).
-It should be noted that `methsim` does perform some post-processing of the `MethylSeekR` output so that the segmentation is disjoint and exhaustive.
-
-The segmentation is a property of the sample, not of any particular haploid methylome in the sample.
+[^other]: `other` roughly corresponds to 'mostly methylated regions', although \citet{Burger:2013kq} do not describe these regions as such.
 
@@
 
-I propose to model $Z_{h, j}$ as a two-state, one-step Markov chain.
-This allows me to incorporate a model of within-fragment co-methylation into my simulation.
+`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $Z_{h, j}$[^methsim_model].
+This allows the incorporation of within-fragment co-methylation into the simulation.
 The choice of a one-step chain is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.
+
+[^methsim_model]: In fact, as we will see, `methsim` does not directly model $Z_{h, j}$, but rather models a kind of hybrid process.
 
@@
 [^context]: Actually, it only depends on the context of the $i^{th}$ locus.
-Most pairs of loci, $(i, i + 1)$ will lie in the same genomic feature.
-For pairs that span the boundary of two different features I have arbitrarily chosen to use the context of the $i^{th}$ locus.
+Most pairs of loci, $(i, i + 1)$, will lie in the same context.
+For pairs that span the boundary of two different contexts I have arbitrarily chosen to use the context of the $i^{th}$ locus.
 
@@
 Call these empirical distributions $\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, respectively.
-These empirical distributions are shown for a single sample in __FIGURE__.
-(__TODO: Do these plots for a single sample__).
+These empirical distributions are shown for the ADS in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots}
 
-__TODO: Describe how a transition probability can be computed from a marginal probability and a log odds ratio__
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
+\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the _ADS_ sample.
+Only CpGs with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:ADS_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/ADS_lor_by_pm_quantile_plots.pdf}
+\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _ADS_ sample.
+Only 2-tuples with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:ADS_lor_by_pm_quantile_plots}
+\end{figure}
+
+These same distributions for all the _Ziller_, _Lister_, _Seisenberger_ and _Ziller_ datasets are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lora_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lora_by_pm_quantile_plots},
+\ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
+\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots}.
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
+\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the EPISCOPE dataset.
+Only CpGs with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:EPISCOPE_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_by_pm_quantile_plots.pdf}
+\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _EPISCOPE_ sample.
+Only 2-tuples with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:EPISCOPE_lor_by_pm_quantile_plots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
+\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the Lister dataset.
+Only CpGs with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Lister_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_lor_by_pm_quantile_plots.pdf}
+\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _Lister_ sample.
+Only 2-tuples with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Lister_lor_by_pm_quantile_plots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
+\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the Seisenberger dataset.
+Only CpGs with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Seisenberger_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_by_pm_quantile_plots.pdf}
+\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _Seisenberger_ sample.
+Only 2-tuples with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Seisenberger_lor_by_pm_quantile_plots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
+\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the Ziller dataset.
+Only CpGs with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Ziller_merged_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_by_pm_quantile_plots.pdf}
+\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _Ziller_ sample.
+Only 2-tuples with at least $10\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Ziller_merged_lor_by_pm_quantile_plots}
+\end{figure}
+
+__UP TO HERE: Describe how a transition probability can be computed from a marginal probability and a log odds ratio__
 
@@
 
-\section{Case study}
+\section{Results}
+
+-  __TODO: Figure showing 'rest of genome' is MMRs. More generally, a figure showing the average methylation level in each of `UMR`, `LMR`, `PMR` and `other`.
+
+\subsection{Case study}
 
@@
 
+\subsubsection{Comparison to other methods}
+
+\citet{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).
+__TODO: Try these scripts, see how long they take__.
+The running times reported in the original publication are shown in __TABLE__ for a single simulation.
+__TODO: To simulate a single RRBS dataset would take HOW LONG__.
+The simulation scripts could be made more user-friendly by creating an R package, rather than supplying the software as a collection of scripts, and would benefit from some documentation.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+Number of samples & Number of sites & Running time (seconds) \\ \midrule
+8                 & 1000            & \textless 1            \\
+8                 & 5000            & 13                     \\
+8                 & 10000           &                        \\ \bottomrule
+\end{tabular}
+\caption{Reported running times for simulation methodology of \citet{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
+\label{tab:Lacey_runtime}
+\end{table}
+
+__TODO: Compare `methsim` to `WGBSSuite`.
+
 # TODO:

### 2015-03-24 Rmarkdown/methsim.Rmd
Begin describing how transition probabilities are formed.

@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/dmr_break.pdf}
@@
 As we have seen, DNA methylation is highly heterogeneous along the genome.
-Nonetheless, there are clear regions of "similarity", regions where we might hope to model by a simple parametric model.
+Nonetheless, there are clear regions of "similarity", regions where we might hope to model it by a simple parametric model.
 `methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together.
 The idea of segmenting a globally heterogenous stochastic process into a series of locally homogeneous processes is not new.
-A hidden Markov model is an example of such a process.
-While the entire process may highly heterogeneous, conditional on the hidden states the process is homogeneous.
+A hidden Markov model is an example of such a process; while the entire process may highly heterogeneous, conditional on the hidden states the process may be homogeneous.
 Hidden Markov models, and other models assuming local similarity in spite of global heterogeneity, have been used with great success in bioinformatics.
@@
 1. Segment the methylome into regions of similarity.
-2. Within each region simulate the methylation states from a simple parametric model.
-3. "Stitch" the resulting regions together.
+2. Within each region simulate the binary methylation states from a simple parametric model.
+3. "Stitch" the resulting region-level simulations together[^stitch].
 
-In practice, the stitching together of segments is not strictly a separate step and can be incorporated into step 2.
+[^stitch]: In practice, the stitching together of segments is incorporated into step 2.
+
+Suppose our sample contains $n_{h}$ haplotypes and, therefore, that our sample can be represented by[^strand_symmetric] $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$.
+We might try to simulate this sample by generating $n_{h}$ independent realisations of $\mathbf{Z_{h}}$.
+However, for a number of reasons, this is a little naive and ambitious.
+For example, under this naive model there is no control across haplotypes of the methylation patterns; although there exists between-haplotype heterogeneity, the $\mathbf{Z_{h}}$ are definitely not independent of one another.
+
+[^strand_symmetric]: Currently, `methsim` can only simulate CpG methylation and assumes that the methylation states of all CpGs are strand symmetric.
+
+Rather than try to explicitly model dependencies between the $\mathbf{Z_{h}}$, `methsim` uses the following heuristic procedure:
+
+1.
+Simulate $n_{\tilde{h}} \ll n_{h}$ independent realisations of $\mathbf{{Z_{\tilde{h}}}}$, where $\mathbf{Z_{\tilde{h}}}$ has the same stochastic properties as $\mathbf{Z_{h}}$.
+`methsim` uses a default value of $n_{\tilde{h}} = 8$.
+2.
+Within each region, $r_{i}$, assign each of $\mathbf{Z_{\tilde{h}}}$ a weight, $W_{r_{i}, \tilde{h}}$, where $\sum_{ \tilde{h} = 1}^{n_{ \tilde{h}}} W_{r_{i}, \tilde{h}} = 1$ and, by construction, $W_{r_{i}, 1} \geq \ldots \geq W_{r_{i}, n_{\tilde{h}}}$.
+
+For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (described below; __TODO: Discuss implications of this assumption somewhere below__).
+
+The weights are interpreted as the proportion of $\{ Z_{h, i} \}_{h = 1}^{h = n_{h}}$ with the methylation pattern given by $Z_{\tilde{h}, i}$ for the $i^{th}$ methylation locus[^weights].
+We write the matrix of weights for all methylation loci in the genome as $\mathbf{W} = \{ \mathbf{W_{r_{i}, \cdot}} \}_{i = 1}^{N_{loci}} = \{ \mathbf{W_{\cdot, \tilde{h}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$.
+In regions of low within-sample heterogeneity, such as those unmethylated regions with $B_{i} \approx 0$, this means we want $\mathbf{W_{r_{i}, \cdot}} \approx (1, 0, \ldots, 0)$, whereas in regions of high within-sample heterogeneity, such as partially methylated regions, we want a more dispersed $\mathbf{W_{r_{i}}, \cdot}$.
+
+[^weights]: Note that these weights are defined at the region-level so all loci in the region have the same weights.
+
+A simulated methylome is therefore described by $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$ and $\mathbf{W}$.
+For example, $B_{i} = \sum_{\tilde{h} = 1}^{n_{\tilde{h}}} W_{r_{i}, \tilde{h}} \times Z_{i, \tilde{h}}$.
+
+I now describe each step in this process in greater detail.
 
@@
 
-`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the methylome into these regions of similarity.
-Initially published to discover regulatory motifs from bisulfite-sequencing data, `MethylSeekR` segments the genome into unmethylated regions (_UMR_s), lowly-methylated regions (_LMR_s) and partially methylated regions[^pmds] (_PMR_s).
+`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the input methylome into these regions of similarity.
+Initially published to discover regulatory motifs from bisulfite-sequencing data, `MethylSeekR` segments the genome into unmethylated regions (\emph{UMR}s), lowly-methylated regions (\emph{LMR}s) and partially methylated regions[^pmds] (\emph{PMR}s).
 It does this using a two-stage algorithm applied to the $\beta$-values from a sample:
 
-[^pmds]: Partially methylated regions are more commonly known as partially methylated domains, but I refer to them as regions for consistency with UMRs and LMRs.
+[^pmds]: Partially methylated regions are also commonly known as partially methylated domains, but I refer to them as regions for consistency with UMRs and LMRs.
 
@@
 A summary statistic, $\alpha$, (__TODO: What is $\alpha$__) which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs.
-Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify _PMR_s and non-_PMR_s
+Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify PMRs and non-PMRs.
 2.
@@
 While a tailored algorithm may improve the segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting into "regions of similarity".
-This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)).
+This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)__).
 It should be noted that `methsim` also performs some post-processing of the `MethylSeekR` output so that the resulting segmentation is disjoint and exhaustive.
@@
 
-\subsubsection{Simulating a single haploid methylome}
+\subsubsection{Simulating $\mathbf{{Z_{\tilde{h}}}}$}
 
@@
 
-As discussed in Chapter \ref{chap:statmodel}, whole-genome bisulfite sequencing  measures DNA methylation from individual DNA fragments from a pool of 100s or 1000s of cells.
-The methylation states of each copy of the haploid genome can be represented by a pair of vectors, $Z_{h, j}^{+} = (Z_{h, 1, j}^{+}, \ldots, Z_{h, n_{loci}^{+}, j})$, for the positive strand, and $Z_{h, j}^{-} = (Z_{h, 1, j}^{-}, \ldots, Z_{h, n_{loci}^{-}, j})$, for the negative strand.
-For simplicity, I will focus on CpG methylation and assume that the methylation state of all CpGs are strand-symmetric so that each haploid genome can be represented by a single vector, $Z_{h, j} = (Z_{h, 1, j}, \ldots, Z_{h, n_{loci}, j})$.
-
-`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $Z_{h, j}$[^methsim_model].
+`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{{Z_{\tilde{h}}}}$.
 This allows the incorporation of within-fragment co-methylation into the simulation.
-The choice of a one-step chain is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.
-
-[^methsim_model]: In fact, as we will see, `methsim` does not directly model $Z_{h, j}$, but rather models a kind of hybrid process.
+The choice of a one-step process is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.
 
@@
 
-For these reasons I allow the transition probabilities to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous.
+For these reasons I allow the transition probabilities to vary with $i$ and $r_{i}$, that is, I allow the Markov chain to be spatially inhomogeneous.
 In particular, I allow the transition probabilities, $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} = b | Z_{h, i, j} = a)$, to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region for this pair of loci[^context], $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} | Z_{h, i, j}, r_{i}, IPD(i, i + 1))$, where $r_{i}$ is the region type (UMR, LMR, PMR or MMR) of the $i^{th}$ methylation locus.
 
-[^context]: Actually, it only depends on the context of the $i^{th}$ locus.
-Most pairs of loci, $(i, i + 1)$, will lie in the same context.
-For pairs that span the boundary of two different contexts I have arbitrarily chosen to use the context of the $i^{th}$ locus.
+[^context]: Actually, it only depends on the region of the $i^{th}$ locus.
+Most pairs of loci, $(i, i + 1)$, will lie in the same region.
+For pairs that span the boundary of two different regions I have arbitrarily chosen to use the region of the $i^{th}$ locus.
 
@@
 However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome.
-For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h, j)}}$, to estimate or otherwise assign.
+For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h)}}$, to estimate or otherwise assign.
+Again, I emphasise that `methsim` uses the same set of parameters for all $n_{\tilde{h}}$ "pseudo-haplotypes".
+Therefore, only one set of transition matrices need to be assigned.
 
-`methsim` assigns these transition probabilities based on empirical the distributions of $\beta_{i, j}$ and $LOR_{IPD(i, i + 1), j}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$.
-Call these empirical distributions $\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, respectively.
-These empirical distributions are shown for the ADS in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots}
+`methsim` assigns these transition probabilities based on $\beta_{i | r_{i}}$ and $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$, the empirical distributions of $\beta_{i}$ and $LOR_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
+These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots} (__TODO: Are these other examples necessary?__)
 
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the _ADS_ sample.
+\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/ADS_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _ADS_ sample.
+\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{ADS} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the EPISCOPE dataset.
+\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{EPISCOPE} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _EPISCOPE_ sample.
+\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{EPISCOPE} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the Lister dataset.
+\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Lister} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/Lister_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _Lister_ sample.
+\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Lister} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the Seisenberger dataset.
+\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Seisenberger} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/Seisenberger_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _Seisenberger_ sample.
+\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Seisenberger} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the Ziller dataset.
+\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Ziller} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \centering
-\centering
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i, j | r_{i}}$ and $LOR_{i, j | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the _Ziller_ sample.
+\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Ziller} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 
-__UP TO HERE: Describe how a transition probability can be computed from a marginal probability and a log odds ratio__
+`methsim` assigns these probabilities as follows:
+
+1.
+For each region, $r_{i}$, sample $B_{r_{i}}$ from $\beta_{i | r_{i}}$.
+This can be interpreted as $Pr(Z_{r_{i}} = 1)$, the marginal probability that a methylation locus in region $r_{i}$ is methylated, or, alternatively, as the average level of methylation for all methylation loci in the region.
+2.
+For every pair of methylation loci, $(i, i + 1)$, $ 1 \leq i \leq (n - 1)$, sample $\psi_{(i, i + 1)}$ from $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}.
+This can be interpreted as the odds that the $(i + 1)^{th}$ methylation locus has the same methylation state as the $i^{th}$ methylation locus.
+
+The second step requires further explanation.
+If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\psi_{(i, i + 1)}$ from a Gaussian distribution with mean equal to __TODO: What?__ and standard deviation equal to __TODO: What?__.
+__TODO: Explain how these means and standard deviations are estimated using loess.__ By default, if there are less than __TOOD: how many?__ $LOR$ with the given $IPD(i, i + 1), r_{i}$ combination then the Normal-approximation is used.
+The user may specify a different method for dealing with these rarer $IPD$-region combinations.
+
+Now that we have $B_{r_{i}}$ and $\psi_{i, i + 1}$ for $1 \leq i \leq n - 1$ we can simulate realisations of $\mathbf{Z_{\tild{h}}}$.
+To begin the process (i.e., when $i = 1$) $Z_{\tilde{h}, 1}$, is simulated from the "marginal" distribution, which is Bernoulli($B_{r_{1}}$).
+Then, we loop over $i = 2, \ldots, n - 1$ to simulate the remainder of the process.
+At each step we compute the transition probability, $p_{i, i + 1} = Pr(Z_{\tilde{h}, i + 1} = 1 | Z_{\tilde{h}, i}) = \frac{Pr(Z_{tilde{h}, i}, Z_{tilde{h}, i + 1} = 1)}{Pr(Z_{tilde{h}})}$, from $B_{r_{i}}, $B_{r_{i + 1}}, \psi_{(i, i + 1)}$ and the observed (simulated) value at the previous iteration, $Z_{\tilde{h}, i}$.
+Then, $Z_{\tilde{h}, i + 1}$ is simulated from the Bernoulli($p_{i, i + 1}$) distribution.
+
+To compute the transition probability, $p_{i, i + 1}$, `methsim` uses the iterative proportional fitting algorithm (IPF; __TODO: CITE IPD paper__) to construct the joint probability matrix, __TODO: Mathematically describe this matrix__.
+Then, the appropriate element of this matrix is divided by the appropriate marginal probability to obtain the transition probability.
+An example of this procedure is illustrative.
+
+The IPD procedure is a general method for __TODO: blah blah__.
+In this case we wish to use the IPF procedure to construct the unique $2 \times 2$ array (the joint probability matrix) with specified margins (the marginal probabilities of each methylation state at each locus) and the specified cross-product ratio (the odds-ratio).
+__UP TO HERE__
+
+\begin{equation}
+% An example of IPF being used to form a transition matrix.
+\end{equation}
 

### 2015-03-25 Rmarkdown/methsim.Rmd
Finished section on simulating Z.

@@
 
-\subsubsection{Simulating $\mathbf{{Z_{\tilde{h}}}}$}
+\subsubsection{Simulating $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$}
 
-- [ ] DONE
+- [x] DONE
 
@@
 `methsim` assigns these transition probabilities based on $\beta_{i | r_{i}}$ and $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$, the empirical distributions of $\beta_{i}$ and $LOR_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
-These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots} (__TODO: Are these other examples necessary?__)
+These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots}.
 
@@
 \includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
+\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/ADS_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{ADS} sample.
+\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{ADS} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 
-These same distributions for all the _Ziller_, _Lister_, _Seisenberger_ and _Ziller_ datasets are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lora_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lora_by_pm_quantile_plots},
+These same distributions for all the _Ziller_, _Lister_, _Seisenberger_ and _Ziller_ datasets are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
 \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
-\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots}.
+\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} (__TODO: Are these other examples necessary?__).
 
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{EPISCOPE} dataset.
+\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{EPISCOPE} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{EPISCOPE} sample.
+\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{EPISCOPE} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Lister} dataset.
+\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Lister} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Lister} sample.
+\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Lister} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Seisenberger} dataset.
+\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Seisenberger} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Seisenberger} sample.
+\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Seisenberger} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Ziller} dataset.
+\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Ziller} dataset.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{i | r_{i}}$ and $LOR_{i | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Ziller} sample.
+\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Ziller} sample.
 Only 2-tuples with at least $10\times$ sequencing coverage are used.
@@
 For each region, $r_{i}$, sample $B_{r_{i}}$ from $\beta_{i | r_{i}}$.
-This can be interpreted as $Pr(Z_{r_{i}} = 1)$, the marginal probability that a methylation locus in region $r_{i}$ is methylated, or, alternatively, as the average level of methylation for all methylation loci in the region.
+This can be interpreted as $Pr(Z_{r_{i}} = 1)$, the marginal probability that a methylation locus in region $r_{i}$ is methylated, or, alternatively, as the average level of methylation for all methylation loci in the region[^extreme_B].
 2.
-For every pair of methylation loci, $(i, i + 1)$, $ 1 \leq i \leq (n - 1)$, sample $\psi_{(i, i + 1)}$ from $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}.
+For every pair of methylation loci, $(i, i + 1)$, $1 \leq i \leq (n - 1)$, sample $\Psi_{(i, i + 1)}$ from $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$.
 This can be interpreted as the odds that the $(i + 1)^{th}$ methylation locus has the same methylation state as the $i^{th}$ methylation locus.
 
+[^extreme_B]: If $B_{r_{i}} = 0$ (resp.
+$B_{r_{i}} = 1$), `methsim` adds (resp.
+subtracts) a small value $\epsilon$.
+The default value is $\epsilon = 0.01$.
+The boundary values of $0$ and $1$ are common in $\beta_{i | r_{i}}$ and, were these retained, would result in methylomes with large regions that are uniformly unmethylated (resp.
+methylated).
+
 The second step requires further explanation.
-If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\psi_{(i, i + 1)}$ from a Gaussian distribution with mean equal to __TODO: What?__ and standard deviation equal to __TODO: What?__.
+If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution with mean equal to __TODO: What?__ and standard deviation equal to __TODO: What?__.
 __TODO: Explain how these means and standard deviations are estimated using loess.__ By default, if there are less than __TOOD: how many?__ $LOR$ with the given $IPD(i, i + 1), r_{i}$ combination then the Normal-approximation is used.
@@
 
-Now that we have $B_{r_{i}}$ and $\psi_{i, i + 1}$ for $1 \leq i \leq n - 1$ we can simulate realisations of $\mathbf{Z_{\tild{h}}}$.
+Now that we have $B_{r_{i}}$ and $\Psi_{i, i + 1}$ for $1 \leq i \leq n - 1$ we can simulate realisations of $\mathbf{Z_{\tilde{h}}}$.
 To begin the process (i.e., when $i = 1$) $Z_{\tilde{h}, 1}$, is simulated from the "marginal" distribution, which is Bernoulli($B_{r_{1}}$).
 Then, we loop over $i = 2, \ldots, n - 1$ to simulate the remainder of the process.
-At each step we compute the transition probability, $p_{i, i + 1} = Pr(Z_{\tilde{h}, i + 1} = 1 | Z_{\tilde{h}, i}) = \frac{Pr(Z_{tilde{h}, i}, Z_{tilde{h}, i + 1} = 1)}{Pr(Z_{tilde{h}})}$, from $B_{r_{i}}, $B_{r_{i + 1}}, \psi_{(i, i + 1)}$ and the observed (simulated) value at the previous iteration, $Z_{\tilde{h}, i}$.
-Then, $Z_{\tilde{h}, i + 1}$ is simulated from the Bernoulli($p_{i, i + 1}$) distribution.
+At each step we compute the transition probability, $p_{(i + 1) | i} = Pr(Z_{\tilde{h}, i + 1} = 1 | Z_{\tilde{h}, i}) = \frac{Pr(Z_{tilde{h}, i}, Z_{tilde{h}, i + 1} = 1)}{Pr(Z_{tilde{h}})}$, from $B_{r_{i}}, B_{r_{i + 1}}, \Psi_{(i, i + 1)}$ and the observed (simulated) value at the previous iteration, $Z_{\tilde{h}, i}$.
+Then, $Z_{\tilde{h}, i + 1}$ is simulated from the Bernoulli($p_{(i + 1) | i}$) distribution.
 
-To compute the transition probability, $p_{i, i + 1}$, `methsim` uses the iterative proportional fitting algorithm (IPF; __TODO: CITE IPD paper__) to construct the joint probability matrix, __TODO: Mathematically describe this matrix__.
-Then, the appropriate element of this matrix is divided by the appropriate marginal probability to obtain the transition probability.
-An example of this procedure is illustrative.
+In order to compute the transition probability, $p_{(i + 1) | i}$, `methsim` first uses the iterative proportional fitting algorithm \citet[see]{Speed:2005tg}[for an overview of the algorithm] to construct the joint probability matrix,
 
-The IPD procedure is a general method for __TODO: blah blah__.
-In this case we wish to use the IPF procedure to construct the unique $2 \times 2$ array (the joint probability matrix) with specified margins (the marginal probabilities of each methylation state at each locus) and the specified cross-product ratio (the odds-ratio).
-__UP TO HERE__
+\begin{equation*}
+P = \bordermatrix{~ & Pr(Z_{i+1} = 0) = 1 - B_{r_{i+1}} & Pr(Z_{i+1} = 1) = B_{r_{i+1}} \cr
+Pr(Z_{i} = 0) = 1 - B_{r_{i}} & Pr(Z_{i} = 0, Z_{i + 1} = 0) & Pr(Z_{i} = 0, Z_{i + 1} = 1) \cr
+Pr(Z_{i} = 1) = B_{r_{i}} & Pr(Z_{i} = 1, Z_{i + 1} = 0) & Pr(Z_{i} = 1, Z_{i + 1} = 1) \cr}
+\end{equation*}
 
-\begin{equation}
-% An example of IPF being used to form a transition matrix.
-\end{equation}
+Iterative proportional fitting is a general method "for constructing tables of numbers satisfying certain constraints" \citep{Speed:2005tg}.
+`methsim` uses iterative proportional fitting to construct the _unique_ $2 \times 2$ array (the joint probability matrix) with specified margins (the marginal probabilities of each methylation state at each locus, $B_{r_{i}} = Pr(Z_{i} = 1), B_{r_{i + 1}} = Pr(Z_{i + 1} = 1)$) and the specified cross-product ratio (the odds-ratio, $\psi_{i, i + 1} = \log_{2}(\Psi_{i, i + 1})$).
 
-To simulate $Z_{h, j}$ the algorithm proceeds as follows (in `R` pseudocode):
+Once $P$ is formed, the appropriate element of it is divided by the appropriate marginal probability to obtain the transition probability. An example of this procedure is illustrative.
 
-```{r, echo = TRUE, eval = FALSE}
-n <- length(z)
-u <- runif(n, 0, 1)
-# Sample from region-specific beta-value distributions
-b <- sample(BETA, n)
-# Sample from region-specific LOR distributions
-l <- sample(LOR, n - 1)
+Let $\Psi_{i, i + 1} = 2$, and therefore $\psi = 2^{2} = 4$, and $B_{r_{i}} = B_{r_{i + 1}} = 0.7$.
+To begin the iterative proportional fitting algorithm, form the matrix $P^{(0)} = \Bigl(\begin{smallmatrix}
+\psi & 1 \\
+   1 & 1
+\end{smallmatrix} \Bigr).$ This matrix has the desired cross-ratio, $\psi$, but not the desired row and column margins.
+At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi$ while the row and column margins converge towards the desired values.
+The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P^{(*)} \equiv P$.
+For our example, $P^{(*)} = \Bigl(\begin{smallmatrix}
+0.122 & 0.178 \\
+0.178 & 0.522
+\end{smallmatrix} \Bigr)$, to 3 decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.
 
-# Sample the first observation from the "stationary/marginal" distribution
-if (u[1] < b[1]) {
-  z[1] <- 0
-} else {
-  z[1] <- 1
-}
+In `methsim`, the routine to simulate $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$ is available as the `simulate,SimulateMethylomeParam-method`.
+It is mostly written in C++ using the wonderful Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time.
 
-# Sample the rest of the z
-for (i in 2:n) {
-  if (z[i] == 0) {
-    # Compute transition probability
-    # Compute z[i]
-
-  } else {
-    # Compute transition probability
-    # Compute z[i]
-  }
-}
-```
-
-\subsubsection{Simulating a single sample}
+\subsubsection{Simulating $\mathbf{W}$}
 
@@
 
-Given a set of transition matrices, $\Big \{ P_{i^{(h, j)}} \Big \}$, we can readily simulate a realisation of the methylation states along a haploid genome, $Z_{h, j}$.
-Naively, we might simulate a single sample by simply generating a realisation for each $h$ of $Z_{h, j}$.
-However, although there is cell-to-cell heterogeneity in the methylation patterns, there is still a degree of dependence for the $i^{th}$ methylation state across the $h$ haploid genomes in the $j^{th}$ sample.
-After all, the cells are from the same individual and are broadly of the same "type".
+For each region we need to assign the vector of weights, $\mathbf{W_{r_{i}, \cdot}}$.
+To do so, `methsim` uses a heuristic procedure.
+For higher-order m-tuples, $m = 3$ or $4$ is recommended, `methsim` computes the relative frequency of the $2^{m}$ possible methylation patterns at each m-tuple with sufficient sequencing coverage.
+`methsim` then samples from this distribution, stratified by region-type; call these empirical distributions
 
-What this means is that we want some degree of dependence between the set of transition matrices for haploid methylome $h$, $\Big \{ P_{i^{(h, j)}} \Big \}$, and for those for haploid methylome $h'$, $\Big \{ P_{i^{(h', j)}} \Big \}$.
-This is asking quite a lot of our simple simulation model.
-Instead, `methsim` uses a heuristic procedure to achieve a similar result.
+__TODO__: Really computing (and plotting) $w_{i | r_{(i, \ldots, i + m - 1)}}, \ldots, w_{i + m - 1 | r_{(i, \ldots, i + m - 1)}}$ not $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$.
+These empirical distributions are shown for the \emph{ADS} sample in Figure \ref{fig:ADS_pattern_freqs_by_pm_boxplots}.
+From this we see that __TODO__.
 
-To begin, we assume that all $H$ haploid genomes in the sample have the same co-methylation structure.
-That is equivalent to defining $\Big \{ P_{i^{(h, j)}} \Big \} := \Big \{ P_{i^{(j)}} \Big \}$ for all $h$.
-__TODO: Comment on this assumption__
+__TODO: Update figures to use $w_{1}, \ldots, w_{8}$ rather than $h1, \ldots h8$.__
 
-Naively, we might simulate a sample by simulating $H$ realisations of the Markov chain with transition matrices $\Big \{ P_{i^{(j)}} \Big \}$.
-However, this gives no control over the dependence between the methylation states at the $i^{th}$ locus over the $H$ haploid genomes in the $j^{th}$ sample.
-In fact, such a model will have the "correct" co-methylation structure but "incorrect" $\beta$-values.
-__TODO: Check this and demonstrate using `methsim`__
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/ADS_pattern_freqs_by_pm_boxplots.pdf}
+\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{ADS} sample.
+Only CpGs with at least $5\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:ADS_beta_by_pm_boxplots}
+\end{figure}
 
-Instead, `methsim` simulates $R << H$ realisations of the Markov chain with transition probabilities given by $\Big \{ P_{i^{(j)}} \Big \}$.
-It then weights these realisations according to the observed heterogeneity for each region.
-__TODO: Describe the inuition of this weighting scheme as being that $w_{r}$ of the cells have identical methylation states for that region.
+__TODO: Limitations of this procedure__
 

### 2015-03-25 Rmarkdown/methsim.Rmd
Section on simulating W.

@@
 
-These same distributions for all the _Ziller_, _Lister_, _Seisenberger_ and _Ziller_ datasets are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
+These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
 \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
@@
 For higher-order m-tuples, $m = 3$ or $4$ is recommended, `methsim` computes the relative frequency of the $2^{m}$ possible methylation patterns at each m-tuple with sufficient sequencing coverage.
-`methsim` then samples from this distribution, stratified by region-type; call these empirical distributions
-
-__TODO__: Really computing (and plotting) $w_{i | r_{(i, \ldots, i + m - 1)}}, \ldots, w_{i + m - 1 | r_{(i, \ldots, i + m - 1)}}$ not $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$.
-These empirical distributions are shown for the \emph{ADS} sample in Figure \ref{fig:ADS_pattern_freqs_by_pm_boxplots}.
-From this we see that __TODO__.
-
-__TODO: Update figures to use $w_{1}, \ldots, w_{8}$ rather than $h1, \ldots h8$.__
+`methsim` then samples from this distribution, stratified by region-type; call these empirical distributions $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$.
+Figure \ref{fig:ADS_pattern_freqs_by_pm_boxplots} shows elementwise boxplots of $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$, that is, it shows the distributions $w_{i | r_{(i, \ldots, i + m - 1)}}$ for $i = 1, \ldots, 2^{m}$.
 
@@
 \includegraphics[width=\textwidth]{../figures/ADS_pattern_freqs_by_pm_boxplots.pdf}
-\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{ADS} sample.
+\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{ADS} sample.
 Only CpGs with at least $5\times$ sequencing coverage are used.
 UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:ADS_beta_by_pm_boxplots}
+\label{fig:ADS_pattern_freqs_by_pm_boxplots}
 \end{figure}
 
-__TODO: Limitations of this procedure__
+These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and  \ref{fig:Ziller_pattern_freqs_by_pm_boxplots} (__TODO: Are these other examples necessary?__).
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_pattern_freqs_by_pm_boxplots.pdf}
+\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{EPISCOPE} dataset.
+Only CpGs with at least $5\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_pattern_freqs_by_pm_boxplots.pdf}
+\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Lister} dataset.
+Only CpGs with at least $5\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Lister_pattern_freqs_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_pattern_freqs_by_pm_boxplots.pdf}
+\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Seisenberger} dataset.
+Only CpGs with at least $5\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Seisenberger_pattern_freqs_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_pattern_freqs_by_pm_boxplots.pdf}
+\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Ziller} dataset.
+Only CpGs with at least $5\times$ sequencing coverage are used.
+UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
+\label{fig:Ziller_pattern_freqs_by_pm_boxplots}
+\end{figure}
+
+From Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and \ref{fig:Ziller_pattern_freqs_by_pm_boxplots} we see that, unsurprisingly, in UMRs that a single methylation pattern dominates, the completely unmethylated pattern.
+Outside of UMRs, we see that while there is generally a dominant methylation pattern at each m-tuple, there may be $2-4$ otheres that are observed at non-negligible frequencies.
+PMRs are the most heterogeneous and LMRs and other regions are quite similar.
+
+The above procedure has several obvious limitations.
+Firstly, we are restricted to analying m-tuples within individual reads and read-pairs.
+Secondly, as $m$ increases, and similarly as $IPD$ increases, there are fewer m-tuples with sufficient sequencing coverage.
+This means our data are heavily biased towards regions of high CpG density, which typically correspond to UMRs.
+Additionally, we know that m-tuples with small $IPDs$ are far more likely to be co-methylated than those with larger $IPDs$ and so we are biased towards sampling entirely unmethylated and entirely methylated methylation patterns.
+It is difficult to know how to adjust for these biases, which are unavoidable given current sequencing technologies.
+These estimates very likely underestimate the degree of within-sample heterogeneity and, consequently, weights based on sampling from the empirical distributions, $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$, likely result in overly homogeneous simulated methylomes.
+
+A second, more subtle, bias is due to sequencing errors, bisulfite-conversion errors and alignment errors.
+Using the simple relative frequency estimation procedure will result in errors like these being inferred as (rare) methylation patterns.
+`MPFE` ([http://bioconductor.org/packages/release/bioc/html/MPFE.html](http://bioconductor.org/packages/release/bioc/html/MPFE.html)) is an R package that uses an estimation procedure designed to deal with some of these errors.
+However, `MPFE` is intended for analysing amplicon bisulfite-sequencing data with high sequencing coverages, and it's more computationally demanding estimation procedure does not scale to whole-genome data.
+
+With all the above limitations in mind, `methsim` still uses the simple strategy of sampling the $\mathbf{W}$ from $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$.
+At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data.
 

### 2015-03-25 Rmarkdown/methsim.Rmd
Fix dodgy figure path

@@
 
-These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and  \ref{fig:Ziller_pattern_freqs_by_pm_boxplots} (__TODO: Are these other examples necessary?__).
+These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and  \ref{fig:Ziller_merged_pattern_freqs_by_pm_boxplots} (__TODO: Are these other examples necessary?__).
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_pattern_freqs_by_pm_boxplots.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_pattern_freqs_by_pm_boxplots.pdf}
 \caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Ziller} dataset.
@@
 UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Ziller_pattern_freqs_by_pm_boxplots}
+\label{fig:Ziller_merged_pattern_freqs_by_pm_boxplots}
 \end{figure}
 
-From Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and \ref{fig:Ziller_pattern_freqs_by_pm_boxplots} we see that, unsurprisingly, in UMRs that a single methylation pattern dominates, the completely unmethylated pattern.
+From Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and \ref{fig:Ziller_merged_pattern_freqs_by_pm_boxplots} we see that, unsurprisingly, in UMRs that a single methylation pattern dominates, the completely unmethylated pattern.
 Outside of UMRs, we see that while there is generally a dominant methylation pattern at each m-tuple, there may be $2-4$ otheres that are observed at non-negligible frequencies.

### 2015-03-25 Rmarkdown/methsim.Rmd
Describe extrapolating LOR distributions.

@@
 Furthermore, no licence file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code.
-(__TODO: Email the developer, Owen Rackham, asking him to add a licence.
+(__TODO: Update if I received a response.
+I emailed the developer, Owen Rackham, asking him to add a licence.
 He is considering my request.__)
@@
 And it is counterproductive because the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci.
-Furthermore, almost all downstream analyses are _reference-based_ (__TODO: Cite relevant section__), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.
+Furthermore, almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.
 
@@
 
-For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (described below; __TODO: Discuss implications of this assumption somewhere below__).
+For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (discussed below; __TODO: Discuss implications of this assumption somewhere below__).
 
@@
 Identify partially methylated regions.
-A summary statistic, $\alpha$, (__TODO: What is $\alpha$__) which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs.
+A summary statistic, $\alpha$, which is based on the $\beta$-values in a sliding window of 100 CpGs, is used to identify PMRs.
 Briefly, a two-state hidden Markov model is fit to the $\alpha$ values to identify PMRs and non-PMRs.
@@
 The second step requires further explanation.
-If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution with mean equal to __TODO: What?__ and standard deviation equal to __TODO: What?__.
-__TODO: Explain how these means and standard deviations are estimated using loess.__ By default, if there are less than __TOOD: how many?__ $LOR$ with the given $IPD(i, i + 1), r_{i}$ combination then the Normal-approximation is used.
-The user may specify a different method for dealing with these rarer $IPD$-region combinations.
+If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution[^loess_params].
+By default, if there are less than __TOOD: how many?__ $LOR$ with the given combination of region-type and IPD then the Gaussian-approximation is used.
+The user may specify a different method for dealing with these rarer $IPD$-region combinations and, indeed, different methods are explored in the __RESULTS SECTION__.
+
+[^loess_params]: The mean, $\mu$, and standard deviation, $\sigma$, of this Gaussian distribution depend on both $IPD$ and $r_{i}$.
+A loess fit of $mean(LOR)$ (resp.
+$sd(LOR)$) as a function of $IPD$ is performed, stratified by region type.
+The values of $\mu$ (resp.
+$\sigma$) are then obtained by extrapolating the loess fit (__TODO: Update if changes made to `methsim` defaults__).
 
@@
 At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data.
+In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}, \ldots, \frac{1}{n_{\bar{h}})$.
 

### 2015-03-30 Rmarkdown/methsim.Rmd
Clarify how LOR is sampled for rare/impossible region-IPD combinations. Add data for claims made in this commit.

@@
 \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
-\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} (__TODO: Are these other examples necessary?__).
+\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots}.
 
@@
 The second step requires further explanation.
-If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution[^loess_params].
-By default, if there are less than __TOOD: how many?__ $LOR$ with the given combination of region-type and IPD then the Gaussian-approximation is used.
-The user may specify a different method for dealing with these rarer $IPD$-region combinations and, indeed, different methods are explored in the __RESULTS SECTION__.
+If there are no observations with the required $IPD$-region combination, then, by definition, `methsim` is unable to sample from the appropriate empirical distribution[^min_n].
+Instead, when this occurs, `methsim` samples from a user-specified parametric distribution.
+__RESULTS SECTION__ explores different choices of this distribution __AND SHOWS WHAT__.
+The default option is a zero-mean Gaussian distribution, chosen so that "on average" the methylation state at adjacent CpGs with large $IPD$s are independent of one another.
+The default standard deviation of this Gaussian distribution is $2$; this value, while fairly arbitrary, reflects the median standard deviation of the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions for each sample, which is remarkably constant across $r_{i}$ and samples[^ipd_stability] (__TODO: Format and reference table__).
 
-[^loess_params]: The mean, $\mu$, and standard deviation, $\sigma$, of this Gaussian distribution depend on both $IPD$ and $r_{i}$.
-A loess fit of $mean(LOR)$ (resp.
-$sd(LOR)$) as a function of $IPD$ is performed, stratified by region type.
-The values of $\mu$ (resp.
-$\sigma$) are then obtained by extrapolating the loess fit (__TODO: Update if changes made to `methsim` defaults__).
+[^min_n]: By default, `methsim` requires at least 100 observations in the  $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ to try to avoid sampling from sparse empirical distributions.
+
+[^ipd_stability]: Figures \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_lor_by_pm_quantile_plots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} show that the variability of these distributions is fairly constant across $IPD$ for a given $r_{i}$ and sample.
+
+```{r, eval = TRUE, echo = FALSE}
+knitr::kable(readRDS("../data/median_sd_LOR_distributions.rds"), align = "c", digits = 2)
+```
 
@@
 At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data.
-In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}, \ldots, \frac{1}{n_{\bar{h}})$.
+In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}}, \ldots, \frac{1}{n_{\bar{h}}})$.
 

### 2015-04-03 Rmarkdown/methsim.Rmd
Section on simulating reads, i.e., methsim::simulate,SimulateBSParam-method.

@@
 
-\subsection{Simulating the truth}
+\subsection{Simulating a single methylome}
 
@@
 
-\subsubsection{Simulating multiple samples from the same group}
+\subsection{Simulating reads}
 
-- [ ] DONE
+- [x] DONE
+
+Once we have our 'true' methylome, we want to simulate an assay of this sample.
+In theory this could be any type of methylation assay, but here we focus on simulating whole-genome bisulfite-sequencing data.
+
+`methsim` uses a simple method for simulating bisulfite-sequencing.
+The user specifies the desired read length and average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
+
+The number of reads required is given by $n_{reads} = \frac{\text{average sequencing coverage}}{\text{read length}} \times \text{size of genome}$.
+The number of reads per chromosome is assigned proportional to the chromosome length.
+Then, the start of each read is sampled uniformly across the chromosome respective chromosome.
+
+[^read_length]: Currently only single-end data are supported.
+This isn't a big issue.
+Most paired-end bisulfite-sequencing datasets have overlapping mates, and so paired-end data can be crudely simulated by simply doubling the read length.
+Also, all reads must have the same length, i.e., no simulation of read trimming.
+I don't consider this feature a priority, but it could easily be implemented by generating read lengths from a probability distribution, e.g., Poisson.
+
+For all reads that overlap a methylation locus in the 'true' methylome, a pseudo-haplotype is sampled with probabilities given by $W_{i, \cdot}$.
+The methylation pattern of the sampled pseudo-haplotype is then copied, with probability $\epsilon$ that the methylation state at each locus is switched.
+
+The result is an object storing the genomic co-ordinates and methylation patterns of every read that contains a methylation locus.
 

### 2015-04-03 Rmarkdown/methsim.Rmd
Section on methsim::asMethPat()

@@
 
+\subsection{Construct the output}
+
+- [x] DONE
+
+While the `SimulatedBS` object contains all the data on the methylation states of the reads, it is not always the most convenient or efficient format with which to work.
+For example, many downstream analyses only make use of $\beta$-values, so we might want to summarise our simulated data this way.
+`methsim` provides the `asMethPat()` function to convert the `SimulatedBS` object to a `MethPat` object containing methylation patterns at m-tuples of a given size; all the functionality of the `MethylationTuples` package is then available to the user, such as computing $\beta$-values with the `methLevel()` method.
+
 \subsubsection{Simulating differential methylation}

### 2015-04-03 Rmarkdown/methsim.Rmd
Section on simulating multiple samples.

@@
 
+\subsection{Simulating multiple samples}
+
+- [x] DONE
+
+The above describes how to simulate a single whole-genome bisulfite-sequencing sample.
+If we wanted to simulate six independent samples then we could simply run this procedure six times, with a different `SimulateMethylomeParam` object for each realisation.
+However, we will generally be interested in simulating experiments where the samples are not  is some relationship amongst the samples.
+For example, suppose we want to simulate a two-group experiment with three samples per group.
+Further, suppose we want to include differentially methylated regions between the two groups.
+How do we simulate such an experiment?
+
+To simulate these type of experiments, we need to be able to control the variation between the `SimulatedBS` objects.
+In this example, we probably want the three samples in each group to be more similar to each other than to the three samples in the other group.
+Essentially, we need to be able to control the within-group and between-group variation.
+`methsim` controls this variation through the workflow show in __FIGURE__.
+If we want two samples two be very different from one another, then we would use two different `SimulateMethylomeParam` objects at the beginning.
+However, if we want to simulate two technical replicates, then we would use the same `SimulateMethylomeParam`, `SimulatedMethylome` and `SimulateBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.
+
+We might also consider more subtle ways of introducing variation.
+For example, we might use the same partitioning of the methylome and the same average methylation level for each region but vary the strength of co-methylation.
+This could be achieved by altering the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions between two `SimulateMethylomeParam` objects while holding the rest constant.
+Or we might vary the weights, $W$, to simulate more or less heterogeneous samples.
+
+In __RESULTS SECTION__ I explore how much variation is created using these different techniques.
+
 \subsubsection{Simulating differential methylation}

### 2015-04-03 Rmarkdown/methsim.Rmd
Scattered thoughts on simulating DMRs

@@
 
-\subsection{Parameter estimation}
+An obvious application of `methsim` is to simulate data to be used in a study comparing methods for identifying differential methylation.
+This requires a particularly controllable method for inducing differences in `SimulatedMethylome` objects.
+The user wants to provide a set of methylation loci, either individual loci or as regions, and the expected difference in methylation levels at these regions.
+
+__Scattered thoughts below__
+
+1. Individual loci: Want to specify $B_{i} = \sum_{\tilde{h} = 1}^{n_{\tilde{h}}} W_{r_{i}, \tilde{h}} \times Z_{i, \tilde{h}}$, i.e., fix $\mathbf{W_{i, \cdot}}$ and $\mathbf{Z_{i, \cdot}}$.
+2.
+Regionwise: Want to specify $mean(B_{r_{i}})$ (i.e., average $B$ across all loci in $r_{i}$).
+Could do this by applying (1) to all loci in the region or by specifying $B_{r_{i}}$ and hoping it just works out in the simulation.
+
+Perhaps both could be achieved by holding $Z$ fixed and varying $W$?
+Is that like assuming that the heterogeneity is causing the DMRs (or will it be interpreted that way)?
+Would tweaking $W$ to guarantee a particular $B$ be a computationally feasible strategy?
+Or would it be easier to specify the desired $B$ and simulate the region (this wouldn't guarantee $B$ is the desired value but should equal it "on average")?
 

### 2015-04-03 Rmarkdown/methsim.Rmd
Some stuff on the implementation in meths.

@@
 
-- [ ] DONE
-
 \subsection{Implementation}
@@
 - [ ] DONE
+
+`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as key R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list.
+__TODO: Add citations of all packages used in thesis__).
+
+`methsim` makes extensive use of the S4 object system.
+The most important classes defined by `methsim` are `SimulateMethylomeParam`, `SimulatedMethylome`, `SimulateBSParam` and `SimulatedBS`.
+The most important methods are the `simulate()` methods defined for the `SimulateMethylomeParam` and `SimulateBSParam` classes.
+__FIGURE__ illustrates the `methsim` workflow to create simulate whole-genome bisulfite-sequencing data.
+
+Because `methsim` simulates data by sampling from empirical distributions of key statistics, it requires that the user supplies data from a relevant sample.
+To help a new user get started, `methsim` includes data for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples in the form of `SimulateMethylomeParam` objects.
+Alternatively, the user may process the BAM file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `SimulateMethylomeParam` object.
+
+The next step is to simulate a 'true' methylome by applying the `simulate()` method to the `SimulateMethylomeParam` object.
+This returns a `SimulatedMethylome`, a schematic of which is __SHOWN BELOW__.
 

### 2015-04-06 Rmarkdown/methsim.Rmd
Complete describing implementation of meths

@@
 
-- [ ] DONE
+- [x] DONE
 
-`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as key R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list.
+`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as several R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list.
 __TODO: Add citations of all packages used in thesis__).
@@
 This returns a `SimulatedMethylome`, a schematic of which is __SHOWN BELOW__.
+The `SimulatedMethylome` object forms part of the `SimulateBSParam` object from which a `SimualtedMethylome` is created by calling the `simulate()` method.
 

### 2015-04-06 Rmarkdown/methsim.Rmd
Add performance of methsim

@@
 
+\subsubsection{Performance}
+
+- [x] DONE
+
+Much of `methsim` is written in C++, using the wonderful Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures.
+Furthermore, several steps of the simulation can be run in parallel for each chromosome.
+Parallelism is implemented via the `BiocParallel` Bioconductor package \citep{BiocParallel}.
+
+The `SimulatedMethylome` and `SimulatedBS` objects are large in memory.
+This is to some extent unavoidable; we are simulating entire methylomes and whole-genome sequencing experiments.
+Some improvements in memory usage are possible by alternative implementations of key classes, however, these are generally at the expense of more complicated code and longer running times of associated methods.
+For example, long and repetitive vectors may be efficiently represented by run length encodings, however, it is slower to extract elements from these than simple numeric vectors.
+
+Shown below are the running times and memory usage involved in simulating a $30\times$ whole-genome bisulfite-sequencing sample using the _ADS_ `SimulateMethylomeParam` object included in `methsim`.
+Using eight CPU cores in parallel, it takes approximately seven minutes to simulate the underlying 'true' methylome and a further nine minutes to simulate a single whole-genome bisulfite-sequencing assay of this methylome.
+
+These results are indicative of simulating high-coverage whole-genome bisulfite-sequencing for human-sized genomes.
+Simulations using a smaller genome or lower sequencing coverage will have faster running times and lower memory usage.
+
+```r
+# Load necessary packages
+suppressPackageStartupMessages(library(methsim))
+suppressPackageStartupMessages(library(BiocParallel))
+suppressPackageStartupMessages(library(pryr))
+
+# Use up to 8 cores in parallel computations.
+BPPARAM <- MulticoreParam(workers = 8L)
+
+# Set seed for reproducibility.
+set.seed(666)
+
+# ADS: the SimulateMethylomeParam object.
+pryr::object_size(ADS)
+#> 22.5 MB
+
+# sm: the SimulatedMethylome object
+system.time(sm <- simulate(ADS, BPPARAM = BPPARAM))
+#>    user  system elapsed
+#> 839.505  19.612 417.271
+pryr::object_size(sm)
+#> 2.93 GB
+
+# sbsp: the SimulatedBSParam object.
+# NOTE: Basically all the memory used by the SimulatedBSParam object is the
+# SimulatedMethylome object.
+sbsp <- SimulateBSParam(SimulatedMethylome = sm,
+                        aveCov = 30,
+                        errorRate = 0.01,
+                        target = NULL)
+pryr::object_size(sbsp)
+#> 2.93 GB
+
+# sbs: the SimulatedBS object.
+system.time(sbs <- simulate(sbsp, BPPARAM = BPPARAM))
+#>     user   system  elapsed
+#> 2716.503  174.591  548.122
+pryr::object_size(sbs)
+#> 10.1 GB
+
+# methpat: the MethylationTuples::MethPat object of one-tuples.
+# NOTE: information is lost when a SimulatedBS object is coerced to a MethPat
+# object.
+system.time(methpat <- asMethPat(sbs,
+                                 sampleName = "sim_ADS",
+                                 size = 1L,
+                                 BPPARAM = BPPARAM))
+#>    user  system elapsed
+#> 557.900  34.779  48.944
+pryr::object_size(methpat)
+#> 451 MB
+```
+
 \section{Results}

### 2015-04-06 Rmarkdown/methsim.Rmd
Begin methsim results section

@@
 -  __TODO: Figure showing 'rest of genome' is MMRs. More generally, a figure showing the average methylation level in each of `UMR`, `LMR`, `PMR` and `other`.
+- [ ] DONE
 
-\subsection{Case study}
+A central pillar of `methsim` is that many of the simulation parameters are sampled from real data.
+At a bare minimum, the first test of `methsim` is that simulated data are similar to the real data on which they are based.
+To assess this, I compare several summary measures between the simulated data and the real data.
+The simplest of these are summaries are the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`.
+In contrast, the spatial correlation of $\beta$-values is not explicitly modelled in `methsim`; rather, we hope that `methsim` recapitulates this phenomenon by appropriate modelling of within-fragment co-methylation and choice of weights, $\mathbf{W}$.
+
+Once it is established that `methsim` passes this minimal test, I use `methsim` to explore models of within-fragment co-methylation, the choice of the weights, $\mathbf{W}$, and their respective effects on the summary measures in order to identify areas of improvement in `methsim`.
+Finally, I demonstrate how `methsim` might be used to simulate data for use in a study comparing methods for identifying differential methylation.
+
+\subsection{Summary measures of simulated data}
+
+- [ ] DONE
+
+\subsection{Models of within-fragment co-methylation}
+
+- [ ] DONE
+
+\subsection{The choice of $\mathbf{W}$}
+
+\subsection{Simulating differential methylation}
 

### 2015-04-06 Rmarkdown/methsim.Rmd
Various methsim TODOs and rejigging

@@
 
+__TODO: Figure showing `other` is MMRs. More generally, a figure showing the average methylation level in each of `UMR`, `LMR`, `PMR` and `other`.__
+
 \subsubsection{Simulating $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$}
@@
 \end{smallmatrix} \Bigr)$, to 3 decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.
-
-In `methsim`, the routine to simulate $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$ is available as the `simulate,SimulateMethylomeParam-method`.
-It is mostly written in C++ using the wonderful Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time.
 
@@
 
--  __TODO: Figure showing 'rest of genome' is MMRs. More generally, a figure showing the average methylation level in each of `UMR`, `LMR`, `PMR` and `other`.
 - [ ] DONE
@@
 
+__TODO: Work the comparison to other methods into my conclusions__
+
 \citet{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).

### 2015-04-06 Rmarkdown/methsim.Rmd
Began re-organising Methods.

@@
 
-1. Simulate the true methylome of each sample.
-2. Simulate reads, including sequencing error and bisulfite-conversion error, by sampling from the true methylome.
-3. Construct the output, be it reads, methylation patterns at m-tuples or $\beta$-values.
+1. Simulating the true methylome of each sample.
+2. Simulating reads, including sequencing error and bisulfite-conversion error, by sampling from the true methylome.
+3. Constructing the output, be it reads, methylation patterns at m-tuples or $\beta$-values.
 
@@
 For the input dataset, `methsim` requires the methylation patterns at various sized m-tuples, which can be produced by the `methtuple` software.
+
+This workflow is illustrated in __FIGURE__ and described in the remainder of this section.
+
+\subsection{Implementation}
+
+- [x] DONE
+
+`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as several R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list.
+__TODO: Add citations of all packages used in thesis__).
+
+`methsim` makes extensive use of the S4 object system in R.
+The most important classes defined by `methsim` are `SimulateMethylomeParam`, `SimulatedMethylome`, `SimulateBSParam` and `SimulatedBS`.
+The most important methods are the `simulate()` methods defined for the `SimulateMethylomeParam` and `SimulateBSParam` classes.
+The role of these classes and methods in the workflow is illustrated in __FIGURE__.
+
+The `SimulateMethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled.
+Therefore, the `SimulateMethylomeParam` should be based on a relevant sample.
+To help a new user get started, `methsim` includes `SimulateMethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset.
+Alternatively, the user may process the BAM file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `SimulateMethylomeParam` object based on their sample of interest.
+
+To simulate a 'true' methylome, the user runs the `simulate()` method on the `SimulateMethylomeParam` object. This returns a `SimulatedMethylome` object.
+
+To simulate bisulfite-sequencing reads, the user first constructs a `SimulateBSParam`, which includes the `SimulatedMethylome`.
+The user then runs the `simulate()` method on the `SimulatedBSParam` object to create a simulated bisulfite-sequencing data in the form of a `SimulatedBS` object.
+
+Finally, we may wish to simplify the `SimulatedBS` object.
+For example, we may only require the $\beta$-values for our downstream analysis and not the  read-level data.
+`methsim` includes the `asMethPat()` function to convert the `SimulatedBS` object to a `MethPat` object (described in __CHAPTER__).
+
+__UP TO HERE: See if the remainder of this section needs re-jigging in light of the above changes to Implementation__
 
@@
 As we have seen, DNA methylation is highly heterogeneous along the genome.
-Nonetheless, there are clear regions of "similarity", regions where we might hope to model it by a simple parametric model.
-`methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together.
+Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains.
+In these locally-similar regions we might hope to model DNA methylation by a simple parametric model.
+`methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together to form the 'true' methylome.
+
 The idea of segmenting a globally heterogenous stochastic process into a series of locally homogeneous processes is not new.
@@
 
-\subsection{Implementation}
-
-- [x] DONE
-
-`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as several R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list.
-__TODO: Add citations of all packages used in thesis__).
-
-`methsim` makes extensive use of the S4 object system.
-The most important classes defined by `methsim` are `SimulateMethylomeParam`, `SimulatedMethylome`, `SimulateBSParam` and `SimulatedBS`.
-The most important methods are the `simulate()` methods defined for the `SimulateMethylomeParam` and `SimulateBSParam` classes.
-__FIGURE__ illustrates the `methsim` workflow to create simulate whole-genome bisulfite-sequencing data.
-
-Because `methsim` simulates data by sampling from empirical distributions of key statistics, it requires that the user supplies data from a relevant sample.
-To help a new user get started, `methsim` includes data for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples in the form of `SimulateMethylomeParam` objects.
-Alternatively, the user may process the BAM file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `SimulateMethylomeParam` object.
-
-The next step is to simulate a 'true' methylome by applying the `simulate()` method to the `SimulateMethylomeParam` object.
-This returns a `SimulatedMethylome`, a schematic of which is __SHOWN BELOW__.
-The `SimulatedMethylome` object forms part of the `SimulateBSParam` object from which a `SimualtedMethylome` is created by calling the `simulate()` method.
-
 \subsubsection{Performance}

### 2015-04-12 Rmarkdown/biology_background.Rmd
Add my edits to Biology Background.

@@
 In this chapter I summarise the basic biology necessary for
-understanding my thesis. I introduce DNA methylation, describe some of its biological roles and assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
+understanding my thesis. I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
 
@@
 Chromatin compactly packages the DNA into _chromosomes_, so that the organism's complete nuclear DNA, its nuclear genome, might fit into the nucleus of its cells[^extra_nuclear_genomes].
-
-The material in this chapter can be found in any good introductory molecular biology text, such as \citet{Alberts:2007vj}.
 
@@
 In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_.
-One strand is called the 5' strand (pronounced "5 prime strand") - also known as the sense strand, Crick strand or top strand - and the other strand called the 3' strand (pronounced "3 prime strand") - also known as the antisense strand, Watson strand or bottom strand.
+One strand is called the 5' strand (pronounced "5 prime strand" and also known as the sense strand, Crick strand or top strand) and the other strand called the 3' strand (pronounced "3 prime strand" and also known as the antisense strand, Watson strand or bottom strand).
 Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_).
@@
 DNA is able to self-replicate.
-This means that, for example, eukaryotic cells created during _mitosis_ contain the same DNA as the 'parent' cell.
+This means that eukaryotic cells created during _mitosis_ contain the same DNA as the 'parent' cell.
 During DNA replication, the two strands are separated and each strand's _complementary_ DNA sequence is copied by an enzyme called _DNA polymerase_.
@@
 Each of these histones has a 'tail' consisting of a string of amino acids.
-These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \citep{Alberts:2007vj} and are discussed in Section \ref{sec:epigenetics} on epigenetics.
+These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \citep{Alberts:2007vj}.
+Histone modifications are discussed in Section \ref{sec:epigenetics}.
 
 The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin.
-The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
+The linker sequences are between $20--60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
 
-Chromatin is frequently described as either _heterochromatin_ or _euchromatin_.
-Heterochromatin, or closed chromatin, is more tightly packed than euchromatin, or open chromatin.
+Chromatin is frequently described as either 'closed' or 'open'.
+Closed chromatin, _heterochromatin_, is more tightly packed than the open _euchromatin_.
 Heterochromatin is associated with transcriptionally repressed regions of the genome because the machinery required to translate DNA to RNA is unable to physically access the DNA.
@@
 The same region of an individual's genome may vary between heterochromatin and euchromatin states at different stages of the organism's life and in different cells of the organism.
-This is one mechanism which by gene expression is regulated.
+This is one mechanism by which gene expression is regulated.
 
@@
 While eukaryotic genomes share the above-described features, and indeed share many regions of common DNA sequence, eukaryotic genomes come in many shapes and sizes.
-I only discuss the genomes of three organisms relevant to my thesis: Human (_Homo sapiens_), Mouse (_Mus musculus_) and Arabidopsis (_Arabidopsis thaliana_).
+I only discuss the genomes of two species relevant to my thesis: _Homo sapiens_ (human) and _Mus musculus_ (house mouse).
 
-\subsubsection{Human genome}\label{sec:human_genome}
+\subsubsection{The human genome}\label{sec:human_genome}
 
@@
 
-\subsubsection{Mouse genome}\label{sec:mouse_genome}
+\subsubsection{The mouse genome}\label{sec:mouse_genome}
 
@@
 This reduced genetic variability, and the control that researchers have over it, make these mice a very powerful tool in identifying the biological cause of a diverse range of phenotypes.
-
-\subsubsection{Arabidopsis thaliana genome}\label{sec:arabidopsis_genome}
-
-Arabidopsis is a small flowering plant.
-The most widely studied species is _A.
-thaliana_, which is diploid and has one of the smallest genomes known in the plant kingdom at 135 million base pairs long, which is packaged as five pairs of chromosomes ([http://plants.ensembl.org/Arabidopsis_thaliana/Info/Index](http://plants.ensembl.org/Arabidopsis_thaliana/Info/Index)).
-The small size of its genome, coupled with its small physical size, rapid development and its tolerance to both cross-pollination and self-pollination[^cross_and_self_pollination], make _A.
-thaliana_ an ideal organism for studying plant development.
-
-[^cross_and_self_pollination]: The advantage in genetic studies of using a plant that is tolerant of self-pollination is that recessive mutations can be quickly made homozygous and thus expressed.
-Cross-pollination allows the creation of new strains with multiple mutations.
 
@@
 
-[^indels]: An indel is a short insertion or deletion of DNA sequence in an organism, usually relative to a reference genome.
+[^indels]: An example of a larger genetic variant is an _indel_, which is a short insertion or deletion of sequence in an organism's DNA, usually with respect to a reference genome.
 
@@
 
-_Epigenetics_, which I describe in the next section, plays a role in determining both the between-individual phenotypic variation as well as the within-individual phenotypic variation.
+_Epigenetics_, which I describe in the next section, plays a role in determining both the between-individual phenotypic variation, as well as the within-individual phenotypic variation.
+DNA methylation, the focus of my thesis, is the prototypical and most well-studied epigenetic modification.
 
@@
 
-_Epigenetics_ is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ...
-it means just what I choose it to mean - neither more nor less" \citep{carroll1897through}.
-I am sure to continue in this grand tradition.
+Interest in epigenetics has grown remarkably in recent years.
+However, _epigenetics_ is also a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, \dots it means just what I choose it to mean --- neither more nor less" \citep{carroll1897through}.
+As noted by \citet{Deans:2015kt}, "the unfortunate fact is that the increased use of the term _epigenetics_ is likely due more to inconsistencies in its definition than to a consensus of interest among scientists".
 
@@
 
-A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs - epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \citep[p1]{russo1996epigenetic}.
+A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs --- epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \citep[pp.
+1]{russo1996epigenetic}.
 The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence.
@@
 
-Sir Adrian Bird, an esteemed British geneticist, and someone who is in favour of the broader definition, attempts to unite these definitions in the following \citep{Bird:2007em}:
+Sir Adrian Bird, an esteemed British geneticist, attempts to unite these definitions \citep{Bird:2007em}:
 
 > "The following could be a unifying definition of epigenetic events: the structural adaptation of chromosomal regions so as to register, signal or perpetuate altered activity states.
-This definition is inclusive of chromosomal marks [e.g.
-histone marks], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify."
+This definition is inclusive of chromosomal marks [e.g., histone modifications], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify."
 
-Regardless of which definition you subscribe to, an _epigenetic mark_ is the modification that causes this "epigenetic change".
-Actually, "causes" may be too strong a phrase, as much of current epigenetics research is in identifying associations rather than causations, and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.
+Regardless of which definition you subscribe to, an _epigenetic mark_ is the modification that causes this 'epigenetic change'.
+In fact, 'causes' may be too strong a claim, as much of current epigenetics research is in identifying associations rather than causations, and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.
 
@@
 In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual.
-Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual as well will see in later chapters.
+Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual (see Chapter \ref{chap:co-methylation}).
 _Epigenomics_ is the study of the epigenome, analogous to genomics being the study of the genome.
@@
 
-One mark that most authors agree is an epigenetic mark is DNA methylation, which I describe in the next section and whose study is the focus of my thesis.
+One mark that most authors agree is an epigenetic mark is DNA methylation, which I describe in the next section and the study of which is the focus of my thesis.
 
@@
 It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying].
-It therefore fits into Arthur Riggs' aforementioned definition of an epigenetic modification \citep[p1]{russo1996epigenetic}.
+It therefore fits into Arthur Riggs' aforementioned definition of an epigenetic modification.
 DNA methylation is found in bacteria, fungi, plants and animals.
@@
 A major reason that DNA methylation is studied is that it is essential for normal development in mammals \citep{Li:1992bp}.
-It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \citep{Razin:1991vj}, silencing of transposable elements\citep{Jones:2001eg}, X chromosome inactivation \citep{Mohandas:1981uf} and tumorigenesis \citep{Ehrlich:2002kz}.
+It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \citep{Razin:1991vj}, silencing of transposable elements \citep{Jones:2001eg}, X chromosome inactivation \citep{Mohandas:1981uf} and tumorigenesis \citep{Ehrlich:2002kz}.
 
@@
 \centering
-\includegraphics[width=0.3\textwidth]{../figures/5-methylcytosine.png}
-\caption{Chemical structure of 5-methylcytosine.
-"5-methylcytosine".
-By Yikrazuul (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg](http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg)}
+\includegraphics[width=0.3\textwidth]{../figures/5-Methylcytosine.pdf}
+\caption{Chemical structure of 5-methylcytosine. "5-methylcytosine". By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
 \label{fig:5mC}
@@
 
-It is common to describe a cytosine as being 'methylated' or 'unmethylated', however, care must be taken in defining these terms.
+It is common to describe a cytosine as being 'methylated' or 'unmethylated', however, care must be taken when using these terms.
 At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
@@
 
-[^palindromic_sequence]: A palindromic DNA sequence is one that is identical on when read in the 5' to 3' direction on both the original strand and the complementary strand of the double helix.
+[^palindromic_sequence]: A palindromic DNA sequence is one that is identical when read in the 5' to 3' direction on both the original strand and the complementary strand of the double helix.
 For example, _CG_ is a palindromic sequence.
@@
 Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other.
-While the former is more common, examples of the latter case, such as _allele-specific methylation_ \citep{Shoemaker:2010jz} and _genomic imprinting_ \citep{Li:1993ea}, are of considerable biological interest.
+While the former is more common, examples of the latter case, such as _allele-specific methylation_ \citep{Shoemaker:2010jz} and _genomic imprinting_ \citep{Li:1993ea}, are important epigenetic phenomena.
 
@@
 CpGs are underrepresented in the human genome.
-The GC-content of each the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is between $41\%$.
-If the nucleotides were uniformly distributed across the human genome we would expect about $4.1\%$ of dinucleotides to be CpGs.
+The _GC-content_ of the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is approximately $41\%$.
+If these nucleotides were uniformly distributed across the genome then we would expect about $4.1\%$ of _dinucleotides_, pairs of nucleotides, to be CpGs.
 Instead, only $1\%$ of dinucleotides are CpGs.
 
-One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}.
+One reason for the relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}.
 Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}).
@@
 \centering
-\includegraphics[width=0.3\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
+\includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
 \caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19).
@@
 Figure \ref{fig:IPD_ECDF_hg19} also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome.
-By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more 'close' pairs and more 'distant' pairs than we would expect by chance.
+By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more 'close' pairs than we would expect by chance.
 
@@
 
-Figure \ref{fig:IPD_frequency_polygon_hg19} is an alternative way to visualise these data, which plots the percentage of pairs of CpGs with a given IPD.
+Figure \ref{fig:IPD_frequency_polygon_hg19} is an alternative way to visualise these data by plotting the percentage of pairs of CpGs with a given IPD.
 Figure \ref{fig:IPD_frequency_polygon_hg19} shows that there is a cluster of CpGs with $IPD < 10$.
@@
 CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}.
-The classical definition of a CpG island, given by \citep{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:
+The classical definition of a CpG island, given by \citet{GardinerGarden:1987we}, and used by the popular UCSC genome browser ([http://genome.ucsc.edu/cgi-bin/hgGateway](http://genome.ucsc.edu/cgi-bin/hgGateway)), is a region of the genome where:
 
@@
 
-This definition was refined by \citet{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands.
+This definition was refined by \citet{Takai:2002jb} to exclude _Alu_-repetitive elements, which are otherwise misclassified as _bona fide_ CpG islands.
 More recently, \citet{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
-I use this definition in my thesis.
+In my thesis, I use the definition given by \citet{Wu:2010do}.
 
@@
 These 'sandy/beachy' metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
-CpG island shores, shelves and the open sea are all defined with respect to CpG islands.
-Specifically,
+CpG island shores, shelves and the open sea are all defined with respect to CpG islands:
 
-- CpG island shores are regions within 2 kb of CpG islands and have been demonstrated to have an increased variability of CpG methylation \citep{Irizarry:2008hg}.
-Shelves have a comparatively low density of CpGs.
+- CpG island shores are regions within 2 kb of CpG islands. These have been demonstrated to have an increased variability of CpG methylation \citep{Irizarry:2009ky}.
 - CpG island shelves are defined as regions within 2 kb of a CpG island shore \citep{Bibikova:2011fw}.
@@
 \citet{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines .
-Pluripotent cells includes embryonic stem cells along with induced pluripotent stem cell lines.
+Pluripotent cells includes embryonic stem (_ES_) cells and induced pluripotent stem (_iPS_) cell lines.
 
-An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, where they found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
+An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, who found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
 
 Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation.
-This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips, only measure CpG methylation.
-However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context.
-This discussed further in Section sec:dna_methylation_assays.
+This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k microarrays, only measure CpG methylation.
+However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context (see Section \ref{sec:dna_methylation_assays}).
 
@@
 Methylation is not the only chemical modification of cytosines, although it is by far the most common.
-These additional modifications may be thought of as 'modifications of a modification' since they extend 5-methylcytosine.
 Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \citep{Plongthongkum:2014jg}.
@@
 \citet{Kriaucionis:2009bm} and \citet{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC.
-This suggests a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.
+This suggests a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks.
 
@@
 
-A common analogy used in describing epigentic marks uses 'writers', 'readers' and 'erasers' \citep{Moore:2013in}.
+A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,]{Moore:2013in}.
 In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
@@
 In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes.
-The DNMTs are commonly split into two groups, namely the maintenance methyltransferase and the _de novo_ methyltransferases.
+The DNMTs are commonly split into two groups, namely the maintenance methyltransferases and the _de novo_ methyltransferases.
+
 DNA methylation is not preserved by the DNA replication machinery and so it is the role of the maintainence methyltransferases to restore the methylation pattern on the daughter strand of DNA following DNA replication.
 In mammals, _DNMT1_ is known as the maintenance DNA methyltransferase.
-_DNMT3a_ and _DNMT3b_ are known as the _de nove_ methyltransferases, although these are also required for the maintenance of DNA methylation \citep{Jones:2009ji}.
-Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \citep{Li:1992bp}.
+
+_DNMT3a_ and _DNMT3b_ are known as the _de novo_ methyltransferases, although these are also required for the maintenance of DNA methylation \citep{Jones:2009ji}.
+Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development since mouse knockouts[^KO] for either gene are embryonically lethal \citep{Li:1992bp}.
 In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \citep{Li:1992bp}.
@@
 
-The removal or erasure of DNA methylation, called _demethylation_, is characterised as _passive_ loss or _active_ removal.
+The removal or erasure of DNA methylation, called _demethylation_, may be characterised as _passive_ loss or _active_ removal.
 Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division.
@@
 5mC oxidation-dependent active DNA demethylation.
-This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ -> _5hmC_ -> _5fC_ -> _5caC_ reaction.
-The removal of _5hmC_, _5fC_ or _5caC_ is biochemically "easier" than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.
+This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ $\to$ _5hmC_ $\to$ _5fC_ $\to$ _5caC_ reaction.
+The removal of _5hmC_, _5fC_ or _5caC_ is biochemically 'easier' than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.
 
 One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
-This is an active area of investigation.
+This is an active area of research.
 
@@
 
-Each of these pre-treatments can then be combined with a second step of amplification or hybridisation. For example,
+Following treatment, DNA methylation can be assayed using standard techniques such as:
 
@@
 
-An exception to this classification scheme are a new class of assay that seek to directly 'read' whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA.
+An exception to this classification scheme are a new class of assays that seek to directly 'read' whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA.
 For example, both \citet{Laszlo:2013kf} and \citet{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine is methylated.
@@
 Almost all assays of DNA methylation measure a population average from a pool of hundreds or thousands of cells.
-For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay.
+For a diploid organism, this is an average over several distinct levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay.
 Hundreds or thousands of cells are required in order to have sufficient material as input for the assay.
@@
 Depending on your definition of 'high', a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample.
-The second definition of throughput is  related to cost, be it money or time, i.e., 'how many samples can I afford to analyse?''.
+The second definition of throughput is  related to cost, be it money or time, i.e., 'how many samples can I afford to analyse?'.
 
@@
 In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays.
-In particular, I will describe in detail the 'gold standard; assay of DNA methylation, _whole-genome bisulfite-sequencing_ (WGBS), that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.
+In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.
 
@@
 Assays based on restriction enzyme were some of the first developed for studying DNA methylation.
-They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \citep{Laird:2010iv}.
-One such genome-wide assays is CHARM (_comprehensive high-throughput arrays for relative methylation_, \citet{Irizarry:2008hg}).
+These were initially developed for studying a small number of loci although they have been extended to genome-scale analysis approaches \citep{Laird:2010iv}.
+One such genome-wide assays is _CHARM_, 'comprehensive high-throughput arrays for relative methylation' \citep{Irizarry:2008hg}).
 CHARM combines a methylation fractionation step --- be it _MeDIP_, _HpaII_, or as in the original publication, _McrBC_ --- with a tiling array and analysis techniques that leverage regional DNA methylation levels.
@@
 
-When bisulfite-treated is amplified by PCR, the uracils are converted to thymines.
-These bisulfite-conversion assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (Figure \ref{fig:bisulfite-conversion}).
+When bisulfite-treated DNA is amplified by PCR, the uracils are converted to thymines.
+Therefore, these bisulfite-conversion assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (Figure \ref{fig:bisulfite-conversion}).
 
@@
 Although subsequent enhancements in the automation of Sanger sequencing improved the throughput of these assays, it was never going to be able to deliver a cost-effective, genome-scale assay of DNA methylation.
-The development of hybridisation microarrays provided an alternative way of analysing bisulfite-treated DNA that could provide cheap, genome-wide measurements of DNA methylation.
+The development of hybridisation microarrays provided cheap, genome-wide measurements of DNA methylation from bisulfite-treated DNA.
 
@@
 
-- The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a largely 3-base alphabet) leads to decreased hybridisation specificity
+- The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a mostly 3-base alphabet) leads to decreased hybridisation specificity
 - Sequences containing multiple cytosines require multiple probe versions in order to assay all possible methylation patterns
@@
 In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology \citep{Cokus:2008fc, Lister:2008bh}.
-Whole-genome bisulfite-sequencing is the gold-standard for analysing genome-wide DNA methylation data.
+Whole-genome bisulfite-sequencing remains the gold standard assay for measuring genome-wide DNA methylation data.
 
-\citet{Cokus:2008fc} termed their approach _BS-seq_ while \citet{Lister:2008bh} called theirs _methylC-seq_.
-From a bioinformatics perspective, the main difference is that the _BS-seq_ protocol produces sequencing reads from four bisulfite-converted DNA strands - the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) - which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results.
-In contrast, the _methylC-seq_ protocol only produces sequencing reads from two bisulfite-converted strands - _OT_ and _OB_ - and so only requires mapping to two different _in silico_ converted reference genomes, followed by a merge.
+\citet{Cokus:2008fc} termed their approach _BS-seq_ while \citet{Lister:2008bh} called their method _methylC-seq_.
+From a bioinformatics perspective, the main difference is that the _BS-seq_ protocol produces sequencing reads from four bisulfite-converted DNA strands --- the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) --- which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results.
+In contrast, the _methylC-seq_ protocol only produces sequencing reads from two bisulfite-converted strands --- _OT_ and _OB_ --- and so only requires mapping to two different _in silico_ converted reference genomes, followed by a merge.
 The methylC-seq protocol is now the standard whole-genome bisulfite-sequencing protocol, due in part to the simpler bioinformatics analysis.
@@
 Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples.
-Furthermore, approximately $40\%$ of reads will not contain any CpGs, which, depending on the experiment, might be considered a gross waste of money.
+Furthermore, depending on the choice of several sequencing parameters, approximately $35--72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_withou_cpgs}), which might be considered a gross waste of resources for some experiments.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}rc@{}}
+\toprule
+Sample              & Reads with no CpG \\ \midrule
+ADS-iPSC            & $35 \%$                \\
+ADS-adipose         & $37 \%$                \\
+ADS                 & $39 \%$                \\
+H9\_Laurent         & $39 \%$                \\
+FF-iPSC\_19.11      & $48 \%$                \\
+FF-iPSC\_19.7       & $49 \%$                \\
+H1+BMP4             & $49 \%$                \\
+FF-iPSC\_6.9        & $50 \%$                \\
+IMR90-iPSC          & $50 \%$                \\
+H9                  & $51 \%$                \\
+FF-iPSC\_19.11+BMP4 & $52 \%$                \\
+FF                  & $53 \%$                \\
+IMR90\_r2           & $57 \%$                \\
+IMR90\_r1           & $58 \%$                \\
+H1\_r2              & $59 \%$                \\
+H1\_r1              & $62 \%$                \\
+HSF1                & $72 \%$                \\ \bottomrule
+\end{tabular}
+\caption{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset.
+Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs.
+See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
+\label{tab:reads_withou_cpgs}
+\end{table}
+
 Several assays have been developed to perform targeted high-throughput bisulfite-sequencing.
-The targeted nature of these assays are their obvious advantage and disadvantage - only a subset of the genome needs be sequenced but you only obtain information about methylation in those regions.
-Depending on the experiment, this tradeoff may be worthwhile, and these assays have been successfully used in a number of studies such as the BLUEPRINT Epigenome project ([http://www.blueprint-epigenome.eu/](http://www.blueprint-epigenome.eu/)) and the NIH Roadmap Epigenomics Mapping Consortium ([http://www.roadmapepigenomics.org/](http://www.roadmapepigenomics.org))
+The targeted nature of these assays are their obvious advantage and disadvantage; only a subset of the genome needs be sequenced but you only obtain information about methylation for that subset.
+Depending on the experiment, this tradeoff may be worthwhile, and these assays have been successfully used in a number of studies such as the BLUEPRINT Epigenome project ([http://www.blueprint-epigenome.eu](http://www.blueprint-epigenome.eu)) and the NIH Roadmap Epigenomics Mapping Consortium ([http://www.roadmapepigenomics.org](http://www.roadmapepigenomics.org))
 
@@
 - Extended reduced representation bisulfite-sequencing, also known as enhanced reduced representation bisulfite-sequencing (eRRBS, \citet{Akalin:2012he}). A modified version of the RRBS protocol.
-- NimbleGen's SeqCap Epi Enrichment System and Agilent's SureSelectxt Human Methyl-Seq: Both of these commercial products use DNA hybridisation to enrich the sequencing library for pre-defined regions of interest.
+- NimbleGen's SeqCap Epi Enrichment System and Agilent's SureSelectXT Human Methyl-Seq: Both of these commercial products use DNA hybridisation to enrich the sequencing library for pre-defined regions of interest.
 This enriched library is then bisulfite-converted and sequenced.
@@
 
-Bisulfite-conversion assays are considered the 'gold standard' for studying DNA methylation since cytosine methylation can be detected at single-base resolution \citep{Stirzaker:2014cu}.
+Bisulfite-conversion assays are considered the gold standard for studying DNA methylation since cytosine methylation can be detected at single-base resolution \citep{Stirzaker:2014cu}.
 In fact, single-molecule, base pair resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].
 
-[^microarrays]: Microarray hybridisation can provide base pair resolution but not single molecule resolution.
-The signal from a microarray-based experiment a sample-wide average since, for a given locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.
+[^microarrays]: Microarray hybridisation assays can provide base pair resolution but not single molecule resolution.
+The signal from a microarray-based experiment a sample-wide average since, for each locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.
 
-Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1-5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
+Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1--5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
 However, the minimal amount of DNA is being reduced with each technological advance.
-Recently, both \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although they are not yet a 'commercial' sequencing assay.
+Recently, both \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although they are not yet a commercially available sequencing assay.
 Furthermore, these are not yet proper genome-wide assays.
@@
 
-A recently discovered disadvantage is that bisulfite-conversion assays are unable to distinguish 5hmC from 5mC because hydroxymethyl similarly protects a cytosine from deamination to uracil.
+A recently discovered disadvantage of bisulfite-conversion assays is that they are unable to distinguish 5hmC from 5mC; 5-hydroxymethyl, like 5-methyl, similarly protects a cytosine from deamination to uracil.
 In effect, the detection of 5mC, and all subsequent inference, is confounded with that of 5hmC.
-For most experiments this isn't much of a problem - most cells have very low levels of 5hmC and so there is little confounding - however, in certain experiments this needs a more careful approach.
+For most experiments this isn't much of a problem --- most cells have very low levels of 5hmC and so there is little confounding --- however, in certain experiments this needs a more careful approach.
 To address this issue, \citet{Booth:2012fh} developed _oxidative bisulfite-sequencing_ (oxBS-seq) and \citet{Yu:2012ej} developed _Tet-assisted bisulfite sequencing_ (TAB-seq) for separate 5mC and 5hmC detection.
@@
 
-There is also potential confounding of 5C, 5mC and 5hmC levels with 5fC and 5caC, although this has received less attention since 5fC and 5caC are believed to exist at far lower quantities than 5mC and 5hmC.
+There is also potential confounding of 5C, 5mC and 5hmC with 5fC and 5caC, although this has received less attention since 5fC and 5caC are believed to exist at far lower quantities than 5mC and 5hmC.
 Nonetheless, sequencing assays of 5fC and 5caC exist based on the idea of treating the DNA with a chemical and performing 'signal subtraction' with traditional bisulfite-sequencing or another assay \citep{Wu:2014jj}.
@@
 The locus in orange is a cytosine in the refererence genome but a thymine in the sample's genome.
-Because the read is compared against the reference genome it is incorrectly inferred to be an unmethylated cytosine.
+Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine.
 The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome.
-Because the read is compared against the reference genome it is incorrectly inferred to be a thymine.}
+Because the read is compared against the reference genome it may incorrectly inferred to be a thymine.}
 \label{fig:bisulfite_CT_SNP.pdf}
@@
 Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_.
-_PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \citep{Warnecke:1997eh}.
-_Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment.
-Finally, treating DNA with sodium bisulfite damages the DNA, resulting in the fragmentation of long molecules.
-This limits the size of the fragments that can be studied using bisulfite-conversion assays to $\approx 500$ bp \citep{Ecker:2010et}
+PCR-bias is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \citep{Warnecke:1997eh}.
+Incomplete bisulfite-conversion leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment.
+DNA degredation occurs because sodium bisulfite damages DNA, resulting in the fragmentation of long molecules.
+This limits the size of the fragments that can be studied using bisulfite-conversion assays to approximately $500$ bp \citep{Ecker:2010et}
 
@@
 
-Ideally, an assay of DNA methylation could naturally 'read' methylated cytosines as a distinct signal from unmethylated cytosines.
+Ideally, an assay of DNA methylation could natively 'read' methylated cytosines as a distinct signal from unmethylated cytosines.
 This idea forms the basis of assays using the kinetics of DNA to infer the presence of DNA methylation and other DNA modifications.
@@
 DNA methylation is an epigenetic modification with important roles in many biological systems.
-In mammals, CpGs are frequently methylated while non-CpG sites are less frequently methylated.
+In mammals, CpGs are frequently methylated while non-CpG cytosines are less frequently methylated.
 Most unmethylated CpGs are found in CpG islands, which are important regulatory elements in the genome.
 
-Whole-genome bisulfite-sequencing provides single-nucleotide resolution data and is the gold-standard assay for studying DNA methylation.
+Whole-genome bisulfite-sequencing provides single-nucleotide resolution data and is the gold standard assay for studying DNA methylation.
 Methods for analysing whole-genome bisulfite-sequencing data are the focus of my thesis.

### 2015-04-12 Rmarkdown/biology_background.Rmd
Add chapter abstract macro

@@
 
-\section{Chapter overview}\label{sec:biology_background_chapter_overview}
-
-In this chapter I summarise the basic biology necessary for
-understanding my thesis. I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
+\begin{chapabstract}
+I summarise the basic biology necessary for understanding my thesis.
+I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
+\begin{end}
 
@@
 Genomics research was revolutionised by the development of cheap high-throughput sequencing technology, and the study of DNA methylation was no exception.
-In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology \citep{Cokus:2008fc, Lister:2008bh}.
+In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Solexa/Illumina sequencing technology \citep{Cokus:2008fc, Lister:2008bh}.
 Whole-genome bisulfite-sequencing remains the gold standard assay for measuring genome-wide DNA methylation data.

### 2015-04-12 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Add my edits to WGBS bioinformatics analysis.

@@
 
-\section{Chapter overview}\label{sec:wgbs_analysis_chapter_overview}
+\begin{chapabstract}
+I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used \emph{methylC-seq} protocol.
+All data used in my thesis were generated using this protocol.
 
-This chapter and Chapter \ref{chap:wgbs_downstream_analyses} describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges.
-I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay.
-All data used in my thesis were generated using this protocol.
-Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.
+There are four fundamental steps in the analysis of bisulfite-sequencing data:
 
-There are four fundamental steps in the analysis of whole-genome bisulfite-sequencing data:
+\begin{enumerate}
+  \item Data quality control checks
+  \item Read mapping and post-processing of mapped reads
+  \item Methylation calling
+  \item Downstream analyses
+\end{enumerate}
 
-1. Data quality control checks
-2. Read mapping and post-processing of mapped reads
-3. Methylation calling
-4. Downstream analyses
+This chapter focuses on steps 1-3, while Chapter \ref{chap:wgbs_downstream_analyses} addresses the wide variety of analyses available at Step 4.
+Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data.
+Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA sequencing.
 
-Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data.
-Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq.
-There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis.
-This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:wgbs_downstream_analyses}.
-
-The chapter concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data.
+The chapter concludes by introducing the `methtuple` software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci.
+\texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
+\end{chapabstract}
 
@@
 $20\%$).
-As noted in the `FastQC` documentation, some a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.
+As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.
 
-More important is the identification and removal of contaminating sequences.
+Perhaps the most important quality control of bisulfite-sequencing data is the identification and removal of contaminating sequences.
 `FastQC` will screen a subset of the reads against a list of known, common contaminants such as adapter sequences.
-Illumina adapter sequences are ligated to each DNA molecule in the library in order to perform Illumina sequencing.
-The sequencer can 'read into' the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those created in bisulfite-sequencing libraries.
-This means that some reads are a chimera that contain the sequence of interest (from the sample) and junk (from the adapters).
+When sequencing is performed using the widely used Illumina technology, adapter sequences must be ligated to the ends of each DNA molecule in the library.
+The adapters do not contain the biological sequence of interest, however, the sequencer can 'read into' the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those frequently created in bisulfite-sequencing libraries.
+This means that some reads are chimeras that contain the biological sequence of interest (from the sample) and junk sequence (from the adapters).
 This contamination needs to be removed for two reasons:
 
-1. Reads containing adapter contamination will generally not map to the reference genome, meaning these reads go to waste.
-2. If they do map, then this will result incorrect inferences - the 'garbage in, garbage out' maxim.
+1. Reads containing adapter contamination will generally not map to the reference genome, meaning these reads are needlessly wasted.
+2. If they do map, then this will result incorrect inferences; the 'garbage in, garbage out' maxim.
 
 Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \citep{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants.
-Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.
+Reads might also be trimmed to remove low quality sequencing cycles, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.
 
@@
 Read mapping is complicated by the bisulfite-treatment of the DNA.
-Following bisulfite-treatment, the DNA fragments are now mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location.
+Following bisulfite-treatment, the DNA fragments are mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location.
 Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome.
 Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.
-Several review articles have tried to summarise and compare the various approaches \citep{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
+Several review articles have summarised and compared the various approaches \citep[e.g.,]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
 
@@
 This biases downstream inference and means that these mappers are generally less popular.
-I will instead focus on the _in silico_ bisulfite-conversion mappers, called as such because of a key step taken by these in the mapping process.
 
@@
 
-Table \ref{tab:bs-seq_mappers} list some popular bisulfite-sequencing read mappers, which have been selected to highlight the variety of back-ends used by these software.
+Table \ref{tab:bs-seq_mappers} list some popular bisulfite-sequencing read mappers, which have been selected to highlight the variety of underlying mapping software used by these tools.
 
@@
 \end{tabulary}
-\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of back-ends used by these software.}
+\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
 \label{tab:bs-seq_mappers}
@@
 
-Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or its binary equivalent, `BAM` \citep{Li:2009ka}.
+Each of these aligners can report the output in the standard Sequence Alignment/Map format, `SAM`, or its binary equivalent, `BAM` \citep{Li:2009ka}.
 However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way.
-This makes it difficult to develop downstream analysis tools.
+This complexity makes it difficult for downstream analysis tools to support the output of different mapping software.
 
@@
 
-In bisulfite-sequencing data, PCR duplicates containing the a methylation locus can result in a biased estimate of the methylation level at that locus.
+In bisulfite-sequencing data, PCR duplicates containing a methylation locus can result in a biased estimate of the methylation level at that locus.
 This is because the sequenced reads do not accurately represent the true methylation levels of the sample.
 
-Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate.
+There is generally no way to tell if a read is truly a PCR duplicated based on the sequencing data alone.
+
+There is generally no way to tell based on sequencing data if a read is truly a PCR duplicate.
 However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as 'PCR duplicates'[^pcr_dup]).
-Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
-These software use mapped reads from a `SAM/BAM` file as input.
-There are subtle differences in how each method calls PCR duplicates.
-As an example, this is how `Picard`'s `MarkDuplicates` function works:
+Software to identify PCR duplicates includes the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
+These software all use mapped reads from a `SAM` or `BAM` file as input.
 
-[^pcr_dup]: The distinction between suspected PCR duplicates and true PCR duplicates is rarely made, probably because the phrase is so clunky.
+[^pcr_dup]: The distinction between suspected PCR duplicates and true PCR duplicates is rarely made, possibly because the phrase is so clunky.
 Suspected PCR duplicates are almost always referred to as PCR duplicates with the implicit assumption that the reader is aware that these very likely include false positive calls.
@@
 
-> "Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair.
-When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment.
-You can thus think of it as determining "if all the bases from the read were aligned, where would the 5' most base have been aligned".
-It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair.
-"Best" is defined as the read pair having the highest sum of base qualities as bases with $Q >= 15$" ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).
-
-Using these software will result in false positive calls because reads may satisfy this (or similar) criteria yet come from separate DNA fragments in the original sample.
-The false positive rate is a particular problem when a small 'genome' is sequenced at high coverage, where the 'genome' might be a targetted region of a genome such as exons or CpG islands.
+Roughly speaking, these tools will flag reads with identical start and end co-ordinates as being suspected PCR duplicates.
+This will inevitably lead to some false positives because reads may have identical co-ordinates yet not be PCR duplicates.
+The false positive rate is a particular problem when a subset of the genome is sequenced at high coverage, such as in RRBS.
 This can be thought of as an example of the pigeonhole principle, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).
@@
 
-In practice, this means that for whole-genome sequencing data we can be fairly confident that possible PCR duplicates are _true_ PCR duplicates.
+In practice, this means that for whole-genome sequencing data we can be fairly confident that suspected PCR duplicates are _true_ PCR duplicates.
 However, for targetted sequencing, such as RRBS or amplicon sequencing, we are much less confident and may remove some of our signal if we remove possible PCR duplicates.
@@
 
-Ideally, the probability that a base is called methylated should be independent of the sequencing cycle.
+Ideally, the probability that a base is called as methylated should be independent of the sequencing cycle.
 \citet{Hansen:2011gu} found that this is not the case and that in fact there is considerable bias towards the start (5') and end (3') of reads.
@@
 M-bias can be identified by plotting the read-position methylation level ($rpml$), which is the proportion of reads that are methylated at each read-position, as a function of read-position.
-These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for _read 1_ and _read 2_.
+These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for _read-1_ and _read-2_.
 If there is no M-bias then this plot should be a horizontal line.
 A 'bend' or 'spike' in this plot is evidence of M-bias.
-Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same height for `read_1` and `read_2`, although we see this is often not quite the case.
-Figure \ref{fig:ADS_M-bias} shows the M-bias plot for the ADS sample from the Lister (2011) dataset.
+Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same level for _read-1_ and _read-2_, although we see this is often not quite the case.
+Figure \ref{fig:ADS_M-bias} is the M-bias plot for the _ADS_ sample from the _Lister_ dataset[^datasets], which shows significant CpG M-bias at the start of _read-2_ and some noise at the start of _read-1_.
+
+[^datasets]: See Chapter \ref{chap:datasets} for a description of this sample.
 
@@
 \includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
-\caption{M-bias plot for the ADS sample from the Lister (2011) dataset.
-Each of read 1 (R1) and read 2 (R2) are plotted separately.
-We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.}
+\caption{M-bias plot for the \emph{ADS} sample from the \emph{Lister} dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.}
 \label{fig:ADS_M-bias}
@@
 
-If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should probably be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
+If a sample is processed over multiple batches, then M-bias estimation (and methylation calling) should probably be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
 For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs.
-Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with.
+Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with in practice.
 
-The strongest source of M-bias in Illumina WGBS data is at the 5' end of _read 2_, which sequences the 3' end of the DNA fragment.
+The strongest source of M-bias in Illumina whole-genome bisulfite-sequencing data is at the 5' end of _read-2_, which sequences the 3' end of the DNA fragment.
 Because the DNA fragment is often shorter than the sum of the read lengths, the 3' end of the fragment often contains adapter sequence and other 'junk' sequence.
@@
 
-\subsubsection{Computing M-bias}\label{sec:estimating_m-bias}
+\subsubsection{Estimating M-bias}\label{sec:estimating_m-bias}
 
-Estimating the M-bias and incorporating the results into the methylation calling can be done using two different strategies:
+Estimating M-bias and incorporating it into the methylation calling can be done using two different strategies:
 
@@
 Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias.
-This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls.
+This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the file containing the methylation calls.
 This is the approach taken by `BSmooth` \citep{Hansen:2012gr}.
 
-I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
+I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{section:methtuple}).
+
 In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.
@@
 A somewhat subtle point is that M-bias should only be estimated from reads that are actually going to be used for methylation calling.
-Suppose M-bias is highly correlated with some other quality metric, such as base quality, so that positions with M-bias also have low base quality[^bq_mbias].
-If you already intend to ignore all positions with a base quality less than 20 in your methylation calling then it makes sense to also ignore these positions when computing M-bias, otherwise you will overestimate the effect of M-bias and unnecessarily exclude read positions in your methylation calling.
-Unfortunately, the software I use for estimating M-bias, `bismark_methylation_extractor` \cite{Krueger:2011eb}, does not allow the user to exclude certain reads or read-positions from its calculations.
+Suppose that M-bias is highly correlated with some other quality metric, such as base quality, so that positions with M-bias also have low base quality[^bq_mbias].
+If you already intend to ignore read-positions with a base quality less than some threshold in your methylation calling, then it makes sense to also ignore these positions when estimating M-bias, otherwise you will overestimate the effect of M-bias and unnecessarily exclude read-positions in your methylation calling.
+Unfortunately, perhaps the most widely used software for estimating M-bias, `bismark_methylation_extractor` \cite{Krueger:2011eb}, does not allow the user to exclude certain reads or read-positions when estimating M-bias.
 
@@
 
-Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read position.
-This causes a minor problem when computing M-bias because we no longer know whether the read position is identical to the sequencing cycle.
+Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read-position.
+This causes a minor problem when computing M-bias because we no longer know whether the read-position is identical to the sequencing cycle.
 Soft-clipping or hard-clipping reads of their adapter sequence _during_ the alignment avoids this issue, because the clipping information (should be) preserved in the `CIGAR` string[^cigar].
@@
 This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice.
-The `bwa-meth` \citep{Pedersen:2014wt} and `LAST` \citep{Kieibasa:2011do} alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark` \citep{Krueger:2011eb}, cannot soft-clip reads and so require that reads are pre-trimmed.
+`bwa-meth` \citep{Pedersen:2014wt} and `LAST` \citep{Kieibasa:2011do} both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark` \citep{Krueger:2011eb}, cannot soft-clip reads and so require that reads are pre-trimmed.
 
 The M-bias plot is based on the read-position from the aligned data and not the sequencing cycle (which isn't directly available in the `SAM/BAM` file).
-If the reads have been pre-trimmed, then each read-position in the M-bias plot will therefore contain data from multiple sequencing cycles, which can amplify or mask M-bias signals.
+If the reads have been pre-trimmed, then each read-position in the M-bias plot will therefore contain data from multiple sequencing cycles, which can amplify or mask the M-bias signal.
 
@@
 Then, read-position $80$ will comprise $10\%$ sequencing cycle $80$ and $90\%$ sequencing cycle $100$.
-It is very likely that sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$.
+It is very likely that sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read-position $80$.
 
@@
 \caption{M-bias plot for the E18VA sample from the EPISCOPE dataset.
-Each of read 1 (R1) and read 2 (R2) are plotted separately.
-For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike.
-For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types.
-The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1.
+Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.
+For CpGs in \emph{read-1}, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike.
+For CpGs in \emph{read-2}, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read-positions for all methylation types.
+The spike at read-position 101 is also evidence, albeit to a lesser extent, in \emph{read-1}.
 This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
@@
 
-[^read_trimming]: Generally, this would also require that methylation calling is performed separately on each read length because most methylation callers are unable to deal with different M-bias profiles for different read lengths.
+[^read_trimming]: This would also require that methylation calling is performed separately for read with different lengths because most methylation callers are unable to deal with different M-bias profiles for different read lengths.
 
@@
 It is rather a subjective decision to make.
-Two problems that I found were:
+Two problems that I found when making these decisions were:
 
-1. Maintaining consistency across samples
+1. Maintaining consistency across samples.
 2. Determining where to draw the line when there is no dramatic 'spike' in the M-bias.
 
-For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both read 1 and read 2, as well as a big problem at read-position 101 in read 2.
-What is less clear is where to draw the line on the gradual decay downwards toward the end of read 1 and towards the start of read 2.
-This motivated me to write a few simple functions to do perform more systematic processing of M-bias results.
-These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.
+For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both _read-1_ and _read-2_, as well as a big problem at read-position 101 in _read-2_.
+What is less clear is where to draw the line on the gradual decay toward the end of _read-1_ and towards the start of _read-2_.
+This motivated me to write a few simple functions to perform more systematic processing of M-bias results.
+These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias` (__TODO: Add this class and methods to `MethylationTuples`__).
 
 For each sample, M-bias is computed separately for each methylation type because the level of methylation varies widely between CG and non-CG methylation types.
-For paired-end sequencing experiments, it is also done separately for each of _read 1_ and _read 2_ because M-bias is very different for these two mates and also because _read 2_ often has a slightly lower average level of methylation than does _read 1_ (e.g., see Figure \ref{fig:ADS_M-bias}).
+For paired-end sequencing experiments, it is also done separately for each of _read-1_ and _read-2_ because M-bias is very different for these two mates and also because _read-2_ often has a slightly lower average level of methylation than does _read-1_ (e.g., see Figure \ref{fig:ADS_M-bias}).
 
-To do this, I use a simple normalisation of the read-position methylation levels ($rpml$).
-Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels, i.e., $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read 1_.
+In `MethylationTuples`, I use a simple normalisation of the read-position methylation levels ($rpml$).
+Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels, i.e., $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read-1_.
 
@@
 
-The `plot` method has a `type` argument, so that the 'raw' M-bias or the 'normalised' M-bias can be visualised.
+The `plot` method has a `type` argument, so that the 'raw' M-bias or the 'normalised' M-bias can be visualised (__TODO__).
 
@@
 Typically, read-positions showing evidence of M-bias are excluded when calling methylation events.
-In fact, a slightly cruder procedure is often used whereby the entire ends of reads are removed.
-For example, for a $100$ bp read with M-bias observed at positions $1-4, 9, 94, 96-100$, then positions $1-9$ and $94-100$ would be ignored when methylation calling.
+In fact, a slightly cruder procedure is typically used whereby the entire ends of reads are removed.
+For example, suppose we have $100$ bp reads with M-bias observed at positions $1-4, 9, 94, 96-100$, then read-positions $1-9$ and $94-100$ would be ignored when methylation calling.
 Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method.
 
-Generally, this strategy is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads.
+This strategy is generally sufficient because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads.
 However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude.
-Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101.
+Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101 in 150 bp reads.
 Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
-My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded, however, this is not without its own problems, as we shall see in Section \ref{sec:cometh_nil_geq0}.
+My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants to exclude[^methtuple_mbias].
 
 In the datasets I have analysed I've had to ignore up to 30 read-positions per read due to M-bias.
-Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read positions entirely.
+Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely.
 One idea is to inversely weight methylation calls by the level of M-bias.
@@
 
-There are several other sources of bias when making methylation calls.
-Some are what I call technical biases - those driven by the sequencing protocol and processing of the data --- while others I call biological biases - those driven by measuring something other than what you thought you were measuring.
+There are several other sources of potential bias in analysing bisulfite-sequencing data.
+These include sequencing and alignment errors, and sequence variation at, or nearby to, methylation loci.
+A particularly interesting source of bias is due to cellular heterogeneity.
 
-Apart from M-bias, technical biases include sequencing and alignment errors, which can both give rise to spurious or otherwise incorrect methylation calls in obvious ways.
-Biological biases include cellular heterogeneity and sequence variation at or nearby to methylation loci.
-
-\subsubsection{Cellular heterogeneity}\label{sec:cellular_heterogeneity}
-
-Strictly speaking, cellular heterogeneity exists once you study DNA from more than a single cell, as shown in a recent papers using single-cell bisulfite-sequencing \citep{Guo:2013ih, Smallwood:2014kn}.
-However, in this context it means when you study DNA from a sample that is made up of different _types_ of cells.
+Recent papers using single-cell bisulfite-sequencing \citep{Guo:2013ih, Smallwood:2014kn} have investigated the extent of this cellular heterogeneity by comparing the methylomes of individuals cells that are nominally of the same 'type'.
+Cellular heterogeneity is particularly problematic when a sample contains multiple cell types.
 
@@
 However, whole blood contains a mixture of cell types, each of which has a distinct methylation profile.
-\citet{Houseman:2012km} exploited this fact to estimate the cell composition of whole blood samples, in particular the differences in white blood cell (leukocyte) proportions between different groups, e.g., cases and controls.
-
-In contrast to \citet{Houseman:2012km}, where this cellular heterogeneity was useful, this cellular heterogeneity can seriously bias the results and must be properly accounted for in any study exploring the relationship between differences in DNA methylation and a phenotype \citep{Jaffe:2014gy,Houseman:2014ds}.
+This cellular heterogeneity can seriously bias downstream analyses and must be properly accounted for in any study exploring the relationship between differences in DNA methylation and a phenotype \citep{Jaffe:2014gy,Houseman:2014ds}.
 For example, \citet{Jaffe:2014gy} provide evidence that several reported relationships between age and DNA methylation are likely due to changes in the cell composition of whole blood with age and not due to DNA methylation changes _per se_.
@@
 Most bisulfite-sequencing alignment software either performs methylation calling during the alignment process, as done by `Bismark`[^bismark_methylation_extractor], or as a separate step after the alignment and post-processing of the `SAM/BAM` file.
-An example of the latter is `Bis-SNP` \citep{Liu:2012ge}, which performs methylation calling, and also variant genotyping, from bisulfite-sequencing data aligned with the user's choice of alignment software.
+An example of the latter is `Bis-SNP` \citep{Liu:2012ge}, which performs methylation calling from bisulfite-sequencing data aligned with the user's choice of alignment software.
 
@@
 All bisulfite-sequencing assays use _reference-based_ methylation calling.
-That is, they require the specification of a DNA reference sequence that the aligned bisulfite-sequencing data are compared against to infer the methylation state of each sequenced locus.
+This means that they require the specification of a reference DNA sequence that the aligned bisulfite-sequencing data are compared against to infer the methylation state of each sequenced locus.
 Care must be taken to correctly handle the orientation and strand of the alignment.
@@
 When using reference-based methylation calling, the position of the methylation locus is with respect to the reference genome, since then all samples will use a common set of co-ordinates.
-Some loci cannot be typed using a reference-based approach.
+Some methylation loci cannot be typed using a reference-based approach.
 For example, unless the genome of the sample is fully known, methylation loci in insertions cannot be distinguished from genetic variation since there is no reference sequence to compare them against.
 
-Each aligned read may be used for methylation calling, although each read should be subjected to some filtering to remove low-quality reads and low-quality bases.
-When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from the estimation of $M$ and $U$[^read_filters].
+Prior to methylation calling, each read should be filtered to remove low-quality reads and low quality bases.
+When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters].
 Strictly speaking each sequenced nucleotide is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all nucleotides within 'surviving' reads.
-In my own work, I routinely use the following filters.
+In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.
 
@@
 
-1. The read is mapped (single-end or paire-end) and mapped in the expected orientation (paired-end only).
+1. The read is mapped (single-end or paired-end) and mapped in the expected orientation (paired-end only).
 2. The read is not marked as a PCR duplicate.
@@
 3.
-The base is a 'bisulfite mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a 'non-bisulfite mismatch; (e.g., the sequenced base is an A or a G at a C in the reference sequence).
+The base is a 'bisulfite mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a 'non-bisulfite mismatch' (e.g., the sequenced base is an A or a G at a C in the reference sequence).
 
@@
 
-The reference sequence is typically the reference genome used in the alignment step, although there are obviously differences between the sample's genome and the reference genome.
-This reference-based approach can be refined to incorporate genetic differences between the sample and the reference genome.
+The reference sequence is typically the reference genome used in the alignment step, in spite of the obvious differences between the sample's genome and the reference genome.
+This reference-based approach may be refined to incorporate genetic differences between the sample and the reference genome.
 This can be done in several ways:
@@
 A cheaper alternative is to genotype the sample on a genome-wide SNP microarray.
-This will give very accurate, very cheap genotypes at a large number of loci ($500,000 - 5,000,000$).
+This will give very accurate, very cheap genotypes at a large number of loci ($500,000--5,000,000$).
 However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.
@@
 The next best approach is that implemented in `Bis-SNP` \citep{Liu:2012ge}, which is to call genetic variation from the bisulfite-sequence data itself and to then define a set of sample-specific methylation loci at which to call methylation events.
-It is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina whole-genome bisulfite-sequencing protocol.
+`Bis-SNP` is designed for _directional_ bisulfite-sequencing libraries such as the widely used Illumina whole-genome bisulfite-sequencing protocol.
 
-Certain variants, in particular heterozygous $C>T$ SNPs, are more difficult to accurately genotype than others.
-Unfortunately, $C>T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants.
-Fortunately, it is possible to distinguish such $C>T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand --- if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
-Other base substitutions are more readily detected, and insertion and deletion events (indels) may also be called.
+Certain genetic variants, in particular heterozygous $C$>$T$ SNPs, are more difficult to accurately genotype than others.
+Unfortunately, $C$>$T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants.
+Fortunately, it is often possible to distinguish $C$>$T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand; if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
+Other base substitutions are more readily detectable, and insertion and deletion events (indels) may also be called.
 
@@
 \includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
-\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
+\caption{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
 For all three loci, the reads mapped to the forward strand contain a thymine.
-However, it is the base on the reverse strand that reveal the true genotype.
+However, it is the base on the reverse strand that reveals the true genotype.
 When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location.
@@
 
-In case it isn't clear, `Bis-SNP` actually provides three important pieces of information that make it almost as good as having whole genome DNA sequencing data on the same sample:
+To emphasise, `Bis-SNP` provides three important pieces of information that make it almost as good as having whole genome DNA sequencing data on the same sample:
 
@@
 2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine nucleotides in the reference genome.
-3. Other genetic variants that may be used in additional analyses, such as identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).
+3. Other genetic variants that may be used in additional analyses, such as in identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).
 
-Genotype calls made using `Bis-SNP` are less accurate than those from equal-depth whole-genome DNA sequencing because of the reduced complexity of bisulfite-converted DNA.
+Genotype calls made using `Bis-SNP` are less accurate than those from whole-genome DNA sequencing because of the reduced complexity of bisulfite-converted DNA.
 However, we essentially get to measure DNA variation 'for free' by using `Bis-SNP`, which makes it my preferred approach for incoporating genetic variation into methylation calling.
 The genetic variant calls made by `Bis-SNP` can also be used to _post-hoc_ filter methylation calls made by other software.
-I use this approach to filter methylation calls at m-tuples made with `methtuple`.
+I use this approach to filter methylation calls made with `methtuple`.
 
@@
 
-The third approach is to call methylation loci using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population.
-For example, exclude all cytosines in the reference genome that are also SNPs in dbSNP \citep{Sherry:2001gh}.
-This approach can only exclude sites from consideration, it cannot add sample-specific methylation loci.
+The third approach, and arguably the bare minimum, is to call methylation loci using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population.
+For example, we might exclude all cytosines in the reference genome that are also SNPs in dbSNP \citep{Sherry:2001gh}.
 
@@
 It also obviously requires a database of known variation for the organism being studied, which is the case for commonly studied organisms such as humans and mice.
-Finally, post-hoc filtering of known genetic variants is arguably a bare minimum requirement for methylation calling in order to prevent spurious methylation calls contaminating the dataset.
 
+This approach can obviously only exclude sites of known variation from consideration, and cannot add sample-specific methylation loci.
 To remove those reference-specific methylation loci that are __not__ found in databases of known genetic variation, we might identify loci in the sample that display a large number of non-C/T bases (resp.
@@
 We would like to use the two upstream nucleotides from the sample to infer the sample-specific methylation context, either inferred from each read separately or from a variant calling procedure such as `Bis-SNP`.
-However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome.
+However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome alone.
 
@@
 
-`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software I wrote to perform methylation calling at m-tuples.
-In order to explain how it works and my motivation for writing it, I first need to explain what an m-tuple is.
+`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software I wrote to perform methylation calling at genomic tuples that I call _m-tuples_.
+Before formally defining m-tuples, I first motivate the need for `methtuple`.
+
+\subsection{Motivation}
+
+Most methylation callers, such as `bismark_methylation_extractor` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
+The output file is a table, where each row records the co-ordinates of a cytosine and the number of methylated ($M$) and unmethylated ($U$) reads at that position.
+Table \ref{tab:1-tuples} is representative of the type of data returned by these programs.
+The file format is generally tab-delimited plain text, the Browser Extensible Data (BED) format or the Variant Call Format (VCF).
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lllll@{}}
+\toprule
+Chromosome & Strand & Position & M & U \\ \midrule
+chr1       & $+$      & 100      & 7 & 1 \\
+chr1       & $-$      & 101      & 5 & 2 \\
+chr2       & $+$      & 400      & 0 & 3 \\
+chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
+\end{tabular}
+\caption{Example of output for methylation calling at 1-tuples.
+Each row records the the number of methylated ($M$) and unmethylated ($U$) reads at a 1-tuple.
+Loci may be stratified by strand, as is done here, in which case most CpGs will have measurements for both the positive and negative strands.}
+\label{tab:1-tuples}
+\end{table}
+
+While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data.
+Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
+Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
+In order to study these concepts, we need software that can perform methylation calling at m-tuples.
+
+[^chimeric_reads]: This ignores chimeric reads, which are created when two DNA fragments ligate to one another during the library preparation.
+Certain bisulfite-sequencing protocols frequently produce chimeric reads.
+For example, using the post-bisulfite adapter tagging (PBAT) protocol \citep{Miura:2012fr} with a low input amount of DNA results in a huge number of chimeric reads (personal communication from Felix Krueger).
+The standard whole-genome bisulfite-sequencing protocol is not known to suffer from this issue.
+
+I wrote `methtuple` because I needed a tool to perform methylation calling at m-tuples and nothing was available that suited my purposes.
+Simultaneous with the development of `methtuple` there have been some software published with similar functionality.
+However, none of these do exactly what I require and some have what I consider to be severe deficiencies (Table \ref{tab:other_m-tuple_methylation_callers}).
+To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome bisulfite-sequencing data.
+
+\begin{table}[h]
+\centering
+\begin{tabulary}{\textwidth}{LLLL}
+\toprule
+Software           & Reference                                & Input                                              & Limitations                                                                \\ \midrule
+methclone & \citet{Li:2014ei}                       & \texttt{Bismark} \texttt{BAM}                                  & Unable to install \\
+methpat  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bme} & Designed for amplicons not whole-genome data.
+\\
+DMEAS    & \citet{He:2013cj}                       & Output of \texttt{bme} & Windows operating system only. Perl code only available as PDF file.             \\
+\bottomrule
+\end{tabulary}
+\caption{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
+\label{tab:other_m-tuple_methylation_callers}
+\end{table}
 
@@
 I refer to $m$ as the _size_ of the tuple.
-In principle, the $m$ loci that make up an m-tuple could come from anywhere in the genome, but it makes more sense to require that the $m$ loci be close to one another.
+In principle, the $m$ loci that make up an m-tuple could come from anywhere in the genome, but it makes most sense to require that the $m$ loci be close to one another.
 In fact, I generally require that an m-tuple consists of $m$ _adjacent_ methylation loci[^adjacent].
-An equivalent way of describing an m-tuple as comprising adjacent methylation loci is that the number of intervening loci is zero ($NIL = 0$).
+An equivalent way of describing an m-tuple as comprising adjacent methylation loci is one where the number of intervening loci is zero ($NIL = 0$).
 There are three reasons that I focus on m-tuples with $NIL = 0$:
@@
 1.
-Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ m-tuples, provided that we restrict ourselves to those m-tuples with $NIL = 0$.
-In contrast, if we allow $NIL \geq 0$ then there are $\binom{l}{\text{m}} \sim$ m-tuples.
-Obviously, $\binom{l}{\text{m}} \geq l - \text{m} + 1$, with strict inequality if $m \neq 1, l$.
-2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}
+Quantity: From a sequence containing $l$ methylation loci there are $l - \text{m} + 1$ $NIL = 0$ m-tuples.
+In contrast, there are $\binom{l}{\text{m}}$ $NIL \geq 0$ m-tuples.
+Obviously, $\binom{l}{\text{m}} \geq l - \text{m} + 1$, with strict inequality if $m \neq 1$ or $m \neq l$.
+2. Interpretability: Results for m-tuples with $NIL = 0$ are simpler to interpret than when allowing $NIL \geq 0$. This is discussed in Chapter \ref{chap:co-methylation}.
 3.
@@
 
-Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
+Generally, when referring to m-tuples I implicitly mean those with $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
 The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combinations` flag is set[^nil0].
 
-[^nil0]: Actually, while `methtuple` tries to produce m-tuples with $NIL = 0$ it can't guarantee this because it would require parsing the reference genome sequence.
-This is only really an issue with paired-end sequencing and is made clear in the example.
+[^nil0]: Actually, while `methtuple` tries to produce m-tuples with $NIL = 0$ it can't guarantee this because it would require looking up the reference genome sequence for each m-tuple (this is avoided for computational simplicity).
+This is only really an issue with paired-end sequencing, as is made clear in the examples of Figures \ref{fig:methtuple_example1}, \ref{fig:methtuple_example2}, \ref{fig:methtuple_example3} and \ref{fig:methtuple_example4}.
 Some _post-hoc_ filtering of the m-tuples will generally be required in order to remove those m-tuples with $NIL > 0$.
 
-Just as each methylation call at a 1-tuple comes from a single read, each methylation call at an m-tuple comes from a single read.
-For an m-tuple there are $2^{m}$ possible methylations calls.
-For example, at a 1-tuple there are $2^{1}$ possible methylation calls -- $M$ or $U$; at a 3-tuple there are $2^{3} = 8$ possible methylation calls -- $MMM, MMU, MUM, MUU, UMM, UMU, UUM$ or $UUU$.
+I require that each methylation call at an m-tuple comes from a single read.
+There are $2^{m}$ possible methylation calls at an m-tuple.
+For example, at a 1-tuple there are $2^{1}$ possible methylation calls --- $M$ or $U$; at a 3-tuple there are $2^{3} = 8$ possible methylation calls --- $MMM$, $MMU$, $MUM$, $MUU$, $UMM$, $UMU$, $UUM$ or $UUU$.
 
-For each m-tuple, I also define the intra-pair distance (_IPD_) as the vector containing the $m - 1$ pair-wise distances (measured in bp) between methylation loci in the m-tuple.
+For each m-tuple, I also define the intra-pair distance (_IPD_) as the vector containing the $(m - 1)$ pair-wise distances (measured in bp) between methylation loci in the m-tuple.
 For example, the 2-tuple (`chr7:+:145`, `chr7:+:163`) has $IPD = (163 - 145) = (18)$.
@@
 The IPD vector of a 1-tuple is undefined.
-In Chapter \ref{chap:co-methylation} I use the IPDs, along with other features such as the genomic context, to define 'similar' m-tuples.
 
@@
 \includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
-\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
+\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.
+The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
 \label{fig:methtuple_example1}
@@
 This happens because m-tuples are created independently from each read-pair; effectively, read-pair `B` is unaware of methylation locus `3`.
-Depending on the downstream analysis, you may want to _post-hoc_ filter out these m-tuples with non-adjacent loci.
-- The twice-sequenced methylation loci in read-pair `C`, `2` and `3`, are not double counted.
+Depending on the downstream analysis, the user may wish to _post-hoc_ filter out these m-tuples with non-adjacent loci.
+- The twice-sequenced methylation loci in read-pair `C`, `2` and `3`, are only counted once.
 
@@
 
-\subsection{Motivation}
-
-Most methylation callers, such as `bismark_methylation_extractor` and `Bis-SNP`, perform methylation calling at single methylation loci, which I refer to as 1-tuples.
-The output file is a table, where each row records the co-ordinates of a cytosine and the number of methylated and unmethylated reads at that position.
-Table \ref{tab:1-tuples} is representative of the type of data returned by these programs.
-The file format is generally tab-delimited plain text, the Browser Extensible Data (BED) format or the Variant Call Format (VCF).
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}lllll@{}}
-\toprule
-Chromosome & Strand & Position & M & U \\ \midrule
-chr1       & $+$      & 100      & 7 & 1 \\
-chr1       & $-$      & 101      & 5 & 2 \\
-chr2       & $+$      & 400      & 0 & 3 \\
-chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
-\end{tabular}
-\caption{Example of output for methylation calling at 1-tuples. Most CpGs will have a measurement from both the positive and negative strands, but not always.}
-\label{tab:1-tuples}
-\end{table}
-
-While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data.
-Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
-Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
-
-[^chimeric_reads]: This ignores chimeric reads, which are created when two DNA fragments ligate to one another during the library preparation.
-Certain bisulfite-sequencing protocols frequently produce chimeric reads.
-For example, using the post-bisulfite adapter tagging (PBAT) protocol \citep{Miura:2012fr} with a low input amount of DNA results in a huge number of chimeric reads (personal communication from Felix Krueger).
-The standard whole-genome bisulfite-sequencing protocol is not known to suffer from this issue.
-
-In order to study these concepts, we need software that can perform methylation calling at m-tuples.
-Table \ref{tab:other_m-tuple_methylation_callers} lists other software that can perform methylation calling at m-tuples.
-However, none of these existed when I started writing `methtuple`.
-Even now, none of these do exactly what I require and have more general limitations.
-To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome data.
-
-\begin{table}[h]
-\centering
-\begin{tabulary}{\textwidth}{LLLL}
-\toprule
-Software           & Reference                                & Input                                              & Limitations                                                                \\ \midrule
-methclone & \citet{Li:2014ei}                       & \texttt{Bismark} \texttt{BAM}                                  & Unable to install \\
-methpat  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bsm} & Designed for amplicons not whole-genome data.
-\\
-DMEAS    & \citet{He:2013cj}                       & Output of \texttt{bsm} & Windows operating system only. Perl code only available as PDF file.             \\
-\bottomrule
-\end{tabulary}
-\caption{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bsm} =  \texttt{bismark\_methylation\_extractor}.}
-\label{tab:other_m-tuple_methylation_callers}
-\end{table}
-
-\subsection{Methods}\label{sec:methtuple_methods}
+\subsection{Implementation}\label{sec:methtuple_methods}
 
@@
 Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting these overlapping bases.
-`methtuple` also allows the user to filter out specific read positions rather than wholesale filtering of the ends of reads.
+`methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads.
 This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}.
@@
 It is compatible with both Python2 and Python3.
+To improve performance, I provide a helper script to split the sample by chromosome and process each chromosome in parallel ([https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh](https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh)).
+This helper script makes extensive use of `GNU parallel` \citep{Tange:2011ty}.
+While Python-level parallel processing is desirable, this `GNU parallel`-based approach was simpler to implement and sufficient for my purposes.
 
-`methtuple` is currently limited to processing files produced by `Bismark` due to its reliance on the `Bismark`-specific tags `XM`, the "methylation call string",  `XR`, the "read conversion state for the alignment", and `XG`, the "genome conversion state for the alignment" ([http://www.bioinformatics.bbsrc.ac.uk/projects/bismark/Bismark_User_Guide_v0.13.0.pdf](http://www.bioinformatics.bbsrc.ac.uk/projects/bismark/Bismark_User_Guide_v0.13.0.pdf)).
+`methtuple` is currently limited to processing files produced by `Bismark` due to its reliance on the `Bismark`-specific tags `XM`, the "methylation call string",  `XR`, the "read conversion state for the alignment", and `XG`, the "genome conversion state for the alignment" ([http://www.bioinformatics.bbsrc.ac.uk/projects/bismark/Bismark_User_Guide.pdf](http://www.bioinformatics.bbsrc.ac.uk/projects/bismark/Bismark_User_Guide.pdf)).
 It could be extended to work with other bisulfite-sequencing aligners.
-However, due to the eccentricities of each aligner, such an extension would have to be aligner-specific and therefore a considerable undertaking.
+However, due to the eccentricities of each aligner, such an extension would have to be aligner-specific and is therefore a considerable undertaking.
 Each extension would require that tags analogous to the `XR`, `XG` and `XM` tags can be generated from the given `BAM` file.
@@
 
-I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples from the EPISCOPE, Lister, Seisenberger and Ziller datasets.
-To improve the performance I  split each sample by chromosome and process each chromosome in parallel.
-This is implemented in a helper script ([https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh](https://github.com/PeteHaitch/methtuple/blob/master/helper_scripts/run_methtuple.sh)) that makes extensive use of `GNU parallel` \citep{Tange:2011ty}.
-
+I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples.
 Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets.
@@
 
-The running time of \texttt{methtuple} is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its ploidy.
-The running time is largely independent of the tuple size (\texttt{-m}).
+The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its ploidy.
+The running time is largely independent of the tuple size (`-m`).
 The variation in running times within a chromosome is due to the number of reads generated per sample and the length of the reads, where the length of a paired-end read is defined as the sum of the mates' lengths.
@@
 It is instead driven by the number and density of CpGs on the chromosome.
-For example, \texttt{chr19}, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than \texttt{chr18}, which has less than half the CpG density of \texttt{chr19} (see Table \ref{tab:cpg_density_hg19}).
-The relationship between the maximum memory usage and the tuple size (\texttt{-m}) is complex; more data have to be retained as \texttt{-m} increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
-Therefore, memory usage is relatively constant across values of \texttt{-m} for a given chromosome.
-The obvious exception is for tuple size \texttt{2ac}, which used the \texttt{--all-combinations} flag.
-This means that all 2-tuples with $NIL \geq 0$ were computed and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the enormous increase in memory usage.
+For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Table \ref{tab:cpg_density_hg19}).
+The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
+Therefore, memory usage is relatively constant across values of `-m` for a given chromosome.
+The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`.
+This means that all 2-tuples with $NIL \geq 0$ were computed and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.
 
@@
 The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.
-Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
+Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
 \label{fig:methtuple_runtime}
@@
 The suffix `ac' means that the option \texttt{--all-combinations} was set.
-Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
+Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
 \label{fig:methtuple_memory}
@@
                 header = TRUE, stringsAsFactors = FALSE)
-xt <- xtable(x, caption = "Percentage of CpGs per chromosome in the human reference genome (hg19).", label = "tab:cpg_density_hg19")
+xt <- xtable(x, caption = "Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).", label = "tab:cpg_density_hg19")
 align(xt) <- rep("c", ncol(x) + 1)
@@
 
-`methtuple` is open-source software release under the MIT licence and available from [https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple).
+`methtuple` is open-source software released under the MIT licence and available from [https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple).

### 2015-04-12 Rmarkdown/datasets.Rmd
Add my edits to Datasets

@@
 
-1. Genetic variants were called using `Bis-SNP (v0.82.2)` via the `bissnp_easy_usage.pl` script.
+1. Genetic variants were called using the `bissnp_easy_usage.pl` script included with `Bis-SNP (v0.82.2)`.
 2.
@@
 4.
-A `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples (2-tuples with `--all-combinations`), 3-tuples and 4-tuples.
+Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see \ref{sec:MethylationTuples}).
+Specifically, a `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples (2-tuples with `--all-combinations`), 3-tuples and 4-tuples.
 I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.
@@
 
-The WGBS libraries from the \cite{Lister:2009hy} paper were the first published WGBS libraries from mammalian DNA.
+The methylC-seq libraries from the \cite{Lister:2009hy} paper were the first published whole-genome bisulfite-sequencing libraries of mammalian DNA.
 A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (`IMR90`), with those from a pluripotent tissue, embryonic stem cells (`H1`).
@@
 
-The WGBS libraries from the \cite{Lister:2011kg} include some created and published by other groups (see Table \ref{tab:Lister_dataset}).
+The methylC-seq libraries from the \cite{Lister:2011kg} include some created by the authors and some published by other groups (see Table \ref{tab:Lister_dataset}).
 These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells.
-There are no replicates for any of the 2011 samples.
+There are no replicates for any of the _Lister 2011_ samples.
 
@@
 The third mini dataset I refer to as the _Lister-IMR90_ data and includes samples _IMR90\_r1_ (2009), _IMR90\_r2_ (2009) and _IMR90-iPSC_ (2011).
-The _IMR90-iPSC_ sample is an induced pluripotent stem cell line derived from the _IMR90_cell line.
+The _IMR90-iPSC_ sample is an induced pluripotent stem cell line derived from the _IMR90_ cell line.
 
@@
 The aligned reads for the \cite{Lister:2011kg} data were downloaded from [http://neomorph.salk.edu/ips_methylomes/](http://neomorph.salk.edu/ips_methylomes/).
-These samples were aligned against the hg18 build of the human reference genome.
+These samples had been aligned against the hg18 build of the human reference genome.
 
-I wrote Python scripts to convert these files to `SAM` files.
+As the aligned reads were in a custom file format, I wrote Python scripts to convert these files to the canonical `SAM` format.
 These scripts are available from [https://github.com/PeteHaitch/Lister2BAM](https://github.com/PeteHaitch/Lister2BAM).
-These `SAM` files were then converted to `BAM` files with `SAMtools` \citep{Li:2009ka} and duplicate reads were marked using Picard's `MarkDuplicates` routine [http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/).
+These `SAM` files were then converted to `BAM` files with `SAMtools` \citep{Li:2009ka} and duplicate reads were marked using Picard's `MarkDuplicates` routine ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)).
 
@@
 
-The sequencing data for these 12 samples were processed and aligned by Aaron Statham (Garvan Institute of Medical Research).
+The sequencing data for these 12 samples were processed and aligned by Aaron Statham (Garvan Institute of Medical Research, Sydney).
 Each sample was aligned to the human reference genome (hg19) using `Bismark` (`v0.8.3`) with the `Bowtie2` backend.
@@
 
-The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \cite{Seisenberger:2012ko}.
+The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \citep{Seisenberger:2012ko}.
 These were a convenience sample provided to me by a colleague, Felix Krueger (Babraham Institude).
@@
 J1\_1             & ESC                        & PE         & $100$ bp    & $12\times$     \\
-E6.5\_epiblast\_1 & Epiblast                   & PE         & $101$ bp    & $13\times$     \\
-E16.5\_male\_1    & Male progenitor germ cells & PE         & $150$ bp    & $12\times$     \\ \bottomrule
+E6.5\_epiblast\_1 & Epiblast                   & PE         & $100$ bp    & $13\times$     \\
+E16.5\_male\_1    & Male progenitor germ cells & PE         & $100$ bp    & $12\times$     \\ \bottomrule
 \end{tabulary}
@@
 Specifically, I use a convenience sample of 8 WGBS libraries.
-These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research), who already had aligned and processed BAM files for these 19 libraries.
+These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research, Sydney), who already had aligned and processed BAM files for these 19 libraries.
 These libraries in fact represent 19 sequencing runs of eight biological samples, which I was not aware of until the data had already been processed.
@@
 
-I have used the CpG island definition from the hidden Markov model proposed by \citet{Wu:2010do}.
+I have used the CpG island definition from the hidden Markov model proposed by \citet{Wu:2010do} and implemented in the `makeCGI` R package (`v1.2`).
 The predicted CpG islands for the human reference genome (both hg18 and hg19) were downloaded from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/) on 29/10/2014.
-I used the supplied software to create the predicted CpG islands for the mouse reference genome (mm10).
+I used the `makeCGI` R package to create the predicted CpG islands for the mouse reference genome (mm10) since these were not available for download.

### 2015-04-12 latex/phd_thesis.tex
Add chapter abstract macro
Update date :(

@@
 \usepackage{etex}
+
+% Chapter abstracts (http://tex.stackexchange.com/a/59138)
+\usepackage{changepage}% http://ctan.org/pkg/changepage
+\usepackage{lipsum}% http://ctan.org/pkg/lipsum
+\makeatletter
+\newenvironment{chapabstract}{%
+    \begin{center}%
+      \bfseries Overview
+    \end{center}}%
+   {\par}
+\makeatother
 
@@
 of the degree of Doctor of Philosophy\\\\
-March 2015\\\\
+April 2015\\\\
 Department of Mathematics and Statistics\\

### 2015-04-12 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Add Terry’s suggested edits to WGBS bioinformatics analysis.

@@
 This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome.
-These multiple-sequenced fragments are called _PCR duplicates_.
+These multiply-sequenced fragments are called _PCR duplicates_.
 
@@
 
-If a sample is processed over multiple batches, then M-bias estimation (and methylation calling) should probably be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
+If a sample is processed over multiple batches, then M-bias estimation (and methylation calling) should be performed separately for each batch and then combined, or in some manner that is 'batch aware'.
 For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs.
@@
 1. The read-position of the base means that it is unlikely to be affected by M-bias.
-2. The base quality score greater than some threshold.
+2. The base quality score is greater than some threshold.
 3.
@@
 
-While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data.
-Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple.
-Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
-In order to study these concepts, we need software that can perform methylation calling at m-tuples.
+While 1-tuples are the basis of most analyses of bisulfite-sequencing data, they do not always give the complete picture of how DNA methylation is acting in the sample.
+To gain a clearer picture, we can exploit the fact that many bisulfite-sequencing reads contain multiple methylation loci (_m-tuples_) and that each read is from a single cell[^chimeric_reads].
+An example of where this is useful is shown in Figure \ref{fig:incomplete_picture_1-tuples} where we have two regions, each with four methylation loci, that have identical methylation calls at 1-tuples yet very different overall methylation patterns.
+Further examples of where m-tuples are useful are in studying _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
 
@@
 
-I wrote `methtuple` because I needed a tool to perform methylation calling at m-tuples and nothing was available that suited my purposes.
-Simultaneous with the development of `methtuple` there have been some software published with similar functionality.
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
+\caption{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
+\label{fig:incomplete_picture_1-tuples}
+\end{figure}
+
+In order to study these phenomena, we firstly need software that can perform methylation calling at m-tuples, which is why I wrote `methtuple`.
+When I began my PhD, there was no software capable of calling methylation patterns at arbitrarily sized m-tuples from whole-genome bisulfite-sequencing data.
+Simultaneous with the development of `methtuple`, there have been some software published with similar functionality.
 However, none of these do exactly what I require and some have what I consider to be severe deficiencies (Table \ref{tab:other_m-tuple_methylation_callers}).
@@
 \includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
-\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.
+\caption{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}).
 The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}

### 2015-04-13 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Finish TODOs in WGBS bioinformatics analysis.

@@
 This motivated me to write a few simple functions to perform more systematic processing of M-bias results.
-These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias` (__TODO: Add this class and methods to `MethylationTuples`__).
+These are included in the `MethylationTuples` R package (described in Section \ref{sec:MethylationTuples}) with the `MBias` class, its associated methods, `plot()` and `filter()`, and the helper function `readMBias()`.
 
@@
 
-The `trim` method identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$.
-While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types.
-I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $rpml_{CpG} < 0.72$ or $rpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias.
+The `filter()` method identifies read-positions where the $rpml$ differs by more than a given value (`threshold`) from $nrpml$.
+While it computes these statistics separately for each methylation type and read type, a common `threshold` is used for all methylation types and read types.
+I tend to use a value of `threshold = 3`, meaning that if the $median(rpml_{CpG}) = 75$, then any read-position with $rpml_{CpG} < 72$ or $rpml_{CpG} > 78$ is flagged as showing evidence of M-bias[^mbias_percentage].
 It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types.
@@
 
-The `plot` method has a `type` argument, so that the 'raw' M-bias or the 'normalised' M-bias can be visualised (__TODO__).
-
-```{r}
-# TODO: Insert code snippet showing plots returned by `plot(MBias)` with both `type = raw` and `type = normalised`
-```
+[^mbias_percentage]: The M-bias files created by `bismark_methylation_extractor --mbias_only` report the methylation levels as percentages rather than proportions.
 

### 2015-04-13 Rmarkdown/biology_background.Rmd
Fix citations

@@
 [^other_DNAm]: Two additional examples of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C).
-N6-methyladenine is a methylated form of adenine, which is found in mRNA \citep[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \citep{Ratel:2006dv}.
-N4-methylcytosine has also been detected in bacterial DNA \citep[e.g.,]{Ehrlich:1985bv, Ratel:2006dv}.
+N6-methyladenine is a methylated form of adenine, which is found in mRNA \citep[e.g.,][]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \citep{Ratel:2006dv}.
+N4-methylcytosine has also been detected in bacterial DNA \citep[e.g.,][]{Ehrlich:1985bv, Ratel:2006dv}.
 
@@
 
-A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,]{Moore:2013in}.
+A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,][]{Moore:2013in}.
 In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
@@
 Hundreds or thousands of cells are required in order to have sufficient material as input for the assay.
-Assays that require only a single cell as input do exist \citep[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.
+Assays that require only a single cell as input do exist \citep[e.g.,][]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.
 

### 2015-04-13 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Fix citations

@@
 Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.
-Several review articles have summarised and compared the various approaches \citep[e.g.,]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
+Several review articles have summarised and compared the various approaches \citep[e.g.,][]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
 
@@
 
-Methods to estimate the cellular heterogeneity bias and adjust for it are available \citep[e.g.,]{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}, although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data.
+Methods to estimate the cellular heterogeneity bias and adjust for it are available \citep[e.g.,][]{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}, although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data.
 This is not to say that these problems don't exist for sequencing data, merely that these have not been as well-explored.

### 2015-04-13 Rmarkdown/wgbs_statistical_framework.Rmd
Working through my and Terry’s edits to WGBS statistical framework.

@@
 ---
-title: "Statistical framework for whole-genome bisulfite-sequencing data"
+title: "A statistical framework for whole-genome bisulfite-sequencing data"
 output:
@@
 
-\chapter{Statistical framework for whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}
+\chapter{A statistical framework for whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}
 
-\section{Chapter overview}\label{sec:wgbs_statistical_framework_overview}
-
-In this chapter I set out a statistical framework for analysing bisulfite-sequencing data.
+\begin{chapabstract}
+I set out a statistical framework for analysing bisulfite-sequencing data.
 The ideas here are simple, however, they have not yet been put into a unified mathematical framework.
@@
 Finally, I describe key variables, common estimators of these and their statistical properties by analysing 40 whole-genome bisulfite-sequencing samples.
+\end{chapabstract}
 
@@
 Each sample starts as a population of cells, with potentially different methylation patterns.
-Each DNA fragment is coloured by its originating cell, although this is unknown in practice.
+Each DNA fragment is coloured by its originating cell (although this is unknown in practice).
 Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
@@
 The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., `chr1:+:666`.
-Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome.
-Not knowing the exact value of $N_{loci}$ is no great concern.
-
-It is sometimes convenient to refer to loci by the subscript $i$ rather than by $pos_{i}$.
-It is important to remember that the number of base pairs between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome.
-Also, depending on whether the data are stranded, a pair of loci may be on opposite strands.
+It will be convenient to refer to loci by the subscript $i$ rather than by $pos_{i}$, although it is important to remember that the number of base pairs between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome.
+Furthermore, depending on whether the data are stranded, a pair of loci may be on opposite strands.
 In a small number of instances the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes, e.g., the last CpG on chromosome 1 and the first CpG on chromosome 2.
@@
 The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state.
-As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile.
+As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell may have a slightly different methylation profile.
 Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states.
@@
 But the real problem is that none of the steps in creating the pool of DNA fragments is perfect.
-In particular, PCR introduces biases -- some molecules are preferentially amplified while others 'drop out'.
+In particular, PCR introduces biases; some molecules are preferentially amplified while others 'drop out'.
 So even if we knew (1), (2) and (3) we cannot simply multiply these together to compute $H_{i}$, although this might at least give us a rough estimate.
@@
 
-By summing over the number of fragments containing the $i^{th}$ locus, we obtain the number of fragments that are methylated at the $i^{th}$ locus ($M_{i}$) and unmethylated at the $i^{th}$ locus ($U_{i}$):
+By summing over the fragments containing the $i^{th}$ locus, we obtain the number of fragments that are methylated at the $i^{th}$ locus ($M_{i}$) and unmethylated at the $i^{th}$ locus ($U_{i}$):
 
 \begin{align*}
-  M_{i} &= \sum_{h = 1}^{H = H_{i}} Z_{h, i} = |\{Z: Z \in \mathcal{H}_{i}, Z = 1 \}| \\
-  U_{i} &= \sum_{h = 1}^{H = H_{i}} (1 - Z_{h, i}) = |\{Z: Z \in \mathcal{H}_{i}, Z = 0 \}|
+  M_{i} &= \sum_{h = 1}^{H_{i}} Z_{h, i} = |\{Z: Z \in \mathcal{H}_{i}, Z = 1 \}| \\
+  U_{i} &= \sum_{h = 1}^{H_{i}} (1 - Z_{h, i}) = |\{Z: Z \in \mathcal{H}_{i}, Z = 0 \}|
 \end{align*}
@@
 \begin{equation*}
-Z_{h, i} = \left\{
+Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)} = \left\{
   \begin{array}{l l}
@@
 There are $2^{\text{m}}$ possible methylation patterns at an m-tuple.
-Rather than describe a methylation pattern by a m-vector of zeros and ones, I also write these using $U$ and $M$; for example, the possible methylation patterns at a 2-tuple are $MM, MU, UM$ and $UU$.
+Rather than describe a methylation pattern by an m-vector of zeros and ones, I also write these using $U$ and $M$; for example, the possible methylation patterns at a 2-tuple are $MM, MU, UM$ and $UU$.
 
@@
 Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments.
-We have a large number(~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
+We have a large number (~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
 Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to $H_{i}$.
 
-At this point I ignore reads that do not contain any methylation loci as these are not relevant to this discussion and I make three simplifying assumptions:
+We can ignore reads that do not contain any methylation loci as these are not relevant to ths discussion. I make three further simplifying assumptions:
 
@@
 When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
-However, when using paired-end sequencing the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur:
+However, when using paired-end sequencing, the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur:
 
@@
 Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g.
-_read 1_ gives us $(i, i + 1)$ and _read 2_ gives us $(i + 1, i + 2, i + 3)$.
-In this example we must use only one of _read 1_ or _read 2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
+_read-1_ gives us $(i, i + 1)$ and _read-2_ gives us $(i + 1, i + 2, i + 3)$.
+In this example we must use only one of _read-1_ or _read-2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
 
 Each read measures the methylation state of one or more loci from a single DNA fragment.
-I denote by $\mathcal{R}_i$ the set of all reads containing the $i^{th}$ locus.
-Therefore, the number of reads containing the $i^{th}$ locus is $|\mathcal{R}_{i}|$, where $|\mathcal{R}_{i}| \leq H_{i}$ with strict inequality for almost all $i$.
+I denote by $\mathcal{R}_i$ the set of all mapped reads containing the $i^{th}$ locus.
+The number of reads containing the $i^{th}$ locus is referred to as the sequencing depth at the $i^{th}$ locus, which I denote by $d_{i} = |\mathcal{R}_{i}|$, where $d_{i} \leq H_{i}$ with strict inequality for almost all $i$.
 
@@
 
-By summing over the number of reads containing the $i^{th}$ locus we obtain the number of reads that are methylated at the $i^{th}$ locus ($m_{i}$) and unmethylated at the $i^{th}$ locus ($u_{i}$):
+By summing over the reads containing the $i^{th}$ locus we obtain the number of reads that are methylated at the $i^{th}$ locus ($m_{i}$) and unmethylated at the $i^{th}$ locus ($u_{i}$):
 
 \begin{align*}
-  m_{i} &= \sum_{z: z \in \mathcal{R}_{i}} z \\
+  m_{i} &= \sum_{r = 1}^{d_{i}} z_{i} \\
         &= |\{z: z \in \mathcal{R}_{i}, z = 1 \}| \\
-  u_{i} &= \sum_{z: z \in \mathcal{R}_{i}} (1 - z) \\
+  u_{i} &= \sum_{r = 1}^{d_{i}} (1 - z_{i}) \\
       &= |\{z: z \in \mathcal{R}_{i}, z = 0 \}
@@
 \begin{equation*}
-  \beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}
+  \beta_{i} = \frac{m_{i}}{d_{i}}
 \end{equation*}
 
-These are the so-called $\beta$-values, which are commonly interpreted as an estimate of the proportion of cells in the sample that are methylated at the $i^{th}$ locus.
-I discuss this interpretation, and other estimators of the  methylation level at a locus, in Section \ref{sec:estimating_B}.
+Here we have assumed that $d_{i} = m_{i} + u_{i}$, meaning that reads that do not have a methylation locus mapped to $pos_{i}$ do not contribute[^non-methylation_reads].
+
+[^non-methylation_reads]: Such reads can occur due to sequencing error, mapping error or genetically heterozygous methylation loci.
+
+This is the so-called $\beta$-value, which is commonly interpreted as an estimate of $B_{i}$, the proportion of cells in the sample that are methylated at the $i^{th}$ locus.
+In Section \ref{sec:estimating_B} I discuss this interpretation and other estimators of the 'methylation level' at a locus.
 
 Again, these definitions can be extended from 1-tuples to m-tuples.
-The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$.
+The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$ and has sequencing depth $d_{(i, i + 1, \ldots, i + \text{m} - 1)} = |\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}|$.
 A single read containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, and the observed methylation state is given by:
@@
 
-Note that we do not know from which haplotype, $h$, each read came from, only that all methylation loci in the read came from the same DNA fragment.
+Note that we do not know from which DNA fragment, $h$, each read came from, only that all methylation loci in the read came from the same DNA fragment.
 
-By counting over the number of reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ we obtain the number of reads contain each methylation pattern at that m-tuple.
+By summing over the reads containing the m-tuple, $(i, i + 1, \ldots, i + \text{m} - 1)$, we obtain the number of reads containing each methylation pattern at that m-tuple.
 Here are the definitions for $\text{m} = 2$:
@@
 
-We can extend the $\beta_{i}$ values to m-tuples. Here are the definitions for $\text{m} = 2$:
+As we did for $B$-values, we can extend the $\beta_{i}$ values to m-tuples. Here are the definitions for $\text{m} = 2$:
 
 \begin{align*}
-  \beta_{(i, i + 1)}^{mm} &= \frac{mm_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{mu} &= \frac{MU_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{um} &= \frac{UM_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{uu} &= \frac{UU_{(i, i + 1)}}{mm_{(i, i + 1)} + mu_{(i, i + 1)} + um_{(i, i + 1)} + uu_{(i, i + 1)}}
+  \beta_{(i, i + 1)}^{mm} &= \frac{mm_{(i, i + 1)}}{d_{(i, i + 1)}} \\
+  \beta_{(i, i + 1)}^{mu} &= \frac{MU_{(i, i + 1)}}{d_{(i, i + 1)}} \\
+  \beta_{(i, i + 1)}^{um} &= \frac{UM_{(i, i + 1)}}{d_{(i, i + 1)}} \\
+  \beta_{(i, i + 1)}^{uu} &= \frac{UU_{(i, i + 1)}}{d_{(i, i + 1)}}
 \end{align*}
@@
 
-One final definition is the average methylation of each read, which is used by \citet{Landan:2012kp}.
+One final definition is the average methylation level of each read, which is used by \citet{Landan:2012kp}.
 For each read, $z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, the average methylation of the read, $\zeta_{z}$, is defined as the proportion of methylation loci in the read that are methylated.
@@
 All methylation loci are defined with respect to the sample's genome.
-The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth}$ are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.
+The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth}$ are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors.
 
@@
 
-The genome of a diploid organism has sites that are heterozygous due to differences between the maternal and paternal chromosomes.
-Such heterozygous loci are sometimes also methylation loci.
-For example, a locus where the maternal chromosome is a CpG and the paternal chromosome is an ApG.
+The genome of a diploid organism has sites that are genetically heterozygous due to differences between the maternal and paternal chromosomes.
+Such heterozygous loci are sometimes also methylation loci; for example, a locus where the maternal chromosome is a CpG and the paternal chromosome is an ApG.
 In effect, the maternal and paternal chromosomes within the sample have different $\mathcal{I}^{truth}$.
@@
 The number of these genetically heterozygous methylation loci is often small enough not to worry about.
-However, in some studies, such as those of allele-specific methylation, these loci can be very important and might be identified by calling heterozygous genetic variants using `Bis-SNP` or from WGS of the sample.
+However, in some studies, such as those of allele-specific methylation, these loci can be very important and should first be identified by calling heterozygous genetic variants using `Bis-SNP` or from whole-genome sequencing of the sample.
 In practice, the existence of such loci is often ignored.
@@
 
-From a purely notational perspective, the move from a single sample to multiple samples simply requires an additional subscript, $j = 1, \ldots, n$, where $n$ is the number of samples.
-This defines the three levels in the hierarchy of a typical experiment -- DNA fragments ($h$), methylation loci ($i$) and samples ($j$).
-For example, $\mathcal{I}_{j}$ is the set of methylation loci in the $j^{th}$ sample and $\beta_{i, j}$ is the beta value for the $i^{th}$ locus in the $j^{th}$ sample.
+From a purely notational perspective, the move from a single sample to multiple samples simply requires an additional subscript, $j = 1, \ldots, N_{samples}$, where $N_{samples}$ is the number of samples.
+This defines the three levels in the hierarchy of a typical experiment: DNA fragments ($h$), methylation loci ($i$) and samples ($j$).
+For example, $Z_{h, i, j}$ is the methylation state on the $h^{th}$ DNA fragment at the $i^{th}$ methylation locus in the $j^{th}$ sample and $\beta_{i, j}$ is the $\beta$-value for the $i^{th}$ locus in the $j^{th}$ sample.
 
@@
 For example, in a two-group experiment $X_{j} = 1$ if the sample is from group $1$ and $X_{j} = 0$ if the sample is from group $2$.
-We can also include covariates in the standard way by allowing $X_{j}$ to be a row vector, $X_{j} = (x_{1, j}, \ldots, x_{P, j})$, where $x_{p, j}$ encodes the information on the $p^{th}$ covariate for the $j^{th}$ sample.
+We may also include covariates in the standard way by allowing $X_{j}$ to be a row vector, $X_{j} = (x_{1, j}, \ldots, x_{P, j})$, where $x_{p, j}$ encodes the information on the $p^{th}$ covariate for the $j^{th}$ sample.
 
@@
 
-Each sample has their own set of methylation loci, that is, $\mathcal{I}_{j}$ differs between $j$.
+Each sample has its own set of methylation loci, that is, $\mathcal{I}_{j}$ differs across $j$.
 Furthermore, sequencing coverage varies from sample-to-sample.
@@
 
-In practice, we might choose to study $\mathcal{I}^{common} = \cap_{j} \mathcal{I}_{j}$ or some other intersection of the $\mathcal{I}_{j}$, such as all methylation loci present in at least some fraction of the $n$ samples.
+In practice, we might choose to study $\mathcal{I}^{common} = \bigcap_{j} \mathcal{I}_{j}$ or some other suitably defined intersection of the $\mathcal{I}_{j}$, such as all methylation loci present in at least some fraction of the $N_{samples}$ samples.
 
@@
 The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}}$.
-To do so it is necessary to estimate $M$ and $U$.
 In this section I review various methods for estimating these key parameters.
@@
 It is rare to need direct estimates of $M_{i}$ or $U_{i}$.
-In order to estimate $M_{i}$ and $U_{i}$, the absolute number of methylated and unmethylated DNA fragments at the $i^{th}$ locus, we would also require an estimate of $|\mathcal{H}_{i}|$, the number of DNA fragments containing the $i^{th}$ locus.
-Rather, we are generally interested in estimating the proportion of reads that are methylated, $B_{i} = \frac{M_{i}}{M_{i} + U_{i}}$.
+In order to estimate $M_{i}$ and $U_{i}$, the absolute number of methylated and unmethylated DNA fragments at the $i^{th}$ locus, we would also require an estimate of the number of DNA fragments containing the $i^{th}$ locus, $H_{i}$.
+Rather, we are generally interested in estimating the proportion of reads that are methylated, $B_{i} = \frac{M_{i}}{M_{i} + U_{i}}$, which does not require an estimate of $H_{i}$.
 
@@
 
-The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,][]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
-The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).
+The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{d_{i}}$, which has been widely used \citep[e.g.,][]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).
+The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling and then counting the number of reads with each methylation state (see Section \ref{sec:methylation_calling}).
 
-Under a Binomial model for the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample, $M_{i, j} | n_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.
+$\beta_{i}$ is the maximum likelihood estimator of $B_{i}$ under a (conditional) binomial model for the number of methylated reads at the $i^{th}$ locus, $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
 
@@
 `BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$.
-Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$.
+Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{d_{i, j}}$.
 Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
@@
 Both `bsseq` and `BiSeq` use a binomial local likelihood smoother.
-In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$.
+In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
 The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.
@@
 In both `bsseq` and `BiSeq` the raw $\beta$-values are weighted according to the binomial likelihood and a kernel function.
-The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the kernel gives greater weight to those $\beta_{i, j}$ near the centre of the window.
+The binomial likelihood weights $\beta_{i}$ inversely to their standard error, $se(\beta_{i})$, and the kernel gives greater weight to those $\beta_{i}$ near the centre of the window.
 \citet{Lacey:2013iy} note that loci with very high sequencing coverage will strongly influence the smoother, potentially biasing estimates at neighbouring loci with lower coverage.
 
-`bsseq` assumes that for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$.
+`bsseq` assumes for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$.
 In contrast, `BiSeq` first creates clusters of CpGs and only assumes that the underlying methylation level is smooth at positions within each cluster.
@@
 \citet{Lacey:2013iy} kept the default minimum window size of 2000 bp but reduced the minimum number of CpGs per window to 50 from the default of 70.
+Nevertheless, the fact remains that `bsseq` is designed for analysing whole-genome bisulfite-sequencing data and not RRBS, which puts it at a disadvantage in these comparisons.
 
@@
 
-Under the binomial model, ($M_{i, j} | n_{i, j}) \eqd Bin(n_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt(\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}})$ \citep{Hansen:2012gr}.
+Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Bin(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}}}$ \citep{Hansen:2012gr}.
 The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample.
@@
 
-\subsection{Empirical distributions}
+\subsection{Empirical distributions of $\beta$-values}
 
@@
 Some values will be missing, either because there was insufficient sequencing coverage to estimate a $\beta$-value or because that locus is not a cytosine for that sample.
-This matrix might be visualised to learn about the distribution of methylation levels, either row-wise, to learn about the variability across samples, or column-wise, to learn about the variability within samples.
+This matrix might be visualised to learn about the distribution of methylation levels, either row-wise (to learn about the variability across samples) or column-wise (to learn about the variability within samples).
 
@@
 
-Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
+Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
 Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.
 
-What is immediately clear is that these distributions are bimodal --- most CpGs are highly methylated or lowly methylated.
-The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels.
+What is immediately clear is that these distributions are bimodal: most CpGs are highly methylated or lowly methylated.
+The exception is the _E16.5\_male\_1_ sample from the _Seisenberger_ data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels.
 The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice.
@@
 
-Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _IMR90\_cell\_line_ from the Ziller dataset --- or cancer cells lines and tissue --- _HepG2\_cell\_line_, _Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_ from the Ziller dataset.
-Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the Seisenberger dataset also displays greater levels of intermediate methylation.
+Almost all the samples with significant intermediate methylation are either somatic cell lines (_ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the _Lister_ dataset; _IMR90\_cell\_line_ from the _Ziller_ dataset) or cancer cells lines and tissue (_HepG2\_cell\_line_, _Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_ from the _Ziller_ dataset).
+Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the _Seisenberger_ dataset also displays greater levels of intermediate methylation.
 This sample was also created by pooling DNA from multiple mice, which may explain the extra variability in the $\beta$-value distribution.
@@
 Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation.
-Likewise, the various frontal cortex samples in the Ziller dataset --- _Frontal\_cortex\_normal\_1_, _Frontal\_cortex\_normal\_2_, _Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_ --- which includes both 'normal' and Alzheimer's samples, have very low levels of intermediate methylation.
-This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_.
-Naively, a cell line is a 'pure' cell population, however, the DNA methylation data clearly reveal widespread differences between cells in these populations.
+Likewise, the various frontal cortex samples in the _Ziller_ dataset, which includes both 'normal' and Alzheimer's samples (_Frontal\_cortex\_normal\_1_, _Frontal\_cortex\_normal\_2_, _Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_), have very low levels of intermediate methylation.
+This raises the question as to whether the widespread partial methylation observed in the somatic samples from the _Lister_ dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_.
+Naively, a cell line is a 'pure' cell population, however, the DNA methylation data clearly reveal widespread cellular heterogeneity of DNA methylation.
 
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
-Densities were normalised so that the maximum value for each sample is 1.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
+Densities are normalised so that the maximum value for each sample is 1.
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data.
-Densities were normalised so that the maximum value for each sample is 1.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
+Densities are normalised so that the maximum value for each sample is 1.
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data.
-Densities were normalised so that the maximum value for each sample is 1.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
+Densities are normalised so that the maximum value for each sample is 1.
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
-Densities were normalised so that the maximum value for each sample is 1.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
+Densities are normalised so that the maximum value for each sample is 1.
 Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
@@
 
-The bimodality of the genome-wide $\beta$-value distributions is driven by the fact that most CpGs is CpG islands are unmethylated whereas those outside of the islands are mostly methylated.
-Figures \ref{fig:EPISCOPE_plotBetaDensity_CGI}, \ref{fig:Lister_plotBetaDensity_CGI}, \ref{fig:Seisenberger_plotBetaDensity_CGI} and \ref{fig:Ziller_merged_plotBetaDensity_CGI} show the kernel density plots of the $\beta$-value distributions stratified by CpG island status for the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively.
+The bimodality of the genome-wide $\beta$-value distributions is driven by the fact that most CpGs in CpG islands are unmethylated whereas those outside of the islands are mostly methylated.
+Figures \ref{fig:EPISCOPE_plotBetaDensity_CGI}, \ref{fig:Lister_plotBetaDensity_CGI}, \ref{fig:Seisenberger_plotBetaDensity_CGI} and \ref{fig:Ziller_merged_plotBetaDensity_CGI} show the kernel density plots of the $\beta$-value distributions stratified by CpG island status for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
 These distributions are normalised so that each density has a maximum value of 1.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _EPISCOPE_ data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:EPISCOPE_plotBetaDensity_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _Lister_ data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Lister_plotBetaDensity_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _Seisenberger_ data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Seisenberger_plotBetaDensity_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data, stratified by whether the CpG is in a CpG island.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _Ziller_ data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10\times$ sequencing coverage are included.
-Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:Ziller_merged_plotBetaDensity_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _EPISCOPE_ data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _Lister_ data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _Seisenberger_ data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Zilelr data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _Ziller_ data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 
-\subsubsection{Strand-specific summaries}
+\subsubsection{Strand-specific $\beta$-values}
 
@@
 
-There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the low strand-correlations at these lower coverage cutoffs.
-Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 - 0.9$, however, there are some notable exceptions.
+There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the smaller strand-correlations at these lower cutoffs.
+Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 -- 0.9$, with some notable exceptions.
 
-Some of the embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$.
+Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5 -- 0.7$.
 The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$).
@@
 
-All three Seisenberger samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell _J1\_1_.
+All three _Seisenberger_ samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell, _J1\_1_.
 However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.
 
-Overall, it seems that with the exception of embryonic stem cells, most samples have very highly correlated strand-specific CpG $\beta$-values, which means that these data can safely be combined across strands.
+Overall, with the exception of embryonic stem cells, it seems that most samples have highly correlated strand-specific CpG $\beta$-values, which means that these data can safely be combined across strands.
 However, it remains a good idea to first check this assumption prior to combining data across strands.
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the EPISCOPE dataset.
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:EPISCOPE_strand_correlations")
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Lister dataset.
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Lister_strand_correlations")
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Seisenberger dataset.
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Seisenberger_strand_correlations")
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Ziller dataset.
+xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_merged_strand_correlations")
@@
 
-\subsection{Interpretation}\label{sec:beta-value_interpretation}
+\subsection{Interpretation of $\beta$-values}\label{sec:beta-value_interpretation}
 
@@
 In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
-However, as can be seen from Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
+However, as can be seen from Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_plotBetaFreqPoly}, the bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
 To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
@@
 
-\citeauthor{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" \citep[Suppmenetary Material]{Lister:2008bh}.
+\citeauthor{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" \citep[Supplementary Material]{Lister:2008bh}.
 In the language of these papers, a "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[Supplementary Material]{Lister:2009hy}.
@@
 
-Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows.
-For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Pr(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
+Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, nor is any code made available, I believe the method is as follows[^mathematical_description].
+For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = d_{i}} Pr(X = k)$, where $X = Binom(d_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
 Any site with an FDR-adjusted P-value below a threshold was declared a "methylcytosine"[^fdr]
@@
 I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated.
-This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A.
-thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
+This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _Arabidopsis thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation).
 This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.
@@
 The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error rates estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}.
-It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.
+It is not clear how the sequencing error rates were estimated, particularly since the base qualities are not included in the data available from the website.
 
@@
 
-Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values.
-For example, in \citet[Supplementary Figure 2a of][]{Lister:2009hy} the authors use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
+Furthermore, referring to CpGs as "methylcytosines" results in a loss of information since two "methylcytosines" may have very different $\beta$-values.
+For example, in \citet[Supplementary Figure 2a of ][]{Lister:2009hy} the authors use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates.
 A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.
@@
 Instead, it is better to summarise the methylation level at a CpG by a $\beta$-value since this has a natural interpretation as the estimated proportion of haploid genomes in the sample that are methylated at that CpG.
-Unfortunately, $\beta$-values are not without their own issues, as discussed below.
+Unfortunately, $\beta$-values are not without their own issues, as discussed in Section \ref{sec:beta_biases}.
 
-\subsection{Correlations}
+\subsection{Spatial correlations of $\beta$-values}
 
-Many researchers have observed that DNA methylation is spatially correlated along the genome, \citep[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}.
-I call this correlation of methylation levels _co-methylation_.
+Many researchers have observed that DNA methylation is spatially correlated along the genome, \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}.
+I call this spatial correlation of methylation levels _co-methylation_.
 
-Just as we can explore the distribution of DNA methylation levels within a sample or between samples, so too can we explore co-methylation within a sample or between samples.
-I examine this in detail in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
+I examine in detail the spatial correlations of $\beta$-values in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
 
-\subsection{Biases}
+\subsection{Sources of bias in $\beta$-values}\label{sec:beta_biases}
 
-The natural interpretation of $\beta_{i, j}$ as the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample will be biased if the probability of sequencing a fragment with a methylated site is different from the probability of sequencing a fragment with an unmethylated site.
+The natural interpretation of $\beta_{i}$ is the average level of methylation at the $i^{th}$ locus.
+However, this will be biased if the probability of sequencing a fragment with a methylated site is different from the probability of sequencing a fragment with an unmethylated site.
 In fact, it has been shown that methylated DNA is overrepresented in bisulfite-sequencing data due, with the problem exacerbated by higher rounds of PCR amplification and dependent on the bisulfite-conversion protocol \citep{Ji:2014fq}.
@@
 
-Lab-based solutions to overcome these biases exist for targeted bisulfite-sequencing, but are technically difficult and expensive to extend to whole-genome studies.
-\citet{Ji:2014fq} propose potential computational corrections for these biases but as yet these have not been implemented.
+Lab-based solutions to overcome these biases exist for targeted bisulfite-sequencing, but are technically difficult and their cost prohibits their extension to whole-genome bisulfite-sequencing \citep{Ji:2014fq}.
+Computational correction for these biases have been proposed \citep{Ji:2014fq}, but as yet these have not been implemented in any available software.
 
-\subsection{Transformations}\label{sec:transformations}
+\subsection{Transformations of $\beta$-values}\label{sec:transformations}
 
@@
 However, they are not necessarily the best unit for statistical inference.
-This is because a $\beta$-value is an estimate of a proportion and there are a well-known challenges when working with proportion data, such as:
+This is because a $\beta$-value is an estimate of a proportion and there are a well-known statistical challenges when working with proportion data, including:
 
 1.
-The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$.
-Taking the derivative of this, we see that the maximum standard error, $\sqrt{\frac{0.25}{m + u}}$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
+The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{d}}$.
+Taking the derivative of this with respect to $\beta$, we see that the maximum standard error, $\sqrt{\frac{0.25}{d}}$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
 2.
-We need to more than just the $\beta$-value to have a sense of how precise an estimate it is.
+We need to know more than just the $\beta$-value to have a sense of how precise an estimate it is.
 Essentially, we need to also know the sequencing coverage of the methylation loci.
@@
 
-However, the general use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \citep{Warton:2011gm}.
+However, the use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \citep{Warton:2011gm}.
 
@@
 An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2} \Big ( \frac{\beta}{1 - \beta} \Big ) = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero.
-Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \citep[e.g., ]{Smyth:2005ta}.
+$\mathcal{M}$-values are also known as log-ratios and are widely used in the analysis of RNA expression two-colour microarrays \citep[e.g.,][]{Smyth:2005ta}.
 
-\citet{Du:2010dc} advocate for the use of $\mathcal{M}$-values in statistical analyses of methylation levels from microarray data.
-The main reason for this is that $\mathcal{M}$-values are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values.
-As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformatics and genomics researchers makes it a favourable choice.
-Nonetheless, \citet{Du:2010dc} also advocate that the results of analyses should be reported as $\beta$-values, owing to their "more intuitive biological interpretation".
+\citet{Du:2010dc} advocate for the use of $\mathcal{M}$-values for conducting differential methylation analysis from microarray data[^interpretation].
+The main reason they advocate for the use of $\mathcal{M}$-values is that they are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values.
+As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformaticians and genomics researchers makes it a favourable choice.
+
+[^interpretation]: \citet{Du:2010dc} also recommend that the results of analyses are reported as $\beta$-values owing to their "more intuitive biological interpretation".
 
@@
 \section{Summary}
+
+__UP TO HERE: Need a strong summary/conclusion. Emphasise the novel aspects of this chapter and the wide variety of datasets used.__
 

### 2015-04-14 Rmarkdown/comethylation.Rmd
Problems with simple log odds-ratios.

@@
 \FloatBarrier
+
 \subsubsection{$NIL \geq 0$}\label{sec:cometh_nil_geq0}
@@
 Until such data are available, we can try to learn more about plausible models of DNA co-methylation through computational methods such as the simulation method described in the next chapter.
+
+# Problem with LOR estimates
+
+The simple estimator of within-fragment co-methylation is biased when the marginal probability of methylation is close to the boundary (0 or 1) unless the 2-tuple is sequenced at very high depth.
+I demonstrate this by simulating methylation patterns at 2-tuples as independent $binomail(1, p)$ random variables and showing that the simple LOR estimator is positively biased.
+
+```{r}
+# Function to simulate n 2-tuples with d sequencing coverage and marginal 
+# probabilities p1, p2
+f <- function(n, d, p1, p2) {
+  sapply(seq_len(n), function(i, d, p1, p2) {
+  z1 <- factor(rbinom(d, 1, p1), levels = c(0, 1))
+  z2 <- factor(rbinom(d, 1, p2), levels = c(0, 1))
+  x <- xtabs(~ z2 + z1, drop.unused.levels = FALSE)
+  log2(x[1, 1] + 0.5) + log2(x[2, 2] + 0.5) - 
+    log2(x[1, 2] + 0.5) - log2(x[2, 1] + 0.5)
+  }, d = d, p1 = p1, p2 = p2)
+}
+
+# Vary p1 with p2 = p1 at 30x coverage
+d <- 30
+n <- 1000
+p <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+x <- lapply(p, function(p) {
+  f(n, d, p, p)
+})
+par(mfrow = c(3, 3))
+mapply(function(x, p) {
+  hist(x, main = paste0("p1 = p2 = ", p), xlim = c(-10, 10), breaks = 20)
+  abline(v = 0, col = "red", lty = 2)
+  }, x = x, p = p)
+title(paste("d = ", d), outer = TRUE, line = -1)
+
+# Vary p1 with p2 = 1 - p1
+d <- 30
+n <- 1000
+p <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+x <- lapply(p, function(p) {
+  f(n, d, p, 1 - p)
+})
+par(mfrow = c(3, 3))
+mapply(function(x, p) {
+  hist(x, main = paste0("p1 = ", p, ", p2 = ", 1 - p), xlim = c(-10, 10), breaks = 20)
+  abline(v = 0, col = "red", lty = 2)
+  }, x = x, p = p)
+title(paste("d = ", d), outer = TRUE, line = -1)
+
+# Vary p1 with p2 = 0.5
+d <- 30
+n <- 1000
+p <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+x <- lapply(p, function(p) {
+  f(n, d, p, 0.5)
+})
+par(mfrow = c(3, 3))
+mapply(function(x, p) {
+  hist(x, main = paste0("p1 = ", p, ", p2 = ", 0.5), xlim = c(-10, 10), breaks = 20)
+  abline(v = 0, col = "red", lty = 2)
+  }, x = x, p = p)
+title(paste("d = ", d), outer = TRUE, line = -1)
+
+# Vary p1 with p2 = p1 at 1000x coverage
+d <- 1000
+n <- 1000
+p <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+x <- lapply(p, function(p) {
+  f(n, d, p, p)
+})
+par(mfrow = c(3, 3))
+mapply(function(x, p) {
+  hist(x, main = paste0("p1 = p2 = ", p), xlim = c(-10, 10), breaks = 20)
+  abline(v = 0, col = "red", lty = 2)
+  }, x = x, p = p)
+title(paste("d = ", d), outer = TRUE, line = -1)
+
+```

### 2015-04-15 Rmarkdown/biology_background.Rmd
Minor formatting

@@
 The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin.
-The linker sequences are between $20--60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
+The linker sequences are between $20$ to $60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
 
@@
 Non-CpG methylation is very common in other organisms, such as plants.
-For example, \citet{Lister:2008bh} found that in the widely-studied _Arabidopsis Thaliana_ that $45\%$ of cytosines that displayed statistically significant evidence of methylation are at CHG or CHH loci.
+For example, \citet{Lister:2008bh} found that in the widely-studied _Arabidopsis thaliana_ that $45\%$ of cytosines that displayed statistically significant evidence of methylation are at CHG or CHH loci.
 However, they also found that the level of methylation at non-CpG loci is typically lower than that observed at CpG dinucleotides.
@@
 Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples.
-Furthermore, depending on the choice of several sequencing parameters, approximately $35--72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_withou_cpgs}), which might be considered a gross waste of resources for some experiments.
+Furthermore, depending on the choice of several sequencing parameters, approximately $35$ to $72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.
 
@@
 See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
-\label{tab:reads_withou_cpgs}
+\label{tab:reads_without_cpgs}
 \end{table}
@@
 
-Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1--5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
+Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1$ to $5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
 However, the minimal amount of DNA is being reduced with each technological advance.

### 2015-04-15 Rmarkdown/paper_review.Rmd
Pernickety.
Minor updates.
Minor formatting

@@
 \cite{Cokus:2008fc} published the BS-seq protocol for performing WGBS.
-Most of the results in this paper used BS-seq of wild-type _Arabidopsis Thaliana_ and DNMT mutants.
+Most of the results in this paper used BS-seq of wild-type _Arabidopsis thaliana_ and DNMT mutants.
 They also report limited results from low-coverage BS-seq of mouse germ cells but these are not relevant to my discussion.
@@
 \cite{Lister:2008bh} published the methylC-seq protocol for performing WGBS.
-Like \cite{Cokus:2008fc}, \cite{Lister:2008bh} studied wild-type and DNMT-defective mutant _Arabidopsis Thaliana_ strains.
+Like \cite{Cokus:2008fc}, \cite{Lister:2008bh} studied wild-type and DNMT-defective mutant _Arabidopsis thaliana_ strains.
 Also like \cite{Cokus:2008fc}, much of the analysis is by visual comparison of summarised data.
@@
 \cite{Lacey:2013iy} details a modeling and inference framework for RRBS data, as well as a method for simulating RRBS data.
-Here I only describe the modelin framework and reserve the description and discussion of the simulation framework for __CHAPTER__.
+Here I only describe the modeling framework and reserve the description and discussion of the simulation framework for __CHAPTER__.
 
@@
 
-## \cite{Rackham:2015bv}
+## \citet{Rackham:2015bv}
 
@@
   - Normal-mixture
+
+### \citet{Rackham:2015bv}
+
+Simulate WGBS.
+
+1.
+Simulate CpG positions using a homogeneous first-order HMM with three states (CGI, shores and deserts).
+According to the transition matrix, it is impossible to go from a CGI to a desert without going via shore (probably reasonable).
+2.
+Simulate the methylation state of each CpG using an inhomogeneous first-order HMM with four states ('methylated', 'transition1', 'transition2' and 'de-methylated').
+The inhomogeneity is to induce some spatial dependence amongst nearby CpGs such that the transition probabilities are modulated by the distance between CpGs.
+According to the transition matrix, it is impossible to go from methylated to de-methylated without going via transition2.
+Similarly, it is impossible to go from de-methylated to methylated without going via transition1.
+It is also impossible to go between the two transition states without going via one of methylated or de-methylated.
+The modulation of transition probabilities as a function of distance is via an exponential decay.
+The rate of decay depends only on the distance between CpGs and not their context (e.g., identical between CGI and desert for a given distance).
+3. Simulate read depth from a Poisson distribution.
+4.
+Simulate read counts from a binomial or (truncated) negative binomial distribution.
+That is, the binomial (or truncated negative binomial) distribution is the emission distribution of the HMM in (2).
+The probability of success is identical within all 'methylated' states, 'de-methylated' states, and within 'transition1' and 'transition2' states, except for some random Gaussian noise (i.i.d.
+for each CpG).
+The average methylation level in 'de-methylated' regions is defined as $p_{de-methylated} = median(\{ \beta_{i}: 0 \leq \beta_{i} < 0.5 \})$; for 'methylated' regions it is defined as $p_{methylated} = median( \{ \beta_{i} : 0.5 \leq \beta_{i} \geq \leq 1 \}); and for 'transition1' and 'transition2' regions it is defined as $p_{transition} = \frac{p_{methylated} + p_{de-methyalted}}{2}.
+$\beta_{i}$ are computed from real data.
+5. DMRs are induced by defining the expected difference in average methylation between cases and controls for randomly sampled regions.

### 2015-04-15 Rmarkdown/comethylation_review.Rmd
Figure fixing.
A few TODOs
Working through edits to co-methylation review

@@
 
-\section{Introduction}
+__TODO: Review co-methylation in "Bell, J. T. et al. DNA methylation patterns associate with genetic and gene expression variation in HapMap cell lines. Genome Biol 12, R10 (2011).__"
 
+\begin{chapabstract}
 Co-methylation is the dependence structure of DNA methylation data.
@@
 I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.
+\end{chapabstract}
 
@@
 I define co-methylation as the dependence structure of DNA methylation data.
-This broad definition encompasses several concepts that have variously been described as 'co-methylation', or similar.
-My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha} or "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; it also includes the "correlation of two [loci] across many sample" \citep{Akulenko:2013fl} and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
-A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.
+This broad definition encompasses several similar concepts previously described in the literature.
+These include: "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}; "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; the "correlation of two [loci] across many samples" \citep{Akulenko:2013fl}; and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
+
+It is important to understand that there exists both within-sample and between-sample co-methylation.
 
 Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample.
-Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}.
-Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.
+Many researchers have observed that there is a strong spatial dependence of DNA methylation along the genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}.
+These studies also found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.
 
@@
 For example, \citet{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples.
-Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.
+It is worth noting that measures of within-sample co-methylation can themselves be compared between samples.
 
-[^concordance]: One measure of between-sample correlation of methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$.
+[^concordance]: One measure of between-sample correlation of aggregate methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$.
 This measure is often reported as evidence for the "concordance" of methylation levels between replicates, but is not what I consider a form of co-methylation.
@@
 For example, we might identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci.
-This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.
+This idea is analogous to the use of tag-SNPs on genotyping microarrays.
 
@@
 
-\section{Correlations of aggregate measurements of methylation}
+\section{Correlations of aggregate methylation levels}
 \label{sec:previous_cor_of_agg_methylation}
 
-To date, most analyses of co-methylation have been based on analysing correlations of aggregate levels of methylation, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
+To date, most analyses of co-methylation have been based on analysing correlatiosn of aggregate methylation levels, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
 These analyses have an implicit assumption of stationarity, that is, that the correlation between two loci depends only on the distance between them and not the actual position of the loci in the genome.
@@
 \citet{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples.
-Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \citep{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \citep{Hansen:2011gu}.
+Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} is often cited as evidence of co-methylation.
+For example, \citet{Hebestreit:2013ko} cite \citet{Eckhardt:2006gh} as evidence that "methylation levels are strongly spatially correlated" and \citet{Hansen:2011gu} cite \citet{Eckhardt:2006gh} as evidence that"proximal CpGs [have] similar methylation levels".
+
 \citet{Eckhardt:2006gh} performed PCR-amplicon bisulfite-sequencing of $2,524$ amplicons on human chromosomes $6$, $20$ and $22$.
 PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data.
-Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays.
-Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.
+Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern whole-genome bisulfite-sequencing assays.
+The samples are a variety of human tissue types from multiple donors of different ages, both male and female.
 
-As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance".
+As part of their analysis, the authors explore co-methylation, which they define as "the relationship between the degree of methylation over distance".
 They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $>2,000$ bp" and reference the plot reprinted in Figure \ref{fig:Eckhardt} in support of this claim.
-In Figure \ref{fig:Eckhardt}, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp, which is not a correlation in the usual sense of the word.
+In Figure \ref{fig:Eckhardt}, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp.
+This is not a correlation in the usual sense of the word.
 Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 
-[^beck]: An email discussion with the senior author, Stephan Beck, was unable to clarify the definition or method.
+[^beck]: An email discussion with the senior author, Professor Stephan Beck (University College London), was unable to clarify the definition or method.
 
@@
 Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels.
-This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$.
+This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h', i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$.
 In this setting we can compute the theoretical $PIM^{\ast}$ as follows:
@@
 
-This result is proven in Appendix \ref{app:bernoulli_equal}.
+The proof of this result is given in Appendix \ref{app:bernoulli_equal}.
 
@@
 2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
-3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, \allowbreak Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
 
@@
 
-Quite apart from what exactly is plotted in Figure \ref{fig:Eckhardt}, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis.
-Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure \ref{fig:Eckhardt} does not represent any one sample but a pool of samples.
+Quite apart from what exactly is plotted in Figure \ref{fig:Eckhardt}, the analysis is based on data from only $2,524$ amplicons.
+These limited data cannot give a genome-wide picture of co-methylation.
+Furthermore, the data come from a variety of tissues and donors, which makes it difficult to know how applicable these results are for a particular tissue.
+Caution is warranted in extrapolating these results to general statements about the correlation of aggregate methylation levels.
 
@@
 \citet{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing.
-They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_.
+They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis thaliana_.
 They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis].
-The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance.
+The autocorrelation computes, for a set of distances ($0-5,000$ bp), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance.
 To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.
@@
 
-The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only).
+The autocorrelation plots showed two periodicities, one of $\approx 167$ bp and the other of $10$ bp (CHH only).
 Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.
 
-[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}
+[^CHH_only]: The $10$ nucleotide period was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}
 
-The $167$ nucleotide periodicity is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al.
+The $167$ bp period is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al.
 hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.
 
-The $10$ nucleotide periodicity is equal to the length of one helical DNA turn.
+The $10$ bp period is equal to the length of one helical DNA turn.
 From this result Cokus et al.
-hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8-10$ nucleotides apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_).
-The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})
+hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8$ to $10$ bp apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_).
+The $10$ bp CHH-only period in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})
 
 [^comethylation_def2]: This is also referred to as "co-methylation" by \citet{Jurkowska:2011dt}, here meaning the methylation of multiple loci in a single event.
-\citet{Jurkowska:2011dt} reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.
+\citet{Jurkowska:2011dt} review a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8$ to $10$ bp; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.
 
-\subsection{\citet{Chodavarapu:2010iqa}\label{sec:chodavarapu}
+\subsection{\citet{Lister:2009hy}}\label{sec:lister_2009}
+
+\citet{Lister:2009hy} report an $8$ to $10$ bp period in the occurrence "methylcytosines".
+Recall that \citeauthor{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a binomial test (see Section \ref{sec:beta-value_interpretation} for details and a criticism of this concept).
+With its reference to "methylcytosines", this may first appear to be a within-fragment analysis of co-methylation, but it is in fact based on $\beta$-values.
+However, unlike other methods described in this section, the analysis is not based on a correlation of these $\beta$-values, but rather on patterns in the spacing of "methylcytosines" along the genome.
+This leads to problems.
+
+\citet{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$.
+These counts are plotted as a bar chart, with $IPD$ along the x-axis and count on the y-axis, along with a cubic spline fit to the counts.
+This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (reproduced in Figure \ref{fig:Lister_2009_Sup_Fig_9}).
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.pdf}
+\caption{Spacing of Adjacent Methylcytosines in Different Contexts.
+Prevalence of mCHG/mCHH sites (y-axis) as a function of the number of bases between adjacent mCHG/mCHH sites (x-axis) based on all non redundant pair-wise distances up to $50$ nt in exons, introns and random sequences.
+The blue line represents smoothing by cubic splines.
+Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Lister:2009hy}), copyright (2009)}
+\label{fig:Lister_2009_Sup_Fig_9}
+\end{figure}
+
+[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Supplementary Figure 9, but I am not certain.
+
+An $8$ to $10$ bp period in the fitted cubic spline is observed in most, but not all, of these nine plots.
+This periodicity is very weak in several contexts and is absent for exonic CpGs.
+The weakness of the signal is partially attributable to the small number of "methylcytosines" in the CHG and CHH contexts in human samples.
+More importantly, the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs.
+Unlike the $8$ to $10$ nucleotide period, the $3$ nucleotide period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it.
+\citeauthor{Lister:2009hy} do not comment on the $3$ bp period.
+
+Based on these nine plots, \citet{Lister:2009hy} draw parallels to the $10$ bp period in the autocorrelation of CHH $\beta$-values observed by \citet{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites".
+However, I suspect that the results in these nine plots are in fact driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines.
+That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis.
+To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$.
+If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s.
+Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.
+
+In fact, there are well known periodicities in the distribution of nucleotides in genomes.
+For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ nucleotide period and the latter the $3$ bp period for exonic CpGs.
+
+\subsection{\citet{Chodavarapu:2010iqa}}\label{sec:chodavarapu}
 
@@
 The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes.
-A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
+A $10$ bp periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
 Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.
 
-In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ nucleotide periodicity.
-In fact, these results revealed that the $10$ nucleotide is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA.
-This led \citet{Chodavarapu:2010iqa} to discard their earlier hypothesis as to the source of the $10$ nucleotide periodicity \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.
+In addition to the genome-wide analyses, the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of the _Arabidopsis_ sample and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ bp periodicity.
+In fact, these results revealed that the $10$ bp period is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA.
+This led \citeauthor{Chodavarapu:2010iqa} to discard their earlier hypothesis[^same_authors] as to the source of the $10$ bp period \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__ [emphasis added]" and not just for the DRM2 methyltransferase.
+
+[^same_authors]: Many of the same authors contributed to \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.
 
@@
 \citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
-The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", citing their results as being consistent with those in \citet{Eckhardt:2006gh}.
-In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements.
+\citeauthor{Li:2010fb} conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", and that their results are consistent with those in \citet{Eckhardt:2006gh}.
+In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated regions, coding DNA sequences and various repetitive elements.
 Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.
@@
 
-Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp.
+Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are estimates of the correlation of aggregate methylation levels over the range $0$ to $1000$ bp.
 Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs.
-If all pairs were used this $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome.
-To emphasise, the correlation shown by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.
+If all pairs were used, $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome.
+To emphasise, the correlation computed by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 
-The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \citet{Cokus:2008fc} (reproduced in Figure \ref{fig:Li7}).
+The plot of genome-wide correlations (panel (a) in Figure \ref{fig:Li6}) contains a period of approximately $170$ bp, very similar to the $167$ bp period identified in _Arabidopsis_ by \citet{Cokus:2008fc}.
 This is further evidence that nucleosome spacing effects the strength of co-methylation.
-However, \citet{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.
+However, \citet{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ bp period reported by \citet{Cokus:2008fc} and \citet{Chodavarapu:2010iqa}.
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation.
+I do not have the data used in \citet{Lyko:2010dr}, but the following example supports my conclusion.
+This example uses the chromosome 22 data from the _FF_ sample in the _Lister_ dataset (see Section \ref{sec:Lister_dataset} for full details of this sample).
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 
-While I don't have the data used in \citet{Lyko:2010dr}, I believe an example using the chromosome 22 data from the _FF_ sample demonstrates the point.
 Let $\bm{\beta^{\ast}}$ be the augmented vector of $\beta$-values, where:
@@
 
-Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$.
-We see that it is remarkably similar to that panel (a) of Figure \ref{fig:Lyko} --- the correlation first increases then decreases and the maximum correlation is very small, close to zero.
+Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$ for the chromosome 22 data from the _FF_ sample.
+We see that it is remarkably similar to that panel (a) of Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum correlation is very small, close to zero.
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}).
-Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp periodicity is more likely due to the spacing of nucleotides in the genome than anything to do with their methylation state.
+Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of nucleotides in the genome than anything to do with their methylation state.
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 To do so, \citeauthor{Lacey:2013iy} fit a Gaussian variogram model to the CpG $\beta$-values from chromosome 11 of a normal myotube cell line (_MTCTL2_).
-The empirical and fitted variograms asymptote for CpGs separated by around $3000$ bp (Figure 1d of \citet{Lacey:2013iy}, unable to be reproduced due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
+The empirical and fitted variograms asymptote for CpGs separated by approximately $3000$ bp (Figure 1d of \citet{Lacey:2013iy}, unable to reproduce here due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
 
@@
 [^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (see Appendix \ref{app:lacey}).
-
-\subsection{\citet{Lister:2009hy}}\label{sec:lister_2009}
-
-\citet{Lister:2009hy} report an $8-10$ nucleotide periodicity in the occurrence "methylcytosines".
-Recall that \citet{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a Binomial test (see Section \ref{sec:beta-value_interpretation} for details and a criticism of this concept).
-Hence, while this may appear to be a within-fragment analysis of co-methylation with it's reference to "methylcytosines", it in fact  based on $\beta$-values.
-However, unlike the other methods described in this section, the analysis is not based on a correlation of these $\beta$-values, but rather on patterns in the spacing of 'methylcytosines' along the genome, which leads to problems.
-
-\citet{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$.
-These counts are plotted as a bar chart, with $IPD$ along the x-axis and count on the y-axis, along with a cubic spline fit to the counts.
-This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (reproduced in Figure \ref{fig:Lister_2009_Sup_Fig_9}).
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.pdf}
-\caption{Spacing of Adjacent Methylcytosines in Different Contexts.
-Prevalence of mCHG/mCHH sites (y-axis) as a function of the number of bases between adjacent mCHG/mCHH sites (x-axis) based on all non redundant pair-wise distances up to $50$ nt in exons, introns and random sequences.
-The blue line represents smoothing by cubic splines.
-Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Lister:2009hy}), copyright (2009)}
-\label{fig:Lister_2009_Sup_Fig_9}
-\end{figure}
-
-[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Supplementary Figure 9, but I am not certain.
-
-An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots.
-This periodicity is very weak in several contexts and is absent for exonic CpGs.
-The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts, since there are very few "methylcytosines" for these contexts in these human samples.
-Furthermore, the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.
-
-Based on these nine plots, \citet{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \citet{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites".
-However, I suspect that the results in these nine plots are in fact driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines.
-That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis.
-To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$.
-If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s.
-Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.
-
-In fact, there are well known periodicities in the distribution of nucleotides in genomes.
-For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.
 
@@
 More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment.
-It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
-For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure.
-We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
-There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.
+It can be thought of as the dependence structure of the binary stochastic process $\mathbf{Z_{h}} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$.
+There is less previous research studying within-fragment co-methylation than there is studying correlations of aggregate methylation levels.
+In Chapter \ref{chap:co-methylation} I extensively explore within-fragment co-methylation and how different variables affect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
 
-If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci.
-However, the most commonly used technology for high-throughput bisulfite-sequencing, using Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum.
-We are therefore limited to studying co-methylation between methylation loci separated by at most $200-250$ bp.
-Note that assembling these reads into longer contigs will not solve this problem since the assembly cannot guarantee that the assembled reads actually originated from the same cell or haplotype.
-
-Long-read sequencing technologies, such as Pacific Biosciences, Roche 454 and Oxford Nanopore, would overcome this limitation but the lower throughput of these technologies, and the reliability with which they can identify methylcytosines, create new complications.
+If we could sequence entire chromosomes by single reads, then we could analyse co-methylation between any set of methylation loci.
+However, the most commonly used technology for high-throughput bisulfite-sequencing, which is based on Illumina's sequencing technology, can only generate reads that span approximately $200$ to $250$ bp.
+We are therefore limited to studying within-fragment co-methylation between methylation loci separated by at most $200$ to $250$ bp[^assembly].
+Long-read sequencing technologies (e.g., Pacific Biosciences, Roche 454 and Oxford Nanopore) would overcome this limitation but the lower throughput of these technologies, and the reliability with which they can identify methylcytosines, create new complications.
 An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo].
 
-[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons, including the fragility of large DNA fragments treated with sodium bisulfite.
+[^assembly]: NB Assembling these reads into longer contigs will not solve this problem since the assembly does not guarantee that the reads assembled into longer fragments actually originated from the same cell or haplotype.
+
+[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technical reasons, including the fragility of large DNA fragments treated with sodium bisulfite.
 
@@
 
-As already discussed in Section \ref{sec:Cokus}, \citet{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_.
-They also reported they they find this same periodicity "when individual reads [were] examined directly".
+As discussed in Section \ref{sec:Cokus}, \citet{Cokus:2008fc} identify a $10$ bp period in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis thaliana_.
+They also report that they find this same period "when individual reads are examined directly".
 
-The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence.
-For the within-fragment co-methylation results (reproduced in Figure \ref{fig:Cokus_2008_Sup_Fig_14}) \citeauthor{Cokus:2008fc} first identify all reads containing multiple CHH loci.
+The sequencing technology at the time of this publication produced very short reads, on average only $31$ bp of usable sequence.
+For the within-fragment co-methylation results, \citeauthor{Cokus:2008fc} first identify all reads containing multiple CHH loci.
 Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated.
@@
 \citeauthor{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
-This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of normality, and a running average of these averages.
-There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
+The plot includes pointwise $95\%$ confidence intervals, computed under an assumption of normality, and a running average of these averages.
+This plot, reproduced in Figure \ref{fig:Cokus_2008_Sup_Fig_14}, shows evidence of an approximately $10$ bp period, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
 
-\begin{figure}[h!]
+\begin{figure}[!htb]
 \centering
@@
 
-I consider these results to be based on a restricted definition of within-fragment co-methylation.
-To explain, these results tell us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment, but only about the probability that both CHH loci are methylated.
+I consider these results as based on a restricted definition of within-fragment co-methylation.
+To explain, these results tell us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment; they tell us only about the probability that both CHH loci are methylated.
 
 Specifically, \citet{Cokus:2008fc} only estimate the marginal probability that a CHH is methylated, $Pr(Z_{h, i} = 1)$, and the probability that two CHH loci on the same fragment are both methylated, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $i < i'$.
-From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$.
+From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y), x, y \in \{0, 1\}$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$.
 It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
@@
 
-\citet{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans.
-The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field where the methylation state of one loci, $Z_{h, i}$, depends on that of its neighbours either side, $Z_{h, i - 1}$ and $Z_{h, i + 1}$.
+\citet{Lacey:2009by} develop stochastic models of DNA methylation replication during mitosis (cell division) for a single double-stranded DNA molecule in humans.
+The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field.
+Under this model, the methylation state of one locus, $Z_{h, i}$, depends on the methylation states of the adjacent loci, $Z_{h, i - 1}$ and $Z_{h, i + 1}$.
 The physical distances between these neighbours, the $IPD$s, are not considered.
@@
 The neighbouring sites model requires an analysis of within-fragment co-methylation in order to estimate the transition probabilities.
-They estimate these parameters from two small regions, each approximately $200$ nucleotides in length and containing 13 and 14 CpGs, from 10 human ovarian carcinoma samples and a pool of somatic controls.
+They estimate these parameters from two small regions of 10 human ovarian carcinoma samples and a pool of somatic controls, each region approximately $200$ bp in length and containing 13 and 14 CpGs, respectively.
 These regions were sequenced using hairpin-bisulfite PCR, an assay that measures DNA methylation on both strands of a double-stranded DNA molecule[^hairpin_bs].
-For each region, $12-16$ clones were sequenced per sample[^clones]
+For each region, $12$ to $16$ clones were sequenced per sample[^clones]
 
@@
 
-For each of these regions, \citet{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours.
+For each of these regions, \citeauthor{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours.
 In other words, they found significant within-fragment co-methylation.
@@
 \citet{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.
-The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs.
-They use ultra-deep (median coverage of each CpG in the target $> 10,000 \times$) bisulfite-sequencing of a 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
-They also use these data to estimate co-methylation.
+The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing (median coverage of each CpG in the target $> 10,000 \times$) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
+\citeauthor{Landan:2012kp} also use these data to estimate co-methylation.
 
-As part of their analysis they compute "methylation linkage diagrams".
-For each pair of CpGs in their targeted regions, they create a $2 \times 2$ contingency table.
+As part of their analysis, citeauthor{Landan:2012kp} construct "methylation linkage diagrams" based on a Pearson correlation estimate of within-fragment co-methylation.
+For each CpG 2-tuple in their targeted regions, they create a $2 \times 2$ contingency table.
 The general form of the table is shown below in Table \ref{tab:landan}.
-They then compute the Pearson correlation of each table[^cor_equivalence], which estimates the correlation of methylation events at that pair of CpGs.
-Importantly, these are within-read measurements from a single sample and so these are estimates of within-read, within-sample co-methylation.
+They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{mm_{i, i'} \times uu_{i, i'} - mu_{i, i'} \times um_{i, i'}}{sqrt{u\cdot_{i, i'} \times m\cdot_{i, i'} \times \cdot u_{i, i'} \times m_{i, i'}}}$.
+Importantly, $r$ is based on within-read measurements from a single sample and so these $r$ are estimates of within-read, within-sample co-methylation.
 
-[^cor_equivalence]: Since this is a correlation of two binary variables, it is equivalent to computing the _phi_ coefficient of each $2 \times 2$ table.
+[^cor_equivalence]: The formula I give for $r$ is also known as the $\phi$ coefficient of a $2 \times 2$ table.
+The $\phi$ coefficient of a $2 \times 2$ table of binary variables, such as Table \ref{tab:landan}, is mathematically equivalent to the Pearson correlation coefficient of the same data.
 
@@
 \centering
-\begin{tabular}{@{}ccccc@{}}
-\cmidrule(l){3-5}
- &  & \multicolumn{3}{c}{Second CpG} \\ \cmidrule(l){3-5}
- &  & $m$ & $u$ & Total \\ \cmidrule(l){2-5}
-\multirow{3}{*}{First CpG} & $m$ & $n_{m, m}$ & $n_{m, u}$ & $n_{m, .}$ \\
- & $u$ & $n_{u, m}$ & $n_{u, u}$ & $n_{u, .}$ \\
- & Total & $n_{., m}$  & $n_{, u}$  & $n$ \\ \cmidrule(l){2-5}
+\begin{tabular}{@{}cccc@{}}
+\cmidrule(l){2-4}
+ & \multicolumn{3}{c}{Second CpG} \\
+ \cmidrule(l){2-4}
+\multirow{3}{*}{First CpG} & & & \\
+ & $uu_{i, i'}$       & $um_{i,i ''}$       & $u\cdot_{i, i'}$ \\
+ & $mu_{i, i'}$       & $mm_{i, i'}$       & $m\cdot_{i, i'}$ \\
+ & $\cdot u_{i, i'}$  & $\cdot m_{i, i'}$  & $d_{i, i'}$      \\ \cmidrule(l){2-4}
 \end{tabular}
-\caption{Summary of read counts at a CpG 2-tuple.}
+\caption{Summary of read counts at a CpG 2-tuple, $(i, i'), i \neq i'$. A dot ($\cdot$) indicates that this index is summed over, e.g., $u\cdot_{i, i + 1} = uu_{i, i + 1} + um_{i, i + 1}$.}
 \label{tab:landan}
@@
 
-These correlations are then plotted like a linkage disequilibrium heatmap to identify blocks of CpGs with significant co-methylation. From the analysis of this small set of regions they conclude that:
+So-called "methylation linkage diagrams" are created by plotting the Pearson correlation of each 2-tuple, $r_{(i, i')}$ as a heatmap (_à la_ heatmaps of linkage disequilibrium) to identify blocks of CpGs with significant co-methylation.
+It should be noted that the concepts of 'individual' and 'population' are rather different for "methylation linkage" than they are for genetic linkage (see Table \ref{tab:linkage}).
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}lll@{}}
+\toprule
+                    & Individual               & Population \\ \midrule
+Genetic linkage     & Person                   & Population \\
+Methylation linkage & (Haploid) genome of cell & Person     \\ \bottomrule
+\end{tabular}
+\caption{Notions of 'individual' and 'population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}. NB 'person' could be any organism, e.g, 'mouse', 'dog', etc.}
+\label{tab:linkage}
+\end{table}
+
+From the methylation linkage analysis of these 45 regions, \citeauthor{Landan:2012kp} conclude that:
 
@@
 
-As we will see in Chapter \ref{chap:co-methylation}, this is contradicted by my analyses of co-methylation.
+As we will see in Chapter \ref{chap:co-methylation}, this result is contradicted by my analyses of within-fragment co-methylation (__TODO__: Update once Chapter \ref{chap-comethylation} is updated.)
 
@@
 
-As is clear from the above, there are multiple definitions of within-sample co-methylation, each capturing variations on the idea of the correlation of methylation levels at pairs of loci within a single sample.
-A different concept is between-sample co-methylation, which is the correlation of methylation levels of a single pair of loci between a set of samples.
+As is now abundantly clear, there are multiple definitions of within-sample co-methylation.
+An almost orthogonal concept is between-sample co-methylation, the correlation of methylation levels for a single pair of loci between a set of samples.
+
+Between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci since it is a locus-specific correlation measured at the population level.
 
 Since it is a loci-specific, population-level correlation, between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci.
-Briefly, for two loci, $(i, i')$, the correlation of aggregate methylation levels is computed across samples using, for example, the Pearson correlation coefficient, $r = cor(\bm{\beta_{i}}, \bm{\beta_{i'}})$, where $\bm{\beta_{i}} = (\beta_{i, 1}, \ldots, \beta_{i, N_{loci}})$.
+Briefly, for two loci, $(i, i')$, the correlation of aggregate methylation levels is computed across samples using, for example, the Pearson correlation coefficient, $r = cor(\bm{\beta_{i}}, \bm{\beta_{i'}})$, where $\bm{\beta_{i}} = (\beta_{i, 1}, \ldots, \beta_{i, N_{samples}})$.
 
-The concept of between-sample co-methylation leads to the attractive idea of identifying "tag-loci", whose methylation level is a good proxy for the methylation level across a larger region.
+The concept of between-sample co-methylation leads to the attractive idea of identifying "methylation tag-loci", whose methylation level is a good proxy for the methylation level across a larger region.
 These tag-loci could act like tag-SNPs used in genome-wide association studies, which would allow researchers to assay a small subset of loci in the methylome and capture much of the methylation variability.
-This idea has been explored by \citet{Barrera:2012cq} with respect to the use of _HpaII_ sites as proxies for the methylation level of CpG islands.
+This idea has also been explored by \citet{Barrera:2012cq} with respect to the use of _HpaII_ sites as proxies for the methylation level of CpG islands.
 
-Unfortunately, unlike linkage disequilibrium, we cannot assume that between-sample co-methylation is identical across different tissues from the same individuals since DNA methylation, unlike DNA sequence, varies across tissues.
+Unfortunately, unlike linkage disequilibrium, we cannot assume that between-sample co-methylation is identical between different tissues from the same individuals since DNA methylation, unlike DNA sequence, varies between tissues.
 Therefore, at best, we might hope to generate tissue-specific maps of between-sample co-methylation.
@@
 
+__UP TO HERE: Work through the logic of this next paragraph__
+
 A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$.
-Then, compute $R = cor(X^{T})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci.
+Then, compute $R = cor(\bm{\beta})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci.
 This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data.

### 2015-04-15 Rmarkdown/comethylation.Rmd
Notes following chat with Terry

@@
 
-__TODO: I need to re-write this section (bold).
+__TODO__: I need to re-write this section.
 I'm really not yet sure if this result is biology or an artefact.
@@
 # probabilities p1, p2
-f <- function(n, d, p1, p2) {
+f <- function(n, d, p1, p2, offset = 0.5) {
   sapply(seq_len(n), function(i, d, p1, p2) {
@@
   x <- xtabs(~ z2 + z1, drop.unused.levels = FALSE)
-  log2(x[1, 1] + 0.5) + log2(x[2, 2] + 0.5) - 
-    log2(x[1, 2] + 0.5) - log2(x[2, 1] + 0.5)
+  log2(x[1, 1] + offset) + log2(x[2, 2] + offset) -
+    log2(x[1, 2] + offset) - log2(x[2, 1] + offset)
   }, d = d, p1 = p1, p2 = p2)
@@
 title(paste("d = ", d), outer = TRUE, line = -1)
+```
 
-```
+## Following chat with Terry
+
+These are my notes following a Skype conversion with Terry to discuss the issue.
+
+- [ ] Try a higher order Taylor expansion of the log odds-ratio to get e.g., $n^{1}$ or $n^{3/2}$ convergence.
+The $+\frac{1}{2}$ fudge-factor is motivated by such an expansion and perhaps it is possible to do better by a higher order expansion.
+  - [ ] First try showing the $c = \frac{1}{2}$ is optimal.
+- [ ] Read texts about this issue
+- [ ] Look for research papers on the general issue of dealing with bias in log odds-ratio estimates
+- [ ] What if I just increase the fudge-factor to increase the shrinkage/smoothing of the estimates? It'll shrink to zero alright, but what is the theoretical justification?
+- [ ] Check out the behaviour of the $\phi$ coefficient, i.e., Pearson's $r$ of the $2 \times 2$ table.
+  - Landan et al.: "Note that correlation values are subject to considerable noise when coverage is low or when m1 or m2 are close to 0 or 1."
+    - They don't define whether 'noise' means bias, variance, skewness, etc.

### 2015-04-17 Rmarkdown/comethylation.Rmd
Notes on estimating odds ratios and improved simulation code

--- 2015-02-18 Rmarkdown/comethylation.Rmd
@@
 Until such data are available, we can try to learn more about plausible models of DNA co-methylation through computational methods such as the simulation method described in the next chapter.
+
+# Problem with LOR estimates
+
+The simple estimator of within-fragment co-methylation is biased when the marginal probability of methylation is close to the boundary (0 or 1) unless the 2-tuple is sequenced at very high depth.
+I demonstrate this by simulating methylation patterns at 2-tuples as independent $binomail(1, p)$ random variables and showing that the simple LOR estimator is positively biased.
+
+```{r}
+# Function to simulate K 2-tuples with Poisson(mean_d) sequencing coverage and marginal
+# probabilities p1, p2
+simTables <- function(K, mean_d, p1, p2) {
+  .simTables <- function(mean_d, p1, p2) {
+
+    d <- rpois(1, mean_d)
+    z1 <- factor(rbinom(d, 1, p1), levels = c(0, 1))
+    z2 <- factor(rbinom(d, 1, p2), levels = c(0, 1))
+    xtabs(~ z2 + z1, drop.unused.levels = FALSE)
+  }
+  replicate(K, .simTables(mean_d, p1, p2))
+}
+
+# Estimator of the (offset-adjusted) unconditional log odds ratios of K 2x2 tables
+LOR <- function(x, offset = 0.5) {
+  MM <- x[2, 2, ] + offset
+  MU <- x[2, 1, ] + offset
+  UM <- x[1, 2, ] + offset
+  UU <- x[1, 1, ] + offset
+  log(MM) + log(UU) - log(MU) - log(UM)
+}
+
+# Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
+MH <- function(x, offset = 0.5) {
+  # TODO: return the standard error
+  MM <- x[2, 2, ]
+  MU <- x[2, 1, ]
+  UM <- x[1, 2, ]
+  UU <- x[1, 1, ]
+  d <- apply(x, 3, sum)
+  or_mh <- sum((UU * MM) / d) / sum((MU * UM) / d)
+  log(or_mh)
+}
+
+# Estimator of the (offset-adjusted) unconditional log odds ratio of the 
+# 2x2xK collapsed over K.
+LORCollapsed <- function(x, offset) {
+  z <- apply(x, c(1, 2), sum)
+  # Artificially augment so it's a 2x2x1 table
+  dim(z) <- c(dim(z), 1)
+  LOR(z)
+}
+
+# Simulate and plot data
+simAndPlot <- function(K, mean_d, p1, p2) {
+  x <- mapply(function(p1, p2, K, mean_d) {
+    simTables(K = K, mean_d = mean_d, p1 = p1, p2 = p2)
+  }, p1 = p1, p2 = p2, MoreArgs = list(K = K, mean_d = mean_d), SIMPLIFY = FALSE)
+
+  par(mfrow = c(3, 3))
+  mapply(function(x, p1, p2) {
+    hist(LOR(x), main = paste0("p1 = ", p1, ", p2 = ", p2), xlim = c(-8, 8), breaks = 16)
+    abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+    abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+    abline(v = 0, col = "black", lty = 2)
+  }, x = x, p1 = p1, p2 = p2)
+  title(paste0("K = ", K, ", mean_d = ", mean_d), outer = TRUE, line = -1)
+}
+
+# Vary p1 with p2 = p1 at 30x coverage across 1000 CpG pairs
+K <- 1000
+mean_d <- 30
+p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+p2 <- p1
+simAndPlot(K, mean_d, p1, p2)
+
+# Vary p1 with p2 = 1 - p1 at 30x coverage across 1000 CpG pairs
+K <- 1000
+mean_d <- 30
+p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+p2 <- 1 - p1
+simAndPlot(K, mean_d, p1, p2)
+
+# Vary p1 with p2 = p1 at 1000x coverage across 1000 CpG pairs
+K <- 1000
+mean_d <- 1000
+p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+p2 <- p1
+simAndPlot(K, mean_d, p1, p2)
+
+# TODO: p1 and p2 distributed U(0, 1) at 30x coverage across 1000 CpG pairs
+# Will need to simulate p1 and p2 within .simTable().
+```
+
+## Following chat with Terry
+
+These are my notes following a Skype conversion with Terry to discuss the issue.
+
+- [ ] Try a higher order Taylor expansion of the log odds-ratio to get e.g., $n^{1}$ or $n^{3/2}$ convergence.
+The $+\frac{1}{2}$ fudge-factor is motivated by such an expansion and perhaps it is possible to do better by a higher order expansion.
+  - [ ] First try showing the $c = \frac{1}{2}$ is optimal.
+- [ ] Read texts about this issue
+- [ ] Look for research papers on the general issue of dealing with bias in log odds-ratio estimates
+- [ ] What if I just increase the fudge-factor to increase the shrinkage/smoothing of the estimates? It'll shrink to zero alright, but what is the theoretical justification?
+- [ ] Check out the behaviour of the $\phi$ coefficient, i.e., Pearson's $r$ of the $2 \times 2$ table.
+  - Landan et al.: "Note that correlation values are subject to considerable noise when coverage is low or when m1 or m2 are close to 0 or 1."
+    - They don't define whether 'noise' means bias, variance, skewness, etc.
+
+## Miscellaneous notes
+
+Agresti in Categorical Data Analysis:
+
+> In terms of bias and mean square error, Gart and Zweiful (1967) and Haldane (1955) showed the amended [i.e, add 0.5 to each count] estimators of the [odds-ratio log odds-ratio] behave well (see also Problem 12.4)
+
+The Mantel-Haenszel estimator might be my best bet. Breslow (1981) says:
+
+> The principal conclusion to be drawn from these results is that one should avoid using two of the standard techniques of odds ratio estimation with sparse data.
+
+> The bias of the empirical logit estimator [i.e., + 0.5 to each cell], on the other hand, is sufficient erratic that the possibility of correction seems remote.
+
+> Breslow (1981, Biometrika 68, 73-84) has shown that the Mantel-Haenszel odds ratio is a consistent estimator of a common odds ratio in sparse stratifications (Greenland, S.
+& Robins, J.
+M.
+Biometrics 41, 55–68 (1985).)
+
+The Mantel-Haenszel estimator is an estimator of the common odds ratio for $2 \times 2 \times K$ tables.
+Therefore, using the MH estimator requires the assumption that all $2 \times 2$ tables have the same underlying odds ratio.
+The strata for co-methylation would be 'sample', 'IPD' and possibly 'region', where region is something like 'UMRs, LMRs, PMRs and other' or 'CGI and non-CGI'.
+
+Some good notes on MH estimator [http://isites.harvard.edu/fs/docs/icb.topic79671.files/stratcch1.pdf](http://isites.harvard.edu/fs/docs/icb.topic79671.files/stratcch1.pdf).
+
+An offset of 0.5 eliminates the first-order bias of the maximum likelihood estimator of the (log) odds ratio (Walter, S. D. Biostatistics (1987).)
+
+Regarding the effect of the homogeneity on the Mantel-Haenszel estimator, Greenland, S. Stat Med 1, 217–227 (1982):
+
+> Under the homogeneity assumption, OR^MH is an almost fully efficient estimator of the common value, and remains a consistent estimator of this value even with sparse data, i.e.
+when the fineness of stratification results in many small cell counts.
+__As shown later, its convergence properties under heterogeneity are not as desirable, although even in this case it remains insensitive to small cells.__ [emphasis added]
+
+The _Miettenen_ estimator is like a generalised form of the Mantel-Haenszel estimator, but assumes each cell count is greater than approximately 4 (Greenland, S. Stat Med 1, 217–227 (1982)).

### 2015-04-20 Rmarkdown/biology_background.Rmd
Replace ‘nucleotide’ with ‘bp’ where appropriate.

@@
 
-[^snvs_snps]: An individual nucleotide that is variable in the population is called a single nucletoide variant (_SNV_).
-A SNV that is frequently variable in the population, say at least $1\%$ of the population are variable at that site, is called a single nucleotide polymorphism (SNP).
+[^snvs_snps]: A base position that is variable in the population is called a single nucleotide variant (_SNV_).
+A SNV that is frequently variable in the population, say at least $1\%$ frequency, is called a single nucleotide polymorphism (SNP).
 All SNPs are SNVs but the converse is not true.

### 2015-04-20 Rmarkdown/comethylation_review.Rmd
Replace ‘nucleotide’ with ‘bp’ where appropriate.

@@
 
-[^CHH_only]: The $10$ nucleotide period was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}
+[^CHH_only]: The $10$ bp period was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}
 
@@
 More importantly, the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs.
-Unlike the $8$ to $10$ nucleotide period, the $3$ nucleotide period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it.
+Unlike the $8$ to $10$ bp period, the $3$ bp period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it.
 \citeauthor{Lister:2009hy} do not comment on the $3$ bp period.
@@
 In fact, there are well known periodicities in the distribution of nucleotides in genomes.
-For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ nucleotide period and the latter the $3$ bp period for exonic CpGs.
+For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ bp period and the latter the $3$ bp period for exonic CpGs.
 
@@
 The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes.
-A $10$ bp periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
+A $10$ bp period is readily observed in these plots, consistent with the discovery of $10$ bp period in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}.
 Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.
@@
 
-To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200-250$ nucleotides.
+To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200-250$ bp.
 
@@
 They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes.
-\citeauthor{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci.
+\citeauthor{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of bp between the CHH loci.
 The plot includes pointwise $95\%$ confidence intervals, computed under an assumption of normality, and a running average of these averages.
 This plot, reproduced in Figure \ref{fig:Cokus_2008_Sup_Fig_14}, shows evidence of an approximately $10$ bp period, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.
+
+[^cokus_clarification]: A read may contain multiple pairs of CHH loci and it is not clear whether \citeauthor{Cokus:2008fc} consider all pairs of CHH loci or only the first pair of CHH loci in a read.
+The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
+Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.
 
@@
 \end{figure}
-
-[^cokus_clarification]: A read may contain multiple pairs of CHH loci and it is not clear whether \citeauthor{Cokus:2008fc} consider all pairs of CHH loci or only the first pair of CHH loci in a read.
-The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
-Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.
 
@@
 
-__UP TO HERE: Work through the logic of this next paragraph__
-
 A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$.
@@
 While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult.
-As I have shown, there are problems with most, if not all, the previous analyses, ranging from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation.
+As I have shown, there are problems with most, if not all, previous analyses.
+These problems range from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation.
 Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Re-writing of WGBS downstream analysis chapter.

@@
 
-This chapter discusses various statistical analyses of bisulfite-sequencing data, which I call _downstream analyses_.
+\begin{chapabstract}
+I discuss methods for the _downstream analysis_ of bisulfite-sequencing data.
+Downstream analyses proceed the processing of the raw data (Chapter \ref{chap:wgbs_bioinformatics_analysis}) to address the scientific questions of interest.
+Most downstream methods analyse methylation counts at 1-tuples, however, there is growing interest in analyses based on methylation patterns at m-tuples.
+I discuss the statistical questions underlying these downstream analyses.
+I pay particular attention to those based on m-tuples (m $> 1$) since these have received less attention in the literature.
 
-The first half of the chapter deals with analyses that are based on aggregate methylation levels, such as $\beta$-values.
-I explain the statistical questions behind the common downstream analyses of identifying differential methylation, variable methylation and partially methylated domains.
+A barrier to analyses based on m-tuples has been a lack of software.
+To overcome this barrier, I introduce \code{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
+\code{MethylationTuples} complements \code{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples.
+I describe methods available in \code{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
+\end{chapabstract}
 
-The second half of the chapter reviews downstream analyses that are based on methylation patterns at m-tuples.
-I introduce `MethylationTuples`, an R package I am developing for managing, analysing and visualising methylation patterns at m-tuples, which is complementary to `methtuples` --- `methtuples` extracts the methylation patterns at m-tuples and `MethylationTuples` provides a framework for analysing these data, which is currently lacking for this type of data.
+\section{Methods based on 1-tuples}
 
-\section{Differential methylation}\label{sec:differential_methylation}
+The majority of downstream analysis methods use methylation patterns at 1-tuples, i.e., $\mathbf{m} = (m_{1}, \ldots, m_{i}, \ldots, m_{N_{loci}})$ and $\mathbf{u} = (u_{1}, \ldots, u_{i}, \ldots, u_{N_{loci}})$[^not_beta_values].
+Methods based on 1-tuples have been developed to address several scientific questions including testing for differential methylation (Section \ref{sec:differential_methylation}), testing for differentially variable methylation (Section \ref{sec:variable_methylation}) and identifying regulatory regions of the genome (Section \ref{sec:regulatory_regions}).
 
-A common downstream analysis is the identification of differential methylation.
-This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
-Many methods have been proposed \citep[e.g.,]{Akalin:2012cm, Chen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
+[^not_beta_values]: It is insufficient to use $\mathbf{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).
 
-During my PhD I needed to analyse data from one-group experiment for differential methylation, admittedly an unusual design.
-Almost all these existing methods are designed for two-group experiments, with some being general enough to extend to multiple-group experiments[^multi-group_extension].
-Since I spent time working on the one-group cases, and also because a review article for the two-group case was recently published \citep{Robinson:2014jz}, I pay somewhat disproportianate attention to the one-group scenario in what follows.
+\subsection{Differential methylation}\label{sec:differential_methylation}
 
-[^multi-group_extension]: The extension to multiple-group experiments isn't always implemented in the software, even if discussed in the paper.
+By far the most common analysis of bisulfite-sequencing data is to identify differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
+Consequently, there has been a flurry of methods proposed for identifying differential methylation \citep[e.g.,][]{Akalin:2012cm, Chen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}.
+\citet{Robinson:2014jz} recently reviewed methods for identifying DMCs and DMRs and so I give but an overview of this important topic.
 
-\subsection{Experimental design}\label{sec:experimental_design}
+\subsubsection{Experimental design}\label{sec:experimental_design}
 
-In any analysis of differential methylation we want the DMCs and DMRs to be both _biologically_ and _statistically_ significant; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
+In any analysis of differential methylation, we want the DMCs and DMRs to be both _biologically_ and _statistically_ significant; it's no good if all the differences are simply due to technical artefacts or random fluctuations.
 Key to ensuring biological relevance is good experimental design, such as the use of _replicates_ in each experimental group.
-A distinction is often made between _technical replicates_ and _biological replicates_.
+A distinction is often made in the literature between _technical replicates_ and _biological replicates_.
 Briefly, biological replicates are experimental units that all undergo the same treatment and are used to estimate the within-group variability of the treatment.
@@
 
-As is clear from this definition, the boundary between biological and technical replication is often a fuzzy one.
-For example, \citet{Lister:2009hy} state that, "For each cell type, two __biological__ replicates were performed with cells of different passage number" (emphasis mine).
+The boundary between biological and technical replication is not always clear.
+For example, \citet{Lister:2009hy} state that, "For each cell type, two __biological__ replicates were performed with cells of different passage number [emphasis added]".
 I contend that these are better defined as technical replicates since each replicate came from the same cell line, underwent passaging under near-identical conditions and differ only by the number of cell passages in each media[^cell_passaging].
@@
 
-Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind, or pooled the replicates prior to analysis, thus ignoring all within-group variability, as is the case of the H1 and IMR90 cell lines in \citet{Lister:2009hy}.
+Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or pooled the replicates prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}).
+Simply pooling replicates and analysing as if they were a single sample ignores all within-group variability and should not be used.
 
-Ideally, technical variability is orthogonal to the biological variability, but this rarely occurs.
+Technical variability is ideally orthogonal to the biological variability, but this rarely occurs in practice.
 Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.
-Again, this emphasises the importance of good experimental design such as randomisation, replication and the use of control, both positive and negative.
+Again, this emphasises the importance of good experimental design such as randomisation, replication and the use of positive and negative controls.
 
-\subsection{DMCs}\label{sec:dmcs}
+\subsection{Differentially methylated cytosines}\label{sec:dmcs}
 
-There have been reports of differential methylation at individual CpGs resulting in a phenotypic difference \citep[e.g.,]{Furst:2012gh}.
-However, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome experiments with sample sizes you can count on one or two hands and an average sequencing depth of $10-30\times$.
-Nonetheless, identifying DMCs, even if not an end in itself, is an important step in identifying DMRs.
+A differentially methylated cytosine (DMC) is one where the true methylation level, $B_{i}$, is different between experimental conditions.
+This is typically framed as a test of the mean levels of methylation at the locus in each group[^mean].
+Suppose we have a two-group experiment and let $B_{i}^{k}$ denote the true methylation level of the $i^{th}$ methylation locus in the $k^{th}$ group ($k = 0, 1$).
+We wish to test the null hypothesis that $B_{i}^{0} = B_{i}^{1}$ against the alternative hypothesis that $B_{i}^{0} \neq B_{i}^{1}$.
+As such, identifying DMCs boils down to identifying differences in means, for which there is an enormous body of statistical literature.
+This problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.
 
-Identifying DMCs boils down to identifying differences in the mean level of methylation, for which there is an enormous body of statistical literature.
-The problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.
+[^mean]: This could alternatively be framed as a test of the median methylation level at the locus in each group (or some other location parameter of a statistical distribution).
 
-Regardless of the test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty.
-Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments[^multiple_testing].
+Regardless of the statistical test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty.
+Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the number of tests, in this case approximately 25 million, is at least an order of magnitude greater than what is commonly tested in other genomics experiments[^multiple_testing].
 Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.
@@
 The classical Benjamini-Hochberg procedure also controls the false discovery rate under certain forms of positive dependence \citep{Benjamini:2001ho}.
-(__TODO: Do these techniques work in the high-dependence setting of DMC testing?__)
 
-\subsubsection{One-group experiments}\label{sec:one_group_dmcs}
+While there have been reports of DMCs resulting in a phenotypic difference \citep[e.g.,]{Furst:2012gh}, DMCs are mostly tested as a prelude to the identification of differentially methylated regions (DMRs).
+Moreover, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome bisulfite-sequencing experiments.
+This will remain true while sample sizes can be counted on one or two hands and the average sequencing depth is $10\times$ to $30\times$.
 
-While an uncommon design, I did work on a one-group whole-genome bisulfite-sequencing experiment during my PhD.
-The data, courtesy of Harry Oey and Emma Whitelaw (La Trobe University), came from whole-genome bisulfite-sequencing of five _isogenic_ mice, meaning that these mice were genetically identical (or close to it).
+Several software packages are now available for identifying DMCs.
+Most are limited to analysing two-group experiments.
+Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of $\mathbf{\beta}$ (Section \ref{sec:transformations}).
 
-In a one-group experiment a DMC is a cytosine where at least one of the samples has a different methylation level to the group average.
-The data for the $i^{th}$ methylation locus can be summarised by a $2 \times n$ contingency table, as shown in Table \ref{tab:one-group_dmc_testing}.
-
-\begin{table}[h]
-\centering
-  \begin{tabular}{@{}lll@{}}
-  \toprule
-  & m          & u          \\ \midrule
-  $Sample_{1}$ & $m_{i, 1}$ & $u_{i, 1}$ \\
-  $\vdots$     & $\vdots$   & $\vdots$   \\
-  $Sample_{n}$ & $m_{i, n}$ &            \\ \bottomrule
-  \end{tabular}
-\caption{$2 \times n$ contingency table summarising the methylation evidence for the $i^{th}$ methylation locus in a one-group experiment.}
-\label{tab:one-group_dmc_testing}
-\end{table}
-
-Such a contingency table can be analysed in many different ways.
-Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$.
-$\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$.
-
-To the best of my knowledge, none of the software specifically written for analysing differential methylation analyses is able to test this hypothesis.
-This is because these software are almost all designed for two-group experiments.
-However, this hypothesis is readily testing using routine statistical methods.
-As an example, consider the following data:
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}lll@{}}
-\toprule
-& m    & u    \\ \midrule
-$Sample_{1}$ & $38$ & $1$  \\
-$Sample_{2}$ & $79$ & $2$  \\
-$Sample_{3}$ & $59$ & $1$  \\
-$Sample_{4}$ & $69$ & $2$  \\
-$Sample_{5}$ & $44$ & $46$ \\ \bottomrule
-\end{tabular}
-\caption{An example of a differentially methylated cytosine in a one-group experiment.}
-\label{tab:one-group_example}
-\end{table}
-
-We could analyse this table using a simple stand-alone test, such as Pearson's $\chi^{2}$ test.
-Alternatively, we could frame this as a generalised linear model, either a log-linear model or a logistic regression.
-The equivalence of these methods is shown in Appendex \ref{app:one_group_equivalence}.
-
-A key advantage of using either of the generalised linear models is the capability to include additional covariates in the model, such as the sex of each sample or terms to account for batch effects.
-More complex regression analyses are also possible, such as empirical Bayes or fully Bayesian models.
-
-\subsubsection{Multi-group experiments}\label{sec:multi_group_dmcs}
-
-Due to the cost of whole-genome bisulfite-sequencing, early experiments using this technology to study differential methylation compared one sample against another \citep[e.g.,]{Lister:2009hy, Berman:2012ga}.
-While this might be considered a two-group experiment, one without replicates, I believe this is better thought of as a one-group experiment with two samples.
-Indeed, the analysis of such an experiment uses techniques from Section \ref{sec:one_group_dmcs}, e.g., Pearson's $\chi^{2}$ test or Fisher's exact test \citep{Fisher:1922bx}, which is better suited to the small sample sizes routinely seen in each $2 \times 2$ table.
-
-The more interesting experiment is one that includes replicates within each group, since we can then estimate the within-group variation.
-Unfortunately, some authors have still used Fisher's exact test to analyse such experiments.
-For example, \citet{Lister:2009hy} test for differential methylation between the _H1_ and _IMR90_ samples using Fisher's exact test by first pooling data from the two replicates they have for each sample.
-This is incorrect, because such an analysis effectievly ignores all within-group variability \citep{Hansen:2012gr}.
-Instead, a test that incorporates the within-group variability should be used.
-
-In a two-group experiment this might be a t-test comparing the the average methylation level in the first group, $\widebar{\beta_{i}^{(1)}}$, to that in the second group, $\widebar{\beta_{i}^{(2)}}$.
-More generally, we can use a regression model to allow for multiple groups (i.e., ANOVA) and to incorporate additional covariates.
-
-Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment.
-Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (see Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (see Section \ref{sec:transformations}).
-
-`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} are software packages that all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
+`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
 `DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
@@
 
-The test used to identify DMCs in these regression models is variously a Wald test (`BiSeq`, `DSS`), a likelihood ratio test (`methlySig`, `RADmeth`) or based on the credible interval (Bayesian confidence interval) of the difference in methylation between the two groups (`MOABS`).
+The statistical test used to identify DMCs in these regression models is variously a Wald test (`BiSeq`, `DSS`), a likelihood ratio test (`methlySig`, `RADmeth`) or based on the Bayesian credible interval of the difference in methylation between the two groups (`MOABS`).
 
-Not all methods for identifying differential methylation are designed for identifying DMCs.
-Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs.
+Not all software for identifying differential methylation are designed for identifying DMCs.
+For example, both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs.
 
-\subsection{DMRs}
+\subsection{Differentially methylated regions}
+
+__UP TO HERE__
 
@@
 
-\subsection{Partially methylated domains}
+\subsection{Partially methylated domains}\label{sec:pmds}
 

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Further revision of WGBS downstream analyses.

@@
 
-__UP TO HERE__
-
-A differentially methylated region is a region of the genome where there are multiple cytosines with evidence of differential methylation.
-Importantly, not all cytosines in the region must be genome-wide significant DMCs.
+A differentially methylated region (DMR) is a region of the genome where there are multiple cytosines with evidence of differential methylation.
+Importantly, not all cytosines in the region need necessarily be genome-wide statistically significant DMCs.
 Rather, the idea is that a DMR might capture a weaker but broader difference in methylation.
@@
 
-There are two very different strategies for identifying differentially methylated regions:
+There are two very different strategies for identifying DMRs:
 
@@
 Using regions that are defined _a priori_, which are then tested for differential methylation.
-Such regions might be CpG islands \citep[e.g.,][]{Huang:1999fs,Doi:2009ky}[^array], MspI restriction fragments \citep[e.g.,][]{Stockwell:2014fq}, any other predefined genomic feature, such as a gene promoters or transcription factor binding sites, or general predefined bins \citep[e.g., 100 bp bins used by][]{Park:2014ho}.
+Such regions might be CpG islands \citep[e.g.,][]{Huang:1999fs,Doi:2009ky}[^array]; MspI restriction fragments \citep[e.g.,][]{Stockwell:2014fq}; a predefined genomic feature, such as a gene promoters or transcription factor binding sites; or general predefined bins \citep[e.g., 100 bp bins used by][]{Park:2014ho}.
 2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
@@
 
-The former is much simpler to analyse but is limited in its ability to discover new differentially methylated regions.
+The former is much simpler to analyse but is limited in its ability to discover novel DMRs.
 It is also hampered because the correct unit or scale for differential methylation may not be known for the experiment.
 
-The latter offers the opportunity to identify previously unknown regions that are subject to differential methylation.
+The latter offers the opportunity to identify novel regions that are subject to differential methylation.
 Included in this is the opportunity to discover the scale over which differential methylation acts.
@@
 
-The idea of testing for differential methylation is relatively straightforward.
-For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region, for example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage.
-These averages are then compared between samples.
-I describe a basic implementation of this for a one-group experiment and then detail a more sophisticated method that is implemented in `BiSeq` \citep{Hebestreit:2013ko}.
+The idea of testing for differential methylation at _a priori_ defined regions is relatively straightforward.
+Suppose we have a two-group experiment and let $\widebar{B_{r}^{k}}$ be the true average level of methylation for the $r^{th}$ region in the $k^{th}$ group ($k = 0, 1$).
+The null hypothesis is that $\widebar{B_{r}^{0}} = \widebar{B_{r}^{1}}$ against the alternative hypothesis that $\widebar{B_{r}^{0}} \neq \widebar{B_{r}^{1}}$.
 
-\subsubsection{One-group experiments}\label{sec:one_group_dmrs}
+We might estimate \widebar{B_{r}^{k}} by the group-wise average of the sample-wise weighted average of $\beta$-values for all methylation loci in the region, where the weights are proportional to the sequencing coverage.
+Then, identifying differential methylation at _a priori_ defined regions simply boils down to identifying differences in means, just as is the case for testing for DMCs.
+Again, this problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.
 
-For a region containing $c$ methylation loci, construct the $n \times 2 \times c$ contingency table, i.e., each 'slice' of the table is a $n \times 2$ table containing the data for a single locus.
-Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions \citep{Good:1987ce}.
-Instead, we could fit the log-linear model with `sample`, `meth_state` and `locus` as the set of factors.
-The saturated model is
+The above description brushes over some technicalities, such as how to handle CpGs with insufficient sequencing coverage.
+An alternative approach for testing _a priori_ defined regions for differential methylation is offered by `BiSeq` \citep{Hebestreit:2013ko}.
 
-\begin{align*}
-\log(F_{a, b, c}) &= \lambda_{0} \\
-  &+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
-  &+ \lambda_{a, b}^{meth\_state, sample} + \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus} \\
-  &+ \lambda_{a, b, c}^{meth\_state, sample, locus}
-\end{align*}
+\subsubsection{A hierarchical procedure for testing _a priori_ defined regions for differential methylation}
 
-We would first need to test whether there is a significant three-way interaction between `meth_state`, `sample` and `locus` by computing the deviance of the model without the three-way interaction
+`BiSeq` \citep{Hebestreit:2013ko} uses a hierarchical procedure to test for differential methylatoin at _a priori_ defined regions.
+To begin, `BiSeq` first defines _CpG clusters_ by identifying CpGs that are within a user-specified genomic distance of one another and that have sufficient sequencing coverage across the set of samples[^biseq_alternative].
+While there is clearly a 'data-driven' component to these cluster definitions, I reserve the use of 'data-driven regions' for those which are based on the methylation levels of loci rather than their genomic co-ordinates.
 
-\begin{align*}
-\log(F_{a, b, c}) &= \lambda_{0} \\
-&+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
-&+ \lambda_{a, b}^{meth\_state, sample} + \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{align*}
+[^biseq_alternative]: As an alternative to creating these CpG clusters, \citet{Hebestreit:2013ko} also suggest using the target regions of the assay, such as MspI fragments in the case of RRBS.
 
-On the one hand, if there is no three-way interaction, then we can test for an interaction between `meth_state` and `sample`, i.e., test for a difference in methylation level across the loci, by comparing the model
+Once these clusters are defined, the $\beta$-values in each cluster are smoothed for each sample using a local binomial likelihood smoother.
+This procedure will create a smoothed $\beta$-value for each CpG, even those with insufficient sequencing coverage.
+Then, for each CpG, `BiSeq` fits a beta regression model[^not_beta-binomial] to the smoothed $\beta$-values, which is tested for evidence of differential methylation at that cytosine (i.e., a test of whether the cytosine is a DMC).
 
-\begin{align*}
-\log(F_{a, b, c}) &= \lambda_{0} \\
-&+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
-&+ \lambda_{a, b}^{meth\_state, sample} + \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{align*}
-
-to the model
-
-\begin{align*}
-\log(F_{a, b, c}) &= \lambda_{0} \\
-&+ \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} \\
-&+ \lambda_{a, c}^{meth\_state, locus} + \lambda_{b, c}^{sample, locus}
-\end{align*}
-
-This test is analogous to the generalised Cochran-Mantel-Haenszel test \citep{Landis:1978vk,MANTEL:1978iv,Birch:1965uj} of the $n \times 2 \times c$ table \citep[p231]{agresti1990categorical}.
-The advantage of the Cochran-Mantel-Haenszel is its applicability for sparse datasets for which the asymptotic theory does not hold for the test based on the log-linear regression  \citep[p232]{agresti1990categorical}.
-
-On the other hand, if there is a significant three-way interaction between `meth_state`, `sample` and `locus` then this requires more careful modelling to determine the nature of this interaction and whether the region is a consistent DMR.
-
-\subsubsection{A more sophisticated multi-group analysis}\label{sec:a_more_complex_example}
-
-`BiSeq` is designed for RRBS data and tests for differential methylation at "CpG clusters".
-These CpG clusters are based on an initical scan of the genome looking for CpGs that are within a given distance of one another and that have sufficient sequencing coverage across the set of samples.
-While there is clearly a data-dependent component to these cluster definitions, I reserve the use of 'data-driven regions' for those which are based on the methylation levels of loci rather than their genomic co-ordinates.
-\citep{Hebestreit:2013ko} alternatively recommend using the target regions of the assay, such as MspI fragments in the case of RRBS, instead of creating these CpG clusters.
-
-One the clusters are defined, the $\beta$-values in each cluster are smoothed using a local binomial likelihood smoother.
-For each CpG `BiSeq` fits a beta regression model[^not_beta-binomial], which is tested for evidence of differential methylation at that cytosine.
 Based on these P-values, `BiSeq` then use a hierarchical testing procedure to control the false discovery rate at both the cluster-level and locus-level.
-This method is based on several papers by Yoav Benjamini and colleagues \citet{Benjamini:1997fh, Benjamini:2001ho, Benjamini:2006gd, Benjamini:2007cm}.
-It aims to first control the false discovery rate at the cluster-level and then refine the signal by trimming non-DMCs from those clusters that have been declared as differentially methylated.
+This method is based on several papers by Yoav Benjamini and colleagues \citep{Benjamini:1997fh, Benjamini:2001ho, Benjamini:2006gd, Benjamini:2007cm}.
+It aims to first control the false discovery rate at the cluster-level and then refines the signal by trimming non-DMCs from those clusters that have been declared as differentially methylated.
 Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_ (see Section \ref{sec:consistency_of_dmrs}).
@@
 
-Based on my understanding of this procedure, only cytosines within differentially methylated clusters can be declared DMCs since it is only those cytosines that are tested in the second-step of the hierarchical testing algorithm.
-Therefore, isolated DMCs will not be detected by this procedure.
-
 \subsection{Using data-driven regions}
 
-The most common type of algorithm for identifying data-driven DMRs, also arguably the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
-The initial identification of DMCs will typically use a relaxed statistical significance threshold.
-For example, \citet{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction".
-This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
-
-These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
+Methods for identifying _data-driven_ DMRs are fairly _ad hoc_.
+The most common approach is to scan the genome for clusters of DMCs and declare these to be DMRs.
+The initial scan for DMCs will typically use a relaxed statistical significance threshold (i.e., below genome-wide significance).
 Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
 
-The challenges of valid statistical inference at such data-driven regions are not unique to the problem of testing for differentially methylated regions.
+For example, \citet{Hansen:2011gu} start by testing all CpGs for differential methylation and retain all those with a P-value in the lowest $5\%$.
+They then declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction"[^consistency].
+These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
+
+[^consistency]: This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
+
+It is challenging to perform valid statistical inference of differential methylation at these data-driven regions.
+We must be careful when 'double-dipping' into the data, whereby the same data are being used to define the regions as are being used to test their significance.
+These regions have been _selected_ because loci in these region display a difference and therefore tests of whether the region has a difference are biased towards rejecting the null hypothesis.
+
+The challenges of valid statistical inference at such data-driven regions are not unique to the problem of testing for DMRs.
 Similar problems arise in the analysis of chromatin immunoprecipitation sequencing (ChIP-seq) experiments \citep{Schwartzman:2011to, Lun:2014gn} and in the field of signal processing \citep{Schwartzman:2011vg}.
-It boils down to the problem of how to assign statistical significance to regions or peaks using the same data that were used to define these regions or peaks.
-These regions are already enriched for differences, which biases the interpretation the results of tests for differences at these regions.
-One way, and I would argue the best way, to avoid this issue is to test these regions using a separate dataset, which completely avoids the issue of statistical 'double-dipping'.
+
+One way to avoid this issue, and I would argue the best, is to test these regions using a separate dataset, which completely avoids the issue of statistical 'double-dipping'.
+Of course, this requires that such a dataset is available or that resources exist to create it, which is frustratingly rare.
+
 If the sample size is large enough, then permutation testing may also be appropriate.
-For example, \citet{Hansen:2013eo} permute the group labels of their samples and re-ran their analysis to determine for each of the observed regions "how often we see another block of similar length and effect size anywhere in the genome and in any of the permutations".
-The chief limitation of the permutation strategy is the restricted number of permutations that are possible from small sample sizes and the time and computational resources it takes to perform analyse the data for a single permutation, which are often substantial.
-For example, \citet{Hansen:2013eo} only performed nine permutations.
+For example, \citet{Hansen:2013eo} permute the group labels of their samples and re-ran the analysis to determine for each of the observed DMRs "how often we see another block of similar length and effect size anywhere in the genome and in any of the permutations".
+The chief limitation of the permutation strategy is the restricted number of permutations that are possible from small sample sizes, along with the often substantial time and computational resources it takes to analyse the data for each permutation[^hansen_permutations].
 
-A somewhat less _ad hoc_ strategy for identify differentially methylated regions is called "bump-hunting" \citep{Jaffe:2012gx}.
-Initially developed for methylation microarrays, and now available for broader use in the R/Bioconductor package `bumphunter` ([http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html](http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html)), bump-hunting is quite a general statistical problem.
-For the case of identifying DMRs, the idea is to plot the value of a statistic used to test for DMCs as a function of the position in the genome[^smoothing] and then to identify bumps in this plot.
-Bumps are identified as contiguous regions of the genome where the signal is above some threshold and may include an error term to account for the spatial correlation of the signal.
-The significance of these peaks may be assessed with a permutation strategy.
+[^hansen_permutations]: For example, \citet{Hansen:2013eo} only performed nine permutations to estimate significance.
 
-Another alternative for combining individual loci into data-driven DMRs is based on the loci-specific P-values.
-These methods can be thought of extensions to Fisher's method for combining P-values \citep{Fisher:1936vv}, by attempting to account for the correlation of tests at nearby methylation loci.
+Related to the method of creating DMRs by forming clusters of DMCs is that of "bump-hunting".
+Initially developed for application to methylation microarray data \citep{Jaffe:2012gx}, and now available for broader use in the R/Bioconductor package `bumphunter` ([http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html](http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html)), "bump-hunting" may be used to identify DMRs.
+The idea is as follows.
+Firstly, each each CpG is tested for differential methylation.
+The resulting test statistic is then considered as a function of the position in the genome[^smoothing] and processed with an algorithm to identify "bumps" in the signal.
+Bumps are defined as contiguous regions of the genome where the signal is above some threshold.
+The algorithm may include an error term to account for the spatial correlation of the signal and the significance of these peaks may be assessed using a permutation strategy.
+
+[^smoothing]: The test statistic may be smoothed to reduce variation at the expense of increasing bias.
+
+Another alternative for combining individual loci into data-driven DMRs uses the locus-specific P-values rather than the locus-specific test statistics.
+These methods can be thought of extensions to Fisher's method for combining P-values \citep{Fisher:1936vv}, that attempt to account for the correlation of tests at nearby methylation loci.
 `comb-p` \citep{Pedersen:2012vl} uses the Stouffer-Liptak-Kechris \citep{Stouffer:1949ua, Kechris:2010fw, Zaykin:2011jq} correction for spatially correlated P-values.
-`SLIM` \citep{Wang:2011cw} is used by `methylKit` \citep{Akalin:2012cm} to do a similar correction.
+`methylKit` \citep{Akalin:2012cm} uses `SLIM` \citep{Wang:2011cw} to do a similar correction.
 
@@
 \subsubsection{Consistency of DMRs}\label{sec:consistency_of_dmrs}
+
+__UP TO HERE__
 

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
More revisions of WGBS downstream analyses.

@@
 \citet{Robinson:2014jz} recently reviewed methods for identifying DMCs and DMRs and so I give but an overview of this important topic.
+
+All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\mathbf{\beta}$.
 
@@
 It aims to first control the false discovery rate at the cluster-level and then refines the signal by trimming non-DMCs from those clusters that have been declared as differentially methylated.
-Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_ (see Section \ref{sec:consistency_of_dmrs}).
+Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_, i.e., that the differences in methylation are in the same direction.
 
@@
 For example, \citet{Hansen:2011gu} start by testing all CpGs for differential methylation and retain all those with a P-value in the lowest $5\%$.
-They then declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction"[^consistency].
+They then declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction" (i.e., the region is consistent).
 These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
-
-[^consistency]: This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).
 
@@
 
-\subsubsection{Consistency of DMRs}\label{sec:consistency_of_dmrs}
+\subsection{Differentially variable methlyation}\label{sec:variable_methylation}
 
-__UP TO HERE__
+It has been hypothesised that increased variability in DNA methylation indicates an epigentic _plasticity_ that may be highly relevant to common diseases, in particular, cancer \citep{Feinberg:2010dy}.
+Differential variability is distinct from differential methylation.
+Whereas the analysis of differential methylation is based on statistical tests of differences in means, the analysis of differential variability is based on statistical tests of differences in variances.
+Analogously, we define variably methylated cytosines (VMCs) and variably methylated regions (VMRs).
+To emphasise, a locus (resp.
+region) may be a DMC (resp.
+DMR) while not a VMC (resp.
+VMR) and _vice versa_.
 
-While not often explicitly stated, it is desirable that a DMR is _consistent_.
-The definition of consistency is best illustrated by a pathological example describing a region that is inconsistent.
-Suppose we have a one-group experiment with five samples[^one_by_five] and are studying a region with five CpGs, all of which are DMCs.
-However, the first CpG is differentially methylated in the first sample, the second CpG is differentially methylated in the second sample, etc.
-While the entire region is 'differentially methylated', the difference is not observed consistently in any one sample.
-A consistent region is therefore one where the differences are 'in the same direction'.
-
-[^one_by_five]: Equivalently, a five group experiment without replicates.
-
-Consistent DMRs are most easily identified in two-group experiments, where we can use the sign of the test statistic to determine the 'direction' of the signal.
-This is exactly what is done by, for example, \citet{Hansen:2011gu}.
-
-In one-group experiments, the idea of consistency is similar, but not quite identical, to testing the homogeneity of odds ratios for all $n \times 2$ contingency tables in the region.
-For example, a region may be consistent but the strength of the effect may vary across the region, thus the odds ratios may vary.
-
-\section{Other downstream analyses using $\beta$-values}
-
-Two other common downstream analyses based on $\beta$-values are the identification of variable methylation and partially methylated domains.
-
-\subsection{Variable methyation}\label{sec:variable_methylation}
-
-Differential methylation is an analysis of the difference in means and seeks to identify differentially methylated regions (DMRs).
-Analogously, we can analyse differences in the variability of methylation profiles, which is known as differential variability, and those regions known as variably methylated regions (VMRs).
-It has been hypothesised that increased variability in DNA methylation indicates an epigentic _plasticity_ that may be highly relevant to common diseases, in particular, cancer \citep{Feinberg:2010dy}.
-
+\citet{Jaffe:2011hy} first developed formal statistical tests for differential variability of methylation for both the one-group and two-group experiments.
+These methods were developed for us with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}).
 In a one-group experiment, a variably methylated region is one that has increased variability compared to 'similar' regions elsewhere in the genome.
 In a two-group experiment, a differentially variable locus is one where the variation in one group is significantly larger than that in the other group.
-\citet{Jaffe:2011hy} first developed formal statistical tests for both the one-group and two-group experiments.
-These methods were developed for us with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}).
 
-Statistical tests of variations are well known to be more difficult than tests of means and require larger sample sizes.
+Statistical tests of variances are well known to be more difficult than tests of means and require larger sample sizes.
 A more subtle difficulty is in dealing with outliers.
-An outlier in one group will greatly increase the variation in that same group, but this does not necessarily mean that the locus is differentially methylated.
-It is not hard to envisage an example where the two groups in fact have very similar variability once the outlier is excluded.
-Tests of differential variability that are based on the F-test \citep[e.g.,][]{Hansen:2011gu} or Bartlett's test \citep[e.g.,][]{Teschendorff:2012hea} will be susceptible to calling such loci as differentially variable.
-By contrast, \citet{Phipson:2014ji} designed `DiffVar` to test for differential variability that is deliberately robust to such outliers by using Levene's test \citep{Olkin:1960uu}.
-Which approach is preferable will depend on the user's perspective as to whether such loci with outliers should be called as variably methylated.
+An outlier in one group will greatly increase the variation in that group, but this does not necessarily mean that the locus is differentially methylated.
+It might, for example, be due to an error in the assay.
+It is not difficult to envisage an example where the two groups in fact have very similar variability when the outlier is excluded.
+
+Tests of differential variability that are based on the F-test \citep[e.g.,][]{Hansen:2011gu} or Bartlett's test \citep[e.g.,][]{Teschendorff:2012hea} will be susceptible to calling loci with such outliers as being differentially variable.
+By contrast, `DiffVar` \citep{Phipson:2014ji} uses Levene's test \citep{Olkin:1960uu}, which is robust to outliers, to test for differential variability.
+
+All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\mathbf{\beta}$.
 
 \subsection{Partially methylated domains}\label{sec:pmds}
+
+__UP TO HERE: Re-write as "Epigenome segmentation"__
 

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Continue revisions of WGBS downstream analyses.

@@
 
-\subsection{Partially methylated domains}\label{sec:pmds}
+\subsection{Epigenome segmentation}\label{sec:epigenome_segmentation}
 
-__UP TO HERE: Re-write as "Epigenome segmentation"__
+Methylation data in the form of the vector of $\beta$-values, $\mathbf{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}.
+One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs).
+Partially methylated domains are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.2$ to $0.7$.
 
-Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$.
-PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}.
-Using a simple sliding window algorithm, \citeauthor{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb.
-They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample.
-This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.
+PMDs were first identified in the _IMR90_ methylome by \citet{Lister:2009hy}.
+Using a simple sliding window algorithm, \citeauthor{Lister:2009hy} found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb.
+They also showed that these PMDs were not simply due to a methylated subpopulation and an unmethylated subpopulation of cells in the sample.
+This did this by showing that individual reads mapped to these PMDs contained both methylated and unmethylated bases.
 
@@
 
-PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_).
+PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu}.
+\citet{Hansen:2011gu} found that these PMDs overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_).
 However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}.
@@
 
-More sophisticated methods have also been proposed to identify these PMDs.
-`methylSeekR` \citep{Burger:2013kq} is one such algorithm.
-It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions.
-While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that whole-genome bisulfite-sequencing provides
+Recently, more sophisticated methods have been proposed to identify these PMDs.
+`methylSeekR` \citep{Burger:2013kq} is one such method.
+It uses a hidden Markov model of the $\beta$-values, combined with other filters, to segment the genome into unmethylated, lowly methylated and partially methylated regions.
 
-\section{Downstream analyses of methylation patterns at m-tuples}\label{sec:downstream_analyses_of_m-tuples}
+\section{Methods based on m-tuples (m $> 2$)}\label{sec:downstream_analyses_of_m-tuples}
 
-None of the aforementioned analyses specifically exploit the fact that bisulfite-sequencing reads measure methylation patterns at multiple loci, i.e., at m-tuples.
-Yet much can be learnt my studying exactly that.
-
-In this section I review the different types of questions that can be addressed when we make use of this extra information that is only available from bisulfite-sequencing data.
+In this section I review the different types of questions that can be addressed by expanding our analysis to use methylation patterns at m-tuples (m $> 2$) rather than just 1-tuples.
+This extra information is only available from sequencing-based assays.
+I also describe existing software that implement some of these methods.
 
@@
 
-There are several definitions and uses of _methylation entropy_ in the literature.
-\citet{Xie:2011cy} introduced the concept for bisulfite-sequencing data as a way of quantifying the variability of methylation patterns at an m-tuple.
-For a given 4-tuple, $(i, \ldots, i + 3)$, \citet{Xie:2011cy} analyse the set of reads aligned to it, $\{z: z \in \mathcal{R}_{(i, \ldots, i + 3} \}$, to compute the methylatione entropy.
-On the one hand, if we only observe a single unique methylation pattern then the m-tuple has the minimum methylation entropy of zero.
-On the other hand, if we observe all possible $2^{4} = 16$ methylation patterns then the m-tuple has the maximum methylation entropy of 1.
-Depending on the frequency of the patterns, we obtain intermediate methylation values ($0 < \text{methylation entropy} < 1)$.
-Of course, methylation entropy need not be limited to 4-tuples.
-The choice of the size of the tuples depends on the regions of interest, the read lengths and the sequencing coverage.
-
-The natural interpretation of methylation entropy as a measure of 'disorder' allows us to quantify how heterogeneous DNA methylation is at a locus \citep{Xie:2011cy, He:2013cj}[^other_definitions].
+The natural interpretation of methylation entropy is a measure of 'disorder'.
+It has been used to quantify how heterogeneous DNA methylation is at a locus \citep[e.g., ][]{Xie:2011cy, He:2013cj}[^other_definitions].
 These methylation entropies can be analysed to identify heterogeneous regions of the genome or perhaps tested for an association with a phenotype.
 
-[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles_and_epipolymorphism}.
-Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since this uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.
+[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles}.
+Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since it uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.
 
-While \citet{Xie:2011cy} do not provide software implementing the method, a subsequent paper from some of the same researchers introduces the `DMEAS` software to do so \citep{He:2013cj} .
-Unfortunately, `DMEAS` is only available as a Windows binary or as a Perl script that itself is only available as PDF file, greatly limiting its usefulness.
+Broadly speaking, on the one hand, if we only observe a single unique methylation pattern, then the m-tuple has the minimum methylation entropy of zero.
+On the other hand, if we observe all possible $2^{m}$ methylation patterns at equal frequency, then the m-tuple has the maximum methylation entropy (typically normalised to one).
+Depending on the frequency of the observed methylation patterns patterns, we obtain intermediate values of the methylation entropy.
 
@@
 
-In a diploid cell, allele-specific methylation occurs when only one of the parental chromosomes is methylated at a particular locus, where the locus may be an individual cytosine or a broader region.
-One particular form of allele-specific methylation occurs at imprinted genes where the silenced copy of the gene is heavily methylated and the active copy is lightly methylated \citep{Shoemaker:2010jz}.
+In a diploid cell, allele-specific methylation occurs when only one of the parental chromosomes is methylated at a particular locus, where the locus may be an individual cytosine or a broader region such as a gene promoter.
+A particularly interesting form of allele specific methylation occurs at imprinted genes, where one copy of the gene is active in a parent-specific manner.
 However, it is now apparent that allele-specific methylation is far more prevalent than at just these imprinted regions \citep{Tycko:2010hc, Shoemaker:2010jz}.
 
-The obvious method to detect allele-specific methylation from bisulfite-sequencing relies on the read containing a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus.
-The heterozygous variant allows the reads to be separated by the observed allele[^phase], which can then be used to test for allele-specific methylation.
+The obvious method to detect allele-specific methylation from bisulfite-sequencing requires reads that contaian a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus.
+The heterozygous variant allows these reads to be separated by the observed allele[^phase], which can then be used to test for allele-specific methylation.
 For example, \citet{Shoemaker:2010jz} construct a $2 \times 2$ contingency table, like that shown in Table \ref{tab:asm}, and test for an association between the allele and the methylation state using Fisher's exact test \citep{Fisher:1922bx}.
+
+[^phase]: This does not give parent-specificity unless the phase of the genotype is also known.
 
@@
 & Allele 1 & Allele 2 \\ \midrule
-$m$ &  $m_{1}$        &  $m_{2}$       \\
-$u$ &  $u_{1}$        &  $u_{2}$       \\ \bottomrule
+$m$ &  $m^{1}$        &  $m^{2}$       \\
+$u$ &  $u^{1}$        &  $u^{2}$       \\ \bottomrule
 \end{tabular}
-\caption{$2 \times 2$ table used to test for allele-specific methylation.}
+\caption{$2 \times 2$ table used to test for allele-specific methylation.
+$m^{1}$ is the number of reads with the first allele and that are also methylated at the methylation locus, $u^{1}$ is the number of reads with the first allele and that are also unmethylated at the methylation locus, etc.}
 \label{tab:asm}
@@
 
-Such an approach, while straightforward, is also obviously limited to a subset of the available methylation loci.
-Both \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require nearby heterozygous genetic variants nearby to the methylation locus of interest.
+While straightforward, this approach is also obviously limited to a subset of the methylation loci in the genome.
+\citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require heterozygous genetic variants nearby to the methylation locus of interest.
 These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data.
@@
 
-\citet{Fang:2012ci} uses reads containing multiple methylation loci and looks for regions of the genome where there are two distinct methylation patterns at the read-level that occur at roughly equal proportions, indicating one pattern comes from one allele and the other pattern from the other allele.
+\citet{Fang:2012ci} use reads containing multiple methylation loci and looks for regions of the genome where there are two distinct methylation patterns at the read-level that occur at roughly equal proportions, indicating one pattern comes from one allele and the other pattern from the other allele.
 The likelihood of allele-specific methylation is computed using an expectation-maximisation algorithm, which assigns reads to one of the two possible alleles.
@@
 
-[^phase]: This does not give parent-specificity unless the phase of the genotype is also known.
-
-\subsection{Epialleles and epipolymorphism}\label{sec:epialleles_and_epipolymorphism}
+\subsection{Epialleles}\label{sec:epialleles}
 
 A DNA sequence may have multiple epigenetic states.
-For example, the cytosine in the sequence `TCGA` may be methylated or unmethylated; each of the methylated and unmethylated versions of that sequence is an _epiallele_.
-Restricting our attention to CpG methylation, a sequence with $m$ CpGs has $2^{m}$ potential epialleles.
-Therefore, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical[^epiallele_definition].
-An epiallele may also be described as an _epimutation_, if it sufficiently rare.
-
-[^epiallele_definition]: \citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
-While most would agree that is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive.
-After all, we refer to alternate forms of a genetic sequence as alleles regardless of whether we know of a phenotypic consequence of the variant.
-
-Epipolymorphism, analogous to genetic polymorphism, is the presence of multiple epialleles in a population.
-However, unlike genetic polymorphism, where the population is typically a set of chromosomes from multiple individuals, the population here is often within an individual, even within a tissue within an individual.
-The 'epipolymorphism' of a locus has also been defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}.
-\citeauthor{Landan:2012kp} define the epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
-
-[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism under a model that assumes sampling with replacement or, equivalently, an infinite population size.
-While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.
-
-\subsubsection{Examples}
-
+For example, the cytosine in the sequence _TCGA_ may be methylated or unmethylated; each of the methylated and unmethylated versions of that sequence is an _epiallele_.
 Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom.
-Even then, the number of such epialleles is small --- a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.
+Even then, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.
 
@@
 The poster child for the potential important of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}.
-Mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic and have an increased suscpetibility to tumours compared to those mice with a methylated version of the allele, which have a _pseudoagouti[^pseudoagouti_vs_agouti]_ (brown) coat and none of the associated health defects.
+Mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic and have an increased susceptibility to tumours compared to those mice with a methylated version of the allele, which have a _pseudoagouti_[^pseudoagouti_vs_agouti] (brown) coat and none of the associated health defects.
 
@@
 
-In humans, there have been several interesting studies using epialleles to infer the clonality and evolution of cancer \citep{Siegmund:2009kk, Li:2014ei}, as well as the evolution of methylation dynamics and the epipolymorphism of various loci in an immortalised cell line \citep{Landan:2012kp}.
+In humans, there have been several interesting studies using putative epialleles to infer the clonality and evolution of cancer \citep{Siegmund:2009kk, Li:2014ei}, as well as to study the evolution of methylation dynamics and the rate of epipolymorphism of various loci in an immortalised cell line \citep{Landan:2012kp}.
 
-Finally, although there is much interest, if not outright hype, surrounding transgenerational epialleles, these are difficult to prove outside of very controlled experimental conditions \citep{Daxinger:2012kl}.
+Regardless of where you draw the line as to what constitutes an epiallele, it has become clear in the analysis of bisulfite-sequencing data that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception.
 
-Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing \citep{Landan:2012kp}.
-There has been some progress in developing software to extract and analyse these epialleles, which I now describe.
+Restricting our attention to CpG methylation, a sequence with $m$ CpGs has $2^{m}$ potential epialleles.
+In other words, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical[^epiallele_definition].
+An epiallele may also be described as an _epimutation_ if it is different from the 'normal' methylation state.
 
-\subsubsection{Methods}
+[^epiallele_definition]: \citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
+While most would agree that it is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive.
+After all, we refer to alternate forms of a genetic sequence as alleles regardless of whether we know of a phenotypic consequence of the variant.
 
-`methclone` \citep{Li:2014ei} is software to estimate the frequency of epialleles at m-tuples and to identify "shifts" in these distributions between two samples.
-`methclone` is limited to pairwise comparisons.
-The input is a pair of `BAM` files created with the `Bismark` aligner, one for each sample.
-For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples[^methclone_output].
-The "foreground" combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples while the "background" combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two stages [samples]".
+The rate of _epipolymorphism_ of a locus is defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}[^epipolymorphism].
+\citeauthor{Landan:2012kp} define the rate of epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
+Unlike genetic polymorphisms, where the population is typically a set of chromosomes from multiple individuals, the population of epipolymorphisms is often within an individual, even within a tissue within an individual.
+
+[^epipolymorphism]: \citet{Landan:2012kp} actually call this the "epipolymorphism" of the locus rather than the "rate of epipolymorphism" of the locus.
+However, I think this is better described as a rate since it refers to the frequency at which we observe epialleles/epipolymorphisms.
+
+[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism of the locus under a model that assumes sampling with replacement or, equivalently, an infinite population size.
+While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.
+
+The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error.
+Another difficulty, perhaps unavoidable with current technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele.
+
+Of the downstream analyses based on methylation patterns at m-tuples, the study of epialleles has received the most attention with respect to methods and software.
+
+`methclone` \citep{Li:2014ei} is a method to estimate the frequency of epialleles at m-tuples (the rate of epipolymorphism) and to identify "shifts" in these distributions between a pair of samples.
+`methclone` is based on computing and comparing two forms of methylation entropy, the "foreground" and "background".
+The foreground combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples.
+The background combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two [samples]".
 The difference in these combinatorial entropies, $\Delta S = S - \tilde{S}$, a kind of observed-to-expected log-ratio, is used to identify shifts in the epiallele distribution between a pair of samples.
@@
 It isn't clear whether the range of $\Delta S$ depends on the size of the m-tuples nor is it clear how to choose the threshold at which to declare a significant shift in the distribution of epialleles.
-The example in the paper shows results for a threshold of $-60, -70$ and $-80$, but these choices appear somewhat arbitrary.
 
-[^methclone_output]: It is not clear from the paper or documentaiton whether `methclone` returns a file containing the number of methylation patterns at each m-tuple, _a ala_ `methtuple`.
-Unfortunately, I haven't been able to install `methclone`, so have not been able to test this out.
-
-The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error.
-Another difficulty, perhaps unavoidable with the present technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele.
-`methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies, which does not attempt to account for these potential sources of bias.
+`methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies and does not attempt to account for potential sources of bias.
 In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.
@@
 Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,][]{Fang:2012ci, Peng:2012dh}, might also adapted to identify epialleles and their associated frequencies.
-It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which strictly speaking is a requirement for the m-tuple to be an epipolymorphic locus.
+It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which, strictly speaking, is a requirement for the m-tuple to be an epipolymorphic locus.
+
+\subsection{Software for analysing methylation patterns at m-tuples}\label{sec:methylation_pattern_software}
+
+Generally speaking, there are fewer software options for analysing methylation patterns at m-tuples (m $> 2$) than there are for analysing methylation patterns at 1-tuples.
+Furthermore, the available options are often difficult to extend since they are typically developed for a specific task and not for general computations with methylation patterns at m-tuples.
+
+I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations.
+Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` __TODO: Give reasons__.
+
+`MethPipe` is perhaps the best documented and potentially extensible software for analysing methylation patterns at m-tuples.
+`MethPipe` is mostly written in C++ and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data.
+As such, it does not feature tools that are particularly amenable to interactive or exploratory analyses.
+In fact, I am unaware of any software that allows easy exploratory analyses of methylation patterns at m-tuples, which in part motivated the development of `MethylationTuples`.
 
@@
 
-A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are often lacking or, at best, very much tailored toward a particular analysis.
-What has been lacking is:
+In order to facilitate the development of downstream analysis methods based on methylation patterns at m-tuples, I saw the need for two pieces of software:
 
-1. A general method for extracting methylation patterns at m-tuples,
-2. Software for manipulating, analysing and visualising these methylatoin patterns.
+1. Software for extracting methylation patterns at m-tuples,
+2. Software for manipulating, analysing and visualising these methylation patterns.
 
@@
 
-`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples released under a Artistic-2.0 licence.
-R is a very popular language for analysing data, particularly in bioinformatics, and is my computational mother tongue, hence my choice of R as the language.
-To improve the performance of the package, parts are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
+`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples, It is released under an Artistic-2.0 licence (__TODO: Why use this instead of MIT/GNU?__).
+I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage.
+R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code.
+To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
 
-While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are also well-suited to analysing methylation entropy, allele-specific methylation, epialleles and epipolymorphism.
+While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are well-suited to any analysis based on methylation patterns of m-tuples such as methylation entropy, allele-specific methylation and the identification of epialleles.
+Of course, `MethylationTuples` can equally be used to develop methods based on 1-tuples, such as identifying differential methylation, since 1-tuples are just a particular type of m-tuple.
 
-`MethylationTuples` is written to work with the Bioconductor project \citep{Gentleman:2004ju}[^bioc].
+`MethylationTuples` is written to work within the Bioconductor project \citep{Gentleman:2004ju}[^bioc].
 Bioconductor makes extensive use of R's S4 object system and encourages developers to re-use existing Bioconductor infrastructure.
-From a developer's perspective, this avoids the need to re-invent the wheel.
-And from the user's perspective, it helps avoid multiple versions of the 'wheel', each that acts slightly differently and that may not be as well-tested.
+From a developer's perspective, this avoids the need to re-invent the wheel when tackling common tasks.
+And from the user's perspective, it helps avoid multiple versions of the 'wheel', each that might otherwise act slightly differently and that may not be as well-tested.
 
@@
 However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges].
-Therefore, I wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)).
+Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)).
 In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations.
-For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., `type = \`equal'`).
+For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \code{type = `equal'}).
 Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.
@@
 
-In the `MethylationTuples` package, I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated frequencies of the methylation patterns.
-The `MethPat` class extends the `SummarizedExperiment` class, also defined in the `GenomicRanges` package, but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples.
+In the `MethylationTuples` package, I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated counts of each methylation pattern.
 A `MethPat` object is as a matrix-like object, where rows represent m-tuples of interest and columns represent samples.
-Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size ($\text{m}$).
-Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing data on $n$ samples for 3-tuples.
+The `MethPat` class extends the `GenomicRanges::SummarizedExperiment`[^namespace] class but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples.
+Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size (i.e., same $\text{m}$).
+
+[^namespace]: This uses the NAMESPACE notation of R: `GenomicRanges::SummarizedExperiment` can be read as "the `SummarizedExperiment` class is part of the `GenomicRanges` package".
+
+Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing methylation patterns at 3-tuples for $n$ samples.
+The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples.
 
@@
 \caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples.
-Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the frequencies of the methylation patterns (grey box) are aligned.
-The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, S_{n}$.
-Some samples may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
+Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the counts of the methylation patterns (grey box) are aligned.
+The counts of each methylation pattern ($MMM, MMU, \ldots, UUU$) are stored as separate matrices where the columns represent samples ($S_{1}, \ldots, S_{n}$).
+Some samples may not have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
 \label{fig:MethPat}
@@
 
-\subsection{Available methods}
-
-The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples.
-A `MethPat` object can be constructed using the `MethPat` constructor function or from the output files of `methtuple` via the `readMethtuple` function.
-
 [^methtuple_output]: See Section \ref{sec:methtuple_methods} for example of the `methtuple` output format.
 
-The `MethPat` object provides fast subsetting by rows (m-tuples), including subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps` methods.
+\subsection{Methods}
+
+A `MethPat` object can be constructed directly using the `MethPat()` constructor function or from the output files of `methtuple` via the `readMethtuple()` function.
+
+The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the `[` method.
+It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods.
 Several other useful utility functions are:
 
-- `collapseStrand`: Collapse strand-specific data by aggregating the counts. Only applicable to CpG methylation loci.
-- `combine`: Combine multiple `MethPat` objects into one.
-- `filterOutVariants`: Remove m-tuples that contain a known variant. Variants must be provided as a VCF.
-- `getCoverage`: Compute the sequencing coverage of the m-tuples in each sample.
-- `IPD`: Compute the IPD vector for each m-tuple.
-- `methLevel`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
-- `MBias` and associated methods and functions: Helper functions to perform a more systematic analysis of M-bias (see Section \ref{sec:identifying_m-bias}).
-- `plotBetaDensity`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
-- `tuples`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.
+__UP TO HERE: Check these exist, update descriptions and add any new functions__
 
-These are in addition to the many useful utility functions inherited from the `SummarizedExperiment` class.
+- `collapseStrand()`: Collapse strand-specific data by aggregating the counts. Only applicable to CpG methylation loci.
+- `combine()`: Combine multiple `MethPat` objects into one.
+- `filterOutVariants()`: Remove m-tuples that contain a known variant. Variants must be provided as a VCF.
+- `getCoverage()`: Compute the sequencing coverage of the m-tuples in each sample.
+- `IPD()`: Compute the IPD vector for each m-tuple.
+- `methLevel()`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
+- `MBias()` and associated methods and functions: Helper functions to perform a more systematic analysis of M-bias (see Section \ref{sec:identifying_m-bias}).
+- `plotBetaDensity()`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
+- `tuples()`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.
+
+These are in addition to the many useful methods inherited from the `GenomicRanges::SummarizedExperiment` class.
 
@@
 
-Being based on core Bioconductor functionality, `MethylationTuples` is highly compatible with existing packages.
+Since `MethylationTuples` is based on core Bioconductor functionality, it is highly compatible with existing Bioconductor packages.
 In particular, `MethPat` objects containing 1-tuples are readily coerced for use in differential methylation calling packages, e.g., `bsseq` and `BiSeq`, or to identify partially methylated domains with `MethylSeekR`.
@@
 The challenges of working with large datasets in R are well-known.
-These are in large part due to R being designed as an 'in-memory' application and it's implementation of 'copy-on-modify' semantics \citep{wickham2014advanced}.
+These are in large part due to R being designed as an 'in-memory' application and its implementation of 'copy-on-modify' semantics \citep{wickham2014advanced}.
 More generally with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity --- a more efficient way of storing the data may be less convenient to work with and _vice versa_.

### 2015-04-20 Rmarkdown/appendix.Rmd
Latex fixes
Complete revisions to WGBS downstream analysis.

@@
 
-\section{Chapter \ref{chap:wgbs_downstream_analyses}}
+\chapter{Appendix}\label{chap:appendix}
 
-\subsection{Equivalence of tests to identify DMCs in a one-group experiment}\label{app:one_group_equivalence}
-
-\subsubsection{Mathematical equivalence}
-
-The saturated log-linear model can be written as $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{a, b}^{meth_state,sample}$, where $F_{a, b}$ is the value of the $(a, b)$ element in the $n \times 2$ contingency table.
-The null hypothesis that the cytosine is not a DMC is then tested by testing for an association between `sample` and `meth_state`, i.e., whether the parameter $\lambda_{a, b}^{meth_state, sample}$ is significantly different from zero.
-
-This log-linear model can be reformulated as a logistic regression by noting that the log-odds (logits) for each row are $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \log(F_{a, 1}) - \log(F_{a, 2})$. Therefore,
-
-\begin{align*}
-\log(\frac{F_{a, 1}}{F_{a, 2}}) &= \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{1}^{sample} + \lambda_{a, 1} \\
-&- (\lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{2}^{sample} + \lambda_{a, 2}) \\
-&= \underbrace{(\lambda_{1}^{sample} - \lambda_{2}^{sample})}_{\alpha} + \underbrace{(\lambda_{a, 1} - \lambda_{a, 2})}_{\gamma_{i}}
-\end{align*}
-
-The expression $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha + \gamma_{i}$ is a logistic regression model.
-We can test the null hypothesis that the cytosine is not a DMC by testing for a `sample`-effect on the log-odds, i.e., whether the parameter $\gamma_{i}$ is significantly different from zero.
-
-Regardless of whether we use the log-linear model or the logistic model, to test the null hypothesis that the cytosine is not a DMC we can compute the deviance[^deviance] of the reduced model and compare it to a $\chi^{2}$ random variable with $(2 - 1) \times (n - 1)$ degrees of freedom.
-For the log-linear regression the reduced model is $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample}$ and for the logistric regression the reduced model is $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha$.
-
-[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.
-
-\subsubsection{Computational equiavelence}
-
-The computational equivalence of using Pearson's $\chi^{2}$ test, the deviance of a log-linear model and the deviance of a logistic regression are shown in the following R code:
-
-```{r}
-# The data
-x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
-names(dimnames(x)) <- c("sample", "meth_state")
-x
-
-# Using Pearson's chi-square test
-chisq.test(x)
-
-# Fitting a log-linear model to perform iterative proportional fitting to the
-# contingency table.
-loglin(x, margin = list("sample", "meth_state"))
-# An alternative formulation using loglm(), a convenient wrapper of loglin(),
-# from the MASS package.
-library(MASS)
-loglm(formula =  ~ sample + meth_state, data = x)
-
-# Fitting a generalised linear model with a binomial link using two different
-# formulations
-# (1) Using the array of frequencies
-glm(x ~ 1, family = binomial)
-# (2) Using relative frequencies (i.e., the beta-values) and weights (i.e.,
-  # the sequencing coverage).
-  glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[, 1] + x[, 2])
-```
 \section{Chapter \ref{chap:co-methylation_review}}\label{app:co-methylation_review}
@@
 
-\subection{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
+\subsection{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
 

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Complete revisions to WGBS downstream analysis.

@@
 A barrier to analyses based on m-tuples has been a lack of software.
-To overcome this barrier, I introduce \code{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
-\code{MethylationTuples} complements \code{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples.
-I describe methods available in \code{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
+To overcome this barrier, I introduce \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
+\texttt{MethylationTuples} complements \texttt{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples.
+I describe methods available in \texttt{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
 \end{chapabstract}
@@
 
-We might estimate \widebar{B_{r}^{k}} by the group-wise average of the sample-wise weighted average of $\beta$-values for all methylation loci in the region, where the weights are proportional to the sequencing coverage.
+We might estimate $\widebar{B_{r}^{k}}$ by the group-wise average of the sample-wise weighted average of $\beta$-values for all methylation loci in the region, where the weights are proportional to the sequencing coverage.
 Then, identifying differential methylation at _a priori_ defined regions simply boils down to identifying differences in means, just as is the case for testing for DMCs.
@@
 I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations.
-Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` __TODO: Give reasons__.
+Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` due to compilation errors.
 
@@
 
-`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples, It is released under an Artistic-2.0 licence (__TODO: Why use this instead of MIT/GNU?__).
+`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples.
+It is released under an Artistic-2.0 license, consistent with core Bioconductor packages.
 I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage.
@@
 In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations.
-For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \code{type = `equal'}).
+For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \texttt{type = `equal'}).
 Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.
@@
 It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods.
-Several other useful utility functions are:
-
-__UP TO HERE: Check these exist, update descriptions and add any new functions__
+Several other useful utility functions for working with `MethPat` objects include:
 
@@
 - `combine()`: Combine multiple `MethPat` objects into one.
-- `filterOutVariants()`: Remove m-tuples that contain a known variant. Variants must be provided as a VCF.
-- `getCoverage()`: Compute the sequencing coverage of the m-tuples in each sample.
+- `filterOutVariants()`: Remove m-tuples that contain a known variant. Variants must be provided as a `VCF` file.
+- `findMTuples()`: Find m-tuples of a given size in a reference genome.
+- `getCoverage()`: Compute the sequencing coverage of each m-tuples in each sample.
 - `IPD()`: Compute the IPD vector for each m-tuple.
 - `methLevel()`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
-- `MBias()` and associated methods and functions: Helper functions to perform a more systematic analysis of M-bias (see Section \ref{sec:identifying_m-bias}).
-- `plotBetaDensity()`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
 - `tuples()`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.
@@
 
-In the `MethPat` class, its associated methods and other utility functions, the `MethylationTuples` package provides a toolbox for manipulating methylation patterns at m-tuples.
-Aside from providing the necessary infrastructure to analyse methylation patterns at m-tuples, `MethylationTuples` currently includes specific methods to analyse and visualise co-methylation (Chapter \ref{chap:co-methylation}) with the `cometh` and `methLevelCor` functions.
+With the `MethPat` class, its associated methods and other utility functions, the `MethylationTuples` package provides a toolbox for manipulating methylation patterns at m-tuples.
+Aside from providing the necessary infrastructure to analyse methylation patterns at m-tuples, `MethylationTuples` currently includes specific methods to analyse and visualise co-methylation (Chapter \ref{chap:co-methylation}) with the `cometh()` and `methLevelCor()` methods.
 I also plan to add methods for estimating epialleles and epipolymorphism.
-And it is my hope that `MethylationTuples` will provide a useful foundation on which others can implement their own methods for analysing methylation patterns at m-tuples.
+It is my hope that `MethylationTuples` will provide a useful foundation on which others can implement their own methods for analysing methylation patterns at m-tuples.
 
@@
 These are in large part due to R being designed as an 'in-memory' application and its implementation of 'copy-on-modify' semantics \citep{wickham2014advanced}.
-More generally with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity --- a more efficient way of storing the data may be less convenient to work with and _vice versa_.
+More generally, with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity; a more efficient way of storing the data may be less convenient to work with and _vice versa_.
 
-I have used the `MethylationTuples` package to analyses datasets containing up to 17 whole-genome bisulfite-sequencing samples (the Lister data), albeit using a server with $512$ Gb of (shared) memory.
-I have found the `MethPat` class to be a very convenient representation of the data, but one that does not coincide with the most efficient storage.
-This is more of a problem as the size of the m-tuples increases and the data become sparser.
+I have used the `MethylationTuples` package to analyse various sized m-tuples from datasets containing up to 17 whole-genome bisulfite-sequencing samples (the _Lister_ data).
+I have found the `MethPat` class to be a very convenient representation of the data, however, it has also raised challenges that will apply for larger datasets.
+
+One such challenge is the size of a `MethPat` object in RAM.
+The `MethPat` class currently favours a simpler implementation at the expense of storage efficiency.
+The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases.
+
+Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
+Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).
 However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.
-Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
-It also highlights the sparsity of the data as the size of the tuples grows --- most entries are $0$ (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).
 
@@
 
-The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases.
-It may be possible to improve this by using a different approach to the internal storage of the data.
-For example, it may be possible to create an index so that the matrices storing the frequencies of methylation patterns can be re-ordered and stored as run-length encodings, a very space-efficient storage scheme.
+It will be possible to improve the storage efficiency by using a different approach to the internal storage of the data.
+For example, it may be possible to use sparse matrices to store the counts or to create an index so that counts can be re-ordered and stored as run-length encodings, a very space-efficient storage scheme.
 
 In the short term, my aim for `MethylationTuples` is to release a version to Bioconductor with the core infrastructure for manipulating data of methylation patterns at m-tuples.
-In the longer term, I would like to extend the set of downstream analyses of m-tuples that are available in the package and for other users to do the same with their own methods.
+In the longer term, I would like to extend the set of downstream analyses of m-tuples that are available in the package and to help users extend the package to add their own methods.
+
+\section{Summary}
+
+Most methods for downstream analyses of bisulfite-sequencing data have focused on the problem of identifying differential methylation using methylation calls at 1-tuples.
+However, there is a growing interest in questions related to the heterogeneity of DNA methylation and these might be better addressed by analyses based on methylation patterns at m-tuples.
+
+A barrier to these type of analyses has been a lack of software for extracting and manipulating these data.
+It is my hope that the `methtuple` and `MethylationTuples` software will prove useful in facilitating the development of methods for these new types of downstream analyses.

### 2015-04-20 Rmarkdown/datasets.Rmd
Fix formatting

@@
 The reported read lengths are prior to any trimming of the reads.
-Abbreviations: \em{iPSC} = induced pluripotent stem cell; \em{ESC} = embryonic stem cell; \em{IVD} = \em{in vitro} differentiated from pluripotent cell line \em{PE} = single-end sequencing; \em{PE} = paired-end sequencing.}
+Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
 \label{tab:Lister_dataset}

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Fix formatting

@@
 
-\subsection{Using \em{a priori} regions}
+\subsection{Using \emph{a priori} regions}
 
@@
 
-One such challenge is the size of a `MethPat` object in RAM.
+One such challenge is the size of a `MethPat` object in memory.
 The `MethPat` class currently favours a simpler implementation at the expense of storage efficiency.
-The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases.
+The main inefficiency with the `MethPat` class is that the matrices storing the counts of each methylation pattern grow increasingly sparse as the size of the tuples increases.
 
-Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples.
+Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the _EPISCOPE_ whole-genome bisulfite-sequencing data with various sized m-tuples.
 Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).
@@
 \toprule
-& \texttt{pryr::object\_size(x)} & \texttt{nrow} & Number of assays & Percentage of \texttt{NA} and $0$ values \\ \midrule
-1-tuples                                 & $5.9$ GB                        & $56,348,522$    & $2$              & $28\%$                                     \\
-2-tuples                                 & $20.1$ GB                        & $100,586,237$   & $4$              & $80\%$                                     \\
-2-tuples (\texttt{--all-combinations}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
-3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
-4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
+& Size (GB) & Number of rows & Number of assays & Percentage of $0$ and \texttt{NA} values \\ \midrule
+1-tuples                                 & $5.9$                        & $56,348,522$    & $2$              & $28\%$                                     \\
+2-tuples                                 & $20.1$                       & $100,586,237$   & $4$              & $80\%$                                     \\
+2-tuples (\texttt{{-}{-}all-combinations}) & $60.0$                          & $299,814,999$   & $4$              & $78\%$                                     \\
+3-tuples                                 & $43.3$                       & $109,376,348$   & $8$              & $93\%$                                     \\
+4-tuples                                 & $80.5$                       & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
 \end{tabulary}
-\caption{Size of \texttt{MethPat} objects for the \em{EPISCOPE} data.
+\caption{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data.
 All m-tuples are stranded.
-The size of the \texttt{MethPat} object is computed using the \texttt{object\_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
-The number of rows (\texttt{nrow}) corresponds to the number of m-tuples in the object.
-The number of assays is $2^{\text{m}}$, where $m$ is the size of the m-tuples.
+The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
+The `number of rows' corresponds to the number of m-tuples in the object.
+The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples.
 The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}

### 2015-04-21 Rmarkdown/biology_background.Rmd
Changes suggested by Belinda to Biology Background

@@
 The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin.
-The linker sequences are between $20$ to $60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
+The linker sequences are between $20$ to $60$ basepairs of DNA (bp) in length, while approximately $147$ bp are wrapped around each nucleosome and a further $20$ bp wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
 
@@
 As is clear from the above description, the genome is a complex three-dimensional structure.
-Nonetheless, in bioinformatics and computational biology, the genome is mostly considered as a single-stranded, one-dimensional string of the nucleobases _A_, _C_, _G_ and _T_.
+Nonetheless, in bioinformatics and computational biology, the genome is mostly considered as a single-stranded, one-dimensional string of the bases _A_, _C_, _G_ and _T_.
 
@@
 
-Variation between individual's genomes, be it at single nucleotides[^snvs_snps] or across larger regions[^indels], is one important source of phenotypic variation.
+Variation between individual's genomes, be it at single base[^snvs_snps] or across larger regions[^indels], is one important source of phenotypic variation.
 Importantly, genetically-driven phenotypic variation is frequently _heritable_, meaning that phenotypes can be passed on from one generation to the generation via the genome.
@@
 
-When people speak of DNA methylation they are generally referring to methylation of the cytosine nucleobase.
+When people speak of DNA methylation they are generally referring to methylation of the cytosine base.
 Even more specifically, they are referring to the molecule _5-methylcytosine_.
@@
 Ruppel, first identified a methylated nucleic acid in 1898.
-Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide \citep{Ruppel:1899hc}.
+Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated base \citep{Ruppel:1899hc}.
 In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \citep{Johnson:1925js}.
@@
 
-The typical site of DNA methylation is at the C5 carbon position of a cytosine nucleotide.
-This modified cytosine is called _5-Methylcytosine_ or _5mC_.
-Figure \ref{fig:5mC} shows the structure of 5mC.
+The typical site of DNA methylation is at the C5 carbon position of a cytosine base. This modified cytosine is called _5-Methylcytosine_ or _5mC_. Figure \ref{fig:5mC} shows the structure of 5mC.
 
@@
 In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_.
-A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence.
+A CpG dinucleotide, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence.
 The 'p' stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation.
 A CpG is a palindromic sequence and is generally symmetrically methylated.
-Approximately 70% of CpGs are methylated in mammals \citep{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.
+Approximately $70\%$ of CpGs are methylated in mammals \citep{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.
 
@@
 CpGs are underrepresented in the human genome.
-The _GC-content_ of the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is approximately $41\%$.
-If these nucleotides were uniformly distributed across the genome then we would expect about $4.1\%$ of _dinucleotides_, pairs of nucleotides, to be CpGs.
+The _GC-content_ of the human genome, which is defined as the percentage of bases that are either guanines or cytosines \citep{Benjamini:2012er}, is approximately $41\%$.
+If these bases were uniformly distributed across the genome then we would expect about $4.1\%$ of dinucleotides to be CpGs.
 Instead, only $1\%$ of dinucleotides are CpGs.
@@
 Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}).
-There are many other evolutionary pressures on the distribution of nucleotides in a genome.
+There are many other evolutionary pressures on the distribution of bases in a genome.
 One effect of this is that the distribution of CpGs is far from uniform.
@@
 \caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19).
-The expected frequency is computed under an 'independence' model, based on the observed frequencies of each nucleotide.}
+The expected frequency is computed under an 'independence' model, based on the observed frequencies of each base.}
 \label{fig:dinucleotide_frequency_hg19}
@@
 
-CpG islands contain the 20-40% of CpGs that are frequently unmethylated in mammalian genomes.
+CpG islands contain the $20\%$ to $40\%$ of CpGs that are frequently unmethylated in mammalian genomes.
 CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}.
@@
 There are, however, certain cell types with widespread non-CpG methylation.
-Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H is the IUPAC code for any nucleotide except G ([http://www.bioinformatics.org/sms2/iupac.html](http://www.bioinformatics.org/sms2/iupac.html)).
+Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H is the IUPAC code for any base except G ([http://www.bioinformatics.org/sms2/iupac.html](http://www.bioinformatics.org/sms2/iupac.html)).
 The rule-of-thumb for mammalian genomes is that non-CpG methylation is rare in somatic cells but common in pluripotent cells.
@@
 Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation.
-This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k microarrays, only measure CpG methylation.
+This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k microarrays, measure almost exclusively CpG methylation.
 However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context (see Section \ref{sec:dna_methylation_assays}).
@@
 The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-cost throughput.
-Experiments that use an assay with high-resolution and high per-sample throughput generally have fewer samples than experiments using a lower resolution assay or an assay with lower per-sample throughput.
+Experiments that use an assay with high resolution and high per-sample throughput generally have fewer samples due to the associated higher costs than experiments using a lower resolution assay or an assay with lower per-sample throughput.
 
 In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays.
-In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.
+In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at single-base resolution.
 
@@
 
-While restriction enzyme assays do not typically provide single nucleotide resolution data, `methylCRF` \citep{Stevens:2013hv} is able to infer single CpG methylation levels by combining data from a restriction enzyme based assay (MRE-seq) with one based on affinity enrichment (MeDIP-seq) in a sophisticated statistical analysis.
+While restriction enzyme assays do not typically provide single-base resolution data, the `methylCRF` software \citep{Stevens:2013hv} is able to infer single CpG methylation levels by combining data from a restriction enzyme based assay (MRE-seq) with one based on affinity enrichment (MeDIP-seq) in a sophisticated statistical analysis.
 
@@
 Bisulfite-conversion assays are considered the gold standard for studying DNA methylation since cytosine methylation can be detected at single-base resolution \citep{Stirzaker:2014cu}.
-In fact, single-molecule, base pair resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].
+In fact, single-molecule, single-base resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].
 
-[^microarrays]: Microarray hybridisation assays can provide base pair resolution but not single molecule resolution.
+[^microarrays]: Microarray hybridisation assays can provide single-base resolution but not single-molecule resolution.
 The signal from a microarray-based experiment a sample-wide average since, for each locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.
@@
 The technique of \citet{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
-And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \citep{Smallwood:2014kn}.
+And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to $48.4\%$ of cpG sites" \citep{Smallwood:2014kn}.
 
@@
 \includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
-\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inferrence about the methylation state of the sample's genome.
-The locus in orange is a cytosine in the refererence genome but a thymine in the sample's genome.
+\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome.
+The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome.
 Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine.
@@
 The only commercially available assay in this class is the Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}.
-It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.
+It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a base, such as 5mC, to the time it takes to read its unmodified form.
 
@@
 
-Whole-genome bisulfite-sequencing provides single-nucleotide resolution data and is the gold standard assay for studying DNA methylation.
+Whole-genome bisulfite-sequencing provides single-base resolution data and is the gold standard assay for studying DNA methylation.
 Methods for analysing whole-genome bisulfite-sequencing data are the focus of my thesis.

### 2015-04-21 Rmarkdown/comethylation.Rmd
Remove CGI abbreviation.

@@
 
-[^hepg2]: The minor exception is the _HepG2\_cell\_line_ where the CGI and non-CGI lines intersect by $IPD = 1500$, but we have already seen that this sample has an unusual methylome compared to the other samples.
+[^hepg2]: The minor exception is the _HepG2\_cell\_line_ where the CpG island and non-island lines intersect by $IPD = 1500$, but we have already seen that this sample has an unusual methylome compared to the other samples.
 
@@
 
-[^hepg2_2]: Again, the minor exception is the _HepG2\_cell\_line_ where the CGI and non-CGI distributions approximately overlap by $IPD = 50$, but we have already seen that this sample has an unusual methylome compared to the other samples.
+[^hepg2_2]: Again, the minor exception is the _HepG2\_cell\_line_ where the CpG island and non-island distributions approximately overlap by $IPD = 50$, but we have already seen that this sample has an unusual methylome compared to the other samples.
 
@@
 Therefore, using the MH estimator requires the assumption that all $2 \times 2$ tables have the same underlying odds ratio.
-The strata for co-methylation would be 'sample', 'IPD' and possibly 'region', where region is something like 'UMRs, LMRs, PMRs and other' or 'CGI and non-CGI'.
+The strata for co-methylation would be 'sample', 'IPD' and possibly 'region', where region is something like 'UMRs, LMRs, PMRs and other' or 'CpG island and non-island'.
 

### 2015-04-21 Rmarkdown/comethylation.Rmd
Substitute ‘base’ for ‘nucleotide’

@@
 Another complexity that these analyses haved brushed over is the fact that the genome is not a 1D line but is rather a complex 3D structure whose shape also varies in time.
-Therefore, the relevant distance between two loci is not necessarily the number of nucleotides between them but may instead be the Euclidean distance in 3D space.
+Therefore, the relevant distance between two loci is not necessarily the number of basepair between them but may instead be the Euclidean distance in 3D space.
 Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the 3D structure of the genome, such as those provided by chromatin conformation capture technologies \citep{Dekker:2013hi}.

### 2015-04-21 Rmarkdown/comethylation_review.Rmd
Fix formatting
Substitute ‘base’ for ‘nucleotide’

@@
 
-In fact, there are well known periodicities in the distribution of nucleotides in genomes.
+In fact, there are well known periodicities in the distribution of bases in genomes.
 For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ bp period and the latter the $3$ bp period for exonic CpGs.
@@
 \includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star.pdf}
-\caption{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \em{FF} sample.}
+\caption{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \emph{FF} sample.}
 \label{fig:FF_chr21_acf_beta_star}
@@
 Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}).
-Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of nucleotides in the genome than anything to do with their methylation state.
+Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of bases in the genome than anything to do with their methylation state.
 
@@
 The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci.
-Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.
+Either method should give similar results since there will be few $31$ bp reads containing multiple pairs of CHH loci.
 

### 2015-04-21 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Minor formatting
Substitute ‘base’ for ‘nucleotide’

@@
 For example, `FastQC` will report a warning (resp.
-error) if the relative frequency of the four nucleotides differ by more than $10\%$ (resp.
+error) if the relative frequency of the four bases differ by more than $10\%$ (resp.
 $20\%$).
@@
 When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters].
-Strictly speaking each sequenced nucleotide is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all nucleotides within 'surviving' reads.
+Strictly speaking each sequenced bases is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all bases within 'surviving' reads.
 In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.
@@
 A cheaper alternative is to genotype the sample on a genome-wide SNP microarray.
-This will give very accurate, very cheap genotypes at a large number of loci ($500,000--5,000,000$).
+This will give very accurate, very cheap genotypes at a large number of loci ($500,000$ to $5,000,000$).
 However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.
@@
 Unfortunately, $C$>$T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants.
-Fortunately, it is often possible to distinguish $C$>$T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand; if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
+Fortunately, it is often possible to distinguish $C$>$T$ SNPs from unmethylated cytosines by examining the base on the opposing strand; if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}).
 Other base substitutions are more readily detectable, and insertion and deletion events (indels) may also be called.
@@
 
-1. Reference-specific methylation loci, i.e., cytosines in the reference genome that are mutated to non-cytosine nucleotides in the sample's genome.
-2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine nucleotides in the reference genome.
+1. Reference-specific methylation loci, i.e., cytosines in the reference genome that are mutated to non-cytosine bases in the sample's genome.
+2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine bases in the reference genome.
 3. Other genetic variants that may be used in additional analyses, such as in identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).
@@
 
-This is done by examining the two nucleotides upstream of the cytosine.
+This is done by examining the two bases upstream of the cytosine.
 It can be done based on the reference sequence, as is done in `Bismark` and `methtuple`, or from the reads themselves.
@@
 
-A further complication occurs when there is a genetic variant in the two nucleotides upstream of the methylation locus.
-We would like to use the two upstream nucleotides from the sample to infer the sample-specific methylation context, either inferred from each read separately or from a variant calling procedure such as `Bis-SNP`.
+A further complication occurs when there is a genetic variant in the two bases upstream of the methylation locus.
+We would like to use the two upstream bases from the sample to infer the sample-specific methylation context, either inferred from each read separately or from a variant calling procedure such as `Bis-SNP`.
 However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome alone.

### 2015-04-21 Rmarkdown/comethylation_review.Rmd
Complete my edits to Co-methylation

@@
 \chapter{A critical review of co-methylation}\label{chap:co-methylation_review}
-
-__TODO: Review co-methylation in "Bell, J. T. et al. DNA methylation patterns associate with genetic and gene expression variation in HapMap cell lines. Genome Biol 12, R10 (2011).__"
 
@@
 
+\subsection{\citet{Ball:2009kh}}
+
+\citet{Ball:2009kh} use bisulfite padlock probes with $36$ bp single-end reads to profile CpG methylation at approximately $7,000$ CpGs in an Epstein-Barr virus transformed B-lymphocyte cell line.
+As part of their study, \citeauthor{Ball:2009kh} sought to investigate whether the co-methylation reported by \citet{Eckhardt:2006gh} (Section \label{sec:eckhardt}) occurs at the single-molecule level.
+
+To do so, \citeauthor{Ball:2009kh} took all reads containing multiple CpGs and mapping to positions with intermediate $\beta$-values ($0.2 < \beta < 0.8$) with at least $100\times$ sequencing coverage.
+From these reads they computed the correlation of within-read methylation states.
+\citeauthor{Ball:2009kh} found that the distribution of these correlations is centred around $0.5$ with very few values below zero, evidence that methylation levels at CpGs on the same DNA fragment "are generally positively correlated".
+
+The obvious limitation to this analysis is the small number of CpG pairs interrogated by the assay.
+The short reads also only allow for the exploration of within-fragment co-methylation at very close CpGs and \citeauthor{Ball:2009kh} do not explore the effect of distance between CpGs ($IPD$) on co-methylation.
+
 \subsection{\citet{Lacey:2009by}}\label{sec:lacey_2009}
@@
 
-Between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci since it is a locus-specific correlation measured at the population level.
-
 Since it is a loci-specific, population-level correlation, between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci.
@@
 
-An example of between-sample co-methylation is given by \citet{Liu:dy}.
-The authors compute the correlation of pairs of $\beta$-values across samples to identify clusters of correlated CpGs and relate methylation changes at these clusters to nearby SNP genotypes.
-The resolution of their data is limited since it uses Illumina HumanMethylation450 array, but they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set".
+\citet{Bell:2011fm} identify evidence of between-sample co-methylation using the Illumina 27k microarray to measure DNA methylation in lymphoblastoid cell lines from 77 HapMap Yoruba individuals.
+Of note, they found that between-sample co-methylation decays as a function of the distance between CpGs ($IPD$) and is stronger within CpG islands than outside of CpG islands.
+However, it must be recognised that the Illumina 27k microarray measures DNA methylation at less than $1\%$ of all CpGs as is biased towards CpG islands.
+
+Another example of between-sample co-methylation is given by \citet{Liu:dy}.
+The authors compute the correlation of pairs of $\beta$-values between samples in three studies ($n = 247, n = 91$ and $n= 305$, respectively) to identify clusters of correlated CpGs and relate methylation changes at these clusters to nearby SNP genotypes.
+The resolution of their data is better than \citet{Bell:2011fm} since they use the Illumina 450k microarray, however it still only captures less than $2\%$ of all CpGs in the genome.
+\citeauthor{Liu:dy} conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set".
 They also note that, "methylation codependency [co-methylation] itself varies across the genome", as is to be expected.
-No software is provided to implement the method.
 
@@
 This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data.
+As far as I am aware, `coMET` is the only software specifically designed to estimate between-sample co-methylation and integrate this with other genomic annotations.
 
@@
 
-It has been difficult to properly account for co-methylation in downstream analyses due to the variety of definitions, the varying rigour and reproducibility of the published analyses and the lack of software to estimate co-methylation..
+The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses.
 Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data.
-Some authors have already sought to do this.
+In fact, some authors have already sought to do this.
 
-For example, the `bsseq` R/Bioconductor package for identifying differentially methylated regions from WGBS data \citep{Hansen:2011gu, Hansen:2012gr} cite \citet{Eckhardt:2006gh} as evidence that methylation levels at nearby CpGs are correlated.
-`bsseq` uses local binomial likelihood smoothing to exploit the spatial correlation of $\beta$-values.
-Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed estimates of the $\beta$-values are, on average, more accurate than the raw $\beta$-values.
-This is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error.
+For example, the R/Bioconductor packages `bsseq` \citep{Hansen:2011gu, Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} both cite \citet{Eckhardt:2006gh} as providing evidence of co-methylation, which can be exploited by spatial smoothing of the $\beta$-values prior to testing for differential methylation.
+Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed $\beta$-values are, on average, more accurate than the raw $\beta$-values.
+Both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis.
 
-The R/Bioconductor package `BiSeq` \citep{Hebestreit:2013ko} also smooths the $\beta$-values, albeit slightly differently to `bsseq` since it is designed for identifying differentially methylated regions from RRBS data.
-The authors also cite \citet{Eckhardt:2006gh} as justification for smoothing the $\beta$-values.
+Smoothing is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error.
+An important parameter when smoothing is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother.
+The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide.
+`BiSeq` uses a much smaller default window size of 80 bp, a decision that is  driven by its focus on RRBS data which has a higher CpG density than whole-genome bisulfite-sequencing data.
+Ideally, a data-driven bandwidth would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice.
+A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs.
+the rest of the genome (see Chapter \ref{chap:co-methylation}) (__TODO: Revisit in light of updated results in Co-methylation chapter__).
 
-To summarise, both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis and are motivated by an analysis of within-sample co-methylation.
-Whenever using statistical smoothing, an important parameter is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother.
-The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide.
-`BiSeq` uses a much smaller default window size of 80 bp, a decision that is again driven by its focus on RRBS data which has a higher CpG density than WGBS data.
-Ideally, a data-driven window size would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice.
-A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs.
-the rest of the genome (see Chapter \ref{chap:co-methylation}).
-
-`BiSeq` also estimates and tries to account for the spatial correlation between test statistics of differential methylation at neighbouring loci.
+In addition to smoothing the $\beta$-values, `BiSeq` also estimates and tries to account for the spatial correlation of test statistics when testing for differential methylation at neighbouring loci.
 This is not the same as looking for correlations between the $\beta$-values, but is rather based on looking for correlations amongst the _differences_ in $\beta$-values, since these differences are what the test statistics of differential methylation are based upon.
@@
 They do this by analysing the distribution of P-values rather than the distribution of the test statistics themselves or the underlying data.
-This makes these methods quite general but more application-specific methods may be more powerful.
+This makes these methods quite general but application-specific methods may be more powerful.
 `comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \citep{Pedersen:2012vl}.
@@
 
-\citet{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it.
-They study the dynamics of DNA methylation in differentiating cell lineages.
-For each locus there is a correlation between its methylation state in the precursor cell and its dependent cell types, which they call "vertical" correlations in contrast to the "horizontal" correlations of within-sample co-methylation.
-\citet{Capra:2014hx} are bringing well-established methods for the study of DNA sequence evolution to the study of DNA methylation dynamics.
-In this specialised setting, where "between-samples" means "between samples in a lineage", they show that a method based on "vertical" correlations is better at imputing the methylation level of CpGs than a method based on "horizontal" within-sample co-methylation.
+\citet{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it, to study the dynamics of DNA methylation in differentiating cell lineages.
+For each locus there is a correlation between its methylation state in the precursor cell and its dependent cell types, which they call "vertical" correlations (in contrast to the "horizontal" correlations of within-sample co-methylation).
+In this specialised setting, where 'between-samples' means "between samples in a lineage", they show that a method based on "vertical" correlations is better at imputing the methylation level of CpGs than a method based on "horizontal" within-sample co-methylation.
 
@@
 While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult.
-As I have shown, there are problems with most, if not all, previous analyses.
+As I have shown, there are problems and limitations with most, if not all, previous analyses.
 These problems range from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation.

### 2015-04-21 Rmarkdown/paper_review.Rmd
Add a couple of paper reviews

@@
 
+## \citet{Bell:2011fm}
+
+> measured methylation levels at 22,290 CpG dinucleotides in lymphoblastoid cell lines from 77 HapMap Yoruba individuals, for which genome-wide gene expression and genotype data were also available.
+
+Cite Eckhardt et al.
+and [http://www.ncbi.nlm.nih.gov/pubmed/19329998](http://www.ncbi.nlm.nih.gov/pubmed/19329998) as previous evidence that "Methylation is often found to be correlated across genomic regions at the scale of 1-2 kb".
+
+> We observed that methylation levels at probes located in close proximity (up to 2 kb apart) were highly correlated (Figure 1c), indicating that variation in methylation levels between individuals is correlated within cell type.
+Figure 1c also shows that pairs of CpG-sites that were both within a CGI showed greater evidence for co-methylation than pairs of CpG sites for which at least one was outside the CGI, controlling for distance, implying differential regulation of DNA methylation for CpGs inside and outside of CGIs.
+
+This is an estimate of between-sample co-methylation.
+
+## \citet{Ball:2009kh}
+
+> As CpGs in close proximity are known to be often co-methylated [cite Ekhardt et al.], we took advantage of the clonal feature of Illumina sequencing to investigate whether co-methylation occurs at the single-molecule level.
+Within probes spanning more than one CpG, sites with intermediate methylation levels (between 20% and 80%) displayed a positive correlation with the methylation states of neighboring CpGs on individual strands (Supplementary Fig.
+3 online).
+
 # Review of simulation methods

### 2015-04-21 Rmarkdown/comethylation.Rmd
Plan of attack for finishing co-methylation chapter.

@@
 The _Miettenen_ estimator is like a generalised form of the Mantel-Haenszel estimator, but assumes each cell count is greater than approximately 4 (Greenland, S. Stat Med 1, 217–227 (1982)).
+
+## Plan of attack
+
+Following discussion with Terry (2015-04-21), this is my plan of attack to finish this chapter and the `methsim` chapter.
+
+- Read up on what happens with the MH estimator when the common odds ratio assumption is invalid.
+  - Is there a better estimator in this setting? The Miettennen estimator?
+- Demonstrate via simulation that the unconditional LOR estimator applied to $2 \times 2$ tables is biased when $d$ is small and the marginal probabilities are near the boundaries.
+  - Corrollary: Don't use the LOR estimates of within-fragment co-methylation from individual $2\times 2$ tables. I know they're biased, so there's no reason to show these results.
+- Demonstrate that the MH estimator applied to $2 \times 2 \times K$ tables is unbiased when $d$ is small (per-table) and the marginal probabilities are near the boundaries.
+  - Need to investigate the setting where the marginal probabilities are heterogeneous across the $K$ tables but with the same odds-ratio.
+  - Need to investigate the setting where the odds ratio __is not__ common to the $K$ tables; what does the MH estimator end up estimating?
+- Compute MH estimates (and standard errors) of within-fragment co-methylation for different stratification strategies.
+  - Stratify by IPD
+  - Stratify by IPD + chromosome
+  - Stratify by IPD + chromosome + chromosome arm
+  - Stratify by IPD + chromosome + chromosome arm + bin (Mb)
+  - Check whether the more-stratified estimates fall within a reasonable distance from the less-stratified estimates (i.e., with respect to standard errors of the less-stratified estimates).
+- Settle on the most appropriate level of stratification and apply to all samples.
+  - Discuss results.
+- `methsim`: Sample LORs from a Normal(mu, sigma^2) distribution, where mu (resp.
+sigma) is the MH estimator (resp.
+standard error?
+empirical variance of stratified MH estimators?) of within-fragment co-methylation for that IPD and region-type.

### 2015-04-22 Rmarkdown/comethylation.Rmd
Massive improvements to simulation model exploring different estimators of the odds ratio.

--- 2015-04-21 Rmarkdown/comethylation.Rmd
@@
 
+__TODO: Move this code chunk to supplementary material and only include results in thesis.__
+
 ```{r}
-# Function to simulate K 2-tuples with Poisson(mean_d) sequencing coverage and marginal
-# probabilities p1, p2
-simTables <- function(K, mean_d, p1, p2) {
-  .simTables <- function(mean_d, p1, p2) {
-
-    d <- rpois(1, mean_d)
-    z1 <- factor(rbinom(d, 1, p1), levels = c(0, 1))
-    z2 <- factor(rbinom(d, 1, p2), levels = c(0, 1))
-    xtabs(~ z2 + z1, drop.unused.levels = FALSE)
-  }
-  replicate(K, .simTables(mean_d, p1, p2))
+# Requires the installation of methsim
+library(mipfp)
+
+# rtpois is a helper function to simulate n truncated Poisson(lambda) with
+# truncation specified by 'trunc'. 
+# It is adapted from a method for simulating zero-truncated Poisson suggested 
+# by Peter Dalgaard https://stat.ethz.ch/pipermail/r-help/2005-May/070680.html
+rtpois <- function(n, lambda, trunc) {
+  qpois(runif(n, ppois(trunc, lambda), 1), lambda) 
+}
+
+# Function to simulate 2x2xK tables.
+# Each 2x2 table represents the counts of unmethylated (0) and methylated (1) states
+# at each CpG for each read.
+# The number of reads per table (i.e., 2x2 table sample size) is simulated 
+# from a truncated Poisson(lambda) distribution; truncation is to simulate
+# minimum coverage cutoffs and is specified by the 'trunc' argument.
+# The marginal probabilities of each table are p1 = Prob(methylated at first CpG)
+# and p2 = Prob(methylated at second CpG).
+# The odds ratio (or) is for each 2x2 table and should be of length K or 1 (in
+# which case all K tables have the same odds ratio).
+simTables <- function(K, lambda, trunc, p1, p2, or) {
+
+  # d is the sample size of each 2x2 table (i.e, the sequencing depth for the 2-tuple)
+  d <- rtpois(n = K, lambda = lambda, trunc = trunc)
+  if (length(p1) == 1L) {
+    p1 <- rep(p1, K)
+  }
+  if (length(p2) == 1L) {
+    p2 <- rep(p2, K)
+  }
+  if (length(or) == 1L) {
+    or <- rep(or, K)
+  }
+  .simTables <- function(k, d, p1, p2, or) {
+    # Compute joint probabilities
+    P <- mipfp::Ipfp(seed = matrix(c(or, 1, 1, 1), ncol = 2), 
+                     target.list = list(1, 2), 
+                     target.data = list(c(1 - p1, p1), c(1 - p2, p2)))$x.hat
+    matrix(rmultinom(1, d, as.vector(P)), ncol = 2)
+  }
+  mapply(.simTables, k = K, d = d, p1 = p1, p2 = p2, or = or, SIMPLIFY = "array")
 }
@@
 LOR <- function(x, offset = 0.5) {
-  MM <- x[2, 2, ] + offset
-  MU <- x[2, 1, ] + offset
-  UM <- x[1, 2, ] + offset
-  UU <- x[1, 1, ] + offset
-  log(MM) + log(UU) - log(MU) - log(UM)
+  x <- x + offset
+  MM <- x[2, 2, ]
+  MU <- x[2, 1, ]
+  UM <- x[1, 2, ]
+  UU <- x[1, 1, ]
+  log2(MM) + log2(UU) - log2(MU) - log2(UM)
 }
@@
 # Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
-MH <- function(x, offset = 0.5) {
+MH <- function(x, offset = 0) {
+  x <- x + offset
   # TODO: return the standard error
@@
   d <- apply(x, 3, sum)
-  or_mh <- sum((UU * MM) / d) / sum((MU * UM) / d)
-  log(or_mh)
+  or <- sum((UU * MM) / d) / sum((MU * UM) / d)
+  # One or more margins must be identically zero so add offset
+  if (is.nan(or) | isTRUE(all.equal(or, 0)) | is.infinite(or)) {
+    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
+    MH(x, 0.5)
+  } else {
+    log2(or)
+  }
 }
@@
 # 2x2xK collapsed over K.
-LORCollapsed <- function(x, offset) {
+LORCollapsed <- function(x, offset = 0.5) {
   z <- apply(x, c(1, 2), sum)
@@
   dim(z) <- c(dim(z), 1)
-  LOR(z)
-}
-
-# Simulate and plot data
-simAndPlot <- function(K, mean_d, p1, p2) {
-  x <- mapply(function(p1, p2, K, mean_d) {
-    simTables(K = K, mean_d = mean_d, p1 = p1, p2 = p2)
-  }, p1 = p1, p2 = p2, MoreArgs = list(K = K, mean_d = mean_d), SIMPLIFY = FALSE)
-
+  lor <- LOR(z, 0)
+  if (is.nan(lor) | is.infinite(lor)) {
+    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
+    LOR(z, 0.5)
+  } else {
+    lor
+  }
+}
+
+# Simulate and plot data when p1 = p2.
+# 100 CpG pairs, 30x average coverage truncated at 10
+# odds ratio = 1, i.e., independence
+set.seed(666)
 par(mfrow = c(3, 3))
-  mapply(function(x, p1, p2) {
-    hist(LOR(x), main = paste0("p1 = ", p1, ", p2 = ", p2), xlim = c(-8, 8), breaks = 16)
+K <- 100
+lambda <- 20
+trunc <- 11
+or <- 1
+p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+p2 <- p1
+for (i in 1:9) {
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i], p2 = p2[i], or = or)
+  hist(LOR(x), main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)), 
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
   abline(v = MH(x), col = "dodgerBlue", lwd = 3)
   abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-    abline(v = 0, col = "black", lty = 2)
-  }, x = x, p1 = p1, p2 = p2)
-  title(paste0("K = ", K, ", mean_d = ", mean_d), outer = TRUE, line = -1)
-}
-
-# Vary p1 with p2 = p1 at 30x coverage across 1000 CpG pairs
+  abline(v = log2(or), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 = p2.
+# 1000 CpG pairs, 30x average coverage truncated at 10
+# odds ratio = 1, i.e., independence
+set.seed(666)
+par(mfrow = c(3, 3))
 K <- 1000
-mean_d <- 30
+lambda <- 20
+trunc <- 10
+or <- 1
 p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
 p2 <- p1
-simAndPlot(K, mean_d, p1, p2)
-
-# Vary p1 with p2 = 1 - p1 at 30x coverage across 1000 CpG pairs
+for (i in 1:9) {
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i], 
+                 p2 = p2[i], or = or)
+  hist(LOR(x), 
+       main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)), 
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  abline(v = log2(or), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# Simulate and plot data when p1, p2 ~ Uniform(0, 1)
+# 100 CpG pairs, 30x average coverage truncated at 10
+# log odds ratio = -4, -3, ..., 3, 4.
+set.seed(666)
+par(mfrow = c(3, 3))
+K <- 100
+lambda <- 20
+trunc <- 11
+or <- 2^seq(-4, 4, 1)
+p1 <- runif(K)
+p2 <- runif(K)
+for (i in 1:9) {
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, 
+                 or = or[i])
+  hist(LOR(x), 
+       main = paste0("p1 = U(0, 1), p2 = U(0, 1)", ", lor = ", log2(or[i])),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  abline(v = log2(or[i]), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), 
+      outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 ~ Uniform(0, 1), p2 = p1
+# 100 CpG pairs, 30x average coverage truncated at 10
+# log odds ratio = -4, -3, ..., 3, 4.
+set.seed(666)
+par(mfrow = c(3, 3))
+K <- 100
+lambda <- 20
+trunc <- 11
+or <- 2^seq(-4, 4, 1)
+p1 <- runif(K)
+p2 <- p1
+for (i in 1:9) {
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
+  hist(LOR(x), main = paste0("p1 = U(0, 1), p2 = p1", ", lor = ", log2(or[i])),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       frequeny = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  abline(v = log2(or[i]), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), 
+      outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
+# 100 CpG pairs, 30x average coverage truncated at 10
+# log odds ratio = -4, -3, ..., 3, 4.
+# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
+# levels (B).
+set.seed(666)
+par(mfrow = c(3, 3))
+K <- 100
+lambda <- 20
+trunc <- 11
+or <- 2^seq(-4, 4, 1)
+p1 <- rbeta(K, 0.3, 0.2)
+p2 <- p1
+for (i in 1:9) {
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
+  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = ", log2(or[i])),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  abline(v = log2(or[i]), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
+# 100 CpG pairs, 30x average coverage truncated at 10
+# log2(or) ~ Gaussian(log2(or_mean), 1)
+# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
+# levels (B).
+set.seed(666)
+par(mfrow = c(3, 3))
+K <- 100
+lambda <- 20
+trunc <- 11
+or_mean <- 2 ^ seq(-4, 4, 1)
+p1 <- rbeta(K, 0.3, 0.2)
+p2 <- p1
+for (i in 1:9) {
+  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
+  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  lines(density(log2(or)), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
+# 1000 CpG pairs, 30x average coverage truncated at 10
+# log2(or) ~ Gaussian(log2(or_mean), 1)
+# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
+# levels (B).
+set.seed(666)
+par(mfrow = c(3, 3))
 K <- 1000
-mean_d <- 30
-p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
-p2 <- 1 - p1
-simAndPlot(K, mean_d, p1, p2)
-
-# Vary p1 with p2 = p1 at 1000x coverage across 1000 CpG pairs
+lambda <- 20
+trunc <- 11
+or_mean <- 2 ^ seq(-4, 4, 1)
+p1 <- rbeta(K, 0.3, 0.2)
+p2 <- p1
+for (i in 1:9) {
+  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
+  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  lines(density(log2(or)), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
+# 1000 CpG pairs, 30x average coverage truncated at 10
+# log2(or) ~ Skewed-Gaussian(log2(or_mean), 1, -10)
+# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
+# levels (B).
+library(sn)
+set.seed(666)
+par(mfrow = c(3, 3))
 K <- 1000
-mean_d <- 1000
-p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
+lambda <- 20
+trunc <- 11
+or_mean <- 2 ^ seq(-4, 4, 1)
+p1 <- rbeta(K, 0.3, 0.2)
 p2 <- p1
-simAndPlot(K, mean_d, p1, p2)
-
-# TODO: p1 and p2 distributed U(0, 1) at 30x coverage across 1000 CpG pairs
-# Will need to simulate p1 and p2 within .simTable().
+for (i in 1:9) {
+  or <- 2 ^ rsn(K, log2(or_mean[i]), 1, -10)
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
+  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = SN(", log2(or_mean[i]), ", 1, -10)"),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  lines(density(log2(or)), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
+# 1000 CpG pairs, 30x average coverage truncated at 10
+# log2(or) ~ 0.5 * Gaussian(log2(or_mean), 1) + 0.5 * Gaussian(-log2(or_mean))
+# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
+# levels (B).
+set.seed(666)
+par(mfrow = c(3, 3))
+K <- 1000
+lambda <- 20
+trunc <- 11
+or_mean <- 2 ^ seq(-4, 4, 1)
+w <- sample(c(-1, 1), K, replace = TRUE)
+p1 <- rbeta(K, 0.3, 0.2)
+p2 <- p1
+for (i in 1:9) {
+  or <- 2 ^ rnorm(K, w * log2(or_mean[i]), 1)
+  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
+  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = NM(", log2(or_mean[i]), ", 1)"),
+       xlim = c(-10, 10), breaks = 20,
+       xlab = expression(paste(log[2], "(odds ratio)")),
+       freq = FALSE)
+  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
+  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
+  lines(density(log2(or)), col = "black", lty = 2)
+}
+title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
 ```
@@
 The _Miettenen_ estimator is like a generalised form of the Mantel-Haenszel estimator, but assumes each cell count is greater than approximately 4 (Greenland, S. Stat Med 1, 217–227 (1982)).
+
+## Plan of attack
+
+Following discussion with Terry (2015-04-21), this is my plan of attack to finish this chapter and the `methsim` chapter.
+
+- [x] Finish reading Greenland (1982).
+  - What happens with the MH estimator when the common odds ratio assumption is invalid. Is there a better estimator in this setting?
+
+  > As a consequence, when there is heterogeneity, the expectation of MH varies with the overall and stratum-specific case-control ratios.
+The usual common odds ratio estimators share this disconcerting property \citep{Greenland:1982uk}.
+
+  In contrast to the MH estimator, the Miettenen estimator consistently estimates meaningful epimiological parameters. However, it "[requires] consideration in the selection of the weights".
+
+- [ ] Demonstrate via simulation that the unconditional LOR estimator applied to $2 \times 2$ tables is biased when $d$ is small and the marginal probabilities are near the boundaries.
+  - Corrollary: Don't use the LOR estimates of within-fragment co-methylation from individual $2\times 2$ tables. I know they're biased, so there's no reason to show these results.
+- [ ] Demonstrate that the MH estimator applied to $2 \times 2 \times K$ tables is unbiased when $d$ is small (per-table) and the marginal probabilities are near the boundaries.
+  - Need to investigate the setting where the marginal probabilities are heterogeneous across the $K$ tables but with the same odds-ratio.
+  - Need to investigate the setting where the odds ratio __is not__ common to the $K$ tables; what does the MH estimator end up estimating?
+
+- Run Woolf's test or Breslow-Day test of homogeneity of odds ratios (noting that these are not powerful tests).
+
+__In fact, Breslow himself advocates for the Tarone statistic in place of the Breslow-Day statistic (Breslow, N.
+E.
+JASA 91, 14–28 (1996)).
+Also, Woolf's test, Breslow-Day and Tarone's statistic are only appropriate in Asymptopia I (a small number of tables with large frequencies) and not in Asymptopia II (a large number of tables with small frequencies); see Liang and Self (1985) for Asymptopia II (the scenario I am in)__
+
+Liang, K.-Y. & Self, S. G. Biometrika (1985) advocate for a conditional test of homogeneity of odds ratios when the data are sparse (Asymptopia II) but note that:
+
+> The problem with the conditional approach is its complexity.
+It is clear that when the sample size of each table increases, the computation of T2, T4 or T5 [ the conditional tests] becomes a burden which, eventually, will be prohibitive.
+This raises the question of how large the sample sizes in each table must be before the unconditional score test T1 is reasonable to use.
+Further simulation results suggest that even when the number of cases and controls are as small as 10 with as many as 10 tables, the large-sample chi-squared approximation is adequate in that the observed sizes are near to the nominal 0 05 level.
+
+I have $K \\g 10$ ($K$ = number of $2 \times 2$ tables) and $d \geq 10$ ($d$ = sequencing coverage), so the unconditional test may well be fine.
+
+__TODO: Look up Liang-Self in Agresti's CDA (2013)__
+
+- Algorithm: If the null of homogeneity of odds ratio is rejected => stratify further(?), otherwise compute MH or Miettenen estimates of the log odds ratio (and standard errors) for different stratification strategies (see [https://stat.ethz.ch/pipermail/r-help/2007-February/126254.html](https://stat.ethz.ch/pipermail/r-help/2007-February/126254.html))
+  - Stratify by IPD
+  - Stratify by IPD + chromosome
+  - Stratify by IPD + chromosome + chromosome arm
+  - Stratify by IPD + chromosome + chromosome arm + bin (Mb)
+  - Check whether the more-stratified estimates fall within a reasonable distance from the less-stratified estimates (i.e., with respect to standard errors of the less-stratified estimates).
+- Settle on the most appropriate level of stratification and apply to all samples.
+  - Discuss results.
+- `methsim`: Sample LORs from a Normal(mu, sigma^2) distribution, where mu (resp.
+sigma) is the MH estimator (resp.
+standard error?
+empirical variance of stratified MH estimators?) of within-fragment co-methylation for that IPD and region-type.

### 2015-04-22 Rmarkdown/comethylation.Rmd
Begin re-write of Co-methylation.

@@
 
-# TODOs
-
-- Try out Matthew Stephen's adaptive hierarchical shrinkage method on the estimated log odds ratios (see [https://github.com/stephens999/ash](https://github.com/stephens999/ash)).
-
-## Chapter overview
-
-In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
-Using both these measures I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci.
+\begin{chapabstract}
+I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
+The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.
+
+Using estimates of both within-fragment co-methylation and correlation of $\beta$-values, I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci.
 I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.
@@
 Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
-I do not explore co-methylation of non-CpG methylation loci because these loci are rarely methylated in the samples that I am studying.
-
-\FloatBarrier
+\end{chapabstract}
+
 \section{Correlations of $\beta$-values}\label{subsec:beta_cor}
@@
 
+\subsection{Summary}
+
+__TODO: Summarise the results for correlation of $\beta$-values.__
+
+A limitation to analysing co-methylation using correlations of $\beta$-values is that the analysis is based on correlations of measurements that are aggregates from many thousands of DNA molecules.
+We are not measuring co-methylation on the scale at which the biological process of DNA methylation acts, namely at the level of individual DNA molecules.
+As noted by \citet{Ball:2009kh}, "the clonal feature of Illumina sequencing [allows us] to investigate whether co-methylation occurs at the single-molecule level".
+Before doing so, I first explore the challenge of estimating meaningful measures of dependence in sparse $2 \times 2 \times K$ contingency tables.
+While perhaps not yet apparent, this will be essential for the analysis of within-fragment co-methylation.
+
 \FloatBarrier
+
+\section{Estimating dependence in sparse $2 \times 2 \times K$ contingency tables}
+
+Consider two binary random variables, $X$ and $Y$.
+Let $\pi_{ij}, i, j = 1, 2$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$.
+We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$.
+There are many reasons why we might wish to estimate the odds ratio (__CITE__), but our interest here is __WHAT__.
+
+Suppose we observe $(X, Y)$ for $n$ samples.
+Denote by $n_{ij}, i, j = 1, 2$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$.
+We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 \cdot} = n_{11} + n_{12}$.
+The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}cccc@{}}
+\cmidrule(l){2-4}
+ & \multicolumn{3}{c}{Column} \\
+ \cmidrule(l){2-4}
+\multirow{3}{*}{Row} & & & \\
+ & $n_{11}$       & $n_{12}$       & $n_{1 \cdot}$ \\
+ & $n_{21}$       & $n_{22}$       & $n_{2 \cdot}$ \\
+ & $n_{\cdot 1}$  & $n_{\cdot 2}$  & $n_{\cdot \cdot} = n$ \\ \cmidrule(l){2-4}
+\end{tabular}
+\caption{Notation for a $2 \times 2$ contingency table.}
+\label{tab:2x2}
+\end{table}
+
+The simplest estimator of the odds ratio is the unconditional maximum likelihood estimator, $\widehat{\psi_{U}} = \frac{n_{11} n_{22}}{n_{12} n_{21}}$.
+
+- Go through the various estimators, their pros and cons.
+   - Conditional maximum likelihood estimator
+- Explain the advantage of the "0.5-adjusted" unconditional estimator.
+- Move to $2 \times 2 \times K$ tables
+  - Go through same comparison of estimators
+- Move to sparseness
+  - Asymptopia I and II
+  - Go through same comparison of estimators
+  - The homogeneity assumption
+
+ ## Setting for studying within-fragment co-methylation
+
+ - Asymptopia II with sparse tables
+ - Homogeneity??? Don't/can't know.
+
+ ## Simulation
+
+-
+
+## Summary
+
 \section{Within-fragment co-methylation}\label{subsec:wf_cometh}
@@
 For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type.
-This method is implemented in the `cometh` function that is part of the `MethylationTuples` software.
-
-The aim of this analysis is to address the question of 'how are methylation states on individual DNA fragments correlated as a function of the distance between loci?'.
+This method is implemented in the `cometh()` function that is part of the `MethylationTuples` software.
+
+The aim of this analysis is to address the questions:
+
+1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
+2. What influences this dependence?
+
 There are two steps in this analysis:

### 2015-04-23 Rmarkdown/comethylation.Rmd
Estimation of odds ratios in 2x2 and 2x2xK tables

@@
 
+We will begin with an overview of the simplest case of estimating dependence in $2 \times 2$ contingency tables, i.e., $K = 1$.
+We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables.
+We conclude with a simulation study highlighting the performance of several difference estimators and determining which is most appropriate for a study of within-fragment co-methylation.
+
+\subsection{$2 \times 2$ contingency tables}
+
 Consider two binary random variables, $X$ and $Y$.
@@
 There are many reasons why we might wish to estimate the odds ratio (__CITE__), but our interest here is __WHAT__.
+Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:
+
+- $\psi = 1$: $X$ is independent of $Y$.
+- $\psi > 1$: Subjects are more likely to have the same level for both $X$ and $Y$, e.g., subjects with $(X, Y) = (1, 1)$ or $(X, Y) = (2, 2)$ are more likely than subjects with $(X, Y) = (1, 2)$ or $(X, Y) = (2, 1)$.
+- $\psi < 1$: Subjects are more likely to have the the opposite level for $X$ than they do for $Y$, e.g., subjects with $(X, Y) = (1, 2)$ or $(X, Y) = (2, 1)$ are more likely than subjects with $(X, Y) = (1, 1)$ or $(X, Y) = (2, 2)$.
+
+The possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios.
+Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\phi = \log{\psi}$.
+The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler.
+I will routinely switch between discussing the estimation of $\psi$ and $\phi$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.
 
@@
 
-The simplest estimator of the odds ratio is the unconditional maximum likelihood estimator, $\widehat{\psi_{U}} = \frac{n_{11} n_{22}}{n_{12} n_{21}}$.
-
-- Go through the various estimators, their pros and cons.
-   - Conditional maximum likelihood estimator
-- Explain the advantage of the "0.5-adjusted" unconditional estimator.
+The simplest estimator of $\phi$ is the unconditional maximum likelihood estimator, $\widehat{\phi_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$.
+The asymptotic standard error of $\widehat{\phi_{U}}$ is given by $\hat{\sigma}(\widehat{\phi_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
+The asymptotic distribution of $\widehat{\phi_{U}}$ as $n \to \infty$ is $Normal(\phi, \hat{\sigma}(\phi)^{2})$.
+
+A problem with the estimator $\widehat{\phi_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities.
+We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\phi_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\phi_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
+\citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\phi_{U}}$ and $\widehat{\phi_{0.5}}$ have the same asymptotic distribution around $\phi$ as $n \to \infty$ but that $\widehat{\phi_{0.5}}$ has reduced first-order bias.
+The modified estimator, $\widehat{\phi_{0.5}}$, is therefore generally recommended.
+
+Another alternative to $\widehat{\phi_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\phi_{C}}$.
+This is computed by first conditioning on ${n_{1 \cdot}, n_{\cdot 1}}$.
+Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 \cdot}, n_{\cdot, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}.
+The maximum likelihood estimator of $\psi$, $\widehat{psi_{C}}$ is solved using iterative methods.
+
+The conditional estimator, $\widehat{\phi_{C}}$, works better than the unconditional estimator, $\widehat{\phi_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+Estimation of the odds ratio is made more difficult when the contingency table is _sparse_.
+We say that a contingency table is sparse when some of the $n_{ij}$ are small.
+Sparsity can occur even when the sample size, $n$, is large.
+Sparsity becomes more of a problem as we move from a single $2 \times 2$ table to $K$ $2 \times 2$ tables or, equivalently, a $2 \times 2 \times K$ table.
+
+\subsection{$2 \times 2 \times K$ contingency tables}
+
+Suppose we now measure the same two binary variables $X$ and $Y$ in $K$ different strata.
+We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum.
+For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{\cdot 1k} = 1 - \pi_{\cdot 2k}, \pi_{1 \cdot k} = 1 - \pi_{2 \cdot k}$ to vary.
+
+We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\phi = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
+
+Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{\cdot \cdot k})}{n_{12k} n_{21k} / n_{\cdot \cdot k}}$.
+\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\phi_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\phi_{MH}})$.
+This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance].
+This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.
+
+[^mh_estimator]: Other estimators had been proposed but were supplanted by the work of \citet{Robins:1986vh}.
+
+The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{\cdot \cdot k}$, grows large.
+The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{\cdot \cdot k}$ remains small.
+The variance estimator proposed by \citet{Robins:1986vh} is consistent under both $A1$ and $A2$.
+For our study of within-fragment co-methylation, $A2$ is the more relevant asymptotic environment.
+
+\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, converge to the true odds ratio.
+In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi_{MH}}$, and the conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, do convert to $\psi$, with $\widehat{\psi_{MH}}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.
+
+\subsection{Homogeneity of odds ratios assumption}
+
+So far we have assumed a common odds ratio for the $K$ $2 \times 2$ tables.
+
+__UP TO HERE__
+
 - Move to $2 \times 2 \times K$ tables
@@
 title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
+
+# TODO: Increase d for each table
 ```

### 2015-04-23 latex/phd_thesis.tex
Estimation of odds ratios in 2x2 and 2x2xK tables

@@
 
+% TODO: Check list of abbreviations
 \chapter{List of Abbreviations}
@@
 HMM & hidden Markov model \\
-CNV & copy number variant \\
 SE & single-end sequencing \\
@@
 indel & insertion or deletion \\
+CGI & CpG island \\
 

### 2015-04-23 Rmarkdown/comethylation.Rmd
Begin simulation section.

@@
 So far we have assumed a common odds ratio for the $K$ $2 \times 2$ tables.
-
-__UP TO HERE__
-
-- Move to $2 \times 2 \times K$ tables
-  - Go through same comparison of estimators
-- Move to sparseness
-  - Asymptopia I and II
-  - Go through same comparison of estimators
-  - The homogeneity assumption
-
- ## Setting for studying within-fragment co-methylation
+This is also known as the _homogeneity of odds ratios_ assumption.
+As noted by \cite{Liang:1985uw}, \citet{MANTEL:1959uf} did not explicitly assume homogeneity of the odds ratio in their original work.
+Nonetheless, it does raise the question of how to test this assumption and the effect on these estimators when this assumption is not true.
+
+Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse, \citet{Liang:1985uw} developed three tests for the assumption of homogeneity of odds ratios and compared these with two tests for the same assumption but designed under $A1$.
+Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$.
+The disadvantage of these conditional tests is the extra computation required to compute them.
+\citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i \cdot} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.
+
+\subsection{Simulation study}
+
+We analyse the results of a simulation study to explore the effects of sparsity and heterogeneity of odds ratios on estimates of odds ratios in $2 \times 2 \times K$ contingency tables.
+The simulation requires that we specify several parameters shown below.
+In order to relate these back to DNA methylation, I also describe each parameter using the statistical framework proposed in Chapter \ref{chap:wgbs_statistical_framework}:
+
+- the number of tables, $K$. The number of 2-tuples.
+- the marginal probabilities, $\pi_{1 \cdot} = Pr(Z_{i})$ and $\pi_{\cdot 1} = Pr(Z_{i'})$.
+Some simulations use a common set of marginal probabilities across all tables but some simulate the marginal probabilities from a specified distribution.
+- the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
+- the sample size of each table, $n_{\cdot \cdot k} = d_{(i, i')}$.
+The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth).
+Most simulations use a truncated negative binomial distribution to model a bisulfite-sequencing experiment with mean sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].
+
+[^nb_sim]: In the simulation study, a larger cutoff results in higher quality data since we can simply keep simulating until we have $K$ tables with sufficient sequencing depth.
+In analysing real data, increasing the cutoff may be detrimental since we could end up with fewer than $K$ tables to analyse.
+
+__UP TO HERE: Insert simulation results__
+
+- Use parameters $n, \pi, \psi, \phi$ and not the ones from Chapter \ref{chap:wgbs_statistical_framework} since these are more succinct.
+
+\subsection{Summary}
+
+__TODO__
+
+\section{Within-fragment co-methylation}\label{sec:wf_cometh}
+
+The key advantage of studying within-fragment co-methylation is that it measures the dependence of DNA methylation at the scale on which the biological process acts.
+I have developed a method based on odds-ratios to quantify the strength of co-methylation for a pair of methylation loci in a single sample.
+
+\subsection{Studying within-fragment co-methylation from sparse $2 \times 2 \times K$ contingency tables}
 
@@
 
- ## Simulation
-
--
-
-## Summary
-
-\section{Within-fragment co-methylation}\label{subsec:wf_cometh}
-
-The key advantage of studying within-fragment co-methylation is that it measures the dependence of DNA methylation at the scale on which the biological process acts.
-I have developed a method based on odds-ratios to quantify the strength of co-methylation for a pair of methylation loci in a single sample.
+- Describe co-methylation in terms of $X$ and $Y$, the sparse $2 \times 2 \times K$ contingency tables, the marginal probabilities, and the odds ratios.
+
+We will be studying within-fragment co-methylation by analysing a large number of sparse $2 \times 2$ contingency tables. This puts us in the realm of asymptotic environment $A2$.
+Furthermore, the marginal probabilities of these tables, $\pi_{\cdot 1} = 1 - \pi_{\cdot 2}$ and $\pi_{1 \cdot} = 1 - \pi_{2 \cdot}$, will be very often close to the boundaries of zero and one.
+This is because in this case the marginal probability represents the probability that the methylation locus is methylated, and we have seen in Chapter \ref{chap:wgbs_statistical_framework} that these values, as estimated by $\beta$-values, have a bimodal distribution with each mode near the boundary.
+
+These extreme marginal probabilities give rise to sparse $2 \times 2$ tables.
+A property of these sparse $2 \times 2$ tables not yet discussed is that estimates of the odds ratio for these tables are very biased.
+We have seen this in __SIMULATION SETION__, but the intuitive explanation is that if both methylation loci are very likely to be methylated or unmethylated __BLAH BLAH BLAH__
 
@@
 
-# rtpois is a helper function to simulate n truncated Poisson(lambda) with
+# rtpois is a helper function to simulate n truncated Poisson(lambda) RVs with
 # truncation specified by 'trunc'.
@@
   qpois(runif(n, ppois(trunc, lambda), 1), lambda)
+}
+
+# rtnbinom is a helper function to simulate n truncated NB(mu, size) RVs with
+# truncation specified by 'trunc'. It uses a more explicit approach to simulate # truncated RVs than rtpois() because the same trick doesn't immediately work
+# for NB-distributed RVs.
+rtnbinom <- function(n, mu, size, trunc) {
+  x <- rnbinom(n, size = size, mu = mu)
+  i <- x <= trunc
+  while (any(i)) {
+    x[i] <- rnbinom(sum(i), size = size, mu = mu)
+    i <- x <= trunc
+  }
+  x
 }

### 2015-04-23 Rmarkdown/comethylation.Rmd
Switch summation notation. Move code from thesis to supplementary material.

@@
 The possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios.
-Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\phi = \log{\psi}$.
+Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$.
 The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler.
-I will routinely switch between discussing the estimation of $\psi$ and $\phi$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.
+I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.
 
@@
 Denote by $n_{ij}, i, j = 1, 2$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$.
-We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 \cdot} = n_{11} + n_{12}$.
+We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$.
 The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.
@@
 \multirow{3}{*}{Row} & & & \\
- & $n_{11}$       & $n_{12}$       & $n_{1 \cdot}$ \\
- & $n_{21}$       & $n_{22}$       & $n_{2 \cdot}$ \\
- & $n_{\cdot 1}$  & $n_{\cdot 2}$  & $n_{\cdot \cdot} = n$ \\ \cmidrule(l){2-4}
+ & $n_{11}$       & $n_{12}$       & $n_{1 +}$ \\
+ & $n_{21}$       & $n_{22}$       & $n_{2 +}$ \\
+ & $n_{+ 1}$  & $n_{+ 2}$  & $n_{+ +} = n$ \\ \cmidrule(l){2-4}
 \end{tabular}
@@
 
-The simplest estimator of $\phi$ is the unconditional maximum likelihood estimator, $\widehat{\phi_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$.
-The asymptotic standard error of $\widehat{\phi_{U}}$ is given by $\hat{\sigma}(\widehat{\phi_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
-The asymptotic distribution of $\widehat{\phi_{U}}$ as $n \to \infty$ is $Normal(\phi, \hat{\sigma}(\phi)^{2})$.
-
-A problem with the estimator $\widehat{\phi_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities.
-We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\phi_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\phi_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
-\citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\phi_{U}}$ and $\widehat{\phi_{0.5}}$ have the same asymptotic distribution around $\phi$ as $n \to \infty$ but that $\widehat{\phi_{0.5}}$ has reduced first-order bias.
-The modified estimator, $\widehat{\phi_{0.5}}$, is therefore generally recommended.
-
-Another alternative to $\widehat{\phi_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\phi_{C}}$.
-This is computed by first conditioning on ${n_{1 \cdot}, n_{\cdot 1}}$.
-Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 \cdot}, n_{\cdot, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}.
+The simplest estimator of $\theta$ is the unconditional maximum likelihood estimator, $\widehat{\theta_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$.
+The asymptotic standard error of $\widehat{\theta_{U}}$ is given by $\hat{\sigma}(\widehat{\theta_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
+The asymptotic distribution of $\widehat{\theta_{U}}$ as $n \to \infty$ is $Normal(\theta, \hat{\sigma}(\theta)^{2})$.
+
+A problem with the estimator $\widehat{\theta_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities.
+We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\theta_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\theta_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
+\citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\theta_{U}}$ and $\widehat{\theta_{0.5}}$ have the same asymptotic distribution around $\theta$ as $n \to \infty$ but that $\widehat{\theta_{0.5}}$ has reduced first-order bias.
+The modified estimator, $\widehat{\theta_{0.5}}$, is therefore generally recommended.
+
+Another alternative to $\widehat{\theta_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$.
+This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$.
+Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}.
 The maximum likelihood estimator of $\psi$, $\widehat{psi_{C}}$ is solved using iterative methods.
 
-The conditional estimator, $\widehat{\phi_{C}}$, works better than the unconditional estimator, $\widehat{\phi_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+The conditional estimator, $\widehat{\theta_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
 Estimation of the odds ratio is made more difficult when the contingency table is _sparse_.
@@
 We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum.
-For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{\cdot 1k} = 1 - \pi_{\cdot 2k}, \pi_{1 \cdot k} = 1 - \pi_{2 \cdot k}$ to vary.
-
-We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\phi = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
-
-Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{\cdot \cdot k})}{n_{12k} n_{21k} / n_{\cdot \cdot k}}$.
-\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\phi_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\phi_{MH}})$.
+For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{+ 1k} = 1 - \pi_{+ 2k}, \pi_{1 + k} = 1 - \pi_{2 + k}$ to vary.
+
+We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
+
+__TODO: Is the unconditional maximum likelihood estimator the same as `LORCollapsed()`?__
+
+Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$.
+\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\theta_{MH}})$.
 This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance].
@@
 
-The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{\cdot \cdot k}$, grows large.
-The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{\cdot \cdot k}$ remains small.
+The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{+ + k}$, grows large.
+The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{+ + k}$ remains small.
 The variance estimator proposed by \citet{Robins:1986vh} is consistent under both $A1$ and $A2$.
@@
 The disadvantage of these conditional tests is the extra computation required to compute them.
-\citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i \cdot} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.
+\citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.
+
+- __TODO: Comment on the utility (or lack thereof) of homogeneity tests; i.e., rejecting the null is good evidence the null is wrong but failing to reject the null is not strong evidence that it is not wrong.__
+
+- __TODO: See how the variance estimator behaves in some of the more pathological simulations__
 
@@
 - the number of tables, $K$. The number of 2-tuples.
-- the marginal probabilities, $\pi_{1 \cdot} = Pr(Z_{i})$ and $\pi_{\cdot 1} = Pr(Z_{i'})$.
+- the marginal probabilities, $\pi_{1 +} = Pr(Z_{i})$ and $\pi_{+ 1} = Pr(Z_{i'})$.
 Some simulations use a common set of marginal probabilities across all tables but some simulate the marginal probabilities from a specified distribution.
 - the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
-- the sample size of each table, $n_{\cdot \cdot k} = d_{(i, i')}$.
+- the sample size of each table, $n_{+ + k} = d_{(i, i')}$.
 The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth).
@@
 
-- Use parameters $n, \pi, \psi, \phi$ and not the ones from Chapter \ref{chap:wgbs_statistical_framework} since these are more succinct.
+- Use parameters $n, \pi, \psi, \theta$ and not the ones from Chapter \ref{chap:wgbs_statistical_framework} since these are more succinct.
 
@@
 We will be studying within-fragment co-methylation by analysing a large number of sparse $2 \times 2$ contingency tables. This puts us in the realm of asymptotic environment $A2$.
-Furthermore, the marginal probabilities of these tables, $\pi_{\cdot 1} = 1 - \pi_{\cdot 2}$ and $\pi_{1 \cdot} = 1 - \pi_{2 \cdot}$, will be very often close to the boundaries of zero and one.
+Furthermore, the marginal probabilities of these tables, $\pi_{+ 1} = 1 - \pi_{+ 2}$ and $\pi_{1 +} = 1 - \pi_{2 +}$, will be very often close to the boundaries of zero and one.
 This is because in this case the marginal probability represents the probability that the methylation locus is methylated, and we have seen in Chapter \ref{chap:wgbs_statistical_framework} that these values, as estimated by $\beta$-values, have a bimodal distribution with each mode near the boundary.
@@
 
-__TODO: Move this code chunk to supplementary material and only include results in thesis.__
-
-```{r}
-# Requires the installation of methsim
-library(mipfp)
-
-# rtpois is a helper function to simulate n truncated Poisson(lambda) RVs with
-# truncation specified by 'trunc'.
-# It is adapted from a method for simulating zero-truncated Poisson suggested
-# by Peter Dalgaard https://stat.ethz.ch/pipermail/r-help/2005-May/070680.html
-rtpois <- function(n, lambda, trunc) {
-  qpois(runif(n, ppois(trunc, lambda), 1), lambda)
-}
-
-# rtnbinom is a helper function to simulate n truncated NB(mu, size) RVs with
-# truncation specified by 'trunc'. It uses a more explicit approach to simulate # truncated RVs than rtpois() because the same trick doesn't immediately work
-# for NB-distributed RVs.
-rtnbinom <- function(n, mu, size, trunc) {
-  x <- rnbinom(n, size = size, mu = mu)
-  i <- x <= trunc
-  while (any(i)) {
-    x[i] <- rnbinom(sum(i), size = size, mu = mu)
-    i <- x <= trunc
-  }
-  x
-}
-
-# Function to simulate 2x2xK tables.
-# Each 2x2 table represents the counts of unmethylated (0) and methylated (1) states
-# at each CpG for each read.
-# The number of reads per table (i.e., 2x2 table sample size) is simulated
-# from a truncated Poisson(lambda) distribution; truncation is to simulate
-# minimum coverage cutoffs and is specified by the 'trunc' argument.
-# The marginal probabilities of each table are p1 = Prob(methylated at first CpG)
-# and p2 = Prob(methylated at second CpG).
-# The odds ratio (or) is for each 2x2 table and should be of length K or 1 (in
-# which case all K tables have the same odds ratio).
-simTables <- function(K, lambda, trunc, p1, p2, or) {
-
-  # d is the sample size of each 2x2 table (i.e, the sequencing depth for the 2-tuple)
-  d <- rtpois(n = K, lambda = lambda, trunc = trunc)
-  if (length(p1) == 1L) {
-    p1 <- rep(p1, K)
-  }
-  if (length(p2) == 1L) {
-    p2 <- rep(p2, K)
-  }
-  if (length(or) == 1L) {
-    or <- rep(or, K)
-  }
-  .simTables <- function(k, d, p1, p2, or) {
-    # Compute joint probabilities
-    P <- mipfp::Ipfp(seed = matrix(c(or, 1, 1, 1), ncol = 2),
-                     target.list = list(1, 2),
-                     target.data = list(c(1 - p1, p1), c(1 - p2, p2)))$x.hat
-    matrix(rmultinom(1, d, as.vector(P)), ncol = 2)
-  }
-  mapply(.simTables, k = K, d = d, p1 = p1, p2 = p2, or = or, SIMPLIFY = "array")
-}
-
-# Estimator of the (offset-adjusted) unconditional log odds ratios of K 2x2 tables
-LOR <- function(x, offset = 0.5) {
-  x <- x + offset
-  MM <- x[2, 2, ]
-  MU <- x[2, 1, ]
-  UM <- x[1, 2, ]
-  UU <- x[1, 1, ]
-  log2(MM) + log2(UU) - log2(MU) - log2(UM)
-}
-
-# Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
-MH <- function(x, offset = 0) {
-  x <- x + offset
-  # TODO: return the standard error
-  MM <- x[2, 2, ]
-  MU <- x[2, 1, ]
-  UM <- x[1, 2, ]
-  UU <- x[1, 1, ]
-  d <- apply(x, 3, sum)
-  or <- sum((UU * MM) / d) / sum((MU * UM) / d)
-  # One or more margins must be identically zero so add offset
-  if (is.nan(or) | isTRUE(all.equal(or, 0)) | is.infinite(or)) {
-    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
-    MH(x, 0.5)
-  } else {
-    log2(or)
-  }
-}
-
-# Estimator of the (offset-adjusted) unconditional log odds ratio of the
-# 2x2xK collapsed over K.
-LORCollapsed <- function(x, offset = 0.5) {
-  z <- apply(x, c(1, 2), sum)
-  # Artificially augment so it's a 2x2x1 table
-  dim(z) <- c(dim(z), 1)
-  lor <- LOR(z, 0)
-  if (is.nan(lor) | is.infinite(lor)) {
-    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
-    LOR(z, 0.5)
-  } else {
-    lor
-  }
-}
-
-# Simulate and plot data when p1 = p2.
-# 100 CpG pairs, 30x average coverage truncated at 10
-# odds ratio = 1, i.e., independence
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 100
-lambda <- 20
-trunc <- 11
-or <- 1
-p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
-p2 <- p1
-for (i in 1:9) {
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i], p2 = p2[i], or = or)
-  hist(LOR(x), main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  abline(v = log2(or), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 = p2.
-# 1000 CpG pairs, 30x average coverage truncated at 10
-# odds ratio = 1, i.e., independence
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 1000
-lambda <- 20
-trunc <- 10
-or <- 1
-p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
-p2 <- p1
-for (i in 1:9) {
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i],
-                 p2 = p2[i], or = or)
-  hist(LOR(x),
-       main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  abline(v = log2(or), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# Simulate and plot data when p1, p2 ~ Uniform(0, 1)
-# 100 CpG pairs, 30x average coverage truncated at 10
-# log odds ratio = -4, -3, ..., 3, 4.
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 100
-lambda <- 20
-trunc <- 11
-or <- 2^seq(-4, 4, 1)
-p1 <- runif(K)
-p2 <- runif(K)
-for (i in 1:9) {
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2,
-                 or = or[i])
-  hist(LOR(x),
-       main = paste0("p1 = U(0, 1), p2 = U(0, 1)", ", lor = ", log2(or[i])),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  abline(v = log2(or[i]), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc),
-      outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 ~ Uniform(0, 1), p2 = p1
-# 100 CpG pairs, 30x average coverage truncated at 10
-# log odds ratio = -4, -3, ..., 3, 4.
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 100
-lambda <- 20
-trunc <- 11
-or <- 2^seq(-4, 4, 1)
-p1 <- runif(K)
-p2 <- p1
-for (i in 1:9) {
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
-  hist(LOR(x), main = paste0("p1 = U(0, 1), p2 = p1", ", lor = ", log2(or[i])),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       frequeny = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  abline(v = log2(or[i]), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc),
-      outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
-# 100 CpG pairs, 30x average coverage truncated at 10
-# log odds ratio = -4, -3, ..., 3, 4.
-# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
-# levels (B).
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 100
-lambda <- 20
-trunc <- 11
-or <- 2^seq(-4, 4, 1)
-p1 <- rbeta(K, 0.3, 0.2)
-p2 <- p1
-for (i in 1:9) {
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
-  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = ", log2(or[i])),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  abline(v = log2(or[i]), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
-# 100 CpG pairs, 30x average coverage truncated at 10
-# log2(or) ~ Gaussian(log2(or_mean), 1)
-# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
-# levels (B).
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 100
-lambda <- 20
-trunc <- 11
-or_mean <- 2 ^ seq(-4, 4, 1)
-p1 <- rbeta(K, 0.3, 0.2)
-p2 <- p1
-for (i in 1:9) {
-  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
-  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  lines(density(log2(or)), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
-# 1000 CpG pairs, 30x average coverage truncated at 10
-# log2(or) ~ Gaussian(log2(or_mean), 1)
-# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
-# levels (B).
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 1000
-lambda <- 20
-trunc <- 11
-or_mean <- 2 ^ seq(-4, 4, 1)
-p1 <- rbeta(K, 0.3, 0.2)
-p2 <- p1
-for (i in 1:9) {
-  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
-  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  lines(density(log2(or)), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
-# 1000 CpG pairs, 30x average coverage truncated at 10
-# log2(or) ~ Skewed-Gaussian(log2(or_mean), 1, -10)
-# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
-# levels (B).
-library(sn)
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 1000
-lambda <- 20
-trunc <- 11
-or_mean <- 2 ^ seq(-4, 4, 1)
-p1 <- rbeta(K, 0.3, 0.2)
-p2 <- p1
-for (i in 1:9) {
-  or <- 2 ^ rsn(K, log2(or_mean[i]), 1, -10)
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
-  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = SN(", log2(or_mean[i]), ", 1, -10)"),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  lines(density(log2(or)), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
-# 1000 CpG pairs, 30x average coverage truncated at 10
-# log2(or) ~ 0.5 * Gaussian(log2(or_mean), 1) + 0.5 * Gaussian(-log2(or_mean))
-# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
-# levels (B).
-set.seed(666)
-par(mfrow = c(3, 3))
-K <- 1000
-lambda <- 20
-trunc <- 11
-or_mean <- 2 ^ seq(-4, 4, 1)
-w <- sample(c(-1, 1), K, replace = TRUE)
-p1 <- rbeta(K, 0.3, 0.2)
-p2 <- p1
-for (i in 1:9) {
-  or <- 2 ^ rnorm(K, w * log2(or_mean[i]), 1)
-  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
-  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = NM(", log2(or_mean[i]), ", 1)"),
-       xlim = c(-10, 10), breaks = 20,
-       xlab = expression(paste(log[2], "(odds ratio)")),
-       freq = FALSE)
-  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
-  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
-  lines(density(log2(or)), col = "black", lty = 2)
-}
-title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
-
-# TODO: Increase d for each table
-```
-
 ## Following chat with Terry
@@
 - [ ] What if I just increase the fudge-factor to increase the shrinkage/smoothing of the estimates? It'll shrink to zero alright, but what is the theoretical justification?
-- [ ] Check out the behaviour of the $\phi$ coefficient, i.e., Pearson's $r$ of the $2 \times 2$ table.
+- [ ] Check out the behaviour of the $\theta$ coefficient, i.e., Pearson's $r$ of the $2 \times 2$ table.
   - Landan et al.: "Note that correlation values are subject to considerable noise when coverage is low or when m1 or m2 are close to 0 or 1."

### 2015-04-24 Rmarkdown/biology_background.Rmd
Pre-lunch commit. Begin re-write of methsim chapter.

@@
 ---
-title: "Biology background"
+title: "Introduction to genomics and DNA methylation"
 output:
@@
 I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
-\begin{end}
+\end{chapabstract}
 

### 2015-04-24 Rmarkdown/methsim.Rmd
Pre-lunch commit. Begin re-write of methsim chapter.

@@
 
-\section{Chapter overview}\label{sec:methsim_overview}
-
-In this chapter I argue for the utility, even necessity, of methodology and software for simulating DNA methylation data.
+\begin{chapabstract}
 I present the `methsim` software to simulate DNA methylation data.
@@
 Finally, I provide examples of how `methsim` can be used to learn about DNA methylation and how it might be used to develop simulation studies to compare competing analysis methods.
+\end{chapabstract}
 
@@
 In fact, the huge variety of experimental factors, and their possible values, can make such a task infeasible.
-But perhaps the bigger hurdle is that there are not attractive experiments to perform; why spend your time and money on an experiment where you "know the answer" when you could be spending that same time and money on investigating something new?
-In these scenarios and more, simulation studies play a vital role in applied statistics, where they can be used in the development, validation and benchmarking of different analysis methodologies.
+But perhaps the bigger hurdle is that there are not attractive experiments to perform; why spend your time and money on an experiment where you 'know the answer' when you could be spending that same time and money on investigating some new biology?
+In these scenarios and others, simulation studies play a vital role in applied statistics, where they can be used in the development, validation and benchmarking of different analysis methodologies.
 
@@
 Simulation studies can provide many insights into the performance of a method.
-At the most basic level, if a method fails to work or performs poorly when applied to simulated data, then it is very unlikely to work well when applied to real data.
-We can also identify which scenarios are "easy", ones where most methods are able to identify the truth, and scenarios under which certain methods perform better than others.
-While simulated data can never fully capture the richness of "real" data, we can learn a lot about a method by studying how it performs when applied to simulated data.
+At the most basic level, if a method fails performs poorly when applied to 'easy' simulated data, then it is very unlikely to work well when applied to more complex real data.
+We can also identify which scenarios are 'easy' (ones where most methods are able to identify the truth) and scenarios under which certain methods perform better than others.
+While simulated data can never fully capture the richness of real data, we can learn a lot about a method by studying how it performs when applied to simulated data.
 
@@
 
-1. Realism: The simulated data must be "similar" to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes "similar enough".
+1. Realism: The simulated data must be 'similar' to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes 'similar enough'.
 2.
@@
 For the input dataset, `methsim` requires the methylation patterns at various sized m-tuples, which can be produced by the `methtuple` software.
-
-This workflow is illustrated in __FIGURE__ and described in the remainder of this section.
+`methsim` can currently only simulate CpG methylation and assumes that the methylation states of all CpGs are strand symmetric.
 
@@
 
-`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as several R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list.
-__TODO: Add citations of all packages used in thesis__).
+`methsim` is written in R and builds on the `MethylationTuples` package described in Chapter \ref{chap:wgbs_downstream_analyses}, as well as several R packages available on Bioconductor and CRAN (see the `DESCRIPTION` file for a complete list.).
 
 `methsim` makes extensive use of the S4 object system in R.
-The most important classes defined by `methsim` are `SimulateMethylomeParam`, `SimulatedMethylome`, `SimulateBSParam` and `SimulatedBS`.
-The most important methods are the `simulate()` methods defined for the `SimulateMethylomeParam` and `SimulateBSParam` classes.
-The role of these classes and methods in the workflow is illustrated in __FIGURE__.
+The most important classes defined by `methsim` are `MethylomeParam`, `SimulatedMethylome`, `WGBSParam` and `SimulatedBS`.
+The most important methods are the `simulate()` methods defined for the `MethylomeParam` and `WGBSParam` classes.
 
-The `SimulateMethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled.
-Therefore, the `SimulateMethylomeParam` should be based on a relevant sample.
-To help a new user get started, `methsim` includes `SimulateMethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset.
-Alternatively, the user may process the BAM file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `SimulateMethylomeParam` object based on their sample of interest.
+The `MethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled.
+Therefore, the `MethylomeParam` should be based on data from a relevant sample.
+To help a new user get started, `methsim` includes `MethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset.
+Alternatively, the user may process the `BAM` file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `MethylomeParam` object based on their sample of interest.
 
@@
 
-To simulate bisulfite-sequencing reads, the user first constructs a `SimulateBSParam`, which includes the `SimulatedMethylome`.
-The user then runs the `simulate()` method on the `SimulatedBSParam` object to create a simulated bisulfite-sequencing data in the form of a `SimulatedBS` object.
+Once we have a 'true' methylome, we can simulate data from it.
+Currently, `methsim` supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method.
+Other assays, such as RRBS or microarrays may be supported in future releases.
+A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated.
+When applied to a `WGBSParam` object, which includes a `SimulatedMethylome` object, the `simulate()` method can return a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2].
+A `SimulatedBS` object contains all simulated reads and is generally much larger than a `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat].
+`methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.
 
-Finally, we may wish to simplify the `SimulatedBS` object.
-For example, we may only require the $\beta$-values for our downstream analysis and not the  read-level data.
-`methsim` includes the `asMethPat()` function to convert the `SimulatedBS` object to a `MethPat` object (described in __CHAPTER__).
+[^namespace2]: This uses the NAMESPACE notation of R: `MethylationTuples::MethPat` can be read as "the `MethPat` class is part of the `MethylationTuples` package".
+See Section \ref{sec:MethylationTuples} for details of this class.
 
-__UP TO HERE: See if the remainder of this section needs re-jigging in light of the above changes to Implementation__
+[^methpat]: Returning a `MethylationTuples::MethPat` object may be appropriate, for example, if all that is required for downstream analyses are methylation counts at 1-tuples.
 
@@
 
-- [x] DONE
+- [ ] DONE
 
@@
 In these locally-similar regions we might hope to model DNA methylation by a simple parametric model.
-`methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together to form the 'true' methylome.
+`methsim` is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.
 
@@
 1. Segment the methylome into regions of similarity.
-2. Within each region simulate the binary methylation states from a simple parametric model.
-3. "Stitch" the resulting region-level simulations together[^stitch].
+2. Sample parameters for each methylation locus, $i$, based on step 1.
 
-[^stitch]: In practice, the stitching together of segments is incorporated into step 2.
-
-Suppose our sample contains $n_{h}$ haplotypes and, therefore, that our sample can be represented by[^strand_symmetric] $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$.
-We might try to simulate this sample by generating $n_{h}$ independent realisations of $\mathbf{Z_{h}}$.
-However, for a number of reasons, this is a little naive and ambitious.
-For example, under this naive model there is no control across haplotypes of the methylation patterns; although there exists between-haplotype heterogeneity, the $\mathbf{Z_{h}}$ are definitely not independent of one another.
-
-[^strand_symmetric]: Currently, `methsim` can only simulate CpG methylation and assumes that the methylation states of all CpGs are strand symmetric.
-
-Rather than try to explicitly model dependencies between the $\mathbf{Z_{h}}$, `methsim` uses the following heuristic procedure:
-
-1.
-Simulate $n_{\tilde{h}} \ll n_{h}$ independent realisations of $\mathbf{{Z_{\tilde{h}}}}$, where $\mathbf{Z_{\tilde{h}}}$ has the same stochastic properties as $\mathbf{Z_{h}}$.
-`methsim` uses a default value of $n_{\tilde{h}} = 8$.
-2.
-Within each region, $r_{i}$, assign each of $\mathbf{Z_{\tilde{h}}}$ a weight, $W_{r_{i}, \tilde{h}}$, where $\sum_{ \tilde{h} = 1}^{n_{ \tilde{h}}} W_{r_{i}, \tilde{h}} = 1$ and, by construction, $W_{r_{i}, 1} \geq \ldots \geq W_{r_{i}, n_{\tilde{h}}}$.
-
-For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (discussed below; __TODO: Discuss implications of this assumption somewhere below__).
-
-The weights are interpreted as the proportion of $\{ Z_{h, i} \}_{h = 1}^{h = n_{h}}$ with the methylation pattern given by $Z_{\tilde{h}, i}$ for the $i^{th}$ methylation locus[^weights].
-We write the matrix of weights for all methylation loci in the genome as $\mathbf{W} = \{ \mathbf{W_{r_{i}, \cdot}} \}_{i = 1}^{N_{loci}} = \{ \mathbf{W_{\cdot, \tilde{h}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$.
-In regions of low within-sample heterogeneity, such as those unmethylated regions with $B_{i} \approx 0$, this means we want $\mathbf{W_{r_{i}, \cdot}} \approx (1, 0, \ldots, 0)$, whereas in regions of high within-sample heterogeneity, such as partially methylated regions, we want a more dispersed $\mathbf{W_{r_{i}}, \cdot}$.
-
-[^weights]: Note that these weights are defined at the region-level so all loci in the region have the same weights.
-
-A simulated methylome is therefore described by $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$ and $\mathbf{W}$.
-For example, $B_{i} = \sum_{\tilde{h} = 1}^{n_{\tilde{h}}} W_{r_{i}, \tilde{h}} \times Z_{i, \tilde{h}}$.
-
-I now describe each step in this process in greater detail.
+I now describe each step in greater detail.
 
@@
 
-`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the input methylome into these regions of similarity.
-Initially published to discover regulatory motifs from bisulfite-sequencing data, `MethylSeekR` segments the genome into unmethylated regions (\emph{UMR}s), lowly-methylated regions (\emph{LMR}s) and partially methylated regions[^pmds] (\emph{PMR}s).
+`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the input methylome into regions of similarity.
+`MethylSeekR` was developed to discover regulatory motifs from bisulfite-sequencing data by segmenting the methylome into unmethylated regions (\emph{UMR}s), lowly-methylated regions (\emph{LMR}s) and partially methylated regions[^pmds] (\emph{PMR}s).
 It does this using a two-stage algorithm applied to the $\beta$-values from a sample:
@@
 
-`methsim` uses the functionality of `MethylSeekR` to partition the genome into "regions of similarity", namely `UMR`, `LMR`, `PMR` or `other`[^other].
-While a tailored algorithm may improve the segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting into "regions of similarity".
-This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)__).
-It should be noted that `methsim` also performs some post-processing of the `MethylSeekR` output so that the resulting segmentation is disjoint and exhaustive.
+`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarly, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output].
+Roughly speaking, _other_ regions are 'mostly methylated regions' (see Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Ziller_beta_by_pm_boxplots}), although \citet{Burger:2013kq} do not describe these regions as such.
 
-[^other]: `other` roughly corresponds to 'mostly methylated regions', although \citet{Burger:2013kq} do not describe these regions as such.
+[^methylseekr_output]: The output returned by `MethylSeekR` does not strictly partition the methylome since it is neither disjoint nor exhaustive.
 
-__TODO: Figure showing `other` is MMRs. More generally, a figure showing the average methylation level in each of `UMR`, `LMR`, `PMR` and `other`.__
+While a tailored algorithm may improve this segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting the methylome into the required 'regions of similarity'.
 
-\subsubsection{Simulating $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$}
+\subsubsection{Parameterising the model}
 
-- [x] DONE
-
-`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{{Z_{\tilde{h}}}}$.
+`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$.
 This allows the incorporation of within-fragment co-methylation into the simulation.
@@
 
-For these reasons I allow the transition probabilities to vary with $i$ and $r_{i}$, that is, I allow the Markov chain to be spatially inhomogeneous.
-In particular, I allow the transition probabilities, $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} = b | Z_{h, i, j} = a)$, to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region for this pair of loci[^context], $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} | Z_{h, i, j}, r_{i}, IPD(i, i + 1))$, where $r_{i}$ is the region type (UMR, LMR, PMR or MMR) of the $i^{th}$ methylation locus.
+For these reasons I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous.
+In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for this pair of loci[^context].
+That is, $p_{i; a, b} = Pr(Z_{h, i + 1} = b | Z_{h, i} = a, IPD(i, i + 1), r_{i})$.
 
@@
 
-The above-described model, being spatially inhomogeneous, is not particularly amenable to analytical calculations.
-However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome.
-For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h)}}$, to estimate or otherwise assign.
-Again, I emphasise that `methsim` uses the same set of parameters for all $n_{\tilde{h}}$ "pseudo-haplotypes".
-Therefore, only one set of transition matrices need to be assigned.
+The above-described model is not particularly amenable to analytical calculations due to the spatial inhomogeneity.
+It is, however, relatively simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome.
+For a chromosome containing $n$ methylation loci, there are $n - 1$ transition matrices to estimate or otherwise assign.
 
-`methsim` assigns these transition probabilities based on $\beta_{i | r_{i}}$ and $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$, the empirical distributions of $\beta_{i}$ and $LOR_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
-These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots}.
+Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised $\mathbf{Z}$ by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$.
+We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$ as I now describe.
+
+For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}, B_{i + 1}$, and the odds ratio, $\psi_{i}.
+The general form of $P_{i, i + 1}$ is shown below:
+
+\begin{equation*}
+P_{i, i + 1} = \bordermatrix{~ & 1 - B_{i+1} & B_{i+1} \cr
+1 - B_{i} & Pr(Z_{i} = 0, Z_{i + 1} = 0) & Pr(Z_{i} = 0, Z_{i + 1} = 1) \cr
+B_{i} & Pr(Z_{i} = 1, Z_{i + 1} = 0) & Pr(Z_{i} = 1, Z_{i + 1} = 1) \cr}
+\end{equation*}
+
+`methsim` computes $P_{i, i + 1}$ using the iterative proportional fitting algorithm.
+Iterative proportional fitting is a general method "for constructing tables of numbers satisfying certain constraints" \citep{Speed:2005tg}.
+In the case of `methsim`, we use iterative proportional fitting to construct the _unique_ $2 \times 2$ array (the joint probability matrix) with specified margins (the marginal probabilities of each methylation state at each locus) and the specified cross-ratio (the odds-ratio).
+An example of this procedure is illustrative.
+
+Let $B_{i} = 0.7$, $B_{i + 1} = 0.6$ and $\psi_{i, i + 1} = 2$. To begin the iterative proportional fitting algorithm, form the matrix $P_{i, i + 1}^{(0)} = \Bigl(\begin{smallmatrix}
+\psi & 1 \\
+   1 & 1
+\end{smallmatrix} \Bigr)$.
+This matrix has the desired cross-ratio, $\psi$, but not the desired row and column margins.
+At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi$ while the row and column margins converge towards the desired values.
+The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P_{i, i + 1}^{(*)} \equiv P_{i, i + 1}$.
+For our example, $P_{i, i + 1}^{(*)} = \Bigl(\begin{smallmatrix}
+0.155 & 0.145 \\
+0.245 & 0.455
+\end{smallmatrix} \Bigr)$, to three decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.
+
+Once $P_{i, i + 1}$ is computed, the desired transition probability is obtained by dividing the appropriate element of $P_{i, i + 1}$ by the appropriate marginal probability.
+To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also to zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ to three decimal places.
+In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.
+
+To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimators, $\mathbf{\beta} and $\mathbf{\widehat{\psi}}$, in the input methylome.
+More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ and $\mathbf{\psi}$ is based on $\widehat{\psi_{IPD(i, i+1) | r_{i}}$, the empirical distributions of $\beta_{i}$ and $\psi_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
+These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and __FIGURE__.
 
@@
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/ADS_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{ADS} sample.
-Only 2-tuples with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:ADS_lor_by_pm_quantile_plots}
-\end{figure}
+__TODO: Figure of ADS within-fragment co-methylation__
 
-These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
-\ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
-\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots}.
+These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in __FIGURES__.
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{EPISCOPE} dataset.
-Only CpGs with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:EPISCOPE_beta_by_pm_boxplots}
-\end{figure}
+There are several ways to simulate values of $\mathbf{B}$ and $\mathbf{\psi}$ based on these empirical distributions.
+The simplest is to sample each $B_{i}$ and $\psi_{i}$ directly from these empirical distributions.
+Alternatively, we could take a more model-based approach.
+For example, we could specify a  parametric model of the $B_{i}$ where all $i$ in a region have the same mean and variance (the mean and variance being based on the empirical distribution of $\beta$-values) or where the $B_{i}$ in each region come from an autoregressive process.
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{EPISCOPE} sample.
-Only 2-tuples with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:EPISCOPE_lor_by_pm_quantile_plots}
-\end{figure}
+The simulation of $\mathbf{\psi}$ requires further explanation.
+The empirical distributions of $\mathbf{\widehat{\psi}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
+When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?
+Again, there are several options.
+We might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\psi = 1$ (corresponding to independence).
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Lister} dataset.
-Only CpGs with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Lister_beta_by_pm_boxplots}
-\end{figure}
+I explore different methods for simulating $\mathbf{B}$ and $\mathbf{\psi}$ in Section \ref{sec:results}.
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Lister} sample.
-Only 2-tuples with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Lister_lor_by_pm_quantile_plots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Seisenberger} dataset.
-Only CpGs with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Seisenberger_beta_by_pm_boxplots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Seisenberger} sample.
-Only 2-tuples with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Seisenberger_lor_by_pm_quantile_plots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Ziller} dataset.
-Only CpGs with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Ziller_merged_beta_by_pm_boxplots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_by_pm_quantile_plots.pdf}
-\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Ziller} sample.
-Only 2-tuples with at least $10\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Ziller_merged_lor_by_pm_quantile_plots}
-\end{figure}
-
-`methsim` assigns these probabilities as follows:
-
-1.
-For each region, $r_{i}$, sample $B_{r_{i}}$ from $\beta_{i | r_{i}}$.
-This can be interpreted as $Pr(Z_{r_{i}} = 1)$, the marginal probability that a methylation locus in region $r_{i}$ is methylated, or, alternatively, as the average level of methylation for all methylation loci in the region[^extreme_B].
-2.
-For every pair of methylation loci, $(i, i + 1)$, $1 \leq i \leq (n - 1)$, sample $\Psi_{(i, i + 1)}$ from $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$.
-This can be interpreted as the odds that the $(i + 1)^{th}$ methylation locus has the same methylation state as the $i^{th}$ methylation locus.
-
-[^extreme_B]: If $B_{r_{i}} = 0$ (resp.
-$B_{r_{i}} = 1$), `methsim` adds (resp.
-subtracts) a small value $\epsilon$.
-The default value is $\epsilon = 0.01$.
-The boundary values of $0$ and $1$ are common in $\beta_{i | r_{i}}$ and, were these retained, would result in methylomes with large regions that are uniformly unmethylated (resp.
-methylated).
-
-The second step requires further explanation.
-If there are no observations with the required $IPD$-region combination, then, by definition, `methsim` is unable to sample from the appropriate empirical distribution[^min_n].
-Instead, when this occurs, `methsim` samples from a user-specified parametric distribution.
-__RESULTS SECTION__ explores different choices of this distribution __AND SHOWS WHAT__.
-The default option is a zero-mean Gaussian distribution, chosen so that "on average" the methylation state at adjacent CpGs with large $IPD$s are independent of one another.
-The default standard deviation of this Gaussian distribution is $2$; this value, while fairly arbitrary, reflects the median standard deviation of the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions for each sample, which is remarkably constant across $r_{i}$ and samples[^ipd_stability] (__TODO: Format and reference table__).
-
-[^min_n]: By default, `methsim` requires at least 100 observations in the  $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ to try to avoid sampling from sparse empirical distributions.
-
-[^ipd_stability]: Figures \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_lor_by_pm_quantile_plots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} show that the variability of these distributions is fairly constant across $IPD$ for a given $r_{i}$ and sample.
-
-```{r, eval = TRUE, echo = FALSE}
-knitr::kable(readRDS("../data/median_sd_LOR_distributions.rds"), align = "c", digits = 2)
-```
-
-Now that we have $B_{r_{i}}$ and $\Psi_{i, i + 1}$ for $1 \leq i \leq n - 1$ we can simulate realisations of $\mathbf{Z_{\tilde{h}}}$.
-To begin the process (i.e., when $i = 1$) $Z_{\tilde{h}, 1}$, is simulated from the "marginal" distribution, which is Bernoulli($B_{r_{1}}$).
-Then, we loop over $i = 2, \ldots, n - 1$ to simulate the remainder of the process.
-At each step we compute the transition probability, $p_{(i + 1) | i} = Pr(Z_{\tilde{h}, i + 1} = 1 | Z_{\tilde{h}, i}) = \frac{Pr(Z_{tilde{h}, i}, Z_{tilde{h}, i + 1} = 1)}{Pr(Z_{tilde{h}})}$, from $B_{r_{i}}, B_{r_{i + 1}}, \Psi_{(i, i + 1)}$ and the observed (simulated) value at the previous iteration, $Z_{\tilde{h}, i}$.
-Then, $Z_{\tilde{h}, i + 1}$ is simulated from the Bernoulli($p_{(i + 1) | i}$) distribution.
-
-In order to compute the transition probability, $p_{(i + 1) | i}$, `methsim` first uses the iterative proportional fitting algorithm \citet[see]{Speed:2005tg}[for an overview of the algorithm] to construct the joint probability matrix,
-
-\begin{equation*}
-P = \bordermatrix{~ & Pr(Z_{i+1} = 0) = 1 - B_{r_{i+1}} & Pr(Z_{i+1} = 1) = B_{r_{i+1}} \cr
-Pr(Z_{i} = 0) = 1 - B_{r_{i}} & Pr(Z_{i} = 0, Z_{i + 1} = 0) & Pr(Z_{i} = 0, Z_{i + 1} = 1) \cr
-Pr(Z_{i} = 1) = B_{r_{i}} & Pr(Z_{i} = 1, Z_{i + 1} = 0) & Pr(Z_{i} = 1, Z_{i + 1} = 1) \cr}
-\end{equation*}
-
-Iterative proportional fitting is a general method "for constructing tables of numbers satisfying certain constraints" \citep{Speed:2005tg}.
-`methsim` uses iterative proportional fitting to construct the _unique_ $2 \times 2$ array (the joint probability matrix) with specified margins (the marginal probabilities of each methylation state at each locus, $B_{r_{i}} = Pr(Z_{i} = 1), B_{r_{i + 1}} = Pr(Z_{i + 1} = 1)$) and the specified cross-product ratio (the odds-ratio, $\psi_{i, i + 1} = \log_{2}(\Psi_{i, i + 1})$).
-
-Once $P$ is formed, the appropriate element of it is divided by the appropriate marginal probability to obtain the transition probability. An example of this procedure is illustrative.
-
-Let $\Psi_{i, i + 1} = 2$, and therefore $\psi = 2^{2} = 4$, and $B_{r_{i}} = B_{r_{i + 1}} = 0.7$.
-To begin the iterative proportional fitting algorithm, form the matrix $P^{(0)} = \Bigl(\begin{smallmatrix}
-\psi & 1 \\
-   1 & 1
-\end{smallmatrix} \Bigr).$ This matrix has the desired cross-ratio, $\psi$, but not the desired row and column margins.
-At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi$ while the row and column margins converge towards the desired values.
-The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P^{(*)} \equiv P$.
-For our example, $P^{(*)} = \Bigl(\begin{smallmatrix}
-0.122 & 0.178 \\
-0.178 & 0.522
-\end{smallmatrix} \Bigr)$, to 3 decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.
-
-\subsubsection{Simulating $\mathbf{W}$}
-
-- [x] DONE
-
-For each region we need to assign the vector of weights, $\mathbf{W_{r_{i}, \cdot}}$.
-To do so, `methsim` uses a heuristic procedure.
-For higher-order m-tuples, $m = 3$ or $4$ is recommended, `methsim` computes the relative frequency of the $2^{m}$ possible methylation patterns at each m-tuple with sufficient sequencing coverage.
-`methsim` then samples from this distribution, stratified by region-type; call these empirical distributions $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$.
-Figure \ref{fig:ADS_pattern_freqs_by_pm_boxplots} shows elementwise boxplots of $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$, that is, it shows the distributions $w_{i | r_{(i, \ldots, i + m - 1)}}$ for $i = 1, \ldots, 2^{m}$.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/ADS_pattern_freqs_by_pm_boxplots.pdf}
-\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{ADS} sample.
-Only CpGs with at least $5\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:ADS_pattern_freqs_by_pm_boxplots}
-\end{figure}
-
-These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and  \ref{fig:Ziller_merged_pattern_freqs_by_pm_boxplots} (__TODO: Are these other examples necessary?__).
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_pattern_freqs_by_pm_boxplots.pdf}
-\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{EPISCOPE} dataset.
-Only CpGs with at least $5\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_pattern_freqs_by_pm_boxplots.pdf}
-\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Lister} dataset.
-Only CpGs with at least $5\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Lister_pattern_freqs_by_pm_boxplots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_pattern_freqs_by_pm_boxplots.pdf}
-\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Seisenberger} dataset.
-Only CpGs with at least $5\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Seisenberger_pattern_freqs_by_pm_boxplots}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_pattern_freqs_by_pm_boxplots.pdf}
-\caption{$w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}, i = 1, \ldots, 2^{m} = 8$, the distribution of the relative frequencies for all CpG 3-tuples in each region type, for the \emph{Ziller} dataset.
-Only CpGs with at least $5\times$ sequencing coverage are used.
-UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
-\label{fig:Ziller_merged_pattern_freqs_by_pm_boxplots}
-\end{figure}
-
-From Figures \ref{fig:EPISCOPE_pattern_freqs_by_pm_boxplots}, \ref{fig:Lister_pattern_freqs_by_pm_boxplots}, \ref{fig:Seisenberger_pattern_freqs_by_pm_boxplots} and \ref{fig:Ziller_merged_pattern_freqs_by_pm_boxplots} we see that, unsurprisingly, in UMRs that a single methylation pattern dominates, the completely unmethylated pattern.
-Outside of UMRs, we see that while there is generally a dominant methylation pattern at each m-tuple, there may be $2-4$ otheres that are observed at non-negligible frequencies.
-PMRs are the most heterogeneous and LMRs and other regions are quite similar.
-
-The above procedure has several obvious limitations.
-Firstly, we are restricted to analying m-tuples within individual reads and read-pairs.
-Secondly, as $m$ increases, and similarly as $IPD$ increases, there are fewer m-tuples with sufficient sequencing coverage.
-This means our data are heavily biased towards regions of high CpG density, which typically correspond to UMRs.
-Additionally, we know that m-tuples with small $IPDs$ are far more likely to be co-methylated than those with larger $IPDs$ and so we are biased towards sampling entirely unmethylated and entirely methylated methylation patterns.
-It is difficult to know how to adjust for these biases, which are unavoidable given current sequencing technologies.
-These estimates very likely underestimate the degree of within-sample heterogeneity and, consequently, weights based on sampling from the empirical distributions, $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$, likely result in overly homogeneous simulated methylomes.
-
-A second, more subtle, bias is due to sequencing errors, bisulfite-conversion errors and alignment errors.
-Using the simple relative frequency estimation procedure will result in errors like these being inferred as (rare) methylation patterns.
-`MPFE` ([http://bioconductor.org/packages/release/bioc/html/MPFE.html](http://bioconductor.org/packages/release/bioc/html/MPFE.html)) is an R package that uses an estimation procedure designed to deal with some of these errors.
-However, `MPFE` is intended for analysing amplicon bisulfite-sequencing data with high sequencing coverages, and it's more computationally demanding estimation procedure does not scale to whole-genome data.
-
-With all the above limitations in mind, `methsim` still uses the simple strategy of sampling the $\mathbf{W}$ from $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$.
-At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data.
-In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}}, \ldots, \frac{1}{n_{\bar{h}}})$.
+The above-described model implicitly assumes that all $n_{h}$ haplotypes have the same marginal probabilities and co-methylation structure.
+We may extend this model to define $\mathbf{Z}$ as a _mixture_ of a small number of first-order Markov chains defined as in the above.
+Such a mixture model may be appropriate when simulating a sample that is a combination of cell types.
 
 \subsection{Simulating reads}
-
-- [x] DONE
 
@@
 
-`methsim` uses a simple method for simulating bisulfite-sequencing.
-The user specifies the desired read length and average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
+`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing.
+The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
 
-The number of reads required is given by $n_{reads} = \frac{\text{average sequencing coverage}}{\text{read length}} \times \text{size of genome}$.
+The number of reads required is computed by $n_{reads} = \frac{\text{average sequencing coverage}}{\text{read length}} \times \text{size of genome}$.
 The number of reads per chromosome is assigned proportional to the chromosome length.
 Then, the start of each read is sampled uniformly across the chromosome respective chromosome.
+We only retain those reads that overlap a methylation locus.
 
@@
 This isn't a big issue.
-Most paired-end bisulfite-sequencing datasets have overlapping mates, and so paired-end data can be crudely simulated by simply doubling the read length.
-Also, all reads must have the same length, i.e., no simulation of read trimming.
-I don't consider this feature a priority, but it could easily be implemented by generating read lengths from a probability distribution, e.g., Poisson.
+Most paired-end bisulfite-sequencing datasets have overlapping mates, and so paired-end data can be approximated by simply doubling the read length.
+Also, all reads must currently have the same length, i.e., no simulation of read trimming.
+I don't consider this feature a priority, but it could easily be implemented by generating read lengths from a given probability distribution.
 
-For all reads that overlap a methylation locus in the 'true' methylome, a pseudo-haplotype is sampled with probabilities given by $W_{i, \cdot}$.
-The methylation pattern of the sampled pseudo-haplotype is then copied, with probability $\epsilon$ that the methylation state at each locus is switched.
+Suppose we have a simulated read, $z$, that overlaps the 3-tuple $(i, i + 1, i + 2)$. The methylation state along the read is simulated as follows:
 
-The result is an object storing the genomic co-ordinates and methylation patterns of every read that contains a methylation locus.
+1. Sample $z_{i}$ from a Bernoulli($B_{i}$) distribution[^mixture].
+2.
+  a. Sample $z_{i + 1}$ from a Bernoulli($p_{i}$) distribution, where $p_{i}$ is the appropriate transition probability given the simulated $z_{i}$.
+  b. Sample $z_{i + 2} from a Bernoulli($p_{i + 1}$) distribution, where $p_{i + 1}$ is the appropriate transition probability given the simulated $z_{i + 1}$.
+3. Each of element of the simulated $z$ is independently flipped with probability $\epsilon$ to simulate sequencing error.
+
+[^mixture]: If $\mathbf{Z}$ is a mixture of Markov chains then we first simulate which component the read comes from by sampling from a multinomial($w$) distribution, where $w$ is the vector of weights of each component.
+All subsequent steps will be simulated according to which component is sampled at this step.
+
+The result of this process is a `SimulatedBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states.
 
 \subsection{Construct the output}
+
+__UP TO HERE: See what below can be salvaged__
 
@@
 
+- Apparent co-methylation arises because of dependence of $B_{i}$, even when $Z_{i}$ are independent!
+E.g., set within-fragment co-methylation to zero but have $B_{i}$ fixed for a region results in independent $Z_{i}$ but dependent $B_{i}$.
+  - What to make of this?
+  - What if we sample $B_{i}$ from $\beta_{i | r_{i}}$ instead of using a fixed value for all loci in a region?
+
 \subsection{The choice of $\mathbf{W}$}

### 2015-04-24 latex/phd_thesis.tex
Pre-lunch commit. Begin re-write of methsim chapter.

@@
 of the degree of Doctor of Philosophy\\\\
-April 2015\\\\
+May 2015\\\\
 Department of Mathematics and Statistics\\
@@
 
+\textbf{TODO}
+
 \begin{itemize}
@@
 \chapter{Acknowledgments}
+
+First and foremost, I would like to thank my PhD supervisors, Terry Speed and  Peter Hall.
+Peter, thank you for supporting my decision to radically change tac very early on in my PhD from a topic in theoretical statistics to a heavily computational and interdisciplinary problem.
+And Terry, thank you for agreeing to take me on as a student after this change of heart.
+I have learnt much from working with you and greatly appreciate your generosity and guidance over the course of my PhD.
+Thanks also to Gordon Smyth, the chair of my PhD committee, for always taking the time to help me.
+
+Thank you to the Walter and Eliza Hall Institute for being such a great place to do research.
+In particular, thank you to my wonderful friends and colleagues in the Bioinformatics Division for your support and advice.
+A special thanks to Keith Satterley and the ITS department who put up with and helped fix my computational blunders.
+
+My PhD would not be possible without the people who provided me with data to analyse.
+I am grateful to Sue Clark, Aaron Statham, Emma Whitelaw and Harry Oey for the opportunity to collaborate with and learn from you.
+Thanks also to Ryan Lister, Kasper Hansen and Felix Krueger for making their data publicly available and answering all my nit-picky questions.
+Open science is the best science.
+
+Equally, my PhD would not be possible without the countless people who contribute to the software I use in my daily research.
+A special thanks to those who helped me as I learnt to write software of my own: Felix Krueger, Toby Sargeant, Martin Morgan, Hervé Pagès, Hadley Wickham and Dirk Eddelbuettel.
+Open software is the best software.
+
+__software__
+
+__funding, inc. melanie__
+
+- WEHI, Unimelb,
+- APA, VLSCI, EMBL, other funders
+- Family and friends
+- Terry and Peter
+- Melanie
+- Bioinf division
+- Proof readers: Belinda, Saskia, Goknur
+- Hannah and Darcy
+- Gordon
+- Bioconductor, Hadley, Dirk, Felix, Toby
+- IT, Keith
+- collaborators
+- Open access
+
+I am grateful to the Walter and Eliza Hall Institute
 
@@
 
-\include{introduction} % introduction.tex
-
 \include{biology_background} % biology_background.tex
@@
 
+\include{concluding_remarks} % concluding_remarks.tex
+
 \backmatter

### 2015-04-24 Rmarkdown/concluding_remarks.Rmd
Update thesis structure.

---
title: "Concluding remarks"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\chapter{Concluding remarks}\label{chap:concluding_remarks}

__TODO: Review each chapter__

### 2015-04-24 latex/phd_thesis.tex
Update thesis structure.

@@
 
-First and foremost, I would like to thank my PhD supervisors, Terry Speed and  Peter Hall.
-Peter, thank you for supporting my decision to radically change tac very early on in my PhD from a topic in theoretical statistics to a heavily computational and interdisciplinary problem.
+I would like to thank my PhD supervisors, Terry Speed and  Peter Hall.
+Peter, thank you for supporting my decision to radically change tack very early on in my PhD from a topic in theoretical statistics to a heavily computational and interdisciplinary problem.
 And Terry, thank you for agreeing to take me on as a student after this change of heart.
@@
 In particular, thank you to my wonderful friends and colleagues in the Bioinformatics Division for your support and advice.
-A special thanks to Keith Satterley and the ITS department who put up with and helped fix my computational blunders.
+A special thanks to Keith Satterley and the ITS department who both tolerated and helped fix my computational blunders.
 
@@
 I am grateful to Sue Clark, Aaron Statham, Emma Whitelaw and Harry Oey for the opportunity to collaborate with and learn from you.
-Thanks also to Ryan Lister, Kasper Hansen and Felix Krueger for making their data publicly available and answering all my nit-picky questions.
+Thanks also to Ryan Lister, Kasper Hansen and Felix Krueger for making their data publicly available and answering all my nitpicky questions.
 Open science is the best science.
@@
 
-__software__
+I am grateful to the Australian tax payer for supporting me financially during my studies with an Australian Postgraduate Award.
+I also thank the Victoria Life Sciences Computing Initiative for additional funding and Melanie Bahlo for allowing me a very flexible work arrangement as a research assistant and her encouragement to pursue a PhD.
 
-__funding, inc. melanie__
+I owe an enourmous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proof readers, Belinda Phipson, Saskia Freytag and Goknur Giner, who have greatly improved this thesis.
 
-- WEHI, Unimelb,
-- APA, VLSCI, EMBL, other funders
-- Family and friends
-- Terry and Peter
-- Melanie
-- Bioinf division
-- Proof readers: Belinda, Saskia, Goknur
-- Hannah and Darcy
-- Gordon
-- Bioconductor, Hadley, Dirk, Felix, Toby
-- IT, Keith
-- collaborators
-- Open access
+\textbf{TODO: final thanks}
 
-I am grateful to the Walter and Eliza Hall Institute
-
-I am grateful to the following people for their support during my PhD...
+Finally, and most of all, thank you to my parents, Gay and John, to my brothers, Jack, Luke and Daniel, and to my friends. Words cannot express
 
@@
 FDR & false discovery rate \\
+DMC & differentially methylated cytosine \\
 DMR & differentially methylated region \\
-DMC & differentially methylated cytosine \\
 HMM & hidden Markov model \\
@@
 
+% TODO: Check list of symbols
 \chapter{List of Symbols}
@@
 
-\chapter{Concluding Comments}\label{chapter:Conclusions}
-
 % 2 - 4 pages

### 2015-04-24 Rmarkdown/methsim.Rmd
Continue work on methsim chapter.

@@
 
-For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}, B_{i + 1}$, and the odds ratio, $\psi_{i}.
+For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}, B_{i + 1}$, and the odds ratio, $\psi_{i}$.
 The general form of $P_{i, i + 1}$ is shown below:
@@
 
-To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimators, $\mathbf{\beta} and $\mathbf{\widehat{\psi}}$, in the input methylome.
-More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ and $\mathbf{\psi}$ is based on $\widehat{\psi_{IPD(i, i+1) | r_{i}}$, the empirical distributions of $\beta_{i}$ and $\psi_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
+To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimators, $\mathbf{\beta}$ and $\mathbf{\widehat{\psi}}$, in the input methylome.
+More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ and $\mathbf{\psi}$ is based on $\widehat{\psi_{IPD(i, i+1) | r_{i}}}$, the empirical distributions of $\beta_{i}$ and $\psi_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
 These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and __FIGURE__.
@@
 
-__UP TO HERE: See what below can be salvaged__
+The `SimulatedWGBS` object contains all the data from the simulation.
+However, it is not always the most convenient or efficient format with which to work.
+For example, many downstream analyses only make use of $\beta$-values, so we might want to summarise our simulated data this way.
+`methsim` provides the `asMethPat()` function to convert the `SimulatedWGBS` object to a `MethylationTuples::MethPat` object containing methylation patterns at m-tuples of a given size[^simplify]; all the functionality of the `MethylationTuples` package is then available to the user, such as computing $\beta$-values with the `methLevel()` method.
 
-- [x] DONE
-
-While the `SimulatedBS` object contains all the data on the methylation states of the reads, it is not always the most convenient or efficient format with which to work.
-For example, many downstream analyses only make use of $\beta$-values, so we might want to summarise our simulated data this way.
-`methsim` provides the `asMethPat()` function to convert the `SimulatedBS` object to a `MethPat` object containing methylation patterns at m-tuples of a given size; all the functionality of the `MethylationTuples` package is then available to the user, such as computing $\beta$-values with the `methLevel()` method.
+[^simplify]: Alternatively, the user may use the `simplify` argument of the `simulate()` method to return an already 'simplified' `MethylationTuples::MethPat` object rather than the `SimulatedWGBS` object.
 
@@
 
-- [x] DONE
-
 The above describes how to simulate a single whole-genome bisulfite-sequencing sample.
-If we wanted to simulate six independent samples then we could simply run this procedure six times, with a different `SimulateMethylomeParam` object for each realisation.
-However, we will generally be interested in simulating experiments where the samples are not  is some relationship amongst the samples.
+If we wanted to simulate six independent samples then we could simply run this procedure six times, with a different `MethylomeParam` object for each realisation.
+However, we will generally be interested in simulating experiments where there is some relationship between the samples.
 For example, suppose we want to simulate a two-group experiment with three samples per group.
-Further, suppose we want to include differentially methylated regions between the two groups.
+Furthermore, suppose we want to include differentially methylated regions between the two groups.
 How do we simulate such an experiment?
 
-To simulate these type of experiments, we need to be able to control the variation between the `SimulatedBS` objects.
+To simulate these type of experiments, we need to be able to control the variation between the `SimulatedWGBS` objects.
 In this example, we probably want the three samples in each group to be more similar to each other than to the three samples in the other group.
 Essentially, we need to be able to control the within-group and between-group variation.
-`methsim` controls this variation through the workflow show in __FIGURE__.
-If we want two samples two be very different from one another, then we would use two different `SimulateMethylomeParam` objects at the beginning.
-However, if we want to simulate two technical replicates, then we would use the same `SimulateMethylomeParam`, `SimulatedMethylome` and `SimulateBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.
+
+Biological variation is controlled through the `MethylomeParam` object and the method with which we simulate $\mathbf{B}$ and $\mathbf{\psi}$ through the `simulate()` method.
+Technical variation is controlled the `WGBSParam` object.
+If we want two samples to be very different from one another, then we would use two different `MethylomeParam` objects at the beginning of the process.
+However, if we want to simulate two technical replicates, then we would use the same `WGBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.
 
 We might also consider more subtle ways of introducing variation.
-For example, we might use the same partitioning of the methylome and the same average methylation level for each region but vary the strength of co-methylation.
+For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\mathbf{\psi}$ between calls to `simulate()`.
+
+and the same average methylation level for each region but vary the strength of co-methylation.
 This could be achieved by altering the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions between two `SimulateMethylomeParam` objects while holding the rest constant.
@@
 
-In __RESULTS SECTION__ I explore how much variation is created using these different techniques.
+In Section ref{sec:results} I explore how much variation is created using these different techniques.
 
-\subsubsection{Simulating differential methylation}
-
-- [ ] DONE
+\subsubsection{Simulating differential methylation}\label{sec:sim_diff_meth}
 
 An obvious application of `methsim` is to simulate data to be used in a study comparing methods for identifying differential methylation.
-This requires a particularly controllable method for inducing differences in `SimulatedMethylome` objects.
-The user wants to provide a set of methylation loci, either individual loci or as regions, and the expected difference in methylation levels at these regions.
 
-__Scattered thoughts below__
+One way of doing this would be to take a `SimulatedMethylome` object and create a modified copy where $\mathbf{B}$ has been perturbed by specified amounts at a set of specified loci.
+We could then simulate bisulfite-sequencing data from each `SimulatedMethylome` and study which analysis methods can identify these simulated DMCs and DMRs.
 
-1. Individual loci: Want to specify $B_{i} = \sum_{\tilde{h} = 1}^{n_{\tilde{h}}} W_{r_{i}, \tilde{h}} \times Z_{i, \tilde{h}}$, i.e., fix $\mathbf{W_{i, \cdot}}$ and $\mathbf{Z_{i, \cdot}}$.
-2.
-Regionwise: Want to specify $mean(B_{r_{i}})$ (i.e., average $B$ across all loci in $r_{i}$).
-Could do this by applying (1) to all loci in the region or by specifying $B_{r_{i}}$ and hoping it just works out in the simulation.
-
-Perhaps both could be achieved by holding $Z$ fixed and varying $W$?
-Is that like assuming that the heterogeneity is causing the DMRs (or will it be interpreted that way)?
-Would tweaking $W$ to guarantee a particular $B$ be a computationally feasible strategy?
-Or would it be easier to specify the desired $B$ and simulate the region (this wouldn't guarantee $B$ is the desired value but should equal it "on average")?
+__UP TO HERE: See what below can be salvaged__
 
@@
 The `SimulatedMethylome` and `SimulatedBS` objects are large in memory.
-This is to some extent unavoidable; we are simulating entire methylomes and whole-genome sequencing experiments.
-Some improvements in memory usage are possible by alternative implementations of key classes, however, these are generally at the expense of more complicated code and longer running times of associated methods.
-For example, long and repetitive vectors may be efficiently represented by run length encodings, however, it is slower to extract elements from these than simple numeric vectors.
+This is to some extent unavoidable; we are simulating whole-genome bisulfite-sequencing sequencing experiments that produce an enormous amount of data.
 
-Shown below are the running times and memory usage involved in simulating a $30\times$ whole-genome bisulfite-sequencing sample using the _ADS_ `SimulateMethylomeParam` object included in `methsim`.
-Using eight CPU cores in parallel, it takes approximately seven minutes to simulate the underlying 'true' methylome and a further nine minutes to simulate a single whole-genome bisulfite-sequencing assay of this methylome.
+Shown below are the running times and memory usage involved in simulating a $30\times$ whole-genome bisulfite-sequencing sample using the _ADS_ `MethylomeParam` object included in `methsim`.
+Using up to 8 CPU cores in parallel, it takes less than 5 minutes to simulate the underlying 'true' methylome and a further __HOW LONG__ to simulate a single $30 \times$ sequencing coverage whole-genome bisulfite-sequencing assay of this methylome[^parallel].
 
-These results are indicative of simulating high-coverage whole-genome bisulfite-sequencing for human-sized genomes.
-Simulations using a smaller genome or lower sequencing coverage will have faster running times and lower memory usage.
+[^parallel]: NB the creation of the `SimulatedMethylome` via `simulate(ADS, BPPARAM)` currently only uses a single CPU.
 
-```r
+__UP TO HERE: simulate-WGBSParam-method seems to be causing a memory leek and resulting crash. Running valgrind in reduced problem to investigate.__
+
+```{r, echo = TRUE, eval = FALSE}
 # Load necessary packages
@@
 
-# ADS: the SimulateMethylomeParam object.
+# ADS: the MethylomeParam object.
 pryr::object_size(ADS)
-#> 22.5 MB
+#> 16.5 MB
 
 # sm: the SimulatedMethylome object
-system.time(sm <- simulate(ADS, BPPARAM = BPPARAM))
+system.time(l_sm <- simulate(ADS, BPPARAM = BPPARAM))
+#> Currently only simulates CpG methylation.
+#> Currently only simulates unstranded methylomes.
+#> Simulating 1 methylome..
 #>    user  system elapsed
-#> 839.505  19.612 417.271
+#> 186.680   9.676 270.780
+sm <- l_sm[[1]]
 pryr::object_size(sm)
-#> 2.93 GB
+#> 676 MB
 
-# sbsp: the SimulatedBSParam object.
-# NOTE: Basically all the memory used by the SimulatedBSParam object is the
+# wgbsp: the WGBSParam object.
+# NOTE: Basically all the memory used by the WGBSParam object is the
 # SimulatedMethylome object.
-sbsp <- SimulateBSParam(SimulatedMethylome = sm,
-                        aveCov = 30,
-                        errorRate = 0.01,
-                        target = NULL)
-pryr::object_size(sbsp)
-#> 2.93 GB
+wgbsp <- WGBSParam(SimulatedMethylome = sm,
+                   AveCov = 30,
+                   ErrorRate = 0.01,
+                   SequencingType = 'SE',
+                   ReadLength = 100)
+pryr::object_size(wgbsp)
+#> 676 MB
 
 # sbs: the SimulatedBS object.
-system.time(sbs <- simulate(sbsp, BPPARAM = BPPARAM))
+system.time(sbs <- simulate(wgbsp, BPPARAM = BPPARAM))
+#> Simulating 1 bisulfite-sequencing sample...
+
 #>     user   system  elapsed
@@
 ```
+
+These results are indicative of simulating high-coverage whole-genome bisulfite-sequencing for human-sized genomes.
+Simulations using a smaller genome or lower sequencing coverage will have faster running times and lower memory usage.
 

### 2015-04-24 Rmarkdown/datasets.Rmd
Changes suggested by Belinda.

@@
 M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set.
-These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level more than $0.03$ from the median, i.e., $|nrplml_{CpG} - median(rplm_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
+These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e., $|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
 3.

### 2015-04-25 Rmarkdown/wgbs_statistical_framework.Rmd
Fix notation typo
Add changes suggested by Saskia to WGBS statistical framework.

@@
 
-I describe these sources of variation, which are illustrated in Figure \ref{fig:bs-seq_protocol}.
+In the following, I describe these sources of variation, which are illustrated in Figure \ref{fig:bs-seq_protocol}.
 
@@
 We have a large number (~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
-Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to $H_{i}$.
+Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to the number of fragmentation containing the $i^{th}$ methylation locus, $H_{i}$.
 
@@
 \begin{align*}
-  m_{i} &= \sum_{r = 1}^{d_{i}} z_{i} \\
+  m_{i} &= \sum_{r = 1}^{d_{i}} z_{r, i} \\
         &= |\{z: z \in \mathcal{R}_{i}, z = 1 \}| \\
-  u_{i} &= \sum_{r = 1}^{d_{i}} (1 - z_{i}) \\
+  u_{i} &= \sum_{r = 1}^{d_{i}} (1 - z_{r, i}) \\
       &= |\{z: z \in \mathcal{R}_{i}, z = 0 \}
@@
   \beta_{(i, i + 1)}^{mm} &= \frac{mm_{(i, i + 1)}}{d_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{mu} &= \frac{MU_{(i, i + 1)}}{d_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{um} &= \frac{UM_{(i, i + 1)}}{d_{(i, i + 1)}} \\
-  \beta_{(i, i + 1)}^{uu} &= \frac{UU_{(i, i + 1)}}{d_{(i, i + 1)}}
+  \beta_{(i, i + 1)}^{mu} &= \frac{mu_{(i, i + 1)}}{d_{(i, i + 1)}} \\
+  \beta_{(i, i + 1)}^{um} &= \frac{um_{(i, i + 1)}}{d_{(i, i + 1)}} \\
+  \beta_{(i, i + 1)}^{uu} &= \frac{uu_{(i, i + 1)}}{d_{(i, i + 1)}}
 \end{align*}
@@
 Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{d_{i, j}}$.
-Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
+Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample.
+This phenomenon, called _co-methylation_ is discussed and analysed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
 
@@
 In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
-The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.
+The smoothing is 'local' to exploit co-methylation, which is considered a local phenomenon.
 
@@
 \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} compare the smoothing results of `BiSeq` to `bsseq`.
-Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`, but the comparison is a little unfair since both use RRBS data, for which `bsseq` is not designed[^rrbs_sim].
+Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`.
+However, these comparison studies use RRBS data, which `bsseq` is not designed for`, and so will favour methods designed for RRBS data[^rrbs_sim].
 
@@
 In a study involving multiple samples, the set of $\beta$-values can be summarised as a matrix where each row is a locus and each column is a sample.
-Some values will be missing, either because there was insufficient sequencing coverage to estimate a $\beta$-value or because that locus is not a cytosine for that sample.
+Some values will be missing, either because there was insufficient sequencing coverage to estimate a $\beta$-value or because that locus is not a cytosine for the sample in question.
 This matrix might be visualised to learn about the distribution of methylation levels, either row-wise (to learn about the variability across samples) or column-wise (to learn about the variability within samples).
@@
 
-It has been previously observed that cancer samples have highly variable DNA methylation \citep{Hansen:2011gu}, which, combined with the possibility of multiple sub-clones, explains these intermediate $\beta$-values in the cancer samples.
+It has previously been observed that cancer samples have highly variable DNA methylation \citep{Hansen:2011gu}, which, combined with the possibility of multiple sub-clones, explains these intermediate $\beta$-values in the cancer samples.
 
@@
 Densities are normalised so that the maximum value for each sample is 1.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.
+Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.
 `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
@@
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
 \label{fig:EPISCOPE_plotBetaFreqPoly}
@@
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
 \label{fig:Lister_plotBetaFreqPoly}
@@
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
 \label{fig:Seisenberger_plotBetaFreqPoly}
@@
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
+Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
 \label{fig:Ziller_plotBetaFreqPoly}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _EPISCOPE_ data, stratified by whether the CpG is in a CpG island.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _Lister_ data, stratified by whether the CpG is in a CpG island.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _Seisenberger_ data, stratified by whether the CpG is in a CpG island.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the _Ziller_ data, stratified by whether the CpG is in a CpG island.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _EPISCOPE_ data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _Lister_ data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _Seisenberger_ data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the _Ziller_ data, stratified by whether the CpG is in a CpG island.
+\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
-Only CpGs with at least $10\times$ sequencing coverage are included.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
 Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
@@
 There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the smaller strand-correlations at these lower cutoffs.
-Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 -- 0.9$, with some notable exceptions.
+Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8$ to $0.9$, with some notable exceptions.
 
@@
 
-In papers about DNA methylation there is often a sentence like this quote from \citet{Laird:2003kw}: "About 70% of the CpG dinucleotides in the mammalian genome are methylated".
+\citet{Laird:2003kw} says in a review paper on DNA methylation that, "about 70% of the CpG dinucleotides in the mammalian genome are methylated".
+Similar statements are made in many papers about DNA methylation, but how should these be interpreted?
+
 In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG.
 However, as can be seen from Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_plotBetaFreqPoly}, the bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG.
-To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.
+To make a useful statement about the methylation level of a particular CpG really requires more information, such as whether it is within a CpG island.
 
 The "$70\%$" statement can also be interpreted as an estimate of the probability that a cytosine randomly sampled from a haploid copy of a mammalian genome is a methylcytosine.
-Because in a whole-genome bisulfite-sequencing experiment we measure methylation from a pool of cells, not a single haploid copy of the genome, I do not think it makes sense to describe a particular cytosine in the genome as being a 'methylcytosine'.
-That is, it only makes sense to describe an observation of a cytosine, $z_{i}$, as being methylated or unmethylated and not to describe a cytosine, really a position in the genome, $i$, as being a methylated or unmethylated.
-However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have sought to use the latter definition, which I believe to be an unnecessary source of confusion.
+Note that this refers to individual cytosines, $Z_{h, i}$, and not the genomic position of the locus, $pos_{i}$.
+Also recall that most assays measure DNA methylation from a pool of cells, not a single haploid copy of the genome.
+The methylation state at the $i^{th}$ locus may vary across the $\mathcal{H}_{i}$ DNA fragments.
+Therefore, I do not think it makes sense to describe a locus, $i$, as being a 'methylcytosine'.
+However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have used this latter definition, which I believe to be an unnecessary source of confusion.
 
@@
 
-__UP TO HERE: Need a strong summary/conclusion. Emphasise the novel aspects of this chapter and the wide variety of datasets used.__
+__TODO: Need a strong summary/conclusion. Emphasise the novel aspects of this chapter and the wide variety of datasets used.__
 

### 2015-04-26 Rmarkdown/comethylation.Rmd
Begin writing up simulation results.

@@
 
-- __TODO: See how the variance estimator behaves in some of the more pathological simulations__
-
 \subsection{Simulation study}
@@
 We analyse the results of a simulation study to explore the effects of sparsity and heterogeneity of odds ratios on estimates of odds ratios in $2 \times 2 \times K$ contingency tables.
-The simulation requires that we specify several parameters shown below.
-In order to relate these back to DNA methylation, I also describe each parameter using the statistical framework proposed in Chapter \ref{chap:wgbs_statistical_framework}:
+The simulation requires that we specify several parameters.
+In order to relate these back to DNA methylation, I describe each parameter using the statistical framework proposed in Chapter \ref{chap:wgbs_statistical_framework}:
 
 - the number of tables, $K$. The number of 2-tuples.
-- the marginal probabilities, $\pi_{1 +} = Pr(Z_{i})$ and $\pi_{+ 1} = Pr(Z_{i'})$.
+- the marginal probabilities, $\pi_{2 +} = Pr(Z_{i})$ and $\pi_{+ 2} = Pr(Z_{i'})$.
 Some simulations use a common set of marginal probabilities across all tables but some simulate the marginal probabilities from a specified distribution.
@@
 The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth).
-Most simulations use a truncated negative binomial distribution to model a bisulfite-sequencing experiment with mean sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].
-
-[^nb_sim]: In the simulation study, a larger cutoff results in higher quality data since we can simply keep simulating until we have $K$ tables with sufficient sequencing depth.
-In analysing real data, increasing the cutoff may be detrimental since we could end up with fewer than $K$ tables to analyse.
-
-__UP TO HERE: Insert simulation results__
-
-- Use parameters $n, \pi, \psi, \theta$ and not the ones from Chapter \ref{chap:wgbs_statistical_framework} since these are more succinct.
+Most simulations use a truncated negative binomial distribution to model the sequencing depth.
+The truncated negative binomial distribution is parameterised by the mean ($mu$), the dispersion parameter (size) and the truncation level (trunc).
+Unless otherwise noted, simulations use $\mu = 30$, $\text{size} = 10$ and $\text{trunc} = 9$, corresponding to an average sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].
+
+[^nb_sim]: In the simulation study, a larger value of 'trunc' results in higher quality data since we can simply keep simulating until we have $K$ tables with sufficient sequencing depth.
+In analysing real data, increasing this cutoff may be detrimental since we end up with fewer tables to analyse.
+
+For each set of parameters, we generate $K$ $2 \times 2$ contingency tables having the desired marginal probabilities and odds ratio. We then compute the following statistics:
+
+1. The $K$ 0.5-adjusted unconditional odds ratio estimates, $\widehat{\psi_{0.5}}$, from each $2 \times 2$ table.
+2. The unconditional odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\psi_{U}}$.
+3. The Mantel-Haenszel odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\psi_{MH}}$.
+
+The results of each simulation are summarised by a plot of $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line[^density].
+
+[^density]: In many simulations $\psi$ is constant across the $K$ tables and therefore the density is degenerate.
+
+\subsubsection{Results}
+
+To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1, Figure \ref{fig:Simulation_1}), and Simulation 2a, $K = 1000$ (Figure \ref{fig:Simulation_2}).
+Under these parameters, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta_{MH}}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
+This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller.
+
+By contrast, $\widehat{\theta_{0.5}}$ are highly biased estimates of $\theta$, even when $K = 1000$.
+The histograms of $\widehat{\theta_{0.5}}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).
+The bias can be eliminated only by having ultra-deep sequencing coverage ($mean(d) = 10000 \times$), although the variation of $\widehat{\theta_{0.5}}$ across the $K = 1000$ tables remains non-negligible (Simulation 2b Figure \ref{fig:Simulation_2d}).
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_1.pdf}
+\caption{Simulation 1.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_1}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_2.pdf}
+\caption{Simulation 2a.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_2}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
+\caption{Simulation 2b.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_2d}
+\end{figure}
+
+Next we explore what happens when we allow the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, to vary across the $K$ tables when $\theta$ is a fixed value (not necessarily zero) identically across the $K$ tables for $K = 100$.
+To do so, we simulate $\pi_{2 +}, \pi_{+ 2}$ \eqd Uniform(0, 1)$ for each $2 \times 2$ table.
+As is expected from the theory, Figure \ref{fig:Simulation_3} shows that the Mantel-Haenszel estimator does an excellent job of estimating the true value of $\theta$.
+By contrast, $\widehat{\theta_{U}}$, is well away from the true value and the distribution of $\widehat{\theta_{0.5}}$ is skewed and widely dispersed.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
+\caption{Simulation 3.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_3}
+\end{figure}
+
+In a sense, the $\pi_{2 +}, \pi_{+ 2}$ \eqd Uniform(0, 1)$ assumption is too easy.
+As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are mostly near the boundaries, not uniformly distributed.
+Furthermore, each methylation locus in a 2-tuple will likely have a similar marginal probability of methylation.
+To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2}$ \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:Beta_marginals}).
+The results of this simulation (Simulation 4, Figure \ref{fig:Simulation_4}) are basically an exaggerated version of Simulation 3.
+The Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, remains an excellent estimator.
+The unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$.
+The distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.
+
+```{r Beta_marginals, echo = FALSE}
+hist(rbeta(1000, 0.3, 0.2))
+```
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_4.pdf}
+\caption{Simulation 4.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_4}
+\end{figure}
+
+We now explore what happens when the odds ratio is heterogeneous across the $K$ tables. __UP TO HERE__
 
@@
 
-__TODO__
+We conclude that we cannot estimate within-fragment co-methylation at individual 2-tuples using $\psi_{0.5}$ unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
+The $\widehat{\theta_{0.5}}$ at typical sequencing depths ($d \approx 30$) are highly biased.
+
+[^ultra-high_depth]: Ultra-deep sequencing has its own problems such as an increased effect of PCR-bias.
 

### 2015-04-26 Rmarkdown/comethylation.Rmd
Begin writing co-methylation data analysis results.

@@
 \begin{chapabstract}
-I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_.
+I describe two measures of within-sample co-methylation, \emph{within-fragment co-methylation} and \emph{correlation of $\beta$-values}.
 The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.
@@
 \includegraphics[width=0.8\textwidth]{../figures/feature_status.pdf}
-\caption{Schematic illustrating the 'feature status' of four pairs of methylation loci.
+\caption{Schematic illustrating the `feature status' of four pairs of methylation loci.
 The circles represent loci and the grey bars a genomic feature, such as CpG islands.
-Each pair's 'feature status' is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside).
-Note that the rightmost pair is 'inside' even though the two loci that comprise the pair are in different elements of the genomic feature.}
+Each pair's `feature status' is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside).
+Note that the rightmost pair is `inside' even though the two loci that comprise the pair are in different elements of the genomic feature.}
 \label{fig:feature_status}
@@
 
-\section{Estimating dependence in sparse $2 \times 2 \times K$ contingency tables}
+\section{Estimating dependence in sparse $2 \times 2 \times K$ contingency tables}\label{sec:sparse_2x2xK_tables}
 
@@
 Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:
+
+__TODO: Some reasons to use the odds ratio.
+Firstly, the ``possible range of values of the [log] OR is not influenced by the marginal distributions of the variables involved'' \citep{Kraemer:2004el}.
+Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints \citep{plackett1981analysis}.__
 
@@
 
-We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
+We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
 
@@
 
-[^mh_estimator]: Other estimators had been proposed but were supplanted by the work of \citet{Robins:1986vh}.
+[^mh_variance]: Other estimators had been proposed but were supplanted by the work of \citet{Robins:1986vh}.
 
@@
 
-\subsection{Simulation study}
+\subsection{Simulation study}\label{subsec:simulation_study}
 
@@
 Next we explore what happens when we allow the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, to vary across the $K$ tables when $\theta$ is a fixed value (not necessarily zero) identically across the $K$ tables for $K = 100$.
-To do so, we simulate $\pi_{2 +}, \pi_{+ 2}$ \eqd Uniform(0, 1)$ for each $2 \times 2$ table.
+To do so, we simulate $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ for each $2 \times 2$ table.
 As is expected from the theory, Figure \ref{fig:Simulation_3} shows that the Mantel-Haenszel estimator does an excellent job of estimating the true value of $\theta$.
@@
 
-In a sense, the $\pi_{2 +}, \pi_{+ 2}$ \eqd Uniform(0, 1)$ assumption is too easy.
+In a sense, the $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ assumption is too easy.
 As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are mostly near the boundaries, not uniformly distributed.
 Furthermore, each methylation locus in a 2-tuple will likely have a similar marginal probability of methylation.
-To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2}$ \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:Beta_marginals}).
+To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:Beta_marginals}).
 The results of this simulation (Simulation 4, Figure \ref{fig:Simulation_4}) are basically an exaggerated version of Simulation 3.
@@
 
-We now explore what happens when the odds ratio is heterogeneous across the $K$ tables. __UP TO HERE__
+We now explore what happens when the odds ratio is heterogeneous across the $K$ tables.
+Although results are shown for $\widehat{\theta_{0.5}}$ and $\widehat{\theta_{U}}$, we do not discuss these estimators further since they perform poorly in even the simpler scenarios.
+
+In these remaining simulations, all parameters are as for Simulation 4 except that the true log odds ratio of each table, $\theta_{k}$, comes from the following distributions:
+
+- Simulation 5 and Simulation 6: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 7:$\theta_{k} \eqd Skewed-Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 8: $\theta_{k} \eqd 0.5 Gaussian(- \theta_{0}, 1) + 0.5 Gaussian(\theta_{0}, 1)$.
+
+Simulations 5, 6 and 7 (Figures \ref{fig:Simulation_5}, \ref{fig:Simulation_6} and \ref{fig:Simulation_7}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, is a reasonable estimator of the centre of the distribution of log odds ratios.
+It performs better in Simulation 5, where the distribution of $\theta$ is symmetric, as opposite to Simulation 7, where the distribution of $\theta$ is skewed.
+Unsurprisingly, $\widehat{\theta_{MH}}$ does a better job as $K \to \infty$.
+It should be emphasised, however, that the asymptotic variance of the estimator, $hat{\sigma}(\widehat{\theta_{MH}})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$.
+We can see that the asymptotic $95 \%$ confidence interval of $\widehat{\theta_{MH}}$ only covers a small amount of the variation in the $\theta$.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_5.pdf}
+\caption{Simulation 5.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_5}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_6.pdf}
+\caption{Simulation 6.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_6}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_7.pdf}
+\caption{Simulation 7.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_7}
+\end{figure}
+
+Finally, Simulation 8 (Figure \ref{fig:Simulation_8}) shows that if the distribution of $\theta$ is multimodel that $\widehat{\theta_{MH}}$ will estimate the 'average' effect.
+Of course, the utility of any point estimate is severely reduced when the true effect is multimodal.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_8.pdf}
+\caption{Simulation 8.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
+See main text for description of other parameters.}
+\label{fig:Simulation_8}
+\end{figure}
 
@@
 
-We conclude that we cannot estimate within-fragment co-methylation at individual 2-tuples using $\psi_{0.5}$ unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
-The $\widehat{\theta_{0.5}}$ at typical sequencing depths ($d \approx 30$) are highly biased.
+We summarise the results of the simulation study with respect to our aim of estimating within-fragment co-methylation from whole-genome bisulfite-sequencing data.
+
+This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual 2-tuples using the simple estimator $\theta_{0.5}$ unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
+The distributions of $\widehat{\theta_{0.5}}$ at typical sequencing depths ($d \approx 30$) are highly biased and dispersed, even under very simple simulation scenarios.
+Therefore, some degree of aggregation is required in order to study within-fragment co-methylation.
 
@@
 
+Suppose we aggregate $K$ 2-tuples to form a $2 \times 2 \times K$ contingency table.
+When little can be assumed about this $2 \times 2 \times K$ contingency table, the Mantel-Haenszel estimator is the most appropriate of those considered.
+Firstly, it is highly robust to different marginal probabilities across the $K$ tables.
+Secondly, when the odds ratios are heterogeneous across the $K$ tables, the Mantel-Haenszel estimator will estimate the 'average' effect.
+
+Of course, the utility of this 'average' will depend on the distribution of $\theta$, which is unknown in practice.
+However, we might explore its variability by computing the Mantel-Haenszel estimate under a range of aggregation strategies (i.e.
+across different $k$).
+
+\FloatBarrier
+
 \section{Within-fragment co-methylation}\label{sec:wf_cometh}
 
-The key advantage of studying within-fragment co-methylation is that it measures the dependence of DNA methylation at the scale on which the biological process acts.
-I have developed a method based on odds-ratios to quantify the strength of co-methylation for a pair of methylation loci in a single sample.
-
-\subsection{Studying within-fragment co-methylation from sparse $2 \times 2 \times K$ contingency tables}
-
-- Asymptopia II with sparse tables
-- Homogeneity??? Don't/can't know.
-
-- Describe co-methylation in terms of $X$ and $Y$, the sparse $2 \times 2 \times K$ contingency tables, the marginal probabilities, and the odds ratios.
-
-We will be studying within-fragment co-methylation by analysing a large number of sparse $2 \times 2$ contingency tables. This puts us in the realm of asymptotic environment $A2$.
-Furthermore, the marginal probabilities of these tables, $\pi_{+ 1} = 1 - \pi_{+ 2}$ and $\pi_{1 +} = 1 - \pi_{2 +}$, will be very often close to the boundaries of zero and one.
-This is because in this case the marginal probability represents the probability that the methylation locus is methylated, and we have seen in Chapter \ref{chap:wgbs_statistical_framework} that these values, as estimated by $\beta$-values, have a bimodal distribution with each mode near the boundary.
-
-These extreme marginal probabilities give rise to sparse $2 \times 2$ tables.
-A property of these sparse $2 \times 2$ tables not yet discussed is that estimates of the odds ratio for these tables are very biased.
-We have seen this in __SIMULATION SETION__, but the intuitive explanation is that if both methylation loci are very likely to be methylated or unmethylated __BLAH BLAH BLAH__
+Here I describe a simple method to study within-fragment co-methylation by analysing methylation patterns at 2-tuples.
+For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type.
+This method is to be implemented in the `cometh()` function that is part of the `MethylationTuples` software[^cometh].
+
+[^cometh]: At the time of writing, this is available as a stand along `mantelhaen()` function, but it will ultimately one of several strategies of estimating within-fragment co-methylation using the `cometh()` function.
+
+The aim of this analysis is to address the questions:
+
+1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
+2. What factors influences this dependence?
 
@@
 
-Here I describe a simple method based on analysing methylation patterns at 2-tuples.
-For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type.
-This method is implemented in the `cometh()` function that is part of the `MethylationTuples` software.
-
-The aim of this analysis is to address the questions:
-
-1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
-2. What influences this dependence?
-
-There are two steps in this analysis:
-
-1. Define the 2-tuples.
-2. For each 2-tuple, estimates the dependence between the methylation state at the first locus and the second locus. For this I use the logarithm of the odds ratio ($LOR$).
-
-Step (2) may be further stratified by the genomic context of the pair of CpGs, such as whether they are in a CpG island, or other variables of interest.
-This is done in the same manner as for pairs of $\beta$-values (Section \ref{sec:pair_stratification}).
-
-A similar analysis could be performed using a different measure of dependence, which I discuss in Section \ref{sec:choice_of_dependence_measure}.
-
-\subsubsection{Constructing the 2-tuples}
+There are three important decisions to make in our analysis:
+
+1. How to define the 2-tuples?
+2. How to aggregate the 2-tuples?
+3. Which estimator to use?
+
+Answers to the latter two are informed by Section \ref{sec:sparse_2x2xK_tables}.
+
+\subsubsection{How to define the 2-tuples}
 
@@
 With current technology this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD \lessapprox 250$.
-
-Again, we can employ either the $NIL = 0$ or the $NIL \geq 0$ strategy.
-Both strategies start with CpGs in the reference genome that, for each sample, are then filtered to remove reference-specific loci.
-Rather than actually remove these loci, I simply set the corresponding counts of the methylation patterns to `NA`.
-
-As in Section \ref{subsec:beta_cor} I consider the $NIL = 0$ and $NIL \geq 0$ strategies for constructing pairs.
-The $NIL \geq 0$ strategy corresponds to using the `--all-combination` flag in `methtuple`.
-Again, the $NIL = 0$ strategy is simpler, both conceptually and computationally, and so I prefer it.
-For example, if we have a pair with $NIL = 3$ then we need to take care to ensure that this $LOR$ is due to a genuine dependence between these two CpGs above and beyond the pairwise dependencies $(CpG_{1}, CpG_{2}), \ldots, (CpG_{4}, CpG_{5})$.
-
-\FloatBarrier
-\subsubsection{Choice of dependence measure}\label{sec:choice_of_dependence_measure}
-
-As discussed, I use the logarithm of the odds ratio to summarise within-fragment co-methylation. Here I explain my reasons for choosing this measure of association.
-
-Recall that for each 2-tuple and sample we can summarise the methylation patterns by a $2 \times 2$ contingency table, like that shown in Table \ref{tab:2x2}.
-The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the logarithm of the odds ratio of this table, $LOR_{(i, i'), j} = \log_{2} \Big (\frac{mm_{i, i'} \times uu_{i, i'}}{mu_{i, i'} \times um_{i, i'}} \Big)$.
-I use base-2 logarithms due to their widespread use and familiarity within biology.
-In the `cometh` function I also add a small offset (default `offset = 1`) to avoid problems arising from zero counts, which frequently occur.
-For the example in Table \ref{tab:2x2} with `offset = 1` we have $LOR = \log_{2} \Big (\frac{(1 + 1) \times (4 + 1)}{(0 + 1) \times (4 + 1)} \Big) = 1.74$.
-
-From a Bayesian perspective, the offset may be interpreted as a prior count from a table with no association, i.e., a table with $LOR = 0$.
-The offset then has the effect of shrinking the raw estimate of the $LOR$, with more shrinkage applied for larger values of `offset`.
-Analyses using different offsets should be compared with caution.
-
-\begin{table}[h]
-\centering
-\begin{tabular}{lcccl}
-                           & \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
-                           &                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
-\multirow{2}{*}{First CpG} & \textbf{M}           & 1              & 2             & 3              \\
-                           & \textbf{U}           & 0              & 4             & 4              \\ \cline{2-5}
-                           & \textbf{Total}       & 1              & 6             & 7
-\end{tabular}
-\caption{Example $2 \times 2$ contingency table summarising methylation patterns at a 2-tuple.}
-\label{tab:2x2}
-\end{table}
-
-A positive $LOR$ is evidence that the methylation states of two CpGs on the same DNA fragment are likely to be the identical, a negative $LOR$ is evidence that the methylation states of two CpGs on the same DNA fragment are likely to be the different, and a value close to zero is evidence that the methylation states of two CpGs on the same DNA fragment are independent of one another.
-More precisely, we may wish to compute a confidence interval for the $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.
-
-There are several alternative measures of association for a $2 \times 2$ table of binary variables.
-An interesting review of these is given by \citet{Kraemer:2006te}.
-The Pearson correlation coefficient, also known as the phi coefficient when applied to binary $2 \times 2$ tables, was used by \citet{Landan:2012kp} to analyse within-fragment co-methylation.
-There are a few reasons that I prefer $LOR$ to Pearson's correlation.
-
-Firstly, the ``possible range of values of the [log] OR is not influenced by the marginal distributions of the variables involved'' \citep{Kraemer:2004el}.
-Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints \citep{plackett1981analysis}.
-In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities \citep{kraemer1986measure}, which I find undesirable.
-
-Secondly, the empirical distribution of $LOR$ for all 2-tuples in a sample  (Figure \ref{fig:EPISCOPE_lor_histogram}) is much nicer than that for Pearson's correlation coefficient (Figure \ref{fig:EPISCOPE_pearson_histogram}).
-Specifically, the distributions of $LOR$ for each sample are approximately unimodal and symmetric whereas the distributions of Pearson's correlation are multimodal, asymmetric, and clearly constrained by the underlying marginal probabilities.
-It is far easier to work with symmetric, unimodal distributions, such as when wanting to summarise these by measures of central tendency and spread, which is important in what follows.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_histogram.pdf}
-\caption{Histograms of $LOR$ for 2-tuples from the EPISCOPE data (binwidth $= 0.5$).}
-\label{fig:EPISCOPE_lor_histogram}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_histogram.pdf}
-\caption{Histograms of Pearson correlation coefficient for 2-tuples from the EPISCOPE data (binwidth $= 0.05$).}
-\label{fig:EPISCOPE_pearson_histogram}
-\end{figure}
-
-Finally, and related to the second point, there are $2 \times 2$ tables that are clear evidence of co-methylation that have a Pearson correlation close to zero.
-An example is shown in Table \ref{tab:2x2_big_lor_small_cor}.
-This table has a $LOR = 3.3$ and Pearson's correlation of $-0.025$, which is due to the table being dominated by one cell's count, $uu = 39$.
-This occurs for all tables where one cell dominates.
-The problem is only partially fixed by adding an offset.
-For example, adding $1$ to all cells gives a Pearson's correlation of $0.29$, which is still a low correlation.
-
-\begin{table}[h]
-\centering
-\begin{tabular}{lcccl}
-& \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
-&                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
-\multirow{2}{*}{First CpG} & \textbf{M}           & 0              & 1             & 1            \\
-& \textbf{U}           & 1              & 39             & 40             \\ \cline{2-5}
-& \textbf{Total}       & 1              & 40             & 41
-\end{tabular}
-\caption{A $2 \times 2$ contingency table with a large $LOR$ but small Pearson's correlation.}
-\label{tab:2x2_big_lor_small_cor}
-\end{table}
-
-\FloatBarrier
+As in Section \ref{subsec:beta_cor}, we will consider 2-tuples with $NIL = 0$ and $NIL \geq 0$.
+
+\subsubsection{How to aggregate the 2-tuples?}
+
+The counts of methylation patterns at each 2-tuple can be summarised by a $2 \times 2$ contingency table.
+The sequencing depth of each 2-tuple is typically low.
+Furthermore, for each locus in the 2-tuple, the marginal probability that it is methylated is very often close to zero or one.
+Therefore, in light of the results in Section \ref{subsec:simulation_study}, we must aggregate these $2 \times 2$ tables in order to perform any useful inference.
+
+The ideal level of aggregation combines those 2-tuples with homogeneous odds ratios.
+Of course, this information is unknown to us.
+Instead, we will aggregate by chromosome to explore the variation of co-methylation across chromosomes and compare these to genome-wide estimates.
+The resulting estimates must be interpreted as the 'average' degree of within-fragment co-methylation across the aggregation levels.
+
+\subsubsection{Which estimator to use?}
+
+We will use the Mantel-Haenszel odds ratio estimator since it is robust to variable marginal probabilities.
+This allows us to combine data for 2-tuples from different regions of the genome, e.g., regions that are hypomethylated, partially methylated and hypermethylated.
+
 \subsection{Results}
 
-For all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets I compute the $LOR$ of all 2-tuples using both the $NIL = 0$ and $NIL \geq 0$ strategies.
+I have analysed all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets.
+For each sample, I compute $\widehat{\theta_{MH}}$ using different levels of aggregation (listed here in decreasing order):
+
+1. $IPD$-only: A $2 \times 2 \times K$ table for each $IPD$.
+2.
+$IPD$-CGI: A $2 \times 2 \times K$ table for each $IPD$ and 'CpG island status' (CGI-status) combination.
+The CGI-status of each 2-tuple is whether it is _inside_ a CpG island or _outside_ a CpG island[^spanning].
+3. $IPD$-chromosome: A $2 \times 2 \times K$ table for each $IPD$ and chromosome combination.
+4. $IPD$-CGI-chromosome: A $2 \times 2 \times K$ table for each $IPD$, chromosome and CGI-status combination.
+
+[^spanning]: 2-tuples spanning the boundary of a CpG island are ignored.
+There are few such 2-tuples when $NIL = 0$.
+While there are considerably more such 2-tuples when $NIL \geq 0$, I have ignored these in favour of simplicity.
+
+No minimum sequencing coverage cutoff is required since the Mantel-Haenszel estimator appropriately downsweights 2-tuples with low sequencing coverage.
+
+We will first compare the results of chromosome-level estimates to genome-level estimates.
+
+\subsubsection{Chromosome-level analyses}
+
+__UP TO HERE: Plots should have $\widehat{\theta_{MH}}$, not $\psi$, on the y-axis__
+
+__FIGURES__ show the results for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+The estimated log odds ratio, $\widehat{\theta_{MH}}$ is plotted as a function of $IPD$.
+The genome-level estiamtes are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+
+#### These are the main points to write about
+
+- chromosome-level and genome-level qualitatively similar.
+- Chromosome-variation indicates uncaptured heterogeneity. Difficult to get at.
+- point out interesting chromosomes
+
+\subsubsection{Genome-wide analyses}
+
+- cgi vs. non-cgi
+
+\subsection{Summary}
+
+__OLD STUFF BELOW HERE__
+
 I compute these genome-wide and after stratifying CpG 2-tuples by whether they are in a CpG island.

### 2015-04-26 Rmarkdown/comethylation.Rmd
Continue writing co-methylation data analysis results. Remove stuff referring to old analysis based on odds ratios of 2x2 tables.

@@
 3. $IPD$-chromosome: A $2 \times 2 \times K$ table for each $IPD$ and chromosome combination.
-4. $IPD$-CGI-chromosome: A $2 \times 2 \times K$ table for each $IPD$, chromosome and CGI-status combination.
 
@@
 
-We will first compare the results of chromosome-level estimates to genome-level estimates.
-
-\subsubsection{Chromosome-level analyses}
-
-__UP TO HERE: Plots should have $\widehat{\theta_{MH}}$, not $\psi$, on the y-axis__
-
-__FIGURES__ show the results for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
-The estimated log odds ratio, $\widehat{\theta_{MH}}$ is plotted as a function of $IPD$.
-The genome-level estiamtes are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
-
-#### These are the main points to write about
-
-- chromosome-level and genome-level qualitatively similar.
-- Chromosome-variation indicates uncaptured heterogeneity. Difficult to get at.
-- point out interesting chromosomes
-
-\subsubsection{Genome-wide analyses}
-
-- cgi vs. non-cgi
+We will first compare the results of chromosome-level estimates to genome-level estimates[^chrY].
+
+[^chrY]: Some female samples have apparent Y chromosome data, e.g., all the _EPISCOPE_ samples. This is indicative of mapping errors and such data should be ignored.
+
+\subsubsection{Chromosome-level $NIL = 0$ analyses}
+
+Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using $NIL = 0$ 2-tuples for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+
+Unsurprisingly, there is variation in the chromosome-level estimates for a given $IPD$.
+Overall, however, we see that the genome-level and chromosome-level estimates follow a similar trend for all samples[^mitochondria]: co-methylation is strongest for smaller $IPD$s, decaying fairly rapidly before levelling out by $IPD = 20$ to $50$.
+As the $IPD$ increases, we see more variation in the chromosome-level and genome-level estimates, but this is due to the smaller sample sizes we have for $NIL = 0$ 2-tuples with larger $IPD$s.
+
+[^mitochondria]: The obvious outlier on each of these plots is the mitochondrial DNA.
+The mitochondria are typically almost completely unmethylated, i.e.
+the marginal probability of methylation at a CpG on the mitochondria is very close to zero, and so the concept of co-methylation is not particularly meaningful here.
+
+Notably, although the genome-level estimates decay as a function of $IPD$, $\widehat{\theta_{MH}}$ is almost entirely positive over the observable range of $IPD$s, suggesting that within-fragment co-methylation extends for at least a few hundred bp.
+Furthermore, in samples with sufficiently long paired-end reads, there is a hint of an upturn in the genome-wide $\widehat{\theta_{MH}}$ at approximately $IPD = 180$.
+This is approximately the distance between two CpGs on adjacent nucleosomes (see Section \ref{sec:nucleosome_and_chromatin}), and may be evidence of the three-dimensional structure of the genome affecting co-methylation.
+
+- __TODO: Figure out why E18BUF is missing chrM data__
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:EPISCOPE_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:Lister_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:Seisenberger_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:Ziller_merged_MH_by_seqnames}
+\end{figure}
+
+\FloatBarrier
+\subsubsection{Chromosome-level $NIL = 0$ analyses}
+
+The equivalent plots for the $NIL \geq 0$ 2-tuples paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_merged_ac_MH_by_seqnames}.
+In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_)
+
+- __TODO: Figure out what's going on with some of the EPISCOPE and Ziller samples that have a high autosome (chr21, I think)__
+
+In analysing $NIL \geq 0$ 2-tuples we are increasing the heterogeneity of the $K$ $2 \times 2$ tables, which makes more difficult the interpretation of these results.
+However, it does give the advantage of allowing the analysis of 2-tuples with larger $IPD$s.
+The short DNA fragments and read lengths of our data make it difficult to perform reliable chromosome-level inference in even the best samples for $IPD$s greater than 250$ to $300$.
+Retreating to a genome-level analysis suggests that within-fragment co-methylation is
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:EPISCOPE_ac_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:Lister_ac_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:Seisenberger_ac_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:Ziller_merged_ac_MH_by_seqnames}
+\end{figure}
+
+\FloatBarrier
+\subsubsection{The effect of CpG islands}
+
+Since we must aggregate 2-tuples in order to perform any meaningful analysis of within-fragment co-methylation, there will obviously be unaccounted-for heterogeneity in our estimates.
+In general, it will be difficult to identify sources of this heterogeneity.
+One obvious candidate, however, are CpG islands.
+
+We now compare estimates of within-fragment co-methylation within CpG islands and outside of CpG islands.
+Figures \ref{fig:EPISCOPE_MH_by_CGI}, \ref{fig:Lister_MH_by_CGI}, \ref{fig:Seisenberger_MH_by_CGI} and \ref{fig:Ziller_merged_MH_by_CGI} show the results for $NIL = 0$ 2-tuples and Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI} show the results for $NIL \geq 0$ 2-tuples.
+In order to simplify these figures, the chromosome-level estimates are not shown.
+We have seen that the genome-wide estimates capture the trend of the chromosome-level estimates, at least for the autosomes, and so these are sufficient for our purpose.
+
+__UP TO HERE: Discuss CGI results__
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:EPISCOPE_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:Lister_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:Seisenberger_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:Ziller_merged_MH_by_CGI}
+\end{figure}
+
+\subsubsection{Variation of within-fragment co-methylation between samples}
+
+__TODO: Do some sample comparisons, e.g., pluripotent vs. somatic, IMR90 decays superfast.__
 
@@
 
-__OLD STUFF BELOW HERE__
-
-I compute these genome-wide and after stratifying CpG 2-tuples by whether they are in a CpG island.
-These analyses are performed separately for each strand and also for strand-collapsed counts of methylation patterns.
-
-For any given $IPD$ there are hundreds of thousands of CpG pairs, and so there is a _distribution_ of co-methylation for any given $IPD$.
-For a given $IPD$ we can visualise the distribution of $LOR$ using a standard plot such as a histogram, kernel density estimate or boxplot.
-However, these $IPD$-specific plots don't give a sense of how $IPD$ affects within-fragment co-methylation.
-For this reason I instead plot selected quantiles --- $10\%, 25\%, 50\%, 75\% and 90\%$ --- of each $IPD$-specific $LOR$ distribution as a function of $IPD$.
-I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$[^quantile].
-These quantile plots may also be stratified by additional annotations such as genomic features.
-
-[^quantile]: Because the distributions of $LOR$ are approximately symmetric and unimodal, these quantiles give a fair summary of each of these distributions.
-The same would not be true of the highly asymmetric and multimodal distributions of Pearson's correlations, hence one reason for my preference of $LOR$ over Pearson's correlation.
-
-\FloatBarrier
-\subsubsection{Can we collapse by strand?}
-
-Figures \ref{fig:EPISCOPE_lor_min_cov_5}, \ref{fig:Lister_lor_min_cov_5}, \ref{fig:Seisenberger_lor_min_cov_5} and \ref{fig:Ziller_lor_min_cov_5} show the $LOR$ quantile plots for $NIL = 0$ CpGs 2-tuples stratified by strand.
-Since the distributions of $LOR$ are very similar for the two strands, I decided to use the strand-collapsed counts of methylation patterns in what follows.
-This is not to discount the possibility of 2-tuples with strand-specific differences in co-methylation, however, it does mean that it is generally safe to aggregate counts across strands.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_5.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the EPISCOPE dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:EPISCOPE_lor_min_cov_5}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_5.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the Lister dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Lister_lor_min_cov_5}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_5.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the Seisenberger dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Seisenberger_lor_min_cov_5}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_5.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-specific $NIL = 0$ CpG 2-tuples for samples from the Ziller dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Ziller_lor_min_cov_5}
-\end{figure}
-
-\FloatBarrier
-\subsubsection{$NIL = 0$}
-
-Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed} show the $LOR$ quantile plots for strand-collapsed $NIL = 0$ CpGs 2-tuples.
-Like for the correlations of $\beta$-values, two things immediately stand out.
-Firstly, for all samples the trend is that the strength of the within-fragment co-methylation decays as a function of $IPD$ --- 2-tuples with a larger $IPD$ are on average less co-methylated than2-tuples with a smaller $IPD$.
-Secondly, the estimated $LOR$s violently jump around for larger $IPD$s.
-The latter is simply due to the aforementioned lack of $NIL = 0$ pairs at larger $IPD$s with which to estimate these $LOR$s.
-Therefore, it is the trend in these plots that should be analysed.
-
-It is also obvious that longer paired-end reads are far more preferable to shorter single-end reads when studying within-fragment co-methylation.
-Even when using paired-end reads, getting to $IPD = 150$ with $NIL = 0$ pairs is challenging.
-This is unfortunate because it doesn't allow us to investigate the potential nucleosome-related $\approx 150-170$ bp periodicity of methylation levels using $NIL = 0$ pairs.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the EPISCOPE dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the Lister dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Lister_lor_min_cov_10_strand_collapsed}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the Seisenberger dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL = 0$ CpG 2-tuples for samples from the Ziller dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Ziller_lor_min_cov_10_strand_collapsed}
-\end{figure}
-
-The are some consistent patterns in comparing these quantile plots across samples.
-Those 2-tuples with a small $IPD$ have a median $LOR = 3-4$ meaning that the loci in these 2-tuples are $8-16\times$ more likely to have the same rather than opposite methylation states.
-However, we also see than for any given $IPD$ that there is considerable variation in $LOR$.
-For example, for all samples and $IPD$s, the $10\%$ and $90\%$ quantiles span $3-5$ (base 2) orders of magnitude.
-
-For each sample the $LOR$ distributions are consistently centred away from zero, meaning that over the observable $IPD$s the vast majority of 2-tuples are co-methylated.
-As $IPD$ increases, the CpG $NIL = 0$ 2-tuples become less co-methylated on average.
-However, the width of each sample's $LOR$ distributions remains relatively constant, so the variation in $LOR$ doesn't greatly depend on $IPD$.
+__OLD BUT POTENTIALLY USEFUL STUFF BELOW HERE__
 
@@
 
-The above analyses are all based on the genome-wide distributions of within-fragment co-methylation.
-Again, we can stratify these by whether the CpG 2-tuples are inside or outside of a CpG island.
-Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed_CGI} are plots of $LOR$ distributions for $NIL = 0$ CpG 2-tuples stratified by whether the pair is inside or outside of a CpG island[^spanning].
-Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.
-
-Again, the message from these plots is clear, $NIL = 0$ CpG 2-tuples inside CpG islands have higher levels of within-fragment co-methylation on average than those 2-tuples outside of islands[^hepg2_2].
-For each sample, the difference between the CpG island and non-island distributions is a vertical translation.
-For many samples this vertical shift means that the median $LOR$ for islands is approximately equal to the $75\%$ percentile for non-islands.
-The width of these distributions is not greatly affected by the CpG island status.
-
-[^hepg2_2]: Again, the minor exception is the _HepG2\_cell\_line_ where the CpG island and non-island distributions approximately overlap by $IPD = 50$, but we have already seen that this sample has an unusual methylome compared to the other samples.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the EPISCOPE dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the Lister dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Lister_lor_min_cov_10_strand_collapsed_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the Seisenberger dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed CpG 2-tuples stratified by CpG island status with $NIL = 0$ for samples from the Ziller dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Ziller_lor_min_cov_10_strand_collapsed_CGI}
-\end{figure}
-
-\FloatBarrier
-
-\subsubsection{$NIL \geq 0$}\label{sec:cometh_nil_geq0}
-
-Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_ac}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed_ac} show the $LOR$ quantile plots for strand-collapsed $NIL \geq 0$ CpGs 2-tuples.
-By using $NIL \geq 0$ 2-tuples we can explore the effect of within-fragment co-methylation at greater $IPD$s, potentially beyond the crucial $IPD = \approx 150-170$ bp distance.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed_ac.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the EPISCOPE dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed_ac.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the Lister dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Lister_lor_min_cov_10_strand_collapsed_ac}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed_ac.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the Seisenberger dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed_ac.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples for samples from the Ziller dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Ziller_lor_min_cov_10_strand_collapsed_ac}
-\end{figure}
-
-__TODO__: I need to re-write this section.
-I'm really not yet sure if this result is biology or an artefact.
-My instinct says artefact but I haven't yet been able to prove this either way to my satisfaction
-These plots appear to show that there is a dramatic increase in within-fragment co-methylation at precisely these $IPD$s for paired-end samples from the Lister dataset.
-However, I am very skeptical as to the biological relevance of these results.
-
-There are three reasons for my skepticism.
-Firstly, similar results are not observed for any of the EPISCOPE samples nor any of the Ziller samples[^mouse_vs_man].
-Secondly, the Lister samples with single-end sequencing have a similar spike in the estimated within-fragment co-methylation at the edge of the observable $IPD$s.
-However, for single-end data this corresponds to an $IPD \approx 60-80$ bp and we see no evidence of an increase in within-fragment co-methylation at these $IPD$s in any of the paired-end samples.
-This leads to my final reason, and what I think is the source of this artefact.
-Namely, estimates of within-fragment co-methylation are highly susceptible bias due to the incomplete removal of M-bias.
-This is particularly true for larger $IPD$s where the loci in the 2-tuples come from the ends of the reads, which are where M-bias is most prevalent.
-
-[^mouse_vs_man]: I exclude the Seisenberger data simply because these are of mice rather than humans.
-The Seisenberger data also display a dramatic increase in within-fragment co-methylattion but since this increase occurs at the edge of the observable $IPD$s, I am concerned it suffers from the same bias as the Lister data.
-
-As discussed in Section \ref{sec:m-bias}, the incomplete removal of M-bias is a real problem for reads that have undergone pre-trimming because this destroys the relationship between read-position and sequencing cycle.
-FIGURE shows that while there is considerable variation in the read lengths of the _ADS_ sample, which is indicative of aggressive pre-trimming of these reads, most reads are full length or close to it (TODO: Histogram of read lengths).
-
-To investigate my suspicion that these results are due to the incomplete removal of M-bias, I used an aggressive filtering strategy that attempts to only use read-positions that are unlikely to suffer from M-bias.
-I applied this to the _ADS_ sample.
-In this analysis I ignored read-positions $1-20$ and $55-80$ of both _read 1_ and _read 2_.
-Unfortunately, this is a situation where the M-bias filtering strategy of `methtuple` is a disadvantage.
-`methtuple` is designed to filter out specific read-positions and cannot 'exclude the first/last $n$ bases'.
-For example, using this aggressive strategy will have a detrimental effect for a read that has been pre-trimmed to $30$ bp, of which the last $10$ bp are affected by M-bias, because this filtering strategy will _retain_ read-positions $21-30$, which are exactly those affected by M-bias.
-Nevertheless, it is the best that can be done with the current version of the software[^methtuple_refactoring] and with such aggressively pre-trimmed data.
-
-[methtuple_refactoring]: We have now seen that both the 'filter specific read-positions' and 'filter the first/last $n$ read-positions' can lead to undesirable behaviour and so switching strategies is no saviour.
-The real solution is not to pre-trim reads and to instead use an aligner such as `bwa-meth` that can softclip these positions.
-Unfortunately, `bwa-meth` was only available after my data were already aligned and after `methtuple` was written.
-To use `methtuple` with `bwa-meth` output requires significant changes to `methtuple` because the `bwa-meth` output does not include anything analogous to the XM-tag of `Bismark`, which `methtuple` uses to do methylation calling.
-
-The results for the aggressively trimmed _ADS_ sample are shown in FIGURE.
-
-- TODO: Comment on these results and how we should interpret the Lister co-methylation results (particularly NIL >= 0)
-- TODO: Summarise the M-bias issue
-__
-
-Comparing CpG islands to non-islands we again, Figures \ref{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac_CGI}, \ref{fig:Lister_lor_min_cov_10_strand_collapsed_ac_CGI}, \ref{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac_CGI} and \ref{fig:Ziller_lor_min_cov_10_strand_collapsed_ac_CGI}, see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the EPISCOPE dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:EPISCOPE_lor_min_cov_10_strand_collapsed_ac_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the Lister dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Lister_lor_min_cov_10_strand_collapsed_ac_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the Seisenberger dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Seisenberger_lor_min_cov_10_strand_collapsed_ac_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_min_cov_10_strand_collapsed_ac_CGI.pdf}
-\caption{$LOR$ as a function of $IPD$ for strand-collapsed $NIL \geq 0$ CpG 2-tuples stratified by CpG island status for samples from the Ziller dataset.
-The dark line plots the median of these distributions, the inner (darker) band the $25\% - 75\%$ percentiles, and the outer (lighter) bands the $10\% - 25\%$ and $75\% - 90\%$ percentiles.
-Both the x-axis and y-axis have been truncated for clarity.}
-\label{fig:Ziller_lor_min_cov_10_strand_collapsed_ac_CGI}
-\end{figure}
-
-\FloatBarrier
 \section{Summary}
@@
 
-# Problem with LOR estimates
-
-The simple estimator of within-fragment co-methylation is biased when the marginal probability of methylation is close to the boundary (0 or 1) unless the 2-tuple is sequenced at very high depth.
-I demonstrate this by simulating methylation patterns at 2-tuples as independent $binomail(1, p)$ random variables and showing that the simple LOR estimator is positively biased.
-
-## Following chat with Terry
-
-These are my notes following a Skype conversion with Terry to discuss the issue.
-
-- [ ] Try a higher order Taylor expansion of the log odds-ratio to get e.g., $n^{1}$ or $n^{3/2}$ convergence.
-The $+\frac{1}{2}$ fudge-factor is motivated by such an expansion and perhaps it is possible to do better by a higher order expansion.
-  - [ ] First try showing the $c = \frac{1}{2}$ is optimal.
-- [ ] Read texts about this issue
-- [ ] Look for research papers on the general issue of dealing with bias in log odds-ratio estimates
-- [ ] What if I just increase the fudge-factor to increase the shrinkage/smoothing of the estimates? It'll shrink to zero alright, but what is the theoretical justification?
-- [ ] Check out the behaviour of the $\theta$ coefficient, i.e., Pearson's $r$ of the $2 \times 2$ table.
-  - Landan et al.: "Note that correlation values are subject to considerable noise when coverage is low or when m1 or m2 are close to 0 or 1."
-    - They don't define whether 'noise' means bias, variance, skewness, etc.
-
 ## Miscellaneous notes
@@
 
-- [ ] Demonstrate via simulation that the unconditional LOR estimator applied to $2 \times 2$ tables is biased when $d$ is small and the marginal probabilities are near the boundaries.
-  - Corrollary: Don't use the LOR estimates of within-fragment co-methylation from individual $2\times 2$ tables. I know they're biased, so there's no reason to show these results.
-- [ ] Demonstrate that the MH estimator applied to $2 \times 2 \times K$ tables is unbiased when $d$ is small (per-table) and the marginal probabilities are near the boundaries.
-  - Need to investigate the setting where the marginal probabilities are heterogeneous across the $K$ tables but with the same odds-ratio.
-  - Need to investigate the setting where the odds ratio __is not__ common to the $K$ tables; what does the MH estimator end up estimating?
-
 - Run Woolf's test or Breslow-Day test of homogeneity of odds ratios (noting that these are not powerful tests).
@@
 I have $K \\g 10$ ($K$ = number of $2 \times 2$ tables) and $d \geq 10$ ($d$ = sequencing coverage), so the unconditional test may well be fine.
-
-__TODO: Look up Liang-Self in Agresti's CDA (2013)__
 
@@
   - Discuss results.
-- `methsim`: Sample LORs from a Normal(mu, sigma^2) distribution, where mu (resp.
+- `methsim`: Sample LORs from a Normal($\mu$, $\sigma^2$) distribution, where mu (resp.
 sigma) is the MH estimator (resp.

### 2015-04-26 Rmarkdown/comethylation.Rmd
Co-methylation CGI results, begin between sample variation

@@
 In general, it will be difficult to identify sources of this heterogeneity.
-One obvious candidate, however, are CpG islands.
-
-We now compare estimates of within-fragment co-methylation within CpG islands and outside of CpG islands.
-Figures \ref{fig:EPISCOPE_MH_by_CGI}, \ref{fig:Lister_MH_by_CGI}, \ref{fig:Seisenberger_MH_by_CGI} and \ref{fig:Ziller_merged_MH_by_CGI} show the results for $NIL = 0$ 2-tuples and Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI} show the results for $NIL \geq 0$ 2-tuples.
-In order to simplify these figures, the chromosome-level estimates are not shown.
+One obvious candidate, however, are CpG islands and so we compare estimates of within-fragment co-methylation within CpG islands and outside of CpG islands.
+In order to simplify the figures for these analyses, the chromosome-level estimates are not shown.
 We have seen that the genome-wide estimates capture the trend of the chromosome-level estimates, at least for the autosomes, and so these are sufficient for our purpose.
 
-__UP TO HERE: Discuss CGI results__
+We observe that there is very little difference between genome-wide estimates of  $\theta$ inside and outside of CpG islands for 2-tuples with $NIL = 0$ (Figures \ref{fig:EPISCOPE_MH_by_CGI}, \ref{fig:Lister_MH_by_CGI}, \ref{fig:Seisenberger_MH_by_CGI}).
+This is perhaps a little surprising; CpG islands are typically more homogeneous in their average methylation levels and so we would reasonably expect their within-fragment methylation states are more dependent than elsewhere in the genome.
+However, the $NIL = 0$ results do not tell the full story.
+
+When we analyse $NIL = 0$ 2-tuples, we are only estimating the 'first-order' within-fragment co-methylation.
+But suppose that the methylation state of the current locus ($Z_{h, i}$) depends not only on the state at the previous locus ($Z_{h, i - 2}$) but also on the states at the two previous states ($Z_{h, i - 2}$ and $Z_{h, i - 3}$).
+Or it might depend on the states at both upstream and downstream loci (i.e.
+$Z_{h, i + i'}$ and $Z_{i - i''}$, $i', i'' > 0$).
+To summarise, the $NIL = 0$ analyses reduce heterogeneity by focusing on adjacent loci but cannot (by themselves) tell us whether within-fragment co-methylation is truly 'first-order'.
 
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:EPISCOPE_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:Lister_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:Seisenberger_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
+\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
+Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\label{fig:Ziller_merged_MH_by_CGI}
+\end{figure}
+
+To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of $NIL \geq 0$ 2-tuples (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}).
+We now see a fairly clear separation of the genome-wide estimates of $\theta$, with CpG islands being more co-methylated than the rest of the genome.
+Moreover, the estimates of $\theta$ within CpG islands are approximately constant with respect to $IPD$.
+Also notable is that the 2-tuples outside of CpG islands have fairly similar estimates of $\theta$ in both the $NIL = 0$ and $NIL \geq 0$ analyses.
+
+Taken together, these results suggests that within-fragment co-methylation in CpG islands depends on multiple loci and that the genomic distance between these loci is of little direct importance[^cpg_density1].
+In contrast, within-fragment co-methylation outside of CpG islands depends mostly on the adjacent locus and the distance to that locus ($IPD$)[^cpg_density2].
+We might try to estimate the order of within-fragment co-methylation in CpG islands by analysing 2-tuples with $NIL = a, a = 1, 2, \ldots$.
+This quickly becomes complicated, however, since there are $2^{a - 1}$ odds ratios to estimate for $NIL = a$ 2-tuples.
+We have not yet pursued refinement.
+
+[^cpg_density1]: The overall density of methylation loci is likely important, which of related to $IPD$s.
+
+[^cpg_density2]: The ratio $\frac{NIL > 0 \text{2-tuples}}{NIL \geq 0 \text{2-tuples}}$ is lower outside of CpG islands than inside CpG islands owing to the decreased density of CpGs outside of CpG islands.
+We therefore can't rule out that the similarity of $NIL = 0$ and $NIL \geq 0$ estimates outside of CpG islands is simply due to a lack of data.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_CGI.pdf}
 \caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
-\label{fig:EPISCOPE_MH_by_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI.pdf}
+\label{fig:EPISCOPE_ac_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_CGI.pdf}
 \caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
-\label{fig:Lister_MH_by_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI.pdf}
+\label{fig:Lister_ac_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_CGI.pdf}
 \caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
-\label{fig:Seisenberger_MH_by_CGI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
+\label{fig:Seisenberger_ac_MH_by_CGI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_ac_merged_MH_by_CGI.pdf}
 \caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
-\label{fig:Ziller_merged_MH_by_CGI}
+\label{fig:Ziller_ac_merged_MH_by_CGI}
 \end{figure}
@@
 
-__TODO: Do some sample comparisons, e.g., pluripotent vs. somatic, IMR90 decays superfast.__
+The analyses so far have highlighted the broad similarities of within-fragment co-methylation between a diverse set of samples.
+Reasuringly, samples done in technical duplicate (_IMR90\_r1_ and _IMR90\_r2_; _H1\_r1_ and _H1\_r2_) have very similar genome-level estimates of $\theta$.
+Furthermore, biological samples assayed by different laboratores (_H9_ and _H9\_Laurent_; _IMR90\_r1, _IMR90\_r2_ and _IMR90\_cell\_line_) also have fairly similar genome-level estimates of $\theta$[^imr90_ac].
+
+[^imr90_ac]: Differences in the $NIL \geq 0$ results between the _IMR90\_r1, _IMR90\_r2_ and _IMR90\_cell\_line_ samples are due to the _IMR90\_cell\_line_ being sequenced with paired-end reads and therefore having many more $NIL > 0$ 2-tuples to analyse.
+
+There are of course differences, and we now turn our attention to these.
+The results of this section should be interpreted with some caution: few samples have replicates and so we will have little evidence to conclude what is driving any observed variation nor do we have sufficient power to definitely rule out large differences in co-methylation.
+Nonetheless, I will offer some hypotheses and support for these as appropriate.
+
+The _EPISCOPE_ dataset is the best dataset for studying between sample variation of within-fragment co-methylation.
+It has a fully crossed design of donors and tissues.
+The genome-level estimates of $\theta$ are similar across all 12 samples.
+Visually, the variation between tissues is greater than that between donors (the plots are more similar by column than by row).
+This is consistent with average methylation levels being more similar within tissues than within donors (__TODO: Cite__).
+
+Turning to the _Ziller_ dataset, there is little difference between the genome-level estimates of $\theta$ for _Colon\_Normal\_Primary_ and _Colon\_Tumour\_Primary_ nor between the various frontal cortex samples.
+This does not exclude the possibility that there exist more local differences in co-methylation but there is no evidence that within-fragment co-methylation is globally affected by the disease status in each case.
+
+The most interesting differences between samples occur in the _Lister_ dataset.
+Recall that the _Lister_ dataset contains four 'mini dataset': _Lister-ADS_, _Lister-FF_, _Lister-IMR90_ and _Lister-H1_ (see section \ref{sec:Lister_dataset}).
+In each mini datasets, the pluripotent cell lines (induced pluripotent stem cells or embryonic stem cells) have larger genome-level estimates of $\theta$ than do their precursor and differentiated forms.
+This suggests the pluripotent cell cells have stronger within-fragment co-methylation than do somatic or differentiated cells.
+
+__UP TO HERE: Comment on other pluripotent samples (H9, Seisenbergers); comments on IMR90 (decays superfast).__
 

### 2015-04-27 Rmarkdown/comethylation.Rmd
Complete revamped draft of co-methylation chapter.

@@
 
-Here I describe two simple methods based on correlations of $\beta$-values. Both of these are implemented in the `methLevelCor` function that is part of the `MethylationTuples` software.
+Here I describe two simple methods based on correlations of $\beta$-values. Both of these are implemented in the `methLevelCor()` function that is part of the `MethylationTuples` software.
 
@@
 
-Step (2) may be further stratified by the genomic context of the pair of CpGs, such as whether they are in a CpG island, or other variables of interest.
-
-The same analysis could be performed using $\mathcal{M}$-values rather than $\beta$-values, and this is available as an option in the `methLevelCor` function.
+The second step may be further stratified by the genomic context of the pair of CpGs, such as whether they are in a CpG island, or other variables of interest.
+
+The same analysis could be performed using $\mathcal{M}$-values rather than $\beta$-values, and this is available as an option in the `methLevelCor()` function.
 
@@
 The first strategy creates all pairs of adjacent CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context.
-For a chromosome with $N$ CpGs there are $(N - 1)$ pairs.
-I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have no intervening loci ($NIL$ = number of intervening loci)^[nil0].
+For a chromosome with $N_{loci}$ CpGs there are $(N_{loci} - 1)$ pairs.
+I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have no intervening loci ($NIL$ = number of intervening loci)[^nil0].
 
@@
 The interpretation of correlations computed using the $NIL \geq 0$ is complicated because the set of pairs are more heterogeneous than those under the $NIL = 0$ strategy.
-For this reason I prefer the $NIL = 0$ strategy, however, it is useful to consider and contrast the two strategies, which is what I have done in my thesis.
+For this reason I prefer the $NIL = 0$ strategy, however, it is useful to consider and contrast the two strategies, which is what I have done in what follows.
 
@@
 Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two regions of the genome that have very different methylation dynamics.
-In other words, the vector of $\beta$-values for a given sample is highly non-stationary --- the correlation between two loci depends on more than just the $IPD$.
+In other words, the vector of $\beta$-values for a given sample is highly non-stationary; the correlation between two loci depends on more than just the $IPD$.
 An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values.
@@
 Figure \ref{fig:feature_status} illustrates the 'feature status' of some pairs of loci.
-Note that a pair where each loci is in a different CpG island is declared to be inside the feature, i.e., elements need not be in the same feature but rather in the same type of feature.
+Note that a pair where each locus is in a distinct feature from the same class is declared to be inside the feature, i.e.
+elements need not be in the same feature but rather in the same type of feature.
 The genomic feature should partition the genome so that each loci is either inside or outside of the feature.
@@
 The circles represent loci and the grey bars a genomic feature, such as CpG islands.
-Each pair's `feature status' is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside).
+Each pair's feature status is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside).
 Note that the rightmost pair is `inside' even though the two loci that comprise the pair are in different elements of the genomic feature.}
@@
 Owing to the size of whole-genome bisulfite-sequencing data, I have chosen to only compute Pearson's $r$ and Spearman's $\rho$, which are much faster to compute than Kendall's $\tau$.
-The `methLevelCor` function in the `MethylationTuples` will compute a confidence interval, $95\%$ by default, for the Pearson correlation.
+The `methLevelCor()` function in the `MethylationTuples` package will compute a confidence interval, $95\%$ by default, for the Pearson correlation.
 This feature is not yet available when using Spearman's or Kendall's correlation coefficient.
@@
 
-Firstly, the bimodality of the distribution of $\beta$-values means that the points concentrate in the corners of the plot --- $(0, 0), (0, 1), (1, 0)$ and $(1, 1)$.
+Firstly, the bimodality of the distribution of $\beta$-values means that the points concentrate in the corners of the plot: $(0, 0), (0, 1), (1, 0)$ and $(1, 1)$.
 A correlation coefficient is not well suited to summarising these distributions of points.
@@
 
-In summary, the correlations of pairs of $\beta$-values should be cautiously interpreted, bearing in mind that these are rather crude summaries of the underyling scatterplots that hide many details.
+In summary, the correlations of pairs of $\beta$-values should be cautiously interpreted, bearing in mind that these are somewhat crude summaries that hide many details of the underyling scatterplots.
 
@@
 The Spearman correlation of each scatterplot is also reported.
-Individual plots created using the \texttt{smoothScatter} function in R.}
+Individual plots created using the \texttt{smoothScatter()} function in R.}
 \label{fig:ADS_beta_scatterplots}
@@
 
-Figures \ref{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} compares the Pearson correlation coefficient to the Spearman correlation coefficient.
+Figures \ref{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed} compare the Pearson correlation coefficient to the Spearman correlation coefficient.
 The two correlation coefficient give qualitatively similar results, however, the Spearman correlation is consistently smaller in magnitude than the Pearson correlation.
@@
 \FloatBarrier
-\subsubsection{$NIL = 0$}
+\subsection{$NIL = 0$ CpG pairs}
 
@@
 Two things immediately stand out.
-Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$ --- pairs of CpGs separated larger distances are on average less correlated than pairs of CpGs separated by shorter distances.
+Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$; pairs of CpGs separated larger distances are on average less correlated than pairs of CpGs separated by shorter distances.
 Secondly, the estimated correlations violently jump around for larger $IPD$s.
@@
 The are some consistent patterns in comparing these trends across samples.
-The correlations very quickly drop well below $0.5$, within approximately $50$ bp for all the EPISCOPE and Seisenberger samples and most of the Lister and Ziller samples.
-Samples that are exceptions to this trend --- _ADS_, _ADS-adipose_, _ADS-iPSC_, _FF_, _IMR90\_r1_, _IMR90\_r2_, _IMR90\_cell\_line_ and _HepG2\_cell\_line_ --- have a slower decay in these correlations and remain more highly correlated over $IPD = 2, \ldots, 1500$.
+The correlations very quickly drop well below $0.5$, within approximately $50$ bp for all the _EPISCOPE_ and _Seisenberger_ samples and most of the _Lister_ and _Ziller_ samples.
+Samples that are exceptions to this trend (_ADS_, _ADS-adipose_, _ADS-iPSC_, _FF_, _IMR90\_r1_, _IMR90\_r2_, _IMR90\_cell\_line_ and _HepG2\_cell\_line_) have a slower decay in these correlations and remain more highly correlated over $IPD = 2, \ldots, 1500$.
 These same samples are also notable because they have significant amounts of intermediate methylation (Section \ref{sec:genome-wide_distribution_of_beta_values}).
-I believe this is what drives the stronger and slowly decaying correlations.
+I believe this is what drives the stronger and more slowly decaying correlations.
 The scatterplots of $\beta$-values for these samples will have more $\beta$-values in the middle of the scatterplot, and fewer in the corners, hence the stronger correlations.
@@
 One final aspect of these genome-wide plots bears mentioning.
-There are hints of an approximately $150-200$ bp periodicity in these plots.
+There are hints of an approximately $150$ to $200$ bp periodicity in these plots.
 This is a similar scale to the previously reported periodicities in correlations of $\beta$-values (see Chapter \ref{chap:co-methylation_review}).
@@
 \FloatBarrier
-\subsubsection{$NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}
+\subsection{$NIL \geq 0$ CpG pairs}\label{sec:nil_geq_0_beta_cors}
 
@@
 
-Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].
+Comparing CpG islands to non-islands, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2] (Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}).
 
@@
 
-The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands.
+The $150$ to $200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands.
 This may be due to  different regulation of DNA methylation outside of CpG islands or simply because CpG islands have such consistently high correlations that these periodicities are not evident.
 
-\subsection{Summary}
-
-__TODO: Summarise the results for correlation of $\beta$-values.__
-
-A limitation to analysing co-methylation using correlations of $\beta$-values is that the analysis is based on correlations of measurements that are aggregates from many thousands of DNA molecules.
+\subsection{Limitations}
+
+A limitation in to these analyses is they are based on correlations of measurements that are aggregates from many thousands of DNA molecules.
 We are not measuring co-methylation on the scale at which the biological process of DNA methylation acts, namely at the level of individual DNA molecules.
 As noted by \citet{Ball:2009kh}, "the clonal feature of Illumina sequencing [allows us] to investigate whether co-methylation occurs at the single-molecule level".
-Before doing so, I first explore the challenge of estimating meaningful measures of dependence in sparse $2 \times 2 \times K$ contingency tables.
+We explore this topic in Section \ref{sec:wf_cometh}, but before doing so we first must first discuss the challenge of estimating meaningful measures of dependence in sparse $2 \times 2 \times K$ contingency tables.
 While perhaps not yet apparent, this will be essential for the analysis of within-fragment co-methylation.
@@
 
-We will begin with an overview of the simplest case of estimating dependence in $2 \times 2$ contingency tables, i.e., $K = 1$.
+We will begin with an overview of the simplest case, estimating dependence in a $2 \times 2$ contingency table ($K = 1$).
 We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables.
-We conclude with a simulation study highlighting the performance of several difference estimators and determining which is most appropriate for a study of within-fragment co-methylation.
+We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determining which is most appropriate for a study of within-fragment co-methylation.
 
@@
 Consider two binary random variables, $X$ and $Y$.
-Let $\pi_{ij}, i, j = 1, 2$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$.
+Let $\pi_{ij}$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$ ($i, j, = 1, 2$).
 We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$.
-There are many reasons why we might wish to estimate the odds ratio (__CITE__), but our interest here is __WHAT__.
-Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:
-
-__TODO: Some reasons to use the odds ratio.
-Firstly, the ``possible range of values of the [log] OR is not influenced by the marginal distributions of the variables involved'' \citep{Kraemer:2004el}.
-Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints \citep{plackett1981analysis}.__
+The odds is a convenient summary of dependence in a $2 \times 2$ table.Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:
 
@@
 
-The possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios.
+The odds ratio is not perfece, however.
+For one, the possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios.
 Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$.
@@
 Suppose we observe $(X, Y)$ for $n$ samples.
-Denote by $n_{ij}, i, j = 1, 2$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$.
-We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$.
+Denote by $n_{ij} the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$.
+We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$.
 The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.
@@
  & \multicolumn{3}{c}{Column} \\
- \cmidrule(l){2-4}
 \multirow{3}{*}{Row} & & & \\
@@
 The asymptotic standard error of $\widehat{\theta_{U}}$ is given by $\hat{\sigma}(\widehat{\theta_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
-The asymptotic distribution of $\widehat{\theta_{U}}$ as $n \to \infty$ is $Normal(\theta, \hat{\sigma}(\theta)^{2})$.
+The asymptotic distribution of $\widehat{\theta_{U}}$ as $n \to \infty$ is $Gaussian(\theta, \hat{\sigma}(\theta)^{2})$.
 
@@
 Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}.
-The maximum likelihood estimator of $\psi$, $\widehat{psi_{C}}$ is solved using iterative methods.
-
-The conditional estimator, $\widehat{\theta_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+The maximum likelihood estimator of $\psi$, $\widehat{\psi_{C}}$ is solved using iterative methods.
+The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi_{C}}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+
 Estimation of the odds ratio is made more difficult when the contingency table is _sparse_.
@@
 
-__TODO: Is the unconditional maximum likelihood estimator the same as `LORCollapsed()`?__
-
 Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$.
@@
 Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$.
-The disadvantage of these conditional tests is the extra computation required to compute them.
+The disadvantage of these conditional tests is the extra computation required.
 \citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.
 
-- __TODO: Comment on the utility (or lack thereof) of homogeneity tests; i.e., rejecting the null is good evidence the null is wrong but failing to reject the null is not strong evidence that it is not wrong.__
+- __TODO: Find the reference (I think it's Agresti) for the discussion on the utility (or lack thereof) of homogeneity tests; i.e., rejecting the null is good evidence the null is wrong but failing to reject the null is not strong evidence that it is not wrong.__
 
@@
 - the marginal probabilities, $\pi_{2 +} = Pr(Z_{i})$ and $\pi_{+ 2} = Pr(Z_{i'})$.
-Some simulations use a common set of marginal probabilities across all tables but some simulate the marginal probabilities from a specified distribution.
+Some simulations use a common set of marginal probabilities across all tables but others simulate the marginal probabilities from a specified distribution.
 - the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
@@
 Most simulations use a truncated negative binomial distribution to model the sequencing depth.
-The truncated negative binomial distribution is parameterised by the mean ($mu$), the dispersion parameter (size) and the truncation level (trunc).
+The truncated negative binomial distribution is parameterised by the mean ($\mu$), the dispersion parameter (size) and the truncation level (trunc).
 Unless otherwise noted, simulations use $\mu = 30$, $\text{size} = 10$ and $\text{trunc} = 9$, corresponding to an average sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].
@@
 
-To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1, Figure \ref{fig:Simulation_1}), and Simulation 2a, $K = 1000$ (Figure \ref{fig:Simulation_2}).
-Under these parameters, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta_{MH}}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
+To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1, Figure \ref{fig:Simulation_1}), and $K = 1000$ (Simulation 2a, Figure \ref{fig:Simulation_2}).
+Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta_{MH}}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
 This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller.
@@
 
-Next we explore what happens when we allow the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, to vary across the $K$ tables when $\theta$ is a fixed value (not necessarily zero) identically across the $K$ tables for $K = 100$.
+Next we explore what happens when we allow the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, to vary across the $K$ tables when $\theta$ is a fixed value identically across the $K$ tables for $K = 100$ ($\theta$ not necessarily zero).
 To do so, we simulate $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ for each $2 \times 2$ table.
@@
 In a sense, the $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ assumption is too easy.
-As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are mostly near the boundaries, not uniformly distributed.
+As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are mostly near the boundaries, not uniformly distributed on the unit interval.
 Furthermore, each methylation locus in a 2-tuple will likely have a similar marginal probability of methylation.
@@
 The results of this simulation (Simulation 4, Figure \ref{fig:Simulation_4}) are basically an exaggerated version of Simulation 3.
-The Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, remains an excellent estimator.
-The unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$.
-The distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.
+The Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.
 
@@
 We now explore what happens when the odds ratio is heterogeneous across the $K$ tables.
-Although results are shown for $\widehat{\theta_{0.5}}$ and $\widehat{\theta_{U}}$, we do not discuss these estimators further since they perform poorly in even the simpler scenarios.
+Although results are shown for $\widehat{\theta_{0.5}}$ and $\widehat{\theta_{U}}$, we do not discuss these estimators further since we have seen that these perform poorly in even the simplest simulation scenarios.
 
@@
 - Simulation 5 and Simulation 6: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
-- Simulation 7:$\theta_{k} \eqd Skewed-Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
-- Simulation 8: $\theta_{k} \eqd 0.5 Gaussian(- \theta_{0}, 1) + 0.5 Gaussian(\theta_{0}, 1)$.
+- Simulation 7:$\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 8: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$.
 
 Simulations 5, 6 and 7 (Figures \ref{fig:Simulation_5}, \ref{fig:Simulation_6} and \ref{fig:Simulation_7}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, is a reasonable estimator of the centre of the distribution of log odds ratios.
-It performs better in Simulation 5, where the distribution of $\theta$ is symmetric, as opposite to Simulation 7, where the distribution of $\theta$ is skewed.
+It performs better in Simulation 5 and 6, where the distribution of $\theta$ is symmetric, as opposed to Simulation 7, where the distribution of $\theta$ is skewed.
 Unsurprisingly, $\widehat{\theta_{MH}}$ does a better job as $K \to \infty$.
-It should be emphasised, however, that the asymptotic variance of the estimator, $hat{\sigma}(\widehat{\theta_{MH}})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$.
+It should be emphasised, however, that the asymptotic variance of the estimator, $\hat{\sigma}(\widehat{\theta_{MH}})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$.
 We can see that the asymptotic $95 \%$ confidence interval of $\widehat{\theta_{MH}}$ only covers a small amount of the variation in the $\theta$.
@@
 
-This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual 2-tuples using the simple estimator $\theta_{0.5}$ unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
+This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual 2-tuples using the simple estimator odds ratio estimator, $\theta_{0.5}$, unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
 The distributions of $\widehat{\theta_{0.5}}$ at typical sequencing depths ($d \approx 30$) are highly biased and dispersed, even under very simple simulation scenarios.
@@
 
-[^ultra-high_depth]: Ultra-deep sequencing has its own problems such as an increased effect of PCR-bias.
-
-Suppose we aggregate $K$ 2-tuples to form a $2 \times 2 \times K$ contingency table.
+[^ultra-high_depth]: Ultra-deep sequencing has its own problems, such as an increased effect of PCR-bias.
+
+Suppose that we aggregate $K$ 2-tuples to form a $2 \times 2 \times K$ contingency table.
 When little can be assumed about this $2 \times 2 \times K$ contingency table, the Mantel-Haenszel estimator is the most appropriate of those considered.
@@
 
-[^cometh]: At the time of writing, this is available as a stand along `mantelhaen()` function, but it will ultimately one of several strategies of estimating within-fragment co-methylation using the `cometh()` function.
+[^cometh]: At the time of writing, this is available as a stand along `mantelhaen()` function, but it will soon be one of several strategies of estimating within-fragment co-methylation using the `cometh()` function.
 
@@
 However, we are now limited to analysing co-methylation of loci that can be captured within a single read.
-With current technology this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD \lessapprox 250$.
+With current technology this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD$ less than approximately $250$.
 As in Section \ref{subsec:beta_cor}, we will consider 2-tuples with $NIL = 0$ and $NIL \geq 0$.
@@
 Of course, this information is unknown to us.
-Instead, we will aggregate by chromosome to explore the variation of co-methylation across chromosomes and compare these to genome-wide estimates.
+Instead, we will aggregate by chromosome to explore the variation of co-methylation across chromosomes and compare these to genome-level estimates.
 The resulting estimates must be interpreted as the 'average' degree of within-fragment co-methylation across the aggregation levels.
@@
 
-Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using $NIL = 0$ 2-tuples for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using $NIL = 0$ CpG 2-tuples for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
 
@@
 Overall, however, we see that the genome-level and chromosome-level estimates follow a similar trend for all samples[^mitochondria]: co-methylation is strongest for smaller $IPD$s, decaying fairly rapidly before levelling out by $IPD = 20$ to $50$.
-As the $IPD$ increases, we see more variation in the chromosome-level and genome-level estimates, but this is due to the smaller sample sizes we have for $NIL = 0$ 2-tuples with larger $IPD$s.
+As the $IPD$ increases, we see more variation in the chromosome-level and genome-level estimates, but this is due to the smaller sample sizes we have for 2-tuples with $NIL = 0$ at larger $IPD$s.
 
@@
 This is approximately the distance between two CpGs on adjacent nucleosomes (see Section \ref{sec:nucleosome_and_chromatin}), and may be evidence of the three-dimensional structure of the genome affecting co-methylation.
-
-- __TODO: Figure out why E18BUF is missing chrM data__
 
@@
 \caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+\textbf{TODO: Figure out why E18BUF is missing chrM data}}
 \label{fig:EPISCOPE_ac_MH_by_seqnames}
@@
 In general, it will be difficult to identify sources of this heterogeneity.
-One obvious candidate, however, are CpG islands and so we compare estimates of within-fragment co-methylation within CpG islands and outside of CpG islands.
+One obvious candidate, however, are CpG islands and so we compare estimates of within-fragment co-methylation inside CpG islands and outside of CpG islands.
 In order to simplify the figures for these analyses, the chromosome-level estimates are not shown.
@@
 
-When we analyse $NIL = 0$ 2-tuples, we are only estimating the 'first-order' within-fragment co-methylation.
-But suppose that the methylation state of the current locus ($Z_{h, i}$) depends not only on the state at the previous locus ($Z_{h, i - 2}$) but also on the states at the two previous states ($Z_{h, i - 2}$ and $Z_{h, i - 3}$).
+When we analyse 2-tuples with $NIL = 0$, we are only estimating the 'first-order' within-fragment co-methylation.
+But suppose that the methylation state of the current locus ($Z_{h, i}$) depends not only on the state at the previous locus ($Z_{h, i - 1}$) but also on the states at the two previous states ($Z_{h, i - 2}$ and $Z_{h, i - 3}$).
 Or it might depend on the states at both upstream and downstream loci (i.e.
@@
 
-To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of $NIL \geq 0$ 2-tuples (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}).
+To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of 2-tuples with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}).
 We now see a fairly clear separation of the genome-wide estimates of $\theta$, with CpG islands being more co-methylated than the rest of the genome.
@@
 In contrast, within-fragment co-methylation outside of CpG islands depends mostly on the adjacent locus and the distance to that locus ($IPD$)[^cpg_density2].
-We might try to estimate the order of within-fragment co-methylation in CpG islands by analysing 2-tuples with $NIL = a, a = 1, 2, \ldots$.
-This quickly becomes complicated, however, since there are $2^{a - 1}$ odds ratios to estimate for $NIL = a$ 2-tuples.
-We have not yet pursued refinement.
+We might try to estimate the order of within-fragment co-methylation in CpG islands by analysing m-tuples with $\text{m} > 2$.
+This quickly becomes complicated, however, since there are $2^{\text{m} - 1}$ odds ratios to estimate for m-tuples.
+We have therefore not yet pursued this refinement.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_ac_merged_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI.pdf}
 \caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
-\label{fig:Ziller_ac_merged_MH_by_CGI}
+\label{fig:Ziller_merged_ac_MH_by_CGI}
 \end{figure}
@@
 There are of course differences, and we now turn our attention to these.
-The results of this section should be interpreted with some caution: few samples have replicates and so we will have little evidence to conclude what is driving any observed variation nor do we have sufficient power to definitely rule out large differences in co-methylation.
+The results of this section should be interpreted with some caution: few samples have replicates and so we will have little evidence to conclude what is driving any observed variation nor do we have sufficient power to definitively rule out large differences in co-methylation.
 Nonetheless, I will offer some hypotheses and support for these as appropriate.
@@
 Visually, the variation between tissues is greater than that between donors (the plots are more similar by column than by row).
-This is consistent with average methylation levels being more similar within tissues than within donors (__TODO: Cite__).
 
@@
 The most interesting differences between samples occur in the _Lister_ dataset.
-Recall that the _Lister_ dataset contains four 'mini dataset': _Lister-ADS_, _Lister-FF_, _Lister-IMR90_ and _Lister-H1_ (see section \ref{sec:Lister_dataset}).
-In each mini datasets, the pluripotent cell lines (induced pluripotent stem cells or embryonic stem cells) have larger genome-level estimates of $\theta$ than do their precursor and differentiated forms.
+Recall that the _Lister_ dataset contains four 'mini datasets': _Lister-ADS_, _Lister-FF_, _Lister-IMR90_ and _Lister-H1_ (see section \ref{sec:Lister_dataset}).
+In each mini datasets, the pluripotent cell lines (induced pluripotent stem cells or embryonic stem cells) have larger genome-level estimates of $\theta$ than do their precursor forms[^bmp4].
+This difference is less pronounced, even absent, in the samples differentiated _in vitro_ with BMP4 from the induced pluripotent stem cell lines (iPSCs).
+
 This suggests the pluripotent cell cells have stronger within-fragment co-methylation than do somatic or differentiated cells.
-
-__UP TO HERE: Comment on other pluripotent samples (H9, Seisenbergers); comments on IMR90 (decays superfast).__
-
-\subsection{Summary}
-
-__OLD BUT POTENTIALLY USEFUL STUFF BELOW HERE__
-
-While the similarities between samples are obvious, the differences are somewhat harder to detect from these genome-wide summaries.
-The ADS and FF mini datasets provide us with an opportunity to discuss some potential biological differences in within-fragment co-methylation.
-In Figure \ref{fig:Lister_lor_min_cov_10_strand_collapsed} we see that the _ADS-iPSC_ sample has a consistently higher level of co-methylation than the _ADS_ and _ADS-adipose_ samples.
-Likewise, the _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ have consistently higher levels of co-methylation than the _FF_  sample, although not the _FF-iPSC\_19.11+BMP4_ sample.
-
-I can think of two possible interpretations of this result.
-Firstly, iPSCs may have a more regulated methylome than do somatic cells.
-The fact that these same iPSCs have very little intermediate methylation could be seen as support of this interpretation.
-However, this same lack of intermediate methylation may simply reflect that the iPSC samples are a more homogeneous population of cells.
-It would be possible to have a sample with no within-fragment co-methylation but where all $\beta$-values are 0 or 1, e.g., a sample where all cells are perfect clones of a cell whose CpG methylation states are independent of one another.
-
-A second interpretation of this result is that it is due to 'reseting' of the methylome upon induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.
+Consistent with this is that the _H9_, _H9\_Laurent_ and _J1_ samples (embryonic stem cell lines), the _E6.5\_epiblast\_1_ sample (epiblasts derived from day 6.5 embryos) and the _E16.5\_male\_1_ (male progenitor germ cells derived from day 16.5 embryos) all have relatively high genome-wide estimates of $\theta$.
+
+[^bmp4]: The difference is less pronounced in the samples differentiated _in vitro_ with BMP4, e.g., comparing _FF-iPSC\_19.11_ with _FF-iPSC\_19.11+BMP4_.
+
+One explanation of this result, at least for the iPSCs, is that it is due to 'reseting' of the methylome during embryonic development induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.
+This explanation does not immediately carry over the the embryonic stem cell lines, although there is extensive reprogramming of DNA methylation during embryogenesis, which may be relevant \citep{Seisenberger:2013jq}.
+
+A second observation on the results for the _Lister_ dataset is that genome-level estimates of $\theta$ decay particularly rapidly in the somatic cell lines (_ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_).
+This is recapituled in the _IMR90\_cell\_line_ sample from the _Ziller_ dataset.
+
+The reduced within-fragment co-methylation in the somatic cell lines may be due to culturing of the cells.
 As a population of cells develops from a single cell, there are errors in copying the DNA methylation patterns from mother to daughter cell.
@@
 If the iPSC and iPSC-derived cell lines have had less (developmental) time between the laying down of the original DNA methylation pattern and the point at which they were sequenced, then there is a lower probability that the co-methylation has been disrupted by replication errors.
-However, high levels of within-fragment co-methylation does not on its own imply that the sample is homogeneous.
-It would be possible to have a sample with high levels of within-fragment co-methylation that also has high levels of intermediate methylation, e.g., a sample with multiple distinct clonal subpopulations where each subpopulation has high levels of co-methylation but  also has a distinct methylome.
-
-Taking the co-methylation results and the distribution of $\beta$-values together, I conclude that for the ADS and FF mini datasets the iPSC and iPSC-derived samples truly have higher levels of co-methylation and are also more homogeneous than their precursor cells.
+There are also reports that DNA methylation profiles differ widely between cell lines and their precursor primary tissues \cite{Nestor:2015db}, which very likely also effect co-methylation.
+
+\subsection{Limitations}
+
+A major challenge in analysing within-fragment co-methylation is the susceptibility of estimates of $\theta$ to M-bias, particularly when the $IPD$ is large and measurements are taken towards the ends of reads.
+As noted in Section \ref{sec:m-bias}, it is difficult to properly account for M-bias in pre-trimmed data such as the _Lister_ dataset.
+Since we can do little about data that has already been pre-trimmed, we are forced to be cautious in our interpretation of within-fragment co-methylation for 2-tuples with larger $IPD$s from these datasets.
+This is particularly relevant to the apparent increase in within-fragment co-methylation for $NIL \geq 0$ CpG 2-tuples with $IPD \approx 180$.
+This result is tantalisingly similar to the periodicity we see in the correlations of $\beta$-values and the approximate spacing of nucleosomes in human DNA (see Section \ref{sec:nucleosome_and_chromatin}).
+However, it is not as apparent in other paired-end samples, and this is only likely to be resolved with longer reads that can span multiple nucleosomes.
 
@@
 
-Correlations of $\beta$-values and estimates of within-fragment co-methylation are complementary methods to better understand the complex dependence structure of bisulfite-sequencing data.
+Correlations of $\beta$-values and estimates of within-fragment co-methylation are complementary methods to better understand the complex dependence structure of DNA methylatin data.
 Which is more relevant depends on the question at hand.
@@
 However, the most common downstream analyses are based on $\beta$-values and so these correlations may be more relevant to analyses of DNA methylation data.
-Ultimately, the two measures are intertwined since the correlations of $\beta$-values are driven by co-methylation on individual DNA fragments.
+Ultimately, the two measures are intertwined since the correlations of $\beta$-values will in part be driven by co-methylation on individual DNA fragments.
 This idea is explored in greater detail in Chapter \ref{chap:methsim}.
 
-A major challenge in interpreting within-fragment co-methylation is its susceptibility to M-bias or, rather, the difficulty of accounting for M-bias in pre-trimmed data.
-Since we can do nothing about data that has already been pre-trimmed, we are forced to be cautious in our interpretation of within-fragment co-methylation for 2-tuples with larger $IPD$s.
-This is particularly relevant to the increase in within-fragment co-methylation for $NIL \geq 0$ CpG 2-tuples with $IPD \approx 150-170$.
-This result is tantalisingly similar to the periodicity we see in the correlations of $\beta$-values and the approximate spacing of nucleosomes in human DNA (see Section \ref{sec:nucleosome_and_chromatin}).
-However, I do not feel comfortable concluding that this is truly what we are observing because of the potential for these systematic biases to create an artificial signal.
-Furthermore, we do not detect this signal any of the EPISCOPE samples, which are some of the best quality data I have and, being samples from human tissues rather than cell lines, are arguably more biologically relevant.
-
 The presented analyses only measure pairwise dependencies.
-More specifically, the $NIL = 0$ results measure the 'first-order' or 'one-step' correlations of $\beta$-values whereas the $NIL \geq 0$ measure a mixture of different orders of dependences but all are done in a pairwise manner.
+More specifically, the $NIL = 0$ results measure the 'first-order' dependence whereas the $NIL \geq 0$ measure a mixture of different orders of dependences but all are done in a pairwise manner.
 I find the $NIL = 0$ results easier interpret but they clearly do not tell the whole story.
 
-By comparing the $NIL = 0$ and $NIL \geq 0$ results we see that the strength of the dependency itself depends on not just the distance between the two CpGs but the number of intervening CpGs.
+By comparing the $NIL = 0$ and $NIL \geq 0$ results, we see that the strength of the dependency itself depends on not just the distance between the two CpGs, but the number of intervening CpGs.
+This is most noticeable for within-fragment co-methylation.
 This indicates that these dependencies are more than 'first-order'.
@@
 In theory, we could explore these higher-order dependencies of within-fragment co-methylation by analysing m-tuples with $\text{m} > 2$.
-However, the complexity of the challenge increases exponentially with m.
-For example, even when analysing 3-tuples there are four $LOR$s for each $2 \times 2 \times 2$ contingency table and the $IPD$ vector now has two dimensions, meaning that the results have to be visualised over a surface.
-
-Another complexity that these analyses haved brushed over is the fact that the genome is not a 1D line but is rather a complex 3D structure whose shape also varies in time.
-Therefore, the relevant distance between two loci is not necessarily the number of basepair between them but may instead be the Euclidean distance in 3D space.
-Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the 3D structure of the genome, such as those provided by chromatin conformation capture technologies \citep{Dekker:2013hi}.
-
-The results of this chapter clearly demonstrate that DNA methylation has a highly complex dependence structure.
-These dependencies exist on individual DNA fragments and also manifest as correlations of aggregate measures of methylation, such as $\beta$-values.
-Bisulfite-sequencing give the highest resolution picture of this dependence structure.
-Nonetheless, this is still a rather limited picture, particularly for within-fragment co-methylation.
-Data from longer reads will be immensely useful in developing a better model of within-fragment co-methylation.
-In particular, reads that can span multiple nucleosomes will allow for co-methylation to be better tied in with and expand upon existing results on the relationship between nucleosomes and DNA methylation.
+However, the complexity of the challenge increases exponentially as $2^{\text{m}}$.
+Moreoever, the $IPD$ vector now has $(m - 1)$ dimensions, which makes it difficult to visualise the strength of co-methylation as a function of $IPD$.
+
+Another complexity that these analyses have brushed over is the fact that the genome is not a one-dimensional line but is rather a complex three-dimensional structure whose shape also varies in time.
+Therefore, the relevant distance between two loci is not necessarily the number of bp between them but may instead be the Euclidean distance in three-dimensional space.
+Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the three-dimensional structure of the genome, such as those provided by chromatin conformation capture technologies \citep[e.g.,][]{Dekker:2013hi}.
+
+In summary, the results of this chapter demonstrate that DNA methylation has a  complex dependence structure and identify several drivers of these complexity.
+Co-methylation exists along individual DNA fragments and also manifests as correlations of aggregate measures of methylation.
+While it is the highest resolution assay for studying DNA methylation, bisulfite-sequencing can still only give a rather limited picture of co-methylation, particularly of within-fragment co-methylation.
+Data from longer reads will be immensely useful in to advancing our understanding of DNA methylation dynamics.
+In particular, reads that can span multiple nucleosomes will allow for co-methylation to be better tied in with, and expand upon, existing results on the relationship between nucleosomes and DNA methylation.
 Until such data are available, we can try to learn more about plausible models of DNA co-methylation through computational methods such as the simulation method described in the next chapter.
-
-## Miscellaneous notes
-
-Agresti in Categorical Data Analysis:
-
-> In terms of bias and mean square error, Gart and Zweiful (1967) and Haldane (1955) showed the amended [i.e, add 0.5 to each count] estimators of the [odds-ratio log odds-ratio] behave well (see also Problem 12.4)
-
-The Mantel-Haenszel estimator might be my best bet. Breslow (1981) says:
-
-> The principal conclusion to be drawn from these results is that one should avoid using two of the standard techniques of odds ratio estimation with sparse data.
-
-> The bias of the empirical logit estimator [i.e., + 0.5 to each cell], on the other hand, is sufficient erratic that the possibility of correction seems remote.
-
-> Breslow (1981, Biometrika 68, 73-84) has shown that the Mantel-Haenszel odds ratio is a consistent estimator of a common odds ratio in sparse stratifications (Greenland, S.
-& Robins, J.
-M.
-Biometrics 41, 55–68 (1985).)
-
-The Mantel-Haenszel estimator is an estimator of the common odds ratio for $2 \times 2 \times K$ tables.
-Therefore, using the MH estimator requires the assumption that all $2 \times 2$ tables have the same underlying odds ratio.
-The strata for co-methylation would be 'sample', 'IPD' and possibly 'region', where region is something like 'UMRs, LMRs, PMRs and other' or 'CpG island and non-island'.
-
-Some good notes on MH estimator [http://isites.harvard.edu/fs/docs/icb.topic79671.files/stratcch1.pdf](http://isites.harvard.edu/fs/docs/icb.topic79671.files/stratcch1.pdf).
-
-An offset of 0.5 eliminates the first-order bias of the maximum likelihood estimator of the (log) odds ratio (Walter, S. D. Biostatistics (1987).)
-
-Regarding the effect of the homogeneity on the Mantel-Haenszel estimator, Greenland, S. Stat Med 1, 217–227 (1982):
-
-> Under the homogeneity assumption, OR^MH is an almost fully efficient estimator of the common value, and remains a consistent estimator of this value even with sparse data, i.e.
-when the fineness of stratification results in many small cell counts.
-__As shown later, its convergence properties under heterogeneity are not as desirable, although even in this case it remains insensitive to small cells.__ [emphasis added]
-
-The _Miettenen_ estimator is like a generalised form of the Mantel-Haenszel estimator, but assumes each cell count is greater than approximately 4 (Greenland, S. Stat Med 1, 217–227 (1982)).
-
-## Plan of attack
-
-Following discussion with Terry (2015-04-21), this is my plan of attack to finish this chapter and the `methsim` chapter.
-
-- [x] Finish reading Greenland (1982).
-  - What happens with the MH estimator when the common odds ratio assumption is invalid. Is there a better estimator in this setting?
-
-  > As a consequence, when there is heterogeneity, the expectation of MH varies with the overall and stratum-specific case-control ratios.
-The usual common odds ratio estimators share this disconcerting property \citep{Greenland:1982uk}.
-
-  In contrast to the MH estimator, the Miettenen estimator consistently estimates meaningful epimiological parameters. However, it "[requires] consideration in the selection of the weights".
-
-- Run Woolf's test or Breslow-Day test of homogeneity of odds ratios (noting that these are not powerful tests).
-
-__In fact, Breslow himself advocates for the Tarone statistic in place of the Breslow-Day statistic (Breslow, N.
-E.
-JASA 91, 14–28 (1996)).
-Also, Woolf's test, Breslow-Day and Tarone's statistic are only appropriate in Asymptopia I (a small number of tables with large frequencies) and not in Asymptopia II (a large number of tables with small frequencies); see Liang and Self (1985) for Asymptopia II (the scenario I am in)__
-
-Liang, K.-Y. & Self, S. G. Biometrika (1985) advocate for a conditional test of homogeneity of odds ratios when the data are sparse (Asymptopia II) but note that:
-
-> The problem with the conditional approach is its complexity.
-It is clear that when the sample size of each table increases, the computation of T2, T4 or T5 [ the conditional tests] becomes a burden which, eventually, will be prohibitive.
-This raises the question of how large the sample sizes in each table must be before the unconditional score test T1 is reasonable to use.
-Further simulation results suggest that even when the number of cases and controls are as small as 10 with as many as 10 tables, the large-sample chi-squared approximation is adequate in that the observed sizes are near to the nominal 0 05 level.
-
-I have $K \\g 10$ ($K$ = number of $2 \times 2$ tables) and $d \geq 10$ ($d$ = sequencing coverage), so the unconditional test may well be fine.
-
-- Algorithm: If the null of homogeneity of odds ratio is rejected => stratify further(?), otherwise compute MH or Miettenen estimates of the log odds ratio (and standard errors) for different stratification strategies (see [https://stat.ethz.ch/pipermail/r-help/2007-February/126254.html](https://stat.ethz.ch/pipermail/r-help/2007-February/126254.html))
-  - Stratify by IPD
-  - Stratify by IPD + chromosome
-  - Stratify by IPD + chromosome + chromosome arm
-  - Stratify by IPD + chromosome + chromosome arm + bin (Mb)
-  - Check whether the more-stratified estimates fall within a reasonable distance from the less-stratified estimates (i.e., with respect to standard errors of the less-stratified estimates).
-- Settle on the most appropriate level of stratification and apply to all samples.
-  - Discuss results.
-- `methsim`: Sample LORs from a Normal($\mu$, $\sigma^2$) distribution, where mu (resp.
-sigma) is the MH estimator (resp.
-standard error?
-empirical variance of stratified MH estimators?) of within-fragment co-methylation for that IPD and region-type.

### 2015-04-30 Rmarkdown/datasets.Rmd
Correct description of EPISCOPE data (closes #10).

@@
 Unlike visceral adipose tissue, subcutaneous adipose tissue is thought to be protective against obesity-related metabolic dysfunction \citep{Chau:2014do}.
-- _VA_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs and is associated with metabolic dysfunction \citep{Chau:2014do}.
-- _VAT_: Visceral adipocytes, which are derived from the VA tissue.
+- _VA_: Visceral adipocytes, which are derived from _VAT_.
+- _VAT_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs and is associated with metabolic dysfunction \citep{Chau:2014do}.
 

### 2015-05-01 Rmarkdown/biology_background.Rmd
Update thesis structure.
Add brief captions

@@
 
-\chapter{Biology background}\label{chap:biology_background}
+\chapter{Introduction to genomics and DNA methylation}\label{chap:biology_background}
 
@@
 These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix.
-This is illustrated in Figure \ref{fig:double_helix}.
+This is illustrated in Figure \ref{fig:dna_simple2}.
 
-![Simple diagram of double-stranded DNA showing complementary base pairings.
-By Forluvoft (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File%3ADNA_simple2.svg](http://commons.wikimedia.org/wiki/File%3ADNA_simple2.svg)\label{fig:double_helix}](../figures/DNA_simple2.pdf)
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=0.7\textwidth]{../figures/DNA_simple2.pdf}
+\caption[DNA double helix]{Simple diagram of double-stranded DNA showing complementary base pairing.
+By Forluvoft (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File\%3ADNA_simple2.svg}}
+\label{fig:double_helix}
+\end{figure}
 
@@
 \includegraphics[width=0.3\textwidth]{../figures/5-Methylcytosine.pdf}
-\caption{Chemical structure of 5-methylcytosine. "5-methylcytosine". By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
+\caption[Chemical structure of 5-methylcytosine]{Chemical structure of 5-methylcytosine.
+"5-methylcytosine".
+By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
 \label{fig:5mC}
@@
 \includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
-\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19).
+\caption[Dinucleotide frequencies in the human genome]{Observed:expected ratio of dinucleotides in the human reference genome (hg19).
 The expected frequency is computed under an 'independence' model, based on the observed frequencies of each base.}
@@
 \includegraphics[width=\textwidth]{../figures/IPD_ECDF_hg19.pdf}
-\caption{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19).
+\caption[ECDF of distances between CpGs in hg19]{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19).
 The observed distances are contrasted with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
@@
 \includegraphics[width=\textwidth]{../figures/IPD_frequency_polygon_hg19.pdf}
-\caption{The frequency of the observed distances between adjacent CpGs in the human reference genome (hg19).}
+\caption[Frequency of distances between CpGs in hg19]{The frequency of the observed distances between adjacent CpGs in the human reference genome (hg19).}
 \label{fig:IPD_frequency_polygon_hg19}
@@
 \includegraphics[width=\textwidth]{../figures/bisulfite-conversion.pdf}
-\caption{The effect of bisulfite-treatment of DNA.
+\caption[Schematic of bisulfite-conversion]{The effect of bisulfite-treatment of DNA.
 The double-stranded DNA is denatured and each strand undergoes bisulfite-treatment.
@@
 \centering
+\caption[Percentage of reads without CpGs]{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset.
+Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs.
+See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
+\label{tab:reads_without_cpgs}
 \begin{tabular}{@{}rc@{}}
@@
 \end{tabular}
-\caption{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset.
-Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs.
-See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
-\label{tab:reads_without_cpgs}
 \end{table}
@@
 \includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
-\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome.
+\caption[Sequence variation confounds methylation calling]{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome.
 The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome.

### 2015-05-01 Rmarkdown/datasets.Rmd
Add brief captions

@@
 \centering
+\caption[Description of \emph{Lister} dataset]{Samples in \emph{Lister} dataset.
+The reported read lengths are prior to any trimming of the reads.
+Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
+\label{tab:Lister_dataset}
 \begin{tabulary}{\textwidth}{LLLLLL}
@@
 \end{tabulary}
-\caption{Samples in Lister dataset.
-The reported read lengths are prior to any trimming of the reads.
-Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
-\label{tab:Lister_dataset}
 \end{table}
@@
 \centering
+\caption[Description of \emph{EPISCOPE} dataset]{Samples in \emph{EPISCOPE} dataset. The reported read lengths are prior to any trimming of the reads.}
+\label{tab:EPISCOPE_dataset}
 \begin{tabulary}{\textwidth}{LLLL}
@@
 \end{tabulary}
-\caption{Samples in EPISCOPE dataset. The reported read lengths are prior to any trimming of the reads.}
-\label{tab:EPISCOPE_dataset}
 \end{table}
@@
 \centering
+\caption[Description of \emph{Seisenberger} dataset]{Samples in \emph{Seisenberger} dataset.
+The reported read lengths are prior to any trimming of the reads.
+All samples were first published in \citet{Seisenberger:2012ko}.}
+\label{tab:Seisenberger_dataset}
 \begin{tabulary}{\textwidth}{LLLLL}
@@
 \end{tabulary}
-\caption{Samples in Seisenberger dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Seisenberger:2012ko}.}
-\label{tab:Seisenberger_dataset}
 \end{table}
@@
 \centering
+\caption[Description of \emph{Ziller} dataset]{Sequencing runs in \emph{Ziller} dataset.
+The reported read lengths are prior to any trimming of the reads.
+All samples were first published in \citet{Ziller:2013fs}.}
+\label{tab:Ziller_dataset_sequencing_runs}
 \begin{tabulary}{\textwidth}{LLLLL}
@@
 \end{tabulary}
-\caption{Sequencing runs in Ziller dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Ziller:2013fs}.}
-\label{tab:Ziller_dataset_sequencing_runs}
 \end{table}

### 2015-05-01 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Add brief captions

@@
 \end{tabulary}
-\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
+\caption[Popular bisulfite-sequencing read mappers and their underlying mapping software]{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
 \label{tab:bs-seq_mappers}
@@
 \includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
-\caption{M-bias plot for the \emph{ADS} sample from the \emph{Lister} dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.}
+\caption[\emph{ADS} M-bias]{M-bias plot for the \emph{ADS} sample from the \emph{Lister} dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.}
 \label{fig:ADS_M-bias}
@@
 \includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
-\caption{M-bias plot for the E18VA sample from the EPISCOPE dataset.
+\caption[\emph{E18VA} M-bias]{M-bias plot for the E18VA sample from the EPISCOPE dataset.
 Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.
@@
 \includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
-\caption{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
+\caption[\texttt{Bis-SNP} method schematic]{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
 For all three loci, the reads mapped to the forward strand contain a thymine.
@@
 \end{tabular}
-\caption{Example of output for methylation calling at 1-tuples.
+\caption[Methylation patterns at 1-tuples]{Example of output for methylation calling at 1-tuples.
 Each row records the the number of methylated ($M$) and unmethylated ($U$) reads at a 1-tuple.
@@
 \includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
-\caption{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
+\caption[Different methylation patterns can lead to identical $\beta$-values]{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
 \label{fig:incomplete_picture_1-tuples}
@@
 \end{tabulary}
-\caption{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
+\caption[Software for calling methylation patterns at m-tuples]{Other software for methylation calling at m-tuples and their limitations.
+Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
 \label{tab:other_m-tuple_methylation_callers}
@@
 \includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
-\caption{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}).
+\caption[\texttt{methtuple} example region]{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}).
 The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
@@
 \includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
-\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
+\caption[\texttt{methtuple} 1-tuples]{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
 \label{fig:methtuple_example2}
@@
 \includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
-\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
+\caption[\texttt{methtuple} 3-tuples]{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
 \label{fig:methtuple_example3}
@@
 \includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
-\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
+\caption[\texttt{methtuple} 3-tuples with \texttt{--all-combinations}]{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
 \label{fig:methtuple_example4}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
-\caption{Number of CpGs per read for the Lister dataset.}
+\caption[Number of CpGs per read in \emph{Lister} dataset]{Number of CpGs per read for the Lister dataset.}
 \label{fig:Lister_CpGs_per_read}
@@
 \includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
-\caption{Example output of \texttt{methtuple} for 3-tuples.}
+\caption[\texttt{methtuple} 3-tuples example output]{Example output of \texttt{methtuple} for 3-tuples.}
 \label{fig:3-tuples}
@@
 \includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
-\caption{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes.
+\caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes.
 The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.
@@
 \includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
-\caption{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
+\caption[\texttt{methtuple} memory usage]{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes.
 These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).

### 2015-05-01 Rmarkdown/wgbs_downstream_analyses.Rmd
Add brief captions

@@
 \centering
+\caption[Proposed multiple hypothesis testing procedures]{Methods proposed for adjusting for the multiple hypothesis testing performed in an analysis of DMCs.
+Several papers use describe their analysing as performing a ``false discovery rate adjustment'' or ``false discovery rate correction'' without explicitly stating what they are doing or citing a reference.
+One paper uses an ultra-conservative Bonferonni correction.}
+\label{tab:multiple_testing}
 \begin{tabulary}{\textwidth}{LL}
@@
 \end{tabulary}
-\caption{Methods proposed for adjusting for the multiple hypothesis testing performed in an analysis of DMCs.
-Several papers use describe their analysing as performing a ``false discovery rate adjustment'' or ``false discovery rate correction'' without explicitly stating what they are doing or citing a reference.
-One paper uses an ultra-conservative Bonferonni correction.}
-\label{tab:multiple_testing}
 \end{table}
@@
 \centering
+\caption[Allele-specific methylation]{$2 \times 2$ table used to test for allele-specific methylation.
+$m^{1}$ is the number of reads with the first allele and that are also methylated at the methylation locus, $u^{1}$ is the number of reads with the first allele and that are also unmethylated at the methylation locus, etc.}
+\label{tab:asm}
 \begin{tabular}{@{}lll@{}}
@@
 \end{tabular}
-\caption{$2 \times 2$ table used to test for allele-specific methylation.
-$m^{1}$ is the number of reads with the first allele and that are also methylated at the methylation locus, $u^{1}$ is the number of reads with the first allele and that are also unmethylated at the methylation locus, etc.}
-\label{tab:asm}
 \end{table}
@@
 \includegraphics[width=\textwidth]{../figures/MethPat.pdf}
-\caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples.
+\caption[Schematic of \texttt{MethPat} class]{Schematic of the \texttt{MethPat} class, shown here for 3-tuples.
 Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the counts of the methylation patterns (grey box) are aligned.
@@
 \centering
+\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data.
+All m-tuples are stranded.
+The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
+The `number of rows' corresponds to the number of m-tuples in the object.
+The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples.
+The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
+\label{tab:methpat_sparsity}
 \begin{tabulary}{\textwidth}{LLLLL}
@@
 \end{tabulary}
-\caption{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data.
-All m-tuples are stranded.
-The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}).
-The `number of rows' corresponds to the number of m-tuples in the object.
-The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples.
-The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
-\label{tab:methpat_sparsity}
 \end{table}

### 2015-05-01 Rmarkdown/wgbs_statistical_framework.Rmd
Add brief captions

@@
 \includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
-\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment.
+\caption[Source of variation in bisulfite-sequencing experiments]{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment.
 Each sample starts as a population of cells, with potentially different methylation patterns.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
+\caption[\emph{EPISCOPE} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
+\caption[\emph{Lister} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
+\caption[\emph{Seisenberger} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
+\caption[\emph{Ziller} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
+\caption[\emph{EPISCOPE} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data.
+\caption[\emph{Lister} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data.
+\caption[\emph{Seisenberger} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
+\caption[\emph{Ziller} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{EPISCOPE} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Lister} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Seisenberger} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
-\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Ziller} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{EPISCOPE} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Lister} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Seisenberger} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
-\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Ziller} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.

### 2015-05-01 Rmarkdown/methsim.Rmd
Continue methsim chapter

@@
 \begin{chapabstract}
-I present the `methsim` software to simulate DNA methylation data.
+This chapter describes the `methsim` software to simulate DNA methylation data.
 `methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software.
-I describe the statistical model behind `methsim` and its software implementation.
-Finally, I provide examples of how `methsim` can be used to learn about DNA methylation and how it might be used to develop simulation studies to compare competing analysis methods.
+We explore different parameter choices and highlight promising directions of research, such as using `methsim` to simulate data in order to compare methods for identifying differential methylation.
 \end{chapabstract}
 
-1. Introduction
-  - Why simulation?
-2. Literature review
-  - Simulating reads vs. $\beta$-values
-  -
-3. Methods
-  - Implementation
-4. Case study
-5. Conclusions
-
 \section{Introduction}\label{sec:methsim_introduction}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
 The simulation methods described here can be thought to lie on a spectrum with model-based methods at one end and re-sampling based methods at the other end[^model-based].
@@
 \subsection{Methods for simulating bisulfite-sequencing reads}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
 The most widely studied downstream analysis problem is that of identifying differential methylation, which is done by comparing summary measures of methylation, such as $\beta$-values, between two or more groups.
@@
 Consequently, the simulation model is often only briefly described and a software implementation is rarely made available.
-In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available[^WGBSSuite].
+In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available.
 As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}.
@@
 [^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software.
-Furthermore, no licence file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code.
-(__TODO: Update if I received a response.
-I emailed the developer, Owen Rackham, asking him to add a licence.
-He is considering my request.__)
+Furthermore, no license file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code.
 
@@
 
-- [x] DONE
-
 \citet{Lacey:2013iy} and \citet{Rackham:2015bv} are notable in that they choose to simulate the locations of CpGs rather than simply using their locations in a reference genome.
@@
 
-I do not think this a useful or necessary step.
-In fact, I think it is counterproductive.
+I do not think this a useful or necessary step and may even be counterproductive.
 While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed.
 If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome.
-And it is counterproductive because the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci.
-Furthermore, almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.
+Furthermore, the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci.
+
+Almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.
 
 \subsubsection{Simulating $B_{i, j}$}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
-The sequencing depth at each methylation loci, $m_{i, j} + u_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.
+The sequencing depth at each methylation loci, $d_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.
 
@@
 
-- [x] DONE
-
 The final step is to simulate the read counts, $M_{i, j}$ and $U_{i, j}$.
-These are based on the true underlying methylation level, $B_{i, j}$, and the sequencing depth, $m_{i, j} + u_{i, j}$.
-In the beta-binomial model this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
+These are based on the true underlying methylation level, $B_{i, j}$, and the sequencing depth, $d_{i, j}$.
+In the beta-binomial model, this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
 In addition to binomial read sampling, the `WGBSSuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.
@@
 \subsubsection{Simulating differentially methylated regions}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
 A key feature ignored by the majority of simulation methods, with the notable exceptions of \citet{Lacey:2013iy} and \citet{Rackham:2015bv}, is co-methylation.
@@
 
-As demonstrated in Chapter \ref{chap:co-methylation}, the physical process of co-methylation is at the level of individual DNA fragments, not at the level of aggregate methylation levels.
-The observed correlations of $\beta$-values are driven by this within-fragment co-methylation.
-However, a simulation model of DNA methylation data need not induce correlations at the level of individual DNA fragments.
-Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation, but it still may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
+Most simulation methods do not simulate individual reads and so cannot simulate within-fragment co-methylation.
+Instead, they capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
 For example, under the Beta-Binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent.
@@
 
-- [x] DONE
-
 A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \citet[e.g.,]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
@@
 \subsection{Methods based on re-sampling real data}
-
-- [x] DONE
 
@@
 
-- [x] DONE
-
 My initial interest in simulation methods was to explore models of co-methylation, particularly at the level of individual DNA fragments. This requires two key capabilities:
@@
 
-- [x] DONE
-
 Simulating data with `methsim` involves 3 steps:
@@
 
+In what follows we will simulate data based on the _ADS_ methylome from the _Lister_ dataset.
+We focus on simulating autosomal data, since the sex chromosomes and mitochondrial DNA have very different methylation dynamics.
+
 \subsection{Implementation}
 
-- [x] DONE
-
 `methsim` is written in R and builds on the `MethylationTuples` package described in Chapter \ref{chap:wgbs_downstream_analyses}, as well as several R packages available on Bioconductor and CRAN (see the `DESCRIPTION` file for a complete list.).
+It is currently a very experimental package and is therefore not yet published on Biocondutor, but its development can be followed at [https://github.com/PeteHaitch/methsim](https://github.com/PeteHaitch/methsim).
 
@@
 
-To simulate a 'true' methylome, the user runs the `simulate()` method on the `SimulateMethylomeParam` object. This returns a `SimulatedMethylome` object.
+To simulate a true methylome, the user runs the `simulate()` method on the `MethylomeParam` object. This returns a `SimulatedMethylome` object.
 
-Once we have a 'true' methylome, we can simulate data from it.
-Currently, `methsim` supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method.
-Other assays, such as RRBS or microarrays may be supported in future releases.
+Once we have a true methylome, we can simulate data from it.
+`methsim` currently supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method.
+Other assays, such as RRBS or microarrays could in principle be supported.
 A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated.
-When applied to a `WGBSParam` object, which includes a `SimulatedMethylome` object, the `simulate()` method can return a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2].
-A `SimulatedBS` object contains all simulated reads and is generally much larger than a `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat].
+When applied to a `WGBSParam` object, the `simulate()` method returns a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2].
+A `SimulatedBS` object contains all simulated reads and is generally a large object (on the order of $10$ GB).
+By contrast, the `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat], is much smaller (on the order of $100$ MB) but does not contain the full information of the simulation since read-level data are lost.
 `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.
@@
 
-- [ ] DONE
-
 As we have seen, DNA methylation is highly heterogeneous along the genome.
@@
 In these locally-similar regions we might hope to model DNA methylation by a simple parametric model.
-`methsim` is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.
+`methsim`, like other simulation methods, is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.
 
@@
 1. Segment the methylome into regions of similarity.
-2. Sample parameters for each methylation locus, $i$, based on step 1.
+2. Sample parameters for each methylation locus, $i$, based on the segmentation.
 
@@
 
-- [x] DONE
-
 `methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the input methylome into regions of similarity.
@@
 
-[^pmds]: Partially methylated regions are also commonly known as partially methylated domains, but I refer to them as regions for consistency with UMRs and LMRs.
+[^pmds]: Partially methylated regions are also commonly known as partially methylated domains, but we will refer to them as 'regions' for consistency with UMRs and LMRs.
 
@@
 
-`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarly, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output].
+`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarity, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output].
 Roughly speaking, _other_ regions are 'mostly methylated regions' (see Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Ziller_beta_by_pm_boxplots}), although \citet{Burger:2013kq} do not describe these regions as such.
@@
 This allows the incorporation of within-fragment co-methylation into the simulation.
-The choice of a one-step process is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.
+The choice of a one-step process is one of computational simplicity and is a reasonable approximation for most of the genom given the available data (see Chapter \ref{chap:co-methylation}).
 
-In Chapter \ref{chap:co-methylation} I showed that the strength of within-fragment co-methylation varies as a function of the intra-pair distance and by the genomic context.
-I also showed that for a pair of CpGs separated by given IPD, and in the same genomic context, that there is a large amount of variability in the strength of within-fragment co-methylation.
-
-For these reasons I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous.
+In Chapter \ref{chap:co-methylation} we saw that the strength of within-fragment co-methylation varies as a function of the intra-pair distance and by the genomic context.
+For this reason I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous.
 In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for this pair of loci[^context].
@@
 
-Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised $\mathbf{Z}$ by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$.
-We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$ as I now describe.
+Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$.
+We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$.
 
@@
 Once $P_{i, i + 1}$ is computed, the desired transition probability is obtained by dividing the appropriate element of $P_{i, i + 1}$ by the appropriate marginal probability.
-To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also to zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ to three decimal places.
+To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ (to three decimal places).
 In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.
 
-To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimators, $\mathbf{\beta}$ and $\mathbf{\widehat{\psi}}$, in the input methylome.
-More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ and $\mathbf{\psi}$ is based on $\widehat{\psi_{IPD(i, i+1) | r_{i}}}$, the empirical distributions of $\beta_{i}$ and $\psi_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively.
-These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and __FIGURE__.
+The above-described model implicitly assumes that all $n_{h}$ haplotypes have the same marginal probabilities and co-methylation structure.
+We may extend this model to define $\mathbf{Z}$ as a _mixture_ of a small number of first-order Markov chains defined as in the above.
+Such a mixture model may be appropriate when simulating a sample that is a combination of cell types.
+
+To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimates, $\mathbf{\beta}$ and $\mathbf{\widehat{\psi}}$, in the input methylome.
+More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\mathbf{\psi}$ is based on $\widehat{\psi_{MH}}$.
+We have already seen many examples of the distribution of $\widehat{\psi_{MH}}$ in Chapter \ref{chap:co-methylation}.
+The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.
 
@@
 
-__TODO: Figure of ADS within-fragment co-methylation__
+There are many ways to simulate values of $\mathbf{B}$ and $\mathbf{\theta} = \mathbf{log{\psi}}$ based on these empirical distributions.
+I have so far only had time to explore a few simple methods and this remains ongoing work.
+We will discuss the results from these five models (m1, m2, m3, m4, m5):
 
-These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in __FIGURES__.
+\begin{description}
+  \item[m1] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
+$\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
+  \item[m2] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this is then perturbed by a $Gaussian(0, \sigma_{B}^2)$ random variable.
+This is similar to what is done by `WGBSsuite`.
+Here we use $\sigma_{B}^{2} = B \times (1 - B)$.
+The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
+$\theta$ simulated as in m1.
+  %\item[m3] $\mathbf{B}$ simulated as in m2 but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5. $\theta$ simulated as in m1.
+  %\item[m4] As for m3 but where the sample is a 50:50 mixture, i.e., two independent simulations of $\mathbf{B}$ and $\mathbf{\theta}$ that are sampled from with equal probability.
+  %\item[m5] As for m3 but with $\mathbf{theta} = 0$ to simulate independence of methylation states within fragments.
+\end{description}
 
-There are several ways to simulate values of $\mathbf{B}$ and $\mathbf{\psi}$ based on these empirical distributions.
-The simplest is to sample each $B_{i}$ and $\psi_{i}$ directly from these empirical distributions.
-Alternatively, we could take a more model-based approach.
-For example, we could specify a  parametric model of the $B_{i}$ where all $i$ in a region have the same mean and variance (the mean and variance being based on the empirical distribution of $\beta$-values) or where the $B_{i}$ in each region come from an autoregressive process.
+The simulation of $\mathbf{\theta} = \mathbf{log{\psi}}$ requires further explanation.
+The empirical distributions of $\mathbf{\widehat{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
+When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?
 
-The simulation of $\mathbf{\psi}$ requires further explanation.
-The empirical distributions of $\mathbf{\widehat{\psi}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
-When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?
 Again, there are several options.
-We might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\psi = 1$ (corresponding to independence).
+For example, we might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence).
+Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta_{MH}}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$[^sigma].
+While obviously a gross simplification, the vast majority of $NIL = 0$ CpG pairs have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.
 
-I explore different methods for simulating $\mathbf{B}$ and $\mathbf{\psi}$ in Section \ref{sec:results}.
+[^sigma]: $\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for $IPD \leq 180$.
 
-The above-described model implicitly assumes that all $n_{h}$ haplotypes have the same marginal probabilities and co-methylation structure.
-We may extend this model to define $\mathbf{Z}$ as a _mixture_ of a small number of first-order Markov chains defined as in the above.
-Such a mixture model may be appropriate when simulating a sample that is a combination of cell types.
+There is an increase in complexity for models m1 through to m4, while m5 is there as a sanity check that the within-fragment co-methylation model is working as intended.
+The limitations of these models may already be clear, but we will discuss these further in Section \ref{sec:methsim_results}.
 
@@
 The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
-
-The number of reads required is computed by $n_{reads} = \frac{\text{average sequencing coverage}}{\text{read length}} \times \text{size of genome}$.
-The number of reads per chromosome is assigned proportional to the chromosome length.
-Then, the start of each read is sampled uniformly across the chromosome respective chromosome.
-We only retain those reads that overlap a methylation locus.
+We will use $200$ bp, single-end reads with an average coverage of $30 \times$ in what follows.
 
@@
 I don't consider this feature a priority, but it could easily be implemented by generating read lengths from a given probability distribution.
+
+The number of reads required is computed by $n_{reads} = \frac{\text{average sequencing coverage}}{\text{read length}} \times \text{size of genome}$.
+The number of reads per chromosome is assigned proportional to the chromosome length.
+Then, the start of each read is sampled uniformly across the chromosome respective chromosome.
+We only retain those reads that overlap a methylation locus.
 
@@
 
-\subsection{Construct the output}
+\subsection{Constructing the output}
 
@@
 
-To simulate these type of experiments, we need to be able to control the variation between the `SimulatedWGBS` objects.
+To simulate these type of experiments, we need to be able to control the variation between the resulting `SimulatedWGBS` objects.
 In this example, we probably want the three samples in each group to be more similar to each other than to the three samples in the other group.
@@
 For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\mathbf{\psi}$ between calls to `simulate()`.
-
-and the same average methylation level for each region but vary the strength of co-methylation.
-This could be achieved by altering the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions between two `SimulateMethylomeParam` objects while holding the rest constant.
-Or we might vary the weights, $W$, to simulate more or less heterogeneous samples.
-
-In Section ref{sec:results} I explore how much variation is created using these different techniques.
+This avenue remains to be explored.
 
@@
 
-__UP TO HERE: See what below can be salvaged__
-
 \subsubsection{Performance}
-
-- [x] DONE
 
@@
 
-Shown below are the running times and memory usage involved in simulating a $30\times$ whole-genome bisulfite-sequencing sample using the _ADS_ `MethylomeParam` object included in `methsim`.
-Using up to 8 CPU cores in parallel, it takes less than 5 minutes to simulate the underlying 'true' methylome and a further __HOW LONG__ to simulate a single $30 \times$ sequencing coverage whole-genome bisulfite-sequencing assay of this methylome[^parallel].
-
-[^parallel]: NB the creation of the `SimulatedMethylome` via `simulate(ADS, BPPARAM)` currently only uses a single CPU.
-
-__UP TO HERE: simulate-WGBSParam-method seems to be causing a memory leek and resulting crash. Running valgrind in reduced problem to investigate.__
-
-```{r, echo = TRUE, eval = FALSE}
-# Load necessary packages
-suppressPackageStartupMessages(library(methsim))
-suppressPackageStartupMessages(library(BiocParallel))
-suppressPackageStartupMessages(library(pryr))
-
-# Use up to 8 cores in parallel computations.
-BPPARAM <- MulticoreParam(workers = 8L)
-
-# Set seed for reproducibility.
-set.seed(666)
-
-# ADS: the MethylomeParam object.
-pryr::object_size(ADS)
-#> 16.5 MB
-
-# sm: the SimulatedMethylome object
-system.time(l_sm <- simulate(ADS, BPPARAM = BPPARAM))
-#> Currently only simulates CpG methylation.
-#> Currently only simulates unstranded methylomes.
-#> Simulating 1 methylome..
-#>    user  system elapsed
-#> 186.680   9.676 270.780
-sm <- l_sm[[1]]
-pryr::object_size(sm)
-#> 676 MB
-
-# wgbsp: the WGBSParam object.
-# NOTE: Basically all the memory used by the WGBSParam object is the
-# SimulatedMethylome object.
-wgbsp <- WGBSParam(SimulatedMethylome = sm,
-                   AveCov = 30,
-                   ErrorRate = 0.01,
-                   SequencingType = 'SE',
-                   ReadLength = 100)
-pryr::object_size(wgbsp)
-#> 676 MB
-
-# sbs: the SimulatedBS object.
-system.time(sbs <- simulate(wgbsp, BPPARAM = BPPARAM))
-#> Simulating 1 bisulfite-sequencing sample...
-
-#>     user   system  elapsed
-#> 2716.503  174.591  548.122
-pryr::object_size(sbs)
-#> 10.1 GB
-
-# methpat: the MethylationTuples::MethPat object of one-tuples.
-# NOTE: information is lost when a SimulatedBS object is coerced to a MethPat
-# object.
-system.time(methpat <- asMethPat(sbs,
-                                 sampleName = "sim_ADS",
-                                 size = 1L,
-                                 BPPARAM = BPPARAM))
-#>    user  system elapsed
-#> 557.900  34.779  48.944
-pryr::object_size(methpat)
-#> 451 MB
-```
-
+Using up to 8 CPU cores in parallel, it takes less than 9 minutes to simulate a single $30 \times$ sequencing coverage whole-genome bisulfite-sequencing assay from a 'true' methylome.
 These results are indicative of simulating high-coverage whole-genome bisulfite-sequencing for human-sized genomes.
@@
 
-- [ ] DONE
+A central feature of `methsim` is that the simulation parameters are sampled from an input sample.
+At a bare minimum, the first test of `methsim` is that simulated data are similar to the real data on which they are based.
+To assess this, we compare several summary measures between the simulated data and the real data.
 
-A central pillar of `methsim` is that many of the simulation parameters are sampled from real data.
-At a bare minimum, the first test of `methsim` is that simulated data are similar to the real data on which they are based.
-To assess this, I compare several summary measures between the simulated data and the real data.
-The simplest of these are summaries are the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`.
-In contrast, the spatial correlation of $\beta$-values is not explicitly modelled in `methsim`; rather, we hope that `methsim` recapitulates this phenomenon by appropriate modelling of within-fragment co-methylation and choice of weights, $\mathbf{W}$.
+__TODO: Add in the _ADS_ data to the simulated data__
 
-Once it is established that `methsim` passes this minimal test, I use `methsim` to explore models of within-fragment co-methylation, the choice of the weights, $\mathbf{W}$, and their respective effects on the summary measures in order to identify areas of improvement in `methsim`.
-Finally, I demonstrate how `methsim` might be used to simulate data for use in a study comparing methods for identifying differential methylation.
+The simplest of these are summaries of the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`.
+We see that all of m1, m2, m3, m4 and m5 sufficiently capture the bimodality of $\beta$-values (__FIGUREs__) and their relationship to CpG islands (__FIGUREs__).
 
-\subsection{Summary measures of simulated data}
+__FIGURE__ plots the estimates of within-fragment co-methylation.
+Those models that include within-fragment co-methylation (m1, m2, m3 and m4) bear close resemblance to the real _ADS_ data, although are perhaps to homogeneous across chromosomes.
+Reassuringly, we see no evidence of within-fragment co-methylation in model 5, which simulates independence of within-fragment methylation states.
 
-- [ ] DONE
+In contrast, the spatial correlation of $\beta$-values is not explicitly modelled in any of models m1 to m5.
+While m3, m4 and m5 do include some spatial dependence of the $\mathbf{B}$, these are independent of $IPD$ and are intended mostly to analyse the effect of the independence condition in m2.
+__TODO: Describe these crap results.__
 
-\subsection{Models of within-fragment co-methylation}
+\section{Summary}
 
-- [ ] DONE
+Simulating DNA methylation data is complicated by its heterogeneity.
+This heterogeneity exists is in multiple 'directions': along the genome, between cells in a sample, and between subjects.
+As more data become available, we can get a better handle on the causes of the variation but, for now, it remains challenging.
 
-- Apparent co-methylation arises because of dependence of $B_{i}$, even when $Z_{i}$ are independent!
-E.g., set within-fragment co-methylation to zero but have $B_{i}$ fixed for a region results in independent $Z_{i}$ but dependent $B_{i}$.
-  - What to make of this?
-  - What if we sample $B_{i}$ from $\beta_{i | r_{i}}$ instead of using a fixed value for all loci in a region?
+`methsim` provides a framework that I have used to experiment with models for simulating DNA methylation data.
+The models I have currently explored capture some key aspects of the data including, for the first time, within-fragment co-methylation.
+Unfortunately, while these results are promising on some fronts, there is clearly much work to be done on others.
 
-\subsection{The choice of $\mathbf{W}$}
-
-\subsection{Simulating differential methylation}
-
-- [ ] DONE
-
-\section{Conclusions}
-
-- [ ] DONE
-
-\subsubsection{Comparison to other methods}
-
-__TODO: Work the comparison to other methods into my conclusions__
-
-\citet{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)).
-__TODO: Try these scripts, see how long they take__.
-The running times reported in the original publication are shown in __TABLE__ for a single simulation.
-__TODO: To simulate a single RRBS dataset would take HOW LONG__.
-The simulation scripts could be made more user-friendly by creating an R package, rather than supplying the software as a collection of scripts, and would benefit from some documentation.
-
-\begin{table}[h]
-\centering
-\begin{tabular}{@{}lll@{}}
-\toprule
-Number of samples & Number of sites & Running time (seconds) \\ \midrule
-8                 & 1000            & \textless 1            \\
-8                 & 5000            & 13                     \\
-8                 & 10000           &                        \\ \bottomrule
-\end{tabular}
-\caption{Reported running times for simulation methodology of \citet{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
-\label{tab:Lacey_runtime}
-\end{table}
-
-__TODO: Compare `methsim` to `WGBSSuite`.
-
-# TODO:
-
-- [ ] Decide on a variable for sequencing depth; $m_{i, j} + u_{i, j}$ is kinda confusing. Perhaps $d_{i, j} / D_{i, j}$?
+Most notably, and frustratingly, I do not yet have an adequate model for the correlations of the true methylation levels, $\mathbf{B}$.
+The model of \citet{Lacey:2013iy} is promising, although may be computationally infeasible for whole-genome data given the timings reported in the original publication.
+The model used by `WGBSsuite` \citet{Rackham:2015bv} is similar to m2, so I suspect it will produce similar results, but this requires further investigation.
+I continue to explore how alternative models of $\mathbf{B}$ might be integrated into `methsim` to improve this aspect of the simulation.

### 2015-05-01 latex/phd_thesis.tex
Tidying main matter

@@
             pdfsubject={Dissertation},
-            pdfkeywords={microarray gene expression statistical models regression},
+            pdfkeywords={DNA methylation bioinformatics applied statistics epigenetics epigenomics},
             pdfproducer={Latex with hyperref, or other system},
@@
 
+\frontmatter
+
 \maketitle
-
-\frontmatter
 
@@
 
-\textbf{TODO}
-
-\begin{itemize}
-  \item Some data in my thesis are mapped to hg18 and some to hg19.
-  \item Include summary of notation?
-\end{itemize}
-
-The data sets used in this thesis were provided by the investigators listed below.\\\\
+The data sets used in this thesis were provided by the investigators listed in Chapter \ref{chap:datasets}.\\\\
 
@@
 Equally, my PhD would not be possible without the countless people who contribute to the software I use in my daily research.
-A special thanks to those who helped me as I learnt to write software of my own: Felix Krueger, Toby Sargeant, Martin Morgan, Hervé Pagès, Hadley Wickham and Dirk Eddelbuettel.
+A special thanks to those who helped me as I learnt to write software of my own: Felix Krueger, Toby Sargeant, Martin Morgan and Hervé Pagès.
 Open software is the best software.
@@
 
-I owe an enourmous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proof readers, Belinda Phipson, Saskia Freytag and Goknur Giner, who have greatly improved this thesis.
+I owe an enormous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proof readers, Belinda Phipson, Saskia Freytag and Goknur Giner, who have greatly improved this thesis.
 
-\textbf{TODO: final thanks}
-
-Finally, and most of all, thank you to my parents, Gay and John, to my brothers, Jack, Luke and Daniel, and to my friends. Words cannot express
+Finally, and most of all, thank you to my parents, Gay and John, to my brothers, Jack, Luke and Daniel, and to my friends. Your support, love and encouragement kept me going.
 
@@
 \listoftables
-
-% TODO: Check list of abbreviations
-\chapter{List of Abbreviations}
-
-\begin{tabular}{ll}
-
-bp & base pair \\
-kb & kilobase = $1000$ bp \\
-Mb & megabase = $1,000,000$ bp \\
-Gb & gigabase = $1,000,000,000$ bp \\
-WGBS & whole genome bisulfite sequencing \\
-RRBS & reduced representation bisulfite sequencing \\
-FDR & false discovery rate \\
-DMC & differentially methylated cytosine \\
-DMR & differentially methylated region \\
-HMM & hidden Markov model \\
-SE & single-end sequencing \\
-PE & paired-end sequencing \\
-indel & insertion or deletion \\
-CGI & CpG island \\
-
-\end{tabular}
-
-% TODO: Check list of symbols
-\chapter{List of Symbols}
-
-\begin{tabular}{ll}
-
-i.e. & that is\\
-
-\end{tabular}
 

### 2015-05-02 Rmarkdown/comethylation.Rmd
Add short captions

@@
 Rather than actually remove these sites, I simply set the corresponding $\beta$-value to `NA`.
-I also set the $\beta$-values of 'missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.
+I also set the $\beta$-values of `missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.
 
@@
 For a chromosome with $N_{loci}$ CpGs there are $(N_{loci} - 1)$ pairs.
-I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have no intervening loci ($NIL$ = number of intervening loci)[^nil0].
+I call this the `adjacent pairs' or $NIL = 0$ strategy, since all pairs have no intervening loci ($NIL$ = number of intervening loci)[^nil0].
 
@@
 \includegraphics[width=0.8\textwidth]{../figures/feature_status.pdf}
-\caption{Schematic illustrating the `feature status' of four pairs of methylation loci.
+\caption[Schematic of `feature status']{Schematic illustrating the `feature status' of four pairs of methylation loci.
 The circles represent loci and the grey bars a genomic feature, such as CpG islands.
@@
 \includegraphics[width=\textwidth]{../figures/ADS_beta_scatterplots.pdf}
-\caption{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the ADS sample.
+\caption[Scatterplots of pairs of $\beta$-values for \emph{ADS} sample]{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the ADS sample.
 $\beta$-values are computed using strand-collapsed counts.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
-\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+%\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
-\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+%\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
 \label{fig:Lister_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
-\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+%\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs %of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
-\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+%\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
-\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\caption[\emph{EPISCOPE}: Pearson vs.
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
-\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\caption[\emph{Lister}: Pearson vs.
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
 \label{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
-\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\caption[\emph{Seisenberger}: Pearson vs.
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
-\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\caption[\emph{Ziller}: Pearson vs.
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\caption[\emph{EPISCOPE}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
 \label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\caption[\emph{Seisenberger}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\caption[\emph{Ziller}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
+\caption[\emph{EPISCOPE}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
 \label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
+\caption[\emph{Seisenberger}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
+\caption[\emph{Ziller}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
+\caption[\emph{EPISCOPE}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
+\caption[\emph{Lister}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Lister} dataset.}
 \label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
+\caption[\emph{Seisenberger}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
+\caption[\emph{Ziller}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
+\caption[\emph{EPISCOPE}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
+\caption[\emph{Lister}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Lister} dataset.}
 \label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
+\caption[\emph{Seisenberger}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
+\caption[\emph{Ziller}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \centering
+\caption[$2 \times 2$ contingency table notation]{Notation for a $2 \times 2$ contingency table.}
+\label{tab:2x2}
 \begin{tabular}{@{}cccc@{}}
@@
 \end{tabular}
-\caption{Notation for a $2 \times 2$ contingency table.}
-\label{tab:2x2}
 \end{table}
@@
 Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$.
-\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\theta_{MH}})$.
+\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$.
 This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance].
@@
 
-The results of each simulation are summarised by a plot of $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+The results of each simulation are summarised by a plot of $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line[^density].
@@
 To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1, Figure \ref{fig:Simulation_1}), and $K = 1000$ (Simulation 2a, Figure \ref{fig:Simulation_2}).
-Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta_{MH}}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
+Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
 This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_1.pdf}
-\caption{Simulation 1.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 1]{Simulation 1.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_2.pdf}
-\caption{Simulation 2a.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 2a]{Simulation 2a.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
-\caption{Simulation 2b.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 2b]{Simulation 2b.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
-\caption{Simulation 3.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 3]{Simulation 3.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 Furthermore, each methylation locus in a 2-tuple will likely have a similar marginal probability of methylation.
-To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:Beta_marginals}).
+To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}).
 The results of this simulation (Simulation 4, Figure \ref{fig:Simulation_4}) are basically an exaggerated version of Simulation 3.
-The Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.
-
-```{r Beta_marginals, echo = FALSE}
-hist(rbeta(1000, 0.3, 0.2))
-```
+The Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/beta_marginals.pdf}
+\caption[$Beta(0.3, 0.2)$ histogram]{Histogram of 1000 points simulated from a $Beta(0.3, 0.2)$ distribution.}
+\label{fig:beta_marginals}
+\end{figure}
 
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_4.pdf}
-\caption{Simulation 4.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 4]{Simulation 4.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 
-Simulations 5, 6 and 7 (Figures \ref{fig:Simulation_5}, \ref{fig:Simulation_6} and \ref{fig:Simulation_7}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta_{MH}}$, is a reasonable estimator of the centre of the distribution of log odds ratios.
+Simulations 5, 6 and 7 (Figures \ref{fig:Simulation_5}, \ref{fig:Simulation_6} and \ref{fig:Simulation_7}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, is a reasonable estimator of the centre of the distribution of log odds ratios.
 It performs better in Simulation 5 and 6, where the distribution of $\theta$ is symmetric, as opposed to Simulation 7, where the distribution of $\theta$ is skewed.
-Unsurprisingly, $\widehat{\theta_{MH}}$ does a better job as $K \to \infty$.
-It should be emphasised, however, that the asymptotic variance of the estimator, $\hat{\sigma}(\widehat{\theta_{MH}})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$.
-We can see that the asymptotic $95 \%$ confidence interval of $\widehat{\theta_{MH}}$ only covers a small amount of the variation in the $\theta$.
+Unsurprisingly, $\widehat{\theta}_{MH}$ does a better job as $K \to \infty$.
+It should be emphasised, however, that the asymptotic variance of the estimator, $\hat{\sigma}(\widehat{\theta}_{MH})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$.
+We can see that the asymptotic $95 \%$ confidence interval of $\widehat{\theta}_{MH}$ only covers a small amount of the variation in the $\theta$.
 
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_5.pdf}
-\caption{Simulation 5.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 5]{Simulation 5.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_6.pdf}
-\caption{Simulation 6.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 6]{Simulation 6.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_7.pdf}
-\caption{Simulation 7.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 7]{Simulation 7.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 
-Finally, Simulation 8 (Figure \ref{fig:Simulation_8}) shows that if the distribution of $\theta$ is multimodel that $\widehat{\theta_{MH}}$ will estimate the 'average' effect.
+Finally, Simulation 8 (Figure \ref{fig:Simulation_8}) shows that if the distribution of $\theta$ is multimodel that $\widehat{\theta}_{MH}$ will estimate the 'average' effect.
 Of course, the utility of any point estimate is severely reduced when the true effect is multimodal.
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_8.pdf}
-\caption{Simulation 8.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{MH}} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+\caption[Simulation 8]{Simulation 8.
+$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
@@
 I have analysed all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets.
-For each sample, I compute $\widehat{\theta_{MH}}$ using different levels of aggregation (listed here in decreasing order):
+For each sample, I compute $\widehat{\theta}_{MH}$ using different levels of aggregation (listed here in decreasing order):
 
@@
 
-Notably, although the genome-level estimates decay as a function of $IPD$, $\widehat{\theta_{MH}}$ is almost entirely positive over the observable range of $IPD$s, suggesting that within-fragment co-methylation extends for at least a few hundred bp.
-Furthermore, in samples with sufficiently long paired-end reads, there is a hint of an upturn in the genome-wide $\widehat{\theta_{MH}}$ at approximately $IPD = 180$.
+Notably, although the genome-level estimates decay as a function of $IPD$, $\widehat{\theta}_{MH}$ is almost entirely positive over the observable range of $IPD$s, suggesting that within-fragment co-methylation extends for at least a few hundred bp.
+Furthermore, in samples with sufficiently long paired-end reads, there is a hint of an upturn in the genome-wide $\widehat{\theta}_{MH}$ at approximately $IPD = 180$.
 This is approximately the distance between two CpGs on adjacent nucleosomes (see Section \ref{sec:nucleosome_and_chromatin}), and may be evidence of the three-dimensional structure of the genome affecting co-methylation.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI.pdf}
-\caption{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta_{MH}}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
 Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}

### 2015-05-02 Rmarkdown/comethylation_review.Rmd
Fix typo
Add short captions

@@
 \includegraphics[width=0.8\textwidth]{../figures/ng1909-F3a.jpg}
-\caption{Correlation of DNA methylation with spatial distance.
+\caption[Co-methylation figure from \citet{Eckhardt:2006gh}]{Correlation of DNA methylation with spatial distance.
 (a) Correlation between comethylation and spatial distance.
@@
 \includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.pdf}
-\caption{Spacing of Adjacent Methylcytosines in Different Contexts.
+\caption[Co-methylation figure from \citet{Lister:2009hy}]{Spacing of Adjacent Methylcytosines in Different Contexts.
 Prevalence of mCHG/mCHH sites (y-axis) as a function of the number of bases between adjacent mCHG/mCHH sites (x-axis) based on all non redundant pair-wise distances up to $50$ nt in exons, introns and random sequences.
@@
 \includegraphics[width=0.8\textwidth]{../figures/{journal.pbio.1000533.s006}.pdf}
-\caption{Co-methylation patterns for different genomic features. A reproduction of Supplementary Figure 6 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
+\caption[Co-methylation by genomic feature from \citet{Li:2010fb}]{Co-methylation patterns for different genomic features.
+A reproduction of Supplementary Figure 6 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
 \label{fig:Li6}
@@
 \includegraphics[width=0.8\textwidth]{../figures/Figure_S7.pdf}
-\caption{Transformation of methylation correlation at nearby CpG cytosines. A reproduction of Supplementary Figure 7 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
+\caption[Co-methylation figure from \citet{Li:2010fb}]{Transformation of methylation correlation at nearby CpG cytosines.
+A reproduction of Supplementary Figure 7 from \citet{Li:2010fb} under the Creative Common Attribution (CCBY) licence.}
 \label{fig:Li7}
@@
 \includegraphics[width=0.8\textwidth]{../figures/Lyko_Figure_S7.pdf}
-\caption{Periodicity of methylation patterns.
+\caption[Co-methylation figure from \citet{Lyko:2010dr}]{Periodicity of methylation patterns.
 (A) Autocorrelation of CpG methylation status over 1 kb.
@@
 \includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star.pdf}
-\caption{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \emph{FF} sample.}
+\caption[Autocorrelation of incorrectly defined $\beta$-values]{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \emph{FF} sample.}
 \label{fig:FF_chr21_acf_beta_star}
@@
 \includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
-\caption{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample.
+\caption[Fourier transformation of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample.
 Fourier transformation computed and plotted using the \texttt{spectrum} function in R.
@@
 \includegraphics[width=0.8\textwidth]{../figures/Cokus_2008_Sup_Fig_14.pdf}
-\caption{Within-read probability of additional methylation of CHH sites within a given distance from a methylated CHH site.
+\caption[Co-methylation figure from \citet{Cokus:2008fc}]{Within-read probability of additional methylation of CHH sites within a given distance from a methylated CHH site.
 Data were derived from individual BS-Seq reads.
@@
 \centering
+\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple, $(i, i'), i \neq i'$.
+A dot ($\cdot$) indicates that this index is summed over, e.g., $u\cdot_{i, i + 1} = uu_{i, i + 1} + um_{i, i + 1}$.}
+\label{tab:landan}
 \begin{tabular}{@{}cccc@{}}
@@
 \multirow{3}{*}{First CpG} & & & \\
- & $uu_{i, i'}$       & $um_{i,i ''}$       & $u\cdot_{i, i'}$ \\
+ & $uu_{i, i'}$       & $um_{i,i'}$       & $u\cdot_{i, i'}$ \\
  & $mu_{i, i'}$       & $mm_{i, i'}$       & $m\cdot_{i, i'}$ \\
@@
 \end{tabular}
-\caption{Summary of read counts at a CpG 2-tuple, $(i, i'), i \neq i'$. A dot ($\cdot$) indicates that this index is summed over, e.g., $u\cdot_{i, i + 1} = uu_{i, i + 1} + um_{i, i + 1}$.}
-\label{tab:landan}
 \end{table}
@@
 So-called "methylation linkage diagrams" are created by plotting the Pearson correlation of each 2-tuple, $r_{(i, i')}$ as a heatmap (_à la_ heatmaps of linkage disequilibrium) to identify blocks of CpGs with significant co-methylation.
-It should be noted that the concepts of 'individual' and 'population' are rather different for "methylation linkage" than they are for genetic linkage (see Table \ref{tab:linkage}).
+It should be noted that the concepts of 'subject' and 'population' are rather different for "methylation linkage" than they are for genetic linkage (see Table \ref{tab:linkage}).
 
@@
 \centering
+\caption[Notions of 'subject' and 'population' in methylation linkage]{Notions of 'subject' and 'population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}.
+NB 'person' could be any organism, e.g, 'mouse', 'dog', etc.}
+\label{tab:linkage}
 \begin{tabular}{@{}lll@{}}
 \toprule
-                    & Individual               & Population \\ \midrule
+                    & Subject               & Population \\ \midrule
 Genetic linkage     & Person                   & Population \\
@@
 \end{tabular}
-\caption{Notions of 'individual' and 'population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}. NB 'person' could be any organism, e.g, 'mouse', 'dog', etc.}
-\label{tab:linkage}
 \end{table}

### 2015-05-02 Rmarkdown/paper_review.Rmd
typo

@@
 The raw $\beta$-values are weighted according to the binomial likelihood and a tricube kernel.
-The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the tricube kernal gives greater weight to those $\beta_{i, j}$ near the centre of the window.
+The binomial likelihood weights points inversely to their standard error, $se(\beta_{i, j})$, and the tricube kernel gives greater weight to those $\beta_{i, j}$ near the centre of the window.
 

### 2015-05-02 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Table captions up top

@@
                 header = TRUE, stringsAsFactors = FALSE)
-xt <- xtable(x, caption = "Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).", label = "tab:cpg_density_hg19")
+xt <- xtable(x, caption = c("Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).", "CpG percentage per chromosome in hg19"), label = "tab:cpg_density_hg19")
 align(xt) <- rep("c", ncol(x) + 1)
 digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, include.rownames = FALSE)
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, include.rownames = FALSE, caption.placement = "top")
 ```
@@
 `methtuple` is open-source software released under the MIT licence and available from [https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple).
+
+\section{Summary}
+
+__TODO: necessary?__

### 2015-05-02 Rmarkdown/wgbs_statistical_framework.Rmd
Table captions up top

@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:EPISCOPE_strand_correlations")
+xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{EPISCOPE}: Correlations of $\\beta$-values across strands"), label = "tab:EPISCOPE_strand_correlations")
 align(xt) <- rep("c", ncol(x) + 1)
 digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = "top")
 ```
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Lister_strand_correlations")
+xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Lister}: Correlations of $\\beta$-values across strands"), label = "tab:Lister_strand_correlations")
 align(xt) <- rep("c", ncol(x) + 1)
 digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
 ```
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Seisenberger_strand_correlations")
+xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Seisenberger}: Correlations of $\\beta$-values across strands"), label = "tab:Seisenberger_strand_correlations")
 align(xt) <- rep("c", ncol(x) + 1)
 digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
 ```
@@
 colnames(x) <- 1:10
-xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_merged_strand_correlations")
+xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset.
+Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Ziller}: Correlations of $\\beta$-values across strands"), label = "tab:Ziller_merged_strand_correlations")
 align(xt) <- rep("c", ncol(x) + 1)
 digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
 ```

### 2015-05-02 Rmarkdown/wgbs_downstream_analyses.Rmd
Saskia’s suggested changes.

@@
 \begin{chapabstract}
-I discuss methods for the _downstream analysis_ of bisulfite-sequencing data.
+This chapter discussed methods for the \emph{downstream analysis} of bisulfite-sequencing data.
 Downstream analyses proceed the processing of the raw data (Chapter \ref{chap:wgbs_bioinformatics_analysis}) to address the scientific questions of interest.
 Most downstream methods analyse methylation counts at 1-tuples, however, there is growing interest in analyses based on methylation patterns at m-tuples.
-I discuss the statistical questions underlying these downstream analyses.
-I pay particular attention to those based on m-tuples (m $> 1$) since these have received less attention in the literature.
+I discuss the statistical questions underlying these downstream analyses, paying particular attention to those based on m-tuples (m $> 1$) since these have received less attention in the literature.
 
 A barrier to analyses based on m-tuples has been a lack of software.
-To overcome this barrier, I introduce \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
+To overcome this barrier, I develop \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
 \texttt{MethylationTuples} complements \texttt{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples.
@@
 The majority of downstream analysis methods use methylation patterns at 1-tuples, i.e., $\mathbf{m} = (m_{1}, \ldots, m_{i}, \ldots, m_{N_{loci}})$ and $\mathbf{u} = (u_{1}, \ldots, u_{i}, \ldots, u_{N_{loci}})$[^not_beta_values].
-Methods based on 1-tuples have been developed to address several scientific questions including testing for differential methylation (Section \ref{sec:differential_methylation}), testing for differentially variable methylation (Section \ref{sec:variable_methylation}) and identifying regulatory regions of the genome (Section \ref{sec:regulatory_regions}).
+Methods based on 1-tuples have been developed to address several scientific questions including testing for differential methylation (Section \ref{sec:differential_methylation}), testing for differentially variable methylation (Section \ref{sec:variable_methylation}) and identifying regulatory regions of the genome (Section \ref{sec:epigenome_segmentation}).
 
@@
 
-All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\mathbf{\beta}$.
+All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$
 
@@
 Suppose we have a two-group experiment and let $B_{i}^{k}$ denote the true methylation level of the $i^{th}$ methylation locus in the $k^{th}$ group ($k = 0, 1$).
-We wish to test the null hypothesis that $B_{i}^{0} = B_{i}^{1}$ against the alternative hypothesis that $B_{i}^{0} \neq B_{i}^{1}$.
+We wish to test the null hypothesis of $H_{0}: B_{i}^{0} = B_{i}^{1}$ against the alternative hypothesis $H_{1}: B_{i}^{0} \neq B_{i}^{1}$.
 As such, identifying DMCs boils down to identifying differences in means, for which there is an enormous body of statistical literature.
@@
 Regardless of the statistical test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty.
-Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the number of tests, in this case approximately 25 million, is at least an order of magnitude greater than what is commonly tested in other genomics experiments[^multiple_testing].
+Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the number of tests, in this case approximately 25 million, is at least one order of magnitude greater than what is commonly tested in other genomics experiments[^multiple_testing].
 Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.
 
-[^multiple_testing]: For example, there are approximately 20,000 tests in studies of differential gene expression and one million tests in genome-wide association studies.
+[^multiple_testing]: For example, there are approximately 20,000 tests in studies of differential gene expression and two million tests in genome-wide association studies.
 
@@
 Rather, the idea is that a DMR might capture a weaker but broader difference in methylation.
-Such a signal may well have a clearer biological interpretation, such as in differences in methylation at a gene promoter.
+For example, it may be more biologically relevant to identify a broad region with a consistent, albeit smaller, difference in methylation than it is to identify individual cytosines with large differences in methylation.
 
@@
 Suppose we have a two-group experiment and let $\widebar{B_{r}^{k}}$ be the true average level of methylation for the $r^{th}$ region in the $k^{th}$ group ($k = 0, 1$).
-The null hypothesis is that $\widebar{B_{r}^{0}} = \widebar{B_{r}^{1}}$ against the alternative hypothesis that $\widebar{B_{r}^{0}} \neq \widebar{B_{r}^{1}}$.
+The null hypothesis is $H_{0}: \widebar{B_{r}^{0}} = \widebar{B_{r}^{1}}$ against the alternative hypothesis $H_{1}: \widebar{B_{r}^{0}} \neq \widebar{B_{r}^{1}}$.
 
@@
 
-\subsubsection{A hierarchical procedure for testing _a priori_ defined regions for differential methylation}
+\subsubsection{A hierarchical procedure for testing \emph{a priori} defined regions for differential methylation}
 
@@
 Related to the method of creating DMRs by forming clusters of DMCs is that of "bump-hunting".
-Initially developed for application to methylation microarray data \citep{Jaffe:2012gx}, and now available for broader use in the R/Bioconductor package `bumphunter` ([http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html](http://www.bioconductor.org/packages/release/bioc/html/bumphunter.html)), "bump-hunting" may be used to identify DMRs.
+Initially developed for application to methylation microarray data \citep{Jaffe:2012gx}, and now available for broader use in the R/Bioconductor package `bumphunter` ([http://bioconductor.org/packages/bumphunter/](http://bioconductor.org/packages/bumphunter/)), bump-hunting may be used to identify DMRs.
 The idea is as follows.
@@
 This is different to using _a priori_ regions, since the regions are still data-defined, but not with respect to differential methylation.
-The only example I know of such a method is `Aclust` \citep{Sofer:2013bk}.
+The only example of such a method that I am aware of is `Aclust` \citep{Sofer:2013bk}.
 `Aclust` first clusters CpGs into candidate regions by performing agglomerative nested clustering of the between-sample co-methylation.
@@
 These clusters are then tested for differential methylation using generalised estimating equations.
+
+Regardless of the statistical method used to identify differential methylation, it remains important to validate these differences.
+This validation should ideally be performed in a new dataset and perhaps using a different assay in order to mitigate potential biases.
 
@@
 A more subtle difficulty is in dealing with outliers.
-An outlier in one group will greatly increase the variation in that group, but this does not necessarily mean that the locus is differentially methylated.
+An outlier in one group will greatly increase the variation in that group, but this does not necessarily mean that the locus is differentially variably methylated.
 It might, for example, be due to an error in the assay.
@@
 
-The obvious method to detect allele-specific methylation from bisulfite-sequencing requires reads that contaian a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus.
+The obvious method to detect allele-specific methylation from bisulfite-sequencing requires reads that contain a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus.
 The heterozygous variant allows these reads to be separated by the observed allele[^phase], which can then be used to test for allele-specific methylation.
@@
 For example, the cytosine in the sequence _TCGA_ may be methylated or unmethylated; each of the methylated and unmethylated versions of that sequence is an _epiallele_.
+\citet{Rakyan:2002wz} define an _epiallele_ as "an allele that can stably exist in more than one epigenetic state, resulting in different phenotypes".
+The latter requirement, while obviously more interesting than the alternative, may be unduly restrictive.
+After all, we refer to alternative forms of a genetic sequence as _alleles_ regardless of whether we know of a phenotypic consequence of the variant.
+
 Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom.
 Even then, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.
-
 In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}.
-The poster child for the potential important of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}.
-Mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic and have an increased susceptibility to tumours compared to those mice with a methylated version of the allele, which have a _pseudoagouti_[^pseudoagouti_vs_agouti] (brown) coat and none of the associated health defects.
+The poster child for the potential importance of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}.
+Genetically identical mice with different versions of the $A^{vy}$ allele are phenotypically distinct.
+Those mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic, and have an increased susceptibility to tumours; those mice with a methylated version of the allele have a _pseudoagouti_[^pseudoagouti_vs_agouti] (brown) coat and none of the associated health defects.
 
@@
 Restricting our attention to CpG methylation, a sequence with $m$ CpGs has $2^{m}$ potential epialleles.
-In other words, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical[^epiallele_definition].
+In other words, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical.
 An epiallele may also be described as an _epimutation_ if it is different from the 'normal' methylation state.
-
-[^epiallele_definition]: \citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes".
-While most would agree that it is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive.
-After all, we refer to alternate forms of a genetic sequence as alleles regardless of whether we know of a phenotypic consequence of the variant.
 
@@
 
-Of the downstream analyses based on methylation patterns at m-tuples, the study of epialleles has received the most attention with respect to methods and software.
+Of the downstream analyses based on methylation patterns at m-tuples, the study of epialleles has received the most attention with respect to methods and software development.
 
@@
 I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations.
-Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` due to compilation errors.
+Two give two examples, `DMEAS` \citep{He:2013cj} is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` due to compilation errors.
 
@@
 R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code.
-To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
+To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
 
@@
 In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations.
+This makes it easy to use for users already familiar with the `GenomicRanges` package.
 For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \texttt{type = `equal'}).
@@
 
-These are in addition to the many useful methods inherited from the `GenomicRanges::SummarizedExperiment` class.
+These are in addition to the many useful methods inherited from the \linebreak `GenomicRanges::SummarizedExperiment` class.
 

### 2015-05-02 Rmarkdown/appendix.Rmd
Pre-lunch commit. Begin re-write of methsim chapter.
Goknur's suggested changes.

@@
 In particular, in the case $n = 2$ we can compute the probability that two
-dependent but identically distributed Bernoulli random variables are equal.
+dependent and identically distributed Bernoulli random variables are equal.
 
@@
 - others
+
+__TODO: Add citations of all packages used in thesis__

### 2015-05-02 Rmarkdown/comethylation_review.Rmd
Goknur's suggested changes.

@@
 Co-methylation is the dependence structure of DNA methylation data.
-This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation.
-In this chapter I critically review these definitions and put them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}.
-I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
-I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.
+This chapter critically reviews different definitions of co-methylation and puts them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}.
+I highlight various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
+I also review how methods to detect differential methylation have sought to account for, and exploit, co-methylation.
 \end{chapabstract}
@@
 
-Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$.
+Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{h, i}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$.
 For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$.
@@
            Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
-        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
-           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
+        &= \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) +
+           \big [ p_{h'} + c (1 - p_{h}) \big ] p_{h} \\
         &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
@@
 1.
-$PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+$PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h', i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
 2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
@@
 
-The $167$ bp period is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al.
+The $167$ bp period "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al.
 hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.
@@
 \citet{Ball:2009kh} use bisulfite padlock probes with $36$ bp single-end reads to profile CpG methylation at approximately $7,000$ CpGs in an Epstein-Barr virus transformed B-lymphocyte cell line.
-As part of their study, \citeauthor{Ball:2009kh} sought to investigate whether the co-methylation reported by \citet{Eckhardt:2006gh} (Section \label{sec:eckhardt}) occurs at the single-molecule level.
+As part of their study, \citeauthor{Ball:2009kh} sought to investigate whether the co-methylation reported by \citet{Eckhardt:2006gh} occurs at the single-molecule level.
 
@@
 \citet{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.
-The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing (median coverage of each CpG in the target $> 10,000 \times$) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
+The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing ($> 10,000 \times$ median sequencing coverage of each CpG in the target) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
 \citeauthor{Landan:2012kp} also use these data to estimate co-methylation.
 
-As part of their analysis, citeauthor{Landan:2012kp} construct "methylation linkage diagrams" based on a Pearson correlation estimate of within-fragment co-methylation.
+As part of their analysis, \citeauthor{Landan:2012kp} construct "methylation linkage diagrams" based on a Pearson correlation estimate of within-fragment co-methylation.
 For each CpG 2-tuple in their targeted regions, they create a $2 \times 2$ contingency table.
 The general form of the table is shown below in Table \ref{tab:landan}.
-They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{mm_{i, i'} \times uu_{i, i'} - mu_{i, i'} \times um_{i, i'}}{sqrt{u\cdot_{i, i'} \times m\cdot_{i, i'} \times \cdot u_{i, i'} \times m_{i, i'}}}$.
+They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{n_{mm} \times n_{uu} - n_{mu} \times n_{um}}{\sqrt{n_{u+} \times n_{m+} \times n_{+u} \times n_{+m}}}$.
 Importantly, $r$ is based on within-read measurements from a single sample and so these $r$ are estimates of within-read, within-sample co-methylation.
@@
 \centering
-\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple, $(i, i'), i \neq i'$.
-A dot ($\cdot$) indicates that this index is summed over, e.g., $u\cdot_{i, i + 1} = uu_{i, i + 1} + um_{i, i + 1}$.}
+\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple.
+For example, $n_{um}$ is the number of reads methylated at the first CpG and unmethylated at the second CpG.
+We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{u+} = n_{uu} + n_{um}$.}
 \label{tab:landan}
@@
 \multirow{3}{*}{First CpG} & & & \\
- & $uu_{i, i'}$       & $um_{i,i'}$       & $u\cdot_{i, i'}$ \\
- & $mu_{i, i'}$       & $mm_{i, i'}$       & $m\cdot_{i, i'}$ \\
- & $\cdot u_{i, i'}$  & $\cdot m_{i, i'}$  & $d_{i, i'}$      \\ \cmidrule(l){2-4}
+ & $n_{uu}$       & $n_{um}$       & $n_{u+}$ \\
+ & $n_{mu}$       & $n_{mm}$       & $n_{m+}$ \\ \cmidrule(l){2-4}
+ & $n_{+u}$  & $\cdot n_{+m}$  & $n_{++} = n$      \\
 \end{tabular}
@@
 \centering
-\caption[Notions of 'subject' and 'population' in methylation linkage]{Notions of 'subject' and 'population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}.
-NB 'person' could be any organism, e.g, 'mouse', 'dog', etc.}
+\caption[Notions of `subject' and `population' in methylation linkage]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}.
+NB `person' could be any organism, e.g, `mouse', `dog', etc.}
 \label{tab:linkage}
@@
 
-As we will see in Chapter \ref{chap:co-methylation}, this result is contradicted by my analyses of within-fragment co-methylation (__TODO__: Update once Chapter \ref{chap-comethylation} is updated.)
+As we will see in Chapter \ref{chap:co-methylation}, this result is contradicted by my analyses of within-fragment co-methylation using whole-genome data.
 
@@
 A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs.
-the rest of the genome (see Chapter \ref{chap:co-methylation}) (__TODO: Revisit in light of updated results in Co-methylation chapter__).
+the rest of the genome (see Chapter \ref{chap:co-methylation}).
 

### 2015-05-02 Rmarkdown/comethylation.Rmd
Address Terry’s suggestions for co-methlation chapter.

@@
 In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy.
-This is more like a traditional autocorrelation analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$, albeit of a vector of unequally spaced observations and with care taken to only use pairs on the same chromosome and strand.
+This is what is called an 'autocorrelation' analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$ by \citet{Cokus:2008fc}, although it is somewhat different to the traditional notion of an autocorrelation analysis.
 I refer to this as the $NIL \geq 0$ strategy.
@@
 This is a similar scale to the previously reported periodicities in correlations of $\beta$-values (see Chapter \ref{chap:co-methylation_review}).
-While this periodicity can readily be detected by eye, it is somewhat more difficult to verify using traditional methods based on Fourier analysis of the signal.
+While this periodicity can readily be detected by eye, it is somewhat more difficult to verify using classical methods from spectral analysis.
 The chief reason is that the data are very noisy beyond the second or third putative cycle due to the limited data at these larger $IPD$s.
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning].
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning].
 Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.
@@
 
-The odds ratio is not perfece, however.
+The odds ratio is a simple summary of the dependence in a $2 \times 2$ table.
+An attractive feature is that it is independent of the marginal probabilities (__TODO: cite__).
+The odds ratio is not perfect, however.
 For one, the possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios.
@@
 The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler.
-I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.
+I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting here that we can always convert one to the other by appropriate exponentiation or taking of logarithms.
 
@@
 Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}.
-The maximum likelihood estimator of $\psi$, $\widehat{\psi_{C}}$ is solved using iterative methods.
-The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi_{C}}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods.
+The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
 
@@
 
-We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
-
-Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$.
-\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$.
+We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
+
+Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi}_{MH} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$.
+\citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi}_{MH}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$.
 This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance].
@@
 
-\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, converge to the true odds ratio.
-In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi_{MH}}$, and the conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, do convert to $\psi$, with $\widehat{\psi_{MH}}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.
+\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, converge to the true odds ratio.
+In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi}_{MH}$, and the conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, do converge to $\psi$, with $\widehat{\psi}_{MH}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.
 
@@
 
-We analyse the results of a simulation study to explore the effects of sparsity and heterogeneity of odds ratios on estimates of odds ratios in $2 \times 2 \times K$ contingency tables.
+We carry out a simulation study to explore the effects of sparsity and heterogeneity of odds ratios on estimates of odds ratios in $2 \times 2 \times K$ contingency tables.
 The simulation requires that we specify several parameters.
@@
 
-- the number of tables, $K$. The number of 2-tuples.
+- the number of tables, $K$. The number of CpG pairs (2-tuples).
 - the marginal probabilities, $\pi_{2 +} = Pr(Z_{i})$ and $\pi_{+ 2} = Pr(Z_{i'})$.
@@
 The truncated negative binomial distribution is parameterised by the mean ($\mu$), the dispersion parameter (size) and the truncation level (trunc).
-Unless otherwise noted, simulations use $\mu = 30$, $\text{size} = 10$ and $\text{trunc} = 9$, corresponding to an average sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].
+Unless otherwise noted, simulations use $\mu = 30$, $\text{size} = 10$ and $\text{trunc} = 9$, corresponding to an average sequencing depth of $30 \times$ but where only pairs with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].
 
@@
 
-1. The $K$ 0.5-adjusted unconditional odds ratio estimates, $\widehat{\psi_{0.5}}$, from each $2 \times 2$ table.
-2. The unconditional odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\psi_{U}}$.
-3. The Mantel-Haenszel odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\psi_{MH}}$.
-
-The results of each simulation are summarised by a plot of $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
+1. The $K$ 0.5-adjusted unconditional odds ratio estimates, $\widehat{\psi}_{0.5}$, from each $2 \times 2$ table.
+2. The unconditional odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\psi}_{U}$.
+3. The Mantel-Haenszel odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\psi}_{MH}$.
+
+The results of each simulation are summarised by a plot of $\widehat{\theta}_{0.5} = \log(\widehat{\psi}_{0.5})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U} = \log(\widehat{\psi}_{U})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi}_{MH})$ (coloured blue).
 The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line[^density].
 
-[^density]: In many simulations $\psi$ is constant across the $K$ tables and therefore the density is degenerate.
+[^density]: In many simulations $\theta$ is constant across the $K$ tables and therefore the density is degenerate.
 
@@
 
-To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1, Figure \ref{fig:Simulation_1}), and $K = 1000$ (Simulation 2a, Figure \ref{fig:Simulation_2}).
+To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1a, Figure \ref{fig:Simulation_1a}), and $K = 1000$ (Simulation 1b, Figure \ref{fig:Simulation_1b}).
 Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
 This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller.
-
-By contrast, $\widehat{\theta_{0.5}}$ are highly biased estimates of $\theta$, even when $K = 1000$.
-The histograms of $\widehat{\theta_{0.5}}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).
-The bias can be eliminated only by having ultra-deep sequencing coverage ($mean(d) = 10000 \times$), although the variation of $\widehat{\theta_{0.5}}$ across the $K = 1000$ tables remains non-negligible (Simulation 2b Figure \ref{fig:Simulation_2d}).
+By contrast, the estimates of $\theta$ from the individual $2 \times 2$ tables, $\widehat{\theta_{0.5}}$, are highly biased, even when $K = 1000$ in Simulate 1b (Figure \ref{fig:Simulation_1b}).
+In both Simulation 1a and 1b (Figures \ref{fig:Simulation_1a} and \ref{fig:Simulation_1b}) the histograms of $\widehat{\theta}_{0.5}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).
 
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_1.pdf}
-\caption[Simulation 1]{Simulation 1.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
-\label{fig:Simulation_1}
+\caption[Simulation 1a]{Results of simulation 1a ($K = 100$).
+Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
+\label{fig:Simulation_1a}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_2.pdf}
-\caption[Simulation 2a]{Simulation 2a.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
+\caption[Simulation 1b]{Results of simulation 1b ($K = 1000$).
+Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
+\label{fig:Simulation_1b}
+\end{figure}
+
+The bias can be eliminated only by having ultra-deep sequencing coverage, such as in Simulation 1c where $mean(d) = 10000 \times$ (Figure \ref{fig:Simulation_1c}).
+Even then, however, the variation of $\widehat{\theta}_{0.5}$ across the $K = 1000$ tables remains non-negligible when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$ are near the boundaries.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
+\caption[Simulation 1c]{Results of simulation 1c ($K = 1000$).
+Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $10,000$.}
+\label{fig:Simulation_1c}
+\end{figure}
+
+Next, we explore what happens when we allow the marginal probabilities, $\pi_{2 +}$ and $\pi_{+ 2}$, to vary across the $K = 100$ tables for a fixed (possibly non-zero) value of $\theta$.
+To do so, we simulate $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ for each $2 \times 2$ table.
+As is expected from the theory, the results of Simulation 2 show that the Mantel-Haenszel estimator does an excellent job of estimating the true value of $\theta$ (Figure \ref{fig:Simulation_3}).
+By contrast, the estimate formed by collapsing over the $K$ tables, $\widehat{\theta}_{U}$, is well away from the true value and the distribution of $\widehat{\theta}_{0.5}$ is skewed and widely dispersed.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
+\caption[Simulation 2]{Results of simulation 2 ($K = 100$).
+Estimating the log odds ratio, $\theta$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Uniform(0, 1)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
 \label{fig:Simulation_2}
@@
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
-\caption[Simulation 2b]{Simulation 2b.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
-\label{fig:Simulation_2d}
-\end{figure}
-
-Next we explore what happens when we allow the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, to vary across the $K$ tables when $\theta$ is a fixed value identically across the $K$ tables for $K = 100$ ($\theta$ not necessarily zero).
-To do so, we simulate $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ for each $2 \times 2$ table.
-As is expected from the theory, Figure \ref{fig:Simulation_3} shows that the Mantel-Haenszel estimator does an excellent job of estimating the true value of $\theta$.
-By contrast, $\widehat{\theta_{U}}$, is well away from the true value and the distribution of $\widehat{\theta_{0.5}}$ is skewed and widely dispersed.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
-\caption[Simulation 3]{Simulation 3.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
-\label{fig:Simulation_3}
-\end{figure}
-
-In a sense, the $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ assumption is too easy.
-As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are mostly near the boundaries, not uniformly distributed on the unit interval.
-Furthermore, each methylation locus in a 2-tuple will likely have a similar marginal probability of methylation.
-To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}).
-The results of this simulation (Simulation 4, Figure \ref{fig:Simulation_4}) are basically an exaggerated version of Simulation 3.
-The Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.
+In a sense, the $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ assumption in Simulation 2 is too easy.
+As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are not uniformly distributed but, rather, are bimodal and mostly near the boundaries.
+Furthermore, each methylation locus in a pair will likely have a similar marginal probability of methylation.
+To simulate this more relevant scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}).
 
@@
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Simulation_4.pdf}
-\caption[Simulation 4]{Simulation 4.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
-\label{fig:Simulation_4}
-\end{figure}
-
-We now explore what happens when the odds ratio is heterogeneous across the $K$ tables.
-Although results are shown for $\widehat{\theta_{0.5}}$ and $\widehat{\theta_{U}}$, we do not discuss these estimators further since we have seen that these perform poorly in even the simplest simulation scenarios.
-
-In these remaining simulations, all parameters are as for Simulation 4 except that the true log odds ratio of each table, $\theta_{k}$, comes from the following distributions:
-
-- Simulation 5 and Simulation 6: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
-- Simulation 7:$\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
-- Simulation 8: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$.
-
-Simulations 5, 6 and 7 (Figures \ref{fig:Simulation_5}, \ref{fig:Simulation_6} and \ref{fig:Simulation_7}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, is a reasonable estimator of the centre of the distribution of log odds ratios.
-It performs better in Simulation 5 and 6, where the distribution of $\theta$ is symmetric, as opposed to Simulation 7, where the distribution of $\theta$ is skewed.
-Unsurprisingly, $\widehat{\theta}_{MH}$ does a better job as $K \to \infty$.
-It should be emphasised, however, that the asymptotic variance of the estimator, $\hat{\sigma}(\widehat{\theta}_{MH})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$.
-We can see that the asymptotic $95 \%$ confidence interval of $\widehat{\theta}_{MH}$ only covers a small amount of the variation in the $\theta$.
+The results of this simulation (Simulation 3, Figure \ref{fig:Simulation_3}) are basically an exaggerated version of Simulation 2.
+The Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta}_{U} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta}_{0.5}$ is even more widely dispersed and typically not centred around $\theta$.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
+\caption[Simulation 3]{Results of simulation 3 ($K = 100$).
+Estimating the log odds ratio, $\theta$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
+\label{fig:Simulation_3}
+\end{figure}
+
+We now turn our attention to what happens when the odds ratio is heterogeneous across the $K$ tables.
+Although results are shown for $\widehat{\theta}_{0.5}$ and $\widehat{\theta}_{U}$, we do not discuss these estimators further since we have seen that these perform poorly in even the simplest simulation scenarios.
+
+In these remaining simulations, all parameters are as for Simulation 3 except that the true log odds ratio of each table, $\theta_{k}$, comes from the following distributions:
+
+- Simulation 4a and Simulation 4b: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 5:$\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 6: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$.
+
+Simulations 4a, 4b and 5 (Figures \ref{fig:Simulation_4a}, \ref{fig:Simulation_4b} and \ref{fig:Simulation_5}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, is a reasonable estimator of the centre of the distribution of the true log odds ratios, $\theta_{k}$.
+It performs better in Simulation 4a and 4b, where the distribution of the $\theta_{k}$ is symmetric, as opposed to Simulation 5, where the distribution of the $\theta_{k}$ is skewed.
+Unsurprisingly, $\widehat{\theta}_{MH}$ does a better job when $K = 1000$ (Simulation 4b, Figure \ref{fig:Simulation_4b}) than when $K = 100$ (Simulation 4a, Figure \ref{fig:Simulation_4a}).
 
@@
 \includegraphics[width=\textwidth]{../figures/Simulation_5.pdf}
-\caption[Simulation 5]{Simulation 5.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
+\caption[Simulation 4a]{Results of simulation 4a ($K = 100$).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
+\label{fig:Simulation_4a}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_6.pdf}
+\caption[Simulation 4b]{Results of simulation 4b ($K = 1000$).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
+\label{fig:Simulation_4b}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_7.pdf}
+\caption[Simulation 5]{Results of simulation 5 ($K = 100$).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
 \label{fig:Simulation_5}
@@
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Simulation_6.pdf}
-\caption[Simulation 6]{Simulation 6.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
+To take this heterogeneity even further, Simulation 6 (Figure \ref{fig:Simulation_6}) shows that if the distribution of $\theta$ is multimodel, then $\widehat{\theta}_{MH}$ will estimate the 'average' effect.
+Of course, the utility of any point estimate is severely reduced when the true effect is multimodal.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Simulation_8.pdf}
+\caption[Simulation 6]{Results of simulation 5 ($K = 100$).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
 \label{fig:Simulation_6}
@@
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Simulation_7.pdf}
-\caption[Simulation 7]{Simulation 7.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
-\label{fig:Simulation_7}
-\end{figure}
-
-Finally, Simulation 8 (Figure \ref{fig:Simulation_8}) shows that if the distribution of $\theta$ is multimodel that $\widehat{\theta}_{MH}$ will estimate the 'average' effect.
-Of course, the utility of any point estimate is severely reduced when the true effect is multimodal.
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Simulation_8.pdf}
-\caption[Simulation 8]{Simulation 8.
-$\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line.
-See main text for description of other parameters.}
-\label{fig:Simulation_8}
-\end{figure}
+One final note.
+When the true log odds ratios, $\theta_{k}$, is heterogeneous, then the asymptotic variance of the estimator Mantel-Haenszel estiamtor, $\hat{\sigma}(\widehat{\theta}_{MH})^{2}$, is __not__ an estimate of the variance of the true odds ratios, $var(\theta_{k})$.
+We can see in Simulations 4a, 4b, 5 and 6 (Figures \ref{fig:Simulation_4a}, \ref{fig:Simulation_4b}, \ref{fig:Simulation_5} and \ref{fig:Simulation_6}) that the asymptotic $95 \%$ confidence interval of $\widehat{\theta}_{MH}$ only covers a small amount of the variation in the $\theta_{k}$.
 
@@
 
-This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual 2-tuples using the simple estimator odds ratio estimator, $\theta_{0.5}$, unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
+This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual pairs of methylation loci using the simple estimator odds ratio estimator, $\theta_{0.5}$, unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
 The distributions of $\widehat{\theta_{0.5}}$ at typical sequencing depths ($d \approx 30$) are highly biased and dispersed, even under very simple simulation scenarios.
@@
 
-Suppose that we aggregate $K$ 2-tuples to form a $2 \times 2 \times K$ contingency table.
+Suppose that we aggregate $K$ CpG pairs to form a $2 \times 2 \times K$ contingency table.
 When little can be assumed about this $2 \times 2 \times K$ contingency table, the Mantel-Haenszel estimator is the most appropriate of those considered.
@@
 
-Here I describe a simple method to study within-fragment co-methylation by analysing methylation patterns at 2-tuples.
-For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type.
+Here I describe a simple method to study within-fragment co-methylation by analysing methylation patterns at pairs of methylation loci (2-tuples).
+For simplicity, I will use pairs of CpGs, but the method is applicable to any methylation type.
 This method is to be implemented in the `cometh()` function that is part of the `MethylationTuples` software[^cometh].
@@
 
-1. How to define the 2-tuples?
-2. How to aggregate the 2-tuples?
+1. How to define the CpG pairs?
+2. How to aggregate the CpG pairs?
 3. Which estimator to use?
@@
 
-\subsubsection{How to define the 2-tuples}
-
-The construction of 2-tuples is very similar to constructing the pairs of methylation levels, described in Section \ref{sec:beta_pair_construction}.
+\subsubsection{How to define the CpG pairs}
+
+The construction of CpG pairs is very similar to constructing the pairs of methylation levels, described in Section \ref{sec:beta_pair_construction}.
 However, we are now limited to analysing co-methylation of loci that can be captured within a single read.
-With current technology this means we are limited to studying within-fragment co-methylation for 2-tuples with $IPD$ less than approximately $250$.
-As in Section \ref{subsec:beta_cor}, we will consider 2-tuples with $NIL = 0$ and $NIL \geq 0$.
-
-\subsubsection{How to aggregate the 2-tuples?}
-
-The counts of methylation patterns at each 2-tuple can be summarised by a $2 \times 2$ contingency table.
-The sequencing depth of each 2-tuple is typically low.
-Furthermore, for each locus in the 2-tuple, the marginal probability that it is methylated is very often close to zero or one.
+With current technology this means we are limited to studying within-fragment co-methylation for pairs with $IPD$ less than approximately $250$.
+As in Section \ref{subsec:beta_cor}, we will consider pairs with $NIL = 0$ and $NIL \geq 0$.
+
+\subsubsection{How to aggregate the CpG pairs?}
+
+The counts of methylation patterns at each CpG pair can be summarised by a $2 \times 2$ contingency table.
+The sequencing depth of each pair is typically low.
+Furthermore, for each locus in the pair, the marginal probability that it is methylated is very often close to zero or one.
 Therefore, in light of the results in Section \ref{subsec:simulation_study}, we must aggregate these $2 \times 2$ tables in order to perform any useful inference.
 
-The ideal level of aggregation combines those 2-tuples with homogeneous odds ratios.
+The ideal level of aggregation combines those pairs with homogeneous odds ratios.
 Of course, this information is unknown to us.
@@
 We will use the Mantel-Haenszel odds ratio estimator since it is robust to variable marginal probabilities.
-This allows us to combine data for 2-tuples from different regions of the genome, e.g., regions that are hypomethylated, partially methylated and hypermethylated.
+This allows us to combine data for pairs from different regions of the genome, e.g., regions that are hypomethylated, partially methylated and hypermethylated.
 
@@
 $IPD$-CGI: A $2 \times 2 \times K$ table for each $IPD$ and 'CpG island status' (CGI-status) combination.
-The CGI-status of each 2-tuple is whether it is _inside_ a CpG island or _outside_ a CpG island[^spanning].
+The CGI-status of each pair is whether it is _inside_ a CpG island or _outside_ a CpG island[^spanning].
 3. $IPD$-chromosome: A $2 \times 2 \times K$ table for each $IPD$ and chromosome combination.
 
-[^spanning]: 2-tuples spanning the boundary of a CpG island are ignored.
-There are few such 2-tuples when $NIL = 0$.
-While there are considerably more such 2-tuples when $NIL \geq 0$, I have ignored these in favour of simplicity.
-
-No minimum sequencing coverage cutoff is required since the Mantel-Haenszel estimator appropriately downsweights 2-tuples with low sequencing coverage.
+[^spanning]: Pairs spanning the boundary of a CpG island are ignored.
+There are few such pairs when $NIL = 0$.
+While there are considerably more such pairs when $NIL \geq 0$, I have ignored these in favour of simplicity.
+
+No minimum sequencing coverage cutoff is required since the Mantel-Haenszel estimator appropriately downweights pairs with low sequencing coverage.
 
@@
 
-Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using $NIL = 0$ CpG 2-tuples for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using $NIL = 0$ CpG pairs for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_seqnames.pdf}
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+All samples were sequenced with paired-end reads.}
+\label{fig:EPISCOPE_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_MH_by_seqnames.pdf}
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+Only the \emph{ADS}, \emph{ADS-adipose}, \emph{ADS-iPSC} and \emph{H9\_Laurent} samples were sequenced with paired-end reads.}
+\label{fig:Lister_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_seqnames.pdf}
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+All samples were sequenced with paired-end reads.}
+\label{fig:Seisenberger_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+All samples were sequenced with paired-end reads.}
+\label{fig:Ziller_merged_MH_by_seqnames}
+\end{figure}
 
@@
 Overall, however, we see that the genome-level and chromosome-level estimates follow a similar trend for all samples[^mitochondria]: co-methylation is strongest for smaller $IPD$s, decaying fairly rapidly before levelling out by $IPD = 20$ to $50$.
-As the $IPD$ increases, we see more variation in the chromosome-level and genome-level estimates, but this is due to the smaller sample sizes we have for 2-tuples with $NIL = 0$ at larger $IPD$s.
+As the $IPD$ increases, we see more variation in the chromosome-level and genome-level estimates, but this is due to the smaller sample sizes we have for pairs with $NIL = 0$ at larger $IPD$s.
 
@@
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_seqnames.pdf}
-\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
-\label{fig:EPISCOPE_MH_by_seqnames}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_MH_by_seqnames.pdf}
-\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
-\label{fig:Lister_MH_by_seqnames}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_seqnames.pdf}
-\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
-\label{fig:Seisenberger_MH_by_seqnames}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
-\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
-\label{fig:Ziller_merged_MH_by_seqnames}
-\end{figure}
-
 \FloatBarrier
-\subsubsection{Chromosome-level $NIL = 0$ analyses}
-
-The equivalent plots for the $NIL \geq 0$ 2-tuples paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_merged_ac_MH_by_seqnames}.
+\subsubsection{Chromosome-level $NIL \geq 0$ analyses}
+
+The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_merged_ac_MH_by_seqnames}.
 In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_)
 
-- __TODO: Figure out what's going on with some of the EPISCOPE and Ziller samples that have a high autosome (chr21, I think)__
-
-In analysing $NIL \geq 0$ 2-tuples we are increasing the heterogeneity of the $K$ $2 \times 2$ tables, which makes more difficult the interpretation of these results.
-However, it does give the advantage of allowing the analysis of 2-tuples with larger $IPD$s.
-The short DNA fragments and read lengths of our data make it difficult to perform reliable chromosome-level inference in even the best samples for $IPD$s greater than 250$ to $300$.
-Retreating to a genome-level analysis suggests that within-fragment co-methylation is
-
 \begin{figure}[!htb]
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames.pdf}
-\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs.
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
-\textbf{TODO: Figure out why E18BUF is missing chrM data}}
+All samples were sequenced with paired-end reads.}
 \label{fig:EPISCOPE_ac_MH_by_seqnames}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames.pdf}
-\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+Only the \emph{ADS}, \emph{ADS-adipose}, \emph{ADS-iPSC} and \emph{H9\_Laurent} samples were sequenced with paired-end reads.}
 \label{fig:Lister_ac_MH_by_seqnames}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_seqnames.pdf}
-\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+All samples were sequenced with paired-end reads.}
 \label{fig:Seisenberger_ac_MH_by_seqnames}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames.pdf}
-\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+All samples were sequenced with paired-end reads.}
 \label{fig:Ziller_merged_ac_MH_by_seqnames}
 \end{figure}
+
+By analysing $NIL \geq 0$ pairs we are increasing the heterogeneity of the $K$ $2 \times 2$ tables, which makes more difficult the interpretation of these results.
+However, it does give the advantage of allowing the analysis of pairs with larger $IPD$s.
+Nonetheless, the short DNA fragments and read lengths of our data make it difficult to perform reliable chromosome-level inference in even the highest-quality samples for $IPD$s greater than $250$ to $300$.
+Retreating to a genome-level analysis suggests that within-fragment co-methylation is active over at least a few hundred basepairs in distance.
+
+__TODO: Figure out what's going on with some of the EPISCOPE and Ziller samples that have a high autosome (chr21, I think); figure out why E18BUF is missing chrM in $NIL \geq 0$ plots.__
 
@@
 
-Since we must aggregate 2-tuples in order to perform any meaningful analysis of within-fragment co-methylation, there will obviously be unaccounted-for heterogeneity in our estimates.
+Since we must aggregate pairs in order to perform any meaningful analysis of within-fragment co-methylation, there will obviously be unaccounted-for heterogeneity in our estimates.
 In general, it will be difficult to identify sources of this heterogeneity.
@@
 
-We observe that there is very little difference between genome-wide estimates of  $\theta$ inside and outside of CpG islands for 2-tuples with $NIL = 0$ (Figures \ref{fig:EPISCOPE_MH_by_CGI}, \ref{fig:Lister_MH_by_CGI}, \ref{fig:Seisenberger_MH_by_CGI}).
+We observe that there is very little difference between genome-wide estimates of  $\theta$ inside and outside of CpG islands for pairs with $NIL = 0$ (Figures \ref{fig:EPISCOPE_MH_by_CGI}, \ref{fig:Lister_MH_by_CGI}, \ref{fig:Seisenberger_MH_by_CGI}).
 This is perhaps a little surprising; CpG islands are typically more homogeneous in their average methylation levels and so we would reasonably expect their within-fragment methylation states are more dependent than elsewhere in the genome.
@@
 
-When we analyse 2-tuples with $NIL = 0$, we are only estimating the 'first-order' within-fragment co-methylation.
+When we analyse pairs with $NIL = 0$, we are only estimating the 'first-order' within-fragment co-methylation.
 But suppose that the methylation state of the current locus ($Z_{h, i}$) depends not only on the state at the previous locus ($Z_{h, i - 1}$) but also on the states at the two previous states ($Z_{h, i - 2}$ and $Z_{h, i - 3}$).
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_CGI.pdf}
-\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+All samples were sequenced with paired-end reads.}
 \label{fig:EPISCOPE_MH_by_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI.pdf}
-\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+Only the \emph{ADS}, \emph{ADS-adipose}, \emph{ADS-iPSC} and \emph{H9\_Laurent} samples were sequenced with paired-end reads.}
 \label{fig:Lister_MH_by_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI.pdf}
-\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+All samples were sequenced with paired-end reads.}
 \label{fig:Seisenberger_MH_by_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
-\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+All samples were sequenced with paired-end reads.}
 \label{fig:Ziller_merged_MH_by_CGI}
@@
 
-To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of 2-tuples with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}).
+To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of pairs with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}).
 We now see a fairly clear separation of the genome-wide estimates of $\theta$, with CpG islands being more co-methylated than the rest of the genome.
 Moreover, the estimates of $\theta$ within CpG islands are approximately constant with respect to $IPD$.
-Also notable is that the 2-tuples outside of CpG islands have fairly similar estimates of $\theta$ in both the $NIL = 0$ and $NIL \geq 0$ analyses.
+Also notable is that the pairs of CpGs outside of CpG islands have fairly similar estimates of $\theta$ in both the $NIL = 0$ and $NIL \geq 0$ analyses.
 
@@
 
-[^cpg_density2]: The ratio $\frac{NIL > 0 \text{2-tuples}}{NIL \geq 0 \text{2-tuples}}$ is lower outside of CpG islands than inside CpG islands owing to the decreased density of CpGs outside of CpG islands.
+[^cpg_density2]: The ratio $\frac{NIL > 0 \text{pairs}}{NIL \geq 0 \text{pairs}}$ is lower outside of CpG islands than inside CpG islands owing to the decreased density of CpGs outside of CpG islands.
 We therefore can't rule out that the similarity of $NIL = 0$ and $NIL \geq 0$ estimates outside of CpG islands is simply due to a lack of data.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_CGI.pdf}
-\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+All samples were sequenced with paired-end reads.}
 \label{fig:EPISCOPE_ac_MH_by_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_CGI.pdf}
-\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+Only the \emph{ADS}, \emph{ADS-adipose}, \emph{ADS-iPSC} and \emph{H9\_Laurent} samples were sequenced with paired-end reads.}
 \label{fig:Lister_ac_MH_by_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_CGI.pdf}
-\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+All samples were sequenced with paired-end reads.}
 \label{fig:Seisenberger_ac_MH_by_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI.pdf}
-\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ 2-tuples.
-Only the genome-level estimates are shown, stratified by whether the 2-tuple is inside a CpG island (CGI).}
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).
+All samples were sequenced with paired-end reads.}
 \label{fig:Ziller_merged_ac_MH_by_CGI}
@@
 
-[^imr90_ac]: Differences in the $NIL \geq 0$ results between the _IMR90\_r1, _IMR90\_r2_ and _IMR90\_cell\_line_ samples are due to the _IMR90\_cell\_line_ being sequenced with paired-end reads and therefore having many more $NIL > 0$ 2-tuples to analyse.
+[^imr90_ac]: Differences in the $NIL \geq 0$ results between the _IMR90\_r1, _IMR90\_r2_ and _IMR90\_cell\_line_ samples are due to the _IMR90\_cell\_line_ being sequenced with paired-end reads and therefore having many more $NIL > 0$ pairs to analyse.
 
@@
 As a population of cells develops from a single cell, there are errors in copying the DNA methylation patterns from mother to daughter cell.
-These errors disrupt the co-methylation of 2-tuples.
+These errors disrupt the co-methylation of pairs.
 If the iPSC and iPSC-derived cell lines have had less (developmental) time between the laying down of the original DNA methylation pattern and the point at which they were sequenced, then there is a lower probability that the co-methylation has been disrupted by replication errors.
@@
 As noted in Section \ref{sec:m-bias}, it is difficult to properly account for M-bias in pre-trimmed data such as the _Lister_ dataset.
-Since we can do little about data that has already been pre-trimmed, we are forced to be cautious in our interpretation of within-fragment co-methylation for 2-tuples with larger $IPD$s from these datasets.
-This is particularly relevant to the apparent increase in within-fragment co-methylation for $NIL \geq 0$ CpG 2-tuples with $IPD \approx 180$.
+Since we can do little about data that has already been pre-trimmed, we are forced to be cautious in our interpretation of within-fragment co-methylation for pairs with larger $IPD$s from these datasets.
+This is particularly relevant to the apparent increase in within-fragment co-methylation for $NIL \geq 0$ CpG pairs with $IPD \approx 180$.
 This result is tantalisingly similar to the periodicity we see in the correlations of $\beta$-values and the approximate spacing of nucleosomes in human DNA (see Section \ref{sec:nucleosome_and_chromatin}).
 However, it is not as apparent in other paired-end samples, and this is only likely to be resolved with longer reads that can span multiple nucleosomes.
+
+In analysing these data at the genome-level, and even the chromosome-level, we are aggregating measurements from heterogeneous regions of the genome.
+Our simulation results of Section \ref{subsec:simulation_study} tell us that the Mantel-Haenszel estimator estimates the 'average' effect across these potentially heterogeneous regions.
+We might reduce this heterogeneity by aggregating over smaller regions, although choosing the resolution is non-trivial, and so we content ourselves with studying an estimate of the average effect.
 
@@
 
-In summary, the results of this chapter demonstrate that DNA methylation has a  complex dependence structure and identify several drivers of these complexity.
+In summary, the results of this chapter demonstrate that DNA methylation has a  complex dependence structure.
 Co-methylation exists along individual DNA fragments and also manifests as correlations of aggregate measures of methylation.
+We have estimated the effect of this co-methylation using two complementary approaches: correlations of $\beta$-values and within-fragment co-methylation.
+Consistent with previous work, our analysis identifies the inter-pair distance between CpGs and the genomic context of these CpGs, in particular, CpG islands, as being important drivers of co-methylation.
+
+Unique to our results are genome-level and chromosome-level estimates of within-fragment co-methylation that reveal a possible role for nucleosome positioning in determining co-methylation.
+Future work integrating these results with data from assays of nucleosome occupancy may clarify the nature of this relationship.
+Furthermore, stratifying these analyses by other genomic features, such as genic versus intergenic regions, way help elucidate what other factors are at play.
+
 While it is the highest resolution assay for studying DNA methylation, bisulfite-sequencing can still only give a rather limited picture of co-methylation, particularly of within-fragment co-methylation.
-Data from longer reads will be immensely useful in to advancing our understanding of DNA methylation dynamics.
+Data from assays with longer reads will be immensely useful in to advancing our understanding of DNA methylation dynamics.
 In particular, reads that can span multiple nucleosomes will allow for co-methylation to be better tied in with, and expand upon, existing results on the relationship between nucleosomes and DNA methylation.

### 2015-05-02 Rmarkdown/methsim.Rmd
Leaving the library commit

@@
 
-\subsubsection{Parameterising the model}
+\subsubsection{Parameterising the model}\label{subsec:parameterising_the_model}
 
@@
 I have so far only had time to explore a few simple methods and this remains ongoing work.
-We will discuss the results from these five models (m1, m2, m3, m4, m5):
+We will discuss the results from three methods for simulating $\mathbf{B}$ (_m1_, _m2_, and _m3_) and two ways for simulating $\mathbf{\theta}$ (_A_ and _B_):
 
 \begin{description}
-  \item[m1] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
-$\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
-  \item[m2] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this is then perturbed by a $Gaussian(0, \sigma_{B}^2)$ random variable.
+  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
+  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this is then perturbed by a $Gaussian(0, \sigma_{B}^2)$ random variable.
 This is similar to what is done by `WGBSsuite`.
@@
 The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
-$\theta$ simulated as in m1.
-  %\item[m3] $\mathbf{B}$ simulated as in m2 but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5. $\theta$ simulated as in m1.
-  %\item[m4] As for m3 but where the sample is a 50:50 mixture, i.e., two independent simulations of $\mathbf{B}$ and $\mathbf{\theta}$ that are sampled from with equal probability.
-  %\item[m5] As for m3 but with $\mathbf{theta} = 0$ to simulate independence of methylation states within fragments.
+  %\item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5.
+  %\item[\emph{A}] $\theta$ identically zero.
+This is to simulate independence of within-fragment methylation states and as a control to check that `methsim` is working as intended for simulating within-fragment co-methylation.
+  %\item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
 \end{description}
+
+Furthermore, we will consider simulations where the sample is 'pure' (\emph{I}) and where the sample is a mixture of sub-populations with relative frequencies __TODO: Check frequencies__ (\emph{II}).
+We do not explore all combinations of these factors but look at five informative combinations: _m1AI_, _m1BI, _m2BI_, _m3BI_ and _m3BII_.
+We compare these five models to the real _ADS_ data.
 
@@
 To assess this, we compare several summary measures between the simulated data and the real data.
+The simplest of these are summaries of the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`.
 
-__TODO: Add in the _ADS_ data to the simulated data__
+First looking at the distribution of $\beta$-values, we see that all of _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_ do a reasonable job of capturing the bimodality of $\beta$-values (Figure \ref{fig:methsim_plotBetaFreqPoly}) and their relationship to CpG islands (Figure \ref{fig:methsim_plotBetaFreqPoly_CGI}).
+The effect of both the independent perturbations (_m2_) and the correlated perturbations (_m3_) appear to be dominated by variation due to variation in sequencing coverage.
+The lack of within-fragment co-methylation in _m1AI_ does not affect the genome-wide distribution of methylation levels.
+The mixture model, _m3BII_, has noticeably more intermediate $\beta$-values owing to the increased heterogeneity at each CpG across the sub-populations.
 
-The simplest of these are summaries of the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`.
-We see that all of m1, m2, m3, m4 and m5 sufficiently capture the bimodality of $\beta$-values (__FIGUREs__) and their relationship to CpG islands (__FIGUREs__).
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly.pdf}
+\caption[\texttt{methsim} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for \texttt{methsim} simulated data and the \emph{ADS} input sample.
+$\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
+Observations have been combined across strands and only CpGs with at least $10 \times$ sequencing coverage are included.}
+\label{fig:methsim_plotBetaFreqPoly}
+\end{figure}
 
-__FIGURE__ plots the estimates of within-fragment co-methylation.
-Those models that include within-fragment co-methylation (m1, m2, m3 and m4) bear close resemblance to the real _ADS_ data, although are perhaps to homogeneous across chromosomes.
-Reassuringly, we see no evidence of within-fragment co-methylation in model 5, which simulates independence of within-fragment methylation states.
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
+\caption[\texttt{methsim} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island.
+Only CpGs with at least $10 \times$ sequencing coverage are included.
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+\label{fig:methsim_plotBetaFreqPoly_CGI}
+\end{figure}
 
-In contrast, the spatial correlation of $\beta$-values is not explicitly modelled in any of models m1 to m5.
-While m3, m4 and m5 do include some spatial dependence of the $\mathbf{B}$, these are independent of $IPD$ and are intended mostly to analyse the effect of the independence condition in m2.
-__TODO: Describe these crap results.__
+Turning our attention to within-fragment co-methylation, Figure \ref{fig:methsim_methsim_MH_by_seqnames} shows the genome-level and chromosome-level Mantel-Haenszel estimates of within-fragment co-methylation.
+Figure \ref{fig:methsim_MH_by_CGI} shows the genome-level estimates of within-fragment co-methylation stratified by CpG islands.
+Reassuringly, we see no evidence of within-fragment co-methylation in _m1AI_, which simulates independence of within-fragment methylation states.
+The trend of the genome-level estimates, by design, very closely matches that of the _ADS_ input sample.
+We see, however, that the chromosome-level estimates of _m1BI_, _m2BI_, _m3BI_ and _m3BII_ are perhaps a little too homogeneous when compared to the _ADS_ input sample.
+We also see in the results for _m1BI_, _m2BI_, _m3BI_ and _m3BII_ the effect of simulating from a $Gaussian(0, \sigma)$ for $IPD > 180$, as outlined in Section \ref{subsec:parameterising_the_model}.
+The _m3BII_ model again stands out, this time with a noticeably higher level of within-fragment co-methylation.
+Each sub-population has its own $\mathbf{\theta}$ and, although these are sampled from the same parametric model, this suggests that such heterogeneity in the sample will artificially inflate estimates of within-fragment co-methylation, even those made at the genome-level.
+While this requires further investigation, it raises the possibilit that increased genome-level estimates of within-fragment co-methylation may in fact be measuring increased heterogeneity of the sample.
+
+These caveats aside, all five models do quite a good job of capturing the average level of CpG methylation and the strength of within-fragment co-methylation.
+Unfortunately, the results are not so promising for the correlations of $\beta$-values.
+
+In contrast to the other two summaries, the intention with `methsim` was to not explicitly model the spatial correlation of $\beta$-values.
+Rather, the idea was to see how the marginal level of methylation and the within-fragment co-methylation affected these correlations[^ar].
+However, this overlooks that we are already imposing some structure, and therefore correlations, on the $\textbf{B}$  by specifying them in a region-specific manner.
+
+[^ar]: While _m3BI_ and _m3BII_ do include some spatial dependence of the $\mathbf{B}$ via the autoregressive perturbation, these perturbations are independent of $IPD$ and are really intended to analyse the effect of the independence condition in _m2BI_.
+
+Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf} shows the overwhelming nature of this modelling decision.
+When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong.
+Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}).
+This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.
 

### 2015-05-02 Rmarkdown/methsim.Rmd
Fixed a few syntax errors, made a few more

@@
 \begin{chapabstract}
-This chapter describes the `methsim` software to simulate DNA methylation data.
-`methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software.
-We explore different parameter choices and highlight promising directions of research, such as using `methsim` to simulate data in order to compare methods for identifying differential methylation.
+This chapter describes the \texttt{methsim} software to simulate DNA methylation data.
+\texttt{methsim} incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software.
+We explore different parameter choices and highlight promising directions of research, such as using \texttt{methsim} to simulate data in order to compare methods for identifying differential methylation.
 \end{chapabstract}
@@
 
-A popular choice of parametric model in this framework is the beta-binomial model \citep[e.g.,]{Feng:2014iq, Lacey:2013iy, Xu:2013eg, Chen:2014jb, Dolzhenko:2014bo}.
+A popular choice of parametric model in this framework is the beta-binomial model \citep[e.g.,][]{Feng:2014iq, Lacey:2013iy, Xu:2013eg, Chen:2014jb, Dolzhenko:2014bo}.
 This model, and others, are now discussed.
@@
 The second step uses a method published by \citet{Zaykin:2002ko}.
-The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi{C \Phi - 1(1 - B_{j_{k}})}$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\dot)$ denotes the standard normal distribution function.
+The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi{C \Phi - 1(1 - B_{j_{k}})})$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\cdot)$ denotes the standard normal distribution function.
 
@@
 
-A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \citet[e.g.,]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
+A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
 These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth.
@@
 
-[^cases_and_controls] The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at "target" CpGs.
+[^cases_and_controls]: The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at "target" CpGs.
 
@@
 To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ (to three decimal places).
-In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.
+In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.
 
@@
 
-To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimates, $\mathbf{\beta}$ and $\mathbf{\widehat{\psi}}$, in the input methylome.
-More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\mathbf{\psi}$ is based on $\widehat{\psi_{MH}}$.
-We have already seen many examples of the distribution of $\widehat{\psi_{MH}}$ in Chapter \ref{chap:co-methylation}.
+To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimates, $\mathbf{\beta}$ and $\widehat{\mathbf{\psi}}$, in the input methylome.
+More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\mathbf{\psi}$ is based on $\widehat{\psi}_{MH}$.
+We have already seen many examples of the distribution of $\widehat{\psi}_{MH}$ in Chapter \ref{chap:co-methylation}.
 The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.
@@
 
-There are many ways to simulate values of $\mathbf{B}$ and $\mathbf{\theta} = \mathbf{log{\psi}}$ based on these empirical distributions.
+There are many ways to simulate values of $\mathbf{B}$ and $\mathbf{\theta} = \log{\mathbf{\psi}}$ based on these empirical distributions.
 I have so far only had time to explore a few simple methods and this remains ongoing work.
@@
 
-The simulation of $\mathbf{\theta} = \mathbf{log{\psi}}$ requires further explanation.
-The empirical distributions of $\mathbf{\widehat{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
+The simulation of $\mathbf{\theta} = \log{\mathbf{\psi}}$ requires further explanation.
+The empirical distributions of $\widehat{\mathbf{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
 When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?
@@
 For example, we might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence).
-Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta_{MH}}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$[^sigma].
+Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta}_{MH}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$.
+$\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for $IPD \leq 180$.
 While obviously a gross simplification, the vast majority of $NIL = 0$ CpG pairs have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.
-
-[^sigma]: $\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for $IPD \leq 180$.
 
@@
 1. Sample $z_{i}$ from a Bernoulli($B_{i}$) distribution[^mixture].
-2.
-  a. Sample $z_{i + 1}$ from a Bernoulli($p_{i}$) distribution, where $p_{i}$ is the appropriate transition probability given the simulated $z_{i}$.
-  b. Sample $z_{i + 2} from a Bernoulli($p_{i + 1}$) distribution, where $p_{i + 1}$ is the appropriate transition probability given the simulated $z_{i + 1}$.
-3. Each of element of the simulated $z$ is independently flipped with probability $\epsilon$ to simulate sequencing error.
+2. Sample $z_{i + 1}$ from a Bernoulli($p_{i}$) distribution, where $p_{i}$ is the appropriate transition probability given the simulated $z_{i}$.
+3. Sample $z_{i + 2}$ from a Bernoulli($p_{i + 1}$) distribution, where $p_{i + 1}$ is the appropriate transition probability given the simulated $z_{i + 1}$.
+4. Each of element of the simulated $z$ is independently flipped with probability $\epsilon$ to simulate sequencing error.
 
@@
 
-\section{Results}
+\section{Results}\label{sec:methsim_results}
 

### 2015-05-02 Rmarkdown/biology_background.Rmd
Minor edit
Editing

@@
 \begin{chapabstract}
-I summarise the basic biology necessary for understanding my thesis.
+This chapter summarises the basic biology necessary for understanding my thesis.
 I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
@@
 These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix.
-This is illustrated in Figure \ref{fig:dna_simple2}.
+This is illustrated in Figure \ref{fig:double_helix}.
 
@@
 The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin.
-The linker sequences are between $20$ to $60$ basepairs of DNA (bp) in length, while approximately $147$ bp are wrapped around each nucleosome and a further $20$ bp wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
+The linker sequences are between $20$ to $60$ base pairs (bp) of DNA in length, while approximately $147$ bp are wrapped around each nucleosome and a further $20$ bp wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.
 
@@
 
-The length of a chromosome is typically reported as the number of DNA base pairs (bp) in a single copy of that chromosome and the _haploid_ length of a genome is the sum of these chromosome lengths.
+The length of a chromosome is typically reported as the number of DNA base pairs in a single copy of that chromosome and the _haploid_ length of a genome is the sum of these chromosome lengths.
 The haploid human genome is approximately 3 billion base pairs long (Golden Path Length [http://asia.ensembl.org/Homo_sapiens/Location/Genome?r=1](http://asia.ensembl.org/Homo_sapiens/Location/Genome?r=1)).
@@
 In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual.
-Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual (see Chapter \ref{chap:co-methylation}).
+Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual.
 _Epigenomics_ is the study of the epigenome, analogous to genomics being the study of the genome.
@@
 \caption[Chemical structure of 5-methylcytosine]{Chemical structure of 5-methylcytosine.
-"5-methylcytosine".
+``5-methylcytosine''.
 By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
@@
 At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
-Double-stranded DNA, at least at _palindromic_ methylation loci[^palindromic_sequence], is generally symmetrically methylated, i.e., the loci on each strand are both methylated or both unmethylated.
+Double-stranded DNA, at least at _palindromic_ methylation loci[^palindromic_sequence], is generally symmetrically methylated, i.e.
+the loci on each strand are both methylated or both unmethylated.
 However, _hemimethylation_, where the methylation loci on one strand is methylated and its partner on the opposite strand is unmethylated, can and does occur.
@@
 \includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
-\caption[Dinucleotide frequencies in the human genome]{Observed:expected ratio of dinucleotides in the human reference genome (hg19).
+\caption[Dinucleotide frequencies in the human genome]{Observed to expected ratio of dinucleotides in the human reference genome (hg19).
 The expected frequency is computed under an 'independence' model, based on the observed frequencies of each base.}
@@
 \caption[ECDF of distances between CpGs in hg19]{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19).
-The observed distances are contrasted with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
+The observed distances are contrasted with those under the `expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
 \label{fig:IPD_ECDF_hg19}
@@
 
-These 'sandy/beachy' metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.
+These 'sandy/beachy' metaphors have been continued (i.e.
+stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the open sea, CpG deserts and CpG canyons.
 CpG island shores, shelves and the open sea are all defined with respect to CpG islands:
@@
 - CpG island shelves are defined as regions within 2 kb of a CpG island shore \citep{Bibikova:2011fw}.
-- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \citep{Sandoval:2011dj}.
+- The open sea contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \citep{Sandoval:2011dj}.
 
@@
 A subsequent paper from the same group extended this result.
-\citet{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines .
+\citet{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20 \%$ to $30 \%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines.
 Pluripotent cells includes embryonic stem (_ES_) cells and induced pluripotent stem (_iPS_) cell lines.
 
-An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, who found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).
+An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, who found that neurons also have non-CpG methylation, albeit at a lower level ($1.3 \%$ to $1.5 \%$ of all non-CpG cytosines were methylated).
 
@@
 Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development since mouse knockouts[^KO] for either gene are embryonically lethal \citep{Li:1992bp}.
-In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \citep{Li:1992bp}.
+In contrast, mouse knockouts for _DNMT3a_ are runted but survive for approximately 4 weeks after birth \citep{Li:1992bp}.
 
@@
 _DNMT2_, now known as _TRDMT1_, was once thought to be a DNA methyltransferase but was shown to in fact methylate a small RNA and not DNA \citep{Goll:2006kb}.
-Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity.
-Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b \citep{Jurkowska:2011dt}.
+Another protein, _DNMT3L_, is homologous to _DNMT3a_ and _DNMT3b_ but does contain catalytic domain that is necessary for methyltransferase activity.
+Instead, DNMT3L is thought to stimulate the activity of _DNMT3a_ and _DNMT3b_ \citep{Jurkowska:2011dt}.
 
@@
 
-Following treatment, DNA methylation can be assayed using standard techniques such as:
+Following a pre-treatment, DNA methylation can be assayed using standard techniques such as:
 
@@
 Depending on your definition of 'high', a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample.
-The second definition of throughput is  related to cost, be it money or time, i.e., 'how many samples can I afford to analyse?'.
+The second definition of throughput is  related to cost, be it money or time, i.e.
+`how many samples can I afford to analyse?'.
 
@@
 One such genome-wide assays is _CHARM_, 'comprehensive high-throughput arrays for relative methylation' \citep{Irizarry:2008hg}).
-CHARM combines a methylation fractionation step --- be it _MeDIP_, _HpaII_, or as in the original publication, _McrBC_ --- with a tiling array and analysis techniques that leverage regional DNA methylation levels.
+CHARM combines a methylation fractionation step (be it _MeDIP_, _HpaII_, or, as in the original publication, _McrBC_) with a tiling array and analysis techniques that leverage regional DNA methylation levels.
 
@@
 Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples.
-Furthermore, depending on the choice of several sequencing parameters, approximately $35$ to $72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.
+Furthermore, depending on the choice of several sequencing parameters, approximately $35 \%$ to $72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.
 

### 2015-05-02 Rmarkdown/comethylation_review.Rmd
Editing

@@
 This broad definition encompasses several similar concepts previously described in the literature.
-These include: "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}; "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; the "correlation of two [loci] across many samples" \citep{Akulenko:2013fl}; and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
+These include: "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}; "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; the "correlation of two [loci] across many samples" \citep{Akulenko:2013fl}; and the "_vertical_ (i.e.
+progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
 
@@
 
-[^concordance]: One measure of between-sample correlation of aggregate methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$.
+[^concordance]: One measure of between-sample correlation of aggregate methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e.
+$cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$.
 This measure is often reported as evidence for the "concordance" of methylation levels between replicates, but is not what I consider a form of co-methylation.
@@
 1.
-$PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h', i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
-2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
-3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, \allowbreak Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+$PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e.
+$Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h', i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
+2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e. $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
+3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e. $Z_{h, i}, \allowbreak Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
 

### 2015-05-02 Rmarkdown/datasets.Rmd
Editing

@@
 M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set.
-These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e., $|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
+These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e.
+$|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
 3.
@@
 
-__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e., going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $> 275$ GB alone.
+__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e.
+going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $> 275$ GB alone.
 
@@
 The _Lister dataset_ refers to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
-The Lister data were the largest publicly available human WGBS datasets until quite recently.
+The Lister data were the largest publicly available human whole-genome bisulfite-sequencing datasets until quite recently.
 All samples are of human DNA.
@@
 The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \citep{Seisenberger:2012ko}.
-These were a convenience sample provided to me by a colleague, Felix Krueger (Babraham Institude).
-I thank Felix who send my the `BAM` files containing processed and aligned reads.
+These were a convenience sample provided to me by a colleague, Felix Krueger (Babraham Institute).
+I thank Felix who sent me the `BAM` files containing processed and aligned reads.
 
@@
 
-I believe that the samples I received labelled _J1\_1_ and _E16.5\_male\_1_  in fact correspond to _J1\_2_ and _E16.5\_male\_2_, respectively, i.e., the second  replicate rather than the first.
+I believe that the samples I received labelled _J1\_1_ and _E16.5\_male\_1_  in fact correspond to _J1\_2_ and _E16.5\_male\_2_, respectively, i.e.
+the second  replicate rather than the first.
 The data I received are all $100$ bp paired-end sequences, which matches replicate 2 rather than replicate 1 for both of these samples \citet[Supplementary Table 1]{Seisenberger:2012ko}.
@@
 The Ziller are a subset of the data used in \cite{Ziller:2013fs}.
-Specifically, I use a convenience sample of 8 WGBS libraries.
-These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research, Sydney), who already had aligned and processed BAM files for these 19 libraries.
-These libraries in fact represent 19 sequencing runs of eight biological samples, which I was not aware of until the data had already been processed.
-Therefore, I have _post hoc_ merged the processed data from sequencing runs.
-This unfortunately reduces the power to detect genetic variants since the coverage of individual sequencing runs is lower than merged data, but does not adversely affect methylation calling since the number of methylated and unmethylated reads can be summed across sequencing runs.
+Specifically, I use a convenience sample of 8 whole-genome bisulfite-sequencing libraries.
+These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research, Sydney).
+I thank Felix who sent me the `BAM` files containing processed and aligned reads.
+
+\subsection{Sample descriptions}
 
@@
 
+\subsection{Creation of \texttt{BAM} files}
+
+I received 19 `BAM` files from Aaron, which represent 19 sequencing runs of the eight biological samples.
+Each of the 19 `BAM` files was processed separately and I then _post hoc_ merged the processed data from sequencing runs.
+This unfortunately reduces the power to detect genetic variants since the coverage of individual sequencing runs is lower than merged data, but does not adversely affect methylation calling since the number of methylated and unmethylated reads can be summed across sequencing runs.
+
 \section{CpG islands}

### 2015-05-02 Rmarkdown/methsim.Rmd
Editing

@@
 Both the means, $\mu_{i, j^{k}}$, and the dispersions, $\phi_{i, j^{k}}$, are group-specific and are allowed to vary across methylation loci ($i$).
-The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e., the within-group _biological variability_ of DNA methylation.
+The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e.
+the within-group _biological variability_ of DNA methylation.
 
@@
 For example, the average methylation level for all "de-methylated" regions is defined as $B_{'de-methylated'} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$.
-Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e., $B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$.
+Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e.
+$B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$.
 Care needs to be taken that $0 \leq B_{i, j} \leq 1$.
@@
 \includegraphics[width=\textwidth]{../figures/dmr_break.pdf}
-\caption{A hypothetical region in a study of differential methylation in a two-group experiment.
+\caption[A hypothetical differentially methylated region]{A hypothetical region in a study of differential methylation in a two-group experiment.
 Plotted is the difference in $\beta$-value between the two groups ($\delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}).
@@
 
-A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e., without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
+A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e.
+without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
 These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth.
@@
 \includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
-\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
+\caption[Distribution of $\beta$-values stratified by region type]{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 Most paired-end bisulfite-sequencing datasets have overlapping mates, and so paired-end data can be approximated by simply doubling the read length.
-Also, all reads must currently have the same length, i.e., no simulation of read trimming.
+Also, all reads must currently have the same length, i.e.
+no simulation of read trimming.
 I don't consider this feature a priority, but it could easily be implemented by generating read lengths from a given probability distribution.

### 2015-05-02 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Editing

@@
 
-In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions.
+In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e.
+read-positions whose methylation levels are 'too far away' from the vast majority of read-positions.
 It is rather a subjective decision to make.
@@
 In `MethylationTuples`, I use a simple normalisation of the read-position methylation levels ($rpml$).
-Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels, i.e., $nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read-1_.
+Specifically, the median level of methylation across all read-positions is subtracted from the read-position methylation level to create normalised read-position methylation levels, i.e.
+$nrpml_{CpG}^{read_1} = rpml_{CpG}^{read_1} - median(rpml_{CpG}^{read_1})$, for CpG methylation in _read-1_.
 
@@
 
-Methylation calling is the process of calling each sequenced methylation locus as being either methylated or unmethylated[^no_call], as well as determining the _context_ or _type_ of each methylation event (i.e., CpG, CHG or CHH) based on the sequencing data and a reference DNA sequence.
+Methylation calling is the process of calling each sequenced methylation locus as being either methylated or unmethylated[^no_call], as well as determining the _context_ or _type_ of each methylation event (i.e.
+CpG, CHG or CHH) based on the sequencing data and a reference DNA sequence.
 In principle, this is a simple process, however, this belies some complications, which I discuss in this section.
@@
 
-1. Reference-specific methylation loci, i.e., cytosines in the reference genome that are mutated to non-cytosine bases in the sample's genome.
-2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine bases in the reference genome.
+1. Reference-specific methylation loci, i.e. cytosines in the reference genome that are mutated to non-cytosine bases in the sample's genome.
+2. Sample-specific methylation loci, i.e. cytosines in the sample's genome that are non-cytosine bases in the reference genome.
 3. Other genetic variants that may be used in additional analyses, such as in identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).

### 2015-05-02 Rmarkdown/wgbs_downstream_analyses.Rmd
Editing

@@
 
-The majority of downstream analysis methods use methylation patterns at 1-tuples, i.e., $\mathbf{m} = (m_{1}, \ldots, m_{i}, \ldots, m_{N_{loci}})$ and $\mathbf{u} = (u_{1}, \ldots, u_{i}, \ldots, u_{N_{loci}})$[^not_beta_values].
+The majority of downstream analysis methods use methylation patterns at 1-tuples, i.e.
+$\mathbf{m} = (m_{1}, \ldots, m_{i}, \ldots, m_{N_{loci}})$ and $\mathbf{u} = (u_{1}, \ldots, u_{i}, \ldots, u_{N_{loci}})$[^not_beta_values].
 Methods based on 1-tuples have been developed to address several scientific questions including testing for differential methylation (Section \ref{sec:differential_methylation}), testing for differentially variable methylation (Section \ref{sec:variable_methylation}) and identifying regulatory regions of the genome (Section \ref{sec:epigenome_segmentation}).
@@
 
-\subsection{Using \emph{a priori} regions}
+\subsubsection{Using \emph{a priori} regions}
 
@@
 This procedure will create a smoothed $\beta$-value for each CpG, even those with insufficient sequencing coverage.
-Then, for each CpG, `BiSeq` fits a beta regression model[^not_beta-binomial] to the smoothed $\beta$-values, which is tested for evidence of differential methylation at that cytosine (i.e., a test of whether the cytosine is a DMC).
+Then, for each CpG, `BiSeq` fits a beta regression model[^not_beta-binomial] to the smoothed $\beta$-values, which is tested for evidence of differential methylation at that cytosine (i.e.
+a test of whether the cytosine is a DMC).
 
@@
 It aims to first control the false discovery rate at the cluster-level and then refines the signal by trimming non-DMCs from those clusters that have been declared as differentially methylated.
-Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_, i.e., that the differences in methylation are in the same direction.
+Finally, these differentially methylated clusters are _post hoc_ filtered to ensure they are _consistent_, i.e.
+that the differences in methylation are in the same direction.
 
@@
 
-\subsection{Using data-driven regions}
+\subsubsection{Using data-driven regions}
 
@@
 The most common approach is to scan the genome for clusters of DMCs and declare these to be DMRs.
-The initial scan for DMCs will typically use a relaxed statistical significance threshold (i.e., below genome-wide significance).
+The initial scan for DMCs will typically use a relaxed statistical significance threshold (i.e.
+below genome-wide significance).
 Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
@@
 For example, \citet{Hansen:2011gu} start by testing all CpGs for differential methylation and retain all those with a P-value in the lowest $5\%$.
-They then declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction" (i.e., the region is consistent).
+They then declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction" (i.e.
+the region is consistent).
 These putative DMRs may be subject to further filtering, such as requiring that they contain a minimum number of CpGs and span a minimum number of bases, and the merging of nearby putative DMRs into a single putative DMR \citep{Hansen:2011gu}.
@@
 The rate of _epipolymorphism_ of a locus is defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}[^epipolymorphism].
-\citeauthor{Landan:2012kp} define the rate of epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
+\citeauthor{Landan:2012kp} define the rate of epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e.
+the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].
 Unlike genetic polymorphisms, where the population is typically a set of chromosomes from multiple individuals, the population of epipolymorphisms is often within an individual, even within a tissue within an individual.
@@
 This makes it easy to use for users already familiar with the `GenomicRanges` package.
-For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \texttt{type = `equal'}).
+For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e.
+\texttt{type = `equal'}).
 Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.
@@
 The `MethPat` class extends the `GenomicRanges::SummarizedExperiment`[^namespace] class but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples.
-Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size (i.e., same $\text{m}$).
+Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size (i.e.
+same $\text{m}$).
 

### 2015-05-02 Rmarkdown/wgbs_statistical_framework.Rmd
Minor re-write
Editing

@@
 \begin{chapabstract}
-I set out a statistical framework for analysing bisulfite-sequencing data.
+This chapter sets out a statistical framework for analysing bisulfite-sequencing data.
 The ideas here are simple, however, they have not yet been put into a unified mathematical framework.
@@
 The effect of the third assumption is minor.
-When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e., $(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
+When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e.
+$(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
 However, when using paired-end sequencing, the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur:
@@
 Furthermore, sequencing coverage varies from sample-to-sample.
-This means that even if the samples have exactly the same $\mathcal{I}_{j}$, i.e., the samples are genetically identical, each sample will have a different set of loci with 'sufficient' sequencing coverage.
+This means that even if the samples have exactly the same $\mathcal{I}_{j}$, i.e.
+the samples are genetically identical, each sample will have a different set of loci with 'sufficient' sequencing coverage.
 Loci without sufficient sequencing coverage are effectively missing data.
@@
 1.
-The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{d}}$.
+The estimate of the standard error depends on the estimate of the mean (i.e.
+$\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{d}}$.
 Taking the derivative of this with respect to $\beta$, we see that the maximum standard error, $\sqrt{\frac{0.25}{d}}$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
@@
 \citet{Du:2010dc} advocate for the use of $\mathcal{M}$-values for conducting differential methylation analysis from microarray data[^interpretation].
-The main reason they advocate for the use of $\mathcal{M}$-values is that they are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values.
+The main reason they advocate for the use of $\mathcal{M}$-values is that they are approximately _homoscedastic_, i.e.
+their variances are approximately constant across the full range of $\mathcal{M}$-values.
 As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformaticians and genomics researchers makes it a favourable choice.

### 2015-05-03 Rmarkdown/appendix.Rmd
Add software to appendix

@@
 
-## Software details
+\section{Software details}\label{app:software_details}
 
-__TODO__
+The analyses described in my thesis made extensive use of R (`R Under development (unstable) (2014-10-29 r66891)` and `R version 3.2.0 (2015-04-16)`) and Python (`v2.6.6`).
+All analyses were run on one of the servers described above.
+The scripts used to prepare the results for each chapter are available from \url{https://github.com/PeteHaitch/phd_thesis_analyses}.
+Further details of software used are available in Appendix \ref{app:software_details}.
 
-- R version
-- `MethylationTuples`
-- `methsim`
-- `methtuples`
-- others
+I developed several pieces of software during my PhD. These are listed in below, along with the version used in for analyses in my thesis:
 
-__TODO: Add citations of all packages used in thesis__
+- `methtuple` (`v1.4.0`)
+- `GenomicTuples` (`v1.2.1`)
+- `MethylationTuples` (`v0.3.0.9007`)
+- `methsim` (`v0.5.0.9013`)
+
+In addition, I made use of the R packages listed in Table \ref{tab:R_packages} and gratefully acknowledge the developers.
+
+```{r R_packages, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
+library(xtable)
+x <- read.table("../data/r_packages.txt",  header = TRUE,
+                stringsAsFactors = FALSE, sep = "\t")
+# Drop packages with "*" column set, which indicates the package is a
+# "dependency of a dependency".
+x <- x[x[, 2] != "*", ]
+x <- x[, -2]
+xt <- xtable(x, caption = c("R packages used in thesis as reported by \texttt{devtools::session\\_info()}", "R packages used in thesis"),
+             label = "tab:R_packages")
+align(xt) <- rep("c", ncol(x) + 1)
+print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable",
+      floating = FALSE, include.rownames = FALSE, caption.placement = "top")
+```

### 2015-05-03 Rmarkdown/comethylation.Rmd
Editing
Knocking off TODOs and formatting glitches

@@
 The odds ratio is a simple summary of the dependence in a $2 \times 2$ table.
-An attractive feature is that it is independent of the marginal probabilities (__TODO: cite__).
+An attractive feature is that it is independent of the marginal probabilities.
 The odds ratio is not perfect, however.
@@
 \citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.
-
-- __TODO: Find the reference (I think it's Agresti) for the discussion on the utility (or lack thereof) of homogeneity tests; i.e., rejecting the null is good evidence the null is wrong but failing to reject the null is not strong evidence that it is not wrong.__
+It has been noted, however, that statistical tests of homogeneity versus heterogeneity typically have low power and are only able to detect large departures form homogeneity \cite[e.g.,][]{Hauck:1989vt}.
 
@@
 \includegraphics[width=\textwidth]{../figures/beta_marginals.pdf}
-\caption[$Beta(0.3, 0.2)$ histogram]{Histogram of 1000 points simulated from a $Beta(0.3, 0.2)$ distribution.}
+\caption[Histogram of $Beta(0.3, 0.2)$ random variables]{Histogram of 1000 points simulated from a $Beta(0.3, 0.2)$ distribution.}
 \label{fig:beta_marginals}
@@
 
-__TODO: Figure out what's going on with some of the EPISCOPE and Ziller samples that have a high autosome (chr21, I think); figure out why E18BUF is missing chrM in $NIL \geq 0$ plots.__
-
 \FloatBarrier

### 2015-05-03 Rmarkdown/comethylation_review.Rmd
Knocking off TODOs and formatting glitches

@@
 It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
-While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
+While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{sec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
 

### 2015-05-03 Rmarkdown/datasets.Rmd
Knocking off TODOs and formatting glitches

@@
 Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see \ref{sec:MethylationTuples}).
-Specifically, a `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples, 2ac-tuples (2-tuples with `--all-combinations`), 3-tuples and 4-tuples.
+Specifically, a `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples and 2ac-tuples (2-tuples with the  `--all-combinations` flag in `methtuple` set), 3-tuples and 4-tuples.
 I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.
 
-__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e.
-going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $> 275$ GB alone.
+The raw data for the _Lister_, _Seisenberger and _EPISCOPE_ datasets are all publicly available.
+The _EPISCOPE_ data are not yet published and I do not have permission to make these publicly available.
+The scripts used to prepare the results for each chapter are available from \url{https://github.com/PeteHaitch/phd_thesis_analyses}.
+Further details of software used are available in Appendix \ref{app:software_details}.
 

### 2015-05-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Knocking off TODOs and formatting glitches

@@
 
-I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{section:methtuple}).
+I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{sec:methtuple}).
 
@@
 In fact, a slightly cruder procedure is typically used whereby the entire ends of reads are removed.
-For example, suppose we have $100$ bp reads with M-bias observed at positions $1-4, 9, 94, 96-100$, then read-positions $1-9$ and $94-100$ would be ignored when methylation calling.
+For example, suppose we have $100$ bp reads with M-bias observed at positions $1, 2, 3, 4, 9, 94, 96, 97, 98, 99, 100$, then read-positions $1-9$ and $94-100$ would be ignored when methylation calling.
 Both `bismark_methylation_extractor` and `Bis-SNP`, two popular methylation callers, use this method.

### 2015-05-03 Rmarkdown/wgbs_downstream_analyses.Rmd
Knocking off TODOs and formatting glitches

@@
 
-[^not_beta_values]: It is insufficient to use $\mathbf{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).
+[^not_beta_values]: It is insufficient to use $\bm{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).
 
@@
 Most are limited to analysing two-group experiments.
-Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of $\mathbf{\beta}$ (Section \ref{sec:transformations}).
+Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of $\bm{\beta}$ (Section \ref{sec:transformations}).
 
@@
 
-All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\mathbf{\beta}$.
+All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\bm{\beta}$.
 
@@
 
-Methylation data in the form of the vector of $\beta$-values, $\mathbf{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}.
+Methylation data in the form of the vector of $\beta$-values, $\bm{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}.
 One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs).

### 2015-05-03 Rmarkdown/methsim.Rmd
Add methsim figures

@@
 In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific.
-That is, we want to specify $B_{i, j^{k}}$, where $j^{k}$ indicates that sample $j$ comes from group $k$.
-For non-differentially methylated loci all $k$ groups have identical $B_{i, j^{k}} = B_{i, j^{0}}$; differentially methylated loci are simulated by setting $B_{i, j^{k}} \neq B_{i, j^{k'}}$ for some $k \neq k'$.
+That is, we want to specify $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$.
+For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}} = B_{i, j^{0}}$; differentially methylated loci are simulated by setting $B_{i, j_{k}} \neq B_{i, j^{k'}}$ for some $k \neq k'$.
 
-Under the beta-binomial model, the true methylation level, $B^{i, j}$, is assumed to follow a $Beta(\mu_{i, j^{k}}, \phi_{i, j^{k}})$ distribution, where $\mu_{i, j^{k}}$ is the mean and $\phi_{i, j^{k}}$ is the dispersion of the beta distribution[^beta_parameterisation].
-Both the means, $\mu_{i, j^{k}}$, and the dispersions, $\phi_{i, j^{k}}$, are group-specific and are allowed to vary across methylation loci ($i$).
+Under the beta-binomial model, the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j_{k}}, \phi_{i, j_{k}})$ distribution, where $\mu_{i, j_{k}}$ is the mean and $\phi_{i, j_{k}}$ is the dispersion of the beta distribution[^beta_parameterisation].
+Both the means, $\mu_{i, j_{k}}$, and the dispersions, $\phi_{i, j_{k}}$, are group-specific and are allowed to vary across methylation loci ($i$).
 The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e.
@@
 Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset.
-For example, the average methylation level for all "de-methylated" regions is defined as $B_{'de-methylated'} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$.
+For example, the average methylation level for all "de-methylated" regions is defined as $B_{de-methylated} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$.
 Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e.
@@
 
-As we have seen, the simulation of a differentially methylated locus is straightforward; for a given locus ($i$) simply vary $B_{i, j_{k}}$ across the $k$ groups.
+As we have seen, the simulation of a differentially methylated locus is straightforward; for the $i^{th}$ locus, simply vary $B_{i, j_{k}}$ across the $k$ groups.
 The simulation of a differentially methylated region is more complex.
@@
 \end{figure}
+
+\FloatBarrier
 
@@
 The second step uses a method published by \citet{Zaykin:2002ko}.
-The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi{C \Phi - 1(1 - B_{j_{k}})})$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\cdot)$ denotes the standard normal distribution function.
+The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi \Big \{ C \Phi^{-1}(1 - B_{j_{k}}) \Big \}$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\cdot)$ denotes the standard normal distribution function.
 
@@
 A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e.
-without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }.
+without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb}.
 These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth.
@@
 \citet{Sofer:2013bk} sample from a dataset of 539 Illumina 450k microarrays and select a small number of "target" CpGs, which are CpGs whose methylation level is highly variable across the 539 samples.
-If a block does not contain a "target" then it is sampled uniformly at random from the 539 samples.
-However, if a block does contain a "target" then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the "target" CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the "target" CpG[^cases_and_controls].
+If a block does not contain a target then it is sampled uniformly at random from the 539 samples.
+However, if a block does contain a target then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the target CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the target CpG[^cases_and_controls].
 This is essentially weighted sampling of the real data to induce differential methylation.
 
-[^cases_and_controls]: The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at "target" CpGs.
+[^cases_and_controls]: The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at target CpGs.
 
@@
 As we have seen, DNA methylation is highly heterogeneous along the genome.
-Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains.
+Nonetheless, there are clear regions of `similarity', such as the unmethylated CpG islands and long partially methylated domains.
 In these locally-similar regions we might hope to model DNA methylation by a simple parametric model.
@@
 
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
+\caption[\emph{EPISCOPE}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{EPISCOPE} dataset.}
+\label{fig:EPISCOPE_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
+\caption[\emph{Lister}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Lister} dataset.}
+\label{fig:Lister_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
+\caption[\emph{Seisenberger}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
+\label{fig:Seisenberger_beta_by_pm_boxplots}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
+\caption[\emph{Ziller}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Ziller} dataset.}
+\label{fig:Ziller_beta_by_pm_boxplots}
+\end{figure}
+
 While a tailored algorithm may improve this segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting the methylome into the required 'regions of similarity'.
+
+\FloatBarrier
 
@@
 
-Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$.
-We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$.
+Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\bm{\psi} = \frac{Pr(Z_{i + 1} = 1 | Z_{i} = 1) \times Pr(Z_{i + 1} = 0 | Z_{i} = 0)}{Pr(Z_{i + 1} = 1 | Z_{i} = 0) \times Pr(Z_{i + 1} = 0 | Z_{i} = 1)} = \frac{Pr(Z_{i} = 1, Z_{i + 1} = 1) \times Pr(Z_{i} = 0 , Z_{i + 1} = 0)}{Pr(Z_{i} = 1 , Z_{i + 1} = 0) \times Pr(Z_{i} = 0 , Z_{i + 1} = 1)}$.
+We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\bm{\psi}$.
 
-For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}, B_{i + 1}$, and the odds ratio, $\psi_{i}$.
+For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}$ and $B_{i + 1}$, and the odds ratio, $\psi_{i}$.
 The general form of $P_{i, i + 1}$ is shown below:
@@
 
-Let $B_{i} = 0.7$, $B_{i + 1} = 0.6$ and $\psi_{i, i + 1} = 2$. To begin the iterative proportional fitting algorithm, form the matrix $P_{i, i + 1}^{(0)} = \Bigl(\begin{smallmatrix}
-\psi & 1 \\
+Let $B_{i} = 0.7$, $B_{i + 1} = 0.6$ and $\psi_{i} = 2$. To begin the iterative proportional fitting algorithm, form the matrix $P_{i, i + 1}^{(0)} = \Bigl(\begin{smallmatrix}
+\psi_{i} & 1 \\
    1 & 1
 \end{smallmatrix} \Bigr)$.
-This matrix has the desired cross-ratio, $\psi$, but not the desired row and column margins.
-At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi$ while the row and column margins converge towards the desired values.
+This matrix has the desired cross-ratio, $\psi_{i}$, but not the desired row and column margins.
+At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi_{i}$ while the row and column margins converge towards the desired values.
 The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P_{i, i + 1}^{(*)} \equiv P_{i, i + 1}$.
@@
 
-To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimates, $\mathbf{\beta}$ and $\widehat{\mathbf{\psi}}$, in the input methylome.
-More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\mathbf{\psi}$ is based on $\widehat{\psi}_{MH}$.
-We have already seen many examples of the distribution of $\widehat{\psi}_{MH}$ in Chapter \ref{chap:co-methylation}.
+To simulate values of $\mathbf{B}$ and $\bm{\psi}$, `methsim` uses the empirical distributions of their estimates, $\bm{\beta}$ and $\widehat{\bm{\theta}} = \log(\widehat{\bm{\psi}})$, in the input methylome.
+More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\bm{\theta}$ is based on $\widehat{\theta}_{MH}$.
+We have already seen many examples of the distribution of $\widehat{\theta}_{MH}$ in Chapter \ref{chap:co-methylation}.
 The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.
@@
 
-There are many ways to simulate values of $\mathbf{B}$ and $\mathbf{\theta} = \log{\mathbf{\psi}}$ based on these empirical distributions.
+There are many ways to simulate values of $\mathbf{B}$ and $\bm{\theta} = \log{\bm{\psi}}$ based on these empirical distributions.
 I have so far only had time to explore a few simple methods and this remains ongoing work.
-We will discuss the results from three methods for simulating $\mathbf{B}$ (_m1_, _m2_, and _m3_) and two ways for simulating $\mathbf{\theta}$ (_A_ and _B_):
+We will discuss the results from three methods for simulating $\mathbf{B}$ (_m1_, _m2_, and _m3_) and two ways for simulating $\bm{\theta}$ (_A_ and _B_):
 
@@
   \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
-  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this is then perturbed by a $Gaussian(0, \sigma_{B}^2)$ random variable.
-This is similar to what is done by `WGBSsuite`.
-Here we use $\sigma_{B}^{2} = B \times (1 - B)$.
+  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this average is then perturbed for each locus by a $Gaussian(0, \sigma_{B}^2)$ random variable.
+This is similar to what is done by \texttt{WGBSsuite}.
+Here I have used $\sigma_{B}^{2} = B \times (1 - B)$.
 The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
@@
 
-Furthermore, we will consider simulations where the sample is 'pure' (\emph{I}) and where the sample is a mixture of sub-populations with relative frequencies __TODO: Check frequencies__ (\emph{II}).
-We do not explore all combinations of these factors but look at five informative combinations: _m1AI_, _m1BI, _m2BI_, _m3BI_ and _m3BII_.
+Furthermore, we will consider simulations where:
+
+\begin{description}
+  \item[\emph{I}] The sample is 'pure' and all haplotypes have the same parameters.
+  \item[\emph{II}] The sample is a mixture of subpopulations with different parameters. We will consider a sample with four subpopulations with relative frequencies equal to $(0.6, 0.25, 0.1, 0.05)$.
+\end{description}
+
+We do not explore all combinations of these factors in the results presented here but look at five informative combinations: _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_.
 We compare these five models to the real _ADS_ data.
+The limitations of these models may already be clear, and we will discuss these further in Section \ref{sec:methsim_results}.
 
-The simulation of $\mathbf{\theta} = \log{\mathbf{\psi}}$ requires further explanation.
-The empirical distributions of $\widehat{\mathbf{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
-When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?
+The simulation of $\bm{\theta} = \log{\bm{\psi}}$ requires further explanation.
+The empirical distributions of $\widehat{\bm{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details).
+When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these.
 
 Again, there are several options.
-For example, we might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence).
+For example, we might assume a parametric form for the distribution of $\bm{\theta}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence).
 Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta}_{MH}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$.
-$\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for $IPD \leq 180$.
-While obviously a gross simplification, the vast majority of $NIL = 0$ CpG pairs have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.
+$\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for pairs with $IPD \leq 180$.
+While obviously a gross simplification, the vast majority of CpG pairs with $NIL = 0$ have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.
 
-There is an increase in complexity for models m1 through to m4, while m5 is there as a sanity check that the within-fragment co-methylation model is working as intended.
-The limitations of these models may already be clear, but we will discuss these further in Section \ref{sec:methsim_results}.
+\FloatBarrier
 
@@
 The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
-We will use $200$ bp, single-end reads with an average coverage of $30 \times$ in what follows.
+We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $30 \times$ in what follows.
 
@@
 
-The result of this process is a `SimulatedBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states.
+The result of this process is a `SimulatedWGBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states.
 
@@
 
-Biological variation is controlled through the `MethylomeParam` object and the method with which we simulate $\mathbf{B}$ and $\mathbf{\psi}$ through the `simulate()` method.
+Biological variation is controlled through the `MethylomeParam` object and the method with which we simulate $\mathbf{B}$ and $\bm{\psi}$ through the `simulate()` method.
 Technical variation is controlled the `WGBSParam` object.
@@
 We might also consider more subtle ways of introducing variation.
-For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\mathbf{\psi}$ between calls to `simulate()`.
+For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\bm{\psi}$ between calls to `simulate()`.
 This avenue remains to be explored.
@@
 
-Much of `methsim` is written in C++, using the wonderful Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures.
+While `methsim` is an R package, much of it is written in C++, using the Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures.
 Furthermore, several steps of the simulation can be run in parallel for each chromosome.
@@
 
-The `SimulatedMethylome` and `SimulatedBS` objects are large in memory.
+The `SimulatedMethylome` and `SimulatedWGBS` objects are large in memory.
 This is to some extent unavoidable; we are simulating whole-genome bisulfite-sequencing sequencing experiments that produce an enormous amount of data.
@@
 
-First looking at the distribution of $\beta$-values, we see that all of _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_ do a reasonable job of capturing the bimodality of $\beta$-values (Figure \ref{fig:methsim_plotBetaFreqPoly}) and their relationship to CpG islands (Figure \ref{fig:methsim_plotBetaFreqPoly_CGI}).
+Firstly, looking at the distribution of $\beta$-values, we see that all of _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_ do a reasonable job of capturing the bimodality of $\beta$-values (Figure \ref{fig:methsim_plotBetaFreqPoly}) and their relationship to CpG islands (Figure \ref{fig:methsim_plotBetaFreqPoly_CGI}).
 The effect of both the independent perturbations (_m2_) and the correlated perturbations (_m3_) appear to be dominated by variation due to variation in sequencing coverage.
@@
 Only CpGs with at least $10 \times$ sequencing coverage are included.
-`Spikes' in the density estimate are due to the discreteness of $\beta$-values.
+`Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
 \label{fig:methsim_plotBetaFreqPoly_CGI}
@@
 
-Turning our attention to within-fragment co-methylation, Figure \ref{fig:methsim_methsim_MH_by_seqnames} shows the genome-level and chromosome-level Mantel-Haenszel estimates of within-fragment co-methylation.
+Turning our attention to within-fragment co-methylation, Figure \ref{fig:methsim_MH_by_seqnames} shows the genome-level and chromosome-level Mantel-Haenszel estimates of within-fragment co-methylation.
 Figure \ref{fig:methsim_MH_by_CGI} shows the genome-level estimates of within-fragment co-methylation stratified by CpG islands.
+These results are only available for CpG pairs with $NIL = 0$ because the `asMethPat()` function does not create $NIL > 0$ pairs.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methsim_MH_by_seqnames.pdf}
+\caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+\label{fig:methsim_MH_by_seqnames}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methsim_MH_by_CGI.pdf}
+\caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs.
+Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).}
+\label{fig:methsim_MH_by_CGI}
+\end{figure}
+
 Reassuringly, we see no evidence of within-fragment co-methylation in _m1AI_, which simulates independence of within-fragment methylation states.
@@
 The _m3BII_ model again stands out, this time with a noticeably higher level of within-fragment co-methylation.
-Each sub-population has its own $\mathbf{\theta}$ and, although these are sampled from the same parametric model, this suggests that such heterogeneity in the sample will artificially inflate estimates of within-fragment co-methylation, even those made at the genome-level.
+Each sub-population has its own $\bm{\theta}$ and, although these are sampled from the same parametric model, this suggests that such heterogeneity in the sample will artificially inflate estimates of within-fragment co-methylation, even those made at the genome-level.
 While this requires further investigation, it raises the possibilit that increased genome-level estimates of within-fragment co-methylation may in fact be measuring increased heterogeneity of the sample.
@@
 
-Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf} shows the overwhelming nature of this modelling decision.
+Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming nature of this modelling decision.
 When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong.
-Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}).
+Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}).
 This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
+\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
+\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
+caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
+\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\end{figure}
+
+\FloatBarrier
 

### 2015-05-03 latex/phd_thesis.tex
Editing
Add abstract

@@
 
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 
@@
 
-% should be between 300 - 500 words
+DNA methylation is an epigenetic modification that plays an important role in X-chromosome inactivation, genomic imprinting and the repression of repetitive elements in the genome.
+It must be tightly regulated for normal mammalian development and aberrant DNA methylation is strongly associated with many forms of cancer.
 
-\textbf{TODO}
+This thesis examines the statistical and computational challenges raised by high-throughput assays of DNA methylation, particularly the current gold standard assay of whole-genome bisulfite-sequencing.
+Using whole-genome bisulfite-sequencing, we can now measure DNA methylation at individual nucleotides across entire genomes.
+These experiments produce vast amounts of data that require new methods and tools to analyse.
+
+The first half of the thesis outlines the biological questions of interest in studying DNA methylation, the bioinformatics analysis of these data, and the statistical questions we seek to address.
+In discussing these bioinformatics challenges, we develop software to facilitate novel analyses of these data.
+We pay particular attention to analyses of methylation patterns along individual DNA fragments, a novel feature of sequencing-based assays.
+
+The second half of the thesis focuses on \emph{co-methylation}, the spatial dependence of DNA methylation.
+We demonstrate that previous analyses of co-methylation have been limited by inadequate data and analysis methods.
+This motivates a study of co-methylation from 40 whole-genome bisulfite-sequencing samples.
+These 40 samples represent a diverse range of tissues, from embryonic and induced pluripotent stem cells, through to somatic cells and tumours.
+Making use of software developed in the first half of the thesis, we explore different measures of co-methylation and relate these to one another.
+We identify genomic features that influence co-methylation and how it varies between different tissues.
+
+In the final chapter, we develop a framework for simulating whole-genome bisulfite-sequencing data.
+Simulation software is valuable when developing new analysis methods since it can generate data on which to assess the performance of the method and benchmark it against competing methods.
+Our simulation model is informed by our analyses of the 40 whole-genome bisulfite-sequencing samples and our study of co-methylation.
 
@@
 
-The data sets used in this thesis were provided by the investigators listed in Chapter \ref{chap:datasets}.\\\\
+The data sets used in this thesis were provided by the investigators
+
+\textbf{TODO}: List investigators.
+
+For full details, please see Chapter \ref{chap:datasets}.\\\\
 

### 2015-05-03 Rmarkdown/datasets.Rmd
Fix formatting

@@
 The methylC-seq libraries from the \cite{Lister:2009hy} paper were the first published whole-genome bisulfite-sequencing libraries of mammalian DNA.
-A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (`IMR90`), with those from a pluripotent tissue, embryonic stem cells (`H1`).
+A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (_IMR90_), with those from a pluripotent tissue, embryonic stem cells (_H1_).
 Each tissue was run in duplicate.
@@
 The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line.
-The _ADS-adipose_ sample are "adipocytes derived from the `ADS` cells through adipogenic differentiation conditions".
+The _ADS-adipose_ sample are "adipocytes derived from the _ADS_ cells through adipogenic differentiation conditions".
 The _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.
@@
 The _FF_ sample, a foreskin fibroblast cell line, is the 'founder' of this this mini dataset.
-The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from `FF`.
+The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from _FF_.
 In fact, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are subclones derived from _FF-iPSC\_19_, whose methylome was not sequenced.

### 2015-05-03 Rmarkdown/methsim.Rmd
Fix formatting

@@
 
-Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming nature of this modelling decision.
+Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming consequences of this modelling decision.
 When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong.
@@
 \includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
-\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
+\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
 \end{figure}

### 2015-05-03 Rmarkdown/wgbs_statistical_framework.Rmd
Fix formatting

@@
 
-__TODO: Need a strong summary/conclusion. Emphasise the novel aspects of this chapter and the wide variety of datasets used.__
-
 In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data.
@@
 
-Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
-The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_).
-The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA.
-Nonetheless, some of the increased variability in the Seisenberger data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.
+Using these data, I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
+The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_).
+The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_).
+The genome-level results for the _Seisenberger_ samples are more difficult to interpret, however, since they are from pooled DNA.
+Nonetheless, some of the increased variability in the _Seisenberger_ data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.
 
-In contrast, the somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples.
+In contrast, the somatic samples, particularly those from tissue samples rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples.
 The may reflect that the DNA methylome is well-established in these samples and relatively static.

### 2015-05-03 Rmarkdown/concluding_remarks.Rmd
Add concluding remarks

@@
 
-__TODO: Review each chapter__
+This thesis has examined the statistical and computational challenges raised by high-throughput assays of DNA methylation.
+The first chapter introduces DNA methylation and then extensively reviews different high-throughput assays for measuring it, focusing on the current gold standard assay of whole-genome bisulfite-sequencing.
+
+The second chapter details the bioinformatics analysis of whole-genome bisulfite-sequencing data.
+Beginning with the `FASTQ` files created by the DNA sequencers, we review the quality control procedures, read mapping and post-processing essential to creating a `BAM` file creating high-quality mapped reads.
+We then describe methods for calling methylation events from bisulfite-sequencing data.
+The chapter concludes by introducing m-tuples and developing the `methtuple` software that can, uniquely, call methylation patterns at m-tuples from whole-genome bisulfite-sequencing data.
+
+The third chapter describes the 40 whole-genome bisulfite-sequencing samples that are analysed throughout the remainder of the thesis.
+
+Chapter \ref{chap:wgbs_statistical_framework} lays out a statistical framework for analysing bisulfite-sequencing data.
+This process illustrates that the commonly used summary of methylation levels, the $\beta$-value, aggregates over many potential sources of variation.
+In light of this, we discuss the interpretation of $\beta$-values and examine the distributions of $\beta$-values in our 40 samples.
+
+Chapter \ref{chap:wgbs_downstream_analyses} examines methods for analysing bisulfite-sequencing data to address key biological questions.
+Most downstream analyses are based on $\beta$-values, such as the identification of differential methylation, and we review these.
+There is, however, a growing interest in analyses based on methylation patterns at m-tuples since these can address novel biological questions.
+A limitation to the development of these methods has been a lack of software.
+To complement to the `methtuple` software, the chapter concludes with the development of the `MethylationTuples` R package, which is designed to facilitate analyses based on methylation patterns at m-tuples.
+
+The sixth and seventh chapters are a comprehensive review and analysis of _co-methylation_, the spatial dependence of DNA methylation.
+Chapter \ref{chap:co-methylation} re-examines previous analyses of co-methylation and finds that these have been limited by inadequate data and deficiencies in the statistical methods applied.
+We address some of these limitations with an extensive analysis of co-methylation in Chapter \ref{chap:co-methylation}.
+We propose a novel analysis of co-methylation using methylation patterns from individual DNA fragments, which we call _within-fragment co-methylation_.
+A simulation study demonstrates that, in general, we cannot estimate this within-fragment co-methylation for individual pairs of CpGs.
+By aggregating pairs of CpG, however, the simulation study shows that we can estimate a summary of this spatial dependence of DNA methylation along individual DNA fragments.
+We apply this method to 40 whole-genome bisulfite-sequencing samples to identify genomic features that influence co-methylation and study how it varies between different tissues.
+
+To conclude, Chapter \ref{chap:methsim} details our efforts to develop the `methsim` software to simulate whole-genome bisulfite-sequencing data.
+Unlike existing software, `methsim` seeks to model co-methylation so as to create more realistic simulated data.
+Such simulated data will be useful in developing and benchmarking methods for the analysis of DNA methylation data.
+While aspects of `methsim` are promising, the overly-strong correlations of $\beta$-values is an obvious inadequacy in the current model.
+We continue to work to improve this and to develop the software more generally.
+
+DNA methylation data have a complex dependence structure, as demonstrated by the analyses presented in this thesis.
+Any analysis must bear in mind the multiple sources of variation in the data and the level of aggregation required in order to extract meaningful results.
+
+The large size of the data generated by these experiments can make it challenging to develop analysis methods.
+This is particularly true when coupled with a lack of software for extracting, manipulating and summarising the relevant features of the data.
+Much of the effort in this thesis has gone into developing software to extract and manage this information (`methtuple` and `MethylationTuples`) so that new tools can be built on a higher foundation (`methsim`).
+It is my hope that, with continued development, these tools may facilitate other analyses to unravel the complexity of DNA methylation and its many roles in biology.

### 2015-05-03 latex/phd_thesis.tex
Fix formatting

@@
 \usepackage{etex}
+
+% To insert blank (unnumbered) page after title page http://tex.stackexchange.com/a/36881
+\usepackage{afterpage}
+\newcommand\blankpage{%
+    \null
+    \thispagestyle{empty}%
+    \addtocounter{page}{-1}%
+    \newpage}
 
@@
 
+\blankpage
+
 \chapter{Abstract}
@@
 The second half of the thesis focuses on \emph{co-methylation}, the spatial dependence of DNA methylation.
-We demonstrate that previous analyses of co-methylation have been limited by inadequate data and analysis methods.
+We demonstrate that previous analyses of co-methylation have been limited by inadequate data and deficiencies in the statistical methods applied.
 This motivates a study of co-methylation from 40 whole-genome bisulfite-sequencing samples.
@@
 
-The data sets used in this thesis were provided by the investigators
+The datasets used in this thesis were provided by the investigators listed below.
 
-\textbf{TODO}: List investigators.
+Aaron Statham and Sue Clark, Garvan Institute of Medical Research. The \emph{EPISCOPE} dataset.
 
-For full details, please see Chapter \ref{chap:datasets}.\\\\
+Several publicly available datasets were also used and are described in Chapter \ref{chap:datasets}.\\\\
 
@@
 
-Finally, and most of all, thank you to my parents, Gay and John, to my brothers, Jack, Luke and Daniel, and to my friends. Your support, love and encouragement kept me going.
+Finally, and most of all, thank you to my parents, Gay and John, to my brothers, Jack, Luke and Daniel, and to my friends. Your support, love, and encouragement kept me going.
 
@@
 
-% 2 - 4 pages
-\textbf{TODO}
-
 \include{appendix} % appendix.tex

### 2015-05-03 Rmarkdown/comethylation_review.Rmd
Edits to front matter and table of contents

@@
 \includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
-\caption[Fourier transformation of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample.
+\caption[Fourier analysis of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample.
 Fourier transformation computed and plotted using the \texttt{spectrum} function in R.
@@
 \centering
-\caption[Notions of `subject' and `population' in methylation linkage]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}.
+\caption[Notions of `subject' and `population' in methylation linkage analyses]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}.
 NB `person' could be any organism, e.g, `mouse', `dog', etc.}

### 2015-05-03 Rmarkdown/methsim.Rmd
Edits to front matter and table of contents

@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
-\caption[\emph{EPISCOPE}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{EPISCOPE} dataset.}
+\caption[\emph{EPISCOPE}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_beta_by_pm_boxplots}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
-\caption[\emph{Lister}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Lister} dataset.}
+\caption[\emph{Lister}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Lister} dataset.}
 \label{fig:Lister_beta_by_pm_boxplots}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
-\caption[\emph{Seisenberger}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
+\caption[\emph{Seisenberger}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_beta_by_pm_boxplots}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
-\caption[\emph{Ziller}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Ziller} dataset.}
+\caption[\emph{Ziller}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Ziller} dataset.}
 \label{fig:Ziller_beta_by_pm_boxplots}
@@
 \includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly.pdf}
-\caption[\texttt{methsim} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for \texttt{methsim} simulated data and the \emph{ADS} input sample.
+\caption[\texttt{methsim}: $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for \texttt{methsim} simulated data and the \emph{ADS} input sample.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
-\caption[\texttt{methsim} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island.
+\caption[\texttt{methsim}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.

### 2015-05-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits to front matter and table of contents

@@
 \includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
-\caption[\texttt{Bis-SNP} method schematic]{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
+\caption[Schematic of \texttt{Bis-SNP} method]{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand.
 For all three loci, the reads mapped to the forward strand contain a thymine.

### 2015-05-03 Rmarkdown/wgbs_statistical_framework.Rmd
Edits to front matter and table of contents

@@
 
-\subsection{Some complications for a single sample}\label{sec:one_sample_complications}
+\subsection{Some complications}\label{sec:one_sample_complications}
 
@@
 
-\subsection{Some complications with multiple samples}\label{sec{multi_sample_complications}}
+\subsection{Some complications}\label{sec{multi_sample_complications}}
 

### 2015-05-03 latex/phd_thesis.tex
Edits to front matter and table of contents

@@
 \title{\huge \bf The statistical analysis of high-throughput assays
-for studying DNA methylation.\\}
+for studying DNA methylation\\}
 
@@
 
-\date{The Walter and Eliza Hall Institute of Medical Research}
+\date{Bioinformatics Division\\
+The Walter and Eliza Hall Institute of Medical Research}
 
@@
 Using whole-genome bisulfite-sequencing, we can now measure DNA methylation at individual nucleotides across entire genomes.
-These experiments produce vast amounts of data that require new methods and tools to analyse.
+These experiments produce vast amounts of data that require new methods and software to analyse.
 
@@
 
-The second half of the thesis focuses on \emph{co-methylation}, the spatial dependence of DNA methylation.
-We demonstrate that previous analyses of co-methylation have been limited by inadequate data and deficiencies in the statistical methods applied.
+The second half of the thesis focuses on \emph{co-methylation}, the spatial dependence of DNA methylation along the genome.
+We demonstrate that previous analyses of co-methylation have been limited by inadequate data and deficiencies in the applied statistical methods.
 This motivates a study of co-methylation from 40 whole-genome bisulfite-sequencing samples.
@@
 Peter, thank you for supporting my decision to radically change tack very early on in my PhD from a topic in theoretical statistics to a heavily computational and interdisciplinary problem.
-And Terry, thank you for agreeing to take me on as a student after this change of heart.
+Terry, thank you for agreeing to take me on as a student after this change of heart.
 I have learnt much from working with you and greatly appreciate your generosity and guidance over the course of my PhD.

### 2015-05-03 Rmarkdown/biology_background.Rmd
Edits to Biology Background

@@
 \begin{chapabstract}
-This chapter summarises the basic biology necessary for understanding my thesis.
-I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
+This chapter summarises the basic biology necessary for understanding this thesis.
+It introduces DNA methylation, describes some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.
 \end{chapabstract}
@@
 In most organisms, the genome is encoded by deoxyribonucleic acid, _DNA_.
-In eukaryotes[^eukaryotes], which includes plants, fungi and animals, the DNA is wound around repeating units of eight _histone_ protein cores to form _nucleosomes_, which are the fundamental unit of _eukaryotic chromatin_.
+In eukaryotes[^eukaryotes], which includes plants, fungi, and animals, the DNA is wound around repeating units of eight _histone_ protein cores to form _nucleosomes_, which are the fundamental unit of _eukaryotic chromatin_.
 Chromatin compactly packages the DNA into _chromosomes_, so that the organism's complete nuclear DNA, its nuclear genome, might fit into the nucleus of its cells[^extra_nuclear_genomes].
@@
 In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_.
-One strand is called the 5' strand (pronounced "5 prime strand" and also known as the sense strand, Crick strand or top strand) and the other strand called the 3' strand (pronounced "3 prime strand" and also known as the antisense strand, Watson strand or bottom strand).
+One strand is called the 5' strand (pronounced "5 prime strand" and also known as the sense strand, Crick strand or top strand) and the other strand is called the 3' strand (pronounced "3 prime strand" and also known as the antisense strand, Watson strand or bottom strand).
 Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_).
@@
 This means that eukaryotic cells created during _mitosis_ contain the same DNA as the 'parent' cell.
-During DNA replication, the two strands are separated and each strand's _complementary_ DNA sequence is copied by an enzyme called _DNA polymerase_.
+During DNA replication, the two strands are separated and each strand's complementary DNA sequence is copied by an enzyme called _DNA polymerase_.
 It is because the two strands of DNA are complementary that ensures the daughter cell contains the same DNA sequence as the parent cell[^dna_replication].
@@
 
-Chromatin is frequently described as either 'closed' or 'open'.
+Chromatin is often described as either 'closed' or 'open'.
 Closed chromatin, _heterochromatin_, is more tightly packed than the open _euchromatin_.
-Heterochromatin is associated with transcriptionally repressed regions of the genome because the machinery required to translate DNA to RNA is unable to physically access the DNA.
+Heterochromatin is associated with transcriptionally repressed regions of the genome because the machinery required to translate DNA to RNA is less able to physically access the DNA.
 In contrast, euchromatin is associated with transcriptionally active regions of the genome since its more open nature allows easier access for the translational machinery.
@@
 However, genomes of any two randomly selected people are identical at approximately $99.9\%$ of sites.
-Furthermore, the vast majority, $98\%$ by some estimates \citep{Elgar:2008dm}, of the human genome is made up of non-coding DNA and upwards of $50\%$ is repetitive sequence \citep{Treangen:2012dn}.
+Furthermore, the vast majority of the human genome, $98\%$ by some estimates \citep{Elgar:2008dm}, is made up of non-coding DNA and upwards of $50\%$ is repetitive sequence \citep{Treangen:2012dn}.
 
@@
 If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell) and a cone cell (a photoreceptor in the retina) all have identical genomes.
-In fact, all cells in an organism, excluding the gametes, have an identical genome yet play very different biological roles.
+In fact, all cells in an organism, excluding the gametes, have an identical genome[^somatic_mutations] yet play very different biological roles.
 This is due to different genes being active in different cells.
 
-_Epigenetics_, which I describe in the next section, plays a role in determining both the between-individual phenotypic variation, as well as the within-individual phenotypic variation.
+[^somatic_mutations]: This ignores somatic mutations.
+
+_Epigenetics_, which I describe in the next section, plays a role in determining the between-individual phenotypic variation, as well as the within-individual phenotypic variation.
 DNA methylation, the focus of my thesis, is the prototypical and most well-studied epigenetic modification.
@@
 
-Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" \citep{Waddington:2012bu}.
+Conrad Waddington coined the phrase in 1942 as a portmanteau of the words 'epigenesis' and 'genetics' \citep{Waddington:2012bu}.
 Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \citep{Waddington:2012bu}.
@@
 
-However, DNA methylation is not found in all organisms, for example, it not detectable in yeast \citep{Capuano:2014cu}, and so is not essential for life.
+However, DNA methylation is not found in all organisms[^yeast], and so is not essential for life.
 Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation \citep{Capuano:2014cu}.
+
+[^yeast]: For example, DNA methylation not detectable in yeast \citep{Capuano:2014cu}
 
@@
 
-The typical site of DNA methylation is at the C5 carbon position of a cytosine base. This modified cytosine is called _5-Methylcytosine_ or _5mC_. Figure \ref{fig:5mC} shows the structure of 5mC.
+The typical site of DNA methylation is at the C5 carbon position of a cytosine base, hence _5-methylcytosine_. Figure \ref{fig:5mC} shows the structure of 5mC.
 
@@
 
-It is common to describe a cytosine as being 'methylated' or 'unmethylated', however, care must be taken when using these terms.
+Acytosine may be described being 'methylated' or 'unmethylated', however, care must be taken when using these terms.
 At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
@@
 \caption[ECDF of distances between CpGs in hg19]{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19).
-The observed distances are contrasted with those under the `expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
+The observed distances are contrasted with those under the `expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of CpGs on each chromosome.}
 \label{fig:IPD_ECDF_hg19}
@@
 CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}.
-The classical definition of a CpG island, given by \citet{GardinerGarden:1987we}, and used by the popular UCSC genome browser ([http://genome.ucsc.edu/cgi-bin/hgGateway](http://genome.ucsc.edu/cgi-bin/hgGateway)), is a region of the genome where:
+The classical definition of a CpG island, given by \citet{GardinerGarden:1987we}, and used by the popular UCSC genome browser ([http://genome.ucsc.edu/cgi-bin/hgGateway](http://genome.ucsc.edu/cgi-bin/hgGateway)), is a region of the genome where the following conditions are satisfied:
 
@@
 This definition was refined by \citet{Takai:2002jb} to exclude _Alu_-repetitive elements, which are otherwise misclassified as _bona fide_ CpG islands.
-More recently, \citet{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.
-In my thesis, I use the definition given by \citet{Wu:2010do}.
+More recently, \citet{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region; this definition is used in the remainder of my thesis.
 
@@
 In contrast, in an embryonic stem cell line, they found that $24.5\%$ of cytosines that displayed statistically significant evidence of methylation occured in a non-CpG context.
-However, it should also be noted that these non-CpG loci that were methylated had on average a much lower level of methylation than their CpG counterparts.
+However, it should also be noted that these non-CpG loci that were methylated had, on average, a much lower level of methylation than their CpG counterparts.
 
@@
 For example, \citet{Lister:2008bh} found that in the widely-studied _Arabidopsis thaliana_ that $45\%$ of cytosines that displayed statistically significant evidence of methylation are at CHG or CHH loci.
-However, they also found that the level of methylation at non-CpG loci is typically lower than that observed at CpG dinucleotides.
+They also found that the level of methylation at non-CpG loci, however, is typically lower than that observed at CpG dinucleotides.
 
@@
 Methylation is not the only chemical modification of cytosines, although it is by far the most common.
-Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \citep{Plongthongkum:2014jg}.
+Listed from most frequent to least frequent, these are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \citep{Plongthongkum:2014jg}.
 The biological significance of these marks is still being determined, in part because the assays for studying these are still in development and because their relatively scarcity in the genome means that experiments to detect these modifications are more difficult and expensive.
@@
 A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,][]{Moore:2013in}.
-In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.
+In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA, and erasers remove the methyl group from the DNA.
 
@@
 Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division.
-This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions.
+This leads to a gradual, stochastic, and genome-wide loss of DNA methylation after multiple cell divisions.
 This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of local tissue-specific differences in DNA methylation \citep{Irizarry:2009ky} nor the two stages of rapid global demethylation that occur during development \citep{Wu:2014gw}.
@@
 
-One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
+One question raised by the third point is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
 This is an active area of research.
@@
 
-The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-cost throughput.
-Experiments that use an assay with high resolution and high per-sample throughput generally have fewer samples due to the associated higher costs than experiments using a lower resolution assay or an assay with lower per-sample throughput.
+The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput, and per-cost throughput.
+Experiments that use an assay with high resolution and high per-sample throughput generally have fewer samples (due to the associated higher costs) than experiments using a lower resolution assay or an assay with lower per-sample throughput.
 
-In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays.
-In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at single-base resolution.
+In this section I describe each of these pre-treatments but focus on the bisulfite-conversion assays.
+In particular, I describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at single-base resolution.
 
@@
 _Restriction endonucleases_ are an important technique in molecular biology.
-These enzymes can preferentially 'cut', 'cleave' or 'digest' DNA at or near to particular sequence motifs.
+These enzymes can preferentially 'cut', 'cleave', or 'digest' DNA at or near to particular sequence motifs.
 The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_.
@@
 Furthermore, the bioinformatic analysis of data from these assays is complicated by the varying CpG density along the genome, which leads to different enrichment affinities for different regions of the genome.
-However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation \citep{Laird:2010iv}.
+However, these assays can provide a relatively cheap and efficient genome-wide assessment of DNA methylation \citep{Laird:2010iv}.
 
@@
 
-Initial experiments based on the sodium bisulfite pre-treatment of DNA used Sanger sequencing of cloned PCR products, a very laborious task, which restricted experiments to studying a limited number of short segments of DNA.
+Initial experiments based on the sodium bisulfite pre-treatment of DNA used Sanger sequencing of cloned PCR products, a very laborious task that restricted experiments to studying a limited number of short segments of DNA.
 Although subsequent enhancements in the automation of Sanger sequencing improved the throughput of these assays, it was never going to be able to deliver a cost-effective, genome-scale assay of DNA methylation.
@@
 
-The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k_) provides a modern implementation of this type of assay for studying human DNA.
+The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k array_) provides a modern implementation of this type of assay for studying human DNA.
 This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \citep{Stirzaker:2014cu}.
@@
 \citet{Cokus:2008fc} termed their approach _BS-seq_ while \citet{Lister:2008bh} called their method _methylC-seq_.
-From a bioinformatics perspective, the main difference is that the _BS-seq_ protocol produces sequencing reads from four bisulfite-converted DNA strands --- the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) --- which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results.
-In contrast, the _methylC-seq_ protocol only produces sequencing reads from two bisulfite-converted strands --- _OT_ and _OB_ --- and so only requires mapping to two different _in silico_ converted reference genomes, followed by a merge.
+From a bioinformatics perspective, the main difference is that the BS-seq protocol produces sequencing reads from four bisulfite-converted DNA strands --- the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) --- which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results.
+In contrast, the methylC-seq protocol only produces sequencing reads from two bisulfite-converted strands --- _OT_ and _OB_ --- and so only requires mapping to two different _in silico_ converted reference genomes, followed by a merge.
 The methylC-seq protocol is now the standard whole-genome bisulfite-sequencing protocol, due in part to the simpler bioinformatics analysis.
@@
 [^microarrays]: Microarray hybridisation assays can provide single-base resolution but not single-molecule resolution.
-The signal from a microarray-based experiment a sample-wide average since, for each locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.
+The signal from a microarray-based experiment is a sample-wide average since, for each locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.
 
-Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1$ to $5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells.
+Almost all bisulfite-based assays require a considerable amount of DNA that is extracted from a population of cells (e.g., $1$ to $5$ $\mu$g for whole-genome bisulfite-sequencing).
 However, the minimal amount of DNA is being reduced with each technological advance.
-Recently, both \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although they are not yet a commercially available sequencing assay.
+Recently, \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although these are not yet a commercially available sequencing assay.
 Furthermore, these are not yet proper genome-wide assays.
 The technique of \citet{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome.
-And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to $48.4\%$ of cpG sites" \citep{Smallwood:2014kn}.
+And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to $48.4\%$ of CpG sites" \citep{Smallwood:2014kn}.
 
@@
 The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome.
-Because the read is compared against the reference genome it may incorrectly inferred to be a thymine.}
+Because the read is compared against the reference genome, it may incorrectly inferred to be a thymine.}
 \label{fig:bisulfite_CT_SNP.pdf}

### 2015-05-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits to WGBS bioinformatics analysis

@@
 \begin{chapabstract}
-I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used \emph{methylC-seq} protocol.
+I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used methylC-seq protocol.
 All data used in my thesis were generated using this protocol.
@@
 
-The chapter concludes by introducing the `methtuple` software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci.
+The chapter concludes by introducing the \texttt{methtuple} software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci.
 \texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
@@
 \centering
+\caption[Popular bisulfite-sequencing read mappers and their underlying mapping software]{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
+\label{tab:bs-seq_mappers}
 \begin{tabulary}{\textwidth}{LLL}
@@
 \end{tabulary}
-\caption[Popular bisulfite-sequencing read mappers and their underlying mapping software]{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
-\label{tab:bs-seq_mappers}
 \end{table}
@@
 
-There is generally no way to tell if a read is truly a PCR duplicated based on the sequencing data alone.
-
 There is generally no way to tell based on sequencing data if a read is truly a PCR duplicate.
-However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as 'PCR duplicates'[^pcr_dup]).
+However, it is relatively easy to identify suspected PCR duplicates (which are almost always referred to as 'PCR duplicates'[^pcr_dup]).
 Software to identify PCR duplicates includes the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}.
@@
 Roughly speaking, these tools will flag reads with identical start and end co-ordinates as being suspected PCR duplicates.
-This will inevitably lead to some false positives because reads may have identical co-ordinates yet not be PCR duplicates.
+This will inevitably lead to some false positives because reads may have identical co-ordinates yet not be true PCR duplicates.
 The false positive rate is a particular problem when a subset of the genome is sequenced at high coverage, such as in RRBS.
@@
 
-In practice, this means that for whole-genome sequencing data we can be fairly confident that suspected PCR duplicates are _true_ PCR duplicates.
+In practice, for whole-genome sequencing data we can be fairly confident that suspected PCR duplicates are _true_ PCR duplicates.
 However, for targetted sequencing, such as RRBS or amplicon sequencing, we are much less confident and may remove some of our signal if we remove possible PCR duplicates.
@@
 The methylation calling should include filters to remove the detected M-bias (along with any other additional filters).
-This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling.
+This strategy requires two passes over the `SAM/BAM` file, one to compute the M-bias and one to do the methylation calling.
 This is the approach used by `bismark_methylation_extractor` \citep{Krueger:2011eb} and `Bis-SNP` \citep{Liu:2012ge}.
@@
 
-I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{sec:methtuple}).
+I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple` (described in Section \ref{sec:methtuple}).
 
@@
 For CpGs in \emph{read-2}, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read-positions for all methylation types.
-The spike at read-position 101 is also evidence, albeit to a lesser extent, in \emph{read-1}.
+The spike at read-position 101 is also evident, albeit to a lesser extent, in \emph{read-1}.
 This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
@@
 In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e.
-read-positions whose methylation levels are 'too far away' from the vast majority of read-positions.
+read-positions whose methylation levels are 'too far away' from the majority of read-positions.
 It is rather a subjective decision to make.
@@
 Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias.
-My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants to exclude[^methtuple_mbias].
+The `methtuple` software avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants to exclude (discussed in Section \ref{sec:methtuple}).
 
-In the datasets I have analysed I've had to ignore up to 30 read-positions per read due to M-bias.
+In the datasets I have analysed, I have had to ignore up to 30 read-positions per read due to M-bias.
 Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely.
@@
 When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters].
-Strictly speaking each sequenced bases is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all bases within 'surviving' reads.
+Strictly speaking, each sequenced bases is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all bases within 'surviving' reads.
 In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.
@@
 This data is then used to form a set of sample-specific methylation loci.
-However, this approach is also very expensive due to the extra sequencing requirements.
+This approach, however, is also very expensive due to the extra sequencing requirements.
 A cheaper alternative is to genotype the sample on a genome-wide SNP microarray.
@@
 
-The third approach, and arguably the bare minimum, is to call methylation loci using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population.
+The third approach, and arguably the bare minimum, is to call methylation events using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population.
 For example, we might exclude all cytosines in the reference genome that are also SNPs in dbSNP \citep{Sherry:2001gh}.
@@
 G) on the forward (resp.
-reverse) strand of the reference genome.
+reverse) strand of the reference genome[^ad_hoc_bissnp]
+
+[^ad_hoc_bissnp]: This is like an _ad hoc_ and limited version of `Bis-SNP`.
 
@@
 \includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
-\caption[Different methylation patterns can lead to identical $\beta$-values]{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
+\caption[Different methylation patterns can lead to identical $\beta$-values]{Two regions, each with four methylation loci that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.
+Each line is a read, a white circle is an unmethylated CpG and a black circle is a methylated CpG.}
 \label{fig:incomplete_picture_1-tuples}
@@
 
-Generally, when referring to m-tuples I implicitly mean those with $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
+When referring to m-tuples I implicitly mean those with $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple.
 The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combinations` flag is set[^nil0].
@@
 \includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
-\caption[\texttt{methtuple} example region]{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}).
+\caption[\texttt{methtuple} example region]{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1}, \texttt{2}, \texttt{3}, \texttt{4} and  \texttt{5}).
 The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
@@
 
-With current sequencing technology we are limited to extracting m-tuples that span no more than $200-250$ bp.
+With current sequencing technology we are limited to extracting m-tuples that span no more than $200$ to $250$ bp.
 This obviously affects the size of m-tuples that we can study.
@@
 This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}.
-Such a 'spike' can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.
+Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.
 
@@
 
-I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples.
+I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for $40$ whole-genome bisulfite-sequencing samples.
 Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets.
@@
 The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
-Therefore, memory usage is relatively constant across values of `-m` for a given chromosome.
+Memory usage is therefore relatively constant across values of `-m` for a given chromosome.
 The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`.
-This means that all 2-tuples with $NIL \geq 0$ were computed and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.
+This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.
 
@@
 \caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes.
-The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.
-Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
+The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.}
 \label{fig:methtuple_runtime}
@@
 These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
-The suffix `ac' means that the option \texttt{--all-combinations} was set.
-Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
+The suffix `ac' means that the option \texttt{--all-combinations} was set.}
 \label{fig:methtuple_memory}
@@
 
-__TODO: necessary?__
+This chapter has detailed the first steps in a bioinformatics analysis of whole-genome bisulfite-sequencing data.
+There are many decisions to be made at each step and these will ultimately affect the quality of the data available for downstream analysis.
+
+This chapter also introduced `methtuple`, the first of several pieces of software that were developed as part of my thesis.
+`methtuple` is software for calling methylation patterns at m-tuples from whole-genome bisulfite-sequencing data.
+It will be essential for our later analysis of within-fragment co-methylation (Chapter \ref{chap:co-methylation}) and has wider utility in facilitating other downstream analyses based on methylation patterns at m-tuples (Chapter \ref{chap:wgbs_downstream_analyses}).

### 2015-05-03 Rmarkdown/datasets.Rmd
Edits to front matter and table of contents
Edits to datasets chapter

@@
 
-\section{Chapter overview}
+\section{Overview of data processing}
 
@@
 These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e.
-$|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (read 1 and read 2 analysed separately where applicable).
+$|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (_read-1_ and _read-2_ analysed separately where applicable).
 3.
@@
 4.
-Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see \ref{sec:MethylationTuples}).
-Specifically, a `MethPat` object was created for each of the EPISCOPE, Lister, Seisenberger and Ziller datasets for 1-tuples, 2-tuples and 2ac-tuples (2-tuples with the  `--all-combinations` flag in `methtuple` set), 3-tuples and 4-tuples.
+Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see Section \ref{sec:MethylationTuples}).
+Specifically, a `MethPat` object was created for each of the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets for 1-tuples, 2-tuples and 2ac-tuples (2-tuples with the  `--all-combinations` flag in `methtuple` set), 3-tuples and 4-tuples.
 I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.
 
-The raw data for the _Lister_, _Seisenberger and _EPISCOPE_ datasets are all publicly available.
+The raw data for the _Lister_, _Seisenberger_ and _EPISCOPE_ datasets are all publicly available.
 The _EPISCOPE_ data are not yet published and I do not have permission to make these publicly available.
@@
 
-\section{Lister dataset}\label{sec:Lister_dataset}
+\section{\emph{Lister} dataset}\label{sec:Lister_dataset}
 
-The _Lister dataset_ refers to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
-The Lister data were the largest publicly available human whole-genome bisulfite-sequencing datasets until quite recently.
-All samples are of human DNA.
+The _Lister_ dataset refers to whole-genome bisulfite-sequencing libraries used in \cite{Lister:2009hy} and \cite{Lister:2011kg}.
+The _Lister_ data were the largest publicly available human whole-genome bisulfite-sequencing datasets until quite recently.
 
@@
 Each tissue was run in duplicate.
-While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing.
+While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described as technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing.
 In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.
@@
 These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells.
-There are no replicates for any of the _Lister 2011_ samples.
+There are no replicates for any of the \cite{Lister:2011kg} samples.
 
@@
 The reported read lengths are prior to any trimming of the reads.
-Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
+Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
 \label{tab:Lister_dataset}
@@
 
-\section{EPISCOPE dataset}\label{sec:EPISCOPE_dataset}
+\section{\emph{EPISCOPE} dataset}\label{sec:EPISCOPE_dataset}
 
-The EPISCOPE data were kindly provided to me by Professor Susan Clark (Garvan Institute of Medical Research, Sydney) and Dr Peter Molloy (CSIRO Animal, Food and Health Sciences).
+The _EPISCOPE_ data were kindly provided to me by Professor Susan Clark (Garvan Institute of Medical Research, Sydney) and Dr Peter Molloy (CSIRO Animal, Food and Health Sciences).
 This dataset is not yet published.
@@
 
-\section{Seisenberger dataset}\label{sec:Seisenberger_dataset}
+\section{\emph{Seisenberger} dataset}\label{sec:Seisenberger_dataset}
 
-The Seisenberger data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \citep{Seisenberger:2012ko}.
+The _Seisenberger_ data are from a study of the dynamics of DNA methylation reprogramming in mouse primordial germ cells \citep{Seisenberger:2012ko}.
 These were a convenience sample provided to me by a colleague, Felix Krueger (Babraham Institute).
@@
 
-\section{Ziller dataset}\label{sec:Ziller_dataset}
+\section{\emph{Ziller} dataset}\label{sec:Ziller_dataset}
 
-The Ziller are a subset of the data used in \cite{Ziller:2013fs}.
+The _Ziller_ are a subset of the data used in \cite{Ziller:2013fs}.
 Specifically, I use a convenience sample of 8 whole-genome bisulfite-sequencing libraries.
 These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research, Sydney).
-I thank Felix who sent me the `BAM` files containing processed and aligned reads.
+I thank Aaron who sent me the `BAM` files containing processed and aligned reads.
 

### 2015-05-03 latex/phd_thesis.tex
Edits to datasets chapter

@@
 
-Aaron Statham and Sue Clark, Garvan Institute of Medical Research. The \emph{EPISCOPE} dataset.
+Aaron Statham and Sue Clark, Garvan Institute of Medical Research; Peter Molloy, CSIRO. The \emph{EPISCOPE} dataset.
 
@@
 My PhD would not be possible without the people who provided me with data to analyse.
-I am grateful to Sue Clark, Aaron Statham, Emma Whitelaw and Harry Oey for the opportunity to collaborate with and learn from you.
+I am grateful to Sue Clark, Peter Molloy, Aaron Statham, Emma Whitelaw and Harry Oey for the opportunity to collaborate with and learn from you.
 Thanks also to Ryan Lister, Kasper Hansen and Felix Krueger for making their data publicly available and answering all my nitpicky questions.

### 2015-05-03 Rmarkdown/wgbs_statistical_framework.Rmd
Edits to WGBS statistical framework

@@
 ---
-title: "A statistical framework for whole-genome bisulfite-sequencing data"
+title: "A statistical framework for analysing whole-genome bisulfite-sequencing data"
 output:
@@
 
-\chapter{A statistical framework for whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}
+\chapter{A statistical framework for analysing whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}
 
@@
 
-The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state.
-As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell may have a slightly different methylation profile.
+The methylation state of a locus can vary within a sample due to cell-to-cell variability of DNA methylation.
+A sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell may have a slightly different methylation profile.
 Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states.
@@
 
-From these we can compute the proportion of fragments that methylated at the $i^{th}$ locus:
+From these we can compute the proportion of fragments that are methylated at the $i^{th}$ locus:
 
@@
 Again, I emphasise that $H_{(i, i + 1, i + \text{m} - 1), j}, Z_{h, (i, i + 1, i + \text{m} - 1), j}, B_{(i, i + 1, i + \text{m} - 1), j}$ and the set of methylation patterns are unobservable.
-However, by sequencing the pools of DNA fragments we aim to estimate these variables.
+We aim to estimate these variables through sequencing the pool of DNA fragments.
 
@@
 
-In a whole-genome bisulfite-sequencing experiment we do not sequence every DNA fragment in the pool.
+A whole-genome bisulfite-sequencing experiment does not sequence every DNA fragment in the pool.
 Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments.
@@
 1.
-There may be gaps due to the insert size being longer than the sum of the read lengths, e.g., $(i, i + 1, i + 3, i + 4))$.
+There may be gaps due to the insert size being longer than the sum of the read lengths, e.g., $(i, i + 1, i + 3, i + 4)$.
 In effect, we have missing data for any intervening methylation loci, the $(i + 2)^{th}$ loci in this example.
@@
 
-Again, these definitions can be extended from 1-tuples to m-tuples.
+These definitions can also be extended from 1-tuples to m-tuples.
 The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$ and has sequencing depth $d_{(i, i + 1, \ldots, i + \text{m} - 1)} = |\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}|$.
@@
 
-Note that we do not know from which DNA fragment, $h$, each read came from, only that all methylation loci in the read came from the same DNA fragment.
+Note that we do not know from which DNA fragment ($h$) each read came from, only that all methylation loci in the read came from the same DNA fragment.
 
@@
 
-The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}}$.
+The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}} = (B_{1,j}, \ldots, B_{N_{loci}, j})$.
 In this section I review various methods for estimating these key parameters.
@@
 
-Recently, more sophisticated methods have been proposed to estimate or model the average methylation level.
+More sophisticated methods have recently been proposed to estimate or model the average methylation level.
 These methods, which are still based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.
@@
 This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data.
-For example, the `edgeR` software \citep{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.
+For example, the `edgeR` software \citep{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-Poisson mixture distribution, to account for both the biological and sampling variability.
 
@@
 Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample.
-This phenomenon, called _co-methylation_ is discussed and analysed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).
+This phenomenon, called _co-methylation_ is discussed and analysed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
 
@@
 
-`bsseq` assumes for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$.
+`bsseq` assumes that for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$.
 In contrast, `BiSeq` first creates clusters of CpGs and only assumes that the underlying methylation level is smooth at positions within each cluster.
@@
 Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`.
-However, these comparison studies use RRBS data, which `bsseq` is not designed for`, and so will favour methods designed for RRBS data[^rrbs_sim].
+However, these comparison studies use RRBS data, which `bsseq` is not designed for`, and their simulation so will favour methods designed for RRBS data[^rrbs_sim].
 
 [^rrbs_sim]: Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} altered the default `bsseq` parameters to try to make them comparable to `BiSeq`.
-\citet{Hebestreit:2013ko} changed the default minimum window size to 80bp but still required at least 20 CpGs per window.
-\citet{Lacey:2013iy} kept the default minimum window size of 2000 bp but reduced the minimum number of CpGs per window to 50 from the default of 70.
+\citet{Hebestreit:2013ko} changed the default minimum window size to $80$ bp but still required at least $20$ CpGs per window.
+\citet{Lacey:2013iy} kept the default minimum window size of $2,000$ bp but reduced the minimum number of CpGs per window to $50$ from the default of $70$.
 Nevertheless, the fact remains that `bsseq` is designed for analysing whole-genome bisulfite-sequencing data and not RRBS, which puts it at a disadvantage in these comparisons.
@@
 
-Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Bin(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}}}$ \citep{Hansen:2012gr}.
+Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Bin(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{d_{i, j}}}$ \citep{Hansen:2012gr}.
 The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample.
@@
 The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice.
-Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.
+Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \citep{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.
 
@@
 These plots show that CpG islands have a more strictly bimodal distribution than do the non-islands.
-While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample, except for the _E16.5\_male\_1_ sample.
+While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample (except for the _E16.5\_male\_1_ sample).
 The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated.
@@
 
+\FloatBarrier
+
 \subsubsection{Strand-specific $\beta$-values}
@@
 
-Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5 -- 0.7$.
+Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5$ to $0.7$.
 The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$).
-This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.
+This suggests that caution may be warranted in combining CpG methylation levels across strands for embryonic stem cell samples.
 
@@
 
+\FloatBarrier
+
 \subsection{Interpretation of $\beta$-values}\label{sec:beta-value_interpretation}
@@
 Also recall that most assays measure DNA methylation from a pool of cells, not a single haploid copy of the genome.
-The methylation state at the $i^{th}$ locus may vary across the $\mathcal{H}_{i}$ DNA fragments.
+The methylation state at the $i^{th}$ locus may vary across the $H_{i}$ DNA fragments.
 Therefore, I do not think it makes sense to describe a locus, $i$, as being a 'methylcytosine'.
@@
 Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, nor is any code made available, I believe the method is as follows[^mathematical_description].
-For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = d_{i}} Pr(X = k)$, where $X = Binom(d_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'.
+For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = d_{i}} Pr(X = k)$, where $X = Binom(d_{i}, \epsilon)$ and $\epsilon$ is the estimated 'error'.
 Any site with an FDR-adjusted P-value below a threshold was declared a "methylcytosine"[^fdr]
@@
 As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformaticians and genomics researchers makes it a favourable choice.
+As with $\beta$-values, the $\mathcal{M}$-values derived from bisulfite-sequencing data cannot generally be directly analysed due to the variable sequencing coverage across loci.
 
@@
 
-As with $\beta$-values, $\mathcal{M}$-values derived from bisulfite-sequencing data generally cannot be directly analysed due to the variable sequencing coverage across loci.
-However, the variable sequencing coverage can be accounted for using the afore-mentioned beta-binomial models and related regression models, which are discussed in the next section.
-
 \section{Summary}
 
-In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data.
-I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured.
+This chapter has defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data.
+It has addressed some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured.
 By using a common statistical framework we can better understand how different statistical methods relate to one another.
-Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties.
-I then examined the empirical distributions of these variables across a diverse set of 40 whole-genome bisulfite-sequencing samples.
+Using this framework, we described common estimators of DNA methylation levels and some of their statistical properties.
+We then examined the empirical distributions of these variables across a diverse set of 40 whole-genome bisulfite-sequencing samples.
 
-Using these data, I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.
+From these analyses we have seen that the CpG islands drive the strong bimodal distribution of $\beta$-values that is observed in almost all samples.
+We have also observed that most intermediate methylation occurs outside of the CpG islands.
 The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_).

### 2015-05-03 Rmarkdown/wgbs_downstream_analyses.Rmd
Edit to WGBS downstream analysis

@@
 A barrier to analyses based on m-tuples has been a lack of software.
-To overcome this barrier, I develop \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
+To help eliminate this barrier, I develop \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples.
 \texttt{MethylationTuples} complements \texttt{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples.
@@
 
-All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$
-
 \subsubsection{Experimental design}\label{sec:experimental_design}
@@
 
-Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or pooled the replicates prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}).
+Initial experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or these were pooled prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}).
 Simply pooling replicates and analysing as if they were a single sample ignores all within-group variability and should not be used.
@@
 Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}.
-Again, this emphasises the importance of good experimental design such as randomisation, replication and the use of positive and negative controls.
+This again emphasises the importance of good experimental design, with randomisation, replication and the use of positive and negative controls.
 
@@
 
-While there have been reports of DMCs resulting in a phenotypic difference \citep[e.g.,]{Furst:2012gh}, DMCs are mostly tested as a prelude to the identification of differentially methylated regions (DMRs).
+While there have been reports of DMCs resulting in a phenotypic difference \citep[e.g.,][]{Furst:2012gh}, DMCs are mostly tested as a prelude to the identification of differentially methylated regions (DMRs).
 Moreover, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome bisulfite-sequencing experiments.
@@
 To begin, `BiSeq` first defines _CpG clusters_ by identifying CpGs that are within a user-specified genomic distance of one another and that have sufficient sequencing coverage across the set of samples[^biseq_alternative].
-While there is clearly a 'data-driven' component to these cluster definitions, I reserve the use of 'data-driven regions' for those which are based on the methylation levels of loci rather than their genomic co-ordinates.
+While there is clearly a 'data-driven' component to these cluster definitions, I reserve the use of 'data-driven regions' for those that are based on the methylation levels of loci rather than their genomic co-ordinates.
 
@@
 
-Methods for identifying _data-driven_ DMRs are fairly _ad hoc_.
+Methods for identifying _data-driven_ DMRs are statistically _ad hoc_.
 The most common approach is to scan the genome for clusters of DMCs and declare these to be DMRs.
 The initial scan for DMCs will typically use a relaxed statistical significance threshold (i.e.
-below genome-wide significance).
+not necessarily genome-wide significant).
 Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.
@@
 \citet{Jaffe:2011hy} first developed formal statistical tests for differential variability of methylation for both the one-group and two-group experiments.
-These methods were developed for us with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}).
+These methods were developed for use with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}).
 In a one-group experiment, a variably methylated region is one that has increased variability compared to 'similar' regions elsewhere in the genome.
@@
 It might, for example, be due to an error in the assay.
-It is not difficult to envisage an example where the two groups in fact have very similar variability when the outlier is excluded.
+It is not difficult to envisage an example where the two groups in fact have very similar variability once the outlier is excluded.
 
@@
 
-All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\bm{\beta}$.
-
 \subsection{Epigenome segmentation}\label{sec:epigenome_segmentation}
 
-Methylation data in the form of the vector of $\beta$-values, $\bm{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}.
+Methylation data in the form of the vectors of $\mathbf{m}$ and $\mathbf{u}$  may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}.
 One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs).
@@
 
-In this section I review the different types of questions that can be addressed by expanding our analysis to use methylation patterns at m-tuples (m $> 2$) rather than just 1-tuples.
+This section reviews the different types of questions that can be addressed by expanding our analysis to use methylation patterns at m-tuples (m $> 2$) rather than just 1-tuples.
 This extra information is only available from sequencing-based assays.
-I also describe existing software that implement some of these methods.
+We also review existing software that implement some of these methods.
 
@@
 
-Broadly speaking, on the one hand, if we only observe a single unique methylation pattern, then the m-tuple has the minimum methylation entropy of zero.
+On the one hand, if we only observe a single unique methylation pattern, then the m-tuple has the minimum methylation entropy of zero.
 On the other hand, if we observe all possible $2^{m}$ methylation patterns at equal frequency, then the m-tuple has the maximum methylation entropy (typically normalised to one).
@@
 
-While straightforward, this approach is also obviously limited to a subset of the methylation loci in the genome.
+While straightforward, this approach is also obviously limited to the small number of methylation loci that are nearby to a heterozygous genetic variant.
 \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require heterozygous genetic variants nearby to the methylation locus of interest.
@@
 
-Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom.
-Even then, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.
+Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom and, even there, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.
 In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}.
@@
 
-[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism of the locus under a model that assumes sampling with replacement or, equivalently, an infinite population size.
-While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.
+[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the rate of epipolymorphism of the locus under a model that assumes sampling with replacement or, equivalently, an infinite population size.
+While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results provided that the sequencing depth is high.
 
@@
 `methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies and does not attempt to account for potential sources of bias.
-In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.
+In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [i.e.
+epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.
 
@@
 The input is a file containing the number of times each methylation pattern was observed at that m-tuple.
-Unfortunately, `MPFE` does not provide a way to create this table.
+Unfortunately, `MPFE` does not provide a way to create this file.
 
@@
 I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations.
-Two give two examples, `DMEAS` \citep{He:2013cj} is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` due to compilation errors.
+To give two examples, `DMEAS` \citep{He:2013cj} is only available as a Windows binary or as a Perl script that itself is only available as a PDF file, and I have been unable to install `methclone` due to compilation errors.
 
@@
 
-[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to [https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples).
+[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to [bioconductor.org/packages/GenomicTupless](bioconductor.org/packages/GenomicTuples).
 
-Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}.
+Bioconductor already has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}.
 However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges].
@@
 Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing methylation patterns at 3-tuples for $n$ samples.
-The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples.
+The similarities to the output format of `methtuple` are clear (see Figure \ref{fig:3-tuples}), with the added advantage that a single `MethPat` object can contain data from multiple samples.
 
@@
 
-The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the `[` method.
+The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the "`[`" method.
 It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods.
@@
 Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the _EPISCOPE_ whole-genome bisulfite-sequencing data with various sized m-tuples.
-Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).
+We see that as the size of the m-tuples increases, the data become sparser: most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).
 However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.
@@
 The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples.
-The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
+The final column is a measure of how sparse the data are: a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
 \label{tab:methpat_sparsity}

### 2015-05-03 Rmarkdown/comethylation_review.Rmd
Edits to co-methylation review

@@
 This chapter critically reviews different definitions of co-methylation and puts them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}.
-I highlight various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
-I also review how methods to detect differential methylation have sought to account for, and exploit, co-methylation.
+It highlights various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
+We also review how methods to detect differential methylation have sought to account for, and exploit, co-methylation.
 \end{chapabstract}
@@
 
-To date, most analyses of co-methylation have been based on analysing correlatiosn of aggregate methylation levels, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
+To date, most analyses of co-methylation have been based on analysing correlations of aggregate methylation levels, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}.
 These analyses have an implicit assumption of stationarity, that is, that the correlation between two loci depends only on the distance between them and not the actual position of the loci in the genome.
@@
 \citet{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples.
-Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} is often cited as evidence of co-methylation.
-For example, \citet{Hebestreit:2013ko} cite \citet{Eckhardt:2006gh} as evidence that "methylation levels are strongly spatially correlated" and \citet{Hansen:2011gu} cite \citet{Eckhardt:2006gh} as evidence that"proximal CpGs [have] similar methylation levels".
+Of particular relevance to this chapter, \citet{Eckhardt:2006gh} is often cited as evidence of co-methylation.
+For example, \citet{Hebestreit:2013ko} cite \citet{Eckhardt:2006gh} as evidence that "methylation levels are strongly spatially correlated" and \citet{Hansen:2011gu} cite \citet{Eckhardt:2006gh} as evidence that "proximal CpGs [have] similar methylation levels".
 
@@
 At distances $>1,000$ bp, we did not detect any correlation between CpG methylation and spatial distance.
-Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Eckhardt:2006gh}), copyright (2006)}
+Adapted by permission from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Eckhardt:2006gh}), copyright (2006).}
 \label{fig:Eckhardt}
@@
 \citet{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing.
-They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis thaliana_.
+They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype _Arabidopsis thaliana_, a small flowering plant whose methylome is widely studied.
 They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis].
@@
 
-The autocorrelation plots showed two periodicities, one of $\approx 167$ bp and the other of $10$ bp (CHH only).
+The autocorrelation plots showed two periodicities, one of approximately $167$ bp and the other of $10$ bp (CHH only).
 Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.
@@
 hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8$ to $10$ bp apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_).
-The $10$ bp CHH-only period in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})
+The $10$ bp CHH-only period in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in Section \ref{sec:previous_within-fragment_analyses})
 
@@
 
-\citet{Lister:2009hy} report an $8$ to $10$ bp period in the occurrence "methylcytosines".
+\citet{Lister:2009hy} report an $8$ to $10$ bp period in the occurrence of "methylcytosines".
 Recall that \citeauthor{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a binomial test (see Section \ref{sec:beta-value_interpretation} for details and a criticism of this concept).
@@
 
-[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Supplementary Figure 9, but I am not certain.
+[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Figure \ref{fig:Lister_2009_Sup_Fig_9}, but I am not certain.
 
@@
 
-In addition to the genome-wide analyses, the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of the _Arabidopsis_ sample and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ bp periodicity.
+In addition to the genome-wide analyses, the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of the _Arabidopsis_ sample, and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ bp periodicity.
 In fact, these results revealed that the $10$ bp period is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA.
-This led \citeauthor{Chodavarapu:2010iqa} to discard their earlier hypothesis[^same_authors] as to the source of the $10$ bp period \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__ [emphasis added]" and not just for the DRM2 methyltransferase.
+This led \citeauthor{Chodavarapu:2010iqa} to discard their earlier hypothesis[^same_authors] as to the source of the $10$ bp period in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__ [emphasis added]" and not just for the DRM2 methyltransferase.
 
@@
 \citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
-\citeauthor{Li:2010fb} conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", and that their results are consistent with those in \citet{Eckhardt:2006gh}.
+\citeauthor{Li:2010fb} conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", and report that their results are consistent with those in \citet{Eckhardt:2006gh}.
 In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated regions, coding DNA sequences and various repetitive elements.
@@
 Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are estimates of the correlation of aggregate methylation levels over the range $0$ to $1000$ bp.
-Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs.
+Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs (i.e.
+$NIL = 0$ pairs).
 If all pairs were used, $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome.
@@
 They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
-However, this analysis is fundamentally flawed.
+This analysis, however, is fundamentally flawed.
 
@@
 To do so, \citeauthor{Lacey:2013iy} fit a Gaussian variogram model to the CpG $\beta$-values from chromosome 11 of a normal myotube cell line (_MTCTL2_).
-The empirical and fitted variograms asymptote for CpGs separated by approximately $3000$ bp (Figure 1d of \citet{Lacey:2013iy}, unable to reproduce here due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
+The empirical and fitted variograms asymptote for CpGs separated by approximately $3000$ bp (Figure 1d of \citet{Lacey:2013iy}; I am unable to reproduce it here due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
 
@@
 There is less previous research studying within-fragment co-methylation than there is studying correlations of aggregate methylation levels.
-In Chapter \ref{chap:co-methylation} I extensively explore within-fragment co-methylation and how different variables affect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
+Chapter \ref{chap:co-methylation} extensively explores within-fragment co-methylation and how different variables affect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.
 
@@
 
-[^assembly]: NB Assembling these reads into longer contigs will not solve this problem since the assembly does not guarantee that the reads assembled into longer fragments actually originated from the same cell or haplotype.
+[^assembly]: NB assembling these reads into longer contigs will not solve this problem since the assembly does not guarantee that the reads assembled into longer fragments actually originated from the same cell or haplotype.
 
@@
 
-To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200-250$ bp.
+To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ to $250$ bp.
 
@@
 Error bars represent $95\%$ confidence intervals via critical values of Student $t$ distributions.
-Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Cokus:2008fc}), copyright (2008)}
+Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Cokus:2008fc}), copyright (2008).}
 \label{fig:Cokus_2008_Sup_Fig_14}
@@
 It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
-While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{sec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
+While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{sec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\psi = \frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.
 
@@
 \citet{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.
-The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing ($> 10,000 \times$ median sequencing coverage of each CpG in the target) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
+The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing (the median sequencing coverage of each CpG in the target is greater than $10,000 \times$) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
 \citeauthor{Landan:2012kp} also use these data to estimate co-methylation.
@@
 The general form of the table is shown below in Table \ref{tab:landan}.
-They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{n_{mm} \times n_{uu} - n_{mu} \times n_{um}}{\sqrt{n_{u+} \times n_{m+} \times n_{+u} \times n_{+m}}}$.
+They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{n_{mm} n_{uu} - n_{mu} n_{um}}{\sqrt{n_{u+} n_{m+} n_{+u} n_{+m}}}$.
 Importantly, $r$ is based on within-read measurements from a single sample and so these $r$ are estimates of within-read, within-sample co-methylation.
@@
 \centering
-\caption[Notions of `subject' and `population' in methylation linkage analyses]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}.
+\caption[Notions of `subject' and `population' in methylation linkage analyses]{Notions of `subject' and `population' in traditional genetic linkage and the ``methylation linkage'' of \citet{Landan:2012kp}.
 NB `person' could be any organism, e.g, `mouse', `dog', etc.}
@@
 Unlike other methods for detecting DMRs, which test for DMCs and then cluster these DMCs into DMRs, `Aclust` clusters loci prior to testing for differential methylation.
-That is, it only tests for DMRs and not DMCs.
+That is, `Aclust` tests for DMRs and not DMCs.
 This has the advantage of reducing the dimensionality of the data.

### 2015-05-03 Rmarkdown/comethylation.Rmd
Edits to front matter and table of contents
Edit to co-methylation

@@
 \begin{chapabstract}
-I describe two measures of within-sample co-methylation, \emph{within-fragment co-methylation} and \emph{correlation of $\beta$-values}.
+This chapter describes two measures of within-sample co-methylation, \emph{within-fragment co-methylation} and \emph{correlation of $\beta$-values}.
 The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.
 
-Using estimates of both within-fragment co-methylation and correlation of $\beta$-values, I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci.
-I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.
-
-Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
+Using estimates of both within-fragment co-methylation and correlation of $\beta$-values, we explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci.
+We apply these methods to 40 whole-genome bisulfite-sequencing samples to study how co-methylation varies between different samples.
+
+This chapter focuses on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
 \end{chapabstract}
@@
 1. Define the pairs of methylation loci.
-2. For each $IPD$, compute the correlation of $\beta$-values for all pairs of methylation loci separated by that $IPD$.
+2. For each intra-pair distance ($IPD$), compute the correlation of $\beta$-values for all pairs of methylation loci separated by that $IPD$.
 
@@
 Rather than actually remove these sites, I simply set the corresponding $\beta$-value to `NA`.
-I also set the $\beta$-values of `missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.
+I also set the $\beta$-values of 'missing' CpGs, those without sufficient sequencing coverage to call a $\beta$-value, to `NA` so that these are still included in the set of CpGs.
 
@@
 The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context.
-In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy.
+In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2$ to $2000$ bp, otherwise the number of pairs becomes unwieldy.
 This is what is called an 'autocorrelation' analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$ by \citet{Cokus:2008fc}, although it is somewhat different to the traditional notion of an autocorrelation analysis.
@@
 
-Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two regions of the genome that have very different methylation dynamics.
+Two pairs of CpGs with an identical $IPD$ may come from two regions, even when using the $NIL = 0$ strategy.
+These regions mayhave very different methylation dynamics.
 In other words, the vector of $\beta$-values for a given sample is highly non-stationary; the correlation between two loci depends on more than just the $IPD$.
-An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values.
+An obvious variable to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values.
 To do this, we can stratify our analysis by a genomic feature.
-While this does not eliminate this non-stationarity, it may help identify factors that drive some of the heterogeneity.
+While this does not eliminate this heterogeneity, it may help identify factors that drive some of the variation.
 
@@
 elements need not be in the same feature but rather in the same type of feature.
-The genomic feature should partition the genome so that each loci is either inside or outside of the feature.
+The genomic feature should partition the genome so that each locus is either inside or outside of the feature.
 Since the number of 'spanning pairs' is substantially fewer than those inside or outside of the feature, and the boundaries of the feature are oftentimes fuzzy, I have excluded these 'spanning pairs' in the results shown below.
@@
 Before computing correlations and trying to interpret these, it is a good idea to look at the data from which these correlations are computed.
-Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200, 2000$ with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample.
+Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200, 2000$ with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample from the _Lister_ dataset.
 It highlights two major problems.
@@
 
-For all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets, I compute the Pearson and Spearman correlations using both the $NIL = 0$ and $NIL \geq 0$ strategies.
+For all 40 samples in the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, I compute the Pearson and Spearman correlations using both the $NIL = 0$ and $NIL \geq 0$ strategies.
 I compute these genome-wide and after stratifying CpG pairs by whether they are in a CpG island.
@@
 Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
-Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.
+Since the trend in the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.
 
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
-%\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
+\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
 \label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
-%\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
+\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
 \label{fig:Lister_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
-%\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs %of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
+\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
 \label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
-%\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of %$\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of %CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
+\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
 \label{fig:Ziller_spearman_beta_cors_min_cov_5}
@@
 \FloatBarrier
+
 \subsubsection{Choice of correlation coefficient}
@@
 \FloatBarrier
-\subsection{$NIL = 0$ CpG pairs}
+\subsection{CpG pairs with $NIL = 0$}
 
@@
 Two things immediately stand out.
-Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$; pairs of CpGs separated larger distances are on average less correlated than pairs of CpGs separated by shorter distances.
+Firstly, for all samples the trend is that the strength of the correlation decays as a function of $IPD$; pairs of CpGs separated by larger distances are on average less correlated than pairs of CpGs separated by shorter distances.
 Secondly, the estimated correlations violently jump around for larger $IPD$s.
@@
 
-The are some consistent patterns in comparing these trends across samples.
+There are some consistent patterns in comparing these trends across samples.
 The correlations very quickly drop well below $0.5$, within approximately $50$ bp for all the _EPISCOPE_ and _Seisenberger_ samples and most of the _Lister_ and _Ziller_ samples.
@@
 The notable exceptions are the embryonic stem cell replicates, _H1\_r1_ and _H1\_r2_ (Lister dataset), which begin with a maximum correlation of only around $0.5$.
-This is not observed in the other embryonic stem cells _H9_, _H9\_Laurent_, _HSF1_ and _J1\_1_), which makes it difficult to know whether $\beta$-values of embryonic stem cells are truly less correlated than other cell types.
+This is not observed in the other embryonic stem cells (_H9_, _H9\_Laurent_, _HSF1_ and _J1\_1_), which makes it difficult to know whether $\beta$-values of embryonic stem cells are truly less correlated than other cell types.
 
 The trend of the correlations for all these samples plateaus out close to zero, slightly higher for those samples with significant intermediate methylation.
-However, it should not be concluded from this that the methylation of CpGs separated by more than a few hundred base pairs are uncorrelated.
+It should not be concluded from this, however, that the methylation of CpGs separated by more than a few hundred base pairs are uncorrelated.
 Recall that these results are only for $NIL = 0$ pairs of CpGs, which are increasingly rare for larger $IPD$s.
-These results tell us about the 'first-order' correlations of $\beta$-values, which decay very quickly but whose importance also decreases quite quickly owing to the vast majority of CpGs being within $100$ bp of another CpG (e.g., Figure \ref{fig:IPD_ECDF_hg19} for humans).
+These results tell us about the 'first-order' correlations of $\beta$-values, which decay very quickly but whose importance also decreases quite quickly owing to the vast majority of CpGs being within $100$ bp of another CpG (e.g., Figure \ref{fig:IPD_ECDF_hg19} shows the distribution of $IPD$s for humans).
 
@@
 
-The above are genome-wide plots, all pairs of CpGs with a given $IPD$ are treated identically.
+\FloatBarrier
+
+Figures ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are genome-wide plots; all pairs of CpGs with a given $IPD$ are treated identically.
 However, there are good reasons to expect that the correlations of $\beta$-values, like the $\beta$-values themselves, might be influenced by their genomic context.
@@
 The message from these plots is clear, $\beta$-values of $NIL = 0$ pairs within CpG islands are much more correlated than those pairs outside of islands.
-Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside islands is less influenced by $IPD$.
+Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside CpG islands is less influenced by $IPD$.
 In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD$.
@@
 \FloatBarrier
-\subsection{$NIL \geq 0$ CpG pairs}\label{sec:nil_geq_0_beta_cors}
+\subsection{CpG pairs with $NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}
 
@@
 However, constructing all $NIL \geq 0$ pairs requires much more time and memory than constructing all $NIL = 0$ pairs because there are many more such pairs.
-For this reason, and for compatability with the $NIL = 0$ results, I have focused on $IPD = 2, \ldots, 2000$ for the $NIL \geq 0$ pairs.
+For this reason I have focused on $IPD = 2, \ldots, 2000$ for the $NIL \geq 0$ pairs.
 
 When we aggregate over as many factors as we are in these genome-wide $NIL \geq 0$ plots, we find that there is little difference in the correlation curves between samples.
-All samples have a very high correlation, close to $1$, for small $IPD$s that steadily decays to a lower correlation of $0.2-0.4$ by $IPD = 2000$.
+All samples have a very high correlation, close to $1$, for small $IPD$s that steadily decays to a lower correlation of $0.2$ to $0.4$ by $IPD = 2000$.
 
@@
 
+\FloatBarrier
+
 Comparing CpG islands to non-islands, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2] (Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}).
@@
 
+\FloatBarrier
+
 \subsection{Limitations}
 
-A limitation in to these analyses is they are based on correlations of measurements that are aggregates from many thousands of DNA molecules.
+A limitation to these analyses is that they are based on correlations of measurements that are aggregates from many thousands of DNA molecules.
 We are not measuring co-methylation on the scale at which the biological process of DNA methylation acts, namely at the level of individual DNA molecules.
@@
 
-We will begin with an overview of the simplest case, estimating dependence in a $2 \times 2$ contingency table ($K = 1$).
+We will begin with an overview of the simplest case, estimating dependence in a $2 \times 2$ contingency table (i.e.
+$K = 1$).
 We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables.
-We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determining which is most appropriate for a study of within-fragment co-methylation.
+We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determinine which is most appropriate for a study of within-fragment co-methylation.
 
@@
 We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$.
-The odds is a convenient summary of dependence in a $2 \times 2$ table.Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:
+The odds ratio is a convenient summary of dependence in a $2 \times 2$ table.
+Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:
 
@@
 Suppose we observe $(X, Y)$ for $n$ samples.
-Denote by $n_{ij} the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$.
+Denote by $n_{ij}$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$.
 We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$.
@@
 
-The simplest estimator of $\theta$ is the unconditional maximum likelihood estimator, $\widehat{\theta_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$.
-The asymptotic standard error of $\widehat{\theta_{U}}$ is given by $\hat{\sigma}(\widehat{\theta_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
-The asymptotic distribution of $\widehat{\theta_{U}}$ as $n \to \infty$ is $Gaussian(\theta, \hat{\sigma}(\theta)^{2})$.
-
-A problem with the estimator $\widehat{\theta_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities.
-We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\theta_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\theta_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
-\citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\theta_{U}}$ and $\widehat{\theta_{0.5}}$ have the same asymptotic distribution around $\theta$ as $n \to \infty$ but that $\widehat{\theta_{0.5}}$ has reduced first-order bias.
-The modified estimator, $\widehat{\theta_{0.5}}$, is therefore generally recommended.
-
-Another alternative to $\widehat{\theta_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$.
+The simplest estimator of $\theta$ is the unconditional maximum likelihood estimator, $\widehat{\theta}_{U} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$.
+The asymptotic standard error of $\widehat{\theta}_{U}$ is given by $\hat{\sigma}(\widehat{\theta}_{U}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
+The asymptotic distribution of $\widehat{\theta}_{U}$ as $n \to \infty$ is $Gaussian(\theta, \hat{\sigma}(\theta)^{2})$.
+
+A problem with the estimator $\widehat{\theta}_{U}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities.
+We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\theta}_{0.5} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $\hat{\sigma}(\widehat{\theta}_{0.5}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$.
+\citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\theta}_{U}$ and $\widehat{\theta}_{0.5}$ have the same asymptotic distribution around $\theta$ as $n \to \infty$ but that $\widehat{\theta}_{0.5}$ has reduced first-order bias.
+The modified estimator, $\widehat{\theta}_{0.5}$, is therefore generally recommended.
+
+Another alternative to $\widehat{\theta}_{U}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$.
 This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$.
@@
 The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods.
-The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
 
@@
 Sparsity can occur even when the sample size, $n$, is large.
-Sparsity becomes more of a problem as we move from a single $2 \times 2$ table to $K$ $2 \times 2$ tables or, equivalently, a $2 \times 2 \times K$ table.
+Sparsity becomes more of a problem as we move from a $2 \times 2$ table to a $2 \times 2 \times K$ table.
 
@@
 We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum.
-For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{+ 1k} = 1 - \pi_{+ 2k}, \pi_{1 + k} = 1 - \pi_{2 + k}$ to vary.
-
-We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.
+We assume for now that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{+ 1k} = 1 - \pi_{+ 2k}, \pi_{1 + k} = 1 - \pi_{2 + k}$ to vary.
+
+We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the odds ratio $\psi$ \citep[see][]{Breslow:1981ta}.
 
@@
 This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance].
-This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.
+This delay was in part due to the existance of two asymptotic forms of $2 \times 2 \times K$ contingency tables.
 
@@
 
-The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{+ + k}$, grows large.
+The first asymptotic environment, $A1$, assumes that $K$ remains small while the the sample sizes in each strata, $n_{+ + k}$, grows large.
 The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{+ + k}$ remains small.
@@
 
-Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse, \citet{Liang:1985uw} developed three tests for the assumption of homogeneity of odds ratios and compared these with two tests for the same assumption but designed under $A1$.
-Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$.
+Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse and $K \to \infty$, \citet{Liang:1985uw} developed three conditional tests of the homogeneity of odds ratios assumption and compared these with two unconditional tests that were designed under $A1$.
+Unsurprisingly, \citeauthor{Liang:1985uw} found that the three tests designed for $A2$ are more appropriate than either of the tests designed for $A1$ when the data are simulated under $A2$.
 The disadvantage of these conditional tests is the extra computation required.
-\citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.
 It has been noted, however, that statistical tests of homogeneity versus heterogeneity typically have low power and are only able to detect large departures form homogeneity \cite[e.g.,][]{Hauck:1989vt}.
+Therefore these tests are not always of great practical use.
 
@@
 [^nb_sim]: In the simulation study, a larger value of 'trunc' results in higher quality data since we can simply keep simulating until we have $K$ tables with sufficient sequencing depth.
-In analysing real data, increasing this cutoff may be detrimental since we end up with fewer tables to analyse.
+In analysing real data, however, increasing this cutoff may be detrimental since we end up with fewer tables to analyse.
 
@@
 
-1. The $K$ 0.5-adjusted unconditional odds ratio estimates, $\widehat{\psi}_{0.5}$, from each $2 \times 2$ table.
-2. The unconditional odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\psi}_{U}$.
-3. The Mantel-Haenszel odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\psi}_{MH}$.
-
-The results of each simulation are summarised by a plot of $\widehat{\theta}_{0.5} = \log(\widehat{\psi}_{0.5})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U} = \log(\widehat{\psi}_{U})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi}_{MH})$ (coloured blue).
-The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line[^density].
+1. The $K$ 0.5-adjusted unconditional log odds ratio estimates, $\widehat{\theta}_{0.5}$, from each $2 \times 2$ table.
+2. The unconditional log odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\theta}_{U}$.
+3. The Mantel-Haenszel log odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\theta}_{MH}$.
+
+The results of each simulation are summarised by a plot of $\widehat{\theta}_{0.5}$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The density of $\theta_{k}$, the value of the true log odds ratio in the $k^{th}$ slice of the $2 \times 2 \times K$ table, is shown by a dashed black line[^density].
 
@@
 To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1a, Figure \ref{fig:Simulation_1a}), and $K = 1000$ (Simulation 1b, Figure \ref{fig:Simulation_1b}).
-Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
-This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller.
-By contrast, the estimates of $\theta$ from the individual $2 \times 2$ tables, $\widehat{\theta_{0.5}}$, are highly biased, even when $K = 1000$ in Simulate 1b (Figure \ref{fig:Simulation_1b}).
+Under this simulation, we expect both $\widehat{\theta}_{U}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
+This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is small.
+By contrast, the estimates of $\theta$ from the individual $2 \times 2$ tables, $\widehat{\theta}_{0.5}$, are highly biased, even when $K = 1000$ in Simulate 1b (Figure \ref{fig:Simulation_1b}).
 In both Simulation 1a and 1b (Figures \ref{fig:Simulation_1a} and \ref{fig:Simulation_1b}) the histograms of $\widehat{\theta}_{0.5}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).
@@
 
-The bias can be eliminated only by having ultra-deep sequencing coverage, such as in Simulation 1c where $mean(d) = 10000 \times$ (Figure \ref{fig:Simulation_1c}).
+The bias can be eliminated only by having ultra-deep sequencing coverage, such as in Simulation 1c where the average sequencing coverage is greater $10,000 \times$.
 Even then, however, the variation of $\widehat{\theta}_{0.5}$ across the $K = 1000$ tables remains non-negligible when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$ are near the boundaries.
@@
 
+\FloatBarrier
+
 Next, we explore what happens when we allow the marginal probabilities, $\pi_{2 +}$ and $\pi_{+ 2}$, to vary across the $K = 100$ tables for a fixed (possibly non-zero) value of $\theta$.
@@
 Furthermore, each methylation locus in a pair will likely have a similar marginal probability of methylation.
-To simulate this more relevant scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}).
+To simulate this more relevant scenario, Simulation 3 uses $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}).
 
@@
 
-- Simulation 4a and Simulation 4b: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
-- Simulation 5:$\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
-- Simulation 6: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$.
+- Simulation 4a and Simulation 4b: $\theta_{k} \eqd Gaussian(\theta_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 5: $\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
+- Simulation 6: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$, i.e. a mixture distribution.
 
@@
 
+\FloatBarrier
+
 \subsection{Summary}
@@
 
-This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual pairs of methylation loci using the simple estimator odds ratio estimator, $\theta_{0.5}$, unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
-The distributions of $\widehat{\theta_{0.5}}$ at typical sequencing depths ($d \approx 30$) are highly biased and dispersed, even under very simple simulation scenarios.
+This simulation study demonstrates that we cannot estimate within-fragment co-methylation at individual pairs of methylation loci using the simple odds ratio estimator, $\theta_{0.5}$, unless ultra-deep[^ultra-high_depth] sequencing is used ($d > 10000 \times$).
+The distributions of $\widehat{\theta}_{0.5}$ at typical sequencing depths ($d \approx 30$) are highly biased and dispersed, even under very simple simulation scenarios.
 Therefore, some degree of aggregation is required in order to study within-fragment co-methylation.
@@
 
-Here I describe a simple method to study within-fragment co-methylation by analysing methylation patterns at pairs of methylation loci (2-tuples).
-For simplicity, I will use pairs of CpGs, but the method is applicable to any methylation type.
+This section describes a simple method to study within-fragment co-methylation by analysing methylation patterns at pairs of methylation loci (2-tuples).
+For simplicity, we will use pairs of CpGs, but the method is applicable to any methylation type.
 This method is to be implemented in the `cometh()` function that is part of the `MethylationTuples` software[^cometh].
 
-[^cometh]: At the time of writing, this is available as a stand along `mantelhaen()` function, but it will soon be one of several strategies of estimating within-fragment co-methylation using the `cometh()` function.
+[^cometh]: At the time of writing, this is available as a stand alone `mantelhaen()` function, but it will soon be one of several strategies of estimating within-fragment co-methylation using the `cometh()` function.
 
@@
 1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
-2. What factors influences this dependence?
+2. What factors influence this dependence?
 
@@
 
-Answers to the latter two are informed by Section \ref{sec:sparse_2x2xK_tables}.
+Answers to the latter two are informed by Section \ref{sec:sparse_2x2xK_tables}, in particular the simulation study.
 
@@
 
-The construction of CpG pairs is very similar to constructing the pairs of methylation levels, described in Section \ref{sec:beta_pair_construction}.
+The construction of CpG pairs is very similar to constructing the pairs of $\beta$-values, described in Section \ref{sec:beta_pair_construction}.
 However, we are now limited to analysing co-methylation of loci that can be captured within a single read.
@@
 
-I have analysed all 40 samples in the EPISCOPE, Lister, Seisenberger and Ziller datasets.
-For each sample, I compute $\widehat{\theta}_{MH}$ using different levels of aggregation (listed here in decreasing order):
+We analyse all 40 samples in the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets.
+For each sample we compute $\widehat{\theta}_{MH}$ using different levels of aggregation (listed here in decreasing order):
 
@@
 
-Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using $NIL = 0$ CpG pairs for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using CpG pairs with $NIL = 0$ for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
 
@@
 The mitochondria are typically almost completely unmethylated, i.e.
-the marginal probability of methylation at a CpG on the mitochondria is very close to zero, and so the concept of co-methylation is not particularly meaningful here.
+the marginal probability of methylation at a CpG on the mitochondria is very close to zero, and so the concept of co-methylation is perhaps not particularly meaningful here.
 
@@
 The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_merged_ac_MH_by_seqnames}.
-In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_)
+In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_; _ADS_ is a female cell line), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_; _IMR90_ is a female cell line).
 
@@
 
-By analysing $NIL \geq 0$ pairs we are increasing the heterogeneity of the $K$ $2 \times 2$ tables, which makes more difficult the interpretation of these results.
+By analysing CpG pairs with $NIL \geq 0$ we are increasing the heterogeneity of the $K$ $2 \times 2$ tables, which consequently makes more difficult the interpretation of these results.
 However, it does give the advantage of allowing the analysis of pairs with larger $IPD$s.
-Nonetheless, the short DNA fragments and read lengths of our data make it difficult to perform reliable chromosome-level inference in even the highest-quality samples for $IPD$s greater than $250$ to $300$.
+Nonetheless, the short DNA fragments and read lengths of our data make it difficult to perform reliable chromosome-level inference in even the highest-quality samples for $IPD$s greater than $250$ to $300$ bp.
 Retreating to a genome-level analysis suggests that within-fragment co-methylation is active over at least a few hundred basepairs in distance.
@@
 $Z_{h, i + i'}$ and $Z_{i - i''}$, $i', i'' > 0$).
-To summarise, the $NIL = 0$ analyses reduce heterogeneity by focusing on adjacent loci but cannot (by themselves) tell us whether within-fragment co-methylation is truly 'first-order'.
+The $NIL = 0$ analyses reduce heterogeneity by focusing on adjacent loci but cannot (by themselves) tell us whether within-fragment co-methylation is truly 'first-order'.
 
@@
 
+\FloatBarrier
+
 \subsubsection{Variation of within-fragment co-methylation between samples}
@@
 Reasuringly, samples done in technical duplicate (_IMR90\_r1_ and _IMR90\_r2_; _H1\_r1_ and _H1\_r2_) have very similar genome-level estimates of $\theta$.
-Furthermore, biological samples assayed by different laboratores (_H9_ and _H9\_Laurent_; _IMR90\_r1, _IMR90\_r2_ and _IMR90\_cell\_line_) also have fairly similar genome-level estimates of $\theta$[^imr90_ac].
+Furthermore, biological samples assayed by different laboratories (_H9_ and _H9\_Laurent_; _IMR90\_r1_, _IMR90\_r2_ and _IMR90\_cell\_line_) also have fairly similar genome-level estimates of $\theta$[^imr90_ac].
 
@@
 Recall that the _Lister_ dataset contains four 'mini datasets': _Lister-ADS_, _Lister-FF_, _Lister-IMR90_ and _Lister-H1_ (see section \ref{sec:Lister_dataset}).
-In each mini datasets, the pluripotent cell lines (induced pluripotent stem cells or embryonic stem cells) have larger genome-level estimates of $\theta$ than do their precursor forms[^bmp4].
+In each mini datasets, the pluripotent cell lines (induced pluripotent stem cells or embryonic stem cells) have larger genome-level estimates of $\theta$ than do their precursor forms.
 This difference is less pronounced, even absent, in the samples differentiated _in vitro_ with BMP4 from the induced pluripotent stem cell lines (iPSCs).
 
-This suggests the pluripotent cell cells have stronger within-fragment co-methylation than do somatic or differentiated cells.
+This suggests that pluripotent cells have stronger within-fragment co-methylation than do somatic or differentiated cells.
 Consistent with this is that the _H9_, _H9\_Laurent_ and _J1_ samples (embryonic stem cell lines), the _E6.5\_epiblast\_1_ sample (epiblasts derived from day 6.5 embryos) and the _E16.5\_male\_1_ (male progenitor germ cells derived from day 16.5 embryos) all have relatively high genome-wide estimates of $\theta$.
-
-[^bmp4]: The difference is less pronounced in the samples differentiated _in vitro_ with BMP4, e.g., comparing _FF-iPSC\_19.11_ with _FF-iPSC\_19.11+BMP4_.
 
@@
 If the iPSC and iPSC-derived cell lines have had less (developmental) time between the laying down of the original DNA methylation pattern and the point at which they were sequenced, then there is a lower probability that the co-methylation has been disrupted by replication errors.
-There are also reports that DNA methylation profiles differ widely between cell lines and their precursor primary tissues \cite{Nestor:2015db}, which very likely also effect co-methylation.
+There are also reports that DNA methylation profiles differ widely between cell lines and their precursor primary tissues \cite{Nestor:2015db}, which very likely also affect co-methylation.
 
@@
 This result is tantalisingly similar to the periodicity we see in the correlations of $\beta$-values and the approximate spacing of nucleosomes in human DNA (see Section \ref{sec:nucleosome_and_chromatin}).
-However, it is not as apparent in other paired-end samples, and this is only likely to be resolved with longer reads that can span multiple nucleosomes.
+However, it is less apparent in other paired-end samples, and this is only likely to be resolved analyses of longer reads that can span multiple nucleosomes.
 
@@
 
-Another complexity that these analyses have brushed over is the fact that the genome is not a one-dimensional line but is rather a complex three-dimensional structure whose shape also varies in time.
-Therefore, the relevant distance between two loci is not necessarily the number of bp between them but may instead be the Euclidean distance in three-dimensional space.
+Another complexity that these analyses have brushed over is the fact that the genome is not a one-dimensional object but is rather a complex three-dimensional structure (whose shape also varies in time).
+Therefore, the relevant distance between two loci is not necessarily the number of base pairs between them but may instead be the Euclidean distance in three-dimensional space.
 Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the three-dimensional structure of the genome, such as those provided by chromatin conformation capture technologies \citep[e.g.,][]{Dekker:2013hi}.

### 2015-05-03 Rmarkdown/methsim.Rmd
Edits to methsim chapter

@@
 \texttt{methsim} incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software.
-We explore different parameter choices and highlight promising directions of research, such as using \texttt{methsim} to simulate data in order to compare methods for identifying differential methylation.
+We explore different parameter choices and highlight promising directions of research, as well as areas requiring further improvement in ongoing work.
 \end{chapabstract}
@@
 In fact, the huge variety of experimental factors, and their possible values, can make such a task infeasible.
-But perhaps the bigger hurdle is that there are not attractive experiments to perform; why spend your time and money on an experiment where you 'know the answer' when you could be spending that same time and money on investigating some new biology?
+But perhaps the bigger hurdle is that these are not attractive experiments to perform; why spend your time and money on an experiment where you 'know the answer' when you could be spending that same time and money on investigating some new biology?
 In these scenarios and others, simulation studies play a vital role in applied statistics, where they can be used in the development, validation and benchmarking of different analysis methodologies.
@@
 The key advantage of simulated data is that we know the truth _a priori_.
-Moreover, we can manipulate the truth via parameters in the simulation model and can examine how a method performs under a variety of scenarios.
+Moreover, we can manipulate the truth via parameters in the simulation model and can examine how an analysis method performs under a variety of scenarios.
 And this manipulation is cheap, a mere matter of changing parameters and re-running a piece of software, which means that we can investigate a broad range of plausible scenarios.
@@
 Simulation studies can provide many insights into the performance of a method.
-At the most basic level, if a method fails performs poorly when applied to 'easy' simulated data, then it is very unlikely to work well when applied to more complex real data.
+At the most basic level, if a method fails or performs poorly when applied to 'easy' simulated data, then it is very unlikely to work well when applied to more complex experimental data.
 We can also identify which scenarios are 'easy' (ones where most methods are able to identify the truth) and scenarios under which certain methods perform better than others.
@@
 The most common use of simulation models in applied statistics is to repeatedly generate datasets under a range of parameter settings.
-This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to analyse the full space of scenarios.
+This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to explore the full space of scenarios.
 An exception to this rule may occur when the simulation model is used to test a proposed mechanistic or stochastic model of a phenomenon, such as in studies of molecular dynamics.
@@
 It is often simpler to simulate from a model-based simulation, particularly if the model is a well-studied parametric distribution.
-However, this simplicity often comes at the cost of increased assumptions, whose validity may be questionable.
+This simplicity often comes at the cost, however, of increased assumptions, whose validity may be questionable.
 Simulations based on sampling of real data may reduce the number of assumptions required.
-However, care must be taken in selecting the units to be sampled so that the sampling process is efficient and so that the sampled data don't grossly distort within- and between-sample dependencies.
+However, care must be taken in selecting the units to be sampled so that the sampling process is efficient and so that the sampled data don't grossly distort within-sample and between-sample dependencies.
 
@@
 All of the currently available methods for simulating bisulfite-sequencing reads are designed for the comparison of alignment strategies and are model-based.
-These are not suitable for comparisons of downstream analysis methods.
+These are not generally suitable for comparisons of downstream analysis methods.
 
-`Sherman` ([http://www.bioinformatics.babraham.ac.uk/projects/sherman/](http://www.bioinformatics.babraham.ac.uk/projects/sherman/)) is software to simulate bisulfite-sequencing reads, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts as `FASTQ` files.
+`Sherman` ([http://www.bioinformatics.babraham.ac.uk/projects/sherman/](http://www.bioinformatics.babraham.ac.uk/projects/sherman/)) is software to simulate bisulfite-sequencing reads as `FASTQ` files, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts.
 The simulated reads are designed for comparing the performance of different alignment strategies.
@@
 
-While suitable for comparing alignment strategies, `Sherman` produces data that is not suitable for use in comparing downstream analysis methods.
+While appropriate for comparing alignment strategies, `Sherman` produces data that is not suitable for use in comparing downstream analysis methods.
 All CG (resp.
@@
 As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}.
-`WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data when comparing methods for identifying differential methylation.
+`WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data for comparing methods for identifying differential methylation.
 `WGBSSuite` is available for download as a collection of R scripts[^WGBSSuite].
 
-[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software.
+[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) which would greatly simplify the installation and use of the software.
 Furthermore, no license file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code.
@@
 
-I do not think this a useful or necessary step and may even be counterproductive.
+I do not think this a useful or necessary step, and it may even be counterproductive.
 While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed.
-If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome.
+If sequence variation is required then it is easily accommodated by sampling from the set of methylation loci in the reference genome.
 Furthermore, the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci.
 
-Almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.
+Almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.
 
@@
 They do this by using a Normal copula to make the set of sequencing depths a jointly dependent set of random variables.
-While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the benefits are unclear.
+While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the cost-benefit trade-off is unclear.
 Moreover, it is simpler, and likely more computationally efficient, to include such correlations by a sensible sampling of sequencing depths from real data.
@@
 These are based on the true underlying methylation level, $B_{i, j}$, and the sequencing depth, $d_{i, j}$.
-In the beta-binomial model, this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$.
+In the beta-binomial model, this is done by binomial sampling where $M_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
 In addition to binomial read sampling, the `WGBSSuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.
@@
 Such parameters include the length of the DMR, the minimal number of loci it must contain, the maximal intra-pair distances of loci within the DMR and how many of the loci in the DMR must themselves be differentially methylated, e.g., should Figure \ref{fig:dmr_break} be considered one DMR or two DMRs?
-Such decisions ultimately have to be made by the user of the simulation software based on the types of events they are interested in.
+Such decisions ultimately have to be made by the user of the simulation software based on the types of events she is interested in analysing.
 
@@
 \caption[A hypothetical differentially methylated region]{A hypothetical region in a study of differential methylation in a two-group experiment.
-Plotted is the difference in $\beta$-value between the two groups ($\delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}).
+Plotted is the difference in $\beta$-value between the two groups ($\Delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}).
 The first three methylation loci are DMLs, as are the last five methylation loci.
@@
 The methylation states of neighbouring loci are highly dependent and, consequently, the $B_{i, j}$ of nearby loci are highly dependent.
-To be clear, the majority of studies using simulated DNA methylation data have failed to model a very important feature of DNA methylation data, co-methylation.
+To be clear, the majority of studies using simulated DNA methylation data do not model an important feature of DNA methylation data, co-methylation.
 
@@
 A simulation may also be based entirely on re-sampling of real data[^synthetic].
-This type of simulated data is attractive because through careful sampling it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model.
+This type of simulation is attractive because through careful sampling it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model.
 At the same time, however, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features or, worse still, introduce artefacts into the simulated data.
@@
 
-\citet{Sofer:2013bk} use a re-sampling based simulation method in the development of their `Aclust` software for identifying differential methylation from microarray data.
+\citet{Sofer:2013bk} use a re-sampling based simulation method in the development of their `Aclust` software (designed for identifying differential methylation from microarray data).
 The idea is adapted from \citet{Gaile:2007bq}, which is a simulation method for array comparative genomic hybridization experiments.
@@
 If a block does not contain a target then it is sampled uniformly at random from the 539 samples.
-However, if a block does contain a target then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the target CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the target CpG[^cases_and_controls].
-This is essentially weighted sampling of the real data to induce differential methylation.
+If a block does contain a target, however, then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the target CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the target CpG[^cases_and_controls].
+This is essentially weighted re-sampling of the real data to induce differential methylation.
 
@@
 As we have seen, DNA methylation is highly heterogeneous along the genome.
-Nonetheless, there are clear regions of `similarity', such as the unmethylated CpG islands and long partially methylated domains.
+Nonetheless, there are clear regions of 'similarity', such as the unmethylated CpG islands and long partially methylated domains.
 In these locally-similar regions we might hope to model DNA methylation by a simple parametric model.
@@
 1. Segment the methylome into regions of similarity.
-2. Sample parameters for each methylation locus, $i$, based on the segmentation.
+2. Sample parameters for the $i^{th}$ methylation locus based on the segmentation and the empirical distributions of key statistics.
 
@@
 
-`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarity, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output].
+`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarity, namely \emph{UMR}, \emph{LMR}, \emph{PMRS} and \emph{other}[^methylseekr_output].
 Roughly speaking, _other_ regions are 'mostly methylated regions' (see Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Ziller_beta_by_pm_boxplots}), although \citet{Burger:2013kq} do not describe these regions as such.
 
-[^methylseekr_output]: The output returned by `MethylSeekR` does not strictly partition the methylome since it is neither disjoint nor exhaustive.
+[^methylseekr_output]: The output returned by `MethylSeekR` does not strictly partition the methylome since it is neither disjoint nor exhaustive, hence the need for post-processing.
 
@@
 
-`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$.
+`methsim` uses a two-state ($1 =$ methylated, $0 =$ unmethylated), first-order Markov chain to model $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$.
 This allows the incorporation of within-fragment co-methylation into the simulation.
-The choice of a one-step process is one of computational simplicity and is a reasonable approximation for most of the genom given the available data (see Chapter \ref{chap:co-methylation}).
+The choice of a first-order Markov process is one of computational simplicity and is a reasonable approximation for most of the genome given the available data (see Chapter \ref{chap:co-methylation}).
 
@@
 For this reason I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous.
-In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for this pair of loci[^context].
-That is, $p_{i; a, b} = Pr(Z_{h, i + 1} = b | Z_{h, i} = a, IPD(i, i + 1), r_{i})$.
+In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for each pair of loci[^context].
 
@@
 
+\FloatBarrier
+
 There are many ways to simulate values of $\mathbf{B}$ and $\bm{\theta} = \log{\bm{\psi}}$ based on these empirical distributions.
@@
 The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
-  %\item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5.
-  %\item[\emph{A}] $\theta$ identically zero.
-This is to simulate independence of within-fragment methylation states and as a control to check that `methsim` is working as intended for simulating within-fragment co-methylation.
-  %\item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
+  \item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to $0.5$.
+  \item[\emph{A}] $\bm{\theta}$ identically zero.
+This is to simulate independence of within-fragment methylation states and as a control to check that \texttt{methsim} is working as intended for simulating within-fragment co-methylation.
+  \item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta}_{MH}$ (autosomes only).
 \end{description}
@@
 `methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing.
-The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
+The user specifies the desired read length[^read_length], the average sequencing coverage, the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
 We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $30 \times$ in what follows.
@@
 The number of reads per chromosome is assigned proportional to the chromosome length.
-Then, the start of each read is sampled uniformly across the chromosome respective chromosome.
+Then, the start of each read is sampled uniformly across the respective chromosome.
 We only retain those reads that overlap a methylation locus.
@@
 
-The result of this process is a `SimulatedWGBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states.
+The result of this process is a `SimulatedBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states of the read.
 
@@
 
-The `SimulatedWGBS` object contains all the data from the simulation.
+The `SimulatedBS` object contains all the data from the simulation.
 However, it is not always the most convenient or efficient format with which to work.
 For example, many downstream analyses only make use of $\beta$-values, so we might want to summarise our simulated data this way.
-`methsim` provides the `asMethPat()` function to convert the `SimulatedWGBS` object to a `MethylationTuples::MethPat` object containing methylation patterns at m-tuples of a given size[^simplify]; all the functionality of the `MethylationTuples` package is then available to the user, such as computing $\beta$-values with the `methLevel()` method.
+`methsim` provides the `asMethPat()` function to convert the `SimulatedBS` object to a `MethylationTuples::MethPat` object containing methylation patterns at m-tuples of a given size[^simplify]; all the functionality of the `MethylationTuples` package is then available to the user, such as computing $\beta$-values with the `methLevel()` method or the correlations of $\beta$-values using the `methLevelCor()` function.
 
-[^simplify]: Alternatively, the user may use the `simplify` argument of the `simulate()` method to return an already 'simplified' `MethylationTuples::MethPat` object rather than the `SimulatedWGBS` object.
+[^simplify]: Alternatively, the user may use the `simplify` argument of the `simulate()` method to return an already 'simplified' `MethylationTuples::MethPat` object rather than the `SimulatedBS` object.
 
@@
 
-The above describes how to simulate a single whole-genome bisulfite-sequencing sample.
-If we wanted to simulate six independent samples then we could simply run this procedure six times, with a different `MethylomeParam` object for each realisation.
+So far we have described how to simulate a single whole-genome bisulfite-sequencing sample.
+If we want to simulate six independent samples then we could simply run this procedure six times, with a different `MethylomeParam` object for each realisation.
 However, we will generally be interested in simulating experiments where there is some relationship between the samples.
@@
 
-To simulate these type of experiments, we need to be able to control the variation between the resulting `SimulatedWGBS` objects.
+To simulate these type of experiments, we need to be able to control the variation between the resulting `SimulatedBS` objects.
 In this example, we probably want the three samples in each group to be more similar to each other than to the three samples in the other group.
@@
 If we want two samples to be very different from one another, then we would use two different `MethylomeParam` objects at the beginning of the process.
-However, if we want to simulate two technical replicates, then we would use the same `WGBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.
+If we want to simulate two technical replicates, however, then we would use the same `WGBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.
 
@@
 
-The `SimulatedMethylome` and `SimulatedWGBS` objects are large in memory.
+The `SimulatedMethylome` and `SimulatedBS` objects are large in memory.
 This is to some extent unavoidable; we are simulating whole-genome bisulfite-sequencing sequencing experiments that produce an enormous amount of data.
@@
 Reassuringly, we see no evidence of within-fragment co-methylation in _m1AI_, which simulates independence of within-fragment methylation states.
-The trend of the genome-level estimates, by design, very closely matches that of the _ADS_ input sample.
+The trend of the genome-level estimates, by design, very closely match that of the _ADS_ input sample.
 We see, however, that the chromosome-level estimates of _m1BI_, _m2BI_, _m3BI_ and _m3BII_ are perhaps a little too homogeneous when compared to the _ADS_ input sample.
@@
 Each sub-population has its own $\bm{\theta}$ and, although these are sampled from the same parametric model, this suggests that such heterogeneity in the sample will artificially inflate estimates of within-fragment co-methylation, even those made at the genome-level.
-While this requires further investigation, it raises the possibilit that increased genome-level estimates of within-fragment co-methylation may in fact be measuring increased heterogeneity of the sample.
+While this requires further investigation, it raises the possibility that increased genome-level estimates of within-fragment co-methylation may in fact be measuring increased heterogeneity of the sample.
 
@@
 Unfortunately, the results are not so promising for the correlations of $\beta$-values.
+
+\FloatBarrier
 
@@
 
-Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming consequences of this modelling decision.
+Figure \ref{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the dramatic consequences of this modelling decision.
 When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong.
-Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}).
+Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}).
 This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.

### 2015-05-03 Rmarkdown/concluding_remarks.Rmd
Edits to concluding remarks and appendix

@@
 The second chapter details the bioinformatics analysis of whole-genome bisulfite-sequencing data.
-Beginning with the `FASTQ` files created by the DNA sequencers, we review the quality control procedures, read mapping and post-processing essential to creating a `BAM` file creating high-quality mapped reads.
+Beginning with the `FASTQ` files created by the DNA sequencers, we review the quality control procedures, read mapping and post-processing steps that are essential to creating a `BAM` file containing high-quality mapped reads.
 We then describe methods for calling methylation events from bisulfite-sequencing data.
-The chapter concludes by introducing m-tuples and developing the `methtuple` software that can, uniquely, call methylation patterns at m-tuples from whole-genome bisulfite-sequencing data.
+The chapter concludes by introducing m-tuples and developing the `methtuple` software that uniquely can call methylation patterns at m-tuples from whole-genome bisulfite-sequencing data.
 
@@
 A limitation to the development of these methods has been a lack of software.
-To complement to the `methtuple` software, the chapter concludes with the development of the `MethylationTuples` R package, which is designed to facilitate analyses based on methylation patterns at m-tuples.
+To complement the `methtuple` software, the chapter concludes with the development of the `MethylationTuples` R package, which, together, are designed to facilitate analyses based on methylation patterns at m-tuples.
 
 The sixth and seventh chapters are a comprehensive review and analysis of _co-methylation_, the spatial dependence of DNA methylation.
-Chapter \ref{chap:co-methylation} re-examines previous analyses of co-methylation and finds that these have been limited by inadequate data and deficiencies in the statistical methods applied.
+Chapter \ref{chap:co-methylation_review} re-examines previous analyses of co-methylation and finds that these have been limited by inadequate data and deficiencies in the statistical methods applied.
 We address some of these limitations with an extensive analysis of co-methylation in Chapter \ref{chap:co-methylation}.
@@
 DNA methylation data have a complex dependence structure, as demonstrated by the analyses presented in this thesis.
-Any analysis must bear in mind the multiple sources of variation in the data and the level of aggregation required in order to extract meaningful results.
+Any analysis must bear in mind the multiple sources of variation in the data and the amount of aggregation required in order to extract meaningful results.
 
@@
 Much of the effort in this thesis has gone into developing software to extract and manage this information (`methtuple` and `MethylationTuples`) so that new tools can be built on a higher foundation (`methsim`).
-It is my hope that, with continued development, these tools may facilitate other analyses to unravel the complexity of DNA methylation and its many roles in biology.
+It is my hope that, with continued development, these tools may facilitate other analyses to unravel the complexity of DNA methylation and its many biological roles.

### 2015-05-04 Rmarkdown/biology_background.Rmd
Edits suggested by Terry

@@
 
-Acytosine may be described being 'methylated' or 'unmethylated', however, care must be taken when using these terms.
+A cytosine may be described being 'methylated' or 'unmethylated', however, care must be taken when using these terms.
 At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated.
@@
 
-Assays based on restriction enzyme were some of the first developed for studying DNA methylation.
+Assays based on restriction enzymes were some of the first developed for studying DNA methylation.
 These were initially developed for studying a small number of loci although they have been extended to genome-scale analysis approaches \citep{Laird:2010iv}.
@@
 
-Ideally, an assay of DNA methylation could natively 'read' methylated cytosines as a distinct signal from unmethylated cytosines.
+Ideally, an assay of DNA methylation could simply 'read' methylated cytosines as a distinct signal from unmethylated cytosines.
 This idea forms the basis of assays using the kinetics of DNA to infer the presence of DNA methylation and other DNA modifications.

### 2015-05-04 Rmarkdown/appendix.Rmd
Fix formatting
Edits to concluding remarks and appendix
Correct a typo in appendix
Correct Python version

@@
  \begin{pmatrix}
-  (1 - p_{h'}) + c p_{h} & p_{h'} - c p_{h} \\
-  \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) & \big [ p_{h'} + c (1 - p_{h}) \big ] p_{h}
+  (1 - p_{h'}) +c p_{h} & p_{h'} -c p_{h} \\
+  (1 - p_{h'}) -c (1 - p_{h}) & p_{h'} + c (1 - p_{h})
  \end{pmatrix}
@@
 
-__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__
-
-\subsection{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
+\subsection{Estimating the number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
 
@@
 
-The analyses described in my thesis made extensive use of R (`R Under development (unstable) (2014-10-29 r66891)` and `R version 3.2.0 (2015-04-16)`) and Python (`v2.6.6`).
+The analyses described in my thesis made extensive use of R (`R Under development (unstable) (2014-10-29 r66891)` and `R version 3.2.0 (2015-04-16)`) and Python (`v2.7`).
 All analyses were run on one of the servers described above.

### 2015-05-04 Rmarkdown/biology_background.Rmd
Remove italics

@@
 
-In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated _5-methylcytosine_ from nucleic acid prepared from cow thymus \citep{HOTCHKISS:1948va}.
+In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated 5-methylcytosine from nucleic acid prepared from cow thymus \citep{HOTCHKISS:1948va}.
 Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.
 
-The typical site of DNA methylation is at the C5 carbon position of a cytosine base, hence _5-methylcytosine_. Figure \ref{fig:5mC} shows the structure of 5mC.
+The typical site of DNA methylation is at the C5 carbon position of a cytosine base, hence 5-methylcytosine. Figure \ref{fig:5mC} shows the structure of 5mC.
 
@@
 Another protein, _DNMT3L_, is homologous to _DNMT3a_ and _DNMT3b_ but does contain catalytic domain that is necessary for methyltransferase activity.
-Instead, DNMT3L is thought to stimulate the activity of _DNMT3a_ and _DNMT3b_ \citep{Jurkowska:2011dt}.
+Instead, _DNMT3L_ is thought to stimulate the activity of _DNMT3a_ and _DNMT3b_ \citep{Jurkowska:2011dt}.
 
@@
 
-1. The direct removal of the methyl group from _5mC_ is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
+1. The direct removal of the methyl group from 5mC is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
 2.
@@
 5mC oxidation-dependent active DNA demethylation.
-This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ $\to$ _5hmC_ $\to$ _5fC_ $\to$ _5caC_ reaction.
-The removal of _5hmC_, _5fC_ or _5caC_ is biochemically 'easier' than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.
+This follows from the observation that the TET enzymes can iteratively oxidate a 5mC $\to$ 5hmC $\to$ 5fC $\to$ 5caC reaction.
+The removal of 5hmC, 5fC or 5caC is biochemically 'easier' than the removal of 5mC and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.
 
-One question raised by the third point is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
+One question raised by the third point is whether 5hmC, 5fC and 5caC are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.
 This is an active area of research.

### 2015-05-04 Rmarkdown/comethylation.Rmd
Add plots highlighting bizarre chr21 results for within-fragment co-methylation

@@
 The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods.
-The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.
+The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep[pp.
+157-158]{Agresti:2007tf}.
 
@@
 
-Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using CpG pairs with $NIL = 0$ for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
+Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_MH_by_seqnames} show the results using CpG pairs with $NIL = 0$ for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
 
@@
 All samples were sequenced with paired-end reads.}
-\label{fig:Ziller_merged_MH_by_seqnames}
+\label{fig:Ziller_MH_by_seqnames}
 \end{figure}
@@
 
-The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_merged_ac_MH_by_seqnames}.
+The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_ac_MH_by_seqnames}.
 In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_; _ADS_ is a female cell line), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_; _IMR90_ is a female cell line).
@@
 The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
-All samples were sequenced with paired-end reads.}
+All samples were sequenced with paired-end reads.
+There is no Y chromosome or mtDNA data for \emph{E18BUF} due to a coding error that meant these chromosomes weren't processed by \texttt{methtuple} for \texttt{2ac} tuples.
+This omission does not affect the other results.}
 \label{fig:EPISCOPE_ac_MH_by_seqnames}
@@
 All samples were sequenced with paired-end reads.}
-\label{fig:Ziller_merged_ac_MH_by_seqnames}
+\label{fig:Ziller_ac_MH_by_seqnames}
 \end{figure}
@@
 Retreating to a genome-level analysis suggests that within-fragment co-methylation is active over at least a few hundred basepairs in distance.
+
+In many of the $NIL \geq 0$ plots there are samples with autosomal $\widehat{\theta}_{MH}$ that are noticeably separated from the main cloud of autosomal estimates.
+For example, this can be readily observed in the _Colon_Primary_Normal_, _Colon_Primary_Tumour_ samples from the _Ziller_ dataset (Figure \ref{fig:Ziller_ac_MH_by_seqnames}), but it is also apparent in other samples from the _Ziller_ dataset and the _EPISCOPE_ dataset (Figure \ref{fig:EPISCOPE_ac_MH_by_seqnames}).
+These plots do not distinguish estimates from different autosomes.
+Further analysis, however, reveals that these outliers almost all come from a single chromosome, chromosome 21.
+
+Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames_chr21} (_EPISCOPE_), \ref{fig:Lister_ac_MH_by_seqnames_chr21} (_Lister_) and \ref{fig:Ziller_ac_MH_by_seqnames_chr21} (_Ziller_) show the exact same plots as before, but with the chromosome 21 data highlighted by a black line.
+The chromosome 21 data stand out remarkably from the rest of the autosomal data for all samples from the _EPISCOPE_ dataset and several samples from the _Ziller_ dataset.
+Just as remarkable, however, is that this phenomenon is not observed in the _Lister_ dataset nor for any of the frontal cortex samples form the _Ziller_ dataset[^mice_autosomes].
+At this time I have no explanation for this result.
+I cannot rule out it being a technical artefact, although preliminary analyses suggest this is not the case (data not shown), and I do not have a hypothesis for a potential biological cause of this phenomenon.
+
+[^mice_autosomes]: This phenomenon cannot occur in the _Seisenberger_ dataset since these samples are mice, and mice only have 19 autosomes.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames_chr21.pdf}
+\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+The chromosome 21 data are shown by the black line.
+All samples were sequenced with paired-end reads.
+There is no Y chromosome or mtDNA data for \emph{E18BUF} due to a coding error that meant these chromosomes weren't processed by \texttt{methtuple} for \texttt{2ac} tuples.
+This omission does not affect the other results.}
+\label{fig:EPISCOPE_ac_MH_by_seqnames_chr21}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames_chr21.pdf}
+\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+The chromosome 21 data are shown by the black line.
+Only the \emph{ADS}, \emph{ADS-adipose}, \emph{ADS-iPSC} and \emph{H9\_Laurent} samples were sequenced with paired-end reads.}
+\label{fig:Lister_ac_MH_by_seqnames_chr21}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames_chr21.pdf}
+\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs.
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.
+The chromosome 21 data are shown by the black line.
+All samples were sequenced with paired-end reads.}
+\label{fig:Ziller_ac_MH_by_seqnames_chr21}
+\end{figure}
 
@@
 All samples were sequenced with paired-end reads.}
-\label{fig:Ziller_merged_MH_by_CGI}
-\end{figure}
-
-To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of pairs with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}).
+\label{fig:Ziller_MH_by_CGI}
+\end{figure}
+
+To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of pairs with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_ac_MH_by_CGI}).
 We now see a fairly clear separation of the genome-wide estimates of $\theta$, with CpG islands being more co-methylated than the rest of the genome.
@@
 All samples were sequenced with paired-end reads.}
-\label{fig:Ziller_merged_ac_MH_by_CGI}
+\label{fig:Ziller_ac_MH_by_CGI}
 \end{figure}

### 2015-05-04 Rmarkdown/appendix.Rmd
Fix table formatting

@@
 \centering
-\begin{tabular}{@{}lllll@{}}
-\toprule
-Machine name & Processors & Cores per processor & Total number of cores & RAM    \\ \midrule
-unix88       & $4\times$ Intel Xeon X7350 @ 2.93GHz & 4                   & 16                    & 128 GB \\
-unix301      & $4\times$ AMD Opteron 8435 @ 2.6GHz  & 6                   & 24                    & 256 GB \\
-unix302      & $4\times$ AMD Opteron 6174 @ 2.2GHz  & 12                  & 48                    & 512 GB \\
-unix303      & $4\times$ AMD Opteron 6176 @ 2.3GHz  & 12                  & 48                    & 512 GB \\
-unix305      & $4\times$ AMD Opteron 6276 @ 2.3GHz  & 16                  & 64                    &        \\ \bottomrule
-\end{tabular}
 \caption{Bioinformatics Division server specifications.}
 \label{tab:unices}
+\begin{tabulary}{\textwidth}{LLLL}
+\toprule
+Machine name & Processors & Number of cores & RAM    \\ \midrule
+unix88       & $4\times$ Intel Xeon X7350 @ 2.93GHz & 16 & 128 GB \\
+unix301      & $4\times$ AMD Opteron 8435 @ 2.6GHz  & 24 & 256 GB \\
+unix302      & $4\times$ AMD Opteron 6174 @ 2.2GHz  & 48 & 512 GB \\
+unix303      & $4\times$ AMD Opteron 6176 @ 2.3GHz  & 48 & 512 GB \\
+unix305      & $4\times$ AMD Opteron 6276 @ 2.3GHz  & 64 & 512 GB \\ \bottomrule
+\end{tabulary}
 \end{table}

### 2015-05-04 Rmarkdown/datasets.Rmd
Fix table formatting

@@
 In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.
-
-The methylC-seq libraries from the \cite{Lister:2011kg} include some created by the authors and some published by other groups (see Table \ref{tab:Lister_dataset}).
-These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells.
-There are no replicates for any of the \cite{Lister:2011kg} samples.
+These samples are detailed in Table \ref{tab:Lister_dataset_pt1}.
 
@@
 \centering
-\caption[Description of \emph{Lister} dataset]{Samples in \emph{Lister} dataset.
+\caption[Description of \emph{Lister} dataset part 1]{Samples in \emph{Lister} dataset from \cite{Lister:2009hy}.
 The reported read lengths are prior to any trimming of the reads.
-Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
-\label{tab:Lister_dataset}
-\begin{tabulary}{\textwidth}{LLLLLL}
+Abbreviations: \emph{ESC} = embryonic stem cell; \emph{SE} = single-end sequencing.}
+\label{tab:Lister_dataset_pt1}
+\begin{tabulary}{\textwidth}{LLLLL}
 \toprule
-Sample              & Tissue type             & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
-ADS                 & Adipose                 & PE         & $75$ bp     & $23\times$    & \citet{Lister:2011kg}       \\
-ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp     & $24\times$    & \citet{Lister:2011kg}       \\
-ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp     & $26\times$    & \citet{Lister:2011kg}       \\
-FF                  & Foreskin fibroblasts    & SE         & $85$ bp     & $16\times$    & \citet{Lister:2011kg}       \\
-FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp     & $10\times$    & \citet{Lister:2011kg}       \\
-FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
-FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp     & $8\times$     & \citet{Lister:2011kg}       \\
-FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp     & $17\times$    & \citet{Lister:2011kg}       \\
-IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
-IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
-IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
-H1\_r1              & ESC                     & SE         & $85$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
-H1\_r2              & ESC                     & SE         & $85$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
-H1+BMP4             & IVD from H1             & SE         & $85$ bp     & $33\times$    & \citet{Lister:2011kg}       \\
-H9                  & ESC                     & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
-H9\_Laurent         & ESC                     & PE         & $75$ bp     & $8\times$     & \citet{Laurent:2010iw}      \\
-HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
+Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
+IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp    & $14\times$ \\
+IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp    & $15\times$ \\
+H1\_r1              & ESC                     & SE         & $85$ bp    & $15\times$ \\
+H1\_r2              & ESC                     & SE         & $85$ bp    & $14\times$ \\
+\bottomrule
 \end{tabulary}
 \end{table}
+
+The methylC-seq libraries from the \cite{Lister:2011kg} include some created by the authors and some published by other groups.
+These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells.
+There are no replicates for any of the \cite{Lister:2011kg} samples.
+These samples are detailed in Table \ref{tab:Lister_dataset_pt2}.
+
+\begin{table}[!htb]
+\centering
+\caption[Description of \emph{Lister} dataset part 2]{Samples in \emph{Lister} dataset from \cite{Lister:2011kg}.
+The reported read lengths are prior to any trimming of the reads.
+Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{SE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
+\label{tab:Lister_dataset_pt2}
+\begin{tabulary}{\textwidth}{LLLLL}
+\toprule
+Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
+ADS                 & Adipose                 & PE         & $75$ bp    & $23\times$ \\
+ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp    & $24\times$ \\
+ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp    & $26\times$ \\
+FF                  & Foreskin fibroblasts    & SE         & $85$ bp    & $16\times$ \\
+FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp    & $10\times$ \\
+FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp    & $9\times$  \\
+FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp    & $8\times$  \\
+FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp    & $17\times$ \\
+IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp    & $9\times$ \\
+H1+BMP4             & IVD from H1             & SE         & $85$ bp    & $33\times$ \\
+H9                  & ESC                     & SE         & $85$ bp    & $9\times$ \\
+H9\_Laurent         & ESC                     & PE         & $75$ bp    & $8\times$ \\
+HSF1                & ESC                     & SE         & $47$ bp    & $5\times$ \\
+\bottomrule
+\end{tabulary}
+\end{table}
+
+\FloatBarrier
 

### 2015-05-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Fix table formatting

@@
 The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples.
-The final column is a measure of how sparse the data are: a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
+The final column is a measure of how sparse the data are: a $0$ value means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
 \label{tab:methpat_sparsity}

### 2015-05-04 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Fix typesetting of --all-combinations

@@
 \centering
+\caption[Software for calling methylation patterns at m-tuples]{Other software for methylation calling at m-tuples and their limitations.
+Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
+\label{tab:other_m-tuple_methylation_callers}
 \begin{tabulary}{\textwidth}{LLLL}
@@
 \end{tabulary}
-\caption[Software for calling methylation patterns at m-tuples]{Other software for methylation calling at m-tuples and their limitations.
-Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
-\label{tab:other_m-tuple_methylation_callers}
 \end{table}
@@
 \includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
-\caption[\texttt{methtuple} 3-tuples with \texttt{--all-combinations}]{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
+\caption[\texttt{methtuple} 3-tuples with \texttt{{-}{-}all-combinations}]{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{{-}{-}all-combinations} flag is set.}
 \label{fig:methtuple_example4}
@@
 \caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes.
-The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.}
+The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set.}
 \label{fig:methtuple_runtime}
@@
 These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
-The suffix `ac' means that the option \texttt{--all-combinations} was set.}
+The suffix `ac' means that the option \texttt{{-}{-}all-combinations} was set.}
 \label{fig:methtuple_memory}
 \end{figure}
+
+\FloatBarrier
 

### 2015-05-04 Rmarkdown/wgbs_statistical_framework.Rmd
Add missing Ziller strand correlations data

@@
 
-There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the smaller strand-correlations at these lower cutoffs.
-Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8$ to $0.9$, with some notable exceptions.
-
-Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5$ to $0.7$.
-The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$).
-This suggests that caution may be warranted in combining CpG methylation levels across strands for embryonic stem cell samples.
-
-All three _Seisenberger_ samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell, _J1\_1_.
-However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.
-
-Overall, with the exception of embryonic stem cells, it seems that most samples have highly correlated strand-specific CpG $\beta$-values, which means that these data can safely be combined across strands.
-However, it remains a good idea to first check this assumption prior to combining data across strands.
-
 ```{r EPISCOPE_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
@@
 ```{r Ziller_merged_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-x <- read.table("../data/strand_correlation/Ziller_merged_strand_correlations.txt",
+x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
 stringsAsFactors = FALSE)
@@
 ```
+
+There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the smaller strand-correlations at these lower cutoffs.
+Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8$ to $0.9$, with some notable exceptions.
+
+Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5$ to $0.7$.
+The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$).
+This suggests that caution may be warranted in combining CpG methylation levels across strands for embryonic stem cell samples.
+
+All three _Seisenberger_ samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell, _J1\_1_.
+However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.
+
+Overall, with the exception of embryonic stem cells, it seems that most samples have highly correlated strand-specific CpG $\beta$-values, which means that these data can safely be combined across strands.
+However, it remains a good idea to first check this assumption prior to combining data across strands.
 

### 2015-05-04 Rmarkdown/appendix.Rmd
Tick off a bunch of TODOs

@@
 - `GenomicTuples` (`v1.2.1`)
-- `MethylationTuples` (`v0.3.0.9007`)
-- `methsim` (`v0.5.0.9013`)
+- `MethylationTuples` (`v0.3.0.9007`, commit `4e127d2`)
+- `methsim` (`v0.5.0.9013`, commit `9162c8b`)
 
@@
 x <- x[x[, 2] != "*", ]
-x <- x[, -2]
-xt <- xtable(x, caption = c("R packages used in thesis as reported by \texttt{devtools::session\\_info()}", "R packages used in thesis"),
+x <- x[, -c(2, 4)]
+# Drop MethylationTuples and methsim which are reported elsewhere
+x <- x[!x$package %in% c("methsim", "MethylationTuples", "GenomicTUples"), ]
+xt <- xtable(x, caption = c("R packages used in thesis as reported by \\texttt{devtools::session\\_info()}", "R packages used in thesis"),
              label = "tab:R_packages")
-align(xt) <- rep("c", ncol(x) + 1)
+align(xt) <- rep("l", 4)
 print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable",

### 2015-05-04 Rmarkdown/methsim.Rmd
Tick off a bunch of TODOs

@@
 CH) loci have an average $\beta$-value of `-CG` (resp.
-`-CG`) regardless of the genomic context, which we know to be incompatible with real data.
+`-CH`) regardless of the genomic context, which we know to be incompatible with real data.
 Furthermore, the methylation state of each methylation locus is independent, which is clearly inconsistent with the strong co-methylation observed in real data.
@@
 A `SimulatedBS` object contains all simulated reads and is generally a large object (on the order of $10$ GB).
-By contrast, the `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat], is much smaller (on the order of $100$ MB) but does not contain the full information of the simulation since read-level data are lost.
+By contrast, the `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat], is much smaller (on the order of $500$ MB) but does not contain the full information of the simulation since read-level data are lost.
 `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.
@@
 
-\subsubsection{Simulating differential methylation}\label{sec:sim_diff_meth}
+\subsection{Simulating differential methylation}\label{sec:sim_diff_meth}
 
@@
 
-\subsubsection{Performance}
+\subsection{Performance}
 
-While `methsim` is an R package, much of it is written in C++, using the Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures.
+While `methsim` is an R package, much of it is written in \CC, using the Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures.
 Furthermore, several steps of the simulation can be run in parallel for each chromosome.
@@
 Most notably, and frustratingly, I do not yet have an adequate model for the correlations of the true methylation levels, $\mathbf{B}$.
-The model of \citet{Lacey:2013iy} is promising, although may be computationally infeasible for whole-genome data given the timings reported in the original publication.
+The model of \citet{Lacey:2013iy} is promising, although may be computationally infeasible for whole-genome data given the timings reported in the original publication ($13$ seconds for $5,000$ CpGs increasing to $80$ seconds for $10,000$ CpGs).
 The model used by `WGBSsuite` \citet{Rackham:2015bv} is similar to m2, so I suspect it will produce similar results, but this requires further investigation.

### 2015-05-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Tick off a bunch of TODOs

@@
 `MethPipe` is perhaps the best documented and potentially extensible software for analysing methylation patterns at m-tuples.
-`MethPipe` is mostly written in C++ and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data.
+`MethPipe` is mostly written in \CC and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data.
 As such, it does not feature tools that are particularly amenable to interactive or exploratory analyses.
@@
 R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code.
-To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
+To improve the performance of key functionality, parts of `MethylationTuples` are written in \CC, making use of the `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.
 

### 2015-05-04 Rmarkdown/wgbs_statistical_framework.Rmd
Tick off a bunch of TODOs

@@
 colnames(x) <- 1:10
+x <- x[, 1:5]
 xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{EPISCOPE}: Correlations of $\\beta$-values across strands"), label = "tab:EPISCOPE_strand_correlations")
-align(xt) <- rep("c", ncol(x) + 1)
+align(xt) <- c("r", rep("c", ncol(x)))
 digits(xt) <- 2
@@
 colnames(x) <- 1:10
+x <- x[, 1:5]
 xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Lister}: Correlations of $\\beta$-values across strands"), label = "tab:Lister_strand_correlations")
-align(xt) <- rep("c", ncol(x) + 1)
+align(xt) <- c("r", rep("c", ncol(x)))
 digits(xt) <- 2
@@
 colnames(x) <- 1:10
+x <- x[, 1:5]
 xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Seisenberger}: Correlations of $\\beta$-values across strands"), label = "tab:Seisenberger_strand_correlations")
-align(xt) <- rep("c", ncol(x) + 1)
+align(xt) <- c("r", rep("c", ncol(x)))
 digits(xt) <- 2
@@
 ```{r Ziller_merged_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
+x <- read.table("../data/strand_correlation/Ziller_merged_strand_correlations.txt",
 stringsAsFactors = FALSE)
 colnames(x) <- 1:10
+x <- x[, 1:5]
 xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset.
 Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Ziller}: Correlations of $\\beta$-values across strands"), label = "tab:Ziller_merged_strand_correlations")
-align(xt) <- rep("c", ncol(x) + 1)
+align(xt) <- c("r", rep("c", ncol(x)))
 digits(xt) <- 2

### 2015-05-05 Rmarkdown/comethylation.Rmd
Fix table formatting
Update correlation of beta-value plots to include loess fit

@@
 These analyses are performed separately for each strand and also for strand-collapsed $\beta$-values.
+The plots of these analyses show the raw estimates as semi-transparent points and are overlaid with a loess fit to these points.
 
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
-\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
+\caption[\emph{EPISCOPE}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
-\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
+\caption[\emph{Lister}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Lister_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
-\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
+\caption[\emph{Seisenberger}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
-\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
+\caption[\emph{Ziller}: Correlations of strand-specific $NIL = 0$ pairs of $\beta$-values]{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Ziller_spearman_beta_cors_min_cov_5}
@@
 \caption[\emph{EPISCOPE}: Pearson vs.
-Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \caption[\emph{Lister}: Pearson vs.
-Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \caption[\emph{Seisenberger}: Pearson vs.
-Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 \caption[\emph{Ziller}: Pearson vs.
-Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
+Spearman correlation]{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} show the Spearman correlations of strand-collapsed $\beta$-values for $NIL = 0$ pairs of CpGs.
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed} show the Spearman correlations of strand-collapsed $\beta$-values for $NIL = 0$ pairs of CpGs.
 Two things immediately stand out.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{EPISCOPE}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
-\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.}
-\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{Seisenberger}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
-\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{Ziller}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
-\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\FloatBarrier
-
-Figures ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are genome-wide plots; all pairs of CpGs with a given $IPD$ are treated identically.
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{EPISCOPE}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{Seisenberger}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{Ziller}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed}
+\end{figure}
+
+\FloatBarrier
+
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed} are genome-wide plots; all pairs of CpGs with a given $IPD$ are treated identically.
 However, there are good reasons to expect that the correlations of $\beta$-values, like the $\beta$-values themselves, might be influenced by their genomic context.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{EPISCOPE}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.}
+\caption[\emph{EPISCOPE}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{EPISCOPE} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
+\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Seisenberger}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.}
+\caption[\emph{Seisenberger}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Seisenberger} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Ziller}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.}
+\caption[\emph{Ziller}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Ziller} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
@@
 
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI} are the plots of CpG $\beta$-value correlations against $IPD$ for $NIL \geq 0$ pairs.
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed} and \ref{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed} are the plots of CpG $\beta$-value correlations against $IPD$ for $NIL \geq 0$ pairs.
 In contrast to the $NIL = 0$ results, the correlations here are relatively stable, at least for $IPD = 2, \ldots, 2000$[^ipd_choice].
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{EPISCOPE}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{EPISCOPE} dataset.}
-\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{Lister}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Lister} dataset.}
-\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{Seisenberger}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Seisenberger} dataset.}
-\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
-\end{figure}
-
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\emph{Ziller}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Ziller} dataset.}
-\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{EPISCOPE}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{EPISCOPE} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{Lister}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Lister} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{Seisenberger}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Seisenberger} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed.pdf}
+\caption[\emph{Ziller}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values]{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Ziller} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{EPISCOPE}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{EPISCOPE} dataset.}
+\caption[\emph{EPISCOPE}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{EPISCOPE} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Lister}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Lister} dataset.}
+\caption[\emph{Lister}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Lister} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Seisenberger}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Seisenberger} dataset.}
+\caption[\emph{Seisenberger}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Seisenberger} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Ziller}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Ziller} dataset.}
+\caption[\emph{Ziller}: Correlations of $NIL \geq 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the \emph{Ziller} dataset.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
@@
 \label{tab:2x2}
-\begin{tabular}{@{}cccc@{}}
-\cmidrule(l){2-4}
- & \multicolumn{3}{c}{Column} \\
-\multirow{3}{*}{Row} & & & \\
- & $n_{11}$       & $n_{12}$       & $n_{1 +}$ \\
- & $n_{21}$       & $n_{22}$       & $n_{2 +}$ \\
- & $n_{+ 1}$  & $n_{+ 2}$  & $n_{+ +} = n$ \\ \cmidrule(l){2-4}
+\begin{tabular}{@{}ccccl@{}}
+\cmidrule(l){2-5}
+                   &       & \multicolumn{2}{c}{Y} &              \\ \cmidrule(lr){3-4}
+                   &       & 1         & 2         & Total        \\ \cmidrule(l){2-5}
+\multirow{2}{*}{X} & 1     & $n_{11}$  & $n_{12}$  & $n_{1+}$     \\
+                   & 2     & $n_{21}$  & $n_{22}$  & $n_{2+}$     \\
+                   & Total & $n_{+1}$  & $n_{+m}$  & $n = n_{++}$ \\ \cmidrule(l){2-5}
 \end{tabular}

### 2015-05-05 Rmarkdown/methsim.Rmd
Update correlation of beta-value plots to include loess fit

@@
 
-Figure \ref{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the dramatic consequences of this modelling decision.
+Figure \ref{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed} shows the dramatic consequences of this modelling decision.
 When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
-\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
-\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
+\includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
+\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values]{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
+\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
+\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.
+The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
 \label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}

### 2015-05-05 Rmarkdown/methsim.Rmd
Fix notation

@@
 That is, we want to specify $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$.
-For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}} = B_{i, j^{0}}$; differentially methylated loci are simulated by setting $B_{i, j_{k}} \neq B_{i, j^{k'}}$ for some $k \neq k'$.
+For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}} = B_{i, j_{0}}$; differentially methylated loci are simulated by setting $B_{i, j_{k}} \neq B_{i, j_{k'}}$ for some $k \neq k'$.
 

### 2015-05-05 Rmarkdown/appendix.Rmd
Knocking off TODOs

@@
 
-\section{Chapter \ref{chap:co-methylation_review}}\label{app:co-methylation_review}
-
-\subsection{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}
+\section{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}
 
@@
 
-\subsection{Estimating the number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}
-
-I want to estimate the number of CpGs used by \citet{Lacey:2013iy} to estimate the correlation of methylation levels.
-They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, _MTCTL2_.
-Unfortunately, there is no URL provided to these data.
-\citet{Lacey:2013iy} cite their data as having come from \citet{Tsumagari:2013gn}, which includes a sample "Normal Human Skeletal Muscle Myoblasts", _HSMM_ ([http://genome-mirror.duhs.duke.edu/ENCODE/cellTypes.html](http://genome-mirror.duhs.duke.edu/ENCODE/cellTypes.html), [http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/](http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/)).
-I believe this to be the same cell line.
-In any case, for my purposes I simply needed some comparable RRBS data, not necessarily the exact same sample as  _MTCTL2_.
-
-There are approximately $60,000$ CpGs on chromosome 11 in these comparable RRBS data. The following R code shows how I computed this using three RRBS replicates of the _HSMM_ sample.
-
-Firstly, I download the data and unzip it:
-```{r, cache = FALSE, message = FALSE}
-library(R.utils)
-library(data.table)
-dl_dir <- "../data/appendix/"
-url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
-files <- paste0('wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep',
-                1:3, '.bed.gz')
-dests <- paste0(dl_dir, files)
-x <- mapply(function(f, d) {
-  if (!file.exists(gsub(".gz", "", paste0(dl_dir, f)))) {
-    download.file(url = paste0(url, f), d)
-    gunzip(filename = d, overwrite = TRUE)
-    }
-  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t',
-        verbose = FALSE)
-  }, files, dests, SIMPLIFY = FALSE)
-```
-
-Then I count the number of CpGs on chromosome 11 in each replicate:
-```{r}
-y <- rbindlist(x)
-y[V1 == "chr11", .N, by = list(V4)]
-```
-
 \section{Computing details}\label{app:computing_details}

### 2015-05-05 Rmarkdown/comethylation_review.Rmd
Fix table formatting
Knocking off TODOs

@@
 
-[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (see Appendix \ref{app:lacey}).
+[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable myoblast RRBS libraries (samples \texttt{wgEncode\-Haib\-Methyl\-Rrbs\-Hsmmtubefshd\-DukeSitesRep1}, \texttt{wgEncode\-Haib\-Methyl\-Rrbs\-Hsmmtubefshd\-DukeSitesRep2}, and \texttt{wgEncode\-Haib\-Methyl\-Rrbs\-Hsmmtubefshd\-DukeSitesRep3} available from \url{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/}).
 
@@
 
-\begin{table}[h]
+\begin{table}[!htb]
 \centering
@@
 \label{tab:landan}
-\begin{tabular}{@{}cccc@{}}
-\cmidrule(l){2-4}
- & \multicolumn{3}{c}{Second CpG} \\
- \cmidrule(l){2-4}
-\multirow{3}{*}{First CpG} & & & \\
- & $n_{uu}$       & $n_{um}$       & $n_{u+}$ \\
- & $n_{mu}$       & $n_{mm}$       & $n_{m+}$ \\ \cmidrule(l){2-4}
- & $n_{+u}$  & $\cdot n_{+m}$  & $n_{++} = n$      \\
+\begin{tabular}{@{}ccccl@{}}
+\cmidrule(l){2-5}
+                           &              & \multicolumn{2}{c}{Second CpG} &              \\ \cmidrule(lr){3-4}
+                           &              & Unmethylated    & Methylated   & Total        \\ \cmidrule(l){2-5}
+\multirow{2}{*}{First CpG} & Unmethylated & $n_{uu}$        & $n_{um}$     & $n_{u+}$     \\
+                           & Methylated   & $n_{mu}$        & $n_{mm}$   & $n_{m+}$     \\
+                           & Total        & $n_{+u}$        & $n_{+m}$     & $n = n_{++}$ \\ \cmidrule(l){2-5}
 \end{tabular}
@@
 Then, compute $R = cor(\bm{\beta})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci.
-This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data.
+This is what the R package, `coMET` \citep{Martin:2015bs}, does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data.
 As far as I am aware, `coMET` is the only software specifically designed to estimate between-sample co-methylation and integrate this with other genomic annotations.

### 2015-05-05 Rmarkdown/comethylation.Rmd
Embiggen simulation study plot labels
Edits suggested by Terry

@@
 Two pairs of CpGs with an identical $IPD$ may come from two regions, even when using the $NIL = 0$ strategy.
-These regions mayhave very different methylation dynamics.
+These regions may have very different methylation dynamics.
 In other words, the vector of $\beta$-values for a given sample is highly non-stationary; the correlation between two loci depends on more than just the $IPD$.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
-\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.
+\caption[\emph{Lister}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the \emph{Lister} dataset.
 The raw estimates of the correlations are shown as semi-transparent points and are overlaid with a loess fit to these points (\texttt{span} = 0.1).}
@@
 We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables.
-We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determinine which is most appropriate for a study of within-fragment co-methylation.
+We conclude with a simulation study highlighting the performance of several different estimators of dependence in sparse $2 \times 2 \times K$ contingency tables and determine which is most appropriate for a study of within-fragment co-methylation.
 
@@
 
+\FloatBarrier
+
 In these remaining simulations, all parameters are as for Simulation 3 except that the true log odds ratio of each table, $\theta_{k}$, comes from the following distributions:
@@
 
-One explanation of this result, at least for the iPSCs, is that it is due to 'reseting' of the methylome during embryonic development induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.
+One explanation of this result, at least for the iPSCs, is that it is due to 'resetting' of the methylome during embryonic development induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.
 This explanation does not immediately carry over the the embryonic stem cell lines, although there is extensive reprogramming of DNA methylation during embryogenesis, which may be relevant \citep{Seisenberger:2013jq}.
@@
 A second observation on the results for the _Lister_ dataset is that genome-level estimates of $\theta$ decay particularly rapidly in the somatic cell lines (_ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_).
-This is recapituled in the _IMR90\_cell\_line_ sample from the _Ziller_ dataset.
+This is recapitulated in the _IMR90\_cell\_line_ sample from the _Ziller_ dataset.
 

### 2015-05-05 Rmarkdown/methsim.Rmd
Knocking off TODOs
Add y-axis label to boxplots of beta by region

@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
-\caption[\emph{Seisenberger}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
+\caption[\emph{Seisenberger}: $\beta$-values by region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.
+\texttt{MethylSeekR} was unable to partition the \emph{E16.5\_male\_1} methylome due to its unusual $\beta$-value distribution, hence no data is shown for this sample.}
 \label{fig:Seisenberger_beta_by_pm_boxplots}
@@
 Furthermore, several steps of the simulation can be run in parallel for each chromosome.
-Parallelism is implemented via the `BiocParallel` Bioconductor package \citep{BiocParallel}.
+Parallelism is implemented via the `BiocParallel` Bioconductor package (\url{http://bioconductor.org/packages/BiocParallel}).
 

### 2015-05-05 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Polish methtuple performance plots

@@
 
-I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for $40$ whole-genome bisulfite-sequencing samples.
+I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for more than $40$ whole-genome bisulfite-sequencing samples.
 Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets.
@@
 \caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes.
-The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set.}
+The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set.
+Note that the total number of samples is $49$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
 \label{fig:methtuple_runtime}
@@
 These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}).
-The suffix `ac' means that the option \texttt{{-}{-}all-combinations} was set.}
+The suffix `ac' means that the option \texttt{{-}{-}all-combinations} was set.
+Note that the total number of samples is $49$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
 \label{fig:methtuple_memory}

### 2015-05-06 latex/phd_thesis.tex
Tick off a bunch of TODOs
Add Dineika to acknowledgements

@@
             pdfcreator={pdflatex, or other tool}]{hyperref}
+
+% Typesetting C++ (http://tex.stackexchange.com/a/4304)
+\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}
 
@@
 In particular, thank you to my wonderful friends and colleagues in the Bioinformatics Division for your support and advice.
-A special thanks to Keith Satterley and the ITS department who both tolerated and helped fix my computational blunders.
+A special thanks to Keith Satterley and the IT department who both tolerated and helped fix my computational blunders.
 
@@
 
-I owe an enormous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proof readers, Belinda Phipson, Saskia Freytag and Goknur Giner, who have greatly improved this thesis.
+I owe an enormous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proof readers, Belinda Phipson, Saskia Freytag, Goknur Giner, and Dineika Chandrananda who have greatly improved this thesis.
 

### 2015-05-06 Rmarkdown/appendix.Rmd
Update R packages list

@@
 x <- x[!x$package %in% c("methsim", "MethylationTuples", "GenomicTUples"), ]
-xt <- xtable(x, caption = c("R packages used in thesis as reported by \\texttt{devtools::session\\_info()}", "R packages used in thesis"),
+xt <- xtable(x, caption = c("R packages used in thesis (as reported by \\texttt{devtools::session\\_info()})", "R packages used in thesis"),
              label = "tab:R_packages")

### 2015-05-06 latex/phd_thesis.tex
Remove proofreaders test

@@
 
-I owe an enormous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proof readers, Belinda Phipson, Saskia Freytag, Goknur Giner, and Dineika Chandrananda who have greatly improved this thesis.
+I owe an enormous debt of gratitude to my thesis writing buddies, Hannah Vanyai and Darcy Moore, who kept me going during tough times and to my thesis proofreaders, Belinda Phipson, Saskia Freytag, Goknur Giner, and Dineika Chandrananda who have greatly improved this thesis.
 

### 2015-05-06 Rmarkdown/comethylation.Rmd
Reword convergence discussion

@@
 
-To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1a, Figure \ref{fig:Simulation_1a}), and $K = 1000$ (Simulation 1b, Figure \ref{fig:Simulation_1b}).
-Under this simulation, we expect both $\widehat{\theta}_{U}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
-This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is small.
-By contrast, the estimates of $\theta$ from the individual $2 \times 2$ tables, $\widehat{\theta}_{0.5}$, are highly biased, even when $K = 1000$ in Simulate 1b (Figure \ref{fig:Simulation_1b}).
-In both Simulation 1a and 1b (Figures \ref{fig:Simulation_1a} and \ref{fig:Simulation_1b}) the histograms of $\widehat{\theta}_{0.5}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).
+To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}$ and $\pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1a, Figure \ref{fig:Simulation_1a}), and $K = 1000$ (Simulation 1b, Figure \ref{fig:Simulation_1b}).
+In this setting we expect both $\widehat{\theta}_{U}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$.
+The rate of convergence, however, also depends on the the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$.
+We see that when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$), that the approximation is poor when $K = 100$ (Simulation 1a, Figure \ref{fig:Simulation_1a}) but is reasonable when $K = 1000$ (Simulation 1b, Figure \ref{fig:Simulation_1b}).
+Regardless of whether $K = 100$ or $K = 1000$, the histograms of $\widehat{\theta}_{0.5}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).
 

### 2015-05-06 Rmarkdown/appendix.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 
-\begin{equation*}
-Pr(X_{1} = x_{1}, \ldots, X_{n} = x_{n}) = Pr(X_{1} = x_{1}) Pr(X_{2} = x_{2} | X_{2} = x_{2}) \cdots Pr(X_{n} = x_{n} | X_{n - 1} = x_{n - 1})
-\end{equation*}
+\begin{align*}
+ &  Pr(X_{1} = x_{1}, \ldots, X_{n} = x_{n}) \\
+=& Pr(X_{1} = x_{1}) Pr(X_{2} = x_{2} | X_{2} = x_{2}) \cdots Pr(X_{n} = x_{n} | X_{n - 1} = x_{n - 1})
+\end{align*}
 
@@
 Pr(Z_{h, i} = Z_{h', i'}) &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
-                  &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(X_{0} = 0) +
-                     Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(X_{0} = 1) \\
+                  &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) \\
+                  &+ Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
                   &= \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) +
@@
 The scripts used to prepare the results for each chapter are available from \url{https://github.com/PeteHaitch/phd_thesis_analyses}.
-Further details of software used are available in Appendix \ref{app:software_details}.
 
-I developed several pieces of software during my PhD. These are listed in below, along with the version used in for analyses in my thesis:
+I developed several pieces of software during my PhD. These are listed below, along with the version used in for analyses in my thesis:
 

### 2015-05-06 Rmarkdown/comethylation.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 This chapter describes two measures of within-sample co-methylation, \emph{within-fragment co-methylation} and \emph{correlation of $\beta$-values}.
-The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.
+In order to analyse within-fragment co-methylation, we first study methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.
 
@@
 Before computing correlations and trying to interpret these, it is a good idea to look at the data from which these correlations are computed.
-Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200, 2000$ with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample from the _Lister_ dataset.
+Figure \ref{fig:ADS_beta_scatterplots} shows kernel density smoothed scatterplots for pairs of $\beta$-values separated by $IPD = 2, 20, 200$ or $2000$ bp with $NIL = 0$ or $NIL \geq 0$ from the _ADS_ sample from the _Lister_ dataset.
 It highlights two major problems.
@@
 
-To address this problem, we might try smoothing the raw correlations, for example, by using LOESS \citep{Cleveland:1979fu}, and then looking for periodicities in the smoothed values.
-However, this is only helpful when the data are corrupted by a small amount of noise, certainly not the extreme noise we see for $IPD > 1000$ in most samples.
-Furthermore, smoothing introduces problems of its own, such as choosing the smoothing bandwidth in such a way as to not introduce or destroy real periodicities.
-Even if we can automatically choose these smoothing parameters in an unbiased manner, we are still stuck with only observing a handful of periods of this size from $NIL = 0$ pairs.
-So while these periodicities are 'obvious' in the plots, they are rather more difficult to detect without visualising the raw data.
-
 \begin{figure}[!htb]
@@
 Stratifying by CpG island status also reveals that the correlation of the methylation levels of CpGs inside CpG islands is less influenced by $IPD$.
-In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD$.
+In fact, for some samples it appears that the most correlated $NIL = 0$ pairs inside CpG islands are not those with the minimum $IPD = 2$, but those with a slightly larger $IPD$.
 From these plots we can now see that the apparent sharp decline in correlations over the first $50$ bp from the genome-wide data is really driven by a shift from pairs that are frequently inside CpG islands to pairs that are mostly outside of islands.
@@
 
-[^hepg2]: The minor exception is the _HepG2\_cell\_line_ where the CpG island and non-island lines intersect by $IPD = 1500$, but we have already seen that this sample has an unusual methylome compared to the other samples.
+[^hepg2]: The minor exception is the _HepG2\_cell\_line_ where the CpG island and non-island lines intersect by $IPD = 1500$, but we have already seen that this sample has an unusual methylome compared to the other samples, particularly with respect to CpG islands (see Figure \ref{fig:Ziller_merged_plotBetaFreqPoly_CGI}).
 
@@
 Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}.
-The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods.
+The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$, is solved using iterative methods.
 The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep[pp.
@@
 
-We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the odds ratio $\psi$ \citep[see][]{Breslow:1981ta}.
+We can again compute an unconditional maximum likelihood estimator, a '0.5-adjusted' unconditional maximum likelihood estimator, and a conditional maximum likelihood estimator of the odds ratio $\psi$ \citep[see][]{Breslow:1981ta}.
 
@@
 It has been noted, however, that statistical tests of homogeneity versus heterogeneity typically have low power and are only able to detect large departures form homogeneity \cite[e.g.,][]{Hauck:1989vt}.
-Therefore these tests are not always of great practical use.
+These tests are therefore not always of great practical use.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_4.pdf}
 \caption[Simulation 3]{Results of simulation 3 ($K = 100$).
@@
 The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_ac_MH_by_seqnames}.
-In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_; _ADS_ is a female cell line), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_; _IMR90_ is a female cell line).
+In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_; _ADS_ is a female cell line), the _H9_-based samples (_H9_ and _H9\_Laurent_; sex of cell line not reported in \citet{Lister:2011kg}), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_; _IMR90_ is a female cell line).
 
@@
 
-[^cpg_density1]: The overall density of methylation loci is likely important, which of related to $IPD$s.
+[^cpg_density1]: The overall density of methylation loci is likely important, which is of course related to $IPD$s.
 
@@
 
-[^imr90_ac]: Differences in the $NIL \geq 0$ results between the _IMR90\_r1, _IMR90\_r2_ and _IMR90\_cell\_line_ samples are due to the _IMR90\_cell\_line_ being sequenced with paired-end reads and therefore having many more $NIL > 0$ pairs to analyse.
+[^imr90_ac]: Differences in the $NIL \geq 0$ results between the _IMR90\_r1_, _IMR90\_r2_ and _IMR90\_cell\_line_ samples are due to the _IMR90\_cell\_line_ being sequenced with paired-end reads and therefore having many more $NIL > 0$ pairs to analyse.
 
@@
 This result is tantalisingly similar to the periodicity we see in the correlations of $\beta$-values and the approximate spacing of nucleosomes in human DNA (see Section \ref{sec:nucleosome_and_chromatin}).
-However, it is less apparent in other paired-end samples, and this is only likely to be resolved analyses of longer reads that can span multiple nucleosomes.
+However, it is less apparent in other paired-end samples, and this is only likely to be resolved by longer reads that can span multiple nucleosomes.
 
@@
 We have estimated the effect of this co-methylation using two complementary approaches: correlations of $\beta$-values and within-fragment co-methylation.
-Consistent with previous work, our analysis identifies the inter-pair distance between CpGs and the genomic context of these CpGs, in particular, CpG islands, as being important drivers of co-methylation.
+Consistent with previous work, our analysis identifies the intra-pair distance between CpGs and the genomic context of these CpGs, in particular, CpG islands, as being important drivers of co-methylation.
 

### 2015-05-06 Rmarkdown/comethylation_review.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 These include: "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}; "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; the "correlation of two [loci] across many samples" \citep{Akulenko:2013fl}; and the "_vertical_ (i.e.
-progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}.
+progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014hx}.
 
@@
 They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype _Arabidopsis thaliana_, a small flowering plant whose methylome is widely studied.
-They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis].
+They performed an "autocorrelation" analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis].
 The autocorrelation computes, for a set of distances ($0-5,000$ bp), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance.
@@
 
-The autocorrelation plots showed two periodicities, one of approximately $167$ bp and the other of $10$ bp (CHH only).
+The autocorrelation plots showed two periodicities, one of approximately $167$ bp (CpG, CHG and CHH) and the other of $10$ bp (CHH only).
 Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.
@@
 The weakness of the signal is partially attributable to the small number of "methylcytosines" in the CHG and CHH contexts in human samples.
-More importantly, the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs.
+More concerning is that the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs.
 Unlike the $8$ to $10$ bp period, the $3$ bp period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it.
@@
 \citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation".
-\citeauthor{Li:2010fb} conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", and report that their results are consistent with those in \citet{Eckhardt:2006gh}.
+\citeauthor{Li:2010fb} report that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", and conclude that their results are consistent with those in \citet{Eckhardt:2006gh}.
 In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated regions, coding DNA sequences and various repetitive elements.
@@
 
-Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are estimates of the correlation of aggregate methylation levels over the range $0$ to $1000$ bp.
+Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0$ to $1000$), which I believe are estimates of the correlation of aggregate methylation levels over the range $0$ to $1000$ bp.
 Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs (i.e.
 $NIL = 0$ pairs).
-If all pairs were used, $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome.
+If all pairs were used, $n$ would be much larger since there are some 25 millions CpGs in the haploid human genome.
 To emphasise, the correlation computed by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.
@@
 Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$ for the chromosome 22 data from the _FF_ sample.
-We see that it is remarkably similar to that panel (a) of Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum correlation is very small, close to zero.
+We see that it is remarkably similar to that of panel (a) in Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum correlation is very small, close to zero.
 
@@
 Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}).
-Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of bases in the genome than anything to do with their methylation state.
+Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of CpGs in the genome than anything to do with their methylation state.
 
@@
 To do so, \citeauthor{Lacey:2013iy} fit a Gaussian variogram model to the CpG $\beta$-values from chromosome 11 of a normal myotube cell line (_MTCTL2_).
-The empirical and fitted variograms asymptote for CpGs separated by approximately $3000$ bp (Figure 1d of \citet{Lacey:2013iy}; I am unable to reproduce it here due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
+The empirical and fitted variograms asymptote for CpGs separated by approximately $3000$ bp (Figure 1d of \citet{Lacey:2013iy}; I am unable to reproduce the figure here due to rights restrictions) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".
 
@@
 To do so, \citeauthor{Ball:2009kh} took all reads containing multiple CpGs and mapping to positions with intermediate $\beta$-values ($0.2 < \beta < 0.8$) with at least $100\times$ sequencing coverage.
-From these reads they computed the correlation of within-read methylation states.
+From these reads they computed the Pearson correlation of within-read methylation states.
 \citeauthor{Ball:2009kh} found that the distribution of these correlations is centred around $0.5$ with very few values below zero, evidence that methylation levels at CpGs on the same DNA fragment "are generally positively correlated".
@@
 \citet{Lacey:2009by} develop stochastic models of DNA methylation replication during mitosis (cell division) for a single double-stranded DNA molecule in humans.
-The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field.
+The most complex model that \citeauthor{Lacey:2009by} consider, called the _neighbouring sites model_, uses a one-step Markov random field.
 Under this model, the methylation state of one locus, $Z_{h, i}$, depends on the methylation states of the adjacent loci, $Z_{h, i - 1}$ and $Z_{h, i + 1}$.
@@
 
-In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a minuscule fraction of the human genome.
-While the methods and models are quite interesting, the data are so limited that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.
+In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a very small fraction of the human genome.
+While the methods and models are quite interesting, the data are limited to such a small region that it is difficult to know whether these results give an accurate genome-wide picture of within-fragment co-methylation.
 
@@
 The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing (the median sequencing coverage of each CpG in the target is greater than $10,000 \times$) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time.
-\citeauthor{Landan:2012kp} also use these data to estimate co-methylation.
 
@@
 
-The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses.
+The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing the proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses.
 Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data.
@@
 More generally, methods such as `SLIM` \citep{Wang:2011cw} and `comb-p` \citep{Pedersen:2012vl} attempt to account for the correlation amongst test statistics in genomics data.
-They do this by analysing the distribution of P-values rather than the distribution of the test statistics themselves or the underlying data.
-This makes these methods quite general but application-specific methods may be more powerful.
+Rather than analysing the distribution of the test statistics themselves or the underlying data, they do this by analysing the distribution of the resulting P-values.
+This makes these methods quite general, although application-specific methods may be more powerful.
 `comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \citep{Pedersen:2012vl}.
@@
 These problems range from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation.
-Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.
+Moreover, despite numerous papers including an analysis of co-methylation, few of these have includes software to reproduce their method, further complicating the assessment of these methods.
 

### 2015-05-06 Rmarkdown/concluding_remarks.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

--- 2015-04-24 Rmarkdown/concluding_remarks.Rmd
@@
 
-__TODO: Review each chapter__
+This thesis has examined the statistical and computational challenges raised by high-throughput assays of DNA methylation.
+The first chapter introduced DNA methylation and then extensively reviewed different high-throughput assays for measuring it, focusing on the current gold standard assay of whole-genome bisulfite-sequencing.
+
+The second chapter detailed the bioinformatics analysis of whole-genome bisulfite-sequencing data.
+Beginning with the `FASTQ` files created by the DNA sequencers, we reviewed the quality control procedures, read mapping, and post-processing steps that are essential to creating a `BAM` file containing high-quality mapped reads.
+We then described methods for calling methylation events from bisulfite-sequencing data.
+The chapter concluded by introducing m-tuples and the `methtuple` software that I wrote.
+Uniquely, `methtuple` can call methylation patterns at m-tuples from whole-genome bisulfite-sequencing data.
+
+The third chapter described the 40 whole-genome bisulfite-sequencing samples that are analysed throughout the remainder of the thesis.
+
+Chapter \ref{chap:wgbs_statistical_framework} laid out a statistical framework for analysing bisulfite-sequencing data.
+This process illustrated that the commonly used summary of methylation levels, the $\beta$-value, aggregates over many potential sources of variation.
+In light of this, we discussed the interpretation of $\beta$-values and examined the distributions of $\beta$-values in our 40 samples.
+
+Chapter \ref{chap:wgbs_downstream_analyses} examined methods for analysing bisulfite-sequencing data to address key biological questions.
+Most downstream analyses are based on $\beta$-values, such as the identification of differential methylation, and we reviewed these.
+There is, however, a growing interest in analyses based on methylation patterns at m-tuples since these can address novel biological questions.
+A limitation to the development of these methods has been a lack of software.
+The chapter concluded with the development of the `MethylationTuples` R package, which, together with `methtuple`, is designed to facilitate analyses based on methylation patterns at m-tuples.
+
+The sixth and seventh chapters are a comprehensive review and analysis of co-methylation, the spatial dependence of DNA methylation.
+Chapter \ref{chap:co-methylation_review} re-examined previous analyses of co-methylation and found that these have been limited by inadequate data and deficiencies in the statistical methods applied.
+We addressed some of these limitations with an extensive analysis of co-methylation in Chapter \ref{chap:co-methylation}.
+We proposed a novel analysis of co-methylation using methylation patterns from individual DNA fragments, which we call within-fragment co-methylation.
+A simulation study demonstrated that, in general, we cannot estimate this within-fragment co-methylation for individual pairs of CpGs.
+By aggregating pairs of CpG, however, the simulation study showed that we can estimate a summary of this spatial dependence of DNA methylation along individual DNA fragments.
+We applied this method to 40 whole-genome bisulfite-sequencing samples to identify genomic features that influence co-methylation and examined how it varies between different tissues.
+
+To conclude, Chapter \ref{chap:methsim} detailed our efforts to develop the `methsim` software to simulate whole-genome bisulfite-sequencing data.
+Unlike existing software, `methsim` seeks to model co-methylation so as to create more realistic simulated data.
+Such simulated data will be useful in developing and benchmarking methods for the analysis of DNA methylation data.
+While aspects of `methsim` are promising, the overly-strong correlations of $\beta$-values is an obvious inadequacy in the current model.
+We continue to work to improve this and to develop the software more generally.
+
+DNA methylation data have a complex dependence structure, as demonstrated by the analyses presented in this thesis.
+Any analysis must bear in mind the multiple sources of variation in the data and the amount of aggregation required in order to extract meaningful results.
+
+The large size of the data generated by these experiments can make it challenging to develop analysis methods.
+This is particularly true when coupled with a lack of software for extracting, manipulating and summarising the relevant features of the data.
+Much of the effort in this thesis has gone into developing software to extract and manage this information (`methtuple` and `MethylationTuples`) so that new tools can be built on a higher foundation (`methsim`).
+It is my hope that, with continued development, these tools may facilitate other analyses to unravel the complexity of DNA methylation and its many biological roles.

### 2015-05-06 Rmarkdown/datasets.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 
-This chapter briefly documents the 40 whole-genome bisulfite-sequencing samples that I use in my thesis. The `BAM` files for each sample underwent the same basic processing:
+This chapter briefly documents the 40 whole-genome bisulfite-sequencing samples that I use in my thesis.
+The `BAM` files containing the aligned reads for each sample underwent the same basic processing:
 
@@
 I have the data for only three samples from the original publication, detailed in Table \ref{tab:Seisenberger_dataset}.
-The _J1\_1_ sample is from an embryonic stem cell line while both the _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ samples are derived from pools of $10-30$ embryos.
+The _J1\_1_ sample is from an embryonic stem cell line while both the _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ samples are derived from pools of $10$ to $30$ embryos.
 Developmentally, the samples are ordered _J1\_1_ (embryonic stem cell), _E6.5\_epiblast\_1_ (embryonic day 6.5 epiblast) and _E16.5\_male\_1_ (embryonic day 16.5 male progenitor germ cells).
@@
 
-The _Ziller_ are a subset of the data used in \cite{Ziller:2013fs}.
+The _Ziller_ data are a subset of the data used in \cite{Ziller:2013fs}.
 Specifically, I use a convenience sample of 8 whole-genome bisulfite-sequencing libraries.
@@
 
-I have used the CpG island definition from the hidden Markov model proposed by \citet{Wu:2010do} and implemented in the `makeCGI` R package (`v1.2`).
+I have used the CpG island definition from the hidden Markov model proposed by \citet{Wu:2010do} and implemented in the `makeCGI` R package (`v1.2`, \url{http://rafalab.jhsph.edu/CGI/}).
 The predicted CpG islands for the human reference genome (both hg18 and hg19) were downloaded from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/) on 29/10/2014.

### 2015-05-06 Rmarkdown/methsim.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 There is a lack of gold-standard datasets in the field of DNA methylation for the benchmarking and validation of analysis methods.
-Therefore, the development of realistic, cheap and user-friendly software to simulate DNA methylation data will be of benefit.
+Therefore, the development of realistic, cheap, and user-friendly software to simulate DNA methylation data will be of benefit.
 
@@
 
-`Sherman` ([http://www.bioinformatics.babraham.ac.uk/projects/sherman/](http://www.bioinformatics.babraham.ac.uk/projects/sherman/)) is software to simulate bisulfite-sequencing reads as `FASTQ` files, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts.
+`Sherman` is software to simulate bisulfite-sequencing reads as `FASTQ` files, including various 'contaminants', such as SNPs, basecall errors and sequence artefacts (\url{http://www.bioinformatics.babraham.ac.uk/projects/sherman/}).
 The simulated reads are designed for comparing the performance of different alignment strategies.
@@
 When a read contains a CG (resp.
-CH) locus, it is randomly assigned as being converted (unmethylated) with probability `-CG` / 100 (resp.
+CH) locus, it is randomly assigned as being converted to a TG (resp.
+TH) with probability `-CG` / 100 (resp.
 `-CH` / 100).
@@
 The most widely studied downstream analysis problem is that of identifying differential methylation, which is done by comparing summary measures of methylation, such as $\beta$-values, between two or more groups.
-It is therefore generally not necessary to simulate individual reads, but is sufficient to directly simulate these aggregated measures, when simulating data for these studies.
+It is therefore generally sufficient to directly simulate these aggregated measures, rather than simulating reads, for these type of simulation studes.
 
-Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section.
+Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation study.
 Generally, such a simulation method exists to support claims about the performance of the proposed method and is not a major feature of the paper.
@@
 
-Almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.
+Almost all downstream analyses are reference-based (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.
 
@@
 
-Recall that $B_{i, j}$ is the underlying _true_ methylation level at locus $i$ in sample $j$.
+Recall that $B_{i, j}$ is the underlying 'true' methylation level at locus $i$ in sample $j$.
 In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific.
@@
 Under the beta-binomial model, the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j_{k}}, \phi_{i, j_{k}})$ distribution, where $\mu_{i, j_{k}}$ is the mean and $\phi_{i, j_{k}}$ is the dispersion of the beta distribution[^beta_parameterisation].
-Both the means, $\mu_{i, j_{k}}$, and the dispersions, $\phi_{i, j_{k}}$, are group-specific and are allowed to vary across methylation loci ($i$).
+Both the means, $\mu_{i, j_{k}}$, and the dispersions, $\phi_{i, j_{k}}$, are group-specific and are allowed to vary across methylation loci (i.e.
+across $i$).
 The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e.
@@
 As noted by \citet{Feng:2014iq}, the beta distribution is a very flexible distribution with support on $[0, 1]$ and has "long been a natural choice to model binomial proportions", particularly as a conjugate prior, as it is used in the empirical Bayes model of \citet{Feng:2014iq}.
-The beta-binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or overdispersed binomial distributions
+The beta-binomial model can also be viewed from a non-Bayesian perspective as a compound distribution or as an overdispersed binomial distribution.
 
@@
 An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$.
-`WGBSSuite`\citep{Rackham:2015bv} uses a modified form of this approach.
+`WGBSSuite` \citep{Rackham:2015bv} uses a modified form of this approach.
 In `WGBSSuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated").
@@
 
-The sequencing depth at each methylation loci, $d_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.
+The sequencing depth at each methylation loci, $d_{i, j}$, may be sampled from real data \citep{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.
 
@@
 In addition to using a mixture of distributions to capture both the low-coverage and high-coverage modes observed in RRBS sequencing coverage, \citet{Lacey:2013iy} model the correlation of sequencing depth across samples for a given region.
-They do this by using a Normal copula to make the set of sequencing depths a jointly dependent set of random variables.
+They do this by using a Gaussian copula to make the set of sequencing depths a jointly dependent set of random variables.
 While this is undoubtably sophisticated, the effect of correlated versus uncorrelated sequencing depths in a simulation model is not explored in the paper and so the cost-benefit trade-off is unclear.
@@
 Plotted is the difference in $\beta$-value between the two groups ($\Delta \beta$) with associated standard error against the position along the genome (\emph{Position (bp)}).
-The first three methylation loci are DMLs, as are the last five methylation loci.
-However, the fourth locus is not a DML.
+The first three methylation loci are DMCs, as are the last five methylation loci.
+However, the fourth locus is not a DMC.
 Should this region be considered as two distinct DMRs or as a single DMR?}
@@
 Instead, they capture correlations of methylation levels by inducing dependence in the $B_{i, j}$.
-For example, under the Beta-Binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent.
+For example, under the beta-binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent.
 This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure via a lag-1 autoregressive model of the simulated $\beta$-values.
@@
 Furthermore, since all loci within each of the four states are assigned the same (perturbed) underlying methylation level, $B_{i, j_{k}} = B_{region-type} + \epsilon_{i, j_{k}}$, neighbouring loci have similar methylation levels.
-Note that the $IPD$ only plays a direct role in the initial segmentation of the genome, not in the assigning of the $B_{i, j_{k}}$.
+Note that the $IPD$ only plays a direct role in the initial segmentation of the genome, not in the assigning of the $B_{i, j}$.
 
@@
 A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e.
-without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb}.
+without simulating sequencing depth \citep{Jaffe:2012gx, Chen:2013eh, Chen:2014jb}.
 These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth.
@@
 A simulation may also be based entirely on re-sampling of real data[^synthetic].
-This type of simulation is attractive because through careful sampling it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model.
+This type of simulation is attractive because, through careful sampling, it can capture behaviour that is otherwise very difficult, if not impossible, to capture in a parametric model.
 At the same time, however, if the sampling units are poorly chosen or the sampling strategy is incorrect, then it may ignore these same features or, worse still, introduce artefacts into the simulated data.
@@
 Re-sampling methods are most easily implemented at the level of $\beta$-values.
-Re-sampling reads is reads is more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis.
+Re-sampling reads is more difficult, except in the special case of down-sampling whereby the positions of reads are held constant but only a sub-sample of them are used in downstream analysis.
 Down-sampling is only really of interest to examine the effects of sequencing coverage on downstream analyses.
@@
 
-Due to the correlation structure of the $\beta$-values, the co-methylation, it is likely, although not guaranteed, that other CpGs in the blocks containing targets also display differential methylation.
+Due to the correlation structure of the $\beta$-values (the co-methylation) it is likely, although not guaranteed, that other CpGs in the blocks containing targets also display differential methylation.
 
@@
 In order to model the co-methylation structure, I decided that `methsim` should simulate at the level of individual DNA fragments.
-An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples or entire sequencing reads.
+An added bonus of this approach is that `methsim` can (in theory) generate data at multiple resolutions: $\beta$-values, methylation patterns at m-tuples, or entire sequencing reads.
 While my initial motivation in developing `methsim` was to explore models of co-methylation, I realised that it could also be used in the development and comparison of methods for the downstream analysis of DNA methylation data.
@@
 2. Simulating reads, including sequencing error and bisulfite-conversion error, by sampling from the true methylome.
-3. Constructing the output, be it reads, methylation patterns at m-tuples or $\beta$-values.
+3. Constructing the output, be it reads, methylation patterns at m-tuples, or $\beta$-values.
 
@@
 `methsim` makes extensive use of the S4 object system in R.
-The most important classes defined by `methsim` are `MethylomeParam`, `SimulatedMethylome`, `WGBSParam` and `SimulatedBS`.
+The most important classes defined in `methsim` are `MethylomeParam`, `SimulatedMethylome`, `WGBSParam` and `SimulatedBS`.
 The most important methods are the `simulate()` methods defined for the `MethylomeParam` and `WGBSParam` classes.
@@
 The `MethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled.
-Therefore, the `MethylomeParam` should be based on data from a relevant sample.
+Therefore, the `MethylomeParam` object should be based on data from a relevant sample.
 To help a new user get started, `methsim` includes `MethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset.
 Alternatively, the user may process the `BAM` file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `MethylomeParam` object based on their sample of interest.
-
-To simulate a true methylome, the user runs the `simulate()` method on the `MethylomeParam` object. This returns a `SimulatedMethylome` object.
+The user runs the `simulate()` method on the `MethylomeParam` object to simulate a true methylome; this returns a `SimulatedMethylome` object.
 
@@
 To simulate values of $\mathbf{B}$ and $\bm{\psi}$, `methsim` uses the empirical distributions of their estimates, $\bm{\beta}$ and $\widehat{\bm{\theta}} = \log(\widehat{\bm{\psi}})$, in the input methylome.
-More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\bm{\theta}$ is based on $\widehat{\theta}_{MH}$.
+More specifically, $\mathbf{B}$ is based on $\bm{\beta_{i | r_{i}}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\bm{\theta}$ is based on $\widehat{\theta}_{MH}$.
 We have already seen many examples of the distribution of $\widehat{\theta}_{MH}$ in Chapter \ref{chap:co-methylation}.
-The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.
+The empirical distributions $\bm{\beta_{i | r_{i}}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.
 
@@
 \includegraphics[width=\textwidth]{../figures/ADS_beta_by_pm_boxplots.pdf}
-\caption[Distribution of $\beta$-values stratified by region type]{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
+\caption[Distribution of $\beta$-values stratified by region type]{$\bm{\beta_{i | r_{i}}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{ADS} sample.
 Only CpGs with at least $10\times$ sequencing coverage are used.
@@
 \begin{description}
-  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
-  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this average is then perturbed for each locus by a $Gaussian(0, \sigma_{B}^2)$ random variable.
+  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\bm{\beta_{i | r_{i}}}$.
+  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\bm{\beta_{i | r_{i}}}$; this average is then perturbed for each locus by a $Gaussian(0, \sigma_{B}^2)$ random variable.
 This is similar to what is done by \texttt{WGBSsuite}.
@@
 The user specifies the desired read length[^read_length], the average sequencing coverage, the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.
-We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $30 \times$ in what follows.
+We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $23 \times$ (mimicking the actual _ADS_ data) in what follows.
 
@@
 \includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
-\caption[\texttt{methsim}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island.
+\caption[\texttt{methsim}: $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs.
-The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
+The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points (only autosomal data are simulated, hence only autosomal data are shown).}
 \label{fig:methsim_MH_by_seqnames}

### 2015-05-06 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 \begin{chapabstract}
-I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used methylC-seq protocol.
+This chapter explains the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used methylC-seq protocol.
 All data used in my thesis were generated using this protocol.
@@
 
-The chapter concludes by introducing the \texttt{methtuple} software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci.
+The chapter concludes by introducing the \texttt{methtuple} software that I wrote, a unique methylation caller for extracting methylation patterns at tuples of methylation loci.
 \texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
@@
 $20\%$).
-As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.
+As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in its sequence composition.
 
@@
 Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.
-Several review articles have summarised and compared the various approaches \citep[e.g.,][]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
+Several review articles have summarised and compared the various approaches \citep{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.
 
@@
 These repetitive sequences will map to multiple locations in the reference genome equally well.
-Furthermore, the number of times these repetitive sequences occur differs between the reference genome and sample's genome.
+Furthermore, the number of times these repetitive sequences occur differs between the reference genome and the sample's genome.
 
-Technically, reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences.
+Reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences.
 Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}, which produces longer reads.
@@
 A sequencing error can transform a uniquely mapping read to one that maps equally well to multiple locations or, worse still, a read that maps uniquely to a single, but incorrect, location.
-Sequencing errors can also corrupt a read so badly that it no longer can be mapped, which leads to that read being a false negative.
+Sequencing errors can also corrupt a read so badly that it no longer can be mapped.
 In practice, most people try to mitigate these problems through their choice of parameters used by the read mapping software.
@@
 PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work.
-Unfortunately, it can introduce biases into the library that result in some molecules being over-represented or under-represented compared to their true frequency.
+Unfortunately, it can introduce biases into the library that results in some molecules being over-represented or under-represented compared to their true frequency.
 This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome.
@@
 
-In the datasets I have analysed, I have had to ignore up to 30 read-positions per read due to M-bias.
+In the datasets I have analysed, I have had to ignore up to 30 read-positions per read due to M-bias, a considerable loss of data.
 Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely.
@@
 CpG, CHG or CHH) based on the sequencing data and a reference DNA sequence.
-In principle, this is a simple process, however, this belies some complications, which I discuss in this section.
+In principle, this is a simple process, however, this belies some complications, which we discuss in this section.
 
@@
 3.
-Measurability: We cannot observe m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology.
+Measurability: We cannot observe methylation patterns from individual reads at m-tuples where the methylation loci are far apart due to the read length limitations of the Illumina sequencing technology.
 This is true even when $NIL = 0$ but is more of an issue if we allow $NIL \geq 0$.
@@
 There are many useful options to filter reads and read-positions.
-Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting these overlapping bases.
+Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting the bases in the overlapping region.
 `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads.
@@
 
-The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its ploidy.
+The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its copy number.
 The running time is largely independent of the tuple size (`-m`).

### 2015-05-06 Rmarkdown/wgbs_downstream_analyses.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 \begin{chapabstract}
-This chapter discussed methods for the \emph{downstream analysis} of bisulfite-sequencing data.
+This chapter discusses methods for the \emph{downstream analysis} of bisulfite-sequencing data.
 Downstream analyses proceed the processing of the raw data (Chapter \ref{chap:wgbs_bioinformatics_analysis}) to address the scientific questions of interest.
-Most downstream methods analyse methylation counts at 1-tuples, however, there is growing interest in analyses based on methylation patterns at m-tuples.
+Most downstream analyses are based on methylation counts at 1-tuples, however, there is growing interest in analyses based on methylation patterns at m-tuples.
 I discuss the statistical questions underlying these downstream analyses, paying particular attention to those based on m-tuples (m $> 1$) since these have received less attention in the literature.
@@
 $\mathbf{m} = (m_{1}, \ldots, m_{i}, \ldots, m_{N_{loci}})$ and $\mathbf{u} = (u_{1}, \ldots, u_{i}, \ldots, u_{N_{loci}})$[^not_beta_values].
-Methods based on 1-tuples have been developed to address several scientific questions including testing for differential methylation (Section \ref{sec:differential_methylation}), testing for differentially variable methylation (Section \ref{sec:variable_methylation}) and identifying regulatory regions of the genome (Section \ref{sec:epigenome_segmentation}).
+Methods based on 1-tuples have been developed to address a variety of scientific questions including testing for differential methylation (Section \ref{sec:differential_methylation}), testing for differentially variable methylation (Section \ref{sec:variable_methylation}) and identifying regulatory regions of the genome (Section \ref{sec:epigenome_segmentation}).
 
-[^not_beta_values]: It is insufficient to use $\bm{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).
+[^not_beta_values]: It is insufficient to use $\bm{\beta} = \frac{\mathbf{m}}{\mathbf{m} + \mathbf{u}}$ because the conversion to $\beta$-values loses information about the precision with which each methylation locus is measured (i.e.
+the sequencing depth, $\mathbf{d} = \mathbf{m} + \mathbf{u}$).
 
@@
 
-Initial experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or these were pooled prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}).
-Simply pooling replicates and analysing as if they were a single sample ignores all within-group variability and should not be used.
+Initial experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or, if they did, these were pooled prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}).
+Simply pooling replicates and analysing as if they were a single sample ignores all variability between replicates and should not be used.
 
@@
 This is typically framed as a test of the mean levels of methylation at the locus in each group[^mean].
-Suppose we have a two-group experiment and let $B_{i}^{k}$ denote the true methylation level of the $i^{th}$ methylation locus in the $k^{th}$ group ($k = 0, 1$).
-We wish to test the null hypothesis of $H_{0}: B_{i}^{0} = B_{i}^{1}$ against the alternative hypothesis $H_{1}: B_{i}^{0} \neq B_{i}^{1}$.
+Suppose we have a two-group experiment and let $B_{i, j_{k}}$ denote the true methylation level of the $i^{th}$ methylation locus for samples in the $k^{th}$ group ($k = 0, 1$).
+We wish to test the null hypothesis of $H_{0}: B_{i, j_{0}} = B_{i, j_{1}}$ against the alternative hypothesis $H_{1}: B_{i, j_{0}} \neq B_{i, j_{1}}$.
 As such, identifying DMCs boils down to identifying differences in means, for which there is an enormous body of statistical literature.
@@
 
-[^mean]: This could alternatively be framed as a test of the median methylation level at the locus in each group (or some other location parameter of a statistical distribution).
+[^mean]: This could alternatively be framed as a test of the median methylation level at the locus in each group (or of some other location parameter of the distribution of methylation levels).
 
@@
 Most are limited to analysing two-group experiments.
-Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of $\bm{\beta}$ (Section \ref{sec:transformations}).
+Rather than directly analysing the $\mathbf{m}$ and $\mathbf{u}$, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_models}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (Section \ref{sec:transformations}).
 
-`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk} `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} all use a Beta-Binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
+`DSS` \citep{Feng:2014iq}, `BiSeq` \citep{Hebestreit:2013ko}, `MOABS` \citep{Sun:2014fk}, `methylSig` \citep{Park:2014ho} and `RADmeth` \citep{Dolzhenko:2014bo} all use a beta-binomial hierarchical model of DNA methylation, although the exact details differ considerably between packages.
 `DSS` and `MOABS` use empirical Bayes methods to estimate parameters whereas `methylSig`, `BiSeq` and `RADmeth` use maximum likelihood estimation.
@@
 Not all software for identifying differential methylation are designed for identifying DMCs.
-For example, both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs.
+For example, both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying differentially methylated regions rather than DMCs.
 
@@
 Rather, the idea is that a DMR might capture a weaker but broader difference in methylation.
-For example, it may be more biologically relevant to identify a broad region with a consistent, albeit smaller, difference in methylation than it is to identify individual cytosines with large differences in methylation.
+For example, it may be more biologically relevant to identify a broad region with a consistent, albeit small, difference in methylation than it is to identify individual cytosines with large differences in methylation.
 
@@
 Such regions might be CpG islands \citep[e.g.,][]{Huang:1999fs,Doi:2009ky}[^array]; MspI restriction fragments \citep[e.g.,][]{Stockwell:2014fq}; a predefined genomic feature, such as a gene promoters or transcription factor binding sites; or general predefined bins \citep[e.g., 100 bp bins used by][]{Park:2014ho}.
-2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.
+2. Using data-driven regions, such as those defined from an analysis of DMCs, which are then tested for differential methylation.
 
@@
 The idea of testing for differential methylation at _a priori_ defined regions is relatively straightforward.
-Suppose we have a two-group experiment and let $\widebar{B_{r}^{k}}$ be the true average level of methylation for the $r^{th}$ region in the $k^{th}$ group ($k = 0, 1$).
-The null hypothesis is $H_{0}: \widebar{B_{r}^{0}} = \widebar{B_{r}^{1}}$ against the alternative hypothesis $H_{1}: \widebar{B_{r}^{0}} \neq \widebar{B_{r}^{1}}$.
+Suppose we have a two-group experiment and let $\widebar{B}_{r, j_{k}}$ be the true average level of methylation for the $r^{th}$ region for samples in the $k^{th}$ group ($k = 0, 1$).
+The null hypothesis is $H_{0}: \widebar{B}_{r, j_{0}} = \widebar{B}_{r, j_{1}}$ against the alternative hypothesis $H_{1}: \widebar{B}_{r, j_{0}} \neq \widebar{B}_{r, j_{1}}$.
 
 We might estimate $\widebar{B_{r}^{k}}$ by the group-wise average of the sample-wise weighted average of $\beta$-values for all methylation loci in the region, where the weights are proportional to the sequencing coverage.
-Then, identifying differential methylation at _a priori_ defined regions simply boils down to identifying differences in means, just as is the case for testing for DMCs.
+Identifying differential methylation at _a priori_ defined regions simply boils down to identifying differences in means, just as is the case for testing for DMCs.
 Again, this problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.
@@
 Tests of differential variability that are based on the F-test \citep[e.g.,][]{Hansen:2011gu} or Bartlett's test \citep[e.g.,][]{Teschendorff:2012hea} will be susceptible to calling loci with such outliers as being differentially variable.
-By contrast, `DiffVar` \citep{Phipson:2014ji} uses Levene's test \citep{Olkin:1960uu}, which is robust to outliers, to test for differential variability.
+By contrast, `DiffVar` \citep{Phipson:2014ji} uses Levene's test \citep{Olkin:1960uu} to test for differential variability since it is robust to outliers.
 
@@
 \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require heterozygous genetic variants nearby to the methylation locus of interest.
-These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data.
+These methods rely on the probabilistic assignment of reads to alleles (which are treated as missing data).
 Unfortunately, there is no publicly available software implementing the method proposed by \citet{Peng:2012dh} and so I do not discuss it further.
@@
 
-The rate of _epipolymorphism_ of a locus is defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}[^epipolymorphism].
+The _rate of epipolymorphism_ of a locus is defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}[^epipolymorphism].
 \citeauthor{Landan:2012kp} define the rate of epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e.
@@
 
-The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error.
+The obvious challenge in estimating the frequency of an epiallele is in distinguishing a 'real' epiallele from a spurious one (perhaps caused by incomplete bisulfite-conversion) sequencing error or mapping error.
 Another difficulty, perhaps unavoidable with current technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele.
@@
 `methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies and does not attempt to account for potential sources of bias.
-In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [i.e.
+In contrast, `MPFE` (\url{http://bioconductor.org/packages/MPFE}, \url{http://f1000.com/posters/browse/summary/1097258}), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [i.e.
 epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.
@@
 
-Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,][]{Fang:2012ci, Peng:2012dh}, might also adapted to identify epialleles and their associated frequencies.
+Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,][]{Fang:2012ci, Peng:2012dh}, might also be adapted to identify epialleles and their associated frequencies.
 It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which, strictly speaking, is a requirement for the m-tuple to be an epipolymorphic locus.
@@
 `MethPipe` is perhaps the best documented and potentially extensible software for analysing methylation patterns at m-tuples.
-`MethPipe` is mostly written in \CC and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data.
+`MethPipe` is mostly written in \CC \ and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data.
 As such, it does not feature tools that are particularly amenable to interactive or exploratory analyses.
@@
 
-While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are well-suited to any analysis based on methylation patterns of m-tuples such as methylation entropy, allele-specific methylation and the identification of epialleles.
-Of course, `MethylationTuples` can equally be used to develop methods based on 1-tuples, such as identifying differential methylation, since 1-tuples are just a particular type of m-tuple.
+While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are well-suited to other analyses based on methylation patterns of m-tuples such as methylation entropy, allele-specific methylation and the identification of epialleles.
+Of course, `MethylationTuples` can also be used to develop methods based on 1-tuples, such as identifying differential methylation, since 1-tuples are just a particular type of m-tuple.
 
@@
 From a developer's perspective, this avoids the need to re-invent the wheel when tackling common tasks.
-And from the user's perspective, it helps avoid multiple versions of the 'wheel', each that might otherwise act slightly differently and that may not be as well-tested.
+And from the user's perspective, it helps avoid multiple versions of the wheel, each that might otherwise act slightly differently and that may not be as well-tested.
 
-[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to [bioconductor.org/packages/GenomicTupless](bioconductor.org/packages/GenomicTuples).
+[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to \url{https://github.com/PeteHaitch/MethylationTuples}).
 
 Bioconductor already has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}.
-However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges].
-Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)).
+Genomic tuples, however, such as the co-ordinates of an m-tuple, do not naturally fit into this framework[^tuples_vs_ranges].
+Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 (\url{http://bioconductor.org/packages/GenomicTuples}).
 In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations.
@@
 
-In the `MethylationTuples` package, I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated counts of each methylation pattern.
-A `MethPat` object is as a matrix-like object, where rows represent m-tuples of interest and columns represent samples.
+In the `MethylationTuples` package I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated counts of each methylation pattern.
+A `MethPat` object is as a matrix-like object, where rows represent m-tuples and columns represent samples.
 The `MethPat` class extends the `GenomicRanges::SummarizedExperiment`[^namespace] class but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples.
@@
 Since `MethylationTuples` is based on core Bioconductor functionality, it is highly compatible with existing Bioconductor packages.
-In particular, `MethPat` objects containing 1-tuples are readily coerced for use in differential methylation calling packages, e.g., `bsseq` and `BiSeq`, or to identify partially methylated domains with `MethylSeekR`.
+In particular, `MethPat` objects containing 1-tuples are readily coerced for use with differential methylation calling packages, e.g., `bsseq` and `BiSeq`, or to identify partially methylated domains with `MethylSeekR`.
 I also make extensive use of `MethylationTuples` in my `methsim` software (Chapter \ref{chap:methsim})
@@
 \centering
-\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data.
+\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data ($N_{samples} = 12$).
 All m-tuples are stranded.

### 2015-05-06 Rmarkdown/wgbs_statistical_framework.Rmd
Edits from final proofread. How do I keep finding freaking typos!?

@@
 
-Beginning with a single sample, I explain the various sources of variation in this statistical model and introduce the mathematical notation that I use throughout my thesis.
+Beginning with a single sample, I explain the various sources of variation in DNA methylation data and introduce the mathematical notation that I use throughout my thesis.
 I then extend this framework to multiple samples.
@@
 
-Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment. I find it convenient to separate these into:
+There are several levels of variation to consider in a bisulfite-sequencing experiment, even with only a single sample. I find it convenient to separate these into:
 
@@
 
-The methylation state of a locus can vary within a sample due to cell-to-cell variability of DNA methylation.
+The methylation state of a locus can vary within a sample due to cell-to-cell heterogeneity of DNA methylation.
 A sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell may have a slightly different methylation profile.
 Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states.
-Therefore, it is also necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.
+It is also therefore necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.
 
@@
 Note that the value of $H_{i}$ is determined following the library preparation, including PCR amplification of the DNA; therefore, it can give a grossly distorted picture of the true representation of the cells.
-I denote by $\mathcal{H}_{i}$ the set of all fragments containing the $i^{th}$ locus.
+We denote by $\mathcal{H}_{i}$ the set of all fragments containing the $i^{th}$ locus.
 
@@
 Although we do not know the number of fragments in the pool, we can define (and measure) the methylation state of a locus on a single DNA fragment.
-I denote by the indicator random variable, $Z_{h, i}$, the methylation state of $i^{th}$ methylation locus on the $h^{th}$ DNA fragment:
+We denote by the indicator random variable, $Z_{h, i}$, the methylation state of $i^{th}$ methylation locus on the $h^{th}$ DNA fragment:
 
@@
 
-I denote by the vector of indicator random variables, $Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)}$, the methylation pattern of the m-tuple $(i, i + 1, \ldots, i + m - 1)$ on the $h^{th}$ DNA fragment containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$:
+We denote the methylation pattern on the $h^{th}$ DNA fragment containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ by the vector of indicator random variables, $Z_{h, (i, i + 1, \ldots, i + \text{m} - 1)}$,:
 
@@
   \begin{array}{l l}
-  (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
-  & \quad \text{ locus on the } h^{th} \text{ fragment}\\
-  (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big ) \\
-  & \quad \text{ locus and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus on the } h^{th} \text{ fragment} \\
+  (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \Big (i^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
+  & \quad \text{locus on the } h^{th} \text{ fragment}\\
+  (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \Big (i^{th}, \ldots, (i + \text{m} - 2)^{th} \Big ) \\
+  & \quad \text{locus and methylated at the } (i + \text{m} - 1)^{th} \\
+  & \quad \text{locus on the } h^{th} \text{ fragment} \\
   \vdots \\
-  (1, 1, \ldots, 1) & \quad \text{if methylated at the } \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
-  & \quad \text{ locus on the } h^{th} \text{ fragment}
+  (1, 1, \ldots, 1) & \quad \text{if methylated at the } \Big (i^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
+  & \quad \text{locus on the } h^{th} \text{ fragment}
   \end{array} \right.
@@
 
-$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$.
+We denote the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$ by $\mathcal{H}_{(i, i + 1, i + m - 1)}$.
 
@@
 We have a large number (~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments.
-Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to the number of fragmentation containing the $i^{th}$ methylation locus, $H_{i}$.
+We can therefore approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to the number of fragmentation containing the $i^{th}$ methylation locus, $H_{i}$.
 
-We can ignore reads that do not contain any methylation loci as these are not relevant to ths discussion. I make three further simplifying assumptions:
+We can ignore reads that do not contain any methylation loci as these are not relevant to ths discussion. We make three further simplifying assumptions:
 
@@
 When using single-end sequencing, the methylation loci from a single read will always form a positively ordered sequence without gaps, i.e.
-$(i, i + 1, i + 2)$ and not $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
+$(i, i + 1, i + 2)$ and not, for example, $(i, i - 1, i - 2)$ nor $(i, i + 1, i + 3)$.
 However, when using paired-end sequencing, the methylation loci from a paired-end read will still be an ordered sequence but one of the following may occur:
@@
 Each read measures the methylation state of one or more loci from a single DNA fragment.
-I denote by $\mathcal{R}_i$ the set of all mapped reads containing the $i^{th}$ locus.
-The number of reads containing the $i^{th}$ locus is referred to as the sequencing depth at the $i^{th}$ locus, which I denote by $d_{i} = |\mathcal{R}_{i}|$, where $d_{i} \leq H_{i}$ with strict inequality for almost all $i$.
+We denote by $\mathcal{R}_i$ the set of all mapped reads containing the $i^{th}$ locus.
+The number of reads containing the $i^{th}$ locus is referred to as the sequencing depth at the $i^{th}$ locus, which we denote by $d_{i} = |\mathcal{R}_{i}|$, where $d_{i} \leq H_{i}$ with strict inequality for almost all $i$.
 
@@
 This is the so-called $\beta$-value, which is commonly interpreted as an estimate of $B_{i}$, the proportion of cells in the sample that are methylated at the $i^{th}$ locus.
-In Section \ref{sec:estimating_B} I discuss this interpretation and other estimators of the 'methylation level' at a locus.
+In Section \ref{sec:estimating_B} we discuss this interpretation and other estimators of the 'methylation level' at a locus.
 
@@
     & \quad \Big ( i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big) \\
-    & \quad \text{ locus}\\
+    & \quad \text{locus}\\
     (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \\
     & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big) \\
-    & \quad \text{ locus} \\
+    & \quad \text{locus} \\
       & \quad \text{and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
@@
     & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
-    & \quad \text{ locus}
+    & \quad \text{locus}
       \end{array} \right.
@@
 
-As we did for $B$-values, we can extend the $\beta_{i}$ values to m-tuples. Here are the definitions for $\text{m} = 2$:
+As we did for $B$-values, we can extend the $\beta$ values to m-tuples. Here are the definitions for $\text{m} = 2$:
 
@@
 
-I now discuss some complications and how this framework might accommodate these issues in practice.
+We now discuss some complications and how this framework might accommodate these issues in practice.
 
@@
 
-A fourth level is how the sample's relate to one another, such as the treatment group of each sample.
+A fourth level is how the samples relate to one another, such as the treatment group of each sample.
 This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between tumour and normal tissue.
@@
 
-A conservative analysis might only analyse those loci where at least some fraction of the $n$ samples have sufficient sequencing coverage.
+A conservative analysis might only analyse those loci where at least some fraction of the samples have sufficient sequencing coverage.
 A less conservative analysis might try to impute the missing values based on methylation levels at neighbouring loci.
@@
 
+In this section we review various methods for estimating these key parameters.
 The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}} = (B_{1,j}, \ldots, B_{N_{loci}, j})$.
-In this section I review various methods for estimating these key parameters.
 When necessary, I have 'translated' the original work into my notation to make these methods more readily comparable.
@@
 Both `bsseq` and `BiSeq` use a binomial local likelihood smoother.
-In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
+In each case this smoother is chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
 The smoothing is 'local' to exploit co-methylation, which is considered a local phenomenon.
@@
 
-Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Bin(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{d_{i, j}}}$ \citep{Hansen:2012gr}.
+Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{d_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{d_{i, j}}}$ \citep{Hansen:2012gr}.
 The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample.
@@
 What is immediately clear is that these distributions are bimodal: most CpGs are highly methylated or lowly methylated.
-The exception is the _E16.5\_male\_1_ sample from the _Seisenberger_ data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels.
-The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice.
+The exception is the _E16.5\_male\_1_ sample from the _Seisenberger_ data which is hypomethylated and with an enormous number of intermediately methylated CpGs.
+The _E16.5\_male\_1_ sample is a progenitor germ cells from a pool of embryonic day 16.5 male mice.
 Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \citep{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.
@@
 The explanation for the somatic cell lines is less clear.
-Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation.
+Notably, all of the 12 EPISCOPE samples, which are tissue samples rather than cell lines, have relatively low levels of intermediate methylation.
 Likewise, the various frontal cortex samples in the _Ziller_ dataset, which includes both 'normal' and Alzheimer's samples (_Frontal\_cortex\_normal\_1_, _Frontal\_cortex\_normal\_2_, _Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_), have very low levels of intermediate methylation.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
-\caption[\emph{EPISCOPE} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data.
+\caption[\emph{EPISCOPE} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
-\caption[\emph{Lister} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data.
+\caption[\emph{Lister} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
-\caption[\emph{Seisenberger} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data.
+\caption[\emph{Seisenberger} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
-\caption[\emph{Ziller} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data.
+\caption[\emph{Ziller} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 
-The bimodality of the genome-wide $\beta$-value distributions is driven by the fact that most CpGs in CpG islands are unmethylated whereas those outside of the islands are mostly methylated.
+\FloatBarrier
+
+The bimodality of the genome-wide $\beta$-value distributions is driven by the fact that most CpGs in CpG islands are unmethylated whereas those outside of the CpG islands are mostly methylated.
 Figures \ref{fig:EPISCOPE_plotBetaDensity_CGI}, \ref{fig:Lister_plotBetaDensity_CGI}, \ref{fig:Seisenberger_plotBetaDensity_CGI} and \ref{fig:Ziller_merged_plotBetaDensity_CGI} show the kernel density plots of the $\beta$-value distributions stratified by CpG island status for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.
@@
 Therefore, Figures \ref{fig:EPISCOPE_plotBetaFreqPoly_CGI}, \ref{fig:Lister_plotBetaFreqPoly_CGI}, \ref{fig:Seisenberger_plotBetaFreqPoly_CGI} and \ref{fig:Ziller_merged_plotBetaFreqPoly_CGI} show the same data but with $\beta$-values grouped into $0.01$-width bins and plotted against the percentage of total CpGs in each bin.
-These plots highlight that the majority of unmethylated CpGs occur in CpG islands and that most of the intermediate methylation occurs outside of CpG islands, owing to most CpGs being outside of an island.
+These plots highlight that the majority of unmethylated CpGs occur in CpG islands and that most of the intermediate methylation occurs outside of CpG islands, owing to most CpGs being outside a CpG island.
 
@@
 
-\subsection{Spatial correlations of $\beta$-values}
-
-Many researchers have observed that DNA methylation is spatially correlated along the genome, \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}.
-I call this spatial correlation of methylation levels _co-methylation_.
-
-I examine in detail the spatial correlations of $\beta$-values in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
-
 \subsection{Sources of bias in $\beta$-values}\label{sec:beta_biases}
@@
 
+\subsection{Spatial correlations of $\beta$-values}
+
+Many researchers have observed that DNA methylation is spatially correlated along the genome, \citep[e.g.,][]{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Lacey:2013iy, Sofer:2013bk, Liu:dy, Lyko:2010dr, Landan:2012kp, Lister:2009hy}.
+I call this spatial correlation of methylation levels _co-methylation_.
+
+I examine in detail the spatial correlations of $\beta$-values in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.
+
 \section{Summary}

### 2015-05-06 Rmarkdown/comethylation.Rmd
Final touches

@@
 
-\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, converge to the true odds ratio.
-In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi}_{MH}$, and the conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, do converge to $\psi$, with $\widehat{\psi}_{MH}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.
+\citet{Breslow:1981ta} showed that under $A2$ that the unconditional maximum likelihood estimator does not converge to the true odds ratio.
+In contrast, both the Mantel-Haenszel estimator and the conditional maximum likelihood estimator do converge to the true odds ratio, with the Mantel-Haenszel estimator having the advantage of a simple closed form expression \citep{Breslow:1996uc}.
 

### 2015-05-07 Rmarkdown/comethylation.Rmd
Fix printing related glitches

@@
 
-The results of each simulation are summarised by a plot of $\widehat{\theta}_{0.5}$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
-The density of $\theta_{k}$, the value of the true log odds ratio in the $k^{th}$ slice of the $2 \times 2 \times K$ table, is shown by a dashed black line[^density].
+The results of each simulation are summarised by a plot of $\widehat{\theta}_{0.5}$ (histogram), the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange), and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+The density of $\theta_{k}$, the value of the true log odds ratio in the $k^{th}$ slice of the $2 \times 2 \times K$ table, is shown by a black curve[^density].
 
@@
 \caption[Simulation 1a]{Results of simulation 1a ($K = 100$).
-Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating the log odds ratio, $\theta = 0$ (shown by the black line), from a $2 \times 2 \times 100$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 1b]{Results of simulation 1b ($K = 1000$).
-Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating the log odds ratio, $\theta = 0$ (shown by the black line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 1c]{Results of simulation 1c ($K = 1000$).
-Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating the log odds ratio, $\theta = 0$ (shown by the black line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $10,000$.}
@@
 \caption[Simulation 2]{Results of simulation 2 ($K = 100$).
-Estimating the log odds ratio, $\theta$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Uniform(0, 1)$ random variables.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating the log odds ratio, $\theta$ (shown by the black line, $-4, -3, \ldots, 3, 4$ across the panels), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Uniform(0, 1)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 3]{Results of simulation 3 ($K = 100$).
-Estimating the log odds ratio, $\theta$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating the log odds ratio, $\theta$ (shown by the black line, $-4, -3, \ldots, 3, 4$ across the panels), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 4a]{Results of simulation 4a ($K = 100$).
-Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the black curve), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 4b]{Results of simulation 4b ($K = 1000$).
-Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the black curve), from a $2 \times 2 \times 1000$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 5]{Results of simulation 5 ($K = 100$).
-Estimating a heterogeneous log odds ratio, $\theta_{k} = Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$ (density shown by the black curve), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
@@
 \caption[Simulation 6]{Results of simulation 5 ($K = 100$).
-Estimating a heterogeneous log odds ratio, $\theta_{k} = 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
-The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue).
+Estimating a heterogeneous log odds ratio, $\theta_{k} = 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$ (density shown by the black curve), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables.
+The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (inner line) and $95 \%$ confidence interval (outer lines) of $\widehat{\theta}_{MH}$ (coloured blue).
 The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}

### 2015-05-07 Rmarkdown/comethylation_review.Rmd
Fix printing related glitches

@@
 \centering
-\includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.pdf}
+\includegraphics[width=0.8\textwidth]{../figures/Lister_2009_Sup_Fig_9.jpg}
 \caption[Co-methylation figure from \citet{Lister:2009hy}]{Spacing of Adjacent Methylcytosines in Different Contexts.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{../figures/Cokus_2008_Sup_Fig_14.pdf}
+\includegraphics[width=0.8\textwidth]{../figures/Cokus_2008_Sup_Fig_14.jpg}
 \caption[Co-methylation figure from \citet{Cokus:2008fc}]{Within-read probability of additional methylation of CHH sites within a given distance from a methylated CHH site.

### 2015-08-24 Rmarkdown/biology_background.Rmd
Change some tables to figures (closes #14).

@@
 Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples.
-Furthermore, depending on the choice of several sequencing parameters, approximately $35 \%$ to $72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.
+Furthermore, depending on the choice of several sequencing parameters, approximately $35 \%$ to $72\%$ of reads will not contain any CpGs (Figure \ref{fig:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.
 
-\begin{table}[h]
-\centering
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../figures/percentage-reads-without-CpGs.pdf}
 \caption[Percentage of reads without CpGs]{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset.
-Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs.
+Paired-end (PE) reads are counted as a single unit.
+Therefore, samples sequenced with paired-end reads have a lower percentage of reads without CpGs than do samples sequenced with single-end (SE) reads.
 See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
-\label{tab:reads_without_cpgs}
-\begin{tabular}{@{}rc@{}}
-\toprule
-Sample              & Reads with no CpG \\ \midrule
-ADS-iPSC            & $35 \%$                \\
-ADS-adipose         & $37 \%$                \\
-ADS                 & $39 \%$                \\
-H9\_Laurent         & $39 \%$                \\
-FF-iPSC\_19.11      & $48 \%$                \\
-FF-iPSC\_19.7       & $49 \%$                \\
-H1+BMP4             & $49 \%$                \\
-FF-iPSC\_6.9        & $50 \%$                \\
-IMR90-iPSC          & $50 \%$                \\
-H9                  & $51 \%$                \\
-FF-iPSC\_19.11+BMP4 & $52 \%$                \\
-FF                  & $53 \%$                \\
-IMR90\_r2           & $57 \%$                \\
-IMR90\_r1           & $58 \%$                \\
-H1\_r2              & $59 \%$                \\
-H1\_r1              & $62 \%$                \\
-HSF1                & $72 \%$                \\ \bottomrule
-\end{tabular}
-\end{table}
+\label{fig:reads_without_cpgs}
+\end{figure}
 
@@
 When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \citep{Liu:2012ge}.
-Figure \ref{fig:bisulfite_CT_SNP.pdf} illustrates such an example.
+Figure \ref{fig:bisulfite_CT_SNP} illustrates such an example.
 
@@
 Because the read is compared against the reference genome, it may incorrectly inferred to be a thymine.}
-\label{fig:bisulfite_CT_SNP.pdf}
+\label{fig:bisulfite_CT_SNP}
 \end{figure}

### 2015-08-24 Rmarkdown/datasets.Rmd
Change some tables to figures (closes #14).

@@
 \toprule
-Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
+Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\ \midrule
 IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp    & $14\times$ \\
@@
 \toprule
-Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
+Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\ \midrule
 ADS                 & Adipose                 & PE         & $75$ bp    & $23\times$ \\
@@
 \centering
-\caption[Description of \emph{EPISCOPE} dataset]{Samples in \emph{EPISCOPE} dataset. The reported read lengths are prior to any trimming of the reads.}
+\caption[Description of \emph{EPISCOPE} dataset]{Samples in \emph{EPISCOPE} dataset.
+The reported read lengths are prior to any trimming of the reads.
+Abbreviations: \emph{PE} = paired-end sequencing; \emph{BUF} = buffy coat; \emph{SA} = subcutaneous adipose; \emph{VA} = visceral adipocytes; \emph{VAT} = visceral adipose tissue.}
 \label{tab:EPISCOPE_dataset}
-\begin{tabulary}{\textwidth}{LLLL}
+\begin{tabulary}{\textwidth}{LLLLL}
 \toprule
-Sample & Sequencing & Read length & Ave. coverage \\ \midrule
-E13BUF & PE         & $101$ bp    & $8\times$     \\
-E13SA  & PE         & $101$ bp    & $28\times$    \\
-E13VA  & PE         & $150$ bp    & $27\times$    \\
-E13VAT & PE         & $101$ bp    & $25\times$    \\
-E18BUF & PE         & $101$ bp    & $21\times$    \\
-E18SA  & PE         & $101$ bp    & $25\times$    \\
-E18VA  & PE         & $150$ bp    & $36\times$    \\
-E18VAT & PE         & $101$ bp    & $26\times$    \\
-E23BUF & PE         & $101$ bp    & $12\times$    \\
-E23SA  & PE         & $101$ bp    & $29\times$    \\
-E23VA  & PE         & $101$ bp    & $32\times$    \\
-E23VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
+Sample & Tissue & Sequencing & Read length & Ave. coverage \\ \midrule
+E13BUF & BUF & PE         & $101$ bp    & $8\times$     \\
+E13SA  & SA  & PE         & $101$ bp    & $28\times$    \\
+E13VA  & VA  & PE         & $150$ bp    & $27\times$    \\
+E13VAT & VAT & PE         & $101$ bp    & $25\times$    \\
+E18BUF & BUF & PE         & $101$ bp    & $21\times$    \\
+E18SA  & SA  & PE         & $101$ bp    & $25\times$    \\
+E18VA  & VA  & PE         & $150$ bp    & $36\times$    \\
+E18VAT & VAT & PE         & $101$ bp    & $26\times$    \\
+E23BUF & BUF & PE         & $101$ bp    & $12\times$    \\
+E23SA  & SA  & PE         & $101$ bp    & $29\times$    \\
+E23VA  & VA  & PE         & $101$ bp    & $32\times$    \\
+E23VAT & VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
 \end{tabulary}
@@
 The reported read lengths are prior to any trimming of the reads.
-All samples were first published in \citet{Seisenberger:2012ko}.}
+All samples were first published in \citet{Seisenberger:2012ko}.
+Abbreviations: \emph{PE} = paired-end sequencing}
 \label{tab:Seisenberger_dataset}
@@
 The reported read lengths are prior to any trimming of the reads.
-All samples were first published in \citet{Ziller:2013fs}.}
+All samples were first published in \citet{Ziller:2013fs}.
+Abbreviations: \emph{PE} = paired-end sequencing}
 \label{tab:Ziller_dataset_sequencing_runs}

### 2015-08-24 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Change some tables to figures (closes #14).

@@
 
-The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its copy number.
-The running time is largely independent of the tuple size (`-m`).
-The variation in running times within a chromosome is due to the number of reads generated per sample and the length of the reads, where the length of a paired-end read is defined as the sum of the mates' lengths.
-Samples with more reads take longer to process and samples sequenced with longer reads take longer because these contain more m-tuples.
-
-The maximum memory usage is not strictly proportional to chromosome length.
-It is instead driven by the number and density of CpGs on the chromosome.
-For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Table \ref{tab:cpg_density_hg19}).
-The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
-Memory usage is therefore relatively constant across values of `-m` for a given chromosome.
-The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`.
-This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.
-
 \begin{figure}[!htb]
@@
 
+The running time of `methtuple` is proportional to the number of reads mapped to the chromosome, which is proportional to the length of the chromosome and its copy number.
+The running time is largely independent of the tuple size (`-m`).
+The variation in running times within a chromosome is due to the number of reads generated per sample and the length of the reads, where the length of a paired-end read is defined as the sum of the mates' lengths.
+Samples with more reads take longer to process and samples sequenced with longer reads take longer because these contain more m-tuples.
+
+The maximum memory usage is not strictly proportional to chromosome length.
+It is instead driven by the number and density of CpGs on the chromosome.
+For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Figure \ref{fig:cpg_density_hg19}).
+The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain.
+Memory usage is therefore relatively constant across values of `-m` for a given chromosome.
+The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`.
+This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/CpG_density_hg19.pdf}
+\caption[CpG percentage per chromosome in hg19]{Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).}
+\label{fig:cpg_density_hg19}
+\end{figure}
+
 \FloatBarrier
-
-```{r autosomal_cpg_density_hg19, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-library(xtable)
-x <- read.table("../data/wgbs_bioinformatics_analysis/CpG_density_hg19.txt",
-                header = TRUE, stringsAsFactors = FALSE)
-xt <- xtable(x, caption = c("Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).", "CpG percentage per chromosome in hg19"), label = "tab:cpg_density_hg19")
-align(xt) <- rep("c", ncol(x) + 1)
-digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, include.rownames = FALSE, caption.placement = "top")
-```
 

### 2015-08-24 Rmarkdown/wgbs_statistical_framework.Rmd
Change some tables to figures (closes #14).

@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
-\caption[\emph{EPISCOPE} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
+\caption[\emph{EPISCOPE}: $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
-\caption[\emph{Lister} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
+\caption[\emph{Lister}: $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
-\caption[\emph{Seisenberger} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
+\caption[\emph{Seisenberger}: $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
-\caption[\emph{Ziller} $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
+\caption[\emph{Ziller}: $\beta$-value kernel density]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
 Densities are normalised so that the maximum value for each sample is 1.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
-\caption[\emph{EPISCOPE} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
+\caption[\emph{EPISCOPE}: $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
-\caption[\emph{Lister} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
+\caption[\emph{Lister}: $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
-\caption[\emph{Seisenberger} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
+\caption[\emph{Seisenberger}: $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
-\caption[\emph{Ziller} $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
+\caption[\emph{Ziller}: $\beta$-value frequency polygon]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
-\caption[\emph{EPISCOPE} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{EPISCOPE}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
-\caption[\emph{Lister} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Lister}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
-\caption[\emph{Seisenberger} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Seisenberger}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
-\caption[\emph{Ziller} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Ziller}: $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
 Only CpGs with at least $10 \times$ sequencing coverage are included.
@@
 \includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
-\caption[\emph{EPISCOPE} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{EPISCOPE}: $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{EPISCOPE} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
-\caption[\emph{Lister} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Lister}: $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Lister} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
-\caption[\emph{Seisenberger} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Seisenberger}: $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Seisenberger} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 \includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
-\caption[\emph{Ziller} $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
+\caption[\emph{Ziller}: $\beta$-value frequency polygon stratified by CpG islands]{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the \emph{Ziller} data, stratified by whether the CpG is in a CpG island.
 $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis.
@@
 To investigate the validity of this assumption we can compute the correlation of strand-specific $\beta$-values.
-Tables \ref{tab:EPISCOPE_strand_correlations}, \ref{tab:Lister_strand_correlations}, \ref{tab:Seisenberger_strand_correlations} and \ref{tab:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
+Figures \ref{fig:EPISCOPE_strand_correlations}, \ref{fig:Lister_strand_correlations}, \ref{fig:Seisenberger_strand_correlations} and \ref{fig:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.
 
-```{r EPISCOPE_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-library(xtable)
-x <- read.table("../data/strand_correlation/EPISCOPE_strand_correlations.txt",
-stringsAsFactors = FALSE)
-colnames(x) <- 1:10
-x <- x[, 1:5]
-xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{EPISCOPE}: Correlations of $\\beta$-values across strands"), label = "tab:EPISCOPE_strand_correlations")
-align(xt) <- c("r", rep("c", ncol(x)))
-digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = "top")
-```
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_strand_correlations.pdf}
+\caption[\emph{EPISCOPE}: Correlations of $\beta$-values across strands]{Correlations of $\beta$-values across strands for the \emph{EPISCOPE} dataset using different minimum sequencing coverage cutoffs.}
+\label{fig:EPISCOPE_strand_correlations}
+\end{figure}
 
-```{r Lister_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-x <- read.table("../data/strand_correlation/Lister_strand_correlations.txt",
-stringsAsFactors = FALSE)
-colnames(x) <- 1:10
-x <- x[, 1:5]
-xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Lister}: Correlations of $\\beta$-values across strands"), label = "tab:Lister_strand_correlations")
-align(xt) <- c("r", rep("c", ncol(x)))
-digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
-```
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Lister_strand_correlations.pdf}
+\caption[\emph{Lister}: Correlations of $\beta$-values across strands]{Correlations of $\beta$-values across strands for the \emph{Lister} dataset using different minimum sequencing coverage cutoffs.}
+\label{fig:Lister_strand_correlations}
+\end{figure}
 
-```{r Seisenberger_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-x <- read.table("../data/strand_correlation/Seisenberger_strand_correlations.txt",
-stringsAsFactors = FALSE)
-colnames(x) <- 1:10
-x <- x[, 1:5]
-xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Seisenberger}: Correlations of $\\beta$-values across strands"), label = "tab:Seisenberger_strand_correlations")
-align(xt) <- c("r", rep("c", ncol(x)))
-digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
-```
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Seisenberger_strand_correlations.pdf}
+\caption[\emph{Seisenberger}: Correlations of $\beta$-values across strands]{Correlations of $\beta$-values across strands for the \emph{Seisenberger} dataset using different minimum sequencing coverage cutoffs.}
+\label{fig:Seisenberger_strand_correlations}
+\end{figure}
 
-```{r Ziller_merged_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
-x <- read.table("../data/strand_correlation/Ziller_merged_strand_correlations.txt",
-stringsAsFactors = FALSE)
-colnames(x) <- 1:10
-x <- x[, 1:5]
-xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset.
-Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Ziller}: Correlations of $\\beta$-values across strands"), label = "tab:Ziller_merged_strand_correlations")
-align(xt) <- c("r", rep("c", ncol(x)))
-digits(xt) <- 2
-print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
-```
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_strand_correlations.pdf}
+\caption[\emph{Ziller: Correlations of $\beta$-values across strands}]{Correlations of $\beta$-values across strands for the \emph{Ziller} dataset using different minimum sequencing coverage cutoffs.}
+\label{fig:Ziller_merged_strand_correlations}
+\end{figure}
 

### 2015-09-15 Rmarkdown/comethylation.Rmd
Update some figure filenames in light of 299242fa79429461861f1b8eb8dc5bf0f66bd028

@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_1.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_1-1.pdf}
 \caption[Simulation 1a]{Results of simulation 1a ($K = 100$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_2.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_2-1.pdf}
 \caption[Simulation 1b]{Results of simulation 1b ($K = 1000$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_2d-1.pdf}
 \caption[Simulation 1c]{Results of simulation 1c ($K = 1000$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_3-1.pdf}
 \caption[Simulation 2]{Results of simulation 2 ($K = 100$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/beta_marginals.pdf}
+\includegraphics[width=\textwidth]{../figures/beta_marginals-1.pdf}
 \caption[Histogram of $Beta(0.3, 0.2)$ random variables]{Histogram of 1000 points simulated from a $Beta(0.3, 0.2)$ distribution.}
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_4.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_4-1.pdf}
 \caption[Simulation 3]{Results of simulation 3 ($K = 100$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_5.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_5-1.pdf}
 \caption[Simulation 4a]{Results of simulation 4a ($K = 100$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_6.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_6-1.pdf}
 \caption[Simulation 4b]{Results of simulation 4b ($K = 1000$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_7.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_7-1.pdf}
 \caption[Simulation 5]{Results of simulation 5 ($K = 100$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Simulation_8.pdf}
+\includegraphics[width=\textwidth]{../figures/Simulation_8-1.pdf}
 \caption[Simulation 6]{Results of simulation 5 ($K = 100$).
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_seqnames-1.pdf}
 \caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Lister_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/Lister_MH_by_seqnames-1.pdf}
 \caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_seqnames-1.pdf}
 \caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames-1.pdf}
 \caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames-1.pdf}
 \caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames-1.pdf}
 \caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_seqnames-1.pdf}
 \caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames-1.pdf}
 \caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames_chr21.pdf}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames_chr21-1.pdf}
 \caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames_chr21.pdf}
+\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames_chr21-1.pdf}
 \caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames_chr21.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames_chr21-1.pdf}
 \caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_MH_by_CGI-1.pdf}
 \caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Lister_MH_by_CGI-1.pdf}
 \caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Seisenberger_MH_by_CGI-1.pdf}
 \caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI-1.pdf}
 \caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_CGI-1.pdf}
 \caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_CGI-1.pdf}
 \caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Seisenberger_ac_MH_by_CGI-1.pdf}
 \caption[\emph{Seisenberger}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Seisenberger} dataset using $NIL \geq 0$ pairs.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI.pdf}
+\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI-1.pdf}
 \caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs.

### 2015-09-15 Rmarkdown/comethylation_review.Rmd
Closes #19

@@
 It highlights various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}.
-We also review how methods to detect differential methylation have sought to account for, and exploit, co-methylation.
+We also review how methods to detect differential methylation have sought to account for, and leverage, co-methylation.
 \end{chapabstract}
@@
 There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data.
-If co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis methods.
+If co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and leverage these dependencies to create more powerful and efficient analysis methods.
 
@@
 CpGs within a neighborhood of 2 bp to 100 bp are thus more likely to share the same methylation status than more distant CpGs.
-(C) Fourrier transform of autocorrelation showing a clear periodicity peak at 33 cycles per 100 bp (every 3 bp).
+(C) Fourrier [sic] transform of autocorrelation showing a clear periodicity peak at 33 cycles per 100 bp (every 3 bp).
 (D) Distribution of codon position of mCs, and distribution of methylation level depending on the position.
@@
 All of this casts serious doubt on the validity and interpretation of the co-methylation results in \citet{Lyko:2010dr}.
+
+\FloatBarrier
 
@@
 
-In order to exploit between-sample co-methylation in this way, we also need to know the size of these co-methylation blocks, that is, the length of the regions over which between-sample co-methylation is sufficiently strong.
+In order to leverage between-sample co-methylation in this way, we also need to know the size of these co-methylation blocks, that is, the length of the regions over which between-sample co-methylation is sufficiently strong.
 The longer these blocks are, the more sparsely we can afford to sample individual methylomes.
@@
 
-\section{Exploiting co-methylation in downstream analyses}
-\label{sec:exploiting_cometh}
+\section{Leveraging co-methylation in downstream analyses}
+\label{sec:levaraging_cometh}
 
 The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing the proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses.
-Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data.
+Ideally, co-methylation would not only be _accounted for_, but _leveraged_, in downstream analyses of DNA methylation data.
 In fact, some authors have already sought to do this.
 
-For example, the R/Bioconductor packages `bsseq` \citep{Hansen:2011gu, Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} both cite \citet{Eckhardt:2006gh} as providing evidence of co-methylation, which can be exploited by spatial smoothing of the $\beta$-values prior to testing for differential methylation.
+For example, the R/Bioconductor packages `bsseq` \citep{Hansen:2011gu, Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} both cite \citet{Eckhardt:2006gh} as providing evidence of co-methylation, which can be leveraged by spatial smoothing of the $\beta$-values prior to testing for differential methylation.
 Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed $\beta$-values are, on average, more accurate than the raw $\beta$-values.
@@
 
-These methods exploiting correlations amongst the test statistics or their P-values are not directly based on co-methylation estimates.
+These methods leveraging correlations amongst the test statistics or their P-values are not directly based on co-methylation estimates.
 However, what drives the correlations amongst the test statistics is a complex mixture of within-sample and between-sample co-methylation.
@@
 
-It is clearly inappropriate to treat methylation at nearby loci as independent and methods have been proposed to account for or exploit this in various ways.
+It is clearly inappropriate to treat methylation at nearby loci as independent and methods have been proposed to account for or leverage this in various ways.
 These methods would be better informed by a more rigorous and reproducible analysis of co-methylation, which I give in Chapter \ref{chap:co-methylation}.

### 2015-09-15 Rmarkdown/methsim.Rmd
Closes #19

@@
 
-1. Realism: The simulated data must be 'similar' to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes 'similar enough'.
+1. Realism: The simulated data must be 'similar' to the real data. While this is obvious, it is also often hard to clearly define or agree upon what constitutes 'similar enough'.
 2.
@@
 \begin{description}
-  \item[\emph{I}] The sample is 'pure' and all haplotypes have the same parameters.
+  \item[\emph{I}] The sample is `pure' and all haplotypes have the same parameters.
   \item[\emph{II}] The sample is a mixture of subpopulations with different parameters. We will consider a sample with four subpopulations with relative frequencies equal to $(0.6, 0.25, 0.1, 0.05)$.

### 2015-09-15 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Closes #19

@@
 While 1-tuples are the basis of most analyses of bisulfite-sequencing data, they do not always give the complete picture of how DNA methylation is acting in the sample.
-To gain a clearer picture, we can exploit the fact that many bisulfite-sequencing reads contain multiple methylation loci (_m-tuples_) and that each read is from a single cell[^chimeric_reads].
+To gain a clearer picture, we can leverage the fact that many bisulfite-sequencing reads contain multiple methylation loci (_m-tuples_) and that each read is from a single cell[^chimeric_reads].
 An example of where this is useful is shown in Figure \ref{fig:incomplete_picture_1-tuples} where we have two regions, each with four methylation loci, that have identical methylation calls at 1-tuples yet very different overall methylation patterns.

### 2015-09-15 Rmarkdown/wgbs_statistical_framework.Rmd
Closes #19

@@
 In each case this smoother is chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | d_{i, j} \eqd Binomial(d_{i, j}, B_{i, j})$.
-The smoothing is 'local' to exploit co-methylation, which is considered a local phenomenon.
+The smoothing is 'local' to leverage co-methylation, which is considered a local phenomenon.
 
@@
 All three _Seisenberger_ samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell, _J1\_1_.
-However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.
+However, since these data are from pooled DNA, the source of this reduced correlation is difficult to identify.
 

### 2015-09-17 Rmarkdown/biology_background.Rmd
Closes #19
Closes #20.

@@
 Environmental variation has clear influence on certain phenotypes.
-However, it often can be difficult to determine whether phenotypic variation is driven by genetic varation or environment variation, particularly in humans where genetically similar individuals typically also grow up in similar environments.
+However, it often can be difficult to determine whether phenotypic variation is driven by genetic variation or environment variation, particularly in humans where genetically similar individuals typically also grow up in similar environments.
 
@@
 But there is also phenotypic variation within the individual.
-If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell) and a cone cell (a photoreceptor in the retina) all have identical genomes.
-In fact, all cells in an organism, excluding the gametes, have an identical genome[^somatic_mutations] yet play very different biological roles.
+If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell), and a cone cell (a photoreceptor in the retina) all have identical genomes.
+In fact, all cells in an organism, excluding the gametes, have an identical genome[^somatic_mutations], yet play very different biological roles.
 This is due to different genes being active in different cells.
 
-[^somatic_mutations]: This ignores somatic mutations.
+[^somatic_mutations]: Even this is a simplification.
+For example, it ignores somatic mutations (the occurrence of a mutation in the somatic tissue of an organism, resulting in a genetically mosaic individual) and V(D)J recombination (which occurs in lymphocytes and is vital for antibody diversity).
 
@@
 
-Conrad Waddington coined the phrase in 1942 as a portmanteau of the words 'epigenesis' and 'genetics' \citep{Waddington:2012bu}.
-Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \citep{Waddington:2012bu}.
+Conrad Waddington coined the phrase in 1942 as a portmanteau of the words 'epigenesis' and 'genetics' \citep{Waddington2012-bq}.
+Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \citep{Waddington2012-bq}.
 
@@
 Nevertheless, the copying of DNA methylation is faithful enough for most biologists to consider it as a mitotically heritable mark, most of the time.
+The enzymes responsible the replication of DNA methylation, the DNA methyltransferases, are discussed in Section \ref{sec:writers_reads_and_erasers}.
 
@@
 
-\subsection{Writers, readers and erasers}\label{sec:writers_reads_and_erasers}
+\subsection{Writers, readers, and erasers}\label{sec:writers_reads_and_erasers}
 
@@
 
-Active demethylation is currently a hot topic in epigenetics research.
+Active demethylation is currently an active area in epigenetics research.
 Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation.
@@
 The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome.
-Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine.
+Because the read is compared against the reference genome, it may be incorrectly inferred to be an unmethylated cytosine.
 The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome.
-Because the read is compared against the reference genome, it may incorrectly inferred to be a thymine.}
+Because the read is compared against the reference genome, it may be incorrectly inferred to be a thymine.}
 \label{fig:bisulfite_CT_SNP}
@@
 
-The bisulfite-treatment of DNA can introduce biases and other problems.
-Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_.
+The bisulfite-treatment of DNA can introduce biases and other problems \citep{Warnecke2002-zl}.
+Four examples are _PCR-bias_, _incomplete bisulfite-conversion_, _bisulfite over-conversion_, and _DNA degradation_.
 PCR-bias is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \citep{Warnecke:1997eh}.
 Incomplete bisulfite-conversion leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment.
+Conversely, bisulfite over-conversion results in a methylated cytosine incorrectly being inferred to be an unmethylated cytosine, although this is uncommon than incomplete bisulfite-conversion \citep{Warnecke2002-zl}.
 DNA degredation occurs because sodium bisulfite damages DNA, resulting in the fragmentation of long molecules.
-This limits the size of the fragments that can be studied using bisulfite-conversion assays to approximately $500$ bp \citep{Ecker:2010et}
+This limits the size of the fragments that can be studied using bisulfite-conversion assays to approximately $500$ bp \citep{Ecker:2010et}.
+The influence of these biases can depend on the experimental setup and their potential effects should be born in mind when interpreting results.
 
@@
 It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a base, such as 5mC, to the time it takes to read its unmodified form.
+This does not require that the DNA is bisulfite-converted prior to sequencing.
+Because the DNA does not undergo bisulfite-conversion (nor the attendant short fragmentation of the DNA), it is in theory possible to analyse DNA methylation from individual, long DNA molecules using Pacific Biosciences SMRT technology[^pacbio].
+However, due to the error rate and cost of SMRT sequencing, it is currently all but unfeasible to study genome-wide DNA methylation in mammalian-sized genomes.
+
+[^pacbio]: While bisulfite-treated DNA could be sequenced using SMRT sequencing, this would eliminate the real advantage of the technology, namely, the long reads generated by this sequencer.
+
+The error rate of Pacific Biosciences SMRT sequencing is currently higher than that of Illumina sequencing.
+This means it is more difficult to make reliable inferences on DNA methylation from individual reads.
+Given sufficient sequencing coverage, it would be possible to reliably estimate the average level of methylation at a given cytosine, however, the cost of SMRT sequencing all but prohibits high-coverage sequencing of mammalian-sized genomes.
 

### 2015-09-24 Rmarkdown/comethylation_review.Rmd
Closes #16

@@
 
+\FloatBarrier
+
 \subsection{\cite{Lyko:2010dr}}\label{sec:lyko}
@@
 They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away".
-This analysis, however, is fundamentally flawed.
+This analysis, however, is potentially flawed.
 
@@
 Note that the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}).
-I strongly suspect that the autocorrelation has been incorrectly computed.
+I suspect that the autocorrelation has been incorrectly computed.
 Specifically, what I believe has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been artificially assigned a $\beta$-value of zero.
 It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation.
-I do not have the data used in \citet{Lyko:2010dr}, but the following example supports my conclusion.
-This example uses the chromosome 22 data from the _FF_ sample in the _Lister_ dataset (see Section \ref{sec:Lister_dataset} for full details of this sample).
 
@@
 
+I do not have the data used in \citet{Lyko:2010dr}, but the following example supports my belief.
+This example uses the chromosome 22 data from the _FF_ sample in the _Lister_ dataset (see Section \ref{sec:Lister_dataset} for full details of this sample).
+
 Let $\bm{\beta^{\ast}}$ be the augmented vector of $\beta$-values, where:
@@
 Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$ for the chromosome 22 data from the _FF_ sample.
-We see that it is remarkably similar to that of panel (a) in Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum correlation is very small, close to zero.
+We see that it is remarkably similar to that of panel (A) in Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum autocorrelation is very small, close to zero.
 
@@
 
-Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}).
-Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of CpGs in the genome than anything to do with their methylation state.
+On the one hand, this suggests that this may be the cause of the negligible autocorrelations in Figure S7 of \citet{Lyko:2010dr}.
+On the other hand, supposing that the maximum autocorrelation of the $\beta$-values is truly less than $0.015$, then this suggests that co-methylation itself is negligible in honeybees, in contrast to results from other organisms.
 
-\begin{figure}[!htb]
-\centering
-\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
-\caption[Fourier analysis of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample.
-Fourier transformation computed and plotted using the \texttt{spectrum} function in R.
-The period is defined as 1 / frequency.
-The maximum frequency is approximately $0.33$, which corresponds to a period of approximately $3$ bp.}
-\label{fig:FF_chr21_acf_beta_star_fourier}
-\end{figure}
-
-All of this casts serious doubt on the validity and interpretation of the co-methylation results in \citet{Lyko:2010dr}.
+The effort required to process the Lyko dataset, and the insufficient detail in the paper to facilitate the reproducibility of the analysis, mean that I decided not to invest further effort into exploring potential interpretations of the co-methylation analysis of \citet{Lyko:2010dr}.
 

### 2015-09-26 Rmarkdown/datasets.Rmd
Closes #22
Closes #21

@@
 CpG 2-tuples were called both with and without the `--all-combinations` flag; all other tuple sizes were called without the `--all-combinations` flag.
+The following `methtuple` flags were also used: `--methylation-type CG --ignore-duplicates --min-mapq 0 --overlap-filter XM_ol --ignore-duplicates`.
 4.
@@
 
-There are a four 'mini datasets' within the Lister data that I make some use of in my thesis.
+There are four 'mini datasets' within the Lister data that I make some use of in my thesis.
 The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper.

### 2015-09-26 Rmarkdown/methsim.Rmd
Closes #21

@@
 
-Almost all downstream analyses are reference-based (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.
+Almost all downstream analyses are reference-based (see Section \ref{sec:methylation_calling}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.
 

### 2015-09-26 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Closes #21

@@
 
-\subsection{Methylated or unmethylated?}\label{sec:methylated_or_unmethylated}
-
 All bisulfite-sequencing assays use _reference-based_ methylation calling.
@@
 For example, unless the genome of the sample is fully known, methylation loci in insertions cannot be distinguished from genetic variation since there is no reference sequence to compare them against.
+
+\subsection{Considerations}
+
+There are several issues that must be carefully considered when performing methylation calling, including filtering of reads and biases, choosing and refining the reference sequence, and determining the context or methylation type of the cytosine.
+
+\subsubsection{Filtering of reads and bases}
 
@@
 
+Although incomplete bisulfite-conversion is a well-recognised issue, most analysis pipelines don't attempt to account for this during methylation calling[^incomplete-conversion].
+An exception is `Bis-SNP` \citep{Liu:2012ge} which has an algorithm to exclude bases suspected of suffering from incomplete bisulfite-conversion at the 5' end of Illumina-generated reads.
+
+[^incomplete-conversion]: Instead, incomplete bisulfite-conversion is usually incorporated into calculations to estimate the methylation level at a given cytosine (see Section \ref{sec:beta-value_interpretation}).
+The bisulfite-conversion rate may be estimated by analysing cytosines that are expected to be unmethylated, such as those in the chloroplast genome \citep{Lister:2008bh} or a spike-in control of lambda phage DNA \citep{Lister:2009hy}.
+
+\subsubsection{Choosing and refining the reference sequence}
+
 The reference sequence is typically the reference genome used in the alignment step, in spite of the obvious differences between the sample's genome and the reference genome.
@@
 
-- Whole-genome sequencing or genotyping of the sample
-- Calling genetic variations directly from the bisulfite-sequencing data
-- Excluding sites of known genetic variation
-
-\subsubsection{Whole-genome sequencing or genotyping of the sample}\label{sec:wgs_or_genotyping}
+1. Whole-genome sequencing or genotyping of the sample
+2. Calling genetic variations directly from the bisulfite-sequencing data
+3. Excluding sites of known genetic variation
 
@@
 However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.
-
-\subsubsection{Calling genetic variations directly from the bisulfite-sequencing data}\label{sec:wgbs_variant_calling}
 
@@
 
-\subsubsection{Excluding sites of known genetic variation}\label{sec:known_genetic_variation}
-
 The third approach, and arguably the bare minimum, is to call methylation events using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population.
@@
 
-\subsection{Determining the context or methylation type}\label{sec:methylation_context}
+\subsubsection{Determining the context or methylation type}
 
@@
 There are many useful options to filter reads and read-positions.
-Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting the bases in the overlapping region.
+Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting the bases in the overlapping region[^overlapping-mates].
 `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads.
@@
 Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.
+
+[^overlapping-mates]: `methtuple` has several options for handling overlapping mates of paired-end reads via the `--overlap-filter` flag.
+The default method is to ignore any read-positions in the overlapping region where the methylation calls disagree.
 
@@
 I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for more than $40$ whole-genome bisulfite-sequencing samples.
-Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets.
+Figure \ref{fig:methtuple_runtime} shows the distribution of running times, Figure \ref{fig:methtuple_memory} the maximum memory usage across, and Figure \ref{fig:methtuple_file_sizes} the output file sizes, for all the samples from the EPISCOPE, Lister, and Ziller datasets.
 For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).
@@
 The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set.
-Note that the total number of samples is $49$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
+Note that the total number of samples is $48$ because each of the Ziller sequencing runs is separately counted (see Chapter \ref{chap:datasets} for details).}
 \label{fig:methtuple_runtime}
@@
 The suffix `ac' means that the option \texttt{{-}{-}all-combinations} was set.
-Note that the total number of samples is $49$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
+Note that the total number of samples is $48$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
 \label{fig:methtuple_memory}
+\end{figure}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=\textwidth]{../figures/methtuple_file_sizes.pdf}
+\caption[\texttt{methtuple} file sizes]{The sizes of the output file when compressed with gzip.
+The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set.
+Note that the total number of samples is $48$ because each of the Ziller sequencing runs is separately counted (see Chapter \ref{chap:datasets} for details).}
+\label{fig:methtuple_file_sizes}
 \end{figure}
@@
 This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.
+
+The regular structure of the output file means that these are particularly compressible. The size of the output file is almost always less than 1 GB when compressed with `gzip`.
 

### 2015-09-28 Rmarkdown/wgbs_statistical_framework.Rmd
Closes #23

@@
 _read-1_ gives us $(i, i + 1)$ and _read-2_ gives us $(i + 1, i + 2, i + 3)$.
-In this example we must use only one of _read-1_ or _read-2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.
+In this example we must use only one of _read-1_ or _read-2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'[^overlapping-mates].
+
+[^overlapping-mates]: We should also check that the overlapping bases agree and, if not, either use the call with the highest base quality or ignore these overlapping positions in both reads.
+`methtuple` has several options for handling overlapping mates of paired-end reads via the `--overlap-filter` flag.
+The default method is to ignore any read-positions in the overlapping region where the methylation calls disagree.
 

### 2015-09-28 Rmarkdown/comethylation.Rmd
Closes #26

@@
 
-In summary, the correlations of pairs of $\beta$-values should be cautiously interpreted, bearing in mind that these are somewhat crude summaries that hide many details of the underyling scatterplots.
+In summary, the correlations of pairs of $\beta$-values should be cautiously interpreted, bearing in mind that these are somewhat crude summaries that hide many details of the underlying scatterplots.
 
@@
 \FloatBarrier
-\subsection{Results}
+\subsection{Results}\label{sec:results-correlation-of-methylation-levels}
 
@@
 
-\subsubsection{Can we collapse by strand?}
-
-Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand.
-Since the trend in the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.
+\subsubsection{The effect of strand}
+
+We have seen that the CpG $\beta$-values are very highly correlated across strands for most samples when the minimum sequencing coverage is at least $5\times$ ($r = 0.8$ to $0.9$, Section \ref{sec:empirical-distributions-of-beta-values}).
+While there are some notable exceptions (the embryo-derived and embryonic stem cell samples), I have opted to collapse by strand in most of the analyses in Section \ref{sec:results-correlation-of-methylation-levels} in order to simplify the plots.
+
+Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} are the only plots to stratify by strand and show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs.
+These correlations are very similar across strands, even for those embyro-derived and embryonic stem cell samples.
 

### 2015-09-28 examination/response-to-examiners.Rmd
Added response to examiners

---
title: "Response to examiners"
author: "Peter Hickey"
date: "28 September 2015"
output: pdf_document
---

Dear Chair of Examiners,

Attached are my responses to the examiners' comments on my PhD thesis. The reviewer's comment is in __bold__ and followed by my response, broken down by reviewer and section.

Regards,

Peter Hickey

\newpage

## Reviewer #1

### Chapter 1

__p6, section 1.1.4, 3rd paragraph: "varation" should be "variation"__

Typo fixed.

__p6, section 1.1.4, 4th paragraph: it is stated that all cells in an organism have identical genomes. Since this thesis focuses on mammalian genomes and most of the datasets were produced in a biomedical context, it should be noted that this is not true for lymphocytes, which undergo [somatic recombination](https://en.wikipedia.org/wiki/V(D)J_recombination).__

Noted in footnote importance of somatic recombination in lymphocytes and that 'all cells in a body have identical genome' is a (useful) simplification. Thank you.

__p7, section 1.2, 2nd paragraph: it is surprising to find a reference to Waddington, 2012 since Waddington died in 1975. Further in this section, the difficulty of defining epigenetics is discussed. If the candidate is interested in reading more about this controversial topic, I would recommend to have a look at the minimalist approach of Mark Ptashne, for instance, [Mark Ptashne, 7101-7103, doi:10.1073/pnas/1305399110](https://paperpile.com/shared/rFMo5e)__

The reference to Waddington (2012) is correct, it refers to a re-print of an article published in 1942 (that is no longer available or difficult to obtain). Thank you for the recommended paper.

__p8, section 1.3, 1st paragraph: it is stated that DNA methylation is heritable, but the mechanisms (maintenance DNMT) are not mentioned until 1.3.6. It would be useful to add a reference to that section here.__

Reference added.

__p27, Figure 1.7: "It may incorrectly inferred" should be "it may be incorrectly inferred".__

Typo fixed.

__p27, last paragraph discusses the various issues associated with bisulfite conversion. One rarely discussed issue that is also ignored here is over-conversion. It would be informative to have a sentence or two saying why this was not considered in this thesis. Also, an important reference that was omitted form this section is [Warnecke, Peter M, Clare Stirzaker, Jenny Song, Christoph Grunau, John R Melki, and Susan J Clark. 2002. “Identification and Resolution of Artifacts in Bisulfite Sequencing.” Methods  27 (2): 101–7.](https://paperpile.com/shared/69shJW)__

Thank you for the excellent reference. I included a sentence describing bisulfite over-conversion and referenced Warnecke et al. (2002) in support that this is less common than incomplete bisulfite-conversion, hence why I have not considered this in my thesis.

__p28, section 1.4.4: here and in a few other section (for instance p33) in this thesis, DNA kinetic assays are presented as an alternative to bisulfite sequencing. I think it would have been important to mention the important limitations of this technique:__
  - __A reliable assessment of the methylation status of a base requires a fairly high coverage (at least 50x from what I hear from the PacBio reps themselves, so probably even more in practice). As a consequence, this would only allow to estimate $\beta$, but not within fragment co-methylation.__
  - __If bisulfite conversion was followed by SMRT sequencing, the fragment size would be limited by the bisulfite treatment itself to a few hundred base pairs__

I have expanded upon the limitations of using PacBio to study DNA methylation in mammalian-sized genomes.

__Finally, although this is admittedly not the focus of the thesis, it would have been useful to mention some of the functions of DNA methylation, at least superficially at the molecular level: inhibition of transcription factor binding in CpG islands around promoters, transposon repression, control of alternative splicing, etc.__

There is an (albeit brief) paragraph on p9 discussing reasons people study DNA methylation (i.e., its biological roles). This includes references to some key papers. I have opted not to expand this section, since, as the examiner notes, this is not the focus of the thesis.

### Chapter 2

__Section 2.2.2: The M-bias is well explained, and a strategy to avoid this problem is presented. However, it would be useful to examine the effect of filtering by quality or neighbour quality standard (NQS) on this bias. I have observed some datasets where doing methylation calls based on NQS virtually remove the M-bias without having to systematically eliminate some positions. Looking at this in detail is obviously a big task, but it might be worth investigating in the future.__

This is an interesting strategy of which I was unaware. As noted by the examiner, this is a considerable undertaking. Furthermore, the largest dataset available to me, the _Lister_ dataset, does not have base-quality scores. For these reasons I have opted not to pursue this avenue in my thesis. I agree that it will be important to consider this in a more focused analysis of M-bias, such as a paper addressing this issue.

__Section 2.3.1: "Methylated or unmethylated" is not a good title as this section is more about which positions and which reads to use__

 Re-named this section "Considerations" and merged with Section 2.3.2 "Determining the context or methylation type".

__Section 2.4.4: In addition to the memory usage and running time it would be useful to know how big is the output of the program, especially with and without the `--all-combination` option.__

Added a plot of output file sizes.

__This chapter could have been improved by a discussion of how to handle overlapping reads, which is not presented until chapter 4, section 4.1.2.__

Added a brief description of how overlapping mates of paired-reads are handled by `methtuple`.

__Also, since this chapter is in great part a review of how to handle whole genome bisulfite sequencing data, it would be useful to discuss how to estimate and deal with incomplete conversion.__

Added a paragraph to section 2.3.1 on how incomplete bisulfite-conversion is handled in practice.

### Chapter 3

__Page 64, 1st line, "There are a four" should be "There are four".__

Fixed.

### Chapter 4

__p76, point 2: it is said that one must use read 1 or read 2 in the region where they overlap. Ideally, one would actually check that they agree, and if not, either use the call with the highest quality, or ignore these reads at that position.__

I have clarified this point and noted how `methtuple` handles such overlapping mates.

__p76, last equation, missing a "|" at the end.__

Fixed, thank you.

__p83, last paragraph: "2000kb" should be "2000bp"__

Fixed, thank you.

__Section 4.4.1: The beta-values have a bimodal distribution in most samples, with a peak around 1 and a peak around 0. It is unclear to me whether the peak at 0 represents a true biological phenomenon, or if this is simply the effect of not taking incomplete bisulfite conversion into account.__

I believe this is a true biological phenomenon. All figures in this section require use a minimum sequencing coverage of 10x. We can compute the probability of observing 10 Cs from 10 reads under the assumption that all Cs are due to incomplete conversion. Let X = the number of Cs, p = the incomplete bisulfite-conversion rate = 0.01 (conversion rates are typically > 99%). We are interested in Pr(X = 10 ) = 0.01^10 = 10^-20. This makes it unlikely these peaks are 0 are simply due to incomplete bisulfite-conversion.

__p108: it is stated that "it does not make sense to refer to individual cytosines as methylated". I think that it can actually be useful to filter out positions that are not methylated and where the signal is driven by incomplete conversion, the question then becomes to identify methylated [PH: cytosines] that are never methylated, which is really the same thing as defining methylated [PH: cytosines] that are methylated (at least in some cell subpopulation of the sample).__

While I understand this line of reasoning, as I argue in my thesis, I believe this terminology can be an unnecessary source of confusion.

### Chapter 5

__If the _MethylationTuples_ R package has any visualisation capabilities, it would be nice to demonstrate them here.__

At the time of submitting my PhD thesis, _MethylationTuples_ did not have any substantial visualisation capabilities. These will be added in due course.

### Chapter 6

__This chapter raises some very important points on the study of correlation in bisulfite sequencing datasets. This includes some fundamental issues with the interpretation of the signal that might come from the underlying sequencing composition rather than the actual methylation pattern. It also raises important issues regarding availability of software implementation. I would encourage the candidate not to give up but to be persistent when requesting software or data from authors, as lack of response or an incomplete answer is more often caused by shortage of time rather than unwillingness to help.__

Noted in agreement.

### Chapter 7

__p163, section 7.1.1. A bs-SNP corrected reference was used. It would be useful to know the impact of not doing this correction. If it turns out that it has little effect, similar studies in the future could be more streamlined.__

I have not attempted to quantify the effect of using a `Bis-SNP`-corrected genome. CpG-SNPs are a well-documented source of bias when estimating methylation levels, as discussed in the Section 2.3.1. Since it is not difficult to address this bias, I do not see great benefit in ignoring it in favour of a more streamlined analysis.

__Section 7.1.2, "Can we collapse the strands". In figure 7.3, 7.4, 7.5 and 7.6, the trend is similar for the two strands, but this does not necessarily mean that at a given locus the two strands have a similar value for $\beta$. Consider for the following example: +strand, pos1: 1M/10U, pos2: 10M/100U; -strand, pos1: 50M/100U, pos2: 5M/10U. Here $\beta$ at the two positions is identical if each strand is taken independently, which would result in a high correlation. However, if the two strands are merged, we would get: pos1: 51M/110U, pos2: 15M/110U, and thus correlation would be low. I might be missing something but it seems to me that scatter plot of + and - strands for $\beta$ would have been more appropriate here.__

Yes, you are correct. I had erroneously conflated two concepts: (1) the consistency of beta-values on opposing strands at a given cytosine (2) the consistency of the correlations-against-IPD trend for beta-values on each strand. It is (1) that we really care about when addressing the question of "can we collapse by strand". Since I addressed in Section 4.4.1 I have re-written this section (re-titled "The effect of strand") to distinguish the two concepts and fix my mistake.

__In this chapter, the genomic features that is used to aggregate to data are: chromosome, and CpG islands vs the rest of the genome. Since the function of DNA methylation seems to be context specific, it would be very informative to look at other types of genomic features. For instance, transposon methylation has a silencing effect, while gene body methylation is often associated with gene expression (see for instance [Ball, Madeleine P, Jin Billy Li, Yuan Gao, Je-Hyuk Lee, Emily M LeProust, In-Hyun Park, Bin Xie, George Q Daley, and George M Church. 2009. “Targeted and Genome-Scale Strategies Reveal Gene-Body Methylation Signatures in Human Cells.” Nature Biotechnology 27 (4). Nature Publishing Group: 361–68.](https://paperpile.com/shared/54En5d). Also, exonic methylation sometimes plays a role in alternative splicing. It would be very interesting to look at these different genomic features in more detail in a future study.__

In some preliminary work (data not shown) I did in fact look at stratifying these analyses by other genomic features. By far the strongest signal I found was for CpG islands vs. the rest of the genome, hence my focus. Now that I have software for automating much of this work, I will likely revisit the effect of other genomic features in a larger dataset in a paper that I am writing.

### Chapter 8

__Even though the software presented in this section is said to be in a "very experimental" state, it looks promising and could be a very important contribution for future rigorous assessment of methods aimed at identifying differential methylation, once the issues associated with the overly high correlations are resolved.__

Thank you.

\newpage

## Reviewer #2

### Figures 4.11-4.17

__It is not clear why are we being shown so many of these.__

I want to show the genome-wide level of methylation for all samples used in my thesis. Also, I visualise these genome-wide methylation levels in two different ways, leading to twice as many figures. This does result in a lot of plots, which are in fact quite similar, but I thought it necessary to show the complete data and believe that the two visualisations show different facets of the data:

- (A) Figures 4.10-4.13 are to show that CpGs in CGIs are typically lowly methylated and those in non-CGIs are typically more highly methylated in these samples. These scale the curves so that the maximum is 1 for each of the CpG island (CGI) and non-CGI data. Without this scaling the CGI data dominate (as seen in Figures 4.14-4.17).
-  (B) Figures 4.14-4.17 figures are to show that because most CpGs are not in CGIs that this means that most CpGs are highly methylated in these samples. These figures do not use scaling of CGI vs. non-CGI.

__using a density estimator here is not appropriate when the distributions are so discrete. The peaks and valleys are artificial.__

I agree that the peaks and valleys are artificial in the kernel density plots. However, I wish to use the kernel density estimator for the reasons listed above.

__since you are using a density estimator, what does %CpGs mean in the y-axis__

This is a misunderstanding. In the figures with %CpGs on the on y-axis (Figures 4.14-4.17) I am not using a density estimator but rather a frequency polygon. The beta-values are binned into 0.01-width bins and so the height of the line corresponds to the percentage of CpGs in that bin, hence the %CpGs genome-wide with that methylation level.

__the orange line in fig 4.17 seems to be 0__

Yes, this is due to the issue of not scaling in this figure to account for the ratio of CpGs in CGIs vs. non-CGIs.

__Summary:__ For the reasons listed above, I believe it is necessary to have all these figures. Different plots, (A) vs. (B), show different features of the data that I believe are important.

### Consider changing the large tables to figures

Thank you, this is a useful suggestion. I have changed the following figures into tables: 1.1, 2.4, 4.1, 4.2, 4.3, 4.4. Other tables were not suitable for turning into figures (e.g., because they contained non-numerical data).

### Figure 6.3

__legend does not specify what colors represent.__

Figure 6.3 is a reproduction of Supplementary Figure 6 of Li Y, Zhu J, Tian G, Li N, Li Q, Ye M, et al. The DNA methylome of human peripheral blood mononuclear cells. _PLoS Biol_. Public Library of Science; 2010;8: e1000533. doi:10.1371/journal.pbio.1000533. There is no figure legend in the original publication, nor are the colours explained in the caption. I am therefore unable to specify what these colours represent.

### Figure 6.5a

__This is not a Fourier transform but a periodogram or spectogram. The Fourier transform is used to obtain power.__

__NB:__ I believe that this comment properly refers to Figure 6.7 rather than Figure 6.5a.

Figure 6.7 is indeed a periodogram and not a Fourier transform. This was due to my misunderstanding of the relationships between the autocovariance/autocorrelation function, the Fourier transform, the power spectrum, and the periodogram/spectogram.

However, in revisiting the code used to generate Figure 6.7, I found more fundamental problems with my analysis than a simple misuse of terminology. I had mistakenly computed the periodogram of the autocorrelation function (Figure 6.7) rather than of the raw data. Once I computed the periodogram of the raw data I found no evidence to support my original hunch about the (strange to me) results of Figure S7 in _Lyko et al._.

To recall, Figures S7a of _Lyko et al._ is a plot of the autocorrelation of beta-values in one of their samples, and Figure S7c is the "Fourier Transform of Autocorrelation". I had two concerns with these figures:

1. I was surprised to see that the maximum value of this autocorrelation was less than 0.015 (i.e., almost zero).
2. I suspected that the 3 bp periodicity of the autocorrelation might be driven by the spacing of CpGs in the genome rather than their methylation levels _per se_.

While the data of _Lyko et al._ are publicly available, it would require significant processing in order to investigate this hypothesis. Instead, I attempted to use existing, processed data to investigate this hunch (specifically, the chr22 data from the _FF_ sample of the _Lister_ dataset. The _FF_ sample is of human DNA, and is obviously very different to the honeybee data of _Lyko et al._, but I thought it might be sufficient to shed light on the matter). Unfortunately there is insufficient information in the paper to reproduce the analysis and no code is available.

#### Figure 6.6 (addressing Figure S7a of _Lyko et al._)

Since the maximum of the autocorrelation was very small, I suspected that there may have been an error in the computations underlying Figure S7a that resulting in an unusual scaling of the autocorrelation function.

I hypothesised that the autocorrelation had been computed using a "padded" vector of beta-values, where non-CpGs had their beta-value (artificially) set to zero. Figure 6.6 shows the autocorrelation of this "padded" vector of beta-values for the chr22 _FF_ data; it is visually very similar to Figure S7a in _Lyko et al._.

#### Figure 6.7 (addressing Figure S7c of _Lyko et al._)

Figure S7c of _Lyko et al._ plots "Power" vs. "Cycles per 100bp" and is titled "Fourier Transform of Autocorrelation". However, it is difficult to know exactly what has been plotted in this figure (it is not, for example, the periodogram returned by the `spectrum()` function in R).

My earlier, incorrect Figure 6.7 (which computed the periodogram of the autocorrelation rather than of the raw data using the `spectrum()` function in R) supported point (2). However, I no longer have evidence to support this interpretation and am less sure of what exactly is plotted in Figure S7c of _Lyko et al._. Since I cannot determine how exactly Figure S7c was created, and therefore cannot apply the method to the chr21 _FF_ data, I have opted to remove claims about the source of periodicity in the _Lyko et al._ data.

#### Summary

My original analysis was flawed due to a computational error and misunderstandings about spectral analysis. In revising this section I have opted to remove claim (2), which I no longer have evidence to support. I do still believe, however, that Figure 6.6 supports my initial concerns about Figure S7a of _Lyko et al._. I have therefore rephrased my concerns with Figure S7 of _Lyko et al._ and removed the inaccurate Figure 6.7.

### Inconsistent fonts in figures

__Fonts across figures is inconsistent. Sometimes they are very small.__

I have checked to ensure that I use a consistent theme for my plots where possible (this is sometimes not possible due to different software being used to generate various plots). Small font sizes have been enlarged.

### Standard errors

__Add standard error bars to trend estimate in correlation versus distance plots (e.g., Figures 7.3). Also, be consistent with colors, and style.__

I understand the reviewer's desire for a standard error on these plots. I did investigate this and found the loess fit to be very narrow for most samples across the IPD = 0-1500. Unfortunately it is difficult to see these at the resolution afforded to a printed page, particularly when considering multiple samples per plot. At this resolution it is difficult to show all of (1) the raw data, (2) the loess fit, and (3) the loess fit's standard error. This is illustrated in the attached figures (using one of the samples with the widest standard error). With that in mind, I opted to prioritise showing (1) + (2) over (2) + (3).

![response-to-reviewer2-1](https://cloud.githubusercontent.com/assets/1049741/10115833/3d4ce8c8-645b-11e5-88b3-9e0b1e399d3a.png)
![response-to-reviewer2-2](https://cloud.githubusercontent.com/assets/1049741/10115832/3d4793f0-645b-11e5-99e2-11daa744bcc4.png)

The differences in colours between plots are in fact important since they signify that a different variable is being stratified upon. For example, Figure 7.3 (red + blue) stratifies by strand whereas Figure 7.7 (green + purple) stratifies by correlation estimator.

### Minor comments

__Avoid colloquialisms such as "hot topic" and hard to interpret relative turns such as "ultra"__

Replaced "hot topic" with "active area". Also replaced some other colloquialisms. Removed some instances of "ultra" and other relative terms that were not qualified by some numerical support (left those instances alone when I felt the term was suitably qualified).

__Figure 6.1 is fuzzy__

Yes, I am aware of this. Unfortunately that is the highest quality freely image available from the publisher. Nature Publishing Group wishes to charge by $223.29 AUD to use a high-resolution version of this figure, which I consider exorbitant and unnecessary.

__change "exploiting" to "leveraging"__

Thank you, I agree this is a better term. Changed.

### 2015-10-01 Rmarkdown/wgbs_downstream_analyses.Rmd
Change some tables to figures (closes #14).
Closes #19
Fix missing journal in reference

@@
 Several papers use describe their analysing as performing a ``false discovery rate adjustment'' or ``false discovery rate correction'' without explicitly stating what they are doing or citing a reference.
-One paper uses an ultra-conservative Bonferonni correction.}
+One paper uses the Bonferonni correction, leading to a very conservative analysis since this correction aims to control the family-wise error rate.}
 \label{tab:multiple_testing}
@@
 The challenges of valid statistical inference at such data-driven regions are not unique to the problem of testing for DMRs.
-Similar problems arise in the analysis of chromatin immunoprecipitation sequencing (ChIP-seq) experiments \citep{Schwartzman:2011to, Lun:2014gn} and in the field of signal processing \citep{Schwartzman:2011vg}.
+Similar problems arise in the analysis of chromatin immunoprecipitation sequencing (ChIP-seq) experiments \citep{Schwartzman2013-eo, Lun:2014gn} and in the field of signal processing \citep{Schwartzman:2011vg}.
 
@@
 \label{tab:methpat_sparsity}
-\begin{tabulary}{\textwidth}{LLLLL}
+\begin{tabulary}{\textwidth}{LRRRR}
 \toprule
